### 2024-09-30 README.qmd
Files generated by PkgTemplates
PkgTemplates version: 0.7.51

---
format:
  commonmark:
    variant: -raw_html+tex_math_dollars
    wrap: none
    mermaid-format: png
crossref:
  fig-prefix: Figure
  tbl-prefix: Table
output: asis
---

# CounterfactualTraining

[![Stable](https://img.shields.io/badge/docs-stable-blue.svg)](https://JuliaTrustworthyAI.github.io/CounterfactualTraining.jl/stable/)
[![Dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://JuliaTrustworthyAI.github.io/CounterfactualTraining.jl/dev/)
[![Build Status](https://github.com/JuliaTrustworthyAI/CounterfactualTraining.jl/actions/workflows/CI.yml/badge.svg?branch=main)](https://github.com/JuliaTrustworthyAI/CounterfactualTraining.jl/actions/workflows/CI.yml?query=branch%3Amain)
[![Coverage](https://codecov.io/gh/JuliaTrustworthyAI/CounterfactualTraining.jl/branch/main/graph/badge.svg)](https://codecov.io/gh/JuliaTrustworthyAI/CounterfactualTraining.jl)
[![Code Style: Blue](https://img.shields.io/badge/code%20style-blue-4495d1.svg)](https://github.com/invenia/BlueStyle)
[![Aqua](https://raw.githubusercontent.com/JuliaTesting/Aqua.jl/master/badge.svg)](https://github.com/JuliaTesting/Aqua.jl)

### 2024-09-30 src/CounterfactualTraining.jl

@@ -1,0 +1,5 @@
+module CounterfactualTraining
+
+# Write your package code here.
+
+end

### 2024-09-30 test/runtests.jl

@@ -1,0 +1,10 @@
+using CounterfactualTraining
+using Test
+using Aqua
+
+@testset "CounterfactualTraining.jl" begin
+    @testset "Code quality (Aqua.jl)" begin
+        Aqua.test_all(CounterfactualTraining)
+    end
+    # Write your tests here.
+end

### 2024-10-01 paper/related_literature.tex
finetuning loss function

\section{Related Literature} \label{sec-lit}

\citet{guo2023counternet}

\citet{ross2021learning}

### 2024-10-01 paper/experiments/prototyping.jl

@@ -3,7 +3,7 @@ using Pkg; Pkg.activate("paper/experiments")
 using CounterfactualExplanations
 using CounterfactualExplanations: counterfactual
 using CounterfactualExplanations.Convergence
-using CounterfactualExplanations.Evaluation: plausibility
+using CounterfactualExplanations.Evaluation: plausibility, validity
 using CounterfactualExplanations.Generators
 using Flux
 using Serialization
@@ -12,11 +12,7 @@ using TaijaParallel
 
 include("utils.jl")
 
-# Countefactual generator:
-generator = ECCoGenerator(opt=Descent(1.0))
-max_iter = 25
-conv =  MaxIterConvergence(max_iter)
-pllr = ThreadsParallelizer()
+################### Setup ###################
 
 # Data and model:
 Xtrain, y = load_mnist(10000)
@@ -34,9 +30,14 @@ model = Chain(
 )
 
 # Loss function:
-function loss(yhat, y, implausibility; λ=0.1)
-    class_loss = Flux.Losses.logitcrossentropy(yhat, y)
-    return class_loss + λ * sum(implausibility)/length(implausibility)
+function loss(yhat, y, implausibility, targets, valids; λ=0.1)
+    if isnothing(valids)
+        return Flux.Losses.logitcrossentropy(yhat, y)
+    else
+        class_loss_valids = Flux.Losses.logitcrossentropy(yhat[valids], targets[valids])
+        class_loss_invalids = Flux.Losses.logitcrossentropy(yhat[valids], y[valids])
+        return class_loss_valids + class_loss_invalids + λ * sum(implausibility)/length(implausibility)
+    end
 end
 function accuracy(model, train_set)
     acc = 0
@@ -48,92 +49,123 @@ function accuracy(model, train_set)
     return acc / (train_set.batchsize * length(train_set))
 end
 
-# Training
-opt_state = Flux.setup(Adam(), model)
+################### Counterfactual Training ###################
 burnin = 40
 nepochs = 50
-my_log = []
-for epoch in 1:nepochs
-    losses = Float32[]
-    M = MLP(model; likelihood=:classification_multi)
-    implausibilities = Float32[]
-    for (i, batch) in enumerate(train_set)
-        input, label = batch
+max_iter = 50
+conv = GeneratorConditionsConvergence(max_iter=max_iter)
+pllr = ThreadsParallelizer()
[…]

### 2024-10-02 paper/related_literature.tex
work on related literature

@@
 
-\citet{guo2023counternet}
+\citet{guo2023counternet} are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce.
+In particular, they propose a specific network architecture that includes a predictor and CE generator network (\textcolor{red}{akin a GAN?}), where the parameters of the CE generator network are learnable.
+Counterfactuals are generated during each training iteration and fed back to the predictor network (\textcolor{red}{here we are aligned}).
+In contrast, we impose no restrictions on the neural network architecture at all.
+\textcolor{red}{NB: to ensure the one-hot encoding of categorical features is maintained, they simple use softmax (might be interesting for CE.jl)}.
+Interestingly, the authors find that their approach is sensitive to the choice of the loss function: only MSE seems to lead to good performance.
+They also demonstrate theoretically, that the objective function is difficult to optimize due to divergent gradients (\textcolor{red}{because partial gradients with respect to the classification loss component and the counterfactual validity component point in opposite directions}) and suffers from poor adversarial robustness.
+To mitigate these issues, the authors use block-wise gradient descent: they first update with respect to classification loss and then use a second update with respect to the other loss components (\textcolor{red}{this might be useful for our task as well}).
 
-\citet{ross2021learning}
+\citet{ross2021learning} propose a way to train models that are guaranteed to provide recourse for individuals with high probability.
+The approach builds on adversarial training (\textcolor{red}{here we are aligned}), where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class.
+The proposed method allows for imposing a set of actionable recourse ex-ante: for example, users can impose mutability constraints for features (\textcolor{red}{here we are aligned}).
+\textcolor{red}{NB: To solve their objective function more efficiently, they use a first-order Taylor approximation to approximate the recourse loss component (might be applicable in our case).}
+
+\subsection{Data Augmentation for Generalization}
+
+\citet{sauer2021counterfactual} generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as \textit{texture}, \textit{shape} and \textit{background}.
+The demonstrate that using these generated counterfactuals during classifier training improves model robustness.
+Similarly, \citet{abbasnejad2020counterfactual} argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
+They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.
+\citet{teney2020learning} propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
+Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as \textit{gradient supervision}) (\textcolor{red}{this might be useful for our task as well}).

### 2024-10-02 paper/related_literature.tex
more on related work

@@
 \section{Related Literature} \label{sec-lit}
+
+\subsection{Background on Counterfactual Explanations}
+
+\subsection{Learning Representations}
+
+\textcolor{red}{For example, joint-energy models ...}
+
+\subsection{Generalization and Robustness}
+
+\citet{sauer2021counterfactual} generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as \textit{texture}, \textit{shape} and \textit{background}.
+The demonstrate that using these generated counterfactuals during classifier training improves model robustness.
+Similarly, \citet{abbasnejad2020counterfactual} argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
+They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.
+\citet{teney2020learning} propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
+Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as \textit{gradient supervision}) (\textcolor{red}{this might be useful for our task as well}).
+In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: \citet{wu2021polyjuice}, propose POLYJUICE, a general-purpose counterfactual generator for language models.
+They demonstrate empirically that augmenting training data through POLYJUICE counterfactuals improves robustness in a number of NLP tasks.
+
+\subsection{Link to Adversarial Training}
+
+\citet{freiesleben2022intriguing} propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE (\textcolor{red}{this might be a useful notion for use to distinguish between adversarials and explanations during training}); secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations.
+\citet{pawelczyk2022exploring} show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.
+
+\subsection{Closely Related}
 
@@
 
-\subsection{Data Augmentation for Generalization}
-
-\citet{sauer2021counterfactual} generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as \textit{texture}, \textit{shape} and \textit{background}.
-The demonstrate that using these generated counterfactuals during classifier training improves model robustness.
-Similarly, \citet{abbasnejad2020counterfactual} argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
-They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.
-\citet{teney2020learning} propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as \textit{gradient supervision}) (\textcolor{red}{this might be useful for our task as well}).
+\citet{luu2023counterfactual} introduce Counterfactual Adversarial Training (CAT) with intention of improving generalization and robustness of language models.
+Specifically, they propose to proceed as follows: firstly, identify training samples that are subject to high predictive uncertainty (entropy); secondly, generate counterfactual explanations for those samples; and, finally, finetune the model on the augmented dataset that includes the generated counterfactuals.

### 2024-11-25 paper/experiments/README.qmd

---
format:
  commonmark:
    variant: -raw_html+tex_math_dollars
    wrap: none
    mermaid-format: png
crossref:
  fig-prefix: Figure
  tbl-prefix: Table
output: asis
engine: julia
julia: 
  exeflags: ["--project=./"]
---

# Experiments

This is the folder that contains all code used to generate the results in the paper. You can also use it to reproduce the results.

## Getting Started

To get started activate the environment in the Julia REPL:

```{julia}
# eval: false

using Pkg; Pkg.activate("paper/experiments")
```

Next, we load the `CTExperiments` package---a small package that ships some additional functionality relevant only to the experiments run for the paper.

```{julia}
using CTExperiments
```

### Running a single experiment

The `CTExperiments` package ships with a helper function that can be used to set up the default experiment configuration:

```{julia}
config_file = generate_template()
```

An experiment can be generated directly from that configuration file:

```{julia}
experiment = Experiment(config_file)
```

### 2024-12-09 paper/experiments/jobs/run_evaluation.sh
updated job scripts

@@ -3,8 +3,8 @@
 #SBATCH --job-name="Run Evaluation"
 #SBATCH --partition=compute
 #SBATCH --time=00:50:00
-#SBATCH --ntasks=48
-#SBATCH --cpus-per-task=4
+#SBATCH --ntasks=192
+#SBATCH --cpus-per-task=1
 #SBATCH --mem-per-cpu=4G
 #SBATCH --account=research-eemcs-insy
 #SBATCH --mail-type=BEGIN,END,FAIL,TIME_LIMIT

### 2024-12-09 paper/experiments/jobs/run_model.sh

@@ -3,8 +3,8 @@
 #SBATCH --job-name="Run Model"
 #SBATCH --partition=compute
 #SBATCH --time=00:10:00
-#SBATCH --ntasks=4
-#SBATCH --cpus-per-task=3
+#SBATCH --ntasks=12
+#SBATCH --cpus-per-task=1
 #SBATCH --mem-per-cpu=4G
 #SBATCH --account=innovation
 #SBATCH --mail-type=END,REQUEUE

### 2024-12-09 paper/experiments/jobs/run_poc_large.sh

@@ -1,0 +1,19 @@
+#!/bin/bash
+#
+#SBATCH --job-name="Run POC"
+#SBATCH --partition=compute
+#SBATCH --time=00:30:00
+#SBATCH --ntasks=100
+#SBATCH --cpus-per-task=1
+#SBATCH --mem-per-cpu=2G
+#SBATCH --account=research-eemcs-insy
+#SBATCH --mail-type=BEGIN,END,FAIL,TIME_LIMIT
+
+module load 2024r1 
+
+source .env
+source $JOB_DIR/slurm_header.sh
+
+srun julia --project=$EXPERIMENT_DIR --threads $SLURM_CPUS_PER_TASK $EXPERIMENT_DIR/run_grid.jl --config=$CONFIG > $LOG_DIR/run_poc.log
+
+srun julia --project=$EXPERIMENT_DIR --threads $SLURM_CPUS_PER_TASK $EXPERIMENT_DIR/run_evaluation.jl --config=$EVAL_CONFIG > $LOG_DIR/run_evaluation.log

### 2024-12-09 paper/experiments/run_evaluation_grid.jl

@@ -38,7 +38,9 @@ eval_list = MPI.bcast(eval_list, comm; root=0)
 
 MPI.Barrier(comm)  # Ensure all processes reach this point before finishing
 
-@assert length(eval_list) >= nprocs  "Ensure there are enough evaluations to distribute across all ranks"
+if length(eval_list) >= nprocs  
+    @warn "There are less evaluations than processes. Check CPU efficiency of job."
+end
 chunks = TaijaParallel.split_obs(eval_list, nprocs)     # distribute across processes
 
 for (i, chunk) in enumerate(chunks)

### 2024-12-09 paper/experiments/run_grid.jl

@@ -33,7 +33,9 @@ exper_list = MPI.bcast(exper_list, comm; root=0)
 
 MPI.Barrier(comm)  # Ensure all processes reach this point before finishing
 
-@assert length(exper_list) >= nprocs "Ensure there are enough experiments to distribute across all ranks"
+if length(exper_list) >= nprocs
+    @warn "There are less experiments than processes. Check CPU efficiency of job."
+end
 chunks = TaijaParallel.split_obs(exper_list, nprocs)     # distribute across processes
 
 for (i, chunk) in enumerate(chunks)

### 2024-12-18 paper/presenations/slides.qmd
trying to run sequentially

---
title: Counterfactual Training
subtitle: Update Meeting Dec 2024
author: 
  - name: "**Patrick Altmeyer**"
    url: https://www.patalt.org/
  - name: Arie van Deursen
  - name: Cynthia C. S. Liem
institute: Delft University of Technology
date: today
format: 
  tudelft-revealjs:
    theme: custom.scss
    self-contained: true
    smaller: false
    scrollable: true
    preview-links: auto
    slide-number: true
    transition: slide
    background-transition: fade
    fig-align: center
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          MathJax = {
            options: {
              menuOptions: {
                settings: {
                  assistiveMml: false
                }
              }
            }
          };
          </script>
revealjs-plugins:
  - pointer
crossref: 
  prp-title: RQ
  prp-prefix: RQ
classoption: "notheorems"
---

## Status

- *Code base*: In place and streamlined for reproducibility and configuration. 
- *Experiments*: Lots of work done and results largely supportive of idea.
    - Ran into problems on DelftBlue, which has set me back about 2 weeks. 
- *Paper*: Still bare-bones. 
- *ICML*: Potentially still possible to submit something, but this will be rushed and not "finished".

## High-Level Idea

$$

$$

### 2024-12-18 paper/presenations/slides.qmd
not sure why sequentially doesn't work

@@
 
+## Problems on Cluster
+
+- Trying to distribute: 
+    1. Models/experiments across processes and then ...
+    2. ... for each model/experiment distribute the counterfactual search across processes.
+- Out-of-memory issues, data races, ...
+- When using multi-processing for models and multi-threading for counterfactual search, I get low CPU efficiency on DelftBlue (jobs get cancelled).
+
 ## High-Level Idea

### 2024-12-18 paper/presenations/slides.qmd
come on

@@
 - Out-of-memory issues, data races, ...
-- When using multi-processing for models and multi-threading for counterfactual search, I get low CPU efficiency on DelftBlue (jobs get cancelled).
+- Multi-processing for models & multi-threading for counterfactual search: low CPU efficiency on DelftBlue (jobs get cancelled).
 
@@
 
-$$
+Combine ideas from Energy-Based Models and Adversarial Training:
 
 $$
+\ell_{\text{clf}}(f(x),y) + \lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x) + \text{reg}
+$$

### 2024-12-18 paper/experiments/jobs/run_grid_sequentially.sh

@@ -3,7 +3,7 @@
 #SBATCH --job-name="Run Grid (Sequentially)"
 #SBATCH --partition=compute
 #SBATCH --time=00:10:00
-#SBATCH --ntasks=42
+#SBATCH --ntasks=10
 #SBATCH --cpus-per-task=1
 #SBATCH --mem-per-cpu=2G
 #SBATCH --account=innovation

### 2024-12-18 paper/presenations/slides.qmd
sequentially might work?

@@
           </script>
-revealjs-plugins:
-  - pointer
 crossref: 
@@
 $$
-\ell_{\text{clf}}(f(x),y) + \lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x) + \text{reg}
+\ell_{\text{clf}}(f(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x) + \lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y) + \text{reg}
 $$

### 2024-12-18 paper/presenations/slides.qmd
try a bit smalelr

@@
 - Trying to distribute: 
-    1. Models/experiments across processes and then ...
-    2. ... for each model/experiment distribute the counterfactual search across processes.
+    1. Models/experiments across processes.
+    2. For each model/experiment distribute the counterfactual search across processes.
 - Out-of-memory issues, data races, ...
@@
 $$
-\ell_{\text{clf}}(f(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x) + \lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y) + \text{reg}
+\ell_{\text{clf}}(f(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x) + \lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y) + \text{r}
 $$
+
+- $x^\prime$ are counterfactuals of $x$.
+- $\ell_{\text{gen}}$ is the difference in energies between observed samples and counterfactuals.
+- Counterfactuals are recycled as adversarial examples.
+
+## Training Details
+
+During each `EPOCH`:
+
+1. Generate `nce` counterfactuals and distribute across mini-batches.
+2. For each batch compute:
+    - Classifier loss: $\ell_{\text{clf}}(f(x),y)$
+    - Generator loss: $\ell_{\text{gen}}(x^\prime,x)$
+    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y)$
+3. Backpropagate all losses and update parameters.

### 2024-12-18 paper/presenations/slides.qmd
uff

@@
 $$
-\ell_{\text{clf}}(f(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x) + \lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y) + \text{r}
+\ell_{\text{clf}}(f_\theta(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x;\theta) + \lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta(x^\prime),y)
 $$
 
-- $x^\prime$ are counterfactuals of $x$.
+- $x^\prime$ are counterfactuals of $x_{s}\subseteq x$.
 - $\ell_{\text{gen}}$ is the difference in energies between observed samples and counterfactuals.
@@
 2. For each batch compute:
-    - Classifier loss: $\ell_{\text{clf}}(f(x),y)$
-    - Generator loss: $\ell_{\text{gen}}(x^\prime,x)$
-    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f(x^\prime),y)$
+    - Classifier loss: $\ell_{\text{clf}}(f_\theta(x),y)$
+    - Generator loss: $\ell_{\text{gen}}(x^\prime,x;\theta)$
+    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta(x^\prime),y)$
 3. Backpropagate all losses and update parameters.
+
+## Motivation and Intuition
+
+- Instead of using SGLD to sample from $p(x|t;\theta)$, we use counterfactual generators.
+- The idea is to align counterfactual explanations with observed data to induce plausibility.
+- This should only work if counterfactuals are generated faithfully (favorable evidence).
+- Approach can be leveraged to impicitly encode mutability and domain constraints in model.

### 2024-12-18 paper/presenations/slides.qmd
upped cpus

@@
 $$
-\ell_{\text{clf}}(f_\theta(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}(x^\prime,x;\theta) + \lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta(x^\prime),y)
+\ell_{\text{clf}}(f_\theta(x),y) + \lambda_{\text{gen}} \ell_{\text{gen}}({x^\prime}_t,x_t;\theta) + \lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta({x^\prime}_t),y)
 $$
 
-- $x^\prime$ are counterfactuals of $x_{s}\subseteq x$.
+- ${x^\prime}_t$ are counterfactuals of $x_{s}\subseteq x$ with target class $t$.
 - $\ell_{\text{gen}}$ is the difference in energies between observed samples and counterfactuals.
@@
     - Classifier loss: $\ell_{\text{clf}}(f_\theta(x),y)$
-    - Generator loss: $\ell_{\text{gen}}(x^\prime,x;\theta)$
-    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta(x^\prime),y)$
+    - Generator loss: $\ell_{\text{gen}}({x^\prime}_t,x_t;\theta)$
+    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta({x^\prime}_t),y)$
 3. Backpropagate all losses and update parameters.
@@
 - Approach can be leveraged to impicitly encode mutability and domain constraints in model.
+
+## Encoding Domain Knowledge
+
+**Example**: Let $f_\theta(x)=\theta^{T}x$ be a linear classifier. Suppose that featue $x_1$ is immutable (e.g. 'gender' or 'faith').
+
+$$
+\begin{aligned}
+\nabla_\theta \ell_{\text{gen}}(x^\prime,x;\theta) &= \nabla_\theta ( \theta^{T}x[t] - \theta^{T}x^\prime[t]) \\
+\frac{\partial \ell_{\text{gen}}}{\partial\theta_1}(x^\prime,x;\theta)&=x_1 - x_1^\prime
+\end{aligned}
+$$
+
+- If $$

### 2024-12-18 paper/presenations/slides.qmd
presentation for friday

@@
 
-Combine ideas from Energy-Based Models and Adversarial Training:
+Counterfactual Training (CT) combines ideas from Energy-Based Models and Adversarial Training:
 
@@
 - ${x^\prime}_t$ are counterfactuals of $x_{s}\subseteq x$ with target class $t$.
-- $\ell_{\text{gen}}$ is the difference in energies between observed samples and counterfactuals.
+- $\ell_{\text{gen}}$ is the difference in energies between observed samples in target class $x_t$ and counterfactuals.
 - Counterfactuals are recycled as adversarial examples.
@@
 
-**Example**: Let $f_\theta(x)=\theta^{T}x$ be a linear classifier. Suppose that featue $x_1$ is immutable (e.g. 'gender' or 'faith').
+Let $f_\theta(x)=\theta^{T}x$ be a linear classifier:
 
@@
 \begin{aligned}
-\nabla_\theta \ell_{\text{gen}}(x^\prime,x;\theta) &= \nabla_\theta ( \theta^{T}x[t] - \theta^{T}x^\prime[t]) \\
-\frac{\partial \ell_{\text{gen}}}{\partial\theta_1}(x^\prime,x;\theta)&=x_1 - x_1^\prime
+\nabla_\theta \ell_{\text{gen}}({x^\prime}_t,x_t;\theta) &= \nabla_\theta ( \theta^{T}x_t - \theta^{T}{x^\prime}_t) \\
+\frac{\partial \ell_{\text{gen}}}{\partial\theta[1]}(x^\prime,x;\theta)&=x_{t}[1] - x_{t}^\prime[1]
 \end{aligned}
@@
 
-- If $$
+Suppose that featue $x[1]$ is immutable (e.g. 'age'), so $x_{t}^\prime[1]=x_s[1]$ where $s \neq t$. If $x_{t}[1] > x_s[1]$:
+
+- $\ell_{\text{gen}}$ induces lower values of $\theta[1]$, acting as a hedge against $\ell_{\text{clf}}$, which favours higher $\theta[1]$.
+
+# Findings
+
+## Moons (Plausibility)
+
+- All counterfactuals at test time generated using *ECCo*.
+- Penalty on energy differential increases from *l.* to *r.*
+
+![Plausibility of faithful counterfactuals $x_{t}^\prime$ measured in terms of their distance from $x_{t}$.](www/moons_plaus.png){#fig-moons-plausib}
+
+## Moons (Example)
+
+::::{.columns}::::
+:::{.column width='50%'}
+![Counterfactual explanations for model trained with CT (*ECCo*).](www/moons_ce_ecco.png){#fig-ce-ecco}
+:::
+:::{.column width='50%'}
+![Counterfactual explanations for conventionally trained model.](www/moons_ce_vanilla.png){#fig-ce-vanilla}
+:::
+::::
+
+## Moons (Validation Accuracy)
+
+![Validation accuracy for different models.](www/moons_acc_val.png){#fig-moons-acc-val}
+
+# Planning Ahead

### 2024-12-18 paper/presenations/slides.qmd
added PDF

@@
 format: 
+  beamer: default
   tudelft-revealjs:

### 2024-12-18 paper/presenations/slides.qmd
uf

@@
 # Planning Ahead
+
+## Planned Contributions
+
+- Focus on enhancing trustworthiness and applicability of small opaque models. 
+    - Add more datasets and neuro-tree models (relatively straightforward).
+- Do not pitch as state-of-the-art approach to generative modelling similar to JEMs, but rather as an extension of JEMs to XAI and AR. 
+- Limit use of image datasets to illustrate arguments:
+    - Example: what happens if digit values can only be increased/decreased?

### 2024-12-19 paper/presenations/slides.qmd
vamos

@@
 - Multi-processing for models & multi-threading for counterfactual search: low CPU efficiency on DelftBlue (jobs get cancelled).
+
+## The Big Oversight
+
+- **Core problem**: `CounterfactualExplanation` objects are *huge*. They store `X` (the input), `y` (the output), etc. in their own memory.
+- Can't easily fit `CounterfactualExplanation` objects on cluster memory and pass them around processes. 
+- **Good news**: Relatively straightforward to add `FlattenedCE` with gazzilion times lower memory footprint.
 

### 2024-12-19 paper/experiments/CTExperiments/src/model_and_data.jl

@@ -41,7 +41,7 @@ function get_data(data::Dataset; n::Union{Nothing,Int}=data.n_train, test_set::B
     else
         idx = 1:n_total
     end
-    X = X[:, idx]
+    X = Float32.(X[:, idx])
     y = y[idx]
     return X, y
 end

### 2024-12-19 paper/experiments/CTExperiments/src/moons.jl

@@ -13,14 +13,16 @@ Keyword container for the `Moons` data set. Can specify the number of samples `n
 """
 Base.@kwdef struct Moons <: Dataset
     n_train::Int = 3000
-    batchsize::Int = 1
+    batchsize::Int = 30
     n_validation::Int = 600
 end
 
 get_domain(d::Moons) = nothing
 
 function get_ce_data(data::Moons, n_total::Int)
-    return CounterfactualData(load_moons(n_total)...)
+    X, y = load_moons(n_total)
+    X = Float32.(X)
+    return CounterfactualData(X, y)
 end
 
 """
@@ -28,4 +30,8 @@ end
 
 Load the Moons data set. Since data is synthetically generated, `test_set` has no effect.
 """
-get_data(data::Moons, test_set::Bool=false) = load_moons()
+function get_data(data::Moons, test_set::Bool=false)
+    X, y = load_moons()
+    X = Float32.(X)
+    return X, y
+end

### 2024-12-19 paper/experiments/jobs/run_grid_sequentially.sh
scatter still causing issues

@@ -2,8 +2,8 @@
 #
 #SBATCH --job-name="Run Grid (Sequentially)"
 #SBATCH --partition=compute
-#SBATCH --time=00:10:00
-#SBATCH --ntasks=50
+#SBATCH --time=01:00:00
+#SBATCH --ntasks=40
 #SBATCH --cpus-per-task=1
 #SBATCH --mem-per-cpu=2G
 #SBATCH --account=research-eemcs-insy

### 2025-01-13 paper/presenations/slides.qmd
scatter still causing issues

@@
 author: 
-  - name: "**Patrick Altmeyer**"
+  - name: Patrick Altmeyer
     url: https://www.patalt.org/
@@
 format: 
-  beamer: default
+  beamer:
+    theme: Boadilla
   tudelft-revealjs:
@@
 - Can't easily fit `CounterfactualExplanation` objects on cluster memory and pass them around processes. 
-- **Good news**: Relatively straightforward to add `FlattenedCE` with gazzilion times lower memory footprint.
+- **Good news**: Relatively straightforward to add `FlattenedCE` with gazzilion times lower memory footprint (done!).
+- Nested parallelization issue remains and I will not spend more time trying to make it work.
+
+# Methodology
 
@@
 2. For each batch compute:
-    - Classifier loss: $\ell_{\text{clf}}(f_\theta(x),y)$
-    - Generator loss: $\ell_{\text{gen}}({x^\prime}_t,x_t;\theta)$
-    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta({x^\prime}_t),y)$
+    - Classifier loss: $\ell_{\text{clf}}(f_\theta(x),y)$.
+    - Generator loss: $\lambda_{\text{gen}} \ell_{\text{gen}}({x^\prime}_t,x_t;\theta)$.
+    - Adversarial loss: $\lambda_{\text{adv}} \ell_{\text{clf}}(f_\theta({x^\prime}_t),y)$.
+    - Regularization term for energies.
 3. Backpropagate all losses and update parameters.
@@
 
-![Plausibility of faithful counterfactuals $x_{t}^\prime$ measured in terms of their distance from $x_{t}$.](www/moons_plaus.png){#fig-moons-plausib}
+![Plausibility of faithful counterfactuals $x_{t}^\prime$ measured in terms of their distance from $x_{t}$. Higher values indicate higher plausibility.](www/moons_plaus.png){#fig-moons-plausib}
 
@@
 
+## MNIST (Vanilla vs *ECCo*)
+
+::::{.columns}::::
+:::{.column width='50%'}
+![Faithful counterfactuals for conventionally trained MLP.](www/poc_model.png){#fig-mnist-model}
+:::
+:::{.column width='50%'}
+![Faithful counterfactuals for same MLP with CT (*ECCo*).](www/poc_model_ct_ecco.png){#fig-mnist-model-ct-ecco}
+:::
+::::
+
+## MNIST (*Generic* and *REVISE*)
+
+::::{.columns}::::
+:::{.column width='50%'}
+![Faithful counterfactuals for same MLP with CT (*Generic*).](www/poc_model_ct_generic.png){#fig-mnist-model-ct-generic}
+:::
+:::{.column width='50%'}
+![Faithful counterfactuals for same MLP with CT (*REVISE*).](www/poc_model_ct_model_revise.png){#fig-mnist-model-ct-model-revise}
+:::
+::::
+
 # Planning Ahead
@@
     - Example: what happens if digit values can only be increased/decreased?
+
+## Timeline
+
+- Code base is already streamlined very well to allow for grid searches with easy configuration.
+- Lots of open tasks and questions (see [issue](https://github.com/JuliaTrustworthyAI/CounterfactualTraining.jl/issues/8))
+    1. Add MMD to measure plausibility.
+    2. Evaluate adversarial robustness and generative capacity. Compare to AT, JEM.
+    3. ... 
+- Still aiming for ICML to commit to a timeline, but it seems unlikely to be in the shape I would like.

### 2025-01-15 paper/experiments/jobs/run_grid.sh
try lin_sep first

@@ -13,4 +13,4 @@ module load 2024r1
 source .env
 source $JOB_DIR/slurm_header.sh
 
-srun julia --project=$EXPERIMENT_DIR --threads $SLURM_CPUS_PER_TASK $EXPERIMENT_DIR/run_grid.jl --config=$CONFIG > $LOG_DIR/run_grid.log
+srun julia --project=$EXPERIMENT_DIR --threads $SLURM_CPUS_PER_TASK $EXPERIMENT_DIR/run_grid.jl --config=$CONFIG $DATA_ARG $MODEL_ARG

### 2025-01-16 paper/experiments/README.qmd
updated job scripts
vamos
try lin_sep first
minor things

@@
 
+## Command Line Usage
+
+```
+sbatch --ntasks=$(determine_resources) --cpus-per-task=$NTHREADS myfile.sh
+```
+
+```
+sbatch --ntasks=$(determine_resources --data=mydata --model=mymodel) --cpus-per-task=$NTHREADS myfile.sh --data=mydata --model=mymodel
+```
+
+:::{.callout-warning}
+The extra arguments `--data=mydata --model=mymodel` should be consistent, i.e. they should be passed consitently to `determine_resources` and the whole `sbatch` command. 
+:::
+
+## Interactive Usage
+
 ### Running a single experiment

### 2025-01-16 paper/experiments/CTExperiments/src/CTExperiments.jl

@@ -52,5 +52,6 @@ export GMSC, MNIST, Moons, LinearlySeparable, Overlapping, Circles, CaliHousing,
 export get_data, get_ce_data, ntotal
 export get_domain, get_mutability
 export default_save_dir
+export has_results, needs_results
 
 end

### 2025-01-16 paper/experiments/determine_resources.jl

@@ -6,10 +6,8 @@ DotEnv.load!()
 
 # Get config and set up grid:
 config_file = get_config_from_args(; save_adjusted=false, return_adjusted=true)
-@info "After `get_config_from_args` call ..."
 root_name = CTExperiments.from_toml(config_file)["name"]
 exper_grid = ExperimentGrid(config_file)
-@info "After setting up grid..."
 
 # Determine number of slurm tasks:
 total_tasks = ntasks(exper_grid)

### 2025-01-23 paper/experiments/CTExperiments/src/presentation.jl
progress on tabulating results

@@ -9,6 +9,8 @@ using TaijaPlotting
 include("plots.jl")
 include("tables.jl")
 
+export tabulate_results
+
 function useful_byvars(df_meta::DataFrame)
     return names(df_meta)[findall([length(unique(x)) != 1 for x in eachcol(df_meta)])]
 end

### 2025-01-23 paper/experiments/CTExperiments/src/tables.jl

@@ -1,20 +1,60 @@
 using ColorSchemes
 using DataFrames
 using PrettyTables
+using StatsBase
 
-function get_highlighter(
+function tabulate_results(
     df::DataFrame,
     value_var::String="mean";
-    scheme::ColorSchemes.ColorScheme=colorschemes[:coolwarm],
+    backend::Val=Val(:text),
+    kwrgs...
+)
+
+    # Highlighters:
+    hls = (value_highlighter(df, value_var; backend=backend, kwrgs...),)
+    if "generator_type" in names(df)
+        df.generator_type = format_generator.(df.generator_type)
+        gen_hl = generator_highlighter(df; backend=backend)
+        hls = (hls..., gen_hl)
+    end
+
+    header = format_header.(names(df))
+    return pretty_table(df; highlighters=hls, backend=backend, header=header)
+end
+
+format_generator(s::AbstractString) = get_generator_name(generator_types[s](), pretty=true)
+
+function format_header(s::String)
+    s = split(s, "_") |>
+        ss -> [s in ["exper", "eval"] ? "($s)" : uppercasefirst(s) for s in ss] |>
+        ss -> join(ss, " ")
+    return s
+end
+
+function value_highlighter(
+    df::DataFrame,
+    value_var::String="mean";
+    scheme::ColorSchemes.ColorScheme=reverse(colorschemes[:rose]),
+    alpha::AbstractFloat=0.5,
+    backend::Val=Val(:text)
 )
     @assert value_var in names(df) "Provided variables $(value_var) is no in the dataframe"
-    col_idx = findall(value_var .== names(df))[1]
-    @info "Highlighting column $col_idx"
+    col_idx = findall(value_var .== names(df))
+    lb, ub = (alpha/2, 1 - alpha/2)
+    @info "Lower bound $(lb), Upper bound $(ub)"
+    lims = (quantile(df[:, value_var], lb), quantile(df[:, value_var], ub))
+    @info "Limits: $(lims)"
 
+    return value_highlighter(lims, col_idx, backend, scheme)
+end
+
+function value_highlighter(
+    lims::Tuple, col_idx::Vector{Int}, backend::Val{:text}, scheme::ColorScheme
+)
     hl = Highlighter(
-        (df, i, j) -> j == col_idx,
+        (df, i, j) -> j in col_idx,
         (h, df, i, j) -> begin
-            color = get(scheme, df[i, j], (0, 20))
+            color = get(scheme, df[i, j], lims)
[…]

### 2025-01-23 paper/experiments/CTExperiments/src/training_params.jl

@@ -23,11 +23,11 @@ struct Gravitational <: AbstractGeneratorType end
 
 struct Omniscient <: AbstractGeneratorType end
 
-get_generator_name(gen::ECCo) = "ecco"
-get_generator_name(gen::Generic) = "generic"
-get_generator_name(gen::REVISE) = "revise"
-get_generator_name(gen::Gravitational) = "gravi"
-get_generator_name(gen::Omniscient) = "omni"
+get_generator_name(gen::ECCo; pretty::Bool=false) = pretty ? "ECCo" : "ecco"
+get_generator_name(gen::Generic; pretty::Bool=false) = pretty ? "Generic" : "generic"
+get_generator_name(gen::REVISE; pretty::Bool=false) = pretty ? "REVISE" : "revise"
+get_generator_name(gen::Gravitational; pretty::Bool=false) = pretty ? "Gravitational" : "gravi"
+get_generator_name(gen::Omniscient; pretty::Bool=false) = pretty ? "Omniscient" : "omni"
 
 """
     generator_types

### 2025-01-23 paper/appendix.qmd
some work on streamline Quarto usage

---
format:
  latex:
    template: bare.tex
---

## Appendix

Here's my awesome appendix.

### 2025-01-23 paper/related.qmd
some work on streamline Quarto usage

# Related Literature {#sec-lit}

## Background on Counterfactual Explanations

## Learning Representations

> For example, joint-energy models

## Generalization and Robustness

@sauer2021counterfactual generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as *texture*, *shape* and *background*. The demonstrate that using these generated counterfactuals during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.

@teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets. Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as *gradient supervision*)

> This might be useful for our task as well...

In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: \cite{wu2021polyjuice}, propose POLYJUICE, a general-purpose counterfactual generator for language models. They demonstrate empirically that augmenting training data through POLYJUICE counterfactuals improves robustness in a number of NLP tasks.

## Link to Adversarial Training

@freiesleben2022intriguing propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE;

> this might be a useful notion for use to distinguish between adversarials and explanations during training

secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations. @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.

## Closely Related

@guo2023counternet are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce. In particular, they propose a specific network architecture that includes a predictor and CE generator network,

> akin a GAN?

where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network (\textcolor{red}{here we are aligned}). In contrast, we impose no restrictions on the neural network architecture at all.

> to ensure the one-hot encoding of categorical features is maintained, they simple use softmax (might be interesting for CE.jl)

Interestingly, the authors find that their approach is sensitive to the choice of the loss function: only MSE seems to lead to good performance. They also demonstrate theoretically, that the objective function is difficult to optimize due to divergent gradients and suffers from poor adversarial robustness.

> because partial gradients with respect to the classification loss component and the counterfactual validity component point in opposite directions

To mitigate these issues, the authors use block-wise gradient descent: they first update with respect to classification loss and then use a second update with respect to the other loss components

> this might be useful for our task as well

@ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals with high probability. The approach builds on adversarial training,

> here we are aligned

where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class. The proposed method allows for imposing a set of actionable recourse ex-ante: for example, users can impose mutability constraints for features.

> here we are aligned

> To solve their objective function more efficiently, they use a first-order Taylor approximation to approximate the recourse loss component (might be applicable in our case

@luu2023counterfactual introduce Counterfactual Adversarial Training (CAT) with intention of improving generalization and robustness of language models. Specifically, they propose to proceed as follows: firstly, identify training samples that are subject to high predictive uncertainty (entropy); secondly, generate counterfactual explanations for those samples; and, finally, finetune the model on the augmented dataset that includes the generated counterfactuals.

### 2025-01-23 paper/related.tex
some work on streamline Quarto usage

\section{Related Literature}\label{sec-lit}

\subsection{Background on Counterfactual
Explanations}\label{background-on-counterfactual-explanations}

\subsection{Learning Representations}\label{learning-representations}

\begin{quote}
For example, joint-energy models
\end{quote}

\subsection{Generalization and
Robustness}\label{generalization-and-robustness}

generate counterfactual images for MNIST and ImageNet through
independent mechanisms (IM): each IM learns class-conditional input
distributions over a specific lower-dimensional, semantically meaningful
factor, such as \emph{texture}, \emph{shape} and \emph{background}. The
demonstrate that using these generated counterfactuals during classifier
training improves model robustness. Similarly, argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization.

propose an approach using counterfactuals in training that does not rely
on data augmentation: they argue that counterfactual pairs typically
already exist in training datasets. Specifically, their approach relies
on, firstly, identifying similar input samples with different
annotations and, secondly, ensuring that the gradient of the classifier
aligns with the vector between pairs of counterfactual inputs using the
cosine distance as a loss function (referred to as \emph{gradient
supervision})

\begin{quote}
This might be useful for our task as well\ldots{}
\end{quote}

In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation:
\cite{wu2021polyjuice}, propose POLYJUICE, a general-purpose
counterfactual generator for language models. They demonstrate
empirically that augmenting training data through POLYJUICE
counterfactuals improves robustness in a number of NLP tasks.

\subsection{Link to Adversarial
Training}\label{link-to-adversarial-training}

propose two definitional differences between Adversarial Examples (AE)
and Counterfactual Explanations (CE): firstly, and more importantly
according to the authors, the term AE implies missclassification, which
is not the case for CE;

\begin{quote}
this might be a useful notion for use to distinguish between
adversarials and explanations during training
\end{quote}

secondly, they argue that closeness plays a more critical role in the
context of CE but confess that even counterfactuals that are not close
might be relevant explanations. show that CE and AE are equivalent under
certain conditions and derive upper bounds on the distances between
them.

\subsection{Closely Related}\label{closely-related}

are the first to propose end-to-end training pipeline that includes
counterfactual explanations as part of the training prodeduce. In
particular, they propose a specific network architecture that includes a
predictor and CE generator network,

\begin{quote}
akin a GAN?
\end{quote}

where the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network (\textcolor{red}{here we are aligned}). In
contrast, we impose no restrictions on the neural network architecture
at all.

\begin{quote}
to ensure the one-hot encoding of categorical features is maintained,
they simple use softmax (might be interesting for CE.jl)
\end{quote}

Interestingly, the authors find that their approach is sensitive to the
choice of the loss function: only MSE seems to lead to good performance.
They also demonstrate theoretically, that the objective function is
difficult to optimize due to divergent gradients and suffers from poor
adversarial robustness.

\begin{quote}
because partial gradients with respect to the classification loss
component and the counterfactual validity component point in opposite
directions
\end{quote}

To mitigate these issues, the authors use block-wise gradient descent:
they first update with respect to classification loss and then use a
second update with respect to the other loss components

\begin{quote}
this might be useful for our task as well
\end{quote}

propose a way to train models that are guaranteed to provide recourse
for individuals with high probability. The approach builds on
adversarial training,

\begin{quote}
here we are aligned
\end{quote}

where in this context adversarial examples are actively encouraged to
exist, but only target attacks with respect to the positive class. The
proposed method allows for imposing a set of actionable recourse
ex-ante: for example, users can impose mutability constraints for
features.

\begin{quote}
here we are aligned
\end{quote}

\begin{quote}
To solve their objective function more efficiently, they use a
first-order Taylor approximation to approximate the recourse loss
component (might be applicable in our case
\end{quote}

introduce Counterfactual Adversarial Training (CAT) with intention of
improving generalization and robustness of language models.
Specifically, they propose to proceed as follows: firstly, identify
training samples that are subject to high predictive uncertainty
(entropy); secondly, generate counterfactual explanations for those
samples; and, finally, finetune the model on the augmented dataset that
includes the generated counterfactuals.

### 2025-01-23 paper/related_literature.tex
some work on streamline Quarto usage

--- 2024-10-02 paper/related_literature.tex
@@
 
-\citet{guo2023counternet} are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce.
+\subsection{Background on Counterfactual Explanations}
+
+\subsection{Learning Representations}
+
+\textcolor{red}{For example, joint-energy models ...}
+
+\subsection{Generalization and Robustness}
+
+\cite{sauer2021counterfactual} generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as \textit{texture}, \textit{shape} and \textit{background}.
+The demonstrate that using these generated counterfactuals during classifier training improves model robustness.
+Similarly, \cite{abbasnejad2020counterfactual} argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
+They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.
+
+\cite{teney2020learning} propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
+Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as \textit{gradient supervision}) (\textcolor{red}{this might be useful for our task as well}).
+
+In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: \cite{wu2021polyjuice}, propose POLYJUICE, a general-purpose counterfactual generator for language models.
+They demonstrate empirically that augmenting training data through POLYJUICE counterfactuals improves robustness in a number of NLP tasks.
+
+\subsection{Link to Adversarial Training}
+
+\cite{freiesleben2022intriguing} propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE (\textcolor{red}{this might be a useful notion for use to distinguish between adversarials and explanations during training}); secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations.
+\cite{pawelczyk2022exploring} show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.
+
+\subsection{Closely Related}
+
+\cite{guo2023counternet} are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce.
 In particular, they propose a specific network architecture that includes a predictor and CE generator network (\textcolor{red}{akin a GAN?}), where the parameters of the CE generator network are learnable.
@@
 
-\citet{ross2021learning} propose a way to train models that are guaranteed to provide recourse for individuals with high probability.
+\cite{ross2021learning} propose a way to train models that are guaranteed to provide recourse for individuals with high probability.
 The approach builds on adversarial training (\textcolor{red}{here we are aligned}), where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class.
@@
 
-\subsection{Data Augmentation for Generalization}
-
-\citet{sauer2021counterfactual} generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as \textit{texture}, \textit{shape} and \textit{background}.
-The demonstrate that using these generated counterfactuals during classifier training improves model robustness.
-Similarly, \citet{abbasnejad2020counterfactual} argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
-They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.
-\citet{teney2020learning} propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as \textit{gradient supervision}) (\textcolor{red}{this might be useful for our task as well}).
+\cite{luu2023counterfactual} introduce Counterfactual Adversarial Training (CAT) with intention of improving generalization and robustness of language models.
+Specifically, they propose to proceed as follows: firstly, identify training samples that are subject to high predictive uncertainty (entropy); secondly, generate counterfactual explanations for those samples; and, finally, finetune the model on the augmented dataset that includes the generated counterfactuals.

### 2025-01-23 paper/related.tex
bibliography sorted and everything adapted to ECML

@@
 Explanations}\label{background-on-counterfactual-explanations}
+
+\cite{wachter2017counterfactual, joshi2019realistic, altmeyer2024faithful}
 
@@
 
-generate counterfactual images for MNIST and ImageNet through
-independent mechanisms (IM): each IM learns class-conditional input
-distributions over a specific lower-dimensional, semantically meaningful
-factor, such as \emph{texture}, \emph{shape} and \emph{background}. The
-demonstrate that using these generated counterfactuals during classifier
-training improves model robustness. Similarly, argue that
+\cite{sauer2021counterfactual} generate counterfactual images for MNIST
+and ImageNet through independent mechanisms (IM): each IM learns
+class-conditional input distributions over a specific lower-dimensional,
+semantically meaningful factor, such as \emph{texture}, \emph{shape} and
+\emph{background}. The demonstrate that using these generated
+counterfactuals during classifier training improves model robustness.
+Similarly, \cite{abbasnejad2020counterfactual} argue that
 counterfactuals represent potentially useful training data in machine
@@
 
-propose an approach using counterfactuals in training that does not rely
-on data augmentation: they argue that counterfactual pairs typically
-already exist in training datasets. Specifically, their approach relies
-on, firstly, identifying similar input samples with different
-annotations and, secondly, ensuring that the gradient of the classifier
-aligns with the vector between pairs of counterfactual inputs using the
-cosine distance as a loss function (referred to as \emph{gradient
-supervision})
+\cite{teney2020learning} propose an approach using counterfactuals in
+training that does not rely on data augmentation: they argue that
+counterfactual pairs typically already exist in training datasets.
+Specifically, their approach relies on, firstly, identifying similar
+input samples with different annotations and, secondly, ensuring that
+the gradient of the classifier aligns with the vector between pairs of
+counterfactual inputs using the cosine distance as a loss function
+(referred to as \emph{gradient supervision})
 
@@
 
-propose two definitional differences between Adversarial Examples (AE)
-and Counterfactual Explanations (CE): firstly, and more importantly
-according to the authors, the term AE implies missclassification, which
-is not the case for CE;
+\cite{freiesleben2022intriguing} propose two definitional differences
+between Adversarial Examples (AE) and Counterfactual Explanations (CE):
+firstly, and more importantly according to the authors, the term AE
+implies missclassification, which is not the case for CE;
 
@@
 context of CE but confess that even counterfactuals that are not close
-might be relevant explanations. show that CE and AE are equivalent under
-certain conditions and derive upper bounds on the distances between
-them.
+might be relevant explanations. \cite{pawelczyk2022exploring} show that
+CE and AE are equivalent under certain conditions and derive upper
+bounds on the distances between them.
 
@@
 
-are the first to propose end-to-end training pipeline that includes
-counterfactual explanations as part of the training prodeduce. In
-particular, they propose a specific network architecture that includes a
-predictor and CE generator network,
+\cite{guo2023counternet} are the first to propose end-to-end training
+pipeline that includes counterfactual explanations as part of the
+training prodeduce. In particular, they propose a specific network
+architecture that includes a predictor and CE generator network,
 
@@
 
-propose a way to train models that are guaranteed to provide recourse
-for individuals with high probability. The approach builds on
-adversarial training,
+\cite{ross2021learning} propose a way to train models that are
+guaranteed to provide recourse for individuals with high probability.
+The approach builds on adversarial training,
 
@@
 
-introduce Counterfactual Adversarial Training (CAT) with intention of
-improving generalization and robustness of language models.
-Specifically, they propose to proceed as follows: firstly, identify
-training samples that are subject to high predictive uncertainty
-(entropy); secondly, generate counterfactual explanations for those
-samples; and, finally, finetune the model on the augmented dataset that
-includes the generated counterfactuals.
+\cite{luu2023counterfactual} introduce Counterfactual Adversarial
+Training (CAT) with intention of improving generalization and robustness
+of language models. Specifically, they propose to proceed as follows:
+firstly, identify training samples that are subject to high predictive
+uncertainty (entropy); secondly, generate counterfactual explanations
+for those samples; and, finally, finetune the model on the augmented
+dataset that includes the generated counterfactuals.

### 2025-01-23 paper/related.qmd
bibliography sorted and everything adapted to ECML
noice

@@
 ## Background on Counterfactual Explanations
+
+[@wachter2017counterfactual;@joshi2019realistic;@altmeyer2024faithful]
 
@@
 @teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as *gradient supervision*)
-
-> This might be useful for our task as well...
-
-In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: \cite{wu2021polyjuice}, propose POLYJUICE, a general-purpose counterfactual generator for language models.
+Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as *gradient supervision*) (***this might be useful for our task as well***).
+In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose POLYJUICE, a general-purpose counterfactual generator for language models.
 They demonstrate empirically that augmenting training data through POLYJUICE counterfactuals improves robustness in a number of NLP tasks.
@@
 
-@freiesleben2022intriguing propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE;
-
-> this might be a useful notion for use to distinguish between adversarials and explanations during training
-
-secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations.
+@freiesleben2022intriguing propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE (***this might be a useful notion for use to distinguish between adversarials and explanations during training***); secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations.
 @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.
@@
 @guo2023counternet are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce.
-In particular, they propose a specific network architecture that includes a predictor and CE generator network,
-
-> akin a GAN?
-
-where the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed back to the predictor network (\textcolor{red}{here we are aligned}).
+In particular, they propose a specific network architecture that includes a predictor and CE generator network (***akin a GAN?***), where the parameters of the CE generator network are learnable.
+Counterfactuals are generated during each training iteration and fed back to the predictor network (***here we are aligned***).
 In contrast, we impose no restrictions on the neural network architecture at all.
-
-> to ensure the one-hot encoding of categorical features is maintained, they simple use softmax (might be interesting for CE.jl)
-
-Interestingly, the authors find that their approach is sensitive to the choice of the loss function: only MSE seems to lead to good performance.
+(***to ensure the one-hot encoding of categorical features is maintained, they simple use softmax (might be interesting for CE.jl)***) Interestingly, the authors find that their approach is sensitive to the choice of the loss function: only MSE seems to lead to good performance.
 They also demonstrate theoretically, that the objective function is difficult to optimize due to divergent gradients and suffers from poor adversarial robustness.
-
-> because partial gradients with respect to the classification loss component and the counterfactual validity component point in opposite directions
-
-To mitigate these issues, the authors use block-wise gradient descent: they first update with respect to classification loss and then use a second update with respect to the other loss components
-
-> this might be useful for our task as well
-
-@ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals with high probability. The approach builds on adversarial training,
-
-> here we are aligned
-
-where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class.
-The proposed method allows for imposing a set of actionable recourse ex-ante: for example, users can impose mutability constraints for features.
-
-> here we are aligned
-
-> To solve their objective function more efficiently, they use a first-order Taylor approximation to approximate the recourse loss component (might be applicable in our case
+(***because partial gradients with respect to the classification loss component and the counterfactual validity component point in opposite directions***).
+To mitigate these issues, the authors use block-wise gradient descent: they first update with respect to classification loss and then use a second update with respect to the other loss components (***this might be useful for our task as well***).
+@ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals with high probability.
+The approach builds on adversarial training (***here we are aligned***), where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class.
+The proposed method allows for imposing a set of actionable recourse ex-ante: for example, users can impose mutability constraints for features (***here we are aligned***).
+(***To solve their objective function more efficiently, they use a first-order Taylor approximation to approximate the recourse loss component (might be applicable in our case)***)
 

### 2025-01-23 paper/related.tex
noice

@@
 counterfactual inputs using the cosine distance as a loss function
-(referred to as \emph{gradient supervision})
-
-\begin{quote}
-This might be useful for our task as well\ldots{}
-\end{quote}
-
-In the natural language processing (NLP) domain, counterfactuals have
-similarly been used to improve models through data augmentation:
-\cite{wu2021polyjuice}, propose POLYJUICE, a general-purpose
-counterfactual generator for language models. They demonstrate
-empirically that augmenting training data through POLYJUICE
+(referred to as \emph{gradient supervision}) (\textbf{\emph{this might
+be useful for our task as well}}). In the natural language processing
+(NLP) domain, counterfactuals have similarly been used to improve models
+through data augmentation: \cite{wu2021polyjuice}, propose POLYJUICE, a
+general-purpose counterfactual generator for language models. They
+demonstrate empirically that augmenting training data through POLYJUICE
 counterfactuals improves robustness in a number of NLP tasks.
@@
 firstly, and more importantly according to the authors, the term AE
-implies missclassification, which is not the case for CE;
-
-\begin{quote}
-this might be a useful notion for use to distinguish between
-adversarials and explanations during training
-\end{quote}
-
-secondly, they argue that closeness plays a more critical role in the
-context of CE but confess that even counterfactuals that are not close
-might be relevant explanations. \cite{pawelczyk2022exploring} show that
-CE and AE are equivalent under certain conditions and derive upper
-bounds on the distances between them.
+implies missclassification, which is not the case for CE
+(\textbf{\emph{this might be a useful notion for use to distinguish
+between adversarials and explanations during training}}); secondly, they
+argue that closeness plays a more critical role in the context of CE but
+confess that even counterfactuals that are not close might be relevant
+explanations. \cite{pawelczyk2022exploring} show that CE and AE are
+equivalent under certain conditions and derive upper bounds on the
+distances between them.
 
@@
 training prodeduce. In particular, they propose a specific network
-architecture that includes a predictor and CE generator network,
-
-\begin{quote}
-akin a GAN?
-\end{quote}
-
-where the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed
-back to the predictor network (\textcolor{red}{here we are aligned}). In
-contrast, we impose no restrictions on the neural network architecture
-at all.
-
-\begin{quote}
-to ensure the one-hot encoding of categorical features is maintained,
-they simple use softmax (might be interesting for CE.jl)
-\end{quote}
-
+architecture that includes a predictor and CE generator network
+(\textbf{\emph{akin a GAN?}}), where the parameters of the CE generator
+network are learnable. Counterfactuals are generated during each
+training iteration and fed back to the predictor network
+(\textbf{\emph{here we are aligned}}). In contrast, we impose no
+restrictions on the neural network architecture at all.
+(\textbf{\emph{to ensure the one-hot encoding of categorical features is
+maintained, they simple use softmax (might be interesting for CE.jl)}})
 Interestingly, the authors find that their approach is sensitive to the
@@
 difficult to optimize due to divergent gradients and suffers from poor
-adversarial robustness.
-
-\begin{quote}
-because partial gradients with respect to the classification loss
-component and the counterfactual validity component point in opposite
-directions
-\end{quote}
-
-To mitigate these issues, the authors use block-wise gradient descent:
-they first update with respect to classification loss and then use a
-second update with respect to the other loss components
-
-\begin{quote}
-this might be useful for our task as well
-\end{quote}
-
-\cite{ross2021learning} propose a way to train models that are
-guaranteed to provide recourse for individuals with high probability.
-The approach builds on adversarial training,
-
-\begin{quote}
-here we are aligned
-\end{quote}
-
-where in this context adversarial examples are actively encouraged to
-exist, but only target attacks with respect to the positive class. The
-proposed method allows for imposing a set of actionable recourse
-ex-ante: for example, users can impose mutability constraints for
-features.
-
-\begin{quote}
-here we are aligned
-\end{quote}
-
-\begin{quote}
-To solve their objective function more efficiently, they use a
-first-order Taylor approximation to approximate the recourse loss
-component (might be applicable in our case
-\end{quote}
+adversarial robustness. (\textbf{\emph{because partial gradients with
+respect to the classification loss component and the counterfactual
+validity component point in opposite directions}}). To mitigate these
+issues, the authors use block-wise gradient descent: they first update
+with respect to classification loss and then use a second update with
+respect to the other loss components (\textbf{\emph{this might be useful
+for our task as well}}). \cite{ross2021learning} propose a way to train
+models that are guaranteed to provide recourse for individuals with high
+probability. The approach builds on adversarial training
+(\textbf{\emph{here we are aligned}}), where in this context adversarial
+examples are actively encouraged to exist, but only target attacks with
+respect to the positive class. The proposed method allows for imposing a
+set of actionable recourse ex-ante: for example, users can impose
+mutability constraints for features (\textbf{\emph{here we are
+aligned}}). (\textbf{\emph{To solve their objective function more
+efficiently, they use a first-order Taylor approximation to approximate
+the recourse loss component (might be applicable in our case)}})
 

### 2025-01-24 paper/appendix.qmd
might need plots after all

@@
----
-format:
-  latex:
-    template: bare.tex
----
+```{julia}
+using CTExperiments
+using CTExperiments.CSV
+using CTExperiments.DataFrames
+using CTExperiments.StatsBase
+using DotEnv
 
-## Appendix
+DotEnv.load!()
+```
 
-Here's my awesome appendix.
+# Appendix
+
+## Initial Grid Search
+
+```{julia}
+results_dir = ENV["INITIAL_RUN"]
+```
+
+### Generator Params
+
+```{julia}
+gen_params_dir = joinpath(results_dir, "gen_params/mlp")
+```
+
+#### Linearly Separable
+
+```{julia}
+df = CSV.read("$(gen_params_dir)/lin_sep/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
+df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
+  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
+  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) |>
+  df -> subset(df, :lambda_energy_exper => x -> x .>= 0.5)
+tabulate_results(df, "value"; alpha=0.9)
+```
+
+#### Moons
+
+```{julia}
+df = CSV.read("$(gen_params_dir)/moons/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
+df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
+  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
+  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) |>
+  df -> subset(df, :lambda_energy_exper => x -> x .>= 0.5)
+tabulate_results(df, "value"; alpha=0.9)
+```
+
+#### Circles
+
+```{julia}
+df = CSV.read("$(gen_params_dir)/circles/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
+df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
+  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
+  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) |>
+  df -> subset(df, :lambda_energy_exper => x -> x .>= 0.5)
+tabulate_results(df, "value"; alpha=0.9)
+```

### 2025-01-24 paper/experiments/CTExperiments/src/tables.jl

@@ -9,6 +9,7 @@ function tabulate_results(
     backend::Val=Val(:text),
     kwrgs...
 )
+    df = deepcopy(df)
 
     # Highlighters:
     hls = (value_highlighter(df, value_var; backend=backend, kwrgs...),)

### 2025-01-27 paper/appendix.qmd
work on tables

--- 2025-01-23 paper/appendix.qmd
@@
 ---
-format:
-  latex:
-    template: bare.tex
+engine: julia
+julia: 
+  exeflags: ["--project=paper/experiments/"]
 ---
 
-## Appendix
+```{julia}
+using CTExperiments
+using CTExperiments.CSV
+using CTExperiments.DataFrames
+using CTExperiments.StatsBase
+using DotEnv
 
-Here's my awesome appendix.
+DotEnv.load!()
+```
+
+# Appendix
+
+## Initial Grid Search
+
+```{julia}
+results_dir = ENV["INITIAL_RUN"]
+```
+
+### Generator Params
+
+```{julia}
+gen_params_dir = joinpath(results_dir, "gen_params/mlp")
+```
+
+#### Linearly Separable
+
+```{julia}
+df = CSV.read("$(gen_params_dir)/lin_sep/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
+df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
+  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
+  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) |>
+  df -> subset(df, :lambda_energy_exper => x -> x .>= 0.5)
+```
+
+::: {.content-hidden unless-format="pdf"}
+
+```{julia}
+get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
+    inputs -> tabulate_results(inputs; label="tab:lin_sep-first_run", title="Results for Linearly Separable data.")
+```
+
+:::
+
+#### Moons
+
+```{julia}
+df = CSV.read("$(gen_params_dir)/moons/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
+df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
+  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
+  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) |>
+  df -> subset(df, :lambda_energy_exper => x -> x .>= 0.5)
+```
+
+::: {.content-hidden unless-format="pdf"}
+
+```{julia}
+tabulate_results(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex))
+```
+
+:::
+
+#### Circles
+
+```{julia}
+df = CSV.read("$(gen_params_dir)/circles/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
+df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
+  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
+  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) |>
+  df -> subset(df, :lambda_energy_exper => x -> x .>= 0.5)
+```
+
+::: {.content-hidden unless-format="pdf"}
+
+```{julia}
+tabulate_results(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex))
+```
+
+:::

### 2025-01-27 paper/experiments/CTExperiments/src/plots.jl

@@ -96,11 +96,7 @@ function boxplot_ce(
     x::Union{Nothing,String}="generator_type",
     y::String="plausibility_distance_from_target",
     byvars::Union{Nothing,String,Vector{String}}=nothing,
-    colorvar::Union{Nothing,String}=nothing,
-    rowvar::Union{Nothing,String}=nothing,
-    colvar::Union{Nothing,String}=nothing,
-    facet=default_facet,
-    axis=default_axis,
+    kwrgs...
 )
     x = isnothing(x) ? "generator_type" : x
 
@@ -109,6 +105,21 @@ function boxplot_ce(
     # Aggregate:
     df_agg = aggregate_ce_evaluation(df, df_meta, df_eval; y=y, byvars=byvars)
 
+    # Plotting:
+    plt = boxplot_ce(df_agg,x;kwrgs...)
+
+    return plt, df_agg
+end
+
+function boxplot_ce(
+    df_agg::DataFrame,
+    x::Union{Nothing,String}="generator_type";
+    colorvar::Union{Nothing,String}=nothing,
+    rowvar::Union{Nothing,String}=nothing,
+    colvar::Union{Nothing,String}=nothing,
+    facet=default_facet,
+    axis=default_axis,
+)
     # Plotting:
     plt = data(df_agg) * mapping(Symbol(x), :mean => "Value") * visual(BoxPlot)
     if !isnothing(colorvar)
@@ -123,7 +134,7 @@ function boxplot_ce(
 
     plt = draw(plt; facet=facet, axis=axis)
 
-    return plt, df_agg
+    return plt
 end
 
 function plot_ce(cfg::EvalConfigOrGrid; save_dir=nothing, kwrgs...)

### 2025-01-27 paper/experiments/CTExperiments/src/presentation.jl

@@ -9,7 +9,7 @@ using TaijaPlotting
 include("plots.jl")
 include("tables.jl")
 
-export tabulate_results
+export tabulate_results, get_table_inputs
 
 function useful_byvars(df_meta::DataFrame)
     return names(df_meta)[findall([length(unique(x)) != 1 for x in eachcol(df_meta)])]

### 2025-01-27 paper/experiments/CTExperiments/src/tables.jl

@@ -4,6 +4,20 @@ using PrettyTables
 using StatsBase
 
 function tabulate_results(
+    inputs;
+    tf=PrettyTables.tf_latex_booktabs,
+    wrap_table=true,
+    table_type=:tabular,
+    kwrgs...,
+)
+    df = inputs[1]
+    other_inputs = inputs[2]
+    return pretty_table(
+        df; tf=tf, wrap_table=wrap_table, table_type=table_type, other_inputs..., kwrgs...
+    )
+end
+
+function get_table_inputs(
     df::DataFrame,
     value_var::String="mean";
     backend::Val=Val(:text),
@@ -12,7 +26,7 @@ function tabulate_results(
     df = deepcopy(df)
 
     # Highlighters:
-    hls = (value_highlighter(df, value_var; backend=backend, kwrgs...),)
+    hls = value_highlighter(df, value_var; backend=backend, kwrgs...)
     if "generator_type" in names(df)
         df.generator_type = format_generator.(df.generator_type)
         gen_hl = generator_highlighter(df; backend=backend)
@@ -20,7 +34,7 @@ function tabulate_results(
     end
 
     header = format_header.(names(df))
-    return pretty_table(df; highlighters=hls, backend=backend, header=header)
+    return df, (; highlighters=hls, backend=backend, header=header)
 end
 
 format_generator(s::AbstractString) = get_generator_name(generator_types[s](), pretty=true)
@@ -35,21 +49,41 @@ end
 function value_highlighter(
     df::DataFrame,
     value_var::String="mean";
-    scheme::ColorSchemes.ColorScheme=reverse(colorschemes[:rose]),
+    scheme::Union{Nothing,ColorSchemes.ColorScheme}=nothing,
     alpha::AbstractFloat=0.5,
-    backend::Val=Val(:text)
+    backend::Val=Val(:text),
+    byvars::Union{Nothing,String,Vector{String}}=nothing,
 )
     @assert value_var in names(df) "Provided variables $(value_var) is no in the dataframe"
     col_idx = findall(value_var .== names(df))
-    lb, ub = (alpha/2, 1 - alpha/2)
-    @info "Lower bound $(lb), Upper bound $(ub)"
-    lims = (quantile(df[:, value_var], lb), quantile(df[:, value_var], ub))
-    @info "Limits: $(lims)"
 
[…]

### 2025-01-27 paper/appendix.qmd
tables

@@
 julia: 
-  exeflags: ["--project=paper/experiments/"]
+  exeflags: ["--project=experiments/"]
 ---
@@
 ```{julia}
+cd("../")
+
 using CTExperiments
@@
 
+::: {#tbl-lin_sep-lambda_energy_exper}
+
 ```{julia}
+#| output: asis
 get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; label="tab:lin_sep-first_run", title="Results for Linearly Separable data.")
+    inputs -> tabulate_results(inputs)
 ```
+
+Results for Linearly Separable data by energy penalty.
+
+:::
 
@@
 ```{julia}
-tabulate_results(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex))
+#| output: asis
+get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
+    inputs -> tabulate_results(inputs; label="tab:moons-lambda_energy_exper", title="Results for Moons data by energy penalty.")
 ```
@@
 ```{julia}
-tabulate_results(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex))
+#| output: asis
+get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
+    inputs -> tabulate_results(inputs; label="tab:circles-lambda_energy_exper", title="Results for Circles data by energy penalty.")
 ```

### 2025-01-27 paper/appendix.tex
some work on streamline Quarto usage
tables

--- 2025-01-23 paper/appendix.qmd
@@
----
-format:
-  latex:
-    template: bare.tex
----
+\section{Appendix}\label{appendix}
 
-## Appendix
+\subsection{Initial Grid Search}\label{initial-grid-search}
 
-Here's my awesome appendix.
+\subsubsection{Generator Params}\label{generator-params}
+
+\paragraph{Linearly Separable}\label{linearly-separable}
+
+\begin{table}
+
+\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly
+Separable data by energy penalty.}
+
+\centering{
+
+  \begin{tabular}{rrrrr}
+    \toprule
+    \textbf{Objective} & \textbf{Lambda Energy (exper)} & \textbf{Generator Type} & \textbf{Value} & \textbf{Std} \\\midrule
+    full & 0.5 & \textit{ECCo} & -11.7601 & 9.83205 \\
+    full & 0.5 & \textit{Generic} & -1.064e18 & 2.41995e18 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5405}} & \color{blue}{\textbf{0.122789}} \\
+    full & 0.5 & \textit{REVISE} & -15.0277 & 12.5588 \\
+    vanilla & 0.5 & \textit{ECCo} & -4.39923 & 3.65268 \\
+    vanilla & 0.5 & \textit{Generic} & -4.38184 & 3.48393 \\
+    vanilla & 0.5 & \textit{Omniscient} & -5.24831 & 4.61237 \\
+    vanilla & 0.5 & \textit{REVISE} & -4.94731 & 4.2233 \\
+    full & 1.0 & \textit{ECCo} & -11.5401 & 11.0622 \\
+    full & 1.0 & \textit{Generic} & -1.70667e11 & 3.88205e11 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{1.0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.58956}} & \color{blue}{\textbf{0.117255}} \\
+    full & 1.0 & \textit{REVISE} & -15.7258 & 13.2676 \\
+    vanilla & 1.0 & \textit{ECCo} & -4.27742 & 3.50817 \\
+    vanilla & 1.0 & \textit{Generic} & -4.44409 & 3.4741 \\
+    vanilla & 1.0 & \textit{Omniscient} & -5.11353 & 4.4628 \\
+    vanilla & 1.0 & \textit{REVISE} & -4.91409 & 4.24885 \\
+    full & 5.0 & \textit{ECCo} & -3.99166 & 3.12284 \\
+    full & 5.0 & \textit{Generic} & -4.88333e17 & 1.11064e18 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{5.0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5325}} & \color{blue}{\textbf{0.117196}} \\
+    full & 5.0 & \textit{REVISE} & -14.5887 & 12.1265 \\
+    vanilla & 5.0 & \textit{ECCo} & -4.39614 & 3.64978 \\
+    vanilla & 5.0 & \textit{Generic} & -4.37909 & 3.48341 \\
+    vanilla & 5.0 & \textit{Omniscient} & -5.24668 & 4.60676 \\
+    vanilla & 5.0 & \textit{REVISE} & -4.94655 & 4.22198 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{10.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.30721}} & \color{blue}{\textbf{0.73475}} \\
+    full & 10.0 & \textit{Generic} & -1.69667e11 & 3.85893e11 \\
+    full & 10.0 & \textit{Omniscient} & -2.53433 & 0.116736 \\
+    full & 10.0 & \textit{REVISE} & -15.5346 & 13.0245 \\
+    vanilla & 10.0 & \textit{ECCo} & -4.28116 & 3.50992 \\
+    vanilla & 10.0 & \textit{Generic} & -4.4428 & 3.47049 \\
+    vanilla & 10.0 & \textit{Omniscient} & -5.11933 & 4.46099 \\
+    vanilla & 10.0 & \textit{REVISE} & -4.91285 & 4.24407 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{15.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.00576}} & \color{blue}{\textbf{0.48751}} \\
+    full & 15.0 & \textit{Generic} & -4.91e17 & 1.11683e18 \\
+    full & 15.0 & \textit{Omniscient} & -2.52833 & 0.11602 \\
+    full & 15.0 & \textit{REVISE} & -14.3763 & 11.7494 \\
+    vanilla & 15.0 & \textit{ECCo} & -4.3957 & 3.65194 \\
+    vanilla & 15.0 & \textit{Generic} & -4.38497 & 3.48359 \\
+    vanilla & 15.0 & \textit{Omniscient} & -5.24893 & 4.60484 \\
+    vanilla & 15.0 & \textit{REVISE} & -4.94518 & 4.22746 \\\bottomrule
+  \end{tabular}
+
+}
+
+\end{table}%
+
+\paragraph{Moons}\label{moons}
+
+\begin{table}
+  \caption{Results for Moons data by energy penalty.}
+  \begin{tabular}{rrrrr}
+    \toprule
+    \textbf{Objective} & \textbf{Lambda Energy (exper)} & \textbf{Generator Type} & \textbf{Value} & \textbf{Std} \\\midrule
+    full & 0.5 & \textit{ECCo} & -7.08577 & 7.51393 \\
+    full & 0.5 & \textit{Generic} & -1.1064e31 & 2.53239e31 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.58057}} & \color{blue}{\textbf{4.8256}} \\
+    full & 0.5 & \textit{REVISE} & -1187.61 & 2643.72 \\
+    vanilla & 0.5 & \textit{ECCo} & -15.4966 & 17.1932 \\
+    vanilla & 0.5 & \textit{Generic} & -11.7071 & 12.8003 \\
+    vanilla & 0.5 & \textit{Omniscient} & -12.3897 & 14.1104 \\
+    vanilla & 0.5 & \textit{REVISE} & -11.2965 & 13.1122 \\
+    full & 1.0 & \textit{ECCo} & -6.06278 & 6.32519 \\
+    full & 1.0 & \textit{Generic} & -1.57758e33 & 3.59342e33 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{1.0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.66436}} & \color{blue}{\textbf{4.88547}} \\
+    full & 1.0 & \textit{REVISE} & -1157.26 & 2585.3 \\
+    vanilla & 1.0 & \textit{ECCo} & -15.4915 & 17.2592 \\
+    vanilla & 1.0 & \textit{Generic} & -10.8969 & 11.888 \\
+    vanilla & 1.0 & \textit{Omniscient} & -12.6685 & 14.4499 \\
+    vanilla & 1.0 & \textit{REVISE} & -11.2874 & 13.1369 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{5.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.56504}} & \color{blue}{\textbf{2.06543}} \\
+    full & 5.0 & \textit{Generic} & -1.16971e28 & 2.66145e28 \\
+    full & 5.0 & \textit{Omniscient} & -4.28955 & 4.30748 \\
+    full & 5.0 & \textit{REVISE} & -530.204 & 1163.55 \\
+    vanilla & 5.0 & \textit{ECCo} & -15.4763 & 17.1877 \\
+    vanilla & 5.0 & \textit{Generic} & -11.6655 & 12.7364 \\
+    vanilla & 5.0 & \textit{Omniscient} & -12.3937 & 14.1141 \\
+    vanilla & 5.0 & \textit{REVISE} & -11.2976 & 13.0533 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{10.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.76439}} & \color{blue}{\textbf{0.973615}} \\
+    full & 10.0 & \textit{Generic} & -1.54318e33 & 3.51163e33 \\
+    full & 10.0 & \textit{Omniscient} & -4.44467 & 4.56008 \\
+    full & 10.0 & \textit{REVISE} & -1515.03 & 3402.96 \\
+    vanilla & 10.0 & \textit{ECCo} & -15.5074 & 17.275 \\
+    vanilla & 10.0 & \textit{Generic} & -10.9077 & 11.8867 \\
+    vanilla & 10.0 & \textit{Omniscient} & -12.6771 & 14.4225 \\
+    vanilla & 10.0 & \textit{REVISE} & -11.2735 & 13.1031 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{15.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.36625}} & \color{blue}{\textbf{0.3652}} \\
+    full & 15.0 & \textit{Generic} & -5.32108e28 & 1.21152e29 \\
+    full & 15.0 & \textit{Omniscient} & -4.34376 & 4.38045 \\
+    full & 15.0 & \textit{REVISE} & -473.027 & 1034.8 \\
+    vanilla & 15.0 & \textit{ECCo} & -15.4703 & 17.1898 \\
+    vanilla & 15.0 & \textit{Generic} & -11.6941 & 12.7669 \\
+    vanilla & 15.0 & \textit{Omniscient} & -12.3895 & 14.0956 \\
+    vanilla & 15.0 & \textit{REVISE} & -11.2868 & 13.0587 \\\bottomrule
+  \end{tabular}
+
+\end{table}
+
+\paragraph{Circles}\label{circles}
+
+\begin{table}
+  \caption{Results for Circles data by energy penalty.}
+  \begin{tabular}{rrrrr}
+    \toprule
+    \textbf{Objective} & \textbf{Lambda Energy (exper)} & \textbf{Generator Type} & \textbf{Value} & \textbf{Std} \\\midrule
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.12388}} & \color{blue}{\textbf{0.216889}} \\
+    full & 0.5 & \textit{Generic} & -1.20782 & 0.352005 \\
+    full & 0.5 & \textit{Omniscient} & -5.09228 & 5.1182 \\
+    full & 0.5 & \textit{REVISE} & -5.97244e27 & 1.36572e28 \\
+    vanilla & 0.5 & \textit{ECCo} & -9.35338 & 7.30155 \\
+    vanilla & 0.5 & \textit{Generic} & -8.89415 & 6.91671 \\
+    vanilla & 0.5 & \textit{Omniscient} & -8.67963 & 6.9307 \\
+    vanilla & 0.5 & \textit{REVISE} & -8.52507 & 6.74796 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{1.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.099}} & \color{blue}{\textbf{0.163365}} \\
+    full & 1.0 & \textit{Generic} & -1.49485 & 0.726287 \\
+    full & 1.0 & \textit{Omniscient} & -5.15975 & 5.20449 \\
+    full & 1.0 & \textit{REVISE} & -3.09069e26 & 7.22344e26 \\
+    vanilla & 1.0 & \textit{ECCo} & -9.33801 & 7.36386 \\
+    vanilla & 1.0 & \textit{Generic} & -8.8619 & 6.85196 \\
+    vanilla & 1.0 & \textit{Omniscient} & -8.69785 & 6.89941 \\
+    vanilla & 1.0 & \textit{REVISE} & -8.69498 & 6.85371 \\
+    full & 5.0 & \textit{ECCo} & -1.75204 & 0.154399 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{5.0}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21285}} & \color{blue}{\textbf{0.362686}} \\
+    full & 5.0 & \textit{Omniscient} & -5.13516 & 5.16338 \\
+    full & 5.0 & \textit{REVISE} & -1.09598e28 & 2.50339e28 \\
+    vanilla & 5.0 & \textit{ECCo} & -9.36397 & 7.32382 \\
+    vanilla & 5.0 & \textit{Generic} & -8.88498 & 6.90503 \\
+    vanilla & 5.0 & \textit{Omniscient} & -8.70333 & 6.9289 \\
+    vanilla & 5.0 & \textit{REVISE} & -8.51631 & 6.72565 \\
+    full & 10.0 & \textit{ECCo} & -1.01708e6 & 2.31516e6 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{10.0}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.48827}} & \color{blue}{\textbf{0.701741}} \\
+    full & 10.0 & \textit{Omniscient} & -5.13432 & 5.15897 \\
+    full & 10.0 & \textit{REVISE} & -3.74376e26 & 9.08858e26 \\
+    vanilla & 10.0 & \textit{ECCo} & -9.31463 & 7.32684 \\
+    vanilla & 10.0 & \textit{Generic} & -8.87348 & 6.86388 \\
+    vanilla & 10.0 & \textit{Omniscient} & -8.7046 & 6.89274 \\
+    vanilla & 10.0 & \textit{REVISE} & -8.68653 & 6.83497 \\
+    full & 15.0 & \textit{ECCo} & -3.31332e13 & 7.53714e13 \\
+    \color{blue}{\textbf{full}} & \color{blue}{\textbf{15.0}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21817}} & \color{blue}{\textbf{0.370377}} \\
+    full & 15.0 & \textit{Omniscient} & -5.19548 & 5.23317 \\
+    full & 15.0 & \textit{REVISE} & -9.01467e27 & 2.0592e28 \\
+    vanilla & 15.0 & \textit{ECCo} & -9.37662 & 7.34277 \\
+    vanilla & 15.0 & \textit{Generic} & -8.86149 & 6.8695 \\
+    vanilla & 15.0 & \textit{Omniscient} & -8.69488 & 6.95691 \\
+    vanilla & 15.0 & \textit{REVISE} & -8.50583 & 6.72685 \\\bottomrule
+  \end{tabular}
+
+\end{table}

### 2025-01-27 paper/appendix.tex
annex

--- 2025-01-23 paper/appendix.qmd
@@
----
-format:
-  latex:
-    template: bare.tex
----
+\section{Appendix}\label{appendix}
 
-## Appendix
+\subsection{Initial Grid Search}\label{initial-grid-search}
 
-Here's my awesome appendix.
+\subsubsection{Generator Params}\label{generator-params}
+
+\paragraph{Linearly Separable}\label{linearly-separable}
+
+\begin{longtable}{rrrrr}
+
+\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly
+Separable data by energy penalty.}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Objective} & \textbf{Lambda Energy (exper)} & \textbf{Generator Type} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Objective} & \textbf{Lambda Energy (exper)} & \textbf{Generator Type} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  full & 0.01 & \textit{ECCo} & -9.91333e11 & 2.25464e12 \\
+  full & 0.01 & \textit{Generic} & -5.71333e17 & 1.29951e18 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54312}} & \color{blue}{\textbf{0.115996}} \\
+  full & 0.01 & \textit{REVISE} & -15.6339 & 13.1813 \\
+  vanilla & 0.01 & \textit{ECCo} & -4.28089 & 3.52021 \\
+  vanilla & 0.01 & \textit{Generic} & -4.4476 & 3.47151 \\
+  vanilla & 0.01 & \textit{Omniscient} & -5.11878 & 4.45562 \\
+  vanilla & 0.01 & \textit{REVISE} & -4.91181 & 4.24403 \\
+  full & 0.05 & \textit{ECCo} & -5.63008e5 & 1.28084e6 \\
+  full & 0.05 & \textit{Generic} & -8.34667e17 & 1.89839e18 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53131}} & \color{blue}{\textbf{0.114159}} \\
+  full & 0.05 & \textit{REVISE} & -15.0237 & 12.5878 \\
+  vanilla & 0.05 & \textit{ECCo} & -4.40037 & 3.6585 \\
+  vanilla & 0.05 & \textit{Generic} & -4.38095 & 3.48267 \\
+  vanilla & 0.05 & \textit{Omniscient} & -5.24923 & 4.6174 \\
+  vanilla & 0.05 & \textit{REVISE} & -4.9445 & 4.21764 \\
+  full & 0.1 & \textit{ECCo} & -674007.0 & 1.53289e6 \\
+  full & 0.1 & \textit{Generic} & -1.71667e11 & 3.90456e11 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.56041}} & \color{blue}{\textbf{0.124295}} \\
+  full & 0.1 & \textit{REVISE} & -15.6368 & 13.2452 \\
+  vanilla & 0.1 & \textit{ECCo} & -4.28229 & 3.51532 \\
+  vanilla & 0.1 & \textit{Generic} & -4.44665 & 3.47681 \\
+  vanilla & 0.1 & \textit{Omniscient} & -5.11633 & 4.464 \\
+  vanilla & 0.1 & \textit{REVISE} & -4.91309 & 4.25104 \\
+  full & 0.5 & \textit{ECCo} & -11.7601 & 9.83205 \\
+  full & 0.5 & \textit{Generic} & -1.064e18 & 2.41995e18 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5405}} & \color{blue}{\textbf{0.122789}} \\
+  full & 0.5 & \textit{REVISE} & -15.0277 & 12.5588 \\
+  vanilla & 0.5 & \textit{ECCo} & -4.39923 & 3.65268 \\
+  vanilla & 0.5 & \textit{Generic} & -4.38184 & 3.48393 \\
+  vanilla & 0.5 & \textit{Omniscient} & -5.24831 & 4.61237 \\
+  vanilla & 0.5 & \textit{REVISE} & -4.94731 & 4.2233 \\
+  full & 1.0 & \textit{ECCo} & -11.5401 & 11.0622 \\
+  full & 1.0 & \textit{Generic} & -1.70667e11 & 3.88205e11 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1.0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.58956}} & \color{blue}{\textbf{0.117255}} \\
+  full & 1.0 & \textit{REVISE} & -15.7258 & 13.2676 \\
+  vanilla & 1.0 & \textit{ECCo} & -4.27742 & 3.50817 \\
+  vanilla & 1.0 & \textit{Generic} & -4.44409 & 3.4741 \\
+  vanilla & 1.0 & \textit{Omniscient} & -5.11353 & 4.4628 \\
+  vanilla & 1.0 & \textit{REVISE} & -4.91409 & 4.24885 \\
+  full & 5.0 & \textit{ECCo} & -3.99166 & 3.12284 \\
+  full & 5.0 & \textit{Generic} & -4.88333e17 & 1.11064e18 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5.0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5325}} & \color{blue}{\textbf{0.117196}} \\
+  full & 5.0 & \textit{REVISE} & -14.5887 & 12.1265 \\
+  vanilla & 5.0 & \textit{ECCo} & -4.39614 & 3.64978 \\
+  vanilla & 5.0 & \textit{Generic} & -4.37909 & 3.48341 \\
+  vanilla & 5.0 & \textit{Omniscient} & -5.24668 & 4.60676 \\
+  vanilla & 5.0 & \textit{REVISE} & -4.94655 & 4.22198 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.30721}} & \color{blue}{\textbf{0.73475}} \\
+  full & 10.0 & \textit{Generic} & -1.69667e11 & 3.85893e11 \\
+  full & 10.0 & \textit{Omniscient} & -2.53433 & 0.116736 \\
+  full & 10.0 & \textit{REVISE} & -15.5346 & 13.0245 \\
+  vanilla & 10.0 & \textit{ECCo} & -4.28116 & 3.50992 \\
+  vanilla & 10.0 & \textit{Generic} & -4.4428 & 3.47049 \\
+  vanilla & 10.0 & \textit{Omniscient} & -5.11933 & 4.46099 \\
+  vanilla & 10.0 & \textit{REVISE} & -4.91285 & 4.24407 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.00576}} & \color{blue}{\textbf{0.48751}} \\
+  full & 15.0 & \textit{Generic} & -4.91e17 & 1.11683e18 \\
+  full & 15.0 & \textit{Omniscient} & -2.52833 & 0.11602 \\
+  full & 15.0 & \textit{REVISE} & -14.3763 & 11.7494 \\
+  vanilla & 15.0 & \textit{ECCo} & -4.3957 & 3.65194 \\
+  vanilla & 15.0 & \textit{Generic} & -4.38497 & 3.48359 \\
+  vanilla & 15.0 & \textit{Omniscient} & -5.24893 & 4.60484 \\
+  vanilla & 15.0 & \textit{REVISE} & -4.94518 & 4.22746 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Moons}\label{moons}
+
+\begin{longtable}{rrrrr}
+
+\caption{\label{tbl-moons-lambda_energy_exper}Results for Moons data by
+energy penalty.}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Objective} & \textbf{Lambda Energy (exper)} & \textbf{Generator Type} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Objective} & \textbf{Lambda Energy (exper)} & \textbf{Generator Type} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  full & 0.01 & \textit{ECCo} & -2.7995e22 & 6.38864e22 \\
+  full & 0.01 & \textit{Generic} & -4.8906e30 & 1.11313e31 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.73844}} & \color{blue}{\textbf{5.0839}} \\
+  full & 0.01 & \textit{REVISE} & -572.158 & 1245.03 \\
+  vanilla & 0.01 & \textit{ECCo} & -15.5007 & 17.2501 \\
+  vanilla & 0.01 & \textit{Generic} & -10.9068 & 11.8954 \\
+  vanilla & 0.01 & \textit{Omniscient} & -12.6633 & 14.3968 \\
+  vanilla & 0.01 & \textit{REVISE} & -11.2239 & 13.0133 \\
+  full & 0.05 & \textit{ECCo} & -1.54854e16 & 3.52394e16 \\
+  full & 0.05 & \textit{Generic} & -2.21669e20 & 5.00175e20 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.41}} & \color{blue}{\textbf{4.48466}} \\
+  full & 0.05 & \textit{REVISE} & -1036.78 & 2300.0 \\
+  vanilla & 0.05 & \textit{ECCo} & -15.4978 & 17.2215 \\
+  vanilla & 0.05 & \textit{Generic} & -11.697 & 12.8139 \\
+  vanilla & 0.05 & \textit{Omniscient} & -12.4116 & 14.1033 \\
+  vanilla & 0.05 & \textit{REVISE} & -11.3225 & 13.1404 \\
+  full & 0.1 & \textit{ECCo} & -3408.28 & 7734.83 \\
+  full & 0.1 & \textit{Generic} & -5.21659e30 & 1.18648e31 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.78488}} & \color{blue}{\textbf{5.1212}} \\
+  full & 0.1 & \textit{REVISE} & -287.874 & 593.988 \\
+  vanilla & 0.1 & \textit{ECCo} & -15.4788 & 17.2353 \\
+  vanilla & 0.1 & \textit{Generic} & -10.9226 & 11.9247 \\
+  vanilla & 0.1 & \textit{Omniscient} & -12.666 & 14.3869 \\
+  vanilla & 0.1 & \textit{REVISE} & -11.2936 & 13.1035 \\
+  full & 0.5 & \textit{ECCo} & -7.08577 & 7.51393 \\
+  full & 0.5 & \textit{Generic} & -1.1064e31 & 2.53239e31 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.58057}} & \color{blue}{\textbf{4.8256}} \\
+  full & 0.5 & \textit{REVISE} & -1187.61 & 2643.72 \\
+  vanilla & 0.5 & \textit{ECCo} & -15.4966 & 17.1932 \\
+  vanilla & 0.5 & \textit{Generic} & -11.7071 & 12.8003 \\
+  vanilla & 0.5 & \textit{Omniscient} & -12.3897 & 14.1104 \\
+  vanilla & 0.5 & \textit{REVISE} & -11.2965 & 13.1122 \\
+  full & 1.0 & \textit{ECCo} & -6.06278 & 6.32519 \\
+  full & 1.0 & \textit{Generic} & -1.57758e33 & 3.59342e33 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1.0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.66436}} & \color{blue}{\textbf{4.88547}} \\
+  full & 1.0 & \textit{REVISE} & -1157.26 & 2585.3 \\
+  vanilla & 1.0 & \textit{ECCo} & -15.4915 & 17.2592 \\
+  vanilla & 1.0 & \textit{Generic} & -10.8969 & 11.888 \\
+  vanilla & 1.0 & \textit{Omniscient} & -12.6685 & 14.4499 \\
+  vanilla & 1.0 & \textit{REVISE} & -11.2874 & 13.1369 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.56504}} & \color{blue}{\textbf{2.06543}} \\
+  full & 5.0 & \textit{Generic} & -1.16971e28 & 2.66145e28 \\
+  full & 5.0 & \textit{Omniscient} & -4.28955 & 4.30748 \\
+  full & 5.0 & \textit{REVISE} & -530.204 & 1163.55 \\
+  vanilla & 5.0 & \textit{ECCo} & -15.4763 & 17.1877 \\
+  vanilla & 5.0 & \textit{Generic} & -11.6655 & 12.7364 \\
+  vanilla & 5.0 & \textit{Omniscient} & -12.3937 & 14.1141 \\
+  vanilla & 5.0 & \textit{REVISE} & -11.2976 & 13.0533 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.76439}} & \color{blue}{\textbf{0.973615}} \\
+  full & 10.0 & \textit{Generic} & -1.54318e33 & 3.51163e33 \\
+  full & 10.0 & \textit{Omniscient} & -4.44467 & 4.56008 \\
+  full & 10.0 & \textit{REVISE} & -1515.03 & 3402.96 \\
+  vanilla & 10.0 & \textit{ECCo} & -15.5074 & 17.275 \\
+  vanilla & 10.0 & \textit{Generic} & -10.9077 & 11.8867 \\
+  vanilla & 10.0 & \textit{Omniscient} & -12.6771 & 14.4225 \\
+  vanilla & 10.0 & \textit{REVISE} & -11.2735 & 13.1031 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.36625}} & \color{blue}{\textbf{0.3652}} \\
+  full & 15.0 & \textit{Generic} & -5.32108e28 & 1.21152e29 \\
+  full & 15.0 & \textit{Omniscient} & -4.34376 & 4.38045 \\
+  full & 15.0 & \textit{REVISE} & -473.027 & 1034.8 \\
+  vanilla & 15.0 & \textit{ECCo} & -15.4703 & 17.1898 \\
+  vanilla & 15.0 & \textit{Generic} & -11.6941 & 12.7669 \\
+  vanilla & 15.0 & \textit{Omniscient} & -12.3895 & 14.0956 \\
+  vanilla & 15.0 & \textit{REVISE} & -11.2868 & 13.0587 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Circles}\label{circles}
+
+\begin{longtable}{rrrrr}
+
+\caption{\label{tbl-circles-lambda_energy_exper}Results for Circles data
+by energy penalty.}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Objective} & \textbf{Lambda Energy (exper)} & \textbf{Generator Type} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Objective} & \textbf{Lambda Energy (exper)} & \textbf{Generator Type} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.26251}} & \color{blue}{\textbf{0.422879}} \\
+  full & 0.01 & \textit{Generic} & -1.49019 & 0.710317 \\
+  full & 0.01 & \textit{Omniscient} & -5.20834 & 5.24965 \\
+  full & 0.01 & \textit{REVISE} & -2.71465e26 & 6.36795e26 \\
+  vanilla & 0.01 & \textit{ECCo} & -9.32869 & 7.33883 \\
+  vanilla & 0.01 & \textit{Generic} & -8.88908 & 6.88126 \\
+  vanilla & 0.01 & \textit{Omniscient} & -8.67329 & 6.86953 \\
+  vanilla & 0.01 & \textit{REVISE} & -8.65247 & 6.79692 \\
+  full & 0.05 & \textit{ECCo} & -1.2868 & 0.396551 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.20911}} & \color{blue}{\textbf{0.356273}} \\
+  full & 0.05 & \textit{Omniscient} & -5.0777 & 5.09456 \\
+  full & 0.05 & \textit{REVISE} & -5.91165e27 & 1.36148e28 \\
+  vanilla & 0.05 & \textit{ECCo} & -9.35176 & 7.31732 \\
+  vanilla & 0.05 & \textit{Generic} & -8.854 & 6.87298 \\
+  vanilla & 0.05 & \textit{Omniscient} & -8.70242 & 6.95789 \\
+  vanilla & 0.05 & \textit{REVISE} & -8.52208 & 6.75727 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.2021}} & \color{blue}{\textbf{0.382734}} \\
+  full & 0.1 & \textit{Generic} & -1.50229 & 0.735365 \\
+  full & 0.1 & \textit{Omniscient} & -5.17466 & 5.22628 \\
+  full & 0.1 & \textit{REVISE} & -3.0594e26 & 7.70324e26 \\
+  vanilla & 0.1 & \textit{ECCo} & -9.32978 & 7.32099 \\
+  vanilla & 0.1 & \textit{Generic} & -8.88085 & 6.86308 \\
+  vanilla & 0.1 & \textit{Omniscient} & -8.68505 & 6.90258 \\
+  vanilla & 0.1 & \textit{REVISE} & -8.67959 & 6.80887 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.12388}} & \color{blue}{\textbf{0.216889}} \\
+  full & 0.5 & \textit{Generic} & -1.20782 & 0.352005 \\
+  full & 0.5 & \textit{Omniscient} & -5.09228 & 5.1182 \\
+  full & 0.5 & \textit{REVISE} & -5.97244e27 & 1.36572e28 \\
+  vanilla & 0.5 & \textit{ECCo} & -9.35338 & 7.30155 \\
+  vanilla & 0.5 & \textit{Generic} & -8.89415 & 6.91671 \\
+  vanilla & 0.5 & \textit{Omniscient} & -8.67963 & 6.9307 \\
+  vanilla & 0.5 & \textit{REVISE} & -8.52507 & 6.74796 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1.0}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.099}} & \color{blue}{\textbf{0.163365}} \\
+  full & 1.0 & \textit{Generic} & -1.49485 & 0.726287 \\
+  full & 1.0 & \textit{Omniscient} & -5.15975 & 5.20449 \\
+  full & 1.0 & \textit{REVISE} & -3.09069e26 & 7.22344e26 \\
+  vanilla & 1.0 & \textit{ECCo} & -9.33801 & 7.36386 \\
+  vanilla & 1.0 & \textit{Generic} & -8.8619 & 6.85196 \\
+  vanilla & 1.0 & \textit{Omniscient} & -8.69785 & 6.89941 \\
+  vanilla & 1.0 & \textit{REVISE} & -8.69498 & 6.85371 \\
+  full & 5.0 & \textit{ECCo} & -1.75204 & 0.154399 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5.0}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21285}} & \color{blue}{\textbf{0.362686}} \\
+  full & 5.0 & \textit{Omniscient} & -5.13516 & 5.16338 \\
+  full & 5.0 & \textit{REVISE} & -1.09598e28 & 2.50339e28 \\
+  vanilla & 5.0 & \textit{ECCo} & -9.36397 & 7.32382 \\
+  vanilla & 5.0 & \textit{Generic} & -8.88498 & 6.90503 \\
+  vanilla & 5.0 & \textit{Omniscient} & -8.70333 & 6.9289 \\
+  vanilla & 5.0 & \textit{REVISE} & -8.51631 & 6.72565 \\
+  full & 10.0 & \textit{ECCo} & -1.01708e6 & 2.31516e6 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10.0}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.48827}} & \color{blue}{\textbf{0.701741}} \\
+  full & 10.0 & \textit{Omniscient} & -5.13432 & 5.15897 \\
+  full & 10.0 & \textit{REVISE} & -3.74376e26 & 9.08858e26 \\
+  vanilla & 10.0 & \textit{ECCo} & -9.31463 & 7.32684 \\
+  vanilla & 10.0 & \textit{Generic} & -8.87348 & 6.86388 \\
+  vanilla & 10.0 & \textit{Omniscient} & -8.7046 & 6.89274 \\
+  vanilla & 10.0 & \textit{REVISE} & -8.68653 & 6.83497 \\
+  full & 15.0 & \textit{ECCo} & -3.31332e13 & 7.53714e13 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15.0}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21817}} & \color{blue}{\textbf{0.370377}} \\
+  full & 15.0 & \textit{Omniscient} & -5.19548 & 5.23317 \\
+  full & 15.0 & \textit{REVISE} & -9.01467e27 & 2.0592e28 \\
+  vanilla & 15.0 & \textit{ECCo} & -9.37662 & 7.34277 \\
+  vanilla & 15.0 & \textit{Generic} & -8.86149 & 6.8695 \\
+  vanilla & 15.0 & \textit{Omniscient} & -8.69488 & 6.95691 \\
+  vanilla & 15.0 & \textit{REVISE} & -8.50583 & 6.72685 \\\bottomrule
+
+\end{longtable}

### 2025-01-27 paper/experiments/CTExperiments/src/tables.jl

@@ -7,13 +7,20 @@ function tabulate_results(
     inputs;
     tf=PrettyTables.tf_latex_booktabs,
     wrap_table=true,
-    table_type=:tabular,
+    table_type=:longtable,
+    longtable_footer="Continuing table below.",
     kwrgs...,
 )
     df = inputs[1]
     other_inputs = inputs[2]
     return pretty_table(
-        df; tf=tf, wrap_table=wrap_table, table_type=table_type, other_inputs..., kwrgs...
+        df;
+        tf=tf,
+        wrap_table=wrap_table,
+        table_type=table_type,
+        longtable_footer=longtable_footer,
+        other_inputs...,
+        kwrgs...,
     )
 end
 

### 2025-01-27 paper/appendix.qmd
fuck latex

@@
 
+- **Energy Penalty**: *ECCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training.
+*Generic* performs poorly accross the board.
+*Omni* seems to have an anchoring effect, in that it never performs terribly but also never as good as the best *ECCo* results.
+*REVISE* performs poorly across the board.
+- **Cost (distance penalty)**: Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged).
+For *ECCo* and *Generic*, higher cost values generally lead to worse results.
+- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
+- **Validity**: *ECCo* almost always valid except for very low values during training and high values at evaluation time. *Generic* often has poor validity.
+- **Accuracy**: Seems largely unaffected.
+
 ```{julia}
@@
   gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
-  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) |>
-  df -> subset(df, :lambda_energy_exper => x -> x .>= 0.5)
+  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type])
 ```
@@
 
-::: {#tbl-lin_sep-lambda_energy_exper}
-
 ```{julia}
@@
 get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs)
+    inputs -> tabulate_results(inputs; label="tbl-moons-lambda_energy_exper", title="Results for Linearly Separable data by energy penalty.")
 ```
-
-Results for Linearly Separable data by energy penalty.
-
-:::
 
@@
   gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
-  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) |>
-  df -> subset(df, :lambda_energy_exper => x -> x .>= 0.5)
+  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) 
 ```
+
+::: {#tbl-moons-lambda_energy_exper}
 
@@
 get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; label="tab:moons-lambda_energy_exper", title="Results for Moons data by energy penalty.")
+    inputs -> tabulate_results(inputs)
 ```
+
+:::
+
+Results for Moons data by energy penalty.
 
@@
   gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
-  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) |>
-  df -> subset(df, :lambda_energy_exper => x -> x .>= 0.5)
+  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) 
 ```
+
+::: {#tbl-circles-lambda_energy_exper}
 
@@
 get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; label="tab:circles-lambda_energy_exper", title="Results for Circles data by energy penalty.")
+    inputs -> tabulate_results(inputs)
 ```
@@
 :::
+
+Results for Circles data by energy penalty.
+
+:::

### 2025-01-27 paper/experiments/CTExperiments/src/tables.jl

@@ -13,15 +13,29 @@ function tabulate_results(
 )
     df = inputs[1]
     other_inputs = inputs[2]
-    return pretty_table(
-        df;
-        tf=tf,
-        wrap_table=wrap_table,
-        table_type=table_type,
-        longtable_footer=longtable_footer,
-        other_inputs...,
-        kwrgs...,
-    )
+    if isa(other_inputs.backend,Val{:latex})
+        formatters = PrettyTables.ft_latex_sn(3)
+        tf = PrettyTables.tf_latex_booktabs
+        tab = pretty_table(
+            df;
+            tf=tf,
+            formatters=formatters,
+            wrap_table=wrap_table,
+            table_type=table_type,
+            longtable_footer=longtable_footer,
+            alignment=:c,
+            other_inputs...,
+            kwrgs...,
+        )
+    else
+        tab = pretty_table(
+            df;
+            alignment=:c,
+            other_inputs...,
+            kwrgs...,
+        )
+    end
+    return tab
 end
 
 function get_table_inputs(
@@ -47,9 +61,12 @@ end
 format_generator(s::AbstractString) = get_generator_name(generator_types[s](), pretty=true)
 
 function format_header(s::String)
-    s = split(s, "_") |>
+    s = replace(s, "_type" => "") |>
+        s -> replace(s, "objective" => "obj.") |>
+        s -> split(s, "_") |>
         ss -> [s in ["exper", "eval"] ? "($s)" : uppercasefirst(s) for s in ss] |>
-        ss -> join(ss, " ")
+        ss -> join(ss, " ") |>
+        s -> replace(s, "Lambda Energy" => "\$\\lambda\$") 
     return s
 end
 

### 2025-01-27 paper/appendix.qmd
waste of bloody time

@@
 
-- **Energy Penalty**: *ECCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training.
+- **Energy Penalty** (@tbl-lin_sep-lambda_energy_exper): *ECCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training.
 *Generic* performs poorly accross the board.
@@
 get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; label="tbl-moons-lambda_energy_exper", title="Results for Linearly Separable data by energy penalty.")
+    inputs -> tabulate_results(inputs; label="tbl-lin_sep-lambda_energy_exper", title="Results for Linearly Separable data by energy penalty.")
 ```
@@
 
-::: {#tbl-moons-lambda_energy_exper}
-
 ::: {.content-hidden unless-format="pdf"}
@@
 get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs)
+    inputs -> tabulate_results(inputs; label="tbl-moons-lambda_energy_exper", title="Results for Moons data by energy penalty.")
 ```
-
-:::
-
-Results for Moons data by energy penalty.
 
@@
 
-::: {#tbl-circles-lambda_energy_exper}
-
 ::: {.content-hidden unless-format="pdf"}
@@
 get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs)
+    inputs -> tabulate_results(inputs; label="tbl-circles-lambda_energy_exper", title="Results for Circles data by energy penalty.")
 ```
@@
 :::
-
-Results for Circles data by energy penalty.
-
-:::

### 2025-01-27 paper/appendix.tex
waste of bloody time

--- 2025-01-23 paper/appendix.qmd
@@
----
-format:
-  latex:
-    template: bare.tex
----
+\section{Appendix}\label{appendix}
 
-## Appendix
+\subsection{Initial Grid Search}\label{initial-grid-search}
 
-Here's my awesome appendix.
+\subsubsection{Generator Params}\label{generator-params}
+
+\paragraph{Linearly Separable}\label{linearly-separable}
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Energy Penalty} (Table~\ref{tbl-lin_sep-lambda_energy_exper}):
+  \emph{ECCo} generally does yield better results than \emph{Vanilla}
+  for higher choices of the energy penalty (10,15) during training.
+  \emph{Generic} performs poorly accross the board. \emph{Omni} seems to
+  have an anchoring effect, in that it never performs terribly but also
+  never as good as the best \emph{ECCo} results. \emph{REVISE} performs
+  poorly across the board.
+\item
+  \textbf{Cost (distance penalty)}: Results for all generators (except
+  \emph{Omni}) are quite bad, which can likely be attributed to
+  extremely bad results for some choices of the \textbf{Energy Penalty}
+  (results here are averaged). For \emph{ECCo} and \emph{Generic},
+  higher cost values generally lead to worse results.
+\item
+  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
+  seems that smaller choices are ok.
+\item
+  \textbf{Validity}: \emph{ECCo} almost always valid except for very low
+  values during training and high values at evaluation time.
+  \emph{Generic} often has poor validity.
+\item
+  \textbf{Accuracy}: Seems largely unaffected.
+\end{itemize}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly Separable data by energy penalty.}
+
+\tabularnewline
+
+  \\
+  \toprule
+  \textbf{Obj.} & \textbf{\textbackslash{}lambda (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Obj.} & \textbf{\textbackslash{}lambda (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  full & 0.01 & \textit{ECCo} & $-9.91 \cdot 10^{11}$ & $2.25 \cdot 10^{12}$ \\
+  full & 0.01 & \textit{Generic} & $-5.71 \cdot 10^{17}$ & $1.3 \cdot 10^{18}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.116}} \\
+  full & 0.01 & \textit{REVISE} & -15.6 & 13.2 \\
+  vanilla & 0.01 & \textit{ECCo} & -4.28 & 3.52 \\
+  vanilla & 0.01 & \textit{Generic} & -4.45 & 3.47 \\
+  vanilla & 0.01 & \textit{Omniscient} & -5.12 & 4.46 \\
+  vanilla & 0.01 & \textit{REVISE} & -4.91 & 4.24 \\
+  full & 0.05 & \textit{ECCo} & $-5.63 \cdot 10^{5}$ & $1.28 \cdot 10^{6}$ \\
+  full & 0.05 & \textit{Generic} & $-8.35 \cdot 10^{17}$ & $1.9 \cdot 10^{18}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.114}} \\
+  full & 0.05 & \textit{REVISE} & -15 & 12.6 \\
+  vanilla & 0.05 & \textit{ECCo} & -4.4 & 3.66 \\
+  vanilla & 0.05 & \textit{Generic} & -4.38 & 3.48 \\
+  vanilla & 0.05 & \textit{Omniscient} & -5.25 & 4.62 \\
+  vanilla & 0.05 & \textit{REVISE} & -4.94 & 4.22 \\
+  full & 0.1 & \textit{ECCo} & $-6.74 \cdot 10^{5}$ & $1.53 \cdot 10^{6}$ \\
+  full & 0.1 & \textit{Generic} & $-1.72 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.56}} & \color{blue}{\textbf{0.124}} \\
+  full & 0.1 & \textit{REVISE} & -15.6 & 13.2 \\
+  vanilla & 0.1 & \textit{ECCo} & -4.28 & 3.52 \\
+  vanilla & 0.1 & \textit{Generic} & -4.45 & 3.48 \\
+  vanilla & 0.1 & \textit{Omniscient} & -5.12 & 4.46 \\
+  vanilla & 0.1 & \textit{REVISE} & -4.91 & 4.25 \\
+  full & 0.5 & \textit{ECCo} & -11.8 & 9.83 \\
+  full & 0.5 & \textit{Generic} & $-1.06 \cdot 10^{18}$ & $2.42 \cdot 10^{18}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.123}} \\
+  full & 0.5 & \textit{REVISE} & -15 & 12.6 \\
+  vanilla & 0.5 & \textit{ECCo} & -4.4 & 3.65 \\
+  vanilla & 0.5 & \textit{Generic} & -4.38 & 3.48 \\
+  vanilla & 0.5 & \textit{Omniscient} & -5.25 & 4.61 \\
+  vanilla & 0.5 & \textit{REVISE} & -4.95 & 4.22 \\
+  full & 1 & \textit{ECCo} & -11.5 & 11.1 \\
+  full & 1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.88 \cdot 10^{11}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.59}} & \color{blue}{\textbf{0.117}} \\
+  full & 1 & \textit{REVISE} & -15.7 & 13.3 \\
+  vanilla & 1 & \textit{ECCo} & -4.28 & 3.51 \\
+  vanilla & 1 & \textit{Generic} & -4.44 & 3.47 \\
+  vanilla & 1 & \textit{Omniscient} & -5.11 & 4.46 \\
+  vanilla & 1 & \textit{REVISE} & -4.91 & 4.25 \\
+  full & 5 & \textit{ECCo} & -3.99 & 3.12 \\
+  full & 5 & \textit{Generic} & $-4.88 \cdot 10^{17}$ & $1.11 \cdot 10^{18}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.117}} \\
+  full & 5 & \textit{REVISE} & -14.6 & 12.1 \\
+  vanilla & 5 & \textit{ECCo} & -4.4 & 3.65 \\
+  vanilla & 5 & \textit{Generic} & -4.38 & 3.48 \\
+  vanilla & 5 & \textit{Omniscient} & -5.25 & 4.61 \\
+  vanilla & 5 & \textit{REVISE} & -4.95 & 4.22 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.31}} & \color{blue}{\textbf{0.735}} \\
+  full & 10 & \textit{Generic} & $-1.7 \cdot 10^{11}$ & $3.86 \cdot 10^{11}$ \\
+  full & 10 & \textit{Omniscient} & -2.53 & 0.117 \\
+  full & 10 & \textit{REVISE} & -15.5 & 13 \\
+  vanilla & 10 & \textit{ECCo} & -4.28 & 3.51 \\
+  vanilla & 10 & \textit{Generic} & -4.44 & 3.47 \\
+  vanilla & 10 & \textit{Omniscient} & -5.12 & 4.46 \\
+  vanilla & 10 & \textit{REVISE} & -4.91 & 4.24 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.01}} & \color{blue}{\textbf{0.488}} \\
+  full & 15 & \textit{Generic} & $-4.91 \cdot 10^{17}$ & $1.12 \cdot 10^{18}$ \\
+  full & 15 & \textit{Omniscient} & -2.53 & 0.116 \\
+  full & 15 & \textit{REVISE} & -14.4 & 11.7 \\
+  vanilla & 15 & \textit{ECCo} & -4.4 & 3.65 \\
+  vanilla & 15 & \textit{Generic} & -4.38 & 3.48 \\
+  vanilla & 15 & \textit{Omniscient} & -5.25 & 4.6 \\
+  vanilla & 15 & \textit{REVISE} & -4.95 & 4.23 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Moons}\label{moons}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-moons-lambda_energy_exper}Results for Moons data by energy penalty.}
+
+\tabularnewline
+
+  \\
+  \toprule
+  \textbf{Obj.} & \textbf{\textbackslash{}lambda (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Obj.} & \textbf{\textbackslash{}lambda (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  full & 0.01 & \textit{ECCo} & $-2.8 \cdot 10^{22}$ & $6.39 \cdot 10^{22}$ \\
+  full & 0.01 & \textit{Generic} & $-4.89 \cdot 10^{30}$ & $1.11 \cdot 10^{31}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.74}} & \color{blue}{\textbf{5.08}} \\
+  full & 0.01 & \textit{REVISE} & -572 & $1.25 \cdot 10^{3}$ \\
+  vanilla & 0.01 & \textit{ECCo} & -15.5 & 17.3 \\
+  vanilla & 0.01 & \textit{Generic} & -10.9 & 11.9 \\
+  vanilla & 0.01 & \textit{Omniscient} & -12.7 & 14.4 \\
+  vanilla & 0.01 & \textit{REVISE} & -11.2 & 13 \\
+  full & 0.05 & \textit{ECCo} & $-1.55 \cdot 10^{16}$ & $3.52 \cdot 10^{16}$ \\
+  full & 0.05 & \textit{Generic} & $-2.22 \cdot 10^{20}$ & $5 \cdot 10^{20}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.41}} & \color{blue}{\textbf{4.48}} \\
+  full & 0.05 & \textit{REVISE} & $-1.04 \cdot 10^{3}$ & $2.3 \cdot 10^{3}$ \\
+  vanilla & 0.05 & \textit{ECCo} & -15.5 & 17.2 \\
+  vanilla & 0.05 & \textit{Generic} & -11.7 & 12.8 \\
+  vanilla & 0.05 & \textit{Omniscient} & -12.4 & 14.1 \\
+  vanilla & 0.05 & \textit{REVISE} & -11.3 & 13.1 \\
+  full & 0.1 & \textit{ECCo} & $-3.41 \cdot 10^{3}$ & $7.73 \cdot 10^{3}$ \\
+  full & 0.1 & \textit{Generic} & $-5.22 \cdot 10^{30}$ & $1.19 \cdot 10^{31}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.78}} & \color{blue}{\textbf{5.12}} \\
+  full & 0.1 & \textit{REVISE} & -288 & 594 \\
+  vanilla & 0.1 & \textit{ECCo} & -15.5 & 17.2 \\
+  vanilla & 0.1 & \textit{Generic} & -10.9 & 11.9 \\
+  vanilla & 0.1 & \textit{Omniscient} & -12.7 & 14.4 \\
+  vanilla & 0.1 & \textit{REVISE} & -11.3 & 13.1 \\
+  full & 0.5 & \textit{ECCo} & -7.09 & 7.51 \\
+  full & 0.5 & \textit{Generic} & $-1.11 \cdot 10^{31}$ & $2.53 \cdot 10^{31}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.58}} & \color{blue}{\textbf{4.83}} \\
+  full & 0.5 & \textit{REVISE} & $-1.19 \cdot 10^{3}$ & $2.64 \cdot 10^{3}$ \\
+  vanilla & 0.5 & \textit{ECCo} & -15.5 & 17.2 \\
+  vanilla & 0.5 & \textit{Generic} & -11.7 & 12.8 \\
+  vanilla & 0.5 & \textit{Omniscient} & -12.4 & 14.1 \\
+  vanilla & 0.5 & \textit{REVISE} & -11.3 & 13.1 \\
+  full & 1 & \textit{ECCo} & -6.06 & 6.33 \\
+  full & 1 & \textit{Generic} & $-1.58 \cdot 10^{33}$ & $3.59 \cdot 10^{33}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.66}} & \color{blue}{\textbf{4.89}} \\
+  full & 1 & \textit{REVISE} & $-1.16 \cdot 10^{3}$ & $2.59 \cdot 10^{3}$ \\
+  vanilla & 1 & \textit{ECCo} & -15.5 & 17.3 \\
+  vanilla & 1 & \textit{Generic} & -10.9 & 11.9 \\
+  vanilla & 1 & \textit{Omniscient} & -12.7 & 14.4 \\
+  vanilla & 1 & \textit{REVISE} & -11.3 & 13.1 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.57}} & \color{blue}{\textbf{2.07}} \\
+  full & 5 & \textit{Generic} & $-1.17 \cdot 10^{28}$ & $2.66 \cdot 10^{28}$ \\
+  full & 5 & \textit{Omniscient} & -4.29 & 4.31 \\
+  full & 5 & \textit{REVISE} & -530 & $1.16 \cdot 10^{3}$ \\
+  vanilla & 5 & \textit{ECCo} & -15.5 & 17.2 \\
+  vanilla & 5 & \textit{Generic} & -11.7 & 12.7 \\
+  vanilla & 5 & \textit{Omniscient} & -12.4 & 14.1 \\
+  vanilla & 5 & \textit{REVISE} & -11.3 & 13.1 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.76}} & \color{blue}{\textbf{0.974}} \\
+  full & 10 & \textit{Generic} & $-1.54 \cdot 10^{33}$ & $3.51 \cdot 10^{33}$ \\
+  full & 10 & \textit{Omniscient} & -4.44 & 4.56 \\
+  full & 10 & \textit{REVISE} & $-1.52 \cdot 10^{3}$ & $3.4 \cdot 10^{3}$ \\
+  vanilla & 10 & \textit{ECCo} & -15.5 & 17.3 \\
+  vanilla & 10 & \textit{Generic} & -10.9 & 11.9 \\
+  vanilla & 10 & \textit{Omniscient} & -12.7 & 14.4 \\
+  vanilla & 10 & \textit{REVISE} & -11.3 & 13.1 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.37}} & \color{blue}{\textbf{0.365}} \\
+  full & 15 & \textit{Generic} & $-5.32 \cdot 10^{28}$ & $1.21 \cdot 10^{29}$ \\
+  full & 15 & \textit{Omniscient} & -4.34 & 4.38 \\
+  full & 15 & \textit{REVISE} & -473 & $1.03 \cdot 10^{3}$ \\
+  vanilla & 15 & \textit{ECCo} & -15.5 & 17.2 \\
+  vanilla & 15 & \textit{Generic} & -11.7 & 12.8 \\
+  vanilla & 15 & \textit{Omniscient} & -12.4 & 14.1 \\
+  vanilla & 15 & \textit{REVISE} & -11.3 & 13.1 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Circles}\label{circles}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-circles-lambda_energy_exper}Results for Circles data by energy penalty.}
+
+\tabularnewline
+
+  \\
+  \toprule
+  \textbf{Obj.} & \textbf{\textbackslash{}lambda (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Obj.} & \textbf{\textbackslash{}lambda (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.26}} & \color{blue}{\textbf{0.423}} \\
+  full & 0.01 & \textit{Generic} & -1.49 & 0.71 \\
+  full & 0.01 & \textit{Omniscient} & -5.21 & 5.25 \\
+  full & 0.01 & \textit{REVISE} & $-2.71 \cdot 10^{26}$ & $6.37 \cdot 10^{26}$ \\
+  vanilla & 0.01 & \textit{ECCo} & -9.33 & 7.34 \\
+  vanilla & 0.01 & \textit{Generic} & -8.89 & 6.88 \\
+  vanilla & 0.01 & \textit{Omniscient} & -8.67 & 6.87 \\
+  vanilla & 0.01 & \textit{REVISE} & -8.65 & 6.8 \\
+  full & 0.05 & \textit{ECCo} & -1.29 & 0.397 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.356}} \\
+  full & 0.05 & \textit{Omniscient} & -5.08 & 5.09 \\
+  full & 0.05 & \textit{REVISE} & $-5.91 \cdot 10^{27}$ & $1.36 \cdot 10^{28}$ \\
+  vanilla & 0.05 & \textit{ECCo} & -9.35 & 7.32 \\
+  vanilla & 0.05 & \textit{Generic} & -8.85 & 6.87 \\
+  vanilla & 0.05 & \textit{Omniscient} & -8.7 & 6.96 \\
+  vanilla & 0.05 & \textit{REVISE} & -8.52 & 6.76 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.2}} & \color{blue}{\textbf{0.383}} \\
+  full & 0.1 & \textit{Generic} & -1.5 & 0.735 \\
+  full & 0.1 & \textit{Omniscient} & -5.17 & 5.23 \\
+  full & 0.1 & \textit{REVISE} & $-3.06 \cdot 10^{26}$ & $7.7 \cdot 10^{26}$ \\
+  vanilla & 0.1 & \textit{ECCo} & -9.33 & 7.32 \\
+  vanilla & 0.1 & \textit{Generic} & -8.88 & 6.86 \\
+  vanilla & 0.1 & \textit{Omniscient} & -8.69 & 6.9 \\
+  vanilla & 0.1 & \textit{REVISE} & -8.68 & 6.81 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.12}} & \color{blue}{\textbf{0.217}} \\
+  full & 0.5 & \textit{Generic} & -1.21 & 0.352 \\
+  full & 0.5 & \textit{Omniscient} & -5.09 & 5.12 \\
+  full & 0.5 & \textit{REVISE} & $-5.97 \cdot 10^{27}$ & $1.37 \cdot 10^{28}$ \\
+  vanilla & 0.5 & \textit{ECCo} & -9.35 & 7.3 \\
+  vanilla & 0.5 & \textit{Generic} & -8.89 & 6.92 \\
+  vanilla & 0.5 & \textit{Omniscient} & -8.68 & 6.93 \\
+  vanilla & 0.5 & \textit{REVISE} & -8.53 & 6.75 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.1}} & \color{blue}{\textbf{0.163}} \\
+  full & 1 & \textit{Generic} & -1.49 & 0.726 \\
+  full & 1 & \textit{Omniscient} & -5.16 & 5.2 \\
+  full & 1 & \textit{REVISE} & $-3.09 \cdot 10^{26}$ & $7.22 \cdot 10^{26}$ \\
+  vanilla & 1 & \textit{ECCo} & -9.34 & 7.36 \\
+  vanilla & 1 & \textit{Generic} & -8.86 & 6.85 \\
+  vanilla & 1 & \textit{Omniscient} & -8.7 & 6.9 \\
+  vanilla & 1 & \textit{REVISE} & -8.69 & 6.85 \\
+  full & 5 & \textit{ECCo} & -1.75 & 0.154 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.363}} \\
+  full & 5 & \textit{Omniscient} & -5.14 & 5.16 \\
+  full & 5 & \textit{REVISE} & $-1.1 \cdot 10^{28}$ & $2.5 \cdot 10^{28}$ \\
+  vanilla & 5 & \textit{ECCo} & -9.36 & 7.32 \\
+  vanilla & 5 & \textit{Generic} & -8.88 & 6.91 \\
+  vanilla & 5 & \textit{Omniscient} & -8.7 & 6.93 \\
+  vanilla & 5 & \textit{REVISE} & -8.52 & 6.73 \\
+  full & 10 & \textit{ECCo} & $-1.02 \cdot 10^{6}$ & $2.32 \cdot 10^{6}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.49}} & \color{blue}{\textbf{0.702}} \\
+  full & 10 & \textit{Omniscient} & -5.13 & 5.16 \\
+  full & 10 & \textit{REVISE} & $-3.74 \cdot 10^{26}$ & $9.09 \cdot 10^{26}$ \\
+  vanilla & 10 & \textit{ECCo} & -9.31 & 7.33 \\
+  vanilla & 10 & \textit{Generic} & -8.87 & 6.86 \\
+  vanilla & 10 & \textit{Omniscient} & -8.7 & 6.89 \\
+  vanilla & 10 & \textit{REVISE} & -8.69 & 6.83 \\
+  full & 15 & \textit{ECCo} & $-3.31 \cdot 10^{13}$ & $7.54 \cdot 10^{13}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.22}} & \color{blue}{\textbf{0.37}} \\
+  full & 15 & \textit{Omniscient} & -5.2 & 5.23 \\
+  full & 15 & \textit{REVISE} & $-9.01 \cdot 10^{27}$ & $2.06 \cdot 10^{28}$ \\
+  vanilla & 15 & \textit{ECCo} & -9.38 & 7.34 \\
+  vanilla & 15 & \textit{Generic} & -8.86 & 6.87 \\
+  vanilla & 15 & \textit{Omniscient} & -8.69 & 6.96 \\
+  vanilla & 15 & \textit{REVISE} & -8.51 & 6.73 \\\bottomrule
+
+\end{longtable}

### 2025-01-28 paper/sections/abstract.qmd
streamlined quarto bs

# Abstract

Counterfactual Explanations (CE) have emerged as a popular method to explain the predictions made by opaque machine learning models in a post-hoc fashion. We propose a novel approach that leverages counterfactuals during the training phase of models.

### 2025-01-28 paper/sections/appendix.qmd
streamlined quarto bs

--- 2025-01-27 paper/appendix.qmd
@@
 
-- **Energy Penalty**: *ECCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training.
-*Generic* performs poorly accross the board.
-*Omni* seems to have an anchoring effect, in that it never performs terribly but also never as good as the best *ECCo* results.
-*REVISE* performs poorly across the board.
-- **Cost (distance penalty)**: Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged).
-For *ECCo* and *Generic*, higher cost values generally lead to worse results.
-- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
-- **Validity**: *ECCo* almost always valid except for very low values during training and high values at evaluation time. *Generic* often has poor validity.
-- **Accuracy**: Seems largely unaffected.
-
 ```{julia}
@@
 
+::: {#tbl-lin_sep-lambda_energy_exper}
+
 ::: {.content-hidden unless-format="pdf"}
@@
 get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; label="tbl-moons-lambda_energy_exper", title="Results for Linearly Separable data by energy penalty.")
+    inputs -> tabulate_results(inputs; wrap_table=false)
 ```
+
+:::
+
+Results for Linearly Separable data by energy penalty.
 

### 2025-01-28 paper/sections/experiments.qmd
streamlined quarto bs

# Experiments {#sec-experiments}

## Experimental Setup

## Experimental Results

### 2025-01-28 paper/ecml/paper.tex
the joys of quarto and julia and latex and shootme

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
% N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
\usepackage{mwe}
\usepackage{color}
\usepackage[numbers,compress]{natbib}
\usepackage{longtable}
\usepackage{hyperref}
% N.B.: you may delete the preceding line. It is used to display an example image in this template.

\begin{document}

\title{Counterfactual Training: Teaching Models Plausible and Actionable Explanations}

\titlerunning{Counterfactual Training}
% If the full title of your paper is short enough to also fit in the running head, you can omit the abbreviated paper title here. You can check as follows: if you comment out the \titlerunning line, something will appear in the header of all odd-numbered pages of your PDF from page 3 onward. This something is either the full title (in which case all is well), or the error message "Title Suppressed Due to Excessive Length". If this error message appears, you're going to want to provide an abbreviated title within the \titlerunning command, because if you won't do it, Springer will do it for you.

%N.B.: Author information (both in the \author{} and \authorrunning{} command) should only be present in the Camera-Ready Version of your paper. The version that you initially submit for review, ought to be double-blind. So, when initially submitting your paper, use:
\author{Author information scrubbed for double-blind reviewing}
\author{Patrick Altmeyer\inst{1}\orcidID{0000-0003-4726-8613} \corr \and
Arie van Deursen\inst{1} \and
Cynthia C. S. Liem\inst{1}}
% You may leave out the orcidID information, if you want to.
% Use \corr to indicate the corresponding author. Note the spacing around the \corr command. Only one author can be the corresponding author.

%N.B.: comment out the \authorrunning{} command for the double-blind version of your paper submitted for review. Later, if your paper is accepted, use the command for the Camera-Ready Version.
% \authorrunning{A.L. Benjamin et al.}
% First names are abbreviated in the running head.
% If there is one author, write 'A.L. Benjamin'.
% If there are two authors, write 'A.L. Benjamin and C.C. Broadus Jr.'
% If there are more than two authors, '[...] et al.' is used.

\institute{Delft University of Technology, Faculty of Electrical Engineering, Mathematics and Computer Science \email{\{p.altmeyer\}@tudelft.net}}

\maketitle              % typeset the header of the contribution

\input{../body_ecml.tex}

\begin{credits}
\subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
used for general acknowledgments, for example: This study was funded
by X (grant number Y).

\subsubsection{\discintname}
It is now necessary to declare any competing interests or to specifically
state that the authors have no competing interests. Please place the
statement with a bold run-in heading in small font size beneath the
(optional) acknowledgments,
for example: The authors have no competing interests to declare that are
relevant to the content of this article. Or: Author A has received research
grants from Company W. Author B has received a speaker honorarium from
Company X and owns stock in Company Y. Author C is a member of committee Z.
\end{credits}

\bibliography{../bib}
\bibliographystyle{splncs04}

% \input{appendix.tex}

\end{document}

### 2025-01-28 paper/sections/appendix.qmd
the joys of quarto and julia and latex and shootme

@@
----
-engine: julia
-julia: 
-  exeflags: ["--project=experiments/"]
----
-
 ```{julia}
-cd("../")
-
 using CTExperiments

### 2025-01-28 paper/preprint/paper.tex
finally working version for preprint and ecml template using quarto

--- 2025-01-27 paper/appendix.tex
@@
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+]{article}
+
+\usepackage{amsmath,amssymb}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else  
+    % xetex/luatex font selection
+    \setmainfont[]{Latin Modern Roman}
+  \setmathfont[]{Latin Modern Math}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{5}
+% Make \paragraph and \subparagraph free-standing
+\makeatletter
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}{
+    \@ifstar
+      \xxxParagraphStar
+      \xxxParagraphNoStar
+  }
+  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
+  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}{
+    \@ifstar
+      \xxxSubParagraphStar
+      \xxxSubParagraphNoStar
+  }
+  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
+  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+\makeatother
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+\usepackage{graphicx}
+\makeatletter
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
+% Set default figure placement to htbp
+\def\fps@figure{htbp}
+\makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
+  \fi
+  % set entry spacing
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\usepackage{arxiv}
+\usepackage{orcidlink}
+\usepackage{amsmath}
+\usepackage[T1]{fontenc}
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+
+\usepackage{bookmark}
+
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdfkeywords={Counterfactual Explanations, Explainable AI},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
+
+\newcommand{\runninghead}{A Preprint }
+\renewcommand{\runninghead}{A Preprint }
+\def\asep{\\\\\\ } % default: all authors on same column
+\author{}
+\date{}
+\begin{document}
+{\bfseries \emph Keywords}
+\def\sep{\textbullet\ }
+Counterfactual Explanations \sep 
+Explainable AI
+
+\section{Abstract}\label{abstract}
+
+Counterfactual Explanations (CE) have emerged as a popular method to
+explain the predictions made by opaque machine learning models in a
+post-hoc fashion. We propose a novel approach that leverages
+counterfactuals during the training phase of models.
+
+\section{Introduction}\label{introduction}
+
+\section{Related Literature}\label{sec-lit}
+
+\subsection{Background on Counterfactual
+Explanations}\label{background-on-counterfactual-explanations}
+
+(Wachter, Mittelstadt, and Russell 2017; Joshi et al. 2019; Altmeyer et
+al. 2024)
+
+\subsection{Learning Representations}\label{learning-representations}
+
+\begin{quote}
+For example, joint-energy models
+\end{quote}
+
+\subsection{Generalization and
+Robustness}\label{generalization-and-robustness}
+
+Sauer and Geiger (2021) generate counterfactual images for MNIST and
+ImageNet through independent mechanisms (IM): each IM learns
+class-conditional input distributions over a specific lower-dimensional,
+semantically meaningful factor, such as \emph{texture}, \emph{shape} and
+\emph{background}. The demonstrate that using these generated
+counterfactuals during classifier training improves model robustness.
+Similarly, Abbasnejad et al. (2020) argue that counterfactuals represent
+potentially useful training data in machine learning, especially in
+supervised settings where inputs may be reasonably mapped to multiple
+outputs. They, too, demonstrate the augmenting the training data of
+image classifiers can improve generalization.
+
+Teney, Abbasnedjad, and Hengel (2020) propose an approach using
+counterfactuals in training that does not rely on data augmentation:
+they argue that counterfactual pairs typically already exist in training
+datasets. Specifically, their approach relies on, firstly, identifying
+similar input samples with different annotations and, secondly, ensuring
+that the gradient of the classifier aligns with the vector between pairs
+of counterfactual inputs using the cosine distance as a loss function
+(referred to as \emph{gradient supervision}) (\textbf{\emph{this might
+be useful for our task as well}}). In the natural language processing
+(NLP) domain, counterfactuals have similarly been used to improve models
+through data augmentation: Wu et al. (2021), propose POLYJUICE, a
+general-purpose counterfactual generator for language models. They
+demonstrate empirically that augmenting training data through POLYJUICE
+counterfactuals improves robustness in a number of NLP tasks.
+
+\subsection{Link to Adversarial
+Training}\label{link-to-adversarial-training}
+
+Freiesleben (2022) propose two definitional differences between
+Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly,
+and more importantly according to the authors, the term AE implies
+missclassification, which is not the case for CE (\textbf{\emph{this
+might be a useful notion for use to distinguish between adversarials and
+explanations during training}}); secondly, they argue that closeness
+plays a more critical role in the context of CE but confess that even
+counterfactuals that are not close might be relevant explanations.
+Pawelczyk et al. (2022) show that CE and AE are equivalent under certain
+conditions and derive upper bounds on the distances between them.
+
+\subsection{Closely Related}\label{closely-related}
+
+Guo, Nguyen, and Yadav (2023) are the first to propose end-to-end
+training pipeline that includes counterfactual explanations as part of
+the training prodeduce. In particular, they propose a specific network
+architecture that includes a predictor and CE generator network
+(\textbf{\emph{akin a GAN?}}), where the parameters of the CE generator
+network are learnable. Counterfactuals are generated during each
+training iteration and fed back to the predictor network
+(\textbf{\emph{here we are aligned}}). In contrast, we impose no
+restrictions on the neural network architecture at all.
+(\textbf{\emph{to ensure the one-hot encoding of categorical features is
+maintained, they simple use softmax (might be interesting for CE.jl)}})
+Interestingly, the authors find that their approach is sensitive to the
+choice of the loss function: only MSE seems to lead to good performance.
+They also demonstrate theoretically, that the objective function is
+difficult to optimize due to divergent gradients and suffers from poor
+adversarial robustness. (\textbf{\emph{because partial gradients with
+respect to the classification loss component and the counterfactual
+validity component point in opposite directions}}). To mitigate these
+issues, the authors use block-wise gradient descent: they first update
+with respect to classification loss and then use a second update with
+respect to the other loss components (\textbf{\emph{this might be useful
+for our task as well}}). Ross, Lakkaraju, and Bastani (2024) propose a
+way to train models that are guaranteed to provide recourse for
+individuals with high probability. The approach builds on adversarial
+training (\textbf{\emph{here we are aligned}}), where in this context
+adversarial examples are actively encouraged to exist, but only target
+attacks with respect to the positive class. The proposed method allows
+for imposing a set of actionable recourse ex-ante: for example, users
+can impose mutability constraints for features (\textbf{\emph{here we
+are aligned}}). (\textbf{\emph{To solve their objective function more
+efficiently, they use a first-order Taylor approximation to approximate
+the recourse loss component (might be applicable in our case)}})
+
+Luu and Inoue (2023) introduce Counterfactual Adversarial Training (CAT)
+with intention of improving generalization and robustness of language
+models. Specifically, they propose to proceed as follows: firstly,
+identify training samples that are subject to high predictive
+uncertainty (entropy); secondly, generate counterfactual explanations
+for those samples; and, finally, finetune the model on the augmented
+dataset that includes the generated counterfactuals.
+
+\section{Counterfactual Training}\label{counterfactual-training}
+
+\section{Experiments}\label{sec-experiments}
+
+\subsection{Experimental Setup}\label{experimental-setup}
+
+\subsection{Experimental Results}\label{experimental-results}
+
+\section{Discussion}\label{sec-discussion}
+
+\section{Conclusion}\label{sec-conclusion}
+
 \section{Appendix}\label{appendix}
@@
 
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-lin_sep-lambda_energy_exper}):
-  \emph{ECCo} generally does yield better results than \emph{Vanilla}
-  for higher choices of the energy penalty (10,15) during training.
-  \emph{Generic} performs poorly accross the board. \emph{Omni} seems to
-  have an anchoring effect, in that it never performs terribly but also
-  never as good as the best \emph{ECCo} results. \emph{REVISE} performs
-  poorly across the board.
-\item
-  \textbf{Cost (distance penalty)}: Results for all generators (except
-  \emph{Omni}) are quite bad, which can likely be attributed to
-  extremely bad results for some choices of the \textbf{Energy Penalty}
-  (results here are averaged). For \emph{ECCo} and \emph{Generic},
-  higher cost values generally lead to worse results.
-\item
-  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
-  seems that smaller choices are ok.
-\item
-  \textbf{Validity}: \emph{ECCo} almost always valid except for very low
-  values during training and high values at evaluation time.
-  \emph{Generic} often has poor validity.
-\item
-  \textbf{Accuracy}: Seems largely unaffected.
-\end{itemize}
-
 \begin{longtable}{ccccc}
 
-\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly Separable data by energy penalty.}
+\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly
+Separable data by energy penalty.}
 
@@
 
-  \\
   \toprule
-  \textbf{Obj.} & \textbf{\textbackslash{}lambda (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \textbf{Obj.} & \textbf{\$\textbackslash{}lambda\$ (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
   \endfirsthead
   \toprule
-  \textbf{Obj.} & \textbf{\textbackslash{}lambda (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \textbf{Obj.} & \textbf{\$\textbackslash{}lambda\$ (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
   \endhead
@@
 
-\caption{\label{tbl-moons-lambda_energy_exper}Results for Moons data by energy penalty.}
+\caption{\label{tbl-moons-lambda_energy_exper}Results for Moons data by
+energy penalty.}
 
@@
 
-\caption{\label{tbl-circles-lambda_energy_exper}Results for Circles data by energy penalty.}
+\caption{\label{tbl-circles-lambda_energy_exper}Results for Circles data
+by energy penalty.}
 
@@
 \end{longtable}
+
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
+Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
+den Hengel. 2020. {``Counterfactual Vision and Language Learning.''} In
+\emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
+Recognition (CVPR)}, 10041--51.
+\url{https://doi.org/10.1109/CVPR42600.2020.01006}.
+
+\bibitem[\citeproctext]{ref-altmeyer2024faithful}
+Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia CS
+Liem. 2024. {``Faithful Model Explanations Through Energy-Constrained
+Conformal Counterfactuals.''} In \emph{Proceedings of the AAAI
+Conference on Artificial Intelligence}, 38:10829--37. 10.
+
+\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
+Freiesleben, Timo. 2022. {``The Intriguing Relation Between
+Counterfactual Explanations and Adversarial Examples.''} \emph{Minds and
+Machines} 32 (1): 77--109.
+
+\bibitem[\citeproctext]{ref-guo2023counternet}
+Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``CounterNet:
+End-to-End Training of Prediction Aware Counterfactual Explanations.''}
+In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
+Discovery and Data Mining}, 577--89. KDD '23. New York, NY, USA:
+Association for Computing Machinery.
+\url{https://doi.org/10.1145/3580305.3599290}.
+
+\bibitem[\citeproctext]{ref-joshi2019realistic}
+Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
+Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
+Actionable Explanations in Black-Box Decision Making Systems.''}
+\url{https://arxiv.org/abs/1907.09615}.
+
+\bibitem[\citeproctext]{ref-luu2023counterfactual}
+Luu, Hoai Linh, and Naoya Inoue. 2023. {``Counterfactual Adversarial
+Training for Improving Robustness of Pre-Trained Language Models.''} In
+\emph{Proceedings of the 37th Pacific Asia Conference on Language,
+Information and Computation}, 881--88.
+
+\bibitem[\citeproctext]{ref-pawelczyk2022exploring}
+Pawelczyk, Martin, Chirag Agarwal, Shalmali Joshi, Sohini Upadhyay, and
+Himabindu Lakkaraju. 2022. {``Exploring Counterfactual Explanations
+Through the Lens of Adversarial Examples: A Theoretical and Empirical
+Analysis.''} In \emph{Proceedings of the 25th International Conference
+on Artificial Intelligence and Statistics}, edited by Gustau
+Camps-Valls, Francisco J. R. Ruiz, and Isabel Valera, 151:4574--94.
+Proceedings of Machine Learning Research. PMLR.
+\url{https://proceedings.mlr.press/v151/pawelczyk22a.html}.
+
+\bibitem[\citeproctext]{ref-ross2021learning}
+Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024. {``Learning
+Models for Actionable Recourse.''} In \emph{Proceedings of the 35th
+International Conference on Neural Information Processing Systems}. NIPS
+'21. Red Hook, NY, USA: Curran Associates Inc.
+
+\bibitem[\citeproctext]{ref-sauer2021counterfactual}
+Sauer, Axel, and Andreas Geiger. 2021. {``Counterfactual Generative
+Networks.''} \url{https://arxiv.org/abs/2101.06046}.
+
+\bibitem[\citeproctext]{ref-teney2020learning}
+Teney, Damien, Ehsan Abbasnedjad, and Anton van den Hengel. 2020.
+{``Learning What Makes a Difference from Counterfactual Examples and
+Gradient Supervision.''} In \emph{Computer Vision--ECCV 2020: 16th
+European Conference, Glasgow, UK, August 23--28, 2020, Proceedings, Part
+x 16}, 580--99. Springer.
+
+\bibitem[\citeproctext]{ref-wachter2017counterfactual}
+Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
+{``Counterfactual Explanations Without Opening the Black Box:
+{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
+841. \url{https://doi.org/10.2139/ssrn.3063289}.
+
+\bibitem[\citeproctext]{ref-wu2021polyjuice}
+Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
+2021. {``Polyjuice: Generating Counterfactuals for Explaining,
+Evaluating, and Improving Models.''} In \emph{Proceedings of the 59th
+Annual Meeting of the Association for Computational Linguistics and the
+11th International Joint Conference on Natural Language Processing
+(Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
+and Roberto Navigli, 6707--23. Online: Association for Computational
+Linguistics. \url{https://doi.org/10.18653/v1/2021.acl-long.523}.
+
+\end{CSLReferences}
+
+\end{document}

### 2025-01-28 paper/sections/appendix.qmd
finally working version for preprint and ecml template using quarto

@@
+```{julia}
+projectdir = splitpath(pwd()) |>
+    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
+cd(projectdir)
+```
+
 ```{julia}
@@
 using CTExperiments.StatsBase
+
 using DotEnv
-
 DotEnv.load!()

### 2025-01-29 paper/preprint/paper.tex
vamos

@@
 
+\subsection{Training Details}\label{training-details}
+
 \subsection{Initial Grid Search}\label{initial-grid-search}
+
+For the initial round of experiments we
 
@@
 \paragraph{Linearly Separable}\label{linearly-separable}
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Energy Penalty} (Table~\ref{tbl-lin_sep-lambda_energy_exper}):
+  \emph{ECCo} generally does yield better results than \emph{Vanilla}
+  for higher choices of the energy penalty (10,15) during training.
+  \emph{Generic} performs poorly accross the board. \emph{Omni} seems to
+  have an anchoring effect, in that it never performs terribly but also
+  never as good as the best \emph{ECCo} results. \emph{REVISE} performs
+  poorly across the board.
+\item
+  \textbf{Cost (distance penalty)}: Results for all generators (except
+  \emph{Omni}) are quite bad, which can likely be attributed to
+  extremely bad results for some choices of the \textbf{Energy Penalty}
+  (results here are averaged). For \emph{ECCo} and \emph{Generic},
+  higher cost values generally lead to worse results.
+\item
+  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
+  seems that smaller choices are ok.
+\item
+  \textbf{Validity}: \emph{ECCo} almost always valid except for very low
+  values during training and high values at evaluation time.
+  \emph{Generic} often has poor validity.
+\item
+  \textbf{Accuracy}: Seems largely unaffected.
+\end{itemize}
 
@@
   \toprule
-  \textbf{Obj.} & \textbf{\$\textbackslash{}lambda\$ (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
   \endfirsthead
   \toprule
-  \textbf{Obj.} & \textbf{\$\textbackslash{}lambda\$ (exper)} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
   \endhead
@@
 
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Energy Penalty} (Table~\ref{tbl-moons-lambda_energy_exper}):
+  \emph{ECCo} consistently yields better results than \emph{Vanilla},
+  except for very low choices of the energy penalty during training for
+  which it performs abismal. \emph{Generic} performs quite badly across
+  the board for high enough choices of the energy penalty at evaluation
+  time. \emph{Omni} has small positive effect. \emph{REVISE} performs
+  poorly across the board.
+\item
+  \textbf{Cost (distance penalty)}: \emph{Generic} generally does better
+  for higher values, while \emph{ECCo} does better for lower values.
+\item
+  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
+  seems that smaller choices are ok.
+\item
+  \textbf{Validity}: \emph{ECCo} generally achieves full validity except
+  for very low choices the energy penalty during training and high
+  choices at evaluation time. \emph{Generic} performs poorly for high
+  choices of the energy penalty during evaluation.
+\item
+  \textbf{Accuracy}: Largely unaffected although \emph{ECCo} suffers a
+  bit for very low choices the energy penalty during training.
+  \emph{REVISE} suffers a lot in general (around 10 percentage points).
+\end{itemize}
+
 \begin{longtable}{ccccc}
@@
 
+\begin{itemize}
+\tightlist
+\item
+  \textbf{Energy Penalty} (Table~\ref{tbl-circles-lambda_energy_exper}):
+  \emph{ECCo} consistently yields better results than \emph{Vanilla},
+  though primarily for low to medium choices of the energy penalty
+  (\textless=5) during training. The same goes for \emph{Generic}, which
+  sometimes outperforms \emph{ECCo} (for small energy penalty at
+  evaluation time). \emph{Omni} does alright for lower energy penalty at
+  evaluation time, but loses out for higher choices. \emph{REVISE}
+  performs poorly across the board (except very low choices at
+  evaluation time).
+\item
+  \textbf{Cost (distance penalty)}: \emph{ECCo} and \emph{Generic}
+  generally achieve the best results when no cost penalty is used during
+  training. Both \emph{Omni} and \emph{REVISE} are largely unaffected.
+\item
+  \textbf{Maximum Iterations}: \emph{ECCo} consistently yields better
+  results for higher numbers of iterations. \emph{Generic} generally
+  does best for a medium number (50). \emph{Omni} is sometimes invalid
+  (\textbf{???}).
+\item
+  \textbf{Validity}: \emph{ECCo} tends to outperform its \emph{Vanilla}
+  counterpart, though primarily for low to medium choices of the energy
+  penalty (\textless=5) during training and evaluation. \emph{Vanilla}
+  typically worse across the board.
+\item
+  \textbf{Accuracy}: Mostly unaffected, but \emph{REVISE} again
+  consistently some deterioration and \emph{ECCo} deteriorates for high
+  choices of energy penalty during training, reflecting other outcomes
+  above.
+\end{itemize}
+
 \begin{longtable}{ccccc}

### 2025-01-29 paper/sections/appendix.qmd
vamos

@@
 
+## Training Details
+
 ## Initial Grid Search
@@
 ```{julia}
-results_dir = ENV["INITIAL_RUN"]
+results_dir = ENV["INITIAL_RUN_RESULTS"]
+config_dir = ENV["INITIAL_RUN_CONFIG"]
 ```
 
+For the initial round of experiments we
+
 ### Generator Params
+
+```{julia}
+dict = CTExperiments.from_toml(joinpath(config_dir,"gen_params.toml")) 
+```
 
@@
 #### Linearly Separable
+
+- **Energy Penalty** (@tbl-lin_sep-lambda_energy_exper): *ECCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training.
+*Generic* performs poorly accross the board.
+*Omni* seems to have an anchoring effect, in that it never performs terribly but also never as good as the best *ECCo* results.
+*REVISE* performs poorly across the board.
+- **Cost (distance penalty)**: Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged).
+For *ECCo* and *Generic*, higher cost values generally lead to worse results.
+- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
+- **Validity**: *ECCo* almost always valid except for very low values during training and high values at evaluation time. *Generic* often has poor validity.
+- **Accuracy**: Seems largely unaffected.
 
@@
 
+- **Energy Penalty** (@tbl-moons-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, except for very low choices of the energy penalty during training for which it performs abismal.
+*Generic* performs quite badly across the board for high enough choices of the energy penalty at evaluation time.
+*Omni* has small positive effect.
+*REVISE* performs poorly across the board.
+- **Cost (distance penalty)**: *Generic* generally does better for higher values, while *ECCo* does better for lower values.
+- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
+- **Validity**: *ECCo* generally achieves full validity except for very low choices the energy penalty during training and high choices at evaluation time.
+*Generic* performs poorly for high choices of the energy penalty during evaluation.
+- **Accuracy**: Largely unaffected although *ECCo* suffers a bit for very low choices the energy penalty during training. *REVISE* suffers a lot in general (around 10 percentage points).
+
 ```{julia}
@@
 
+- **Energy Penalty** (@tbl-circles-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, though primarily for low to medium choices of the energy penalty (<=5) during training.
+The same goes for *Generic*, which sometimes outperforms *ECCo* (for small energy penalty at evaluation time).
+*Omni* does alright for lower energy penalty at evaluation time, but loses out for higher choices.
+*REVISE* performs poorly across the board (except very low choices at evaluation time).
+- **Cost (distance penalty)**: *ECCo* and *Generic* generally achieve the best results when no cost penalty is used during training. Both *Omni* and *REVISE* are largely unaffected.
+- **Maximum Iterations**: *ECCo* consistently yields better results for higher numbers of iterations. *Generic* generally does best for a medium number (50). *Omni* is sometimes invalid (**???**).
+- **Validity**: *ECCo* tends to outperform its *Vanilla* counterpart, though primarily for low to medium choices of the energy penalty (<=5) during training and evaluation.
+*Vanilla* typically worse across the board.
+- **Accuracy**: Mostly unaffected, but *REVISE* again consistently some deterioration and *ECCo* deteriorates for high choices of energy penalty during training, reflecting other outcomes above.
+
 ```{julia}

### 2025-01-29 paper/experiments/CTExperiments/src/config.jl

@@ -1,3 +1,4 @@
+using Markdown
 using TaijaParallel
 using TOML
 
@@ -125,3 +126,6 @@ function has_results(cfg::AbstractConfiguration)
 end
 
 needs_results(cfg::AbstractConfiguration) = !has_results(cfg)
+
+function to_mkd(dict::Dict)
+end

### 2025-01-29 paper/experiments/CTExperiments/src/config.jl
nice

@@ -127,17 +127,73 @@ end
 
 needs_results(cfg::AbstractConfiguration) = !has_results(cfg)
 
-function to_mkd(dict::Dict, header::Union{Nothing,String}=nothing)
-    drop_fields = ["name"]
-    dict = filter(((k,v),) -> length(v) > 0 && !(k in drop_fields), dict)
-    strs = []
-    for (i,(k,v)) in enumerate(dict)
-        if !isa(v, Dict)
-            str = isnothing(header) ? "**$(k)**" : "$(header) **$(k)**"
-        else
-            str = to_mkd(v, k)
-        end
-        push!(strs, str)
+# Function to check if a value is effectively empty
+function is_empty_value(v::Any)
+    if v isa String
+        return isempty(v)
+    elseif v isa Vector
+        return isempty(v)
+    elseif v isa Dict
+        # A dictionary is empty if it's empty itself or if all its filtered values would be empty
+        filtered = filter_dict(v)
+        return isempty(filtered)
+    else
+        return false
     end
-    str = join(strs, "\n")
+end
+
+# Function to filter dictionary
+function filter_dict(dict::Dict)
+    # Filter out empty values and specified fields
+    drop_fields = ["name"]
+    return filter(dict) do (k, v)
+        !is_empty_value(v) && !(k in drop_fields)
+    end
+end
+
+function to_mkd(dict::Dict, level::Int=0)
+
+    drop_fields = ["name"]
+    dict = filter(((k, v),) -> length(v) > 0 && !(k in drop_fields), dict)
+
+    # Create indent string based on level
+    indent = repeat("    ", level)
+
+    # Initialize array to store markdown lines
+    lines = String[]
+
+    # Sort dictionary keys for consistent output
+    for key in sort(collect(keys(dict)))
+        value = dict[key]
+
+        if value isa Dict
+            # Handle nested dictionary
+            push!(lines, "$(indent)- **$(key)**:")
+            # Recursively process nested dictionary with increased indentation
+            nested_lines = to_mkd(value, level + 1)
+            push!(lines, nested_lines)
+        elseif value isa Vector
+            # Handle vector values by joining with commas
+            value_str = join(value, ", ")
+            push!(lines, "$(indent)- **$(key)**: $(value_str)")
+        else
+            # Handle single values
+            push!(lines, "$(indent)- **$(key)**: $(value)")
+        end
+    end
+
[…]

### 2025-01-29 paper/preprint/paper.tex
nice
ready to start writing this up

@@
 \newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{exercise}{Parameters}[section]
+\theoremstyle{remark}
+\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
+\newtheorem*{remark}{Remark}
+\newtheorem*{solution}{Solution}
+\newtheorem{refremark}{Remark}[section]
+\newtheorem{refsolution}{Solution}[section]
 \makeatother
@@
 \hypersetup{
+  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
+  pdfauthor={Patrick Altmeyer; Arie van Deursen; Cynthia C. S. Liem},
   pdfkeywords={Counterfactual Explanations, Explainable AI},
@@
 
+\usepackage{lineno}
+\linenumbers
 \newcommand{\runninghead}{A Preprint }
-\renewcommand{\runninghead}{A Preprint }
+\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
+\title{Counterfactual Training: Teaching Models Plausible and Actionable
+Explanations}
 \def\asep{\\\\\\ } % default: all authors on same column
-\author{}
+\def\asep{\And }
+\author{\textbf{Patrick
+Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
+Engineering, Mathematics and Computer Science\\Delft University of
+Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
+van Deursen}\\Faculty of Electrical Engineering, Mathematics and
+Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
+C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
+Science\\Delft University of Technology\\\\}
 \date{}
 \begin{document}
+\maketitle
+\begin{abstract}
+Counterfactual Explanations (CE) have emerged as a popular method to
+explain the predictions made by opaque machine learning models in a
+post-hoc fashion. We propose a novel approach that leverages
+counterfactuals during the training phase of models.
+\end{abstract}
 {\bfseries \emph Keywords}
@@
 Explainable AI
-
-\section{Abstract}\label{abstract}
-
-Counterfactual Explanations (CE) have emerged as a popular method to
-explain the predictions made by opaque machine learning models in a
-post-hoc fashion. We propose a novel approach that leverages
-counterfactuals during the training phase of models.
 
@@
 
-\subsubsection{Generator Params}\label{generator-params}
+\subsubsection{Generator Parameters}\label{generator-parameters}
+
+The hyperparameter choices are shown in
+Parameters~\ref{exr-gen-params-first-run}:
+
+\begin{exercise}[Parameters]\protect\hypertarget{exr-gen-params-first-run}{}\label{exr-gen-params-first-run}
+
+~
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{generator\_params}:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \textbf{lambda\_cost}: 0.0, 0.001, 0.1
+  \item
+    \textbf{lambda\_energy}: 0.01, 0.05, 0.1, 0.5, 1.0, 5.0, 10.0, 15.0
+  \item
+    \textbf{lr}: 1.0
+  \item
+    \textbf{maxiter}: 20, 50, 100
+  \item
+    \textbf{opt}: sgd
+  \end{itemize}
+\item
+  \textbf{generator\_type}: ecco, generic, omni, revise
+\item
+  \textbf{training\_params}:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \textbf{objective}: full, vanilla
+  \end{itemize}
+\end{itemize}
+
+\end{exercise}
 

### 2025-01-29 paper/sections/appendix.qmd
nice
ready to start writing this up

@@
 
-### Generator Params
+### Generator Parameters
+
+The hyperparameter choices are shown in @exr-gen-params-first-run:
+
+::: {#exr-gen-params-first-run}
+
+## Parameters
 
 ```{julia}
+#| output: asis
 dict = CTExperiments.from_toml(joinpath(config_dir,"gen_params.toml")) 
+println(CTExperiments.dict_to_quarto_markdown(dict))
 ```
+
+:::
 

### 2025-01-30 paper/preprint/paper.tex
vamos
progress on appendix

@@
 \usepackage[T1]{fontenc}
+\usepackage{placeins}
 \makeatletter
@@
 
-\section{Appendix}\label{appendix}
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
 
-\subsection{Training Details}\label{training-details}
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
+Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
+den Hengel. 2020. {``Counterfactual Vision and Language Learning.''} In
+\emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
+Recognition (CVPR)}, 10041--51.
+\url{https://doi.org/10.1109/CVPR42600.2020.01006}.
+
+\bibitem[\citeproctext]{ref-altmeyer2024faithful}
+Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia CS
+Liem. 2024. {``Faithful Model Explanations Through Energy-Constrained
+Conformal Counterfactuals.''} In \emph{Proceedings of the AAAI
+Conference on Artificial Intelligence}, 38:10829--37. 10.
+
+\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
+Freiesleben, Timo. 2022. {``The Intriguing Relation Between
+Counterfactual Explanations and Adversarial Examples.''} \emph{Minds and
+Machines} 32 (1): 77--109.
+
+\bibitem[\citeproctext]{ref-guo2023counternet}
+Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``CounterNet:
+End-to-End Training of Prediction Aware Counterfactual Explanations.''}
+In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
+Discovery and Data Mining}, 577--89. KDD '23. New York, NY, USA:
+Association for Computing Machinery.
+\url{https://doi.org/10.1145/3580305.3599290}.
+
+\bibitem[\citeproctext]{ref-joshi2019realistic}
+Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
+Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
+Actionable Explanations in Black-Box Decision Making Systems.''}
+\url{https://arxiv.org/abs/1907.09615}.
+
+\bibitem[\citeproctext]{ref-luu2023counterfactual}
+Luu, Hoai Linh, and Naoya Inoue. 2023. {``Counterfactual Adversarial
+Training for Improving Robustness of Pre-Trained Language Models.''} In
+\emph{Proceedings of the 37th Pacific Asia Conference on Language,
+Information and Computation}, 881--88.
+
+\bibitem[\citeproctext]{ref-pawelczyk2022exploring}
+Pawelczyk, Martin, Chirag Agarwal, Shalmali Joshi, Sohini Upadhyay, and
+Himabindu Lakkaraju. 2022. {``Exploring Counterfactual Explanations
+Through the Lens of Adversarial Examples: A Theoretical and Empirical
+Analysis.''} In \emph{Proceedings of the 25th International Conference
+on Artificial Intelligence and Statistics}, edited by Gustau
+Camps-Valls, Francisco J. R. Ruiz, and Isabel Valera, 151:4574--94.
+Proceedings of Machine Learning Research. PMLR.
+\url{https://proceedings.mlr.press/v151/pawelczyk22a.html}.
+
+\bibitem[\citeproctext]{ref-ross2021learning}
+Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024. {``Learning
+Models for Actionable Recourse.''} In \emph{Proceedings of the 35th
+International Conference on Neural Information Processing Systems}. NIPS
+'21. Red Hook, NY, USA: Curran Associates Inc.
+
+\bibitem[\citeproctext]{ref-sauer2021counterfactual}
+Sauer, Axel, and Andreas Geiger. 2021. {``Counterfactual Generative
+Networks.''} \url{https://arxiv.org/abs/2101.06046}.
+
+\bibitem[\citeproctext]{ref-teney2020learning}
+Teney, Damien, Ehsan Abbasnedjad, and Anton van den Hengel. 2020.
+{``Learning What Makes a Difference from Counterfactual Examples and
+Gradient Supervision.''} In \emph{Computer Vision--ECCV 2020: 16th
+European Conference, Glasgow, UK, August 23--28, 2020, Proceedings, Part
+x 16}, 580--99. Springer.
+
+\bibitem[\citeproctext]{ref-wachter2017counterfactual}
+Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
+{``Counterfactual Explanations Without Opening the Black Box:
+{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
+841. \url{https://doi.org/10.2139/ssrn.3063289}.
+
+\bibitem[\citeproctext]{ref-wu2021polyjuice}
+Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
+2021. {``Polyjuice: Generating Counterfactuals for Explaining,
+Evaluating, and Improving Models.''} In \emph{Proceedings of the 59th
+Annual Meeting of the Association for Computational Linguistics and the
+11th International Joint Conference on Natural Language Processing
+(Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
+and Roberto Navigli, 6707--23. Online: Association for Computational
+Linguistics. \url{https://doi.org/10.18653/v1/2021.acl-long.523}.
+
+\end{CSLReferences}
+
+\newpage{}
+
+\FloatBarrier
+
+\setcounter{section}{0}
+\renewcommand{\thesection}{\Alph{section}}
+
+\setcounter{table}{0}
+\renewcommand{\thetable}{A\arabic{table}}
+
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{A\arabic{figure}}
+
+\section{Training Details}\label{training-details}
 
@@
 
-The hyperparameter choices are shown in
-Parameters~\ref{exr-gen-params-first-run}:
+The hyperparameter grids for the first investigation of the effect of
+generator parameters are shown in
+Parameters~\ref{exr-gen-params-first-run-train} and
+Parameters~\ref{exr-gen-params-first-run-eval}.
 
-\begin{exercise}[Parameters]\protect\hypertarget{exr-gen-params-first-run}{}\label{exr-gen-params-first-run}
+\begin{exercise}[Training
+Phase]\protect\hypertarget{exr-gen-params-first-run-train}{}\label{exr-gen-params-first-run-train}
 
@@
 \item
-  \textbf{generator\_params}:
+  Generator Parameters:
 
@@
   \item
-    \textbf{lambda\_cost}: 0.0, 0.001, 0.1
+    \(\lambda_{\text{cost}}\): \texttt{0.0,\ 0.001,\ 0.1}
   \item
-    \textbf{lambda\_energy}: 0.01, 0.05, 0.1, 0.5, 1.0, 5.0, 10.0, 15.0
+    \(\lambda_{\text{div}}\):
+    \texttt{0.01,\ 0.05,\ 0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 15.0}
   \item
-    \textbf{lr}: 1.0
+    Learning Rate: \texttt{1.0}
   \item
-    \textbf{maxiter}: 20, 50, 100
+    Maximum Iterations: \texttt{20,\ 50,\ 100}
   \item
-    \textbf{opt}: sgd
+    Optimizerimizer: \texttt{sgd}
   \end{itemize}
 \item
-  \textbf{generator\_type}: ecco, generic, omni, revise
+  Generator: \texttt{ecco,\ generic,\ omni,\ revise}
 \item
-  \textbf{training\_params}:
+  Training Parameters:
 
@@
   \item
-    \textbf{objective}: full, vanilla
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{exercise}
+
+\begin{exercise}[Evaluation
+Phase]\protect\hypertarget{exr-gen-params-first-run-eval}{}\label{exr-gen-params-first-run-eval}
+
+~
+
+\begin{itemize}
+\tightlist
+\item
+  Counterfactual Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Convergence: \texttt{max\_iter}
+  \item
+    Maximum Iterations: \texttt{100}
+  \item
+    No.~Individuals: \texttt{100}
+  \item
+    No.~Runs: \texttt{5}
+  \end{itemize}
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{cost}}\): \texttt{0.0}
+  \item
+    \(\lambda_{\text{div}}\):
+    \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 20.0}
+  \item
+    Learning Rate: \texttt{1.0}
+  \item
+    Maximum Iterations: \texttt{50}
+  \item
+    Optimizerimizer: \texttt{sgd}
   \end{itemize}
@@
 \item
-  \textbf{Cost (distance penalty)}: Results for all generators (except
-  \emph{Omni}) are quite bad, which can likely be attributed to
-  extremely bad results for some choices of the \textbf{Energy Penalty}
-  (results here are averaged). For \emph{ECCo} and \emph{Generic},
-  higher cost values generally lead to worse results.
+  \textbf{Cost} (Table~\ref{tbl-lin_sep-lambda_cost_exper}): Results for
+  all generators (except \emph{Omni}) are quite bad, which can likely be
+  attributed to extremely bad results for some choices of the
+  \textbf{Energy Penalty} (results here are averaged). For \emph{ECCo}
+  and \emph{Generic}, higher cost values generally lead to worse
+  results.
 \item
@@
   vanilla & 15 & \textit{REVISE} & -4.95 & 4.23 \\\bottomrule
+
+\end{longtable}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-lin_sep-lambda_cost_exper}Results for Linearly
+Separable data by cost penalty.}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  full & 0 & \textit{ECCo} & $-5.32 \cdot 10^{3}$ & $1.21 \cdot 10^{4}$ \\
+  full & 0 & \textit{Generic} & $-1.03 \cdot 10^{18}$ & $2.34 \cdot 10^{18}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.64}} & \color{blue}{\textbf{0.125}} \\
+  full & 0 & \textit{REVISE} & -15.4 & 12.9 \\
+  vanilla & 0 & \textit{ECCo} & -4.34 & 3.58 \\
+  vanilla & 0 & \textit{Generic} & -4.41 & 3.48 \\
+  vanilla & 0 & \textit{Omniscient} & -5.18 & 4.54 \\
+  vanilla & 0 & \textit{REVISE} & -4.93 & 4.23 \\
+  full & 0.001 & \textit{ECCo} & -362 & 811 \\
+  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.03 \cdot 10^{17}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.49}} & \color{blue}{\textbf{0.115}} \\
+  full & 0.001 & \textit{REVISE} & -15.5 & 13 \\
+  vanilla & 0.001 & \textit{ECCo} & -4.34 & 3.58 \\
+  vanilla & 0.001 & \textit{Generic} & -4.41 & 3.48 \\
+  vanilla & 0.001 & \textit{Omniscient} & -5.18 & 4.53 \\
+  vanilla & 0.001 & \textit{REVISE} & -4.93 & 4.23 \\
+  full & 0.1 & \textit{ECCo} & $-3.72 \cdot 10^{11}$ & $8.46 \cdot 10^{11}$ \\
+  full & 0.1 & \textit{Generic} & $-4.49 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5}} & \color{blue}{\textbf{0.112}} \\
+  full & 0.1 & \textit{REVISE} & -14.6 & 12.2 \\
+  vanilla & 0.1 & \textit{ECCo} & -4.34 & 3.58 \\
+  vanilla & 0.1 & \textit{Generic} & -4.41 & 3.48 \\
+  vanilla & 0.1 & \textit{Omniscient} & -5.18 & 4.54 \\
+  vanilla & 0.1 & \textit{REVISE} & -4.93 & 4.24 \\\bottomrule
 
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
-Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
-den Hengel. 2020. {``Counterfactual Vision and Language Learning.''} In
-\emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
-Recognition (CVPR)}, 10041--51.
-\url{https://doi.org/10.1109/CVPR42600.2020.01006}.
-
-\bibitem[\citeproctext]{ref-altmeyer2024faithful}
-Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia CS
-Liem. 2024. {``Faithful Model Explanations Through Energy-Constrained
-Conformal Counterfactuals.''} In \emph{Proceedings of the AAAI
-Conference on Artificial Intelligence}, 38:10829--37. 10.
-
-\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
-Freiesleben, Timo. 2022. {``The Intriguing Relation Between
-Counterfactual Explanations and Adversarial Examples.''} \emph{Minds and
-Machines} 32 (1): 77--109.
-
-\bibitem[\citeproctext]{ref-guo2023counternet}
-Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``CounterNet:
-End-to-End Training of Prediction Aware Counterfactual Explanations.''}
-In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
-Discovery and Data Mining}, 577--89. KDD '23. New York, NY, USA:
-Association for Computing Machinery.
-\url{https://doi.org/10.1145/3580305.3599290}.
-
-\bibitem[\citeproctext]{ref-joshi2019realistic}
-Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
-Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
-Actionable Explanations in Black-Box Decision Making Systems.''}
-\url{https://arxiv.org/abs/1907.09615}.
-
-\bibitem[\citeproctext]{ref-luu2023counterfactual}
-Luu, Hoai Linh, and Naoya Inoue. 2023. {``Counterfactual Adversarial
-Training for Improving Robustness of Pre-Trained Language Models.''} In
-\emph{Proceedings of the 37th Pacific Asia Conference on Language,
-Information and Computation}, 881--88.
-
-\bibitem[\citeproctext]{ref-pawelczyk2022exploring}
-Pawelczyk, Martin, Chirag Agarwal, Shalmali Joshi, Sohini Upadhyay, and
-Himabindu Lakkaraju. 2022. {``Exploring Counterfactual Explanations
-Through the Lens of Adversarial Examples: A Theoretical and Empirical
-Analysis.''} In \emph{Proceedings of the 25th International Conference
-on Artificial Intelligence and Statistics}, edited by Gustau
-Camps-Valls, Francisco J. R. Ruiz, and Isabel Valera, 151:4574--94.
-Proceedings of Machine Learning Research. PMLR.
-\url{https://proceedings.mlr.press/v151/pawelczyk22a.html}.
-
-\bibitem[\citeproctext]{ref-ross2021learning}
-Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024. {``Learning
-Models for Actionable Recourse.''} In \emph{Proceedings of the 35th
-International Conference on Neural Information Processing Systems}. NIPS
-'21. Red Hook, NY, USA: Curran Associates Inc.
-
-\bibitem[\citeproctext]{ref-sauer2021counterfactual}
-Sauer, Axel, and Andreas Geiger. 2021. {``Counterfactual Generative
-Networks.''} \url{https://arxiv.org/abs/2101.06046}.
-
-\bibitem[\citeproctext]{ref-teney2020learning}
-Teney, Damien, Ehsan Abbasnedjad, and Anton van den Hengel. 2020.
-{``Learning What Makes a Difference from Counterfactual Examples and
-Gradient Supervision.''} In \emph{Computer Vision--ECCV 2020: 16th
-European Conference, Glasgow, UK, August 23--28, 2020, Proceedings, Part
-x 16}, 580--99. Springer.
-
-\bibitem[\citeproctext]{ref-wachter2017counterfactual}
-Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
-{``Counterfactual Explanations Without Opening the Black Box:
-{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
-841. \url{https://doi.org/10.2139/ssrn.3063289}.
-
-\bibitem[\citeproctext]{ref-wu2021polyjuice}
-Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
-2021. {``Polyjuice: Generating Counterfactuals for Explaining,
-Evaluating, and Improving Models.''} In \emph{Proceedings of the 59th
-Annual Meeting of the Association for Computational Linguistics and the
-11th International Joint Conference on Natural Language Processing
-(Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
-and Roberto Navigli, 6707--23. Online: Association for Computational
-Linguistics. \url{https://doi.org/10.18653/v1/2021.acl-long.523}.
-
-\end{CSLReferences}
-
 \end{document}

### 2025-01-30 paper/sections/appendix.qmd
vamos
progress on appendix

@@
 
-# Appendix
+\FloatBarrier
 
-## Training Details
+\setcounter{section}{0}
+\renewcommand{\thesection}{\Alph{section}}
+
+\setcounter{table}{0}
+\renewcommand{\thetable}{A\arabic{table}}
+
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{A\arabic{figure}}
+
+<!-- # Supplementary Material {.appendix} -->
+
+# Training Details {.appendix}
 
@@
 
-The hyperparameter choices are shown in @exr-gen-params-first-run:
+The hyperparameter grids for the first investigation of the effect of generator parameters are shown in @exr-gen-params-first-run-train and @exr-gen-params-first-run-eval.
 
-::: {#exr-gen-params-first-run}
+::: {#exr-gen-params-first-run-train}
 
-## Parameters
+## Training Phase
 
@@
 dict = CTExperiments.from_toml(joinpath(config_dir,"gen_params.toml")) 
+println(CTExperiments.dict_to_quarto_markdown(dict))
+```
+
+:::
+
+::: {#exr-gen-params-first-run-eval}
+
+## Evaluation Phase
+
+```{julia}
+#| output: asis
+dict = CTExperiments.from_toml(joinpath(results_dir, "gen_params/mlp/lin_sep/evaluation/evaluation_grid_config.toml"))
 println(CTExperiments.dict_to_quarto_markdown(dict))
@@
 *REVISE* performs poorly across the board.
-- **Cost (distance penalty)**: Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged).
+- **Cost** (@tbl-lin_sep-lambda_cost_exper): Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged).
 For *ECCo* and *Generic*, higher cost values generally lead to worse results.
@@
 Results for Linearly Separable data by energy penalty.
+
+:::
+
+<!-- Cost -->
+
+```{julia}
+df = CSV.read("$(gen_params_dir)/lin_sep/evaluation/results/ce/objective---lambda_cost_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
+df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
+  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
+  df -> sort(df, [:lambda_cost_exper, :objective, :generator_type])
+```
+
+::: {#tbl-lin_sep-lambda_cost_exper}
+
+::: {.content-hidden unless-format="pdf"}
+
+```{julia}
+#| output: asis
+get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_cost_exper", backend=Val(:latex)) |>
+    inputs -> tabulate_results(inputs; wrap_table=false)
+```
+
+:::
+
+Results for Linearly Separable data by cost penalty.
 

### 2025-01-31 paper/preprint/paper.tex
uff this is going too slow

@@
   pdfauthor={Patrick Altmeyer; Arie van Deursen; Cynthia C. S. Liem},
-  pdfkeywords={Counterfactual Explanations, Explainable AI},
+  pdfkeywords={Counterfactual Explanations, Explainable
+AI, Representation Learning},
   colorlinks=true,
@@
 \begin{abstract}
-Counterfactual Explanations (CE) have emerged as a popular method to
-explain the predictions made by opaque machine learning models in a
-post-hoc fashion. We propose a novel approach that leverages
-counterfactuals during the training phase of models.
+Counterfactual Explanations (CE) have emerged as a popular tool to
+explain predictions made by opaque machine learning models: they explain
+how factual inputs need to change in order for some fitted model to
+produce some desired output. Much existing research has focused on
+identifying explanations that are not only valid but also deemed
+desirable with respect to the underlying data and stakeholder
+requirements. Recent work has shown that under this premise, the task of
+learning desirable explanations is effectively reassigned from the model
+itself to the (post-hoc) counterfactual explainer. Building on that
+work, we propose a novel model objective that leverages counterfactuals
+during the training phase (ad-hoc) in order to minimize the divergence
+between learned representations and desirable explanations. Through
+extensive experiments, we demonstrate that our proposed methodology
+facilitates training models that inherently deliver desirable
+explanations while maintaining high predictive performance.
 \end{abstract}
@@
 \def\sep{\textbullet\ }
-Counterfactual Explanations \sep 
-Explainable AI
+Counterfactual Explanations \sep Explainable AI \sep 
+Representation Learning
 
 \section{Introduction}\label{introduction}
+
+Today's prominence of artificial intelligence (AI) has largely been
+driven by advances in \textbf{representation learning}: instead of
+relying on features and rules that are carefully hand-crafted by humans,
+modern AIs are tasked with learning these representations from scratch,
+guided by narrow objectives such as predictive accuracy
+(\citeproc{ref-goodfellow2016deep}{I. Goodfellow, Bengio, and Courville
+2016}). Modern advances in computing have made it possible to provide
+such AIs with ever greater degrees of freedom to achieve that task,
+which has often led them to outperform traditionally more parsimonious
+models. Unfortunately, in doing so they have also learned increasingly
+complex representations that we can no longer easily interpret.
+
+This trend towards complexity for the sake of performance has come under
+serious scrutiny in recent years. At the very cusp of the deep learning
+revolution, I. J. Goodfellow, Shlens, and Szegedy
+(\citeproc{ref-goodfellow2014explaining}{2014}) showed that artificial
+neural networks (ANN) are highly sensitive to adversarial examples (AE):
+counterfactuals of model inputs that yield vastly different model
+predictions despite being semantically indifferent from their factual
+counterparts. Despite continued research efforts towards making models
+more robust to such counterfactuals, the issue remains unresolved even
+for models that are considered to small by today's standards
+(\citeproc{ref-kolter2023keynote}{Kolter 2023}). It seems that robust
+representation learning is fundamentally at odds with large degrees of
+freedom.
+
+Wilson (\citeproc{ref-wilson2020case}{2020})
+
+Rudin (\citeproc{ref-rudin2019stop}{2019})
 
@@
 
-(Wachter, Mittelstadt, and Russell 2017; Joshi et al. 2019; Altmeyer et
-al. 2024)
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}; \citeproc{ref-joshi2019realistic}{Joshi et al. 2019};
+\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024})
 
@@
 
-Sauer and Geiger (2021) generate counterfactual images for MNIST and
-ImageNet through independent mechanisms (IM): each IM learns
-class-conditional input distributions over a specific lower-dimensional,
-semantically meaningful factor, such as \emph{texture}, \emph{shape} and
-\emph{background}. The demonstrate that using these generated
-counterfactuals during classifier training improves model robustness.
-Similarly, Abbasnejad et al. (2020) argue that counterfactuals represent
-potentially useful training data in machine learning, especially in
-supervised settings where inputs may be reasonably mapped to multiple
-outputs. They, too, demonstrate the augmenting the training data of
-image classifiers can improve generalization.
+Sauer and Geiger (\citeproc{ref-sauer2021counterfactual}{2021}) generate
+counterfactual images for MNIST and ImageNet through independent
+mechanisms (IM): each IM learns class-conditional input distributions
+over a specific lower-dimensional, semantically meaningful factor, such
+as \emph{texture}, \emph{shape} and \emph{background}. The demonstrate
+that using these generated counterfactuals during classifier training
+improves model robustness. Similarly, Abbasnejad et al.
+(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
+counterfactuals represent potentially useful training data in machine
+learning, especially in supervised settings where inputs may be
+reasonably mapped to multiple outputs. They, too, demonstrate the
+augmenting the training data of image classifiers can improve
+generalization.
 
-Teney, Abbasnedjad, and Hengel (2020) propose an approach using
-counterfactuals in training that does not rely on data augmentation:
-they argue that counterfactual pairs typically already exist in training
-datasets. Specifically, their approach relies on, firstly, identifying
-similar input samples with different annotations and, secondly, ensuring
-that the gradient of the classifier aligns with the vector between pairs
-of counterfactual inputs using the cosine distance as a loss function
-(referred to as \emph{gradient supervision}) (\textbf{\emph{this might
-be useful for our task as well}}). In the natural language processing
-(NLP) domain, counterfactuals have similarly been used to improve models
-through data augmentation: Wu et al. (2021), propose POLYJUICE, a
-general-purpose counterfactual generator for language models. They
-demonstrate empirically that augmenting training data through POLYJUICE
-counterfactuals improves robustness in a number of NLP tasks.
+Teney, Abbasnedjad, and Hengel (\citeproc{ref-teney2020learning}{2020})
+propose an approach using counterfactuals in training that does not rely
+on data augmentation: they argue that counterfactual pairs typically
+already exist in training datasets. Specifically, their approach relies
+on, firstly, identifying similar input samples with different
+annotations and, secondly, ensuring that the gradient of the classifier
+aligns with the vector between pairs of counterfactual inputs using the
+cosine distance as a loss function (referred to as \emph{gradient
+supervision}) (\textbf{\emph{this might be useful for our task as
+well}}). In the natural language processing (NLP) domain,
+counterfactuals have similarly been used to improve models through data
+augmentation: Wu et al. (\citeproc{ref-wu2021polyjuice}{2021}), propose
+POLYJUICE, a general-purpose counterfactual generator for language
+models. They demonstrate empirically that augmenting training data
+through POLYJUICE counterfactuals improves robustness in a number of NLP
+tasks.
 
@@
 
-Freiesleben (2022) propose two definitional differences between
-Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly,
-and more importantly according to the authors, the term AE implies
-missclassification, which is not the case for CE (\textbf{\emph{this
-might be a useful notion for use to distinguish between adversarials and
-explanations during training}}); secondly, they argue that closeness
-plays a more critical role in the context of CE but confess that even
-counterfactuals that are not close might be relevant explanations.
-Pawelczyk et al. (2022) show that CE and AE are equivalent under certain
-conditions and derive upper bounds on the distances between them.
+Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) propose two
+definitional differences between Adversarial Examples (AE) and
+Counterfactual Explanations (CE): firstly, and more importantly
+according to the authors, the term AE implies missclassification, which
+is not the case for CE (\textbf{\emph{this might be a useful notion for
+use to distinguish between adversarials and explanations during
+training}}); secondly, they argue that closeness plays a more critical
+role in the context of CE but confess that even counterfactuals that are
+not close might be relevant explanations. Pawelczyk et al.
+(\citeproc{ref-pawelczyk2022exploring}{2022}) show that CE and AE are
+equivalent under certain conditions and derive upper bounds on the
+distances between them.
 
@@
 
-Guo, Nguyen, and Yadav (2023) are the first to propose end-to-end
-training pipeline that includes counterfactual explanations as part of
-the training prodeduce. In particular, they propose a specific network
-architecture that includes a predictor and CE generator network
-(\textbf{\emph{akin a GAN?}}), where the parameters of the CE generator
-network are learnable. Counterfactuals are generated during each
-training iteration and fed back to the predictor network
-(\textbf{\emph{here we are aligned}}). In contrast, we impose no
-restrictions on the neural network architecture at all.
-(\textbf{\emph{to ensure the one-hot encoding of categorical features is
-maintained, they simple use softmax (might be interesting for CE.jl)}})
-Interestingly, the authors find that their approach is sensitive to the
-choice of the loss function: only MSE seems to lead to good performance.
-They also demonstrate theoretically, that the objective function is
-difficult to optimize due to divergent gradients and suffers from poor
-adversarial robustness. (\textbf{\emph{because partial gradients with
-respect to the classification loss component and the counterfactual
-validity component point in opposite directions}}). To mitigate these
-issues, the authors use block-wise gradient descent: they first update
-with respect to classification loss and then use a second update with
-respect to the other loss components (\textbf{\emph{this might be useful
-for our task as well}}). Ross, Lakkaraju, and Bastani (2024) propose a
+Guo, Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the
+first to propose end-to-end training pipeline that includes
+counterfactual explanations as part of the training prodeduce. In
+particular, they propose a specific network architecture that includes a
+predictor and CE generator network (\textbf{\emph{akin a GAN?}}), where
+the parameters of the CE generator network are learnable.
+Counterfactuals are generated during each training iteration and fed
+back to the predictor network (\textbf{\emph{here we are aligned}}). In
+contrast, we impose no restrictions on the neural network architecture
+at all. (\textbf{\emph{to ensure the one-hot encoding of categorical
+features is maintained, they simple use softmax (might be interesting
+for CE.jl)}}) Interestingly, the authors find that their approach is
+sensitive to the choice of the loss function: only MSE seems to lead to
+good performance. They also demonstrate theoretically, that the
+objective function is difficult to optimize due to divergent gradients
+and suffers from poor adversarial robustness. (\textbf{\emph{because
+partial gradients with respect to the classification loss component and
+the counterfactual validity component point in opposite directions}}).
+To mitigate these issues, the authors use block-wise gradient descent:
+they first update with respect to classification loss and then use a
+second update with respect to the other loss components
+(\textbf{\emph{this might be useful for our task as well}}). Ross,
+Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024}) propose a
 way to train models that are guaranteed to provide recourse for
@@
 
-Luu and Inoue (2023) introduce Counterfactual Adversarial Training (CAT)
-with intention of improving generalization and robustness of language
-models. Specifically, they propose to proceed as follows: firstly,
-identify training samples that are subject to high predictive
-uncertainty (entropy); secondly, generate counterfactual explanations
-for those samples; and, finally, finetune the model on the augmented
-dataset that includes the generated counterfactuals.
+Luu and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
+Counterfactual Adversarial Training (CAT) with intention of improving
+generalization and robustness of language models. Specifically, they
+propose to proceed as follows: firstly, identify training samples that
+are subject to high predictive uncertainty (entropy); secondly, generate
+counterfactual explanations for those samples; and, finally, finetune
+the model on the augmented dataset that includes the generated
+counterfactuals.
 
@@
 
+\bibitem[\citeproctext]{ref-goodfellow2014explaining}
+Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
+{``Explaining and Harnessing Adversarial Examples.''}
+\url{https://arxiv.org/abs/1412.6572}.
+
+\bibitem[\citeproctext]{ref-goodfellow2016deep}
+Goodfellow, Ian, Yoshua Bengio, and Aaron Courville. 2016. \emph{Deep
+{Learning}}. {MIT Press}.
+
 \bibitem[\citeproctext]{ref-guo2023counternet}
@@
 \url{https://arxiv.org/abs/1907.09615}.
+
+\bibitem[\citeproctext]{ref-kolter2023keynote}
+Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
+IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)},
+xvi--. Los Alamitos, CA, USA: IEEE Computer Society.
+\url{https://doi.org/10.1109/SaTML54575.2023.00009}.
 
@@
 
+\bibitem[\citeproctext]{ref-rudin2019stop}
+Rudin, Cynthia. 2019. {``Stop Explaining Black Box Machine Learning
+Models for High Stakes Decisions and Use Interpretable Models
+Instead.''} \emph{Nature Machine Intelligence} 1 (5): 206--15.
+\url{https://doi.org/10.1038/s42256-019-0048-x}.
+
 \bibitem[\citeproctext]{ref-sauer2021counterfactual}
@@
 841. \url{https://doi.org/10.2139/ssrn.3063289}.
+
+\bibitem[\citeproctext]{ref-wilson2020case}
+Wilson, Andrew Gordon. 2020. {``The Case for Bayesian Deep Learning.''}
+\url{https://arxiv.org/abs/2001.10995}.
 

### 2025-01-31 paper/sections/abstract.qmd
uff this is going too slow

@@
 
-Counterfactual Explanations (CE) have emerged as a popular method to explain the predictions made by opaque machine learning models in a post-hoc fashion.
-We propose a novel approach that leverages counterfactuals during the training phase of models.
+Counterfactual Explanations (CE) have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output.
+Much existing research has focused on identifying explanations that are not only valid but also deemed desirable with respect to the underlying data and stakeholder requirements.
+Recent work has shown that under this premise, the task of learning desirable explanations is effectively reassigned from the model itself to the (post-hoc) counterfactual explainer.
+Building on that work, we propose a novel model objective that leverages counterfactuals during the training phase (ad-hoc) in order to minimize the divergence between learned representations and desirable explanations.
+Through extensive experiments, we demonstrate that our proposed methodology facilitates training models that inherently deliver desirable explanations while maintaining high predictive performance.

### 2025-01-31 paper/sections/intro.qmd
uff this is going too slow

# Introduction

Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern AIs are tasked with learning these representations from scratch, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep]. Modern advances in computing have made it possible to provide such AIs with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models. Unfortunately, in doing so they have also learned increasingly complex representations that we can no longer easily interpret.

This trend towards complexity for the sake of performance has come under serious scrutiny in recent years. At the very cusp of the deep learning revolution, @goodfellow2014explaining showed that artificial neural networks (ANN) are highly sensitive to adversarial examples (AE): counterfactuals of model inputs that yield vastly different model predictions despite being semantically indifferent from their factual counterparts. Despite continued research efforts towards making models more robust to such counterfactuals, the issue remains unresolved even for models that are considered to small by today's standards [@kolter2023keynote]. It seems that robust representation learning is fundamentally at odds with large degrees of freedom.

@wilson2020case

@rudin2019stop

### 2025-02-01 paper/preprint/paper.tex
intro

@@
 which has often led them to outperform traditionally more parsimonious
-models. Unfortunately, in doing so they have also learned increasingly
-complex representations that we can no longer easily interpret.
+models. Unfortunately, in doing so they also learn increasingly complex
+and highly sensitive representations that we can no longer easily
+interpret.
 
@@
 (\citeproc{ref-goodfellow2014explaining}{2014}) showed that artificial
-neural networks (ANN) are highly sensitive to adversarial examples (AE):
+neural networks (ANN) are sensitive to adversarial examples (AE):
 counterfactuals of model inputs that yield vastly different model
 predictions despite being semantically indifferent from their factual
-counterparts. Despite continued research efforts towards making models
-more robust to such counterfactuals, the issue remains unresolved even
-for models that are considered to small by today's standards
-(\citeproc{ref-kolter2023keynote}{Kolter 2023}). It seems that robust
-representation learning is fundamentally at odds with large degrees of
-freedom.
-
-Wilson (\citeproc{ref-wilson2020case}{2020})
-
-Rudin (\citeproc{ref-rudin2019stop}{2019})
+counterparts. Despite partially effective mitigation strategies such as
+\textbf{adversarial training}, truly robust deep learning (DL) remains
+unattainable even for models that are considered shallow by today's
+standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).
+
+Part of the problem is that great degrees of freedom provide room for
+many locally optimal solutions when using narrow objectives
+(\citeproc{ref-wilson2020case}{Wilson 2020}). Based purely on predictive
+performance, these solutions may seem to provide compelling explanations
+for the data at hand, even though they are in fact not grounded in
+meaningful semantics. In other words, the greatest strength of modern
+representation learning is also its greatest pitfall.
+
+This opaqueness has another dire consquence: since we cannot easily
+interpret the mapping from inputs to outputs, deploying such models in
+practive effectively means blindly relying on model predictions. On
+countless occasions, this has already caused real harm to people who
+were affected adversely and often unfairly by automated decision-making
+systems involving opaque models (\citeproc{ref-oneil2016weapons}{O'Neil
+2016}). Prominent voices have therefore argued to completely abolish the
+use of such ``black boxes'' along with any attempts to explain their
+predictions (\citeproc{ref-rudin2019stop}{Rudin 2019}): after all, if we
+had some interpretable abstraction of a ``black box'' that explains its
+decisions with full fidelity, there would be no need for the ``black
+box'' at all, because we could simply rely on the abstraction for
+decision-making.
+
+While we sympathise with that stance and share the aforementioned
+concerns, the reality is that the trend towards complexity has not been
+reversed. If anything, it has accelerated on the back of unprecented
+investments in computing infrastructure. With opaque AI here to stay, we
+believe that \textbf{explainable AI} (XAI) will conintue to play an
+important role in dealing with ``black boxes''.
 
@@
 Information and Computation}, 881--88.
+
+\bibitem[\citeproctext]{ref-oneil2016weapons}
+O'Neil, Cathy. 2016. \emph{Weapons of Math Destruction: {How} Big Data
+Increases Inequality and Threatens Democracy}. {Crown}.
 

### 2025-02-01 paper/sections/intro.qmd
intro

@@
 Modern advances in computing have made it possible to provide such AIs with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models.
-Unfortunately, in doing so they have also learned increasingly complex representations that we can no longer easily interpret.
+Unfortunately, in doing so they also learn increasingly complex and highly sensitive representations that we can no longer easily interpret.
 
 This trend towards complexity for the sake of performance has come under serious scrutiny in recent years.
-At the very cusp of the deep learning revolution, @goodfellow2014explaining showed that artificial neural networks (ANN) are highly sensitive to adversarial examples (AE): counterfactuals of model inputs that yield vastly different model predictions despite being semantically indifferent from their factual counterparts.
-Despite continued research efforts towards making models more robust to such counterfactuals, the issue remains unresolved even for models that are considered to small by today's standards [@kolter2023keynote].
-It seems that robust representation learning is fundamentally at odds with large degrees of freedom.
+At the very cusp of the deep learning revolution, @goodfellow2014explaining showed that artificial neural networks (ANN) are sensitive to adversarial examples (AE): counterfactuals of model inputs that yield vastly different model predictions despite being semantically indifferent from their factual counterparts.
+Despite partially effective mitigation strategies such as **adversarial training**, truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].
 
-@wilson2020case
+Part of the problem is that great degrees of freedom provide room for many locally optimal solutions when using narrow objectives [@wilson2020case].
+Based purely on predictive performance, these solutions may seem to provide compelling explanations for the data at hand, even though they are in fact not grounded in meaningful semantics.
+In other words, the greatest strength of modern representation learning is also its greatest pitfall.
 
-@rudin2019stop
+This opaqueness has another dire consquence: since we cannot easily interpret the mapping from inputs to outputs, deploying such models in practive effectively means blindly relying on model predictions.
+On countless occasions, this has already caused real harm to people who were affected adversely and often unfairly by automated decision-making systems involving opaque models [@oneil2016weapons].
+Prominent voices have therefore argued to completely abolish the use of such "black boxes" along with any attempts to explain their predictions [@rudin2019stop]: after all, if we had some interpretable abstraction of a "black box" that explains its decisions with full fidelity, there would be no need for the "black box" at all, because we could simply rely on the abstraction for decision-making.
+
+While we sympathise with that stance and share the aforementioned concerns, the reality is that the trend towards complexity has not been reversed.
+If anything, it has accelerated on the back of unprecented investments in computing infrastructure.
+With opaque AI here to stay, we believe that **explainable AI** (XAI) will conintue to play an important role in dealing with "black boxes".

### 2025-02-03 paper/preprint/paper.tex
intro basically done

@@
 
-Part of the problem is that great degrees of freedom provide room for
-many locally optimal solutions when using narrow objectives
-(\citeproc{ref-wilson2020case}{Wilson 2020}). Based purely on predictive
-performance, these solutions may seem to provide compelling explanations
-for the data at hand, even though they are in fact not grounded in
-meaningful semantics. In other words, the greatest strength of modern
-representation learning is also its greatest pitfall.
-
-This opaqueness has another dire consquence: since we cannot easily
-interpret the mapping from inputs to outputs, deploying such models in
-practive effectively means blindly relying on model predictions. On
-countless occasions, this has already caused real harm to people who
-were affected adversely and often unfairly by automated decision-making
-systems involving opaque models (\citeproc{ref-oneil2016weapons}{O'Neil
-2016}). Prominent voices have therefore argued to completely abolish the
-use of such ``black boxes'' along with any attempts to explain their
-predictions (\citeproc{ref-rudin2019stop}{Rudin 2019}): after all, if we
-had some interpretable abstraction of a ``black box'' that explains its
-decisions with full fidelity, there would be no need for the ``black
-box'' at all, because we could simply rely on the abstraction for
-decision-making.
-
-While we sympathise with that stance and share the aforementioned
-concerns, the reality is that the trend towards complexity has not been
-reversed. If anything, it has accelerated on the back of unprecented
-investments in computing infrastructure. With opaque AI here to stay, we
-believe that \textbf{explainable AI} (XAI) will conintue to play an
-important role in dealing with ``black boxes''.
+Part of the problem is that high degrees of freedom provide room for
+many solutions that are locally optimal with respect to narrow
+objectives (\citeproc{ref-wilson2020case}{Wilson 2020}). Based purely on
+predictive performance, these solutions may seem to provide compelling
+explanations for the data, when in fact they are based on purely
+associative, semantically meaningless patterns. This poses two related
+challenges: firstly, it makes these models inherently opaque, since
+humans cannot simply interpret what type of explanation the complex
+learned representations correspond to; secondly, even if we could
+resolve the first challenge, it is not obvious how to mitigate models
+from learning representations that correspond to meaningless and
+undesirable explanations.
+
+The first challenge has attracted an abundance of research on
+\textbf{explainable AI} (XAI) which aims to develop tools to derive
+explanations from complex model representations. This can mitigate a
+scenario in which we deploy opaque models and blindly rely on their
+predictions. On countless occasions, this scenario has already occurred
+in practice and caused real harm to people who were affected adversely
+and often unfairly by automated decision-making systems involving opaque
+models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI
+tools can aide us in monitoring models and providing affected
+individuals with recourse
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}).
+
+To our surprise, the second challenge has not yet attracted any
+consolidated research effort. Specifically, there has been no concerted
+effort towards improving model \textbf{explainability}, which we define
+here as the degree to which learned representations correspond to
+explanations that are deemed desirable by humans. Instead, the choice
+has typically been to improve the capacity of XAI tools to identify the
+subset explanations that are both desirable and valid for any given
+model, independent of whether the learned representations are also
+compatible with undesirable explanations
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+Fortunately, recent findings indicate that explainability can arise as
+byproduct of regularization techniques aimed at other objectives such as
+robustness, generalization and generative capacity Altmeyer et al.
+(\citeproc{ref-altmeyer2024faithful}{2024}).
+
+In this work, we build on these findings and introduce a
+\textbf{counterfactual training}: a novel approach geared explicitly
+towards aligning model representations with desirable explanations.
+
+From this perspective, adversarial training induces models to
+``unlearn'' representations that are susceptible to the semantically
+most meaningless explanations---adversarial examples.
 
@@
 over a specific lower-dimensional, semantically meaningful factor, such
-as \emph{texture}, \emph{shape} and \emph{background}. The demonstrate
+as \emph{texture}, \emph{shape} and \emph{background}. They demonstrate
 that using these generated counterfactuals during classifier training
@@
 
-\bibitem[\citeproctext]{ref-rudin2019stop}
-Rudin, Cynthia. 2019. {``Stop Explaining Black Box Machine Learning
-Models for High Stakes Decisions and Use Interpretable Models
-Instead.''} \emph{Nature Machine Intelligence} 1 (5): 206--15.
-\url{https://doi.org/10.1038/s42256-019-0048-x}.
-
 \bibitem[\citeproctext]{ref-sauer2021counterfactual}
@@
 Networks.''} \url{https://arxiv.org/abs/2101.06046}.
+
+\bibitem[\citeproctext]{ref-schut2021generating}
+Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
+Sacaleanu, Yarin Gal, et al. 2021. {``Generating {Interpretable
+Counterfactual Explanations By Implicit Minimisation} of {Epistemic} and
+{Aleatoric Uncertainties}.''} In \emph{International {Conference} on
+{Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.
 

### 2025-02-03 paper/sections/intro.qmd
intro basically done

--- 2025-01-31 paper/sections/intro.qmd
@@
 Modern advances in computing have made it possible to provide such AIs with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models.
-Unfortunately, in doing so they have also learned increasingly complex representations that we can no longer easily interpret.
+Unfortunately, in doing so they also learn increasingly complex and highly sensitive representations that we can no longer easily interpret.
 
 This trend towards complexity for the sake of performance has come under serious scrutiny in recent years.
-At the very cusp of the deep learning revolution, @goodfellow2014explaining showed that artificial neural networks (ANN) are highly sensitive to adversarial examples (AE): counterfactuals of model inputs that yield vastly different model predictions despite being semantically indifferent from their factual counterparts.
-Despite continued research efforts towards making models more robust to such counterfactuals, the issue remains unresolved even for models that are considered to small by today's standards [@kolter2023keynote].
-It seems that robust representation learning is fundamentally at odds with large degrees of freedom.
+At the very cusp of the deep learning revolution, @goodfellow2014explaining showed that artificial neural networks (ANN) are sensitive to adversarial examples (AE): counterfactuals of model inputs that yield vastly different model predictions despite being semantically indifferent from their factual counterparts.
+Despite partially effective mitigation strategies such as **adversarial training**, truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].
 
-@wilson2020case
+Part of the problem is that high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].
+Based purely on predictive performance, these solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns.
+This poses two related challenges: firstly, it makes these models inherently opaque, since humans cannot simply interpret what type of explanation the complex learned representations correspond to; secondly, even if we could resolve the first challenge, it is not obvious how to mitigate models from learning representations that correspond to meaningless and undesirable explanations.
 
-@rudin2019stop
+The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations.
+This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions.
+On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems involving opaque models [@oneil2016weapons].
+Effective XAI tools can aide us in monitoring models and providing affected individuals with recourse [@wachter2017counterfactual].
+
+To our surprise, the second challenge has not yet attracted any consolidated research effort.
+Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are deemed desirable by humans.
+Instead, the choice has typically been to improve the capacity of XAI tools to identify the subset explanations that are both desirable and valid for any given model, independent of whether the learned representations are also compatible with undesirable explanations [@altmeyer2024faithful].
+Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization and generative capacity [@schut2021generating, @augustin2020adversarial, @altmeyer2024faithful].
+
+Building on these findings, we introduce a **counterfactual training**: a novel approach geared explicitly towards aligning model representations with desirable explanations.
+
+From this perspective, adversarial training induces models to "unlearn" representations that are susceptible to the semantically most meaningless explanations---adversarial examples.

### 2025-02-03 paper/body_ecml.tex
beautiful

--- 2025-02-03 paper/preprint/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-]{article}
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
-    \setmainfont[]{Latin Modern Roman}
-  \setmathfont[]{Latin Modern Math}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
-% Make \paragraph and \subparagraph free-standing
-\makeatletter
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-\makeatother
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
-% Set default figure placement to htbp
-\def\fps@figure{htbp}
-\makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\usepackage{arxiv}
-\usepackage{orcidlink}
-\usepackage{amsmath}
-\usepackage[T1]{fontenc}
-\usepackage{placeins}
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
-\theoremstyle{remark}
-\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
-\newtheorem*{remark}{Remark}
-\newtheorem*{solution}{Solution}
-\newtheorem{refremark}{Remark}[section]
-\newtheorem{refsolution}{Solution}[section]
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-
-\usepackage{bookmark}
-
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
-  pdfauthor={Patrick Altmeyer; Arie van Deursen; Cynthia C. S. Liem},
-  pdfkeywords={Counterfactual Explanations, Explainable
-AI, Representation Learning},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-\usepackage{lineno}
-\linenumbers
-\newcommand{\runninghead}{A Preprint }
-\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
-\title{Counterfactual Training: Teaching Models Plausible and Actionable
-Explanations}
-\def\asep{\\\\\\ } % default: all authors on same column
-\def\asep{\And }
-\author{\textbf{Patrick
-Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
-Engineering, Mathematics and Computer Science\\Delft University of
-Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
-van Deursen}\\Faculty of Electrical Engineering, Mathematics and
-Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
-C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\}
-\date{}
-\begin{document}
-\maketitle
 \begin{abstract}
@@
 explanations while maintaining high predictive performance.
+\keywords{Counterfactual Explanations \and Explainable
+AI \and Representation Learning}
 \end{abstract}
-{\bfseries \emph Keywords}
-\def\sep{\textbullet\ }
-Counterfactual Explanations \sep Explainable AI \sep 
-Representation Learning
 
@@
 guided by narrow objectives such as predictive accuracy
-(\citeproc{ref-goodfellow2016deep}{I. Goodfellow, Bengio, and Courville
-2016}). Modern advances in computing have made it possible to provide
-such AIs with ever greater degrees of freedom to achieve that task,
-which has often led them to outperform traditionally more parsimonious
-models. Unfortunately, in doing so they also learn increasingly complex
-and highly sensitive representations that we can no longer easily
-interpret.
+\cite{goodfellow2016deep}. Modern advances in computing have made it
+possible to provide such AIs with ever greater degrees of freedom to
+achieve that task, which has often led them to outperform traditionally
+more parsimonious models. Unfortunately, in doing so they also learn
+increasingly complex and highly sensitive representations that we can no
+longer easily interpret.
 
@@
 serious scrutiny in recent years. At the very cusp of the deep learning
-revolution, I. J. Goodfellow, Shlens, and Szegedy
-(\citeproc{ref-goodfellow2014explaining}{2014}) showed that artificial
+revolution, \cite{goodfellow2014explaining} showed that artificial
 neural networks (ANN) are sensitive to adversarial examples (AE):
@@
 unattainable even for models that are considered shallow by today's
-standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).
+standards \cite{kolter2023keynote}.
 
@@
 many solutions that are locally optimal with respect to narrow
-objectives (\citeproc{ref-wilson2020case}{Wilson 2020}). Based purely on
-predictive performance, these solutions may seem to provide compelling
-explanations for the data, when in fact they are based on purely
-associative, semantically meaningless patterns. This poses two related
-challenges: firstly, it makes these models inherently opaque, since
-humans cannot simply interpret what type of explanation the complex
-learned representations correspond to; secondly, even if we could
-resolve the first challenge, it is not obvious how to mitigate models
-from learning representations that correspond to meaningless and
-undesirable explanations.
+objectives \cite{wilson2020case}. Based purely on predictive
+performance, these solutions may seem to provide compelling explanations
+for the data, when in fact they are based on purely associative,
+semantically meaningless patterns. This poses two related challenges:
+firstly, it makes these models inherently opaque, since humans cannot
+simply interpret what type of explanation the complex learned
+representations correspond to; secondly, even if we could resolve the
+first challenge, it is not obvious how to mitigate models from learning
+representations that correspond to meaningless and undesirable
+explanations.
 
@@
 and often unfairly by automated decision-making systems involving opaque
-models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI
-tools can aide us in monitoring models and providing affected
-individuals with recourse
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}).
+models \cite{oneil2016weapons}. Effective XAI tools can aide us in
+monitoring models and providing affected individuals with recourse
+\cite{wachter2017counterfactual}.
 
@@
 model, independent of whether the learned representations are also
-compatible with undesirable explanations
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+compatible with undesirable explanations \cite{altmeyer2024faithful}.
 Fortunately, recent findings indicate that explainability can arise as
 byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization and generative capacity Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}).
+robustness, generalization and generative capacity
+\cite{schut2021generating}.
 
-In this work, we build on these findings and introduce a
-\textbf{counterfactual training}: a novel approach geared explicitly
-towards aligning model representations with desirable explanations.
+Building on these findings, we introduce \textbf{counterfactual
+training}: a novel regularization technique geared explicitly towards
+aligning model representations with desirable explanations. Our
+contributions are as follows:
 
-From this perspective, adversarial training induces models to
-``unlearn'' representations that are susceptible to the semantically
-most meaningless explanations---adversarial examples.
+\begin{itemize}
+\tightlist
+\item
+  We discuss existing related work on improving models and consolidate
+  it through the lens of counterfactual explanations
+  (Section~\ref{sec-lit}).
+\item
+  We present our proposed methodological framework that leverages
+  faithful counterfactual explanations during the training phase of
+  models to achieve the explainability objective
+  (Section~\ref{sec-method}).
+\item
+  Through extensive experiments we demonstrate the counterfactual
+  training improve model explainability while maintaining high
+  predictive performance. We run ablation studies and grid searches to
+  understand how the underlying model components and hyperparameters
+  affect outcomes. (Section~\ref{sec-experiments}).
+\end{itemize}
+
+Despite limitations of our approach discussed in
+Section~\ref{sec-discussion}, we conclude that counterfactual training
+provides a practical framework for researchers and practitioners
+interested in making opaque models more trustworthy
+Section~\ref{sec-conclusion}. We also believe that this work serves as
+an opportunity for XAI researchers to reevaluate the premise of
+improving XAI tools without improving models.
 
@@
 
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}; \citeproc{ref-joshi2019realistic}{Joshi et al. 2019};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024})
+\cite{wachter2017counterfactual, joshi2019realistic, altmeyer2024faithful}
 
@@
 
-Sauer and Geiger (\citeproc{ref-sauer2021counterfactual}{2021}) generate
-counterfactual images for MNIST and ImageNet through independent
-mechanisms (IM): each IM learns class-conditional input distributions
-over a specific lower-dimensional, semantically meaningful factor, such
-as \emph{texture}, \emph{shape} and \emph{background}. They demonstrate
-that using these generated counterfactuals during classifier training
-improves model robustness. Similarly, Abbasnejad et al.
-(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
+\cite{sauer2021counterfactual} generate counterfactual images for MNIST
+and ImageNet through independent mechanisms (IM): each IM learns
+class-conditional input distributions over a specific lower-dimensional,
+semantically meaningful factor, such as \emph{texture}, \emph{shape} and
+\emph{background}. They demonstrate that using these generated
+counterfactuals during classifier training improves model robustness.
+Similarly, \cite{abbasnejad2020counterfactual} argue that
 counterfactuals represent potentially useful training data in machine
@@
 
-Teney, Abbasnedjad, and Hengel (\citeproc{ref-teney2020learning}{2020})
-propose an approach using counterfactuals in training that does not rely
-on data augmentation: they argue that counterfactual pairs typically
-already exist in training datasets. Specifically, their approach relies
-on, firstly, identifying similar input samples with different
-annotations and, secondly, ensuring that the gradient of the classifier
-aligns with the vector between pairs of counterfactual inputs using the
-cosine distance as a loss function (referred to as \emph{gradient
-supervision}) (\textbf{\emph{this might be useful for our task as
-well}}). In the natural language processing (NLP) domain,
-counterfactuals have similarly been used to improve models through data
-augmentation: Wu et al. (\citeproc{ref-wu2021polyjuice}{2021}), propose
-POLYJUICE, a general-purpose counterfactual generator for language
-models. They demonstrate empirically that augmenting training data
-through POLYJUICE counterfactuals improves robustness in a number of NLP
-tasks.
+\cite{teney2020learning} propose an approach using counterfactuals in
+training that does not rely on data augmentation: they argue that
+counterfactual pairs typically already exist in training datasets.
+Specifically, their approach relies on, firstly, identifying similar
+input samples with different annotations and, secondly, ensuring that
+the gradient of the classifier aligns with the vector between pairs of
+counterfactual inputs using the cosine distance as a loss function
+(referred to as \emph{gradient supervision}) (\textbf{\emph{this might
+be useful for our task as well}}). In the natural language processing
+(NLP) domain, counterfactuals have similarly been used to improve models
+through data augmentation: \cite{wu2021polyjuice}, propose POLYJUICE, a
+general-purpose counterfactual generator for language models. They
+demonstrate empirically that augmenting training data through POLYJUICE
+counterfactuals improves robustness in a number of NLP tasks.
 
@@
 
-Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) propose two
-definitional differences between Adversarial Examples (AE) and
-Counterfactual Explanations (CE): firstly, and more importantly
-according to the authors, the term AE implies missclassification, which
-is not the case for CE (\textbf{\emph{this might be a useful notion for
-use to distinguish between adversarials and explanations during
-training}}); secondly, they argue that closeness plays a more critical
-role in the context of CE but confess that even counterfactuals that are
-not close might be relevant explanations. Pawelczyk et al.
-(\citeproc{ref-pawelczyk2022exploring}{2022}) show that CE and AE are
+From this perspective, adversarial training induces models to
+``unlearn'' representations that are susceptible to the semantically
+most meaningless explanations---adversarial examples.
+
+\cite{freiesleben2022intriguing} propose two definitional differences
+between Adversarial Examples (AE) and Counterfactual Explanations (CE):
+firstly, and more importantly according to the authors, the term AE
+implies missclassification, which is not the case for CE
+(\textbf{\emph{this might be a useful notion for use to distinguish
+between adversarials and explanations during training}}); secondly, they
+argue that closeness plays a more critical role in the context of CE but
+confess that even counterfactuals that are not close might be relevant
+explanations. \cite{pawelczyk2022exploring} show that CE and AE are
 equivalent under certain conditions and derive upper bounds on the
@@
 
-Guo, Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the
-first to propose end-to-end training pipeline that includes
-counterfactual explanations as part of the training prodeduce. In
-particular, they propose a specific network architecture that includes a
-predictor and CE generator network (\textbf{\emph{akin a GAN?}}), where
-the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed
-back to the predictor network (\textbf{\emph{here we are aligned}}). In
-contrast, we impose no restrictions on the neural network architecture
-at all. (\textbf{\emph{to ensure the one-hot encoding of categorical
-features is maintained, they simple use softmax (might be interesting
-for CE.jl)}}) Interestingly, the authors find that their approach is
-sensitive to the choice of the loss function: only MSE seems to lead to
-good performance. They also demonstrate theoretically, that the
-objective function is difficult to optimize due to divergent gradients
-and suffers from poor adversarial robustness. (\textbf{\emph{because
-partial gradients with respect to the classification loss component and
-the counterfactual validity component point in opposite directions}}).
-To mitigate these issues, the authors use block-wise gradient descent:
-they first update with respect to classification loss and then use a
-second update with respect to the other loss components
-(\textbf{\emph{this might be useful for our task as well}}). Ross,
-Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024}) propose a
-way to train models that are guaranteed to provide recourse for
-individuals with high probability. The approach builds on adversarial
-training (\textbf{\emph{here we are aligned}}), where in this context
-adversarial examples are actively encouraged to exist, but only target
-attacks with respect to the positive class. The proposed method allows
-for imposing a set of actionable recourse ex-ante: for example, users
-can impose mutability constraints for features (\textbf{\emph{here we
-are aligned}}). (\textbf{\emph{To solve their objective function more
+\cite{guo2023counternet} are the first to propose end-to-end training
+pipeline that includes counterfactual explanations as part of the
+training prodeduce. In particular, they propose a specific network
+architecture that includes a predictor and CE generator network
+(\textbf{\emph{akin a GAN?}}), where the parameters of the CE generator
+network are learnable. Counterfactuals are generated during each
+training iteration and fed back to the predictor network
+(\textbf{\emph{here we are aligned}}). In contrast, we impose no
+restrictions on the neural network architecture at all.
+(\textbf{\emph{to ensure the one-hot encoding of categorical features is
+maintained, they simple use softmax (might be interesting for CE.jl)}})
+Interestingly, the authors find that their approach is sensitive to the
+choice of the loss function: only MSE seems to lead to good performance.
+They also demonstrate theoretically, that the objective function is
+difficult to optimize due to divergent gradients and suffers from poor
+adversarial robustness. (\textbf{\emph{because partial gradients with
+respect to the classification loss component and the counterfactual
+validity component point in opposite directions}}). To mitigate these
+issues, the authors use block-wise gradient descent: they first update
+with respect to classification loss and then use a second update with
+respect to the other loss components (\textbf{\emph{this might be useful
+for our task as well}}). \cite{ross2021learning} propose a way to train
+models that are guaranteed to provide recourse for individuals with high
+probability. The approach builds on adversarial training
+(\textbf{\emph{here we are aligned}}), where in this context adversarial
+examples are actively encouraged to exist, but only target attacks with
+respect to the positive class. The proposed method allows for imposing a
+set of actionable recourse ex-ante: for example, users can impose
+mutability constraints for features (\textbf{\emph{here we are
+aligned}}). (\textbf{\emph{To solve their objective function more
 efficiently, they use a first-order Taylor approximation to approximate
@@
 
-Luu and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
-Counterfactual Adversarial Training (CAT) with intention of improving
-generalization and robustness of language models. Specifically, they
-propose to proceed as follows: firstly, identify training samples that
-are subject to high predictive uncertainty (entropy); secondly, generate
-counterfactual explanations for those samples; and, finally, finetune
-the model on the augmented dataset that includes the generated
-counterfactuals.
+\cite{luu2023counterfactual} introduce Counterfactual Adversarial
+Training (CAT) with intention of improving generalization and robustness
+of language models. Specifically, they propose to proceed as follows:
+firstly, identify training samples that are subject to high predictive
+uncertainty (entropy); secondly, generate counterfactual explanations
+for those samples; and, finally, finetune the model on the augmented
+dataset that includes the generated counterfactuals.
 
-\section{Counterfactual Training}\label{counterfactual-training}
+\section{Counterfactual Training}\label{sec-method}
 
@@
 \section{Conclusion}\label{sec-conclusion}
-
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
-Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
-den Hengel. 2020. {``Counterfactual Vision and Language Learning.''} In
-\emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
-Recognition (CVPR)}, 10041--51.
-\url{https://doi.org/10.1109/CVPR42600.2020.01006}.
-
-\bibitem[\citeproctext]{ref-altmeyer2024faithful}
-Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia CS
-Liem. 2024. {``Faithful Model Explanations Through Energy-Constrained
-Conformal Counterfactuals.''} In \emph{Proceedings of the AAAI
-Conference on Artificial Intelligence}, 38:10829--37. 10.
-
-\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
-Freiesleben, Timo. 2022. {``The Intriguing Relation Between
-Counterfactual Explanations and Adversarial Examples.''} \emph{Minds and
-Machines} 32 (1): 77--109.
-
-\bibitem[\citeproctext]{ref-goodfellow2014explaining}
-Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
-{``Explaining and Harnessing Adversarial Examples.''}
-\url{https://arxiv.org/abs/1412.6572}.
-
-\bibitem[\citeproctext]{ref-goodfellow2016deep}
-Goodfellow, Ian, Yoshua Bengio, and Aaron Courville. 2016. \emph{Deep
-{Learning}}. {MIT Press}.
-
-\bibitem[\citeproctext]{ref-guo2023counternet}
-Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``CounterNet:
-End-to-End Training of Prediction Aware Counterfactual Explanations.''}
-In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
-Discovery and Data Mining}, 577--89. KDD '23. New York, NY, USA:
-Association for Computing Machinery.
-\url{https://doi.org/10.1145/3580305.3599290}.
-
-\bibitem[\citeproctext]{ref-joshi2019realistic}
-Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
-Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
-Actionable Explanations in Black-Box Decision Making Systems.''}
-\url{https://arxiv.org/abs/1907.09615}.
-
-\bibitem[\citeproctext]{ref-kolter2023keynote}
-Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
-IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)},
-xvi--. Los Alamitos, CA, USA: IEEE Computer Society.
-\url{https://doi.org/10.1109/SaTML54575.2023.00009}.
-
-\bibitem[\citeproctext]{ref-luu2023counterfactual}
-Luu, Hoai Linh, and Naoya Inoue. 2023. {``Counterfactual Adversarial
-Training for Improving Robustness of Pre-Trained Language Models.''} In
-\emph{Proceedings of the 37th Pacific Asia Conference on Language,
-Information and Computation}, 881--88.
-
-\bibitem[\citeproctext]{ref-oneil2016weapons}
-O'Neil, Cathy. 2016. \emph{Weapons of Math Destruction: {How} Big Data
-Increases Inequality and Threatens Democracy}. {Crown}.
-
-\bibitem[\citeproctext]{ref-pawelczyk2022exploring}
-Pawelczyk, Martin, Chirag Agarwal, Shalmali Joshi, Sohini Upadhyay, and
-Himabindu Lakkaraju. 2022. {``Exploring Counterfactual Explanations
-Through the Lens of Adversarial Examples: A Theoretical and Empirical
-Analysis.''} In \emph{Proceedings of the 25th International Conference
-on Artificial Intelligence and Statistics}, edited by Gustau
-Camps-Valls, Francisco J. R. Ruiz, and Isabel Valera, 151:4574--94.
-Proceedings of Machine Learning Research. PMLR.
-\url{https://proceedings.mlr.press/v151/pawelczyk22a.html}.
-
-\bibitem[\citeproctext]{ref-ross2021learning}
-Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024. {``Learning
-Models for Actionable Recourse.''} In \emph{Proceedings of the 35th
-International Conference on Neural Information Processing Systems}. NIPS
-'21. Red Hook, NY, USA: Curran Associates Inc.
-
-\bibitem[\citeproctext]{ref-sauer2021counterfactual}
-Sauer, Axel, and Andreas Geiger. 2021. {``Counterfactual Generative
-Networks.''} \url{https://arxiv.org/abs/2101.06046}.
-
-\bibitem[\citeproctext]{ref-schut2021generating}
-Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
-Sacaleanu, Yarin Gal, et al. 2021. {``Generating {Interpretable
-Counterfactual Explanations By Implicit Minimisation} of {Epistemic} and
-{Aleatoric Uncertainties}.''} In \emph{International {Conference} on
-{Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.
-
-\bibitem[\citeproctext]{ref-teney2020learning}
-Teney, Damien, Ehsan Abbasnedjad, and Anton van den Hengel. 2020.
-{``Learning What Makes a Difference from Counterfactual Examples and
-Gradient Supervision.''} In \emph{Computer Vision--ECCV 2020: 16th
-European Conference, Glasgow, UK, August 23--28, 2020, Proceedings, Part
-x 16}, 580--99. Springer.
-
-\bibitem[\citeproctext]{ref-wachter2017counterfactual}
-Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
-{``Counterfactual Explanations Without Opening the Black Box:
-{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
-841. \url{https://doi.org/10.2139/ssrn.3063289}.
-
-\bibitem[\citeproctext]{ref-wilson2020case}
-Wilson, Andrew Gordon. 2020. {``The Case for Bayesian Deep Learning.''}
-\url{https://arxiv.org/abs/2001.10995}.
-
-\bibitem[\citeproctext]{ref-wu2021polyjuice}
-Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
-2021. {``Polyjuice: Generating Counterfactuals for Explaining,
-Evaluating, and Improving Models.''} In \emph{Proceedings of the 59th
-Annual Meeting of the Association for Computational Linguistics and the
-11th International Joint Conference on Natural Language Processing
-(Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
-and Roberto Navigli, 6707--23. Online: Association for Computational
-Linguistics. \url{https://doi.org/10.18653/v1/2021.acl-long.523}.
-
-\end{CSLReferences}
-
-\newpage{}
-
-\FloatBarrier
-
-\setcounter{section}{0}
-\renewcommand{\thesection}{\Alph{section}}
-
-\setcounter{table}{0}
-\renewcommand{\thetable}{A\arabic{table}}
-
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{A\arabic{figure}}
-
-\section{Training Details}\label{training-details}
-
-\subsection{Initial Grid Search}\label{initial-grid-search}
-
-For the initial round of experiments we
-
-\subsubsection{Generator Parameters}\label{generator-parameters}
-
-The hyperparameter grids for the first investigation of the effect of
-generator parameters are shown in
-Parameters~\ref{exr-gen-params-first-run-train} and
-Parameters~\ref{exr-gen-params-first-run-eval}.
-
-\begin{exercise}[Training
-Phase]\protect\hypertarget{exr-gen-params-first-run-train}{}\label{exr-gen-params-first-run-train}
-
-~
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{cost}}\): \texttt{0.0,\ 0.001,\ 0.1}
-  \item
-    \(\lambda_{\text{div}}\):
-    \texttt{0.01,\ 0.05,\ 0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 15.0}
-  \item
-    Learning Rate: \texttt{1.0}
-  \item
-    Maximum Iterations: \texttt{20,\ 50,\ 100}
-  \item
-    Optimizerimizer: \texttt{sgd}
-  \end{itemize}
-\item
-  Generator: \texttt{ecco,\ generic,\ omni,\ revise}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{exercise}
-
-\begin{exercise}[Evaluation
-Phase]\protect\hypertarget{exr-gen-params-first-run-eval}{}\label{exr-gen-params-first-run-eval}
-
-~
-
-\begin{itemize}
-\tightlist
-\item
-  Counterfactual Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Convergence: \texttt{max\_iter}
-  \item
-    Maximum Iterations: \texttt{100}
-  \item
-    No.~Individuals: \texttt{100}
-  \item
-    No.~Runs: \texttt{5}
-  \end{itemize}
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{cost}}\): \texttt{0.0}
-  \item
-    \(\lambda_{\text{div}}\):
-    \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 20.0}
-  \item
-    Learning Rate: \texttt{1.0}
-  \item
-    Maximum Iterations: \texttt{50}
-  \item
-    Optimizerimizer: \texttt{sgd}
-  \end{itemize}
-\end{itemize}
-
-\end{exercise}
-
-\paragraph{Linearly Separable}\label{linearly-separable}
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-lin_sep-lambda_energy_exper}):
-  \emph{ECCo} generally does yield better results than \emph{Vanilla}
-  for higher choices of the energy penalty (10,15) during training.
-  \emph{Generic} performs poorly accross the board. \emph{Omni} seems to
-  have an anchoring effect, in that it never performs terribly but also
-  never as good as the best \emph{ECCo} results. \emph{REVISE} performs
-  poorly across the board.
-\item
-  \textbf{Cost} (Table~\ref{tbl-lin_sep-lambda_cost_exper}): Results for
-  all generators (except \emph{Omni}) are quite bad, which can likely be
-  attributed to extremely bad results for some choices of the
-  \textbf{Energy Penalty} (results here are averaged). For \emph{ECCo}
-  and \emph{Generic}, higher cost values generally lead to worse
-  results.
-\item
-  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
-  seems that smaller choices are ok.
-\item
-  \textbf{Validity}: \emph{ECCo} almost always valid except for very low
-  values during training and high values at evaluation time.
-  \emph{Generic} often has poor validity.
-\item
-  \textbf{Accuracy}: Seems largely unaffected.
-\end{itemize}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly
-Separable data by energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0.01 & \textit{ECCo} & $-9.91 \cdot 10^{11}$ & $2.25 \cdot 10^{12}$ \\
-  full & 0.01 & \textit{Generic} & $-5.71 \cdot 10^{17}$ & $1.3 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.116}} \\
-  full & 0.01 & \textit{REVISE} & -15.6 & 13.2 \\
-  vanilla & 0.01 & \textit{ECCo} & -4.28 & 3.52 \\
-  vanilla & 0.01 & \textit{Generic} & -4.45 & 3.47 \\
-  vanilla & 0.01 & \textit{Omniscient} & -5.12 & 4.46 \\
-  vanilla & 0.01 & \textit{REVISE} & -4.91 & 4.24 \\
-  full & 0.05 & \textit{ECCo} & $-5.63 \cdot 10^{5}$ & $1.28 \cdot 10^{6}$ \\
-  full & 0.05 & \textit{Generic} & $-8.35 \cdot 10^{17}$ & $1.9 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.114}} \\
-  full & 0.05 & \textit{REVISE} & -15 & 12.6 \\
-  vanilla & 0.05 & \textit{ECCo} & -4.4 & 3.66 \\
-  vanilla & 0.05 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 0.05 & \textit{Omniscient} & -5.25 & 4.62 \\
-  vanilla & 0.05 & \textit{REVISE} & -4.94 & 4.22 \\
-  full & 0.1 & \textit{ECCo} & $-6.74 \cdot 10^{5}$ & $1.53 \cdot 10^{6}$ \\
-  full & 0.1 & \textit{Generic} & $-1.72 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.56}} & \color{blue}{\textbf{0.124}} \\
-  full & 0.1 & \textit{REVISE} & -15.6 & 13.2 \\
-  vanilla & 0.1 & \textit{ECCo} & -4.28 & 3.52 \\
-  vanilla & 0.1 & \textit{Generic} & -4.45 & 3.48 \\
-  vanilla & 0.1 & \textit{Omniscient} & -5.12 & 4.46 \\
-  vanilla & 0.1 & \textit{REVISE} & -4.91 & 4.25 \\
-  full & 0.5 & \textit{ECCo} & -11.8 & 9.83 \\
-  full & 0.5 & \textit{Generic} & $-1.06 \cdot 10^{18}$ & $2.42 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.123}} \\
-  full & 0.5 & \textit{REVISE} & -15 & 12.6 \\
-  vanilla & 0.5 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 0.5 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 0.5 & \textit{Omniscient} & -5.25 & 4.61 \\
-  vanilla & 0.5 & \textit{REVISE} & -4.95 & 4.22 \\
-  full & 1 & \textit{ECCo} & -11.5 & 11.1 \\
-  full & 1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.88 \cdot 10^{11}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.59}} & \color{blue}{\textbf{0.117}} \\
-  full & 1 & \textit{REVISE} & -15.7 & 13.3 \\
-  vanilla & 1 & \textit{ECCo} & -4.28 & 3.51 \\
-  vanilla & 1 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 1 & \textit{Omniscient} & -5.11 & 4.46 \\
-  vanilla & 1 & \textit{REVISE} & -4.91 & 4.25 \\
-  full & 5 & \textit{ECCo} & -3.99 & 3.12 \\
-  full & 5 & \textit{Generic} & $-4.88 \cdot 10^{17}$ & $1.11 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.117}} \\
-  full & 5 & \textit{REVISE} & -14.6 & 12.1 \\
-  vanilla & 5 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 5 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 5 & \textit{Omniscient} & -5.25 & 4.61 \\
-  vanilla & 5 & \textit{REVISE} & -4.95 & 4.22 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.31}} & \color{blue}{\textbf{0.735}} \\
-  full & 10 & \textit{Generic} & $-1.7 \cdot 10^{11}$ & $3.86 \cdot 10^{11}$ \\
-  full & 10 & \textit{Omniscient} & -2.53 & 0.117 \\
-  full & 10 & \textit{REVISE} & -15.5 & 13 \\
-  vanilla & 10 & \textit{ECCo} & -4.28 & 3.51 \\
-  vanilla & 10 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 10 & \textit{Omniscient} & -5.12 & 4.46 \\
-  vanilla & 10 & \textit{REVISE} & -4.91 & 4.24 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.01}} & \color{blue}{\textbf{0.488}} \\
-  full & 15 & \textit{Generic} & $-4.91 \cdot 10^{17}$ & $1.12 \cdot 10^{18}$ \\
-  full & 15 & \textit{Omniscient} & -2.53 & 0.116 \\
-  full & 15 & \textit{REVISE} & -14.4 & 11.7 \\
-  vanilla & 15 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 15 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 15 & \textit{Omniscient} & -5.25 & 4.6 \\
-  vanilla & 15 & \textit{REVISE} & -4.95 & 4.23 \\\bottomrule
-
-\end{longtable}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-lin_sep-lambda_cost_exper}Results for Linearly
-Separable data by cost penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0 & \textit{ECCo} & $-5.32 \cdot 10^{3}$ & $1.21 \cdot 10^{4}$ \\
-  full & 0 & \textit{Generic} & $-1.03 \cdot 10^{18}$ & $2.34 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.64}} & \color{blue}{\textbf{0.125}} \\
-  full & 0 & \textit{REVISE} & -15.4 & 12.9 \\
-  vanilla & 0 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0 & \textit{Omniscient} & -5.18 & 4.54 \\
-  vanilla & 0 & \textit{REVISE} & -4.93 & 4.23 \\
-  full & 0.001 & \textit{ECCo} & -362 & 811 \\
-  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.03 \cdot 10^{17}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.49}} & \color{blue}{\textbf{0.115}} \\
-  full & 0.001 & \textit{REVISE} & -15.5 & 13 \\
-  vanilla & 0.001 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0.001 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0.001 & \textit{Omniscient} & -5.18 & 4.53 \\
-  vanilla & 0.001 & \textit{REVISE} & -4.93 & 4.23 \\
-  full & 0.1 & \textit{ECCo} & $-3.72 \cdot 10^{11}$ & $8.46 \cdot 10^{11}$ \\
-  full & 0.1 & \textit{Generic} & $-4.49 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5}} & \color{blue}{\textbf{0.112}} \\
-  full & 0.1 & \textit{REVISE} & -14.6 & 12.2 \\
-  vanilla & 0.1 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0.1 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0.1 & \textit{Omniscient} & -5.18 & 4.54 \\
-  vanilla & 0.1 & \textit{REVISE} & -4.93 & 4.24 \\\bottomrule
-
-\end{longtable}
-
-\paragraph{Moons}\label{moons}
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-moons-lambda_energy_exper}):
-  \emph{ECCo} consistently yields better results than \emph{Vanilla},
-  except for very low choices of the energy penalty during training for
-  which it performs abismal. \emph{Generic} performs quite badly across
-  the board for high enough choices of the energy penalty at evaluation
-  time. \emph{Omni} has small positive effect. \emph{REVISE} performs
-  poorly across the board.
-\item
-  \textbf{Cost (distance penalty)}: \emph{Generic} generally does better
-  for higher values, while \emph{ECCo} does better for lower values.
-\item
-  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
-  seems that smaller choices are ok.
-\item
-  \textbf{Validity}: \emph{ECCo} generally achieves full validity except
-  for very low choices the energy penalty during training and high
-  choices at evaluation time. \emph{Generic} performs poorly for high
-  choices of the energy penalty during evaluation.
-\item
-  \textbf{Accuracy}: Largely unaffected although \emph{ECCo} suffers a
-  bit for very low choices the energy penalty during training.
-  \emph{REVISE} suffers a lot in general (around 10 percentage points).
-\end{itemize}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-moons-lambda_energy_exper}Results for Moons data by
-energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0.01 & \textit{ECCo} & $-2.8 \cdot 10^{22}$ & $6.39 \cdot 10^{22}$ \\
-  full & 0.01 & \textit{Generic} & $-4.89 \cdot 10^{30}$ & $1.11 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.74}} & \color{blue}{\textbf{5.08}} \\
-  full & 0.01 & \textit{REVISE} & -572 & $1.25 \cdot 10^{3}$ \\
-  vanilla & 0.01 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 0.01 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.01 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 0.01 & \textit{REVISE} & -11.2 & 13 \\
-  full & 0.05 & \textit{ECCo} & $-1.55 \cdot 10^{16}$ & $3.52 \cdot 10^{16}$ \\
-  full & 0.05 & \textit{Generic} & $-2.22 \cdot 10^{20}$ & $5 \cdot 10^{20}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.41}} & \color{blue}{\textbf{4.48}} \\
-  full & 0.05 & \textit{REVISE} & $-1.04 \cdot 10^{3}$ & $2.3 \cdot 10^{3}$ \\
-  vanilla & 0.05 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.05 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 0.05 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 0.05 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 0.1 & \textit{ECCo} & $-3.41 \cdot 10^{3}$ & $7.73 \cdot 10^{3}$ \\
-  full & 0.1 & \textit{Generic} & $-5.22 \cdot 10^{30}$ & $1.19 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.78}} & \color{blue}{\textbf{5.12}} \\
-  full & 0.1 & \textit{REVISE} & -288 & 594 \\
-  vanilla & 0.1 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.1 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 0.1 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 0.5 & \textit{ECCo} & -7.09 & 7.51 \\
-  full & 0.5 & \textit{Generic} & $-1.11 \cdot 10^{31}$ & $2.53 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.58}} & \color{blue}{\textbf{4.83}} \\
-  full & 0.5 & \textit{REVISE} & $-1.19 \cdot 10^{3}$ & $2.64 \cdot 10^{3}$ \\
-  vanilla & 0.5 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.5 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 0.5 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 0.5 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 1 & \textit{ECCo} & -6.06 & 6.33 \\
-  full & 1 & \textit{Generic} & $-1.58 \cdot 10^{33}$ & $3.59 \cdot 10^{33}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.66}} & \color{blue}{\textbf{4.89}} \\
-  full & 1 & \textit{REVISE} & $-1.16 \cdot 10^{3}$ & $2.59 \cdot 10^{3}$ \\
-  vanilla & 1 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 1 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 1 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.57}} & \color{blue}{\textbf{2.07}} \\
-  full & 5 & \textit{Generic} & $-1.17 \cdot 10^{28}$ & $2.66 \cdot 10^{28}$ \\
-  full & 5 & \textit{Omniscient} & -4.29 & 4.31 \\
-  full & 5 & \textit{REVISE} & -530 & $1.16 \cdot 10^{3}$ \\
-  vanilla & 5 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 5 & \textit{Generic} & -11.7 & 12.7 \\
-  vanilla & 5 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 5 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.76}} & \color{blue}{\textbf{0.974}} \\
-  full & 10 & \textit{Generic} & $-1.54 \cdot 10^{33}$ & $3.51 \cdot 10^{33}$ \\
-  full & 10 & \textit{Omniscient} & -4.44 & 4.56 \\
-  full & 10 & \textit{REVISE} & $-1.52 \cdot 10^{3}$ & $3.4 \cdot 10^{3}$ \\
-  vanilla & 10 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 10 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 10 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 10 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.37}} & \color{blue}{\textbf{0.365}} \\
-  full & 15 & \textit{Generic} & $-5.32 \cdot 10^{28}$ & $1.21 \cdot 10^{29}$ \\
-  full & 15 & \textit{Omniscient} & -4.34 & 4.38 \\
-  full & 15 & \textit{REVISE} & -473 & $1.03 \cdot 10^{3}$ \\
-  vanilla & 15 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 15 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 15 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 15 & \textit{REVISE} & -11.3 & 13.1 \\\bottomrule
-
-\end{longtable}
-
-\paragraph{Circles}\label{circles}
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-circles-lambda_energy_exper}):
-  \emph{ECCo} consistently yields better results than \emph{Vanilla},
-  though primarily for low to medium choices of the energy penalty
-  (\textless=5) during training. The same goes for \emph{Generic}, which
-  sometimes outperforms \emph{ECCo} (for small energy penalty at
-  evaluation time). \emph{Omni} does alright for lower energy penalty at
-  evaluation time, but loses out for higher choices. \emph{REVISE}
-  performs poorly across the board (except very low choices at
-  evaluation time).
-\item
-  \textbf{Cost (distance penalty)}: \emph{ECCo} and \emph{Generic}
-  generally achieve the best results when no cost penalty is used during
-  training. Both \emph{Omni} and \emph{REVISE} are largely unaffected.
-\item
-  \textbf{Maximum Iterations}: \emph{ECCo} consistently yields better
-  results for higher numbers of iterations. \emph{Generic} generally
-  does best for a medium number (50). \emph{Omni} is sometimes invalid
-  (\textbf{???}).
-\item
-  \textbf{Validity}: \emph{ECCo} tends to outperform its \emph{Vanilla}
-  counterpart, though primarily for low to medium choices of the energy
-  penalty (\textless=5) during training and evaluation. \emph{Vanilla}
-  typically worse across the board.
-\item
-  \textbf{Accuracy}: Mostly unaffected, but \emph{REVISE} again
-  consistently some deterioration and \emph{ECCo} deteriorates for high
-  choices of energy penalty during training, reflecting other outcomes
-  above.
-\end{itemize}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-circles-lambda_energy_exper}Results for Circles data
-by energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.26}} & \color{blue}{\textbf{0.423}} \\
-  full & 0.01 & \textit{Generic} & -1.49 & 0.71 \\
-  full & 0.01 & \textit{Omniscient} & -5.21 & 5.25 \\
-  full & 0.01 & \textit{REVISE} & $-2.71 \cdot 10^{26}$ & $6.37 \cdot 10^{26}$ \\
-  vanilla & 0.01 & \textit{ECCo} & -9.33 & 7.34 \\
-  vanilla & 0.01 & \textit{Generic} & -8.89 & 6.88 \\
-  vanilla & 0.01 & \textit{Omniscient} & -8.67 & 6.87 \\
-  vanilla & 0.01 & \textit{REVISE} & -8.65 & 6.8 \\
-  full & 0.05 & \textit{ECCo} & -1.29 & 0.397 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.356}} \\
-  full & 0.05 & \textit{Omniscient} & -5.08 & 5.09 \\
-  full & 0.05 & \textit{REVISE} & $-5.91 \cdot 10^{27}$ & $1.36 \cdot 10^{28}$ \\
-  vanilla & 0.05 & \textit{ECCo} & -9.35 & 7.32 \\
-  vanilla & 0.05 & \textit{Generic} & -8.85 & 6.87 \\
-  vanilla & 0.05 & \textit{Omniscient} & -8.7 & 6.96 \\
-  vanilla & 0.05 & \textit{REVISE} & -8.52 & 6.76 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.2}} & \color{blue}{\textbf{0.383}} \\
-  full & 0.1 & \textit{Generic} & -1.5 & 0.735 \\
-  full & 0.1 & \textit{Omniscient} & -5.17 & 5.23 \\
-  full & 0.1 & \textit{REVISE} & $-3.06 \cdot 10^{26}$ & $7.7 \cdot 10^{26}$ \\
-  vanilla & 0.1 & \textit{ECCo} & -9.33 & 7.32 \\
-  vanilla & 0.1 & \textit{Generic} & -8.88 & 6.86 \\
-  vanilla & 0.1 & \textit{Omniscient} & -8.69 & 6.9 \\
-  vanilla & 0.1 & \textit{REVISE} & -8.68 & 6.81 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.12}} & \color{blue}{\textbf{0.217}} \\
-  full & 0.5 & \textit{Generic} & -1.21 & 0.352 \\
-  full & 0.5 & \textit{Omniscient} & -5.09 & 5.12 \\
-  full & 0.5 & \textit{REVISE} & $-5.97 \cdot 10^{27}$ & $1.37 \cdot 10^{28}$ \\
-  vanilla & 0.5 & \textit{ECCo} & -9.35 & 7.3 \\
-  vanilla & 0.5 & \textit{Generic} & -8.89 & 6.92 \\
-  vanilla & 0.5 & \textit{Omniscient} & -8.68 & 6.93 \\
-  vanilla & 0.5 & \textit{REVISE} & -8.53 & 6.75 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.1}} & \color{blue}{\textbf{0.163}} \\
-  full & 1 & \textit{Generic} & -1.49 & 0.726 \\
-  full & 1 & \textit{Omniscient} & -5.16 & 5.2 \\
-  full & 1 & \textit{REVISE} & $-3.09 \cdot 10^{26}$ & $7.22 \cdot 10^{26}$ \\
-  vanilla & 1 & \textit{ECCo} & -9.34 & 7.36 \\
-  vanilla & 1 & \textit{Generic} & -8.86 & 6.85 \\
-  vanilla & 1 & \textit{Omniscient} & -8.7 & 6.9 \\
-  vanilla & 1 & \textit{REVISE} & -8.69 & 6.85 \\
-  full & 5 & \textit{ECCo} & -1.75 & 0.154 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.363}} \\
-  full & 5 & \textit{Omniscient} & -5.14 & 5.16 \\
-  full & 5 & \textit{REVISE} & $-1.1 \cdot 10^{28}$ & $2.5 \cdot 10^{28}$ \\
-  vanilla & 5 & \textit{ECCo} & -9.36 & 7.32 \\
-  vanilla & 5 & \textit{Generic} & -8.88 & 6.91 \\
-  vanilla & 5 & \textit{Omniscient} & -8.7 & 6.93 \\
-  vanilla & 5 & \textit{REVISE} & -8.52 & 6.73 \\
-  full & 10 & \textit{ECCo} & $-1.02 \cdot 10^{6}$ & $2.32 \cdot 10^{6}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.49}} & \color{blue}{\textbf{0.702}} \\
-  full & 10 & \textit{Omniscient} & -5.13 & 5.16 \\
-  full & 10 & \textit{REVISE} & $-3.74 \cdot 10^{26}$ & $9.09 \cdot 10^{26}$ \\
-  vanilla & 10 & \textit{ECCo} & -9.31 & 7.33 \\
-  vanilla & 10 & \textit{Generic} & -8.87 & 6.86 \\
-  vanilla & 10 & \textit{Omniscient} & -8.7 & 6.89 \\
-  vanilla & 10 & \textit{REVISE} & -8.69 & 6.83 \\
-  full & 15 & \textit{ECCo} & $-3.31 \cdot 10^{13}$ & $7.54 \cdot 10^{13}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.22}} & \color{blue}{\textbf{0.37}} \\
-  full & 15 & \textit{Omniscient} & -5.2 & 5.23 \\
-  full & 15 & \textit{REVISE} & $-9.01 \cdot 10^{27}$ & $2.06 \cdot 10^{28}$ \\
-  vanilla & 15 & \textit{ECCo} & -9.38 & 7.34 \\
-  vanilla & 15 & \textit{Generic} & -8.86 & 6.87 \\
-  vanilla & 15 & \textit{Omniscient} & -8.69 & 6.96 \\
-  vanilla & 15 & \textit{REVISE} & -8.51 & 6.73 \\\bottomrule
-
-\end{longtable}
-
-\end{document}

### 2025-02-03 paper/preprint/paper.tex
beautiful

@@
 models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI
-tools can aide us in monitoring models and providing affected
-individuals with recourse
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}).
+tools can aide us in monitoring models and providing recourse to
+affected individuals (\citeproc{ref-wachter2017counterfactual}{Wachter,
+Mittelstadt, and Russell 2017}).
 
@@
 
-In this work, we build on these findings and introduce a
-\textbf{counterfactual training}: a novel approach geared explicitly
-towards aligning model representations with desirable explanations.
-
-From this perspective, adversarial training induces models to
-``unlearn'' representations that are susceptible to the semantically
-most meaningless explanations---adversarial examples.
+Building on these findings, we introduce \textbf{counterfactual
+training}: a novel regularization technique geared explicitly towards
+aligning model representations with desirable explanations. Our
+contributions are as follows:
+
+\begin{itemize}
+\tightlist
+\item
+  We discuss existing related work on improving models and consolidate
+  it through the lens of counterfactual explanations
+  (Section~\ref{sec-lit}).
+\item
+  We present our proposed methodological framework that leverages
+  faithful counterfactual explanations during the training phase of
+  models to achieve the explainability objective
+  (Section~\ref{sec-method}).
+\item
+  Through extensive experiments we demonstrate the counterfactual
+  training improve model explainability while maintaining high
+  predictive performance. We run ablation studies and grid searches to
+  understand how the underlying model components and hyperparameters
+  affect outcomes. (Section~\ref{sec-experiments}).
+\end{itemize}
+
+Despite limitations of our approach discussed in
+Section~\ref{sec-discussion}, we conclude that counterfactual training
+provides a practical framework for researchers and practitioners
+interested in making opaque models more trustworthy
+Section~\ref{sec-conclusion}. We also believe that this work serves as
+an opportunity for XAI researchers to reevaluate the premise of
+improving XAI tools without improving models.
 
@@
 Training}\label{link-to-adversarial-training}
+
+From this perspective, adversarial training induces models to
+``unlearn'' representations that are susceptible to the semantically
+most meaningless explanations---adversarial examples.
 
@@
 
-\section{Counterfactual Training}\label{counterfactual-training}
+\section{Counterfactual Training}\label{sec-method}
 
@@
 Conference on Artificial Intelligence}, 38:10829--37. 10.
+
+\bibitem[\citeproctext]{ref-augustin2020adversarial}
+Augustin, Maximilian, Alexander Meinke, and Matthias Hein. 2020.
+{``Adversarial Robustness on in-and Out-Distribution Improves
+Explainability.''} In \emph{European Conference on Computer Vision},
+228--45. Springer.
 

### 2025-02-03 paper/sections/intro.qmd
beautiful

@@
 On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems involving opaque models [@oneil2016weapons].
-Effective XAI tools can aide us in monitoring models and providing affected individuals with recourse [@wachter2017counterfactual].
+Effective XAI tools can aide us in monitoring models and providing recourse to affected individuals [@wachter2017counterfactual].
 
@@
 
-Building on these findings, we introduce a **counterfactual training**: a novel approach geared explicitly towards aligning model representations with desirable explanations.
+Building on these findings, we introduce **counterfactual training**: a novel regularization technique geared explicitly towards aligning model representations with desirable explanations.
+Our contributions are as follows:
 
-From this perspective, adversarial training induces models to "unlearn" representations that are susceptible to the semantically most meaningless explanations---adversarial examples.
+- We discuss existing related work on improving models and consolidate it through the lens of counterfactual explanations (@sec-lit).
+- We present our proposed methodological framework that leverages faithful counterfactual explanations during the training phase of models to achieve the explainability objective (@sec-method).
+- Through extensive experiments we demonstrate the counterfactual training improve model explainability while maintaining high predictive performance.
+We run ablation studies and grid searches to understand how the underlying model components and hyperparameters affect outcomes.
+(@sec-experiments).
+
+Despite limitations of our approach discussed in @sec-discussion, we conclude that counterfactual training provides a practical framework for researchers and practitioners interested in making opaque models more trustworthy [@sec-conclusion].
+We also believe that this work serves as an opportunity for XAI researchers to reevaluate the premise of improving XAI tools without improving models.

### 2025-02-03 paper/sections/related.qmd
streamlined quarto bs
intro basically done
beautiful

--- 2025-01-23 paper/related.qmd
@@
 @sauer2021counterfactual generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as *texture*, *shape* and *background*.
-The demonstrate that using these generated counterfactuals during classifier training improves model robustness.
+They demonstrate that using these generated counterfactuals during classifier training improves model robustness.
 Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
@@
 ## Link to Adversarial Training
+
+From this perspective, adversarial training induces models to "unlearn" representations that are susceptible to the semantically most meaningless explanations---adversarial examples.
 

### 2025-02-03 paper/sections/intro.qmd
slowly slowly

@@
 To our surprise, the second challenge has not yet attracted any consolidated research effort.
-Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are deemed desirable by humans.
+Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are interpretable and deemed desirable by humans.
 Instead, the choice has typically been to improve the capacity of XAI tools to identify the subset explanations that are both desirable and valid for any given model, independent of whether the learned representations are also compatible with undesirable explanations [@altmeyer2024faithful].

### 2025-02-03 paper/sections/related.qmd
slowly slowly

@@
 # Related Literature {#sec-lit}
+
+To the best of our knowledge, our proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability.
+In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed desirable with respect to the underlying data and stakeholder requirements.
+To make this more concrete, we follow @augustin2020adversarial in tieing the concept of explainability to the quality of counterfactual explanations that we can generate for a given model.
+The authors show that counterfactual explanations---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples.
+This finding is intuitive when we look at adversarial training (AT) as a means of regularization in the context of representation learning with high degrees of freedom: it reduces the space of viable solutions by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e.
+adversarial examples).
+@szegedy2013intriguing
+
+The link between counterfactual explanations and adversarial examples has been studied elsewhere: @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.
+
+::: {#def-explainability}
+
+## Model Explainability
+
+:::
+
+@altmeyer2024faithful find that
 
@@
 
-From this perspective, adversarial training induces models to "unlearn" representations that are susceptible to the semantically most meaningless explanations---adversarial examples.
-
 @freiesleben2022intriguing propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE (***this might be a useful notion for use to distinguish between adversarials and explanations during training***); secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations.
-@pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.
 

### 2025-02-03 paper/body_ecml.tex
progress baby

@@
 \begin{abstract}
-Counterfactual Explanations (CE) have emerged as a popular tool to
-explain predictions made by opaque machine learning models: they explain
-how factual inputs need to change in order for some fitted model to
-produce some desired output. Much existing research has focused on
-identifying explanations that are not only valid but also deemed
-desirable with respect to the underlying data and stakeholder
-requirements. Recent work has shown that under this premise, the task of
-learning desirable explanations is effectively reassigned from the model
-itself to the (post-hoc) counterfactual explainer. Building on that
-work, we propose a novel model objective that leverages counterfactuals
-during the training phase (ad-hoc) in order to minimize the divergence
-between learned representations and desirable explanations. Through
-extensive experiments, we demonstrate that our proposed methodology
-facilitates training models that inherently deliver desirable
-explanations while maintaining high predictive performance.
+Counterfactual Explanations have emerged as a popular tool to explain
+predictions made by opaque machine learning models: they explain how
+factual inputs need to change in order for some fitted model to produce
+some desired output. Much existing research has focused on identifying
+explanations that are not only valid but also deemed desirable with
+respect to the underlying data and stakeholder requirements. Recent work
+has shown that under this premise, the task of learning desirable
+explanations is effectively reassigned from the model itself to the
+(post-hoc) counterfactual explainer. Building on that work, we propose a
+novel model objective that leverages counterfactuals during the training
+phase (ad-hoc) in order to minimize the divergence between learned
+representations and desirable explanations. Through extensive
+experiments, we demonstrate that our proposed methodology facilitates
+training models that inherently deliver desirable explanations while
+maintaining high predictive performance.
 \keywords{Counterfactual Explanations \and Explainable
@@
 serious scrutiny in recent years. At the very cusp of the deep learning
-revolution, \cite{goodfellow2014explaining} showed that artificial
-neural networks (ANN) are sensitive to adversarial examples (AE):
-counterfactuals of model inputs that yield vastly different model
-predictions despite being semantically indifferent from their factual
-counterparts. Despite partially effective mitigation strategies such as
-\textbf{adversarial training}, truly robust deep learning (DL) remains
-unattainable even for models that are considered shallow by today's
-standards \cite{kolter2023keynote}.
+revolution, \cite{szegedy2013intriguing} showed that artificial neural
+networks (ANN) are sensitive to adversarial examples: counterfactuals of
+model inputs that yield vastly different model predictions despite being
+``imperceptible'' in that they are semantically indifferent from their
+factual counterparts. Despite partially effective mitigation strategies
+such as \textbf{adversarial training} \cite{goodfellow2014explaining},
+truly robust deep learning (DL) remains unattainable even for models
+that are considered shallow by today's standards
+\cite{kolter2023keynote}.
 
@@
 in practice and caused real harm to people who were affected adversely
-and often unfairly by automated decision-making systems involving opaque
-models \cite{oneil2016weapons}. Effective XAI tools can aide us in
-monitoring models and providing affected individuals with recourse
-\cite{wachter2017counterfactual}.
+and often unfairly by automated decision-making systems (ADMS) involving
+opaque models \cite{oneil2016weapons}. Effective XAI tools can aide us
+in monitoring models and providing recourse to individuals to turn
+adverse outcomes (e.g.~``loan application rejected'') into positive ones
+(``application accepted''). \cite{wachter2017counterfactual} propose
+\textbf{counterfactual explanations} as an effective approach to achieve
+this: they explain how factual inputs need to change in order for some
+fitted model to produce some desired output, typically involving minimal
+perturbations.
 
@@
 here as the degree to which learned representations correspond to
-explanations that are deemed desirable by humans. Instead, the choice
-has typically been to improve the capacity of XAI tools to identify the
-subset explanations that are both desirable and valid for any given
-model, independent of whether the learned representations are also
-compatible with undesirable explanations \cite{altmeyer2024faithful}.
-Fortunately, recent findings indicate that explainability can arise as
-byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization and generative capacity
-\cite{schut2021generating}.
+explanations that are interpretable and deemed desirable by humans.
+Instead, the choice has typically been to improve the capacity of XAI
+tools to identify the subset explanations that are both desirable and
+valid for any given model, independent of whether the learned
+representations are also compatible with undesirable explanations
+\cite{altmeyer2024faithful}. Fortunately, recent findings indicate that
+explainability can arise as byproduct of regularization techniques aimed
+at other objectives such as robustness, generalization and generative
+capacity \cite{schut2021generating}.
 
@@
 
-\subsection{Background on Counterfactual
-Explanations}\label{background-on-counterfactual-explanations}
+To the best of our knowledge, our proposed framework for counterfactual
+training represents the first attempt to use counterfactual explanations
+during training to improve model explainability. In high-level terms, we
+define model explainability as the extent to which valid explanations
+derived for an opaque model are also deemed desirable with respect to
+the underlying data and stakeholder requirements. To make this more
+concrete, we follow \cite{augustin2020adversarial} in tieing the concept
+of explainability to the quality of counterfactual explanations that we
+can generate for a given model. The authors show that counterfactual
+explanations---understood here as minimal input perturbations that yield
+some desired model prediction---are generally more meaningful if the
+underlying model is more robust to adversarial examples. We can make
+intuitive sense of this finding when looking at adversarial training
+(AT) through the lens of representation learning with high degrees of
+freedom: by inducing models to ``unlearn'' representations that are
+susceptible to worst-case counterfactuals (i.e.~adversarial examples),
+AT effectively removes some undesirable explanations from the solution
+space.
 
-\cite{wachter2017counterfactual, joshi2019realistic, altmeyer2024faithful}
+\subsection{Adversarial Examples are Counterfactual
+Explanations}\label{adversarial-examples-are-counterfactual-explanations}
 
-\subsection{Learning Representations}\label{learning-representations}
+This interpretation of the link between explainability through
+counterfactuals on one side, and robustness to adversarial examples on
+the other, is backed by empirical evidence.
+\cite{sauer2021counterfactual} demonstrate that using counterfactual
+images during classifier training improves model robustness. Similarly,
+\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
+potentially useful training data in machine learning, especially in
+supervised settings where inputs may be reasonably mapped to multiple
+outputs. They, too, demonstrate the augmenting the training data of
+image classifiers can improve generalization. \cite{teney2020learning}
+propose an approach using counterfactuals in training that does not rely
+on data augmentation: they argue that counterfactual pairs typically
+already exist in training datasets. Specifically, their approach relies
+on, firstly, identifying similar input samples with different
+annotations and, secondly, ensuring that the gradient of the classifier
+aligns with the vector between pairs of counterfactual inputs using the
+cosine distance as a loss function. In the natural language processing
+(NLP) domain, counterfactuals have similarly been used to improve models
+through data augmentation: \cite{wu2021polyjuice}, propose
+\emph{POLYJUICE}, a general-purpose counterfactual generator for
+language models. They demonstrate empirically that augmenting training
+data through \emph{POLYJUICE} counterfactuals improves robustness in a
+number of NLP tasks. \cite{luu2023counterfactual} introduce
+Counterfactual Adversarial Training (CAT), which also aims at improving
+generalization and robustness of language models. Specifically, they
+propose to proceed as follows: firstly, they identify training samples
+that are subject to high predictive uncertainty; secondly, they generate
+counterfactual explanations for those samples; and, finally, they
+fine-tune the given language model on the augmented dataset that
+includes the generated counterfactuals.
 
-\begin{quote}
-For example, joint-energy models
-\end{quote}
+There have also been several attempts at formalizing the relationship
+between counterfactual explanations (CE) and adversarial examples (AE).
+Pointing to clear similarities in how CE and AE are generated,
+\cite{freiesleben2022intriguing} makes the case for jointly studying the
+opaqueness and robustness problem in representation learning. Formally,
+AE can be seen as the subset of CE, for which misclassification is
+achieved \cite{freiesleben2022intriguing}. Similarly,
+\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
+certain conditions and derive theoretical upper bounds on the distances
+between them.
 
-\subsection{Generalization and
-Robustness}\label{generalization-and-robustness}
+Two recent works are closely related to ours in that they use
+counterfactuals during training with the explicit goal of affecting
+certain properties of post-hoc counterfactual explanations. Firstly,
+\cite{ross2021learning} propose a way to train models that are
+guaranteed to provide recourse for individuals to move from an adverse
+outcome to some positive target class with high probability. The
+approach proposed by \cite{ross2021learning} builds on adversarial
+training, where in this context susceptibility to targeted adversarial
+examples for the positive class is explicitly induced. The proposed
+method allows for imposing a set of actionability constraints ex-ante:
+for example, users can specify that certain features (e.g.~\emph{age},
+\emph{gender}, \ldots) are immutable. Secondly, \cite{guo2023counternet}
+are the first to propose an end-to-end training pipeline that includes
+counterfactual explanations as part of the training procedure. In
+particular, they propose a specific network architecture that includes a
+predictor and CE generator network, where the parameters of the CE
+generator network are learnable. Counterfactuals are generated during
+each training iteration and fed back to the predictor network. In
+contrast to \cite{guo2023counternet}, we impose no restrictions on the
+neural network architecture at all.
 
-\cite{sauer2021counterfactual} generate counterfactual images for MNIST
-and ImageNet through independent mechanisms (IM): each IM learns
-class-conditional input distributions over a specific lower-dimensional,
-semantically meaningful factor, such as \emph{texture}, \emph{shape} and
-\emph{background}. They demonstrate that using these generated
-counterfactuals during classifier training improves model robustness.
-Similarly, \cite{abbasnejad2020counterfactual} argue that
-counterfactuals represent potentially useful training data in machine
-learning, especially in supervised settings where inputs may be
-reasonably mapped to multiple outputs. They, too, demonstrate the
-augmenting the training data of image classifiers can improve
-generalization.
+\subsection{Beyond Robustness}\label{beyond-robustness}
 
-\cite{teney2020learning} propose an approach using counterfactuals in
-training that does not rely on data augmentation: they argue that
-counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on, firstly, identifying similar
-input samples with different annotations and, secondly, ensuring that
-the gradient of the classifier aligns with the vector between pairs of
-counterfactual inputs using the cosine distance as a loss function
-(referred to as \emph{gradient supervision}) (\textbf{\emph{this might
-be useful for our task as well}}). In the natural language processing
-(NLP) domain, counterfactuals have similarly been used to improve models
-through data augmentation: \cite{wu2021polyjuice}, propose POLYJUICE, a
-general-purpose counterfactual generator for language models. They
-demonstrate empirically that augmenting training data through POLYJUICE
-counterfactuals improves robustness in a number of NLP tasks.
+Improving the adversarial robustness of models is not the only path
+towards aligning representations with desirable explanations. In a work
+closely related to this one, \cite{altmeyer2024faithful} show that
+explainability can be improved through model averaging and refined model
+objectives. The authors propose a way to generate counterfactuals that
+are maximally \textbf{faithful} to the model in that they are consistent
+with what the model has learned about the underlying data. Formally,
+they rely on tools from energy-based modelling to minimize the
+divergence between the distribution of counterfactuals and the
+conditional posterior over inputs learned by the model. Their proposed
+counterfactual explainer, \emph{ECCCo}, yields desirable (or
+\textbf{plausible}) explanations if and only if the underlying model has
+learned representations that align with them. They find that both deep
+ensembles \cite{lakshminarayanan2016simple} and joint energy-based
+models (JEMs) \cite{grathwohl2020your} tend to do well in this regard.
 
-\subsection{Link to Adversarial
-Training}\label{link-to-adversarial-training}
-
-From this perspective, adversarial training induces models to
-``unlearn'' representations that are susceptible to the semantically
-most meaningless explanations---adversarial examples.
-
-\cite{freiesleben2022intriguing} propose two definitional differences
-between Adversarial Examples (AE) and Counterfactual Explanations (CE):
-firstly, and more importantly according to the authors, the term AE
-implies missclassification, which is not the case for CE
-(\textbf{\emph{this might be a useful notion for use to distinguish
-between adversarials and explanations during training}}); secondly, they
-argue that closeness plays a more critical role in the context of CE but
-confess that even counterfactuals that are not close might be relevant
-explanations. \cite{pawelczyk2022exploring} show that CE and AE are
-equivalent under certain conditions and derive upper bounds on the
-distances between them.
-
-\subsection{Closely Related}\label{closely-related}
-
-\cite{guo2023counternet} are the first to propose end-to-end training
-pipeline that includes counterfactual explanations as part of the
-training prodeduce. In particular, they propose a specific network
-architecture that includes a predictor and CE generator network
-(\textbf{\emph{akin a GAN?}}), where the parameters of the CE generator
-network are learnable. Counterfactuals are generated during each
-training iteration and fed back to the predictor network
-(\textbf{\emph{here we are aligned}}). In contrast, we impose no
-restrictions on the neural network architecture at all.
-(\textbf{\emph{to ensure the one-hot encoding of categorical features is
-maintained, they simple use softmax (might be interesting for CE.jl)}})
-Interestingly, the authors find that their approach is sensitive to the
-choice of the loss function: only MSE seems to lead to good performance.
-They also demonstrate theoretically, that the objective function is
-difficult to optimize due to divergent gradients and suffers from poor
-adversarial robustness. (\textbf{\emph{because partial gradients with
-respect to the classification loss component and the counterfactual
-validity component point in opposite directions}}). To mitigate these
-issues, the authors use block-wise gradient descent: they first update
-with respect to classification loss and then use a second update with
-respect to the other loss components (\textbf{\emph{this might be useful
-for our task as well}}). \cite{ross2021learning} propose a way to train
-models that are guaranteed to provide recourse for individuals with high
-probability. The approach builds on adversarial training
-(\textbf{\emph{here we are aligned}}), where in this context adversarial
-examples are actively encouraged to exist, but only target attacks with
-respect to the positive class. The proposed method allows for imposing a
-set of actionable recourse ex-ante: for example, users can impose
-mutability constraints for features (\textbf{\emph{here we are
-aligned}}). (\textbf{\emph{To solve their objective function more
-efficiently, they use a first-order Taylor approximation to approximate
-the recourse loss component (might be applicable in our case)}})
-
-\cite{luu2023counterfactual} introduce Counterfactual Adversarial
-Training (CAT) with intention of improving generalization and robustness
-of language models. Specifically, they propose to proceed as follows:
-firstly, identify training samples that are subject to high predictive
-uncertainty (entropy); secondly, generate counterfactual explanations
-for those samples; and, finally, finetune the model on the augmented
-dataset that includes the generated counterfactuals.
+Once again it helps to look at these findings through the lens of
+representation learning with high degrees of freedom. Deep ensembles are
+approximate Bayesian model averages, which are most called for when
+models are underspecified by the available data \cite{wilson2020case}.
+Averaging across solutions mitigates the aforementioned risk of relying
+on a single locally optimal representations that corresponds to
+semantically meaningless explanations for the data. Previous work by
+\cite{schut2021generating} similarly found that generating desirable
+(``interpretable'') counterfactual explanations is almost trivial for
+deep ensembles that have also undergone adversarial training. The case
+for JEMs is even clearer: they involve a hybrid objective that induces
+both high predictive performance and generative capacity
+\cite{grathwohl2020your}. This is closely related to the idea of
+aligning models with desirable explanations and has inspired our
+proposed counterfactual training objective, as we explain in
+Section~\ref{sec-method}.
 
 \section{Counterfactual Training}\label{sec-method}
+
+\begin{definition}[Model
+Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}
+
+~
+
+\end{definition}
 

### 2025-02-03 paper/preprint/paper.tex
progress baby

@@
 \theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
 \newtheorem{exercise}{Parameters}[section]
@@
 \begin{abstract}
-Counterfactual Explanations (CE) have emerged as a popular tool to
-explain predictions made by opaque machine learning models: they explain
-how factual inputs need to change in order for some fitted model to
-produce some desired output. Much existing research has focused on
-identifying explanations that are not only valid but also deemed
-desirable with respect to the underlying data and stakeholder
-requirements. Recent work has shown that under this premise, the task of
-learning desirable explanations is effectively reassigned from the model
-itself to the (post-hoc) counterfactual explainer. Building on that
-work, we propose a novel model objective that leverages counterfactuals
-during the training phase (ad-hoc) in order to minimize the divergence
-between learned representations and desirable explanations. Through
-extensive experiments, we demonstrate that our proposed methodology
-facilitates training models that inherently deliver desirable
-explanations while maintaining high predictive performance.
+Counterfactual Explanations have emerged as a popular tool to explain
+predictions made by opaque machine learning models: they explain how
+factual inputs need to change in order for some fitted model to produce
+some desired output. Much existing research has focused on identifying
+explanations that are not only valid but also deemed desirable with
+respect to the underlying data and stakeholder requirements. Recent work
+has shown that under this premise, the task of learning desirable
+explanations is effectively reassigned from the model itself to the
+(post-hoc) counterfactual explainer. Building on that work, we propose a
+novel model objective that leverages counterfactuals during the training
+phase (ad-hoc) in order to minimize the divergence between learned
+representations and desirable explanations. Through extensive
+experiments, we demonstrate that our proposed methodology facilitates
+training models that inherently deliver desirable explanations while
+maintaining high predictive performance.
 \end{abstract}
@@
 serious scrutiny in recent years. At the very cusp of the deep learning
-revolution, I. J. Goodfellow, Shlens, and Szegedy
-(\citeproc{ref-goodfellow2014explaining}{2014}) showed that artificial
-neural networks (ANN) are sensitive to adversarial examples (AE):
-counterfactuals of model inputs that yield vastly different model
-predictions despite being semantically indifferent from their factual
-counterparts. Despite partially effective mitigation strategies such as
-\textbf{adversarial training}, truly robust deep learning (DL) remains
+revolution, Szegedy et al. (\citeproc{ref-szegedy2013intriguing}{2013})
+showed that artificial neural networks (ANN) are sensitive to
+adversarial examples: counterfactuals of model inputs that yield vastly
+different model predictions despite being ``imperceptible'' in that they
+are semantically indifferent from their factual counterparts. Despite
+partially effective mitigation strategies such as \textbf{adversarial
+training} (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow,
+Shlens, and Szegedy 2014}), truly robust deep learning (DL) remains
 unattainable even for models that are considered shallow by today's
@@
 in practice and caused real harm to people who were affected adversely
-and often unfairly by automated decision-making systems involving opaque
-models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI
-tools can aide us in monitoring models and providing recourse to
-affected individuals (\citeproc{ref-wachter2017counterfactual}{Wachter,
-Mittelstadt, and Russell 2017}).
+and often unfairly by automated decision-making systems (ADMS) involving
+opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective
+XAI tools can aide us in monitoring models and providing recourse to
+individuals to turn adverse outcomes (e.g.~``loan application
+rejected'') into positive ones (``application accepted''). Wachter,
+Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}) propose
+\textbf{counterfactual explanations} as an effective approach to achieve
+this: they explain how factual inputs need to change in order for some
+fitted model to produce some desired output, typically involving minimal
+perturbations.
 
@@
 here as the degree to which learned representations correspond to
-explanations that are deemed desirable by humans. Instead, the choice
-has typically been to improve the capacity of XAI tools to identify the
-subset explanations that are both desirable and valid for any given
-model, independent of whether the learned representations are also
-compatible with undesirable explanations
+explanations that are interpretable and deemed desirable by humans.
+Instead, the choice has typically been to improve the capacity of XAI
+tools to identify the subset explanations that are both desirable and
+valid for any given model, independent of whether the learned
+representations are also compatible with undesirable explanations
 (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
@@
 
-\subsection{Background on Counterfactual
-Explanations}\label{background-on-counterfactual-explanations}
-
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}; \citeproc{ref-joshi2019realistic}{Joshi et al. 2019};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024})
-
-\subsection{Learning Representations}\label{learning-representations}
-
-\begin{quote}
-For example, joint-energy models
-\end{quote}
-
-\subsection{Generalization and
-Robustness}\label{generalization-and-robustness}
-
-Sauer and Geiger (\citeproc{ref-sauer2021counterfactual}{2021}) generate
-counterfactual images for MNIST and ImageNet through independent
-mechanisms (IM): each IM learns class-conditional input distributions
-over a specific lower-dimensional, semantically meaningful factor, such
-as \emph{texture}, \emph{shape} and \emph{background}. They demonstrate
-that using these generated counterfactuals during classifier training
-improves model robustness. Similarly, Abbasnejad et al.
+To the best of our knowledge, our proposed framework for counterfactual
+training represents the first attempt to use counterfactual explanations
+during training to improve model explainability. In high-level terms, we
+define model explainability as the extent to which valid explanations
+derived for an opaque model are also deemed desirable with respect to
+the underlying data and stakeholder requirements. To make this more
+concrete, we follow Augustin, Meinke, and Hein
+(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
+explainability to the quality of counterfactual explanations that we can
+generate for a given model. The authors show that counterfactual
+explanations---understood here as minimal input perturbations that yield
+some desired model prediction---are generally more meaningful if the
+underlying model is more robust to adversarial examples. We can make
+intuitive sense of this finding when looking at adversarial training
+(AT) through the lens of representation learning with high degrees of
+freedom: by inducing models to ``unlearn'' representations that are
+susceptible to worst-case counterfactuals (i.e.~adversarial examples),
+AT effectively removes some undesirable explanations from the solution
+space.
+
+\subsection{Adversarial Examples are Counterfactual
+Explanations}\label{adversarial-examples-are-counterfactual-explanations}
+
+This interpretation of the link between explainability through
+counterfactuals on one side, and robustness to adversarial examples on
+the other, is backed by empirical evidence. Sauer and Geiger
+(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
+counterfactual images during classifier training improves model
+robustness. Similarly, Abbasnejad et al.
 (\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
@@
 augmenting the training data of image classifiers can improve
-generalization.
-
-Teney, Abbasnedjad, and Hengel (\citeproc{ref-teney2020learning}{2020})
-propose an approach using counterfactuals in training that does not rely
-on data augmentation: they argue that counterfactual pairs typically
-already exist in training datasets. Specifically, their approach relies
-on, firstly, identifying similar input samples with different
-annotations and, secondly, ensuring that the gradient of the classifier
-aligns with the vector between pairs of counterfactual inputs using the
-cosine distance as a loss function (referred to as \emph{gradient
-supervision}) (\textbf{\emph{this might be useful for our task as
-well}}). In the natural language processing (NLP) domain,
-counterfactuals have similarly been used to improve models through data
-augmentation: Wu et al. (\citeproc{ref-wu2021polyjuice}{2021}), propose
-POLYJUICE, a general-purpose counterfactual generator for language
-models. They demonstrate empirically that augmenting training data
-through POLYJUICE counterfactuals improves robustness in a number of NLP
-tasks.
-
-\subsection{Link to Adversarial
-Training}\label{link-to-adversarial-training}
-
-From this perspective, adversarial training induces models to
-``unlearn'' representations that are susceptible to the semantically
-most meaningless explanations---adversarial examples.
-
-Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) propose two
-definitional differences between Adversarial Examples (AE) and
-Counterfactual Explanations (CE): firstly, and more importantly
-according to the authors, the term AE implies missclassification, which
-is not the case for CE (\textbf{\emph{this might be a useful notion for
-use to distinguish between adversarials and explanations during
-training}}); secondly, they argue that closeness plays a more critical
-role in the context of CE but confess that even counterfactuals that are
-not close might be relevant explanations. Pawelczyk et al.
-(\citeproc{ref-pawelczyk2022exploring}{2022}) show that CE and AE are
-equivalent under certain conditions and derive upper bounds on the
-distances between them.
-
-\subsection{Closely Related}\label{closely-related}
-
-Guo, Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the
-first to propose end-to-end training pipeline that includes
-counterfactual explanations as part of the training prodeduce. In
-particular, they propose a specific network architecture that includes a
-predictor and CE generator network (\textbf{\emph{akin a GAN?}}), where
-the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed
-back to the predictor network (\textbf{\emph{here we are aligned}}). In
-contrast, we impose no restrictions on the neural network architecture
-at all. (\textbf{\emph{to ensure the one-hot encoding of categorical
-features is maintained, they simple use softmax (might be interesting
-for CE.jl)}}) Interestingly, the authors find that their approach is
-sensitive to the choice of the loss function: only MSE seems to lead to
-good performance. They also demonstrate theoretically, that the
-objective function is difficult to optimize due to divergent gradients
-and suffers from poor adversarial robustness. (\textbf{\emph{because
-partial gradients with respect to the classification loss component and
-the counterfactual validity component point in opposite directions}}).
-To mitigate these issues, the authors use block-wise gradient descent:
-they first update with respect to classification loss and then use a
-second update with respect to the other loss components
-(\textbf{\emph{this might be useful for our task as well}}). Ross,
-Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024}) propose a
-way to train models that are guaranteed to provide recourse for
-individuals with high probability. The approach builds on adversarial
-training (\textbf{\emph{here we are aligned}}), where in this context
-adversarial examples are actively encouraged to exist, but only target
-attacks with respect to the positive class. The proposed method allows
-for imposing a set of actionable recourse ex-ante: for example, users
-can impose mutability constraints for features (\textbf{\emph{here we
-are aligned}}). (\textbf{\emph{To solve their objective function more
-efficiently, they use a first-order Taylor approximation to approximate
-the recourse loss component (might be applicable in our case)}})
-
-Luu and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
-Counterfactual Adversarial Training (CAT) with intention of improving
-generalization and robustness of language models. Specifically, they
-propose to proceed as follows: firstly, identify training samples that
-are subject to high predictive uncertainty (entropy); secondly, generate
-counterfactual explanations for those samples; and, finally, finetune
-the model on the augmented dataset that includes the generated
-counterfactuals.
+generalization. Teney, Abbasnedjad, and Hengel
+(\citeproc{ref-teney2020learning}{2020}) propose an approach using
+counterfactuals in training that does not rely on data augmentation:
+they argue that counterfactual pairs typically already exist in training
+datasets. Specifically, their approach relies on, firstly, identifying
+similar input samples with different annotations and, secondly, ensuring
+that the gradient of the classifier aligns with the vector between pairs
+of counterfactual inputs using the cosine distance as a loss function.
+In the natural language processing (NLP) domain, counterfactuals have
+similarly been used to improve models through data augmentation: Wu et
+al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{POLYJUICE}, a
+general-purpose counterfactual generator for language models. They
+demonstrate empirically that augmenting training data through
+\emph{POLYJUICE} counterfactuals improves robustness in a number of NLP
+tasks. Luu and Inoue (\citeproc{ref-luu2023counterfactual}{2023})
+introduce Counterfactual Adversarial Training (CAT), which also aims at
+improving generalization and robustness of language models.
+Specifically, they propose to proceed as follows: firstly, they identify
+training samples that are subject to high predictive uncertainty;
+secondly, they generate counterfactual explanations for those samples;
+and, finally, they fine-tune the given language model on the augmented
+dataset that includes the generated counterfactuals.
+
+There have also been several attempts at formalizing the relationship
+between counterfactual explanations (CE) and adversarial examples (AE).
+Pointing to clear similarities in how CE and AE are generated,
+Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
+case for jointly studying the opaqueness and robustness problem in
+representation learning. Formally, AE can be seen as the subset of CE,
+for which misclassification is achieved
+(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
+Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
+CE and AE are equivalent under certain conditions and derive theoretical
+upper bounds on the distances between them.
+
+Two recent works are closely related to ours in that they use
+counterfactuals during training with the explicit goal of affecting
+certain properties of post-hoc counterfactual explanations. Firstly,
+Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
+propose a way to train models that are guaranteed to provide recourse
+for individuals to move from an adverse outcome to some positive target
+class with high probability. The approach proposed by Ross, Lakkaraju,
+and Bastani (\citeproc{ref-ross2021learning}{2024}) builds on
+adversarial training, where in this context susceptibility to targeted
+adversarial examples for the positive class is explicitly induced. The
+proposed method allows for imposing a set of actionability constraints
+ex-ante: for example, users can specify that certain features
+(e.g.~\emph{age}, \emph{gender}, \ldots) are immutable. Secondly, Guo,
+Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the first
+to propose an end-to-end training pipeline that includes counterfactual
+explanations as part of the training procedure. In particular, they
+propose a specific network architecture that includes a predictor and CE
+generator network, where the parameters of the CE generator network are
+learnable. Counterfactuals are generated during each training iteration
+and fed back to the predictor network. In contrast to Guo, Nguyen, and
+Yadav (\citeproc{ref-guo2023counternet}{2023}), we impose no
+restrictions on the neural network architecture at all.
+
+\subsection{Beyond Robustness}\label{beyond-robustness}
+
+Improving the adversarial robustness of models is not the only path
+towards aligning representations with desirable explanations. In a work
+closely related to this one, Altmeyer et al.
+(\citeproc{ref-altmeyer2024faithful}{2024}) show that explainability can
+be improved through model averaging and refined model objectives. The
+authors propose a way to generate counterfactuals that are maximally
+\textbf{faithful} to the model in that they are consistent with what the
+model has learned about the underlying data. Formally, they rely on
+tools from energy-based modelling to minimize the divergence between the
+distribution of counterfactuals and the conditional posterior over
+inputs learned by the model. Their proposed counterfactual explainer,
+\emph{ECCCo}, yields desirable (or \textbf{plausible}) explanations if
+and only if the underlying model has learned representations that align
+with them. They find that both deep ensembles
+(\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
+and Blundell 2017}) and joint energy-based models (JEMs)
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
+well in this regard.
+
+Once again it helps to look at these findings through the lens of
+representation learning with high degrees of freedom. Deep ensembles are
+approximate Bayesian model averages, which are most called for when
+models are underspecified by the available data
+(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
+mitigates the aforementioned risk of relying on a single locally optimal
+representations that corresponds to semantically meaningless
+explanations for the data. Previous work by Schut et al.
+(\citeproc{ref-schut2021generating}{2021}) similarly found that
+generating desirable (``interpretable'') counterfactual explanations is
+almost trivial for deep ensembles that have also undergone adversarial
+training. The case for JEMs is even clearer: they involve a hybrid
+objective that induces both high predictive performance and generative
+capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
+is closely related to the idea of aligning models with desirable
+explanations and has inspired our proposed counterfactual training
+objective, as we explain in Section~\ref{sec-method}.
 
 \section{Counterfactual Training}\label{sec-method}
+
+\begin{definition}[Model
+Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}
+
+~
+
+\end{definition}
 
@@
 
+\bibitem[\citeproctext]{ref-grathwohl2020your}
+Grathwohl, Will, Kuan-Chieh Wang, Joern-Henrik Jacobsen, David Duvenaud,
+Mohammad Norouzi, and Kevin Swersky. 2020. {``Your Classifier Is
+Secretly an Energy Based Model and You Should Treat It Like One.''} In
+\emph{International Conference on Learning Representations}.
+
 \bibitem[\citeproctext]{ref-guo2023counternet}
@@
 
-\bibitem[\citeproctext]{ref-joshi2019realistic}
-Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
-Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
-Actionable Explanations in Black-Box Decision Making Systems.''}
-\url{https://arxiv.org/abs/1907.09615}.
-
 \bibitem[\citeproctext]{ref-kolter2023keynote}
@@
 \url{https://doi.org/10.1109/SaTML54575.2023.00009}.
+
+\bibitem[\citeproctext]{ref-lakshminarayanan2016simple}
+Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2017.
+{``Simple and Scalable Predictive Uncertainty Estimation Using Deep
+Ensembles.''} \emph{Advances in Neural Information Processing Systems}
+30.
 
@@
 {Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.
+
+\bibitem[\citeproctext]{ref-szegedy2013intriguing}
+Szegedy, Christian, Wojciech Zaremba, Ilya Sutskever, Joan Bruna,
+Dumitru Erhan, Ian Goodfellow, and Rob Fergus. 2013. {``Intriguing
+Properties of Neural Networks.''} \url{https://arxiv.org/abs/1312.6199}.
 

### 2025-02-03 paper/sections/abstract.qmd
progress baby

@@
 
-Counterfactual Explanations (CE) have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output.
+Counterfactual Explanations have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output.
 Much existing research has focused on identifying explanations that are not only valid but also deemed desirable with respect to the underlying data and stakeholder requirements.

### 2025-02-03 paper/sections/intro.qmd
progress baby

@@
 This trend towards complexity for the sake of performance has come under serious scrutiny in recent years.
-At the very cusp of the deep learning revolution, @goodfellow2014explaining showed that artificial neural networks (ANN) are sensitive to adversarial examples (AE): counterfactuals of model inputs that yield vastly different model predictions despite being semantically indifferent from their factual counterparts.
-Despite partially effective mitigation strategies such as **adversarial training**, truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].
+At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: counterfactuals of model inputs that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts.
+Despite partially effective mitigation strategies such as **adversarial training** [@goodfellow2014explaining], truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].
 
@@
 This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions.
-On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems involving opaque models [@oneil2016weapons].
-Effective XAI tools can aide us in monitoring models and providing recourse to affected individuals [@wachter2017counterfactual].
+On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems (ADMS) involving opaque models [@oneil2016weapons].
+Effective XAI tools can aide us in monitoring models and providing recourse to individuals to turn adverse outcomes (e.g.
+"loan application rejected") into positive ones ("application accepted").
+@wachter2017counterfactual propose **counterfactual explanations** as an effective approach to achieve this: they explain how factual inputs need to change in order for some fitted model to produce some desired output, typically involving minimal perturbations.
 

### 2025-02-03 paper/sections/method.qmd
progress baby

--- 2025-01-23 paper/related.qmd
@@
-# Related Literature {#sec-lit}
+# Counterfactual Training {#sec-method}
 
-## Background on Counterfactual Explanations
+::: {#def-explainability}
 
-[@wachter2017counterfactual;@joshi2019realistic;@altmeyer2024faithful]
+## Model Explainability
 
-## Learning Representations
-
-> For example, joint-energy models
-
-## Generalization and Robustness
-
-@sauer2021counterfactual generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as *texture*, *shape* and *background*.
-The demonstrate that using these generated counterfactuals during classifier training improves model robustness.
-Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
-They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.
-
-@teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as *gradient supervision*) (***this might be useful for our task as well***).
-In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose POLYJUICE, a general-purpose counterfactual generator for language models.
-They demonstrate empirically that augmenting training data through POLYJUICE counterfactuals improves robustness in a number of NLP tasks.
-
-## Link to Adversarial Training
-
-@freiesleben2022intriguing propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE (***this might be a useful notion for use to distinguish between adversarials and explanations during training***); secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations.
-@pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.
-
-## Closely Related
-
-@guo2023counternet are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce.
-In particular, they propose a specific network architecture that includes a predictor and CE generator network (***akin a GAN?***), where the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed back to the predictor network (***here we are aligned***).
-In contrast, we impose no restrictions on the neural network architecture at all.
-(***to ensure the one-hot encoding of categorical features is maintained, they simple use softmax (might be interesting for CE.jl)***) Interestingly, the authors find that their approach is sensitive to the choice of the loss function: only MSE seems to lead to good performance.
-They also demonstrate theoretically, that the objective function is difficult to optimize due to divergent gradients and suffers from poor adversarial robustness.
-(***because partial gradients with respect to the classification loss component and the counterfactual validity component point in opposite directions***).
-To mitigate these issues, the authors use block-wise gradient descent: they first update with respect to classification loss and then use a second update with respect to the other loss components (***this might be useful for our task as well***).
-@ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals with high probability.
-The approach builds on adversarial training (***here we are aligned***), where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class.
-The proposed method allows for imposing a set of actionable recourse ex-ante: for example, users can impose mutability constraints for features (***here we are aligned***).
-(***To solve their objective function more efficiently, they use a first-order Taylor approximation to approximate the recourse loss component (might be applicable in our case)***)
-
-@luu2023counterfactual introduce Counterfactual Adversarial Training (CAT) with intention of improving generalization and robustness of language models.
-Specifically, they propose to proceed as follows: firstly, identify training samples that are subject to high predictive uncertainty (entropy); secondly, generate counterfactual explanations for those samples; and, finally, finetune the model on the augmented dataset that includes the generated counterfactuals.
+:::

### 2025-02-03 paper/sections/related.qmd
progress baby

@@
 The authors show that counterfactual explanations---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples.
-This finding is intuitive when we look at adversarial training (AT) as a means of regularization in the context of representation learning with high degrees of freedom: it reduces the space of viable solutions by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e.
-adversarial examples).
-@szegedy2013intriguing
+We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom: by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e.
+adversarial examples), AT effectively removes some undesirable explanations from the solution space.
 
-The link between counterfactual explanations and adversarial examples has been studied elsewhere: @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.
+## Adversarial Examples are Counterfactual Explanations
 
-::: {#def-explainability}
-
-## Model Explainability
-
-:::
-
-@altmeyer2024faithful find that
-
-## Background on Counterfactual Explanations
-
-[@wachter2017counterfactual;@joshi2019realistic;@altmeyer2024faithful]
-
-## Learning Representations
-
-> For example, joint-energy models
-
-## Generalization and Robustness
-
-@sauer2021counterfactual generate counterfactual images for MNIST and ImageNet through independent mechanisms (IM): each IM learns class-conditional input distributions over a specific lower-dimensional, semantically meaningful factor, such as *texture*, *shape* and *background*.
-They demonstrate that using these generated counterfactuals during classifier training improves model robustness.
+This interpretation of the link between explainability through counterfactuals on one side, and robustness to adversarial examples on the other, is backed by empirical evidence.
+@sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness.
 Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
 They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.
+@teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
+Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function.
+In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose *POLYJUICE*, a general-purpose counterfactual generator for language models.
+They demonstrate empirically that augmenting training data through *POLYJUICE* counterfactuals improves robustness in a number of NLP tasks.
+@luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models.
+Specifically, they propose to proceed as follows: firstly, they identify training samples that are subject to high predictive uncertainty; secondly, they generate counterfactual explanations for those samples; and, finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.
 
-@teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as *gradient supervision*) (***this might be useful for our task as well***).
-In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose POLYJUICE, a general-purpose counterfactual generator for language models.
-They demonstrate empirically that augmenting training data through POLYJUICE counterfactuals improves robustness in a number of NLP tasks.
+There have also been several attempts at formalizing the relationship between counterfactual explanations (CE) and adversarial examples (AE).
+Pointing to clear similarities in how CE and AE are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problem in representation learning.
+Formally, AE can be seen as the subset of CE, for which misclassification is achieved [@freiesleben2022intriguing].
+Similarly, @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive theoretical upper bounds on the distances between them.
 
-## Link to Adversarial Training
+Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of post-hoc counterfactual explanations.
+Firstly, @ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals to move from an adverse outcome to some positive target class with high probability.
+The approach proposed by @ross2021learning builds on adversarial training, where in this context susceptibility to targeted adversarial examples for the positive class is explicitly induced.
+The proposed method allows for imposing a set of actionability constraints ex-ante: for example, users can specify that certain features (e.g.
+*age*, *gender*, ...) are immutable.
+Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes counterfactual explanations as part of the training procedure.
+In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable.
+Counterfactuals are generated during each training iteration and fed back to the predictor network.
+In contrast to @guo2023counternet, we impose no restrictions on the neural network architecture at all.
 
-@freiesleben2022intriguing propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE (***this might be a useful notion for use to distinguish between adversarials and explanations during training***); secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations.
+## Beyond Robustness
 
-## Closely Related
+Improving the adversarial robustness of models is not the only path towards aligning representations with desirable explanations.
+In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives.
+The authors propose a way to generate counterfactuals that are maximally **faithful** to the model in that they are consistent with what the model has learned about the underlying data.
+Formally, they rely on tools from energy-based modelling to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
+Their proposed counterfactual explainer, *ECCCo*, yields desirable (or **plausible**) explanations if and only if the underlying model has learned representations that align with them.
+They find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.
 
-@guo2023counternet are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce.
-In particular, they propose a specific network architecture that includes a predictor and CE generator network (***akin a GAN?***), where the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed back to the predictor network (***here we are aligned***).
-In contrast, we impose no restrictions on the neural network architecture at all.
-(***to ensure the one-hot encoding of categorical features is maintained, they simple use softmax (might be interesting for CE.jl)***) Interestingly, the authors find that their approach is sensitive to the choice of the loss function: only MSE seems to lead to good performance.
-They also demonstrate theoretically, that the objective function is difficult to optimize due to divergent gradients and suffers from poor adversarial robustness.
-(***because partial gradients with respect to the classification loss component and the counterfactual validity component point in opposite directions***).
-To mitigate these issues, the authors use block-wise gradient descent: they first update with respect to classification loss and then use a second update with respect to the other loss components (***this might be useful for our task as well***).
-@ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals with high probability.
-The approach builds on adversarial training (***here we are aligned***), where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class.
-The proposed method allows for imposing a set of actionable recourse ex-ante: for example, users can impose mutability constraints for features (***here we are aligned***).
-(***To solve their objective function more efficiently, they use a first-order Taylor approximation to approximate the recourse loss component (might be applicable in our case)***)
-
-@luu2023counterfactual introduce Counterfactual Adversarial Training (CAT) with intention of improving generalization and robustness of language models.
-Specifically, they propose to proceed as follows: firstly, identify training samples that are subject to high predictive uncertainty (entropy); secondly, generate counterfactual explanations for those samples; and, finally, finetune the model on the augmented dataset that includes the generated counterfactuals.
+Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom.
+Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case].
+Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data.
+Previous work by @schut2021generating similarly found that generating desirable ("interpretable") counterfactual explanations is almost trivial for deep ensembles that have also undergone adversarial training.
+The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your].
+This is closely related to the idea of aligning models with desirable explanations and has inspired our proposed counterfactual training objective, as we explain in @sec-method.

### 2025-02-04 paper/body_ecml.tex
progress on methodology

@@
 some desired output. Much existing research has focused on identifying
-explanations that are not only valid but also deemed desirable with
-respect to the underlying data and stakeholder requirements. Recent work
-has shown that under this premise, the task of learning desirable
-explanations is effectively reassigned from the model itself to the
-(post-hoc) counterfactual explainer. Building on that work, we propose a
-novel model objective that leverages counterfactuals during the training
-phase (ad-hoc) in order to minimize the divergence between learned
-representations and desirable explanations. Through extensive
-experiments, we demonstrate that our proposed methodology facilitates
-training models that inherently deliver desirable explanations while
-maintaining high predictive performance.
+explanations that are not only valid but also deemed plausible and
+desirable with respect to the underlying data and stakeholder
+requirements. Recent work has shown that under this premise, the task of
+learning plausible explanations is effectively reassigned from the model
+itself to the (post-hoc) counterfactual explainer. Building on that
+work, we propose a novel model objective that leverages counterfactuals
+during the training phase (ad-hoc) in order to minimize the divergence
+between learned representations and plausible explanations. Through
+extensive experiments, we demonstrate that our proposed methodology
+facilitates training models that inherently deliver plausible
+explanations while maintaining high predictive performance.
 \keywords{Counterfactual Explanations \and Explainable
@@
 relying on features and rules that are carefully hand-crafted by humans,
-modern AIs are tasked with learning these representations from scratch,
-guided by narrow objectives such as predictive accuracy
-\cite{goodfellow2016deep}. Modern advances in computing have made it
-possible to provide such AIs with ever greater degrees of freedom to
-achieve that task, which has often led them to outperform traditionally
-more parsimonious models. Unfortunately, in doing so they also learn
-increasingly complex and highly sensitive representations that we can no
-longer easily interpret.
+modern machine learning (ML) models are tasked with learning these
+representations from scratch, guided by narrow objectives such as
+predictive accuracy \cite{goodfellow2016deep}. Modern advances in
+computing have made it possible to provide such models with ever greater
+degrees of freedom to achieve that task, which has often led them to
+outperform traditionally more parsimonious models. Unfortunately, in
+doing so they also learn increasingly complex and highly sensitive
+representations that we can no longer easily interpret.
 
@@
 many solutions that are locally optimal with respect to narrow
-objectives \cite{wilson2020case}. Based purely on predictive
+objectives \cite{wilson2020case}\footnote{For clarity: we follow
+  standard ML convention in using ``degrees of freedom'' to refer to the
+  number of parameters estimated from data.}. Based purely on predictive
 performance, these solutions may seem to provide compelling explanations
@@
 first challenge, it is not obvious how to mitigate models from learning
-representations that correspond to meaningless and undesirable
+representations that correspond to meaningless and implausible
 explanations.
@@
 here as the degree to which learned representations correspond to
-explanations that are interpretable and deemed desirable by humans.
-Instead, the choice has typically been to improve the capacity of XAI
-tools to identify the subset explanations that are both desirable and
-valid for any given model, independent of whether the learned
-representations are also compatible with undesirable explanations
-\cite{altmeyer2024faithful}. Fortunately, recent findings indicate that
-explainability can arise as byproduct of regularization techniques aimed
-at other objectives such as robustness, generalization and generative
-capacity \cite{schut2021generating}.
+explanations that are interpretable and deemed \textbf{plausible} by
+humans (see Definition~\ref{def-explainability}). Instead, the choice
+has typically been to improve the capacity of XAI tools to identify the
+subset explanations that are both plausible and valid for any given
+model, independent of whether the learned representations are also
+compatible with implausible explanations \cite{altmeyer2024faithful}.
+Fortunately, recent findings indicate that explainability can arise as
+byproduct of regularization techniques aimed at other objectives such as
+robustness, generalization and generative capacity
+\cite{schut2021generating}.
 
@@
 training}: a novel regularization technique geared explicitly towards
-aligning model representations with desirable explanations. Our
+aligning model representations with plausible explanations. Our
 contributions are as follows:
@@
 define model explainability as the extent to which valid explanations
-derived for an opaque model are also deemed desirable with respect to
+derived for an opaque model are also deemed plausible with respect to
 the underlying data and stakeholder requirements. To make this more
@@
 susceptible to worst-case counterfactuals (i.e.~adversarial examples),
-AT effectively removes some undesirable explanations from the solution
+AT effectively removes some implausible explanations from the solution
 space.
@@
 Improving the adversarial robustness of models is not the only path
-towards aligning representations with desirable explanations. In a work
+towards aligning representations with plausible explanations. In a work
 closely related to this one, \cite{altmeyer2024faithful} show that
@@
 conditional posterior over inputs learned by the model. Their proposed
-counterfactual explainer, \emph{ECCCo}, yields desirable (or
-\textbf{plausible}) explanations if and only if the underlying model has
-learned representations that align with them. They find that both deep
-ensembles \cite{lakshminarayanan2016simple} and joint energy-based
-models (JEMs) \cite{grathwohl2020your} tend to do well in this regard.
+counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
+and only if the underlying model has learned representations that align
+with them. They find that both deep ensembles
+\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
+\cite{grathwohl2020your} tend to do well in this regard.
 
@@
 semantically meaningless explanations for the data. Previous work by
-\cite{schut2021generating} similarly found that generating desirable
+\cite{schut2021generating} similarly found that generating plausible
 (``interpretable'') counterfactual explanations is almost trivial for
@@
 \cite{grathwohl2020your}. This is closely related to the idea of
-aligning models with desirable explanations and has inspired our
+aligning models with plausible explanations and has inspired our
 proposed counterfactual training objective, as we explain in
@@
 
+Counterfactual training combines ideas from adversarial training,
+energy-based modelling and counterfactuals explanations with the
+explicit objective of aligning representations with plausible
+explanations that comply with user requirements. In the context of CE,
+plausibility has broadly been defined as the degree to which
+counterfactuals comply with the underlying data generating process
+\cite{poyiadzi2020face, guidotti2022counterfactual, altmeyer2024faithful}.
+Plausibility is a necessary but insufficient condition for using CE to
+provide algorithmic recourse (AR) to individuals affected by opaque
+models in practice. This is because for recourse recommendations to be
+\textbf{actionable}, they need to not only result in plausible
+counterfactuals but also be attainable. A plausible CE for a rejected
+20-year-old loan applicant, for example, might reveal that their
+application would have been accepted, if only they were 20 years older.
+Ignoring all other features, this complies with the definition of
+plausibility if 40-year-old individuals are in fact more credit-worthy
+on average than young adults. But of course this CE does not qualify for
+providing actionable recourse to the applicant. For our intents and
+purposes, counterfactual training aims at improving model explainability
+by aligning models with counterfactuals that meet both desiderata,
+plausibility and actionability. Formally, we define explainability as
+follows:
+
 \begin{definition}[Model
@@
 
-~
+Let \(M_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a supervised
+classification model that maps from intuts to representations
+\(\phi(x;theta)\) and finally to outputs. Assume that for any given
+input-output pair \(\{\mathbf{x},\mathbf{y}\}\) there exists a
+counterfactual
+\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})\)
+where \(\mathbf{y}^{+}\) denotes some target output. We say that
+\(M_\theta\) is \textbf{explainable} to the extent that faithfully
+generated counterfactuals are:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Plausible: \(\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}\)
+  with \(p \rightarrow 1\) as defined in \cite{altmeyer2024faithful}.
+\item
+  Actionable: permutations \(\Delta\) are subject to actionability
+  constraints.
+\end{enumerate}
 
 \end{definition}
+
+Actionability constraints in Definition~\ref{def-explainability} vary
+and depend on the context in which \(M_\theta\) is deployed. In this
+work, we focus on domain and mutability constraints for individual
+features. We also limit ourselves to classification tasks in this work,
+a limitation that we discuss in Section~\ref{sec-discussion}.
+
+\subsection{Our Proposed Objective}\label{our-proposed-objective}
+
+To train models with high explainability as defined in
+Definition~\ref{def-explainability}, we propose the following objective,
+
+\begin{equation}\phantomsection\label{eq-obj}{
+\text{yloss}(M_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(M_\theta({x^\prime}),y)
+}\end{equation}
+
+where \(\text{yloss}(\cdot)\) denotes any conventional classification
+loss function (e.g.~crossentropy) that induces discriminative
+(predictive) performance. The two additional components in
+Equation~\ref{eq-obj} are explained in more detail below. For now they
+can be sufficiently described as inducing explainability directly and
+indirectly by penalizing: 1) the contrastive divergence,
+\(\text{div}(\cdot)\), between counterfactuals \(x^\prime\) and observed
+samples \(x\) and, 2) the adversarial loss, \(\text{advloss}(.)\), with
+respect to counterfactuals. The tradeoff between the different
+components can be governed by adjusting the strengths of the penalties
+\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).
+
+\subsubsection{Directly Inducing Explainability through Contrastive
+Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}
+
+\subsubsection{Indirectly Inducing Explainability through Adversarial
+Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}
+
+A reasonable choice for the latter is to define
+\(\text{advloss}(M_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(M_\theta({x^\prime}),y)\)
+
+\subsection{Encoding Domain Knowledge}\label{encoding-domain-knowledge}
 

### 2025-02-04 paper/preprint/paper.tex
progress on methodology

@@
 \theoremstyle{definition}
+\newtheorem{exercise}{Parameters}[section]
+\theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
 \theoremstyle{remark}
@@
 some desired output. Much existing research has focused on identifying
-explanations that are not only valid but also deemed desirable with
-respect to the underlying data and stakeholder requirements. Recent work
-has shown that under this premise, the task of learning desirable
-explanations is effectively reassigned from the model itself to the
-(post-hoc) counterfactual explainer. Building on that work, we propose a
-novel model objective that leverages counterfactuals during the training
-phase (ad-hoc) in order to minimize the divergence between learned
-representations and desirable explanations. Through extensive
-experiments, we demonstrate that our proposed methodology facilitates
-training models that inherently deliver desirable explanations while
-maintaining high predictive performance.
+explanations that are not only valid but also deemed plausible and
+desirable with respect to the underlying data and stakeholder
+requirements. Recent work has shown that under this premise, the task of
+learning plausible explanations is effectively reassigned from the model
+itself to the (post-hoc) counterfactual explainer. Building on that
+work, we propose a novel model objective that leverages counterfactuals
+during the training phase (ad-hoc) in order to minimize the divergence
+between learned representations and plausible explanations. Through
+extensive experiments, we demonstrate that our proposed methodology
+facilitates training models that inherently deliver plausible
+explanations while maintaining high predictive performance.
 \end{abstract}
@@
 relying on features and rules that are carefully hand-crafted by humans,
-modern AIs are tasked with learning these representations from scratch,
-guided by narrow objectives such as predictive accuracy
-(\citeproc{ref-goodfellow2016deep}{I. Goodfellow, Bengio, and Courville
-2016}). Modern advances in computing have made it possible to provide
-such AIs with ever greater degrees of freedom to achieve that task,
-which has often led them to outperform traditionally more parsimonious
-models. Unfortunately, in doing so they also learn increasingly complex
-and highly sensitive representations that we can no longer easily
-interpret.
+modern machine learning (ML) models are tasked with learning these
+representations from scratch, guided by narrow objectives such as
+predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
+Bengio, and Courville 2016}). Modern advances in computing have made it
+possible to provide such models with ever greater degrees of freedom to
+achieve that task, which has often led them to outperform traditionally
+more parsimonious models. Unfortunately, in doing so they also learn
+increasingly complex and highly sensitive representations that we can no
+longer easily interpret.
 
@@
 many solutions that are locally optimal with respect to narrow
-objectives (\citeproc{ref-wilson2020case}{Wilson 2020}). Based purely on
-predictive performance, these solutions may seem to provide compelling
-explanations for the data, when in fact they are based on purely
-associative, semantically meaningless patterns. This poses two related
-challenges: firstly, it makes these models inherently opaque, since
-humans cannot simply interpret what type of explanation the complex
-learned representations correspond to; secondly, even if we could
-resolve the first challenge, it is not obvious how to mitigate models
-from learning representations that correspond to meaningless and
-undesirable explanations.
+objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
+  clarity: we follow standard ML convention in using ``degrees of
+  freedom'' to refer to the number of parameters estimated from data.}.
+Based purely on predictive performance, these solutions may seem to
+provide compelling explanations for the data, when in fact they are
+based on purely associative, semantically meaningless patterns. This
+poses two related challenges: firstly, it makes these models inherently
+opaque, since humans cannot simply interpret what type of explanation
+the complex learned representations correspond to; secondly, even if we
+could resolve the first challenge, it is not obvious how to mitigate
+models from learning representations that correspond to meaningless and
+implausible explanations.
 
@@
 here as the degree to which learned representations correspond to
-explanations that are interpretable and deemed desirable by humans.
-Instead, the choice has typically been to improve the capacity of XAI
-tools to identify the subset explanations that are both desirable and
-valid for any given model, independent of whether the learned
-representations are also compatible with undesirable explanations
+explanations that are interpretable and deemed \textbf{plausible} by
+humans (see Definition~\ref{def-explainability}). Instead, the choice
+has typically been to improve the capacity of XAI tools to identify the
+subset explanations that are both plausible and valid for any given
+model, independent of whether the learned representations are also
+compatible with implausible explanations
 (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
@@
 Improving the adversarial robustness of models is not the only path
-towards aligning representations with desirable explanations. In a work
+towards aligning representations with plausible explanations. In a work
 closely related to this one, Altmeyer et al.
@@
 inputs learned by the model. Their proposed counterfactual explainer,
-\emph{ECCCo}, yields desirable (or \textbf{plausible}) explanations if
-and only if the underlying model has learned representations that align
-with them. They find that both deep ensembles
+\emph{ECCCo}, yields plausible explanations if and only if the
+underlying model has learned representations that align with them. They
+find that both deep ensembles
 (\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
@@
 (\citeproc{ref-schut2021generating}{2021}) similarly found that
-generating desirable (``interpretable'') counterfactual explanations is
+generating plausible (``interpretable'') counterfactual explanations is
 almost trivial for deep ensembles that have also undergone adversarial
@@
 capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
-is closely related to the idea of aligning models with desirable
+is closely related to the idea of aligning models with plausible
 explanations and has inspired our proposed counterfactual training
@@
 
+Counterfactual training combines ideas from adversarial training,
+energy-based modelling and counterfactuals explanations with the
+explicit objective of aligning representations with plausible
+explanations that comply with user requirements. In the context of CE,
+plausibility has broadly been defined as the degree to which
+counterfactuals comply with the underlying data generating process
+(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
+\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
+\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
+is a necessary but insufficient condition for using CE to provide
+algorithmic recourse (AR) to individuals affected by opaque models in
+practice. This is because for recourse recommendations to be
+\textbf{actionable}, they need to not only result in plausible
+counterfactuals but also be attainable. A plausible CE for a rejected
+20-year-old loan applicant, for example, might reveal that their
+application would have been accepted, if only they were 20 years older.
+Ignoring all other features, this complies with the definition of
+plausibility if 40-year-old individuals are in fact more credit-worthy
+on average than young adults. But of course this CE does not qualify for
+providing actionable recourse to the applicant. For our intents and
+purposes, counterfactual training aims at improving model explainability
+by aligning models with counterfactuals that meet both desiderata,
+plausibility and actionability. Formally, we define explainability as
+follows:
+
 \begin{definition}[Model
@@
 
-~
+Let \(M_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a supervised
+classification model that maps from intuts to representations
+\(\phi(x;theta)\) and finally to outputs. Assume that for any given
+input-output pair \(\{\mathbf{x},\mathbf{y}\}\) there exists a
+counterfactual
+\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})\)
+where \(\mathbf{y}^{+}\) denotes some target output. We say that
+\(M_\theta\) is \textbf{explainable} to the extent that faithfully
+generated counterfactuals are:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Plausible: \(\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}\)
+  with \(p \rightarrow 1\) as defined in Altmeyer et al.
+  (\citeproc{ref-altmeyer2024faithful}{2024}).
+\item
+  Actionable: permutations \(\Delta\) are subject to actionability
+  constraints.
+\end{enumerate}
 
 \end{definition}
+
+Actionability constraints in Definition~\ref{def-explainability} vary
+and depend on the context in which \(M_\theta\) is deployed. In this
+work, we focus on domain and mutability constraints for individual
+features. We also limit ourselves to classification tasks in this work,
+a limitation that we discuss in Section~\ref{sec-discussion}.
+
+\subsection{Our Proposed Objective}\label{our-proposed-objective}
+
+To train models with high explainability as defined in
+Definition~\ref{def-explainability}, we propose the following objective,
+
+\begin{equation}\phantomsection\label{eq-obj}{
+\text{yloss}(M_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(M_\theta({x^\prime}),y)
+}\end{equation}
+
+where \(\text{yloss}(\cdot)\) denotes any conventional classification
+loss function (e.g.~crossentropy) that induces discriminative
+(predictive) performance. The two additional components in
+Equation~\ref{eq-obj} are explained in more detail below. For now they
+can be sufficiently described as inducing explainability directly and
+indirectly by penalizing: 1) the contrastive divergence,
+\(\text{div}(\cdot)\), between counterfactuals \(x^\prime\) and observed
+samples \(x\) and, 2) the adversarial loss, \(\text{advloss}(.)\), with
+respect to counterfactuals. The tradeoff between the different
+components can be governed by adjusting the strengths of the penalties
+\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).
+
+\subsubsection{Directly Inducing Explainability through Contrastive
+Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}
+
+\subsubsection{Indirectly Inducing Explainability through Adversarial
+Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}
+
+A reasonable choice for the latter is to define
+\(\text{advloss}(M_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(M_\theta({x^\prime}),y)\)
+
+\subsection{Encoding Domain Knowledge}\label{encoding-domain-knowledge}
 
@@
 \emph{International Conference on Learning Representations}.
+
+\bibitem[\citeproctext]{ref-guidotti2022counterfactual}
+Guidotti, Riccardo. 2022. {``Counterfactual Explanations and How to Find
+Them: Literature Review and Benchmarking.''} \emph{Data Mining and
+Knowledge Discovery}, 1--55.
 
@@
 \url{https://proceedings.mlr.press/v151/pawelczyk22a.html}.
+
+\bibitem[\citeproctext]{ref-poyiadzi2020face}
+Poyiadzi, Rafael, Kacper Sokol, Raul Santos-Rodriguez, Tijl De Bie, and
+Peter Flach. 2020. {``{FACE}: {Feasible} and Actionable Counterfactual
+Explanations.''} In \emph{Proceedings of the {AAAI}/{ACM Conference} on
+{AI}, {Ethics}, and {Society}}, 344--50.
 

### 2025-02-04 paper/sections/abstract.qmd
progress on methodology

--- 2025-01-28 paper/sections/abstract.qmd
@@
 
-Counterfactual Explanations (CE) have emerged as a popular method to explain the predictions made by opaque machine learning models in a post-hoc fashion.
-We propose a novel approach that leverages counterfactuals during the training phase of models.
+Counterfactual Explanations have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output.
+Much existing research has focused on identifying explanations that are not only valid but also deemed plausible and desirable with respect to the underlying data and stakeholder requirements.
+Recent work has shown that under this premise, the task of learning plausible explanations is effectively reassigned from the model itself to the (post-hoc) counterfactual explainer.
+Building on that work, we propose a novel model objective that leverages counterfactuals during the training phase (ad-hoc) in order to minimize the divergence between learned representations and plausible explanations.
+Through extensive experiments, we demonstrate that our proposed methodology facilitates training models that inherently deliver plausible explanations while maintaining high predictive performance.

### 2025-02-04 paper/sections/method.qmd
progress on methodology

@@
 # Counterfactual Training {#sec-method}
+
+Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit objective of aligning representations with plausible explanations that comply with user requirements.
+In the context of CE, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful].
+Plausibility is a necessary but insufficient condition for using CE to provide algorithmic recourse (AR) to individuals affected by opaque models in practice.
+This is because for recourse recommendations to be **actionable**, they need to not only result in plausible counterfactuals but also be attainable.
+A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older.
+Ignoring all other features, this complies with the definition of plausibility if 40-year-old individuals are in fact more credit-worthy on average than young adults.
+But of course this CE does not qualify for providing actionable recourse to the applicant.
+For our intents and purposes, counterfactual training aims at improving model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability.
+Formally, we define explainability as follows:
 
@@
 
+Let $M_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from intuts to representations $\phi(x;theta)$ and finally to outputs.
+Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})$ where $\mathbf{y}^{+}$ denotes some target output.
+We say that $M_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are:
+
+1. Plausible: $\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}$ with $p \rightarrow 1$  as defined in @altmeyer2024faithful.
+2. Actionable: permutations $\Delta$ are subject to actionability constraints.
+
 :::
+
+Actionability constraints in @def-explainability vary and depend on the context in which $M_\theta$ is deployed.
+In this work, we focus on domain and mutability constraints for individual features.
+We also limit ourselves to classification tasks in this work, a limitation that we discuss in @sec-discussion.
+
+## Our Proposed Objective
+
+To train models with high explainability as defined in @def-explainability, we propose the following objective,
+
+$$
+\text{yloss}(M_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(M_\theta({x^\prime}),y)
+$$ {#eq-obj}
+
+where $\text{yloss}(\cdot)$ denotes any conventional classification loss function (e.g.
+crossentropy) that induces discriminative (predictive) performance.
+The two additional components in @eq-obj are explained in more detail below.
+For now they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $x^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to counterfactuals.
+The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.
+
+### Directly Inducing Explainability through Contrastive Divergence
+
+### Indirectly Inducing Explainability through Adversarial Robustness
+
+A reasonable choice for the latter is to define $\text{advloss}(M_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(M_\theta({x^\prime}),y)$
+
+## Encoding Domain Knowledge

### 2025-02-04 paper/preprint/paper.tex
of

@@
 where \(\text{yloss}(\cdot)\) denotes any conventional classification
-loss function (e.g.~crossentropy) that induces discriminative
+loss function (e.g.~cross-entropy) that induces discriminative
 (predictive) performance. The two additional components in
@@
 Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}
+
+Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
+any classifier can be re-interpreted as a joint energy-based model (JEM)
+that learns to discriminate output classes conditional on inputs and
+generate inputs. They show that JEMs can perform well at both tasks by
+directly optimizing the joint distribution
+\(\log p_\theta(x,y)=\log p_\theta(y|x) + \log p_\theta(x)\). The first
+factor The distribution over inputs is learned via contrastive
+divergence:
+
+Considered through the lens of explainability, JEMs learn plausible
+explanations.
 

### 2025-02-04 paper/sections/method.qmd
of

@@
 where $\text{yloss}(\cdot)$ denotes any conventional classification loss function (e.g.
-crossentropy) that induces discriminative (predictive) performance.
+cross-entropy) that induces discriminative (predictive) performance.
 The two additional components in @eq-obj are explained in more detail below.
@@
 
+@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on inputs and generate inputs.
+They show that JEMs can perform well at both tasks by directly optimizing the joint distribution $\log p_\theta(x,y)=\log p_\theta(y|x) + \log p_\theta(x)$.
+The first factor
+The distribution over inputs is learned via contrastive divergence:
+
+Considered through the lens of explainability, JEMs learn plausible explanations.
+
 ### Indirectly Inducing Explainability through Adversarial Robustness

### 2025-02-05 paper/body_ecml.tex
vamos

@@
 
-Let \(M_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a supervised
-classification model that maps from intuts to representations
-\(\phi(x;theta)\) and finally to outputs. Assume that for any given
-input-output pair \(\{\mathbf{x},\mathbf{y}\}\) there exists a
-counterfactual
-\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})\)
+Let \(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a
+supervised classification model that maps from the \(D\)-dimensional
+input space \(\mathcal{X}\) to representations
+\(\phi(\mathbf{x};\theta)\) and finally to the \(K\)-dimensional output
+space \(\mathcal{Y}\). Assume that for any given input-output pair
+\(\{\mathbf{x},\mathbf{y}\}_i\) there exists a counterfactual
+\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
 where \(\mathbf{y}^{+}\) denotes some target output. We say that
-\(M_\theta\) is \textbf{explainable} to the extent that faithfully
-generated counterfactuals are:
+\(\mathbf{M}_\theta\) is \textbf{explainable} to the extent that
+faithfully generated counterfactuals are plausible (i.e.~consistent with
+the data) and actionable. Formally, we define these properties as
+follows:
 
@@
 \item
-  Plausible: \(\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}\)
-  with \(p \rightarrow 1\) as defined in \cite{altmeyer2024faithful}.
+  (Plausibility)
+  \(\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
+  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
 \item
-  Actionable: permutations \(\Delta\) are subject to actionability
+  (Actionability) Permutations \(\Delta\) are subject to actionability
   constraints.
+\end{enumerate}
+
+We consider counterfactuals as faithful to the extent that they are
+consistent with what the model has learned about the input data. Let
+\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
+over inputs, then formally:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  (Faithfulness)
+  \(\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
+  where \(A\) is defined as above.
 \end{enumerate}
@@
 
-Actionability constraints in Definition~\ref{def-explainability} vary
-and depend on the context in which \(M_\theta\) is deployed. In this
-work, we focus on domain and mutability constraints for individual
-features. We also limit ourselves to classification tasks in this work,
-a limitation that we discuss in Section~\ref{sec-discussion}.
+The definitions of faithfulness and plausibility in
+Definition~\ref{def-explainability} are the same as in
+\cite{altmeyer2024faithful}, with adapted notation. Actionability
+constraints in Definition~\ref{def-explainability} vary and depend on
+the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
+focus on domain and mutability constraints for individual features
+\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
+for reasons discussed in Section~\ref{sec-discussion}.
 
@@
 \begin{equation}\phantomsection\label{eq-obj}{
-\text{yloss}(M_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(M_\theta({x^\prime}),y)
+\text{yloss}(\mathbf{M}_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta({x^\prime}),y)
 }\end{equation}
@@
 where \(\text{yloss}(\cdot)\) denotes any conventional classification
-loss function (e.g.~crossentropy) that induces discriminative
+loss function (e.g.~cross-entropy) that induces discriminative
 (predictive) performance. The two additional components in
-Equation~\ref{eq-obj} are explained in more detail below. For now they
+Equation~\ref{eq-obj} are explained in more detail below. For now, they
 can be sufficiently described as inducing explainability directly and
@@
 
+\cite{grathwohl2020your} observe that any classifier can be
+re-interpreted as a joint energy-based model (JEM) that learns to
+discriminate output classes conditional on inputs and generate inputs.
+They show that JEMs can be trained to perform well at both tasks by
+directly maximizing the joint log-likelihood factorized as
+\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
+The first factor can be optimized using conventional cross-entropy as in
+Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
+\cite{grathwohl2020your} minimize the contrastive divergence between
+samples drawn from \(p_\theta(\mathbf{x})\) and training observations,
+i.e.~samples from \(p(\mathbf{x})\).
+
+A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
+tend to do well with respect to the plausibility objective in
+Definition~\ref{def-explainability}. If we consider samples drawn from
+\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
+finding, because the JEM objective effectively minimizes the divergence
+between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
+To generate samples, \cite{grathwohl2020your} rely on Stochastic
+Gradient Langevin Dynamics (SGLD) using an uninformative prior for
+initialization. This is where we depart from their methodology: instead
+of generating samples through SGLD, we propose using counterfactual
+explainers to generate counterfactuals for observed training samples.
+
+This departure allows us to tap into the vast repertoire of explainers
+that have been proposed in the literature to meet different desiderata.
+Typically, these methods facilitate the imposition of domain and
+mutability constraints, for example. In principle, any existing approach
+for generating counterfactual explanations is viable, so long as it does
+not violate the faithfulness condition.
+
 \subsubsection{Indirectly Inducing Explainability through Adversarial
@@
 A reasonable choice for the latter is to define
-\(\text{advloss}(M_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(M_\theta({x^\prime}),y)\)
+\(\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)\)
 

### 2025-02-05 paper/preprint/paper.tex
vamos

@@
 \item
-  Plausible: \(\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}\)
-  with \(p \rightarrow 1\) as defined in Altmeyer et al.
-  (\citeproc{ref-altmeyer2024faithful}{2024}).
-\item
-  Actionable: permutations \(\Delta\) are subject to actionability
+  (Plausibility)
+  \(\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
+  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
+\item
+  (Actionability) Permutations \(\Delta\) are subject to actionability
   constraints.
@@
 
+We consider counterfactuals as faithful to the extent that they are
+consistent with what the model has learned about the input data. Let
+\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
+over inputs, then formally:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
+  (Faithfulness)
+  \(\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
+  where \(A\) is defined as above.
+\end{enumerate}
+
 \end{definition}
 
+The definitions of faithfulness and plausibility in
+Definition~\ref{def-explainability} are the same as in Altmeyer et al.
+(\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
 Actionability constraints in Definition~\ref{def-explainability} vary
-and depend on the context in which \(M_\theta\) is deployed. In this
-work, we focus on domain and mutability constraints for individual
-features. We also limit ourselves to classification tasks in this work,
-a limitation that we discuss in Section~\ref{sec-discussion}.
+and depend on the context in which \(\mathbf{M}_\theta\) is deployed. In
+this work, we focus on domain and mutability constraints for individual
+features \(x_d\) for \(d=1,...,D\). We limit ourselves to classification
+tasks for reasons discussed in Section~\ref{sec-discussion}.
 
@@
 (predictive) performance. The two additional components in
-Equation~\ref{eq-obj} are explained in more detail below. For now they
+Equation~\ref{eq-obj} are explained in more detail below. For now, they
 can be sufficiently described as inducing explainability directly and
@@
 that learns to discriminate output classes conditional on inputs and
-generate inputs. They show that JEMs can perform well at both tasks by
-directly optimizing the joint distribution
-\(\log p_\theta(x,y)=\log p_\theta(y|x) + \log p_\theta(x)\). The first
-factor The distribution over inputs is learned via contrastive
-divergence:
-
-Considered through the lens of explainability, JEMs learn plausible
-explanations.
+generate inputs. They show that JEMs can be trained to perform well at
+both tasks by directly maximizing the joint log-likelihood factorized as
+\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
+The first factor can be optimized using conventional cross-entropy as in
+Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
+Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
+contrastive divergence between samples drawn from
+\(p_\theta(\mathbf{x})\) and training observations, i.e.~samples from
+\(p(\mathbf{x})\).
+
+A key empirical finding in Altmeyer et al.
+(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
+well with respect to the plausibility objective in
+Definition~\ref{def-explainability}. If we consider samples drawn from
+\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
+finding, because the JEM objective effectively minimizes the divergence
+between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
+To generate samples, Grathwohl et al.
+(\citeproc{ref-grathwohl2020your}{2020}) rely on Stochastic Gradient
+Langevin Dynamics (SGLD) using an uninformative prior for
+initialization. This is where we depart from their methodology: instead
+of generating samples through SGLD, we propose using counterfactual
+explainers to generate counterfactuals for observed training samples.
+
+This departure allows us to tap into the vast repertoire of explainers
+that have been proposed in the literature to meet different desiderata.
+Typically, these methods facilitate the imposition of domain and
+mutability constraints, for example. In principle, any existing approach
+for generating counterfactual explanations is viable, so long as it does
+not violate the faithfulness condition.
 

### 2025-02-05 paper/sections/method.qmd
vamos

@@
 
-Let $M_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from intuts to representations $\phi(x;theta)$ and finally to outputs.
-Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: M_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = M_\theta(\mathbf{x})$ where $\mathbf{y}^{+}$ denotes some target output.
-We say that $M_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are:
+Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$.
+Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\mathbf{y}^{+}$ denotes some target output.
+We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e.
+consistent with the data) and actionable.
+Formally, we define these properties as follows:
 
-1. Plausible: $\mathbf{x}^{\prime} \sim_{p} \mathcal{X}|\mathbf{y}^{+}$ with $p \rightarrow 1$  as defined in @altmeyer2024faithful.
-2. Actionable: permutations $\Delta$ are subject to actionability constraints.
+1. (Plausibility) $\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
+2. (Actionability) Permutations $\Delta$ are subject to actionability constraints.
+
+We consider counterfactuals as faithful to the extent that they are consistent with what the model has learned about the input data.
+Let $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denote the conditional posterior over inputs, then formally:
+
+3. (Faithfulness) $\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
 
@@
 
-Actionability constraints in @def-explainability vary and depend on the context in which $M_\theta$ is deployed.
-In this work, we focus on domain and mutability constraints for individual features.
-We also limit ourselves to classification tasks in this work, a limitation that we discuss in @sec-discussion.
+The definitions of faithfulness and plausibility in @def-explainability are the same as in @altmeyer2024faithful, with adapted notation.
+Actionability constraints in @def-explainability vary and depend on the context in which $\mathbf{M}_\theta$ is deployed.
+In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$.
+We limit ourselves to classification tasks for reasons discussed in @sec-discussion.
 
@@
 $$
-\text{yloss}(M_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(M_\theta({x^\prime}),y)
+\text{yloss}(\mathbf{M}_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta({x^\prime}),y)
 $$ {#eq-obj}
@@
 The two additional components in @eq-obj are explained in more detail below.
-For now they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $x^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to counterfactuals.
+For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $x^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to counterfactuals.
 The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.
@@
 @grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on inputs and generate inputs.
-They show that JEMs can perform well at both tasks by directly optimizing the joint distribution $\log p_\theta(x,y)=\log p_\theta(y|x) + \log p_\theta(x)$.
-The first factor
-The distribution over inputs is learned via contrastive divergence:
+They show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$.
+The first factor can be optimized using conventional cross-entropy as in @eq-obj.
+To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between samples drawn from $p_\theta(\mathbf{x})$ and training observations, i.e.
+samples from $p(\mathbf{x})$.
 
-Considered through the lens of explainability, JEMs learn plausible explanations.
+A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability.
+If we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals, this is an expected finding, because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$.
+To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization.
+This is where we depart from their methodology: instead of generating samples through SGLD, we propose using counterfactual explainers to generate counterfactuals for observed training samples.
+
+This departure allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
+Typically, these methods facilitate the imposition of domain and mutability constraints, for example.
+In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition.
 
@@
 
-A reasonable choice for the latter is to define $\text{advloss}(M_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(M_\theta({x^\prime}),y)$
+A reasonable choice for the latter is to define $\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)$
 

### 2025-02-05 paper/body_ecml.tex
progress

@@
 \begin{equation}\phantomsection\label{eq-obj}{
-\text{yloss}(\mathbf{M}_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta({x^\prime}),y)
+\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),y) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime};\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),y)
 }\end{equation}
@@
 explainers to generate counterfactuals for observed training samples.
+Specifically, we have
 
-This departure allows us to tap into the vast repertoire of explainers
-that have been proposed in the literature to meet different desiderata.
-Typically, these methods facilitate the imposition of domain and
-mutability constraints, for example. In principle, any existing approach
-for generating counterfactual explanations is viable, so long as it does
-not violate the faithfulness condition.
+\begin{equation}\phantomsection\label{eq-div}{
+\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
+}\end{equation}
+
+where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
+generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
+target class \(\mathbf{y}^+ \sim p(\mathbf{y})\) and then generating a
+counterfactual explanation for that target, similar to how conditional
+sampling can be used for JEMs \cite{grathwohl2020your}. Intuitively, the
+gradient of Equation~\ref{eq-div} decreases the energy of observed
+training samples (positive samples) while at same time increasing the
+energy of counterfactuals (negative samples) \cite{du2019implicit}. As
+the generated counterfactuals get more plausible
+(Definition~\ref{def-explainability}) over the cause of training, these
+two opposing effects gradually balance each out \cite{lippe2024uvadlc}.
+
+The departure from SGLD allows us to tap into the vast repertoire of
+explainers that have been proposed in the literature to meet different
+desiderata. Typically, these methods facilitate the imposition of domain
+and mutability constraints, for example. In principle, any existing
+approach for generating counterfactual explanations is viable, so long
+as it does not violate the faithfulness condition. Like JEMs
+\cite{murphy2022probabilistic}, counterfactual training can be
+considered as a form of contrastive representation learning.
 
@@
 
-A reasonable choice for the latter is to define
-\(\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)\)
+Based on our analysis in Section~\ref{sec-lit}, counterfactuals
+\(\mathbf{x}^\prime\) can be repurposed as additional training samples
+\cite{luu2023counterfactual} or adversarial examples
+\cite{freiesleben2022intriguing, pawelczyk2022exploring}. This leaves
+some flexibility with respect to the exact choice for
+\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
+functional form to use, though likely not the only reasonable choice, is
+inspired by adversarial training:
+
+\begin{equation}\phantomsection\label{eq-adv}{
+\begin{aligned}
+\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
+\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
+0 & \text{otherwise.}
+\end{cases}
+\end{aligned}
+}\end{equation}
+
+Under this choice we treat the counterfactual \(\mathbf{x}^\prime\) as
+an adversarial example iff it is imperceptible, i.e.~the magnitude of
+the perturbation of any individual feature is upper-bounded at
+\(\varepsilon\).
 

### 2025-02-05 paper/preprint/paper.tex
progress

@@
 explainers to generate counterfactuals for observed training samples.
-
-This departure allows us to tap into the vast repertoire of explainers
-that have been proposed in the literature to meet different desiderata.
-Typically, these methods facilitate the imposition of domain and
-mutability constraints, for example. In principle, any existing approach
-for generating counterfactual explanations is viable, so long as it does
-not violate the faithfulness condition.
+Specifically, we have
+
+\begin{equation}\phantomsection\label{eq-div}{
+\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
+}\end{equation}
+
+where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
+generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
+target class \(\mathbf{y}^+ \sim p(\mathbf{y})\) and then generating a
+counterfactual explanation for that target, similar to how conditional
+sampling can be used for JEMs
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). Intuitively,
+the gradient of Equation~\ref{eq-div} decreases the energy of observed
+training samples (positive samples) while at same time increasing the
+energy of counterfactuals (negative samples)
+(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
+counterfactuals get more plausible (Definition~\ref{def-explainability})
+over the cause of training, these two opposing effects gradually balance
+each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
+
+The departure from SGLD allows us to tap into the vast repertoire of
+explainers that have been proposed in the literature to meet different
+desiderata. Typically, these methods facilitate the imposition of domain
+and mutability constraints, for example. In principle, any existing
+approach for generating counterfactual explanations is viable, so long
+as it does not violate the faithfulness condition. Like JEMs
+(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), counterfactual
+training can be considered as a form of contrastive representation
+learning.
 
@@
 
-A reasonable choice for the latter is to define
-\(\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)\)
+Based on our analysis in Section~\ref{sec-lit}, counterfactuals
+\(\mathbf{x}^\prime\) can be repurposed as additional training samples
+(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023}) or
+adversarial examples
+(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
+\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
+leaves some flexibility with respect to the exact choice for
+\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
+functional form to use, though likely not the only reasonable choice, is
+inspired by adversarial training:
+
+\begin{equation}\phantomsection\label{eq-adv}{
+\begin{aligned}
+\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
+\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
+0 & \text{otherwise.}
+\end{cases}
+\end{aligned}
+}\end{equation}
+
+Under this choice we treat the counterfactual \(\mathbf{x}^\prime\) as
+an adversarial example iff it is imperceptible, i.e.~the magnitude of
+the perturbation of any individual feature is upper-bounded at
+\(\varepsilon\).
 
@@
 
+\bibitem[\citeproctext]{ref-du2019implicit}
+Du, Yilun, and Igor Mordatch. 2020. {``Implicit Generation and
+Generalization in Energy-Based Models.''}
+\url{https://arxiv.org/abs/1903.08689}.
+
 \bibitem[\citeproctext]{ref-freiesleben2022intriguing}
@@
 
+\bibitem[\citeproctext]{ref-lippe2024uvadlc}
+Lippe, Phillip. 2024. {``{UvA Deep Learning Tutorials}.''}
+\url{https://uvadlc-notebooks.readthedocs.io/en/latest/}.
+
 \bibitem[\citeproctext]{ref-luu2023counterfactual}
@@
 Information and Computation}, 881--88.
+
+\bibitem[\citeproctext]{ref-murphy2022probabilistic}
+Murphy, Kevin P. 2022. \emph{Probabilistic {Machine Learning}: {An}
+Introduction}. {MIT Press}.
 

### 2025-02-05 paper/sections/method.qmd
progress

@@
 $$
-\text{yloss}(\mathbf{M}_\theta(x),y) + \lambda_{\text{div}} \text{div}({x^\prime},x;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta({x^\prime}),y)
+\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),y) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime};\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),y)
 $$ {#eq-obj}
@@
 This is where we depart from their methodology: instead of generating samples through SGLD, we propose using counterfactual explainers to generate counterfactuals for observed training samples.
+Specifically, we have
 
-This departure allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
+$$
+\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
+$$ {#eq-div}
+
+where $\mathcal{E}_\theta(\cdot)$ denotes the energy function.
+We generate samples $\mathbf{x}^\prime$ by first randomly sampling the target class $\mathbf{y}^+ \sim p(\mathbf{y})$ and then generating a counterfactual explanation for that target, similar to how conditional sampling can be used for JEMs [@grathwohl2020your].
+Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit].
+As the generated counterfactuals get more plausible (@def-explainability) over the cause of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].
+
+The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
 Typically, these methods facilitate the imposition of domain and mutability constraints, for example.
 In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition.
+Like JEMs [@murphy2022probabilistic], counterfactual training can be considered as a form of contrastive representation learning.
 
@@
 
-A reasonable choice for the latter is to define $\text{advloss}(\mathbf{M}_\theta({x^\prime}),y;\varepsilon):=\text{yloss}(\mathbf{M}_\theta({x^\prime}),y)$
+Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring].
+This leaves some flexibility with respect to the exact choice for $\text{advloss}(\cdot)$ in @eq-obj.
+An intuitive functional form to use, though likely not the only reasonable choice, is inspired by adversarial training:
+
+$$
+\begin{aligned}
+\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
+\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
+0 & \text{otherwise.}
+\end{cases}
+\end{aligned}
+$$ {#eq-adv}
+
+Under this choice we treat the counterfactual $\mathbf{x}^\prime$ as an adversarial example iff it is imperceptible, i.e.
+the magnitude of the perturbation of any individual feature is upper-bounded at $\varepsilon$.
 

### 2025-02-06 paper/sections/other/constraints.qmd
vamos

```{julia}
projectdir = splitpath(pwd()) |>
    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
cd(projectdir)
```

```{julia}
using CTExperiments
using CTExperiments.CounterfactualExplanations
using CTExperiments.CounterfactualTraining
using CTExperiments.Flux
using CTExperiments.Plots
using CTExperiments.TaijaPlotting
```

# Linear Model

First, we define the mutability constraints and load the data and model:

```{julia}
using CTExperiments: 
    get_ce_data, 
    train_val_split, 
    build_model, 
    LinearModel, 
    get_input_encoder

# Data:
# constraints = ["none", "both"]        # x1 cannot be mutated
constraints = ["both", "none"]        # x2 cannot be mutated
data = LinearlySeparable(
    n_train=500,
    batchsize=50,
    mutability=constraints
)
ce_data = get_ce_data(data)
val_size = data.n_validation / (data.n_validation + data.n_train)
train_set, val_set, _ = train_val_split(data, ce_data, val_size)

# Model:
nin = size(first(train_set)[1], 1)
nout = size(first(train_set)[2], 1)
model = build_model(LinearModel(), nin, nout)
```

Next, we defined the training objective and the counterfactual generator. Finally, we initialize the optimization state.

```{julia}
obj = EnergyDifferentialObjective(lambda=[1.0,0.1,0.0])
# obj = VanillaObjective()
generator = GenericGenerator()
opt_state = Flux.setup(Descent(), model)
```

Then we train the model:

```{julia}
using Random
Random.seed!(42)

model, logs = counterfactual_training(
    obj,
    model,
    generator,
    train_set,
    opt_state;
    val_set = val_set,
    nepochs = 50,
    mutability = Symbol.(constraints)
)
```

Finally, we visualize the results:

```{julia}
coeff = get_log_reg_params(model)
db = get_decision_boundary(coeff)
Plots.scatter(ce_data)
Plots.abline!(db.slope, db.intercept; lw=5, label="Dec. Boundary")
```

### 2025-02-06 paper/experiments/CTExperiments/src/experiment.jl

@@ -389,8 +389,12 @@ end
 function get_log_reg_params(exper::Experiment)
     @assert isa(exper.model_type,LinearModel) "Model needs to be linear model."
     M = load_results(exper)[3]
-    coeffs =
-        Flux.params(M.model) |>
+    coeffs = get_log_reg_params(M.model)
+    return coeffs
+end
+
+function get_log_reg_params(model)
+    coeffs = Flux.params(model) |>
         W -> (β₀=W[2][1] - W[2][2], β₁=W[1][1, 1] - W[1][2, 1], β₂=W[1][1, 2] - W[1][2, 2])
     return coeffs
 end
@@ -405,7 +409,11 @@ end
 function get_decision_boundary(exper::Experiment; flip_axis=false)
     @assert input_dim(exper.data) == 2 "Decision boundary is only defined for 2D data."
     coeffs = get_log_reg_params(exper)
-    if !flip_axis 
+    return get_decision_boundary(coeffs; flip_axis=flip_axis)
+end
+
+function get_decision_boundary(coeffs; flip_axis=false)
+    if !flip_axis
         dec_boundary = DecisionBoundary(-(coeffs.β₀ / coeffs.β₂), -(coeffs.β₁ / coeffs.β₂))
     else
         dec_boundary = DecisionBoundary(-(coeffs.β₀ / coeffs.β₁), -(coeffs.β₂ / coeffs.β₁))

### 2025-02-06 src/implausibility.jl

@@ -8,7 +8,8 @@ Compute the implausibility (contrastive divergence) of the counterfactuals (`per
 """
 function implausibility(model, perturbed_input, samples, targets)
     E(x) = -model(x)                                    # energy
-    x = ((E(samples)) - (E(perturbed_input)))'targets   # contrastive divergence
+    # x = ((E(samples)) - (E(perturbed_input)))'targets   # contrastive divergence
+    x = ((E(perturbed_input)) - (E(samples)))'targets   # contrastive divergence
     return diag(x[:, :])
 end
 

### 2025-02-06 paper/sections/other/constraints.qmd
gotta revisit this

@@
 # constraints = ["none", "both"]        # x1 cannot be mutated
-constraints = ["both", "none"]        # x2 cannot be mutated
+constraints = ["both", "both"]        # x2 cannot be mutated
 data = LinearlySeparable(

### 2025-02-06 paper/sections/other/constraints.qmd
uhhh

--- 2025-02-06 paper/sections/other/constraints.qmd
@@
 ```{julia}
+using CounterfactualExplanations.Convergence
 obj = EnergyDifferentialObjective(lambda=[1.0,0.1,0.0])
@@
 opt_state = Flux.setup(Descent(), model)
+conv = MaxIterConvergence()
+```
+
+```{julia}
+function plot_db(model, x)
+    coeff = get_log_reg_params(model)
+    db = get_decision_boundary(coeff)
+    plt = Plots.scatter(ce_data)
+    yhat = [argmax(y) for y in eachcol(model(x))]
+    Plots.scatter!(x[1,:], x[2,:], label="Counterfactuals", ms=15, shape=:star, color=yhat)
+    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
+    display(plt)
+end
 ```
@@
     val_set = val_set,
-    nepochs = 50,
-    mutability = Symbol.(constraints)
+    nepochs = 20,
+    mutability = Symbol.(constraints),
+    callback = plot_db,
+    nce=50,
+    convergence=conv, 
 )

### 2025-02-06 paper/experiments/evaluations/poc/mlp/mnist/evaluation/present_results.jl

@@ -1,0 +1,45 @@
+using CTExperiments
+using CTExperiments.CairoMakie
+using CTExperiments.DataFrames
+using DotEnv
+
+DotEnv.load!()
+
+# Get config and set up grid:
+eval_grid = EvaluationGrid(get_config_from_args())
+exper_grid = ExperimentGrid(eval_grid.grid_file)
+
+local_save_dir = get_work_dir(eval_grid, ENV["EVAL_WORK_DIR"], ENV["OUTPUT_DIR"])
+output_dir = results_dir(eval_grid)
+
+# Visualize logs:
+prefix = "logs"
+valid_y = CTExperiments.valid_y_logs(eval_grid)
+
+params = PlotParams(; colvar="generator_type", rowvar="objective")
+final_save_dir = save_dir(params, output_dir; prefix)
+for y in valid_y
+    plt = plot_errorbar_logs(eval_grid; y=y, params()...)
+    display(plt)
+    save(joinpath(final_save_dir, "$y.png"), plt; px_per_unit=3)
+end
+@info "Images stored in $final_save_dir/"
+
+# Visualize CE:
+prefix = "ce"
+all_data = CTExperiments.merge_with_meta(
+    eval_grid, CTExperiments.load_ce_evaluation(eval_grid)
+)
+valid_y = CTExperiments.valid_y_ce(all_data[1])
+
+params = PlotParams(; colvar="lambda_energy_eval", rowvar="objective")
+final_save_dir = save_dir(params, output_dir; prefix)
+for y in valid_y
+    plt = boxplot_ce(all_data...; y=y, params()..., facet=(; linkyaxes=:none))
+    display(plt)
+    save(joinpath(final_save_dir, "$y.png"), plt; px_per_unit=3)
+end
+
+# Plot images:
+plot_ce(eval_grid; save_dir=final_save_dir, byvars=["objective"])
+@info "Images stored in $final_save_dir/"

### 2025-02-06 src/training.jl

@@ -29,6 +29,7 @@ function counterfactual_training(
     mutability=nothing,
     verbose::Int=1,
     checkpoint_dir::Union{Nothing,String}=nothing,
+    callback::Union{Nothing,Function}=nothing,
     kwrgs...,
 )
 
@@ -148,6 +149,11 @@ function counterfactual_training(
             Flux.update!(opt_state, model, grads[1])
         end
 
+        if !isnothing(callback)
+            counterfactuals = reduce(hcat, [x[1] for x in counterfactual_dl])
+            callback(model, counterfactuals)
+        end
+
         # Logging:
         time_taken = time() - start
         acc = accuracy(model, train_set)
@@ -261,6 +267,7 @@ function counterfactual_training(
             @info msg_adv
             @info msg_valid
         end
+
     end
     return model, log
 end

### 2025-02-06 paper/sections/other/constraints.qmd
ok that trick works

--- 2025-02-06 paper/sections/other/constraints.qmd
@@
 # Data:
-# constraints = ["none", "both"]        # x1 cannot be mutated
-constraints = ["both", "both"]        # x2 cannot be mutated
+constraints = ["none", "both"]        # x1 cannot be mutated
+# constraints = ["both", "none"]        # x2 cannot be mutated
 data = LinearlySeparable(
@@
 ```{julia}
-obj = EnergyDifferentialObjective(lambda=[1.0,0.1,0.0])
+using CounterfactualExplanations.Convergence
+obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
 # obj = VanillaObjective()
@@
 opt_state = Flux.setup(Descent(), model)
+conv = MaxIterConvergence()
+```
+
+```{julia}
+using CTExperiments: get_log_reg_params, get_decision_boundary
+
+function plot_db(model, x)
+    coeff = get_log_reg_params(model)
+    db = get_decision_boundary(coeff)
+    plt = Plots.scatter(ce_data)
+    yhat = [argmax(y) for y in eachcol(model(x))]
+    Plots.scatter!(x[1,:], x[2,:], label="Counterfactuals", ms=15, shape=:star, color=yhat)
+    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
+    display(plt)
+end
 ```
@@
     nepochs = 50,
-    mutability = Symbol.(constraints)
+    mutability = Symbol.(constraints),
+    callback = plot_db,
+    nce=50,
+    convergence=conv, 
 )
@@
 coeff = get_log_reg_params(model)
+display(coeff)
 db = get_decision_boundary(coeff)

### 2025-02-06 paper/experiments/CTExperiments/src/plots.jl

@@ -96,6 +96,9 @@ function boxplot_ce(
     x::Union{Nothing,String}="generator_type",
     y::String="plausibility_distance_from_target",
     byvars::Union{Nothing,String,Vector{String}}=nothing,
+    colorvar::Union{Nothing,String}=nothing,
+    rowvar::Union{Nothing,String}=nothing,
+    colvar::Union{Nothing,String}="generator_type",
     kwrgs...
 )
     x = isnothing(x) ? "generator_type" : x
@@ -106,7 +109,7 @@ function boxplot_ce(
     df_agg = aggregate_ce_evaluation(df, df_meta, df_eval; y=y, byvars=byvars)
 
     # Plotting:
-    plt = boxplot_ce(df_agg,x;kwrgs...)
+    plt = boxplot_ce(df_agg,x;colorvar, rowvar, colvar, kwrgs...)
 
     return plt, df_agg
 end

### 2025-02-06 src/counterfactuals.jl

@@ -69,6 +69,19 @@ function generate!(
             ]
         ).(ces)
 
+    # Adjust for mutability:
+    mtblty = counterfactual_data.mutability
+    if !isnothing(mtblty)
+        for (i, ce) in enumerate(counterfactuals)
+            immtble = findall(mtblty .!= :both)
+            for j in immtble
+                # println(counterfactuals[i])
+                counterfactuals[i][j,:] = neighbours[i][j,:]
+                # println(counterfactuals[i])
+            end
+        end
+    end
+
     aversarial_targets = []
     targets_enc = []
     percent_valid = 0.0

### 2025-02-06 src/implausibility.jl

@@ -2,18 +2,18 @@ using EnergySamplers: EnergySamplers
 using LinearAlgebra
 
 @doc raw"""
-    implausibility(model, perturbed_input, samples, targets)
+    implausibility(model, counterfactual, samples, targets)
 
-Compute the implausibility (contrastive divergence) of the counterfactuals (`perturbed_input`) with respect to `samples` in the target class. This is computed as the difference between negative logits indexed at the target class for the `samples` and the `perturbed_input`.
+Compute the implausibility (contrastive divergence) of the counterfactuals (`counterfactual`) with respect to `samples` in the target class. This is computed as the difference between negative logits indexed at the target class for the `samples` and the `counterfactual`.
 """
-function implausibility(model, perturbed_input, samples, targets)
+function implausibility(model, counterfactual, samples, targets)
     E(x) = -model(x)                                    # energy
-    # x = ((E(samples)) - (E(perturbed_input)))'targets   # contrastive divergence
-    x = ((E(perturbed_input)) - (E(samples)))'targets   # contrastive divergence
+    x = ((E(samples)) - (E(counterfactual)))'targets   # contrastive divergence
+    # x = ((E(counterfactual)) - (E(samples)))'targets   # contrastive divergence
     return diag(x[:, :])
 end
 
-function reg_loss(model, perturbed_input, samples, targets)
-    x = (abs2.(model(samples)) + abs2.(model(perturbed_input)))'targets
+function reg_loss(model, counterfactual, samples, targets)
+    x = (abs2.(model(samples)) + abs2.(model(counterfactual)))'targets
     return diag(x[:, :])
 end

### 2025-02-06 paper/body_ecml.tex
uh

@@
 
-\subsection{Encoding Domain Knowledge}\label{encoding-domain-knowledge}
+\subsection{Encoding Actionability
+Constraints}\label{encoding-actionability-constraints}
+
+Many existing counterfactual explainers support domain and mutability
+constraints out-of-the-box. In fact, both types of constraints can be
+implemented for any counterfactual explainer that relies on gradient
+descent in the feature space for optimization
+\cite{altmeyer2023explaining}. In this context, domain constraints can
+be imposed by simply projecting counterfactuals back to the specified
+domain, if the previous gradient step moved resulted in updated feature
+values out of the domain. Mutability constraints can similarly be
+enforced by simply clipping gradients to ensure that features are only
+mutated in the allowed direction, if at all.
+
+Since both types of constraints directly affect \(\mathbf{x}^\prime\),
+they also influence model outcomes through Equation~\ref{eq-div}.
+
+Mutability constraints
 

### 2025-02-06 paper/preprint/paper.tex
uh

@@
 
-\subsection{Encoding Domain Knowledge}\label{encoding-domain-knowledge}
+\subsection{Encoding Actionability
+Constraints}\label{encoding-actionability-constraints}
+
+Many existing counterfactual explainers support domain and mutability
+constraints out-of-the-box. In fact, both types of constraints can be
+implemented for any counterfactual explainer that relies on gradient
+descent in the feature space for optimization
+(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
+In this context, domain constraints can be imposed by simply projecting
+counterfactuals back to the specified domain, if the previous gradient
+step moved resulted in updated feature values out of the domain.
+Mutability constraints can similarly be enforced by simply clipping
+gradients to ensure that features are only mutated in the allowed
+direction, if at all.
+
+Since both types of constraints directly affect \(\mathbf{x}^\prime\),
+they also influence model outcomes through Equation~\ref{eq-div}.
+
+Mutability constraints
 
@@
 \url{https://doi.org/10.1109/CVPR42600.2020.01006}.
+
+\bibitem[\citeproctext]{ref-altmeyer2023explaining}
+Altmeyer, Patrick, Arie van Deursen, et al. 2023. {``Explaining
+Black-Box Models Through Counterfactuals.''} In \emph{Proceedings of the
+JuliaCon Conferences}, 1:130. 1.
 

### 2025-02-06 paper/sections/method.qmd
vamos
uh

@@
 
-## Encoding Domain Knowledge
+## Encoding Actionability Constraints
+
+Many existing counterfactual explainers support domain and mutability constraints out-of-the-box.
+In fact, both types of constraints can be implemented for any counterfactual explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining].
+In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step moved resulted in updated feature values out of the domain.
+Mutability constraints can similarly be enforced by simply clipping gradients to ensure that features are only mutated in the allowed direction, if at all.
+
+Since both types of constraints directly affect $\mathbf{x}^\prime$, they also influence model outcomes through @eq-div.
+
+Mutability constraints

### 2025-02-07 paper/body_ecml.tex
slowly slowly

--- 2025-02-05 paper/body_ecml.tex
@@
 on average than young adults. But of course this CE does not qualify for
-providing actionable recourse to the applicant. For our intents and
-purposes, counterfactual training aims at improving model explainability
-by aligning models with counterfactuals that meet both desiderata,
-plausibility and actionability. Formally, we define explainability as
-follows:
+providing actionable recourse to the applicant since \emph{age} is not a
+mutable feature. For our intents and purposes, counterfactual training
+aims at improving model explainability by aligning models with
+counterfactuals that meet both desiderata, plausibility and
+actionability. Formally, we define explainability as follows:
 
@@
 \begin{equation}\phantomsection\label{eq-obj}{
-\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),y) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime};\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),y)
+\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y})
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-div}{
-\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
+\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
 }\end{equation}
@@
 generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
-target class \(\mathbf{y}^+ \sim p(\mathbf{y})\) and then generating a
-counterfactual explanation for that target, similar to how conditional
-sampling can be used for JEMs \cite{grathwohl2020your}. Intuitively, the
-gradient of Equation~\ref{eq-div} decreases the energy of observed
-training samples (positive samples) while at same time increasing the
-energy of counterfactuals (negative samples) \cite{du2019implicit}. As
-the generated counterfactuals get more plausible
-(Definition~\ref{def-explainability}) over the cause of training, these
-two opposing effects gradually balance each out \cite{lippe2024uvadlc}.
+target class \(y^+ \sim p(y)\) and then generating a counterfactual
+explanation for that target, similar to how conditional sampling is used
+to draw from \(p_\theta(\mathbf{x})\) in \cite{grathwohl2020your}. In
+particular, we set
+\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
+where \(y^+\) denotes the index of the target class.
+
+Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
+of observed training samples (positive samples) while at same time
+increasing the energy of counterfactuals (negative samples)
+\cite{du2019implicit}. As the generated counterfactuals get more
+plausible (Definition~\ref{def-explainability}) over the cause of
+training, these two opposing effects gradually balance each out
+\cite{lippe2024uvadlc}.
 
@@
 
-\subsection{Encoding Domain Knowledge}\label{encoding-domain-knowledge}
+\subsection{Encoding Actionability
+Constraints}\label{encoding-actionability-constraints}
+
+Many existing counterfactual explainers support domain and mutability
+constraints out-of-the-box. In fact, both types of constraints can be
+implemented for any counterfactual explainer that relies on gradient
+descent in the feature space for optimization
+\cite{altmeyer2023explaining}. In this context, domain constraints can
+be imposed by simply projecting counterfactuals back to the specified
+domain, if the previous gradient step resulted in updated feature values
+that were out-of-domain. Mutability constraints can similarly be
+enforced by setting partial derivatives to zero to ensure that features
+are only mutated in the allowed direction, if at all.
+
+Since actionability constraints are binding at test time, we should also
+impose them when generating \(\mathbf{x}^\prime\) during each training
+iteration to align model representations with user requirements. Through
+their effect on \(\mathbf{x}^\prime\), both types of constraints
+influence model outcomes through Equation~\ref{eq-div}. This is best
+explained through a simple example involving a linear classifier:
+\(\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}\). Then for
+\(y^+ \sim p(y)\) we have:
+
+\begin{equation}\phantomsection\label{eq-grad}{
+\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
+}\end{equation}
+
+In other words, contrastive divergence will push down (up) on the
+coefficient \(\theta[d]\) corresponding to a feature \(d\) that is lower
+(higher) for counterfactuals than samples. This has the intended effect
+of aligning counterfactual explanations with the data, as discussed
+above.
+
+Crucially though, we want to avoid penalizing implausibility that arises
+due to mutability constraints. Suppose, for example, that feature \(d\)
+represents \emph{age} and our model predicts the likelihood of credit
+default.
 

### 2025-02-07 paper/preprint/paper.tex
slowly slowly

@@
 generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
-target class \(\mathbf{y}^+ \sim p(\mathbf{y})\) and then generating a
-counterfactual explanation for that target, similar to how conditional
-sampling can be used for JEMs
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). Intuitively,
-the gradient of Equation~\ref{eq-div} decreases the energy of observed
-training samples (positive samples) while at same time increasing the
-energy of counterfactuals (negative samples)
+target class \(y^+ \sim p(y)\) and then generating a counterfactual
+explanation for that target, similar to how conditional sampling is used
+to draw from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
+(\citeproc{ref-grathwohl2020your}{2020}). In particular, we set
+\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
+where \(y^+\) denotes the index of the target class.
+
+Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
+of observed training samples (positive samples) while at same time
+increasing the energy of counterfactuals (negative samples)
 (\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
@@
 counterfactuals back to the specified domain, if the previous gradient
-step moved resulted in updated feature values out of the domain.
-Mutability constraints can similarly be enforced by simply clipping
-gradients to ensure that features are only mutated in the allowed
-direction, if at all.
-
-Since both types of constraints directly affect \(\mathbf{x}^\prime\),
-they also influence model outcomes through Equation~\ref{eq-div}.
-
-Mutability constraints
+step resulted in updated feature values that were out-of-domain.
+Mutability constraints can similarly be enforced by setting partial
+derivatives to zero to ensure that features are only mutated in the
+allowed direction, if at all.
+
+Since actionability constraints are binding at test time, we should also
+impose them when generating \(\mathbf{x}^\prime\) during each training
+iteration to align model representations with user requirements. Through
+their effect on \(\mathbf{x}^\prime\), both types of constraints
+influence model outcomes through Equation~\ref{eq-div}. This is best
+explained through a simple example involving a linear classifier:
+\(\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}\). Then for
+\(y^+ \sim p(y)\) we have:
+
+\begin{equation}\phantomsection\label{eq-grad}{
+\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
+}\end{equation}
+
+In other words, contrastive divergence will push down (up) on the
+coefficient \(\theta[d]\) corresponding to a feature \(d\) that is lower
+(higher) for counterfactuals than samples. This has the intended effect
+of aligning counterfactual explanations with the data, as discussed
+above.
+
+Crucially though, we want to avoid penalizing implausibility that arises
+due to mutability constraints. Suppose, for example, that feature \(d\)
+represents \emph{age} and our model predicts the likelihood of credit
+default.
 
@@
 \renewcommand{\thefigure}{A\arabic{figure}}
+
+\section{Notation}\label{notation}
+
+\begin{itemize}
+\tightlist
+\item
+  \(y^+\): The target class and also the index of the target class.
+\item
+  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
+  class.
+\end{itemize}
 

### 2025-02-07 paper/sections/method.qmd
slowly slowly

--- 2025-02-05 paper/sections/method.qmd
@@
 Ignoring all other features, this complies with the definition of plausibility if 40-year-old individuals are in fact more credit-worthy on average than young adults.
-But of course this CE does not qualify for providing actionable recourse to the applicant.
+But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a mutable feature.
 For our intents and purposes, counterfactual training aims at improving model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability.
@@
 $$
-\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),y) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime};\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),y)
+\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y})
 $$ {#eq-obj}
@@
 $$
-\text{div}(\mathbf{x},\mathbf{x^\prime};\theta) = \mathcal{E}_\theta(\mathbf{x}) - \mathcal{E}_\theta(\mathbf{x}^\prime)
+\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
 $$ {#eq-div}
@@
 where $\mathcal{E}_\theta(\cdot)$ denotes the energy function.
-We generate samples $\mathbf{x}^\prime$ by first randomly sampling the target class $\mathbf{y}^+ \sim p(\mathbf{y})$ and then generating a counterfactual explanation for that target, similar to how conditional sampling can be used for JEMs [@grathwohl2020your].
+We generate samples $\mathbf{x}^\prime$ by first randomly sampling the target class $y^+ \sim p(y)$ and then generating a counterfactual explanation for that target, similar to how conditional sampling is used to draw from $p_\theta(\mathbf{x})$ in @grathwohl2020your.
+In particular, we set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the target class.
+
 Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit].
@@
 
-## Encoding Domain Knowledge
+## Encoding Actionability Constraints
+
+Many existing counterfactual explainers support domain and mutability constraints out-of-the-box.
+In fact, both types of constraints can be implemented for any counterfactual explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining].
+In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain.
+Mutability constraints can similarly be enforced by setting partial derivatives to zero to ensure that features are only mutated in the allowed direction, if at all.
+
+Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to align model representations with user requirements.
+Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div.
+This is best explained through a simple example involving a linear classifier: $\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}$.
+Then for $y^+ \sim p(y)$ we have:
+
+$$
+\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
+$$ {#eq-grad}
+
+In other words, contrastive divergence will push down (up) on the coefficient $\theta[d]$ corresponding to a feature $d$ that is lower (higher) for counterfactuals than samples.
+This has the intended effect of aligning counterfactual explanations with the data, as discussed above.
+
+Crucially though, we want to avoid penalizing implausibility that arises due to mutability constraints.
+Suppose, for example, that feature $d$ represents *age* and our model predicts the likelihood of credit default.

### 2025-02-10 paper/preprint/paper.tex
progress

@@
 
-\subsection{Encoding Actionability
-Constraints}\label{encoding-actionability-constraints}
+\subsection{Encoding Actionability Constraints}\label{sec-constraints}
 
@@
 their effect on \(\mathbf{x}^\prime\), both types of constraints
-influence model outcomes through Equation~\ref{eq-div}. This is best
-explained through a simple example involving a linear classifier:
-\(\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}\). Then for
-\(y^+ \sim p(y)\) we have:
-
-\begin{equation}\phantomsection\label{eq-grad}{
-\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
-}\end{equation}
-
-In other words, contrastive divergence will push down (up) on the
-coefficient \(\theta[d]\) corresponding to a feature \(d\) that is lower
-(higher) for counterfactuals than samples. This has the intended effect
-of aligning counterfactual explanations with the data, as discussed
-above.
-
-Crucially though, we want to avoid penalizing implausibility that arises
-due to mutability constraints. Suppose, for example, that feature \(d\)
-represents \emph{age} and our model predicts the likelihood of credit
-default.
+influence model outcomes through Equation~\ref{eq-div}. Here it is
+crucial that avoid penalizing implausibility that arises due to
+mutability constraints. For any mutability-constrained feature \(d\)
+this can be achieved by enforcing
+\(\mathbf{x}^\prime[d] = \mathbf{x}[d]\) whenever the constraint is
+violated, by adjusting either one to the other. Setting
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) intuitively corresponds to
+imposing user requirements on the target domain; from a Bayesian
+perspective it can be understood as assuming a point mass prior for
+\(p(\mathbf{x})\) with respect to feature \(d\)\footnote{For additional
+  detail, see Section~\ref{sec-app-constraints} of the supplementary
+  appendix.}.
 
@@
   class.
+\item
+  \(\theta\): Model parameters (unspecified).
+\item
+  \(\Theta\): Matrix of parameters.
 \end{itemize}
@@
 \section{Training Details}\label{training-details}
+
+\subsection{Details on Encoding Mutability
+Constraints}\label{sec-app-constraints}
+
+In Section~\ref{sec-constraints} we explained that to avoid penalizing
+implausibility that arises due to mutability constraints for any feature
+\(d\), we impose a point mass prior on \(p(\mathbf{x})\) for that
+features. This is best explained through a simple example involving a
+binary linear classifier.
+
+Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\), with
+\(y \in \{1,2\}\). Then for \(y^+ \sim p(y)=2\) we have:
+
+\begin{equation}\phantomsection\label{eq-grad}{
+\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \begin{pmatrix} 0 & 0 \\ \mathbf{x}^\prime[1] - \mathbf{x}[1] & \mathbf{x}^\prime[2] - \mathbf{x}[2] \end{pmatrix}
+}\end{equation}
+
+Contrastive divergence pushes down (up) on the coefficient
+\(\Theta[y^+,d]\) by the amount that feature \(d\) that is lower
+(higher) for counterfactuals than samples. This has the intended effect
+of aligning counterfactual explanations with the data, as discussed
+above.
+
+Suppose, for example, that feature \(d=1\) represents \emph{age} and our
+model predicts the likelihood of credit default:
+\(y \in \{1:=\text{'no default'},2:=\text{'default'}\}\). Then if
+\(\mathbf{x}^\prime[1] < \mathbf{x}[1]\), contrastive divergence will
+push down on the corresponding coefficient,
+\(\Theta[\text{'default'},\text{'age'}]\). Intuitively, this has the
+effect of decreasing the effect of \emph{age} that the model attributes
+to the log odds of default, i.e.~the model learns to higher default risk
+younger individuals. But if \emph{age} is immutable, then this
+corresponds to penalizing younger individuals for implausibility that
+arises due to a feature that they can change about themselves. This is
+precisely what we want to avoid, and therefore we impose no penalty at
+all.
+
+\subsection{Gradient}\label{gradient}
+
+\subsection{Full Training Loop}\label{full-training-loop}
+
+Next, we defined the training objective and the counterfactual
+generator. Finally, we initialize the optimization state.
+
+Finally, we visualize the results:
+
+\section{Detailed Results}\label{detailed-results}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{cost}}\): \texttt{0.0,\ 0.001,\ 0.1}
-  \item
-    \(\lambda_{\text{div}}\):
-    \texttt{0.01,\ 0.05,\ 0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 15.0}
-  \item
-    Learning Rate: \texttt{1.0}
-  \item
-    Maximum Iterations: \texttt{20,\ 50,\ 100}
-  \item
-    Optimizerimizer: \texttt{sgd}
-  \end{itemize}
-\item
-  Generator: \texttt{ecco,\ generic,\ omni,\ revise}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
 \end{exercise}
@@
 ~
-
-\begin{itemize}
-\tightlist
-\item
-  Counterfactual Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Convergence: \texttt{max\_iter}
-  \item
-    Maximum Iterations: \texttt{100}
-  \item
-    No.~Individuals: \texttt{100}
-  \item
-    No.~Runs: \texttt{5}
-  \end{itemize}
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{cost}}\): \texttt{0.0}
-  \item
-    \(\lambda_{\text{div}}\):
-    \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 20.0}
-  \item
-    Learning Rate: \texttt{1.0}
-  \item
-    Maximum Iterations: \texttt{50}
-  \item
-    Optimizerimizer: \texttt{sgd}
-  \end{itemize}
-\end{itemize}
 
@@
 
-\begin{longtable}{ccccc}
+\begin{table}
 
@@
 
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0.01 & \textit{ECCo} & $-9.91 \cdot 10^{11}$ & $2.25 \cdot 10^{12}$ \\
-  full & 0.01 & \textit{Generic} & $-5.71 \cdot 10^{17}$ & $1.3 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.116}} \\
-  full & 0.01 & \textit{REVISE} & -15.6 & 13.2 \\
-  vanilla & 0.01 & \textit{ECCo} & -4.28 & 3.52 \\
-  vanilla & 0.01 & \textit{Generic} & -4.45 & 3.47 \\
-  vanilla & 0.01 & \textit{Omniscient} & -5.12 & 4.46 \\
-  vanilla & 0.01 & \textit{REVISE} & -4.91 & 4.24 \\
-  full & 0.05 & \textit{ECCo} & $-5.63 \cdot 10^{5}$ & $1.28 \cdot 10^{6}$ \\
-  full & 0.05 & \textit{Generic} & $-8.35 \cdot 10^{17}$ & $1.9 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.114}} \\
-  full & 0.05 & \textit{REVISE} & -15 & 12.6 \\
-  vanilla & 0.05 & \textit{ECCo} & -4.4 & 3.66 \\
-  vanilla & 0.05 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 0.05 & \textit{Omniscient} & -5.25 & 4.62 \\
-  vanilla & 0.05 & \textit{REVISE} & -4.94 & 4.22 \\
-  full & 0.1 & \textit{ECCo} & $-6.74 \cdot 10^{5}$ & $1.53 \cdot 10^{6}$ \\
-  full & 0.1 & \textit{Generic} & $-1.72 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.56}} & \color{blue}{\textbf{0.124}} \\
-  full & 0.1 & \textit{REVISE} & -15.6 & 13.2 \\
-  vanilla & 0.1 & \textit{ECCo} & -4.28 & 3.52 \\
-  vanilla & 0.1 & \textit{Generic} & -4.45 & 3.48 \\
-  vanilla & 0.1 & \textit{Omniscient} & -5.12 & 4.46 \\
-  vanilla & 0.1 & \textit{REVISE} & -4.91 & 4.25 \\
-  full & 0.5 & \textit{ECCo} & -11.8 & 9.83 \\
-  full & 0.5 & \textit{Generic} & $-1.06 \cdot 10^{18}$ & $2.42 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.123}} \\
-  full & 0.5 & \textit{REVISE} & -15 & 12.6 \\
-  vanilla & 0.5 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 0.5 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 0.5 & \textit{Omniscient} & -5.25 & 4.61 \\
-  vanilla & 0.5 & \textit{REVISE} & -4.95 & 4.22 \\
-  full & 1 & \textit{ECCo} & -11.5 & 11.1 \\
-  full & 1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.88 \cdot 10^{11}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.59}} & \color{blue}{\textbf{0.117}} \\
-  full & 1 & \textit{REVISE} & -15.7 & 13.3 \\
-  vanilla & 1 & \textit{ECCo} & -4.28 & 3.51 \\
-  vanilla & 1 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 1 & \textit{Omniscient} & -5.11 & 4.46 \\
-  vanilla & 1 & \textit{REVISE} & -4.91 & 4.25 \\
-  full & 5 & \textit{ECCo} & -3.99 & 3.12 \\
-  full & 5 & \textit{Generic} & $-4.88 \cdot 10^{17}$ & $1.11 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.117}} \\
-  full & 5 & \textit{REVISE} & -14.6 & 12.1 \\
-  vanilla & 5 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 5 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 5 & \textit{Omniscient} & -5.25 & 4.61 \\
-  vanilla & 5 & \textit{REVISE} & -4.95 & 4.22 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.31}} & \color{blue}{\textbf{0.735}} \\
-  full & 10 & \textit{Generic} & $-1.7 \cdot 10^{11}$ & $3.86 \cdot 10^{11}$ \\
-  full & 10 & \textit{Omniscient} & -2.53 & 0.117 \\
-  full & 10 & \textit{REVISE} & -15.5 & 13 \\
-  vanilla & 10 & \textit{ECCo} & -4.28 & 3.51 \\
-  vanilla & 10 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 10 & \textit{Omniscient} & -5.12 & 4.46 \\
-  vanilla & 10 & \textit{REVISE} & -4.91 & 4.24 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.01}} & \color{blue}{\textbf{0.488}} \\
-  full & 15 & \textit{Generic} & $-4.91 \cdot 10^{17}$ & $1.12 \cdot 10^{18}$ \\
-  full & 15 & \textit{Omniscient} & -2.53 & 0.116 \\
-  full & 15 & \textit{REVISE} & -14.4 & 11.7 \\
-  vanilla & 15 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 15 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 15 & \textit{Omniscient} & -5.25 & 4.6 \\
-  vanilla & 15 & \textit{REVISE} & -4.95 & 4.23 \\\bottomrule
-
-\end{longtable}
-
-\begin{longtable}{ccccc}
+\centering{
+
+}
+
+\end{table}%
+
+\begin{table}
 
@@
 
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0 & \textit{ECCo} & $-5.32 \cdot 10^{3}$ & $1.21 \cdot 10^{4}$ \\
-  full & 0 & \textit{Generic} & $-1.03 \cdot 10^{18}$ & $2.34 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.64}} & \color{blue}{\textbf{0.125}} \\
-  full & 0 & \textit{REVISE} & -15.4 & 12.9 \\
-  vanilla & 0 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0 & \textit{Omniscient} & -5.18 & 4.54 \\
-  vanilla & 0 & \textit{REVISE} & -4.93 & 4.23 \\
-  full & 0.001 & \textit{ECCo} & -362 & 811 \\
-  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.03 \cdot 10^{17}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.49}} & \color{blue}{\textbf{0.115}} \\
-  full & 0.001 & \textit{REVISE} & -15.5 & 13 \\
-  vanilla & 0.001 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0.001 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0.001 & \textit{Omniscient} & -5.18 & 4.53 \\
-  vanilla & 0.001 & \textit{REVISE} & -4.93 & 4.23 \\
-  full & 0.1 & \textit{ECCo} & $-3.72 \cdot 10^{11}$ & $8.46 \cdot 10^{11}$ \\
-  full & 0.1 & \textit{Generic} & $-4.49 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5}} & \color{blue}{\textbf{0.112}} \\
-  full & 0.1 & \textit{REVISE} & -14.6 & 12.2 \\
-  vanilla & 0.1 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0.1 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0.1 & \textit{Omniscient} & -5.18 & 4.54 \\
-  vanilla & 0.1 & \textit{REVISE} & -4.93 & 4.24 \\\bottomrule
-
-\end{longtable}
+\centering{
+
+}
+
+\end{table}%
 
@@
 
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0.01 & \textit{ECCo} & $-2.8 \cdot 10^{22}$ & $6.39 \cdot 10^{22}$ \\
-  full & 0.01 & \textit{Generic} & $-4.89 \cdot 10^{30}$ & $1.11 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.74}} & \color{blue}{\textbf{5.08}} \\
-  full & 0.01 & \textit{REVISE} & -572 & $1.25 \cdot 10^{3}$ \\
-  vanilla & 0.01 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 0.01 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.01 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 0.01 & \textit{REVISE} & -11.2 & 13 \\
-  full & 0.05 & \textit{ECCo} & $-1.55 \cdot 10^{16}$ & $3.52 \cdot 10^{16}$ \\
-  full & 0.05 & \textit{Generic} & $-2.22 \cdot 10^{20}$ & $5 \cdot 10^{20}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.41}} & \color{blue}{\textbf{4.48}} \\
-  full & 0.05 & \textit{REVISE} & $-1.04 \cdot 10^{3}$ & $2.3 \cdot 10^{3}$ \\
-  vanilla & 0.05 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.05 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 0.05 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 0.05 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 0.1 & \textit{ECCo} & $-3.41 \cdot 10^{3}$ & $7.73 \cdot 10^{3}$ \\
-  full & 0.1 & \textit{Generic} & $-5.22 \cdot 10^{30}$ & $1.19 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.78}} & \color{blue}{\textbf{5.12}} \\
-  full & 0.1 & \textit{REVISE} & -288 & 594 \\
-  vanilla & 0.1 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.1 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 0.1 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 0.5 & \textit{ECCo} & -7.09 & 7.51 \\
-  full & 0.5 & \textit{Generic} & $-1.11 \cdot 10^{31}$ & $2.53 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.58}} & \color{blue}{\textbf{4.83}} \\
-  full & 0.5 & \textit{REVISE} & $-1.19 \cdot 10^{3}$ & $2.64 \cdot 10^{3}$ \\
-  vanilla & 0.5 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.5 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 0.5 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 0.5 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 1 & \textit{ECCo} & -6.06 & 6.33 \\
-  full & 1 & \textit{Generic} & $-1.58 \cdot 10^{33}$ & $3.59 \cdot 10^{33}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.66}} & \color{blue}{\textbf{4.89}} \\
-  full & 1 & \textit{REVISE} & $-1.16 \cdot 10^{3}$ & $2.59 \cdot 10^{3}$ \\
-  vanilla & 1 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 1 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 1 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.57}} & \color{blue}{\textbf{2.07}} \\
-  full & 5 & \textit{Generic} & $-1.17 \cdot 10^{28}$ & $2.66 \cdot 10^{28}$ \\
-  full & 5 & \textit{Omniscient} & -4.29 & 4.31 \\
-  full & 5 & \textit{REVISE} & -530 & $1.16 \cdot 10^{3}$ \\
-  vanilla & 5 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 5 & \textit{Generic} & -11.7 & 12.7 \\
-  vanilla & 5 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 5 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.76}} & \color{blue}{\textbf{0.974}} \\
-  full & 10 & \textit{Generic} & $-1.54 \cdot 10^{33}$ & $3.51 \cdot 10^{33}$ \\
-  full & 10 & \textit{Omniscient} & -4.44 & 4.56 \\
-  full & 10 & \textit{REVISE} & $-1.52 \cdot 10^{3}$ & $3.4 \cdot 10^{3}$ \\
-  vanilla & 10 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 10 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 10 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 10 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.37}} & \color{blue}{\textbf{0.365}} \\
-  full & 15 & \textit{Generic} & $-5.32 \cdot 10^{28}$ & $1.21 \cdot 10^{29}$ \\
-  full & 15 & \textit{Omniscient} & -4.34 & 4.38 \\
-  full & 15 & \textit{REVISE} & -473 & $1.03 \cdot 10^{3}$ \\
-  vanilla & 15 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 15 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 15 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 15 & \textit{REVISE} & -11.3 & 13.1 \\\bottomrule
-
-\end{longtable}
+\centering{
+
+}
+
+\end{table}%
 
@@
 
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.26}} & \color{blue}{\textbf{0.423}} \\
-  full & 0.01 & \textit{Generic} & -1.49 & 0.71 \\
-  full & 0.01 & \textit{Omniscient} & -5.21 & 5.25 \\
-  full & 0.01 & \textit{REVISE} & $-2.71 \cdot 10^{26}$ & $6.37 \cdot 10^{26}$ \\
-  vanilla & 0.01 & \textit{ECCo} & -9.33 & 7.34 \\
-  vanilla & 0.01 & \textit{Generic} & -8.89 & 6.88 \\
-  vanilla & 0.01 & \textit{Omniscient} & -8.67 & 6.87 \\
-  vanilla & 0.01 & \textit{REVISE} & -8.65 & 6.8 \\
-  full & 0.05 & \textit{ECCo} & -1.29 & 0.397 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.356}} \\
-  full & 0.05 & \textit{Omniscient} & -5.08 & 5.09 \\
-  full & 0.05 & \textit{REVISE} & $-5.91 \cdot 10^{27}$ & $1.36 \cdot 10^{28}$ \\
-  vanilla & 0.05 & \textit{ECCo} & -9.35 & 7.32 \\
-  vanilla & 0.05 & \textit{Generic} & -8.85 & 6.87 \\
-  vanilla & 0.05 & \textit{Omniscient} & -8.7 & 6.96 \\
-  vanilla & 0.05 & \textit{REVISE} & -8.52 & 6.76 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.2}} & \color{blue}{\textbf{0.383}} \\
-  full & 0.1 & \textit{Generic} & -1.5 & 0.735 \\
-  full & 0.1 & \textit{Omniscient} & -5.17 & 5.23 \\
-  full & 0.1 & \textit{REVISE} & $-3.06 \cdot 10^{26}$ & $7.7 \cdot 10^{26}$ \\
-  vanilla & 0.1 & \textit{ECCo} & -9.33 & 7.32 \\
-  vanilla & 0.1 & \textit{Generic} & -8.88 & 6.86 \\
-  vanilla & 0.1 & \textit{Omniscient} & -8.69 & 6.9 \\
-  vanilla & 0.1 & \textit{REVISE} & -8.68 & 6.81 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.12}} & \color{blue}{\textbf{0.217}} \\
-  full & 0.5 & \textit{Generic} & -1.21 & 0.352 \\
-  full & 0.5 & \textit{Omniscient} & -5.09 & 5.12 \\
-  full & 0.5 & \textit{REVISE} & $-5.97 \cdot 10^{27}$ & $1.37 \cdot 10^{28}$ \\
-  vanilla & 0.5 & \textit{ECCo} & -9.35 & 7.3 \\
-  vanilla & 0.5 & \textit{Generic} & -8.89 & 6.92 \\
-  vanilla & 0.5 & \textit{Omniscient} & -8.68 & 6.93 \\
-  vanilla & 0.5 & \textit{REVISE} & -8.53 & 6.75 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.1}} & \color{blue}{\textbf{0.163}} \\
-  full & 1 & \textit{Generic} & -1.49 & 0.726 \\
-  full & 1 & \textit{Omniscient} & -5.16 & 5.2 \\
-  full & 1 & \textit{REVISE} & $-3.09 \cdot 10^{26}$ & $7.22 \cdot 10^{26}$ \\
-  vanilla & 1 & \textit{ECCo} & -9.34 & 7.36 \\
-  vanilla & 1 & \textit{Generic} & -8.86 & 6.85 \\
-  vanilla & 1 & \textit{Omniscient} & -8.7 & 6.9 \\
-  vanilla & 1 & \textit{REVISE} & -8.69 & 6.85 \\
-  full & 5 & \textit{ECCo} & -1.75 & 0.154 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.363}} \\
-  full & 5 & \textit{Omniscient} & -5.14 & 5.16 \\
-  full & 5 & \textit{REVISE} & $-1.1 \cdot 10^{28}$ & $2.5 \cdot 10^{28}$ \\
-  vanilla & 5 & \textit{ECCo} & -9.36 & 7.32 \\
-  vanilla & 5 & \textit{Generic} & -8.88 & 6.91 \\
-  vanilla & 5 & \textit{Omniscient} & -8.7 & 6.93 \\
-  vanilla & 5 & \textit{REVISE} & -8.52 & 6.73 \\
-  full & 10 & \textit{ECCo} & $-1.02 \cdot 10^{6}$ & $2.32 \cdot 10^{6}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.49}} & \color{blue}{\textbf{0.702}} \\
-  full & 10 & \textit{Omniscient} & -5.13 & 5.16 \\
-  full & 10 & \textit{REVISE} & $-3.74 \cdot 10^{26}$ & $9.09 \cdot 10^{26}$ \\
-  vanilla & 10 & \textit{ECCo} & -9.31 & 7.33 \\
-  vanilla & 10 & \textit{Generic} & -8.87 & 6.86 \\
-  vanilla & 10 & \textit{Omniscient} & -8.7 & 6.89 \\
-  vanilla & 10 & \textit{REVISE} & -8.69 & 6.83 \\
-  full & 15 & \textit{ECCo} & $-3.31 \cdot 10^{13}$ & $7.54 \cdot 10^{13}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.22}} & \color{blue}{\textbf{0.37}} \\
-  full & 15 & \textit{Omniscient} & -5.2 & 5.23 \\
-  full & 15 & \textit{REVISE} & $-9.01 \cdot 10^{27}$ & $2.06 \cdot 10^{28}$ \\
-  vanilla & 15 & \textit{ECCo} & -9.38 & 7.34 \\
-  vanilla & 15 & \textit{Generic} & -8.86 & 6.87 \\
-  vanilla & 15 & \textit{Omniscient} & -8.69 & 6.96 \\
-  vanilla & 15 & \textit{REVISE} & -8.51 & 6.73 \\\bottomrule
-
-\end{longtable}
+\centering{
+
+}
+
+\end{table}%
 

### 2025-02-10 paper/sections/method.qmd
progress

@@
 
-## Encoding Actionability Constraints
+## Encoding Actionability Constraints {#sec-constraints}
 
@@
 Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div.
-This is best explained through a simple example involving a linear classifier: $\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}$.
-Then for $y^+ \sim p(y)$ we have:
-
-$$
-\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
-$$ {#eq-grad}
-
-In other words, contrastive divergence will push down (up) on the coefficient $\theta[d]$ corresponding to a feature $d$ that is lower (higher) for counterfactuals than samples.
-This has the intended effect of aligning counterfactual explanations with the data, as discussed above.
-
-Crucially though, we want to avoid penalizing implausibility that arises due to mutability constraints.
-Suppose, for example, that feature $d$ represents *age* and our model predicts the likelihood of credit default.
+Here it is crucial that avoid penalizing implausibility that arises due to mutability constraints.
+For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}^\prime[d] = \mathbf{x}[d]$ whenever the constraint is violated, by adjusting either one to the other.
+Setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ intuitively corresponds to imposing user requirements on the target domain; from a Bayesian perspective it can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$^[For additional detail, see @sec-app-constraints of the supplementary appendix.].

### 2025-02-10 paper/sections/other/constraints.qmd
progress

@@
+---
+execute: 
+  eval: false
+---
+
 ```{julia}
@@
 
-# Linear Model
+## Details on Encoding Mutability Constraints {#sec-app-constraints}
 
-First, we define the mutability constraints and load the data and model:
+In @sec-constraints we explained that to avoid penalizing implausibility that arises due to mutability constraints for any feature $d$, we impose a point mass prior on $p(\mathbf{x})$ for that features.
+This is best explained through a simple example involving a binary linear classifier.
+
+:::{exm-constraints}
+
+Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$, with $y \in \{1,2\}$. Then for $y^+ \sim p(y)=2$ we have:
+
+$$
+\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \begin{pmatrix} 0 & 0 \\ \mathbf{x}^\prime[1] - \mathbf{x}[1] & \mathbf{x}^\prime[2] - \mathbf{x}[2] \end{pmatrix}
+$$ {#eq-grad}
+
+Contrastive divergence pushes down (up) on the coefficient $\Theta[y^+,d]$ by the amount that feature $d$ that is lower (higher) for counterfactuals than samples.
+This has the intended effect of aligning counterfactual explanations with the data, as discussed above.
+
+Suppose, for example, that feature $d=1$ represents *age* and our model predicts the likelihood of credit default: $y \in \{1:=\text{'no default'},2:=\text{'default'}\}$.
+Then if $\mathbf{x}^\prime[1] < \mathbf{x}[1]$, contrastive divergence will push down on the corresponding coefficient, $\Theta[\text{'default'},\text{'age'}]$.
+Intuitively, this has the effect of decreasing the effect of *age* that the model attributes to the log odds of default, i.e.
+the model learns to higher default risk younger individuals.
+But if *age* is immutable, then this corresponds to penalizing younger individuals for implausibility that arises due to a feature that they can change about themselves.
+This is precisely what we want to avoid, and therefore we impose no penalty at all.
+
+:::
 
@@
 
+## Gradient
+
+```{julia}
+X1 = ce_data.X[:,ce_data.output_encoder.labels.==1]
+X2 = ce_data.X[:,ce_data.output_encoder.labels.==2]
+factual = X1[:,1]
+target = 2
+target_enc = Flux.onehotbatch([target], [1,2])
+neighbor = X2[:,1]
+```
+
+## Full Training Loop
+
 Next, we defined the training objective and the counterfactual generator. Finally, we initialize the optimization state.
@@
 obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
-# obj = VanillaObjective()
 generator = GenericGenerator()
@@
 ```
-
-Then we train the model:
 

### 2025-02-10 src/implausibility.jl

@@ -7,8 +7,8 @@ using LinearAlgebra
 Compute the implausibility (contrastive divergence) of the counterfactuals (`counterfactual`) with respect to `samples` in the target class. This is computed as the difference between negative logits indexed at the target class for the `samples` and the `counterfactual`.
 """
 function implausibility(model, counterfactual, samples, targets)
-    E(x) = -model(x)                                    # energy
-    x = ((E(samples)) - (E(counterfactual)))'targets   # contrastive divergence
+    E(x) = -model(x)                                            # energy
+    x = ((E(samples)) - (E(counterfactual)))[:,:]'targets       # contrastive divergence
     # x = ((E(counterfactual)) - (E(samples)))'targets   # contrastive divergence
     return diag(x[:, :])
 end

### 2025-02-10 paper/body_ecml.tex
pfhfhhf

@@
 their effect on \(\mathbf{x}^\prime\), both types of constraints
-influence model outcomes through Equation~\ref{eq-div}. This is best
-explained through a simple example involving a linear classifier:
-\(\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}\). Then for
-\(y^+ \sim p(y)\) we have:
+influence model outcomes through Equation~\ref{eq-div}. Here it is
+crucial that we avoid penalizing implausibility that arises due to
+mutability constraints. For any mutability-constrained feature \(d\)
+this can be achieved by enforcing
+\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
+\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
+violate mutability constraints. Specifically, we set
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if
 
-\begin{equation}\phantomsection\label{eq-grad}{
-\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
-}\end{equation}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Feature \(d\) is strictly immutable in practice.
+\item
+  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\) but feature \(d\) can
+  only be decreased in practice.
+\item
+  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\) but feature \(d\) can
+  only be increased in practice.
+\end{enumerate}
 
-In other words, contrastive divergence will push down (up) on the
-coefficient \(\theta[d]\) corresponding to a feature \(d\) that is lower
-(higher) for counterfactuals than samples. This has the intended effect
-of aligning counterfactual explanations with the data, as discussed
-above.
-
-Crucially though, we want to avoid penalizing implausibility that arises
-due to mutability constraints. Suppose, for example, that feature \(d\)
-represents \emph{age} and our model predicts the likelihood of credit
-default.
+From a Bayesian perspective, setting
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
+a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
+Intuitively, we think of this simply in terms ignoring implausibility
+costs with respect to immutable features, which effectively forces the
+model to instead seek plausibility with respect to the remaining
+features. This in turn results in lower overall sensitivity to immutable
+features, which we demonstrate empirically bellow and analytically for
+the linear case in \textbf{?@sec-app-constraints} of the supplementary
+appendix.
 

### 2025-02-10 paper/preprint/paper.tex
pfhfhhf

@@
 \makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
+\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
+\definecolor{quarto-callout-color}{HTML}{909090}
+\definecolor{quarto-callout-note-color}{HTML}{0758E5}
+\definecolor{quarto-callout-important-color}{HTML}{CC1914}
+\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
+\definecolor{quarto-callout-tip-color}{HTML}{00A047}
+\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
+\definecolor{quarto-callout-color-frame}{HTML}{acacac}
+\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
+\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
+\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
+\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
+\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \newtheorem{exercise}{Parameters}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{definition}
@@
 influence model outcomes through Equation~\ref{eq-div}. Here it is
-crucial that avoid penalizing implausibility that arises due to
+crucial that we avoid penalizing implausibility that arises due to
 mutability constraints. For any mutability-constrained feature \(d\)
 this can be achieved by enforcing
-\(\mathbf{x}^\prime[d] = \mathbf{x}[d]\) whenever the constraint is
-violated, by adjusting either one to the other. Setting
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) intuitively corresponds to
-imposing user requirements on the target domain; from a Bayesian
-perspective it can be understood as assuming a point mass prior for
-\(p(\mathbf{x})\) with respect to feature \(d\)\footnote{For additional
-  detail, see Section~\ref{sec-app-constraints} of the supplementary
-  appendix.}.
+\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
+\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
+violate mutability constraints. Specifically, we set
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Feature \(d\) is strictly immutable in practice.
+\item
+  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\) but feature \(d\) can
+  only be decreased in practice.
+\item
+  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\) but feature \(d\) can
+  only be increased in practice.
+\end{enumerate}
+
+From a Bayesian perspective, setting
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
+a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
+Intuitively, we think of this simply in terms ignoring implausibility
+costs with respect to immutable features, which effectively forces the
+model to instead seek plausibility with respect to the remaining
+features. This in turn results in lower overall sensitivity to immutable
+features, which we demonstrate empirically bellow and analytically for
+the linear case in Section~\ref{sec-app-constraints} of the
+supplementary appendix.
 
@@
 Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\), with
-\(y \in \{1,2\}\). Then for \(y^+ \sim p(y)=2\) we have:
+\(\mathbf{x} \in \mathbb{R}^2\) and \(y \in \{0,1\}\). Then for any
+given counterfactual with \(y^+ \sim p(y)=1\) we have:
 
@@
 
-Contrastive divergence pushes down (up) on the coefficient
-\(\Theta[y^+,d]\) by the amount that feature \(d\) that is lower
-(higher) for counterfactuals than samples. This has the intended effect
-of aligning counterfactual explanations with the data, as discussed
-above.
-
-Suppose, for example, that feature \(d=1\) represents \emph{age} and our
-model predicts the likelihood of credit default:
-\(y \in \{1:=\text{'no default'},2:=\text{'default'}\}\). Then if
+For the given counterfactual, contrastive divergence pushes down (up) on
+the coefficient \(\Theta[y^+,d]\) by the amount that feature \(d\) that
+is lower (higher) for counterfactuals than samples. This has the
+intended effect of aligning counterfactual explanations with the data,
+as discussed above.
+
+\begin{example}[]\protect\hypertarget{exm-grad}{}\label{exm-grad}
+
+Let feature \(d=1\) in Equation~\ref{eq-grad} represents \emph{age} and
+our model predicts the likelihood of credit default:
+\(y \in \{0:=\text{'no default'},1:=\text{'default'}\}\). Then if
 \(\mathbf{x}^\prime[1] < \mathbf{x}[1]\), contrastive divergence will
@@
 effect of decreasing the effect of \emph{age} that the model attributes
-to the log odds of default, i.e.~the model learns to higher default risk
-younger individuals. But if \emph{age} is immutable, then this
-corresponds to penalizing younger individuals for implausibility that
-arises due to a feature that they can change about themselves. This is
-precisely what we want to avoid, and therefore we impose no penalty at
-all.
-
-\subsection{Gradient}\label{gradient}
-
-\subsection{Full Training Loop}\label{full-training-loop}
-
-Next, we defined the training objective and the counterfactual
-generator. Finally, we initialize the optimization state.
-
-Finally, we visualize the results:
+to the log odds of default, i.e.~the model learns to predict higher
+default risk for younger individuals.
+
+If \(\Theta[y^+,d]<0\) even in absence of this force, then contrastive
+divergence exacerbates the existing effect. But if \emph{age} is
+immutable, then this corresponds to penalizing younger individuals for
+implausibility that arises due to a feature that they can change about
+themselves. This is precisely what we want to avoid, and therefore we
+impose \(\mathbf{x}^\prime[1] - \mathbf{x}[1]:=0\) in such cases. In
+doing so, we induce lower sensitivity with respect to \emph{age}
+compared to other mutable features (Figure~\ref{fig-mtblty}).
+
+\end{example}
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/app_mtblty.pdf}}
+
+}
+
+\caption{\label{fig-mtblty}Visual illustration of the effect of imposing
+mutability constraints.}
+
+\end{figure}%
+
+\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, breakable, colback=white, toprule=.15mm, bottomrule=.15mm, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, colframe=quarto-callout-warning-color-frame, opacityback=0, left=2mm, toptitle=1mm, bottomtitle=1mm, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black]
+
+@Cynthia, @Arie, the above still needs to be polished. I'm not sure how
+to demonstrate/prove this last point analytically.
+
+\end{tcolorbox}
 

### 2025-02-10 paper/sections/method.qmd
buff
pfhfhhf

@@
 Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div.
-Here it is crucial that avoid penalizing implausibility that arises due to mutability constraints.
-For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}^\prime[d] = \mathbf{x}[d]$ whenever the constraint is violated, by adjusting either one to the other.
-Setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ intuitively corresponds to imposing user requirements on the target domain; from a Bayesian perspective it can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$^[For additional detail, see @sec-app-constraints of the supplementary appendix.].
+Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints.
+For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}[d]$ would violate mutability constraints.
+Specifically, we set $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ if
+
+1. Feature $d$ is strictly immutable in practice.
+2. We have $\mathbf{x}[d]>\mathbf{x}^\prime[d]$ but feature $d$ can only be decreased in practice.
+3. We have $\mathbf{x}[d]<\mathbf{x}^\prime[d]$ but feature $d$ can only be increased in practice.
+
+From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$.
+Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features.
+This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically bellow and analytically for the linear case in @sec-app-constraints of the supplementary appendix.

### 2025-02-10 paper/sections/other/constraints.qmd
pfhfhhf

@@
 
-:::{exm-constraints}
-
-Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$, with $y \in \{1,2\}$. Then for $y^+ \sim p(y)=2$ we have:
+Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$, with $\mathbf{x} \in \mathbb{R}^2$ and $y \in \{0,1\}$. Then for any given counterfactual with $y^+ \sim p(y)=1$ we have:
 
@@
 
-Contrastive divergence pushes down (up) on the coefficient $\Theta[y^+,d]$ by the amount that feature $d$ that is lower (higher) for counterfactuals than samples.
+For the given counterfactual, contrastive divergence pushes down (up) on the coefficient $\Theta[y^+,d]$ by the amount that feature $d$ that is lower (higher) for counterfactuals than samples.
 This has the intended effect of aligning counterfactual explanations with the data, as discussed above.
 
-Suppose, for example, that feature $d=1$ represents *age* and our model predicts the likelihood of credit default: $y \in \{1:=\text{'no default'},2:=\text{'default'}\}$.
+::: {#exm-grad}
+
+Let feature $d=1$ in @eq-grad represents *age* and our model predicts the likelihood of credit default: $y \in \{0:=\text{'no default'},1:=\text{'default'}\}$.
 Then if $\mathbf{x}^\prime[1] < \mathbf{x}[1]$, contrastive divergence will push down on the corresponding coefficient, $\Theta[\text{'default'},\text{'age'}]$.
 Intuitively, this has the effect of decreasing the effect of *age* that the model attributes to the log odds of default, i.e.
-the model learns to higher default risk younger individuals.
+the model learns to predict higher default risk for younger individuals.
+
+If $\Theta[y^+,d]<0$ even in absence of this force, then contrastive divergence exacerbates the existing effect.
 But if *age* is immutable, then this corresponds to penalizing younger individuals for implausibility that arises due to a feature that they can change about themselves.
-This is precisely what we want to avoid, and therefore we impose no penalty at all.
+This is precisely what we want to avoid, and therefore we impose $\mathbf{x}^\prime[1] - \mathbf{x}[1]:=0$ in such cases.
+In doing so, we induce lower sensitivity with respect to *age* compared to other mutable features (@fig-mtblty).
+
+:::
+
+![Visual illustration of the effect of imposing mutability constraints.](/paper/figures/app_mtblty.svg){#fig-mtblty}
+
+::: {.callout-warning}
+
+\@Cynthia, \@Arie, the above still needs to be polished. I'm not sure how to demonstrate/prove this last point analytically.
 
@@
 
-# Data:
-constraints = ["none", "both"]        # x1 cannot be mutated
-# constraints = ["both", "none"]        # x2 cannot be mutated
-data = LinearlySeparable(
-    n_train=500,
-    batchsize=50,
-    mutability=constraints
-)
-ce_data = get_ce_data(data)
-val_size = data.n_validation / (data.n_validation + data.n_train)
-train_set, val_set, _ = train_val_split(data, ce_data, val_size)
-
-# Model:
-nin = size(first(train_set)[1], 1)
-nout = size(first(train_set)[2], 1)
-model = build_model(LinearModel(), nin, nout)
-```
-
-## Gradient
-
-```{julia}
-X1 = ce_data.X[:,ce_data.output_encoder.labels.==1]
-X2 = ce_data.X[:,ce_data.output_encoder.labels.==2]
-factual = X1[:,1]
-target = 2
-target_enc = Flux.onehotbatch([target], [1,2])
-neighbor = X2[:,1]
-```
-
-## Full Training Loop
-
-Next, we defined the training objective and the counterfactual generator. Finally, we initialize the optimization state.
-
-```{julia}
-using CounterfactualExplanations.Convergence
-obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
-generator = GenericGenerator()
-opt_state = Flux.setup(Descent(), model)
-conv = MaxIterConvergence()
-```
-
-```{julia}
+# Callback:
 using CTExperiments: get_log_reg_params, get_decision_boundary
@@
 end
+
+# Data:
+constraints = [
+    ["both", "both"],
+    ["none", "both"],
+    ["both", "none"]
+]
 ```
@@
 ```{julia}
+using CounterfactualExplanations.Convergence
 using Random
 Random.seed!(42)
+
+models = []
+
+for constraint in constraints
+    # Data:
+    data = LinearlySeparable(
+        n_train=500,
+        batchsize=50,
+        mutability=constraint
+    )
+    ce_data = get_ce_data(data)
+    val_size = data.n_validation / (data.n_validation + data.n_train)
+    train_set, val_set, _ = train_val_split(data, ce_data, val_size)
+
+    # Model:
+    nin = size(first(train_set)[1], 1)
+    nout = size(first(train_set)[2], 1)
+    model = build_model(LinearModel(), nin, nout)
+
+    # Objective:
+    obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
+    generator = GenericGenerator()
+    opt_state = Flux.setup(Descent(), model)
+    conv = MaxIterConvergence()
 
@@
         nepochs = 50,
-    mutability = Symbol.(constraints),
+        mutability = Symbol.(constraint),
         callback = plot_db,
@@
     )
+
+    push!(models, model)
+end
 ```
 
-Finally, we visualize the results:
+```{julia}
+using Plots.PlotMeasures
+plts = []
+for (i,model) in enumerate(models)
+    coeff = get_log_reg_params(model)
+    db = get_decision_boundary(coeff)
+    xlab = constraints[i][1] == "both" ? "mutable" : "immutable"
+    ylab = constraints[i][2] == "both" ? "mutable" : "immutable"
+    plt = Plots.scatter(ce_data; xlab=xlab, ylab=ylab, bottom_margin = 5mm, left_margin = 5mm)
+    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
 
-```{julia}
-coeff = get_log_reg_params(model)
-display(coeff)
-db = get_decision_boundary(coeff)
-Plots.scatter(ce_data)
-Plots.abline!(db.slope, db.intercept; lw=5, label="Dec. Boundary")
+    push!(plts, plt)
+end
+
+Plots.plot(plts..., layout=(1,3), size=(900,300))
 ```

### 2025-02-11 paper/preprint/paper.tex
just experimental setup and stuff left now and then sending off the first draft

--- 2025-02-07 paper/preprint/paper.tex
@@
 \theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
 \newtheorem{exercise}{Parameters}[section]
 \theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
+\newtheorem{example}{Example}[section]
 \theoremstyle{remark}
@@
 their effect on \(\mathbf{x}^\prime\), both types of constraints
-influence model outcomes through Equation~\ref{eq-div}. This is best
-explained through a simple example involving a linear classifier:
-\(\mathbf{M}_\theta(\mathbf{x})=\theta^\intercal\mathbf{x}\). Then for
-\(y^+ \sim p(y)\) we have:
-
-\begin{equation}\phantomsection\label{eq-grad}{
-\nabla_\theta \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \nabla_\theta \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) = \mathbf{x}^\prime - \mathbf{x}
-}\end{equation}
-
-In other words, contrastive divergence will push down (up) on the
-coefficient \(\theta[d]\) corresponding to a feature \(d\) that is lower
-(higher) for counterfactuals than samples. This has the intended effect
-of aligning counterfactual explanations with the data, as discussed
-above.
-
-Crucially though, we want to avoid penalizing implausibility that arises
-due to mutability constraints. Suppose, for example, that feature \(d\)
-represents \emph{age} and our model predicts the likelihood of credit
-default.
+influence model outcomes through Equation~\ref{eq-div}. Here it is
+crucial that we avoid penalizing implausibility that arises due to
+mutability constraints. For any mutability-constrained feature \(d\)
+this can be achieved by enforcing
+\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
+\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
+violate mutability constraints. Specifically, we set
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Feature \(d\) is strictly immutable in practice.
+\item
+  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\) but feature \(d\) can
+  only be decreased in practice.
+\item
+  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\) but feature \(d\) can
+  only be increased in practice.
+\end{enumerate}
+
+From a Bayesian perspective, setting
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
+a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
+Intuitively, we think of this simply in terms ignoring implausibility
+costs with respect to immutable features, which effectively forces the
+model to instead seek plausibility with respect to the remaining
+features. This in turn results in lower overall sensitivity to immutable
+features, which we demonstrate empirically below. Under certain
+assumptions this can also be shown to hold analytically
+Section~\ref{sec-app-constraints}.
 
@@
 \url{https://doi.org/10.1145/3580305.3599290}.
+
+\bibitem[\citeproctext]{ref-hastie2009elements}
+Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
+Elements of Statistical Learning}. Springer New York.
+\url{https://doi.org/10.1007/978-0-387-84858-7}.
 
@@
 \section{Training Details}\label{training-details}
+
+\subsection{Details on Encoding Mutability
+Constraints}\label{sec-app-constraints}
+
+In Section~\ref{sec-constraints} we explain that to avoid penalizing
+implausibility that arises due to mutability constraints, we impose a
+point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
+argue in Section~\ref{sec-constraints} that this approach induces models
+to be less sensitive to feature \(d\) vis-à-vis some other mutable
+feature \(m\) and demonstrate this empirically in
+Section~\ref{sec-experiments}. Below we derive this result analytically
+for linear classifier.
+
+\subsubsection{Linear Classifier}\label{linear-classifier}
+
+Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
+classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
+\(\mathbf{x} \in \mathbb{R}^D\). Let \(d_{\text{mtbl}}\) and
+\(d_{\text{immtbl}}\) denote some mutable and immutable feature,
+respectively. Suppose that
+\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
+\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
+\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
+class \(k\).
+
+Assuming multivariate Gaussian class densities with common diagonal
+covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), we
+have for the log likelihood ratio between any two classes
+\(k,m \in \mathcal{K}\) (\citeproc{ref-hastie2009elements}{Hastie,
+Tibshirani, and Friedman 2009}):
+
+\begin{equation}\phantomsection\label{eq-loglike}{
+\log \frac{p(k|x_{d})}{p(m|x_{d})}=\frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
+}\end{equation}
+
+By Equation~\ref{eq-loglike} we have the
+\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\)
+and \(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\),
+where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
+\(d\) for class \(k\).
+
+Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
+class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
+target class. Then the partial derivative of the contrastive divergence
+penalty Equation~\ref{eq-div} with respect to coefficient
+\(\theta_{y^+,d}\) is equal to
+
+\begin{equation}\phantomsection\label{eq-grad}{
+\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
+}\end{equation}
+
+and equal to zero everywhere else.
+
+Since \(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) we
+are more likely to have
+\(x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}} < 0\) than vice
+versa at initialization. Similarly, we are more likely to have
+\(x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}} > 0\) by
+\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\).
+
+This implies that if we do not protect feature \(d_{\text{immtbl}}\),
+the contrastive divergence penalty will push down on
+\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
+effect
+\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\).
+In words, not protecting the immutable feature would have the
+undesirable effect of making the classifier more sensitive to this
+feature, in that it would be more likely to predict class \(y^-\) as
+opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).
+
+By the same rationale, the contrastive divergence penalty can generally
+be expected to push up on \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
+\(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\). In
+words, this has the effect of making the classifier more sensitive to
+the mutable feature, in that it would be more likely to predict class
+\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).
+
+Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
+has the net affect of decreasing the classifier's sensitivity to the
+immutable feature relative to the mutable feature (i.e.~no change in
+sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
+for \(d_{\text{mtbl}}\)).
+
+We provide an illustrative example in Example~\ref{exm-grad}.
+
+\begin{example}[Prediction of Consumer Credit
+Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}
+
+Suppose now that \(d_{\text{immtbl}}\) represents an individual's
+\emph{age} and \(d_{\text{mtbl}}\) represents an individual's existing
+level of credit card debt. Assume that these two features are
+independent, the class conditional densities are Gaussian and we use no
+other features to predict the risk of individuals defaulting on a
+consumer loan using a linear classifier. We have simulated this scenario
+using synthetic data in Figure~\ref{fig-mtblty}.
+
+In panel (a) of Figure~\ref{fig-mtblty}, we have trained the linear
+classifier using counterfactual training, treating both features as
+mutable. The linear decision boundary is roughly equally sensitive to
+both \emph{age} and existing levels of \emph{debt}.
+
+Conversely, in panel (b) of Figure~\ref{fig-mtblty}, we have trained the
+same classifier using counterfactual training, but this time treating
+\emph{age} as an immutable feature. The result is a new decision
+boundary that has tilted in favor of higher sensitivity to the mutable
+feature (existing \emph{debt}) and lower sensitivity to the immutable
+feature (\emph{age}).
+
+\end{example}
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/app_mtblty.pdf}}
+
+}
+
+\caption{\label{fig-mtblty}Visual illustration of the effect of imposing
+mutability constraints. See Example~\ref{exm-grad} for details.}
+
+\end{figure}%
+
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, opacityback=0, rightrule=.15mm, leftrule=.75mm, colback=white, left=2mm, colframe=quarto-callout-warning-color-frame, opacitybacktitle=0.6, breakable, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, bottomtitle=1mm]
+
+@Cynthia, @Arie, let me know if you want to phrase the above in terms of
+a theorem and proof. This is something I've so far shied away from, but
+I can see how it might be appropriate here.
+
+\end{tcolorbox}
+
+\section{Detailed Results}\label{detailed-results}
 

### 2025-02-11 paper/sections/method.qmd
just experimental setup and stuff left now and then sending off the first draft

@@
 Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features.
-This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically bellow and analytically for the linear case in @sec-app-constraints of the supplementary appendix.
+This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically below.
+Under certain assumptions this can also be shown to hold analytically [@sec-app-constraints].

### 2025-02-11 paper/sections/other/constraints.qmd
just experimental setup and stuff left now and then sending off the first draft

--- 2025-02-06 paper/sections/other/constraints.qmd
@@
 ```{julia}
+#| eval: false
+
 projectdir = splitpath(pwd()) |>
@@
 cd(projectdir)
-```
 
-```{julia}
 using CTExperiments
@@
 
-# Linear Model
+## Details on Encoding Mutability Constraints {#sec-app-constraints}
 
-First, we define the mutability constraints and load the data and model:
+In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature.
+We argue in @sec-constraints that this approach induces models to be less sensitive to feature $d$ vis-à-vis some other mutable feature $m$ and demonstrate this empirically in @sec-experiments.
+Below we derive this result analytically for linear classifier.
+
+### Linear Classifier
+
+Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$ denote a linear classifier with $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
+Let $d_{\text{mtbl}}$ and $d_{\text{immtbl}}$ denote some mutable and immutable feature, respectively.
+Suppose that $\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}$ and $\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}$, where $\mu_{k,d}$ denotes the conditional sample mean of feature $d$ in class $k$.
+
+Assuming multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, we have for the log likelihood ratio between any two classes $k,m \in \mathcal{K}$ [@hastie2009elements]:
+
+$$
+\log \frac{p(k|x_{d})}{p(m|x_{d})}=\frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
+$$ {#eq-loglike}
+
+By @eq-loglike we have the $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$ and $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$, where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.
+
+Let $\mathbf{x}^\prime$ denote some randomly chosen individual from class $y^-$ and let $y^+ \sim p(y)$ denote the randomly chosen target class.
+Then the partial derivative of the contrastive divergence penalty [@eq-div] with respect to coefficient $\theta_{y^+,d}$ is equal to
+
+$$
+\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
+$$ {#eq-grad}
+
+and equal to zero everywhere else.
+
+Since $\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}$ we are more likely to have $x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}} < 0$ than vice versa at initialization.
+Similarly, we are more likely to have $x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}} > 0$ by $\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}$.
+
+This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will push down on $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$.
+In words, not protecting the immutable feature would have the undesirable effect of making the classifier more sensitive to this feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for lower values of $d_{\text{immtbl}}$.
+
+By the same rationale, the contrastive divergence penalty can generally be expected to push up on $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$.
+In words, this has the effect of making the classifier more sensitive to the mutable feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for higher values of $d_{\text{mtbl}}$.
+
+Thus, our proposed approach of protecting feature $d_{\text{immtbl}}$ has the net affect of decreasing the classifier's sensitivity to the immutable feature relative to the mutable feature (i.e.
+no change in sensitivity for $d_{\text{immtbl}}$ relative to increased sensitivity for $d_{\text{mtbl}}$).
+
+We provide an illustrative example in @exm-grad.
+
+::: {#exm-grad}
+
+## Prediction of Consumer Credit Default
+
+Suppose now that $d_{\text{immtbl}}$ represents an individual's *age* and $d_{\text{mtbl}}$ represents an individual's existing level of credit card debt.
+Assume that these two features are independent, the class conditional densities are Gaussian and we use no other features to predict the risk of individuals defaulting on a consumer loan using a linear classifier.
+We have simulated this scenario using synthetic data in @fig-mtblty.
+
+In panel (a) of @fig-mtblty, we have trained the linear classifier using counterfactual training, treating both features as mutable.
+The linear decision boundary is roughly equally sensitive to both *age* and existing levels of *debt*.
+
+Conversely, in panel (b) of @fig-mtblty, we have trained the same classifier using counterfactual training, but this time treating *age* as an immutable feature.
+The result is a new decision boundary that has tilted in favor of higher sensitivity to the mutable feature (existing *debt*) and lower sensitivity to the immutable feature (*age*).
+
+:::
+
+![Visual illustration of the effect of imposing mutability constraints. See @exm-grad for details.](/paper/figures/app_mtblty.svg){#fig-mtblty}
+
+::: {.callout-warning}
+
+\@Cynthia, \@Arie, let me know if you want to phrase the above in terms of a theorem and proof. This is something I've so far shied away from, but I can see how it might be appropriate here.
+
+:::
 
 ```{julia}
+#| eval: false
+
 using CTExperiments: 
@@
 
-# Data:
-constraints = ["none", "both"]        # x1 cannot be mutated
-# constraints = ["both", "none"]        # x2 cannot be mutated
-data = LinearlySeparable(
-    n_train=500,
-    batchsize=50,
-    mutability=constraints
-)
-ce_data = get_ce_data(data)
-val_size = data.n_validation / (data.n_validation + data.n_train)
-train_set, val_set, _ = train_val_split(data, ce_data, val_size)
-
-# Model:
-nin = size(first(train_set)[1], 1)
-nout = size(first(train_set)[2], 1)
-model = build_model(LinearModel(), nin, nout)
-```
-
-Next, we defined the training objective and the counterfactual generator. Finally, we initialize the optimization state.
-
-```{julia}
-using CounterfactualExplanations.Convergence
-obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
-# obj = VanillaObjective()
-generator = GenericGenerator()
-opt_state = Flux.setup(Descent(), model)
-conv = MaxIterConvergence()
-```
-
-```{julia}
+# Callback:
 using CTExperiments: get_log_reg_params, get_decision_boundary
@@
 end
+
+# Data:
+constraints = [
+    ["both", "both"],
+    ["both", "none"]
+]
 ```
 
-Then we train the model:
+```{julia}
+#| eval: false
 
-```{julia}
+using CounterfactualExplanations.Convergence
 using Random
 Random.seed!(42)
+
+models = []
+
+for constraint in constraints
+    # Data:
+    data = LinearlySeparable(
+        n_train=500,
+        batchsize=50,
+        mutability=constraint
+    )
+    ce_data = get_ce_data(data)
+    val_size = data.n_validation / (data.n_validation + data.n_train)
+    train_set, val_set, _ = train_val_split(data, ce_data, val_size)
+
+    # Model:
+    nin = size(first(train_set)[1], 1)
+    nout = size(first(train_set)[2], 1)
+    model = build_model(LinearModel(), nin, nout)
+
+    # Objective:
+    obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
+    generator = GenericGenerator()
+    opt_state = Flux.setup(Descent(), model)
+    conv = MaxIterConvergence()
 
@@
     )
+
+    push!(models, model)
+end
 ```
 
-Finally, we visualize the results:
+```{julia}
+#| eval: false
 
-```{julia}
+using Plots.PlotMeasures
+plts = []
+titles = ["(a)", "(b)"]
+_xlab = "Existing Debt"
+_ylab = "Age"
+for (i,model) in enumerate(models)
     coeff = get_log_reg_params(model)
-display(coeff)
     db = get_decision_boundary(coeff)
-Plots.scatter(ce_data)
-Plots.abline!(db.slope, db.intercept; lw=5, label="Dec. Boundary")
+    xlab = constraints[i][1] == "both" ? "$(_xlab) (mutable)" : "$(_xlab) (immutable)"
+    ylab = constraints[i][2] == "both" ? "$(_ylab) (mutable)" : "$(_ylab) (immutable)"
+    plt = Plots.scatter(ce_data; xlab=xlab, ylab=ylab, bottom_margin = 5mm, left_margin = 5mm, label=["Default" "No Default"], title=titles[i])
+    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
+
+    push!(plts, plt)
+end
+
+plt = Plots.plot(plts..., layout=(1,2), size=(600,300))
+Plots.savefig(plt, "paper/figures/app_mtblty.svg")
 ```

### 2025-02-11 paper/body_ecml.tex
methodology done

@@
 features. This in turn results in lower overall sensitivity to immutable
-features, which we demonstrate empirically bellow and analytically for
-the linear case in \textbf{?@sec-app-constraints} of the supplementary
-appendix.
+features, which we demonstrate empirically for different classifiers in
+Section~\ref{sec-experiments}. Under certain conditions, this results
+holds theoretically\footnote{For the proof, see the supplementary
+  appendix.}:
+
+\begin{theorem}[Protecting Immutable
+Features]\protect\hypertarget{thm-mtblty}{}\label{thm-mtblty}
+
+Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
+classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
+\(\mathbf{x} \in \mathbb{R}^D\). If we assume multivariate Gaussian
+class densities with common diagonal covariance matrix
+\(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), then protecting an
+immutable feature from the contrastive divergence penalty
+Equation~\ref{eq-div} will result in lower classifier sensitivity to
+that feature, provided that at least one other feature is mutable.
+
+\end{theorem}
+
+It is worth highlighting that Theorem~\ref{thm-mtblty} assumes
+independence of features. This raises a valid concern about the effect
+of protecting immutable features in the presence of proxy features that
+remain unprotected. We discuss this limitation in
+Section~\ref{sec-discussion}.
 
@@
 
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Limited to classification models.
+\item
+  Proxy attributes of immutable features.
+\end{enumerate}
+
 \section{Conclusion}\label{sec-conclusion}

### 2025-02-11 paper/preprint/paper.tex
methodology done

@@
 \usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{theorem}{Theorem}[section]
+\theoremstyle{definition}
+\newtheorem{exercise}{Parameters}[section]
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
 \theoremstyle{definition}
@@
 features. This in turn results in lower overall sensitivity to immutable
-features, which we demonstrate empirically below. Under certain
-assumptions this can also be shown to hold analytically
-Section~\ref{sec-app-constraints}.
+features, which we demonstrate empirically for different classifiers in
+Section~\ref{sec-experiments}. Under certain conditions, this results
+holds theoretically\footnote{For the proof, see the supplementary
+  appendix.}:
+
+\begin{theorem}[Protecting Immutable
+Features]\protect\hypertarget{thm-mtblty}{}\label{thm-mtblty}
+
+Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
+classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
+\(\mathbf{x} \in \mathbb{R}^D\). If we assume multivariate Gaussian
+class densities with common diagonal covariance matrix
+\(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), then protecting an
+immutable feature from the contrastive divergence penalty
+Equation~\ref{eq-div} will result in lower classifier sensitivity to
+that feature, provided that at least one other feature is mutable.
+
+\end{theorem}
+
+It is worth highlighting that Theorem~\ref{thm-mtblty} assumes
+independence of features. This raises a valid concern about the effect
+of protecting immutable features in the presence of proxy features that
+remain unprotected. We discuss this limitation in
+Section~\ref{sec-discussion}.
 
@@
 \section{Discussion}\label{sec-discussion}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Limited to classification models.
+\item
+  Proxy attributes of immutable features.
+\end{enumerate}
 
@@
 \item
+  \(y^-\): The non-target class and also the index of non-the target
+  class.
+\item
   \(\mathbf{y}^+\): The one-hot encoded output vector for the target
@@
 
-\subsection{Details on Encoding Mutability
-Constraints}\label{sec-app-constraints}
+\subsection{Protecting Mutability Constraints with Linear
+Classifiers}\label{sec-app-constraints}
 
@@
 argue in Section~\ref{sec-constraints} that this approach induces models
-to be less sensitive to feature \(d\) vis-à-vis some other mutable
-feature \(m\) and demonstrate this empirically in
-Section~\ref{sec-experiments}. Below we derive this result analytically
-for linear classifier.
-
-\subsubsection{Linear Classifier}\label{linear-classifier}
-
-Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
-classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
-\(\mathbf{x} \in \mathbb{R}^D\). Let \(d_{\text{mtbl}}\) and
-\(d_{\text{immtbl}}\) denote some mutable and immutable feature,
-respectively. Suppose that
+to be less sensitive to immutable features and demonstrate this
+empirically in Section~\ref{sec-experiments}. Below we derive the
+analytical results in Theorem~\ref{thm-mtblty}.
+
+\begin{proof}
+Let \(d_{\text{mtbl}}\) and \(d_{\text{immtbl}}\) denote some mutable
+and immutable feature, respectively. Suppose that
 \(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
@@
 \(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
-class \(k\).
+class \(k\). In words, we assume that the immutable feature tends to
+take lower values for samples in the non-target class \(y^-\) than in
+the target class \(y^+\). We assume the opposite to hold for the mutable
+feature.
 
@@
 \begin{equation}\phantomsection\label{eq-loglike}{
-\log \frac{p(k|x_{d})}{p(m|x_{d})}=\frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
 }\end{equation}
 
-By Equation~\ref{eq-loglike} we have the
+By independence of \(x_1,...,x_D\), the full log-likelihood ratio
+decomposes into:
+
+\begin{equation}\phantomsection\label{eq-loglike-decomp}{
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
+}\end{equation}
+
+By Equation~\ref{eq-loglike-decomp} we have the
 \(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\)
@@
 for \(d_{\text{mtbl}}\)).
+\end{proof}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, opacityback=0, rightrule=.15mm, leftrule=.75mm, colback=white, left=2mm, colframe=quarto-callout-warning-color-frame, opacitybacktitle=0.6, breakable, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, bottomtitle=1mm]
-
-@Cynthia, @Arie, let me know if you want to phrase the above in terms of
-a theorem and proof. This is something I've so far shied away from, but
-I can see how it might be appropriate here.
+\begin{tcolorbox}[enhanced jigsaw, bottomtitle=1mm, titlerule=0mm, breakable, toprule=.15mm, left=2mm, bottomrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, opacityback=0, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, colback=white]
+
+@Cynthia, @Arie, I have tentatively phrased the above in terms of a
+theorem and proof. This is something I've so far shied away from because
+I feel a bit out of my depth when it comes to mathematical proofs. The
+above makes intuitive sense to me, but I don't know for sure if it's
+correct.
 

### 2025-02-11 paper/sections/discussion.qmd
methodology done

# Discussion {#sec-discussion}

1. Limited to classification models.
2. Proxy attributes of immutable features.

### 2025-02-11 paper/sections/method.qmd
methodology done

--- 2025-02-10 paper/sections/method.qmd
@@
 Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features.
-This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically bellow and analytically for the linear case in @sec-app-constraints of the supplementary appendix.
+This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments.
+Under certain conditions, this results holds theoretically^[For the proof, see the supplementary appendix.]:
+
+::: {#thm-mtblty}
+
+## Protecting Immutable Features
+
+Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$ denote a linear classifier with $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
+If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty [@eq-div] will result in lower classifier sensitivity to that feature, provided that at least one other feature is mutable.
+
+:::
+
+It is worth highlighting that @thm-mtblty assumes independence of features.
+This raises a valid concern about the effect of protecting immutable features in the presence of proxy features that remain unprotected.
+We discuss this limitation in @sec-discussion.

### 2025-02-11 paper/sections/other/constraints.qmd
methodology done

@@
 
-## Details on Encoding Mutability Constraints {#sec-app-constraints}
+## Protecting Mutability Constraints with Linear Classifiers {#sec-app-constraints}
 
 In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature.
-We argue in @sec-constraints that this approach induces models to be less sensitive to feature $d$ vis-à-vis some other mutable feature $m$ and demonstrate this empirically in @sec-experiments.
-Below we derive this result analytically for linear classifier.
+We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments.
+Below we derive the analytical results in @thm-mtblty.
 
-### Linear Classifier
+::: {.proof}
 
-Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$ denote a linear classifier with $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
 Let $d_{\text{mtbl}}$ and $d_{\text{immtbl}}$ denote some mutable and immutable feature, respectively.
 Suppose that $\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}$ and $\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}$, where $\mu_{k,d}$ denotes the conditional sample mean of feature $d$ in class $k$.
+In words, we assume that the immutable feature tends to take lower values for samples in the non-target class $y^-$ than in the target class $y^+$.
+We assume the opposite to hold for the mutable feature.
 
@@
 $$
-\log \frac{p(k|x_{d})}{p(m|x_{d})}=\frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
 $$ {#eq-loglike}
 
-By @eq-loglike we have the $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$ and $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$, where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.
+By independence of $x_1,...,x_D$, the full log-likelihood ratio decomposes into:
+
+$$
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
+$$ {#eq-loglike-decomp}
+
+By @eq-loglike-decomp we have the $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$ and $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$, where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.
 
@@
 no change in sensitivity for $d_{\text{immtbl}}$ relative to increased sensitivity for $d_{\text{mtbl}}$).
+
+:::
 
@@
 
-\@Cynthia, \@Arie, let me know if you want to phrase the above in terms of a theorem and proof. This is something I've so far shied away from, but I can see how it might be appropriate here.
+\@Cynthia, \@Arie, I have tentatively phrased the above in terms of a theorem and proof.
+This is something I've so far shied away from because I feel a bit out of my depth when it comes to mathematical proofs.
+The above makes intuitive sense to me, but I don't know for sure if it's correct.
 

### 2025-02-11 paper/preprint/paper.tex
uh

@@
 \usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{definition}
+\newtheorem{exercise}{Parameters}[section]
 \theoremstyle{plain}
@@
 \theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
-\theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
 \theoremstyle{remark}
@@
 
-\section{Training Details}\label{training-details}
+\section{Technical Details of Our
+Approach}\label{technical-details-of-our-approach}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomtitle=1mm, titlerule=0mm, breakable, toprule=.15mm, left=2mm, bottomrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, opacityback=0, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toptitle=1mm, opacitybacktitle=0.6, breakable, colbacktitle=quarto-callout-warning-color!10!white, opacityback=0, colframe=quarto-callout-warning-color-frame, arc=.35mm, bottomtitle=1mm, rightrule=.15mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, titlerule=0mm, leftrule=.75mm, coltitle=black, bottomrule=.15mm, toprule=.15mm, colback=white]
 

### 2025-02-11 paper/sections/appendix.qmd
slowly slowly
progress
methodology done
uh

@@
 
-# Training Details {.appendix}
+# Notation {.appendix}
+
+- $y^+$: The target class and also the index of the target class.
+- $y^-$: The non-target class and also the index of non-the target class.
+- $\mathbf{y}^+$: The one-hot encoded output vector for the target class. 
+- $\theta$: Model parameters (unspecified).
+- $\Theta$: Matrix of parameters.
+
+# Technical Details of Our Approach {.appendix}
+
+{{< include /paper/sections/other/constraints.qmd >}}
+
+# Detailed Results {.appendix}
 

### 2025-02-11 paper/body_ecml.tex
first version of research questions defined

@@
 
+In this section, we present experiments that we have conducted in order
+to answer the following research questions:
+
+\begin{corollary}[Plausibility]\protect\hypertarget{cor-plaus}{}\label{cor-plaus}
+
+Does our proposed counterfactual training objective
+(Equation~\ref{eq-obj}) induce models to learn plausible explanations?
+
+\end{corollary}
+
+\begin{corollary}[Actionability]\protect\hypertarget{cor-action}{}\label{cor-action}
+
+Does our proposed counterfactual training objective
+(Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
+outcomes in the presence of actionability constraints?
+
+\end{corollary}
+
+Beyond this, we are also interested in understanding how robust our
+answers to RQ~\ref{cor-plaus} and RQ~\ref{cor-action} are:
+
+\begin{corollary}[Hyperparameters]\protect\hypertarget{cor-hyper}{}\label{cor-hyper}
+
+What are the effects of different hyperparameter choices with respect to
+Equation~\ref{eq-obj}?
+
+\end{corollary}
+
 \subsection{Experimental Setup}\label{experimental-setup}

### 2025-02-11 paper/preprint/paper.tex
first version of research questions defined

@@
 \theoremstyle{definition}
-\newtheorem{example}{Example}[section]
+\newtheorem{exercise}{Parameters}[section]
 \theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
+\newtheorem{definition}{Definition}[section]
 \theoremstyle{plain}
@@
 \theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
+\newtheorem{example}{Example}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, toptitle=1mm, opacitybacktitle=0.6, breakable, colbacktitle=quarto-callout-warning-color!10!white, opacityback=0, colframe=quarto-callout-warning-color-frame, arc=.35mm, bottomtitle=1mm, rightrule=.15mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, titlerule=0mm, leftrule=.75mm, coltitle=black, bottomrule=.15mm, toprule=.15mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, colframe=quarto-callout-warning-color-frame, opacityback=0, coltitle=black, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, bottomrule=.15mm, colback=white, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, rightrule=.15mm, titlerule=0mm, leftrule=.75mm, opacitybacktitle=0.6, left=2mm]
 

### 2025-02-11 paper/sections/experiments.qmd
first version of research questions defined

@@
 # Experiments {#sec-experiments}
+
+In this section, we present experiments that we have conducted in order to answer the following research questions:
+
+::: {#cor-plaus}
+
+## Plausibility
+
+Does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?
+
+:::
+
+::: {#cor-action}
+
+## Actionability
+
+Does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+
+:::
+
+Beyond this, we are also interested in understanding how robust our answers to @cor-plaus and @cor-action are:
+
+::: {#cor-hyper}
+
+## Hyperparameters
+
+What are the effects of different hyperparameter choices with respect to @eq-obj?
+
+:::
 

### 2025-02-12 paper/preprint/paper.tex
small clarification

@@
 \newtheorem{exercise}{Parameters}[section]
+\theoremstyle{plain}
+\newtheorem{theorem}{Theorem}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
 \theoremstyle{definition}
 \newtheorem{example}{Example}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 Section~\ref{sec-experiments}. Under certain conditions, this results
-holds theoretically\footnote{For the proof, see the supplementary
-  appendix.}:
+holds theoretically{[}For the proof, see the supplementary appendix.{]}:
 
@@
 
-Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
-classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
-\(\mathbf{x} \in \mathbb{R}^D\). If we assume multivariate Gaussian
-class densities with common diagonal covariance matrix
-\(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), then protecting an
-immutable feature from the contrastive divergence penalty
-Equation~\ref{eq-div} will result in lower classifier sensitivity to
-that feature, provided that at least one other feature is mutable.
+Let
+\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
+denote a linear classifier with softmax activation \(\mathcal{S}\)
+(i.e.~\emph{multinomial logistic regression}) where
+\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
+we assume multivariate Gaussian class densities with common diagonal
+covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
+then protecting an immutable feature from the contrastive divergence
+penalty (Equation~\ref{eq-div}) will result in lower classifier
+sensitivity to that feature relative to the remaining features, provided
+that at least one of those is mutable.
 
@@
 
-By Equation~\ref{eq-loglike-decomp} we have the
-\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\)
-and \(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\),
+By the properties of our classifier (\emph{multinomial logistic
+regression}), we have:
+
+\begin{equation}\phantomsection\label{eq-multi}{
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
+}\end{equation}
+
 where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
 \(d\) for class \(k\).
+
+Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
+can identify that
+\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
+the assumptions we made above. Hence, we have that
+\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
+and
+\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)
 
@@
 
-Since \(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) we
+Since \((\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})\) we
 are more likely to have
-\(x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}} < 0\) than vice
+\((x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0\) than vice
 versa at initialization. Similarly, we are more likely to have
-\(x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}} > 0\) by
-\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\).
+\((x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0\) since
+\((\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})\).
 
@@
 effect
-\(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0\).
+\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
 In words, not protecting the immutable feature would have the
@@
 be expected to push up on \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
-\(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0\). In
-words, this has the effect of making the classifier more sensitive to
+\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
+In words, this has the effect of making the classifier more sensitive to
 the mutable feature, in that it would be more likely to predict class
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, colframe=quarto-callout-warning-color-frame, opacityback=0, coltitle=black, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, bottomrule=.15mm, colback=white, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, rightrule=.15mm, titlerule=0mm, leftrule=.75mm, opacitybacktitle=0.6, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]
 
@@
 
-\subsection{Initial Grid Search}\label{initial-grid-search}
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]
+
+@Cynthia, @Arie, I'm including some preliminary results here but will
+rerun experiments in the coming days. You'll notice some odd outlier
+results (huge values) for the initial grid search
+(Section~\ref{sec-app-initial}). My belief is that this is once again
+due to \emph{ECCo} overshooting for some hyperparameter choices, that we
+have discussed before. A simple solution to this story is to actually
+impose domain constraints. Let's see what the final results show us (I
+also still plan to make slight changes to the implementation), but in
+any case I think it might even be worth to report results for the
+initial grid search in the final appendix (they do include good results
+for CT for certain hyperparameter ranges and highlight limitations
+inherited from energy-based modelling).
+
+\end{tcolorbox}
+
+\subsection{Qualitative Findings for Image
+Data}\label{qualitative-findings-for-image-data}
+
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]
+
+@Cynthia, @Arie, Figure~\ref{fig-mnist} shows much more plausible
+(faithful) counterfactuals for a model with CT than the model with
+conventional training (Figure~\ref{fig-mnist-vanilla}). In fact, this is
+not even using \emph{ECCo+} and still showing better results than the
+best results we achieved in our AAAI paper for JEM ensembles.
+
+\end{tcolorbox}
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp.png}}
+
+}
+
+\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
+counterfactual training. The underlying generator, \emph{ECCo}, aims to
+generate counterfactuals that are faithful to the model
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp_vanilla.png}}
+
+}
+
+\caption{\label{fig-mnist-vanilla}Counterfactual images for \emph{MLP}
+with conventional training. The underlying generator, \emph{ECCo}, aims
+to generate counterfactuals that are faithful to the model
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
+
+\end{figure}%
+
+\subsection{Initial Grid Search}\label{sec-app-initial}
 

### 2025-02-12 paper/sections/appendix.qmd
small clarification

@@
 
-## Initial Grid Search
+::: {.callout-warning}
+
+\@Cynthia, \@Arie, I'm including some preliminary results here but will rerun experiments in the coming days.
+You'll notice some odd outlier results (huge values) for the initial grid search (@sec-app-initial).
+My belief is that this is once again due to *ECCo* overshooting for some hyperparameter choices, that we have discussed before.
+A simple solution to this story is to actually impose domain constraints.
+Let's see what the final results show us (I also still plan to make slight changes to the implementation), but in any case I think it might even be worth to report results for the initial grid search in the final appendix (they do include good results for CT for certain hyperparameter ranges and highlight limitations inherited from energy-based modelling).
+
+:::
+
+## Qualitative Findings for Image Data
+
+::: {.callout-note}
+
+\@Cynthia, \@Arie, @fig-mnist shows much more plausible (faithful) counterfactuals for a model with CT than the model with conventional training (@fig-mnist-vanilla).
+In fact, this is not even using *ECCo+* and still showing better results than the best results we achieved in our AAAI paper for JEM ensembles.
+
+:::
+
+![Counterfactual images for *MLP* with counterfactual training.
+The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}
+
+![Counterfactual images for *MLP* with conventional training.
+The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp_vanilla.png){#fig-mnist-vanilla}
+
+## Initial Grid Search {#sec-app-initial}
 

### 2025-02-12 paper/ECML_PKDD_2025_Author_Kit/paper.tex

\documentclass[runningheads]{llncs}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
% N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
\usepackage{mwe}
% N.B.: you may delete the preceding line. It is used to display an example image in this template.

\begin{document}

\title{Recent Advances in Underwater Basket Weaving Under the Extreme Pressure of the Mariana Trench}

\titlerunning{Underwater Basket Weaving Under Extreme Pressure}
% If the full title of your paper is short enough to also fit in the running head, you can omit the abbreviated paper title here. You can check as follows: if you comment out the \titlerunning line, something will appear in the header of all odd-numbered pages of your PDF from page 3 onward. This something is either the full title (in which case all is well), or the error message "Title Suppressed Due to Excessive Length". If this error message appears, you're going to want to provide an abbreviated title within the \titlerunning command, because if you won't do it, Springer will do it for you.

%N.B.: Author information (both in the \author{} and \authorrunning{} command) should only be present in the Camera-Ready Version of your paper. The version that you initially submit for review, ought to be double-blind. So, when initially submitting your paper, use:
%\author{Author information scrubbed for double-blind reviewing}
\author{Andr\'e Lauren Benjamin\inst{1} \and
Calvin Cordozar Broadus Jr.\inst{2,3} \corr \and
Antwan Andr\'e Patton\inst{1}\orcidID{0000-1111-2222-3333}}
% You may leave out the orcidID information, if you want to.
% Use \corr to indicate the corresponding author. Note the spacing around the \corr command. Only one author can be the corresponding author.

%N.B.: comment out the \authorrunning{} command for the double-blind version of your paper submitted for review. Later, if your paper is accepted, use the command for the Camera-Ready Version.
\authorrunning{A.L. Benjamin et al.}
% First names are abbreviated in the running head.
% If there is one author, write 'A.L. Benjamin'.
% If there are two authors, write 'A.L. Benjamin and C.C. Broadus Jr.'
% If there are more than two authors, '[...] et al.' is used.

\institute{Fictional Southern University, Savannah GA 31404, USA \email{\{a.l.benjamin,a.a.patton\}@fsu.fake}
\and
Fictional West Coast University, Long Beach CA 90840, USA \email{ccb@fwcu.fake}
\and
Secondary European Affiliation, Tiergartenstr. 17, 69121 Heidelberg, Germany
\email{lncs@springer.com}}

\maketitle              % typeset the header of the contribution

\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}

\section{Introduction}
\subsection{Main Contributions}
Please note that the first paragraph of a section or subsection is
not indented. The first paragraph that follows a table, figure,
equation etc. does not need an indent, either.

Subsequent paragraphs, however, are indented.

\section{Related Work}
\subsection{Basket Weaving}
\subsubsection{Underwater Basket Weaving} Only two levels of
headings should be numbered. Lower level headings remain unnumbered;
they are formatted as run-in headings.

\paragraph{Underwater Basket Weaving Under Difficult Circumstances}
The contribution should contain no more than four levels of
headings. Table~\ref{tab1} gives a summary of all heading levels.

\begin{table}[t]
\caption{Table captions should be placed above the
tables.}\label{tab1}
\begin{tabular}{lll}
\toprule
Heading level &  Example & Font size and style\\
\midrule
Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
\bottomrule
\end{tabular}
\end{table}

\section{Recent Advances from the Mariana Trench}

Displayed equations are centered and set on a separate
line.
\begin{equation}
x + y = z
\end{equation}
Please try to avoid rasterized images for line-art diagrams and
schemas. Whenever possible, use vector graphics instead (see
Fig.\@ \ref{fig1}).

\begin{figure}[t]
\includegraphics[width=\textwidth]{example-image-duck}
\caption{A figure caption is always placed below the illustration.
Please note that short captions are centered, while long ones are
justified by the macro package automatically.} \label{fig1}
\end{figure}
%Note that the \includegraphics command is commented out in this example figure, so that we can compile this document without sending sample figures around.

\begin{theorem}
This is a sample theorem. The run-in heading is set in bold, while
the following text appears in italics. Definitions, lemmas,
propositions, and corollaries are styled the same way.
\end{theorem}
%
% the environments 'definition', 'lemma', 'proposition', 'corollary',
% 'remark', and 'example' are defined in the LLNCS documentclass as well.
%
\begin{proof}
Proofs, examples, and remarks have the initial word in italics,
while the following text appears in normal font.
\end{proof}
For citations of references, we prefer the use of square brackets
and consecutive numbers. Citations using labels or the author/year
convention are also acceptable. The following bibliography provides
a sample reference list with entries for journal
articles~\cite{ref_article1}, an LNCS chapter~\cite{ref_lncs1}, a
book~\cite{ref_book1}, proceedings without editors~\cite{ref_proc1},
and a homepage~\cite{ref_url1}. Multiple citations are grouped
\cite{ref_article1,ref_lncs1,ref_book1},
\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.

\section{Experiments}
\subsection{Experimental Setup}
\subsection{Experimental Results}

\section{Discussion}

\section{Conclusion}

Of course, authors have complete freedom on how they choose to structure their paper. Section headers from Introduction up to and including Conclusions are completely up to the discretion of the authors; use whichever structure you see fit. Title, Abstract, the credits environment, and References, however, are mandatory.

\begin{credits}
\subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
used for general acknowledgments, for example: This study was funded
by X (grant number Y).

\subsubsection{\discintname}
It is now necessary to declare any competing interests or to specifically
state that the authors have no competing interests. Please place the
statement with a bold run-in heading in small font size beneath the
(optional) acknowledgments,
for example: The authors have no competing interests to declare that are
relevant to the content of this article. Or: Author A has received research
grants from Company W. Author B has received a speaker honorarium from
Company X and owns stock in Company Y. Author C is a member of committee Z.
\end{credits}
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%% Note that this preceding line implies that you store your BibTeX references in a file called 'mybibliography.bib'. If you instead store your references in a file with a different name, for instance 'references.bib', the preceding line should read '\bibliography{references}'. Whatever you do, DO NOT put the file name extension .bib inside the \bibliography command; this will trip up LaTeX compilers. 
%
% If you do not want to use BibTeX, you can also type up the bibliography exactly as you see fit, using the following structure:
\begin{thebibliography}{8}
% Note that this number 8 reserves an amount of space (equal to the natural width of the given number) for the label of your references; if you have more than 9 references, you will want to change this number to 18. If you have more than 19 references, this number is best changed to 88. If you have more than 99 references, I salute you.
\bibitem{ref_article1}
Author, F.: Article title. Journal \textbf{2}(5), 99--110 (2016)

\bibitem{ref_lncs1}
Author, F., Author, S.: Title of a proceedings paper. In: Editor,
F., Editor, S. (eds.) CONFERENCE 2016, LNCS, vol. 9999, pp. 1--13.
Springer, Heidelberg (2016). \doi{10.10007/1234567890}

\bibitem{ref_book1}
Author, F., Author, S., Author, T.: Book title. 2nd edn. Publisher,
Location (1999)

\bibitem{ref_proc1}
Author, A.-B.: Contribution title. In: 9th International Proceedings
on Proceedings, pp. 1--2. Publisher, Location (2010)

\bibitem{ref_url1}
LNCS Homepage, \url{http://www.springer.com/lncs}, last accessed 2023/10/25
\end{thebibliography}
\end{document}

### 2025-02-12 paper/sections/intro.qmd

@@
 
-Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern AIs are tasked with learning these representations from scratch, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
-Modern advances in computing have made it possible to provide such AIs with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models.
+Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning these representations from scratch, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
+Modern advances in computing have made it possible to provide such models with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models.
 Unfortunately, in doing so they also learn increasingly complex and highly sensitive representations that we can no longer easily interpret.
@@
 
-Part of the problem is that high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].
+Part of the problem is that high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case]^[For clarity: we follow standard ML convention in using "degrees of freedom" to refer to the number of parameters estimated from data.].
 Based purely on predictive performance, these solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns.
-This poses two related challenges: firstly, it makes these models inherently opaque, since humans cannot simply interpret what type of explanation the complex learned representations correspond to; secondly, even if we could resolve the first challenge, it is not obvious how to mitigate models from learning representations that correspond to meaningless and undesirable explanations.
+This poses two related challenges: firstly, it makes these models inherently opaque, since humans cannot simply interpret what type of explanation the complex learned representations correspond to; secondly, even if we could resolve the first challenge, it is not obvious how to mitigate models from learning representations that correspond to meaningless and implausible explanations.
 
@@
 To our surprise, the second challenge has not yet attracted any consolidated research effort.
-Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are interpretable and deemed desirable by humans.
-Instead, the choice has typically been to improve the capacity of XAI tools to identify the subset explanations that are both desirable and valid for any given model, independent of whether the learned representations are also compatible with undesirable explanations [@altmeyer2024faithful].
+Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are interpretable and deemed **plausible** by humans (see @def-explainability).
+Instead, the choice has typically been to improve the capacity of XAI tools to identify the subset explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with implausible explanations [@altmeyer2024faithful].
 Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization and generative capacity [@schut2021generating, @augustin2020adversarial, @altmeyer2024faithful].
 
-Building on these findings, we introduce **counterfactual training**: a novel regularization technique geared explicitly towards aligning model representations with desirable explanations.
+Building on these findings, we introduce **counterfactual training**: a novel regularization technique geared explicitly towards aligning model representations with plausible explanations.
 Our contributions are as follows:

### 2025-02-12 paper/sections/method.qmd

@@
 This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments.
-Under certain conditions, this results holds theoretically^[For the proof, see the supplementary appendix.]:
+Under certain conditions, this results holds theoretically[For the proof, see the supplementary appendix.]:
 
@@
 
-Let $\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}$ denote a linear classifier with $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
-If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty [@eq-div] will result in lower classifier sensitivity to that feature, provided that at least one other feature is mutable.
+Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e.
+*multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
+If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is mutable.
 

### 2025-02-12 paper/sections/other/constraints.qmd

@@
 
-By @eq-loglike-decomp we have the $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$ and $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$, where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.
+By the properties of our classifier (*multinomial logistic regression*), we have:
+
+$$
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
+$$ {#eq-multi}
+
+where $\theta_{k,d}=\Theta[k,d]$ denotes the coefficient on feature $d$ for class $k$.
+
+Based on @eq-loglike-decomp and @eq-multi we can identify that $(\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})$ under the assumptions we made above.
+Hence, we have that $(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0$ and $(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0$
 
@@
 
-Since $\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}$ we are more likely to have $x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}} < 0$ than vice versa at initialization.
-Similarly, we are more likely to have $x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}} > 0$ by $\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}$.
+Since $(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})$ we are more likely to have $(x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0$ than vice versa at initialization.
+Similarly, we are more likely to have $(x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0$ since $(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})$.
 
-This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will push down on $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}} < 0$.
+This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will push down on $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0$.
 In words, not protecting the immutable feature would have the undesirable effect of making the classifier more sensitive to this feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for lower values of $d_{\text{immtbl}}$.
 
-By the same rationale, the contrastive divergence penalty can generally be expected to push up on $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}} > 0$.
+By the same rationale, the contrastive divergence penalty can generally be expected to push up on $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0$.
 In words, this has the effect of making the classifier more sensitive to the mutable feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for higher values of $d_{\text{mtbl}}$.

### 2025-02-12 paper/sections/related.qmd

@@
 To the best of our knowledge, our proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability.
-In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed desirable with respect to the underlying data and stakeholder requirements.
+In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and stakeholder requirements.
 To make this more concrete, we follow @augustin2020adversarial in tieing the concept of explainability to the quality of counterfactual explanations that we can generate for a given model.
@@
 We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom: by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e.
-adversarial examples), AT effectively removes some undesirable explanations from the solution space.
+adversarial examples), AT effectively removes some implausible explanations from the solution space.
 
@@
 
-Improving the adversarial robustness of models is not the only path towards aligning representations with desirable explanations.
+Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations.
 In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives.
@@
 Formally, they rely on tools from energy-based modelling to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
-Their proposed counterfactual explainer, *ECCCo*, yields desirable (or **plausible**) explanations if and only if the underlying model has learned representations that align with them.
+Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
 They find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.
@@
 Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data.
-Previous work by @schut2021generating similarly found that generating desirable ("interpretable") counterfactual explanations is almost trivial for deep ensembles that have also undergone adversarial training.
+Previous work by @schut2021generating similarly found that generating plausible ("interpretable") counterfactual explanations is almost trivial for deep ensembles that have also undergone adversarial training.
 The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your].
-This is closely related to the idea of aligning models with desirable explanations and has inspired our proposed counterfactual training objective, as we explain in @sec-method.
+This is closely related to the idea of aligning models with plausible explanations and has inspired our proposed counterfactual training objective, as we explain in @sec-method.

### 2025-02-14 paper/sections/other/constraints.qmd
working on key chart

@@
 
-function plot_db(model, x)
+function plot_db(model, x; _xlab="Existing Debt", _ylab="Age")
+
+    xlab = _constraint[1] == "both" ? "$(_xlab) (mutable)" : "$(_xlab) (immutable)"
+    ylab = _constraint[2] == "both" ? "$(_ylab) (mutable)" : "$(_ylab) (immutable)"
+
+    # Data and decision boundary:
     coeff = get_log_reg_params(model)
     db = get_decision_boundary(coeff)
-    plt = Plots.scatter(ce_data)
+    plt = Plots.scatter(ce_data, label=["1=Default" "2=No Default"], xlab=xlab, ylab=ylab, title="Iteration $(_itercount)")
+    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
+
+    # Counterfactuals:
+    if !any(isnothing.(x))
         yhat = [argmax(y) for y in eachcol(model(x))]
-    Plots.scatter!(x[1,:], x[2,:], label="Counterfactuals", ms=15, shape=:star, color=yhat)
-    Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
+        Plots.scatter!(x[1,:], x[2,:], label=["CE (y⁺=1)" "CE (y⁺=2)"], ms=15, shape=:star, color=yhat, group=yhat)
+    end
     display(plt)
+
+    if _itercount == 1
+        push!(_plts, plt)
+    end
+    global _itercount += 1
 end
@@
 # Data:
-constraints = [
-    ["both", "both"],
-    ["both", "none"]
+specs = [
+    (VanillaObjective(needs_ce=true), ["both", "both"]),
+    (EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
+    (EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),
 ]
@@
 models = []
+plts = []
 
-for constraint in constraints
+for (obj, constraint) in specs
+
+    # Globals for the callback:
+    global _constraint = constraint
+    global _plts = []
+    global _itercount = 1
+
     # Data:
@@
     )
-    ce_data = get_ce_data(data)
+    global ce_data = get_ce_data(data)
     val_size = data.n_validation / (data.n_validation + data.n_train)
@@
     # Objective:
-    obj = EnergyDifferentialObjective(lambda=[1.0,1.0,0.0])
     generator = GenericGenerator()
@@
         val_set = val_set,
-        nepochs = 50,
+        nepochs = 10,
         mutability = Symbol.(constraint),
@@
     push!(models, model)
+    push!(_plts, current())
+    push!(plts, _plts)
 end

### 2025-02-14 src/counterfactuals.jl

@@ -7,8 +7,6 @@ using Flux
 using StatsBase
 using TaijaParallel
 
-function generate!() end
-
 """
     generate!(
         model,
@@ -56,35 +54,15 @@ function generate!(
         verbose=verbose > 1,
     )
 
-    counterfactuals = (ce -> ce.counterfactual).(ces)                               # get actual counterfactuals
-    targets = (ce -> ce.target).(ces)
-
-    # Get neighbours in target class: find `nneighbours` potential neighbours than randomly choose one for each counterfactual.
-    neighbours =
-        (
-            ce -> find_potential_neighbours(ce, counterfactual_data, nneighbours)[
-                :, rand(1:end)
-            ]
-        ).(ces)
-
-    # Adjust for mutability:
-    mtblty = counterfactual_data.mutability
-    if !isnothing(mtblty)
-        for (i, ce) in enumerate(counterfactuals)
-            immtble = findall(mtblty .!= :both)
-            for j in immtble
-                # println(counterfactuals[i])
-                counterfactuals[i][j, :] = neighbours[i][j, :]
-                # println(counterfactuals[i])
-            end
-        end
-    end
+    counterfactuals = (ce -> ce.counterfactual).(ces)                                   # get actual counterfactuals
+    targets = (ce -> ce.target).(ces)                                                   # get targets
+    neighbours = (ce -> find_potential_neighbours(ce, counterfactual_data, 1)).(ces)    # randomly draw a sample from the target class
+    protect_immutable!(neighbours, counterfactuals, counterfactual_data.mutability)     # adjust for mutability
 
     aversarial_targets = []
     targets_enc = []
     percent_valid = 0.0
     idx = sample(1:length(ces), 10; replace=false)
-    # @info "Index: $idx"
     for (i, ce) in enumerate(ces)
         target_enc = target_encoded(ce, counterfactual_data)
         push!(targets_enc, target_enc)
@@ -99,7 +77,6 @@ function generate!(
     end
 
     n_total = length(counterfactuals)
-    # @info "Counter: $(percent_valid)"
     percent_valid = percent_valid / n_total
     group_indices = TaijaParallel.split_obs(1:n_total, length(data))
 
@@ -117,6 +94,54 @@ function generate!(
     return dl, percent_valid
 end
 
+"""
+    protect_immutable!(
+        samples::AbstractArray,
+        counterfactuals::AbstractArray,
+        mutability::Union{Nothing,AbstractArray},
[…]

### 2025-02-14 paper/sections/other/constraints.qmd
more work on plot

@@
 using CTExperiments.TaijaPlotting
+using Plots.PlotMeasures
 ```
@@
     db = get_decision_boundary(coeff)
-    plt = Plots.scatter(ce_data, label=["1=Default" "2=No Default"], xlab=xlab, ylab=ylab, title="Iteration $(_itercount)")
+    plt = Plots.scatter(
+        ce_data, 
+        title="Iteration $(_itercount)",
+        # label=["1=Default" "2=No Default"], 
+        # legend_position=:topright, 
+        bottom_margin = 5mm, 
+        left_margin = 5mm, 
+        xlab=xlab, ylab=ylab,  
+        axis=nothing, 
+        legend=false
+    )
     Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
+
+    # # Gradient field:
+    # u = []
+    # v = []
+    # for (x, y) in inputs
+    #     g = -loss_grads(generator, M, ce, [x, y][:,:])
+    #     push!(u, η * g[1])
+    #     push!(v, η * g[2])
+    # end
+    # Plots.quiver!(x1, x2, quiver=(u, v))
 
@@
 
-    if _itercount == 1
+    if _itercount == 5
         push!(_plts, plt)
@@
 specs = [
-    (VanillaObjective(needs_ce=true), ["both", "both"]),
-    (EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
-    (EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),
+    ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
+    ("(b)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
+    ("(c)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),
 ]
@@
 
-for (obj, constraint) in specs
+for (title, obj, constraint) in specs
 
@@
         val_set = val_set,
-        nepochs = 10,
+        nepochs = 50,
         mutability = Symbol.(constraint),
@@
     push!(_plts, current())
-    push!(plts, _plts)
+    plt = Plots.plot(_plts..., plot_title = title, layout=(2,1))
+    push!(plts, plt)
 end
@@
 
-using Plots.PlotMeasures
 plts = []

### 2025-02-14 paper/sections/other/constraints.qmd
mmm trying to do grads with respect to implausibility but not clear how to plot

@@
 
-    # # Gradient field:
-    # u = []
-    # v = []
-    # for (x, y) in inputs
-    #     g = -loss_grads(generator, M, ce, [x, y][:,:])
-    #     push!(u, η * g[1])
-    #     push!(v, η * g[2])
-    # end
-    # Plots.quiver!(x1, x2, quiver=(u, v))
+    # Gradient field:
+    targets = Flux.onehotbatch(fill(2,size(x,2)), [1,2])
+    samples = ce_data.X[:,ce_data.output_encoder.labels.==2][:,rand(1:end,size(x,2))]
+    u = []
+    v = []
+    η = 0.1
+    for (i,ce) in enumerate(eachcol(x))
+        grads = Flux.gradient((m,x,s,t) -> implausibility(m,x,s,t)[1], model, ce, samples[:,i], targets[:,i])
+        g = grads[1].layers[1].weight
+        push!(u, η * g[1])
+        push!(v, η * g[2])
+    end
+    Plots.quiver!(x1, x2, quiver=(u, v))
 
@@
         yhat = [argmax(y) for y in eachcol(model(x))]
-        Plots.scatter!(x[1,:], x[2,:], label=["CE (y⁺=1)" "CE (y⁺=2)"], ms=15, shape=:star, color=yhat, group=yhat)
+        if any(yhat.==2)
+            Plots.scatter!(x[1,yhat.==2], x[2,yhat.==2], label=["CE (y⁺=1)" "CE (y⁺=2)"], ms=15, shape=:star, color=yhat[yhat.==2], group=yhat[yhat.==2])
+        end
     end
@@
         val_set = val_set,
-        nepochs = 50,
+        nepochs = 5,
         mutability = Symbol.(constraint),

### 2025-02-14 src/implausibility.jl

@@ -7,9 +7,8 @@ using LinearAlgebra
 Compute the implausibility (contrastive divergence) of the counterfactuals (`counterfactual`) with respect to `samples` in the target class. This is computed as the difference between negative logits indexed at the target class for the `samples` and the `counterfactual`.
 """
 function implausibility(model, counterfactual, samples, targets)
-    E(x) = -model(x)                                            # energy
-    x = ((E(samples)) - (E(counterfactual)))[:, :]'targets       # contrastive divergence
-    # x = ((E(counterfactual)) - (E(samples)))'targets   # contrastive divergence
+    E(x) = -model(x)                                        # energy
+    x = ((E(samples)) - (E(counterfactual)))[:, :]'targets  # contrastive divergence
     return diag(x[:, :])
 end
 

### 2025-02-14 paper/sections/other/constraints.qmd
uf

@@
     samples = ce_data.X[:,ce_data.output_encoder.labels.==2][:,rand(1:end,size(x,2))]
+    CounterfactualTraining.protect_immutable!(eachcol(samples), eachcol(x), ce_data.mutability)
     u = []
     v = []
-    η = 0.1
+    η = 0.5
     for (i,ce) in enumerate(eachcol(x))
         grads = Flux.gradient((m,x,s,t) -> implausibility(m,x,s,t)[1], model, ce, samples[:,i], targets[:,i])
-        g = grads[1].layers[1].weight
-        push!(u, η * g[1])
-        push!(v, η * g[2])
+        g = CTExperiments.get_log_reg_params(grads[1].layers[1])
+        push!(u, η * g[2])
+        push!(v, η * g[3])
     end
-    Plots.quiver!(x1, x2, quiver=(u, v))
+    Plots.quiver!(x[1,:], x[2,:], quiver=(u, v))
 
@@
 specs = [
-    ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
-    ("(b)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
+    # ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
+    # ("(b)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
     ("(c)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),
@@
         val_set = val_set,
-        nepochs = 5,
+        nepochs = 25,
         mutability = Symbol.(constraint),

### 2025-02-14 paper/experiments/CTExperiments/src/experiment.jl

@@ -393,11 +393,15 @@ function get_log_reg_params(exper::Experiment)
     return coeffs
 end
 
-function get_log_reg_params(model)
+function get_log_reg_params(model::Flux.Chain)
     coeffs =
         Flux.params(model) |>
-        W -> (β₀=W[2][1] - W[2][2], β₁=W[1][1, 1] - W[1][2, 1], β₂=W[1][1, 2] - W[1][2, 2])
+        get_log_reg_params
     return coeffs
+end
+
+function get_log_reg_params(W)
+    return (β₀=W[2][1] - W[2][2], β₁=W[1][1, 1] - W[1][2, 1], β₂=W[1][1, 2] - W[1][2, 2])
 end
 
 struct DecisionBoundary

### 2025-02-14 paper/sections/other/constraints.qmd
don't think this is worth it

@@
 specs = [
-    # ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
-    # ("(b)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
+    ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
+    ("(b)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
     ("(c)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),

### 2025-02-14 paper/body_ecml.tex
fucking latex

@@
 Section~\ref{sec-discussion}.
+
+\subsection{Illustration}\label{illustration}
+
+To better convey the intuition underlying our proposed method, we
+illustrate different model outcomes in Example~\ref{exm-grad}.
+
+\begin{example}[Prediction of Consumer Credit
+Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}
+
+Suppose we are interested in predicting the likelihood that loan
+applicants default on their credit. We have access to historical data on
+previous loan takers comprised of a binary outcome variable
+(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
+1) the subjects' \emph{age}, which we define as immutable, and 2) the
+subjects' existing level of \emph{debt}, which we define as mutable.
+
+We have simulated this scenario using synthetic data with independent
+features and Gaussian class-conditional densities in
+Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
+outcomes for different training procedures using the same model
+architecture each time (a linear classifier). In each case, we show the
+linear decision boundary (green) and the training data colored according
+to their ground-truth label: orange points belong to the target class,
+\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
+indicate counterfactuals in the target class generated at test time
+using generic gradient descent for a fixed number of iterations.
+
+In panel (a), we have trained our model conventionally, and we do not
+impose mutability constraints at test time. The generated
+counterfactuals are all valid, but not plausible: they are clearly
+distinguishable from the ground-truth data. In panel (b), we have
+trained our model with counterfactual training, once again not imposing
+mutability constraints at test time. We observe that the counterfactuals
+are clearly plausible, therefore meeting the first objective of
+Definition~\ref{def-explainability}.
+
+In panel (c), we have used conventional training again, this time
+imposing the mutability constraint on \emph{age} at test time.
+Counterfactuals are valid but involve some substantial reductions in
+\emph{debt} for some individuals. By comparison, counterfactual paths
+are shorter on average in panel (d), where we have used counterfactual
+training and protected immutable features as described in
+Section~\ref{sec-constraints}. The counterfactuals are also plausible
+with respect to the mutable feature. Thus, we consider the model in
+panel (d) as the most explainable according to
+Definition~\ref{def-explainability}.
+
+\end{example}
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{body_ecml_files/mediabag/../paper/figures/poc.pdf}}
+
+}
+
+\caption{\label{fig-poc}Visual illustration of how counterfactual
+training improves explainability. See Example~\ref{exm-grad} for
+details.}
+
+\end{figure}%
 

### 2025-02-14 paper/preprint/paper.tex
fucking latex

@@
 \usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{plain}
 \newtheorem{theorem}{Theorem}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{definition}
@@
 \theoremstyle{definition}
-\newtheorem{example}{Example}[section]
+\newtheorem{exercise}{Parameters}[section]
 \theoremstyle{remark}
@@
 
+\subsection{Illustration}\label{illustration}
+
+To better convey the intuition underlying our proposed method, we
+illustrate different model outcomes in Example~\ref{exm-grad}.
+
+\begin{example}[Prediction of Consumer Credit
+Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}
+
+Suppose we are interested in predicting the likelihood that loan
+applicants default on their credit. We have access to historical data on
+previous loan takers comprised of a binary outcome variable
+(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
+1) the subjects' \emph{age}, which we define as immutable, and 2) the
+subjects' existing level of \emph{debt}, which we define as mutable.
+
+We have simulated this scenario using synthetic data with independent
+features and Gaussian class-conditional densities in
+Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
+outcomes for different training procedures using the same model
+architecture each time (a linear classifier). In each case, we show the
+linear decision boundary (green) and the training data colored according
+to their ground-truth label: orange points belong to the target class,
+\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
+indicate counterfactuals in the target class generated at test time
+using generic gradient descent for a fixed number of iterations.
+
+In panel (a), we have trained our model conventionally, and we do not
+impose mutability constraints at test time. The generated
+counterfactuals are all valid, but not plausible: they are clearly
+distinguishable from the ground-truth data. In panel (b), we have
+trained our model with counterfactual training, once again not imposing
+mutability constraints at test time. We observe that the counterfactuals
+are clearly plausible, therefore meeting the first objective of
+Definition~\ref{def-explainability}.
+
+In panel (c), we have used conventional training again, this time
+imposing the mutability constraint on \emph{age} at test time.
+Counterfactuals are valid but involve some substantial reductions in
+\emph{debt} for some individuals. By comparison, counterfactual paths
+are shorter on average in panel (d), where we have used counterfactual
+training and protected immutable features as described in
+Section~\ref{sec-constraints}. The counterfactuals are also plausible
+with respect to the mutable feature. Thus, we consider the model in
+panel (d) as the most explainable according to
+Definition~\ref{def-explainability}.
+
+\end{example}
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
+
+}
+
+\caption{\label{fig-poc}Visual illustration of how counterfactual
+training improves explainability. See Example~\ref{exm-grad} for
+details.}
+
+\end{figure}%
+
 \section{Experiments}\label{sec-experiments}
@@
 
-We provide an illustrative example in Example~\ref{exm-grad}.
-
-\begin{example}[Prediction of Consumer Credit
-Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}
-
-Suppose now that \(d_{\text{immtbl}}\) represents an individual's
-\emph{age} and \(d_{\text{mtbl}}\) represents an individual's existing
-level of credit card debt. Assume that these two features are
-independent, the class conditional densities are Gaussian and we use no
-other features to predict the risk of individuals defaulting on a
-consumer loan using a linear classifier. We have simulated this scenario
-using synthetic data in Figure~\ref{fig-mtblty}.
-
-In panel (a) of Figure~\ref{fig-mtblty}, we have trained the linear
-classifier using counterfactual training, treating both features as
-mutable. The linear decision boundary is roughly equally sensitive to
-both \emph{age} and existing levels of \emph{debt}.
-
-Conversely, in panel (b) of Figure~\ref{fig-mtblty}, we have trained the
-same classifier using counterfactual training, but this time treating
-\emph{age} as an immutable feature. The result is a new decision
-boundary that has tilted in favor of higher sensitivity to the mutable
-feature (existing \emph{debt}) and lower sensitivity to the immutable
-feature (\emph{age}).
-
-\end{example}
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/app_mtblty.pdf}}
-
-}
-
-\caption{\label{fig-mtblty}Visual illustration of the effect of imposing
-mutability constraints. See Example~\ref{exm-grad} for details.}
-
-\end{figure}%
-
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomtitle=1mm, coltitle=black, opacitybacktitle=0.6, rightrule=.15mm, toprule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, colback=white, breakable, left=2mm, leftrule=.75mm, titlerule=0mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, bottomrule=.15mm, opacityback=0, colframe=quarto-callout-warning-color-frame]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomtitle=1mm, coltitle=black, opacitybacktitle=0.6, rightrule=.15mm, toprule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, colback=white, breakable, left=2mm, leftrule=.75mm, titlerule=0mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, bottomrule=.15mm, opacityback=0, colframe=quarto-callout-warning-color-frame]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colback=white, opacityback=0, coltitle=black, arc=.35mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, breakable, opacitybacktitle=0.6, rightrule=.15mm, left=2mm, bottomtitle=1mm, titlerule=0mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomtitle=1mm, coltitle=black, opacitybacktitle=0.6, rightrule=.15mm, toprule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, leftrule=.75mm, titlerule=0mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, bottomrule=.15mm, opacityback=0, colframe=quarto-callout-note-color-frame]
 
@@
   full & 0.1 & \textit{ECCo} & $-6.74 \cdot 10^{5}$ & $1.53 \cdot 10^{6}$ \\
-  full & 0.1 & \textit{Generic} & $-1.72 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\
+  full & 0.1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\
   \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.56}} & \color{blue}{\textbf{0.124}} \\
@@
   full & 0.001 & \textit{ECCo} & -362 & 811 \\
-  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.03 \cdot 10^{17}$ \\
+  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.04 \cdot 10^{17}$ \\
   \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.49}} & \color{blue}{\textbf{0.115}} \\
@@
   full & 0.1 & \textit{ECCo} & $-3.72 \cdot 10^{11}$ & $8.46 \cdot 10^{11}$ \\
-  full & 0.1 & \textit{Generic} & $-4.49 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
+  full & 0.1 & \textit{Generic} & $-4.48 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
   \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5}} & \color{blue}{\textbf{0.112}} \\

### 2025-02-14 paper/sections/method.qmd
fucking latex

@@
 We discuss this limitation in @sec-discussion.
+
+## Illustration
+
+To better convey the intuition underlying our proposed method, we illustrate different model outcomes in @exm-grad.
+
+::: {#exm-grad}
+
+## Prediction of Consumer Credit Default
+
+Suppose we are interested in predicting the likelihood that loan applicants default on their credit.
+We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) two input features: 1) the subjects' *age*, which we define as immutable, and 2) the subjects' existing level of *debt*, which we define as mutable.
+
+We have simulated this scenario using synthetic data with independent features and Gaussian class-conditional densities in @fig-poc.
+The four panels in @fig-poc show the outcomes for different training procedures using the same model architecture each time (a linear classifier).
+In each case, we show the linear decision boundary (green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$.
+Stars indicate counterfactuals in the target class generated at test time using generic gradient descent for a fixed number of iterations.
+
+In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time.
+The generated counterfactuals are all valid, but not plausible: they are clearly distinguishable from the ground-truth data.
+In panel (b), we have trained our model with counterfactual training, once again not imposing mutability constraints at test time.
+We observe that the counterfactuals are clearly plausible, therefore meeting the first objective of @def-explainability.
+
+In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time.
+Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals.
+By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected immutable features as described in @sec-constraints.
+The counterfactuals are also plausible with respect to the mutable feature.
+Thus, we consider the model in panel (d) as the most explainable according to @def-explainability.
+
+:::
+
+![Visual illustration of how counterfactual training improves explainability. See @exm-grad for details.](/paper/figures/poc.svg){#fig-poc}

### 2025-02-14 paper/sections/other/constraints.qmd
fucking latex

--- 2025-02-14 paper/sections/other/constraints.qmd
@@
 
-We provide an illustrative example in @exm-grad.
-
-::: {#exm-grad}
-
-## Prediction of Consumer Credit Default
-
-Suppose now that $d_{\text{immtbl}}$ represents an individual's *age* and $d_{\text{mtbl}}$ represents an individual's existing level of credit card debt.
-Assume that these two features are independent, the class conditional densities are Gaussian and we use no other features to predict the risk of individuals defaulting on a consumer loan using a linear classifier.
-We have simulated this scenario using synthetic data in @fig-mtblty.
-
-In panel (a) of @fig-mtblty, we have trained the linear classifier using counterfactual training, treating both features as mutable.
-The linear decision boundary is roughly equally sensitive to both *age* and existing levels of *debt*.
-
-Conversely, in panel (b) of @fig-mtblty, we have trained the same classifier using counterfactual training, but this time treating *age* as an immutable feature.
-The result is a new decision boundary that has tilted in favor of higher sensitivity to the mutable feature (existing *debt*) and lower sensitivity to the immutable feature (*age*).
-
-:::
-
-![Visual illustration of the effect of imposing mutability constraints. See @exm-grad for details.](/paper/figures/app_mtblty.svg){#fig-mtblty}
-
 ::: {.callout-warning}
@@
 
-function plot_db(model, x; _xlab="Existing Debt", _ylab="Age")
+function plot_db(model, ces; _xlab="Debt", _ylab="Age")
+
+    x = (ce -> ce.counterfactual).(ces) |> xs -> reduce(hcat, xs)
+    x0 = (ce -> ce.factual).(ces) |> xs -> reduce(hcat, xs)
 
@@
         ce_data, 
-        title="Iteration $(_itercount)",
         # label=["1=Default" "2=No Default"], 
         # legend_position=:topright, 
-        bottom_margin = 5mm, 
-        left_margin = 5mm, 
         xlab=xlab, ylab=ylab,  
         axis=nothing, 
-        legend=false
+        legend=false,
+        title=_title
     )
     Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
-
-    # Gradient field:
-    targets = Flux.onehotbatch(fill(2,size(x,2)), [1,2])
-    samples = ce_data.X[:,ce_data.output_encoder.labels.==2][:,rand(1:end,size(x,2))]
-    u = []
-    v = []
-    η = 0.1
-    for (i,ce) in enumerate(eachcol(x))
-        grads = Flux.gradient((m,x,s,t) -> implausibility(m,x,s,t)[1], model, ce, samples[:,i], targets[:,i])
-        g = grads[1].layers[1].weight
-        push!(u, η * g[1])
-        push!(v, η * g[2])
-    end
-    Plots.quiver!(x1, x2, quiver=(u, v))
 
@@
         yhat = [argmax(y) for y in eachcol(model(x))]
+        yhat0 = [argmax(y) for y in eachcol(model(x0))]
         if any(yhat.==2)
-            Plots.scatter!(x[1,yhat.==2], x[2,yhat.==2], label=["CE (y⁺=1)" "CE (y⁺=2)"], ms=15, shape=:star, color=yhat[yhat.==2], group=yhat[yhat.==2])
+
+            # Paths:
+            u = []
+            v = []
+            for (i,ce) in enumerate(eachcol(x))
+                Δ = ce - x0[:,i]
+                push!(u, Δ[1])
+                push!(v, Δ[2])
+            end
+            Plots.quiver!(x0[1,yhat.==2], x0[2,yhat.==2], quiver=(u[yhat.==2], v[yhat.==2]), color=1)
+
+            # End points:
+            Plots.scatter!(x[1,yhat.==2], x[2,yhat.==2], label=["CE (y⁺=1)" "CE (y⁺=2)"], ms=15, shape=:star, color=yhat[yhat.==2], group=yhat[yhat.==2], mscolor=yhat0[yhat.==2])
+
         end
@@
     display(plt)
-
-    if _itercount == 5
-        push!(_plts, plt)
-    end
-    global _itercount += 1
 end
@@
     ("(b)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
-    ("(c)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),
+    ("(c)", VanillaObjective(needs_ce=true), ["both", "none"]),
+    ("(d)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),
 ]
@@
     # Globals for the callback:
+    global _title = title
     global _constraint = constraint
-    global _plts = []
-    global _itercount = 1
 
@@
         val_set = val_set,
-        nepochs = 5,
+        nepochs = 50,
         mutability = Symbol.(constraint),
@@
     push!(models, model)
-    push!(_plts, current())
-    plt = Plots.plot(_plts..., plot_title = title, layout=(2,1))
-    push!(plts, plt)
+    push!(plts, current())
 end
+```
+
+```{julia}
+#| eval: false
+
+plt = Plots.plot(
+    plts..., 
+    layout=(1,4), 
+    size=(1150,250),  
+    left_margin = 10mm, 
+    bottom_margin = 5mm,
+    top_margin = 3mm,
+    right_margin = 10mm,
+)
+display(plt)
+Plots.savefig(plt, "paper/figures/poc.svg")
 ```

### 2025-02-14 src/counterfactuals.jl

@@ -91,7 +91,7 @@ function generate!(
     ]
     @assert length(dl) == length(data)
 
-    return dl, percent_valid
+    return dl, percent_valid, ces
 end
 
 """

### 2025-02-15 paper/preprint/paper.tex
slowly slowly

@@
 \theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
-\theoremstyle{plain}
 \newtheorem{theorem}{Theorem}[section]
+\theoremstyle{definition}
+\newtheorem{exercise}{Parameters}[section]
 \theoremstyle{definition}
@@
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomtitle=1mm, coltitle=black, opacitybacktitle=0.6, rightrule=.15mm, toprule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, colback=white, breakable, left=2mm, leftrule=.75mm, titlerule=0mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, bottomrule=.15mm, opacityback=0, colframe=quarto-callout-warning-color-frame]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, breakable, left=2mm, coltitle=black, colbacktitle=quarto-callout-warning-color!10!white, titlerule=0mm, opacitybacktitle=0.6, colback=white, bottomtitle=1mm, toptitle=1mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toprule=.15mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, bottomtitle=1mm, coltitle=black, opacitybacktitle=0.6, rightrule=.15mm, toprule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, leftrule=.75mm, titlerule=0mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, bottomrule=.15mm, opacityback=0, colframe=quarto-callout-note-color-frame]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, breakable, left=2mm, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, opacitybacktitle=0.6, colback=white, bottomtitle=1mm, toptitle=1mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, toprule=.15mm, leftrule=.75mm]
 

### 2025-02-15 src/training.jl

@@ -106,7 +106,7 @@ function counterfactual_training(
 
             # Unpack:
             input, label = batch
-            perturbed_input, targets_enc, neighbours, adversarial_targets = counterfactual_dl[i]
+            perturbed_input, targets_enc, neighbours, perturbations, validities = counterfactual_dl[i]
 
             val, grads = Flux.withgradient(model) do m
 
@@ -119,9 +119,7 @@ function counterfactual_training(
                     regs = reg_loss(m, perturbed_input, neighbours, targets_enc)
                     # Validity loss (counterfactual):
                     yhat_ce = m(perturbed_input)
-                    adversarial_loss = loss.class_loss(
-                        yhat_ce, adversarial_targets
-                    )
+                    adversarial_loss = loss.class_loss(yhat_ce, targets_enc)
                 else
                     implaus = [0.0f0]
                     regs = [0.0f0]

### 2025-02-15 paper/preprint/paper.tex
interesting

@@
 \theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{definition}
@@
 \theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
+\newtheorem{theorem}{Theorem}[section]
 \theoremstyle{remark}
@@
 Counterfactuals are valid but involve some substantial reductions in
-\emph{debt} for some individuals. By comparison, counterfactual paths
-are shorter on average in panel (d), where we have used counterfactual
-training and protected immutable features as described in
-Section~\ref{sec-constraints}. The counterfactuals are also plausible
-with respect to the mutable feature. Thus, we consider the model in
-panel (d) as the most explainable according to
-Definition~\ref{def-explainability}.
+\emph{debt} for some individuals, in particular very young applicants.
+By comparison, counterfactual paths are shorter on average in panel (d),
+where we have used counterfactual training and protected immutable
+features as described in Section~\ref{sec-constraints}. In particular,
+we observe that due to the classifier's lower sensitivity to \emph{age},
+recourse recommendations with respect to \emph{debt} are much more
+homogenous, in that they do not disproportionately punish younger
+individuals. The counterfactuals are also plausible with respect to the
+mutable feature. Thus, we consider the model in panel (d) as the most
+explainable according to Definition~\ref{def-explainability}.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, breakable, left=2mm, coltitle=black, colbacktitle=quarto-callout-warning-color!10!white, titlerule=0mm, opacitybacktitle=0.6, colback=white, bottomtitle=1mm, toptitle=1mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toprule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, toprule=.15mm, opacityback=0, left=2mm, breakable, colback=white, bottomtitle=1mm, rightrule=.15mm, titlerule=0mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-warning-color!10!white, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, coltitle=black, toptitle=1mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, breakable, left=2mm, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, opacitybacktitle=0.6, colback=white, bottomtitle=1mm, toptitle=1mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, toprule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, opacityback=0, left=2mm, breakable, colback=white, bottomtitle=1mm, rightrule=.15mm, titlerule=0mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, coltitle=black, toptitle=1mm, arc=.35mm, bottomrule=.15mm]
 

### 2025-02-15 paper/sections/method.qmd
interesting

@@
 In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time.
-Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals.
+Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals, in particular very young applicants.
 By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected immutable features as described in @sec-constraints.
+In particular, we observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous, in that they do not disproportionately punish younger individuals.
 The counterfactuals are also plausible with respect to the mutable feature.

### 2025-02-15 paper/sections/other/constraints.qmd
interesting

@@
     # Counterfactuals:
+    target = 2
+    factual = 1
     if !any(isnothing.(x))
@@
         yhat0 = [argmax(y) for y in eachcol(model(x0))]
-        if any(yhat.==2)
+        idx_plotted = yhat.==target .&& yhat0.==factual
+        if any(idx_plotted)
 
@@
             end
-            Plots.quiver!(x0[1,yhat.==2], x0[2,yhat.==2], quiver=(u[yhat.==2], v[yhat.==2]), color=1)
+            Plots.quiver!(x0[1,idx_plotted], x0[2,idx_plotted], quiver=(u[idx_plotted], v[idx_plotted]), color=factual)
 
             # End points:
-            Plots.scatter!(x[1,yhat.==2], x[2,yhat.==2], label=["CE (y⁺=1)" "CE (y⁺=2)"], ms=15, shape=:star, color=yhat[yhat.==2], group=yhat[yhat.==2], mscolor=yhat0[yhat.==2])
+            Plots.scatter!(x[1,idx_plotted], x[2,idx_plotted], label=["CE (y⁺=1)" "CE (y⁺=2)"], ms=15, shape=:star, color=yhat[idx_plotted], group=yhat[idx_plotted], mscolor=yhat0[idx_plotted])
 
@@
     ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
-    ("(b)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "both"]),
+    ("(b)", FullObjective(lambda=[1.0,1.0,0.0,1.0]), ["both", "both"]),
     ("(c)", VanillaObjective(needs_ce=true), ["both", "none"]),
-    ("(d)", EnergyDifferentialObjective(lambda=[1.0,1.0,0.0]), ["both", "none"]),
+    ("(d)", FullObjective(lambda=[1.0,1.0,0.0,1.0]), ["both", "none"]),
 ]
@@
     opt_state = Flux.setup(Descent(), model)
-    conv = MaxIterConvergence()
+    conv = MaxIterConvergence(5)
 
@@
         convergence=conv, 
+        verbose=3
     )

### 2025-02-15 src/CounterfactualTraining.jl

@@ -1,7 +1,7 @@
 module CounterfactualTraining
 
 include("utils.jl")
-include("implausibility.jl")
+include("loss.jl")
 include("objectives.jl")
 include("counterfactuals.jl")
 export EnergyDifferentialObjective, AdversarialObjective, FullObjective, VanillaObjective

### 2025-02-15 src/implausibility.jl

@@ -1,18 +1,0 @@
-using EnergySamplers: EnergySamplers
-using LinearAlgebra
-
-@doc raw"""
-    implausibility(model, counterfactual, samples, targets)
-
-Compute the implausibility (contrastive divergence) of the counterfactuals (`counterfactual`) with respect to `samples` in the target class. This is computed as the difference between negative logits indexed at the target class for the `samples` and the `counterfactual`.
-"""
-function implausibility(model, counterfactual, samples, targets)
-    E(x) = -model(x)                                        # energy
-    x = ((E(samples)) - (E(counterfactual)))[:, :]'targets  # contrastive divergence
-    return diag(x[:, :])
-end
-
-function reg_loss(model, counterfactual, samples, targets)
-    x = (abs2.(model(samples)) + abs2.(model(counterfactual)))'targets
-    return diag(x[:, :])
-end

### 2025-02-15 src/loss.jl

@@ -1,0 +1,47 @@
+using EnergySamplers: EnergySamplers
+using Flux
+using LinearAlgebra
+
+@doc raw"""
+    implausibility(model, counterfactual, samples, targets)
+
+Compute the implausibility (contrastive divergence) of the counterfactuals (`counterfactual`) with respect to `samples` in the target class. This is computed as the difference between negative logits indexed at the target class for the `samples` and the `counterfactual`.
+"""
+function implausibility(model, counterfactual, samples, targets)
+    E(x) = -model(x)                                        # energy
+    x = ((E(samples)) - (E(counterfactual)))[:, :]'targets  # contrastive divergence
+    return diag(x[:, :])
+end
+
+"""
+    reg_loss(model, counterfactual, samples, targets)
+
+Compute the regularization loss for the contrastice divergence.
+"""
+function reg_loss(model, counterfactual, samples, targets)
+    x = (abs2.(model(samples)) + abs2.(model(counterfactual)))'targets
+    return diag(x[:, :])
+end
+
+"""
+    adv_loss(
+        model, counterfactual, perturbations, targets; epsilon=2.0, p::Real=Inf, validities=nothing
+    )
+
+Adversarial loss function.
+"""
+function adv_loss(
+    model, counterfactual, perturbations, targets; epsilon=0.5, p::Real=Inf, validities=nothing
+)
+    # Identify adversarial examples
+    idx_advexm = [isadvexm(perturbation, epsilon, p) for perturbation in eachcol(perturbations)]
+    if sum(idx_advexm) > 0
+        println("Percent AE: $(sum(idx_advexm)/length(idx_advexm))")
+        yhat_ce = model(counterfactual[:,idx_advexm])   # predictions
+        return Flux.logitcrossentropy(yhat_ce, targets[:,idx_advexm])
+    else 
+        return 0.0f0
+    end
+end
+
+isadvexm(perturbation, epsilon, p) = abs(norm(perturbation, p)) <= epsilon

### 2025-02-15 paper/sections/other/constraints.qmd
very interesting

@@
         yhat0 = [argmax(y) for y in eachcol(model(x0))]
-        idx_plotted = yhat.==target .&& yhat0.==factual
+        idx_plotted = yhat0.==factual
         if any(idx_plotted)

### 2025-02-16 paper/sections/method.qmd
final tweaks

@@
 $$
-\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y})
+\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
 $$ {#eq-obj}

### 2025-02-16 src/training.jl

@@ -106,7 +106,7 @@ function counterfactual_training(
 
             # Unpack:
             input, label = batch
-            perturbed_input, targets_enc, neighbours, perturbations, validities = counterfactual_dl[i]
+            perturbed_input, advexms, targets_enc, neighbours, perturbations, validities = counterfactual_dl[i]
 
             val, grads = Flux.withgradient(model) do m
 
@@ -117,13 +117,7 @@ function counterfactual_training(
                 if !isnothing(perturbed_input)
                     implaus = implausibility(m, perturbed_input, neighbours, targets_enc)
                     regs = reg_loss(m, perturbed_input, neighbours, targets_enc)
-                    adversarial_loss = adv_loss(
-                        m,
-                        perturbed_input,
-                        perturbations,
-                        targets_enc;
-                        validities=validities,
-                    )
+                    adversarial_loss = loss.class_loss(m(advexms), label)
                 else
                     implaus = [0.0f0]
                     regs = [0.0f0]

### 2025-02-16 paper/sections/other/constraints.qmd
final tweaks
beautiful

@@
     opt_state = Flux.setup(Descent(), model)
-    conv = MaxIterConvergence(5)
+    conv = MaxIterConvergence(25)
 
@@
         val_set = val_set,
-        nepochs = 50,
+        nepochs = 100,
         mutability = Symbol.(constraint),

### 2025-02-16 src/counterfactuals.jl

@@ -58,7 +58,7 @@ function generate!(
     )
 
     counterfactuals = (ce -> ce.counterfactual).(ces)                                   # get actual counterfactuals
-    advexms = (ce -> ce.search[:last_valid_ce])                                         # get adversarial example
+    advexms = (ce -> ce.search[:last_valid_ae]).(ces)                                   # get adversarial example
     perturbations = (ce -> ce.counterfactual - ce.factual).(ces)                        # get perturbations
     targets = (ce -> ce.target).(ces)                                                   # get targets
     neighbours = (ce -> find_potential_neighbours(ce, counterfactual_data, 1)).(ces)    # randomly draw a sample from the target class
@@ -77,8 +77,7 @@ function generate!(
             stack(hcat(advexms[i]...)),
             stack(hcat(targets_enc[i]...)),
             stack(hcat(neighbours[i]...)),
-            stack(hcat(perturbations[i]...)),
-            stack(hcat(validities[i]...)),
+            stack(hcat(eachcol(factual_enc)[i]...)),
         ) for i in group_indices
     ]
     @assert length(dl) == length(data)
@@ -94,9 +93,9 @@ A callback function used to store the last counterfactual that is also a valid a
 function get_last_valid_ae(ce::CounterfactualExplanation)
     # Find last counterfactual that meets imperceptability criterium:
     xs = ce.search[:path]
+    perturbations = [x - ce.factual for x in xs]
     aecrit = get_global_ae_criterium()
-    idx_advexm = [aecrit(x) for x in xs]
-    println(idx_advexm)
+    idx_advexm = [aecrit(x) for x in perturbations]
     last_valid_ae = xs[idx_advexm][end]
     ce.search[:last_valid_ae] = last_valid_ae
 end

### 2025-02-16 src/training.jl

@@ -106,7 +106,7 @@ function counterfactual_training(
 
             # Unpack:
             input, label = batch
-            perturbed_input, advexms, targets_enc, neighbours, perturbations, validities = counterfactual_dl[i]
+            perturbed_input, advexms, targets_enc, neighbours, factual_enc = counterfactual_dl[i]
 
             val, grads = Flux.withgradient(model) do m
 
@@ -117,7 +117,7 @@ function counterfactual_training(
                 if !isnothing(perturbed_input)
                     implaus = implausibility(m, perturbed_input, neighbours, targets_enc)
                     regs = reg_loss(m, perturbed_input, neighbours, targets_enc)
-                    adversarial_loss = loss.class_loss(m(advexms), label)
+                    adversarial_loss = loss.class_loss(m(advexms), factual_enc)
                 else
                     implaus = [0.0f0]
                     regs = [0.0f0]

### 2025-02-16 paper/preprint/paper.tex
nice

@@
 \usepackage{amsthm}
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{definition}
@@
 \newtheorem{theorem}{Theorem}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 
-To train models with high explainability as defined in
-Definition~\ref{def-explainability}, we propose the following objective,
+Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote iteratively
+generated counterfactuals. To train models with high explainability as
+defined in Definition~\ref{def-explainability}, we propose to leverage
+counterfactuals in the following objective,
 
 \begin{equation}\phantomsection\label{eq-obj}{
-\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y})
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
 }\end{equation}
 
-where \(\text{yloss}(\cdot)\) denotes any conventional classification
-loss function (e.g.~cross-entropy) that induces discriminative
-(predictive) performance. The two additional components in
-Equation~\ref{eq-obj} are explained in more detail below. For now, they
-can be sufficiently described as inducing explainability directly and
-indirectly by penalizing: 1) the contrastive divergence,
-\(\text{div}(\cdot)\), between counterfactuals \(x^\prime\) and observed
-samples \(x\) and, 2) the adversarial loss, \(\text{advloss}(.)\), with
-respect to counterfactuals. The tradeoff between the different
-components can be governed by adjusting the strengths of the penalties
-\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).
+where \(\text{yloss}(\cdot)\) is any conventional classification loss
+that induces discriminative performance (e.g.~cross-entropy). The two
+additional components in Equation~\ref{eq-obj} are explained in more
+detail below. For now, they can be sufficiently described as inducing
+explainability directly and indirectly by penalizing: 1) the contrastive
+divergence, \(\text{div}(\cdot)\), between counterfactuals
+\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, 2) the
+adversarial loss, \(\text{advloss}(.)\), with respect to nascent
+counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
+the different components can be governed by adjusting the strengths of
+the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, toprule=.15mm, opacityback=0, left=2mm, breakable, colback=white, bottomtitle=1mm, rightrule=.15mm, titlerule=0mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-warning-color!10!white, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, coltitle=black, toptitle=1mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, breakable, leftrule=.75mm, coltitle=black, bottomtitle=1mm, toprule=.15mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-warning-color!10!white, arc=.35mm, bottomrule=.15mm, left=2mm, titlerule=0mm, opacityback=0, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, opacityback=0, left=2mm, breakable, colback=white, bottomtitle=1mm, rightrule=.15mm, titlerule=0mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, coltitle=black, toptitle=1mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, breakable, leftrule=.75mm, coltitle=black, bottomtitle=1mm, toprule=.15mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, bottomrule=.15mm, left=2mm, titlerule=0mm, opacityback=0, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm]
 

### 2025-02-16 paper/sections/method.qmd
nice

--- 2025-02-15 paper/sections/method.qmd
@@
 
-To train models with high explainability as defined in @def-explainability, we propose the following objective,
+Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote iteratively generated counterfactuals.
+To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective,
 
 $$
-\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y})
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
 $$ {#eq-obj}
 
-where $\text{yloss}(\cdot)$ denotes any conventional classification loss function (e.g.
-cross-entropy) that induces discriminative (predictive) performance.
+where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g.
+cross-entropy).
 The two additional components in @eq-obj are explained in more detail below.
-For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $x^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to counterfactuals.
+For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
 The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.

### 2025-02-16 paper/sections/method.qmd
nearly there here

@@
 $$
-\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
+\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
 $$ {#eq-div}
@@
 where $\mathcal{E}_\theta(\cdot)$ denotes the energy function.
-We generate samples $\mathbf{x}^\prime$ by first randomly sampling the target class $y^+ \sim p(y)$ and then generating a counterfactual explanation for that target, similar to how conditional sampling is used to draw from $p_\theta(\mathbf{x})$ in @grathwohl2020your.
 In particular, we set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the target class.
+We generate samples $\mathbf{x}_T^\prime$ by first randomly sampling the target class $y^+ \sim p(y)$ and then generating a counterfactual explanation for that target over $T$ iterations using a gradient-based counterfactual generator.
+This is similar to how conditional sampling is used to draw from $p_\theta(\mathbf{x})$ in @grathwohl2020your.
 
@@
 \begin{aligned}
-\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
-\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
-0 & \text{otherwise.}
-\end{cases}
+\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
+t_\varepsilon &= \max_t \{t: ||\Delta||_\infty < \varepsilon\}
 \end{aligned}

### 2025-02-18 paper/body_ecml.tex
latex is going to kill me one day

@@
 counterfactuals comply with the underlying data generating process
-\cite{poyiadzi2020face, guidotti2022counterfactual, altmeyer2024faithful}.
+\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
 Plausibility is a necessary but insufficient condition for using CE to
@@
 \begin{equation}\phantomsection\label{eq-div}{
-\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
+\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
 }\end{equation}
 
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
-generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
-target class \(y^+ \sim p(y)\) and then generating a counterfactual
-explanation for that target, similar to how conditional sampling is used
-to draw from \(p_\theta(\mathbf{x})\) in \cite{grathwohl2020your}. In
+where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
 particular, we set
 \(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
-where \(y^+\) denotes the index of the target class.
+where \(y^+\) denotes the index of the target class. We generate samples
+\(\mathbf{x}_T^\prime\) by first randomly sampling the target class
+\(y^+ \sim p(y)\) and then generating a counterfactual explanation for
+that target over \(T\) iterations using a gradient-based counterfactual
+generator. This is similar to how conditional sampling is used to draw
+from \(p_\theta(\mathbf{x})\) in \cite{grathwohl2020your}.
 
@@
 \cite{luu2023counterfactual} or adversarial examples
-\cite{freiesleben2022intriguing, pawelczyk2022exploring}. This leaves
+\cite{freiesleben2022intriguing,pawelczyk2022exploring}. This leaves
 some flexibility with respect to the exact choice for
@@
   Proxy attributes of immutable features.
+\item
+  Increased training time.
 \end{enumerate}
@@
 \section{Conclusion}\label{sec-conclusion}
+
+\begin{credits}
+\subsubsection{\ackname}
+
+- Aleks
+- DelftBlue
+- Julia packages: Julia paper itself, PrettyTables, Flux, Plots, AlgebraOfGraphics
+
+\subsubsection{\discintname}
+
+The authors have no competing interests to declare that are relevant to the content of this article.
+\end{credits}

### 2025-02-18 paper/sections/conclusion.qmd
latex is going to kill me one day

--- 2025-02-11 paper/body_ecml.tex
@@
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+# Conclusion {#sec-conclusion}
 
-\begin{abstract}
-Counterfactual Explanations have emerged as a popular tool to explain
-predictions made by opaque machine learning models: they explain how
-factual inputs need to change in order for some fitted model to produce
-some desired output. Much existing research has focused on identifying
-explanations that are not only valid but also deemed plausible and
-desirable with respect to the underlying data and stakeholder
-requirements. Recent work has shown that under this premise, the task of
-learning plausible explanations is effectively reassigned from the model
-itself to the (post-hoc) counterfactual explainer. Building on that
-work, we propose a novel model objective that leverages counterfactuals
-during the training phase (ad-hoc) in order to minimize the divergence
-between learned representations and plausible explanations. Through
-extensive experiments, we demonstrate that our proposed methodology
-facilitates training models that inherently deliver plausible
-explanations while maintaining high predictive performance.
-\keywords{Counterfactual Explanations \and Explainable
-AI \and Representation Learning}
-\end{abstract}
+\begin{credits}
+\subsubsection{\ackname}
 
-\section{Introduction}\label{introduction}
+- Aleks
+- DelftBlue
+- Julia packages: Julia paper itself, PrettyTables, Flux, Plots, AlgebraOfGraphics
 
-Today's prominence of artificial intelligence (AI) has largely been
-driven by advances in \textbf{representation learning}: instead of
-relying on features and rules that are carefully hand-crafted by humans,
-modern machine learning (ML) models are tasked with learning these
-representations from scratch, guided by narrow objectives such as
-predictive accuracy \cite{goodfellow2016deep}. Modern advances in
-computing have made it possible to provide such models with ever greater
-degrees of freedom to achieve that task, which has often led them to
-outperform traditionally more parsimonious models. Unfortunately, in
-doing so they also learn increasingly complex and highly sensitive
-representations that we can no longer easily interpret.
+\subsubsection{\discintname}
 
-This trend towards complexity for the sake of performance has come under
-serious scrutiny in recent years. At the very cusp of the deep learning
-revolution, \cite{szegedy2013intriguing} showed that artificial neural
-networks (ANN) are sensitive to adversarial examples: counterfactuals of
-model inputs that yield vastly different model predictions despite being
-``imperceptible'' in that they are semantically indifferent from their
-factual counterparts. Despite partially effective mitigation strategies
-such as \textbf{adversarial training} \cite{goodfellow2014explaining},
-truly robust deep learning (DL) remains unattainable even for models
-that are considered shallow by today's standards
-\cite{kolter2023keynote}.
-
-Part of the problem is that high degrees of freedom provide room for
-many solutions that are locally optimal with respect to narrow
-objectives \cite{wilson2020case}\footnote{For clarity: we follow
-  standard ML convention in using ``degrees of freedom'' to refer to the
-  number of parameters estimated from data.}. Based purely on predictive
-performance, these solutions may seem to provide compelling explanations
-for the data, when in fact they are based on purely associative,
-semantically meaningless patterns. This poses two related challenges:
-firstly, it makes these models inherently opaque, since humans cannot
-simply interpret what type of explanation the complex learned
-representations correspond to; secondly, even if we could resolve the
-first challenge, it is not obvious how to mitigate models from learning
-representations that correspond to meaningless and implausible
-explanations.
-
-The first challenge has attracted an abundance of research on
-\textbf{explainable AI} (XAI) which aims to develop tools to derive
-explanations from complex model representations. This can mitigate a
-scenario in which we deploy opaque models and blindly rely on their
-predictions. On countless occasions, this scenario has already occurred
-in practice and caused real harm to people who were affected adversely
-and often unfairly by automated decision-making systems (ADMS) involving
-opaque models \cite{oneil2016weapons}. Effective XAI tools can aide us
-in monitoring models and providing recourse to individuals to turn
-adverse outcomes (e.g.~``loan application rejected'') into positive ones
-(``application accepted''). \cite{wachter2017counterfactual} propose
-\textbf{counterfactual explanations} as an effective approach to achieve
-this: they explain how factual inputs need to change in order for some
-fitted model to produce some desired output, typically involving minimal
-perturbations.
-
-To our surprise, the second challenge has not yet attracted any
-consolidated research effort. Specifically, there has been no concerted
-effort towards improving model \textbf{explainability}, which we define
-here as the degree to which learned representations correspond to
-explanations that are interpretable and deemed \textbf{plausible} by
-humans (see Definition~\ref{def-explainability}). Instead, the choice
-has typically been to improve the capacity of XAI tools to identify the
-subset explanations that are both plausible and valid for any given
-model, independent of whether the learned representations are also
-compatible with implausible explanations \cite{altmeyer2024faithful}.
-Fortunately, recent findings indicate that explainability can arise as
-byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization and generative capacity
-\cite{schut2021generating}.
-
-Building on these findings, we introduce \textbf{counterfactual
-training}: a novel regularization technique geared explicitly towards
-aligning model representations with plausible explanations. Our
-contributions are as follows:
-
-\begin{itemize}
-\tightlist
-\item
-  We discuss existing related work on improving models and consolidate
-  it through the lens of counterfactual explanations
-  (Section~\ref{sec-lit}).
-\item
-  We present our proposed methodological framework that leverages
-  faithful counterfactual explanations during the training phase of
-  models to achieve the explainability objective
-  (Section~\ref{sec-method}).
-\item
-  Through extensive experiments we demonstrate the counterfactual
-  training improve model explainability while maintaining high
-  predictive performance. We run ablation studies and grid searches to
-  understand how the underlying model components and hyperparameters
-  affect outcomes. (Section~\ref{sec-experiments}).
-\end{itemize}
-
-Despite limitations of our approach discussed in
-Section~\ref{sec-discussion}, we conclude that counterfactual training
-provides a practical framework for researchers and practitioners
-interested in making opaque models more trustworthy
-Section~\ref{sec-conclusion}. We also believe that this work serves as
-an opportunity for XAI researchers to reevaluate the premise of
-improving XAI tools without improving models.
-
-\section{Related Literature}\label{sec-lit}
-
-To the best of our knowledge, our proposed framework for counterfactual
-training represents the first attempt to use counterfactual explanations
-during training to improve model explainability. In high-level terms, we
-define model explainability as the extent to which valid explanations
-derived for an opaque model are also deemed plausible with respect to
-the underlying data and stakeholder requirements. To make this more
-concrete, we follow \cite{augustin2020adversarial} in tieing the concept
-of explainability to the quality of counterfactual explanations that we
-can generate for a given model. The authors show that counterfactual
-explanations---understood here as minimal input perturbations that yield
-some desired model prediction---are generally more meaningful if the
-underlying model is more robust to adversarial examples. We can make
-intuitive sense of this finding when looking at adversarial training
-(AT) through the lens of representation learning with high degrees of
-freedom: by inducing models to ``unlearn'' representations that are
-susceptible to worst-case counterfactuals (i.e.~adversarial examples),
-AT effectively removes some implausible explanations from the solution
-space.
-
-\subsection{Adversarial Examples are Counterfactual
-Explanations}\label{adversarial-examples-are-counterfactual-explanations}
-
-This interpretation of the link between explainability through
-counterfactuals on one side, and robustness to adversarial examples on
-the other, is backed by empirical evidence.
-\cite{sauer2021counterfactual} demonstrate that using counterfactual
-images during classifier training improves model robustness. Similarly,
-\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
-potentially useful training data in machine learning, especially in
-supervised settings where inputs may be reasonably mapped to multiple
-outputs. They, too, demonstrate the augmenting the training data of
-image classifiers can improve generalization. \cite{teney2020learning}
-propose an approach using counterfactuals in training that does not rely
-on data augmentation: they argue that counterfactual pairs typically
-already exist in training datasets. Specifically, their approach relies
-on, firstly, identifying similar input samples with different
-annotations and, secondly, ensuring that the gradient of the classifier
-aligns with the vector between pairs of counterfactual inputs using the
-cosine distance as a loss function. In the natural language processing
-(NLP) domain, counterfactuals have similarly been used to improve models
-through data augmentation: \cite{wu2021polyjuice}, propose
-\emph{POLYJUICE}, a general-purpose counterfactual generator for
-language models. They demonstrate empirically that augmenting training
-data through \emph{POLYJUICE} counterfactuals improves robustness in a
-number of NLP tasks. \cite{luu2023counterfactual} introduce
-Counterfactual Adversarial Training (CAT), which also aims at improving
-generalization and robustness of language models. Specifically, they
-propose to proceed as follows: firstly, they identify training samples
-that are subject to high predictive uncertainty; secondly, they generate
-counterfactual explanations for those samples; and, finally, they
-fine-tune the given language model on the augmented dataset that
-includes the generated counterfactuals.
-
-There have also been several attempts at formalizing the relationship
-between counterfactual explanations (CE) and adversarial examples (AE).
-Pointing to clear similarities in how CE and AE are generated,
-\cite{freiesleben2022intriguing} makes the case for jointly studying the
-opaqueness and robustness problem in representation learning. Formally,
-AE can be seen as the subset of CE, for which misclassification is
-achieved \cite{freiesleben2022intriguing}. Similarly,
-\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
-certain conditions and derive theoretical upper bounds on the distances
-between them.
-
-Two recent works are closely related to ours in that they use
-counterfactuals during training with the explicit goal of affecting
-certain properties of post-hoc counterfactual explanations. Firstly,
-\cite{ross2021learning} propose a way to train models that are
-guaranteed to provide recourse for individuals to move from an adverse
-outcome to some positive target class with high probability. The
-approach proposed by \cite{ross2021learning} builds on adversarial
-training, where in this context susceptibility to targeted adversarial
-examples for the positive class is explicitly induced. The proposed
-method allows for imposing a set of actionability constraints ex-ante:
-for example, users can specify that certain features (e.g.~\emph{age},
-\emph{gender}, \ldots) are immutable. Secondly, \cite{guo2023counternet}
-are the first to propose an end-to-end training pipeline that includes
-counterfactual explanations as part of the training procedure. In
-particular, they propose a specific network architecture that includes a
-predictor and CE generator network, where the parameters of the CE
-generator network are learnable. Counterfactuals are generated during
-each training iteration and fed back to the predictor network. In
-contrast to \cite{guo2023counternet}, we impose no restrictions on the
-neural network architecture at all.
-
-\subsection{Beyond Robustness}\label{beyond-robustness}
-
-Improving the adversarial robustness of models is not the only path
-towards aligning representations with plausible explanations. In a work
-closely related to this one, \cite{altmeyer2024faithful} show that
-explainability can be improved through model averaging and refined model
-objectives. The authors propose a way to generate counterfactuals that
-are maximally \textbf{faithful} to the model in that they are consistent
-with what the model has learned about the underlying data. Formally,
-they rely on tools from energy-based modelling to minimize the
-divergence between the distribution of counterfactuals and the
-conditional posterior over inputs learned by the model. Their proposed
-counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
-and only if the underlying model has learned representations that align
-with them. They find that both deep ensembles
-\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
-\cite{grathwohl2020your} tend to do well in this regard.
-
-Once again it helps to look at these findings through the lens of
-representation learning with high degrees of freedom. Deep ensembles are
-approximate Bayesian model averages, which are most called for when
-models are underspecified by the available data \cite{wilson2020case}.
-Averaging across solutions mitigates the aforementioned risk of relying
-on a single locally optimal representations that corresponds to
-semantically meaningless explanations for the data. Previous work by
-\cite{schut2021generating} similarly found that generating plausible
-(``interpretable'') counterfactual explanations is almost trivial for
-deep ensembles that have also undergone adversarial training. The case
-for JEMs is even clearer: they involve a hybrid objective that induces
-both high predictive performance and generative capacity
-\cite{grathwohl2020your}. This is closely related to the idea of
-aligning models with plausible explanations and has inspired our
-proposed counterfactual training objective, as we explain in
-Section~\ref{sec-method}.
-
-\section{Counterfactual Training}\label{sec-method}
-
-Counterfactual training combines ideas from adversarial training,
-energy-based modelling and counterfactuals explanations with the
-explicit objective of aligning representations with plausible
-explanations that comply with user requirements. In the context of CE,
-plausibility has broadly been defined as the degree to which
-counterfactuals comply with the underlying data generating process
-\cite{poyiadzi2020face, guidotti2022counterfactual, altmeyer2024faithful}.
-Plausibility is a necessary but insufficient condition for using CE to
-provide algorithmic recourse (AR) to individuals affected by opaque
-models in practice. This is because for recourse recommendations to be
-\textbf{actionable}, they need to not only result in plausible
-counterfactuals but also be attainable. A plausible CE for a rejected
-20-year-old loan applicant, for example, might reveal that their
-application would have been accepted, if only they were 20 years older.
-Ignoring all other features, this complies with the definition of
-plausibility if 40-year-old individuals are in fact more credit-worthy
-on average than young adults. But of course this CE does not qualify for
-providing actionable recourse to the applicant since \emph{age} is not a
-mutable feature. For our intents and purposes, counterfactual training
-aims at improving model explainability by aligning models with
-counterfactuals that meet both desiderata, plausibility and
-actionability. Formally, we define explainability as follows:
-
-\begin{definition}[Model
-Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}
-
-Let \(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a
-supervised classification model that maps from the \(D\)-dimensional
-input space \(\mathcal{X}\) to representations
-\(\phi(\mathbf{x};\theta)\) and finally to the \(K\)-dimensional output
-space \(\mathcal{Y}\). Assume that for any given input-output pair
-\(\{\mathbf{x},\mathbf{y}\}_i\) there exists a counterfactual
-\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
-where \(\mathbf{y}^{+}\) denotes some target output. We say that
-\(\mathbf{M}_\theta\) is \textbf{explainable} to the extent that
-faithfully generated counterfactuals are plausible (i.e.~consistent with
-the data) and actionable. Formally, we define these properties as
-follows:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  (Plausibility)
-  \(\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
-  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
-\item
-  (Actionability) Permutations \(\Delta\) are subject to actionability
-  constraints.
-\end{enumerate}
-
-We consider counterfactuals as faithful to the extent that they are
-consistent with what the model has learned about the input data. Let
-\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
-over inputs, then formally:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  (Faithfulness)
-  \(\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
-  where \(A\) is defined as above.
-\end{enumerate}
-
-\end{definition}
-
-The definitions of faithfulness and plausibility in
-Definition~\ref{def-explainability} are the same as in
-\cite{altmeyer2024faithful}, with adapted notation. Actionability
-constraints in Definition~\ref{def-explainability} vary and depend on
-the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
-focus on domain and mutability constraints for individual features
-\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
-for reasons discussed in Section~\ref{sec-discussion}.
-
-\subsection{Our Proposed Objective}\label{our-proposed-objective}
-
-To train models with high explainability as defined in
-Definition~\ref{def-explainability}, we propose the following objective,
-
-\begin{equation}\phantomsection\label{eq-obj}{
-\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y})
-}\end{equation}
-
-where \(\text{yloss}(\cdot)\) denotes any conventional classification
-loss function (e.g.~cross-entropy) that induces discriminative
-(predictive) performance. The two additional components in
-Equation~\ref{eq-obj} are explained in more detail below. For now, they
-can be sufficiently described as inducing explainability directly and
-indirectly by penalizing: 1) the contrastive divergence,
-\(\text{div}(\cdot)\), between counterfactuals \(x^\prime\) and observed
-samples \(x\) and, 2) the adversarial loss, \(\text{advloss}(.)\), with
-respect to counterfactuals. The tradeoff between the different
-components can be governed by adjusting the strengths of the penalties
-\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).
-
-\subsubsection{Directly Inducing Explainability through Contrastive
-Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}
-
-\cite{grathwohl2020your} observe that any classifier can be
-re-interpreted as a joint energy-based model (JEM) that learns to
-discriminate output classes conditional on inputs and generate inputs.
-They show that JEMs can be trained to perform well at both tasks by
-directly maximizing the joint log-likelihood factorized as
-\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
-The first factor can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
-\cite{grathwohl2020your} minimize the contrastive divergence between
-samples drawn from \(p_\theta(\mathbf{x})\) and training observations,
-i.e.~samples from \(p(\mathbf{x})\).
-
-A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
-tend to do well with respect to the plausibility objective in
-Definition~\ref{def-explainability}. If we consider samples drawn from
-\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
-finding, because the JEM objective effectively minimizes the divergence
-between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
-To generate samples, \cite{grathwohl2020your} rely on Stochastic
-Gradient Langevin Dynamics (SGLD) using an uninformative prior for
-initialization. This is where we depart from their methodology: instead
-of generating samples through SGLD, we propose using counterfactual
-explainers to generate counterfactuals for observed training samples.
-Specifically, we have
-
-\begin{equation}\phantomsection\label{eq-div}{
-\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
-}\end{equation}
-
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
-generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
-target class \(y^+ \sim p(y)\) and then generating a counterfactual
-explanation for that target, similar to how conditional sampling is used
-to draw from \(p_\theta(\mathbf{x})\) in \cite{grathwohl2020your}. In
-particular, we set
-\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
-where \(y^+\) denotes the index of the target class.
-
-Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
-of observed training samples (positive samples) while at same time
-increasing the energy of counterfactuals (negative samples)
-\cite{du2019implicit}. As the generated counterfactuals get more
-plausible (Definition~\ref{def-explainability}) over the cause of
-training, these two opposing effects gradually balance each out
-\cite{lippe2024uvadlc}.
-
-The departure from SGLD allows us to tap into the vast repertoire of
-explainers that have been proposed in the literature to meet different
-desiderata. Typically, these methods facilitate the imposition of domain
-and mutability constraints, for example. In principle, any existing
-approach for generating counterfactual explanations is viable, so long
-as it does not violate the faithfulness condition. Like JEMs
-\cite{murphy2022probabilistic}, counterfactual training can be
-considered as a form of contrastive representation learning.
-
-\subsubsection{Indirectly Inducing Explainability through Adversarial
-Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}
-
-Based on our analysis in Section~\ref{sec-lit}, counterfactuals
-\(\mathbf{x}^\prime\) can be repurposed as additional training samples
-\cite{luu2023counterfactual} or adversarial examples
-\cite{freiesleben2022intriguing, pawelczyk2022exploring}. This leaves
-some flexibility with respect to the exact choice for
-\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
-functional form to use, though likely not the only reasonable choice, is
-inspired by adversarial training:
-
-\begin{equation}\phantomsection\label{eq-adv}{
-\begin{aligned}
-\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
-\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
-0 & \text{otherwise.}
-\end{cases}
-\end{aligned}
-}\end{equation}
-
-Under this choice we treat the counterfactual \(\mathbf{x}^\prime\) as
-an adversarial example iff it is imperceptible, i.e.~the magnitude of
-the perturbation of any individual feature is upper-bounded at
-\(\varepsilon\).
-
-\subsection{Encoding Actionability Constraints}\label{sec-constraints}
-
-Many existing counterfactual explainers support domain and mutability
-constraints out-of-the-box. In fact, both types of constraints can be
-implemented for any counterfactual explainer that relies on gradient
-descent in the feature space for optimization
-\cite{altmeyer2023explaining}. In this context, domain constraints can
-be imposed by simply projecting counterfactuals back to the specified
-domain, if the previous gradient step resulted in updated feature values
-that were out-of-domain. Mutability constraints can similarly be
-enforced by setting partial derivatives to zero to ensure that features
-are only mutated in the allowed direction, if at all.
-
-Since actionability constraints are binding at test time, we should also
-impose them when generating \(\mathbf{x}^\prime\) during each training
-iteration to align model representations with user requirements. Through
-their effect on \(\mathbf{x}^\prime\), both types of constraints
-influence model outcomes through Equation~\ref{eq-div}. Here it is
-crucial that we avoid penalizing implausibility that arises due to
-mutability constraints. For any mutability-constrained feature \(d\)
-this can be achieved by enforcing
-\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
-\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
-violate mutability constraints. Specifically, we set
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Feature \(d\) is strictly immutable in practice.
-\item
-  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\) but feature \(d\) can
-  only be decreased in practice.
-\item
-  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\) but feature \(d\) can
-  only be increased in practice.
-\end{enumerate}
-
-From a Bayesian perspective, setting
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
-a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
-Intuitively, we think of this simply in terms ignoring implausibility
-costs with respect to immutable features, which effectively forces the
-model to instead seek plausibility with respect to the remaining
-features. This in turn results in lower overall sensitivity to immutable
-features, which we demonstrate empirically for different classifiers in
-Section~\ref{sec-experiments}. Under certain conditions, this results
-holds theoretically\footnote{For the proof, see the supplementary
-  appendix.}:
-
-\begin{theorem}[Protecting Immutable
-Features]\protect\hypertarget{thm-mtblty}{}\label{thm-mtblty}
-
-Let \(\mathbf{M}_\theta(\mathbf{x})=\Theta\mathbf{x}\) denote a linear
-classifier with \(y\in\{1,...,K\}=\mathcal{K}\) and
-\(\mathbf{x} \in \mathbb{R}^D\). If we assume multivariate Gaussian
-class densities with common diagonal covariance matrix
-\(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), then protecting an
-immutable feature from the contrastive divergence penalty
-Equation~\ref{eq-div} will result in lower classifier sensitivity to
-that feature, provided that at least one other feature is mutable.
-
-\end{theorem}
-
-It is worth highlighting that Theorem~\ref{thm-mtblty} assumes
-independence of features. This raises a valid concern about the effect
-of protecting immutable features in the presence of proxy features that
-remain unprotected. We discuss this limitation in
-Section~\ref{sec-discussion}.
-
-\section{Experiments}\label{sec-experiments}
-
-In this section, we present experiments that we have conducted in order
-to answer the following research questions:
-
-\begin{corollary}[Plausibility]\protect\hypertarget{cor-plaus}{}\label{cor-plaus}
-
-Does our proposed counterfactual training objective
-(Equation~\ref{eq-obj}) induce models to learn plausible explanations?
-
-\end{corollary}
-
-\begin{corollary}[Actionability]\protect\hypertarget{cor-action}{}\label{cor-action}
-
-Does our proposed counterfactual training objective
-(Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
-outcomes in the presence of actionability constraints?
-
-\end{corollary}
-
-Beyond this, we are also interested in understanding how robust our
-answers to RQ~\ref{cor-plaus} and RQ~\ref{cor-action} are:
-
-\begin{corollary}[Hyperparameters]\protect\hypertarget{cor-hyper}{}\label{cor-hyper}
-
-What are the effects of different hyperparameter choices with respect to
-Equation~\ref{eq-obj}?
-
-\end{corollary}
-
-\subsection{Experimental Setup}\label{experimental-setup}
-
-\subsection{Experimental Results}\label{experimental-results}
-
-\section{Discussion}\label{sec-discussion}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Limited to classification models.
-\item
-  Proxy attributes of immutable features.
-\end{enumerate}
-
-\section{Conclusion}\label{sec-conclusion}
+The authors have no competing interests to declare that are relevant to the content of this article.
+\end{credits}

### 2025-02-18 paper/body_ecml.tex
vamos

@@
 \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
-t_\varepsilon &= \max_t \{t: ||\Delta||_\infty < \varepsilon\}
+t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
 \end{aligned}
@@
 
-Under this choice we treat the counterfactual \(\mathbf{x}^\prime\) as
-an adversarial example iff it is imperceptible, i.e.~the magnitude of
-the perturbation of any individual feature is upper-bounded at
-\(\varepsilon\).
+Under this choice, we consider nascent counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
+magnitude of the perturbation to any individual feature is at most
+\(\varepsilon\). This is closely aligned with
+\cite{szegedy2013intriguing}, who define an adversarial attack as an
+``imperceptible non-random perturbation''. Thus, we choose to work with
+a different distinction between CE and AE than
+\cite{freiesleben2022intriguing}, who considers misclassification as the
+key distinguishing feature of AE. One of the key observations in this
+work is that we can leverage counterfactual explanations during training
+and get adversarial examples, essentially for free.
 
@@
 
-\begin{theorem}[Protecting Immutable
-Features]\protect\hypertarget{thm-mtblty}{}\label{thm-mtblty}
+\begin{proposition}[Protecting Immutable
+Features]\protect\hypertarget{prp-mtblty}{}\label{prp-mtblty}
 
@@
 sensitivity to that feature relative to the remaining features, provided
-that at least one of those is mutable.
+that at least one of those is mutable and discriminative.
 
-\end{theorem}
+\end{proposition}
 
-It is worth highlighting that Theorem~\ref{thm-mtblty} assumes
+It is worth highlighting that Proposition~\ref{prp-mtblty} assumes
 independence of features. This raises a valid concern about the effect
@@
 We have simulated this scenario using synthetic data with independent
-features and Gaussian class-conditional densities in
-Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
-outcomes for different training procedures using the same model
-architecture each time (a linear classifier). In each case, we show the
-linear decision boundary (green) and the training data colored according
-to their ground-truth label: orange points belong to the target class,
-\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
-indicate counterfactuals in the target class generated at test time
-using generic gradient descent for a fixed number of iterations.
+features and Gaussian class-conditional densities in \textbf{?@fig-poc}.
+The four panels in \textbf{?@fig-poc} show the outcomes for different
+training procedures using the same model architecture each time (a
+linear classifier). In each case, we show the linear decision boundary
+(green) and the training data colored according to their ground-truth
+label: orange points belong to the target class, \(y^+=2\), blue points
+belong to the non-target class, \(y^-=1\). Stars indicate
+counterfactuals in the target class generated at test time using generic
+gradient descent for a fixed number of iterations.
 
@@
 \end{example}
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{body_ecml_files/mediabag/../paper/figures/poc.pdf}}
-
-}
-
-\caption{\label{fig-poc}Visual illustration of how counterfactual
-training improves explainability. See Example~\ref{exm-grad} for
-details.}
-
-\end{figure}%
 
@@
 \section{Conclusion}\label{sec-conclusion}
-
-\begin{credits}
-\subsubsection{\ackname}
-
-- Aleks
-- DelftBlue
-- Julia packages: Julia paper itself, PrettyTables, Flux, Plots, AlgebraOfGraphics
-
-\subsubsection{\discintname}
-
-The authors have no competing interests to declare that are relevant to the content of this article.
-\end{credits}

### 2025-02-18 paper/ecml/paper.tex
progress on methodology
slowly slowly
first version of research questions defined
fucking latex
latex is going to kill me one day
vamos

@@
 % N.B.: do not change anything above this line. If you require additional packages, please load them directly after this line.
-\usepackage{mwe}
 \usepackage{color}
-\usepackage[numbers,compress]{natbib}
 \usepackage{longtable}
 \usepackage{hyperref}
+\usepackage{amsmath,amsfonts,amssymb}
+\usepackage{graphicx}
+\usepackage{svg}
 % N.B.: you may delete the preceding line. It is used to display an example image in this template.
@@
 
+\let\corollary\relax
+\let\endcorollary\relax
+\newtheorem{corollary}{Research Question}
+
 \input{../body_ecml.tex}
@@
 \begin{credits}
-\subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
-used for general acknowledgments, for example: This study was funded
-by X (grant number Y).
+\subsubsection{\ackname}
+
+- Aleks
+- DelftBlue
+- Julia packages: Julia paper itself, PrettyTables, Flux, Plots, AlgebraOfGraphics
 
 \subsubsection{\discintname}
-It is now necessary to declare any competing interests or to specifically
-state that the authors have no competing interests. Please place the
-statement with a bold run-in heading in small font size beneath the
-(optional) acknowledgments,
-for example: The authors have no competing interests to declare that are
-relevant to the content of this article. Or: Author A has received research
-grants from Company W. Author B has received a speaker honorarium from
-Company X and owns stock in Company Y. Author C is a member of committee Z.
+
+The authors have no competing interests to declare that are relevant to the content of this article.
 \end{credits}

### 2025-02-18 paper/preprint/paper.tex
vamos

@@
 \theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
 \newtheorem{exercise}{Parameters}[section]
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{definition}
 \newtheorem{example}{Example}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{theorem}{Theorem}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 of counterfactual inputs using the cosine distance as a loss function.
+
 In the natural language processing (NLP) domain, counterfactuals have
 similarly been used to improve models through data augmentation: Wu et
-al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{POLYJUICE}, a
+al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{Polyjuice}, a
 general-purpose counterfactual generator for language models. They
 demonstrate empirically that augmenting training data through
-\emph{POLYJUICE} counterfactuals improves robustness in a number of NLP
-tasks. Luu and Inoue (\citeproc{ref-luu2023counterfactual}{2023})
-introduce Counterfactual Adversarial Training (CAT), which also aims at
-improving generalization and robustness of language models.
-Specifically, they propose to proceed as follows: firstly, they identify
-training samples that are subject to high predictive uncertainty;
-secondly, they generate counterfactual explanations for those samples;
-and, finally, they fine-tune the given language model on the augmented
-dataset that includes the generated counterfactuals.
+\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
+tasks. Balashankar et al.
+(\citeproc{ref-balashankar2023improving}{2023}) also use
+\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
+and show that classifier robustness improves up to 20\%. Finally, Luu
+and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
+Counterfactual Adversarial Training (CAT), which also aims at improving
+generalization and robustness of language models. Specifically, they
+propose to proceed as follows: firstly, they identify training samples
+that are subject to high predictive uncertainty; secondly, they generate
+counterfactual explanations for those samples; and, finally, they
+fine-tune the given language model on the augmented dataset that
+includes the generated counterfactuals.
 
@@
 \begin{equation}\phantomsection\label{eq-div}{
-\text{div}(\mathbf{x},\mathbf{x^\prime},y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}^\prime,y)
+\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
 }\end{equation}
 
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We
-generate samples \(\mathbf{x}^\prime\) by first randomly sampling the
-target class \(y^+ \sim p(y)\) and then generating a counterfactual
-explanation for that target, similar to how conditional sampling is used
-to draw from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
-(\citeproc{ref-grathwohl2020your}{2020}). In particular, we set
+where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
+particular, we set
 \(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
-where \(y^+\) denotes the index of the target class.
+where \(y^+\) denotes the index of the target class. We generate samples
+\(\mathbf{x}_T^\prime\) by first randomly sampling the target class
+\(y^+ \sim p(y)\) and then generating a counterfactual explanation for
+that target over \(T\) iterations using a gradient-based counterfactual
+generator. This is similar to how conditional sampling is used to draw
+from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
+(\citeproc{ref-grathwohl2020your}{2020}).
 
@@
 \(\mathbf{x}^\prime\) can be repurposed as additional training samples
-(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023}) or
+(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023};
+\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023}) or
 adversarial examples
@@
 \begin{aligned}
-\text{advloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y};\varepsilon)&=\begin{cases}
-\text{yloss}(\mathbf{M}_\theta(\mathbf{x^\prime}),\mathbf{y}) & \text{if} \ ||\Delta||_\infty \leq \varepsilon \\
-0 & \text{otherwise.}
-\end{cases}
+\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
+t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
 \end{aligned}
@@
 
-Under this choice we treat the counterfactual \(\mathbf{x}^\prime\) as
-an adversarial example iff it is imperceptible, i.e.~the magnitude of
-the perturbation of any individual feature is upper-bounded at
-\(\varepsilon\).
+Under this choice, we consider nascent counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
+magnitude of the perturbation to any individual feature is at most
+\(\varepsilon\). This is closely aligned with Szegedy et al.
+(\citeproc{ref-szegedy2013intriguing}{2013}), who define an adversarial
+attack as an ``imperceptible non-random perturbation''. Thus, we choose
+to work with a different distinction between CE and AE than Freiesleben
+(\citeproc{ref-freiesleben2022intriguing}{2022}), who considers
+misclassification as the key distinguishing feature of AE. One of the
+key observations in this work is that we can leverage counterfactual
+explanations during training and get adversarial examples, essentially
+for free.
 
@@
 sensitivity to that feature relative to the remaining features, provided
-that at least one of those is mutable.
-
-\end{theorem}
-
-It is worth highlighting that Theorem~\ref{thm-mtblty} assumes
+that at least one of those is mutable and discriminative.
+
+\end{proposition}
+
+It is worth highlighting that Proposition~\ref{prp-mtblty} assumes
 independence of features. This raises a valid concern about the effect
@@
 228--45. Springer.
+
+\bibitem[\citeproctext]{ref-balashankar2023improving}
+Balashankar, Ananth, Xuezhi Wang, Yao Qin, Ben Packer, Nithum Thain, Ed
+Chi, Jilin Chen, and Alex Beutel. 2023. {``Improving Classifier
+Robustness Through Active Generative Counterfactual Data
+Augmentation.''} In \emph{Findings of the Association for Computational
+Linguistics: EMNLP 2023}, 127--39.
 
@@
 empirically in Section~\ref{sec-experiments}. Below we derive the
-analytical results in Theorem~\ref{thm-mtblty}.
+analytical results in Proposition~\ref{prp-mtblty}.
 
@@
 This implies that if we do not protect feature \(d_{\text{immtbl}}\),
-the contrastive divergence penalty will push down on
+the contrastive divergence penalty will decrease
 \(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
@@
 By the same rationale, the contrastive divergence penalty can generally
-be expected to push up on \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
+be expected to increase \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
 \((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, breakable, leftrule=.75mm, coltitle=black, bottomtitle=1mm, toprule=.15mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-warning-color!10!white, arc=.35mm, bottomrule=.15mm, left=2mm, titlerule=0mm, opacityback=0, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, colback=white, titlerule=0mm, opacityback=0, coltitle=black, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, opacitybacktitle=0.6, rightrule=.15mm, bottomtitle=1mm, toptitle=1mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, breakable, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, breakable, leftrule=.75mm, coltitle=black, bottomtitle=1mm, toprule=.15mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, bottomrule=.15mm, left=2mm, titlerule=0mm, opacityback=0, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, colback=white, titlerule=0mm, opacityback=0, coltitle=black, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, rightrule=.15mm, bottomtitle=1mm, toptitle=1mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, arc=.35mm, breakable, leftrule=.75mm]
 

### 2025-02-18 paper/sections/method.qmd
vamos

@@
 
-Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring].
+Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring].
 This leaves some flexibility with respect to the exact choice for $\text{advloss}(\cdot)$ in @eq-obj.
@@
 
-Under this choice we treat the counterfactual $\mathbf{x}^\prime$ as an adversarial example iff it is imperceptible, i.e.
-the magnitude of the perturbation of any individual feature is upper-bounded at $\varepsilon$.
+Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as adversarial examples as long as the magnitude of the perturbation to any individual feature is at most $\varepsilon$.
+This is closely aligned with @szegedy2013intriguing, who define an adversarial attack as an "imperceptible non-random perturbation".
+Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing, who considers misclassification as the key distinguishing feature of AE.
+One of the key observations in this work is that we can leverage counterfactual explanations during training and get adversarial examples, essentially for free.
 
@@
 
-::: {#thm-mtblty}
+::: {#prp-mtblty}
 
@@
 *multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
-If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is mutable.
+If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is mutable and discriminative.
 
@@
 
-It is worth highlighting that @thm-mtblty assumes independence of features.
+It is worth highlighting that @prp-mtblty assumes independence of features.
 This raises a valid concern about the effect of protecting immutable features in the presence of proxy features that remain unprotected.

### 2025-02-18 paper/sections/other/constraints.qmd
vamos

@@
 We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments.
-Below we derive the analytical results in @thm-mtblty.
+Below we derive the analytical results in @prp-mtblty.
 
@@
 
-This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will push down on $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0$.
+This implies that if we do not protect feature $d_{\text{immtbl}}$, the contrastive divergence penalty will decrease $\theta_{y^-,d_{\text{immtbl}}}$ thereby exacerbating the existing effect $(\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0$.
 In words, not protecting the immutable feature would have the undesirable effect of making the classifier more sensitive to this feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for lower values of $d_{\text{immtbl}}$.
 
-By the same rationale, the contrastive divergence penalty can generally be expected to push up on $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0$.
+By the same rationale, the contrastive divergence penalty can generally be expected to increase $\theta_{y^-,d_{\text{mtbl}}}$ exacerbating $(\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0$.
 In words, this has the effect of making the classifier more sensitive to the mutable feature, in that it would be more likely to predict class $y^-$ as opposed to $y^+$ for higher values of $d_{\text{mtbl}}$.

### 2025-02-18 paper/sections/related.qmd
vamos

@@
 Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function.
-In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose *POLYJUICE*, a general-purpose counterfactual generator for language models.
-They demonstrate empirically that augmenting training data through *POLYJUICE* counterfactuals improves robustness in a number of NLP tasks.
-@luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models.
+
+In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose *Polyjuice*, a general-purpose counterfactual generator for language models.
+They demonstrate empirically that augmenting training data through *Polyjuice* counterfactuals improves robustness in a number of NLP tasks.
+@balashankar2023improving also use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves up to 20%.
+Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models.
 Specifically, they propose to proceed as follows: firstly, they identify training samples that are subject to high predictive uncertainty; secondly, they generate counterfactual explanations for those samples; and, finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.

### 2025-02-19 paper/preprint/paper.tex
added note on domain constraints to appendix

@@
 \theoremstyle{definition}
+\newtheorem{exercise}{Parameters}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{plain}
 \newtheorem{proposition}{Proposition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
 \theoremstyle{remark}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, colback=white, titlerule=0mm, opacityback=0, coltitle=black, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, opacitybacktitle=0.6, rightrule=.15mm, bottomtitle=1mm, toptitle=1mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, breakable, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, rightrule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, toprule=.15mm, left=2mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, bottomtitle=1mm, breakable, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}]
 
@@
 
+\subsection{Domain Constraints}\label{domain-constraints}
+
+We apply domain constraints on counterfactuals during training and
+evaluation. There are at least two good reasons for doing so. Firstly,
+within the context of explainability and algorithmic recourse,
+real-world attributes are often domain constrained: the \emph{age}
+feature, for example, is lower bounded by zero and upper bounded by the
+maximum human lifespan. Secondly, domain constraints help mitigate
+training instabilities commonly associated with energy-based modelling
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020};
+\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+
+For our image datasets, features are pixel values and hence the domain
+is constrained by the lower and upper bound of values that pixels can
+take depending on how they are scaled (in our case \([-1,1]\)). For all
+other features \(d\) in our synthetic and tabular datasets, we
+automatically infer domain constraints
+\([x_d^{\text{LB}},x_d^{\text{UB}}]\) as follows,
+
+\begin{equation}\phantomsection\label{eq-domain}{
+\begin{aligned}
+x_d^{\text{LB}} &= \arg\min_{x_d} \{\mu_d - n_{\sigma_d}\sigma_d, \arg \min_{x_d} x_d\} \\
+x_d^{\text{UB}} &= \arg\max_{x_d} \{\mu_d + n_{\sigma_d}\sigma_d, \arg \max_{x_d} x_d\} 
+\end{aligned}
+}\end{equation}
+
+where \(\mu_d\) and \(\sigma_d\) denote the sample mean and standard
+deviation of feature \(d\). We set \(n_{\sigma_d}=3\) across the board
+but higher values and hence wider bounds may be appropriate depending on
+the application.
+
 \section{Detailed Results}\label{detailed-results}
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, colback=white, titlerule=0mm, opacityback=0, coltitle=black, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, opacitybacktitle=0.6, rightrule=.15mm, bottomtitle=1mm, toptitle=1mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, arc=.35mm, breakable, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, rightrule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, toprule=.15mm, left=2mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, bottomtitle=1mm, breakable, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, colback=white, titlerule=0mm, opacityback=0, coltitle=black, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, rightrule=.15mm, bottomtitle=1mm, toptitle=1mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, arc=.35mm, breakable, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, rightrule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, toprule=.15mm, left=2mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, breakable, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}]
 

### 2025-02-19 paper/sections/other/constraints.qmd
added note on domain constraints to appendix

@@
 ```
+
+## Domain Constraints
+
+We apply domain constraints on counterfactuals during training and evaluation.
+There are at least two good reasons for doing so.
+Firstly, within the context of explainability and algorithmic recourse, real-world attributes are often domain constrained: the *age* feature, for example, is lower bounded by zero and upper bounded by the maximum human lifespan.
+Secondly, domain constraints help mitigate training instabilities commonly associated with energy-based modelling [@grathwohl2020your;@altmeyer2024faithful].
+
+For our image datasets, features are pixel values and hence the domain is constrained by the lower and upper bound of values that pixels can take depending on how they are scaled (in our case $[-1,1]$).
+For all other features $d$ in our synthetic and tabular datasets, we automatically infer domain constraints $[x_d^{\text{LB}},x_d^{\text{UB}}]$  as follows,
+
+$$
+\begin{aligned}
+x_d^{\text{LB}} &= \arg\min_{x_d} \{\mu_d - n_{\sigma_d}\sigma_d, \arg \min_{x_d} x_d\} \\
+x_d^{\text{UB}} &= \arg\max_{x_d} \{\mu_d + n_{\sigma_d}\sigma_d, \arg \max_{x_d} x_d\} 
+\end{aligned}
+$$ {#eq-domain}
+
+where $\mu_d$ and $\sigma_d$ denote the sample mean and standard deviation of feature $d$.
+We set $n_{\sigma_d}=3$ across the board but higher values and hence wider bounds may be appropriate depending on the application.

### 2025-02-19 paper/sections/other/constraints.qmd
convergence

@@
 
-function plot_db(model, ces; _xlab="Debt", _ylab="Age")
+function plot_db(model, ces; _xlab="Debt", _ylab="Age", plot_contour::Bool=false)
 
@@
     opt_state = Flux.setup(Descent(), model)
-    conv = MaxIterConvergence(25)
+    conv = DecisionThresholdConvergence(decision_threshold=0.99)
 

### 2025-02-19 src/counterfactuals.jl

@@ -105,7 +105,7 @@ end
 
 Checks if the label has been flipped.
 """
-function isvalid(ce::CounterfactualExplanation, model, data)
+function isvalid(ce::AbstractCounterfactualExplanation, model, data)
     return argmax(vec(model(ce.counterfactual))) == argmax(vec(target_encoded(ce, data)))
 end
 

### 2025-02-19 paper/preprint/paper.tex
noice

@@
 \usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{definition}
@@
 \newtheorem{example}{Example}[section]
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{remark}
@@
 \(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
-where \(\mathbf{y}^{+}\) denotes some target output. We say that
-\(\mathbf{M}_\theta\) is \textbf{explainable} to the extent that
-faithfully generated counterfactuals are plausible (i.e.~consistent with
-the data) and actionable. Formally, we define these properties as
+where \(\arg\max_y{\mathbf{y}^{+}}=y^+\) and \(y^+\) denotes the index
+of the target class.
+
+We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
+that faithfully generated counterfactuals are plausible (i.e.~consistent
+with the data) and actionable. Formally, we define these properties as
 follows:
@@
   (Plausibility)
-  \(\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
+  \(\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
   where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
@@
   (Faithfulness)
-  \(\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
+  \(\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
   where \(A\) is defined as above.
@@
 
-Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote iteratively
-generated counterfactuals. To train models with high explainability as
-defined in Definition~\ref{def-explainability}, we propose to leverage
+Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
+explanation generated through gradient descent over \(T\) iterations as
+initially proposed by Wachter, Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
+let \(T\) vary and consider the counterfactual search as converged as
+soon as the predicted probability for the target class has reached a
+pre-determined threshold, \(\tau\):
+\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
+  detailed background information on gradient-based counterfactual
+  search and convergence see Section~\ref{sec-app-ce}.}.
+
+To train models with high explainability as defined in
+Definition~\ref{def-explainability}, we propose to leverage
 counterfactuals in the following objective,
@@
 indicate counterfactuals in the target class generated at test time
-using generic gradient descent for a fixed number of iterations.
+using generic gradient descent until convergence.
 
@@
 
+\subsection{Generating Counterfactuals through Gradient
+Descent}\label{sec-app-ce}
+
+In this section, we provide some background on gradient-based
+counterfactual generators (Section~\ref{sec-app-ce-background}) and
+discuss how we define convergence in this context
+(Section~\ref{sec-app-conv}).
+
+\subsubsection{Background}\label{sec-app-ce-background}
+
+\subsubsection{Convergence}\label{sec-app-conv}
+
+An important consideration when generating counterfactual explanations
+using gradient-based methods is how to defined convergence.
+
 \subsection{Protecting Mutability Constraints with Linear
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, rightrule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, toprule=.15mm, left=2mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, bottomtitle=1mm, breakable, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, leftrule=.75mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, opacityback=0, coltitle=black, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, left=2mm, arc=.35mm, toptitle=1mm, breakable, toprule=.15mm, titlerule=0mm, bottomtitle=1mm, colframe=quarto-callout-warning-color-frame, opacitybacktitle=0.6, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, rightrule=.15mm, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, toprule=.15mm, left=2mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, breakable, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, opacityback=0, coltitle=black, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, left=2mm, arc=.35mm, toptitle=1mm, breakable, toprule=.15mm, titlerule=0mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, colback=white]
 

### 2025-02-19 paper/sections/other/constraints.qmd
noice

@@
 specs = [
-    ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
-    ("(b)", FullObjective(lambda=[1.0,1.0,0.0,1.0]), ["both", "both"]),
-    ("(c)", VanillaObjective(needs_ce=true), ["both", "none"]),
+    # ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
+    # ("(b)", FullObjective(lambda=[1.0,1.0,0.0,1.0]), ["both", "both"]),
+    # ("(c)", VanillaObjective(needs_ce=true), ["both", "none"]),
     ("(d)", FullObjective(lambda=[1.0,1.0,0.0,1.0]), ["both", "none"]),
@@
     # Objective:
-    generator = GenericGenerator()
+    generator = ECCoGenerator(λ=[0.1,0.1])
     opt_state = Flux.setup(Descent(), model)
-    conv = DecisionThresholdConvergence(decision_threshold=0.99)
+    conv = DecisionThresholdConvergence(decision_threshold=0.90)
 
@@
         val_set = val_set,
-        nepochs = 100,
+        nepochs = 50,
         mutability = Symbol.(constraint),

### 2025-02-19 paper/sections/other/counterfactuals.qmd
noice

## Generating Counterfactuals through Gradient Descent {#sec-app-ce}

In this section, we provide some background on gradient-based counterfactual generators (@sec-app-ce-background) and discuss how we define convergence in this context (@sec-app-conv).

### Background {#sec-app-ce-background}

### Convergence {#sec-app-conv}

An important consideration when generating counterfactual explanations using gradient-based methods is how to define convergence.

### 2025-02-19 src/utils.jl

@@ -32,7 +32,7 @@ function accuracy(model, train_set)
         y = Flux.onecold(y)
         acc += sum(yhat .== y)
     end
-    return acc / (train_set.batchsize * length(train_set))
+    return acc / size(train_set.data[1],2)
 end
 
 """

### 2025-02-19 paper/preprint/paper.tex
storing average number of iterations

@@
 \usepackage{amsthm}
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{definition}
 \newtheorem{exercise}{Parameters}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
 \theoremstyle{plain}
@@
 \newtheorem{definition}{Definition}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 An important consideration when generating counterfactual explanations
-using gradient-based methods is how to defined convergence.
+using gradient-based methods is how to define convergence. Two common
+choices are to 1) perform gradient descent over a fixed number of
+iterations \(T\), or 2) conclude the search as soon as the predicted
+probability for the target class has reached a pre-determined threshold,
+\(\tau\):
+\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\). We
+prefer the latter for our purposes, because it explicitly defines
+convergence in terms of the black-box model, \(\mathbf{M}(\mathbf{x})\).
+
+Defining convergence in this way allows for a more intuitive
+interpretation of the resulting counterfactual outcomes than with fixed
+\(T\). Specifically, it allows us to think of counterfactuals as
+explaining `high-confidence' predictions by the model for the target
+class \(y^+\). Depending on the context and application, different
+choices of \(\tau\) can be considered as representing `high-confidence'
+predictions.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, leftrule=.75mm, colbacktitle=quarto-callout-warning-color!10!white, bottomrule=.15mm, opacityback=0, coltitle=black, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, left=2mm, arc=.35mm, toptitle=1mm, breakable, toprule=.15mm, titlerule=0mm, bottomtitle=1mm, colframe=quarto-callout-warning-color-frame, opacitybacktitle=0.6, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, bottomrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, arc=.35mm, toptitle=1mm, toprule=.15mm, bottomtitle=1mm, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, opacityback=0, coltitle=black, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, left=2mm, arc=.35mm, toptitle=1mm, breakable, toprule=.15mm, titlerule=0mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, arc=.35mm, toptitle=1mm, toprule=.15mm, bottomtitle=1mm, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, left=2mm]
 

### 2025-02-19 paper/experiments/CTExperiments/src/training_params.jl

@@ -199,7 +199,7 @@ Base.@kwdef struct TrainingParams <: AbstractConfiguration
     generator_params::GeneratorParams = GeneratorParams()
     nce::Int = get_global_param("nce", 100)
     nneighbours::Int = 100
-    conv::AbstractString = "max_iter"
+    conv::AbstractString = "threshold"
     lr::AbstractFloat = 0.001
     opt::AbstractString = "adam"
     parallelizer::AbstractString = ""

### 2025-02-19 paper/sections/method.qmd

@@
 Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$.
-Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\mathbf{y}^{+}$ denotes some target output.
+Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.
+
 We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e.
@@
 
-1. (Plausibility) $\int^{A} p(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
+1. (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
 2. (Actionability) Permutations $\Delta$ are subject to actionability constraints.
@@
 
-3. (Faithfulness) $\int^{A} p_\theta(\mathbf{x}|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
+3. (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
 
@@
 
-Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote iteratively generated counterfactuals.
+Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual.
+For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.].
+
 To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective,
@@
 In each case, we show the linear decision boundary (green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$.
-Stars indicate counterfactuals in the target class generated at test time using generic gradient descent for a fixed number of iterations.
+Stars indicate counterfactuals in the target class generated at test time using generic gradient descent until convergence.
 

### 2025-02-19 paper/sections/other/counterfactuals.qmd

@@
 An important consideration when generating counterfactual explanations using gradient-based methods is how to define convergence.
+Two common choices are to 1) perform gradient descent over a fixed number of iterations $T$, or 2) conclude the search as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$.
+We prefer the latter for our purposes, because it explicitly defines convergence in terms of the black-box model, $\mathbf{M}(\mathbf{x})$.
+
+Defining convergence in this way allows for a more intuitive interpretation of the resulting counterfactual outcomes than with fixed $T$.
+Specifically, it allows us to think of counterfactuals as explaining 'high-confidence' predictions by the model for the target class $y^+$.
+Depending on the context and application, different choices of $\tau$ can be considered as representing 'high-confidence' predictions.

### 2025-02-20 paper/sections/appendix.qmd
noice
phew

@@
 # Technical Details of Our Approach {.appendix}
+
+{{< include /paper/sections/other/counterfactuals.qmd >}}
 
@@
 :::
+
+# Compute {.appendix}
+
+<!-- [@DHPC2022] -->
+
+We performed our experiments on a high-performance cluster.
+Details about the cluster will be disclosed upon publication to avoid revealing information that might interfere with the double-blind review process.
+Since our experiments have largely involved highly parallel tasks and small models, we have made extensive use of multi-processing.

### 2025-02-20 paper/sections/other/constraints.qmd
phew

--- 2025-02-19 paper/sections/other/constraints.qmd
@@
     LinearModel, 
+    MLPModel,
     get_input_encoder
@@
     # Data and decision boundary:
-    coeff = get_log_reg_params(model)
-    db = get_decision_boundary(coeff)
     plt = Plots.scatter(
@@
     )
+    if length(model) == 1
+        coeff = get_log_reg_params(model)
+        db = get_decision_boundary(coeff)
         Plots.abline!(plt, db.slope, db.intercept; lw=5, label="Dec. Boundary")
+    end
 
@@
 specs = [
-    ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
-    ("(b)", FullObjective(lambda=[1.0,1.0,0.0,1.0]), ["both", "both"]),
-    ("(c)", VanillaObjective(needs_ce=true), ["both", "none"]),
-    ("(d)", FullObjective(lambda=[1.0,1.0,0.0,1.0]), ["both", "none"]),
+    # ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
+    ("(b)", FullObjective(lambda=[1.0,0.5,0.01,1.0]), ["both", "both"]),
+    # ("(c)", VanillaObjective(needs_ce=true), ["both", "none"]),
+    # ("(d)", FullObjective(lambda=[1.0,1.0,0.0,1.0]), ["both", "none"]),
 ]
@@
     data = LinearlySeparable(
-        n_train=500,
+        n_train=5000,
         batchsize=50,
@@
     nout = size(first(train_set)[2], 1)
-    model = build_model(LinearModel(), nin, nout)
+    model = build_model(MLPModel(), nin, nout)
 
     # Objective:
-    generator = GenericGenerator()
+    generator = ECCoGenerator()
     opt_state = Flux.setup(Descent(), model)
-    conv = DecisionThresholdConvergence(decision_threshold=0.99)
+    conv = DecisionThresholdConvergence(decision_threshold=0.95, max_iter=25)
 
@@
         val_set = val_set,
-        nepochs = 100,
+        nepochs = 50,
         mutability = Symbol.(constraint),
         callback = plot_db,
-        nce=50,
+        nce=1000,
         convergence=conv, 

### 2025-02-20 src/training.jl

@@ -9,7 +9,7 @@ using ProgressMeter
 using TaijaParallel
 using UnicodePlots
 
-global _min_nce_ratio = 0.1
+global _min_nce_ratio = 0.5
 
 function counterfactual_training(
     loss::AbstractObjective,
@@ -38,7 +38,7 @@ function counterfactual_training(
     nce = isnothing(nce) ? length(train_set) : nce
     nce_per_batch = Int(ceil(nce / length(train_set)))
     nce_batch_ratio = nce_per_batch / train_set.batchsize
-    if nce_batch_ratio < 0.1
+    if nce_batch_ratio < _min_nce_ratio
         @warn "The ratio of counterfactuals to training examples is less than $(_min_nce_ratio * 100)% ($(nce_batch_ratio * 100)%). Consider increasing  the `nce` parameter." maxlog =
             1
     end
@@ -79,9 +79,10 @@ function counterfactual_training(
         reg_losses = Float32[]
         validity_losses = Float32[]
         start = time()
-
+        
         # Generate counterfactuals:
         if epoch > burnin && needs_counterfactuals(loss)
+
             counterfactual_dl, percent_valid, ces = generate!(
                 model,
                 train_set,

### 2025-02-20 paper/sections/other/constraints.qmd
solid observations and progress today

@@
     data = LinearlySeparable(
-        n_train=5000,
+        n_train=3000,
         batchsize=50,
@@
     opt_state = Flux.setup(Descent(), model)
-    conv = DecisionThresholdConvergence(decision_threshold=0.95, max_iter=25)
+    conv = DecisionThresholdConvergence(decision_threshold=0.95, max_iter=30)
 

### 2025-02-21 paper/sections/other/constraints.qmd
good starting parameters creating robust results across datasets and models

@@
         yhat0 = [argmax(y) for y in eachcol(model(x0))]
-        idx_plotted = yhat0.==factual
+        idx_plotted = yhat0.==factual .|| yhat0.==target
         if any(idx_plotted)
 
-            # Paths:
-            u = []
-            v = []
-            for (i,ce) in enumerate(eachcol(x))
-                Δ = ce - x0[:,i]
-                push!(u, Δ[1])
-                push!(v, Δ[2])
-            end
-            Plots.quiver!(x0[1,idx_plotted], x0[2,idx_plotted], quiver=(u[idx_plotted], v[idx_plotted]), color=factual)
+            # # Paths:
+            # u = []
+            # v = []
+            # for (i,ce) in enumerate(eachcol(x))
+            #     Δ = ce - x0[:,i]
+            #     push!(u, Δ[1])
+            #     push!(v, Δ[2])
+            # end
+            # Plots.quiver!(x0[1,idx_plotted], x0[2,idx_plotted], quiver=(u[idx_plotted], v[idx_plotted]), color=yhat0[idx_plotted])
 
@@
     # ("(a)", VanillaObjective(needs_ce=true), ["both", "both"]),
-    ("(b)", FullObjective(lambda=[1.0,0.5,0.01,1.0]), ["both", "both"]),
+    # ("(b)", FullObjective(lambda=[1.0,0.5,0.01,0.1]), ["both", "both"]),
     # ("(c)", VanillaObjective(needs_ce=true), ["both", "none"]),
-    # ("(d)", FullObjective(lambda=[1.0,1.0,0.0,1.0]), ["both", "none"]),
+    ("(d)", FullObjective(lambda=[1.0,0.5,0.01,0.1]), ["both", "none"]),
 ]
@@
     # Objective:
-    generator = ECCoGenerator()
-    opt_state = Flux.setup(Descent(), model)
-    conv = DecisionThresholdConvergence(decision_threshold=0.95, max_iter=30)
+    opt = AMSGrad()
+    generator = GenericGenerator(opt=Descent(0.1))
+    opt_state = Flux.setup(opt, model)
+    conv = DecisionThresholdConvergence(decision_threshold=0.75, max_iter=30)
 

### 2025-02-22 paper/experiments/CTExperiments/src/utils.jl
vamos

@@ -155,8 +155,10 @@ function get_config_from_args(;
         subdirname = ARGS[(x -> contains(x, "--subdir=")).(ARGS)]
         @assert length(subdirname) == 1 "Please provide exactly one name for the subdirectory."
         subdirname = replace(subdirname[1], "--subdir=" => "")
-    else
+    elseif haskey(ENV, "OUTPUT_SUBDIR")
         subdirname = ENV["OUTPUT_SUBDIR"]
+    else
+        subdirname = ""
     end
     new_save_dir = mkpath(joinpath(new_save_dir, subdirname))
     @info "Storing results in $(new_save_dir)."

### 2025-02-24 paper/preprint/paper.tex
added Aleks as author to preprint

@@
 \theoremstyle{definition}
+\newtheorem{exercise}{Parameters}[section]
+\theoremstyle{definition}
 \newtheorem{example}{Example}[section]
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{plain}
@@
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
   Increased training time.
+\item
+  Training instabilities
+\item
+  Fairness and caveats (aware it's not a classical approach in this
+  context, but there is a clear link).
 \end{enumerate}
@@
 
+\bibitem[\citeproctext]{ref-bezanson2017julia}
+Bezanson, Jeff, Alan Edelman, Stefan Karpinski, and Viral B Shah. 2017.
+{``Julia: A Fresh Approach to Numerical Computing.''} \emph{SIAM Review}
+59 (1): 65--98. \url{https://doi.org/10.1137/141000671}.
+
+\bibitem[\citeproctext]{ref-milan2023dataframes}
+Bouchet-Valat, Milan, and Bogumił Kamiński. 2023. {``DataFrames.jl:
+Flexible and Fast Tabular Data in Julia.''} \emph{Journal of Statistical
+Software} 107 (4): 1--32. \url{https://doi.org/10.18637/jss.v107.i04}.
+
+\bibitem[\citeproctext]{ref-chagas2024pretty}
+Chagas, Ronan Arraes Jardim, Ben Baumgold, Glen Hertz, Hendrik Ranocha,
+Mark Wells, Nathan Boyer, Nicholas Ritchie, et al. 2024.
+{``Ronisbr/PrettyTables.jl: V2.4.0.''} Zenodo.
+\url{https://doi.org/10.5281/zenodo.13835553}.
+
+\bibitem[\citeproctext]{ref-PlotsJL}
+Christ, Simon, Daniel Schwabeneder, Christopher Rackauckas, Michael
+Krabbe Borregaard, and Thomas Breloff. 2023. {``Plots.jl -- a User
+Extendable Plotting API for the Julia Programming Language.''}
+https://doi.org/\url{https://doi.org/10.5334/jors.431}.
+
+\bibitem[\citeproctext]{ref-danisch2021makie}
+Danisch, Simon, and Julius Krumbiegel. 2021. {``{Makie.jl}: Flexible
+High-Performance Data Visualization for {Julia}.''} \emph{Journal of
+Open Source Software} 6 (65): 3349.
+\url{https://doi.org/10.21105/joss.03349}.
+
 \bibitem[\citeproctext]{ref-du2019implicit}
@@
 \url{https://doi.org/10.1007/978-0-387-84858-7}.
+
+\bibitem[\citeproctext]{ref-innes2018fashionable}
+Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco
+Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and
+Viral Shah. 2018. {``Fashionable Modelling with Flux.''}
+\url{https://arxiv.org/abs/1811.01457}.
+
+\bibitem[\citeproctext]{ref-innes2018flux}
+Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
+\emph{Journal of Open Source Software} 3 (25): 602.
+\url{https://doi.org/10.21105/joss.00602}.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, bottomrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, arc=.35mm, toptitle=1mm, toprule=.15mm, bottomtitle=1mm, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, breakable, coltitle=black, colframe=quarto-callout-warning-color-frame, bottomtitle=1mm, toptitle=1mm, titlerule=0mm, leftrule=.75mm, left=2mm, colbacktitle=quarto-callout-warning-color!10!white, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacityback=0, bottomrule=.15mm]
 
@@
 
+\subsection{Training Details}\label{sec-app-training}
+
+In this section, we describe the training procedure in detail. While the
+details laid out here are not crucial for understanding our proposed
+approach, they are of importance to anyone looking to implement
+counterfactual training.
+
 \section{Detailed Results}\label{detailed-results}
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, bottomrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, arc=.35mm, toptitle=1mm, toprule=.15mm, bottomtitle=1mm, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, breakable, coltitle=black, colframe=quarto-callout-warning-color-frame, bottomtitle=1mm, toptitle=1mm, titlerule=0mm, leftrule=.75mm, left=2mm, colbacktitle=quarto-callout-warning-color!10!white, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacityback=0, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, arc=.35mm, toptitle=1mm, toprule=.15mm, bottomtitle=1mm, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, breakable, coltitle=black, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, titlerule=0mm, leftrule=.75mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, opacityback=0, bottomrule=.15mm]
 
@@
 
-For the initial round of experiments we
+For the initial round of experiments we employed a different training
+objective and procedure that led to promising results for some
+hyperparameter choices, but suffered from training instabilities.
 
@@
   \item
-    \(\lambda_{\text{cost}}\): \texttt{0.0,\ 0.001,\ 0.1}
-  \item
-    \(\lambda_{\text{div}}\):
-    \texttt{0.01,\ 0.05,\ 0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 15.0}
-  \item
-    Learning Rate: \texttt{1.0}
-  \item
-    Maximum Iterations: \texttt{20,\ 50,\ 100}
+    Decisiono. Threshold: \texttt{0.75,\ 0.9,\ 0.95}
+  \item
+    \(\lambda_{\text{cost}}\): \texttt{0.0,\ 0.1}
+  \item
+    \(\lambda_{\text{div}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 10.0}
+  \item
+    Maximum Iterations: \texttt{5,\ 25,\ 50}
   \item
@@
 \item
-  Generator: \texttt{ecco,\ generic,\ omni,\ revise}
+  Generator: \texttt{ecco,\ generic,\ revise}
 \item
@@
   full & 0.01 & \textit{Generic} & $-5.71 \cdot 10^{17}$ & $1.3 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.116}} \\
-  full & 0.01 & \textit{REVISE} & -15.6 & 13.2 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-15.6}} & \color{blue}{\textbf{13.2}} \\
   vanilla & 0.01 & \textit{ECCo} & -4.28 & 3.52 \\
   vanilla & 0.01 & \textit{Generic} & -4.45 & 3.47 \\
-  vanilla & 0.01 & \textit{Omniscient} & -5.12 & 4.46 \\
   vanilla & 0.01 & \textit{REVISE} & -4.91 & 4.24 \\
@@
   full & 0.05 & \textit{Generic} & $-8.35 \cdot 10^{17}$ & $1.9 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.114}} \\
   full & 0.05 & \textit{REVISE} & -15 & 12.6 \\
   vanilla & 0.05 & \textit{ECCo} & -4.4 & 3.66 \\
-  vanilla & 0.05 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 0.05 & \textit{Omniscient} & -5.25 & 4.62 \\
+  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-4.38}} & \color{blue}{\textbf{3.48}} \\
   vanilla & 0.05 & \textit{REVISE} & -4.94 & 4.22 \\
@@
   full & 0.1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.56}} & \color{blue}{\textbf{0.124}} \\
   full & 0.1 & \textit{REVISE} & -15.6 & 13.2 \\
@@
   vanilla & 0.1 & \textit{Generic} & -4.45 & 3.48 \\
-  vanilla & 0.1 & \textit{Omniscient} & -5.12 & 4.46 \\
   vanilla & 0.1 & \textit{REVISE} & -4.91 & 4.25 \\
-  full & 0.5 & \textit{ECCo} & -11.8 & 9.83 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-11.8}} & \color{blue}{\textbf{9.83}} \\
   full & 0.5 & \textit{Generic} & $-1.06 \cdot 10^{18}$ & $2.42 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.123}} \\
   full & 0.5 & \textit{REVISE} & -15 & 12.6 \\
@@
   vanilla & 0.5 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 0.5 & \textit{Omniscient} & -5.25 & 4.61 \\
   vanilla & 0.5 & \textit{REVISE} & -4.95 & 4.22 \\
@@
   full & 1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.88 \cdot 10^{11}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.59}} & \color{blue}{\textbf{0.117}} \\
-  full & 1 & \textit{REVISE} & -15.7 & 13.3 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-15.7}} & \color{blue}{\textbf{13.3}} \\
   vanilla & 1 & \textit{ECCo} & -4.28 & 3.51 \\
   vanilla & 1 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 1 & \textit{Omniscient} & -5.11 & 4.46 \\
   vanilla & 1 & \textit{REVISE} & -4.91 & 4.25 \\
@@
   full & 5 & \textit{Generic} & $-4.88 \cdot 10^{17}$ & $1.11 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.117}} \\
   full & 5 & \textit{REVISE} & -14.6 & 12.1 \\
   vanilla & 5 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 5 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 5 & \textit{Omniscient} & -5.25 & 4.61 \\
+  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-4.38}} & \color{blue}{\textbf{3.48}} \\
   vanilla & 5 & \textit{REVISE} & -4.95 & 4.22 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.31}} & \color{blue}{\textbf{0.735}} \\
+  full & 10 & \textit{ECCo} & -2.31 & 0.735 \\
   full & 10 & \textit{Generic} & $-1.7 \cdot 10^{11}$ & $3.86 \cdot 10^{11}$ \\
-  full & 10 & \textit{Omniscient} & -2.53 & 0.117 \\
   full & 10 & \textit{REVISE} & -15.5 & 13 \\
@@
   vanilla & 10 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 10 & \textit{Omniscient} & -5.12 & 4.46 \\
   vanilla & 10 & \textit{REVISE} & -4.91 & 4.24 \\
@@
   full & 15 & \textit{Generic} & $-4.91 \cdot 10^{17}$ & $1.12 \cdot 10^{18}$ \\
-  full & 15 & \textit{Omniscient} & -2.53 & 0.116 \\
   full & 15 & \textit{REVISE} & -14.4 & 11.7 \\
@@
   vanilla & 15 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 15 & \textit{Omniscient} & -5.25 & 4.6 \\
   vanilla & 15 & \textit{REVISE} & -4.95 & 4.23 \\\bottomrule
@@
   full & 0 & \textit{Generic} & $-1.03 \cdot 10^{18}$ & $2.34 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.64}} & \color{blue}{\textbf{0.125}} \\
-  full & 0 & \textit{REVISE} & -15.4 & 12.9 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-15.4}} & \color{blue}{\textbf{12.9}} \\
   vanilla & 0 & \textit{ECCo} & -4.34 & 3.58 \\
   vanilla & 0 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0 & \textit{Omniscient} & -5.18 & 4.54 \\
   vanilla & 0 & \textit{REVISE} & -4.93 & 4.23 \\
@@
   full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.04 \cdot 10^{17}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.49}} & \color{blue}{\textbf{0.115}} \\
   full & 0.001 & \textit{REVISE} & -15.5 & 13 \\
   vanilla & 0.001 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0.001 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0.001 & \textit{Omniscient} & -5.18 & 4.53 \\
+  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-4.41}} & \color{blue}{\textbf{3.48}} \\
   vanilla & 0.001 & \textit{REVISE} & -4.93 & 4.23 \\
@@
   full & 0.1 & \textit{Generic} & $-4.48 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5}} & \color{blue}{\textbf{0.112}} \\
   full & 0.1 & \textit{REVISE} & -14.6 & 12.2 \\
@@
   vanilla & 0.1 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0.1 & \textit{Omniscient} & -5.18 & 4.54 \\
   vanilla & 0.1 & \textit{REVISE} & -4.93 & 4.24 \\\bottomrule
@@
   full & 0.01 & \textit{Generic} & $-4.89 \cdot 10^{30}$ & $1.11 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.74}} & \color{blue}{\textbf{5.08}} \\
-  full & 0.01 & \textit{REVISE} & -572 & $1.25 \cdot 10^{3}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-572}} & \color{blue}{\textbf{$1.25 \cdot 10^{3}$}} \\
   vanilla & 0.01 & \textit{ECCo} & -15.5 & 17.3 \\
   vanilla & 0.01 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.01 & \textit{Omniscient} & -12.7 & 14.4 \\
   vanilla & 0.01 & \textit{REVISE} & -11.2 & 13 \\
@@
   full & 0.05 & \textit{Generic} & $-2.22 \cdot 10^{20}$ & $5 \cdot 10^{20}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.41}} & \color{blue}{\textbf{4.48}} \\
   full & 0.05 & \textit{REVISE} & $-1.04 \cdot 10^{3}$ & $2.3 \cdot 10^{3}$ \\
   vanilla & 0.05 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.05 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 0.05 & \textit{Omniscient} & -12.4 & 14.1 \\
+  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-11.7}} & \color{blue}{\textbf{12.8}} \\
   vanilla & 0.05 & \textit{REVISE} & -11.3 & 13.1 \\
@@
   full & 0.1 & \textit{Generic} & $-5.22 \cdot 10^{30}$ & $1.19 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.78}} & \color{blue}{\textbf{5.12}} \\
   full & 0.1 & \textit{REVISE} & -288 & 594 \\
@@
   vanilla & 0.1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.1 & \textit{Omniscient} & -12.7 & 14.4 \\
   vanilla & 0.1 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 0.5 & \textit{ECCo} & -7.09 & 7.51 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-7.09}} & \color{blue}{\textbf{7.51}} \\
   full & 0.5 & \textit{Generic} & $-1.11 \cdot 10^{31}$ & $2.53 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.58}} & \color{blue}{\textbf{4.83}} \\
   full & 0.5 & \textit{REVISE} & $-1.19 \cdot 10^{3}$ & $2.64 \cdot 10^{3}$ \\
@@
   vanilla & 0.5 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 0.5 & \textit{Omniscient} & -12.4 & 14.1 \\
   vanilla & 0.5 & \textit{REVISE} & -11.3 & 13.1 \\
@@
   full & 1 & \textit{Generic} & $-1.58 \cdot 10^{33}$ & $3.59 \cdot 10^{33}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.66}} & \color{blue}{\textbf{4.89}} \\
-  full & 1 & \textit{REVISE} & $-1.16 \cdot 10^{3}$ & $2.59 \cdot 10^{3}$ \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{$-1.16 \cdot 10^{3}$}} & \color{blue}{\textbf{$2.59 \cdot 10^{3}$}} \\
   vanilla & 1 & \textit{ECCo} & -15.5 & 17.3 \\
   vanilla & 1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 1 & \textit{Omniscient} & -12.7 & 14.4 \\
   vanilla & 1 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.57}} & \color{blue}{\textbf{2.07}} \\
+  full & 5 & \textit{ECCo} & -2.57 & 2.07 \\
   full & 5 & \textit{Generic} & $-1.17 \cdot 10^{28}$ & $2.66 \cdot 10^{28}$ \\
-  full & 5 & \textit{Omniscient} & -4.29 & 4.31 \\
   full & 5 & \textit{REVISE} & -530 & $1.16 \cdot 10^{3}$ \\
   vanilla & 5 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 5 & \textit{Generic} & -11.7 & 12.7 \\
-  vanilla & 5 & \textit{Omniscient} & -12.4 & 14.1 \\
+  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-11.7}} & \color{blue}{\textbf{12.7}} \\
   vanilla & 5 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.76}} & \color{blue}{\textbf{0.974}} \\
+  full & 10 & \textit{ECCo} & -1.76 & 0.974 \\
   full & 10 & \textit{Generic} & $-1.54 \cdot 10^{33}$ & $3.51 \cdot 10^{33}$ \\
-  full & 10 & \textit{Omniscient} & -4.44 & 4.56 \\
   full & 10 & \textit{REVISE} & $-1.52 \cdot 10^{3}$ & $3.4 \cdot 10^{3}$ \\
   vanilla & 10 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 10 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 10 & \textit{Omniscient} & -12.7 & 14.4 \\
+  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-10.9}} & \color{blue}{\textbf{11.9}} \\
   vanilla & 10 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.37}} & \color{blue}{\textbf{0.365}} \\
+  full & 15 & \textit{ECCo} & -1.37 & 0.365 \\
   full & 15 & \textit{Generic} & $-5.32 \cdot 10^{28}$ & $1.21 \cdot 10^{29}$ \\
-  full & 15 & \textit{Omniscient} & -4.34 & 4.38 \\
   full & 15 & \textit{REVISE} & -473 & $1.03 \cdot 10^{3}$ \\
@@
   vanilla & 15 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 15 & \textit{Omniscient} & -12.4 & 14.1 \\
   vanilla & 15 & \textit{REVISE} & -11.3 & 13.1 \\\bottomrule
@@
   full & 0.01 & \textit{Generic} & -1.49 & 0.71 \\
-  full & 0.01 & \textit{Omniscient} & -5.21 & 5.25 \\
   full & 0.01 & \textit{REVISE} & $-2.71 \cdot 10^{26}$ & $6.37 \cdot 10^{26}$ \\
@@
   vanilla & 0.01 & \textit{Generic} & -8.89 & 6.88 \\
-  vanilla & 0.01 & \textit{Omniscient} & -8.67 & 6.87 \\
   vanilla & 0.01 & \textit{REVISE} & -8.65 & 6.8 \\
   full & 0.05 & \textit{ECCo} & -1.29 & 0.397 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.356}} \\
-  full & 0.05 & \textit{Omniscient} & -5.08 & 5.09 \\
+  full & 0.05 & \textit{Generic} & -1.21 & 0.356 \\
   full & 0.05 & \textit{REVISE} & $-5.91 \cdot 10^{27}$ & $1.36 \cdot 10^{28}$ \\
-  vanilla & 0.05 & \textit{ECCo} & -9.35 & 7.32 \\
+  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-9.35}} & \color{blue}{\textbf{7.32}} \\
   vanilla & 0.05 & \textit{Generic} & -8.85 & 6.87 \\
-  vanilla & 0.05 & \textit{Omniscient} & -8.7 & 6.96 \\
   vanilla & 0.05 & \textit{REVISE} & -8.52 & 6.76 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.2}} & \color{blue}{\textbf{0.383}} \\
+  full & 0.1 & \textit{ECCo} & -1.2 & 0.383 \\
   full & 0.1 & \textit{Generic} & -1.5 & 0.735 \\
-  full & 0.1 & \textit{Omniscient} & -5.17 & 5.23 \\
   full & 0.1 & \textit{REVISE} & $-3.06 \cdot 10^{26}$ & $7.7 \cdot 10^{26}$ \\
   vanilla & 0.1 & \textit{ECCo} & -9.33 & 7.32 \\
-  vanilla & 0.1 & \textit{Generic} & -8.88 & 6.86 \\
-  vanilla & 0.1 & \textit{Omniscient} & -8.69 & 6.9 \\
+  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-8.88}} & \color{blue}{\textbf{6.86}} \\
   vanilla & 0.1 & \textit{REVISE} & -8.68 & 6.81 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.12}} & \color{blue}{\textbf{0.217}} \\
+  full & 0.5 & \textit{ECCo} & -1.12 & 0.217 \\
   full & 0.5 & \textit{Generic} & -1.21 & 0.352 \\
-  full & 0.5 & \textit{Omniscient} & -5.09 & 5.12 \\
   full & 0.5 & \textit{REVISE} & $-5.97 \cdot 10^{27}$ & $1.37 \cdot 10^{28}$ \\
@@
   vanilla & 0.5 & \textit{Generic} & -8.89 & 6.92 \\
-  vanilla & 0.5 & \textit{Omniscient} & -8.68 & 6.93 \\
   vanilla & 0.5 & \textit{REVISE} & -8.53 & 6.75 \\
@@
   full & 1 & \textit{Generic} & -1.49 & 0.726 \\
-  full & 1 & \textit{Omniscient} & -5.16 & 5.2 \\
   full & 1 & \textit{REVISE} & $-3.09 \cdot 10^{26}$ & $7.22 \cdot 10^{26}$ \\
@@
   vanilla & 1 & \textit{Generic} & -8.86 & 6.85 \\
-  vanilla & 1 & \textit{Omniscient} & -8.7 & 6.9 \\
   vanilla & 1 & \textit{REVISE} & -8.69 & 6.85 \\
   full & 5 & \textit{ECCo} & -1.75 & 0.154 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.363}} \\
-  full & 5 & \textit{Omniscient} & -5.14 & 5.16 \\
-  full & 5 & \textit{REVISE} & $-1.1 \cdot 10^{28}$ & $2.5 \cdot 10^{28}$ \\
+  full & 5 & \textit{Generic} & -1.21 & 0.363 \\
+  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{$-1.1 \cdot 10^{28}$}} & \color{blue}{\textbf{$2.5 \cdot 10^{28}$}} \\
   vanilla & 5 & \textit{ECCo} & -9.36 & 7.32 \\
   vanilla & 5 & \textit{Generic} & -8.88 & 6.91 \\
-  vanilla & 5 & \textit{Omniscient} & -8.7 & 6.93 \\
   vanilla & 5 & \textit{REVISE} & -8.52 & 6.73 \\
   full & 10 & \textit{ECCo} & $-1.02 \cdot 10^{6}$ & $2.32 \cdot 10^{6}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.49}} & \color{blue}{\textbf{0.702}} \\
-  full & 10 & \textit{Omniscient} & -5.13 & 5.16 \\
+  full & 10 & \textit{Generic} & -1.49 & 0.702 \\
   full & 10 & \textit{REVISE} & $-3.74 \cdot 10^{26}$ & $9.09 \cdot 10^{26}$ \\
@@
   vanilla & 10 & \textit{Generic} & -8.87 & 6.86 \\
-  vanilla & 10 & \textit{Omniscient} & -8.7 & 6.89 \\
-  vanilla & 10 & \textit{REVISE} & -8.69 & 6.83 \\
+  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-8.69}} & \color{blue}{\textbf{6.83}} \\
   full & 15 & \textit{ECCo} & $-3.31 \cdot 10^{13}$ & $7.54 \cdot 10^{13}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.22}} & \color{blue}{\textbf{0.37}} \\
-  full & 15 & \textit{Omniscient} & -5.2 & 5.23 \\
+  full & 15 & \textit{Generic} & -1.22 & 0.37 \\
   full & 15 & \textit{REVISE} & $-9.01 \cdot 10^{27}$ & $2.06 \cdot 10^{28}$ \\
@@
   vanilla & 15 & \textit{Generic} & -8.86 & 6.87 \\
-  vanilla & 15 & \textit{Omniscient} & -8.69 & 6.96 \\
   vanilla & 15 & \textit{REVISE} & -8.51 & 6.73 \\\bottomrule
@@
 
+\section{Computation Details}\label{computation-details}
+
+\subsection{Hardware}\label{hardware}
+
+We performed our experiments on a high-performance cluster. Details
+about the cluster will be disclosed upon publication to avoid revealing
+information that might interfere with the double-blind review process.
+Since our experiments have largely involved highly parallel tasks and
+small models, we have made extensive use of multi-processing.
+
+\subsection{Software}\label{software}
+
+All computations were performed in the Julia Programming Language
+(\citeproc{ref-bezanson2017julia}{Bezanson et al. 2017}). We have
+developed a package for counterfactual training that leverages and
+extends the functionality provided by several existing packages, most
+notably
+\href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{CounterfactualExplanations.jl}
+(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023})
+and the \href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library for
+deep learning (\citeproc{ref-innes2018fashionable}{Michael Innes et al.
+2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). For
+data-wrangling and presentation-ready tables we relied on
+\href{https://dataframes.juliadata.org/v1.7/}{DataFrames.jl}
+(\citeproc{ref-milan2023dataframes}{Bouchet-Valat and Kamiński 2023})
+and
+\href{https://ronisbr.github.io/PrettyTables.jl/v2.4/}{PrettyTables.jl}
+(\citeproc{ref-chagas2024pretty}{Chagas et al. 2024}), respectively. For
+plots and visualizations we used both
+\href{https://docs.juliaplots.org/v1.40/}{Plots.jl}
+(\citeproc{ref-PlotsJL}{Christ et al. 2023}) and
+\href{https://docs.makie.org/v0.22/}{Makie.jl}
+(\citeproc{ref-danisch2021makie}{Danisch and Krumbiegel 2021}), in
+particular \href{https://aog.makie.org/v0.9.3/}{AlgebraOfGraphics.jl}.
+
 \end{document}

### 2025-02-24 paper/sections/appendix.qmd
added Aleks as author to preprint

@@
 
+{{< include /paper/sections/other/training.qmd >}}
+
 # Detailed Results {.appendix}
@@
 
-For the initial round of experiments we
+For the initial round of experiments we employed a different training objective and procedure that led to promising results for some hyperparameter choices, but suffered from training instabilities.
 
@@
 
-# Compute {.appendix}
+# Computation Details {.appendix}
 
 <!-- [@DHPC2022] -->
+## Hardware
 
@@
 Since our experiments have largely involved highly parallel tasks and small models, we have made extensive use of multi-processing.
+
+## Software
+
+All computations were performed in the Julia Programming Language [@bezanson2017julia].
+We have developed a package for counterfactual training that leverages and extends the functionality provided by several existing packages, most notably [CounterfactualExplanations.jl](https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl) [@altmeyer2023explaining] and the [Flux.jl](https://fluxml.ai/Flux.jl/v0.16/) library for deep learning [@innes2018fashionable;@innes2018flux].
+For data-wrangling and presentation-ready tables we relied on [DataFrames.jl](https://dataframes.juliadata.org/v1.7/) [@milan2023dataframes] and [PrettyTables.jl](https://ronisbr.github.io/PrettyTables.jl/v2.4/) [@chagas2024pretty], respectively.
+For plots and visualizations we used both [Plots.jl](https://docs.juliaplots.org/v1.40/) [@PlotsJL] and [Makie.jl](https://docs.makie.org/v0.22/) [@danisch2021makie], in particular [AlgebraOfGraphics.jl](https://aog.makie.org/v0.9.3/).

### 2025-02-24 paper/experiments/CTExperiments/src/tables.jl

@@ -41,6 +41,8 @@ function get_table_inputs(
     # Highlighters:
     hls = value_highlighter(df, value_var; backend=backend, kwrgs...)
     if "generator_type" in names(df)
+        # Filter out "omni":
+        df = df[df.generator_type .!= "omni",:]
         df.generator_type = format_generator.(df.generator_type)
         gen_hl = generator_highlighter(df; backend=backend)
         hls = (hls..., gen_hl)

### 2025-02-24 paper/preprint/paper.tex

@@
 \usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
 \theoremstyle{definition}
 \newtheorem{exercise}{Parameters}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{plain}
@@
 \theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
+\newtheorem{example}{Example}[section]
 \theoremstyle{remark}
@@
   pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
-  pdfauthor={Patrick Altmeyer; Arie van Deursen; Cynthia C. S. Liem},
+  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
   pdfkeywords={Counterfactual Explanations, Explainable
@@
 Engineering, Mathematics and Computer Science\\Delft University of
-Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
-van Deursen}\\Faculty of Electrical Engineering, Mathematics and
-Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
-C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
+Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
+Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
+Science\\Delft University of Technology\\\\\asep\textbf{Arie van
+Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
+Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
+Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
 Science\\Delft University of Technology\\\\}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, breakable, coltitle=black, colframe=quarto-callout-warning-color-frame, bottomtitle=1mm, toptitle=1mm, titlerule=0mm, leftrule=.75mm, left=2mm, colbacktitle=quarto-callout-warning-color!10!white, arc=.35mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacityback=0, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colbacktitle=quarto-callout-warning-color!10!white, opacitybacktitle=0.6, breakable, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, coltitle=black, opacityback=0, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toprule=.15mm, toptitle=1mm, left=2mm, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, arc=.35mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, breakable, coltitle=black, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, titlerule=0mm, leftrule=.75mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, opacityback=0, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, breakable, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, coltitle=black, opacityback=0, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, toprule=.15mm, toptitle=1mm, left=2mm, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, arc=.35mm, rightrule=.15mm]
 

### 2025-02-24 paper/sections/other/training.qmd

--- 2025-02-19 paper/preprint/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-]{article}
+## Training Details {#sec-app-training}
 
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
-    \setmainfont[]{Latin Modern Roman}
-  \setmathfont[]{Latin Modern Math}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
-% Make \paragraph and \subparagraph free-standing
-\makeatletter
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-\makeatother
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
-% Set default figure placement to htbp
-\def\fps@figure{htbp}
-\makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\usepackage{arxiv}
-\usepackage{orcidlink}
-\usepackage{amsmath}
-\usepackage[T1]{fontenc}
-\usepackage{placeins}
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
-\definecolor{quarto-callout-color}{HTML}{909090}
-\definecolor{quarto-callout-note-color}{HTML}{0758E5}
-\definecolor{quarto-callout-important-color}{HTML}{CC1914}
-\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
-\definecolor{quarto-callout-tip-color}{HTML}{00A047}
-\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
-\definecolor{quarto-callout-color-frame}{HTML}{acacac}
-\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
-\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
-\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
-\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
-\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Figure}
-\else
-  \newcommand\figurename{Figure}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
-\theoremstyle{remark}
-\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
-\newtheorem*{remark}{Remark}
-\newtheorem*{solution}{Solution}
-\newtheorem{refremark}{Remark}[section]
-\newtheorem{refsolution}{Solution}[section]
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-
-\usepackage{bookmark}
-
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
-  pdfauthor={Patrick Altmeyer; Arie van Deursen; Cynthia C. S. Liem},
-  pdfkeywords={Counterfactual Explanations, Explainable
-AI, Representation Learning},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-\usepackage{lineno}
-\linenumbers
-\newcommand{\runninghead}{A Preprint }
-\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
-\title{Counterfactual Training: Teaching Models Plausible and Actionable
-Explanations}
-\def\asep{\\\\\\ } % default: all authors on same column
-\def\asep{\And }
-\author{\textbf{Patrick
-Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
-Engineering, Mathematics and Computer Science\\Delft University of
-Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
-van Deursen}\\Faculty of Electrical Engineering, Mathematics and
-Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
-C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\}
-\date{}
-\begin{document}
-\maketitle
-\begin{abstract}
-Counterfactual Explanations have emerged as a popular tool to explain
-predictions made by opaque machine learning models: they explain how
-factual inputs need to change in order for some fitted model to produce
-some desired output. Much existing research has focused on identifying
-explanations that are not only valid but also deemed plausible and
-desirable with respect to the underlying data and stakeholder
-requirements. Recent work has shown that under this premise, the task of
-learning plausible explanations is effectively reassigned from the model
-itself to the (post-hoc) counterfactual explainer. Building on that
-work, we propose a novel model objective that leverages counterfactuals
-during the training phase (ad-hoc) in order to minimize the divergence
-between learned representations and plausible explanations. Through
-extensive experiments, we demonstrate that our proposed methodology
-facilitates training models that inherently deliver plausible
-explanations while maintaining high predictive performance.
-\end{abstract}
-{\bfseries \emph Keywords}
-\def\sep{\textbullet\ }
-Counterfactual Explanations \sep Explainable AI \sep 
-Representation Learning
-
-\section{Introduction}\label{introduction}
-
-Today's prominence of artificial intelligence (AI) has largely been
-driven by advances in \textbf{representation learning}: instead of
-relying on features and rules that are carefully hand-crafted by humans,
-modern machine learning (ML) models are tasked with learning these
-representations from scratch, guided by narrow objectives such as
-predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
-Bengio, and Courville 2016}). Modern advances in computing have made it
-possible to provide such models with ever greater degrees of freedom to
-achieve that task, which has often led them to outperform traditionally
-more parsimonious models. Unfortunately, in doing so they also learn
-increasingly complex and highly sensitive representations that we can no
-longer easily interpret.
-
-This trend towards complexity for the sake of performance has come under
-serious scrutiny in recent years. At the very cusp of the deep learning
-revolution, Szegedy et al. (\citeproc{ref-szegedy2013intriguing}{2013})
-showed that artificial neural networks (ANN) are sensitive to
-adversarial examples: counterfactuals of model inputs that yield vastly
-different model predictions despite being ``imperceptible'' in that they
-are semantically indifferent from their factual counterparts. Despite
-partially effective mitigation strategies such as \textbf{adversarial
-training} (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow,
-Shlens, and Szegedy 2014}), truly robust deep learning (DL) remains
-unattainable even for models that are considered shallow by today's
-standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).
-
-Part of the problem is that high degrees of freedom provide room for
-many solutions that are locally optimal with respect to narrow
-objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
-  clarity: we follow standard ML convention in using ``degrees of
-  freedom'' to refer to the number of parameters estimated from data.}.
-Based purely on predictive performance, these solutions may seem to
-provide compelling explanations for the data, when in fact they are
-based on purely associative, semantically meaningless patterns. This
-poses two related challenges: firstly, it makes these models inherently
-opaque, since humans cannot simply interpret what type of explanation
-the complex learned representations correspond to; secondly, even if we
-could resolve the first challenge, it is not obvious how to mitigate
-models from learning representations that correspond to meaningless and
-implausible explanations.
-
-The first challenge has attracted an abundance of research on
-\textbf{explainable AI} (XAI) which aims to develop tools to derive
-explanations from complex model representations. This can mitigate a
-scenario in which we deploy opaque models and blindly rely on their
-predictions. On countless occasions, this scenario has already occurred
-in practice and caused real harm to people who were affected adversely
-and often unfairly by automated decision-making systems (ADMS) involving
-opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective
-XAI tools can aide us in monitoring models and providing recourse to
-individuals to turn adverse outcomes (e.g.~``loan application
-rejected'') into positive ones (``application accepted''). Wachter,
-Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}) propose
-\textbf{counterfactual explanations} as an effective approach to achieve
-this: they explain how factual inputs need to change in order for some
-fitted model to produce some desired output, typically involving minimal
-perturbations.
-
-To our surprise, the second challenge has not yet attracted any
-consolidated research effort. Specifically, there has been no concerted
-effort towards improving model \textbf{explainability}, which we define
-here as the degree to which learned representations correspond to
-explanations that are interpretable and deemed \textbf{plausible} by
-humans (see Definition~\ref{def-explainability}). Instead, the choice
-has typically been to improve the capacity of XAI tools to identify the
-subset explanations that are both plausible and valid for any given
-model, independent of whether the learned representations are also
-compatible with implausible explanations
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-Fortunately, recent findings indicate that explainability can arise as
-byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization and generative capacity Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}).
-
-Building on these findings, we introduce \textbf{counterfactual
-training}: a novel regularization technique geared explicitly towards
-aligning model representations with plausible explanations. Our
-contributions are as follows:
-
-\begin{itemize}
-\tightlist
-\item
-  We discuss existing related work on improving models and consolidate
-  it through the lens of counterfactual explanations
-  (Section~\ref{sec-lit}).
-\item
-  We present our proposed methodological framework that leverages
-  faithful counterfactual explanations during the training phase of
-  models to achieve the explainability objective
-  (Section~\ref{sec-method}).
-\item
-  Through extensive experiments we demonstrate the counterfactual
-  training improve model explainability while maintaining high
-  predictive performance. We run ablation studies and grid searches to
-  understand how the underlying model components and hyperparameters
-  affect outcomes. (Section~\ref{sec-experiments}).
-\end{itemize}
-
-Despite limitations of our approach discussed in
-Section~\ref{sec-discussion}, we conclude that counterfactual training
-provides a practical framework for researchers and practitioners
-interested in making opaque models more trustworthy
-Section~\ref{sec-conclusion}. We also believe that this work serves as
-an opportunity for XAI researchers to reevaluate the premise of
-improving XAI tools without improving models.
-
-\section{Related Literature}\label{sec-lit}
-
-To the best of our knowledge, our proposed framework for counterfactual
-training represents the first attempt to use counterfactual explanations
-during training to improve model explainability. In high-level terms, we
-define model explainability as the extent to which valid explanations
-derived for an opaque model are also deemed plausible with respect to
-the underlying data and stakeholder requirements. To make this more
-concrete, we follow Augustin, Meinke, and Hein
-(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
-explainability to the quality of counterfactual explanations that we can
-generate for a given model. The authors show that counterfactual
-explanations---understood here as minimal input perturbations that yield
-some desired model prediction---are generally more meaningful if the
-underlying model is more robust to adversarial examples. We can make
-intuitive sense of this finding when looking at adversarial training
-(AT) through the lens of representation learning with high degrees of
-freedom: by inducing models to ``unlearn'' representations that are
-susceptible to worst-case counterfactuals (i.e.~adversarial examples),
-AT effectively removes some implausible explanations from the solution
-space.
-
-\subsection{Adversarial Examples are Counterfactual
-Explanations}\label{adversarial-examples-are-counterfactual-explanations}
-
-This interpretation of the link between explainability through
-counterfactuals on one side, and robustness to adversarial examples on
-the other, is backed by empirical evidence. Sauer and Geiger
-(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
-counterfactual images during classifier training improves model
-robustness. Similarly, Abbasnejad et al.
-(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
-counterfactuals represent potentially useful training data in machine
-learning, especially in supervised settings where inputs may be
-reasonably mapped to multiple outputs. They, too, demonstrate the
-augmenting the training data of image classifiers can improve
-generalization. Teney, Abbasnedjad, and Hengel
-(\citeproc{ref-teney2020learning}{2020}) propose an approach using
-counterfactuals in training that does not rely on data augmentation:
-they argue that counterfactual pairs typically already exist in training
-datasets. Specifically, their approach relies on, firstly, identifying
-similar input samples with different annotations and, secondly, ensuring
-that the gradient of the classifier aligns with the vector between pairs
-of counterfactual inputs using the cosine distance as a loss function.
-
-In the natural language processing (NLP) domain, counterfactuals have
-similarly been used to improve models through data augmentation: Wu et
-al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{Polyjuice}, a
-general-purpose counterfactual generator for language models. They
-demonstrate empirically that augmenting training data through
-\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
-tasks. Balashankar et al.
-(\citeproc{ref-balashankar2023improving}{2023}) also use
-\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
-and show that classifier robustness improves up to 20\%. Finally, Luu
-and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
-Counterfactual Adversarial Training (CAT), which also aims at improving
-generalization and robustness of language models. Specifically, they
-propose to proceed as follows: firstly, they identify training samples
-that are subject to high predictive uncertainty; secondly, they generate
-counterfactual explanations for those samples; and, finally, they
-fine-tune the given language model on the augmented dataset that
-includes the generated counterfactuals.
-
-There have also been several attempts at formalizing the relationship
-between counterfactual explanations (CE) and adversarial examples (AE).
-Pointing to clear similarities in how CE and AE are generated,
-Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
-case for jointly studying the opaqueness and robustness problem in
-representation learning. Formally, AE can be seen as the subset of CE,
-for which misclassification is achieved
-(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
-Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
-CE and AE are equivalent under certain conditions and derive theoretical
-upper bounds on the distances between them.
-
-Two recent works are closely related to ours in that they use
-counterfactuals during training with the explicit goal of affecting
-certain properties of post-hoc counterfactual explanations. Firstly,
-Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
-propose a way to train models that are guaranteed to provide recourse
-for individuals to move from an adverse outcome to some positive target
-class with high probability. The approach proposed by Ross, Lakkaraju,
-and Bastani (\citeproc{ref-ross2021learning}{2024}) builds on
-adversarial training, where in this context susceptibility to targeted
-adversarial examples for the positive class is explicitly induced. The
-proposed method allows for imposing a set of actionability constraints
-ex-ante: for example, users can specify that certain features
-(e.g.~\emph{age}, \emph{gender}, \ldots) are immutable. Secondly, Guo,
-Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the first
-to propose an end-to-end training pipeline that includes counterfactual
-explanations as part of the training procedure. In particular, they
-propose a specific network architecture that includes a predictor and CE
-generator network, where the parameters of the CE generator network are
-learnable. Counterfactuals are generated during each training iteration
-and fed back to the predictor network. In contrast to Guo, Nguyen, and
-Yadav (\citeproc{ref-guo2023counternet}{2023}), we impose no
-restrictions on the neural network architecture at all.
-
-\subsection{Beyond Robustness}\label{beyond-robustness}
-
-Improving the adversarial robustness of models is not the only path
-towards aligning representations with plausible explanations. In a work
-closely related to this one, Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}) show that explainability can
-be improved through model averaging and refined model objectives. The
-authors propose a way to generate counterfactuals that are maximally
-\textbf{faithful} to the model in that they are consistent with what the
-model has learned about the underlying data. Formally, they rely on
-tools from energy-based modelling to minimize the divergence between the
-distribution of counterfactuals and the conditional posterior over
-inputs learned by the model. Their proposed counterfactual explainer,
-\emph{ECCCo}, yields plausible explanations if and only if the
-underlying model has learned representations that align with them. They
-find that both deep ensembles
-(\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
-and Blundell 2017}) and joint energy-based models (JEMs)
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
-well in this regard.
-
-Once again it helps to look at these findings through the lens of
-representation learning with high degrees of freedom. Deep ensembles are
-approximate Bayesian model averages, which are most called for when
-models are underspecified by the available data
-(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
-mitigates the aforementioned risk of relying on a single locally optimal
-representations that corresponds to semantically meaningless
-explanations for the data. Previous work by Schut et al.
-(\citeproc{ref-schut2021generating}{2021}) similarly found that
-generating plausible (``interpretable'') counterfactual explanations is
-almost trivial for deep ensembles that have also undergone adversarial
-training. The case for JEMs is even clearer: they involve a hybrid
-objective that induces both high predictive performance and generative
-capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
-is closely related to the idea of aligning models with plausible
-explanations and has inspired our proposed counterfactual training
-objective, as we explain in Section~\ref{sec-method}.
-
-\section{Counterfactual Training}\label{sec-method}
-
-Counterfactual training combines ideas from adversarial training,
-energy-based modelling and counterfactuals explanations with the
-explicit objective of aligning representations with plausible
-explanations that comply with user requirements. In the context of CE,
-plausibility has broadly been defined as the degree to which
-counterfactuals comply with the underlying data generating process
-(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
-\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
-is a necessary but insufficient condition for using CE to provide
-algorithmic recourse (AR) to individuals affected by opaque models in
-practice. This is because for recourse recommendations to be
-\textbf{actionable}, they need to not only result in plausible
-counterfactuals but also be attainable. A plausible CE for a rejected
-20-year-old loan applicant, for example, might reveal that their
-application would have been accepted, if only they were 20 years older.
-Ignoring all other features, this complies with the definition of
-plausibility if 40-year-old individuals are in fact more credit-worthy
-on average than young adults. But of course this CE does not qualify for
-providing actionable recourse to the applicant since \emph{age} is not a
-mutable feature. For our intents and purposes, counterfactual training
-aims at improving model explainability by aligning models with
-counterfactuals that meet both desiderata, plausibility and
-actionability. Formally, we define explainability as follows:
-
-\begin{definition}[Model
-Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}
-
-Let \(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a
-supervised classification model that maps from the \(D\)-dimensional
-input space \(\mathcal{X}\) to representations
-\(\phi(\mathbf{x};\theta)\) and finally to the \(K\)-dimensional output
-space \(\mathcal{Y}\). Assume that for any given input-output pair
-\(\{\mathbf{x},\mathbf{y}\}_i\) there exists a counterfactual
-\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
-where \(\arg\max_y{\mathbf{y}^{+}}=y^+\) and \(y^+\) denotes the index
-of the target class.
-
-We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
-that faithfully generated counterfactuals are plausible (i.e.~consistent
-with the data) and actionable. Formally, we define these properties as
-follows:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  (Plausibility)
-  \(\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
-  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
-\item
-  (Actionability) Permutations \(\Delta\) are subject to actionability
-  constraints.
-\end{enumerate}
-
-We consider counterfactuals as faithful to the extent that they are
-consistent with what the model has learned about the input data. Let
-\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
-over inputs, then formally:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  (Faithfulness)
-  \(\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
-  where \(A\) is defined as above.
-\end{enumerate}
-
-\end{definition}
-
-The definitions of faithfulness and plausibility in
-Definition~\ref{def-explainability} are the same as in Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
-Actionability constraints in Definition~\ref{def-explainability} vary
-and depend on the context in which \(\mathbf{M}_\theta\) is deployed. In
-this work, we focus on domain and mutability constraints for individual
-features \(x_d\) for \(d=1,...,D\). We limit ourselves to classification
-tasks for reasons discussed in Section~\ref{sec-discussion}.
-
-\subsection{Our Proposed Objective}\label{our-proposed-objective}
-
-Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
-explanation generated through gradient descent over \(T\) iterations as
-initially proposed by Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
-let \(T\) vary and consider the counterfactual search as converged as
-soon as the predicted probability for the target class has reached a
-pre-determined threshold, \(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
-  detailed background information on gradient-based counterfactual
-  search and convergence see Section~\ref{sec-app-ce}.}.
-
-To train models with high explainability as defined in
-Definition~\ref{def-explainability}, we propose to leverage
-counterfactuals in the following objective,
-
-\begin{equation}\phantomsection\label{eq-obj}{
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
-}\end{equation}
-
-where \(\text{yloss}(\cdot)\) is any conventional classification loss
-that induces discriminative performance (e.g.~cross-entropy). The two
-additional components in Equation~\ref{eq-obj} are explained in more
-detail below. For now, they can be sufficiently described as inducing
-explainability directly and indirectly by penalizing: 1) the contrastive
-divergence, \(\text{div}(\cdot)\), between counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, 2) the
-adversarial loss, \(\text{advloss}(.)\), with respect to nascent
-counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
-the different components can be governed by adjusting the strengths of
-the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).
-
-\subsubsection{Directly Inducing Explainability through Contrastive
-Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}
-
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
-any classifier can be re-interpreted as a joint energy-based model (JEM)
-that learns to discriminate output classes conditional on inputs and
-generate inputs. They show that JEMs can be trained to perform well at
-both tasks by directly maximizing the joint log-likelihood factorized as
-\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
-The first factor can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
-contrastive divergence between samples drawn from
-\(p_\theta(\mathbf{x})\) and training observations, i.e.~samples from
-\(p(\mathbf{x})\).
-
-A key empirical finding in Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
-well with respect to the plausibility objective in
-Definition~\ref{def-explainability}. If we consider samples drawn from
-\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
-finding, because the JEM objective effectively minimizes the divergence
-between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
-To generate samples, Grathwohl et al.
-(\citeproc{ref-grathwohl2020your}{2020}) rely on Stochastic Gradient
-Langevin Dynamics (SGLD) using an uninformative prior for
-initialization. This is where we depart from their methodology: instead
-of generating samples through SGLD, we propose using counterfactual
-explainers to generate counterfactuals for observed training samples.
-Specifically, we have
-
-\begin{equation}\phantomsection\label{eq-div}{
-\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
-}\end{equation}
-
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
-particular, we set
-\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
-where \(y^+\) denotes the index of the target class. We generate samples
-\(\mathbf{x}_T^\prime\) by first randomly sampling the target class
-\(y^+ \sim p(y)\) and then generating a counterfactual explanation for
-that target over \(T\) iterations using a gradient-based counterfactual
-generator. This is similar to how conditional sampling is used to draw
-from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
-(\citeproc{ref-grathwohl2020your}{2020}).
-
-Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
-of observed training samples (positive samples) while at same time
-increasing the energy of counterfactuals (negative samples)
-(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
-counterfactuals get more plausible (Definition~\ref{def-explainability})
-over the cause of training, these two opposing effects gradually balance
-each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
-
-The departure from SGLD allows us to tap into the vast repertoire of
-explainers that have been proposed in the literature to meet different
-desiderata. Typically, these methods facilitate the imposition of domain
-and mutability constraints, for example. In principle, any existing
-approach for generating counterfactual explanations is viable, so long
-as it does not violate the faithfulness condition. Like JEMs
-(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), counterfactual
-training can be considered as a form of contrastive representation
-learning.
-
-\subsubsection{Indirectly Inducing Explainability through Adversarial
-Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}
-
-Based on our analysis in Section~\ref{sec-lit}, counterfactuals
-\(\mathbf{x}^\prime\) can be repurposed as additional training samples
-(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023};
-\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023}) or
-adversarial examples
-(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
-\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
-leaves some flexibility with respect to the exact choice for
-\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
-functional form to use, though likely not the only reasonable choice, is
-inspired by adversarial training:
-
-\begin{equation}\phantomsection\label{eq-adv}{
-\begin{aligned}
-\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
-t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
-\end{aligned}
-}\end{equation}
-
-Under this choice, we consider nascent counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
-magnitude of the perturbation to any individual feature is at most
-\(\varepsilon\). This is closely aligned with Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2013}), who define an adversarial
-attack as an ``imperceptible non-random perturbation''. Thus, we choose
-to work with a different distinction between CE and AE than Freiesleben
-(\citeproc{ref-freiesleben2022intriguing}{2022}), who considers
-misclassification as the key distinguishing feature of AE. One of the
-key observations in this work is that we can leverage counterfactual
-explanations during training and get adversarial examples, essentially
-for free.
-
-\subsection{Encoding Actionability Constraints}\label{sec-constraints}
-
-Many existing counterfactual explainers support domain and mutability
-constraints out-of-the-box. In fact, both types of constraints can be
-implemented for any counterfactual explainer that relies on gradient
-descent in the feature space for optimization
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
-In this context, domain constraints can be imposed by simply projecting
-counterfactuals back to the specified domain, if the previous gradient
-step resulted in updated feature values that were out-of-domain.
-Mutability constraints can similarly be enforced by setting partial
-derivatives to zero to ensure that features are only mutated in the
-allowed direction, if at all.
-
-Since actionability constraints are binding at test time, we should also
-impose them when generating \(\mathbf{x}^\prime\) during each training
-iteration to align model representations with user requirements. Through
-their effect on \(\mathbf{x}^\prime\), both types of constraints
-influence model outcomes through Equation~\ref{eq-div}. Here it is
-crucial that we avoid penalizing implausibility that arises due to
-mutability constraints. For any mutability-constrained feature \(d\)
-this can be achieved by enforcing
-\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
-\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
-violate mutability constraints. Specifically, we set
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Feature \(d\) is strictly immutable in practice.
-\item
-  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\) but feature \(d\) can
-  only be decreased in practice.
-\item
-  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\) but feature \(d\) can
-  only be increased in practice.
-\end{enumerate}
-
-From a Bayesian perspective, setting
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
-a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
-Intuitively, we think of this simply in terms ignoring implausibility
-costs with respect to immutable features, which effectively forces the
-model to instead seek plausibility with respect to the remaining
-features. This in turn results in lower overall sensitivity to immutable
-features, which we demonstrate empirically for different classifiers in
-Section~\ref{sec-experiments}. Under certain conditions, this results
-holds theoretically{[}For the proof, see the supplementary appendix.{]}:
-
-\begin{proposition}[Protecting Immutable
-Features]\protect\hypertarget{prp-mtblty}{}\label{prp-mtblty}
-
-Let
-\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
-denote a linear classifier with softmax activation \(\mathcal{S}\)
-(i.e.~\emph{multinomial logistic regression}) where
-\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
-we assume multivariate Gaussian class densities with common diagonal
-covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
-then protecting an immutable feature from the contrastive divergence
-penalty (Equation~\ref{eq-div}) will result in lower classifier
-sensitivity to that feature relative to the remaining features, provided
-that at least one of those is mutable and discriminative.
-
-\end{proposition}
-
-It is worth highlighting that Proposition~\ref{prp-mtblty} assumes
-independence of features. This raises a valid concern about the effect
-of protecting immutable features in the presence of proxy features that
-remain unprotected. We discuss this limitation in
-Section~\ref{sec-discussion}.
-
-\subsection{Illustration}\label{illustration}
-
-To better convey the intuition underlying our proposed method, we
-illustrate different model outcomes in Example~\ref{exm-grad}.
-
-\begin{example}[Prediction of Consumer Credit
-Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}
-
-Suppose we are interested in predicting the likelihood that loan
-applicants default on their credit. We have access to historical data on
-previous loan takers comprised of a binary outcome variable
-(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
-1) the subjects' \emph{age}, which we define as immutable, and 2) the
-subjects' existing level of \emph{debt}, which we define as mutable.
-
-We have simulated this scenario using synthetic data with independent
-features and Gaussian class-conditional densities in
-Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
-outcomes for different training procedures using the same model
-architecture each time (a linear classifier). In each case, we show the
-linear decision boundary (green) and the training data colored according
-to their ground-truth label: orange points belong to the target class,
-\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
-indicate counterfactuals in the target class generated at test time
-using generic gradient descent until convergence.
-
-In panel (a), we have trained our model conventionally, and we do not
-impose mutability constraints at test time. The generated
-counterfactuals are all valid, but not plausible: they are clearly
-distinguishable from the ground-truth data. In panel (b), we have
-trained our model with counterfactual training, once again not imposing
-mutability constraints at test time. We observe that the counterfactuals
-are clearly plausible, therefore meeting the first objective of
-Definition~\ref{def-explainability}.
-
-In panel (c), we have used conventional training again, this time
-imposing the mutability constraint on \emph{age} at test time.
-Counterfactuals are valid but involve some substantial reductions in
-\emph{debt} for some individuals, in particular very young applicants.
-By comparison, counterfactual paths are shorter on average in panel (d),
-where we have used counterfactual training and protected immutable
-features as described in Section~\ref{sec-constraints}. In particular,
-we observe that due to the classifier's lower sensitivity to \emph{age},
-recourse recommendations with respect to \emph{debt} are much more
-homogenous, in that they do not disproportionately punish younger
-individuals. The counterfactuals are also plausible with respect to the
-mutable feature. Thus, we consider the model in panel (d) as the most
-explainable according to Definition~\ref{def-explainability}.
-
-\end{example}
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
-
-}
-
-\caption{\label{fig-poc}Visual illustration of how counterfactual
-training improves explainability. See Example~\ref{exm-grad} for
-details.}
-
-\end{figure}%
-
-\section{Experiments}\label{sec-experiments}
-
-In this section, we present experiments that we have conducted in order
-to answer the following research questions:
-
-\begin{corollary}[Plausibility]\protect\hypertarget{cor-plaus}{}\label{cor-plaus}
-
-Does our proposed counterfactual training objective
-(Equation~\ref{eq-obj}) induce models to learn plausible explanations?
-
-\end{corollary}
-
-\begin{corollary}[Actionability]\protect\hypertarget{cor-action}{}\label{cor-action}
-
-Does our proposed counterfactual training objective
-(Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
-outcomes in the presence of actionability constraints?
-
-\end{corollary}
-
-Beyond this, we are also interested in understanding how robust our
-answers to RQ~\ref{cor-plaus} and RQ~\ref{cor-action} are:
-
-\begin{corollary}[Hyperparameters]\protect\hypertarget{cor-hyper}{}\label{cor-hyper}
-
-What are the effects of different hyperparameter choices with respect to
-Equation~\ref{eq-obj}?
-
-\end{corollary}
-
-\subsection{Experimental Setup}\label{experimental-setup}
-
-\subsection{Experimental Results}\label{experimental-results}
-
-\section{Discussion}\label{sec-discussion}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Limited to classification models.
-\item
-  Proxy attributes of immutable features.
-\item
-  Increased training time.
-\end{enumerate}
-
-\section{Conclusion}\label{sec-conclusion}
-
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
-Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
-den Hengel. 2020. {``Counterfactual Vision and Language Learning.''} In
-\emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
-Recognition (CVPR)}, 10041--51.
-\url{https://doi.org/10.1109/CVPR42600.2020.01006}.
-
-\bibitem[\citeproctext]{ref-altmeyer2023explaining}
-Altmeyer, Patrick, Arie van Deursen, et al. 2023. {``Explaining
-Black-Box Models Through Counterfactuals.''} In \emph{Proceedings of the
-JuliaCon Conferences}, 1:130. 1.
-
-\bibitem[\citeproctext]{ref-altmeyer2024faithful}
-Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia CS
-Liem. 2024. {``Faithful Model Explanations Through Energy-Constrained
-Conformal Counterfactuals.''} In \emph{Proceedings of the AAAI
-Conference on Artificial Intelligence}, 38:10829--37. 10.
-
-\bibitem[\citeproctext]{ref-augustin2020adversarial}
-Augustin, Maximilian, Alexander Meinke, and Matthias Hein. 2020.
-{``Adversarial Robustness on in-and Out-Distribution Improves
-Explainability.''} In \emph{European Conference on Computer Vision},
-228--45. Springer.
-
-\bibitem[\citeproctext]{ref-balashankar2023improving}
-Balashankar, Ananth, Xuezhi Wang, Yao Qin, Ben Packer, Nithum Thain, Ed
-Chi, Jilin Chen, and Alex Beutel. 2023. {``Improving Classifier
-Robustness Through Active Generative Counterfactual Data
-Augmentation.''} In \emph{Findings of the Association for Computational
-Linguistics: EMNLP 2023}, 127--39.
-
-\bibitem[\citeproctext]{ref-du2019implicit}
-Du, Yilun, and Igor Mordatch. 2020. {``Implicit Generation and
-Generalization in Energy-Based Models.''}
-\url{https://arxiv.org/abs/1903.08689}.
-
-\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
-Freiesleben, Timo. 2022. {``The Intriguing Relation Between
-Counterfactual Explanations and Adversarial Examples.''} \emph{Minds and
-Machines} 32 (1): 77--109.
-
-\bibitem[\citeproctext]{ref-goodfellow2014explaining}
-Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
-{``Explaining and Harnessing Adversarial Examples.''}
-\url{https://arxiv.org/abs/1412.6572}.
-
-\bibitem[\citeproctext]{ref-goodfellow2016deep}
-Goodfellow, Ian, Yoshua Bengio, and Aaron Courville. 2016. \emph{Deep
-{Learning}}. {MIT Press}.
-
-\bibitem[\citeproctext]{ref-grathwohl2020your}
-Grathwohl, Will, Kuan-Chieh Wang, Joern-Henrik Jacobsen, David Duvenaud,
-Mohammad Norouzi, and Kevin Swersky. 2020. {``Your Classifier Is
-Secretly an Energy Based Model and You Should Treat It Like One.''} In
-\emph{International Conference on Learning Representations}.
-
-\bibitem[\citeproctext]{ref-guidotti2022counterfactual}
-Guidotti, Riccardo. 2022. {``Counterfactual Explanations and How to Find
-Them: Literature Review and Benchmarking.''} \emph{Data Mining and
-Knowledge Discovery}, 1--55.
-
-\bibitem[\citeproctext]{ref-guo2023counternet}
-Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``CounterNet:
-End-to-End Training of Prediction Aware Counterfactual Explanations.''}
-In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
-Discovery and Data Mining}, 577--89. KDD '23. New York, NY, USA:
-Association for Computing Machinery.
-\url{https://doi.org/10.1145/3580305.3599290}.
-
-\bibitem[\citeproctext]{ref-hastie2009elements}
-Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
-Elements of Statistical Learning}. Springer New York.
-\url{https://doi.org/10.1007/978-0-387-84858-7}.
-
-\bibitem[\citeproctext]{ref-kolter2023keynote}
-Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
-IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)},
-xvi--. Los Alamitos, CA, USA: IEEE Computer Society.
-\url{https://doi.org/10.1109/SaTML54575.2023.00009}.
-
-\bibitem[\citeproctext]{ref-lakshminarayanan2016simple}
-Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2017.
-{``Simple and Scalable Predictive Uncertainty Estimation Using Deep
-Ensembles.''} \emph{Advances in Neural Information Processing Systems}
-30.
-
-\bibitem[\citeproctext]{ref-lippe2024uvadlc}
-Lippe, Phillip. 2024. {``{UvA Deep Learning Tutorials}.''}
-\url{https://uvadlc-notebooks.readthedocs.io/en/latest/}.
-
-\bibitem[\citeproctext]{ref-luu2023counterfactual}
-Luu, Hoai Linh, and Naoya Inoue. 2023. {``Counterfactual Adversarial
-Training for Improving Robustness of Pre-Trained Language Models.''} In
-\emph{Proceedings of the 37th Pacific Asia Conference on Language,
-Information and Computation}, 881--88.
-
-\bibitem[\citeproctext]{ref-murphy2022probabilistic}
-Murphy, Kevin P. 2022. \emph{Probabilistic {Machine Learning}: {An}
-Introduction}. {MIT Press}.
-
-\bibitem[\citeproctext]{ref-oneil2016weapons}
-O'Neil, Cathy. 2016. \emph{Weapons of Math Destruction: {How} Big Data
-Increases Inequality and Threatens Democracy}. {Crown}.
-
-\bibitem[\citeproctext]{ref-pawelczyk2022exploring}
-Pawelczyk, Martin, Chirag Agarwal, Shalmali Joshi, Sohini Upadhyay, and
-Himabindu Lakkaraju. 2022. {``Exploring Counterfactual Explanations
-Through the Lens of Adversarial Examples: A Theoretical and Empirical
-Analysis.''} In \emph{Proceedings of the 25th International Conference
-on Artificial Intelligence and Statistics}, edited by Gustau
-Camps-Valls, Francisco J. R. Ruiz, and Isabel Valera, 151:4574--94.
-Proceedings of Machine Learning Research. PMLR.
-\url{https://proceedings.mlr.press/v151/pawelczyk22a.html}.
-
-\bibitem[\citeproctext]{ref-poyiadzi2020face}
-Poyiadzi, Rafael, Kacper Sokol, Raul Santos-Rodriguez, Tijl De Bie, and
-Peter Flach. 2020. {``{FACE}: {Feasible} and Actionable Counterfactual
-Explanations.''} In \emph{Proceedings of the {AAAI}/{ACM Conference} on
-{AI}, {Ethics}, and {Society}}, 344--50.
-
-\bibitem[\citeproctext]{ref-ross2021learning}
-Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024. {``Learning
-Models for Actionable Recourse.''} In \emph{Proceedings of the 35th
-International Conference on Neural Information Processing Systems}. NIPS
-'21. Red Hook, NY, USA: Curran Associates Inc.
-
-\bibitem[\citeproctext]{ref-sauer2021counterfactual}
-Sauer, Axel, and Andreas Geiger. 2021. {``Counterfactual Generative
-Networks.''} \url{https://arxiv.org/abs/2101.06046}.
-
-\bibitem[\citeproctext]{ref-schut2021generating}
-Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
-Sacaleanu, Yarin Gal, et al. 2021. {``Generating {Interpretable
-Counterfactual Explanations By Implicit Minimisation} of {Epistemic} and
-{Aleatoric Uncertainties}.''} In \emph{International {Conference} on
-{Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.
-
-\bibitem[\citeproctext]{ref-szegedy2013intriguing}
-Szegedy, Christian, Wojciech Zaremba, Ilya Sutskever, Joan Bruna,
-Dumitru Erhan, Ian Goodfellow, and Rob Fergus. 2013. {``Intriguing
-Properties of Neural Networks.''} \url{https://arxiv.org/abs/1312.6199}.
-
-\bibitem[\citeproctext]{ref-teney2020learning}
-Teney, Damien, Ehsan Abbasnedjad, and Anton van den Hengel. 2020.
-{``Learning What Makes a Difference from Counterfactual Examples and
-Gradient Supervision.''} In \emph{Computer Vision--ECCV 2020: 16th
-European Conference, Glasgow, UK, August 23--28, 2020, Proceedings, Part
-x 16}, 580--99. Springer.
-
-\bibitem[\citeproctext]{ref-wachter2017counterfactual}
-Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
-{``Counterfactual Explanations Without Opening the Black Box:
-{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
-841. \url{https://doi.org/10.2139/ssrn.3063289}.
-
-\bibitem[\citeproctext]{ref-wilson2020case}
-Wilson, Andrew Gordon. 2020. {``The Case for Bayesian Deep Learning.''}
-\url{https://arxiv.org/abs/2001.10995}.
-
-\bibitem[\citeproctext]{ref-wu2021polyjuice}
-Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
-2021. {``Polyjuice: Generating Counterfactuals for Explaining,
-Evaluating, and Improving Models.''} In \emph{Proceedings of the 59th
-Annual Meeting of the Association for Computational Linguistics and the
-11th International Joint Conference on Natural Language Processing
-(Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
-and Roberto Navigli, 6707--23. Online: Association for Computational
-Linguistics. \url{https://doi.org/10.18653/v1/2021.acl-long.523}.
-
-\end{CSLReferences}
-
-\newpage{}
-
-\FloatBarrier
-
-\setcounter{section}{0}
-\renewcommand{\thesection}{\Alph{section}}
-
-\setcounter{table}{0}
-\renewcommand{\thetable}{A\arabic{table}}
-
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{A\arabic{figure}}
-
-\section{Notation}\label{notation}
-
-\begin{itemize}
-\tightlist
-\item
-  \(y^+\): The target class and also the index of the target class.
-\item
-  \(y^-\): The non-target class and also the index of non-the target
-  class.
-\item
-  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
-  class.
-\item
-  \(\theta\): Model parameters (unspecified).
-\item
-  \(\Theta\): Matrix of parameters.
-\end{itemize}
-
-\section{Technical Details of Our
-Approach}\label{technical-details-of-our-approach}
-
-\subsection{Generating Counterfactuals through Gradient
-Descent}\label{sec-app-ce}
-
-In this section, we provide some background on gradient-based
-counterfactual generators (Section~\ref{sec-app-ce-background}) and
-discuss how we define convergence in this context
-(Section~\ref{sec-app-conv}).
-
-\subsubsection{Background}\label{sec-app-ce-background}
-
-\subsubsection{Convergence}\label{sec-app-conv}
-
-An important consideration when generating counterfactual explanations
-using gradient-based methods is how to define convergence. Two common
-choices are to 1) perform gradient descent over a fixed number of
-iterations \(T\), or 2) conclude the search as soon as the predicted
-probability for the target class has reached a pre-determined threshold,
-\(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\). We
-prefer the latter for our purposes, because it explicitly defines
-convergence in terms of the black-box model, \(\mathbf{M}(\mathbf{x})\).
-
-Defining convergence in this way allows for a more intuitive
-interpretation of the resulting counterfactual outcomes than with fixed
-\(T\). Specifically, it allows us to think of counterfactuals as
-explaining `high-confidence' predictions by the model for the target
-class \(y^+\). Depending on the context and application, different
-choices of \(\tau\) can be considered as representing `high-confidence'
-predictions.
-
-\subsection{Protecting Mutability Constraints with Linear
-Classifiers}\label{sec-app-constraints}
-
-In Section~\ref{sec-constraints} we explain that to avoid penalizing
-implausibility that arises due to mutability constraints, we impose a
-point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
-argue in Section~\ref{sec-constraints} that this approach induces models
-to be less sensitive to immutable features and demonstrate this
-empirically in Section~\ref{sec-experiments}. Below we derive the
-analytical results in Proposition~\ref{prp-mtblty}.
-
-\begin{proof}
-Let \(d_{\text{mtbl}}\) and \(d_{\text{immtbl}}\) denote some mutable
-and immutable feature, respectively. Suppose that
-\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
-\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
-\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
-class \(k\). In words, we assume that the immutable feature tends to
-take lower values for samples in the non-target class \(y^-\) than in
-the target class \(y^+\). We assume the opposite to hold for the mutable
-feature.
-
-Assuming multivariate Gaussian class densities with common diagonal
-covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), we
-have for the log likelihood ratio between any two classes
-\(k,m \in \mathcal{K}\) (\citeproc{ref-hastie2009elements}{Hastie,
-Tibshirani, and Friedman 2009}):
-
-\begin{equation}\phantomsection\label{eq-loglike}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
-}\end{equation}
-
-By independence of \(x_1,...,x_D\), the full log-likelihood ratio
-decomposes into:
-
-\begin{equation}\phantomsection\label{eq-loglike-decomp}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
-}\end{equation}
-
-By the properties of our classifier (\emph{multinomial logistic
-regression}), we have:
-
-\begin{equation}\phantomsection\label{eq-multi}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
-}\end{equation}
-
-where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
-\(d\) for class \(k\).
-
-Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
-can identify that
-\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
-the assumptions we made above. Hence, we have that
-\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
-and
-\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)
-
-Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
-class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
-target class. Then the partial derivative of the contrastive divergence
-penalty Equation~\ref{eq-div} with respect to coefficient
-\(\theta_{y^+,d}\) is equal to
-
-\begin{equation}\phantomsection\label{eq-grad}{
-\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
-}\end{equation}
-
-and equal to zero everywhere else.
-
-Since \((\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})\) we
-are more likely to have
-\((x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0\) than vice
-versa at initialization. Similarly, we are more likely to have
-\((x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0\) since
-\((\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})\).
-
-This implies that if we do not protect feature \(d_{\text{immtbl}}\),
-the contrastive divergence penalty will decrease
-\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
-effect
-\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
-In words, not protecting the immutable feature would have the
-undesirable effect of making the classifier more sensitive to this
-feature, in that it would be more likely to predict class \(y^-\) as
-opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).
-
-By the same rationale, the contrastive divergence penalty can generally
-be expected to increase \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
-\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
-In words, this has the effect of making the classifier more sensitive to
-the mutable feature, in that it would be more likely to predict class
-\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).
-
-Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
-has the net affect of decreasing the classifier's sensitivity to the
-immutable feature relative to the mutable feature (i.e.~no change in
-sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
-for \(d_{\text{mtbl}}\)).
-\end{proof}
-
-\begin{tcolorbox}[enhanced jigsaw, breakable, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, bottomrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, arc=.35mm, toptitle=1mm, toprule=.15mm, bottomtitle=1mm, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, left=2mm]
-
-@Cynthia, @Arie, I have tentatively phrased the above in terms of a
-theorem and proof. This is something I've so far shied away from because
-I feel a bit out of my depth when it comes to mathematical proofs. The
-above makes intuitive sense to me, but I don't know for sure if it's
-correct.
-
-\end{tcolorbox}
-
-\subsection{Domain Constraints}\label{domain-constraints}
-
-We apply domain constraints on counterfactuals during training and
-evaluation. There are at least two good reasons for doing so. Firstly,
-within the context of explainability and algorithmic recourse,
-real-world attributes are often domain constrained: the \emph{age}
-feature, for example, is lower bounded by zero and upper bounded by the
-maximum human lifespan. Secondly, domain constraints help mitigate
-training instabilities commonly associated with energy-based modelling
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-
-For our image datasets, features are pixel values and hence the domain
-is constrained by the lower and upper bound of values that pixels can
-take depending on how they are scaled (in our case \([-1,1]\)). For all
-other features \(d\) in our synthetic and tabular datasets, we
-automatically infer domain constraints
-\([x_d^{\text{LB}},x_d^{\text{UB}}]\) as follows,
-
-\begin{equation}\phantomsection\label{eq-domain}{
-\begin{aligned}
-x_d^{\text{LB}} &= \arg\min_{x_d} \{\mu_d - n_{\sigma_d}\sigma_d, \arg \min_{x_d} x_d\} \\
-x_d^{\text{UB}} &= \arg\max_{x_d} \{\mu_d + n_{\sigma_d}\sigma_d, \arg \max_{x_d} x_d\} 
-\end{aligned}
-}\end{equation}
-
-where \(\mu_d\) and \(\sigma_d\) denote the sample mean and standard
-deviation of feature \(d\). We set \(n_{\sigma_d}=3\) across the board
-but higher values and hence wider bounds may be appropriate depending on
-the application.
-
-\section{Detailed Results}\label{detailed-results}
-
-\begin{tcolorbox}[enhanced jigsaw, breakable, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, bottomrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, arc=.35mm, toptitle=1mm, toprule=.15mm, bottomtitle=1mm, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-warning-color-frame, left=2mm]
-
-@Cynthia, @Arie, I'm including some preliminary results here but will
-rerun experiments in the coming days. You'll notice some odd outlier
-results (huge values) for the initial grid search
-(Section~\ref{sec-app-initial}). My belief is that this is once again
-due to \emph{ECCo} overshooting for some hyperparameter choices, that we
-have discussed before. A simple solution to this story is to actually
-impose domain constraints. Let's see what the final results show us (I
-also still plan to make slight changes to the implementation), but in
-any case I think it might even be worth to report results for the
-initial grid search in the final appendix (they do include good results
-for CT for certain hyperparameter ranges and highlight limitations
-inherited from energy-based modelling).
-
-\end{tcolorbox}
-
-\subsection{Qualitative Findings for Image
-Data}\label{qualitative-findings-for-image-data}
-
-\begin{tcolorbox}[enhanced jigsaw, breakable, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, opacitybacktitle=0.6, titlerule=0mm, colback=white, opacityback=0, bottomrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, arc=.35mm, toptitle=1mm, toprule=.15mm, bottomtitle=1mm, rightrule=.15mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, left=2mm]
-
-@Cynthia, @Arie, Figure~\ref{fig-mnist} shows much more plausible
-(faithful) counterfactuals for a model with CT than the model with
-conventional training (Figure~\ref{fig-mnist-vanilla}). In fact, this is
-not even using \emph{ECCo+} and still showing better results than the
-best results we achieved in our AAAI paper for JEM ensembles.
-
-\end{tcolorbox}
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp.png}}
-
-}
-
-\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
-counterfactual training. The underlying generator, \emph{ECCo}, aims to
-generate counterfactuals that are faithful to the model
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp_vanilla.png}}
-
-}
-
-\caption{\label{fig-mnist-vanilla}Counterfactual images for \emph{MLP}
-with conventional training. The underlying generator, \emph{ECCo}, aims
-to generate counterfactuals that are faithful to the model
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
-
-\end{figure}%
-
-\subsection{Initial Grid Search}\label{sec-app-initial}
-
-For the initial round of experiments we
-
-\subsubsection{Generator Parameters}\label{generator-parameters}
-
-The hyperparameter grids for the first investigation of the effect of
-generator parameters are shown in
-Parameters~\ref{exr-gen-params-first-run-train} and
-Parameters~\ref{exr-gen-params-first-run-eval}.
-
-\begin{exercise}[Training
-Phase]\protect\hypertarget{exr-gen-params-first-run-train}{}\label{exr-gen-params-first-run-train}
-
-~
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{cost}}\): \texttt{0.0,\ 0.001,\ 0.1}
-  \item
-    \(\lambda_{\text{div}}\):
-    \texttt{0.01,\ 0.05,\ 0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 15.0}
-  \item
-    Learning Rate: \texttt{1.0}
-  \item
-    Maximum Iterations: \texttt{20,\ 50,\ 100}
-  \item
-    Optimizerimizer: \texttt{sgd}
-  \end{itemize}
-\item
-  Generator: \texttt{ecco,\ generic,\ omni,\ revise}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{exercise}
-
-\begin{exercise}[Evaluation
-Phase]\protect\hypertarget{exr-gen-params-first-run-eval}{}\label{exr-gen-params-first-run-eval}
-
-~
-
-\begin{itemize}
-\tightlist
-\item
-  Counterfactual Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Convergence: \texttt{max\_iter}
-  \item
-    Maximum Iterations: \texttt{100}
-  \item
-    No.~Individuals: \texttt{100}
-  \item
-    No.~Runs: \texttt{5}
-  \end{itemize}
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{cost}}\): \texttt{0.0}
-  \item
-    \(\lambda_{\text{div}}\):
-    \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 20.0}
-  \item
-    Learning Rate: \texttt{1.0}
-  \item
-    Maximum Iterations: \texttt{50}
-  \item
-    Optimizerimizer: \texttt{sgd}
-  \end{itemize}
-\end{itemize}
-
-\end{exercise}
-
-\paragraph{Linearly Separable}\label{linearly-separable}
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-lin_sep-lambda_energy_exper}):
-  \emph{ECCo} generally does yield better results than \emph{Vanilla}
-  for higher choices of the energy penalty (10,15) during training.
-  \emph{Generic} performs poorly accross the board. \emph{Omni} seems to
-  have an anchoring effect, in that it never performs terribly but also
-  never as good as the best \emph{ECCo} results. \emph{REVISE} performs
-  poorly across the board.
-\item
-  \textbf{Cost} (Table~\ref{tbl-lin_sep-lambda_cost_exper}): Results for
-  all generators (except \emph{Omni}) are quite bad, which can likely be
-  attributed to extremely bad results for some choices of the
-  \textbf{Energy Penalty} (results here are averaged). For \emph{ECCo}
-  and \emph{Generic}, higher cost values generally lead to worse
-  results.
-\item
-  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
-  seems that smaller choices are ok.
-\item
-  \textbf{Validity}: \emph{ECCo} almost always valid except for very low
-  values during training and high values at evaluation time.
-  \emph{Generic} often has poor validity.
-\item
-  \textbf{Accuracy}: Seems largely unaffected.
-\end{itemize}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly
-Separable data by energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0.01 & \textit{ECCo} & $-9.91 \cdot 10^{11}$ & $2.25 \cdot 10^{12}$ \\
-  full & 0.01 & \textit{Generic} & $-5.71 \cdot 10^{17}$ & $1.3 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.116}} \\
-  full & 0.01 & \textit{REVISE} & -15.6 & 13.2 \\
-  vanilla & 0.01 & \textit{ECCo} & -4.28 & 3.52 \\
-  vanilla & 0.01 & \textit{Generic} & -4.45 & 3.47 \\
-  vanilla & 0.01 & \textit{Omniscient} & -5.12 & 4.46 \\
-  vanilla & 0.01 & \textit{REVISE} & -4.91 & 4.24 \\
-  full & 0.05 & \textit{ECCo} & $-5.63 \cdot 10^{5}$ & $1.28 \cdot 10^{6}$ \\
-  full & 0.05 & \textit{Generic} & $-8.35 \cdot 10^{17}$ & $1.9 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.114}} \\
-  full & 0.05 & \textit{REVISE} & -15 & 12.6 \\
-  vanilla & 0.05 & \textit{ECCo} & -4.4 & 3.66 \\
-  vanilla & 0.05 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 0.05 & \textit{Omniscient} & -5.25 & 4.62 \\
-  vanilla & 0.05 & \textit{REVISE} & -4.94 & 4.22 \\
-  full & 0.1 & \textit{ECCo} & $-6.74 \cdot 10^{5}$ & $1.53 \cdot 10^{6}$ \\
-  full & 0.1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.56}} & \color{blue}{\textbf{0.124}} \\
-  full & 0.1 & \textit{REVISE} & -15.6 & 13.2 \\
-  vanilla & 0.1 & \textit{ECCo} & -4.28 & 3.52 \\
-  vanilla & 0.1 & \textit{Generic} & -4.45 & 3.48 \\
-  vanilla & 0.1 & \textit{Omniscient} & -5.12 & 4.46 \\
-  vanilla & 0.1 & \textit{REVISE} & -4.91 & 4.25 \\
-  full & 0.5 & \textit{ECCo} & -11.8 & 9.83 \\
-  full & 0.5 & \textit{Generic} & $-1.06 \cdot 10^{18}$ & $2.42 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.54}} & \color{blue}{\textbf{0.123}} \\
-  full & 0.5 & \textit{REVISE} & -15 & 12.6 \\
-  vanilla & 0.5 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 0.5 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 0.5 & \textit{Omniscient} & -5.25 & 4.61 \\
-  vanilla & 0.5 & \textit{REVISE} & -4.95 & 4.22 \\
-  full & 1 & \textit{ECCo} & -11.5 & 11.1 \\
-  full & 1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.88 \cdot 10^{11}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.59}} & \color{blue}{\textbf{0.117}} \\
-  full & 1 & \textit{REVISE} & -15.7 & 13.3 \\
-  vanilla & 1 & \textit{ECCo} & -4.28 & 3.51 \\
-  vanilla & 1 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 1 & \textit{Omniscient} & -5.11 & 4.46 \\
-  vanilla & 1 & \textit{REVISE} & -4.91 & 4.25 \\
-  full & 5 & \textit{ECCo} & -3.99 & 3.12 \\
-  full & 5 & \textit{Generic} & $-4.88 \cdot 10^{17}$ & $1.11 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.53}} & \color{blue}{\textbf{0.117}} \\
-  full & 5 & \textit{REVISE} & -14.6 & 12.1 \\
-  vanilla & 5 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 5 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 5 & \textit{Omniscient} & -5.25 & 4.61 \\
-  vanilla & 5 & \textit{REVISE} & -4.95 & 4.22 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.31}} & \color{blue}{\textbf{0.735}} \\
-  full & 10 & \textit{Generic} & $-1.7 \cdot 10^{11}$ & $3.86 \cdot 10^{11}$ \\
-  full & 10 & \textit{Omniscient} & -2.53 & 0.117 \\
-  full & 10 & \textit{REVISE} & -15.5 & 13 \\
-  vanilla & 10 & \textit{ECCo} & -4.28 & 3.51 \\
-  vanilla & 10 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 10 & \textit{Omniscient} & -5.12 & 4.46 \\
-  vanilla & 10 & \textit{REVISE} & -4.91 & 4.24 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.01}} & \color{blue}{\textbf{0.488}} \\
-  full & 15 & \textit{Generic} & $-4.91 \cdot 10^{17}$ & $1.12 \cdot 10^{18}$ \\
-  full & 15 & \textit{Omniscient} & -2.53 & 0.116 \\
-  full & 15 & \textit{REVISE} & -14.4 & 11.7 \\
-  vanilla & 15 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 15 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 15 & \textit{Omniscient} & -5.25 & 4.6 \\
-  vanilla & 15 & \textit{REVISE} & -4.95 & 4.23 \\\bottomrule
-
-\end{longtable}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-lin_sep-lambda_cost_exper}Results for Linearly
-Separable data by cost penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0 & \textit{ECCo} & $-5.32 \cdot 10^{3}$ & $1.21 \cdot 10^{4}$ \\
-  full & 0 & \textit{Generic} & $-1.03 \cdot 10^{18}$ & $2.34 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.64}} & \color{blue}{\textbf{0.125}} \\
-  full & 0 & \textit{REVISE} & -15.4 & 12.9 \\
-  vanilla & 0 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0 & \textit{Omniscient} & -5.18 & 4.54 \\
-  vanilla & 0 & \textit{REVISE} & -4.93 & 4.23 \\
-  full & 0.001 & \textit{ECCo} & -362 & 811 \\
-  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.04 \cdot 10^{17}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.49}} & \color{blue}{\textbf{0.115}} \\
-  full & 0.001 & \textit{REVISE} & -15.5 & 13 \\
-  vanilla & 0.001 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0.001 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0.001 & \textit{Omniscient} & -5.18 & 4.53 \\
-  vanilla & 0.001 & \textit{REVISE} & -4.93 & 4.23 \\
-  full & 0.1 & \textit{ECCo} & $-3.72 \cdot 10^{11}$ & $8.46 \cdot 10^{11}$ \\
-  full & 0.1 & \textit{Generic} & $-4.48 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-2.5}} & \color{blue}{\textbf{0.112}} \\
-  full & 0.1 & \textit{REVISE} & -14.6 & 12.2 \\
-  vanilla & 0.1 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0.1 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0.1 & \textit{Omniscient} & -5.18 & 4.54 \\
-  vanilla & 0.1 & \textit{REVISE} & -4.93 & 4.24 \\\bottomrule
-
-\end{longtable}
-
-\paragraph{Moons}\label{moons}
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-moons-lambda_energy_exper}):
-  \emph{ECCo} consistently yields better results than \emph{Vanilla},
-  except for very low choices of the energy penalty during training for
-  which it performs abismal. \emph{Generic} performs quite badly across
-  the board for high enough choices of the energy penalty at evaluation
-  time. \emph{Omni} has small positive effect. \emph{REVISE} performs
-  poorly across the board.
-\item
-  \textbf{Cost (distance penalty)}: \emph{Generic} generally does better
-  for higher values, while \emph{ECCo} does better for lower values.
-\item
-  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
-  seems that smaller choices are ok.
-\item
-  \textbf{Validity}: \emph{ECCo} generally achieves full validity except
-  for very low choices the energy penalty during training and high
-  choices at evaluation time. \emph{Generic} performs poorly for high
-  choices of the energy penalty during evaluation.
-\item
-  \textbf{Accuracy}: Largely unaffected although \emph{ECCo} suffers a
-  bit for very low choices the energy penalty during training.
-  \emph{REVISE} suffers a lot in general (around 10 percentage points).
-\end{itemize}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-moons-lambda_energy_exper}Results for Moons data by
-energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0.01 & \textit{ECCo} & $-2.8 \cdot 10^{22}$ & $6.39 \cdot 10^{22}$ \\
-  full & 0.01 & \textit{Generic} & $-4.89 \cdot 10^{30}$ & $1.11 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.74}} & \color{blue}{\textbf{5.08}} \\
-  full & 0.01 & \textit{REVISE} & -572 & $1.25 \cdot 10^{3}$ \\
-  vanilla & 0.01 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 0.01 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.01 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 0.01 & \textit{REVISE} & -11.2 & 13 \\
-  full & 0.05 & \textit{ECCo} & $-1.55 \cdot 10^{16}$ & $3.52 \cdot 10^{16}$ \\
-  full & 0.05 & \textit{Generic} & $-2.22 \cdot 10^{20}$ & $5 \cdot 10^{20}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.41}} & \color{blue}{\textbf{4.48}} \\
-  full & 0.05 & \textit{REVISE} & $-1.04 \cdot 10^{3}$ & $2.3 \cdot 10^{3}$ \\
-  vanilla & 0.05 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.05 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 0.05 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 0.05 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 0.1 & \textit{ECCo} & $-3.41 \cdot 10^{3}$ & $7.73 \cdot 10^{3}$ \\
-  full & 0.1 & \textit{Generic} & $-5.22 \cdot 10^{30}$ & $1.19 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.78}} & \color{blue}{\textbf{5.12}} \\
-  full & 0.1 & \textit{REVISE} & -288 & 594 \\
-  vanilla & 0.1 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.1 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 0.1 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 0.5 & \textit{ECCo} & -7.09 & 7.51 \\
-  full & 0.5 & \textit{Generic} & $-1.11 \cdot 10^{31}$ & $2.53 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.58}} & \color{blue}{\textbf{4.83}} \\
-  full & 0.5 & \textit{REVISE} & $-1.19 \cdot 10^{3}$ & $2.64 \cdot 10^{3}$ \\
-  vanilla & 0.5 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.5 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 0.5 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 0.5 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 1 & \textit{ECCo} & -6.06 & 6.33 \\
-  full & 1 & \textit{Generic} & $-1.58 \cdot 10^{33}$ & $3.59 \cdot 10^{33}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{Omniscient}} & \color{blue}{\textbf{-4.66}} & \color{blue}{\textbf{4.89}} \\
-  full & 1 & \textit{REVISE} & $-1.16 \cdot 10^{3}$ & $2.59 \cdot 10^{3}$ \\
-  vanilla & 1 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 1 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 1 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.57}} & \color{blue}{\textbf{2.07}} \\
-  full & 5 & \textit{Generic} & $-1.17 \cdot 10^{28}$ & $2.66 \cdot 10^{28}$ \\
-  full & 5 & \textit{Omniscient} & -4.29 & 4.31 \\
-  full & 5 & \textit{REVISE} & -530 & $1.16 \cdot 10^{3}$ \\
-  vanilla & 5 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 5 & \textit{Generic} & -11.7 & 12.7 \\
-  vanilla & 5 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 5 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.76}} & \color{blue}{\textbf{0.974}} \\
-  full & 10 & \textit{Generic} & $-1.54 \cdot 10^{33}$ & $3.51 \cdot 10^{33}$ \\
-  full & 10 & \textit{Omniscient} & -4.44 & 4.56 \\
-  full & 10 & \textit{REVISE} & $-1.52 \cdot 10^{3}$ & $3.4 \cdot 10^{3}$ \\
-  vanilla & 10 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 10 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 10 & \textit{Omniscient} & -12.7 & 14.4 \\
-  vanilla & 10 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.37}} & \color{blue}{\textbf{0.365}} \\
-  full & 15 & \textit{Generic} & $-5.32 \cdot 10^{28}$ & $1.21 \cdot 10^{29}$ \\
-  full & 15 & \textit{Omniscient} & -4.34 & 4.38 \\
-  full & 15 & \textit{REVISE} & -473 & $1.03 \cdot 10^{3}$ \\
-  vanilla & 15 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 15 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 15 & \textit{Omniscient} & -12.4 & 14.1 \\
-  vanilla & 15 & \textit{REVISE} & -11.3 & 13.1 \\\bottomrule
-
-\end{longtable}
-
-\paragraph{Circles}\label{circles}
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-circles-lambda_energy_exper}):
-  \emph{ECCo} consistently yields better results than \emph{Vanilla},
-  though primarily for low to medium choices of the energy penalty
-  (\textless=5) during training. The same goes for \emph{Generic}, which
-  sometimes outperforms \emph{ECCo} (for small energy penalty at
-  evaluation time). \emph{Omni} does alright for lower energy penalty at
-  evaluation time, but loses out for higher choices. \emph{REVISE}
-  performs poorly across the board (except very low choices at
-  evaluation time).
-\item
-  \textbf{Cost (distance penalty)}: \emph{ECCo} and \emph{Generic}
-  generally achieve the best results when no cost penalty is used during
-  training. Both \emph{Omni} and \emph{REVISE} are largely unaffected.
-\item
-  \textbf{Maximum Iterations}: \emph{ECCo} consistently yields better
-  results for higher numbers of iterations. \emph{Generic} generally
-  does best for a medium number (50). \emph{Omni} is sometimes invalid
-  (\textbf{???}).
-\item
-  \textbf{Validity}: \emph{ECCo} tends to outperform its \emph{Vanilla}
-  counterpart, though primarily for low to medium choices of the energy
-  penalty (\textless=5) during training and evaluation. \emph{Vanilla}
-  typically worse across the board.
-\item
-  \textbf{Accuracy}: Mostly unaffected, but \emph{REVISE} again
-  consistently some deterioration and \emph{ECCo} deteriorates for high
-  choices of energy penalty during training, reflecting other outcomes
-  above.
-\end{itemize}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-circles-lambda_energy_exper}Results for Circles data
-by energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.26}} & \color{blue}{\textbf{0.423}} \\
-  full & 0.01 & \textit{Generic} & -1.49 & 0.71 \\
-  full & 0.01 & \textit{Omniscient} & -5.21 & 5.25 \\
-  full & 0.01 & \textit{REVISE} & $-2.71 \cdot 10^{26}$ & $6.37 \cdot 10^{26}$ \\
-  vanilla & 0.01 & \textit{ECCo} & -9.33 & 7.34 \\
-  vanilla & 0.01 & \textit{Generic} & -8.89 & 6.88 \\
-  vanilla & 0.01 & \textit{Omniscient} & -8.67 & 6.87 \\
-  vanilla & 0.01 & \textit{REVISE} & -8.65 & 6.8 \\
-  full & 0.05 & \textit{ECCo} & -1.29 & 0.397 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.356}} \\
-  full & 0.05 & \textit{Omniscient} & -5.08 & 5.09 \\
-  full & 0.05 & \textit{REVISE} & $-5.91 \cdot 10^{27}$ & $1.36 \cdot 10^{28}$ \\
-  vanilla & 0.05 & \textit{ECCo} & -9.35 & 7.32 \\
-  vanilla & 0.05 & \textit{Generic} & -8.85 & 6.87 \\
-  vanilla & 0.05 & \textit{Omniscient} & -8.7 & 6.96 \\
-  vanilla & 0.05 & \textit{REVISE} & -8.52 & 6.76 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.2}} & \color{blue}{\textbf{0.383}} \\
-  full & 0.1 & \textit{Generic} & -1.5 & 0.735 \\
-  full & 0.1 & \textit{Omniscient} & -5.17 & 5.23 \\
-  full & 0.1 & \textit{REVISE} & $-3.06 \cdot 10^{26}$ & $7.7 \cdot 10^{26}$ \\
-  vanilla & 0.1 & \textit{ECCo} & -9.33 & 7.32 \\
-  vanilla & 0.1 & \textit{Generic} & -8.88 & 6.86 \\
-  vanilla & 0.1 & \textit{Omniscient} & -8.69 & 6.9 \\
-  vanilla & 0.1 & \textit{REVISE} & -8.68 & 6.81 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.12}} & \color{blue}{\textbf{0.217}} \\
-  full & 0.5 & \textit{Generic} & -1.21 & 0.352 \\
-  full & 0.5 & \textit{Omniscient} & -5.09 & 5.12 \\
-  full & 0.5 & \textit{REVISE} & $-5.97 \cdot 10^{27}$ & $1.37 \cdot 10^{28}$ \\
-  vanilla & 0.5 & \textit{ECCo} & -9.35 & 7.3 \\
-  vanilla & 0.5 & \textit{Generic} & -8.89 & 6.92 \\
-  vanilla & 0.5 & \textit{Omniscient} & -8.68 & 6.93 \\
-  vanilla & 0.5 & \textit{REVISE} & -8.53 & 6.75 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.1}} & \color{blue}{\textbf{0.163}} \\
-  full & 1 & \textit{Generic} & -1.49 & 0.726 \\
-  full & 1 & \textit{Omniscient} & -5.16 & 5.2 \\
-  full & 1 & \textit{REVISE} & $-3.09 \cdot 10^{26}$ & $7.22 \cdot 10^{26}$ \\
-  vanilla & 1 & \textit{ECCo} & -9.34 & 7.36 \\
-  vanilla & 1 & \textit{Generic} & -8.86 & 6.85 \\
-  vanilla & 1 & \textit{Omniscient} & -8.7 & 6.9 \\
-  vanilla & 1 & \textit{REVISE} & -8.69 & 6.85 \\
-  full & 5 & \textit{ECCo} & -1.75 & 0.154 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.21}} & \color{blue}{\textbf{0.363}} \\
-  full & 5 & \textit{Omniscient} & -5.14 & 5.16 \\
-  full & 5 & \textit{REVISE} & $-1.1 \cdot 10^{28}$ & $2.5 \cdot 10^{28}$ \\
-  vanilla & 5 & \textit{ECCo} & -9.36 & 7.32 \\
-  vanilla & 5 & \textit{Generic} & -8.88 & 6.91 \\
-  vanilla & 5 & \textit{Omniscient} & -8.7 & 6.93 \\
-  vanilla & 5 & \textit{REVISE} & -8.52 & 6.73 \\
-  full & 10 & \textit{ECCo} & $-1.02 \cdot 10^{6}$ & $2.32 \cdot 10^{6}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.49}} & \color{blue}{\textbf{0.702}} \\
-  full & 10 & \textit{Omniscient} & -5.13 & 5.16 \\
-  full & 10 & \textit{REVISE} & $-3.74 \cdot 10^{26}$ & $9.09 \cdot 10^{26}$ \\
-  vanilla & 10 & \textit{ECCo} & -9.31 & 7.33 \\
-  vanilla & 10 & \textit{Generic} & -8.87 & 6.86 \\
-  vanilla & 10 & \textit{Omniscient} & -8.7 & 6.89 \\
-  vanilla & 10 & \textit{REVISE} & -8.69 & 6.83 \\
-  full & 15 & \textit{ECCo} & $-3.31 \cdot 10^{13}$ & $7.54 \cdot 10^{13}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-1.22}} & \color{blue}{\textbf{0.37}} \\
-  full & 15 & \textit{Omniscient} & -5.2 & 5.23 \\
-  full & 15 & \textit{REVISE} & $-9.01 \cdot 10^{27}$ & $2.06 \cdot 10^{28}$ \\
-  vanilla & 15 & \textit{ECCo} & -9.38 & 7.34 \\
-  vanilla & 15 & \textit{Generic} & -8.86 & 6.87 \\
-  vanilla & 15 & \textit{Omniscient} & -8.69 & 6.96 \\
-  vanilla & 15 & \textit{REVISE} & -8.51 & 6.73 \\\bottomrule
-
-\end{longtable}
-
-\end{document}
+In this section, we describe the training procedure in detail.
+While the details laid out here are not crucial for understanding our proposed approach, they are of importance to anyone looking to implement counterfactual training.

### 2025-02-24 paper/preprint/paper.tex
appendix

@@
 \usepackage{amsthm}
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
 \theoremstyle{definition}
 \newtheorem{exercise}{Parameters}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{plain}
@@
 \theoremstyle{definition}
-\newtheorem{example}{Example}[section]
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 
+\bibitem[\citeproctext]{ref-byrne2021mpi}
+Byrne, Simon, Lucas C. Wilcox, and Valentin Churavy. 2021. {``MPI.jl:
+Julia Bindings for the Message Passing Interface.''} \emph{Proceedings
+of the JuliaCon Conferences} 1 (1): 68.
+\url{https://doi.org/10.21105/jcon.00068}.
+
 \bibitem[\citeproctext]{ref-chagas2024pretty}
@@
 \url{https://doi.org/10.21105/joss.00602}.
+
+\bibitem[\citeproctext]{ref-joshi2019realistic}
+Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
+Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
+Actionable Explanations in Black-Box Decision Making Systems.''}
+\url{https://arxiv.org/abs/1907.09615}.
 
@@
 \subsubsection{Background}\label{sec-app-ce-background}
+
+Gradient-based counterfactual search was originally proposed by Wachter,
+Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}). It generally solves
+the following unconstrained objective,
+
+\[
+\begin{aligned}
+\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
+\end{aligned} 
+\]
+
+where \(g: \mathcal{Z} \mapsto \mathcal{X}\) is an invertible function
+that maps from the \(L\)-dimensional counterfactual state space to the
+feature space and \(\text{cost}(\cdot)\) denotes one or more penalties
+that are used to induce certain properties of the counterfactual
+outcome. As above, \(\mathbf{y}^+\) denotes the target output and
+\(\mathbf{M}_{\theta}(\mathbf{x})\) returns the logit predictions of the
+underlying classifier for \(\mathbf{x}=g(\mathbf{z})\).
+
+For all generators used in this work we use standard logit crossentropy
+loss for \(\text{yloss}(\cdot)\). All generators also penalize the
+distance (\(\ell_1\)-norm) of counterfactuals from their original
+factual state. For \emph{Generic} and \emph{ECCo}, we have
+\(\mathcal{Z}:=\mathcal{X}\) and
+\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
+are searched directly in the feature space. Conversely, \emph{REVISE}
+traverses the latent space of a variational autoencoder (VAE) fitted to
+the training data, where \(g(\cdot)\) corresponds to the decoder
+(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
+the distance penalty, \emph{ECCo} uses an additional penalty component
+that regularizes the energy associated with the counterfactual,
+\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
+al. 2024}).
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colbacktitle=quarto-callout-warning-color!10!white, opacitybacktitle=0.6, breakable, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, coltitle=black, opacityback=0, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, toprule=.15mm, toptitle=1mm, left=2mm, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, arc=.35mm, rightrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, breakable, left=2mm, colback=white, coltitle=black, bottomtitle=1mm, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, toprule=.15mm, bottomrule=.15mm, arc=.35mm, titlerule=0mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacityback=0, leftrule=.75mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, breakable, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, coltitle=black, opacityback=0, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, toprule=.15mm, toptitle=1mm, left=2mm, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, arc=.35mm, rightrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, breakable, left=2mm, colback=white, coltitle=black, bottomtitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, bottomrule=.15mm, arc=.35mm, titlerule=0mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, opacityback=0, leftrule=.75mm, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 information that might interfere with the double-blind review process.
-Since our experiments have largely involved highly parallel tasks and
-small models, we have made extensive use of multi-processing.
+Since our experiments involve highly parallel tasks and rather small
+models by today's standard, we have relied on distributed computing
+across multiple central processing units (CPU). Graphical processing
+units (GPU) were not required. For larger grid searches we used up to
+but typically less than 100 CPUs.
 
@@
 particular \href{https://aog.makie.org/v0.9.3/}{AlgebraOfGraphics.jl}.
+To distribute computational tasks across multiple processors, we have
+relied on \href{https://juliaparallel.org/MPI.jl/v0.20/}{MPI.jl}
+(\citeproc{ref-byrne2021mpi}{Byrne, Wilcox, and Churavy 2021}).
 

### 2025-02-24 paper/sections/appendix.qmd

@@
 Details about the cluster will be disclosed upon publication to avoid revealing information that might interfere with the double-blind review process.
-Since our experiments have largely involved highly parallel tasks and small models, we have made extensive use of multi-processing.
+Since our experiments involve highly parallel tasks and rather small models by today's standard, we have relied on distributed computing across multiple central processing units (CPU).
+Graphical processing units (GPU) were not required.
+For larger grid searches we used up to but typically less than 100 CPUs.
 
@@
 For plots and visualizations we used both [Plots.jl](https://docs.juliaplots.org/v1.40/) [@PlotsJL] and [Makie.jl](https://docs.makie.org/v0.22/) [@danisch2021makie], in particular [AlgebraOfGraphics.jl](https://aog.makie.org/v0.9.3/).
+To distribute computational tasks across multiple processors, we have relied on [MPI.jl](https://juliaparallel.org/MPI.jl/v0.20/) [@byrne2021mpi].

### 2025-02-24 paper/sections/other/counterfactuals.qmd

@@
 ### Background {#sec-app-ce-background}
+
+Gradient-based counterfactual search was originally proposed by @wachter2017counterfactual. It generally solves the following unconstrained objective,
+
+$$
+\begin{aligned}
+\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
+\end{aligned} 
+$$
+
+where $g: \mathcal{Z} \mapsto \mathcal{X}$ is an invertible function that maps from the $L$-dimensional counterfactual state space to the feature space and $\text{cost}(\cdot)$ denotes one or more penalties that are used to induce certain properties of the counterfactual outcome.
+As above, $\mathbf{y}^+$ denotes the target output and $\mathbf{M}_{\theta}(\mathbf{x})$ returns the logit predictions of the underlying classifier for $\mathbf{x}=g(\mathbf{z})$.
+
+For all generators used in this work we use standard logit crossentropy loss for $\text{yloss}(\cdot)$.
+All generators also penalize the distance ($\ell_1$-norm) of counterfactuals from their original factual state.
+For *Generic* and *ECCo*, we have $\mathcal{Z}:=\mathcal{X}$ and $g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}$, that is counterfactual are searched directly in the feature space.
+Conversely, *REVISE* traverses the latent space of a variational autoencoder (VAE) fitted to the training data, where $g(\cdot)$ corresponds to the decoder [@joshi2019realistic].
+In addition to the distance penalty, *ECCo* uses an additional penalty component that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful].
 

### 2025-02-24 paper/sections/appendix.qmd
buff

--- 2025-02-20 paper/sections/appendix.qmd
@@
 
-# Notation {.appendix}
-
-- $y^+$: The target class and also the index of the target class.
-- $y^-$: The non-target class and also the index of non-the target class.
-- $\mathbf{y}^+$: The one-hot encoded output vector for the target class. 
-- $\theta$: Model parameters (unspecified).
-- $\Theta$: Matrix of parameters.
+{{< include /paper/sections/other/notation.qmd >}}
 
@@
 
-\@Cynthia, \@Arie, I'm including some preliminary results here but will rerun experiments in the coming days.
+I'm including some preliminary results here but will rerun experiments in the coming days.
 You'll notice some odd outlier results (huge values) for the initial grid search (@sec-app-initial).
@@
 
-\@Cynthia, \@Arie, @fig-mnist shows much more plausible (faithful) counterfactuals for a model with CT than the model with conventional training (@fig-mnist-vanilla).
+@fig-mnist shows much more plausible (faithful) counterfactuals for a model with CT than the model with conventional training (@fig-mnist-vanilla).
 In fact, this is not even using *ECCo+* and still showing better results than the best results we achieved in our AAAI paper for JEM ensembles.
@@
 
-# Compute {.appendix}
-
-<!-- [@DHPC2022] -->
-
-We performed our experiments on a high-performance cluster.
-Details about the cluster will be disclosed upon publication to avoid revealing information that might interfere with the double-blind review process.
-Since our experiments have largely involved highly parallel tasks and small models, we have made extensive use of multi-processing.
+{{< include /paper/sections/other/compute.qmd >}}

### 2025-02-24 paper/sections/other/compute.qmd
buff

# Computation Details {.appendix}

<!-- [@DHPC2022] -->
## Hardware

We performed our experiments on a high-performance cluster. Details about the cluster will be disclosed upon publication to avoid revealing information that might interfere with the double-blind review process. Since our experiments involve highly parallel tasks and rather small models by today's standard, we have relied on distributed computing across multiple central processing units (CPU). Graphical processing units (GPU) were not required. For larger grid searches we used up to but typically less than 100 CPUs.

## Software

All computations were performed in the Julia Programming Language [@bezanson2017julia]. We have developed a package for counterfactual training that leverages and extends the functionality provided by several existing packages, most notably [CounterfactualExplanations.jl](https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl) [@altmeyer2023explaining] and the [Flux.jl](https://fluxml.ai/Flux.jl/v0.16/) library for deep learning [@innes2018fashionable;@innes2018flux]. For data-wrangling and presentation-ready tables we relied on [DataFrames.jl](https://dataframes.juliadata.org/v1.7/) [@milan2023dataframes] and [PrettyTables.jl](https://ronisbr.github.io/PrettyTables.jl/v2.4/) [@chagas2024pretty], respectively. For plots and visualizations we used both [Plots.jl](https://docs.juliaplots.org/v1.40/) [@PlotsJL] and [Makie.jl](https://docs.makie.org/v0.22/) [@danisch2021makie], in particular [AlgebraOfGraphics.jl](https://aog.makie.org/v0.9.3/). To distribute computational tasks across multiple processors, we have relied on [MPI.jl](https://juliaparallel.org/MPI.jl/v0.20/) [@byrne2021mpi].

### 2025-02-24 paper/sections/other/notation.qmd
buff

# Notation {.appendix}

- $y^+$: The target class and also the index of the target class.
- $y^-$: The non-target class and also the index of non-the target class.
- $\mathbf{y}^+$: The one-hot encoded output vector for the target class. 
- $\theta$: Model parameters (unspecified).
- $\Theta$: Matrix of parameters.

### 2025-02-24 paper/preprint/paper.tex
First review commit
Fixes some typos and minor inconsistencies to initialize review

--- 2025-02-24 paper/preprint/paper.tex
@@
 \begin{abstract}
-Counterfactual Explanations have emerged as a popular tool to explain
+Counterfactual explanations have emerged as a popular tool to explain
 predictions made by opaque machine learning models: they explain how
@@
 many solutions that are locally optimal with respect to narrow
-objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
+objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{For
   clarity: we follow standard ML convention in using ``degrees of
-  freedom'' to refer to the number of parameters estimated from data.}.
+  freedom'' to refer to the number of parameters estimated from data.}
 Based purely on predictive performance, these solutions may seem to
@@
 providing actionable recourse to the applicant since \emph{age} is not a
-mutable feature. For our intents and purposes, counterfactual training
+(directly) mutable feature. For our intents and purposes, counterfactual training
 aims at improving model explainability by aligning models with
@@
 \item
-  (Actionability) Permutations \(\Delta\) are subject to actionability
+  (Actionability) Permutations \(\Delta\) are subject to some actionability
   constraints.
@@
 pre-determined threshold, \(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
+\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\), where \(\mathcal{S}\) is the softmax function.\footnote{For
   detailed background information on gradient-based counterfactual
-  search and convergence see Section~\ref{sec-app-ce}.}.
+  search and convergence see Section~\ref{sec-app-ce}.}
 
@@
 counterfactuals get more plausible (Definition~\ref{def-explainability})
-over the cause of training, these two opposing effects gradually balance
+over the course of training, these two opposing effects gradually balance
 each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
@@
 key observations in this work is that we can leverage counterfactual
-explanations during training and get adversarial examples, essentially
+explanations during training and get adversarial examples essentially
 for free.

### 2025-03-06 paper/experiments/CTExperiments/src/presentation.jl
small things

@@ -17,6 +17,19 @@ global _lnstyvar = nothing
 global _dodgevar = nothing
 global _sidevar = nothing
 
+function adjust_plot_var(x::Union{Nothing,String}, cfg::CTExperiments.EvalConfigOrGrid)
+    if isnothing(x)
+        return x
+    else
+        if typeof(cfg) == EvaluationConfig
+            x = replace(x, "_exper" => "")
+            return x
+        else
+            return x
+        end
+    end
+end
+
 include("plots.jl")
 include("tables.jl")
 
@@ -247,8 +260,13 @@ function aggregate_ce_evaluation(
 
     # Aggregate:
     if "run" in names(df)
+        byvars_must_include = ["run", "lambda_energy_eval", "objective"]
+        byvars_must_include = byvars_must_include[[x in names(df) for x in byvars_must_include]]
         df_agg = aggregate_data(
-            df, y, byvars; byvars_must_include=["run", "lambda_energy_eval", "objective"]
+            df,
+            y,
+            byvars;
+            byvars_must_include=byvars_must_include,
         )
         if agg_runs
             # Compute mean of means and std of means:

### 2025-03-06 paper/experiments/present_results.jl

@@ -2,12 +2,14 @@ using CTExperiments
 using CTExperiments.CairoMakie
 using CTExperiments.CSV
 using CTExperiments.DataFrames
+using CTExperiments: adjust_plot_var
 using DotEnv
 
 DotEnv.load!()
 
 # Get config and set up grid:
 eval_grid = EvaluationGrid(get_config_from_args(; new_save_dir=ENV["OUTPUT_DIR"]))
+eval_list = load_list(eval_grid)
 exper_grid = ExperimentGrid(eval_grid.grid_file)
 
 local_save_dir = get_work_dir(eval_grid, ENV["EVAL_WORK_DIR"], ENV["OUTPUT_DIR"])
@@ -54,16 +56,17 @@ end
 
 # Visualize CE:
 prefix = "ce"
+cfg = eval_grid
 all_data = CTExperiments.merge_with_meta(
-    eval_grid, CTExperiments.load_ce_evaluation(eval_grid)
+    cfg, CTExperiments.load_ce_evaluation(cfg)
 )
 valid_y = CTExperiments.valid_y_ce(all_data[1])
 params = PlotParams(;
-    colorvar=get_global_param("colorvar_ce", colorvar_ce),
-    rowvar=get_global_param("rowvar_ce", rowvar_ce),
-    colvar=get_global_param("colvar_ce", colvar_ce),
-    sidevar=get_global_param("sidevar", sidevar),
-    dodgevar=get_global_param("dodgevar", dodgevar),
+    colorvar=get_global_param("colorvar_ce", colorvar_ce) |> x -> adjust_plot_var(x,cfg),
+    rowvar=get_global_param("rowvar_ce", rowvar_ce) |> x -> adjust_plot_var(x, cfg),
+    colvar=get_global_param("colvar_ce", colvar_ce) |> x -> adjust_plot_var(x, cfg),
+    sidevar=get_global_param("sidevar", sidevar) |> x -> adjust_plot_var(x, cfg),
+    dodgevar=get_global_param("dodgevar", dodgevar) |> x -> adjust_plot_var(x, cfg),
 )
 @info "CE Boxplots"
 @info params

### 2025-03-06 paper/preprint/paper.tex
more work on appendix

--- 2025-02-24 paper/preprint/paper.tex
@@
 \usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
 \theoremstyle{plain}
@@
 \theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
+\newtheorem{example}{Example}[section]
+\theoremstyle{definition}
+\newtheorem{exercise}{Parameters}[section]
 \theoremstyle{plain}
 \newtheorem{corollary}{Research Question}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
 \theoremstyle{remark}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, breakable, left=2mm, colback=white, coltitle=black, bottomtitle=1mm, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, toprule=.15mm, bottomrule=.15mm, arc=.35mm, titlerule=0mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacityback=0, leftrule=.75mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white]
-
-@Cynthia, @Arie, I have tentatively phrased the above in terms of a
-theorem and proof. This is something I've so far shied away from because
-I feel a bit out of my depth when it comes to mathematical proofs. The
-above makes intuitive sense to me, but I don't know for sure if it's
-correct.
-
-\end{tcolorbox}
-
 \subsection{Domain Constraints}\label{domain-constraints}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, breakable, left=2mm, colback=white, coltitle=black, bottomtitle=1mm, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, toprule=.15mm, bottomrule=.15mm, arc=.35mm, titlerule=0mm, title=\textcolor{quarto-callout-warning-color}{\faExclamationTriangle}\hspace{0.5em}{Warning}, opacityback=0, leftrule=.75mm, toptitle=1mm, colbacktitle=quarto-callout-warning-color!10!white]
-
-@Cynthia, @Arie, I'm including some preliminary results here but will
-rerun experiments in the coming days. You'll notice some odd outlier
-results (huge values) for the initial grid search
-(Section~\ref{sec-app-initial}). My belief is that this is once again
-due to \emph{ECCo} overshooting for some hyperparameter choices, that we
-have discussed before. A simple solution to this story is to actually
-impose domain constraints. Let's see what the final results show us (I
-also still plan to make slight changes to the implementation), but in
-any case I think it might even be worth to report results for the
-initial grid search in the final appendix (they do include good results
-for CT for certain hyperparameter ranges and highlight limitations
-inherited from energy-based modelling).
-
-\end{tcolorbox}
-
 \subsection{Qualitative Findings for Image
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, breakable, left=2mm, colback=white, coltitle=black, bottomtitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, bottomrule=.15mm, arc=.35mm, titlerule=0mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, opacityback=0, leftrule=.75mm, toptitle=1mm, colbacktitle=quarto-callout-note-color!10!white]
-
-@Cynthia, @Arie, Figure~\ref{fig-mnist} shows much more plausible
-(faithful) counterfactuals for a model with CT than the model with
-conventional training (Figure~\ref{fig-mnist-vanilla}). In fact, this is
-not even using \emph{ECCo+} and still showing better results than the
-best results we achieved in our AAAI paper for JEM ensembles.
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, titlerule=0mm, coltitle=black, left=2mm, bottomtitle=1mm, breakable, opacityback=0, rightrule=.15mm, toptitle=1mm, arc=.35mm, bottomrule=.15mm, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6]
+
+Figure~\ref{fig-mnist} shows much more plausible (faithful)
+counterfactuals for a model with CT than the model with conventional
+training (Figure~\ref{fig-mnist-vanilla}). In fact, this is not even
+using \emph{ECCo+} and still showing better results than the best
+results we achieved in our AAAI paper for JEM ensembles.
 
@@
 
-\subsection{Initial Grid Search}\label{sec-app-initial}
-
-For the initial round of experiments we employed a different training
-objective and procedure that led to promising results for some
-hyperparameter choices, but suffered from training instabilities.
+\subsection{Grid Searches}\label{sec-app-grid}
 
@@
 
-The hyperparameter grids for the first investigation of the effect of
-generator parameters are shown in
-Parameters~\ref{exr-gen-params-first-run-train} and
-Parameters~\ref{exr-gen-params-first-run-eval}.
+The hyperparameter grid with varying generator parameters during
+training is shown in Parameters~\ref{exr-gen-params-final-run-train}.
+The corresponding evaluation grid used for these experiments is shown in
+Parameters~\ref{exr-gen-params-final-run-eval}.
 
 \begin{exercise}[Training
-Phase]\protect\hypertarget{exr-gen-params-first-run-train}{}\label{exr-gen-params-first-run-train}
+Phase]\protect\hypertarget{exr-gen-params-final-run-train}{}\label{exr-gen-params-final-run-train}
 
@@
   \item
-    Decisiono. Threshold: \texttt{0.75,\ 0.9,\ 0.95}
-  \item
-    \(\lambda_{\text{cost}}\): \texttt{0.0,\ 0.1}
-  \item
-    \(\lambda_{\text{div}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 10.0}
+    Decision Threshold: \texttt{0.75,\ 0.9,\ 0.95}
+  \item
+    \(\lambda_{\text{cost}}\): \texttt{0.001}
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 5.0,\ 10.0,\ 20.0}
+  \item
+    Learning Rate: \texttt{0.25}
   \item
@@
   \item
-    Optimizerimizer: \texttt{sgd}
+    Optimizer: \texttt{sgd}
   \end{itemize}
@@
 \item
-  Training Parameters:
+  Model Parameters:
 
@@
   \item
+    Activation: \texttt{relu}
+  \item
+    No.~Hidden: \texttt{32}
+  \item
+    No.~Layers: \texttt{1}
+  \end{itemize}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Burnin: \texttt{0.0}
+  \item
+    Class Loss: \texttt{logitcrossentropy}
+  \item
+    Convergence: \texttt{threshold}
+  \item
+    \(\lambda_{\text{adv}}\): \texttt{0.25}
+  \item
+    \(\lambda_{\text{yloss}}\): \texttt{1.0}
+  \item
+    \(\lambda_{\text{div}}\): \texttt{0.5}
+  \item
+    \(\lambda_{\text{reg}}\): \texttt{0.01}
+  \item
+    Learning Rate: \texttt{0.001}
+  \item
+    No.~Counterfactuals: \texttt{1000}
+  \item
+    No.~Epochs: \texttt{100}
+  \item
+    No.~Neighbours: \texttt{100}
+  \item
     Objective: \texttt{full,\ vanilla}
+  \item
+    Optimizer: \texttt{adam}
   \end{itemize}
@@
 \begin{exercise}[Evaluation
-Phase]\protect\hypertarget{exr-gen-params-first-run-eval}{}\label{exr-gen-params-first-run-eval}
+Phase]\protect\hypertarget{exr-gen-params-final-run-eval}{}\label{exr-gen-params-final-run-eval}
 
@@
   \item
-    Convergence: \texttt{max\_iter}
-  \item
-    Maximum Iterations: \texttt{100}
+    Convergence: \texttt{threshold}
+  \item
+    Decision Threshold: \texttt{0.95}
+  \item
+    Maximum Iterations: \texttt{50}
   \item
@@
   \item
+    Decision Threshold: \texttt{0.75}
+  \item
     \(\lambda_{\text{cost}}\): \texttt{0.0}
   \item
-    \(\lambda_{\text{div}}\):
-    \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0,\ 20.0}
-  \item
-    Learning Rate: \texttt{1.0}
-  \item
-    Maximum Iterations: \texttt{50}
-  \item
-    Optimizerimizer: \texttt{sgd}
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \item
+    Learning Rate: \texttt{0.25}
+  \item
+    Maximum Iterations: \texttt{30}
+  \item
+    Optimizer: \texttt{sgd}
   \end{itemize}
@@
 
+\subsubsection{High-Level Findings}\label{high-level-findings}
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/percent_valid.png}}
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/acc_val.png}}
+
+\subsubsection{Detailed Findings}\label{detailed-findings}
+
 \paragraph{Linearly Separable}\label{linearly-separable}
 
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-lin_sep-lambda_energy_exper}):
-  \emph{ECCo} generally does yield better results than \emph{Vanilla}
-  for higher choices of the energy penalty (10,15) during training.
-  \emph{Generic} performs poorly accross the board. \emph{Omni} seems to
-  have an anchoring effect, in that it never performs terribly but also
-  never as good as the best \emph{ECCo} results. \emph{REVISE} performs
-  poorly across the board.
-\item
-  \textbf{Cost} (Table~\ref{tbl-lin_sep-lambda_cost_exper}): Results for
-  all generators (except \emph{Omni}) are quite bad, which can likely be
-  attributed to extremely bad results for some choices of the
-  \textbf{Energy Penalty} (results here are averaged). For \emph{ECCo}
-  and \emph{Generic}, higher cost values generally lead to worse
-  results.
-\item
-  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
-  seems that smaller choices are ok.
-\item
-  \textbf{Validity}: \emph{ECCo} almost always valid except for very low
-  values during training and high values at evaluation time.
-  \emph{Generic} often has poor validity.
-\item
-  \textbf{Accuracy}: Seems largely unaffected.
-\end{itemize}
-
-\begin{longtable}{ccccc}
+\begin{longtable}{cccccc}
 
@@
   \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \textbf{Decision Threshold} & \textbf{Lambda Energy} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
   \endfirsthead
   \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \textbf{Decision Threshold} & \textbf{Lambda Energy} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
   \endhead
   \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
+  \multicolumn{6}{r}{Continuing table below.}\\
   \bottomrule
@@
   \endlastfoot
-  full & 0.01 & \textit{ECCo} & $-9.91 \cdot 10^{11}$ & $2.25 \cdot 10^{12}$ \\
-  full & 0.01 & \textit{Generic} & $-5.71 \cdot 10^{17}$ & $1.3 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-15.6}} & \color{blue}{\textbf{13.2}} \\
-  vanilla & 0.01 & \textit{ECCo} & -4.28 & 3.52 \\
-  vanilla & 0.01 & \textit{Generic} & -4.45 & 3.47 \\
-  vanilla & 0.01 & \textit{REVISE} & -4.91 & 4.24 \\
-  full & 0.05 & \textit{ECCo} & $-5.63 \cdot 10^{5}$ & $1.28 \cdot 10^{6}$ \\
-  full & 0.05 & \textit{Generic} & $-8.35 \cdot 10^{17}$ & $1.9 \cdot 10^{18}$ \\
-  full & 0.05 & \textit{REVISE} & -15 & 12.6 \\
-  vanilla & 0.05 & \textit{ECCo} & -4.4 & 3.66 \\
-  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-4.38}} & \color{blue}{\textbf{3.48}} \\
-  vanilla & 0.05 & \textit{REVISE} & -4.94 & 4.22 \\
-  full & 0.1 & \textit{ECCo} & $-6.74 \cdot 10^{5}$ & $1.53 \cdot 10^{6}$ \\
-  full & 0.1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.9 \cdot 10^{11}$ \\
-  full & 0.1 & \textit{REVISE} & -15.6 & 13.2 \\
-  vanilla & 0.1 & \textit{ECCo} & -4.28 & 3.52 \\
-  vanilla & 0.1 & \textit{Generic} & -4.45 & 3.48 \\
-  vanilla & 0.1 & \textit{REVISE} & -4.91 & 4.25 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-11.8}} & \color{blue}{\textbf{9.83}} \\
-  full & 0.5 & \textit{Generic} & $-1.06 \cdot 10^{18}$ & $2.42 \cdot 10^{18}$ \\
-  full & 0.5 & \textit{REVISE} & -15 & 12.6 \\
-  vanilla & 0.5 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 0.5 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 0.5 & \textit{REVISE} & -4.95 & 4.22 \\
-  full & 1 & \textit{ECCo} & -11.5 & 11.1 \\
-  full & 1 & \textit{Generic} & $-1.71 \cdot 10^{11}$ & $3.88 \cdot 10^{11}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-15.7}} & \color{blue}{\textbf{13.3}} \\
-  vanilla & 1 & \textit{ECCo} & -4.28 & 3.51 \\
-  vanilla & 1 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 1 & \textit{REVISE} & -4.91 & 4.25 \\
-  full & 5 & \textit{ECCo} & -3.99 & 3.12 \\
-  full & 5 & \textit{Generic} & $-4.88 \cdot 10^{17}$ & $1.11 \cdot 10^{18}$ \\
-  full & 5 & \textit{REVISE} & -14.6 & 12.1 \\
-  vanilla & 5 & \textit{ECCo} & -4.4 & 3.65 \\
-  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-4.38}} & \color{blue}{\textbf{3.48}} \\
-  vanilla & 5 & \textit{REVISE} & -4.95 & 4.22 \\
-  full & 10 & \textit{ECCo} & -2.31 & 0.735 \\
-  full & 10 & \textit{Generic} & $-1.7 \cdot 10^{11}$ & $3.86 \cdot 10^{11}$ \\
-  full & 10 & \textit{REVISE} & -15.5 & 13 \\
-  vanilla & 10 & \textit{ECCo} & -4.28 & 3.51 \\
-  vanilla & 10 & \textit{Generic} & -4.44 & 3.47 \\
-  vanilla & 10 & \textit{REVISE} & -4.91 & 4.24 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{15}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-2.01}} & \color{blue}{\textbf{0.488}} \\
-  full & 15 & \textit{Generic} & $-4.91 \cdot 10^{17}$ & $1.12 \cdot 10^{18}$ \\
-  full & 15 & \textit{REVISE} & -14.4 & 11.7 \\
-  vanilla & 15 & \textit{ECCo} & -4.4 & 3.65 \\
-  vanilla & 15 & \textit{Generic} & -4.38 & 3.48 \\
-  vanilla & 15 & \textit{REVISE} & -4.95 & 4.23 \\\bottomrule
-
-\end{longtable}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-lin_sep-lambda_cost_exper}Results for Linearly
-Separable data by cost penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{cost}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0 & \textit{ECCo} & $-5.32 \cdot 10^{3}$ & $1.21 \cdot 10^{4}$ \\
-  full & 0 & \textit{Generic} & $-1.03 \cdot 10^{18}$ & $2.34 \cdot 10^{18}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-15.4}} & \color{blue}{\textbf{12.9}} \\
-  vanilla & 0 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0 & \textit{REVISE} & -4.93 & 4.23 \\
-  full & 0.001 & \textit{ECCo} & -362 & 811 \\
-  full & 0.001 & \textit{Generic} & $-2.65 \cdot 10^{17}$ & $6.04 \cdot 10^{17}$ \\
-  full & 0.001 & \textit{REVISE} & -15.5 & 13 \\
-  vanilla & 0.001 & \textit{ECCo} & -4.34 & 3.58 \\
-  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.001}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-4.41}} & \color{blue}{\textbf{3.48}} \\
-  vanilla & 0.001 & \textit{REVISE} & -4.93 & 4.23 \\
-  full & 0.1 & \textit{ECCo} & $-3.72 \cdot 10^{11}$ & $8.46 \cdot 10^{11}$ \\
-  full & 0.1 & \textit{Generic} & $-4.48 \cdot 10^{14}$ & $1.02 \cdot 10^{15}$ \\
-  full & 0.1 & \textit{REVISE} & -14.6 & 12.2 \\
-  vanilla & 0.1 & \textit{ECCo} & -4.34 & 3.58 \\
-  vanilla & 0.1 & \textit{Generic} & -4.41 & 3.48 \\
-  vanilla & 0.1 & \textit{REVISE} & -4.93 & 4.24 \\\bottomrule
-
-\end{longtable}
-
-\paragraph{Moons}\label{moons}
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-moons-lambda_energy_exper}):
-  \emph{ECCo} consistently yields better results than \emph{Vanilla},
-  except for very low choices of the energy penalty during training for
-  which it performs abismal. \emph{Generic} performs quite badly across
-  the board for high enough choices of the energy penalty at evaluation
-  time. \emph{Omni} has small positive effect. \emph{REVISE} performs
-  poorly across the board.
-\item
-  \textbf{Cost (distance penalty)}: \emph{Generic} generally does better
-  for higher values, while \emph{ECCo} does better for lower values.
-\item
-  \textbf{Maximum Iterations}: No clear patterns recognizable, so it
-  seems that smaller choices are ok.
-\item
-  \textbf{Validity}: \emph{ECCo} generally achieves full validity except
-  for very low choices the energy penalty during training and high
-  choices at evaluation time. \emph{Generic} performs poorly for high
-  choices of the energy penalty during evaluation.
-\item
-  \textbf{Accuracy}: Largely unaffected although \emph{ECCo} suffers a
-  bit for very low choices the energy penalty during training.
-  \emph{REVISE} suffers a lot in general (around 10 percentage points).
-\end{itemize}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-moons-lambda_energy_exper}Results for Moons data by
-energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  full & 0.01 & \textit{ECCo} & $-2.8 \cdot 10^{22}$ & $6.39 \cdot 10^{22}$ \\
-  full & 0.01 & \textit{Generic} & $-4.89 \cdot 10^{30}$ & $1.11 \cdot 10^{31}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-572}} & \color{blue}{\textbf{$1.25 \cdot 10^{3}$}} \\
-  vanilla & 0.01 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 0.01 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.01 & \textit{REVISE} & -11.2 & 13 \\
-  full & 0.05 & \textit{ECCo} & $-1.55 \cdot 10^{16}$ & $3.52 \cdot 10^{16}$ \\
-  full & 0.05 & \textit{Generic} & $-2.22 \cdot 10^{20}$ & $5 \cdot 10^{20}$ \\
-  full & 0.05 & \textit{REVISE} & $-1.04 \cdot 10^{3}$ & $2.3 \cdot 10^{3}$ \\
-  vanilla & 0.05 & \textit{ECCo} & -15.5 & 17.2 \\
-  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-11.7}} & \color{blue}{\textbf{12.8}} \\
-  vanilla & 0.05 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 0.1 & \textit{ECCo} & $-3.41 \cdot 10^{3}$ & $7.73 \cdot 10^{3}$ \\
-  full & 0.1 & \textit{Generic} & $-5.22 \cdot 10^{30}$ & $1.19 \cdot 10^{31}$ \\
-  full & 0.1 & \textit{REVISE} & -288 & 594 \\
-  vanilla & 0.1 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 0.1 & \textit{REVISE} & -11.3 & 13.1 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.5}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-7.09}} & \color{blue}{\textbf{7.51}} \\
-  full & 0.5 & \textit{Generic} & $-1.11 \cdot 10^{31}$ & $2.53 \cdot 10^{31}$ \\
-  full & 0.5 & \textit{REVISE} & $-1.19 \cdot 10^{3}$ & $2.64 \cdot 10^{3}$ \\
-  vanilla & 0.5 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 0.5 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 0.5 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 1 & \textit{ECCo} & -6.06 & 6.33 \\
-  full & 1 & \textit{Generic} & $-1.58 \cdot 10^{33}$ & $3.59 \cdot 10^{33}$ \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{$-1.16 \cdot 10^{3}$}} & \color{blue}{\textbf{$2.59 \cdot 10^{3}$}} \\
-  vanilla & 1 & \textit{ECCo} & -15.5 & 17.3 \\
-  vanilla & 1 & \textit{Generic} & -10.9 & 11.9 \\
-  vanilla & 1 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 5 & \textit{ECCo} & -2.57 & 2.07 \\
-  full & 5 & \textit{Generic} & $-1.17 \cdot 10^{28}$ & $2.66 \cdot 10^{28}$ \\
-  full & 5 & \textit{REVISE} & -530 & $1.16 \cdot 10^{3}$ \\
-  vanilla & 5 & \textit{ECCo} & -15.5 & 17.2 \\
-  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-11.7}} & \color{blue}{\textbf{12.7}} \\
-  vanilla & 5 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 10 & \textit{ECCo} & -1.76 & 0.974 \\
-  full & 10 & \textit{Generic} & $-1.54 \cdot 10^{33}$ & $3.51 \cdot 10^{33}$ \\
-  full & 10 & \textit{REVISE} & $-1.52 \cdot 10^{3}$ & $3.4 \cdot 10^{3}$ \\
-  vanilla & 10 & \textit{ECCo} & -15.5 & 17.3 \\
-  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-10.9}} & \color{blue}{\textbf{11.9}} \\
-  vanilla & 10 & \textit{REVISE} & -11.3 & 13.1 \\
-  full & 15 & \textit{ECCo} & -1.37 & 0.365 \\
-  full & 15 & \textit{Generic} & $-5.32 \cdot 10^{28}$ & $1.21 \cdot 10^{29}$ \\
-  full & 15 & \textit{REVISE} & -473 & $1.03 \cdot 10^{3}$ \\
-  vanilla & 15 & \textit{ECCo} & -15.5 & 17.2 \\
-  vanilla & 15 & \textit{Generic} & -11.7 & 12.8 \\
-  vanilla & 15 & \textit{REVISE} & -11.3 & 13.1 \\\bottomrule
-
-\end{longtable}
-
-\paragraph{Circles}\label{circles}
-
-\begin{itemize}
-\tightlist
-\item
-  \textbf{Energy Penalty} (Table~\ref{tbl-circles-lambda_energy_exper}):
-  \emph{ECCo} consistently yields better results than \emph{Vanilla},
-  though primarily for low to medium choices of the energy penalty
-  (\textless=5) during training. The same goes for \emph{Generic}, which
-  sometimes outperforms \emph{ECCo} (for small energy penalty at
-  evaluation time). \emph{Omni} does alright for lower energy penalty at
-  evaluation time, but loses out for higher choices. \emph{REVISE}
-  performs poorly across the board (except very low choices at
-  evaluation time).
-\item
-  \textbf{Cost (distance penalty)}: \emph{ECCo} and \emph{Generic}
-  generally achieve the best results when no cost penalty is used during
-  training. Both \emph{Omni} and \emph{REVISE} are largely unaffected.
-\item
-  \textbf{Maximum Iterations}: \emph{ECCo} consistently yields better
-  results for higher numbers of iterations. \emph{Generic} generally
-  does best for a medium number (50). \emph{Omni} is sometimes invalid
-  (\textbf{???}).
-\item
-  \textbf{Validity}: \emph{ECCo} tends to outperform its \emph{Vanilla}
-  counterpart, though primarily for low to medium choices of the energy
-  penalty (\textless=5) during training and evaluation. \emph{Vanilla}
-  typically worse across the board.
-\item
-  \textbf{Accuracy}: Mostly unaffected, but \emph{REVISE} again
-  consistently some deterioration and \emph{ECCo} deteriorates for high
-  choices of energy penalty during training, reflecting other outcomes
-  above.
-\end{itemize}
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-circles-lambda_energy_exper}Results for Circles data
-by energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Objective} & \textbf{$\lambda_{\text{div}} (\text{train})$} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{0.01}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.26}} & \color{blue}{\textbf{0.423}} \\
-  full & 0.01 & \textit{Generic} & -1.49 & 0.71 \\
-  full & 0.01 & \textit{REVISE} & $-2.71 \cdot 10^{26}$ & $6.37 \cdot 10^{26}$ \\
-  vanilla & 0.01 & \textit{ECCo} & -9.33 & 7.34 \\
-  vanilla & 0.01 & \textit{Generic} & -8.89 & 6.88 \\
-  vanilla & 0.01 & \textit{REVISE} & -8.65 & 6.8 \\
-  full & 0.05 & \textit{ECCo} & -1.29 & 0.397 \\
-  full & 0.05 & \textit{Generic} & -1.21 & 0.356 \\
-  full & 0.05 & \textit{REVISE} & $-5.91 \cdot 10^{27}$ & $1.36 \cdot 10^{28}$ \\
-  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.05}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-9.35}} & \color{blue}{\textbf{7.32}} \\
-  vanilla & 0.05 & \textit{Generic} & -8.85 & 6.87 \\
-  vanilla & 0.05 & \textit{REVISE} & -8.52 & 6.76 \\
-  full & 0.1 & \textit{ECCo} & -1.2 & 0.383 \\
-  full & 0.1 & \textit{Generic} & -1.5 & 0.735 \\
-  full & 0.1 & \textit{REVISE} & $-3.06 \cdot 10^{26}$ & $7.7 \cdot 10^{26}$ \\
-  vanilla & 0.1 & \textit{ECCo} & -9.33 & 7.32 \\
-  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{0.1}} & \color{blue}{\textbf{Generic}} & \color{blue}{\textbf{-8.88}} & \color{blue}{\textbf{6.86}} \\
-  vanilla & 0.1 & \textit{REVISE} & -8.68 & 6.81 \\
-  full & 0.5 & \textit{ECCo} & -1.12 & 0.217 \\
-  full & 0.5 & \textit{Generic} & -1.21 & 0.352 \\
-  full & 0.5 & \textit{REVISE} & $-5.97 \cdot 10^{27}$ & $1.37 \cdot 10^{28}$ \\
-  vanilla & 0.5 & \textit{ECCo} & -9.35 & 7.3 \\
-  vanilla & 0.5 & \textit{Generic} & -8.89 & 6.92 \\
-  vanilla & 0.5 & \textit{REVISE} & -8.53 & 6.75 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{1}} & \color{blue}{\textbf{ECCo}} & \color{blue}{\textbf{-1.1}} & \color{blue}{\textbf{0.163}} \\
-  full & 1 & \textit{Generic} & -1.49 & 0.726 \\
-  full & 1 & \textit{REVISE} & $-3.09 \cdot 10^{26}$ & $7.22 \cdot 10^{26}$ \\
-  vanilla & 1 & \textit{ECCo} & -9.34 & 7.36 \\
-  vanilla & 1 & \textit{Generic} & -8.86 & 6.85 \\
-  vanilla & 1 & \textit{REVISE} & -8.69 & 6.85 \\
-  full & 5 & \textit{ECCo} & -1.75 & 0.154 \\
-  full & 5 & \textit{Generic} & -1.21 & 0.363 \\
-  \color{blue}{\textbf{full}} & \color{blue}{\textbf{5}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{$-1.1 \cdot 10^{28}$}} & \color{blue}{\textbf{$2.5 \cdot 10^{28}$}} \\
-  vanilla & 5 & \textit{ECCo} & -9.36 & 7.32 \\
-  vanilla & 5 & \textit{Generic} & -8.88 & 6.91 \\
-  vanilla & 5 & \textit{REVISE} & -8.52 & 6.73 \\
-  full & 10 & \textit{ECCo} & $-1.02 \cdot 10^{6}$ & $2.32 \cdot 10^{6}$ \\
-  full & 10 & \textit{Generic} & -1.49 & 0.702 \\
-  full & 10 & \textit{REVISE} & $-3.74 \cdot 10^{26}$ & $9.09 \cdot 10^{26}$ \\
-  vanilla & 10 & \textit{ECCo} & -9.31 & 7.33 \\
-  vanilla & 10 & \textit{Generic} & -8.87 & 6.86 \\
-  \color{blue}{\textbf{vanilla}} & \color{blue}{\textbf{10}} & \color{blue}{\textbf{REVISE}} & \color{blue}{\textbf{-8.69}} & \color{blue}{\textbf{6.83}} \\
-  full & 15 & \textit{ECCo} & $-3.31 \cdot 10^{13}$ & $7.54 \cdot 10^{13}$ \\
-  full & 15 & \textit{Generic} & -1.22 & 0.37 \\
-  full & 15 & \textit{REVISE} & $-9.01 \cdot 10^{27}$ & $2.06 \cdot 10^{28}$ \\
-  vanilla & 15 & \textit{ECCo} & -9.38 & 7.34 \\
-  vanilla & 15 & \textit{Generic} & -8.86 & 6.87 \\
-  vanilla & 15 & \textit{REVISE} & -8.51 & 6.73 \\\bottomrule
+  0.75 & 0.1 & 5 & \textit{ECCo} & -15.9 & 16.7 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{41.9}} & \color{green}{\textbf{8.13}} \\
+  0.75 & 0.1 & 5 & \textit{REVISE} & -64.4 & 3.58 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{28.9}} & \color{green}{\textbf{2.28}} \\
+  0.75 & 0.1 & 25 & \textit{Generic} & -11.3 & 11.5 \\
+  0.75 & 0.1 & 25 & \textit{REVISE} & -55.3 & 4.18 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{10.9}} & \color{green}{\textbf{7.47}} \\
+  0.75 & 0.1 & 50 & \textit{Generic} & -32.3 & 16.6 \\
+  0.75 & 0.1 & 50 & \textit{REVISE} & -44.1 & 4.24 \\
+  0.9 & 0.1 & 5 & \textit{ECCo} & -13.7 & 13.7 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{22.3}} & \color{green}{\textbf{4.87}} \\
+  0.9 & 0.1 & 5 & \textit{REVISE} & -27.8 & 6.26 \\
+  0.9 & 0.1 & 25 & \textit{ECCo} & 8.84 & 7.13 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{30.7}} & \color{green}{\textbf{5.77}} \\
+  0.9 & 0.1 & 25 & \textit{REVISE} & -72.7 & 34.9 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{17.1}} & \color{green}{\textbf{7.98}} \\
+  0.9 & 0.1 & 50 & \textit{Generic} & 3.82 & 8.86 \\
+  0.9 & 0.1 & 50 & \textit{REVISE} & -45.8 & 2.79 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{35.7}} & \color{green}{\textbf{7.56}} \\
+  0.95 & 0.1 & 5 & \textit{Generic} & -13.3 & 5.73 \\
+  0.95 & 0.1 & 5 & \textit{REVISE} & -54.9 & 5.43 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{37}} & \color{green}{\textbf{5.31}} \\
+  0.95 & 0.1 & 25 & \textit{Generic} & 29 & 8.6 \\
+  0.95 & 0.1 & 25 & \textit{REVISE} & -40.5 & 41 \\
+  0.95 & 0.1 & 50 & \textit{ECCo} & 13.2 & 4.89 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{35.5}} & \color{green}{\textbf{4.27}} \\
+  0.95 & 0.1 & 50 & \textit{REVISE} & -60.4 & 45.8 \\
+  \color{red}{\textbf{0.75}} & \color{red}{\textbf{0.5}} & \color{red}{\textbf{5}} & \color{red}{\textbf{ECCo}} & \color{red}{\textbf{-0.94}} & \color{red}{\textbf{17.6}} \\
+  0.75 & 0.5 & 5 & \textit{Generic} & -11.7 & 17.4 \\
+  0.75 & 0.5 & 5 & \textit{REVISE} & -37.7 & 3.92 \\
+  0.75 & 0.5 & 25 & \textit{ECCo} & 14.7 & 6.32 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{32.3}} & \color{green}{\textbf{3.49}} \\
+  0.75 & 0.5 & 25 & \textit{REVISE} & -39.9 & 7.21 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{28.3}} & \color{green}{\textbf{5.05}} \\
+  0.75 & 0.5 & 50 & \textit{Generic} & -9.72 & 9.19 \\
+  0.75 & 0.5 & 50 & \textit{REVISE} & -30 & 5.22 \\
+  0.9 & 0.5 & 5 & \textit{ECCo} & -7.38 & 13.6 \\
+  \color{red}{\textbf{0.9}} & \color{red}{\textbf{0.5}} & \color{red}{\textbf{5}} & \color{red}{\textbf{Generic}} & \color{red}{\textbf{-0.0952}} & \color{red}{\textbf{17.1}} \\
+  0.9 & 0.5 & 5 & \textit{REVISE} & -50.6 & 4.77 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{39.1}} & \color{green}{\textbf{8.3}} \\
+  0.9 & 0.5 & 25 & \textit{Generic} & 38.6 & 6.03 \\
+  0.9 & 0.5 & 25 & \textit{REVISE} & -56.9 & 4.28 \\
+  0.9 & 0.5 & 50 & \textit{ECCo} & -6.53 & 10.9 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{13.6}} & \color{green}{\textbf{12.8}} \\
+  0.9 & 0.5 & 50 & \textit{REVISE} & -24.1 & 12.5 \\
+  0.95 & 0.5 & 5 & \textit{ECCo} & 0.596 & 1.11 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{4.57}} & \color{green}{\textbf{5.88}} \\
+  0.95 & 0.5 & 5 & \textit{REVISE} & -34.3 & 6.09 \\
+  0.95 & 0.5 & 25 & \textit{ECCo} & 16.5 & 3.53 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{19.1}} & \color{green}{\textbf{4.43}} \\
+  0.95 & 0.5 & 25 & \textit{REVISE} & -37.2 & 39.2 \\
+  0.95 & 0.5 & 50 & \textit{ECCo} & -31 & 14.2 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{24.8}} & \color{green}{\textbf{4.78}} \\
+  0.95 & 0.5 & 50 & \textit{REVISE} & -1.28 & 2.7 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{5}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{37}} & \color{green}{\textbf{9.21}} \\
+  0.75 & 5 & 5 & \textit{Generic} & 0.977 & 21.1 \\
+  0.75 & 5 & 5 & \textit{REVISE} & -55.2 & 4.75 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{21.8}} & \color{green}{\textbf{5.97}} \\
+  0.75 & 5 & 25 & \textit{Generic} & 17.8 & 5.33 \\
+  0.75 & 5 & 25 & \textit{REVISE} & -68.1 & 8.69 \\
+  0.75 & 5 & 50 & \textit{ECCo} & 0.864 & 9.51 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{8.56}} & \color{green}{\textbf{13.6}} \\
+  0.75 & 5 & 50 & \textit{REVISE} & -71.8 & 3.18 \\
+  0.9 & 5 & 5 & \textit{ECCo} & -3.59 & 20.3 \\
+  \color{red}{\textbf{0.9}} & \color{red}{\textbf{5}} & \color{red}{\textbf{5}} & \color{red}{\textbf{Generic}} & \color{red}{\textbf{-1.18}} & \color{red}{\textbf{16.7}} \\
+  0.9 & 5 & 5 & \textit{REVISE} & -58.6 & 3.55 \\
+  0.9 & 5 & 25 & \textit{ECCo} & -1.73 & 9.07 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{21.9}} & \color{green}{\textbf{7.48}} \\
+  0.9 & 5 & 25 & \textit{REVISE} & -44.4 & 4.23 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{30.8}} & \color{green}{\textbf{10.7}} \\
+  0.9 & 5 & 50 & \textit{Generic} & 16.3 & 11.1 \\
+  0.9 & 5 & 50 & \textit{REVISE} & -53.3 & 6.73 \\
+  \color{red}{\textbf{0.95}} & \color{red}{\textbf{5}} & \color{red}{\textbf{5}} & \color{red}{\textbf{ECCo}} & \color{red}{\textbf{-5.29}} & \color{red}{\textbf{24.5}} \\
+  0.95 & 5 & 5 & \textit{Generic} & -5.71 & 17.4 \\
+  0.95 & 5 & 5 & \textit{REVISE} & -36.5 & 4.01 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{26.9}} & \color{green}{\textbf{4.08}} \\
+  0.95 & 5 & 25 & \textit{Generic} & 19.5 & 5.49 \\
+  0.95 & 5 & 25 & \textit{REVISE} & -25 & 5 \\
+  0.95 & 5 & 50 & \textit{ECCo} & 4.6 & 7.46 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{16.6}} & \color{green}{\textbf{4.3}} \\
+  0.95 & 5 & 50 & \textit{REVISE} & -50.4 & 46.4 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{10}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{2.47}} & \color{green}{\textbf{8.57}} \\
+  0.75 & 10 & 5 & \textit{Generic} & -7.23 & 9.13 \\
+  0.75 & 10 & 5 & \textit{REVISE} & -67.2 & 5.75 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{10}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{29.9}} & \color{green}{\textbf{15.8}} \\
+  0.75 & 10 & 25 & \textit{Generic} & 4.47 & 8.22 \\
+  0.75 & 10 & 25 & \textit{REVISE} & -52 & 5.51 \\
+  0.75 & 10 & 50 & \textit{ECCo} & 6.23 & 6.11 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{10}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{15.2}} & \color{green}{\textbf{10.2}} \\
+  0.75 & 10 & 50 & \textit{REVISE} & -17 & 5.46 \\
+  0.9 & 10 & 5 & \textit{ECCo} & -6.93 & 14.7 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{10}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{6.88}} & \color{green}{\textbf{5.11}} \\
+  0.9 & 10 & 5 & \textit{REVISE} & -45.6 & 47.8 \\
+  0.9 & 10 & 25 & \textit{ECCo} & 6.2 & 6.75 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{10}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{34.7}} & \color{green}{\textbf{9.74}} \\
+  0.9 & 10 & 25 & \textit{REVISE} & -55.7 & 6.04 \\
+  0.9 & 10 & 50 & \textit{ECCo} & 30.1 & 14.2 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{10}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{35}} & \color{green}{\textbf{7.11}} \\
+  0.9 & 10 & 50 & \textit{REVISE} & -48.8 & 3.3 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{10}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{35.7}} & \color{green}{\textbf{3.83}} \\
+  0.95 & 10 & 5 & \textit{Generic} & 2.85 & 7.9 \\
+  0.95 & 10 & 5 & \textit{REVISE} & -34.9 & 4.56 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{10}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{33}} & \color{green}{\textbf{6.45}} \\
+  0.95 & 10 & 25 & \textit{Generic} & 24.8 & 4.59 \\
+  0.95 & 10 & 25 & \textit{REVISE} & -55.7 & 44 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{10}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{8.07}} & \color{green}{\textbf{7.68}} \\
+  0.95 & 10 & 50 & \textit{Generic} & -7.11 & 9.2 \\
+  0.95 & 10 & 50 & \textit{REVISE} & -26.1 & 18.6 \\
+  0.75 & 20 & 5 & \textit{ECCo} & -10.8 & 20.8 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{20}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{12.6}} & \color{green}{\textbf{17.6}} \\
+  0.75 & 20 & 5 & \textit{REVISE} & -21.6 & 3.34 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{20}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{17.2}} & \color{green}{\textbf{4.25}} \\
+  0.75 & 20 & 25 & \textit{Generic} & -10.3 & 9.59 \\
+  0.75 & 20 & 25 & \textit{REVISE} & -59 & 3.66 \\
+  \color{green}{\textbf{0.75}} & \color{green}{\textbf{20}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{15.6}} & \color{green}{\textbf{8.38}} \\
+  0.75 & 20 & 50 & \textit{Generic} & 3.92 & 6.2 \\
+  0.75 & 20 & 50 & \textit{REVISE} & -70.9 & 5.16 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{20}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{26.3}} & \color{green}{\textbf{6.62}} \\
+  0.9 & 20 & 5 & \textit{Generic} & 2.35 & 7.04 \\
+  0.9 & 20 & 5 & \textit{REVISE} & -64.8 & 4.31 \\
+  0.9 & 20 & 25 & \textit{ECCo} & 21.1 & 11.6 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{20}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{39}} & \color{green}{\textbf{5.98}} \\
+  0.9 & 20 & 25 & \textit{REVISE} & -50.7 & 13.8 \\
+  \color{green}{\textbf{0.9}} & \color{green}{\textbf{20}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{15.1}} & \color{green}{\textbf{7.46}} \\
+  0.9 & 20 & 50 & \textit{Generic} & 8.53 & 6.95 \\
+  0.9 & 20 & 50 & \textit{REVISE} & -36 & 6.14 \\
+  0.95 & 20 & 5 & \textit{ECCo} & -23.8 & 17.8 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{20}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{7.23}} & \color{green}{\textbf{4.74}} \\
+  0.95 & 20 & 5 & \textit{REVISE} & -34 & 4.11 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{20}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{19.3}} & \color{green}{\textbf{14.6}} \\
+  0.95 & 20 & 25 & \textit{Generic} & -1.35 & 7.71 \\
+  0.95 & 20 & 25 & \textit{REVISE} & -52.8 & 5.11 \\
+  0.95 & 20 & 50 & \textit{ECCo} & 18.7 & 3.84 \\
+  \color{green}{\textbf{0.95}} & \color{green}{\textbf{20}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{37.3}} & \color{green}{\textbf{7.78}} \\
+  0.95 & 20 & 50 & \textit{REVISE} & -28.3 & 15 \\\bottomrule
 

### 2025-03-06 paper/sections/appendix.qmd
more work on appendix

@@
-```{julia}
-projectdir = splitpath(pwd()) |>
-    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
-cd(projectdir)
-```
-
-```{julia}
-using CTExperiments
-using CTExperiments.CSV
-using CTExperiments.DataFrames
-using CTExperiments.StatsBase
-
-using DotEnv
-DotEnv.load!()
-```
-
 \FloatBarrier
@@
 
-::: {.callout-warning}
-
-I'm including some preliminary results here but will rerun experiments in the coming days.
-You'll notice some odd outlier results (huge values) for the initial grid search (@sec-app-initial).
-My belief is that this is once again due to *ECCo* overshooting for some hyperparameter choices, that we have discussed before.
-A simple solution to this story is to actually impose domain constraints.
-Let's see what the final results show us (I also still plan to make slight changes to the implementation), but in any case I think it might even be worth to report results for the initial grid search in the final appendix (they do include good results for CT for certain hyperparameter ranges and highlight limitations inherited from energy-based modelling).
-
-:::
-
 ## Qualitative Findings for Image Data
@@
 
-## Initial Grid Search {#sec-app-initial}
+{{< include /paper/sections/other/grids.qmd >}}
 
-```{julia}
-results_dir = ENV["INITIAL_RUN_RESULTS"]
-config_dir = ENV["INITIAL_RUN_CONFIG"]
-```
-
-For the initial round of experiments we employed a different training objective and procedure that led to promising results for some hyperparameter choices, but suffered from training instabilities.
-
-### Generator Parameters
-
-The hyperparameter grids for the first investigation of the effect of generator parameters are shown in @exr-gen-params-first-run-train and @exr-gen-params-first-run-eval.
-
-::: {#exr-gen-params-first-run-train}
-
-## Training Phase
-
-```{julia}
-#| output: asis
-dict = CTExperiments.from_toml(joinpath(config_dir,"gen_params.toml")) 
-println(CTExperiments.dict_to_quarto_markdown(dict))
-```
-
-:::
-
-::: {#exr-gen-params-first-run-eval}
-
-## Evaluation Phase
-
-```{julia}
-#| output: asis
-dict = CTExperiments.from_toml(joinpath(results_dir, "gen_params/mlp/lin_sep/evaluation/evaluation_grid_config.toml"))
-println(CTExperiments.dict_to_quarto_markdown(dict))
-```
-
-:::
-
-```{julia}
-gen_params_dir = joinpath(results_dir, "gen_params/mlp")
-```
-
-#### Linearly Separable
-
-- **Energy Penalty** (@tbl-lin_sep-lambda_energy_exper): *ECCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training.
-*Generic* performs poorly accross the board.
-*Omni* seems to have an anchoring effect, in that it never performs terribly but also never as good as the best *ECCo* results.
-*REVISE* performs poorly across the board.
-- **Cost** (@tbl-lin_sep-lambda_cost_exper): Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged).
-For *ECCo* and *Generic*, higher cost values generally lead to worse results.
-- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
-- **Validity**: *ECCo* almost always valid except for very low values during training and high values at evaluation time. *Generic* often has poor validity.
-- **Accuracy**: Seems largely unaffected.
-
-```{julia}
-df = CSV.read("$(gen_params_dir)/lin_sep/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
-df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
-  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
-  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type])
-```
-
-::: {#tbl-lin_sep-lambda_energy_exper}
-
-::: {.content-hidden unless-format="pdf"}
-
-```{julia}
-#| output: asis
-get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; wrap_table=false)
-```
-
-:::
-
-Results for Linearly Separable data by energy penalty.
-
-:::
-
-<!-- Cost -->
-
-```{julia}
-df = CSV.read("$(gen_params_dir)/lin_sep/evaluation/results/ce/objective---lambda_cost_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
-df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
-  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
-  df -> sort(df, [:lambda_cost_exper, :objective, :generator_type])
-```
-
-::: {#tbl-lin_sep-lambda_cost_exper}
-
-::: {.content-hidden unless-format="pdf"}
-
-```{julia}
-#| output: asis
-get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_cost_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; wrap_table=false)
-```
-
-:::
-
-Results for Linearly Separable data by cost penalty.
-
-:::
-
-#### Moons
-
-- **Energy Penalty** (@tbl-moons-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, except for very low choices of the energy penalty during training for which it performs abismal.
-*Generic* performs quite badly across the board for high enough choices of the energy penalty at evaluation time.
-*Omni* has small positive effect.
-*REVISE* performs poorly across the board.
-- **Cost (distance penalty)**: *Generic* generally does better for higher values, while *ECCo* does better for lower values.
-- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
-- **Validity**: *ECCo* generally achieves full validity except for very low choices the energy penalty during training and high choices at evaluation time.
-*Generic* performs poorly for high choices of the energy penalty during evaluation.
-- **Accuracy**: Largely unaffected although *ECCo* suffers a bit for very low choices the energy penalty during training. *REVISE* suffers a lot in general (around 10 percentage points).
-
-```{julia}
-df = CSV.read("$(gen_params_dir)/moons/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
-df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
-  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
-  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) 
-```
-
-::: {#tbl-moons-lambda_energy_exper}
-
-::: {.content-hidden unless-format="pdf"}
-
-```{julia}
-#| output: asis
-get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs)
-```
-
-:::
-
-Results for Moons data by energy penalty.
-
-:::
-
-#### Circles
-
-- **Energy Penalty** (@tbl-circles-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, though primarily for low to medium choices of the energy penalty (<=5) during training.
-The same goes for *Generic*, which sometimes outperforms *ECCo* (for small energy penalty at evaluation time).
-*Omni* does alright for lower energy penalty at evaluation time, but loses out for higher choices.
-*REVISE* performs poorly across the board (except very low choices at evaluation time).
-- **Cost (distance penalty)**: *ECCo* and *Generic* generally achieve the best results when no cost penalty is used during training. Both *Omni* and *REVISE* are largely unaffected.
-- **Maximum Iterations**: *ECCo* consistently yields better results for higher numbers of iterations. *Generic* generally does best for a medium number (50). *Omni* is sometimes invalid (**???**).
-- **Validity**: *ECCo* tends to outperform its *Vanilla* counterpart, though primarily for low to medium choices of the energy penalty (<=5) during training and evaluation.
-*Vanilla* typically worse across the board.
-- **Accuracy**: Mostly unaffected, but *REVISE* again consistently some deterioration and *ECCo* deteriorates for high choices of energy penalty during training, reflecting other outcomes above.
-
-```{julia}
-df = CSV.read("$(gen_params_dir)/circles/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
-df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
-  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
-  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) 
-```
-
-::: {#tbl-circles-lambda_energy_exper}
-
-::: {.content-hidden unless-format="pdf"}
-
-```{julia}
-#| output: asis
-get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs)
-```
-
-:::
-
-Results for Circles data by energy penalty.
-
-:::
+<!-- {{< include /paper/sections/other/initial_grids.qmd >}} -->
 

### 2025-03-06 paper/sections/other/grids.qmd
more work on appendix

## Grid Searches {#sec-app-grid}

```{julia}
projectdir = splitpath(pwd()) |>
    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
cd(projectdir)

using CTExperiments
using CTExperiments.CSV
using CTExperiments.DataFrames
using CTExperiments.StatsBase

using DotEnv
DotEnv.load!()
```

```{julia}
res_dir = ENV["FINAL_GRID_RESULTS"]
```

### Generator Parameters

The hyperparameter grid with varying generator parameters during training is shown in @exr-gen-params-final-run-train. The corresponding evaluation grid used for these experiments is shown in @exr-gen-params-final-run-eval.

::: {#exr-gen-params-final-run-train}

## Training Phase

```{julia}
#| output: asis
dict = CTExperiments.from_toml(joinpath(res_dir,"gen_params/mlp/lin_sep/grid_config.toml")) 
println(CTExperiments.dict_to_quarto_markdown(dict))
```

:::

::: {#exr-gen-params-final-run-eval}

## Evaluation Phase

```{julia}
#| output: asis
dict = CTExperiments.from_toml(joinpath(res_dir, "gen_params/mlp/lin_sep/evaluation/evaluation_grid_config.toml"))
println(CTExperiments.dict_to_quarto_markdown(dict))
```

:::

### High-Level Findings

![](/paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png)

![](/paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png)

![](/paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/percent_valid.png)

![](/paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/acc_val.png)

### Detailed Findings

```{julia}
gen_params_dir = joinpath(res_dir, "gen_params/mlp")
```

#### Linearly Separable

```{julia}
df = CSV.read("$(gen_params_dir)/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.csv", DataFrame)
select!(df, Not(:objective))
df = groupby(df, Not(:run, :mean, :lambda_energy_eval, :is_pct)) |> 
  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
  df -> sort(df, [:lambda_energy_exper, :decision_threshold_exper, :maxiter_exper, :generator_type])
```

::: {#tbl-lin_sep-lambda_energy_exper}

::: {.content-hidden unless-format="pdf"}

```{julia}
#| output: asis

using CTExperiments.ColorSchemes
get_table_inputs(df, "value"; alpha=0.9, byvars=["lambda_energy_exper", "decision_threshold_exper", "maxiter_exper"], backend=Val(:latex)) |>
    inputs -> tabulate_results(inputs; wrap_table=false)
```

:::

Results for Linearly Separable data by energy penalty.

:::

### 2025-03-06 paper/sections/other/initial_grids.qmd
more work on appendix

```{julia}
projectdir = splitpath(pwd()) |>
    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
cd(projectdir)
```

```{julia}
using CTExperiments
using CTExperiments.CSV
using CTExperiments.DataFrames
using CTExperiments.StatsBase

using DotEnv
DotEnv.load!()
```

## Initial Grid Search {#sec-app-initial}

```{julia}
res_dir = ENV["INITIAL_RUN_RESULTS"]
```

For the initial round of experiments we employed a different training objective and procedure that led to promising results for some hyperparameter choices, but suffered from training instabilities.

### Generator Parameters

The hyperparameter grids for the first investigation of the effect of generator parameters are shown in @exr-gen-params-first-run-train and @exr-gen-params-first-run-eval.

::: {#exr-gen-params-first-run-train}

## Training Phase

```{julia}
#| output: asis
dict = CTExperiments.from_toml(joinpath(res_dir,"gen_params/mlp/lin_sep/grid_config.toml")) 
println(CTExperiments.dict_to_quarto_markdown(dict))
```

:::

::: {#exr-gen-params-first-run-eval}

## Evaluation Phase

```{julia}
#| output: asis
dict = CTExperiments.from_toml(joinpath(res_dir, "gen_params/mlp/lin_sep/evaluation/evaluation_grid_config.toml"))
println(CTExperiments.dict_to_quarto_markdown(dict))
```

:::

```{julia}
gen_params_dir = joinpath(res_dir, "gen_params/mlp")
```

#### Linearly Separable

- **Energy Penalty** (@tbl-lin_sep-lambda_energy_exper): *ECCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training. *Generic* performs poorly accross the board. *Omni* seems to have an anchoring effect, in that it never performs terribly but also never as good as the best *ECCo* results. *REVISE* performs poorly across the board.
- **Cost** (@tbl-lin_sep-lambda_cost_exper): Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged). For *ECCo* and *Generic*, higher cost values generally lead to worse results.
- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
- **Validity**: *ECCo* almost always valid except for very low values during training and high values at evaluation time. *Generic* often has poor validity.
- **Accuracy**: Seems largely unaffected.

```{julia}
df = CSV.read("$(gen_params_dir)/lin_sep/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type])
```

::: {#tbl-lin_sep-lambda_energy_exper}

::: {.content-hidden unless-format="pdf"}

```{julia}
#| output: asis
get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
    inputs -> tabulate_results(inputs; wrap_table=false)
```

:::

Results for Linearly Separable data by energy penalty.

:::

<!-- Cost -->

```{julia}
df = CSV.read("$(gen_params_dir)/lin_sep/evaluation/results/ce/objective---lambda_cost_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
  df -> sort(df, [:lambda_cost_exper, :objective, :generator_type])
```

::: {#tbl-lin_sep-lambda_cost_exper}

::: {.content-hidden unless-format="pdf"}

```{julia}
#| output: asis
get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_cost_exper", backend=Val(:latex)) |>
    inputs -> tabulate_results(inputs; wrap_table=false)
```

:::

Results for Linearly Separable data by cost penalty.

:::

#### Moons

- **Energy Penalty** (@tbl-moons-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, except for very low choices of the energy penalty during training for which it performs abismal. *Generic* performs quite badly across the board for high enough choices of the energy penalty at evaluation time. *Omni* has small positive effect. *REVISE* performs poorly across the board.
- **Cost (distance penalty)**: *Generic* generally does better for higher values, while *ECCo* does better for lower values.
- **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
- **Validity**: *ECCo* generally achieves full validity except for very low choices the energy penalty during training and high choices at evaluation time. *Generic* performs poorly for high choices of the energy penalty during evaluation.
- **Accuracy**: Largely unaffected although *ECCo* suffers a bit for very low choices the energy penalty during training. *REVISE* suffers a lot in general (around 10 percentage points).

```{julia}
df = CSV.read("$(gen_params_dir)/moons/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) 
```

::: {#tbl-moons-lambda_energy_exper}

::: {.content-hidden unless-format="pdf"}

```{julia}
#| output: asis
get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
    inputs -> tabulate_results(inputs)
```

:::

Results for Moons data by energy penalty.

:::

#### Circles

- **Energy Penalty** (@tbl-circles-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, though primarily for low to medium choices of the energy penalty (<=5) during training. The same goes for *Generic*, which sometimes outperforms *ECCo* (for small energy penalty at evaluation time). *Omni* does alright for lower energy penalty at evaluation time, but loses out for higher choices. *REVISE* performs poorly across the board (except very low choices at evaluation time).
- **Cost (distance penalty)**: *ECCo* and *Generic* generally achieve the best results when no cost penalty is used during training. Both *Omni* and *REVISE* are largely unaffected.
- **Maximum Iterations**: *ECCo* consistently yields better results for higher numbers of iterations. *Generic* generally does best for a medium number (50). *Omni* is sometimes invalid (**???**).
- **Validity**: *ECCo* tends to outperform its *Vanilla* counterpart, though primarily for low to medium choices of the energy penalty (<=5) during training and evaluation. *Vanilla* typically worse across the board.
- **Accuracy**: Mostly unaffected, but *REVISE* again consistently some deterioration and *ECCo* deteriorates for high choices of energy penalty during training, reflecting other outcomes above.

```{julia}
df = CSV.read("$(gen_params_dir)/circles/evaluation/results/ce/objective---lambda_energy_exper---lambda_energy_eval/plausibility_distance_from_target.csv", DataFrame)
df = groupby(df, Not(:run, :std, :mean, :lambda_energy_eval)) |> 
  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
  df -> sort(df, [:lambda_energy_exper, :objective, :generator_type]) 
```

::: {#tbl-circles-lambda_energy_exper}

::: {.content-hidden unless-format="pdf"}

```{julia}
#| output: asis
get_table_inputs(df, "value"; alpha=0.9, byvars="lambda_energy_exper", backend=Val(:latex)) |>
    inputs -> tabulate_results(inputs)
```

:::

Results for Circles data by energy penalty.

:::

### 2025-03-06 paper/experiments/CTExperiments/src/tables.jl

@@ -43,7 +43,7 @@ function get_table_inputs(
     if "generator_type" in names(df)
         # Filter out "omni":
         df = df[df.generator_type .!= "omni", :]
-        df.generator_type = format_generator.(df.generator_type)
+        # df.generator_type = format_generator.(df.generator_type)
         gen_hl = generator_highlighter(df; backend=backend)
         hls = (hls..., gen_hl)
     end
@@ -78,13 +78,15 @@ function value_highlighter(
     if !isnothing(byvars)
         byvars = isa(byvars, String) ? [byvars] : byvars
         df.row .= 1:nrow(df)
-        max_idx = combine(groupby(df, byvars...)) do sdf
+        max_idx = combine(groupby(df, byvars)) do sdf
             (max_idx=sdf.row[argmax(sdf[:, value_var])],)
         end
         max_idx = max_idx.max_idx
         select!(df, Not(:row))
         hl = bolden_max_hl(max_idx, col_idx, backend)
         push!(hls, hl)
+        hl_bad = bolden_max_hl_bad(max_idx, col_idx, backend)
+        push!(hls, hl_bad)
     end
 
     # Color scale:
@@ -136,7 +138,11 @@ function color_scale_hl(
 end
 
 function bolden_max_hl(max_idx::Vector{Int}, col_idx::Vector{Int}, backend::Val{:latex})
-    return hl = LatexHighlighter((df, i, j) -> i in max_idx, ["color{blue}", "textbf"])
+    return hl = LatexHighlighter((df, i, j) -> (i in max_idx) && df[i,:value]>0, ["color{Green}", "textbf"])
+end
+
+function bolden_max_hl_bad(max_idx::Vector{Int}, col_idx::Vector{Int}, backend::Val{:latex})
+    return hl = LatexHighlighter((df, i, j) -> (i in max_idx) && df[i,:value]<0, ["color{Red}", "textbf"])
 end
 
 function generator_highlighter(df::DataFrame; backend::Val=Val(:text))

### 2025-03-06 paper/experiments/present_results.jl

@@ -57,12 +57,10 @@ end
 # Visualize CE:
 prefix = "ce"
 cfg = eval_grid
-all_data = CTExperiments.merge_with_meta(
-    cfg, CTExperiments.load_ce_evaluation(cfg)
-)
+all_data = CTExperiments.merge_with_meta(cfg, CTExperiments.load_ce_evaluation(cfg))
 valid_y = CTExperiments.valid_y_ce(all_data[1])
 params = PlotParams(;
-    colorvar=get_global_param("colorvar_ce", colorvar_ce) |> x -> adjust_plot_var(x,cfg),
+    colorvar=get_global_param("colorvar_ce", colorvar_ce) |> x -> adjust_plot_var(x, cfg),
     rowvar=get_global_param("rowvar_ce", rowvar_ce) |> x -> adjust_plot_var(x, cfg),
     colvar=get_global_param("colvar_ce", colvar_ce) |> x -> adjust_plot_var(x, cfg),
     sidevar=get_global_param("sidevar", sidevar) |> x -> adjust_plot_var(x, cfg),

### 2025-03-06 paper/preprint/paper.tex
ok final final runs?

@@
 \usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{plain}
@@
 \theoremstyle{definition}
-\newtheorem{example}{Example}[section]
+\newtheorem{exercise}{Parameters}[section]
 \theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
+\newtheorem{definition}{Definition}[section]
 \theoremstyle{plain}
 \newtheorem{corollary}{Research Question}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
 \theoremstyle{remark}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, titlerule=0mm, coltitle=black, left=2mm, bottomtitle=1mm, breakable, opacityback=0, rightrule=.15mm, toptitle=1mm, arc=.35mm, bottomrule=.15mm, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, opacitybacktitle=0.6, arc=.35mm, toptitle=1mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, colback=white, breakable, toprule=.15mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, titlerule=0mm, coltitle=black]
 
@@
   0.75 & 0.1 & 5 & \textit{ECCo} & -15.9 & 16.7 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{41.9}} & \color{green}{\textbf{8.13}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{41.9}} & \color{Green}{\textbf{8.13}} \\
   0.75 & 0.1 & 5 & \textit{REVISE} & -64.4 & 3.58 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{28.9}} & \color{green}{\textbf{2.28}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{28.9}} & \color{Green}{\textbf{2.28}} \\
   0.75 & 0.1 & 25 & \textit{Generic} & -11.3 & 11.5 \\
   0.75 & 0.1 & 25 & \textit{REVISE} & -55.3 & 4.18 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{10.9}} & \color{green}{\textbf{7.47}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{10.9}} & \color{Green}{\textbf{7.47}} \\
   0.75 & 0.1 & 50 & \textit{Generic} & -32.3 & 16.6 \\
@@
   0.9 & 0.1 & 5 & \textit{ECCo} & -13.7 & 13.7 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{22.3}} & \color{green}{\textbf{4.87}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{22.3}} & \color{Green}{\textbf{4.87}} \\
   0.9 & 0.1 & 5 & \textit{REVISE} & -27.8 & 6.26 \\
   0.9 & 0.1 & 25 & \textit{ECCo} & 8.84 & 7.13 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{30.7}} & \color{green}{\textbf{5.77}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{30.7}} & \color{Green}{\textbf{5.77}} \\
   0.9 & 0.1 & 25 & \textit{REVISE} & -72.7 & 34.9 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{17.1}} & \color{green}{\textbf{7.98}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{17.1}} & \color{Green}{\textbf{7.98}} \\
   0.9 & 0.1 & 50 & \textit{Generic} & 3.82 & 8.86 \\
   0.9 & 0.1 & 50 & \textit{REVISE} & -45.8 & 2.79 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{35.7}} & \color{green}{\textbf{7.56}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{35.7}} & \color{Green}{\textbf{7.56}} \\
   0.95 & 0.1 & 5 & \textit{Generic} & -13.3 & 5.73 \\
   0.95 & 0.1 & 5 & \textit{REVISE} & -54.9 & 5.43 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{37}} & \color{green}{\textbf{5.31}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{37}} & \color{Green}{\textbf{5.31}} \\
   0.95 & 0.1 & 25 & \textit{Generic} & 29 & 8.6 \\
@@
   0.95 & 0.1 & 50 & \textit{ECCo} & 13.2 & 4.89 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.1}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{35.5}} & \color{green}{\textbf{4.27}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{35.5}} & \color{Green}{\textbf{4.27}} \\
   0.95 & 0.1 & 50 & \textit{REVISE} & -60.4 & 45.8 \\
-  \color{red}{\textbf{0.75}} & \color{red}{\textbf{0.5}} & \color{red}{\textbf{5}} & \color{red}{\textbf{ECCo}} & \color{red}{\textbf{-0.94}} & \color{red}{\textbf{17.6}} \\
+  \color{Red}{\textbf{0.75}} & \color{Red}{\textbf{0.5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{ECCo}} & \color{Red}{\textbf{-0.94}} & \color{Red}{\textbf{17.6}} \\
   0.75 & 0.5 & 5 & \textit{Generic} & -11.7 & 17.4 \\
@@
   0.75 & 0.5 & 25 & \textit{ECCo} & 14.7 & 6.32 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{32.3}} & \color{green}{\textbf{3.49}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{32.3}} & \color{Green}{\textbf{3.49}} \\
   0.75 & 0.5 & 25 & \textit{REVISE} & -39.9 & 7.21 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{28.3}} & \color{green}{\textbf{5.05}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{28.3}} & \color{Green}{\textbf{5.05}} \\
   0.75 & 0.5 & 50 & \textit{Generic} & -9.72 & 9.19 \\
@@
   0.9 & 0.5 & 5 & \textit{ECCo} & -7.38 & 13.6 \\
-  \color{red}{\textbf{0.9}} & \color{red}{\textbf{0.5}} & \color{red}{\textbf{5}} & \color{red}{\textbf{Generic}} & \color{red}{\textbf{-0.0952}} & \color{red}{\textbf{17.1}} \\
+  \color{Red}{\textbf{0.9}} & \color{Red}{\textbf{0.5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{Generic}} & \color{Red}{\textbf{-0.0952}} & \color{Red}{\textbf{17.1}} \\
   0.9 & 0.5 & 5 & \textit{REVISE} & -50.6 & 4.77 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{39.1}} & \color{green}{\textbf{8.3}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{39.1}} & \color{Green}{\textbf{8.3}} \\
   0.9 & 0.5 & 25 & \textit{Generic} & 38.6 & 6.03 \\
@@
   0.9 & 0.5 & 50 & \textit{ECCo} & -6.53 & 10.9 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{13.6}} & \color{green}{\textbf{12.8}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{13.6}} & \color{Green}{\textbf{12.8}} \\
   0.9 & 0.5 & 50 & \textit{REVISE} & -24.1 & 12.5 \\
   0.95 & 0.5 & 5 & \textit{ECCo} & 0.596 & 1.11 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{4.57}} & \color{green}{\textbf{5.88}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{4.57}} & \color{Green}{\textbf{5.88}} \\
   0.95 & 0.5 & 5 & \textit{REVISE} & -34.3 & 6.09 \\
   0.95 & 0.5 & 25 & \textit{ECCo} & 16.5 & 3.53 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{19.1}} & \color{green}{\textbf{4.43}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{19.1}} & \color{Green}{\textbf{4.43}} \\
   0.95 & 0.5 & 25 & \textit{REVISE} & -37.2 & 39.2 \\
   0.95 & 0.5 & 50 & \textit{ECCo} & -31 & 14.2 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{0.5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{24.8}} & \color{green}{\textbf{4.78}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{24.8}} & \color{Green}{\textbf{4.78}} \\
   0.95 & 0.5 & 50 & \textit{REVISE} & -1.28 & 2.7 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{5}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{37}} & \color{green}{\textbf{9.21}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{37}} & \color{Green}{\textbf{9.21}} \\
   0.75 & 5 & 5 & \textit{Generic} & 0.977 & 21.1 \\
   0.75 & 5 & 5 & \textit{REVISE} & -55.2 & 4.75 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{21.8}} & \color{green}{\textbf{5.97}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{21.8}} & \color{Green}{\textbf{5.97}} \\
   0.75 & 5 & 25 & \textit{Generic} & 17.8 & 5.33 \\
@@
   0.75 & 5 & 50 & \textit{ECCo} & 0.864 & 9.51 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{8.56}} & \color{green}{\textbf{13.6}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{8.56}} & \color{Green}{\textbf{13.6}} \\
   0.75 & 5 & 50 & \textit{REVISE} & -71.8 & 3.18 \\
   0.9 & 5 & 5 & \textit{ECCo} & -3.59 & 20.3 \\
-  \color{red}{\textbf{0.9}} & \color{red}{\textbf{5}} & \color{red}{\textbf{5}} & \color{red}{\textbf{Generic}} & \color{red}{\textbf{-1.18}} & \color{red}{\textbf{16.7}} \\
+  \color{Red}{\textbf{0.9}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{Generic}} & \color{Red}{\textbf{-1.18}} & \color{Red}{\textbf{16.7}} \\
   0.9 & 5 & 5 & \textit{REVISE} & -58.6 & 3.55 \\
   0.9 & 5 & 25 & \textit{ECCo} & -1.73 & 9.07 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{21.9}} & \color{green}{\textbf{7.48}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{21.9}} & \color{Green}{\textbf{7.48}} \\
   0.9 & 5 & 25 & \textit{REVISE} & -44.4 & 4.23 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{30.8}} & \color{green}{\textbf{10.7}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{30.8}} & \color{Green}{\textbf{10.7}} \\
   0.9 & 5 & 50 & \textit{Generic} & 16.3 & 11.1 \\
   0.9 & 5 & 50 & \textit{REVISE} & -53.3 & 6.73 \\
-  \color{red}{\textbf{0.95}} & \color{red}{\textbf{5}} & \color{red}{\textbf{5}} & \color{red}{\textbf{ECCo}} & \color{red}{\textbf{-5.29}} & \color{red}{\textbf{24.5}} \\
+  \color{Red}{\textbf{0.95}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{ECCo}} & \color{Red}{\textbf{-5.29}} & \color{Red}{\textbf{24.5}} \\
   0.95 & 5 & 5 & \textit{Generic} & -5.71 & 17.4 \\
   0.95 & 5 & 5 & \textit{REVISE} & -36.5 & 4.01 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{5}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{26.9}} & \color{green}{\textbf{4.08}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{26.9}} & \color{Green}{\textbf{4.08}} \\
   0.95 & 5 & 25 & \textit{Generic} & 19.5 & 5.49 \\
@@
   0.95 & 5 & 50 & \textit{ECCo} & 4.6 & 7.46 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{5}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{16.6}} & \color{green}{\textbf{4.3}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{16.6}} & \color{Green}{\textbf{4.3}} \\
   0.95 & 5 & 50 & \textit{REVISE} & -50.4 & 46.4 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{10}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{2.47}} & \color{green}{\textbf{8.57}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{2.47}} & \color{Green}{\textbf{8.57}} \\
   0.75 & 10 & 5 & \textit{Generic} & -7.23 & 9.13 \\
   0.75 & 10 & 5 & \textit{REVISE} & -67.2 & 5.75 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{10}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{29.9}} & \color{green}{\textbf{15.8}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{29.9}} & \color{Green}{\textbf{15.8}} \\
   0.75 & 10 & 25 & \textit{Generic} & 4.47 & 8.22 \\
@@
   0.75 & 10 & 50 & \textit{ECCo} & 6.23 & 6.11 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{10}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{15.2}} & \color{green}{\textbf{10.2}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{15.2}} & \color{Green}{\textbf{10.2}} \\
   0.75 & 10 & 50 & \textit{REVISE} & -17 & 5.46 \\
   0.9 & 10 & 5 & \textit{ECCo} & -6.93 & 14.7 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{10}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{6.88}} & \color{green}{\textbf{5.11}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{6.88}} & \color{Green}{\textbf{5.11}} \\
   0.9 & 10 & 5 & \textit{REVISE} & -45.6 & 47.8 \\
   0.9 & 10 & 25 & \textit{ECCo} & 6.2 & 6.75 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{10}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{34.7}} & \color{green}{\textbf{9.74}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{34.7}} & \color{Green}{\textbf{9.74}} \\
   0.9 & 10 & 25 & \textit{REVISE} & -55.7 & 6.04 \\
   0.9 & 10 & 50 & \textit{ECCo} & 30.1 & 14.2 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{10}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{35}} & \color{green}{\textbf{7.11}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{35}} & \color{Green}{\textbf{7.11}} \\
   0.9 & 10 & 50 & \textit{REVISE} & -48.8 & 3.3 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{10}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{35.7}} & \color{green}{\textbf{3.83}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{35.7}} & \color{Green}{\textbf{3.83}} \\
   0.95 & 10 & 5 & \textit{Generic} & 2.85 & 7.9 \\
   0.95 & 10 & 5 & \textit{REVISE} & -34.9 & 4.56 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{10}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{33}} & \color{green}{\textbf{6.45}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{33}} & \color{Green}{\textbf{6.45}} \\
   0.95 & 10 & 25 & \textit{Generic} & 24.8 & 4.59 \\
   0.95 & 10 & 25 & \textit{REVISE} & -55.7 & 44 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{10}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{8.07}} & \color{green}{\textbf{7.68}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{8.07}} & \color{Green}{\textbf{7.68}} \\
   0.95 & 10 & 50 & \textit{Generic} & -7.11 & 9.2 \\
@@
   0.75 & 20 & 5 & \textit{ECCo} & -10.8 & 20.8 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{20}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{12.6}} & \color{green}{\textbf{17.6}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{12.6}} & \color{Green}{\textbf{17.6}} \\
   0.75 & 20 & 5 & \textit{REVISE} & -21.6 & 3.34 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{20}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{17.2}} & \color{green}{\textbf{4.25}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{17.2}} & \color{Green}{\textbf{4.25}} \\
   0.75 & 20 & 25 & \textit{Generic} & -10.3 & 9.59 \\
   0.75 & 20 & 25 & \textit{REVISE} & -59 & 3.66 \\
-  \color{green}{\textbf{0.75}} & \color{green}{\textbf{20}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{15.6}} & \color{green}{\textbf{8.38}} \\
+  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{15.6}} & \color{Green}{\textbf{8.38}} \\
   0.75 & 20 & 50 & \textit{Generic} & 3.92 & 6.2 \\
   0.75 & 20 & 50 & \textit{REVISE} & -70.9 & 5.16 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{20}} & \color{green}{\textbf{5}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{26.3}} & \color{green}{\textbf{6.62}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{26.3}} & \color{Green}{\textbf{6.62}} \\
   0.9 & 20 & 5 & \textit{Generic} & 2.35 & 7.04 \\
@@
   0.9 & 20 & 25 & \textit{ECCo} & 21.1 & 11.6 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{20}} & \color{green}{\textbf{25}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{39}} & \color{green}{\textbf{5.98}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{39}} & \color{Green}{\textbf{5.98}} \\
   0.9 & 20 & 25 & \textit{REVISE} & -50.7 & 13.8 \\
-  \color{green}{\textbf{0.9}} & \color{green}{\textbf{20}} & \color{green}{\textbf{50}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{15.1}} & \color{green}{\textbf{7.46}} \\
+  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{15.1}} & \color{Green}{\textbf{7.46}} \\
   0.9 & 20 & 50 & \textit{Generic} & 8.53 & 6.95 \\
@@
   0.95 & 20 & 5 & \textit{ECCo} & -23.8 & 17.8 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{20}} & \color{green}{\textbf{5}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{7.23}} & \color{green}{\textbf{4.74}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{7.23}} & \color{Green}{\textbf{4.74}} \\
   0.95 & 20 & 5 & \textit{REVISE} & -34 & 4.11 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{20}} & \color{green}{\textbf{25}} & \color{green}{\textbf{ECCo}} & \color{green}{\textbf{19.3}} & \color{green}{\textbf{14.6}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{19.3}} & \color{Green}{\textbf{14.6}} \\
   0.95 & 20 & 25 & \textit{Generic} & -1.35 & 7.71 \\
@@
   0.95 & 20 & 50 & \textit{ECCo} & 18.7 & 3.84 \\
-  \color{green}{\textbf{0.95}} & \color{green}{\textbf{20}} & \color{green}{\textbf{50}} & \color{green}{\textbf{Generic}} & \color{green}{\textbf{37.3}} & \color{green}{\textbf{7.78}} \\
+  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{37.3}} & \color{Green}{\textbf{7.78}} \\
   0.95 & 20 & 50 & \textit{REVISE} & -28.3 & 15 \\\bottomrule

### 2025-03-06 paper/sections/abstract.qmd
ok final final runs?

--- 2025-02-03 paper/sections/abstract.qmd
@@
 
-Counterfactual Explanations have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output.
-Much existing research has focused on identifying explanations that are not only valid but also deemed desirable with respect to the underlying data and stakeholder requirements.
-Recent work has shown that under this premise, the task of learning desirable explanations is effectively reassigned from the model itself to the (post-hoc) counterfactual explainer.
-Building on that work, we propose a novel model objective that leverages counterfactuals during the training phase (ad-hoc) in order to minimize the divergence between learned representations and desirable explanations.
-Through extensive experiments, we demonstrate that our proposed methodology facilitates training models that inherently deliver desirable explanations while maintaining high predictive performance.
+We propose a novel model objective that leverages counterfactual explanations to increase the explanatory capacity of trained models.
+Counterfactual explanations have emerged as a popular tool to explain predictions made by opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output.
+Much existing research has focused on generating counterfactuals that are not only valid but also deemed acceptable with respect to the stakeholder requirements and plausible with respect to the underlying data.
+Recent work has shown that under this premise, the task of learning plausible explanations---ones that respect the conditional distribution of the target class---is effectively reassigned from the model itself to the post-hoc counterfactual explainer.
+Building on that work, we propose a novel objective function that employs counterfactuals (ad-hoc) during the training phase to minimize the divergence between the learned representations and the explanations.
+Through extensive experiments, we demonstrate that our proposed method facilitates training models that deliver inherently plausible explanations while maintaining high predictive performance.

### 2025-03-06 paper/experiments/CTExperiments/src/config.jl

@@ -146,7 +146,7 @@ function is_empty_value(v::Any)
     if v isa String
         return isempty(v)
     elseif v isa Vector
-        return isempty(v)
+        return isempty(v) || length(v) == 1
     elseif v isa Dict
         # A dictionary is empty if it's empty itself or if all its filtered values would be empty
         filtered = filter_dict(v)
@@ -160,6 +160,11 @@ end
 function filter_dict(dict::Dict; drop_fields=["name", "data", "data_params"])
     # Filter out empty values and specified fields
     return filter(dict) do (k, v)
+        if v isa Dict
+            v = filter_dict(v; drop_fields)
+        end
+        @info "$k"
+        println(v)
         !is_empty_value(v) && !(k in drop_fields)
     end
 end

### 2025-03-06 paper/preprint/paper.tex
tables sorted

@@
 \theoremstyle{definition}
-\newtheorem{example}{Example}[section]
+\newtheorem{definition}{Definition}[section]
 \theoremstyle{plain}
 \newtheorem{proposition}{Proposition}[section]
-\theoremstyle{definition}
-\newtheorem{exercise}{Parameters}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
 \theoremstyle{plain}
 \newtheorem{corollary}{Research Question}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{remark}
@@
 \makeatother
+\newcounter{quartocalloutnteno}
+\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
 
@@
 \begin{abstract}
-Counterfactual Explanations have emerged as a popular tool to explain
-predictions made by opaque machine learning models: they explain how
-factual inputs need to change in order for some fitted model to produce
-some desired output. Much existing research has focused on identifying
-explanations that are not only valid but also deemed plausible and
-desirable with respect to the underlying data and stakeholder
-requirements. Recent work has shown that under this premise, the task of
-learning plausible explanations is effectively reassigned from the model
-itself to the (post-hoc) counterfactual explainer. Building on that
-work, we propose a novel model objective that leverages counterfactuals
-during the training phase (ad-hoc) in order to minimize the divergence
-between learned representations and plausible explanations. Through
-extensive experiments, we demonstrate that our proposed methodology
-facilitates training models that inherently deliver plausible
+We propose a novel model objective that leverages counterfactual
+explanations to increase the explanatory capacity of trained models.
+Counterfactual explanations have emerged as a popular tool to explain
+predictions made by opaque machine learning models: they inform how
+factual inputs would need to change in order for a model to produce some
+desired output. Much existing research has focused on generating
+counterfactuals that are not only valid but also deemed acceptable with
+respect to the stakeholder requirements and plausible with respect to
+the underlying data. Recent work has shown that under this premise, the
+task of learning plausible explanations---ones that respect the
+conditional distribution of the target class---is effectively reassigned
+from the model itself to the post-hoc counterfactual explainer. Building
+on that work, we propose a novel objective function that employs
+counterfactuals (ad-hoc) during the training phase to minimize the
+divergence between the learned representations and the explanations.
+Through extensive experiments, we demonstrate that our proposed method
+facilitates training models that deliver inherently plausible
 explanations while maintaining high predictive performance.
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, opacitybacktitle=0.6, arc=.35mm, toptitle=1mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame, colback=white, breakable, toprule=.15mm, title=\textcolor{quarto-callout-note-color}{\faInfo}\hspace{0.5em}{Note}, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, titlerule=0mm, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]
 
@@
 The hyperparameter grid with varying generator parameters during
-training is shown in Parameters~\ref{exr-gen-params-final-run-train}.
-The corresponding evaluation grid used for these experiments is shown in
-Parameters~\ref{exr-gen-params-final-run-eval}.
-
-\begin{exercise}[Training
-Phase]\protect\hypertarget{exr-gen-params-final-run-train}{}\label{exr-gen-params-final-run-train}
-
-~
+training is shown in Note~\ref{nte-gen-params-final-run-train}. The
+corresponding evaluation grid used for these experiments is shown in
+Note~\ref{nte-gen-params-final-run-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]
+
+\quartocalloutnte{nte-gen-params-final-run-train}
 
@@
   \item
-    \(\lambda_{\text{cost}}\): \texttt{0.001}
-  \item
     \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 5.0,\ 10.0,\ 20.0}
   \item
-    Learning Rate: \texttt{0.25}
-  \item
     Maximum Iterations: \texttt{5,\ 25,\ 50}
-  \item
-    Optimizer: \texttt{sgd}
   \end{itemize}
@@
 \item
-  Model Parameters:
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
 
@@
   \item
-    Activation: \texttt{relu}
-  \item
-    No.~Hidden: \texttt{32}
-  \item
-    No.~Layers: \texttt{1}
+    Objective: \texttt{full,\ vanilla}
   \end{itemize}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]
+
+\quartocalloutnte{nte-gen-params-final-run-eval}
 
@@
 \item
-    Burnin: \texttt{0.0}
-  \item
-    Class Loss: \texttt{logitcrossentropy}
-  \item
-    Convergence: \texttt{threshold}
-  \item
-    \(\lambda_{\text{adv}}\): \texttt{0.25}
-  \item
-    \(\lambda_{\text{yloss}}\): \texttt{1.0}
-  \item
-    \(\lambda_{\text{div}}\): \texttt{0.5}
-  \item
-    \(\lambda_{\text{reg}}\): \texttt{0.01}
-  \item
-    Learning Rate: \texttt{0.001}
-  \item
-    No.~Counterfactuals: \texttt{1000}
-  \item
-    No.~Epochs: \texttt{100}
-  \item
-    No.~Neighbours: \texttt{100}
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \item
-    Optimizer: \texttt{adam}
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
   \end{itemize}
@@
 
-\end{exercise}
-
-\begin{exercise}[Evaluation
-Phase]\protect\hypertarget{exr-gen-params-final-run-eval}{}\label{exr-gen-params-final-run-eval}
-
-~
-
-\begin{itemize}
-\tightlist
-\item
-  Counterfactual Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Convergence: \texttt{threshold}
-  \item
-    Decision Threshold: \texttt{0.95}
-  \item
-    Maximum Iterations: \texttt{50}
-  \item
-    No.~Individuals: \texttt{100}
-  \item
-    No.~Runs: \texttt{5}
-  \end{itemize}
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Decision Threshold: \texttt{0.75}
-  \item
-    \(\lambda_{\text{cost}}\): \texttt{0.0}
-  \item
-    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \item
-    Learning Rate: \texttt{0.25}
-  \item
-    Maximum Iterations: \texttt{30}
-  \item
-    Optimizer: \texttt{sgd}
-  \end{itemize}
-\end{itemize}
-
-\end{exercise}
+\end{tcolorbox}
 
@@
   \toprule
-  \textbf{Decision Threshold} & \textbf{Lambda Energy} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \textbf{Decision Threshold} & \textbf{$\lambda_{\text{energy}}$} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
   \endfirsthead
   \toprule
-  \textbf{Decision Threshold} & \textbf{Lambda Energy} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
+  \textbf{Decision Threshold} & \textbf{$\lambda_{\text{energy}}$} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
   \endhead

### 2025-03-06 paper/sections/other/grids.qmd
tables sorted

@@
 
-The hyperparameter grid with varying generator parameters during training is shown in @exr-gen-params-final-run-train.
-The corresponding evaluation grid used for these experiments is shown in @exr-gen-params-final-run-eval.
+The hyperparameter grid with varying generator parameters during training is shown in @nte-gen-params-final-run-train.
+The corresponding evaluation grid used for these experiments is shown in @nte-gen-params-final-run-eval.
 
-::: {#exr-gen-params-final-run-train}
+::: {#nte-gen-params-final-run-train .callout-note}
 
@@
 
-::: {#exr-gen-params-final-run-eval}
+::: {#nte-gen-params-final-run-eval .callout-note}
 

### 2025-03-06 paper/experiments/CTExperiments/src/config.jl

@@ -158,15 +158,20 @@ end
 
 # Function to filter dictionary
 function filter_dict(dict::Dict; drop_fields=["name", "data", "data_params"])
-    # Filter out empty values and specified fields
-    return filter(dict) do (k, v)
+
+    # Take care of nested dicts:
+    for (k, v) in dict
         if v isa Dict
-            v = filter_dict(v; drop_fields)
+            dict[k] = filter_dict(v; drop_fields)
         end
-        @info "$k"
-        println(v)
+    end
+
+    # Filter out empty values and specified fields
+    dict = filter(dict) do (k, v)
         !is_empty_value(v) && !(k in drop_fields)
     end
+
+    return dict
 end
 
 global LatexReplacements = Dict(

### 2025-03-06 paper/experiments/CTExperiments/src/tables.jl

@@ -54,11 +54,14 @@ end
 
 format_generator(s::AbstractString) = get_generator_name(generator_types[s](); pretty=true)
 
-global LatexHeaderReplacements = Dict(
-    "lambda_energy_exper" => LatexCell("\$\\lambda_{\\text{div}} (\\text{train})\$"),
-    "lambda_energy_eval" => LatexCell("\$\\lambda_{\\text{div}} (\\text{eval})\$"),
-    "lambda_cost_exper" => LatexCell("\$\\lambda_{\\text{cost}} (\\text{train})\$"),
-    "lambda_cost_eval" => LatexCell("\$\\lambda_{\\text{cost}} (\\text{eval})\$"),
+global LatexHeaderReplacements = merge(
+    Dict(
+        "lambda_energy_exper" => LatexCell("\$\\lambda_{\\text{div}} (\\text{train})\$"),
+        "lambda_energy_eval" => LatexCell("\$\\lambda_{\\text{div}} (\\text{eval})\$"),
+        "lambda_cost_exper" => LatexCell("\$\\lambda_{\\text{cost}} (\\text{train})\$"),
+        "lambda_cost_eval" => LatexCell("\$\\lambda_{\\text{cost}} (\\text{eval})\$"),
+    ),
+    Dict(k => LatexCell(v) for (k, v) in CTExperiments.LatexReplacements),
 )
 
 function value_highlighter(

### 2025-03-06 paper/quarto_ecml/paper.tex
getting closer with quarto template

--- 2025-03-06 paper/preprint/paper.tex
@@
 %
-\documentclass[
-]{article}
+\documentclass[runningheads]{llncs}
 
@@
     % xetex/luatex font selection
-    \setmainfont[]{Latin Modern Roman}
-  \setmathfont[]{Latin Modern Math}
 \fi
@@
 \IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
+  \usepackage[activate=false]{microtype}
   \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
 \usepackage{xcolor}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
-% Make \paragraph and \subparagraph free-standing
-\makeatletter
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}{
-    \@ifstar
-      \xxxParagraphStar
-      \xxxParagraphNoStar
-  }
-  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
-  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}{
-    \@ifstar
-      \xxxSubParagraphStar
-      \xxxSubParagraphNoStar
-  }
-  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
-  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-\makeatother
+\setcounter{secnumdepth}{2}
 
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-\usepackage{arxiv}
-\usepackage{orcidlink}
-\usepackage{amsmath}
+
 \usepackage[T1]{fontenc}
+\usepackage{graphicx}
+\usepackage{booktabs}
+\usepackage[misc]{ifsym}
+\newcommand{\corr}{(\Letter)}
 \usepackage{placeins}
@@
 \ifdefined\figurename
-  \renewcommand*\figurename{Figure}
+  \renewcommand*\figurename{Fig.}
 \else
-  \newcommand\figurename{Figure}
+  \newcommand\figurename{Fig.}
 \fi
@@
 \newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{remark}
-\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
-\newtheorem*{remark}{Remark}
-\newtheorem*{solution}{Solution}
-\newtheorem{refremark}{Remark}[section]
-\newtheorem{refsolution}{Solution}[section]
+% \usepackage{amsthm}
+% \theoremstyle{plain}
+% \newtheorem{corollary}{Research Question}[section]
+% \theoremstyle{plain}
+% \newtheorem{proposition}{Proposition}[section]
+% \theoremstyle{definition}
+% \newtheorem{example}{Example}[section]
+% \theoremstyle{definition}
+% \newtheorem{definition}{Definition}[section]
+% \theoremstyle{remark}
+% \AtBeginDocument{\renewcommand*{\proofname}{Proof}}
+% \newtheorem*{remark}{Remark}
+% \newtheorem*{solution}{Solution}
+% \newtheorem{refremark}{Remark}[section]
+% \newtheorem{refsolution}{Solution}[section]
 \makeatother
@@
   colorlinks=true,
-  linkcolor={blue},
+  linkcolor={black},
   filecolor={Maroon},
@@
 
-\usepackage{lineno}
-\linenumbers
-\newcommand{\runninghead}{A Preprint }
-\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
+\hypersetup{nolinks=true}
+
 \title{Counterfactual Training: Teaching Models Plausible and Actionable
 Explanations}
-\def\asep{\\\\\\ } % default: all authors on same column
-\def\asep{\And }
-\author{\textbf{Patrick
-Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
-Engineering, Mathematics and Computer Science\\Delft University of
-Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
-Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\\asep\textbf{Arie van
-Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
-Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\}
-\date{}
+
+\titlerunning{Counterfactual Training}
+
+\author{Author information scrubbed for double-blind reviewing}
+
 \begin{document}
 \maketitle
+
 \begin{abstract}
@@
 explanations while maintaining high predictive performance.
+
+\keywords{Counterfactual Explanations \and Explainable
+AI \and Representation Learning}
+
 \end{abstract}
-{\bfseries \emph Keywords}
-\def\sep{\textbullet\ }
-Counterfactual Explanations \sep Explainable AI \sep 
-Representation Learning
 
@@
 representations from scratch, guided by narrow objectives such as
-predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
-Bengio, and Courville 2016}). Modern advances in computing have made it
-possible to provide such models with ever greater degrees of freedom to
-achieve that task, which has often led them to outperform traditionally
-more parsimonious models. Unfortunately, in doing so they also learn
-increasingly complex and highly sensitive representations that we can no
-longer easily interpret.
+predictive accuracy \cite{goodfellow2016deep}. Modern advances in
+computing have made it possible to provide such models with ever greater
+degrees of freedom to achieve that task, which has often led them to
+outperform traditionally more parsimonious models. Unfortunately, in
+doing so they also learn increasingly complex and highly sensitive
+representations that we can no longer easily interpret.
 
@@
 serious scrutiny in recent years. At the very cusp of the deep learning
-revolution, Szegedy et al. (\citeproc{ref-szegedy2013intriguing}{2013})
-showed that artificial neural networks (ANN) are sensitive to
-adversarial examples: counterfactuals of model inputs that yield vastly
-different model predictions despite being ``imperceptible'' in that they
-are semantically indifferent from their factual counterparts. Despite
-partially effective mitigation strategies such as \textbf{adversarial
-training} (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow,
-Shlens, and Szegedy 2014}), truly robust deep learning (DL) remains
-unattainable even for models that are considered shallow by today's
-standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).
+revolution, \cite{szegedy2013intriguing} showed that artificial neural
+networks (ANN) are sensitive to adversarial examples: counterfactuals of
+model inputs that yield vastly different model predictions despite being
+``imperceptible'' in that they are semantically indifferent from their
+factual counterparts. Despite partially effective mitigation strategies
+such as \textbf{adversarial training} \cite{goodfellow2014explaining},
+truly robust deep learning (DL) remains unattainable even for models
+that are considered shallow by today's standards
+\cite{kolter2023keynote}.
 
@@
 many solutions that are locally optimal with respect to narrow
-objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
-  clarity: we follow standard ML convention in using ``degrees of
-  freedom'' to refer to the number of parameters estimated from data.}.
-Based purely on predictive performance, these solutions may seem to
-provide compelling explanations for the data, when in fact they are
-based on purely associative, semantically meaningless patterns. This
-poses two related challenges: firstly, it makes these models inherently
-opaque, since humans cannot simply interpret what type of explanation
-the complex learned representations correspond to; secondly, even if we
-could resolve the first challenge, it is not obvious how to mitigate
-models from learning representations that correspond to meaningless and
-implausible explanations.
+objectives \cite{wilson2020case}\footnote{For clarity: we follow
+  standard ML convention in using ``degrees of freedom'' to refer to the
+  number of parameters estimated from data.}. Based purely on predictive
+performance, these solutions may seem to provide compelling explanations
+for the data, when in fact they are based on purely associative,
+semantically meaningless patterns. This poses two related challenges:
+firstly, it makes these models inherently opaque, since humans cannot
+simply interpret what type of explanation the complex learned
+representations correspond to; secondly, even if we could resolve the
+first challenge, it is not obvious how to mitigate models from learning
+representations that correspond to meaningless and implausible
+explanations.
 
@@
 and often unfairly by automated decision-making systems (ADMS) involving
-opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective
-XAI tools can aide us in monitoring models and providing recourse to
-individuals to turn adverse outcomes (e.g.~``loan application
-rejected'') into positive ones (``application accepted''). Wachter,
-Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}) propose
+opaque models \cite{oneil2016weapons}. Effective XAI tools can aide us
+in monitoring models and providing recourse to individuals to turn
+adverse outcomes (e.g.~``loan application rejected'') into positive ones
+(``application accepted''). \cite{wachter2017counterfactual} propose
 \textbf{counterfactual explanations} as an effective approach to achieve
@@
 model, independent of whether the learned representations are also
-compatible with implausible explanations
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+compatible with implausible explanations \cite{altmeyer2024faithful}.
 Fortunately, recent findings indicate that explainability can arise as
 byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization and generative capacity Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}).
+robustness, generalization and generative capacity
+\cite{schut2021generating}.
 
@@
 the underlying data and stakeholder requirements. To make this more
-concrete, we follow Augustin, Meinke, and Hein
-(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
-explainability to the quality of counterfactual explanations that we can
-generate for a given model. The authors show that counterfactual
+concrete, we follow \cite{augustin2020adversarial} in tieing the concept
+of explainability to the quality of counterfactual explanations that we
+can generate for a given model. The authors show that counterfactual
 explanations---understood here as minimal input perturbations that yield
@@
 counterfactuals on one side, and robustness to adversarial examples on
-the other, is backed by empirical evidence. Sauer and Geiger
-(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
-counterfactual images during classifier training improves model
-robustness. Similarly, Abbasnejad et al.
-(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
-counterfactuals represent potentially useful training data in machine
-learning, especially in supervised settings where inputs may be
-reasonably mapped to multiple outputs. They, too, demonstrate the
-augmenting the training data of image classifiers can improve
-generalization. Teney, Abbasnedjad, and Hengel
-(\citeproc{ref-teney2020learning}{2020}) propose an approach using
-counterfactuals in training that does not rely on data augmentation:
-they argue that counterfactual pairs typically already exist in training
-datasets. Specifically, their approach relies on, firstly, identifying
-similar input samples with different annotations and, secondly, ensuring
-that the gradient of the classifier aligns with the vector between pairs
-of counterfactual inputs using the cosine distance as a loss function.
+the other, is backed by empirical evidence.
+\cite{sauer2021counterfactual} demonstrate that using counterfactual
+images during classifier training improves model robustness. Similarly,
+\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
+potentially useful training data in machine learning, especially in
+supervised settings where inputs may be reasonably mapped to multiple
+outputs. They, too, demonstrate the augmenting the training data of
+image classifiers can improve generalization. \cite{teney2020learning}
+propose an approach using counterfactuals in training that does not rely
+on data augmentation: they argue that counterfactual pairs typically
+already exist in training datasets. Specifically, their approach relies
+on, firstly, identifying similar input samples with different
+annotations and, secondly, ensuring that the gradient of the classifier
+aligns with the vector between pairs of counterfactual inputs using the
+cosine distance as a loss function.
 
 In the natural language processing (NLP) domain, counterfactuals have
-similarly been used to improve models through data augmentation: Wu et
-al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{Polyjuice}, a
-general-purpose counterfactual generator for language models. They
-demonstrate empirically that augmenting training data through
-\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
-tasks. Balashankar et al.
-(\citeproc{ref-balashankar2023improving}{2023}) also use
-\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
-and show that classifier robustness improves up to 20\%. Finally, Luu
-and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
-Counterfactual Adversarial Training (CAT), which also aims at improving
-generalization and robustness of language models. Specifically, they
-propose to proceed as follows: firstly, they identify training samples
-that are subject to high predictive uncertainty; secondly, they generate
-counterfactual explanations for those samples; and, finally, they
-fine-tune the given language model on the augmented dataset that
-includes the generated counterfactuals.
+similarly been used to improve models through data augmentation:
+\cite{wu2021polyjuice}, propose \emph{Polyjuice}, a general-purpose
+counterfactual generator for language models. They demonstrate
+empirically that augmenting training data through \emph{Polyjuice}
+counterfactuals improves robustness in a number of NLP tasks.
+\cite{balashankar2023improving} also use \emph{Polyjuice} to augment NLP
+datasets through diverse counterfactuals and show that classifier
+robustness improves up to 20\%. Finally, \cite{luu2023counterfactual}
+introduce Counterfactual Adversarial Training (CAT), which also aims at
+improving generalization and robustness of language models.
+Specifically, they propose to proceed as follows: firstly, they identify
+training samples that are subject to high predictive uncertainty;
+secondly, they generate counterfactual explanations for those samples;
+and, finally, they fine-tune the given language model on the augmented
+dataset that includes the generated counterfactuals.
 
@@
 Pointing to clear similarities in how CE and AE are generated,
-Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
-case for jointly studying the opaqueness and robustness problem in
-representation learning. Formally, AE can be seen as the subset of CE,
-for which misclassification is achieved
-(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
-Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
-CE and AE are equivalent under certain conditions and derive theoretical
-upper bounds on the distances between them.
+\cite{freiesleben2022intriguing} makes the case for jointly studying the
+opaqueness and robustness problem in representation learning. Formally,
+AE can be seen as the subset of CE, for which misclassification is
+achieved \cite{freiesleben2022intriguing}. Similarly,
+\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
+certain conditions and derive theoretical upper bounds on the distances
+between them.
 
@@
 certain properties of post-hoc counterfactual explanations. Firstly,
-Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
-propose a way to train models that are guaranteed to provide recourse
-for individuals to move from an adverse outcome to some positive target
-class with high probability. The approach proposed by Ross, Lakkaraju,
-and Bastani (\citeproc{ref-ross2021learning}{2024}) builds on
-adversarial training, where in this context susceptibility to targeted
-adversarial examples for the positive class is explicitly induced. The
-proposed method allows for imposing a set of actionability constraints
-ex-ante: for example, users can specify that certain features
-(e.g.~\emph{age}, \emph{gender}, \ldots) are immutable. Secondly, Guo,
-Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the first
-to propose an end-to-end training pipeline that includes counterfactual
-explanations as part of the training procedure. In particular, they
-propose a specific network architecture that includes a predictor and CE
-generator network, where the parameters of the CE generator network are
-learnable. Counterfactuals are generated during each training iteration
-and fed back to the predictor network. In contrast to Guo, Nguyen, and
-Yadav (\citeproc{ref-guo2023counternet}{2023}), we impose no
-restrictions on the neural network architecture at all.
+\cite{ross2021learning} propose a way to train models that are
+guaranteed to provide recourse for individuals to move from an adverse
+outcome to some positive target class with high probability. The
+approach proposed by \cite{ross2021learning} builds on adversarial
+training, where in this context susceptibility to targeted adversarial
+examples for the positive class is explicitly induced. The proposed
+method allows for imposing a set of actionability constraints ex-ante:
+for example, users can specify that certain features (e.g.~\emph{age},
+\emph{gender}, \ldots) are immutable. Secondly, \cite{guo2023counternet}
+are the first to propose an end-to-end training pipeline that includes
+counterfactual explanations as part of the training procedure. In
+particular, they propose a specific network architecture that includes a
+predictor and CE generator network, where the parameters of the CE
+generator network are learnable. Counterfactuals are generated during
+each training iteration and fed back to the predictor network. In
+contrast to \cite{guo2023counternet}, we impose no restrictions on the
+neural network architecture at all.
 
@@
 towards aligning representations with plausible explanations. In a work
-closely related to this one, Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}) show that explainability can
-be improved through model averaging and refined model objectives. The
-authors propose a way to generate counterfactuals that are maximally
-\textbf{faithful} to the model in that they are consistent with what the
-model has learned about the underlying data. Formally, they rely on
-tools from energy-based modelling to minimize the divergence between the
-distribution of counterfactuals and the conditional posterior over
-inputs learned by the model. Their proposed counterfactual explainer,
-\emph{ECCCo}, yields plausible explanations if and only if the
-underlying model has learned representations that align with them. They
-find that both deep ensembles
-(\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
-and Blundell 2017}) and joint energy-based models (JEMs)
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
-well in this regard.
+closely related to this one, \cite{altmeyer2024faithful} show that
+explainability can be improved through model averaging and refined model
+objectives. The authors propose a way to generate counterfactuals that
+are maximally \textbf{faithful} to the model in that they are consistent
+with what the model has learned about the underlying data. Formally,
+they rely on tools from energy-based modelling to minimize the
+divergence between the distribution of counterfactuals and the
+conditional posterior over inputs learned by the model. Their proposed
+counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
+and only if the underlying model has learned representations that align
+with them. They find that both deep ensembles
+\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
+\cite{grathwohl2020your} tend to do well in this regard.
 
@@
 approximate Bayesian model averages, which are most called for when
-models are underspecified by the available data
-(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
-mitigates the aforementioned risk of relying on a single locally optimal
-representations that corresponds to semantically meaningless
-explanations for the data. Previous work by Schut et al.
-(\citeproc{ref-schut2021generating}{2021}) similarly found that
-generating plausible (``interpretable'') counterfactual explanations is
-almost trivial for deep ensembles that have also undergone adversarial
-training. The case for JEMs is even clearer: they involve a hybrid
-objective that induces both high predictive performance and generative
-capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
-is closely related to the idea of aligning models with plausible
-explanations and has inspired our proposed counterfactual training
-objective, as we explain in Section~\ref{sec-method}.
+models are underspecified by the available data \cite{wilson2020case}.
+Averaging across solutions mitigates the aforementioned risk of relying
+on a single locally optimal representations that corresponds to
+semantically meaningless explanations for the data. Previous work by
+\cite{schut2021generating} similarly found that generating plausible
+(``interpretable'') counterfactual explanations is almost trivial for
+deep ensembles that have also undergone adversarial training. The case
+for JEMs is even clearer: they involve a hybrid objective that induces
+both high predictive performance and generative capacity
+\cite{grathwohl2020your}. This is closely related to the idea of
+aligning models with plausible explanations and has inspired our
+proposed counterfactual training objective, as we explain in
+Section~\ref{sec-method}.
 
@@
 counterfactuals comply with the underlying data generating process
-(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
-\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
-is a necessary but insufficient condition for using CE to provide
-algorithmic recourse (AR) to individuals affected by opaque models in
-practice. This is because for recourse recommendations to be
+\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
+Plausibility is a necessary but insufficient condition for using CE to
+provide algorithmic recourse (AR) to individuals affected by opaque
+models in practice. This is because for recourse recommendations to be
 \textbf{actionable}, they need to not only result in plausible
@@
 The definitions of faithfulness and plausibility in
-Definition~\ref{def-explainability} are the same as in Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
-Actionability constraints in Definition~\ref{def-explainability} vary
-and depend on the context in which \(\mathbf{M}_\theta\) is deployed. In
-this work, we focus on domain and mutability constraints for individual
-features \(x_d\) for \(d=1,...,D\). We limit ourselves to classification
-tasks for reasons discussed in Section~\ref{sec-discussion}.
+Definition~\ref{def-explainability} are the same as in
+\cite{altmeyer2024faithful}, with adapted notation. Actionability
+constraints in Definition~\ref{def-explainability} vary and depend on
+the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
+focus on domain and mutability constraints for individual features
+\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
+for reasons discussed in Section~\ref{sec-discussion}.
 
@@
 explanation generated through gradient descent over \(T\) iterations as
-initially proposed by Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
-let \(T\) vary and consider the counterfactual search as converged as
-soon as the predicted probability for the target class has reached a
-pre-determined threshold, \(\tau\):
+initially proposed by \cite{wachter2017counterfactual}. For our
+purposes, we let \(T\) vary and consider the counterfactual search as
+converged as soon as the predicted probability for the target class has
+reached a pre-determined threshold, \(\tau\):
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
@@
 
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
-any classifier can be re-interpreted as a joint energy-based model (JEM)
-that learns to discriminate output classes conditional on inputs and
-generate inputs. They show that JEMs can be trained to perform well at
-both tasks by directly maximizing the joint log-likelihood factorized as
+\cite{grathwohl2020your} observe that any classifier can be
+re-interpreted as a joint energy-based model (JEM) that learns to
+discriminate output classes conditional on inputs and generate inputs.
+They show that JEMs can be trained to perform well at both tasks by
+directly maximizing the joint log-likelihood factorized as
 \(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
@@
 Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
-contrastive divergence between samples drawn from
-\(p_\theta(\mathbf{x})\) and training observations, i.e.~samples from
-\(p(\mathbf{x})\).
-
-A key empirical finding in Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
-well with respect to the plausibility objective in
+\cite{grathwohl2020your} minimize the contrastive divergence between
+samples drawn from \(p_\theta(\mathbf{x})\) and training observations,
+i.e.~samples from \(p(\mathbf{x})\).
+
+A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
+tend to do well with respect to the plausibility objective in
 Definition~\ref{def-explainability}. If we consider samples drawn from
@@
 between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
-To generate samples, Grathwohl et al.
-(\citeproc{ref-grathwohl2020your}{2020}) rely on Stochastic Gradient
-Langevin Dynamics (SGLD) using an uninformative prior for
+To generate samples, \cite{grathwohl2020your} rely on Stochastic
+Gradient Langevin Dynamics (SGLD) using an uninformative prior for
 initialization. This is where we depart from their methodology: instead
@@
 generator. This is similar to how conditional sampling is used to draw
-from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
-(\citeproc{ref-grathwohl2020your}{2020}).
+from \(p_\theta(\mathbf{x})\) in \cite{grathwohl2020your}.
 
@@
 increasing the energy of counterfactuals (negative samples)
-(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
-counterfactuals get more plausible (Definition~\ref{def-explainability})
-over the cause of training, these two opposing effects gradually balance
-each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
+\cite{du2019implicit}. As the generated counterfactuals get more
+plausible (Definition~\ref{def-explainability}) over the cause of
+training, these two opposing effects gradually balance each out
+\cite{lippe2024uvadlc}.
 
@@
 as it does not violate the faithfulness condition. Like JEMs
-(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), counterfactual
-training can be considered as a form of contrastive representation
-learning.
+\cite{murphy2022probabilistic}, counterfactual training can be
+considered as a form of contrastive representation learning.
 
@@
 \(\mathbf{x}^\prime\) can be repurposed as additional training samples
-(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023};
-\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023}) or
-adversarial examples
-(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
-\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
+\cite{luu2023counterfactual,balashankar2023improving} or adversarial
+examples \cite{freiesleben2022intriguing,pawelczyk2022exploring}. This
 leaves some flexibility with respect to the exact choice for
@@
 magnitude of the perturbation to any individual feature is at most
-\(\varepsilon\). This is closely aligned with Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2013}), who define an adversarial
-attack as an ``imperceptible non-random perturbation''. Thus, we choose
-to work with a different distinction between CE and AE than Freiesleben
-(\citeproc{ref-freiesleben2022intriguing}{2022}), who considers
-misclassification as the key distinguishing feature of AE. One of the
-key observations in this work is that we can leverage counterfactual
-explanations during training and get adversarial examples, essentially
-for free.
+\(\varepsilon\). This is closely aligned with
+\cite{szegedy2013intriguing}, who define an adversarial attack as an
+``imperceptible non-random perturbation''. Thus, we choose to work with
+a different distinction between CE and AE than
+\cite{freiesleben2022intriguing}, who considers misclassification as the
+key distinguishing feature of AE. One of the key observations in this
+work is that we can leverage counterfactual explanations during training
+and get adversarial examples, essentially for free.
 
@@
 descent in the feature space for optimization
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
-In this context, domain constraints can be imposed by simply projecting
-counterfactuals back to the specified domain, if the previous gradient
-step resulted in updated feature values that were out-of-domain.
-Mutability constraints can similarly be enforced by setting partial
-derivatives to zero to ensure that features are only mutated in the
-allowed direction, if at all.
+\cite{altmeyer2023explaining}. In this context, domain constraints can
+be imposed by simply projecting counterfactuals back to the specified
+domain, if the previous gradient step resulted in updated feature values
+that were out-of-domain. Mutability constraints can similarly be
+enforced by setting partial derivatives to zero to ensure that features
+are only mutated in the allowed direction, if at all.
 
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
-Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
-den Hengel. 2020. {``Counterfactual Vision and Language Learning.''} In
-\emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
-Recognition (CVPR)}, 10041--51.
-\url{https://doi.org/10.1109/CVPR42600.2020.01006}.
-
-\bibitem[\citeproctext]{ref-altmeyer2023explaining}
-Altmeyer, Patrick, Arie van Deursen, et al. 2023. {``Explaining
-Black-Box Models Through Counterfactuals.''} In \emph{Proceedings of the
-JuliaCon Conferences}, 1:130. 1.
-
-\bibitem[\citeproctext]{ref-altmeyer2024faithful}
-Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia CS
-Liem. 2024. {``Faithful Model Explanations Through Energy-Constrained
-Conformal Counterfactuals.''} In \emph{Proceedings of the AAAI
-Conference on Artificial Intelligence}, 38:10829--37. 10.
-
-\bibitem[\citeproctext]{ref-augustin2020adversarial}
-Augustin, Maximilian, Alexander Meinke, and Matthias Hein. 2020.
-{``Adversarial Robustness on in-and Out-Distribution Improves
-Explainability.''} In \emph{European Conference on Computer Vision},
-228--45. Springer.
-
-\bibitem[\citeproctext]{ref-balashankar2023improving}
-Balashankar, Ananth, Xuezhi Wang, Yao Qin, Ben Packer, Nithum Thain, Ed
-Chi, Jilin Chen, and Alex Beutel. 2023. {``Improving Classifier
-Robustness Through Active Generative Counterfactual Data
-Augmentation.''} In \emph{Findings of the Association for Computational
-Linguistics: EMNLP 2023}, 127--39.
-
-\bibitem[\citeproctext]{ref-bezanson2017julia}
-Bezanson, Jeff, Alan Edelman, Stefan Karpinski, and Viral B Shah. 2017.
-{``Julia: A Fresh Approach to Numerical Computing.''} \emph{SIAM Review}
-59 (1): 65--98. \url{https://doi.org/10.1137/141000671}.
-
-\bibitem[\citeproctext]{ref-milan2023dataframes}
-Bouchet-Valat, Milan, and Bogumił Kamiński. 2023. {``DataFrames.jl:
-Flexible and Fast Tabular Data in Julia.''} \emph{Journal of Statistical
-Software} 107 (4): 1--32. \url{https://doi.org/10.18637/jss.v107.i04}.
-
-\bibitem[\citeproctext]{ref-byrne2021mpi}
-Byrne, Simon, Lucas C. Wilcox, and Valentin Churavy. 2021. {``MPI.jl:
-Julia Bindings for the Message Passing Interface.''} \emph{Proceedings
-of the JuliaCon Conferences} 1 (1): 68.
-\url{https://doi.org/10.21105/jcon.00068}.
-
-\bibitem[\citeproctext]{ref-chagas2024pretty}
-Chagas, Ronan Arraes Jardim, Ben Baumgold, Glen Hertz, Hendrik Ranocha,
-Mark Wells, Nathan Boyer, Nicholas Ritchie, et al. 2024.
-{``Ronisbr/PrettyTables.jl: V2.4.0.''} Zenodo.
-\url{https://doi.org/10.5281/zenodo.13835553}.
-
-\bibitem[\citeproctext]{ref-PlotsJL}
-Christ, Simon, Daniel Schwabeneder, Christopher Rackauckas, Michael
-Krabbe Borregaard, and Thomas Breloff. 2023. {``Plots.jl -- a User
-Extendable Plotting API for the Julia Programming Language.''}
-https://doi.org/\url{https://doi.org/10.5334/jors.431}.
-
-\bibitem[\citeproctext]{ref-danisch2021makie}
-Danisch, Simon, and Julius Krumbiegel. 2021. {``{Makie.jl}: Flexible
-High-Performance Data Visualization for {Julia}.''} \emph{Journal of
-Open Source Software} 6 (65): 3349.
-\url{https://doi.org/10.21105/joss.03349}.
-
-\bibitem[\citeproctext]{ref-du2019implicit}
-Du, Yilun, and Igor Mordatch. 2020. {``Implicit Generation and
-Generalization in Energy-Based Models.''}
-\url{https://arxiv.org/abs/1903.08689}.
-
-\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
-Freiesleben, Timo. 2022. {``The Intriguing Relation Between
-Counterfactual Explanations and Adversarial Examples.''} \emph{Minds and
-Machines} 32 (1): 77--109.
-
-\bibitem[\citeproctext]{ref-goodfellow2014explaining}
-Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
-{``Explaining and Harnessing Adversarial Examples.''}
-\url{https://arxiv.org/abs/1412.6572}.
-
-\bibitem[\citeproctext]{ref-goodfellow2016deep}
-Goodfellow, Ian, Yoshua Bengio, and Aaron Courville. 2016. \emph{Deep
-{Learning}}. {MIT Press}.
-
-\bibitem[\citeproctext]{ref-grathwohl2020your}
-Grathwohl, Will, Kuan-Chieh Wang, Joern-Henrik Jacobsen, David Duvenaud,
-Mohammad Norouzi, and Kevin Swersky. 2020. {``Your Classifier Is
-Secretly an Energy Based Model and You Should Treat It Like One.''} In
-\emph{International Conference on Learning Representations}.
-
-\bibitem[\citeproctext]{ref-guidotti2022counterfactual}
-Guidotti, Riccardo. 2022. {``Counterfactual Explanations and How to Find
-Them: Literature Review and Benchmarking.''} \emph{Data Mining and
-Knowledge Discovery}, 1--55.
-
-\bibitem[\citeproctext]{ref-guo2023counternet}
-Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``CounterNet:
-End-to-End Training of Prediction Aware Counterfactual Explanations.''}
-In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
-Discovery and Data Mining}, 577--89. KDD '23. New York, NY, USA:
-Association for Computing Machinery.
-\url{https://doi.org/10.1145/3580305.3599290}.
-
-\bibitem[\citeproctext]{ref-hastie2009elements}
-Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
-Elements of Statistical Learning}. Springer New York.
-\url{https://doi.org/10.1007/978-0-387-84858-7}.
-
-\bibitem[\citeproctext]{ref-innes2018fashionable}
-Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco
-Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and
-Viral Shah. 2018. {``Fashionable Modelling with Flux.''}
-\url{https://arxiv.org/abs/1811.01457}.
-
-\bibitem[\citeproctext]{ref-innes2018flux}
-Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
-\emph{Journal of Open Source Software} 3 (25): 602.
-\url{https://doi.org/10.21105/joss.00602}.
-
-\bibitem[\citeproctext]{ref-joshi2019realistic}
-Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
-Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
-Actionable Explanations in Black-Box Decision Making Systems.''}
-\url{https://arxiv.org/abs/1907.09615}.
-
-\bibitem[\citeproctext]{ref-kolter2023keynote}
-Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
-IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)},
-xvi--. Los Alamitos, CA, USA: IEEE Computer Society.
-\url{https://doi.org/10.1109/SaTML54575.2023.00009}.
-
-\bibitem[\citeproctext]{ref-lakshminarayanan2016simple}
-Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2017.
-{``Simple and Scalable Predictive Uncertainty Estimation Using Deep
-Ensembles.''} \emph{Advances in Neural Information Processing Systems}
-30.
-
-\bibitem[\citeproctext]{ref-lippe2024uvadlc}
-Lippe, Phillip. 2024. {``{UvA Deep Learning Tutorials}.''}
-\url{https://uvadlc-notebooks.readthedocs.io/en/latest/}.
-
-\bibitem[\citeproctext]{ref-luu2023counterfactual}
-Luu, Hoai Linh, and Naoya Inoue. 2023. {``Counterfactual Adversarial
-Training for Improving Robustness of Pre-Trained Language Models.''} In
-\emph{Proceedings of the 37th Pacific Asia Conference on Language,
-Information and Computation}, 881--88.
-
-\bibitem[\citeproctext]{ref-murphy2022probabilistic}
-Murphy, Kevin P. 2022. \emph{Probabilistic {Machine Learning}: {An}
-Introduction}. {MIT Press}.
-
-\bibitem[\citeproctext]{ref-oneil2016weapons}
-O'Neil, Cathy. 2016. \emph{Weapons of Math Destruction: {How} Big Data
-Increases Inequality and Threatens Democracy}. {Crown}.
-
-\bibitem[\citeproctext]{ref-pawelczyk2022exploring}
-Pawelczyk, Martin, Chirag Agarwal, Shalmali Joshi, Sohini Upadhyay, and
-Himabindu Lakkaraju. 2022. {``Exploring Counterfactual Explanations
-Through the Lens of Adversarial Examples: A Theoretical and Empirical
-Analysis.''} In \emph{Proceedings of the 25th International Conference
-on Artificial Intelligence and Statistics}, edited by Gustau
-Camps-Valls, Francisco J. R. Ruiz, and Isabel Valera, 151:4574--94.
-Proceedings of Machine Learning Research. PMLR.
-\url{https://proceedings.mlr.press/v151/pawelczyk22a.html}.
-
-\bibitem[\citeproctext]{ref-poyiadzi2020face}
-Poyiadzi, Rafael, Kacper Sokol, Raul Santos-Rodriguez, Tijl De Bie, and
-Peter Flach. 2020. {``{FACE}: {Feasible} and Actionable Counterfactual
-Explanations.''} In \emph{Proceedings of the {AAAI}/{ACM Conference} on
-{AI}, {Ethics}, and {Society}}, 344--50.
-
-\bibitem[\citeproctext]{ref-ross2021learning}
-Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024. {``Learning
-Models for Actionable Recourse.''} In \emph{Proceedings of the 35th
-International Conference on Neural Information Processing Systems}. NIPS
-'21. Red Hook, NY, USA: Curran Associates Inc.
-
-\bibitem[\citeproctext]{ref-sauer2021counterfactual}
-Sauer, Axel, and Andreas Geiger. 2021. {``Counterfactual Generative
-Networks.''} \url{https://arxiv.org/abs/2101.06046}.
-
-\bibitem[\citeproctext]{ref-schut2021generating}
-Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
-Sacaleanu, Yarin Gal, et al. 2021. {``Generating {Interpretable
-Counterfactual Explanations By Implicit Minimisation} of {Epistemic} and
-{Aleatoric Uncertainties}.''} In \emph{International {Conference} on
-{Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.
-
-\bibitem[\citeproctext]{ref-szegedy2013intriguing}
-Szegedy, Christian, Wojciech Zaremba, Ilya Sutskever, Joan Bruna,
-Dumitru Erhan, Ian Goodfellow, and Rob Fergus. 2013. {``Intriguing
-Properties of Neural Networks.''} \url{https://arxiv.org/abs/1312.6199}.
-
-\bibitem[\citeproctext]{ref-teney2020learning}
-Teney, Damien, Ehsan Abbasnedjad, and Anton van den Hengel. 2020.
-{``Learning What Makes a Difference from Counterfactual Examples and
-Gradient Supervision.''} In \emph{Computer Vision--ECCV 2020: 16th
-European Conference, Glasgow, UK, August 23--28, 2020, Proceedings, Part
-x 16}, 580--99. Springer.
-
-\bibitem[\citeproctext]{ref-wachter2017counterfactual}
-Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
-{``Counterfactual Explanations Without Opening the Black Box:
-{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
-841. \url{https://doi.org/10.2139/ssrn.3063289}.
-
-\bibitem[\citeproctext]{ref-wilson2020case}
-Wilson, Andrew Gordon. 2020. {``The Case for Bayesian Deep Learning.''}
-\url{https://arxiv.org/abs/2001.10995}.
-
-\bibitem[\citeproctext]{ref-wu2021polyjuice}
-Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
-2021. {``Polyjuice: Generating Counterfactuals for Explaining,
-Evaluating, and Improving Models.''} In \emph{Proceedings of the 59th
-Annual Meeting of the Association for Computational Linguistics and the
-11th International Joint Conference on Natural Language Processing
-(Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
-and Roberto Navigli, 6707--23. Online: Association for Computational
-Linguistics. \url{https://doi.org/10.18653/v1/2021.acl-long.523}.
-
-\end{CSLReferences}
-
 \newpage{}
@@
 
-Gradient-based counterfactual search was originally proposed by Wachter,
-Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}). It generally solves
-the following unconstrained objective,
+Gradient-based counterfactual search was originally proposed by
+\cite{wachter2017counterfactual}. It generally solves the following
+unconstrained objective,
 
@@
 the training data, where \(g(\cdot)\) corresponds to the decoder
-(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
-the distance penalty, \emph{ECCo} uses an additional penalty component
-that regularizes the energy associated with the counterfactual,
-\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
-al. 2024}).
+\cite{joshi2019realistic}. In addition to the distance penalty,
+\emph{ECCo} uses an additional penalty component that regularizes the
+energy associated with the counterfactual, \(\mathbf{x}^\prime\)
+\cite{altmeyer2024faithful}.
 
@@
 have for the log likelihood ratio between any two classes
-\(k,m \in \mathcal{K}\) (\citeproc{ref-hastie2009elements}{Hastie,
-Tibshirani, and Friedman 2009}):
+\(k,m \in \mathcal{K}\) \cite{hastie2009elements}:
 
@@
 training instabilities commonly associated with energy-based modelling
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+\cite{grathwohl2020your,altmeyer2024faithful}.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, bottomrule=.15mm, coltitle=black, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomtitle=1mm, leftrule=.75mm, breakable, colframe=quarto-callout-note-color-frame, left=2mm, colback=white, toptitle=1mm, toprule=.15mm, rightrule=.15mm, opacitybacktitle=0.6, title={Note}]
 
@@
 generate counterfactuals that are faithful to the model
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
+\cite{altmeyer2024faithful}.}
 
@@
 to generate counterfactuals that are faithful to the model
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
+\cite{altmeyer2024faithful}.}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, bottomrule=.15mm, coltitle=black, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomtitle=1mm, leftrule=.75mm, breakable, colframe=quarto-callout-note-color-frame, left=2mm, colback=white, toptitle=1mm, toprule=.15mm, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, bottomrule=.15mm, coltitle=black, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomtitle=1mm, leftrule=.75mm, breakable, colframe=quarto-callout-note-color-frame, left=2mm, colback=white, toptitle=1mm, toprule=.15mm, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]
 
@@
 All computations were performed in the Julia Programming Language
-(\citeproc{ref-bezanson2017julia}{Bezanson et al. 2017}). We have
-developed a package for counterfactual training that leverages and
-extends the functionality provided by several existing packages, most
-notably
+\cite{bezanson2017julia}. We have developed a package for counterfactual
+training that leverages and extends the functionality provided by
+several existing packages, most notably
 \href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{CounterfactualExplanations.jl}
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023})
-and the \href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library for
-deep learning (\citeproc{ref-innes2018fashionable}{Michael Innes et al.
-2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). For
-data-wrangling and presentation-ready tables we relied on
+\cite{altmeyer2023explaining} and the
+\href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library for deep
+learning \cite{innes2018fashionable,innes2018flux}. For data-wrangling
+and presentation-ready tables we relied on
 \href{https://dataframes.juliadata.org/v1.7/}{DataFrames.jl}
-(\citeproc{ref-milan2023dataframes}{Bouchet-Valat and Kamiński 2023})
-and
+\cite{milan2023dataframes} and
 \href{https://ronisbr.github.io/PrettyTables.jl/v2.4/}{PrettyTables.jl}
-(\citeproc{ref-chagas2024pretty}{Chagas et al. 2024}), respectively. For
-plots and visualizations we used both
-\href{https://docs.juliaplots.org/v1.40/}{Plots.jl}
-(\citeproc{ref-PlotsJL}{Christ et al. 2023}) and
-\href{https://docs.makie.org/v0.22/}{Makie.jl}
-(\citeproc{ref-danisch2021makie}{Danisch and Krumbiegel 2021}), in
-particular \href{https://aog.makie.org/v0.9.3/}{AlgebraOfGraphics.jl}.
-To distribute computational tasks across multiple processors, we have
+\cite{chagas2024pretty}, respectively. For plots and visualizations we
+used both \href{https://docs.juliaplots.org/v1.40/}{Plots.jl}
+\cite{PlotsJL} and \href{https://docs.makie.org/v0.22/}{Makie.jl}
+\cite{danisch2021makie}, in particular
+\href{https://aog.makie.org/v0.9.3/}{AlgebraOfGraphics.jl}. To
+distribute computational tasks across multiple processors, we have
 relied on \href{https://juliaparallel.org/MPI.jl/v0.20/}{MPI.jl}
-(\citeproc{ref-byrne2021mpi}{Byrne, Wilcox, and Churavy 2021}).
+\cite{byrne2021mpi}.
+
+\bibliographystyle{splncs04}
+\bibliography{bibliography}
 

### 2025-03-06 paper/quarto_ecml/paper.tex
long road ahead still

@@
 % \usepackage{amsthm}
+% \theoremstyle{definition}
+% \newtheorem{example}{Example}[section]
+% \theoremstyle{plain}
+% \newtheorem{proposition}{Proposition}[section]
 % \theoremstyle{plain}
 % \newtheorem{corollary}{Research Question}[section]
-% \theoremstyle{plain}
-% \newtheorem{proposition}{Proposition}[section]
-% \theoremstyle{definition}
-% \newtheorem{example}{Example}[section]
 % \theoremstyle{definition}
@@
   pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
-  pdfkeywords={Counterfactual Explanations, Explainable
-AI, Representation Learning},
+  pdfkeywords={Counterfactual Explanations, Algorithmic
+Recourse, Explainable AI, Adversarial Robustness, Representation
+Learning},
   colorlinks=true,
@@
 
-\keywords{Counterfactual Explanations \and Explainable
-AI \and Representation Learning}
+\keywords{Counterfactual Explanations \and Algorithmic
+Recourse \and Explainable AI \and Adversarial
+Robustness \and Representation Learning}
 
@@
 
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
+
 \newpage{}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, bottomrule=.15mm, coltitle=black, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomtitle=1mm, leftrule=.75mm, breakable, colframe=quarto-callout-note-color-frame, left=2mm, colback=white, toptitle=1mm, toprule=.15mm, rightrule=.15mm, opacitybacktitle=0.6, title={Note}]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, leftrule=.75mm, arc=.35mm, titlerule=0mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, title={Note}, bottomtitle=1mm, bottomrule=.15mm, toprule=.15mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, bottomrule=.15mm, coltitle=black, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomtitle=1mm, leftrule=.75mm, breakable, colframe=quarto-callout-note-color-frame, left=2mm, colback=white, toptitle=1mm, toprule=.15mm, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, leftrule=.75mm, arc=.35mm, titlerule=0mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, bottomtitle=1mm, bottomrule=.15mm, toprule=.15mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, bottomrule=.15mm, coltitle=black, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomtitle=1mm, leftrule=.75mm, breakable, colframe=quarto-callout-note-color-frame, left=2mm, colback=white, toptitle=1mm, toprule=.15mm, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, leftrule=.75mm, arc=.35mm, titlerule=0mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, bottomrule=.15mm, toprule=.15mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable]
 

### 2025-03-07 paper/preprint/paper.tex
some notes from meeting and updates to appendix

@@
 \newtheorem{proposition}{Proposition}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{plain}
 \newtheorem{corollary}{Research Question}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
 \theoremstyle{remark}
@@
   pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
-  pdfkeywords={Counterfactual Explanations, Explainable
-AI, Representation Learning},
+  pdfkeywords={Counterfactual Training, Counterfactual
+Explanations, Algorithmic Recourse, Explainable AI, Representation
+Learning},
   colorlinks=true,
@@
 \begin{abstract}
-We propose a novel model objective that leverages counterfactual
-explanations to increase the explanatory capacity of trained models.
-Counterfactual explanations have emerged as a popular tool to explain
-predictions made by opaque machine learning models: they inform how
-factual inputs would need to change in order for a model to produce some
-desired output. Much existing research has focused on generating
-counterfactuals that are not only valid but also deemed acceptable with
-respect to the stakeholder requirements and plausible with respect to
-the underlying data. Recent work has shown that under this premise, the
-task of learning plausible explanations---ones that respect the
-conditional distribution of the target class---is effectively reassigned
-from the model itself to the post-hoc counterfactual explainer. Building
-on that work, we propose a novel objective function that employs
-counterfactuals (ad-hoc) during the training phase to minimize the
-divergence between the learned representations and the explanations.
-Through extensive experiments, we demonstrate that our proposed method
-facilitates training models that deliver inherently plausible
-explanations while maintaining high predictive performance.
+We propose a novel training regime called Counterfactual Training that
+leverages counterfactual explanations to increase the explanatory
+capacity of models. Counterfactual explanations have emerged as a
+popular post-hoc explanation method for opaque machine learning models:
+they inform how factual inputs would need to change in order for a model
+to produce some desired output. To be useful in real-word
+decision-making systems, counterfactuals should be plausible with
+respect to the underlying data and actionable with respect to
+stakeholder requirements. Much existing research has therefore focused
+on developing post-hoc methods to generate counterfactuals that meet
+these desiderata. In this work, we instead hold models directly
+accountable for this desired end goal: Counterfactual Training employs
+counterfactuals ad-hoc during the training phase to minimize the
+divergence between learned representations and plausible, actionable
+explanations. We demonstrate empirically and theoretically that our
+proposed method facilitates training models that deliver inherently
+desirable explanations while maintaining high predictive performance.
 \end{abstract}
@@
 \def\sep{\textbullet\ }
-Counterfactual Explanations \sep Explainable AI \sep 
+Counterfactual Training \sep Counterfactual
+Explanations \sep Algorithmic Recourse \sep Explainable AI \sep 
 Representation Learning
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, coltitle=black, titlerule=0mm, left=2mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colback=white, rightrule=.15mm, toprule=.15mm, bottomtitle=1mm, opacitybacktitle=0.6, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
 
@@
 
-\subsubsection{High-Level Findings}\label{high-level-findings}
+\paragraph{Plausibility}\label{plausibility-1}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-grid-gen_params-plaus-circles} to
+Figure~\ref{fig-grid-gen_params-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-circles}Average outcomes for
+the plausibility measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-lin_sep}Average outcomes for
+the plausibility measure across hyperparameters. Data: Linearly
+Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
 
@@
 
+}
+
+\caption{\label{fig-grid-gen_params-plaus-moons}Average outcomes for the
+plausibility measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-over}Average outcomes for the
+plausibility measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Cost}\label{cost}
+
+The results with respect to the cost measure are shown in
+Figure~\ref{fig-grid-gen_params-cost-circles} to
+Figure~\ref{fig-grid-gen_params-cost-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-circles}Average outcomes for
+the cost measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-lin_sep}Average outcomes for
+the cost measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
 \pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/percent_valid.png}}
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/acc_val.png}}
-
-\subsubsection{Detailed Findings}\label{detailed-findings}
-
-\paragraph{Linearly Separable}\label{linearly-separable}
-
-\begin{longtable}{cccccc}
-
-\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly
-Separable data by energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Decision Threshold} & \textbf{$\lambda_{\text{energy}}$} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Decision Threshold} & \textbf{$\lambda_{\text{energy}}$} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{6}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  0.75 & 0.1 & 5 & \textit{ECCo} & -15.9 & 16.7 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{41.9}} & \color{Green}{\textbf{8.13}} \\
-  0.75 & 0.1 & 5 & \textit{REVISE} & -64.4 & 3.58 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{28.9}} & \color{Green}{\textbf{2.28}} \\
-  0.75 & 0.1 & 25 & \textit{Generic} & -11.3 & 11.5 \\
-  0.75 & 0.1 & 25 & \textit{REVISE} & -55.3 & 4.18 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{10.9}} & \color{Green}{\textbf{7.47}} \\
-  0.75 & 0.1 & 50 & \textit{Generic} & -32.3 & 16.6 \\
-  0.75 & 0.1 & 50 & \textit{REVISE} & -44.1 & 4.24 \\
-  0.9 & 0.1 & 5 & \textit{ECCo} & -13.7 & 13.7 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{22.3}} & \color{Green}{\textbf{4.87}} \\
-  0.9 & 0.1 & 5 & \textit{REVISE} & -27.8 & 6.26 \\
-  0.9 & 0.1 & 25 & \textit{ECCo} & 8.84 & 7.13 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{30.7}} & \color{Green}{\textbf{5.77}} \\
-  0.9 & 0.1 & 25 & \textit{REVISE} & -72.7 & 34.9 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{17.1}} & \color{Green}{\textbf{7.98}} \\
-  0.9 & 0.1 & 50 & \textit{Generic} & 3.82 & 8.86 \\
-  0.9 & 0.1 & 50 & \textit{REVISE} & -45.8 & 2.79 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{35.7}} & \color{Green}{\textbf{7.56}} \\
-  0.95 & 0.1 & 5 & \textit{Generic} & -13.3 & 5.73 \\
-  0.95 & 0.1 & 5 & \textit{REVISE} & -54.9 & 5.43 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{37}} & \color{Green}{\textbf{5.31}} \\
-  0.95 & 0.1 & 25 & \textit{Generic} & 29 & 8.6 \\
-  0.95 & 0.1 & 25 & \textit{REVISE} & -40.5 & 41 \\
-  0.95 & 0.1 & 50 & \textit{ECCo} & 13.2 & 4.89 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{35.5}} & \color{Green}{\textbf{4.27}} \\
-  0.95 & 0.1 & 50 & \textit{REVISE} & -60.4 & 45.8 \\
-  \color{Red}{\textbf{0.75}} & \color{Red}{\textbf{0.5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{ECCo}} & \color{Red}{\textbf{-0.94}} & \color{Red}{\textbf{17.6}} \\
-  0.75 & 0.5 & 5 & \textit{Generic} & -11.7 & 17.4 \\
-  0.75 & 0.5 & 5 & \textit{REVISE} & -37.7 & 3.92 \\
-  0.75 & 0.5 & 25 & \textit{ECCo} & 14.7 & 6.32 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{32.3}} & \color{Green}{\textbf{3.49}} \\
-  0.75 & 0.5 & 25 & \textit{REVISE} & -39.9 & 7.21 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{28.3}} & \color{Green}{\textbf{5.05}} \\
-  0.75 & 0.5 & 50 & \textit{Generic} & -9.72 & 9.19 \\
-  0.75 & 0.5 & 50 & \textit{REVISE} & -30 & 5.22 \\
-  0.9 & 0.5 & 5 & \textit{ECCo} & -7.38 & 13.6 \\
-  \color{Red}{\textbf{0.9}} & \color{Red}{\textbf{0.5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{Generic}} & \color{Red}{\textbf{-0.0952}} & \color{Red}{\textbf{17.1}} \\
-  0.9 & 0.5 & 5 & \textit{REVISE} & -50.6 & 4.77 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{39.1}} & \color{Green}{\textbf{8.3}} \\
-  0.9 & 0.5 & 25 & \textit{Generic} & 38.6 & 6.03 \\
-  0.9 & 0.5 & 25 & \textit{REVISE} & -56.9 & 4.28 \\
-  0.9 & 0.5 & 50 & \textit{ECCo} & -6.53 & 10.9 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{13.6}} & \color{Green}{\textbf{12.8}} \\
-  0.9 & 0.5 & 50 & \textit{REVISE} & -24.1 & 12.5 \\
-  0.95 & 0.5 & 5 & \textit{ECCo} & 0.596 & 1.11 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{4.57}} & \color{Green}{\textbf{5.88}} \\
-  0.95 & 0.5 & 5 & \textit{REVISE} & -34.3 & 6.09 \\
-  0.95 & 0.5 & 25 & \textit{ECCo} & 16.5 & 3.53 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{19.1}} & \color{Green}{\textbf{4.43}} \\
-  0.95 & 0.5 & 25 & \textit{REVISE} & -37.2 & 39.2 \\
-  0.95 & 0.5 & 50 & \textit{ECCo} & -31 & 14.2 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{24.8}} & \color{Green}{\textbf{4.78}} \\
-  0.95 & 0.5 & 50 & \textit{REVISE} & -1.28 & 2.7 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{37}} & \color{Green}{\textbf{9.21}} \\
-  0.75 & 5 & 5 & \textit{Generic} & 0.977 & 21.1 \\
-  0.75 & 5 & 5 & \textit{REVISE} & -55.2 & 4.75 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{21.8}} & \color{Green}{\textbf{5.97}} \\
-  0.75 & 5 & 25 & \textit{Generic} & 17.8 & 5.33 \\
-  0.75 & 5 & 25 & \textit{REVISE} & -68.1 & 8.69 \\
-  0.75 & 5 & 50 & \textit{ECCo} & 0.864 & 9.51 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{8.56}} & \color{Green}{\textbf{13.6}} \\
-  0.75 & 5 & 50 & \textit{REVISE} & -71.8 & 3.18 \\
-  0.9 & 5 & 5 & \textit{ECCo} & -3.59 & 20.3 \\
-  \color{Red}{\textbf{0.9}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{Generic}} & \color{Red}{\textbf{-1.18}} & \color{Red}{\textbf{16.7}} \\
-  0.9 & 5 & 5 & \textit{REVISE} & -58.6 & 3.55 \\
-  0.9 & 5 & 25 & \textit{ECCo} & -1.73 & 9.07 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{21.9}} & \color{Green}{\textbf{7.48}} \\
-  0.9 & 5 & 25 & \textit{REVISE} & -44.4 & 4.23 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{30.8}} & \color{Green}{\textbf{10.7}} \\
-  0.9 & 5 & 50 & \textit{Generic} & 16.3 & 11.1 \\
-  0.9 & 5 & 50 & \textit{REVISE} & -53.3 & 6.73 \\
-  \color{Red}{\textbf{0.95}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{ECCo}} & \color{Red}{\textbf{-5.29}} & \color{Red}{\textbf{24.5}} \\
-  0.95 & 5 & 5 & \textit{Generic} & -5.71 & 17.4 \\
-  0.95 & 5 & 5 & \textit{REVISE} & -36.5 & 4.01 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{26.9}} & \color{Green}{\textbf{4.08}} \\
-  0.95 & 5 & 25 & \textit{Generic} & 19.5 & 5.49 \\
-  0.95 & 5 & 25 & \textit{REVISE} & -25 & 5 \\
-  0.95 & 5 & 50 & \textit{ECCo} & 4.6 & 7.46 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{16.6}} & \color{Green}{\textbf{4.3}} \\
-  0.95 & 5 & 50 & \textit{REVISE} & -50.4 & 46.4 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{2.47}} & \color{Green}{\textbf{8.57}} \\
-  0.75 & 10 & 5 & \textit{Generic} & -7.23 & 9.13 \\
-  0.75 & 10 & 5 & \textit{REVISE} & -67.2 & 5.75 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{29.9}} & \color{Green}{\textbf{15.8}} \\
-  0.75 & 10 & 25 & \textit{Generic} & 4.47 & 8.22 \\
-  0.75 & 10 & 25 & \textit{REVISE} & -52 & 5.51 \\
-  0.75 & 10 & 50 & \textit{ECCo} & 6.23 & 6.11 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{15.2}} & \color{Green}{\textbf{10.2}} \\
-  0.75 & 10 & 50 & \textit{REVISE} & -17 & 5.46 \\
-  0.9 & 10 & 5 & \textit{ECCo} & -6.93 & 14.7 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{6.88}} & \color{Green}{\textbf{5.11}} \\
-  0.9 & 10 & 5 & \textit{REVISE} & -45.6 & 47.8 \\
-  0.9 & 10 & 25 & \textit{ECCo} & 6.2 & 6.75 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{34.7}} & \color{Green}{\textbf{9.74}} \\
-  0.9 & 10 & 25 & \textit{REVISE} & -55.7 & 6.04 \\
-  0.9 & 10 & 50 & \textit{ECCo} & 30.1 & 14.2 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{35}} & \color{Green}{\textbf{7.11}} \\
-  0.9 & 10 & 50 & \textit{REVISE} & -48.8 & 3.3 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{35.7}} & \color{Green}{\textbf{3.83}} \\
-  0.95 & 10 & 5 & \textit{Generic} & 2.85 & 7.9 \\
-  0.95 & 10 & 5 & \textit{REVISE} & -34.9 & 4.56 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{33}} & \color{Green}{\textbf{6.45}} \\
-  0.95 & 10 & 25 & \textit{Generic} & 24.8 & 4.59 \\
-  0.95 & 10 & 25 & \textit{REVISE} & -55.7 & 44 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{8.07}} & \color{Green}{\textbf{7.68}} \\
-  0.95 & 10 & 50 & \textit{Generic} & -7.11 & 9.2 \\
-  0.95 & 10 & 50 & \textit{REVISE} & -26.1 & 18.6 \\
-  0.75 & 20 & 5 & \textit{ECCo} & -10.8 & 20.8 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{12.6}} & \color{Green}{\textbf{17.6}} \\
-  0.75 & 20 & 5 & \textit{REVISE} & -21.6 & 3.34 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{17.2}} & \color{Green}{\textbf{4.25}} \\
-  0.75 & 20 & 25 & \textit{Generic} & -10.3 & 9.59 \\
-  0.75 & 20 & 25 & \textit{REVISE} & -59 & 3.66 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{15.6}} & \color{Green}{\textbf{8.38}} \\
-  0.75 & 20 & 50 & \textit{Generic} & 3.92 & 6.2 \\
-  0.75 & 20 & 50 & \textit{REVISE} & -70.9 & 5.16 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{26.3}} & \color{Green}{\textbf{6.62}} \\
-  0.9 & 20 & 5 & \textit{Generic} & 2.35 & 7.04 \\
-  0.9 & 20 & 5 & \textit{REVISE} & -64.8 & 4.31 \\
-  0.9 & 20 & 25 & \textit{ECCo} & 21.1 & 11.6 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{39}} & \color{Green}{\textbf{5.98}} \\
-  0.9 & 20 & 25 & \textit{REVISE} & -50.7 & 13.8 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{15.1}} & \color{Green}{\textbf{7.46}} \\
-  0.9 & 20 & 50 & \textit{Generic} & 8.53 & 6.95 \\
-  0.9 & 20 & 50 & \textit{REVISE} & -36 & 6.14 \\
-  0.95 & 20 & 5 & \textit{ECCo} & -23.8 & 17.8 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{7.23}} & \color{Green}{\textbf{4.74}} \\
-  0.95 & 20 & 5 & \textit{REVISE} & -34 & 4.11 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{19.3}} & \color{Green}{\textbf{14.6}} \\
-  0.95 & 20 & 25 & \textit{Generic} & -1.35 & 7.71 \\
-  0.95 & 20 & 25 & \textit{REVISE} & -52.8 & 5.11 \\
-  0.95 & 20 & 50 & \textit{ECCo} & 18.7 & 3.84 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{37.3}} & \color{Green}{\textbf{7.78}} \\
-  0.95 & 20 & 50 & \textit{REVISE} & -28.3 & 15 \\\bottomrule
-
-\end{longtable}
+}
+
+\caption{\label{fig-grid-gen_params-cost-moons}Average outcomes for the
+cost measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-over}Average outcomes for the
+cost measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\subsubsection{Penalty Strengths}\label{penalty-strengths}
+
+The hyperparameter grid with varying penalty strengths during training
+is shown in Note~\ref{nte-pen-final-run-train}. The corresponding
+evaluation grid used for these experiments is shown in
+Note~\ref{nte-pen-final-run-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
+
+\quartocalloutnte{nte-pen-final-run-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator: \texttt{ecco,\ generic,\ revise}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{adv}}\): \texttt{0.1,\ 0.25,\ 1.0}
+  \item
+    \(\lambda_{\text{div}}\): \texttt{0.01,\ 0.1,\ 1.0}
+  \item
+    \(\lambda_{\text{reg}}\): \texttt{0.0,\ 0.01,\ 0.1}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
+
+\quartocalloutnte{nte-pen-final-run-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Plausibility}\label{plausibility-2}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-grid-pen-plaus-circles} to
+Figure~\ref{fig-grid-pen-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-circles}Average outcomes for the
+plausibility measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-lin_sep}Average outcomes for the
+plausibility measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-moons}Average outcomes for the
+plausibility measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-over}Average outcomes for the
+plausibility measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Cost}\label{cost-1}
+
+The results with respect to the cost measure are shown in
+Figure~\ref{fig-grid-pen-cost-circles} to
+Figure~\ref{fig-grid-pen-cost-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-circles}Average outcomes for the cost
+measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-lin_sep}Average outcomes for the cost
+measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-moons}Average outcomes for the cost
+measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-over}Average outcomes for the cost
+measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
 

### 2025-03-07 paper/sections/other/grids.qmd
some notes from meeting and updates to appendix

@@
 
+```{julia}
+grid_dir = joinpath(res_dir, "gen_params/mlp")
+data_dirs = readdir(grid_dir) |> x -> joinpath.(grid_dir, x) |> x -> x[isdir.(x)]
+data_names = basename.(data_dirs)
+suffix = "evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/"
+```
+
 The hyperparameter grid with varying generator parameters during training is shown in @nte-gen-params-final-run-train.
@@
 #| output: asis
-dict = CTExperiments.from_toml(joinpath(res_dir,"gen_params/mlp/lin_sep/grid_config.toml")) 
+dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/grid_config.toml")) 
 println(CTExperiments.dict_to_quarto_markdown(dict))
@@
 #| output: asis
-dict = CTExperiments.from_toml(joinpath(res_dir, "gen_params/mlp/lin_sep/evaluation/evaluation_grid_config.toml"))
+dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/evaluation/evaluation_grid_config.toml"))
 println(CTExperiments.dict_to_quarto_markdown(dict))
@@
 
-### High-Level Findings
-
-![](/paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png)
-
-![](/paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png)
-
-![](/paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/percent_valid.png)
-
-![](/paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/acc_val.png)
-
-### Detailed Findings
-
-```{julia}
-gen_params_dir = joinpath(res_dir, "gen_params/mlp")
-```
-
-#### Linearly Separable
-
-```{julia}
-df = CSV.read("$(gen_params_dir)/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.csv", DataFrame)
-select!(df, Not(:objective))
-df = groupby(df, Not(:run, :mean, :lambda_energy_eval, :is_pct)) |> 
-  gdf -> combine(gdf, :mean => (x -> [(mean(x),std(x))]) => [:value,:std]) |>
-  df -> sort(df, [:lambda_energy_exper, :decision_threshold_exper, :maxiter_exper, :generator_type])
-```
-
-::: {#tbl-lin_sep-lambda_energy_exper}
-
-::: {.content-hidden unless-format="pdf"}
+#### Plausibility
 
@@
 
-using CTExperiments.ColorSchemes
-get_table_inputs(df, "value"; alpha=0.9, byvars=["lambda_energy_exper", "decision_threshold_exper", "maxiter_exper"], backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; wrap_table=false)
+fig_label_prefix = "grid-gen_params-plaus"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the plausibility measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "plausibility_distance_from_target.png"
+fig_caption = "Average outcomes for the plausibility measure across hyperparameters."
+full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```
+
+#### Cost
+
+```{julia}
+#| output: asis
+
+fig_label_prefix = "grid-gen_params-cost"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the cost measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "distance.png"
+fig_caption = "Average outcomes for the cost measure across hyperparameters."
+full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption)
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```
+
+### Penalty Strengths
+
+```{julia}
+grid_dir = joinpath(res_dir, "penalties/mlp")
+data_dirs = readdir(grid_dir) |> x -> joinpath.(grid_dir, x) |> x -> x[isdir.(x)]
+data_names = basename.(data_dirs)
+suffix = "evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/"
+```
+
+The hyperparameter grid with varying penalty strengths during training is shown in @nte-pen-final-run-train.
+The corresponding evaluation grid used for these experiments is shown in @nte-pen-final-run-eval.
+
+::: {#nte-pen-final-run-train .callout-note}
+
+## Training Phase
+
+```{julia}
+#| output: asis
+dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/grid_config.toml")) 
+println(CTExperiments.dict_to_quarto_markdown(dict))
 ```
@@
 
-Results for Linearly Separable data by energy penalty.
+::: {#nte-pen-final-run-eval .callout-note}
+
+## Evaluation Phase
+
+```{julia}
+#| output: asis
+dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/evaluation/evaluation_grid_config.toml"))
+println(CTExperiments.dict_to_quarto_markdown(dict))
+```
 
 :::
+
+#### Plausibility
+
+```{julia}
+#| output: asis
+
+fig_label_prefix = "grid-pen-plaus"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the plausibility measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "plausibility_distance_from_target.png"
+fig_caption = "Average outcomes for the plausibility measure across hyperparameters."
+full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```
+
+#### Cost
+
+```{julia}
+#| output: asis
+
+fig_label_prefix = "grid-pen-cost"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the cost measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "distance.png"
+fig_caption = "Average outcomes for the cost measure across hyperparameters."
+full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption)
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```

### 2025-03-07 paper/experiments/CTExperiments/src/evaluate_counterfactuals.jl

@@ -230,7 +230,7 @@ function load_data_models_generators(cfg::AbstractEvaluationConfig)
 
     # Counterfactual generators (same for each exper):
     gen_params = cfg.counterfactual_params.generator_params
-    _gen_name = get_generator_name(gen_params)
+    _gen_name = get_name(gen_params)
     _generator = get_generator(gen_params)
     generators = Dict(_gen_name => _generator)
 

### 2025-03-07 paper/experiments/CTExperiments/src/presentation.jl

@@ -365,3 +365,8 @@ function aggregate_counterfactuals(
         df, "ce", byvars; byvars_must_include=byvars_must_include, agg_fun=:identity
     )
 end
+
+function get_img_command(data_names, full_paths, fig_labels; fig_caption="")
+    fig_cap = fig_caption == "" ? fig_caption : "$fig_caption "
+    return ["![$(fig_cap)Data: $(CTExperiments.get_data_name(nm)).](/$pth){#$(lbl)}" for (nm, pth, lbl) in zip(data_names,full_paths,fig_labels)]
+end

### 2025-03-07 paper/experiments/CTExperiments/src/synthetic.jl

@@ -91,3 +91,12 @@ dname(d::Circles) = "circles"
 function load_data(d::Circles, n::Int; seed=TaijaData.data_seed)
     return load_circles(n; seed=seed)
 end
+
+get_data_name(str::String; pretty=true) = get_name(data_sets[str](); pretty)
+
+get_name(d::Moons; pretty::Bool=false) = pretty ? "Moons" : "moons"
+function get_name(d::LinearlySeparable; pretty::Bool=false)
+    return pretty ? "Linearly Separable" : "lin_sep"
+end
+get_name(d::Circles; pretty::Bool=false) = pretty ? "Circles" : "circles"
+get_name(d::Overlapping; pretty::Bool=false) = pretty ? "Overlapping" : "over"

### 2025-03-07 paper/experiments/CTExperiments/src/tables.jl

@@ -52,7 +52,7 @@ function get_table_inputs(
     return df, (; highlighters=hls, backend=backend, header=header)
 end
 
-format_generator(s::AbstractString) = get_generator_name(generator_types[s](); pretty=true)
+format_generator(s::AbstractString) = get_name(generator_types[s](); pretty=true)
 
 global LatexHeaderReplacements = merge(
     Dict(

### 2025-03-07 paper/experiments/CTExperiments/src/training_params.jl

@@ -22,10 +22,10 @@ struct Generic <: AbstractGeneratorType end
 "Type for the GravitationalGenerator."
 struct Gravitational <: AbstractGeneratorType end
 
-get_generator_name(gen::ECCo; pretty::Bool=false) = pretty ? "ECCo" : "ecco"
-get_generator_name(gen::Generic; pretty::Bool=false) = pretty ? "Generic" : "generic"
-get_generator_name(gen::REVISE; pretty::Bool=false) = pretty ? "REVISE" : "revise"
-function get_generator_name(gen::Gravitational; pretty::Bool=false)
+get_name(gen::ECCo; pretty::Bool=false) = pretty ? "ECCo" : "ecco"
+get_name(gen::Generic; pretty::Bool=false) = pretty ? "Generic" : "generic"
+get_name(gen::REVISE; pretty::Bool=false) = pretty ? "REVISE" : "revise"
+function get_name(gen::Gravitational; pretty::Bool=false)
     return pretty ? "Gravitational" : "gravi"
 end
 
@@ -35,10 +35,10 @@ end
 Catalogue of available generator types.
 """
 const generator_types = Dict(
-    get_generator_name(ECCo()) => ECCo,
-    get_generator_name(Generic()) => Generic,
-    get_generator_name(REVISE()) => REVISE,
-    get_generator_name(Gravitational()) => Gravitational,
+    get_name(ECCo()) => ECCo,
+    get_name(Generic()) => Generic,
+    get_name(REVISE()) => REVISE,
+    get_name(Gravitational()) => Gravitational,
 )
 
 """
@@ -84,7 +84,7 @@ Base.@kwdef struct GeneratorParams <: AbstractGeneratorParams
     lambda_energy::AbstractFloat = 5.0
 end
 
-get_generator_name(params::GeneratorParams) = get_generator_name(params.type)
+get_name(params::GeneratorParams) = get_name(params.type)
 
 """
     get_generator(params::GeneratorParams)

### 2025-03-07 paper/preprint/paper.tex
more work on app

@@
 \usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{plain}
@@
 \theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
 \newtheorem{example}{Example}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 
+\subsection{Approach/Future Directions}\label{approachfuture-directions}
+
 \begin{enumerate}
@@
 \item
+  Training instabilities.
+\item
+  Hyperparameter sensitivity -\textgreater{} can we do better than grid
+  search? (Bayes opt, \ldots)
+\end{enumerate}
+
+\subsection{Limitations}\label{limitations}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{2}
+\tightlist
+\item
   Proxy attributes of immutable features.
@@
   Increased training time.
-\item
-  Training instabilities
 \item
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note}, opacityback=0, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, left=2mm, coltitle=black, toptitle=1mm, breakable, bottomtitle=1mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, titlerule=0mm, arc=.35mm, toprule=.15mm, rightrule=.15mm, opacityback=0, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, leftrule=.75mm]
 
@@
 \caption{\label{fig-grid-pen-cost-over}Average outcomes for the cost
+measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\subsubsection{Training Parameters}\label{training-parameters}
+
+The hyperparameter grid with other varying training parameters is shown
+in Note~\ref{nte-train-final-run-train}. The corresponding evaluation
+grid used for these experiments is shown in
+Note~\ref{nte-train-final-run-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, leftrule=.75mm]
+
+\quartocalloutnte{nte-train-final-run-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator: \texttt{ecco,\ generic,\ revise}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Burnin: \texttt{0.0,\ 0.5}
+  \item
+    No.~Counterfactuals: \texttt{100,\ 1000}
+  \item
+    No.~Epochs: \texttt{50,\ 100}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, leftrule=.75mm]
+
+\quartocalloutnte{nte-train-final-run-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Plausibility}\label{plausibility-3}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-grid-train-plaus-circles} to
+Figure~\ref{fig-grid-train-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-circles}Average outcomes for the
+plausibility measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-lin_sep}Average outcomes for the
+plausibility measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-moons}Average outcomes for the
+plausibility measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-over}Average outcomes for the
+plausibility measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Cost}\label{cost-2}
+
+The results with respect to the cost measure are shown in
+Figure~\ref{fig-grid-train-cost-circles} to
+Figure~\ref{fig-grid-train-cost-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-circles}Average outcomes for the
+cost measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-lin_sep}Average outcomes for the
+cost measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-moons}Average outcomes for the cost
+measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-over}Average outcomes for the cost
 measure across hyperparameters. Data: Overlapping.}

### 2025-03-07 paper/sections/other/grids.qmd
more work on app

@@
 ```
+
+### Training Parameters
+
+```{julia}
+grid_dir = joinpath(res_dir, "training_params/mlp")
+data_dirs = readdir(grid_dir) |> x -> joinpath.(grid_dir, x) |> x -> x[isdir.(x)]
+data_names = basename.(data_dirs)
+suffix = "evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/"
+```
+
+The hyperparameter grid with other varying training parameters is shown in @nte-train-final-run-train.
+The corresponding evaluation grid used for these experiments is shown in @nte-train-final-run-eval.
+
+::: {#nte-train-final-run-train .callout-note}
+
+## Training Phase
+
+```{julia}
+#| output: asis
+dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/grid_config.toml")) 
+println(CTExperiments.dict_to_quarto_markdown(dict))
+```
+
+:::
+
+::: {#nte-train-final-run-eval .callout-note}
+
+## Evaluation Phase
+
+```{julia}
+#| output: asis
+dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/evaluation/evaluation_grid_config.toml"))
+println(CTExperiments.dict_to_quarto_markdown(dict))
+```
+
+:::
+
+#### Plausibility
+
+```{julia}
+#| output: asis
+
+fig_label_prefix = "grid-train-plaus"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the plausibility measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "plausibility_distance_from_target.png"
+fig_caption = "Average outcomes for the plausibility measure across hyperparameters."
+full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```
+
+#### Cost
+
+```{julia}
+#| output: asis
+
+fig_label_prefix = "grid-train-cost"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the cost measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "distance.png"
+fig_caption = "Average outcomes for the cost measure across hyperparameters."
+full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption)
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```

### 2025-03-09 paper/preprint/paper.tex
rerendered

@@
 \theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
 \theoremstyle{remark}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note}, opacityback=0, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
 
@@
   \item
-    \(\lambda_{\text{reg}}\): \texttt{0.0,\ 0.01,\ 0.1}
+    \(\lambda_{\text{reg}}\): \texttt{0.0,\ 0.01,\ 0.1,\ 0.25,\ 0.5}
   \item
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colbacktitle=quarto-callout-note-color!10!white, colback=white, breakable, left=2mm, coltitle=black, rightrule=.15mm, bottomtitle=1mm, toprule=.15mm, toptitle=1mm, titlerule=0mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
 
@@
 across multiple central processing units (CPU). Graphical processing
-units (GPU) were not required. For larger grid searches we used up to
-but typically less than 100 CPUs.
+units (GPU) were not required. Model training for the largest grid
+searches with 270 unique parameter combinations was parallelized across
+34 CPUs with 2GB memory each. The time to completion varied by dataset
+for reasons discussed in Section~\ref{sec-discussion}: 0h49m
+(\emph{Moons}), 1h4m (\emph{Linearly Separable}), 1h49m
+(\emph{Circles}), 3h52m (\emph{Overlapping}). Model evaluations for
+large grid searches were parallelized across 20 CPUs with 3GB memory
+each. Evaluations for all data sets took less than one hour
+(\textless1h) to complete.
 

### 2025-03-09 paper/preprint/paper.tex
okk

@@
 \usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{plain}
 \newtheorem{corollary}{Research Question}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
 \theoremstyle{remark}
@@
 
-\setcounter{section}{0}
 \renewcommand{\thesection}{\Alph{section}}
 
-\setcounter{table}{0}
 \renewcommand{\thetable}{A\arabic{table}}
 
-\setcounter{figure}{0}
 \renewcommand{\thefigure}{A\arabic{figure}}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note}]
 
@@
 
-\subsubsection{Generator Parameters}\label{generator-parameters}
+To assess the hyperparameter sensitivity of our proposed training regime
+we ran multiple large grid searches for all of our synthetic datasets.
+We have grouped these grid searches into multiple categories:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \textbf{Generator Parameters} (Section~\ref{sec-app-grid-gen}):
+  Investigates the effect of changing hyperparameters that affect the
+  counterfactual outcomes during the training phase.
+\item
+  \textbf{Penalty Strengths} (Section~\ref{sec-app-grid-pen}):
+  Investigates the effect of changing the penalty strengths in out
+  proposed objective (Equation~\ref{eq-obj}).
+\item
+  \textbf{Other Parameters} (Section~\ref{sec-app-grid-train}):
+  Investigates the effect of changing other training parameters,
+  including the total number of generated counterfactuals in each epoch.
+\end{enumerate}
+
+We begin by summarizing the high-level findings in
+Section~\ref{sec-app-grid-hl}. For each of the categories,
+Section~\ref{sec-app-grid-gen} to Section~\ref{sec-app-grid-train} then
+present all details including the parameter grids, average performance
+outcomes and key evaluation metrics for the generated counterfactuals.
+
+\subsubsection{High-Level Findings}\label{sec-app-grid-hl}
+
+Overall, we find that Counterfactual Training (CT) achieves it key
+objectives consistently across all hyperparameter settings and also
+broadly across datasets. We do observe strong sensitivity to certain
+hyperparameters, but clear and manageable patterns emerge in those
+cases. With respect to the underlying training data distribution, we
+find that outcomes for CT seem to depend to some degree on class
+separability.
+
+\paragraph{Predictive Performance}\label{predictive-performance}
+
+We find that CT is associated with little to no decrease in average
+predictive performance for our synthetic datasets: test accuracy and
+F1-scores decrease by at most \textasciitilde1 percentage point, but
+generally much less (Table~\ref{tbl-acc-gen}, Table~\ref{tbl-acc-pen},
+Table~\ref{tbl-acc-train}).
+
+\paragraph{Key Counterfactual
+Outcomes}\label{key-counterfactual-outcomes}
+
+\subsubsection{Generator Parameters}\label{sec-app-grid-gen}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]
 
@@
 
+\paragraph{Accuracy}\label{accuracy}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-acc-gen}Predictive performance measures by dataset
+and objective averaged across training-phase parameters
+(Note~\ref{nte-gen-params-final-run-train}) and evaluation-phase
+parameters (Note~\ref{nte-gen-params-final-run-eval}).}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Circles & Accuracy & Full & 0.997 & 0.00309 \\
+  Circles & Accuracy & Vanilla & 0.998 & 0.000557 \\
+  Circles & F1-score & Full & 0.997 & 0.00309 \\
+  Circles & F1-score & Vanilla & 0.998 & 0.000558 \\
+  Lin Sep & Accuracy & Full & 0.999 & 0.00201 \\
+  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
+  Lin Sep & F1-score & Full & 0.999 & 0.00201 \\
+  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Moons & Accuracy & Full & 0.999 & 0.000696 \\
+  Moons & Accuracy & Vanilla & 1 & 0.00111 \\
+  Moons & F1-score & Full & 0.999 & 0.000696 \\
+  Moons & F1-score & Vanilla & 1 & 0.00111 \\
+  Over & Accuracy & Full & 0.915 & 0.00477 \\
+  Over & Accuracy & Vanilla & 0.917 & 0.00123 \\
+  Over & F1-score & Full & 0.915 & 0.00478 \\
+  Over & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule
+
+\end{longtable}
+
 \paragraph{Plausibility}\label{plausibility-1}
@@
 
-\subsubsection{Penalty Strengths}\label{penalty-strengths}
+\subsubsection{Penalty Strengths}\label{sec-app-grid-pen}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}]
 
@@
 
+\paragraph{Accuracy}\label{accuracy-1}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-acc-pen}Predictive performance measures by dataset
+and objective averaged across training-phase parameters
+(Note~\ref{nte-pen-final-run-train}) and evaluation-phase parameters
+(Note~\ref{nte-pen-final-run-eval}).}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Circles & Accuracy & Full & 0.994 & 0.0144 \\
+  Circles & Accuracy & Vanilla & 0.998 & 0.000875 \\
+  Circles & F1-score & Full & 0.994 & 0.0145 \\
+  Circles & F1-score & Vanilla & 0.998 & 0.000875 \\
+  Lin Sep & Accuracy & Full & 0.998 & 0.00772 \\
+  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
+  Lin Sep & F1-score & Full & 0.998 & 0.00773 \\
+  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Moons & Accuracy & Full & 0.987 & 0.0351 \\
+  Moons & Accuracy & Vanilla & 0.998 & 0.0101 \\
+  Moons & F1-score & Full & 0.987 & 0.0352 \\
+  Moons & F1-score & Vanilla & 0.998 & 0.0102 \\
+  Over & Accuracy & Full & 0.911 & 0.0217 \\
+  Over & Accuracy & Vanilla & 0.916 & 0.00236 \\
+  Over & F1-score & Full & 0.911 & 0.0219 \\
+  Over & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule
+
+\end{longtable}
+
 \paragraph{Plausibility}\label{plausibility-2}
@@
 
-\subsubsection{Training Parameters}\label{training-parameters}
+\subsubsection{Other Parameters}\label{sec-app-grid-train}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}]
 
@@
 \end{tcolorbox}
+
+\paragraph{Accuracy}\label{accuracy-2}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-acc-train}Predictive performance measures by dataset
+and objective averaged across training-phase parameters
+(Note~\ref{nte-train-final-run-train}) and evaluation-phase parameters
+(Note~\ref{nte-train-final-run-eval}).}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Circles & Accuracy & Full & 0.995 & 0.00431 \\
+  Circles & Accuracy & Vanilla & 0.998 & 0.000566 \\
+  Circles & F1-score & Full & 0.995 & 0.00432 \\
+  Circles & F1-score & Vanilla & 0.998 & 0.000566 \\
+  Lin Sep & Accuracy & Full & 0.999 & 0.00231 \\
+  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
+  Lin Sep & F1-score & Full & 0.999 & 0.00231 \\
+  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Moons & Accuracy & Full & 0.996 & 0.0136 \\
+  Moons & Accuracy & Vanilla & 0.988 & 0.022 \\
+  Moons & F1-score & Full & 0.996 & 0.0136 \\
+  Moons & F1-score & Vanilla & 0.988 & 0.022 \\
+  Over & Accuracy & Full & 0.914 & 0.00563 \\
+  Over & Accuracy & Vanilla & 0.918 & 0.00116 \\
+  Over & F1-score & Full & 0.914 & 0.0057 \\
+  Over & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule
+
+\end{longtable}
 

### 2025-03-09 paper/sections/other/grids.qmd
okk

@@
 
-### Generator Parameters
+To assess the hyperparameter sensitivity of our proposed training regime we ran multiple large grid searches for all of our synthetic datasets.
+We have grouped these grid searches into multiple categories:
+
+1. **Generator Parameters** (@sec-app-grid-gen): Investigates the effect of changing hyperparameters that affect the counterfactual outcomes during the training phase.
+2. **Penalty Strengths** (@sec-app-grid-pen): Investigates the effect of changing the penalty strengths in out proposed objective (@eq-obj).
+3. **Other Parameters** (@sec-app-grid-train): Investigates the effect of changing other training parameters, including the total number of generated counterfactuals in each epoch.
+
+We begin by summarizing the high-level findings in @sec-app-grid-hl.
+For each of the categories, @sec-app-grid-gen to @sec-app-grid-train then present all details including the parameter grids, average performance outcomes and key evaluation metrics for the generated counterfactuals.
+
+### High-Level Findings {#sec-app-grid-hl}
+
+Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets.
+We do observe strong sensitivity to certain hyperparameters, but clear and manageable patterns emerge in those cases.
+With respect to the underlying training data distribution, we find that outcomes for CT seem to depend to some degree on class separability.
+
+#### Predictive Performance
+
+We find that CT is associated with little to no decrease in average predictive performance for our synthetic datasets: test accuracy and F1-scores decrease by at most ~1 percentage point, but generally much less (@tbl-acc-gen, @tbl-acc-pen, @tbl-acc-train).
+
+#### Key Counterfactual Outcomes
+
+### Generator Parameters {#sec-app-grid-gen}
 
@@
 data_dirs = readdir(grid_dir) |> x -> joinpath.(grid_dir, x) |> x -> x[isdir.(x)]
+eval_grids = (p -> EvaluationGrid(joinpath(p,"grid_config.toml"))).(data_dirs)
 data_names = basename.(data_dirs)
@@
 ```
+
+:::
+
+#### Accuracy
+
+::: {#tbl-acc-gen}
+
+::: {.content-hidden unless-format="pdf"}
+
+```{julia}
+#| output: asis
+df = CTExperiments.aggregate_performance(eval_grids; byvars=["objective"]) 
+get_table_inputs(df, "mean";backend=Val(:latex)) |>
+    inputs -> tabulate_results(inputs; wrap_table=false)
+```
+
+:::
+
+Predictive performance measures by dataset and objective averaged across training-phase parameters (@nte-gen-params-final-run-train) and evaluation-phase parameters (@nte-gen-params-final-run-eval).
 
@@
 
-### Penalty Strengths
+### Penalty Strengths {#sec-app-grid-pen}
 
@@
 ```
+
+:::
+
+#### Accuracy
+
+::: {#tbl-acc-pen}
+
+::: {.content-hidden unless-format="pdf"}
+
+```{julia}
+#| output: asis
+df = CTExperiments.aggregate_performance(eval_grids; byvars=["objective"]) 
+get_table_inputs(df, "mean";backend=Val(:latex)) |>
+    inputs -> tabulate_results(inputs; wrap_table=false)
+```
+
+:::
+
+Predictive performance measures by dataset and objective averaged across training-phase parameters (@nte-pen-final-run-train) and evaluation-phase parameters (@nte-pen-final-run-eval).
 
@@
 
-### Training Parameters
+### Other Parameters {#sec-app-grid-train}
 
@@
 ```
+
+:::
+
+#### Accuracy
+
+::: {#tbl-acc-train}
+
+::: {.content-hidden unless-format="pdf"}
+
+```{julia}
+#| output: asis
+df = CTExperiments.aggregate_performance(eval_grids; byvars=["objective"]) 
+get_table_inputs(df, "mean";backend=Val(:latex)) |>
+    inputs -> tabulate_results(inputs; wrap_table=false)
+```
+
+:::
+
+Predictive performance measures by dataset and objective averaged across training-phase parameters (@nte-train-final-run-train) and evaluation-phase parameters (@nte-train-final-run-eval).
 

### 2025-03-09 paper/experiments/CTExperiments/src/presentation.jl

@@ -146,10 +146,34 @@ end
 
 """
     aggregate_performance(
-        cfg::EvaluationConfig;
-        y::String="acc_val",
-        byvars::Union{Nothing,String,Vector{String}}=nothing,
-    )
+        eval_grids::Vector{EvalConfigOrGrid}; byvars=["objective"], kwrgs...
+    )   
+
+Aggregate performance measures across multiple experiments.
+"""
+function aggregate_performance(
+    eval_grids::Vector{<:EvalConfigOrGrid}; byvars=["objective"], kwrgs...
+)   
+    @assert "objective" in byvars "Need to specify 'objective' as a byvar to aggregate performance measures."
+
+    df = Logging.with_logger(Logging.NullLogger()) do 
+        map(eval_grids) do cfg
+            df = CTExperiments.aggregate_performance(cfg; byvars=byvars, kwrgs...)
+            exper_grid = ExperimentGrid(cfg.grid_file)
+            df.dataset .= CTExperiments.format_header(exper_grid.data)
+            df.objective .= CTExperiments.format_header.(df.objective)
+            keyvars = [:dataset, :variable, :objective]
+            select!(df, keyvars, Not(keyvars))
+            sort!(df, keyvars)
+            return df
+        end |>
+            dfs -> reduce(vcat, dfs) 
+    end
+    return df
+end
+
+"""
+    aggregate_performance(cfg::EvalConfigOrGrid; kwrgs...)
 
 Aggregate performance variable `y` from an experiment grid by columns specified in `byvars`.
 """
@@ -166,12 +190,10 @@ function aggregate_performance(
     df::DataFrame,
     df_meta::DataFrame,
     df_perf::DataFrame;
-    y::String="accuracy",
+    y::Union{Nothing,String}=nothing,
     byvars::Union{Nothing,String,Vector{String}}=nothing,
 )
     # Assertions:
-    valid_y = valid_y_perf(df_perf)
-    @assert y in valid_y "Variable `y` must be one of the following: $valid_y."
     if isa(byvars, String)
         byvars = [byvars]
     end
@@ -179,8 +201,14 @@ function aggregate_performance(
 
     # Aggregate data:
     df = aggregate_data(df, "value", byvars; byvars_must_include=["variable"])
-    df = df[df.variable .== y,:]
-    return select!(df,Not(:variable))
+    if !isnothing(y)
+        valid_y = valid_y_perf(df_perf)
[…]

### 2025-03-10 paper/preprint/paper.tex

@@
 \theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{plain}
 \newtheorem{proposition}{Proposition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note}]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note}, rightrule=.15mm, titlerule=0mm]
 
@@
 generally much less (Table~\ref{tbl-acc-gen}, Table~\ref{tbl-acc-pen},
-Table~\ref{tbl-acc-train}).
+Table~\ref{tbl-acc-train}). Variation across hyperparameters is
+negligible as indicated by small standard deviations for these metrics
+across the board.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, opacitybacktitle=0.6, toptitle=1mm, toprule=.15mm, leftrule=.75mm, rightrule=.15mm, coltitle=black, breakable, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, left=2mm, arc=.35mm, opacityback=0, bottomtitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]
 

### 2025-03-10 paper/sections/experiments.qmd
try that again

@@
 
+### Evaluation
+
+Our key outcome of interest is how well models perform with respect to explainability (@def-explainability): to this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
+To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.
+For plausibility, we assess how similar counterfactuals are to observed samples in the target domain.
+We rely on previously used distance-based metrics [@giudot]
+
+We also assess their predictive performance using conventional performance metrics including accuracy and the F1-score.
+
 ## Experimental Results

### 2025-03-10 paper/sections/other/grids.qmd
work on paper
try that again

@@
 We begin by summarizing the high-level findings in @sec-app-grid-hl.
-For each of the categories, @sec-app-grid-gen to @sec-app-grid-train then present all details including the parameter grids, average performance outcomes and key evaluation metrics for the generated counterfactuals.
+For each of the categories, @sec-app-grid-gen to @sec-app-grid-train then present all details including the exact parameter grids, average predictive performance outcomes and key evaluation metrics for the generated counterfactuals.
+
+### Evaluation Details
+
+To measure predictive performance, we compute the accuracy and F1-score for all models on test data (@tbl-acc-gen, @tbl-acc-pen, @tbl-acc-train).
+
+We evaluated
 
@@
 
-Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets.
-We do observe strong sensitivity to certain hyperparameters, but clear and manageable patterns emerge in those cases.
+Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved  We do observe strong sensitivity to certain hyperparameters, but clear and manageable patterns emerge in those cases.
 With respect to the underlying training data distribution, we find that outcomes for CT seem to depend to some degree on class separability.
@@
 We find that CT is associated with little to no decrease in average predictive performance for our synthetic datasets: test accuracy and F1-scores decrease by at most ~1 percentage point, but generally much less (@tbl-acc-gen, @tbl-acc-pen, @tbl-acc-train).
+Variation across hyperparameters is negligible as indicated by small standard deviations for these metrics across the board.
 

### 2025-03-10 paper/preprint/paper.tex

--- 2025-03-09 paper/preprint/paper.tex
@@
 \makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
-\definecolor{quarto-callout-color}{HTML}{909090}
-\definecolor{quarto-callout-note-color}{HTML}{0758E5}
-\definecolor{quarto-callout-important-color}{HTML}{CC1914}
-\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
-\definecolor{quarto-callout-tip-color}{HTML}{00A047}
-\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
-\definecolor{quarto-callout-color-frame}{HTML}{acacac}
-\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
-\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
-\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
-\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
-\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
-\makeatother
-\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \newtheorem{corollary}{Research Question}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{definition}
@@
 \newtheorem{definition}{Definition}[section]
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{remark}
@@
 \makeatother
-\newcounter{quartocalloutnteno}
-\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
 
@@
 relying on features and rules that are carefully hand-crafted by humans,
-modern machine learning (ML) models are tasked with learning these
-representations from scratch, guided by narrow objectives such as
+modern machine learning (ML) models are tasked with learning the
+representations directly from data, guided by narrow objectives such as
 predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
 Bengio, and Courville 2016}). Modern advances in computing have made it
-possible to provide such models with ever greater degrees of freedom to
+possible to provide such models with ever growing degrees of freedom to
 achieve that task, which has often led them to outperform traditionally
-more parsimonious models. Unfortunately, in doing so they also learn
+more parsimonious models. Unfortunately, in doing so, the models learn
 increasingly complex and highly sensitive representations that we can no
@@
 
-This trend towards complexity for the sake of performance has come under
+The trend towards complexity for the sake of performance has come under
 serious scrutiny in recent years. At the very cusp of the deep learning
@@
 different model predictions despite being ``imperceptible'' in that they
-are semantically indifferent from their factual counterparts. Despite
-partially effective mitigation strategies such as \textbf{adversarial
-training} (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow,
-Shlens, and Szegedy 2014}), truly robust deep learning (DL) remains
-unattainable even for models that are considered shallow by today's
-standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).
-
-Part of the problem is that high degrees of freedom provide room for
+are semantically indifferent from their factual counterparts. Although
+some partially effective mitigation strategies have been proposed, for
+example \textbf{adversarial training}
+(\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow, Shlens, and
+Szegedy 2014}), truly robust deep learning (DL) remains unattainable
+even for models that are considered shallow by today's standards
+(\citeproc{ref-kolter2023keynote}{Kolter 2023}).
+
+Part of the problem is that the high degrees of freedom provide room for
 many solutions that are locally optimal with respect to narrow
-objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
-  clarity: we follow standard ML convention in using ``degrees of
-  freedom'' to refer to the number of parameters estimated from data.}.
-Based purely on predictive performance, these solutions may seem to
-provide compelling explanations for the data, when in fact they are
-based on purely associative, semantically meaningless patterns. This
-poses two related challenges: firstly, it makes these models inherently
-opaque, since humans cannot simply interpret what type of explanation
-the complex learned representations correspond to; secondly, even if we
-could resolve the first challenge, it is not obvious how to mitigate
-models from learning representations that correspond to meaningless and
-implausible explanations.
+objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
+  follow the standard ML convention, where ``degrees of freedom'' refer
+  to the number of parameters estimated from data.} Indeed, recent work
+on the so called ``lottery tickets'' suggests that modern neural
+networks can be pruned by up to 90\% while preserving their predictive
+performance (\citeproc{ref-frankle2018lottery}{Frankle and Carbin 2019})
+and generalizability (\citeproc{ref-morcos2019success}{Morcos et al.
+2019}). Similarly, Zhang et al.
+(\citeproc{ref-zhang2021understanding}{2021}) showed that
+state-of-the-art neural networks are so expressive that they can fit
+randomly labeled data. Thus, looking at the predictive performance, the
+solutions may seem to provide compelling explanations for the data, when
+in fact they are based on purely associative, semantically meaningless
+patterns. This poses two related challenges. Firstly, there is no
+dependable way to verify if such complex representations correspond to
+meaningful and plausible explanations. Secondly, even if we could
+resolve the first challenge, it remains undecided how to ensure that
+models can \emph{only} learn valuable explanations.
 
@@
 scenario in which we deploy opaque models and blindly rely on their
-predictions. On countless occasions, this scenario has already occurred
-in practice and caused real harm to people who were affected adversely
-and often unfairly by automated decision-making systems (ADMS) involving
-opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective
-XAI tools can aide us in monitoring models and providing recourse to
-individuals to turn adverse outcomes (e.g.~``loan application
-rejected'') into positive ones (``application accepted''). Wachter,
-Mittelstadt, and Russell
+predictions. On countless occasions, this scenario has occurred in
+practice and caused real harm to people who were affected adversely and
+often unfairly by automated decision-making (ADM) systems involving
+opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016};
+\citeproc{ref-mcgregor2021preventing}{McGregor 2021}). Effective XAI
+tools can aid us in monitoring models and providing recourse to
+individuals to turn adverse outcomes (e.g., ``loan application
+rejected'') into positive ones (e.g., ``application accepted'').
+Wachter, Mittelstadt, and Russell
 (\citeproc{ref-wachter2017counterfactual}{2017}) propose
-\textbf{counterfactual explanations} as an effective approach to achieve
-this: they explain how factual inputs need to change in order for some
-fitted model to produce some desired output, typically involving minimal
-perturbations.
-
-To our surprise, the second challenge has not yet attracted any
+\textbf{counterfactual explanations} (CE) as an effective approach to
+achieve this goal: CEs explain how factual inputs need to change in
+order for some fitted model to produce some desired output, typically
+involving minimal perturbations.
+
+To our surprise, the second challenge has not yet attracted any major
 consolidated research effort. Specifically, there has been no concerted
-effort towards improving model \textbf{explainability}, which we define
-here as the degree to which learned representations correspond to
-explanations that are interpretable and deemed \textbf{plausible} by
-humans (see Definition~\ref{def-explainability}). Instead, the choice
-has typically been to improve the capacity of XAI tools to identify the
-subset explanations that are both plausible and valid for any given
-model, independent of whether the learned representations are also
-compatible with implausible explanations
+effort towards improving improving models' explanatory capacity, which
+we will henceforth simply call ``explainability'', defined as the degree
+to which learned representations correspond to explanations that are
+interpretable and deemed \textbf{plausible} by humans (see
+Definition~\ref{def-explainability}). Instead, the choice has typically
+been to improve the ability of XAI tools to identify the subset
+explanations that are both plausible and valid for any given model,
+independent of whether the learned representations are also compatible
+with implausible explanations
 (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
@@
 byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization and generative capacity Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}).
+robustness, generalization, and generative capacity
+(\citeproc{ref-schut2021generating}{Schut et al. 2021};
+\citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein 2020};
+\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
 
 Building on these findings, we introduce \textbf{counterfactual
-training}: a novel regularization technique geared explicitly towards
-aligning model representations with plausible explanations. Our
-contributions are as follows:
+training}: a novel training regime explicitly geared towards aligning
+model representations with plausible explanations. Our contributions are
+as follows:
 
@@
 
-Despite limitations of our approach discussed in
-Section~\ref{sec-discussion}, we conclude that counterfactual training
-provides a practical framework for researchers and practitioners
-interested in making opaque models more trustworthy
-Section~\ref{sec-conclusion}. We also believe that this work serves as
-an opportunity for XAI researchers to reevaluate the premise of
-improving XAI tools without improving models.
+Despite some limitations of our approach discussed in
+Section~\ref{sec-discussion}, we conclude in
+Section~\ref{sec-conclusion} that counterfactual training provides a
+practical framework for researchers and practitioners interested in
+making opaque models more trustworthy. We also believe that this work
+serves as an opportunity for XAI researchers to re-evaluate the trend of
+improving XAI tools without improving the underlying models.
 
@@
 
-To the best of our knowledge, our proposed framework for counterfactual
+To the best of our knowledge, our proposed framework of counterfactual
 training represents the first attempt to use counterfactual explanations
@@
 concrete, we follow Augustin, Meinke, and Hein
-(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
+(\citeproc{ref-augustin2020adversarial}{2020}) in tying the concept of
 explainability to the quality of counterfactual explanations that we can
@@
 freedom: by inducing models to ``unlearn'' representations that are
-susceptible to worst-case counterfactuals (i.e.~adversarial examples),
+susceptible to worst-case counterfactuals (i.e., adversarial examples),
 AT effectively removes some implausible explanations from the solution
@@
 they argue that counterfactual pairs typically already exist in training
-datasets. Specifically, their approach relies on, firstly, identifying
-similar input samples with different annotations and, secondly, ensuring
-that the gradient of the classifier aligns with the vector between pairs
-of counterfactual inputs using the cosine distance as a loss function.
+datasets. Specifically, their approach relies on identifying similar
+input samples with different annotations and ensuring that the gradient
+of the classifier aligns with the vector between such pairs of
+counterfactual inputs using the cosine distance as the loss function.
 
@@
 There have also been several attempts at formalizing the relationship
-between counterfactual explanations (CE) and adversarial examples (AE).
+between counterfactual explanations and adversarial examples (AE).
 Pointing to clear similarities in how CE and AE are generated,
@@
 case for jointly studying the opaqueness and robustness problem in
-representation learning. Formally, AE can be seen as the subset of CE,
+representation learning. Formally, AE can be seen as the subset of CE
 for which misclassification is achieved
@@
 for individuals to move from an adverse outcome to some positive target
-class with high probability. The approach proposed by Ross, Lakkaraju,
-and Bastani (\citeproc{ref-ross2021learning}{2024}) builds on
-adversarial training, where in this context susceptibility to targeted
-adversarial examples for the positive class is explicitly induced. The
-proposed method allows for imposing a set of actionability constraints
-ex-ante: for example, users can specify that certain features
-(e.g.~\emph{age}, \emph{gender}, \ldots) are immutable. Secondly, Guo,
-Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the first
-to propose an end-to-end training pipeline that includes counterfactual
-explanations as part of the training procedure. In particular, they
-propose a specific network architecture that includes a predictor and CE
-generator network, where the parameters of the CE generator network are
-learnable. Counterfactuals are generated during each training iteration
-and fed back to the predictor network. In contrast to Guo, Nguyen, and
-Yadav (\citeproc{ref-guo2023counternet}{2023}), we impose no
-restrictions on the neural network architecture at all.
+class with high probability. Their approach builds on adversarial
+training, where in this context susceptibility to targeted adversarial
+examples for the positive class is explicitly induced. The proposed
+method allows for imposing a set of actionability constraints ex-ante:
+for example, users can specify that certain features (e.g., \emph{age},
+\emph{gender}, \ldots) are immutable. Secondly, Guo, Nguyen, and Yadav
+(\citeproc{ref-guo2023counternet}{2023}) are the first to propose an
+end-to-end training pipeline that includes counterfactual explanations
+as part of the training procedure. In particular, they propose a
+specific network architecture that includes a predictor and CE generator
+network, where the parameters of the CE generator network are learnable.
+Counterfactuals are generated during each training iteration and fed
+back to the predictor network. In contrast to Guo, Nguyen, and Yadav
+(\citeproc{ref-guo2023counternet}{2023}), we impose no restrictions on
+the neural network architecture at all.
 
@@
 explicit objective of aligning representations with plausible
-explanations that comply with user requirements. In the context of CE,
+explanations that comply with user requirements. In the context of CEs,
 plausibility has broadly been defined as the degree to which
@@
 \citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
-is a necessary but insufficient condition for using CE to provide
+is a necessary but insufficient condition for using CEs to provide
 algorithmic recourse (AR) to individuals affected by opaque models in
@@
 application would have been accepted, if only they were 20 years older.
-Ignoring all other features, this complies with the definition of
-plausibility if 40-year-old individuals are in fact more credit-worthy
+Ignoring all other features, this would comply with the definition of
+plausibility if 40-year-old individuals were in fact more credit-worthy
 on average than young adults. But of course this CE does not qualify for
 providing actionable recourse to the applicant since \emph{age} is not a
-mutable feature. For our intents and purposes, counterfactual training
-aims at improving model explainability by aligning models with
+(directly) mutable feature. For our intents and purposes, counterfactual
+training aims to improve model explainability by aligning models with
 counterfactuals that meet both desiderata, plausibility and
@@
 \item
-  (Actionability) Permutations \(\Delta\) are subject to actionability
-  constraints.
+  (Actionability) Permutations \(\Delta\) are subject to some
+  actionability constraints.
 \end{enumerate}
@@
 pre-determined threshold, \(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
-  detailed background information on gradient-based counterfactual
-  search and convergence see Section~\ref{sec-app-ce}.}.
+\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
+where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
+  background information on gradient-based counterfactual search and
+  convergence see \textbf{?@sec-app-ce}.}
 
@@
 Definition~\ref{def-explainability}, we propose to leverage
-counterfactuals in the following objective,
+counterfactuals in the following objective:
 
@@
 where \(\text{yloss}(\cdot)\) is any conventional classification loss
-that induces discriminative performance (e.g.~cross-entropy). The two
+that induces discriminative performance (e.g., cross-entropy). The two
 additional components in Equation~\ref{eq-obj} are explained in more
 detail below. For now, they can be sufficiently described as inducing
-explainability directly and indirectly by penalizing: 1) the contrastive
-divergence, \(\text{div}(\cdot)\), between counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, 2) the
+explainability directly and indirectly by penalizing: (1) the
+contrastive divergence, \(\text{div}(\cdot)\), between counterfactuals
+\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
 adversarial loss, \(\text{advloss}(.)\), with respect to nascent
@@
 contrastive divergence between samples drawn from
-\(p_\theta(\mathbf{x})\) and training observations, i.e.~samples from
+\(p_\theta(\mathbf{x})\) and training observations, i.e., samples from
 \(p(\mathbf{x})\).
@@
 counterfactuals get more plausible (Definition~\ref{def-explainability})
-over the cause of training, these two opposing effects gradually balance
-each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
+over the course of training, these two opposing effects gradually
+balance each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
 
@@
 descent in the feature space for optimization
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
-In this context, domain constraints can be imposed by simply projecting
-counterfactuals back to the specified domain, if the previous gradient
-step resulted in updated feature values that were out-of-domain.
-Mutability constraints can similarly be enforced by setting partial
-derivatives to zero to ensure that features are only mutated in the
-allowed direction, if at all.
+(\citeproc{ref-altmeyer2023explaining}{Altmeyer, van Deursen, and Liem
+2023}). In this context, domain constraints can be imposed by simply
+projecting counterfactuals back to the specified domain, if the previous
+gradient step resulted in updated feature values that were
+out-of-domain. Mutability constraints can similarly be enforced by
+setting partial derivatives to zero to ensure that features are only
+mutated in the allowed direction, if at all.
 
@@
 violate mutability constraints. Specifically, we set
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
 
@@
 Section~\ref{sec-experiments}. Under certain conditions, this results
-holds theoretically{[}For the proof, see the supplementary appendix.{]}:
+holds theoretically:\footnote{For the proof, see the supplementary
+  appendix.}
 
@@
 denote a linear classifier with softmax activation \(\mathcal{S}\)
-(i.e.~\emph{multinomial logistic regression}) where
+(i.e., \emph{multinomial logistic regression}) where
 \(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
@@
 (\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
-1) the subjects' \emph{age}, which we define as immutable, and 2) the
+(1) the subjects' \emph{age}, which we define as immutable, and (2) the
 subjects' existing level of \emph{debt}, which we define as mutable.
@@
 Counterfactuals are valid but involve some substantial reductions in
-\emph{debt} for some individuals, in particular very young applicants.
-By comparison, counterfactual paths are shorter on average in panel (d),
-where we have used counterfactual training and protected immutable
-features as described in Section~\ref{sec-constraints}. In particular,
-we observe that due to the classifier's lower sensitivity to \emph{age},
-recourse recommendations with respect to \emph{debt} are much more
-homogenous, in that they do not disproportionately punish younger
-individuals. The counterfactuals are also plausible with respect to the
-mutable feature. Thus, we consider the model in panel (d) as the most
-explainable according to Definition~\ref{def-explainability}.
+\emph{debt} for some individuals (very young applicants). By comparison,
+counterfactual paths are shorter on average in panel (d), where we have
+used counterfactual training and protected immutable features as
+described in Section~\ref{sec-constraints}. In particular, we observe
+that due to the classifier's lower sensitivity to \emph{age}, recourse
+recommendations with respect to \emph{debt} are much more homogenous, in
+that they do not disproportionately punish younger individuals. The
+counterfactuals are also plausible with respect to the mutable feature.
+Thus, we consider the model in panel (d) as the most explainable
+according to Definition~\ref{def-explainability}.
 
@@
 
-\bibitem[\citeproctext]{ref-altmeyer2023explaining}
-Altmeyer, Patrick, Arie van Deursen, et al. 2023. {``Explaining
-Black-Box Models Through Counterfactuals.''} In \emph{Proceedings of the
-JuliaCon Conferences}, 1:130. 1.
-
 \bibitem[\citeproctext]{ref-altmeyer2024faithful}
@@
 Conference on Artificial Intelligence}, 38:10829--37. 10.
+
+\bibitem[\citeproctext]{ref-altmeyer2023explaining}
+Altmeyer, Patrick, Arie van Deursen, and Cynthia C. S. Liem. 2023.
+{``Explaining Black-Box Models Through Counterfactuals.''} In
+\emph{Proceedings of the JuliaCon Conferences}, 1:130. 1.
 
@@
 
-\bibitem[\citeproctext]{ref-bezanson2017julia}
-Bezanson, Jeff, Alan Edelman, Stefan Karpinski, and Viral B Shah. 2017.
-{``Julia: A Fresh Approach to Numerical Computing.''} \emph{SIAM Review}
-59 (1): 65--98. \url{https://doi.org/10.1137/141000671}.
-
-\bibitem[\citeproctext]{ref-milan2023dataframes}
-Bouchet-Valat, Milan, and Bogumił Kamiński. 2023. {``DataFrames.jl:
-Flexible and Fast Tabular Data in Julia.''} \emph{Journal of Statistical
-Software} 107 (4): 1--32. \url{https://doi.org/10.18637/jss.v107.i04}.
-
-\bibitem[\citeproctext]{ref-byrne2021mpi}
-Byrne, Simon, Lucas C. Wilcox, and Valentin Churavy. 2021. {``MPI.jl:
-Julia Bindings for the Message Passing Interface.''} \emph{Proceedings
-of the JuliaCon Conferences} 1 (1): 68.
-\url{https://doi.org/10.21105/jcon.00068}.
-
-\bibitem[\citeproctext]{ref-chagas2024pretty}
-Chagas, Ronan Arraes Jardim, Ben Baumgold, Glen Hertz, Hendrik Ranocha,
-Mark Wells, Nathan Boyer, Nicholas Ritchie, et al. 2024.
-{``Ronisbr/PrettyTables.jl: V2.4.0.''} Zenodo.
-\url{https://doi.org/10.5281/zenodo.13835553}.
-
-\bibitem[\citeproctext]{ref-PlotsJL}
-Christ, Simon, Daniel Schwabeneder, Christopher Rackauckas, Michael
-Krabbe Borregaard, and Thomas Breloff. 2023. {``Plots.jl -- a User
-Extendable Plotting API for the Julia Programming Language.''}
-https://doi.org/\url{https://doi.org/10.5334/jors.431}.
-
-\bibitem[\citeproctext]{ref-danisch2021makie}
-Danisch, Simon, and Julius Krumbiegel. 2021. {``{Makie.jl}: Flexible
-High-Performance Data Visualization for {Julia}.''} \emph{Journal of
-Open Source Software} 6 (65): 3349.
-\url{https://doi.org/10.21105/joss.03349}.
-
 \bibitem[\citeproctext]{ref-du2019implicit}
@@
 \url{https://arxiv.org/abs/1903.08689}.
+
+\bibitem[\citeproctext]{ref-frankle2018lottery}
+Frankle, Jonathan, and Michael Carbin. 2019. {``The Lottery Ticket
+Hypothesis: Finding Sparse, Trainable Neural Networks.''} In
+\emph{International Conference on Learning Representations}.
+\url{https://openreview.net/forum?id=rJl-b3RcF7}.
 
@@
 
-\bibitem[\citeproctext]{ref-hastie2009elements}
-Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
-Elements of Statistical Learning}. Springer New York.
-\url{https://doi.org/10.1007/978-0-387-84858-7}.
-
-\bibitem[\citeproctext]{ref-innes2018fashionable}
-Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco
-Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and
-Viral Shah. 2018. {``Fashionable Modelling with Flux.''}
-\url{https://arxiv.org/abs/1811.01457}.
-
-\bibitem[\citeproctext]{ref-innes2018flux}
-Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
-\emph{Journal of Open Source Software} 3 (25): 602.
-\url{https://doi.org/10.21105/joss.00602}.
-
-\bibitem[\citeproctext]{ref-joshi2019realistic}
-Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
-Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
-Actionable Explanations in Black-Box Decision Making Systems.''}
-\url{https://arxiv.org/abs/1907.09615}.
-
 \bibitem[\citeproctext]{ref-kolter2023keynote}
@@
 Information and Computation}, 881--88.
+
+\bibitem[\citeproctext]{ref-mcgregor2021preventing}
+McGregor, Sean. 2021. {``{Preventing repeated real world AI failures by
+cataloging incidents: The AI incident database}.''} In \emph{Proceedings
+of the AAAI Conference on Artificial Intelligence}, 35:15458--63. 17.
+
+\bibitem[\citeproctext]{ref-morcos2019success}
+Morcos, Ari S., Haonan Yu, Michela Paganini, and Yuandong Tian. 2019.
+{``One Ticket to Win Them All: Generalizing Lottery Ticket
+Initializations Across Datasets and Optimizers.''} In \emph{Proceedings
+of the 33rd International Conference on Neural Information Processing
+Systems}. Red Hook, NY, USA: Curran Associates Inc.
 
@@
 
+\bibitem[\citeproctext]{ref-zhang2021understanding}
+Zhang, Chiyuan, Samy Bengio, Moritz Hardt, Benjamin Recht, and Oriol
+Vinyals. 2021. {``Understanding Deep Learning (Still) Requires
+Rethinking Generalization.''} \emph{Commun. ACM} 64 (3): 107--15.
+\url{https://doi.org/10.1145/3446776}.
+
 \end{CSLReferences}
@@
 
-\FloatBarrier
-
-\setcounter{section}{0}
-\renewcommand{\thesection}{\Alph{section}}
-
-\setcounter{table}{0}
-\renewcommand{\thetable}{A\arabic{table}}
-
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{A\arabic{figure}}
-
-\section{Notation}\label{notation}
-
-\begin{itemize}
-\tightlist
-\item
-  \(y^+\): The target class and also the index of the target class.
-\item
-  \(y^-\): The non-target class and also the index of non-the target
-  class.
-\item
-  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
-  class.
-\item
-  \(\theta\): Model parameters (unspecified).
-\item
-  \(\Theta\): Matrix of parameters.
-\end{itemize}
-
-\section{Technical Details of Our
-Approach}\label{technical-details-of-our-approach}
-
-\subsection{Generating Counterfactuals through Gradient
-Descent}\label{sec-app-ce}
-
-In this section, we provide some background on gradient-based
-counterfactual generators (Section~\ref{sec-app-ce-background}) and
-discuss how we define convergence in this context
-(Section~\ref{sec-app-conv}).
-
-\subsubsection{Background}\label{sec-app-ce-background}
-
-Gradient-based counterfactual search was originally proposed by Wachter,
-Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}). It generally solves
-the following unconstrained objective,
-
-\[
-\begin{aligned}
-\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
-\end{aligned} 
-\]
-
-where \(g: \mathcal{Z} \mapsto \mathcal{X}\) is an invertible function
-that maps from the \(L\)-dimensional counterfactual state space to the
-feature space and \(\text{cost}(\cdot)\) denotes one or more penalties
-that are used to induce certain properties of the counterfactual
-outcome. As above, \(\mathbf{y}^+\) denotes the target output and
-\(\mathbf{M}_{\theta}(\mathbf{x})\) returns the logit predictions of the
-underlying classifier for \(\mathbf{x}=g(\mathbf{z})\).
-
-For all generators used in this work we use standard logit crossentropy
-loss for \(\text{yloss}(\cdot)\). All generators also penalize the
-distance (\(\ell_1\)-norm) of counterfactuals from their original
-factual state. For \emph{Generic} and \emph{ECCo}, we have
-\(\mathcal{Z}:=\mathcal{X}\) and
-\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
-are searched directly in the feature space. Conversely, \emph{REVISE}
-traverses the latent space of a variational autoencoder (VAE) fitted to
-the training data, where \(g(\cdot)\) corresponds to the decoder
-(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
-the distance penalty, \emph{ECCo} uses an additional penalty component
-that regularizes the energy associated with the counterfactual,
-\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
-al. 2024}).
-
-\subsubsection{Convergence}\label{sec-app-conv}
-
-An important consideration when generating counterfactual explanations
-using gradient-based methods is how to define convergence. Two common
-choices are to 1) perform gradient descent over a fixed number of
-iterations \(T\), or 2) conclude the search as soon as the predicted
-probability for the target class has reached a pre-determined threshold,
-\(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\). We
-prefer the latter for our purposes, because it explicitly defines
-convergence in terms of the black-box model, \(\mathbf{M}(\mathbf{x})\).
-
-Defining convergence in this way allows for a more intuitive
-interpretation of the resulting counterfactual outcomes than with fixed
-\(T\). Specifically, it allows us to think of counterfactuals as
-explaining `high-confidence' predictions by the model for the target
-class \(y^+\). Depending on the context and application, different
-choices of \(\tau\) can be considered as representing `high-confidence'
-predictions.
-
-\subsection{Protecting Mutability Constraints with Linear
-Classifiers}\label{sec-app-constraints}
-
-In Section~\ref{sec-constraints} we explain that to avoid penalizing
-implausibility that arises due to mutability constraints, we impose a
-point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
-argue in Section~\ref{sec-constraints} that this approach induces models
-to be less sensitive to immutable features and demonstrate this
-empirically in Section~\ref{sec-experiments}. Below we derive the
-analytical results in Proposition~\ref{prp-mtblty}.
-
-\begin{proof}
-Let \(d_{\text{mtbl}}\) and \(d_{\text{immtbl}}\) denote some mutable
-and immutable feature, respectively. Suppose that
-\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
-\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
-\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
-class \(k\). In words, we assume that the immutable feature tends to
-take lower values for samples in the non-target class \(y^-\) than in
-the target class \(y^+\). We assume the opposite to hold for the mutable
-feature.
-
-Assuming multivariate Gaussian class densities with common diagonal
-covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), we
-have for the log likelihood ratio between any two classes
-\(k,m \in \mathcal{K}\) (\citeproc{ref-hastie2009elements}{Hastie,
-Tibshirani, and Friedman 2009}):
-
-\begin{equation}\phantomsection\label{eq-loglike}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
-}\end{equation}
-
-By independence of \(x_1,...,x_D\), the full log-likelihood ratio
-decomposes into:
-
-\begin{equation}\phantomsection\label{eq-loglike-decomp}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
-}\end{equation}
-
-By the properties of our classifier (\emph{multinomial logistic
-regression}), we have:
-
-\begin{equation}\phantomsection\label{eq-multi}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
-}\end{equation}
-
-where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
-\(d\) for class \(k\).
-
-Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
-can identify that
-\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
-the assumptions we made above. Hence, we have that
-\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
-and
-\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)
-
-Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
-class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
-target class. Then the partial derivative of the contrastive divergence
-penalty Equation~\ref{eq-div} with respect to coefficient
-\(\theta_{y^+,d}\) is equal to
-
-\begin{equation}\phantomsection\label{eq-grad}{
-\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
-}\end{equation}
-
-and equal to zero everywhere else.
-
-Since \((\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})\) we
-are more likely to have
-\((x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0\) than vice
-versa at initialization. Similarly, we are more likely to have
-\((x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0\) since
-\((\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})\).
-
-This implies that if we do not protect feature \(d_{\text{immtbl}}\),
-the contrastive divergence penalty will decrease
-\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
-effect
-\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
-In words, not protecting the immutable feature would have the
-undesirable effect of making the classifier more sensitive to this
-feature, in that it would be more likely to predict class \(y^-\) as
-opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).
-
-By the same rationale, the contrastive divergence penalty can generally
-be expected to increase \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
-\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
-In words, this has the effect of making the classifier more sensitive to
-the mutable feature, in that it would be more likely to predict class
-\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).
-
-Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
-has the net affect of decreasing the classifier's sensitivity to the
-immutable feature relative to the mutable feature (i.e.~no change in
-sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
-for \(d_{\text{mtbl}}\)).
-\end{proof}
-
-\subsection{Domain Constraints}\label{domain-constraints}
-
-We apply domain constraints on counterfactuals during training and
-evaluation. There are at least two good reasons for doing so. Firstly,
-within the context of explainability and algorithmic recourse,
-real-world attributes are often domain constrained: the \emph{age}
-feature, for example, is lower bounded by zero and upper bounded by the
-maximum human lifespan. Secondly, domain constraints help mitigate
-training instabilities commonly associated with energy-based modelling
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-
-For our image datasets, features are pixel values and hence the domain
-is constrained by the lower and upper bound of values that pixels can
-take depending on how they are scaled (in our case \([-1,1]\)). For all
-other features \(d\) in our synthetic and tabular datasets, we
-automatically infer domain constraints
-\([x_d^{\text{LB}},x_d^{\text{UB}}]\) as follows,
-
-\begin{equation}\phantomsection\label{eq-domain}{
-\begin{aligned}
-x_d^{\text{LB}} &= \arg\min_{x_d} \{\mu_d - n_{\sigma_d}\sigma_d, \arg \min_{x_d} x_d\} \\
-x_d^{\text{UB}} &= \arg\max_{x_d} \{\mu_d + n_{\sigma_d}\sigma_d, \arg \max_{x_d} x_d\} 
-\end{aligned}
-}\end{equation}
-
-where \(\mu_d\) and \(\sigma_d\) denote the sample mean and standard
-deviation of feature \(d\). We set \(n_{\sigma_d}=3\) across the board
-but higher values and hence wider bounds may be appropriate depending on
-the application.
-
-\subsection{Training Details}\label{sec-app-training}
-
-In this section, we describe the training procedure in detail. While the
-details laid out here are not crucial for understanding our proposed
-approach, they are of importance to anyone looking to implement
-counterfactual training.
-
-\section{Detailed Results}\label{detailed-results}
-
-\subsection{Qualitative Findings for Image
-Data}\label{qualitative-findings-for-image-data}
-
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
-
-Figure~\ref{fig-mnist} shows much more plausible (faithful)
-counterfactuals for a model with CT than the model with conventional
-training (Figure~\ref{fig-mnist-vanilla}). In fact, this is not even
-using \emph{ECCo+} and still showing better results than the best
-results we achieved in our AAAI paper for JEM ensembles.
-
-\end{tcolorbox}
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp.png}}
-
-}
-
-\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
-counterfactual training. The underlying generator, \emph{ECCo}, aims to
-generate counterfactuals that are faithful to the model
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp_vanilla.png}}
-
-}
-
-\caption{\label{fig-mnist-vanilla}Counterfactual images for \emph{MLP}
-with conventional training. The underlying generator, \emph{ECCo}, aims
-to generate counterfactuals that are faithful to the model
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
-
-\end{figure}%
-
-\subsection{Grid Searches}\label{sec-app-grid}
-
-\subsubsection{Generator Parameters}\label{generator-parameters}
-
-The hyperparameter grid with varying generator parameters during
-training is shown in Note~\ref{nte-gen-params-final-run-train}. The
-corresponding evaluation grid used for these experiments is shown in
-Note~\ref{nte-gen-params-final-run-eval}.
-
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
-
-\quartocalloutnte{nte-gen-params-final-run-train}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Decision Threshold: \texttt{0.75,\ 0.9,\ 0.95}
-  \item
-    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 5.0,\ 10.0,\ 20.0}
-  \item
-    Maximum Iterations: \texttt{5,\ 25,\ 50}
-  \end{itemize}
-\item
-  Generator: \texttt{ecco,\ generic,\ revise}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
-
-\quartocalloutnte{nte-gen-params-final-run-eval}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\paragraph{Plausibility}\label{plausibility-1}
-
-The results with respect to the plausibility measure are shown in
-Figure~\ref{fig-grid-gen_params-plaus-circles} to
-Figure~\ref{fig-grid-gen_params-plaus-over}.
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-gen_params-plaus-circles}Average outcomes for
-the plausibility measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-gen_params-plaus-lin_sep}Average outcomes for
-the plausibility measure across hyperparameters. Data: Linearly
-Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-gen_params-plaus-moons}Average outcomes for the
-plausibility measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-gen_params-plaus-over}Average outcomes for the
-plausibility measure across hyperparameters. Data: Overlapping.}
-
-\end{figure}%
-
-\paragraph{Cost}\label{cost}
-
-The results with respect to the cost measure are shown in
-Figure~\ref{fig-grid-gen_params-cost-circles} to
-Figure~\ref{fig-grid-gen_params-cost-over}.
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
-
-}
-
-\caption{\label{fig-grid-gen_params-cost-circles}Average outcomes for
-the cost measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
-
-}
-
-\caption{\label{fig-grid-gen_params-cost-lin_sep}Average outcomes for
-the cost measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
-
-}
-
-\caption{\label{fig-grid-gen_params-cost-moons}Average outcomes for the
-cost measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
-
-}
-
-\caption{\label{fig-grid-gen_params-cost-over}Average outcomes for the
-cost measure across hyperparameters. Data: Overlapping.}
-
-\end{figure}%
-
-\subsubsection{Penalty Strengths}\label{penalty-strengths}
-
-The hyperparameter grid with varying penalty strengths during training
-is shown in Note~\ref{nte-pen-final-run-train}. The corresponding
-evaluation grid used for these experiments is shown in
-Note~\ref{nte-pen-final-run-eval}.
-
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
-
-\quartocalloutnte{nte-pen-final-run-train}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator: \texttt{ecco,\ generic,\ revise}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{adv}}\): \texttt{0.1,\ 0.25,\ 1.0}
-  \item
-    \(\lambda_{\text{div}}\): \texttt{0.01,\ 0.1,\ 1.0}
-  \item
-    \(\lambda_{\text{reg}}\): \texttt{0.0,\ 0.01,\ 0.1,\ 0.25,\ 0.5}
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
-
-\quartocalloutnte{nte-pen-final-run-eval}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\paragraph{Plausibility}\label{plausibility-2}
-
-The results with respect to the plausibility measure are shown in
-Figure~\ref{fig-grid-pen-plaus-circles} to
-Figure~\ref{fig-grid-pen-plaus-over}.
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-pen-plaus-circles}Average outcomes for the
-plausibility measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-pen-plaus-lin_sep}Average outcomes for the
-plausibility measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-pen-plaus-moons}Average outcomes for the
-plausibility measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-pen-plaus-over}Average outcomes for the
-plausibility measure across hyperparameters. Data: Overlapping.}
-
-\end{figure}%
-
-\paragraph{Cost}\label{cost-1}
-
-The results with respect to the cost measure are shown in
-Figure~\ref{fig-grid-pen-cost-circles} to
-Figure~\ref{fig-grid-pen-cost-over}.
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
-
-}
-
-\caption{\label{fig-grid-pen-cost-circles}Average outcomes for the cost
-measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
-
-}
-
-\caption{\label{fig-grid-pen-cost-lin_sep}Average outcomes for the cost
-measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
-
-}
-
-\caption{\label{fig-grid-pen-cost-moons}Average outcomes for the cost
-measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
-
-}
-
-\caption{\label{fig-grid-pen-cost-over}Average outcomes for the cost
-measure across hyperparameters. Data: Overlapping.}
-
-\end{figure}%
-
-\subsubsection{Training Parameters}\label{training-parameters}
-
-The hyperparameter grid with other varying training parameters is shown
-in Note~\ref{nte-train-final-run-train}. The corresponding evaluation
-grid used for these experiments is shown in
-Note~\ref{nte-train-final-run-eval}.
-
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
-
-\quartocalloutnte{nte-train-final-run-train}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator: \texttt{ecco,\ generic,\ revise}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Burnin: \texttt{0.0,\ 0.5}
-  \item
-    No.~Counterfactuals: \texttt{100,\ 1000}
-  \item
-    No.~Epochs: \texttt{50,\ 100}
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, bottomrule=.15mm, coltitle=black, colback=white, titlerule=0mm, opacityback=0, rightrule=.15mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, toptitle=1mm, left=2mm]
-
-\quartocalloutnte{nte-train-final-run-eval}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\paragraph{Plausibility}\label{plausibility-3}
-
-The results with respect to the plausibility measure are shown in
-Figure~\ref{fig-grid-train-plaus-circles} to
-Figure~\ref{fig-grid-train-plaus-over}.
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-train-plaus-circles}Average outcomes for the
-plausibility measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-train-plaus-lin_sep}Average outcomes for the
-plausibility measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-train-plaus-moons}Average outcomes for the
-plausibility measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
-
-}
-
-\caption{\label{fig-grid-train-plaus-over}Average outcomes for the
-plausibility measure across hyperparameters. Data: Overlapping.}
-
-\end{figure}%
-
-\paragraph{Cost}\label{cost-2}
-
-The results with respect to the cost measure are shown in
-Figure~\ref{fig-grid-train-cost-circles} to
-Figure~\ref{fig-grid-train-cost-over}.
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
-
-}
-
-\caption{\label{fig-grid-train-cost-circles}Average outcomes for the
-cost measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
-
-}
-
-\caption{\label{fig-grid-train-cost-lin_sep}Average outcomes for the
-cost measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
-
-}
-
-\caption{\label{fig-grid-train-cost-moons}Average outcomes for the cost
-measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
-
-}
-
-\caption{\label{fig-grid-train-cost-over}Average outcomes for the cost
-measure across hyperparameters. Data: Overlapping.}
-
-\end{figure}%
-
-\section{Computation Details}\label{computation-details}
-
-\subsection{Hardware}\label{hardware}
-
-We performed our experiments on a high-performance cluster. Details
-about the cluster will be disclosed upon publication to avoid revealing
-information that might interfere with the double-blind review process.
-Since our experiments involve highly parallel tasks and rather small
-models by today's standard, we have relied on distributed computing
-across multiple central processing units (CPU). Graphical processing
-units (GPU) were not required. Model training for the largest grid
-searches with 270 unique parameter combinations was parallelized across
-34 CPUs with 2GB memory each. The time to completion varied by dataset
-for reasons discussed in Section~\ref{sec-discussion}: 0h49m
-(\emph{Moons}), 1h4m (\emph{Linearly Separable}), 1h49m
-(\emph{Circles}), 3h52m (\emph{Overlapping}). Model evaluations for
-large grid searches were parallelized across 20 CPUs with 3GB memory
-each. Evaluations for all data sets took less than one hour
-(\textless1h) to complete.
-
-\subsection{Software}\label{software}
-
-All computations were performed in the Julia Programming Language
-(\citeproc{ref-bezanson2017julia}{Bezanson et al. 2017}). We have
-developed a package for counterfactual training that leverages and
-extends the functionality provided by several existing packages, most
-notably
-\href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{CounterfactualExplanations.jl}
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023})
-and the \href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library for
-deep learning (\citeproc{ref-innes2018fashionable}{Michael Innes et al.
-2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). For
-data-wrangling and presentation-ready tables we relied on
-\href{https://dataframes.juliadata.org/v1.7/}{DataFrames.jl}
-(\citeproc{ref-milan2023dataframes}{Bouchet-Valat and Kamiński 2023})
-and
-\href{https://ronisbr.github.io/PrettyTables.jl/v2.4/}{PrettyTables.jl}
-(\citeproc{ref-chagas2024pretty}{Chagas et al. 2024}), respectively. For
-plots and visualizations we used both
-\href{https://docs.juliaplots.org/v1.40/}{Plots.jl}
-(\citeproc{ref-PlotsJL}{Christ et al. 2023}) and
-\href{https://docs.makie.org/v0.22/}{Makie.jl}
-(\citeproc{ref-danisch2021makie}{Danisch and Krumbiegel 2021}), in
-particular \href{https://aog.makie.org/v0.9.3/}{AlgebraOfGraphics.jl}.
-To distribute computational tasks across multiple processors, we have
-relied on \href{https://juliaparallel.org/MPI.jl/v0.20/}{MPI.jl}
-(\citeproc{ref-byrne2021mpi}{Byrne, Wilcox, and Churavy 2021}).
-
 \end{document}

### 2025-03-10 paper/sections/abstract.qmd

--- 2025-02-03 paper/sections/abstract.qmd
@@
 
-Counterfactual Explanations have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output.
-Much existing research has focused on identifying explanations that are not only valid but also deemed desirable with respect to the underlying data and stakeholder requirements.
-Recent work has shown that under this premise, the task of learning desirable explanations is effectively reassigned from the model itself to the (post-hoc) counterfactual explainer.
-Building on that work, we propose a novel model objective that leverages counterfactuals during the training phase (ad-hoc) in order to minimize the divergence between learned representations and desirable explanations.
-Through extensive experiments, we demonstrate that our proposed methodology facilitates training models that inherently deliver desirable explanations while maintaining high predictive performance.
+We propose a novel training regime called Counterfactual Training that leverages counterfactual explanations to increase the explanatory capacity of models.
+Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output.
+To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to stakeholder requirements.
+Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata.
+In this work, we instead hold models directly accountable for this desired end goal: Counterfactual Training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
+We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

### 2025-03-10 paper/sections/intro.qmd

--- 2025-02-03 paper/sections/intro.qmd
@@
 
-Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern AIs are tasked with learning these representations from scratch, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
-Modern advances in computing have made it possible to provide such AIs with ever greater degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models.
-Unfortunately, in doing so they also learn increasingly complex and highly sensitive representations that we can no longer easily interpret.
+Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning the representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
+Modern advances in computing have made it possible to provide such models with ever growing degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models.
+Unfortunately, in doing so, the models learn increasingly complex and highly sensitive representations that we can no longer easily interpret.
 
-This trend towards complexity for the sake of performance has come under serious scrutiny in recent years.
+The trend towards complexity for the sake of performance has come under serious scrutiny in recent years.
 At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: counterfactuals of model inputs that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts.
-Despite partially effective mitigation strategies such as **adversarial training** [@goodfellow2014explaining], truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].
+Although some partially effective mitigation strategies have been proposed, for example **adversarial training** [@goodfellow2014explaining], truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].
 
-Part of the problem is that high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].
-Based purely on predictive performance, these solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns.
-This poses two related challenges: firstly, it makes these models inherently opaque, since humans cannot simply interpret what type of explanation the complex learned representations correspond to; secondly, even if we could resolve the first challenge, it is not obvious how to mitigate models from learning representations that correspond to meaningless and undesirable explanations.
+Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery tickets" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success].
+Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data.
+Thus, looking at the predictive performance, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns.
+This poses two related challenges.
+Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations.
+Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.
 
@@
 This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions.
-On countless occasions, this scenario has already occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making systems (ADMS) involving opaque models [@oneil2016weapons].
-Effective XAI tools can aide us in monitoring models and providing recourse to individuals to turn adverse outcomes (e.g.
-"loan application rejected") into positive ones ("application accepted").
-@wachter2017counterfactual propose **counterfactual explanations** as an effective approach to achieve this: they explain how factual inputs need to change in order for some fitted model to produce some desired output, typically involving minimal perturbations.
+On countless occasions, this scenario has occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making (ADM) systems involving opaque models [@oneil2016weapons; @mcgregor2021preventing].
+Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn adverse outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted").
+@wachter2017counterfactual propose **counterfactual explanations** (CE) as an effective approach to achieve this goal: CEs explain how factual inputs need to change in order for some fitted model to produce some desired output, typically involving minimal perturbations.
 
-To our surprise, the second challenge has not yet attracted any consolidated research effort.
-Specifically, there has been no concerted effort towards improving model **explainability**, which we define here as the degree to which learned representations correspond to explanations that are interpretable and deemed desirable by humans.
-Instead, the choice has typically been to improve the capacity of XAI tools to identify the subset explanations that are both desirable and valid for any given model, independent of whether the learned representations are also compatible with undesirable explanations [@altmeyer2024faithful].
-Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization and generative capacity [@schut2021generating, @augustin2020adversarial, @altmeyer2024faithful].
+To our surprise, the second challenge has not yet attracted any major consolidated research effort.
+Specifically, there has been no concerted effort towards improving improving models' explanatory capacity, which we will henceforth simply call "explainability", defined as the degree to which learned representations correspond to explanations that are interpretable and deemed **plausible** by humans (see @def-explainability).
+Instead, the choice has typically been to improve the ability of XAI tools to identify the subset explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with implausible explanations [@altmeyer2024faithful].
+Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful].
 
-Building on these findings, we introduce **counterfactual training**: a novel regularization technique geared explicitly towards aligning model representations with desirable explanations.
+Building on these findings, we introduce **counterfactual training**: a novel training regime explicitly geared towards aligning model representations with plausible explanations.
 Our contributions are as follows:
@@
 
-Despite limitations of our approach discussed in @sec-discussion, we conclude that counterfactual training provides a practical framework for researchers and practitioners interested in making opaque models more trustworthy [@sec-conclusion].
-We also believe that this work serves as an opportunity for XAI researchers to reevaluate the premise of improving XAI tools without improving models.
+Despite some limitations of our approach discussed in @sec-discussion, we conclude in @sec-conclusion that counterfactual training provides a practical framework for researchers and practitioners interested in making opaque models more trustworthy.
+We also believe that this work serves as an opportunity for XAI researchers to re-evaluate the trend of improving XAI tools without improving the underlying models.

### 2025-03-10 paper/sections/method.qmd

@@
 Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit objective of aligning representations with plausible explanations that comply with user requirements.
-In the context of CE, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful].
-Plausibility is a necessary but insufficient condition for using CE to provide algorithmic recourse (AR) to individuals affected by opaque models in practice.
+In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful].
+Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals affected by opaque models in practice.
 This is because for recourse recommendations to be **actionable**, they need to not only result in plausible counterfactuals but also be attainable.
 A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older.
-Ignoring all other features, this complies with the definition of plausibility if 40-year-old individuals are in fact more credit-worthy on average than young adults.
-But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a mutable feature.
-For our intents and purposes, counterfactual training aims at improving model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability.
+Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults.
+But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature.
+For our intents and purposes, counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability.
 Formally, we define explainability as follows:
@@
 1. (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
-2. (Actionability) Permutations $\Delta$ are subject to actionability constraints.
+2. (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
 
@@
 Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual.
-For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.].
+For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]
 
-To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective,
+To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective:
 
@@
 
-where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g.
-cross-entropy).
+where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g., cross-entropy).
 The two additional components in @eq-obj are explained in more detail below.
-For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
+For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
 The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.
@@
 The first factor can be optimized using conventional cross-entropy as in @eq-obj.
-To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between samples drawn from $p_\theta(\mathbf{x})$ and training observations, i.e.
-samples from $p(\mathbf{x})$.
+To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between samples drawn from $p_\theta(\mathbf{x})$ and training observations, i.e., samples from $p(\mathbf{x})$.
 
@@
 This is where we depart from their methodology: instead of generating samples through SGLD, we propose using counterfactual explainers to generate counterfactuals for observed training samples.
-Specifically, we have
+Specifically, we have:
 
@@
 Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit].
-As the generated counterfactuals get more plausible (@def-explainability) over the cause of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].
+As the generated counterfactuals get more plausible (@def-explainability) over the course of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].
 
@@
 For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}[d]$ would violate mutability constraints.
-Specifically, we set $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ if
+Specifically, we set $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ if:
 
@@
 This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments.
-Under certain conditions, this results holds theoretically[For the proof, see the supplementary appendix.]:
+Under certain conditions, this results holds theoretically:^[For the proof, see the supplementary appendix.]
 
@@
 
-Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e.
-*multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
-If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is mutable and discriminative.
+Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e., *multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
+If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
 
@@
 Suppose we are interested in predicting the likelihood that loan applicants default on their credit.
-We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) two input features: 1) the subjects' *age*, which we define as immutable, and 2) the subjects' existing level of *debt*, which we define as mutable.
+We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.
 
@@
 In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time.
-Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals, in particular very young applicants.
+Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants).
 By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected immutable features as described in @sec-constraints.

### 2025-03-10 paper/sections/other/compute.qmd

@@
 Graphical processing units (GPU) were not required.
-For larger grid searches we used up to but typically less than 100 CPUs.
+Model training for the largest grid searches with 270 unique parameter combinations was parallelized across 34 CPUs with 2GB memory each.
+The time to completion varied by dataset for reasons discussed in @sec-discussion: 0h49m (*Moons*), 1h4m (*Linearly Separable*), 1h49m (*Circles*), 3h52m (*Overlapping*).
+Model evaluations for large grid searches were parallelized across 20 CPUs with 3GB memory each.
+Evaluations for all data sets took less than one hour (<1h) to complete.
 

### 2025-03-10 paper/preprint/paper.tex
done with grid search appendix

--- 2025-03-10 paper/preprint/paper.tex
@@
 \newtheorem{definition}{Definition}[section]
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{definition}
@@
 \newtheorem{corollary}{Research Question}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{remark}
@@
 explainability directly and indirectly by penalizing: 1) the contrastive
-divergence, \(\text{div}(\cdot)\), between counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, 2) the
+divergence, \(\text{div}(\cdot)\), between mature counterfactuals
+\(\mathbf{x}_T^\prime\) and observed samples \(\mathbf{x}\) and, 2) the
 adversarial loss, \(\text{advloss}(.)\), with respect to nascent
@@
 particular, we set
-\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
-where \(y^+\) denotes the index of the target class. We generate samples
-\(\mathbf{x}_T^\prime\) by first randomly sampling the target class
-\(y^+ \sim p(y)\) and then generating a counterfactual explanation for
-that target over \(T\) iterations using a gradient-based counterfactual
-generator. This is similar to how conditional sampling is used to draw
-from \(p_\theta(\mathbf{x})\) in Grathwohl et al.
-(\citeproc{ref-grathwohl2020your}{2020}).
+\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
+where \(y^+\) denotes the index of the randomly drawn target class,
+\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed data point
+sampled from target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\).
+Conditional on the target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes
+a mature counterfactual for a randomly sampled factual from a non-target
+class generated through a gradient-based counterfactual generator for at
+most \(T\) iterations. We define mature counterfactuals as those that
+have either exhausted \(T\) or reached convergence in terms of the
+pre-determined decision threshold earlier.
 
@@
 
+\subsubsection{Evaluation}\label{evaluation}
+
+Our key outcome of interest is how well models perform with respect to
+explainability (Definition~\ref{def-explainability}): to this end, we
+focus primarily on the plausibility and cost of faithfully generated
+counterfactuals at test time. To measure the cost of counterfactuals, we
+follow the standard convention of using distances (\(\ell_1\)-norm)
+between factuals and counterfactuals as a proxy. For plausibility, we
+assess how similar counterfactuals are to observed samples in the target
+domain. We rely on the distance-based metric used by Altmeyer et al.
+(\citeproc{ref-altmeyer2024faithful}{2024}),
+
+\begin{equation}\phantomsection\label{eq-impl-dist}{
+\text{implaus}_{\text{dist}}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+}\end{equation}
+
+and introduce a novel divergence metric,
+
+\begin{equation}\phantomsection\label{eq-impl-div}{
+\text{implaus}_{\text{div}}(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
+}\end{equation}
+
+where \(\mathbf{X}^\prime\) denotes a set of multiple counterfactuals
+and \(\text{MMD}(\cdot)\) is an unbiased estimate of the squared
+population maximum mean discrepancy
+(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
+Equation~\ref{eq-impl-div} is equal to zero iff
+\(\mathbf{X}^\prime=\mathbf{X}^+\).
+
+In addition to cost and plausibility, we also compute other standard
+metrics to evaluate counterfactuals at test time including validity and
+redundancy. Finally, we also assess the predictive performance of models
+using standard metrics.
+
 \subsection{Experimental Results}\label{experimental-results}
@@
 \emph{International Conference on Learning Representations}.
+
+\bibitem[\citeproctext]{ref-gretton2012kernel}
+Gretton, Arthur, Karsten M Borgwardt, Malte J Rasch, Bernhard Schölkopf,
+and Alexander Smola. 2012. {``A Kernel Two-Sample Test.''} \emph{The
+Journal of Machine Learning Research} 13 (1): 723--73.
 
@@
 \end{itemize}
+
+\subsection{Other Technical Details}\label{other-technical-details}
+
+\begin{equation}\phantomsection\label{eq-mmd}{
+\begin{aligned}
+MMD({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
+\end{aligned}
+}\end{equation}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note}, rightrule=.15mm, titlerule=0mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]
 
@@
 Section~\ref{sec-app-grid-gen} to Section~\ref{sec-app-grid-train} then
-present all details including the parameter grids, average performance
-outcomes and key evaluation metrics for the generated counterfactuals.
-
-\subsubsection{High-Level Findings}\label{sec-app-grid-hl}
-
-Overall, we find that Counterfactual Training (CT) achieves it key
-objectives consistently across all hyperparameter settings and also
-broadly across datasets. We do observe strong sensitivity to certain
-hyperparameters, but clear and manageable patterns emerge in those
-cases. With respect to the underlying training data distribution, we
-find that outcomes for CT seem to depend to some degree on class
-separability.
+present all details including the exact parameter grids, average
+predictive performance outcomes and key evaluation metrics for the
+generated counterfactuals.
+
+\subsubsection{Evaluation Details}\label{evaluation-details}
+
+To measure predictive performance, we compute the accuracy and F1-score
+for all models on test data (Table~\ref{tbl-acc-gen},
+Table~\ref{tbl-acc-pen}, Table~\ref{tbl-acc-train}). With respect to
+explanatory performance, we report here our findings for the
+(im)plausibility and cost of counterfactuals at test time. Since the
+computation of our proposed divergence metric
+(Equation~\ref{eq-impl-div}) is memory-intensive, we rely on the
+distance-based metric for the grid searches. For the counterfactual
+evaluation, we draw factual samples from the training data for the grid
+searches to avoid data leakage with respect to our final results
+reported in the body of the paper. Specifically, we want to avoid
+choosing our default hyperparameters based on results on the test data.
+Since we are optimizing for explainability, not predictive performance,
+we still present test accuracy and F1-scores.
 
@@
 
-\paragraph{Key Counterfactual
-Outcomes}\label{key-counterfactual-outcomes}
+\paragraph{Counterfactual Outcomes}\label{sec-app-grid-hl}
+
+Overall, we find that Counterfactual Training (CT) achieves it key
+objectives consistently across all hyperparameter settings and also
+broadly across datasets: plausibility is improved by up to
+\textasciitilde60 percentage points (ppts) for the \emph{Circles} data
+(e.g. Figure~\ref{fig-grid-gen_params-plaus-circles}),
+\textasciitilde25-30ppts for the \emph{Moons} data (e.g.
+Figure~\ref{fig-grid-gen_params-plaus-moons}) and
+\textasciitilde10-20ppts for the \emph{Linearly Separable} data (e.g.
+Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
+average costs of faithful counterfactuals are reduced in many cases by
+around \textasciitilde20-25ppts for \emph{Circles} (e.g.
+Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
+\textasciitilde50ppts for \emph{Moons} (e.g.
+Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
+Separable} data, costs are generally increased although typically by
+less than 10ppts (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
+which reflects a common tradeoff between costs and plausibility
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+
+We do observe strong sensitivity to certain hyperparameters, with clear
+an manageable patterns. Concerning generator parameters, we firstly find
+that using \emph{REVISE} to generate counterfactuals during training
+typically yields the worst outcomes out of all generators, often leading
+to a substantial decrease in plausibility. This finding can be
+attributed to the fact that \emph{REVISE} effectively assigns the task
+of learning plausible explanations from the model itself to a surrogate
+VAE. In other words, counterfactuals generated by \emph{REVISE} are less
+faithful to the model that \emph{ECCo} and \emph{Generic}, and hence we
+would expect them to be a less effective and, in fact, potentially
+detrimental role in our training regime. Secondly, we observe that
+allowing for a higher number of maximum steps \(T\) for the
+counterfactual search generally yields better outcomes. This is
+intuitive, because it allows more counterfactuals to reach maturity in
+any given iteration. Looking in particular at the results for
+\emph{Linearly Separable}, it seems that higher values for \(T\) in
+combination with higher decision thresholds (\(\tau\)) yields the best
+results when using \emph{ECCo}. But depending on the degree of class
+separability of the underlying data, a high decision-threshold can also
+affect results adversely, as evident from the results for the
+\emph{Overlapping} data (Figure~\ref{fig-grid-gen_params-plaus-over}):
+here we find that CT generally fails to achieve its objective because
+only a tiny proportion of counterfactuals ever reaches maturity.
+
+Regarding penalty strengths, we find that the strength of the energy
+regularization, \(\lambda_{\text{reg}}\) is a key hyperparameter, while
+sensitivity with respect to \(\lambda_{\text{div}}\) and
+\(\lambda_{\text{adv}}\) is much less evident. In particular, we observe
+that not regularizing energy enough or at all typically leads to poor
+performance in terms of decreased plausibility and increased costs, in
+particular for \emph{Circles} (Figure~\ref{fig-grid-pen-plaus-circles}),
+\emph{Linearly Separable} (Figure~\ref{fig-grid-pen-plaus-lin_sep}) and
+\emph{Overlapping} (Figure~\ref{fig-grid-pen-plaus-over}). High values
+of \(\lambda_{\text{reg}}\) can increase the variability in outcomes, in
+particular when combined with high values for \(\lambda_{\text{div}}\)
+and \(\lambda_{\text{adv}}\), but this effect is less pronounced.
+
+Finally, concerning other hyperparameters we observe that the
+effectiveness and stability of CT is positively associated with the
+number of counterfactuals generated during each training epoch, in
+particular for \emph{Circles}
+(Figure~\ref{fig-grid-train-plaus-circles}) and \emph{Moons}
+(Figure~\ref{fig-grid-train-plaus-moons}). We further find that a higher
+number of training epochs is beneficial as expected, where we tested
+training models for 50 and 100 epochs. Interestingly, we find that it is
+not necessary to employ CT during the entire training phase to achieve
+the desired improvements in explainability: specifically, we have tested
+training models conventionally during the first half of training before
+switching to CT after this initial burn-in period.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, rightrule=.15mm, titlerule=0mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, colframe=quarto-callout-note-color-frame, toptitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacitybacktitle=0.6, coltitle=black, bottomtitle=1mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomrule=.15mm, arc=.35mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, rightrule=.15mm, titlerule=0mm]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, opacitybacktitle=0.6, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, rightrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, toptitle=1mm, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, arc=.35mm, left=2mm, coltitle=black]
 

### 2025-03-10 paper/sections/method.qmd
done with grid search appendix

--- 2025-02-19 paper/sections/method.qmd
@@
 The two additional components in @eq-obj are explained in more detail below.
-For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
+For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: 1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}$ and, 2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
 The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.
@@
 where $\mathcal{E}_\theta(\cdot)$ denotes the energy function.
-In particular, we set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the target class.
-We generate samples $\mathbf{x}_T^\prime$ by first randomly sampling the target class $y^+ \sim p(y)$ and then generating a counterfactual explanation for that target over $T$ iterations using a gradient-based counterfactual generator.
-This is similar to how conditional sampling is used to draw from $p_\theta(\mathbf{x})$ in @grathwohl2020your.
+In particular, we set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$, and $\mathbf{x^+}$ denotes an observed data point sampled from target domain: $\mathbf{X}^+=\{\mathbf{x}:y=y^+\}$.
+Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated through a gradient-based counterfactual generator for at most $T$ iterations.
+We define mature counterfactuals as those that have either exhausted $T$ or reached convergence in terms of the pre-determined decision threshold earlier.
 

### 2025-03-10 paper/sections/other/grids.qmd
done with grid search appendix

@@
 To measure predictive performance, we compute the accuracy and F1-score for all models on test data (@tbl-acc-gen, @tbl-acc-pen, @tbl-acc-train).
-
-We evaluated
-
-### High-Level Findings {#sec-app-grid-hl}
-
-Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved  We do observe strong sensitivity to certain hyperparameters, but clear and manageable patterns emerge in those cases.
-With respect to the underlying training data distribution, we find that outcomes for CT seem to depend to some degree on class separability.
+With respect to explanatory performance, we report here our findings for the (im)plausibility and cost of counterfactuals at test time.
+Since the computation of our proposed divergence metric (@eq-impl-div) is memory-intensive, we rely on the distance-based metric for the grid searches.
+For the counterfactual evaluation, we draw factual samples from the training data for the grid searches to avoid data leakage with respect to our final results reported in the body of the paper.
+Specifically, we want to avoid choosing our default hyperparameters based on results on the test data.
+Since we are optimizing for explainability, not predictive performance, we still present test accuracy and F1-scores.
 
@@
 
-#### Key Counterfactual Outcomes
+#### Counterfactual Outcomes {#sec-app-grid-hl}
+
+Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to ~60 percentage points (ppts) for the *Circles* data (e.g.
+@fig-grid-gen_params-plaus-circles), ~25-30ppts for the *Moons* data (e.g.
+@fig-grid-gen_params-plaus-moons) and ~10-20ppts for the *Linearly Separable* data (e.g.
+@fig-grid-gen_params-plaus-lin_sep).
+At the same time, the average costs of faithful counterfactuals are reduced in many cases by around ~20-25ppts for  *Circles* (e.g.
+@fig-grid-gen_params-cost-circles) and up to ~50ppts for *Moons* (e.g.
+@fig-grid-gen_params-cost-moons).
+For the *Linearly Separable* data, costs are generally increased although typically by less than 10ppts (e.g.
+@fig-grid-gen_params-cost-lin_sep), which reflects a common tradeoff between costs and plausibility [@altmeyer2024faithful].
+
+We do observe strong sensitivity to certain hyperparameters, with clear an manageable patterns.
+Concerning generator parameters, we firstly find that using *REVISE* to generate counterfactuals during training typically yields the worst outcomes out of all generators, often leading to a substantial decrease in plausibility.
+This finding can be attributed to the fact that *REVISE* effectively assigns the task of learning plausible explanations from the model itself to a surrogate VAE.
+In other words, counterfactuals generated by *REVISE* are less faithful to the model that *ECCo* and *Generic*, and hence we would expect them to be a less effective and, in fact, potentially detrimental role in our training regime.
+Secondly, we observe that allowing for a higher number of maximum steps $T$ for the counterfactual search generally yields better outcomes.
+This is intuitive, because it allows more counterfactuals to reach maturity in any given iteration.
+Looking in particular at the results for *Linearly Separable*, it seems that higher values for $T$ in combination with higher decision thresholds ($\tau$) yields the best results when using *ECCo*.
+But depending on the degree of class separability of the underlying data, a high decision-threshold can also affect results adversely, as evident from the results for the *Overlapping* data (@fig-grid-gen_params-plaus-over): here we find that CT generally fails to achieve its objective because only a tiny proportion of counterfactuals ever reaches maturity.
+
+Regarding penalty strengths, we find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is a key hyperparameter, while sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
+In particular, we observe that not regularizing energy enough or at all typically leads to poor performance in terms of decreased plausibility and increased costs, in particular for *Circles* (@fig-grid-pen-plaus-circles), *Linearly Separable* (@fig-grid-pen-plaus-lin_sep) and *Overlapping* (@fig-grid-pen-plaus-over).
+High values of $\lambda_{\text{reg}}$ can increase the variability in outcomes, in particular when combined with high values for $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$, but this effect is less pronounced.
+
+Finally, concerning other hyperparameters we observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch, in particular for *Circles* (@fig-grid-train-plaus-circles) and *Moons* (@fig-grid-train-plaus-moons).
+We further find that a higher number of training epochs is beneficial as expected, where we tested training models for 50 and 100 epochs.
+Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability: specifically, we have tested training models conventionally during the first half of training before switching to CT after this initial burn-in period.
 

### 2025-03-10 paper/preprint/paper.tex
Improve wording in 3.1.1

--- 2025-03-10 paper/preprint/paper.tex
@@
 \theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{plain}
 \newtheorem{corollary}{Research Question}[section]
 \theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{remark}
@@
 any classifier can be re-interpreted as a joint energy-based model (JEM)
-that learns to discriminate output classes conditional on inputs and
-generate inputs. They show that JEMs can be trained to perform well at
-both tasks by directly maximizing the joint log-likelihood factorized as
+that learns to discriminate output classes conditional on the observed
+(training) samples from \(p(\mathbf{x})\) and the generated samples from
+\(p_\theta(\mathbf{x})\). They show that JEMs can be trained to perform
+well at both tasks by directly maximizing the joint log-likelihood
+factorized as
 \(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
 The first factor can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
+Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
 Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
-contrastive divergence between samples drawn from
-\(p_\theta(\mathbf{x})\) and training observations, i.e., samples from
-\(p(\mathbf{x})\).
+contrastive divergence between these observed samples from
+\(p(\mathbf{x})\) and generated samples from \(p_\theta(\mathbf{x})\).
 
@@
 initialization. This is where we depart from their methodology: instead
-of generating samples through SGLD, we propose using counterfactual
-explainers to generate counterfactuals for observed training samples.
-Specifically, we have
+of SGLD, we propose to use counterfactual explainers to generate
+counterfactuals of observed training samples. Specifically, we have:
 
@@
 sensitivity to that feature relative to the remaining features, provided
-that at least one of those is mutable and discriminative.
+that at least one of those is discriminative and mutable.
 

### 2025-03-10 paper/sections/method.qmd
Improve wording in 3.1.1

--- 2025-03-10 paper/sections/method.qmd
@@
 
-@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on inputs and generate inputs.
+@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
 They show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$.
 The first factor can be optimized using conventional cross-entropy as in @eq-obj.
-To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between samples drawn from $p_\theta(\mathbf{x})$ and training observations, i.e., samples from $p(\mathbf{x})$.
+Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.
 
@@
 To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization.
-This is where we depart from their methodology: instead of generating samples through SGLD, we propose using counterfactual explainers to generate counterfactuals for observed training samples.
+This is where we depart from their methodology: instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples.
 Specifically, we have:

### 2025-03-10 paper/preprint/paper.tex

@@
 \theoremstyle{definition}
-\newtheorem{example}{Example}[section]
+\newtheorem{definition}{Definition}[section]
 \theoremstyle{plain}
 \newtheorem{corollary}{Research Question}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
 \theoremstyle{remark}
@@
 explainability directly and indirectly by penalizing: (1) the
-contrastive divergence, \(\text{div}(\cdot)\), between counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
-adversarial loss, \(\text{advloss}(.)\), with respect to nascent
+contrastive divergence, \(\text{div}(\cdot)\), between mature
+counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
+(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
 counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between

### 2025-03-10 paper/sections/experiments.qmd
ffs

@@
 For plausibility, we assess how similar counterfactuals are to observed samples in the target domain.
-We rely on previously used distance-based metrics [@giudot]
+We rely on the distance-based metric used by @altmeyer2024faithful,
 
-We also assess their predictive performance using conventional performance metrics including accuracy and the F1-score.
+$$
+\text{implaus}_{\text{dist}}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+$$ {#eq-impl-dist}
+
+and introduce a novel divergence metric,
+
+$$
+\text{implaus}_{\text{div}}(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
+$$ {#eq-impl-div}
+
+where $\mathbf{X}^\prime$ denotes a set of multiple counterfactuals and $\text{MMD}(\cdot)$ is an unbiased estimate of the squared population maximum mean discrepancy [@gretton2012kernel].
+The metric in @eq-impl-div is equal to zero iff $\mathbf{X}^\prime=\mathbf{X}^+$.
+
+In addition to cost and plausibility, we also compute other standard metrics to evaluate counterfactuals at test time including validity and redundancy.
+Finally, we also assess the predictive performance of models using standard metrics.
 

### 2025-03-10 paper/sections/method.qmd

@@
 The two additional components in @eq-obj are explained in more detail below.
-For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
+For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
 The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.

### 2025-03-10 paper/sections/other/compute.qmd
present tunign results

@@
 Graphical processing units (GPU) were not required.
+
+### Grid Searches
+
 Model training for the largest grid searches with 270 unique parameter combinations was parallelized across 34 CPUs with 2GB memory each.
@@
 Evaluations for all data sets took less than one hour (<1h) to complete.
+
+### Tuning
+
+For tuning of selected hyperparameters, we distributed the task of generating counterfactuals during training across 40 CPUs with 2GB memory each for all tabular datasets.
+Except for the *Adult* dataset, all training runs were completed in less that half an hour (<0h30m).
+The *Adult* dataset took around 0h35m to complete.
+For *MNIST*, we relied on 100 CPUs with 2GB memory each.
 

### 2025-03-10 paper/sections/other/tuning.qmd
present tunign results

--- 2025-03-09 paper/sections/other/grids.qmd
@@
-## Grid Searches {#sec-app-grid}
+## Hyperparameter Tuning {#sec-app-tune}
 
@@
 
-To assess the hyperparameter sensitivity of our proposed training regime we ran multiple large grid searches for all of our synthetic datasets.
-We have grouped these grid searches into multiple categories:
-
-1. **Generator Parameters** (@sec-app-grid-gen): Investigates the effect of changing hyperparameters that affect the counterfactual outcomes during the training phase.
-2. **Penalty Strengths** (@sec-app-grid-pen): Investigates the effect of changing the penalty strengths in out proposed objective (@eq-obj).
-3. **Other Parameters** (@sec-app-grid-train): Investigates the effect of changing other training parameters, including the total number of generated counterfactuals in each epoch.
-
-We begin by summarizing the high-level findings in @sec-app-grid-hl.
-For each of the categories, @sec-app-grid-gen to @sec-app-grid-train then present all details including the parameter grids, average performance outcomes and key evaluation metrics for the generated counterfactuals.
-
-### High-Level Findings {#sec-app-grid-hl}
-
-Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets.
-We do observe strong sensitivity to certain hyperparameters, but clear and manageable patterns emerge in those cases.
-With respect to the underlying training data distribution, we find that outcomes for CT seem to depend to some degree on class separability.
-
-#### Predictive Performance
-
-We find that CT is associated with little to no decrease in average predictive performance for our synthetic datasets: test accuracy and F1-scores decrease by at most ~1 percentage point, but generally much less (@tbl-acc-gen, @tbl-acc-pen, @tbl-acc-train).
-
-#### Key Counterfactual Outcomes
-
-### Generator Parameters {#sec-app-grid-gen}
-
-```{julia}
-grid_dir = joinpath(res_dir, "gen_params/mlp")
-data_dirs = readdir(grid_dir) |> x -> joinpath.(grid_dir, x) |> x -> x[isdir.(x)]
-eval_grids = (p -> EvaluationGrid(joinpath(p,"grid_config.toml"))).(data_dirs)
-data_names = basename.(data_dirs)
-suffix = "evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/"
-```
-
-The hyperparameter grid with varying generator parameters during training is shown in @nte-gen-params-final-run-train.
-The corresponding evaluation grid used for these experiments is shown in @nte-gen-params-final-run-eval.
-
-::: {#nte-gen-params-final-run-train .callout-note}
-
-## Training Phase
-
-```{julia}
-#| output: asis
-dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/grid_config.toml")) 
-println(CTExperiments.dict_to_quarto_markdown(dict))
-```
-
-:::
-
-::: {#nte-gen-params-final-run-eval .callout-note}
-
-## Evaluation Phase
-
-```{julia}
-#| output: asis
-dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/evaluation/evaluation_grid_config.toml"))
-println(CTExperiments.dict_to_quarto_markdown(dict))
-```
-
-:::
-
-#### Accuracy
-
-::: {#tbl-acc-gen}
-
-::: {.content-hidden unless-format="pdf"}
-
-```{julia}
-#| output: asis
-df = CTExperiments.aggregate_performance(eval_grids; byvars=["objective"]) 
-get_table_inputs(df, "mean";backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; wrap_table=false)
-```
-
-:::
-
-Predictive performance measures by dataset and objective averaged across training-phase parameters (@nte-gen-params-final-run-train) and evaluation-phase parameters (@nte-gen-params-final-run-eval).
-
-:::
-
-#### Plausibility
-
-```{julia}
-#| output: asis
-
-fig_label_prefix = "grid-gen_params-plaus"
-fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
-_str = "The results with respect to the plausibility measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
-println(_str)
-```
-
-```{julia}
-#| output: asis
-
-imgfname = "plausibility_distance_from_target.png"
-fig_caption = "Average outcomes for the plausibility measure across hyperparameters."
-full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
-include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
-_str = join(include_img_commands, "\n\n")
-println(_str)
-```
-
-#### Cost
-
-```{julia}
-#| output: asis
-
-fig_label_prefix = "grid-gen_params-cost"
-fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
-_str = "The results with respect to the cost measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
-println(_str)
-```
-
-```{julia}
-#| output: asis
-
-imgfname = "distance.png"
-fig_caption = "Average outcomes for the cost measure across hyperparameters."
-full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
-include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption)
-_str = join(include_img_commands, "\n\n")
-println(_str)
-```
-
-### Penalty Strengths {#sec-app-grid-pen}
-
-```{julia}
-grid_dir = joinpath(res_dir, "penalties/mlp")
-data_dirs = readdir(grid_dir) |> x -> joinpath.(grid_dir, x) |> x -> x[isdir.(x)]
-eval_grids = (p -> EvaluationGrid(joinpath(p,"grid_config.toml"))).(data_dirs)
-data_names = basename.(data_dirs)
-suffix = "evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/"
-```
-
-The hyperparameter grid with varying penalty strengths during training is shown in @nte-pen-final-run-train.
-The corresponding evaluation grid used for these experiments is shown in @nte-pen-final-run-eval.
-
-::: {#nte-pen-final-run-train .callout-note}
-
-## Training Phase
-
-```{julia}
-#| output: asis
-dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/grid_config.toml")) 
-println(CTExperiments.dict_to_quarto_markdown(dict))
-```
-
-:::
-
-::: {#nte-pen-final-run-eval .callout-note}
-
-## Evaluation Phase
-
-```{julia}
-#| output: asis
-dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/evaluation/evaluation_grid_config.toml"))
-println(CTExperiments.dict_to_quarto_markdown(dict))
-```
-
-:::
-
-#### Accuracy
-
-::: {#tbl-acc-pen}
-
-::: {.content-hidden unless-format="pdf"}
-
-```{julia}
-#| output: asis
-df = CTExperiments.aggregate_performance(eval_grids; byvars=["objective"]) 
-get_table_inputs(df, "mean";backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; wrap_table=false)
-```
-
-:::
-
-Predictive performance measures by dataset and objective averaged across training-phase parameters (@nte-pen-final-run-train) and evaluation-phase parameters (@nte-pen-final-run-eval).
-
-:::
-
-#### Plausibility
-
-```{julia}
-#| output: asis
-
-fig_label_prefix = "grid-pen-plaus"
-fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
-_str = "The results with respect to the plausibility measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
-println(_str)
-```
-
-```{julia}
-#| output: asis
-
-imgfname = "plausibility_distance_from_target.png"
-fig_caption = "Average outcomes for the plausibility measure across hyperparameters."
-full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
-include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
-_str = join(include_img_commands, "\n\n")
-println(_str)
-```
-
-#### Cost
-
-```{julia}
-#| output: asis
-
-fig_label_prefix = "grid-pen-cost"
-fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
-_str = "The results with respect to the cost measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
-println(_str)
-```
-
-```{julia}
-#| output: asis
-
-imgfname = "distance.png"
-fig_caption = "Average outcomes for the cost measure across hyperparameters."
-full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
-include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption)
-_str = join(include_img_commands, "\n\n")
-println(_str)
-```
-
-### Other Parameters {#sec-app-grid-train}
-
-```{julia}
-grid_dir = joinpath(res_dir, "training_params/mlp")
-data_dirs = readdir(grid_dir) |> x -> joinpath.(grid_dir, x) |> x -> x[isdir.(x)]
-eval_grids = (p -> EvaluationGrid(joinpath(p,"grid_config.toml"))).(data_dirs)
-data_names = basename.(data_dirs)
-suffix = "evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/"
-```
-
-The hyperparameter grid with other varying training parameters is shown in @nte-train-final-run-train.
-The corresponding evaluation grid used for these experiments is shown in @nte-train-final-run-eval.
-
-::: {#nte-train-final-run-train .callout-note}
-
-## Training Phase
-
-```{julia}
-#| output: asis
-dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/grid_config.toml")) 
-println(CTExperiments.dict_to_quarto_markdown(dict))
-```
-
-:::
-
-::: {#nte-train-final-run-eval .callout-note}
-
-## Evaluation Phase
-
-```{julia}
-#| output: asis
-dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/evaluation/evaluation_grid_config.toml"))
-println(CTExperiments.dict_to_quarto_markdown(dict))
-```
-
-:::
-
-#### Accuracy
-
-::: {#tbl-acc-train}
-
-::: {.content-hidden unless-format="pdf"}
-
-```{julia}
-#| output: asis
-df = CTExperiments.aggregate_performance(eval_grids; byvars=["objective"]) 
-get_table_inputs(df, "mean";backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; wrap_table=false)
-```
-
-:::
-
-Predictive performance measures by dataset and objective averaged across training-phase parameters (@nte-train-final-run-train) and evaluation-phase parameters (@nte-train-final-run-eval).
-
-:::
-
-#### Plausibility
-
-```{julia}
-#| output: asis
-
-fig_label_prefix = "grid-train-plaus"
-fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
-_str = "The results with respect to the plausibility measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
-println(_str)
-```
-
-```{julia}
-#| output: asis
-
-imgfname = "plausibility_distance_from_target.png"
-fig_caption = "Average outcomes for the plausibility measure across hyperparameters."
-full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
-include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
-_str = join(include_img_commands, "\n\n")
-println(_str)
-```
-
-#### Cost
-
-```{julia}
-#| output: asis
-
-fig_label_prefix = "grid-train-cost"
-fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
-_str = "The results with respect to the cost measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
-println(_str)
-```
-
-```{julia}
-#| output: asis
-
-imgfname = "distance.png"
-fig_caption = "Average outcomes for the cost measure across hyperparameters."
-full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
-include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption)
-_str = join(include_img_commands, "\n\n")
-println(_str)
-```
+Based on the findings from our initial large grid searches (@sec-app-grid), we tune selected hyperparameters for all datasets.

### 2025-03-10 paper/experiments/present_results.jl

@@ -22,6 +22,7 @@ colvar = get_global_param("colvar", CTExperiments._colvar)
 colorvar_logs = get_global_param("colorvar_logs", colorvar)
 rowvar_logs = get_global_param("rowvar_logs", rowvar)
 colvar_logs = get_global_param("colvar_logs", colvar)
+xvar_ce = get_global_param("xvar_ce", nothing)
 colorvar_ce = get_global_param("colorvar_ce", colorvar)
 rowvar_ce = get_global_param("rowvar_ce", CTExperiments._rowvar_ce)
 colvar_ce = get_global_param("colvar_ce", CTExperiments._colvar_ce)
@@ -60,6 +61,7 @@ cfg = eval_grid
 all_data = CTExperiments.merge_with_meta(cfg, CTExperiments.load_ce_evaluation(cfg))
 valid_y = CTExperiments.valid_y_ce(all_data[1])
 params = PlotParams(;
+    x=get_global_param("xvar_ce", xvar_ce) |> x -> adjust_plot_var(x, cfg),
     colorvar=get_global_param("colorvar_ce", colorvar_ce) |> x -> adjust_plot_var(x, cfg),
     rowvar=get_global_param("rowvar_ce", rowvar_ce) |> x -> adjust_plot_var(x, cfg),
     colvar=get_global_param("colvar_ce", colvar_ce) |> x -> adjust_plot_var(x, cfg),

### 2025-03-10 paper/preprint/paper.tex
Add shortened discussion of hyper-parameters to the body

@@
 
+We run the experiments with three CE generators: \emph{Generic} of
+Wachter, Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}) as a simple baseline
+approach, \emph{REVISE} (\citeproc{ref-joshi2019realistic}{Joshi et al.
+2019}) that aims to generate plausible counterfactuals using a surrogate
+Variational Autoencoder (VAE), and \emph{ECCo}---the generator of
+Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023}) but without
+the conformal prediction component---as a method that directly targets
+both faithfulness and plausibility of the CEs.
+
 \subsection{Experimental Results}\label{experimental-results}
+
+\subsubsection{Plausibility}\label{plausibility-1}
+
+\subsubsection{Actionability}\label{actionability-1}
+
+\subsubsection{Impact of hyperparameter
+settings}\label{impact-of-hyperparameter-settings}
+
+We extensively test the impact of three types of hyperparameters on the
+proposed training regime. Our complete results are available in the
+technical appendix; this section focuses on the main findings.
+
+\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
+that CT is highly sensitive to hyperparameter settings but (a) there are
+manageable patterns and (b) we can typically identify settings that
+improve either plausibility or cost, and commonly both of them at the
+same time. Second, we note that the choice of a CE generator has a major
+impact on the results. For example, \emph{REVISE} tends to perform the
+worst, most likely because it attempts to explain the decisions of a VAE
+that is an imperfect surrogate of the original model. Third, increasing
+\(T\), the maximum number of steps, generally yields better outcomes
+because more CEs can mature in any iteration. Fourth, the impact of
+\(\tau\), the required decision threshold is more difficult to predict.
+On ``harder'' datasets it may be difficult to satisfy high \(\tau\) for
+any given sample (i.e., also factuals) and so increasing this threshold
+does not seem to correlate with better outcomes.
+
+\textbf{\emph{Hyperparameters for penalties}} We find that the strength
+of the energy regularization, \(\lambda_{\text{reg}}\) is highly
+impactful; energy must be sufficiently regularized to avoid poor
+performance in terms of decreased plausibility and increased costs. The
+sensitivity with respect to \(\lambda_{\text{div}}\) and
+\(\lambda_{\text{adv}}\) is much less evident. While high values of
+\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
+combined with high values for Equation~\ref{eq-obj} penalties, this
+effect is not very pronounced.
+
+\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
+and stability of CT is positively associated with the number of
+counterfactuals generated during each training epoch. We also confirm
+that a higher number of training epochs is beneficial. Interestingly, we
+find that it is not necessary to employ CT during the entire training
+phase to achieve the desired improvements in explainability. We have
+tested training models conventionally during the first half of training
+before switching to CT after this initial ``burn-in'' period and
+observed positive results. Put differently, CT may be a way to improve
+the explainability of trained models in a fine-tuning manner.
 
@@
 
+\bibitem[\citeproctext]{ref-altmeyer2023faithful}
+Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia C.
+S. Liem. 2023. {``Faithful Model Explanations Through Energy-Constrained
+Conformal Counterfactuals.''} \url{https://arxiv.org/abs/2312.10648}.
+
 \bibitem[\citeproctext]{ref-altmeyer2024faithful}
@@
 \url{https://doi.org/10.1145/3580305.3599290}.
+
+\bibitem[\citeproctext]{ref-joshi2019realistic}
+Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
+Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
+Actionable Explanations in Black-Box Decision Making Systems.''}
+\url{https://arxiv.org/abs/1907.09615}.
 

### 2025-03-10 paper/sections/experiments.qmd
Add shortened discussion of hyper-parameters to the body

@@
 
+We run the experiments with three CE generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2023faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the CEs.
+
 ## Experimental Results
+
+### Plausibility
+
+### Actionability
+
+### Impact of hyperparameter settings
+We extensively test the impact of three types of hyperparameters on the proposed training regime.
+Our complete results are available in the technical appendix; this section focuses on the main findings.
+
+***Hyperparameters of the CE generators.*** First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time.
+Second, we note that the choice of a CE generator has a major impact on the results.
+For example, *REVISE* tends to perform the worst, most likely because it attempts to explain the decisions of a VAE that is an imperfect surrogate of the original model.
+Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in any iteration.
+Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict.
+On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes.
+
+***Hyperparameters for penalties*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs.
+The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
+While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values for @eq-obj penalties, this effect is not very pronounced.
+
+***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
+We also confirm that a higher number of training epochs is beneficial.
+Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability.
+We have tested training models conventionally during the first half of training before switching to CT after this initial "burn-in" period and observed positive results.
+Put differently, CT may be a way to improve the explainability of trained models in a fine-tuning manner.

### 2025-03-11 paper/preprint/paper.tex
Add first part of discussion (Section 5.1)

@@
 \theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
 \newtheorem{example}{Example}[section]
 \theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{plain}
 \newtheorem{proposition}{Proposition}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 \subsubsection{Impact of hyperparameter
-settings}\label{impact-of-hyperparameter-settings}
+settings}\label{sec-hyperparameters}
 
@@
 
-\subsection{Approach/Future Directions}\label{approachfuture-directions}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Limited to classification models.
-\item
-  Training instabilities.
-\item
-  Hyperparameter sensitivity -\textgreater{} can we do better than grid
-  search? (Bayes opt, \ldots)
-\end{enumerate}
-
-\subsection{Limitations}\label{limitations}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  Proxy attributes of immutable features.
-\item
-  Increased training time.
-\item
-  Fairness and caveats (aware it's not a classical approach in this
-  context, but there is a clear link).
-\end{enumerate}
+We begin the discussion by addressing the direct extensions of the
+counterfactual training approach in Section~\ref{sec-approach}. Then, we
+look at its broader limitations and challenges in
+Section~\ref{sec-limitations}.
+
+\subsection{Future research}\label{sec-approach}
+
+\textbf{\emph{CT is defined only for classification settings.}} Our
+formulation relies on the distinction between non-target class(es)
+\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
+through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
+arbitrarily defined by the user, CT requires the output space
+\(\mathcal{Y}\) to be discrete. Thus, it applies to binary and
+multi-class classification but it is not well-defined for other ML tasks
+where the change in outcome with respect to a decision threshold
+\(\tau\) cannot be readily quantified. In fact, this is a common
+restriction in research on CEs and AR that predominantly focuses on
+classification models. Although other settings have attracted some
+interest (e.g., regression in
+\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021};
+\citeproc{ref-zhao2023counterfactual}{Zhao, Broelemann, and Kasneci
+2023}), there is still no consensus on what constitutes a counterfactual
+in such settings.
+
+\textbf{\emph{CT is subject to training instabilities.}} Joint
+energy-based models are susceptible to instabilities during training
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
+though we depart from the SGLD-based sampling, we still encounter major
+variability in the outcomes. CT is exposed to two potential sources of
+instabilities: (1) the energy-based contrastive divergence term in
+Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
+For example, Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023})
+recognize this to be a challenge for \emph{ECCCo} and so it may have
+downstream impacts on our proposed method. Still, we find that training
+instabilities can be largely counteracted with strong regularization, in
+particular through \(\lambda_{\text{reg}}\) penalty on the energy.
+
+\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
+discussed in Section~\ref{sec-hyperparameters}, our method requires
+careful selection of the hyperparameters. All experiments relied on
+grid-search tuning but we recognize that other algorithms may be more
+efficient. Notably, counterfactual training is iterative which makes a
+variety of methods applicable, including Bayesian (e.g.,
+\citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
+gradient-based (e.g., \citeproc{ref-franceschi2017forward}{Franceschi et
+al. 2017}) optimization.
+
+\subsection{Limitations and challenges}\label{sec-limitations}
+
+\textbf{\emph{CT increases the training time of models.}}
+
+\textbf{\emph{Immutable features may have proxies.}}
+
+\textbf{\emph{Interventions on features may have downstream impacts on
+fairness.}}
 
@@
 \url{https://arxiv.org/abs/1903.08689}.
+
+\bibitem[\citeproctext]{ref-franceschi2017forward}
+Franceschi, Luca, Michele Donini, Paolo Frasconi, and Massimiliano
+Pontil. 2017. {``{Forward and Reverse Gradient-Based Hyperparameter
+Optimization}.''} In \emph{Proceedings of the 34th International
+Conference on Machine Learning}, edited by Doina Precup and Yee Whye
+Teh, 70:1165--73. Proceedings of Machine Learning Research. PMLR.
+\url{https://proceedings.mlr.press/v70/franceschi17a.html}.
 
@@
 
+\bibitem[\citeproctext]{ref-snoek2012practical}
+Snoek, Jasper, Hugo Larochelle, and Ryan P. Adams. 2012. {``{Practical
+Bayesian Optimization of Machine Learning Algorithms}.''} In
+\emph{Advances in Neural Information Processing Systems}, edited by F.
+Pereira, C. J. Burges, L. Bottou, and K. Q. Weinberger. Vol. 25. Curran
+Associates, Inc.
+\url{https://proceedings.neurips.cc/paper_files/paper/2012/file/05311655a15b75fab86956663e1819cd-Paper.pdf}.
+
+\bibitem[\citeproctext]{ref-spooner2021counterfactual}
+Spooner, Thomas, Danial Dervovic, Jason Long, Jon Shepard, Jiahao Chen,
+and Daniele Magazzeni. 2021. {``{Counterfactual Explanations for
+Arbitrary Regression Models}.''} \emph{CoRR} abs/2106.15212.
+\url{https://arxiv.org/abs/2106.15212}.
+
 \bibitem[\citeproctext]{ref-szegedy2013intriguing}
@@
 
+\bibitem[\citeproctext]{ref-zhao2023counterfactual}
+Zhao, Xuan, Klaus Broelemann, and Gjergji Kasneci. 2023.
+{``{Counterfactual Explanation for Regression via Disentanglement in
+Latent Space}.''} In \emph{2023 IEEE International Conference on Data
+Mining Workshops (ICDMW)}, 976--84. Los Alamitos, CA, USA: IEEE Computer
+Society. \url{https://doi.org/10.1109/ICDMW60847.2023.00130}.
+
 \end{CSLReferences}

### 2025-03-11 paper/sections/discussion.qmd
latex is going to kill me one day
phew
vamos
some notes from meeting and updates to appendix
Add first part of discussion (Section 5.1)

@@
 
-1. Limited to classification models.
-2. Proxy attributes of immutable features.
+We begin the discussion by addressing the direct extensions of the counterfactual training approach in @sec-approach. Then, we look at its broader limitations and challenges in @sec-limitations.
+
+## Future research {#sec-approach}
+
+***CT is defined only for classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj.
+While $y^{-}$ and $y^{+}$ can be arbitrarily defined by the user, CT requires the output space $\mathcal{Y}$ to be discrete.
+Thus, it applies to binary and multi-class classification but it is not well-defined for other ML tasks where the change in outcome with respect to a decision threshold $\tau$ cannot be readily quantified.
+In fact, this is a common restriction in research on CEs and AR that predominantly focuses on classification models.
+Although other settings have attracted some interest [e.g., regression in @spooner2021counterfactual; @zhao2023counterfactual], there is still no consensus on what constitutes a counterfactual in such settings.
+
+***CT is subject to training instabilities.*** Joint energy-based models are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes.
+CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers.
+For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method.
+Still, we find that training instabilities can be largely counteracted with strong regularization, in particular through $\lambda_{\text{reg}}$ penalty on the energy.
+
+***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method requires careful selection of the hyperparameters.
+All experiments relied on grid-search tuning but we recognize that other algorithms may be more efficient.
+Notably, counterfactual training is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.
+
+## Limitations and challenges {#sec-limitations}
+
+***CT increases the training time of models.***
+
+***Immutable features may have proxies.***
+
+***Interventions on features may have downstream impacts on fairness.***

### 2025-03-11 paper/preprint/paper.tex
Add two points in 5.2 limitations

@@
 \usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{definition}
@@
 
-\textbf{\emph{CT increases the training time of models.}}
-
-\textbf{\emph{Immutable features may have proxies.}}
+\textbf{\emph{CT increases the training time of models.}} Counterfactual
+training promotes explainability through CEs and robustness through AEs
+at the cost of longer training times compared to conventional training
+regimes. While higher numbers of iterations and counterfactuals per
+iteration positively impact the quality of found solutions, they also
+increase the required amount of computations. We find that relatively
+small grids with 270 settings can take almost four hours for more
+demanding datasets on a high-performance computing cluster with 34 2GB
+CPUs (see details in \textbf{?@sec-hardware}). However, there are three
+factors that attenuate the impact of this limitation. First, CT provides
+counterfactual explanations for the training samples essentially for
+free, which may be beneficial in many ADM systems. Second, we find that
+CT can retain its value when used as a ``fine-tuning'' training regime
+for conventionally-trained models. Third, in principle, CT yields itself
+to multi-threaded execution.
+
+\textbf{\emph{Immutable features may have proxies.}} In
+Proposition~\ref{prp-mtblty} we define an approach to protect immutable
+features and thus increase the actionability of the generated
+counterfactuals. However, this approach requires that model owners
+define the mutability constraints for (all) features considered by the
+model. Even with sufficient domain knowledge to protect all immutable
+features---ones that cannot be changed at all and ones that cannot be
+reasonably expected to change---there may exist proxies that are
+theoretically mutable (and hence should not be protected) but preserve
+enough information about the principals to counteract the protections.
+As one example, consider the Adult dataset used in our experiments where
+the mutable education status is a proxy for the immutable age, in that
+the attainment of degrees is correlated with age. Delineating
+actionability is a major undecided challenge in the AR literature (see,
+e.g.,
+\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
+and Alfano 2020}) impacting the capacity of CT to increase the
+explainability of the model.
 
 \textbf{\emph{Interventions on features may have downstream impacts on
-fairness.}}
+fairness.}}\\
+- \href{https://arxiv.org/abs/2401.16088}{Equality of opportunity}?\\
+- \href{https://proceedings.mlr.press/v202/gao23d/gao23d.pdf}{Social
+segregation}?\\
+- Supporting AR within one context/system may still unfairly target
+recourse-affected individuals within other contexts/systems? In Example
+3.1 younger individuals gain access to a loan but their age could still
+lead to, e.g., enforcing stronger supervisory mechanisms?
 
@@
 
+\bibitem[\citeproctext]{ref-venkatasubramanian2020philosophical}
+Venkatasubramanian, Suresh, and Mark Alfano. 2020. {``The Philosophical
+Basis of Algorithmic Recourse.''} In \emph{Proceedings of the 2020
+Conference on Fairness, Accountability, and Transparency}, 284--93. FAT*
+'20. New York, NY, USA: Association for Computing Machinery.
+\url{https://doi.org/10.1145/3351095.3372876}.
+
 \bibitem[\citeproctext]{ref-wachter2017counterfactual}

### 2025-03-11 paper/sections/discussion.qmd
Add two points in 5.2 limitations

@@
 
-***CT increases the training time of models.***
+***CT increases the training time of models.*** Counterfactual training promotes explainability through CEs and robustness through AEs at the cost of longer training times compared to conventional training regimes.
+While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations.
+We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see details in @sec-hardware).
+However, there are three factors that attenuate the impact of this limitation.
+First, CT provides counterfactual explanations for the training samples essentially for free, which may be beneficial in many ADM systems.
+Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models.
+Third, in principle, CT yields itself to multi-threaded execution.
 
-***Immutable features may have proxies.***
+***Immutable features may have proxies.*** In @prp-mtblty we define an approach to protect immutable features and thus increase the actionability of the generated counterfactuals.
+However, this approach requires that model owners define the mutability constraints for (all) features considered by the model.
+Even with sufficient domain knowledge to protect all immutable features---ones that cannot be changed at all and ones that cannot be reasonably expected to change---there may exist proxies that are theoretically mutable (and hence should not be protected) but preserve enough information about the principals to counteract the protections.
+As one example, consider the Adult dataset used in our experiments where the mutable education status is a proxy for the immutable age, in that the attainment of degrees is correlated with age.
+Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to increase the explainability of the model.
 
 ***Interventions on features may have downstream impacts on fairness.***  
+- [Equality of opportunity](https://arxiv.org/abs/2401.16088)?  
+- [Social segregation](https://proceedings.mlr.press/v202/gao23d/gao23d.pdf)?  
+- Supporting AR within one context/system may still unfairly target recourse-affected individuals within other contexts/systems?
+In Example 3.1 younger individuals gain access to a loan but their age could still lead to, e.g., enforcing stronger supervisory mechanisms?

### 2025-03-12 paper/sections/other/compute.qmd
Add two points in 5.2 limitations

@@
 The *Adult* dataset took around 0h35m to complete.
+Evaluations across 20 CPUs with 3GB memory each generally took less than 0h30m to complete.
 For *MNIST*, we relied on 100 CPUs with 2GB memory each.
+For the *MLP*, training of all models could be completed in 1h30m, while the evaluation across 20 CPUs (6GB memory) took 4h12m.
+For the *CNN*, training of all models took ~8h, with conventionally trained models taking ~0h15m each and model with CT taking ~0h30m-0h45m each.
 

### 2025-03-12 paper/sections/other/tuning.qmd

@@
 
-Based on the findings from our initial large grid searches (@sec-app-grid), we tune selected hyperparameters for all datasets.
+Based on the findings from our initial large grid searches (@sec-app-grid), we tune selected hyperparameters for all datasets: namely, the decision threshold $\tau$ and the strength of the energy regularization $\lambda_{\text{reg}}$.
+The final hyperparameter choices for each dataset are presented in **ADD TABLE**.
+Detailed results for each data set are shown in **ADD FIGURES**.
+From **ADD TABLE**, we notice that the same decision threshold of $\tau=0.5$ is optimal for all but on dataset.
+We attribute this to the fact that a low decision threshold results in a higher share of mature counterfactuals and hence more opportunities for the model to learn from examples.
+This has played a role in particular for our real-world tabular datasets and MNIST, which suffered from low levels of maturity for higher decision thresholds.
+In cases where maturity is not an issue, as for *Moons*, higher decision thresholds lead to better outcomes, which may have to do with the fact that the resulting counterfactuals are more faithful to the model.
+Concerning the regularization strength, we find somewhat high variation across datasets.
+Most notably, we find that relatively low levels of regularization are optimal for MNIST.
+We hypothesize that this finding may be attributed to the uniform scaling of all input features (digits).
+Finally, to increase the proportion of mature counterfactuals for some datasets, we have also investigated the effect on the learning rate $\eta$ for the counterfactual search, but found little effect on the results.

### 2025-03-12 paper/preprint/paper.tex
added tuning section to appendix

--- 2025-03-10 paper/preprint/paper.tex
@@
 \theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{definition}
 \newtheorem{example}{Example}[section]
@@
 \newtheorem{proposition}{Proposition}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 relying on features and rules that are carefully hand-crafted by humans,
-modern machine learning (ML) models are tasked with learning the
-representations directly from data, guided by narrow objectives such as
+modern machine learning (ML) models are tasked with learning these
+representations from scratch, guided by narrow objectives such as
 predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
 Bengio, and Courville 2016}). Modern advances in computing have made it
-possible to provide such models with ever growing degrees of freedom to
+possible to provide such models with ever greater degrees of freedom to
 achieve that task, which has often led them to outperform traditionally
-more parsimonious models. Unfortunately, in doing so, the models learn
+more parsimonious models. Unfortunately, in doing so they also learn
 increasingly complex and highly sensitive representations that we can no
@@
 
-The trend towards complexity for the sake of performance has come under
+This trend towards complexity for the sake of performance has come under
 serious scrutiny in recent years. At the very cusp of the deep learning
@@
 different model predictions despite being ``imperceptible'' in that they
-are semantically indifferent from their factual counterparts. Although
-some partially effective mitigation strategies have been proposed, for
-example \textbf{adversarial training}
-(\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow, Shlens, and
-Szegedy 2014}), truly robust deep learning (DL) remains unattainable
-even for models that are considered shallow by today's standards
-(\citeproc{ref-kolter2023keynote}{Kolter 2023}).
-
-Part of the problem is that the high degrees of freedom provide room for
+are semantically indifferent from their factual counterparts. Despite
+partially effective mitigation strategies such as \textbf{adversarial
+training} (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow,
+Shlens, and Szegedy 2014}), truly robust deep learning (DL) remains
+unattainable even for models that are considered shallow by today's
+standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).
+
+Part of the problem is that high degrees of freedom provide room for
 many solutions that are locally optimal with respect to narrow
-objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
-  follow the standard ML convention, where ``degrees of freedom'' refer
-  to the number of parameters estimated from data.} Indeed, recent work
-on the so called ``lottery tickets'' suggests that modern neural
-networks can be pruned by up to 90\% while preserving their predictive
-performance (\citeproc{ref-frankle2018lottery}{Frankle and Carbin 2019})
-and generalizability (\citeproc{ref-morcos2019success}{Morcos et al.
-2019}). Similarly, Zhang et al.
-(\citeproc{ref-zhang2021understanding}{2021}) showed that
-state-of-the-art neural networks are so expressive that they can fit
-randomly labeled data. Thus, looking at the predictive performance, the
-solutions may seem to provide compelling explanations for the data, when
-in fact they are based on purely associative, semantically meaningless
-patterns. This poses two related challenges. Firstly, there is no
-dependable way to verify if such complex representations correspond to
-meaningful and plausible explanations. Secondly, even if we could
-resolve the first challenge, it remains undecided how to ensure that
-models can \emph{only} learn valuable explanations.
+objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
+  clarity: we follow standard ML convention in using ``degrees of
+  freedom'' to refer to the number of parameters estimated from data.}.
+Based purely on predictive performance, these solutions may seem to
+provide compelling explanations for the data, when in fact they are
+based on purely associative, semantically meaningless patterns. This
+poses two related challenges: firstly, it makes these models inherently
+opaque, since humans cannot simply interpret what type of explanation
+the complex learned representations correspond to; secondly, even if we
+could resolve the first challenge, it is not obvious how to mitigate
+models from learning representations that correspond to meaningless and
+implausible explanations.
 
@@
 scenario in which we deploy opaque models and blindly rely on their
-predictions. On countless occasions, this scenario has occurred in
-practice and caused real harm to people who were affected adversely and
-often unfairly by automated decision-making (ADM) systems involving
-opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016};
-\citeproc{ref-mcgregor2021preventing}{McGregor 2021}). Effective XAI
-tools can aid us in monitoring models and providing recourse to
-individuals to turn adverse outcomes (e.g., ``loan application
-rejected'') into positive ones (e.g., ``application accepted'').
-Wachter, Mittelstadt, and Russell
+predictions. On countless occasions, this scenario has already occurred
+in practice and caused real harm to people who were affected adversely
+and often unfairly by automated decision-making systems (ADMS) involving
+opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective
+XAI tools can aide us in monitoring models and providing recourse to
+individuals to turn adverse outcomes (e.g.~``loan application
+rejected'') into positive ones (``application accepted''). Wachter,
+Mittelstadt, and Russell
 (\citeproc{ref-wachter2017counterfactual}{2017}) propose
-\textbf{counterfactual explanations} (CE) as an effective approach to
-achieve this goal: CEs explain how factual inputs need to change in
-order for some fitted model to produce some desired output, typically
-involving minimal perturbations.
-
-To our surprise, the second challenge has not yet attracted any major
+\textbf{counterfactual explanations} as an effective approach to achieve
+this: they explain how factual inputs need to change in order for some
+fitted model to produce some desired output, typically involving minimal
+perturbations.
+
+To our surprise, the second challenge has not yet attracted any
 consolidated research effort. Specifically, there has been no concerted
-effort towards improving improving models' explanatory capacity, which
-we will henceforth simply call ``explainability'', defined as the degree
-to which learned representations correspond to explanations that are
-interpretable and deemed \textbf{plausible} by humans (see
-Definition~\ref{def-explainability}). Instead, the choice has typically
-been to improve the ability of XAI tools to identify the subset
-explanations that are both plausible and valid for any given model,
-independent of whether the learned representations are also compatible
-with implausible explanations
+effort towards improving model \textbf{explainability}, which we define
+here as the degree to which learned representations correspond to
+explanations that are interpretable and deemed \textbf{plausible} by
+humans (see Definition~\ref{def-explainability}). Instead, the choice
+has typically been to improve the capacity of XAI tools to identify the
+subset explanations that are both plausible and valid for any given
+model, independent of whether the learned representations are also
+compatible with implausible explanations
 (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
@@
 byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization, and generative capacity
-(\citeproc{ref-schut2021generating}{Schut et al. 2021};
-\citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein 2020};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+robustness, generalization and generative capacity Altmeyer et al.
+(\citeproc{ref-altmeyer2024faithful}{2024}).
 
 Building on these findings, we introduce \textbf{counterfactual
-training}: a novel training regime explicitly geared towards aligning
-model representations with plausible explanations. Our contributions are
-as follows:
+training}: a novel regularization technique geared explicitly towards
+aligning model representations with plausible explanations. Our
+contributions are as follows:
 
@@
 
-Despite some limitations of our approach discussed in
-Section~\ref{sec-discussion}, we conclude in
-Section~\ref{sec-conclusion} that counterfactual training provides a
-practical framework for researchers and practitioners interested in
-making opaque models more trustworthy. We also believe that this work
-serves as an opportunity for XAI researchers to re-evaluate the trend of
-improving XAI tools without improving the underlying models.
+Despite limitations of our approach discussed in
+Section~\ref{sec-discussion}, we conclude that counterfactual training
+provides a practical framework for researchers and practitioners
+interested in making opaque models more trustworthy
+Section~\ref{sec-conclusion}. We also believe that this work serves as
+an opportunity for XAI researchers to reevaluate the premise of
+improving XAI tools without improving models.
 
@@
 
-To the best of our knowledge, our proposed framework of counterfactual
+To the best of our knowledge, our proposed framework for counterfactual
 training represents the first attempt to use counterfactual explanations
@@
 concrete, we follow Augustin, Meinke, and Hein
-(\citeproc{ref-augustin2020adversarial}{2020}) in tying the concept of
+(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
 explainability to the quality of counterfactual explanations that we can
@@
 freedom: by inducing models to ``unlearn'' representations that are
-susceptible to worst-case counterfactuals (i.e., adversarial examples),
+susceptible to worst-case counterfactuals (i.e.~adversarial examples),
 AT effectively removes some implausible explanations from the solution
@@
 they argue that counterfactual pairs typically already exist in training
-datasets. Specifically, their approach relies on identifying similar
-input samples with different annotations and ensuring that the gradient
-of the classifier aligns with the vector between such pairs of
-counterfactual inputs using the cosine distance as the loss function.
+datasets. Specifically, their approach relies on, firstly, identifying
+similar input samples with different annotations and, secondly, ensuring
+that the gradient of the classifier aligns with the vector between pairs
+of counterfactual inputs using the cosine distance as a loss function.
 
@@
 There have also been several attempts at formalizing the relationship
-between counterfactual explanations and adversarial examples (AE).
+between counterfactual explanations (CE) and adversarial examples (AE).
 Pointing to clear similarities in how CE and AE are generated,
@@
 case for jointly studying the opaqueness and robustness problem in
-representation learning. Formally, AE can be seen as the subset of CE
+representation learning. Formally, AE can be seen as the subset of CE,
 for which misclassification is achieved
@@
 for individuals to move from an adverse outcome to some positive target
-class with high probability. Their approach builds on adversarial
-training, where in this context susceptibility to targeted adversarial
-examples for the positive class is explicitly induced. The proposed
-method allows for imposing a set of actionability constraints ex-ante:
-for example, users can specify that certain features (e.g., \emph{age},
-\emph{gender}, \ldots) are immutable. Secondly, Guo, Nguyen, and Yadav
-(\citeproc{ref-guo2023counternet}{2023}) are the first to propose an
-end-to-end training pipeline that includes counterfactual explanations
-as part of the training procedure. In particular, they propose a
-specific network architecture that includes a predictor and CE generator
-network, where the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed
-back to the predictor network. In contrast to Guo, Nguyen, and Yadav
-(\citeproc{ref-guo2023counternet}{2023}), we impose no restrictions on
-the neural network architecture at all.
+class with high probability. The approach proposed by Ross, Lakkaraju,
+and Bastani (\citeproc{ref-ross2021learning}{2024}) builds on
+adversarial training, where in this context susceptibility to targeted
+adversarial examples for the positive class is explicitly induced. The
+proposed method allows for imposing a set of actionability constraints
+ex-ante: for example, users can specify that certain features
+(e.g.~\emph{age}, \emph{gender}, \ldots) are immutable. Secondly, Guo,
+Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the first
+to propose an end-to-end training pipeline that includes counterfactual
+explanations as part of the training procedure. In particular, they
+propose a specific network architecture that includes a predictor and CE
+generator network, where the parameters of the CE generator network are
+learnable. Counterfactuals are generated during each training iteration
+and fed back to the predictor network. In contrast to Guo, Nguyen, and
+Yadav (\citeproc{ref-guo2023counternet}{2023}), we impose no
+restrictions on the neural network architecture at all.
 
@@
 explicit objective of aligning representations with plausible
-explanations that comply with user requirements. In the context of CEs,
+explanations that comply with user requirements. In the context of CE,
 plausibility has broadly been defined as the degree to which
@@
 \citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
-is a necessary but insufficient condition for using CEs to provide
+is a necessary but insufficient condition for using CE to provide
 algorithmic recourse (AR) to individuals affected by opaque models in
@@
 application would have been accepted, if only they were 20 years older.
-Ignoring all other features, this would comply with the definition of
-plausibility if 40-year-old individuals were in fact more credit-worthy
+Ignoring all other features, this complies with the definition of
+plausibility if 40-year-old individuals are in fact more credit-worthy
 on average than young adults. But of course this CE does not qualify for
 providing actionable recourse to the applicant since \emph{age} is not a
-(directly) mutable feature. For our intents and purposes, counterfactual
-training aims to improve model explainability by aligning models with
+mutable feature. For our intents and purposes, counterfactual training
+aims at improving model explainability by aligning models with
 counterfactuals that meet both desiderata, plausibility and
@@
 \item
-  (Actionability) Permutations \(\Delta\) are subject to some
-  actionability constraints.
+  (Actionability) Permutations \(\Delta\) are subject to actionability
+  constraints.
 \end{enumerate}
@@
 pre-determined threshold, \(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
-where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
-  background information on gradient-based counterfactual search and
-  convergence see \textbf{?@sec-app-ce}.}
+\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
+  detailed background information on gradient-based counterfactual
+  search and convergence see Section~\ref{sec-app-ce}.}.
 
@@
 Definition~\ref{def-explainability}, we propose to leverage
-counterfactuals in the following objective:
+counterfactuals in the following objective,
 
@@
 where \(\text{yloss}(\cdot)\) is any conventional classification loss
-that induces discriminative performance (e.g., cross-entropy). The two
+that induces discriminative performance (e.g.~cross-entropy). The two
 additional components in Equation~\ref{eq-obj} are explained in more
 detail below. For now, they can be sufficiently described as inducing
-explainability directly and indirectly by penalizing: (1) the
-contrastive divergence, \(\text{div}(\cdot)\), between mature
-counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
-(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
+explainability directly and indirectly by penalizing: 1) the contrastive
+divergence, \(\text{div}(\cdot)\), between mature counterfactuals
+\(\mathbf{x}_T^\prime\) and observed samples \(\mathbf{x}\) and, 2) the
+adversarial loss, \(\text{advloss}(.)\), with respect to nascent
 counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
@@
 any classifier can be re-interpreted as a joint energy-based model (JEM)
-that learns to discriminate output classes conditional on the observed
-(training) samples from \(p(\mathbf{x})\) and the generated samples from
-\(p_\theta(\mathbf{x})\). They show that JEMs can be trained to perform
-well at both tasks by directly maximizing the joint log-likelihood
-factorized as
+that learns to discriminate output classes conditional on inputs and
+generate inputs. They show that JEMs can be trained to perform well at
+both tasks by directly maximizing the joint log-likelihood factorized as
 \(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
 The first factor can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
+Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
 Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
-contrastive divergence between these observed samples from
-\(p(\mathbf{x})\) and generated samples from \(p_\theta(\mathbf{x})\).
+contrastive divergence between samples drawn from
+\(p_\theta(\mathbf{x})\) and training observations, i.e.~samples from
+\(p(\mathbf{x})\).
 
@@
 initialization. This is where we depart from their methodology: instead
-of SGLD, we propose to use counterfactual explainers to generate
-counterfactuals of observed training samples. Specifically, we have:
+of generating samples through SGLD, we propose using counterfactual
+explainers to generate counterfactuals for observed training samples.
+Specifically, we have
 
@@
 counterfactuals get more plausible (Definition~\ref{def-explainability})
-over the course of training, these two opposing effects gradually
-balance each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
+over the cause of training, these two opposing effects gradually balance
+each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
 
@@
 descent in the feature space for optimization
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, van Deursen, and Liem
-2023}). In this context, domain constraints can be imposed by simply
-projecting counterfactuals back to the specified domain, if the previous
-gradient step resulted in updated feature values that were
-out-of-domain. Mutability constraints can similarly be enforced by
-setting partial derivatives to zero to ensure that features are only
-mutated in the allowed direction, if at all.
+(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
+In this context, domain constraints can be imposed by simply projecting
+counterfactuals back to the specified domain, if the previous gradient
+step resulted in updated feature values that were out-of-domain.
+Mutability constraints can similarly be enforced by setting partial
+derivatives to zero to ensure that features are only mutated in the
+allowed direction, if at all.
 
@@
 violate mutability constraints. Specifically, we set
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if
 
@@
 Section~\ref{sec-experiments}. Under certain conditions, this results
-holds theoretically:\footnote{For the proof, see the supplementary
-  appendix.}
+holds theoretically{[}For the proof, see the supplementary appendix.{]}:
 
@@
 denote a linear classifier with softmax activation \(\mathcal{S}\)
-(i.e., \emph{multinomial logistic regression}) where
+(i.e.~\emph{multinomial logistic regression}) where
 \(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
@@
 sensitivity to that feature relative to the remaining features, provided
-that at least one of those is discriminative and mutable.
+that at least one of those is mutable and discriminative.
 
@@
 (\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
-(1) the subjects' \emph{age}, which we define as immutable, and (2) the
+1) the subjects' \emph{age}, which we define as immutable, and 2) the
 subjects' existing level of \emph{debt}, which we define as mutable.
@@
 Counterfactuals are valid but involve some substantial reductions in
-\emph{debt} for some individuals (very young applicants). By comparison,
-counterfactual paths are shorter on average in panel (d), where we have
-used counterfactual training and protected immutable features as
-described in Section~\ref{sec-constraints}. In particular, we observe
-that due to the classifier's lower sensitivity to \emph{age}, recourse
-recommendations with respect to \emph{debt} are much more homogenous, in
-that they do not disproportionately punish younger individuals. The
-counterfactuals are also plausible with respect to the mutable feature.
-Thus, we consider the model in panel (d) as the most explainable
-according to Definition~\ref{def-explainability}.
+\emph{debt} for some individuals, in particular very young applicants.
+By comparison, counterfactual paths are shorter on average in panel (d),
+where we have used counterfactual training and protected immutable
+features as described in Section~\ref{sec-constraints}. In particular,
+we observe that due to the classifier's lower sensitivity to \emph{age},
+recourse recommendations with respect to \emph{debt} are much more
+homogenous, in that they do not disproportionately punish younger
+individuals. The counterfactuals are also plausible with respect to the
+mutable feature. Thus, we consider the model in panel (d) as the most
+explainable according to Definition~\ref{def-explainability}.
 
@@
 
-We run the experiments with three CE generators: \emph{Generic} of
-Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}) as a simple baseline
-approach, \emph{REVISE} (\citeproc{ref-joshi2019realistic}{Joshi et al.
-2019}) that aims to generate plausible counterfactuals using a surrogate
-Variational Autoencoder (VAE), and \emph{ECCo}---the generator of
-Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023}) but without
-the conformal prediction component---as a method that directly targets
-both faithfulness and plausibility of the CEs.
-
 \subsection{Experimental Results}\label{experimental-results}
-
-\subsubsection{Plausibility}\label{plausibility-1}
-
-\subsubsection{Actionability}\label{actionability-1}
-
-\subsubsection{Impact of hyperparameter
-settings}\label{impact-of-hyperparameter-settings}
-
-We extensively test the impact of three types of hyperparameters on the
-proposed training regime. Our complete results are available in the
-technical appendix; this section focuses on the main findings.
-
-\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
-that CT is highly sensitive to hyperparameter settings but (a) there are
-manageable patterns and (b) we can typically identify settings that
-improve either plausibility or cost, and commonly both of them at the
-same time. Second, we note that the choice of a CE generator has a major
-impact on the results. For example, \emph{REVISE} tends to perform the
-worst, most likely because it attempts to explain the decisions of a VAE
-that is an imperfect surrogate of the original model. Third, increasing
-\(T\), the maximum number of steps, generally yields better outcomes
-because more CEs can mature in any iteration. Fourth, the impact of
-\(\tau\), the required decision threshold is more difficult to predict.
-On ``harder'' datasets it may be difficult to satisfy high \(\tau\) for
-any given sample (i.e., also factuals) and so increasing this threshold
-does not seem to correlate with better outcomes.
-
-\textbf{\emph{Hyperparameters for penalties}} We find that the strength
-of the energy regularization, \(\lambda_{\text{reg}}\) is highly
-impactful; energy must be sufficiently regularized to avoid poor
-performance in terms of decreased plausibility and increased costs. The
-sensitivity with respect to \(\lambda_{\text{div}}\) and
-\(\lambda_{\text{adv}}\) is much less evident. While high values of
-\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
-combined with high values for Equation~\ref{eq-obj} penalties, this
-effect is not very pronounced.
-
-\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
-and stability of CT is positively associated with the number of
-counterfactuals generated during each training epoch. We also confirm
-that a higher number of training epochs is beneficial. Interestingly, we
-find that it is not necessary to employ CT during the entire training
-phase to achieve the desired improvements in explainability. We have
-tested training models conventionally during the first half of training
-before switching to CT after this initial ``burn-in'' period and
-observed positive results. Put differently, CT may be a way to improve
-the explainability of trained models in a fine-tuning manner.
 
@@
 
-\bibitem[\citeproctext]{ref-altmeyer2023faithful}
-Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia C.
-S. Liem. 2023. {``Faithful Model Explanations Through Energy-Constrained
-Conformal Counterfactuals.''} \url{https://arxiv.org/abs/2312.10648}.
+\bibitem[\citeproctext]{ref-altmeyer2023explaining}
+Altmeyer, Patrick, Arie van Deursen, et al. 2023. {``Explaining
+Black-Box Models Through Counterfactuals.''} In \emph{Proceedings of the
+JuliaCon Conferences}, 1:130. 1.
 
@@
 Conference on Artificial Intelligence}, 38:10829--37. 10.
-
-\bibitem[\citeproctext]{ref-altmeyer2023explaining}
-Altmeyer, Patrick, Arie van Deursen, and Cynthia C. S. Liem. 2023.
-{``Explaining Black-Box Models Through Counterfactuals.''} In
-\emph{Proceedings of the JuliaCon Conferences}, 1:130. 1.
 
@@
 
+\bibitem[\citeproctext]{ref-bezanson2017julia}
+Bezanson, Jeff, Alan Edelman, Stefan Karpinski, and Viral B Shah. 2017.
+{``Julia: A Fresh Approach to Numerical Computing.''} \emph{SIAM Review}
+59 (1): 65--98. \url{https://doi.org/10.1137/141000671}.
+
+\bibitem[\citeproctext]{ref-milan2023dataframes}
+Bouchet-Valat, Milan, and Bogumił Kamiński. 2023. {``DataFrames.jl:
+Flexible and Fast Tabular Data in Julia.''} \emph{Journal of Statistical
+Software} 107 (4): 1--32. \url{https://doi.org/10.18637/jss.v107.i04}.
+
+\bibitem[\citeproctext]{ref-byrne2021mpi}
+Byrne, Simon, Lucas C. Wilcox, and Valentin Churavy. 2021. {``MPI.jl:
+Julia Bindings for the Message Passing Interface.''} \emph{Proceedings
+of the JuliaCon Conferences} 1 (1): 68.
+\url{https://doi.org/10.21105/jcon.00068}.
+
+\bibitem[\citeproctext]{ref-chagas2024pretty}
+Chagas, Ronan Arraes Jardim, Ben Baumgold, Glen Hertz, Hendrik Ranocha,
+Mark Wells, Nathan Boyer, Nicholas Ritchie, et al. 2024.
+{``Ronisbr/PrettyTables.jl: V2.4.0.''} Zenodo.
+\url{https://doi.org/10.5281/zenodo.13835553}.
+
+\bibitem[\citeproctext]{ref-PlotsJL}
+Christ, Simon, Daniel Schwabeneder, Christopher Rackauckas, Michael
+Krabbe Borregaard, and Thomas Breloff. 2023. {``Plots.jl -- a User
+Extendable Plotting API for the Julia Programming Language.''}
+https://doi.org/\url{https://doi.org/10.5334/jors.431}.
+
+\bibitem[\citeproctext]{ref-danisch2021makie}
+Danisch, Simon, and Julius Krumbiegel. 2021. {``{Makie.jl}: Flexible
+High-Performance Data Visualization for {Julia}.''} \emph{Journal of
+Open Source Software} 6 (65): 3349.
+\url{https://doi.org/10.21105/joss.03349}.
+
 \bibitem[\citeproctext]{ref-du2019implicit}
@@
 \url{https://arxiv.org/abs/1903.08689}.
-
-\bibitem[\citeproctext]{ref-frankle2018lottery}
-Frankle, Jonathan, and Michael Carbin. 2019. {``The Lottery Ticket
-Hypothesis: Finding Sparse, Trainable Neural Networks.''} In
-\emph{International Conference on Learning Representations}.
-\url{https://openreview.net/forum?id=rJl-b3RcF7}.
 
@@
 
+\bibitem[\citeproctext]{ref-hastie2009elements}
+Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
+Elements of Statistical Learning}. Springer New York.
+\url{https://doi.org/10.1007/978-0-387-84858-7}.
+
+\bibitem[\citeproctext]{ref-innes2018fashionable}
+Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco
+Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and
+Viral Shah. 2018. {``Fashionable Modelling with Flux.''}
+\url{https://arxiv.org/abs/1811.01457}.
+
+\bibitem[\citeproctext]{ref-innes2018flux}
+Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
+\emph{Journal of Open Source Software} 3 (25): 602.
+\url{https://doi.org/10.21105/joss.00602}.
+
 \bibitem[\citeproctext]{ref-joshi2019realistic}
@@
 Information and Computation}, 881--88.
-
-\bibitem[\citeproctext]{ref-mcgregor2021preventing}
-McGregor, Sean. 2021. {``{Preventing repeated real world AI failures by
-cataloging incidents: The AI incident database}.''} In \emph{Proceedings
-of the AAAI Conference on Artificial Intelligence}, 35:15458--63. 17.
-
-\bibitem[\citeproctext]{ref-morcos2019success}
-Morcos, Ari S., Haonan Yu, Michela Paganini, and Yuandong Tian. 2019.
-{``One Ticket to Win Them All: Generalizing Lottery Ticket
-Initializations Across Datasets and Optimizers.''} In \emph{Proceedings
-of the 33rd International Conference on Neural Information Processing
-Systems}. Red Hook, NY, USA: Curran Associates Inc.
 
@@
 
-\bibitem[\citeproctext]{ref-zhang2021understanding}
-Zhang, Chiyuan, Samy Bengio, Moritz Hardt, Benjamin Recht, and Oriol
-Vinyals. 2021. {``Understanding Deep Learning (Still) Requires
-Rethinking Generalization.''} \emph{Commun. ACM} 64 (3): 107--15.
-\url{https://doi.org/10.1145/3446776}.
-
 \end{CSLReferences}
@@
 
+\FloatBarrier
+
+\renewcommand{\thesection}{\Alph{section}}
+
+\renewcommand{\thetable}{A\arabic{table}}
+
+\renewcommand{\thefigure}{A\arabic{figure}}
+
+\section{Notation}\label{notation}
+
+\begin{itemize}
+\tightlist
+\item
+  \(y^+\): The target class and also the index of the target class.
+\item
+  \(y^-\): The non-target class and also the index of non-the target
+  class.
+\item
+  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
+  class.
+\item
+  \(\theta\): Model parameters (unspecified).
+\item
+  \(\Theta\): Matrix of parameters.
+\end{itemize}
+
+\subsection{Other Technical Details}\label{other-technical-details}
+
+\begin{equation}\phantomsection\label{eq-mmd}{
+\begin{aligned}
+MMD({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
+\end{aligned}
+}\end{equation}
+
+\section{Technical Details of Our
+Approach}\label{technical-details-of-our-approach}
+
+\subsection{Generating Counterfactuals through Gradient
+Descent}\label{sec-app-ce}
+
+In this section, we provide some background on gradient-based
+counterfactual generators (Section~\ref{sec-app-ce-background}) and
+discuss how we define convergence in this context
+(Section~\ref{sec-app-conv}).
+
+\subsubsection{Background}\label{sec-app-ce-background}
+
+Gradient-based counterfactual search was originally proposed by Wachter,
+Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}). It generally solves
+the following unconstrained objective,
+
+\[
+\begin{aligned}
+\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
+\end{aligned} 
+\]
+
+where \(g: \mathcal{Z} \mapsto \mathcal{X}\) is an invertible function
+that maps from the \(L\)-dimensional counterfactual state space to the
+feature space and \(\text{cost}(\cdot)\) denotes one or more penalties
+that are used to induce certain properties of the counterfactual
+outcome. As above, \(\mathbf{y}^+\) denotes the target output and
+\(\mathbf{M}_{\theta}(\mathbf{x})\) returns the logit predictions of the
+underlying classifier for \(\mathbf{x}=g(\mathbf{z})\).
+
+For all generators used in this work we use standard logit crossentropy
+loss for \(\text{yloss}(\cdot)\). All generators also penalize the
+distance (\(\ell_1\)-norm) of counterfactuals from their original
+factual state. For \emph{Generic} and \emph{ECCo}, we have
+\(\mathcal{Z}:=\mathcal{X}\) and
+\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
+are searched directly in the feature space. Conversely, \emph{REVISE}
+traverses the latent space of a variational autoencoder (VAE) fitted to
+the training data, where \(g(\cdot)\) corresponds to the decoder
+(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
+the distance penalty, \emph{ECCo} uses an additional penalty component
+that regularizes the energy associated with the counterfactual,
+\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
+al. 2024}).
+
+\subsubsection{Convergence}\label{sec-app-conv}
+
+An important consideration when generating counterfactual explanations
+using gradient-based methods is how to define convergence. Two common
+choices are to 1) perform gradient descent over a fixed number of
+iterations \(T\), or 2) conclude the search as soon as the predicted
+probability for the target class has reached a pre-determined threshold,
+\(\tau\):
+\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\). We
+prefer the latter for our purposes, because it explicitly defines
+convergence in terms of the black-box model, \(\mathbf{M}(\mathbf{x})\).
+
+Defining convergence in this way allows for a more intuitive
+interpretation of the resulting counterfactual outcomes than with fixed
+\(T\). Specifically, it allows us to think of counterfactuals as
+explaining `high-confidence' predictions by the model for the target
+class \(y^+\). Depending on the context and application, different
+choices of \(\tau\) can be considered as representing `high-confidence'
+predictions.
+
+\subsection{Protecting Mutability Constraints with Linear
+Classifiers}\label{sec-app-constraints}
+
+In Section~\ref{sec-constraints} we explain that to avoid penalizing
+implausibility that arises due to mutability constraints, we impose a
+point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
+argue in Section~\ref{sec-constraints} that this approach induces models
+to be less sensitive to immutable features and demonstrate this
+empirically in Section~\ref{sec-experiments}. Below we derive the
+analytical results in Proposition~\ref{prp-mtblty}.
+
+\begin{proof}
+Let \(d_{\text{mtbl}}\) and \(d_{\text{immtbl}}\) denote some mutable
+and immutable feature, respectively. Suppose that
+\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
+\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
+\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
+class \(k\). In words, we assume that the immutable feature tends to
+take lower values for samples in the non-target class \(y^-\) than in
+the target class \(y^+\). We assume the opposite to hold for the mutable
+feature.
+
+Assuming multivariate Gaussian class densities with common diagonal
+covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), we
+have for the log likelihood ratio between any two classes
+\(k,m \in \mathcal{K}\) (\citeproc{ref-hastie2009elements}{Hastie,
+Tibshirani, and Friedman 2009}):
+
+\begin{equation}\phantomsection\label{eq-loglike}{
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
+}\end{equation}
+
+By independence of \(x_1,...,x_D\), the full log-likelihood ratio
+decomposes into:
+
+\begin{equation}\phantomsection\label{eq-loglike-decomp}{
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
+}\end{equation}
+
+By the properties of our classifier (\emph{multinomial logistic
+regression}), we have:
+
+\begin{equation}\phantomsection\label{eq-multi}{
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
+}\end{equation}
+
+where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
+\(d\) for class \(k\).
+
+Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
+can identify that
+\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
+the assumptions we made above. Hence, we have that
+\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
+and
+\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)
+
+Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
+class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
+target class. Then the partial derivative of the contrastive divergence
+penalty Equation~\ref{eq-div} with respect to coefficient
+\(\theta_{y^+,d}\) is equal to
+
+\begin{equation}\phantomsection\label{eq-grad}{
+\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
+}\end{equation}
+
+and equal to zero everywhere else.
+
+Since \((\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})\) we
+are more likely to have
+\((x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0\) than vice
+versa at initialization. Similarly, we are more likely to have
+\((x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0\) since
+\((\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})\).
+
+This implies that if we do not protect feature \(d_{\text{immtbl}}\),
+the contrastive divergence penalty will decrease
+\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
+effect
+\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
+In words, not protecting the immutable feature would have the
+undesirable effect of making the classifier more sensitive to this
+feature, in that it would be more likely to predict class \(y^-\) as
+opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).
+
+By the same rationale, the contrastive divergence penalty can generally
+be expected to increase \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
+\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
+In words, this has the effect of making the classifier more sensitive to
+the mutable feature, in that it would be more likely to predict class
+\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).
+
+Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
+has the net affect of decreasing the classifier's sensitivity to the
+immutable feature relative to the mutable feature (i.e.~no change in
+sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
+for \(d_{\text{mtbl}}\)).
+\end{proof}
+
+\subsection{Domain Constraints}\label{domain-constraints}
+
+We apply domain constraints on counterfactuals during training and
+evaluation. There are at least two good reasons for doing so. Firstly,
+within the context of explainability and algorithmic recourse,
+real-world attributes are often domain constrained: the \emph{age}
+feature, for example, is lower bounded by zero and upper bounded by the
+maximum human lifespan. Secondly, domain constraints help mitigate
+training instabilities commonly associated with energy-based modelling
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020};
+\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+
+For our image datasets, features are pixel values and hence the domain
+is constrained by the lower and upper bound of values that pixels can
+take depending on how they are scaled (in our case \([-1,1]\)). For all
+other features \(d\) in our synthetic and tabular datasets, we
+automatically infer domain constraints
+\([x_d^{\text{LB}},x_d^{\text{UB}}]\) as follows,
+
+\begin{equation}\phantomsection\label{eq-domain}{
+\begin{aligned}
+x_d^{\text{LB}} &= \arg\min_{x_d} \{\mu_d - n_{\sigma_d}\sigma_d, \arg \min_{x_d} x_d\} \\
+x_d^{\text{UB}} &= \arg\max_{x_d} \{\mu_d + n_{\sigma_d}\sigma_d, \arg \max_{x_d} x_d\} 
+\end{aligned}
+}\end{equation}
+
+where \(\mu_d\) and \(\sigma_d\) denote the sample mean and standard
+deviation of feature \(d\). We set \(n_{\sigma_d}=3\) across the board
+but higher values and hence wider bounds may be appropriate depending on
+the application.
+
+\subsection{Training Details}\label{sec-app-training}
+
+In this section, we describe the training procedure in detail. While the
+details laid out here are not crucial for understanding our proposed
+approach, they are of importance to anyone looking to implement
+counterfactual training.
+
+\section{Detailed Results}\label{detailed-results}
+
+\subsection{Qualitative Findings for Image
+Data}\label{qualitative-findings-for-image-data}
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+
+Figure~\ref{fig-mnist} shows much more plausible (faithful)
+counterfactuals for a model with CT than the model with conventional
+training (Figure~\ref{fig-mnist-vanilla}). In fact, this is not even
+using \emph{ECCo+} and still showing better results than the best
+results we achieved in our AAAI paper for JEM ensembles.
+
+\end{tcolorbox}
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp.png}}
+
+}
+
+\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
+counterfactual training. The underlying generator, \emph{ECCo}, aims to
+generate counterfactuals that are faithful to the model
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp_vanilla.png}}
+
+}
+
+\caption{\label{fig-mnist-vanilla}Counterfactual images for \emph{MLP}
+with conventional training. The underlying generator, \emph{ECCo}, aims
+to generate counterfactuals that are faithful to the model
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
+
+\end{figure}%
+
+\subsection{Grid Searches}\label{sec-app-grid}
+
+To assess the hyperparameter sensitivity of our proposed training regime
+we ran multiple large grid searches for all of our synthetic datasets.
+We have grouped these grid searches into multiple categories:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \textbf{Generator Parameters} (Section~\ref{sec-app-grid-gen}):
+  Investigates the effect of changing hyperparameters that affect the
+  counterfactual outcomes during the training phase.
+\item
+  \textbf{Penalty Strengths} (Section~\ref{sec-app-grid-pen}):
+  Investigates the effect of changing the penalty strengths in out
+  proposed objective (Equation~\ref{eq-obj}).
+\item
+  \textbf{Other Parameters} (Section~\ref{sec-app-grid-train}):
+  Investigates the effect of changing other training parameters,
+  including the total number of generated counterfactuals in each epoch.
+\end{enumerate}
+
+We begin by summarizing the high-level findings in
+Section~\ref{sec-app-grid-hl}. For each of the categories,
+Section~\ref{sec-app-grid-gen} to Section~\ref{sec-app-grid-train} then
+present all details including the exact parameter grids, average
+predictive performance outcomes and key evaluation metrics for the
+generated counterfactuals.
+
+\subsubsection{Evaluation Details}\label{evaluation-details}
+
+To measure predictive performance, we compute the accuracy and F1-score
+for all models on test data (Table~\ref{tbl-acc-gen},
+Table~\ref{tbl-acc-pen}, Table~\ref{tbl-acc-train}). With respect to
+explanatory performance, we report here our findings for the
+(im)plausibility and cost of counterfactuals at test time. Since the
+computation of our proposed divergence metric
+(Equation~\ref{eq-impl-div}) is memory-intensive, we rely on the
+distance-based metric for the grid searches. For the counterfactual
+evaluation, we draw factual samples from the training data for the grid
+searches to avoid data leakage with respect to our final results
+reported in the body of the paper. Specifically, we want to avoid
+choosing our default hyperparameters based on results on the test data.
+Since we are optimizing for explainability, not predictive performance,
+we still present test accuracy and F1-scores.
+
+\paragraph{Predictive Performance}\label{predictive-performance}
+
+We find that CT is associated with little to no decrease in average
+predictive performance for our synthetic datasets: test accuracy and
+F1-scores decrease by at most \textasciitilde1 percentage point, but
+generally much less (Table~\ref{tbl-acc-gen}, Table~\ref{tbl-acc-pen},
+Table~\ref{tbl-acc-train}). Variation across hyperparameters is
+negligible as indicated by small standard deviations for these metrics
+across the board.
+
+\paragraph{Counterfactual Outcomes}\label{sec-app-grid-hl}
+
+Overall, we find that Counterfactual Training (CT) achieves it key
+objectives consistently across all hyperparameter settings and also
+broadly across datasets: plausibility is improved by up to
+\textasciitilde60 percentage points (ppts) for the \emph{Circles} data
+(e.g. Figure~\ref{fig-grid-gen_params-plaus-circles}),
+\textasciitilde25-30ppts for the \emph{Moons} data (e.g.
+Figure~\ref{fig-grid-gen_params-plaus-moons}) and
+\textasciitilde10-20ppts for the \emph{Linearly Separable} data (e.g.
+Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
+average costs of faithful counterfactuals are reduced in many cases by
+around \textasciitilde20-25ppts for \emph{Circles} (e.g.
+Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
+\textasciitilde50ppts for \emph{Moons} (e.g.
+Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
+Separable} data, costs are generally increased although typically by
+less than 10ppts (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
+which reflects a common tradeoff between costs and plausibility
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+
+We do observe strong sensitivity to certain hyperparameters, with clear
+an manageable patterns. Concerning generator parameters, we firstly find
+that using \emph{REVISE} to generate counterfactuals during training
+typically yields the worst outcomes out of all generators, often leading
+to a substantial decrease in plausibility. This finding can be
+attributed to the fact that \emph{REVISE} effectively assigns the task
+of learning plausible explanations from the model itself to a surrogate
+VAE. In other words, counterfactuals generated by \emph{REVISE} are less
+faithful to the model that \emph{ECCo} and \emph{Generic}, and hence we
+would expect them to be a less effective and, in fact, potentially
+detrimental role in our training regime. Secondly, we observe that
+allowing for a higher number of maximum steps \(T\) for the
+counterfactual search generally yields better outcomes. This is
+intuitive, because it allows more counterfactuals to reach maturity in
+any given iteration. Looking in particular at the results for
+\emph{Linearly Separable}, it seems that higher values for \(T\) in
+combination with higher decision thresholds (\(\tau\)) yields the best
+results when using \emph{ECCo}. But depending on the degree of class
+separability of the underlying data, a high decision-threshold can also
+affect results adversely, as evident from the results for the
+\emph{Overlapping} data (Figure~\ref{fig-grid-gen_params-plaus-over}):
+here we find that CT generally fails to achieve its objective because
+only a tiny proportion of counterfactuals ever reaches maturity.
+
+Regarding penalty strengths, we find that the strength of the energy
+regularization, \(\lambda_{\text{reg}}\) is a key hyperparameter, while
+sensitivity with respect to \(\lambda_{\text{div}}\) and
+\(\lambda_{\text{adv}}\) is much less evident. In particular, we observe
+that not regularizing energy enough or at all typically leads to poor
+performance in terms of decreased plausibility and increased costs, in
+particular for \emph{Circles} (Figure~\ref{fig-grid-pen-plaus-circles}),
+\emph{Linearly Separable} (Figure~\ref{fig-grid-pen-plaus-lin_sep}) and
+\emph{Overlapping} (Figure~\ref{fig-grid-pen-plaus-over}). High values
+of \(\lambda_{\text{reg}}\) can increase the variability in outcomes, in
+particular when combined with high values for \(\lambda_{\text{div}}\)
+and \(\lambda_{\text{adv}}\), but this effect is less pronounced.
+
+Finally, concerning other hyperparameters we observe that the
+effectiveness and stability of CT is positively associated with the
+number of counterfactuals generated during each training epoch, in
+particular for \emph{Circles}
+(Figure~\ref{fig-grid-train-plaus-circles}) and \emph{Moons}
+(Figure~\ref{fig-grid-train-plaus-moons}). We further find that a higher
+number of training epochs is beneficial as expected, where we tested
+training models for 50 and 100 epochs. Interestingly, we find that it is
+not necessary to employ CT during the entire training phase to achieve
+the desired improvements in explainability: specifically, we have tested
+training models conventionally during the first half of training before
+switching to CT after this initial burn-in period.
+
+\subsubsection{Generator Parameters}\label{sec-app-grid-gen}
+
+The hyperparameter grid with varying generator parameters during
+training is shown in Note~\ref{nte-gen-params-final-run-train}. The
+corresponding evaluation grid used for these experiments is shown in
+Note~\ref{nte-gen-params-final-run-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-gen-params-final-run-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Decision Threshold: \texttt{0.75,\ 0.9,\ 0.95}
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 5.0,\ 10.0,\ 20.0}
+  \item
+    Maximum Iterations: \texttt{5,\ 25,\ 50}
+  \end{itemize}
+\item
+  Generator: \texttt{ecco,\ generic,\ revise}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-gen-params-final-run-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Accuracy}\label{accuracy}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-acc-gen}Predictive performance measures by dataset
+and objective averaged across training-phase parameters
+(Note~\ref{nte-gen-params-final-run-train}) and evaluation-phase
+parameters (Note~\ref{nte-gen-params-final-run-eval}).}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Circles & Accuracy & Full & 0.997 & 0.00309 \\
+  Circles & Accuracy & Vanilla & 0.998 & 0.000557 \\
+  Circles & F1-score & Full & 0.997 & 0.00309 \\
+  Circles & F1-score & Vanilla & 0.998 & 0.000558 \\
+  Lin Sep & Accuracy & Full & 0.999 & 0.00201 \\
+  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
+  Lin Sep & F1-score & Full & 0.999 & 0.00201 \\
+  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Moons & Accuracy & Full & 0.999 & 0.000696 \\
+  Moons & Accuracy & Vanilla & 1 & 0.00111 \\
+  Moons & F1-score & Full & 0.999 & 0.000696 \\
+  Moons & F1-score & Vanilla & 1 & 0.00111 \\
+  Over & Accuracy & Full & 0.915 & 0.00477 \\
+  Over & Accuracy & Vanilla & 0.917 & 0.00123 \\
+  Over & F1-score & Full & 0.915 & 0.00478 \\
+  Over & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Plausibility}\label{plausibility-1}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-grid-gen_params-plaus-circles} to
+Figure~\ref{fig-grid-gen_params-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-circles}Average outcomes for
+the plausibility measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-lin_sep}Average outcomes for
+the plausibility measure across hyperparameters. Data: Linearly
+Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-moons}Average outcomes for the
+plausibility measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-over}Average outcomes for the
+plausibility measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Cost}\label{cost}
+
+The results with respect to the cost measure are shown in
+Figure~\ref{fig-grid-gen_params-cost-circles} to
+Figure~\ref{fig-grid-gen_params-cost-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-circles}Average outcomes for
+the cost measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-lin_sep}Average outcomes for
+the cost measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-moons}Average outcomes for the
+cost measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-over}Average outcomes for the
+cost measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\subsubsection{Penalty Strengths}\label{sec-app-grid-pen}
+
+The hyperparameter grid with varying penalty strengths during training
+is shown in Note~\ref{nte-pen-final-run-train}. The corresponding
+evaluation grid used for these experiments is shown in
+Note~\ref{nte-pen-final-run-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-pen-final-run-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator: \texttt{ecco,\ generic,\ revise}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{adv}}\): \texttt{0.1,\ 0.25,\ 1.0}
+  \item
+    \(\lambda_{\text{div}}\): \texttt{0.01,\ 0.1,\ 1.0}
+  \item
+    \(\lambda_{\text{reg}}\): \texttt{0.0,\ 0.01,\ 0.1,\ 0.25,\ 0.5}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-pen-final-run-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Accuracy}\label{accuracy-1}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-acc-pen}Predictive performance measures by dataset
+and objective averaged across training-phase parameters
+(Note~\ref{nte-pen-final-run-train}) and evaluation-phase parameters
+(Note~\ref{nte-pen-final-run-eval}).}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Circles & Accuracy & Full & 0.994 & 0.0144 \\
+  Circles & Accuracy & Vanilla & 0.998 & 0.000875 \\
+  Circles & F1-score & Full & 0.994 & 0.0145 \\
+  Circles & F1-score & Vanilla & 0.998 & 0.000875 \\
+  Lin Sep & Accuracy & Full & 0.998 & 0.00772 \\
+  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
+  Lin Sep & F1-score & Full & 0.998 & 0.00773 \\
+  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Moons & Accuracy & Full & 0.987 & 0.0351 \\
+  Moons & Accuracy & Vanilla & 0.998 & 0.0101 \\
+  Moons & F1-score & Full & 0.987 & 0.0352 \\
+  Moons & F1-score & Vanilla & 0.998 & 0.0102 \\
+  Over & Accuracy & Full & 0.911 & 0.0217 \\
+  Over & Accuracy & Vanilla & 0.916 & 0.00236 \\
+  Over & F1-score & Full & 0.911 & 0.0219 \\
+  Over & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Plausibility}\label{plausibility-2}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-grid-pen-plaus-circles} to
+Figure~\ref{fig-grid-pen-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-circles}Average outcomes for the
+plausibility measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-lin_sep}Average outcomes for the
+plausibility measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-moons}Average outcomes for the
+plausibility measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-over}Average outcomes for the
+plausibility measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Cost}\label{cost-1}
+
+The results with respect to the cost measure are shown in
+Figure~\ref{fig-grid-pen-cost-circles} to
+Figure~\ref{fig-grid-pen-cost-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-circles}Average outcomes for the cost
+measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-lin_sep}Average outcomes for the cost
+measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-moons}Average outcomes for the cost
+measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-over}Average outcomes for the cost
+measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\subsubsection{Other Parameters}\label{sec-app-grid-train}
+
+The hyperparameter grid with other varying training parameters is shown
+in Note~\ref{nte-train-final-run-train}. The corresponding evaluation
+grid used for these experiments is shown in
+Note~\ref{nte-train-final-run-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-train-final-run-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator: \texttt{ecco,\ generic,\ revise}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Burnin: \texttt{0.0,\ 0.5}
+  \item
+    No.~Counterfactuals: \texttt{100,\ 1000}
+  \item
+    No.~Epochs: \texttt{50,\ 100}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-train-final-run-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Accuracy}\label{accuracy-2}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-acc-train}Predictive performance measures by dataset
+and objective averaged across training-phase parameters
+(Note~\ref{nte-train-final-run-train}) and evaluation-phase parameters
+(Note~\ref{nte-train-final-run-eval}).}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Circles & Accuracy & Full & 0.995 & 0.00431 \\
+  Circles & Accuracy & Vanilla & 0.998 & 0.000566 \\
+  Circles & F1-score & Full & 0.995 & 0.00432 \\
+  Circles & F1-score & Vanilla & 0.998 & 0.000566 \\
+  Lin Sep & Accuracy & Full & 0.999 & 0.00231 \\
+  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
+  Lin Sep & F1-score & Full & 0.999 & 0.00231 \\
+  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Moons & Accuracy & Full & 0.996 & 0.0136 \\
+  Moons & Accuracy & Vanilla & 0.988 & 0.022 \\
+  Moons & F1-score & Full & 0.996 & 0.0136 \\
+  Moons & F1-score & Vanilla & 0.988 & 0.022 \\
+  Over & Accuracy & Full & 0.914 & 0.00563 \\
+  Over & Accuracy & Vanilla & 0.918 & 0.00116 \\
+  Over & F1-score & Full & 0.914 & 0.0057 \\
+  Over & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Plausibility}\label{plausibility-3}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-grid-train-plaus-circles} to
+Figure~\ref{fig-grid-train-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-circles}Average outcomes for the
+plausibility measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-lin_sep}Average outcomes for the
+plausibility measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-moons}Average outcomes for the
+plausibility measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-over}Average outcomes for the
+plausibility measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Cost}\label{cost-2}
+
+The results with respect to the cost measure are shown in
+Figure~\ref{fig-grid-train-cost-circles} to
+Figure~\ref{fig-grid-train-cost-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-circles}Average outcomes for the
+cost measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-lin_sep}Average outcomes for the
+cost measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-moons}Average outcomes for the cost
+measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-over}Average outcomes for the cost
+measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\subsection{Hyperparameter Tuning}\label{sec-app-tune}
+
+Based on the findings from our initial large grid searches
+(Section~\ref{sec-app-grid}), we tune selected hyperparameters for all
+datasets: namely, the decision threshold \(\tau\) and the strength of
+the energy regularization \(\lambda_{\text{reg}}\). The final
+hyperparameter choices for each dataset are presented in \textbf{ADD
+TABLE}. Detailed results for each data set are shown in
+Figure~\ref{fig-tune-plaus-adult} to Figure~\ref{fig-tune-mat-over}.
+From \textbf{ADD TABLE}, we notice that the same decision threshold of
+\(\tau=0.5\) is optimal for all but on dataset. We attribute this to the
+fact that a low decision threshold results in a higher share of mature
+counterfactuals and hence more opportunities for the model to learn from
+examples (Figure~\ref{fig-tune-mat-adult} to
+Figure~\ref{fig-tune-mat-over}). This has played a role in particular
+for our real-world tabular datasets and MNIST, which suffered from low
+levels of maturity for higher decision thresholds. In cases where
+maturity is not an issue, as for \emph{Moons}, higher decision
+thresholds lead to better outcomes, which may have to do with the fact
+that the resulting counterfactuals are more faithful to the model.
+Concerning the regularization strength, we find somewhat high variation
+across datasets. Most notably, we find that relatively low levels of
+regularization are optimal for MNIST. We hypothesize that this finding
+may be attributed to the uniform scaling of all input features (digits).
+Finally, to increase the proportion of mature counterfactuals for some
+datasets, we have also investigated the effect on the learning rate
+\(\eta\) for the counterfactual search, but found little effect on the
+results.
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Package Version (Reproducibility)}, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm, opacityback=0]
+
+Tuning was run using \texttt{v1.1.3} of \texttt{TaijaData}. The
+follow-up version \texttt{v1.1.4} introduced an option to split
+real-world tabular datasets into train and test set, ensuring that
+pre-processing steps like standardization is fit on the training set
+only. If you are rerunning the tuning experiments with a version of
+\texttt{TaijaData} that is higher than \texttt{v1.1.3}, than for the
+default parameters specified in the configuration files, you may end up
+with slightly different results, although we would not expect any
+changes in terms of qualitative findings. For exact reproducibility,
+please use \texttt{v1.1.3}.
+
+\end{tcolorbox}
+
+\subsubsection{Key Parameters}\label{sec-app-tune}
+
+The hyperparameter grid for tuning key parameters is shown in
+Note~\ref{nte-tune-train}. The corresponding evaluation grid used for
+these experiments is shown in Note~\ref{nte-tune-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-tune-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Decision Threshold: \texttt{0.5,\ 0.75,\ 0.9}
+  \end{itemize}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{reg}}\): \texttt{0.1,\ 0.25,\ 0.5}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-tune-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Plausibility}\label{plausibility-4}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-tune-plaus-adult} to Figure~\ref{fig-tune-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-adult}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-cali}Average outcomes for the
+plausibility measure across key hyperparameters. Data: California
+Housing.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-circles}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-credit}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-gmsc}Average outcomes for the
+plausibility measure across key hyperparameters. Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-lin_sep}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Linearly
+Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-mnist}Average outcomes for the
+plausibility measure across key hyperparameters. Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-moons}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-over}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Proportion of Mature CE}\label{proportion-of-mature-ce}
+
+The results with respect to the proportion of mature counterfactuals in
+each epoch are shown in Figure~\ref{fig-tune-mat-adult} to
+Figure~\ref{fig-tune-mat-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-adult}Proportion of mature counterfactuals
+in each epoch. Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-cali}Proportion of mature counterfactuals
+in each epoch. Data: California Housing.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-circles}Proportion of mature
+counterfactuals in each epoch. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-credit}Proportion of mature counterfactuals
+in each epoch. Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-gmsc}Proportion of mature counterfactuals
+in each epoch. Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-lin_sep}Proportion of mature
+counterfactuals in each epoch. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-mnist}Proportion of mature counterfactuals
+in each epoch. Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-moons}Proportion of mature counterfactuals
+in each epoch. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-over}Proportion of mature counterfactuals
+in each epoch. Data: Overlapping.}
+
+\end{figure}%
+
+\section{Computation Details}\label{computation-details}
+
+\subsection{Hardware}\label{hardware}
+
+We performed our experiments on a high-performance cluster. Details
+about the cluster will be disclosed upon publication to avoid revealing
+information that might interfere with the double-blind review process.
+Since our experiments involve highly parallel tasks and rather small
+models by today's standard, we have relied on distributed computing
+across multiple central processing units (CPU). Graphical processing
+units (GPU) were not required.
+
+\subsubsection{Grid Searches}\label{grid-searches}
+
+Model training for the largest grid searches with 270 unique parameter
+combinations was parallelized across 34 CPUs with 2GB memory each. The
+time to completion varied by dataset for reasons discussed in
+Section~\ref{sec-discussion}: 0h49m (\emph{Moons}), 1h4m (\emph{Linearly
+Separable}), 1h49m (\emph{Circles}), 3h52m (\emph{Overlapping}). Model
+evaluations for large grid searches were parallelized across 20 CPUs
+with 3GB memory each. Evaluations for all data sets took less than one
+hour (\textless1h) to complete.
+
+\subsubsection{Tuning}\label{tuning}
+
+For tuning of selected hyperparameters, we distributed the task of
+generating counterfactuals during training across 40 CPUs with 2GB
+memory each for all tabular datasets. Except for the \emph{Adult}
+dataset, all training runs were completed in less that half an hour
+(\textless0h30m). The \emph{Adult} dataset took around 0h35m to
+complete. Evaluations across 20 CPUs with 3GB memory each generally took
+less than 0h30m to complete. For \emph{MNIST}, we relied on 100 CPUs
+with 2GB memory each. For the \emph{MLP}, training of all models could
+be completed in 1h30m, while the evaluation across 20 CPUs (6GB memory)
+took 4h12m. For the \emph{CNN}, training of all models took
+\textasciitilde8h, with conventionally trained models taking
+\textasciitilde0h15m each and model with CT taking
+\textasciitilde0h30m-0h45m each.
+
+\subsection{Software}\label{software}
+
+All computations were performed in the Julia Programming Language
+(\citeproc{ref-bezanson2017julia}{Bezanson et al. 2017}). We have
+developed a package for counterfactual training that leverages and
+extends the functionality provided by several existing packages, most
+notably
+\href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{CounterfactualExplanations.jl}
+(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023})
+and the \href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library for
+deep learning (\citeproc{ref-innes2018fashionable}{Michael Innes et al.
+2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). For
+data-wrangling and presentation-ready tables we relied on
+\href{https://dataframes.juliadata.org/v1.7/}{DataFrames.jl}
+(\citeproc{ref-milan2023dataframes}{Bouchet-Valat and Kamiński 2023})
+and
+\href{https://ronisbr.github.io/PrettyTables.jl/v2.4/}{PrettyTables.jl}
+(\citeproc{ref-chagas2024pretty}{Chagas et al. 2024}), respectively. For
+plots and visualizations we used both
+\href{https://docs.juliaplots.org/v1.40/}{Plots.jl}
+(\citeproc{ref-PlotsJL}{Christ et al. 2023}) and
+\href{https://docs.makie.org/v0.22/}{Makie.jl}
+(\citeproc{ref-danisch2021makie}{Danisch and Krumbiegel 2021}), in
+particular \href{https://aog.makie.org/v0.9.3/}{AlgebraOfGraphics.jl}.
+To distribute computational tasks across multiple processors, we have
+relied on \href{https://juliaparallel.org/MPI.jl/v0.20/}{MPI.jl}
+(\citeproc{ref-byrne2021mpi}{Byrne, Wilcox, and Churavy 2021}).
+
 \end{document}

### 2025-03-12 paper/experiments/CTExperiments/src/tabular.jl

@@ -7,12 +7,45 @@ using Random
 using StatsBase
 using TaijaData
 
+abstract type TabularData <: Dataset end
+
+"""
+    get_data(data::TabularData, test_set::Bool=false)
+
+
+"""
+function get_data(data::TabularData; n::Union{Nothing,Int}=nothing, test_set::Bool=false)
+
+    if exceeds_max(data)
+        @warn "Requesting more data than available (using oversampling)."
+    end
+    navailable = if isinf(nmax(data))
+        100_000
+    else
+        nmax(data)
+    end
+    X, y, Xtest, ytest = load_data(data, navailable)    # load all available data
+
+    # Get data:
+    if test_set
+        X, y = Xtest, ytest
+    end
+    X = Float32.(X)
+
+    # Subset:
+    if !isnothing(n)
+        X, y = take_subset(X, y, n; rng=get_rng(data))
+    end
+
+    return X, y
+end
+
 """
     GMSC
 
 Keyword container for the `GMSC` data set. Can specify the number of samples `n`, the batch size `batchsize`.
 """
-Base.@kwdef mutable struct GMSC <: Dataset
+Base.@kwdef mutable struct GMSC <: TabularData
     n_train::Int = 12371
     batchsize::Int = 1000
     n_validation::Int = 1000
@@ -25,16 +58,23 @@ end
 
 dname(d::GMSC) = "gmsc"
 
+get_name(d::GMSC; pretty::Bool=false) = pretty ? "GMSC" : dname(d)
+
 nmax(d::GMSC) = 16714
 
-load_data(d::GMSC, n::Int) = load_gmsc(n)
+function load_data(d::GMSC, n::Int)
+    data = Logging.with_logger(Logging.NullLogger()) do
+        load_gmsc(n; train_test_split=d.train_test_ratio)
+    end
+    return data
+end
 
 """
     CaliHousing
 
 Keyword container for the `CaliHousing` (california housing) data set. Can specify the number of samples `n`, the batch size `batchsize`.
 """
-Base.@kwdef mutable struct CaliHousing <: Dataset
+Base.@kwdef mutable struct CaliHousing <: TabularData
     n_train::Int = 15504
     batchsize::Int = 1000
     n_validation::Int = 1000
@@ -47,16 +87,23 @@ end
 
 dname(d::CaliHousing) = "cali"
 
+get_name(d::CaliHousing; pretty::Bool=false) = pretty ? "California Housing" : dname(d)
+
[…]

### 2025-03-12 paper/sections/other/tuning.qmd

@@
 The final hyperparameter choices for each dataset are presented in **ADD TABLE**.
-Detailed results for each data set are shown in **ADD FIGURES**.
+Detailed results for each data set are shown in @fig-tune-plaus-adult to @fig-tune-mat-over.
 From **ADD TABLE**, we notice that the same decision threshold of $\tau=0.5$ is optimal for all but on dataset.
-We attribute this to the fact that a low decision threshold results in a higher share of mature counterfactuals and hence more opportunities for the model to learn from examples.
+We attribute this to the fact that a low decision threshold results in a higher share of mature counterfactuals and hence more opportunities for the model to learn from examples (@fig-tune-mat-adult to @fig-tune-mat-over).
 This has played a role in particular for our real-world tabular datasets and MNIST, which suffered from low levels of maturity for higher decision thresholds.
@@
 Finally, to increase the proportion of mature counterfactuals for some datasets, we have also investigated the effect on the learning rate $\eta$ for the counterfactual search, but found little effect on the results.
+
+::: {.callout-warning}
+
+## Package Version (Reproducibility)
+
+Tuning was run using `v1.1.3` of `TaijaData`.
+The follow-up version `v1.1.4` introduced an option to split real-world tabular datasets into train and test set, ensuring that pre-processing steps like standardization is fit on the training set only.
+If you are rerunning the tuning experiments with a version of `TaijaData` that is higher than `v1.1.3`, than for the default parameters specified in the configuration files, you may end up with slightly different results, although we would not expect any changes in terms of qualitative findings.
+For exact reproducibility, please use `v1.1.3`.
+
+:::
+
+### Key Parameters {#sec-app-tune-key}
+
+```{julia}
+grid_dir = joinpath(res_dir, "tune/mlp")
+data_dirs = readdir(grid_dir) |> x -> joinpath.(grid_dir, x) |> x -> x[isdir.(x)]
+eval_grids = (p -> EvaluationGrid(joinpath(p,"grid_config.toml"))).(data_dirs)
+data_names = basename.(data_dirs)
+ce_suffix = "evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/"
+logs_suffix = "evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/"
+```
+
+The hyperparameter grid for tuning key parameters is shown in @nte-tune-train. The corresponding evaluation grid used for these experiments is shown in @nte-tune-eval.
+
+::: {#nte-tune-train .callout-note}
+
+## Training Phase
+
+```{julia}
+#| output: asis
+dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/grid_config.toml")) 
+println(CTExperiments.dict_to_quarto_markdown(dict))
+```
+
+:::
+
+::: {#nte-tune-eval .callout-note}
+
+## Evaluation Phase
+
+```{julia}
+#| output: asis
+dict = CTExperiments.from_toml(joinpath(grid_dir, "lin_sep/evaluation/evaluation_grid_config.toml"))
+println(CTExperiments.dict_to_quarto_markdown(dict))
+```
+
+:::
+
+#### Plausibility
+
+```{julia}
+#| output: asis
+
+fig_label_prefix = "tune-plaus"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the plausibility measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "plausibility_distance_from_target.png"
+fig_caption = "Average outcomes for the plausibility measure across key hyperparameters."
+full_paths = joinpath.(data_dirs, joinpath(ce_suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```
+
+#### Proportion of Mature CE
+
+```{julia}
+#| output: asis
+
+fig_label_prefix = "tune-mat"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the proportion of mature counterfactuals in each epoch are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "percent_valid.png"
+fig_caption = "Proportion of mature counterfactuals in each epoch."
+full_paths = joinpath.(data_dirs, joinpath(logs_suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```

### 2025-03-12 paper/preprint/paper.tex
vamos

@@
 \usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{definition}
@@
 \theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 may be attributed to the uniform scaling of all input features (digits).
+
 Finally, to increase the proportion of mature counterfactuals for some
 datasets, we have also investigated the effect on the learning rate
-\(\eta\) for the counterfactual search, but found little effect on the
-results.
-
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-warning-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Package Version (Reproducibility)}, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm, opacityback=0]
+\(\eta\) for the counterfactual search and even smaller regularization
+strengths for a fixed decision threshold of 0.5
+(Figure~\ref{fig-tune_lr-plaus-adult} to
+Figure~\ref{fig-tune_lr-plaus-over}). For the given low decision
+threshold, we find that the learning rate has no discernable impact on
+the proportion of mature counterfactuals
+(Figure~\ref{fig-tune_lr-mat-adult} to
+Figure~\ref{fig-tune_lr-mat-over}). We do notice, however, that the
+results for MNIST are much improved when using a low value
+\(\lambda_{\text{reg}}\), the strength for the engery regularization:
+plausibility is increased by up to \textasciitilde10ppt
+(Figure~\ref{fig-tune_lr-plaus-mnist}) and the proportion of mature
+counterfactuals reaches 100\%.
+
+One consideration worth exploring is to combine high decision thresholds
+with high learning rates and low energy regularization strengths, which
+we have not investigated here.
+
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Package Version (Reproducibility)}, colbacktitle=quarto-callout-warning-color!10!white, colback=white, colframe=quarto-callout-warning-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 
-\subsubsection{Key Parameters}\label{sec-app-tune}
+\subsubsection{Key Parameters}\label{sec-app-tune-key}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune-train}: Training Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, breakable, toptitle=1mm, rightrule=.15mm, leftrule=.75mm, colback=white, coltitle=black, left=2mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toprule=.15mm, opacitybacktitle=0.6, arc=.35mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
 
@@
 in each epoch. Data: Overlapping.}
+
+\end{figure}%
+
+\subsubsection{Learning Rate}\label{sec-app-tune-lr}
+
+The hyperparameter grid for tuning the learning rate is shown in
+Note~\ref{nte-tune_lr-train}. The corresponding evaluation grid used for
+these experiments is shown in Note~\ref{nte-tune_lr-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-tune_lr-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Learning Rate: \texttt{0.1,\ 0.5,\ 1.0}
+  \end{itemize}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{reg}}\): \texttt{0.01,\ 0.1,\ 0.5}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colbacktitle=quarto-callout-note-color!10!white, colback=white, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, rightrule=.15mm, opacitybacktitle=0.6, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, breakable, bottomrule=.15mm, leftrule=.75mm, toprule=.15mm, opacityback=0]
+
+\quartocalloutnte{nte-tune_lr-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Plausibility}\label{plausibility-5}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-tune_lr-plaus-adult} to
+Figure~\ref{fig-tune_lr-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-adult}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-credit}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-gmsc}Average outcomes for the
+plausibility measure across key hyperparameters. Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-mnist}Average outcomes for the
+plausibility measure across key hyperparameters. Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-over}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Proportion of Mature CE}\label{proportion-of-mature-ce-1}
+
+The results with respect to the proportion of mature counterfactuals in
+each epoch are shown in Figure~\ref{fig-tune_lr-mat-adult} to
+Figure~\ref{fig-tune_lr-mat-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-adult}Proportion of mature
+counterfactuals in each epoch. Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-credit}Proportion of mature
+counterfactuals in each epoch. Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-gmsc}Proportion of mature
+counterfactuals in each epoch. Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-mnist}Proportion of mature
+counterfactuals in each epoch. Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-over}Proportion of mature
+counterfactuals in each epoch. Data: Overlapping.}
 

### 2025-03-12 paper/experiments/CTExperiments/src/evaluate_counterfactuals.jl

@@ -30,7 +30,7 @@ Base.@kwdef struct CounterfactualParams <: AbstractConfiguration
     n_runs::Int = get_global_param("n_runs", 5)
     conv::AbstractString = "threshold"
     decision_threshold::AbstractFloat = 0.95
-    maxiter::Int = get_global_params("maxiter_eval", 50)
+    maxiter::Int = get_global_param("maxiter_eval", 50)
     vertical_splits::Int = get_global_param("vertical_splits", 100)
     store_ce::Bool = false
     parallelizer::AbstractString = "mpi"

### 2025-03-12 paper/sections/other/tuning.qmd

@@
 We hypothesize that this finding may be attributed to the uniform scaling of all input features (digits).
-Finally, to increase the proportion of mature counterfactuals for some datasets, we have also investigated the effect on the learning rate $\eta$ for the counterfactual search, but found little effect on the results.
+
+Finally, to increase the proportion of mature counterfactuals for some datasets, we have also investigated the effect on the learning rate $\eta$ for the counterfactual search and even smaller regularization strengths for a fixed decision threshold of 0.5 (@fig-tune_lr-plaus-adult to @fig-tune_lr-plaus-over).
+For the given low decision threshold, we find that the learning rate has no discernable impact on the proportion of mature counterfactuals (@fig-tune_lr-mat-adult to @fig-tune_lr-mat-over).
+We do notice, however, that the results for MNIST are much improved when using a low value $\lambda_{\text{reg}}$, the strength for the engery regularization: plausibility is increased by up to ~10ppt (@fig-tune_lr-plaus-mnist) and the proportion of mature counterfactuals reaches 100%.
+
+One consideration worth exploring is to combine high decision thresholds with high learning rates, which we have not investigated here.
 
@@
 ```
+
+### Learning Rate {#sec-app-tune-lr}
+
+```{julia}
+grid_dir = joinpath(res_dir, "tune_lr/mlp")
+data_dirs = readdir(grid_dir) |> x -> joinpath.(grid_dir, x) |> x -> x[isdir.(x)]
+eval_grids = (p -> EvaluationGrid(joinpath(p,"grid_config.toml"))).(data_dirs)
+data_names = basename.(data_dirs)
+ce_suffix = "evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/"
+logs_suffix = "evaluation/results/logs/objective---lr---lambda_energy_reg/"
+```
+
+The hyperparameter grid for tuning the learning rate is shown in @nte-tune_lr-train. The corresponding evaluation grid used for these experiments is shown in @nte-tune_lr-eval.
+
+::: {#nte-tune_lr-train .callout-note}
+
+## Training Phase
+
+```{julia}
+#| output: asis
+dict = CTExperiments.from_toml(joinpath(grid_dir, "over/grid_config.toml")) 
+println(CTExperiments.dict_to_quarto_markdown(dict))
+```
+
+:::
+
+::: {#nte-tune_lr-eval .callout-note}
+
+## Evaluation Phase
+
+```{julia}
+#| output: asis
+dict = CTExperiments.from_toml(joinpath(grid_dir, "over/evaluation/evaluation_grid_config.toml"))
+println(CTExperiments.dict_to_quarto_markdown(dict))
+```
+
+:::
+
+#### Plausibility
+
+```{julia}
+#| output: asis
+
+fig_label_prefix = "tune_lr-plaus"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the plausibility measure are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "plausibility_distance_from_target.png"
+fig_caption = "Average outcomes for the plausibility measure across key hyperparameters."
+full_paths = joinpath.(data_dirs, joinpath(ce_suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```
+
+#### Proportion of Mature CE
+
+```{julia}
+#| output: asis
+
+fig_label_prefix = "tune_lr-mat"
+fig_labels = (nm -> "fig-$(fig_label_prefix)-$nm").(data_names)
+_str = "The results with respect to the proportion of mature counterfactuals in each epoch are shown in @$(fig_labels[1]) to @$(fig_labels[end])."
+println(_str)
+```
+
+```{julia}
+#| output: asis
+
+imgfname = "percent_valid.png"
+fig_caption = "Proportion of mature counterfactuals in each epoch."
+full_paths = joinpath.(data_dirs, joinpath(logs_suffix,imgfname))
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
+_str = join(include_img_commands, "\n\n")
+println(_str)
+```

### 2025-03-12 paper/preprint/paper.tex
just updated bib and rerendered

--- 2025-03-11 paper/preprint/paper.tex
@@
 \usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
 \theoremstyle{plain}
 \newtheorem{proposition}{Proposition}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
 \theoremstyle{plain}
@@
 
+\bibitem[\citeproctext]{ref-altmeyer2023explaining}
+Altmeyer, Patrick, Arie van Deursen, et al. 2023. {``Explaining
+Black-Box Models Through Counterfactuals.''} In \emph{Proceedings of the
+JuliaCon Conferences}, 1:130. 1.
+
 \bibitem[\citeproctext]{ref-altmeyer2023faithful}

### 2025-03-12 paper/preprint/paper.tex
more edits

@@
 \usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
 \theoremstyle{definition}
 \newtheorem{example}{Example}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
 \theoremstyle{plain}
 \newtheorem{proposition}{Proposition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 \begin{equation}\phantomsection\label{eq-obj}{
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
 }\end{equation}
@@
 where \(\text{yloss}(\cdot)\) is any conventional classification loss
-that induces discriminative performance (e.g., cross-entropy). The two
-additional components in Equation~\ref{eq-obj} are explained in more
+that induces discriminative performance (e.g., cross-entropy). The
+second and third terms in Equation~\ref{eq-obj} are explained in more
 detail below. For now, they can be sufficiently described as inducing
@@
 (2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
-counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
-the different components can be governed by adjusting the strengths of
-the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).
+counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
+\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
+regularises the magnitude of the energy terms involved in
+\(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and Mordatch
+2020}). The tradeoff between the different components can be governed by
+adjusting the strengths of the penalties \(\lambda_{\text{div}}\),
+\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).
 
@@
 impact on the results. For example, \emph{REVISE} tends to perform the
-worst, most likely because it attempts to explain the decisions of a VAE
-that is an imperfect surrogate of the original model. Third, increasing
-\(T\), the maximum number of steps, generally yields better outcomes
-because more CEs can mature in any iteration. Fourth, the impact of
-\(\tau\), the required decision threshold is more difficult to predict.
-On ``harder'' datasets it may be difficult to satisfy high \(\tau\) for
-any given sample (i.e., also factuals) and so increasing this threshold
-does not seem to correlate with better outcomes.
-
-\textbf{\emph{Hyperparameters for penalties}} We find that the strength
+worst, most likely because it uses a surrogate VAE to generate
+counterfactuals which impedes faithfulness
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Third,
+increasing \(T\), the maximum number of steps, generally yields better
+outcomes because more CEs can mature in each training epoch. Fourth, the
+impact of \(\tau\), the required decision threshold is more difficult to
+predict. On ``harder'' datasets it may be difficult to satisfy high
+\(\tau\) for any given sample (i.e., also factuals) and so increasing
+this threshold does not seem to correlate with better outcomes. In fact,
+we have generally found that a choice of \(\tau=0.5\) leads to optimal
+results because it is associated with high proportions of mature
+counterfactuals.
+
+\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
 of the energy regularization, \(\lambda_{\text{reg}}\) is highly
@@
 \(\lambda_{\text{reg}}\) may increase the variability in outcomes when
-combined with high values for Equation~\ref{eq-obj} penalties, this
-effect is not very pronounced.
+combined with high values for any of the other penalties in
+Equation~\ref{eq-obj}, this effect is not very pronounced.
 
@@
 downstream impacts on our proposed method. Still, we find that training
-instabilities can be largely counteracted with strong regularization, in
-particular through \(\lambda_{\text{reg}}\) penalty on the energy.
+instabilities can be successfully mitigated by regularizing energy
+(\(\lambda_{\text{reg}}\)), generating a sufficiently large number of
+counterfactuals during each training epoch and including only mature
+counterfactuals for contrastive divergence.
 
 \textbf{\emph{CT is sensitive to hyperparameter selection.}} As
-discussed in Section~\ref{sec-hyperparameters}, our method requires
-careful selection of the hyperparameters. All experiments relied on
-grid-search tuning but we recognize that other algorithms may be more
-efficient. Notably, counterfactual training is iterative which makes a
-variety of methods applicable, including Bayesian (e.g.,
+discussed in Section~\ref{sec-hyperparameters}, our method benefits from
+tuning certain key hyperparameters. In this work, we have relied
+exclusively on grid search for this task. Future work on CT could
+benefit from investigating more sophisticated approaches towards
+hyperparameter tuning. Notably, counterfactual training is iterative
+which makes a variety of methods applicable, including Bayesian (e.g.,
 \citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
@@
 for conventionally-trained models. Third, in principle, CT yields itself
-to multi-threaded execution.
+to parallel execution, which we have leveraged for our own experiments.
 

### 2025-03-12 paper/sections/discussion.qmd
more edits

@@
 For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method.
-Still, we find that training instabilities can be largely counteracted with strong regularization, in particular through $\lambda_{\text{reg}}$ penalty on the energy.
+Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating a sufficiently large number of counterfactuals during each training epoch and including only mature counterfactuals for contrastive divergence.
 
-***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method requires careful selection of the hyperparameters.
-All experiments relied on grid-search tuning but we recognize that other algorithms may be more efficient.
+***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method benefits from tuning certain key hyperparameters.
+In this work, we have relied exclusively on grid search for this task.
+Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning.
 Notably, counterfactual training is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.
@@
 Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models.
-Third, in principle, CT yields itself to multi-threaded execution.
+Third, in principle, CT yields itself to parallel execution, which we have leveraged for our own experiments.
 

### 2025-03-12 paper/preprint/paper.tex
only minor changes, happy to merge

@@
 \newtheorem{example}{Example}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{plain}
@@
 relying on features and rules that are carefully hand-crafted by humans,
-modern machine learning (ML) models are tasked with learning the
+modern machine learning (ML) models are tasked with learning
 representations directly from data, guided by narrow objectives such as
@@
 Bengio, and Courville 2016}). Modern advances in computing have made it
-possible to provide such models with ever growing degrees of freedom to
+possible to provide such models with ever-growing degrees of freedom to
 achieve that task, which has often led them to outperform traditionally
-more parsimonious models. Unfortunately, in doing so, the models learn
+more parsimonious models. Unfortunately, in doing so, models learn
 increasingly complex and highly sensitive representations that we can no
@@
   to the number of parameters estimated from data.} Indeed, recent work
-on the so called ``lottery tickets'' suggests that modern neural
-networks can be pruned by up to 90\% while preserving their predictive
-performance (\citeproc{ref-frankle2018lottery}{Frankle and Carbin 2019})
-and generalizability (\citeproc{ref-morcos2019success}{Morcos et al.
-2019}). Similarly, Zhang et al.
-(\citeproc{ref-zhang2021understanding}{2021}) showed that
+on the so called ``lottery ticket hypothesis'' suggests that modern
+neural networks can be pruned by up to 90\% while preserving their
+predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
+Carbin 2019}) and generalizability
+(\citeproc{ref-morcos2019success}{Morcos et al. 2019}). Similarly, Zhang
+et al. (\citeproc{ref-zhang2021understanding}{2021}) showed that
 state-of-the-art neural networks are so expressive that they can fit

### 2025-03-12 paper/sections/method.qmd

@@
 $$
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
 $$ {#eq-obj}
@@
 where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g., cross-entropy).
-The two additional components in @eq-obj are explained in more detail below.
+The second and third terms in @eq-obj are explained in more detail below.
 For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
-The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.
+Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularises the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit].
+The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.
 

### 2025-03-12 paper/sections/related.qmd

@@
 
-To the best of our knowledge, our proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability.
+To the best of our knowledge, our proposed framework of counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability.
 In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and stakeholder requirements.
-To make this more concrete, we follow @augustin2020adversarial in tieing the concept of explainability to the quality of counterfactual explanations that we can generate for a given model.
+To make this more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of counterfactual explanations that we can generate for a given model.
 The authors show that counterfactual explanations---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples.
-We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom: by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e.
-adversarial examples), AT effectively removes some implausible explanations from the solution space.
+We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom: by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e., adversarial examples), AT effectively removes some implausible explanations from the solution space.
 
@@
 @teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function.
+Specifically, their approach relies on identifying similar input samples with different annotations and ensuring that the gradient of the classifier aligns with the vector between such pairs of counterfactual inputs using the cosine distance as the loss function.
 
@@
 
-There have also been several attempts at formalizing the relationship between counterfactual explanations (CE) and adversarial examples (AE).
+There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples (AE).
 Pointing to clear similarities in how CE and AE are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problem in representation learning.
-Formally, AE can be seen as the subset of CE, for which misclassification is achieved [@freiesleben2022intriguing].
+Formally, AE can be seen as the subset of CE for which misclassification is achieved [@freiesleben2022intriguing].
 Similarly, @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive theoretical upper bounds on the distances between them.
@@
 Firstly, @ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals to move from an adverse outcome to some positive target class with high probability.
-The approach proposed by @ross2021learning builds on adversarial training, where in this context susceptibility to targeted adversarial examples for the positive class is explicitly induced.
-The proposed method allows for imposing a set of actionability constraints ex-ante: for example, users can specify that certain features (e.g.
-*age*, *gender*, ...) are immutable.
+Their approach builds on adversarial training, where in this context susceptibility to targeted adversarial examples for the positive class is explicitly induced.
+The proposed method allows for imposing a set of actionability constraints ex-ante: for example, users can specify that certain features (e.g., *age*, *gender*, ...) are immutable.
 Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes counterfactual explanations as part of the training procedure.

### 2025-03-12 paper/sections/discussion.qmd

@@
 While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations.
-We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see details in @sec-hardware).
+We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs^[See supplementary appendix for computational details.].
 However, there are three factors that attenuate the impact of this limitation.

### 2025-03-12 paper/preprint/paper.tex

@@
 \usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
 \theoremstyle{remark}
@@
   background information on gradient-based counterfactual search and
-  convergence see \textbf{?@sec-app-ce}.}
+  convergence see supplementary appendix.}
 
@@
 demanding datasets on a high-performance computing cluster with 34 2GB
-CPUs (see details in \textbf{?@sec-hardware}). However, there are three
-factors that attenuate the impact of this limitation. First, CT provides
-counterfactual explanations for the training samples essentially for
-free, which may be beneficial in many ADM systems. Second, we find that
-CT can retain its value when used as a ``fine-tuning'' training regime
-for conventionally-trained models. Third, in principle, CT yields itself
-to parallel execution, which we have leveraged for our own experiments.
+CPUs\footnote{See supplementary appendix for computational details.}.
+However, there are three factors that attenuate the impact of this
+limitation. First, CT provides counterfactual explanations for the
+training samples essentially for free, which may be beneficial in many
+ADM systems. Second, we find that CT can retain its value when used as a
+``fine-tuning'' training regime for conventionally-trained models.
+Third, in principle, CT yields itself to parallel execution, which we
+have leveraged for our own experiments.
 
@@
 \url{https://doi.org/10.1145/3580305.3599290}.
+
+\bibitem[\citeproctext]{ref-hastie2009elements}
+Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
+Elements of Statistical Learning}. Springer New York.
+\url{https://doi.org/10.1007/978-0-387-84858-7}.
+
+\bibitem[\citeproctext]{ref-innes2018fashionable}
+Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco
+Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and
+Viral Shah. 2018. {``Fashionable Modelling with Flux.''}
+\url{https://arxiv.org/abs/1811.01457}.
+
+\bibitem[\citeproctext]{ref-innes2018flux}
+Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
+\emph{Journal of Open Source Software} 3 (25): 602.
+\url{https://doi.org/10.21105/joss.00602}.
 
@@
 
+\FloatBarrier
+
+\renewcommand{\thesection}{\Alph{section}}
+
+\renewcommand{\thetable}{A\arabic{table}}
+
+\renewcommand{\thefigure}{A\arabic{figure}}
+
+\section{Notation}\label{notation}
+
+\begin{itemize}
+\tightlist
+\item
+  \(y^+\): The target class and also the index of the target class.
+\item
+  \(y^-\): The non-target class and also the index of non-the target
+  class.
+\item
+  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
+  class.
+\item
+  \(\theta\): Model parameters (unspecified).
+\item
+  \(\Theta\): Matrix of parameters.
+\end{itemize}
+
+\subsection{Other Technical Details}\label{other-technical-details}
+
+\begin{equation}\phantomsection\label{eq-mmd}{
+\begin{aligned}
+MMD({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
+\end{aligned}
+}\end{equation}
+
+\section{Technical Details of Our
+Approach}\label{technical-details-of-our-approach}
+
+\subsection{Generating Counterfactuals through Gradient
+Descent}\label{sec-app-ce}
+
+In this section, we provide some background on gradient-based
+counterfactual generators (Section~\ref{sec-app-ce-background}) and
+discuss how we define convergence in this context
+(Section~\ref{sec-app-conv}).
+
+\subsubsection{Background}\label{sec-app-ce-background}
+
+Gradient-based counterfactual search was originally proposed by Wachter,
+Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}). It generally solves
+the following unconstrained objective,
+
+\[
+\begin{aligned}
+\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
+\end{aligned} 
+\]
+
+where \(g: \mathcal{Z} \mapsto \mathcal{X}\) is an invertible function
+that maps from the \(L\)-dimensional counterfactual state space to the
+feature space and \(\text{cost}(\cdot)\) denotes one or more penalties
+that are used to induce certain properties of the counterfactual
+outcome. As above, \(\mathbf{y}^+\) denotes the target output and
+\(\mathbf{M}_{\theta}(\mathbf{x})\) returns the logit predictions of the
+underlying classifier for \(\mathbf{x}=g(\mathbf{z})\).
+
+For all generators used in this work we use standard logit crossentropy
+loss for \(\text{yloss}(\cdot)\). All generators also penalize the
+distance (\(\ell_1\)-norm) of counterfactuals from their original
+factual state. For \emph{Generic} and \emph{ECCo}, we have
+\(\mathcal{Z}:=\mathcal{X}\) and
+\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
+are searched directly in the feature space. Conversely, \emph{REVISE}
+traverses the latent space of a variational autoencoder (VAE) fitted to
+the training data, where \(g(\cdot)\) corresponds to the decoder
+(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
+the distance penalty, \emph{ECCo} uses an additional penalty component
+that regularizes the energy associated with the counterfactual,
+\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
+al. 2024}).
+
+\subsubsection{Convergence}\label{sec-app-conv}
+
+An important consideration when generating counterfactual explanations
+using gradient-based methods is how to define convergence. Two common
+choices are to 1) perform gradient descent over a fixed number of
+iterations \(T\), or 2) conclude the search as soon as the predicted
+probability for the target class has reached a pre-determined threshold,
+\(\tau\):
+\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\). We
+prefer the latter for our purposes, because it explicitly defines
+convergence in terms of the black-box model, \(\mathbf{M}(\mathbf{x})\).
+
+Defining convergence in this way allows for a more intuitive
+interpretation of the resulting counterfactual outcomes than with fixed
+\(T\). Specifically, it allows us to think of counterfactuals as
+explaining `high-confidence' predictions by the model for the target
+class \(y^+\). Depending on the context and application, different
+choices of \(\tau\) can be considered as representing `high-confidence'
+predictions.
+
+\subsection{Protecting Mutability Constraints with Linear
+Classifiers}\label{sec-app-constraints}
+
+In Section~\ref{sec-constraints} we explain that to avoid penalizing
+implausibility that arises due to mutability constraints, we impose a
+point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
+argue in Section~\ref{sec-constraints} that this approach induces models
+to be less sensitive to immutable features and demonstrate this
+empirically in Section~\ref{sec-experiments}. Below we derive the
+analytical results in Proposition~\ref{prp-mtblty}.
+
+\begin{proof}
+Let \(d_{\text{mtbl}}\) and \(d_{\text{immtbl}}\) denote some mutable
+and immutable feature, respectively. Suppose that
+\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
+\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
+\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
+class \(k\). In words, we assume that the immutable feature tends to
+take lower values for samples in the non-target class \(y^-\) than in
+the target class \(y^+\). We assume the opposite to hold for the mutable
+feature.
+
+Assuming multivariate Gaussian class densities with common diagonal
+covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), we
+have for the log likelihood ratio between any two classes
+\(k,m \in \mathcal{K}\) (\citeproc{ref-hastie2009elements}{Hastie,
+Tibshirani, and Friedman 2009}):
+
+\begin{equation}\phantomsection\label{eq-loglike}{
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
+}\end{equation}
+
+By independence of \(x_1,...,x_D\), the full log-likelihood ratio
+decomposes into:
+
+\begin{equation}\phantomsection\label{eq-loglike-decomp}{
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
+}\end{equation}
+
+By the properties of our classifier (\emph{multinomial logistic
+regression}), we have:
+
+\begin{equation}\phantomsection\label{eq-multi}{
+\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
+}\end{equation}
+
+where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
+\(d\) for class \(k\).
+
+Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
+can identify that
+\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
+the assumptions we made above. Hence, we have that
+\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
+and
+\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)
+
+Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
+class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
+target class. Then the partial derivative of the contrastive divergence
+penalty Equation~\ref{eq-div} with respect to coefficient
+\(\theta_{y^+,d}\) is equal to
+
+\begin{equation}\phantomsection\label{eq-grad}{
+\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
+}\end{equation}
+
+and equal to zero everywhere else.
+
+Since \((\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})\) we
+are more likely to have
+\((x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0\) than vice
+versa at initialization. Similarly, we are more likely to have
+\((x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0\) since
+\((\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})\).
+
+This implies that if we do not protect feature \(d_{\text{immtbl}}\),
+the contrastive divergence penalty will decrease
+\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
+effect
+\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
+In words, not protecting the immutable feature would have the
+undesirable effect of making the classifier more sensitive to this
+feature, in that it would be more likely to predict class \(y^-\) as
+opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).
+
+By the same rationale, the contrastive divergence penalty can generally
+be expected to increase \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
+\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
+In words, this has the effect of making the classifier more sensitive to
+the mutable feature, in that it would be more likely to predict class
+\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).
+
+Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
+has the net affect of decreasing the classifier's sensitivity to the
+immutable feature relative to the mutable feature (i.e.~no change in
+sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
+for \(d_{\text{mtbl}}\)).
+\end{proof}
+
+\subsection{Domain Constraints}\label{domain-constraints}
+
+We apply domain constraints on counterfactuals during training and
+evaluation. There are at least two good reasons for doing so. Firstly,
+within the context of explainability and algorithmic recourse,
+real-world attributes are often domain constrained: the \emph{age}
+feature, for example, is lower bounded by zero and upper bounded by the
+maximum human lifespan. Secondly, domain constraints help mitigate
+training instabilities commonly associated with energy-based modelling
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020};
+\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+
+For our image datasets, features are pixel values and hence the domain
+is constrained by the lower and upper bound of values that pixels can
+take depending on how they are scaled (in our case \([-1,1]\)). For all
+other features \(d\) in our synthetic and tabular datasets, we
+automatically infer domain constraints
+\([x_d^{\text{LB}},x_d^{\text{UB}}]\) as follows,
+
+\begin{equation}\phantomsection\label{eq-domain}{
+\begin{aligned}
+x_d^{\text{LB}} &= \arg\min_{x_d} \{\mu_d - n_{\sigma_d}\sigma_d, \arg \min_{x_d} x_d\} \\
+x_d^{\text{UB}} &= \arg\max_{x_d} \{\mu_d + n_{\sigma_d}\sigma_d, \arg \max_{x_d} x_d\} 
+\end{aligned}
+}\end{equation}
+
+where \(\mu_d\) and \(\sigma_d\) denote the sample mean and standard
+deviation of feature \(d\). We set \(n_{\sigma_d}=3\) across the board
+but higher values and hence wider bounds may be appropriate depending on
+the application.
+
+\subsection{Training Details}\label{sec-app-training}
+
+In this section, we describe the training procedure in detail. While the
+details laid out here are not crucial for understanding our proposed
+approach, they are of importance to anyone looking to implement
+counterfactual training.
+
+\section{Detailed Results}\label{detailed-results}
+
+\subsection{Qualitative Findings for Image
+Data}\label{qualitative-findings-for-image-data}
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+Figure~\ref{fig-mnist} shows much more plausible (faithful)
+counterfactuals for a model with CT than the model with conventional
+training (Figure~\ref{fig-mnist-vanilla}). In fact, this is not even
+using \emph{ECCo+} and still showing better results than the best
+results we achieved in our AAAI paper for JEM ensembles.
+
+\end{tcolorbox}
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp.png}}
+
+}
+
+\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
+counterfactual training. The underlying generator, \emph{ECCo}, aims to
+generate counterfactuals that are faithful to the model
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp_vanilla.png}}
+
+}
+
+\caption{\label{fig-mnist-vanilla}Counterfactual images for \emph{MLP}
+with conventional training. The underlying generator, \emph{ECCo}, aims
+to generate counterfactuals that are faithful to the model
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
+
+\end{figure}%
+
+\subsection{Grid Searches}\label{sec-app-grid}
+
+To assess the hyperparameter sensitivity of our proposed training regime
+we ran multiple large grid searches for all of our synthetic datasets.
+We have grouped these grid searches into multiple categories:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \textbf{Generator Parameters} (Section~\ref{sec-app-grid-gen}):
+  Investigates the effect of changing hyperparameters that affect the
+  counterfactual outcomes during the training phase.
+\item
+  \textbf{Penalty Strengths} (Section~\ref{sec-app-grid-pen}):
+  Investigates the effect of changing the penalty strengths in out
+  proposed objective (Equation~\ref{eq-obj}).
+\item
+  \textbf{Other Parameters} (Section~\ref{sec-app-grid-train}):
+  Investigates the effect of changing other training parameters,
+  including the total number of generated counterfactuals in each epoch.
+\end{enumerate}
+
+We begin by summarizing the high-level findings in
+Section~\ref{sec-app-grid-hl}. For each of the categories,
+Section~\ref{sec-app-grid-gen} to Section~\ref{sec-app-grid-train} then
+present all details including the exact parameter grids, average
+predictive performance outcomes and key evaluation metrics for the
+generated counterfactuals.
+
+\subsubsection{Evaluation Details}\label{evaluation-details}
+
+To measure predictive performance, we compute the accuracy and F1-score
+for all models on test data (Table~\ref{tbl-acc-gen},
+Table~\ref{tbl-acc-pen}, Table~\ref{tbl-acc-train}). With respect to
+explanatory performance, we report here our findings for the
+(im)plausibility and cost of counterfactuals at test time. Since the
+computation of our proposed divergence metric
+(Equation~\ref{eq-impl-div}) is memory-intensive, we rely on the
+distance-based metric for the grid searches. For the counterfactual
+evaluation, we draw factual samples from the training data for the grid
+searches to avoid data leakage with respect to our final results
+reported in the body of the paper. Specifically, we want to avoid
+choosing our default hyperparameters based on results on the test data.
+Since we are optimizing for explainability, not predictive performance,
+we still present test accuracy and F1-scores.
+
+\paragraph{Predictive Performance}\label{predictive-performance}
+
+We find that CT is associated with little to no decrease in average
+predictive performance for our synthetic datasets: test accuracy and
+F1-scores decrease by at most \textasciitilde1 percentage point, but
+generally much less (Table~\ref{tbl-acc-gen}, Table~\ref{tbl-acc-pen},
+Table~\ref{tbl-acc-train}). Variation across hyperparameters is
+negligible as indicated by small standard deviations for these metrics
+across the board.
+
+\paragraph{Counterfactual Outcomes}\label{sec-app-grid-hl}
+
+Overall, we find that Counterfactual Training (CT) achieves it key
+objectives consistently across all hyperparameter settings and also
+broadly across datasets: plausibility is improved by up to
+\textasciitilde60 percentage points (ppts) for the \emph{Circles} data
+(e.g. Figure~\ref{fig-grid-gen_params-plaus-circles}),
+\textasciitilde25-30ppts for the \emph{Moons} data (e.g.
+Figure~\ref{fig-grid-gen_params-plaus-moons}) and
+\textasciitilde10-20ppts for the \emph{Linearly Separable} data (e.g.
+Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
+average costs of faithful counterfactuals are reduced in many cases by
+around \textasciitilde20-25ppts for \emph{Circles} (e.g.
+Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
+\textasciitilde50ppts for \emph{Moons} (e.g.
+Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
+Separable} data, costs are generally increased although typically by
+less than 10ppts (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
+which reflects a common tradeoff between costs and plausibility
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
+
+We do observe strong sensitivity to certain hyperparameters, with clear
+an manageable patterns. Concerning generator parameters, we firstly find
+that using \emph{REVISE} to generate counterfactuals during training
+typically yields the worst outcomes out of all generators, often leading
+to a substantial decrease in plausibility. This finding can be
+attributed to the fact that \emph{REVISE} effectively assigns the task
+of learning plausible explanations from the model itself to a surrogate
+VAE. In other words, counterfactuals generated by \emph{REVISE} are less
+faithful to the model that \emph{ECCo} and \emph{Generic}, and hence we
+would expect them to be a less effective and, in fact, potentially
+detrimental role in our training regime. Secondly, we observe that
+allowing for a higher number of maximum steps \(T\) for the
+counterfactual search generally yields better outcomes. This is
+intuitive, because it allows more counterfactuals to reach maturity in
+any given iteration. Looking in particular at the results for
+\emph{Linearly Separable}, it seems that higher values for \(T\) in
+combination with higher decision thresholds (\(\tau\)) yields the best
+results when using \emph{ECCo}. But depending on the degree of class
+separability of the underlying data, a high decision-threshold can also
+affect results adversely, as evident from the results for the
+\emph{Overlapping} data (Figure~\ref{fig-grid-gen_params-plaus-over}):
+here we find that CT generally fails to achieve its objective because
+only a tiny proportion of counterfactuals ever reaches maturity.
+
+Regarding penalty strengths, we find that the strength of the energy
+regularization, \(\lambda_{\text{reg}}\) is a key hyperparameter, while
+sensitivity with respect to \(\lambda_{\text{div}}\) and
+\(\lambda_{\text{adv}}\) is much less evident. In particular, we observe
+that not regularizing energy enough or at all typically leads to poor
+performance in terms of decreased plausibility and increased costs, in
+particular for \emph{Circles} (Figure~\ref{fig-grid-pen-plaus-circles}),
+\emph{Linearly Separable} (Figure~\ref{fig-grid-pen-plaus-lin_sep}) and
+\emph{Overlapping} (Figure~\ref{fig-grid-pen-plaus-over}). High values
+of \(\lambda_{\text{reg}}\) can increase the variability in outcomes, in
+particular when combined with high values for \(\lambda_{\text{div}}\)
+and \(\lambda_{\text{adv}}\), but this effect is less pronounced.
+
+Finally, concerning other hyperparameters we observe that the
+effectiveness and stability of CT is positively associated with the
+number of counterfactuals generated during each training epoch, in
+particular for \emph{Circles}
+(Figure~\ref{fig-grid-train-plaus-circles}) and \emph{Moons}
+(Figure~\ref{fig-grid-train-plaus-moons}). We further find that a higher
+number of training epochs is beneficial as expected, where we tested
+training models for 50 and 100 epochs. Interestingly, we find that it is
+not necessary to employ CT during the entire training phase to achieve
+the desired improvements in explainability: specifically, we have tested
+training models conventionally during the first half of training before
+switching to CT after this initial burn-in period.
+
+\subsubsection{Generator Parameters}\label{sec-app-grid-gen}
+
+The hyperparameter grid with varying generator parameters during
+training is shown in Note~\ref{nte-gen-params-final-run-train}. The
+corresponding evaluation grid used for these experiments is shown in
+Note~\ref{nte-gen-params-final-run-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-gen-params-final-run-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Decision Threshold: \texttt{0.75,\ 0.9,\ 0.95}
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 5.0,\ 10.0,\ 20.0}
+  \item
+    Maximum Iterations: \texttt{5,\ 25,\ 50}
+  \end{itemize}
+\item
+  Generator: \texttt{ecco,\ generic,\ revise}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-gen-params-final-run-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Accuracy}\label{accuracy}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-acc-gen}Predictive performance measures by dataset
+and objective averaged across training-phase parameters
+(Note~\ref{nte-gen-params-final-run-train}) and evaluation-phase
+parameters (Note~\ref{nte-gen-params-final-run-eval}).}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Circles & Accuracy & Full & 0.997 & 0.00309 \\
+  Circles & Accuracy & Vanilla & 0.998 & 0.000557 \\
+  Circles & F1-score & Full & 0.997 & 0.00309 \\
+  Circles & F1-score & Vanilla & 0.998 & 0.000558 \\
+  Lin Sep & Accuracy & Full & 0.999 & 0.00201 \\
+  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
+  Lin Sep & F1-score & Full & 0.999 & 0.00201 \\
+  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Moons & Accuracy & Full & 0.999 & 0.000696 \\
+  Moons & Accuracy & Vanilla & 1 & 0.00111 \\
+  Moons & F1-score & Full & 0.999 & 0.000696 \\
+  Moons & F1-score & Vanilla & 1 & 0.00111 \\
+  Over & Accuracy & Full & 0.915 & 0.00477 \\
+  Over & Accuracy & Vanilla & 0.917 & 0.00123 \\
+  Over & F1-score & Full & 0.915 & 0.00478 \\
+  Over & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Plausibility}\label{plausibility-2}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-grid-gen_params-plaus-circles} to
+Figure~\ref{fig-grid-gen_params-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-circles}Average outcomes for
+the plausibility measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-lin_sep}Average outcomes for
+the plausibility measure across hyperparameters. Data: Linearly
+Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-moons}Average outcomes for the
+plausibility measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-plaus-over}Average outcomes for the
+plausibility measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Cost}\label{cost}
+
+The results with respect to the cost measure are shown in
+Figure~\ref{fig-grid-gen_params-cost-circles} to
+Figure~\ref{fig-grid-gen_params-cost-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-circles}Average outcomes for
+the cost measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-lin_sep}Average outcomes for
+the cost measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-moons}Average outcomes for the
+cost measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+
+}
+
+\caption{\label{fig-grid-gen_params-cost-over}Average outcomes for the
+cost measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\subsubsection{Penalty Strengths}\label{sec-app-grid-pen}
+
+The hyperparameter grid with varying penalty strengths during training
+is shown in Note~\ref{nte-pen-final-run-train}. The corresponding
+evaluation grid used for these experiments is shown in
+Note~\ref{nte-pen-final-run-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-pen-final-run-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator: \texttt{ecco,\ generic,\ revise}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{adv}}\): \texttt{0.1,\ 0.25,\ 1.0}
+  \item
+    \(\lambda_{\text{div}}\): \texttt{0.01,\ 0.1,\ 1.0}
+  \item
+    \(\lambda_{\text{reg}}\): \texttt{0.0,\ 0.01,\ 0.1,\ 0.25,\ 0.5}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-pen-final-run-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Accuracy}\label{accuracy-1}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-acc-pen}Predictive performance measures by dataset
+and objective averaged across training-phase parameters
+(Note~\ref{nte-pen-final-run-train}) and evaluation-phase parameters
+(Note~\ref{nte-pen-final-run-eval}).}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Circles & Accuracy & Full & 0.994 & 0.0144 \\
+  Circles & Accuracy & Vanilla & 0.998 & 0.000875 \\
+  Circles & F1-score & Full & 0.994 & 0.0145 \\
+  Circles & F1-score & Vanilla & 0.998 & 0.000875 \\
+  Lin Sep & Accuracy & Full & 0.998 & 0.00772 \\
+  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
+  Lin Sep & F1-score & Full & 0.998 & 0.00773 \\
+  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Moons & Accuracy & Full & 0.987 & 0.0351 \\
+  Moons & Accuracy & Vanilla & 0.998 & 0.0101 \\
+  Moons & F1-score & Full & 0.987 & 0.0352 \\
+  Moons & F1-score & Vanilla & 0.998 & 0.0102 \\
+  Over & Accuracy & Full & 0.911 & 0.0217 \\
+  Over & Accuracy & Vanilla & 0.916 & 0.00236 \\
+  Over & F1-score & Full & 0.911 & 0.0219 \\
+  Over & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Plausibility}\label{plausibility-3}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-grid-pen-plaus-circles} to
+Figure~\ref{fig-grid-pen-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-circles}Average outcomes for the
+plausibility measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-lin_sep}Average outcomes for the
+plausibility measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-moons}Average outcomes for the
+plausibility measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-pen-plaus-over}Average outcomes for the
+plausibility measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Cost}\label{cost-1}
+
+The results with respect to the cost measure are shown in
+Figure~\ref{fig-grid-pen-cost-circles} to
+Figure~\ref{fig-grid-pen-cost-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-circles}Average outcomes for the cost
+measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-lin_sep}Average outcomes for the cost
+measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-moons}Average outcomes for the cost
+measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+
+}
+
+\caption{\label{fig-grid-pen-cost-over}Average outcomes for the cost
+measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\subsubsection{Other Parameters}\label{sec-app-grid-train}
+
+The hyperparameter grid with other varying training parameters is shown
+in Note~\ref{nte-train-final-run-train}. The corresponding evaluation
+grid used for these experiments is shown in
+Note~\ref{nte-train-final-run-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-train-final-run-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator: \texttt{ecco,\ generic,\ revise}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Burnin: \texttt{0.0,\ 0.5}
+  \item
+    No.~Counterfactuals: \texttt{100,\ 1000}
+  \item
+    No.~Epochs: \texttt{50,\ 100}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-train-final-run-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Accuracy}\label{accuracy-2}
+
+\begin{longtable}{ccccc}
+
+\caption{\label{tbl-acc-train}Predictive performance measures by dataset
+and objective averaged across training-phase parameters
+(Note~\ref{nte-train-final-run-train}) and evaluation-phase parameters
+(Note~\ref{nte-train-final-run-eval}).}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{5}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Circles & Accuracy & Full & 0.995 & 0.00431 \\
+  Circles & Accuracy & Vanilla & 0.998 & 0.000566 \\
+  Circles & F1-score & Full & 0.995 & 0.00432 \\
+  Circles & F1-score & Vanilla & 0.998 & 0.000566 \\
+  Lin Sep & Accuracy & Full & 0.999 & 0.00231 \\
+  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
+  Lin Sep & F1-score & Full & 0.999 & 0.00231 \\
+  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Moons & Accuracy & Full & 0.996 & 0.0136 \\
+  Moons & Accuracy & Vanilla & 0.988 & 0.022 \\
+  Moons & F1-score & Full & 0.996 & 0.0136 \\
+  Moons & F1-score & Vanilla & 0.988 & 0.022 \\
+  Over & Accuracy & Full & 0.914 & 0.00563 \\
+  Over & Accuracy & Vanilla & 0.918 & 0.00116 \\
+  Over & F1-score & Full & 0.914 & 0.0057 \\
+  Over & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule
+
+\end{longtable}
+
+\paragraph{Plausibility}\label{plausibility-4}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-grid-train-plaus-circles} to
+Figure~\ref{fig-grid-train-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-circles}Average outcomes for the
+plausibility measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-lin_sep}Average outcomes for the
+plausibility measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-moons}Average outcomes for the
+plausibility measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-grid-train-plaus-over}Average outcomes for the
+plausibility measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Cost}\label{cost-2}
+
+The results with respect to the cost measure are shown in
+Figure~\ref{fig-grid-train-cost-circles} to
+Figure~\ref{fig-grid-train-cost-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-circles}Average outcomes for the
+cost measure across hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-lin_sep}Average outcomes for the
+cost measure across hyperparameters. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-moons}Average outcomes for the cost
+measure across hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+
+}
+
+\caption{\label{fig-grid-train-cost-over}Average outcomes for the cost
+measure across hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\subsection{Hyperparameter Tuning}\label{sec-app-tune}
+
+Based on the findings from our initial large grid searches
+(Section~\ref{sec-app-grid}), we tune selected hyperparameters for all
+datasets: namely, the decision threshold \(\tau\) and the strength of
+the energy regularization \(\lambda_{\text{reg}}\). The final
+hyperparameter choices for each dataset are presented in \textbf{ADD
+TABLE}. Detailed results for each data set are shown in
+Figure~\ref{fig-tune-plaus-adult} to Figure~\ref{fig-tune-mat-over}.
+From \textbf{ADD TABLE}, we notice that the same decision threshold of
+\(\tau=0.5\) is optimal for all but on dataset. We attribute this to the
+fact that a low decision threshold results in a higher share of mature
+counterfactuals and hence more opportunities for the model to learn from
+examples (Figure~\ref{fig-tune-mat-adult} to
+Figure~\ref{fig-tune-mat-over}). This has played a role in particular
+for our real-world tabular datasets and MNIST, which suffered from low
+levels of maturity for higher decision thresholds. In cases where
+maturity is not an issue, as for \emph{Moons}, higher decision
+thresholds lead to better outcomes, which may have to do with the fact
+that the resulting counterfactuals are more faithful to the model.
+Concerning the regularization strength, we find somewhat high variation
+across datasets. Most notably, we find that relatively low levels of
+regularization are optimal for MNIST. We hypothesize that this finding
+may be attributed to the uniform scaling of all input features (digits).
+
+Finally, to increase the proportion of mature counterfactuals for some
+datasets, we have also investigated the effect on the learning rate
+\(\eta\) for the counterfactual search and even smaller regularization
+strengths for a fixed decision threshold of 0.5
+(Figure~\ref{fig-tune_lr-plaus-adult} to
+Figure~\ref{fig-tune_lr-plaus-over}). For the given low decision
+threshold, we find that the learning rate has no discernable impact on
+the proportion of mature counterfactuals
+(Figure~\ref{fig-tune_lr-mat-adult} to
+Figure~\ref{fig-tune_lr-mat-over}). We do notice, however, that the
+results for MNIST are much improved when using a low value
+\(\lambda_{\text{reg}}\), the strength for the engery regularization:
+plausibility is increased by up to \textasciitilde10ppt
+(Figure~\ref{fig-tune_lr-plaus-mnist}) and the proportion of mature
+counterfactuals reaches 100\%.
+
+One consideration worth exploring is to combine high decision thresholds
+with high learning rates, which we have not investigated here.
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-warning-color-frame, title={Package Version (Reproducibility)}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-warning-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+Tuning was run using \texttt{v1.1.3} of \texttt{TaijaData}. The
+follow-up version \texttt{v1.1.4} introduced an option to split
+real-world tabular datasets into train and test set, ensuring that
+pre-processing steps like standardization is fit on the training set
+only. If you are rerunning the tuning experiments with a version of
+\texttt{TaijaData} that is higher than \texttt{v1.1.3}, than for the
+default parameters specified in the configuration files, you may end up
+with slightly different results, although we would not expect any
+changes in terms of qualitative findings. For exact reproducibility,
+please use \texttt{v1.1.3}.
+
+\end{tcolorbox}
+
+\subsubsection{Key Parameters}\label{sec-app-tune-key}
+
+The hyperparameter grid for tuning key parameters is shown in
+Note~\ref{nte-tune-train}. The corresponding evaluation grid used for
+these experiments is shown in Note~\ref{nte-tune-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-tune-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Decision Threshold: \texttt{0.5,\ 0.75,\ 0.9}
+  \end{itemize}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{reg}}\): \texttt{0.1,\ 0.25,\ 0.5}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-tune-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Plausibility}\label{plausibility-5}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-tune-plaus-adult} to Figure~\ref{fig-tune-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-adult}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-cali}Average outcomes for the
+plausibility measure across key hyperparameters. Data: California
+Housing.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-circles}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-credit}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-gmsc}Average outcomes for the
+plausibility measure across key hyperparameters. Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-lin_sep}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Linearly
+Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-mnist}Average outcomes for the
+plausibility measure across key hyperparameters. Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-moons}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune-plaus-over}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Proportion of Mature CE}\label{proportion-of-mature-ce}
+
+The results with respect to the proportion of mature counterfactuals in
+each epoch are shown in Figure~\ref{fig-tune-mat-adult} to
+Figure~\ref{fig-tune-mat-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-adult}Proportion of mature counterfactuals
+in each epoch. Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-cali}Proportion of mature counterfactuals
+in each epoch. Data: California Housing.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-circles}Proportion of mature
+counterfactuals in each epoch. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-credit}Proportion of mature counterfactuals
+in each epoch. Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-gmsc}Proportion of mature counterfactuals
+in each epoch. Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-lin_sep}Proportion of mature
+counterfactuals in each epoch. Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-mnist}Proportion of mature counterfactuals
+in each epoch. Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-moons}Proportion of mature counterfactuals
+in each epoch. Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune-mat-over}Proportion of mature counterfactuals
+in each epoch. Data: Overlapping.}
+
+\end{figure}%
+
+\subsubsection{Learning Rate}\label{sec-app-tune-lr}
+
+The hyperparameter grid for tuning the learning rate is shown in
+Note~\ref{nte-tune_lr-train}. The corresponding evaluation grid used for
+these experiments is shown in Note~\ref{nte-tune_lr-eval}.
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-tune_lr-train}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Learning Rate: \texttt{0.1,\ 0.5,\ 1.0}
+  \end{itemize}
+\item
+  Model: \texttt{mlp}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{reg}}\): \texttt{0.01,\ 0.1,\ 0.5}
+  \item
+    Objective: \texttt{full,\ vanilla}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+
+\quartocalloutnte{nte-tune_lr-eval}
+
+\begin{itemize}
+\tightlist
+\item
+  Generator Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\paragraph{Plausibility}\label{plausibility-6}
+
+The results with respect to the plausibility measure are shown in
+Figure~\ref{fig-tune_lr-plaus-adult} to
+Figure~\ref{fig-tune_lr-plaus-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-adult}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-credit}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-gmsc}Average outcomes for the
+plausibility measure across key hyperparameters. Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-mnist}Average outcomes for the
+plausibility measure across key hyperparameters. Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-over}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Overlapping.}
+
+\end{figure}%
+
+\paragraph{Proportion of Mature CE}\label{proportion-of-mature-ce-1}
+
+The results with respect to the proportion of mature counterfactuals in
+each epoch are shown in Figure~\ref{fig-tune_lr-mat-adult} to
+Figure~\ref{fig-tune_lr-mat-over}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-adult}Proportion of mature
+counterfactuals in each epoch. Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-credit}Proportion of mature
+counterfactuals in each epoch. Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-gmsc}Proportion of mature
+counterfactuals in each epoch. Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-mnist}Proportion of mature
+counterfactuals in each epoch. Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-over}Proportion of mature
+counterfactuals in each epoch. Data: Overlapping.}
+
+\end{figure}%
+
+\section{Computation Details}\label{computation-details}
+
+\subsection{Hardware}\label{sec-app-hardware}
+
+We performed our experiments on a high-performance cluster. Details
+about the cluster will be disclosed upon publication to avoid revealing
+information that might interfere with the double-blind review process.
+Since our experiments involve highly parallel tasks and rather small
+models by today's standard, we have relied on distributed computing
+across multiple central processing units (CPU). Graphical processing
+units (GPU) were not required.
+
+\subsubsection{Grid Searches}\label{grid-searches}
+
+Model training for the largest grid searches with 270 unique parameter
+combinations was parallelized across 34 CPUs with 2GB memory each. The
+time to completion varied by dataset for reasons discussed in
+Section~\ref{sec-discussion}: 0h49m (\emph{Moons}), 1h4m (\emph{Linearly
+Separable}), 1h49m (\emph{Circles}), 3h52m (\emph{Overlapping}). Model
+evaluations for large grid searches were parallelized across 20 CPUs
+with 3GB memory each. Evaluations for all data sets took less than one
+hour (\textless1h) to complete.
+
+\subsubsection{Tuning}\label{tuning}
+
+For tuning of selected hyperparameters, we distributed the task of
+generating counterfactuals during training across 40 CPUs with 2GB
+memory each for all tabular datasets. Except for the \emph{Adult}
+dataset, all training runs were completed in less that half an hour
+(\textless0h30m). The \emph{Adult} dataset took around 0h35m to
+complete. Evaluations across 20 CPUs with 3GB memory each generally took
+less than 0h30m to complete. For \emph{MNIST}, we relied on 100 CPUs
+with 2GB memory each. For the \emph{MLP}, training of all models could
+be completed in 1h30m, while the evaluation across 20 CPUs (6GB memory)
+took 4h12m. For the \emph{CNN}, training of all models took
+\textasciitilde8h, with conventionally trained models taking
+\textasciitilde0h15m each and model with CT taking
+\textasciitilde0h30m-0h45m each.
+
+\subsection{Software}\label{software}
+
+All computations were performed in the Julia Programming Language
+(\citeproc{ref-bezanson2017julia}{Bezanson et al. 2017}). We have
+developed a package for counterfactual training that leverages and
+extends the functionality provided by several existing packages, most
+notably
+\href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{CounterfactualExplanations.jl}
+(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023})
+and the \href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library for
+deep learning (\citeproc{ref-innes2018fashionable}{Michael Innes et al.
+2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). For
+data-wrangling and presentation-ready tables we relied on
+\href{https://dataframes.juliadata.org/v1.7/}{DataFrames.jl}
+(\citeproc{ref-milan2023dataframes}{Bouchet-Valat and Kamiński 2023})
+and
+\href{https://ronisbr.github.io/PrettyTables.jl/v2.4/}{PrettyTables.jl}
+(\citeproc{ref-chagas2024pretty}{Chagas et al. 2024}), respectively. For
+plots and visualizations we used both
+\href{https://docs.juliaplots.org/v1.40/}{Plots.jl}
+(\citeproc{ref-PlotsJL}{Christ et al. 2023}) and
+\href{https://docs.makie.org/v0.22/}{Makie.jl}
+(\citeproc{ref-danisch2021makie}{Danisch and Krumbiegel 2021}), in
+particular \href{https://aog.makie.org/v0.9.3/}{AlgebraOfGraphics.jl}.
+To distribute computational tasks across multiple processors, we have
+relied on \href{https://juliaparallel.org/MPI.jl/v0.20/}{MPI.jl}
+(\citeproc{ref-byrne2021mpi}{Byrne, Wilcox, and Churavy 2021}).
+
 \end{document}

### 2025-03-12 paper/quarto_ecml/paper.tex

--- 2025-03-06 paper/quarto_ecml/paper.tex
@@
 \usepackage{placeins}
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
-\definecolor{quarto-callout-color}{HTML}{909090}
-\definecolor{quarto-callout-note-color}{HTML}{0758E5}
-\definecolor{quarto-callout-important-color}{HTML}{CC1914}
-\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
-\definecolor{quarto-callout-tip-color}{HTML}{00A047}
-\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
-\definecolor{quarto-callout-color-frame}{HTML}{acacac}
-\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
-\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
-\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
-\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
-\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
-\makeatother
 \makeatletter
@@
 % \theoremstyle{plain}
+% \newtheorem{proposition}{Proposition}[section]
+% \theoremstyle{plain}
 % \newtheorem{corollary}{Research Question}[section]
-% \theoremstyle{plain}
-% \newtheorem{proposition}{Proposition}[section]
+% \theoremstyle{definition}
+% \newtheorem{definition}{Definition}[section]
 % \theoremstyle{definition}
 % \newtheorem{example}{Example}[section]
-% \theoremstyle{definition}
-% \newtheorem{definition}{Definition}[section]
 % \theoremstyle{remark}
@@
 \begin{abstract}
-We propose a novel model objective that leverages counterfactual
-explanations to increase the explanatory capacity of trained models.
-Counterfactual explanations have emerged as a popular tool to explain
-predictions made by opaque machine learning models: they inform how
-factual inputs would need to change in order for a model to produce some
-desired output. Much existing research has focused on generating
-counterfactuals that are not only valid but also deemed acceptable with
-respect to the stakeholder requirements and plausible with respect to
-the underlying data. Recent work has shown that under this premise, the
-task of learning plausible explanations---ones that respect the
-conditional distribution of the target class---is effectively reassigned
-from the model itself to the post-hoc counterfactual explainer. Building
-on that work, we propose a novel objective function that employs
-counterfactuals (ad-hoc) during the training phase to minimize the
-divergence between the learned representations and the explanations.
-Through extensive experiments, we demonstrate that our proposed method
-facilitates training models that deliver inherently plausible
-explanations while maintaining high predictive performance.
-
-\keywords{Counterfactual Explanations \and Explainable
+We propose a novel training regime called Counterfactual Training that
+leverages counterfactual explanations to increase the explanatory
+capacity of models. Counterfactual explanations have emerged as a
+popular post-hoc explanation method for opaque machine learning models:
+they inform how factual inputs would need to change in order for a model
+to produce some desired output. To be useful in real-word
+decision-making systems, counterfactuals should be plausible with
+respect to the underlying data and actionable with respect to
+stakeholder requirements. Much existing research has therefore focused
+on developing post-hoc methods to generate counterfactuals that meet
+these desiderata. In this work, we instead hold models directly
+accountable for this desired end goal: Counterfactual Training employs
+counterfactuals ad-hoc during the training phase to minimize the
+divergence between learned representations and plausible, actionable
+explanations. We demonstrate empirically and theoretically that our
+proposed method facilitates training models that deliver inherently
+desirable explanations while maintaining high predictive performance.
+
+\keywords{Counterfactual Training \and Counterfactual
+Explanations \and Algorithmic Recourse \and Explainable
 AI \and Representation Learning}
@@
 relying on features and rules that are carefully hand-crafted by humans,
-modern machine learning (ML) models are tasked with learning these
-representations from scratch, guided by narrow objectives such as
+modern machine learning (ML) models are tasked with learning
+representations directly from data, guided by narrow objectives such as
 predictive accuracy \cite{goodfellow2016deep}. Modern advances in
-computing have made it possible to provide such models with ever greater
+computing have made it possible to provide such models with ever-growing
 degrees of freedom to achieve that task, which has often led them to
 outperform traditionally more parsimonious models. Unfortunately, in
-doing so they also learn increasingly complex and highly sensitive
+doing so, models learn increasingly complex and highly sensitive
 representations that we can no longer easily interpret.
 
-This trend towards complexity for the sake of performance has come under
+The trend towards complexity for the sake of performance has come under
 serious scrutiny in recent years. At the very cusp of the deep learning
@@
 ``imperceptible'' in that they are semantically indifferent from their
-factual counterparts. Despite partially effective mitigation strategies
-such as \textbf{adversarial training} \cite{goodfellow2014explaining},
-truly robust deep learning (DL) remains unattainable even for models
-that are considered shallow by today's standards
-\cite{kolter2023keynote}.
-
-Part of the problem is that high degrees of freedom provide room for
+factual counterparts. Although some partially effective mitigation
+strategies have been proposed, for example \textbf{adversarial training}
+\cite{goodfellow2014explaining}, truly robust deep learning (DL) remains
+unattainable even for models that are considered shallow by today's
+standards \cite{kolter2023keynote}.
+
+Part of the problem is that the high degrees of freedom provide room for
 many solutions that are locally optimal with respect to narrow
-objectives \cite{wilson2020case}\footnote{For clarity: we follow
-  standard ML convention in using ``degrees of freedom'' to refer to the
-  number of parameters estimated from data.}. Based purely on predictive
-performance, these solutions may seem to provide compelling explanations
-for the data, when in fact they are based on purely associative,
-semantically meaningless patterns. This poses two related challenges:
-firstly, it makes these models inherently opaque, since humans cannot
-simply interpret what type of explanation the complex learned
-representations correspond to; secondly, even if we could resolve the
-first challenge, it is not obvious how to mitigate models from learning
-representations that correspond to meaningless and implausible
-explanations.
+objectives \cite{wilson2020case}.\footnote{We follow the standard ML
+  convention, where ``degrees of freedom'' refer to the number of
+  parameters estimated from data.} Indeed, recent work on the so called
+``lottery ticket hypothesis'' suggests that modern neural networks can
+be pruned by up to 90\% while preserving their predictive performance
+\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
+Similarly, \cite{zhang2021understanding} showed that state-of-the-art
+neural networks are so expressive that they can fit randomly labeled
+data. Thus, looking at the predictive performance, the solutions may
+seem to provide compelling explanations for the data, when in fact they
+are based on purely associative, semantically meaningless patterns. This
+poses two related challenges. Firstly, there is no dependable way to
+verify if such complex representations correspond to meaningful and
+plausible explanations. Secondly, even if we could resolve the first
+challenge, it remains undecided how to ensure that models can
+\emph{only} learn valuable explanations.
 
@@
 scenario in which we deploy opaque models and blindly rely on their
-predictions. On countless occasions, this scenario has already occurred
-in practice and caused real harm to people who were affected adversely
-and often unfairly by automated decision-making systems (ADMS) involving
-opaque models \cite{oneil2016weapons}. Effective XAI tools can aide us
-in monitoring models and providing recourse to individuals to turn
-adverse outcomes (e.g.~``loan application rejected'') into positive ones
-(``application accepted''). \cite{wachter2017counterfactual} propose
-\textbf{counterfactual explanations} as an effective approach to achieve
-this: they explain how factual inputs need to change in order for some
-fitted model to produce some desired output, typically involving minimal
+predictions. On countless occasions, this scenario has occurred in
+practice and caused real harm to people who were affected adversely and
+often unfairly by automated decision-making (ADM) systems involving
+opaque models \cite{oneil2016weapons,mcgregor2021preventing}. Effective
+XAI tools can aid us in monitoring models and providing recourse to
+individuals to turn adverse outcomes (e.g., ``loan application
+rejected'') into positive ones (e.g., ``application accepted'').
+\cite{wachter2017counterfactual} propose \textbf{counterfactual
+explanations} (CE) as an effective approach to achieve this goal: CEs
+explain how factual inputs need to change in order for some fitted model
+to produce some desired output, typically involving minimal
 perturbations.
 
-To our surprise, the second challenge has not yet attracted any
+To our surprise, the second challenge has not yet attracted any major
 consolidated research effort. Specifically, there has been no concerted
-effort towards improving model \textbf{explainability}, which we define
-here as the degree to which learned representations correspond to
-explanations that are interpretable and deemed \textbf{plausible} by
-humans (see Definition~\ref{def-explainability}). Instead, the choice
-has typically been to improve the capacity of XAI tools to identify the
-subset explanations that are both plausible and valid for any given
-model, independent of whether the learned representations are also
-compatible with implausible explanations \cite{altmeyer2024faithful}.
-Fortunately, recent findings indicate that explainability can arise as
-byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization and generative capacity
-\cite{schut2021generating}.
+effort towards improving improving models' explanatory capacity, which
+we will henceforth simply call ``explainability'', defined as the degree
+to which learned representations correspond to explanations that are
+interpretable and deemed \textbf{plausible} by humans (see
+Definition~\ref{def-explainability}). Instead, the choice has typically
+been to improve the ability of XAI tools to identify the subset
+explanations that are both plausible and valid for any given model,
+independent of whether the learned representations are also compatible
+with implausible explanations \cite{altmeyer2024faithful}. Fortunately,
+recent findings indicate that explainability can arise as byproduct of
+regularization techniques aimed at other objectives such as robustness,
+generalization, and generative capacity
+\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
 
 Building on these findings, we introduce \textbf{counterfactual
-training}: a novel regularization technique geared explicitly towards
-aligning model representations with plausible explanations. Our
-contributions are as follows:
+training}: a novel training regime explicitly geared towards aligning
+model representations with plausible explanations. Our contributions are
+as follows:
 
@@
 the underlying data and stakeholder requirements. To make this more
-concrete, we follow \cite{augustin2020adversarial} in tieing the concept
+concrete, we follow \cite{augustin2020adversarial} in tying the concept
 of explainability to the quality of counterfactual explanations that we
@@
 already exist in training datasets. Specifically, their approach relies
-on, firstly, identifying similar input samples with different
-annotations and, secondly, ensuring that the gradient of the classifier
-aligns with the vector between pairs of counterfactual inputs using the
-cosine distance as a loss function.
+on identifying similar input samples with different annotations and
+ensuring that the gradient of the classifier aligns with the vector
+between such pairs of counterfactual inputs using the cosine distance as
+the loss function.
 
@@
 opaqueness and robustness problem in representation learning. Formally,
-AE can be seen as the subset of CE, for which misclassification is
+AE can be seen as the subset of CE for which misclassification is
 achieved \cite{freiesleben2022intriguing}. Similarly,
@@
 guaranteed to provide recourse for individuals to move from an adverse
-outcome to some positive target class with high probability. The
-approach proposed by \cite{ross2021learning} builds on adversarial
-training, where in this context susceptibility to targeted adversarial
-examples for the positive class is explicitly induced. The proposed
-method allows for imposing a set of actionability constraints ex-ante:
-for example, users can specify that certain features (e.g.~\emph{age},
-\emph{gender}, \ldots) are immutable. Secondly, \cite{guo2023counternet}
-are the first to propose an end-to-end training pipeline that includes
-counterfactual explanations as part of the training procedure. In
-particular, they propose a specific network architecture that includes a
-predictor and CE generator network, where the parameters of the CE
-generator network are learnable. Counterfactuals are generated during
-each training iteration and fed back to the predictor network. In
-contrast to \cite{guo2023counternet}, we impose no restrictions on the
-neural network architecture at all.
+outcome to some positive target class with high probability. Their
+approach builds on adversarial training, where in this context
+susceptibility to targeted adversarial examples for the positive class
+is explicitly induced. The proposed method allows for imposing a set of
+actionability constraints ex-ante: for example, users can specify that
+certain features (e.g., \emph{age}, \emph{gender}, \ldots) are
+immutable. Secondly, \cite{guo2023counternet} are the first to propose
+an end-to-end training pipeline that includes counterfactual
+explanations as part of the training procedure. In particular, they
+propose a specific network architecture that includes a predictor and CE
+generator network, where the parameters of the CE generator network are
+learnable. Counterfactuals are generated during each training iteration
+and fed back to the predictor network. In contrast to
+\cite{guo2023counternet}, we impose no restrictions on the neural
+network architecture at all.
 
@@
 \cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
-Plausibility is a necessary but insufficient condition for using CE to
+Plausibility is a necessary but insufficient condition for using CEs to
 provide algorithmic recourse (AR) to individuals affected by opaque
@@
 reached a pre-determined threshold, \(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
-  detailed background information on gradient-based counterfactual
-  search and convergence see Section~\ref{sec-app-ce}.}.
+\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
+where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
+  background information on gradient-based counterfactual search and
+  convergence see supplementary appendix.}
 
@@
 Definition~\ref{def-explainability}, we propose to leverage
-counterfactuals in the following objective,
+counterfactuals in the following objective:
 
 \begin{equation}\phantomsection\label{eq-obj}{
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
 }\end{equation}
@@
 where \(\text{yloss}(\cdot)\) is any conventional classification loss
-that induces discriminative performance (e.g.~cross-entropy). The two
-additional components in Equation~\ref{eq-obj} are explained in more
+that induces discriminative performance (e.g., cross-entropy). The
+second and third terms in Equation~\ref{eq-obj} are explained in more
 detail below. For now, they can be sufficiently described as inducing
-explainability directly and indirectly by penalizing: 1) the contrastive
-divergence, \(\text{div}(\cdot)\), between counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, 2) the
-adversarial loss, \(\text{advloss}(.)\), with respect to nascent
-counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
-the different components can be governed by adjusting the strengths of
-the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).
+explainability directly and indirectly by penalizing: (1) the
+contrastive divergence, \(\text{div}(\cdot)\), between mature
+counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
+(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
+counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
+\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
+regularises the magnitude of the energy terms involved in
+\(\text{div}(\cdot)\) \cite{du2019implicit}. The tradeoff between the
+different components can be governed by adjusting the strengths of the
+penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
+\(\lambda_{\text{reg}}\).
 
@@
 re-interpreted as a joint energy-based model (JEM) that learns to
-discriminate output classes conditional on inputs and generate inputs.
-They show that JEMs can be trained to perform well at both tasks by
-directly maximizing the joint log-likelihood factorized as
+discriminate output classes conditional on the observed (training)
+samples from \(p(\mathbf{x})\) and the generated samples from
+\(p_\theta(\mathbf{x})\). They show that JEMs can be trained to perform
+well at both tasks by directly maximizing the joint log-likelihood
+factorized as
 \(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
 The first factor can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
+Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
 \cite{grathwohl2020your} minimize the contrastive divergence between
-samples drawn from \(p_\theta(\mathbf{x})\) and training observations,
-i.e.~samples from \(p(\mathbf{x})\).
+these observed samples from \(p(\mathbf{x})\) and generated samples from
+\(p_\theta(\mathbf{x})\).
 
@@
 particular, we set
-\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
-where \(y^+\) denotes the index of the target class. We generate samples
-\(\mathbf{x}_T^\prime\) by first randomly sampling the target class
-\(y^+ \sim p(y)\) and then generating a counterfactual explanation for
-that target over \(T\) iterations using a gradient-based counterfactual
-generator. This is similar to how conditional sampling is used to draw
-from \(p_\theta(\mathbf{x})\) in \cite{grathwohl2020your}.
+\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
+where \(y^+\) denotes the index of the randomly drawn target class,
+\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed data point
+sampled from target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\).
+Conditional on the target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes
+a mature counterfactual for a randomly sampled factual from a non-target
+class generated through a gradient-based counterfactual generator for at
+most \(T\) iterations. We define mature counterfactuals as those that
+have either exhausted \(T\) or reached convergence in terms of the
+pre-determined decision threshold earlier.
 
@@
 \cite{du2019implicit}. As the generated counterfactuals get more
-plausible (Definition~\ref{def-explainability}) over the cause of
+plausible (Definition~\ref{def-explainability}) over the course of
 training, these two opposing effects gradually balance each out
@@
 
+\subsubsection{Evaluation}\label{evaluation}
+
+Our key outcome of interest is how well models perform with respect to
+explainability (Definition~\ref{def-explainability}): to this end, we
+focus primarily on the plausibility and cost of faithfully generated
+counterfactuals at test time. To measure the cost of counterfactuals, we
+follow the standard convention of using distances (\(\ell_1\)-norm)
+between factuals and counterfactuals as a proxy. For plausibility, we
+assess how similar counterfactuals are to observed samples in the target
+domain. We rely on the distance-based metric used by
+\cite{altmeyer2024faithful},
+
+\begin{equation}\phantomsection\label{eq-impl-dist}{
+\text{implaus}_{\text{dist}}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+}\end{equation}
+
+and introduce a novel divergence metric,
+
+\begin{equation}\phantomsection\label{eq-impl-div}{
+\text{implaus}_{\text{div}}(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
+}\end{equation}
+
+where \(\mathbf{X}^\prime\) denotes a set of multiple counterfactuals
+and \(\text{MMD}(\cdot)\) is an unbiased estimate of the squared
+population maximum mean discrepancy \cite{gretton2012kernel}. The metric
+in Equation~\ref{eq-impl-div} is equal to zero iff
+\(\mathbf{X}^\prime=\mathbf{X}^+\).
+
+In addition to cost and plausibility, we also compute other standard
+metrics to evaluate counterfactuals at test time including validity and
+redundancy. Finally, we also assess the predictive performance of models
+using standard metrics.
+
+We run the experiments with three CE generators: \emph{Generic} of
+\cite{wachter2017counterfactual} as a simple baseline approach,
+\emph{REVISE} \cite{joshi2019realistic} that aims to generate plausible
+counterfactuals using a surrogate Variational Autoencoder (VAE), and
+\emph{ECCo}---the generator of \cite{altmeyer2023faithful} but without
+the conformal prediction component---as a method that directly targets
+both faithfulness and plausibility of the CEs.
+
 \subsection{Experimental Results}\label{experimental-results}
 
+\subsubsection{Plausibility}\label{plausibility-1}
+
+\subsubsection{Actionability}\label{actionability-1}
+
+\subsubsection{Impact of hyperparameter
+settings}\label{sec-hyperparameters}
+
+We extensively test the impact of three types of hyperparameters on the
+proposed training regime. Our complete results are available in the
+technical appendix; this section focuses on the main findings.
+
+\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
+that CT is highly sensitive to hyperparameter settings but (a) there are
+manageable patterns and (b) we can typically identify settings that
+improve either plausibility or cost, and commonly both of them at the
+same time. Second, we note that the choice of a CE generator has a major
+impact on the results. For example, \emph{REVISE} tends to perform the
+worst, most likely because it uses a surrogate VAE to generate
+counterfactuals which impedes faithfulness \cite{altmeyer2024faithful}.
+Third, increasing \(T\), the maximum number of steps, generally yields
+better outcomes because more CEs can mature in each training epoch.
+Fourth, the impact of \(\tau\), the required decision threshold is more
+difficult to predict. On ``harder'' datasets it may be difficult to
+satisfy high \(\tau\) for any given sample (i.e., also factuals) and so
+increasing this threshold does not seem to correlate with better
+outcomes. In fact, we have generally found that a choice of \(\tau=0.5\)
+leads to optimal results because it is associated with high proportions
+of mature counterfactuals.
+
+\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
+of the energy regularization, \(\lambda_{\text{reg}}\) is highly
+impactful; energy must be sufficiently regularized to avoid poor
+performance in terms of decreased plausibility and increased costs. The
+sensitivity with respect to \(\lambda_{\text{div}}\) and
+\(\lambda_{\text{adv}}\) is much less evident. While high values of
+\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
+combined with high values for any of the other penalties in
+Equation~\ref{eq-obj}, this effect is not very pronounced.
+
+\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
+and stability of CT is positively associated with the number of
+counterfactuals generated during each training epoch. We also confirm
+that a higher number of training epochs is beneficial. Interestingly, we
+find that it is not necessary to employ CT during the entire training
+phase to achieve the desired improvements in explainability. We have
+tested training models conventionally during the first half of training
+before switching to CT after this initial ``burn-in'' period and
+observed positive results. Put differently, CT may be a way to improve
+the explainability of trained models in a fine-tuning manner.
+
 \section{Discussion}\label{sec-discussion}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Limited to classification models.
-\item
-  Proxy attributes of immutable features.
-\item
-  Increased training time.
-\item
-  Training instabilities
-\item
-  Fairness and caveats (aware it's not a classical approach in this
-  context, but there is a clear link).
-\end{enumerate}
+We begin the discussion by addressing the direct extensions of the
+counterfactual training approach in Section~\ref{sec-approach}. Then, we
+look at its broader limitations and challenges in
+Section~\ref{sec-limitations}.
+
+\subsection{Future research}\label{sec-approach}
+
+\textbf{\emph{CT is defined only for classification settings.}} Our
+formulation relies on the distinction between non-target class(es)
+\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
+through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
+arbitrarily defined by the user, CT requires the output space
+\(\mathcal{Y}\) to be discrete. Thus, it applies to binary and
+multi-class classification but it is not well-defined for other ML tasks
+where the change in outcome with respect to a decision threshold
+\(\tau\) cannot be readily quantified. In fact, this is a common
+restriction in research on CEs and AR that predominantly focuses on
+classification models. Although other settings have attracted some
+interest \cite{spooner2021counterfactual,zhao2023counterfactual}, there
+is still no consensus on what constitutes a counterfactual in such
+settings.
+
+\textbf{\emph{CT is subject to training instabilities.}} Joint
+energy-based models are susceptible to instabilities during training
+\cite{grathwohl2020your} and even though we depart from the SGLD-based
+sampling, we still encounter major variability in the outcomes. CT is
+exposed to two potential sources of instabilities: (1) the energy-based
+contrastive divergence term in Equation~\ref{eq-div}, and (2) the
+underlying counterfactual explainers. For example,
+\cite{altmeyer2023faithful} recognize this to be a challenge for
+\emph{ECCCo} and so it may have downstream impacts on our proposed
+method. Still, we find that training instabilities can be successfully
+mitigated by regularizing energy (\(\lambda_{\text{reg}}\)), generating
+a sufficiently large number of counterfactuals during each training
+epoch and including only mature counterfactuals for contrastive
+divergence.
+
+\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
+discussed in Section~\ref{sec-hyperparameters}, our method benefits from
+tuning certain key hyperparameters. In this work, we have relied
+exclusively on grid search for this task. Future work on CT could
+benefit from investigating more sophisticated approaches towards
+hyperparameter tuning. Notably, counterfactual training is iterative
+which makes a variety of methods applicable, including Bayesian
+\cite{snoek2012practical} or gradient-based \cite{franceschi2017forward}
+optimization.
+
+\subsection{Limitations and challenges}\label{sec-limitations}
+
+\textbf{\emph{CT increases the training time of models.}} Counterfactual
+training promotes explainability through CEs and robustness through AEs
+at the cost of longer training times compared to conventional training
+regimes. While higher numbers of iterations and counterfactuals per
+iteration positively impact the quality of found solutions, they also
+increase the required amount of computations. We find that relatively
+small grids with 270 settings can take almost four hours for more
+demanding datasets on a high-performance computing cluster with 34 2GB
+CPUs\footnote{See supplementary appendix for computational details.}.
+However, there are three factors that attenuate the impact of this
+limitation. First, CT provides counterfactual explanations for the
+training samples essentially for free, which may be beneficial in many
+ADM systems. Second, we find that CT can retain its value when used as a
+``fine-tuning'' training regime for conventionally-trained models.
+Third, in principle, CT yields itself to parallel execution, which we
+have leveraged for our own experiments.
+
+\textbf{\emph{Immutable features may have proxies.}} In
+Proposition~\ref{prp-mtblty} we define an approach to protect immutable
+features and thus increase the actionability of the generated
+counterfactuals. However, this approach requires that model owners
+define the mutability constraints for (all) features considered by the
+model. Even with sufficient domain knowledge to protect all immutable
+features---ones that cannot be changed at all and ones that cannot be
+reasonably expected to change---there may exist proxies that are
+theoretically mutable (and hence should not be protected) but preserve
+enough information about the principals to counteract the protections.
+As one example, consider the Adult dataset used in our experiments where
+the mutable education status is a proxy for the immutable age, in that
+the attainment of degrees is correlated with age. Delineating
+actionability is a major undecided challenge in the AR literature
+\cite{venkatasubramanian2020philosophical} impacting the capacity of CT
+to increase the explainability of the model.
+
+\textbf{\emph{Interventions on features may have downstream impacts on
+fairness.}}\\
+- \href{https://arxiv.org/abs/2401.16088}{Equality of opportunity}?\\
+- \href{https://proceedings.mlr.press/v202/gao23d/gao23d.pdf}{Social
+segregation}?\\
+- Supporting AR within one context/system may still unfairly target
+recourse-affected individuals within other contexts/systems? In Example
+3.1 younger individuals gain access to a loan but their age could still
+lead to, e.g., enforcing stronger supervisory mechanisms?
 
@@
 
-\newpage{}
-
-\FloatBarrier
-
-\setcounter{section}{0}
-\renewcommand{\thesection}{\Alph{section}}
-
-\setcounter{table}{0}
-\renewcommand{\thetable}{A\arabic{table}}
-
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{A\arabic{figure}}
-
-\section{Notation}\label{notation}
-
-\begin{itemize}
-\tightlist
-\item
-  \(y^+\): The target class and also the index of the target class.
-\item
-  \(y^-\): The non-target class and also the index of non-the target
-  class.
-\item
-  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
-  class.
-\item
-  \(\theta\): Model parameters (unspecified).
-\item
-  \(\Theta\): Matrix of parameters.
-\end{itemize}
-
-\section{Technical Details of Our
-Approach}\label{technical-details-of-our-approach}
-
-\subsection{Generating Counterfactuals through Gradient
-Descent}\label{sec-app-ce}
-
-In this section, we provide some background on gradient-based
-counterfactual generators (Section~\ref{sec-app-ce-background}) and
-discuss how we define convergence in this context
-(Section~\ref{sec-app-conv}).
-
-\subsubsection{Background}\label{sec-app-ce-background}
-
-Gradient-based counterfactual search was originally proposed by
-\cite{wachter2017counterfactual}. It generally solves the following
-unconstrained objective,
-
-\[
-\begin{aligned}
-\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
-\end{aligned} 
-\]
-
-where \(g: \mathcal{Z} \mapsto \mathcal{X}\) is an invertible function
-that maps from the \(L\)-dimensional counterfactual state space to the
-feature space and \(\text{cost}(\cdot)\) denotes one or more penalties
-that are used to induce certain properties of the counterfactual
-outcome. As above, \(\mathbf{y}^+\) denotes the target output and
-\(\mathbf{M}_{\theta}(\mathbf{x})\) returns the logit predictions of the
-underlying classifier for \(\mathbf{x}=g(\mathbf{z})\).
-
-For all generators used in this work we use standard logit crossentropy
-loss for \(\text{yloss}(\cdot)\). All generators also penalize the
-distance (\(\ell_1\)-norm) of counterfactuals from their original
-factual state. For \emph{Generic} and \emph{ECCo}, we have
-\(\mathcal{Z}:=\mathcal{X}\) and
-\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
-are searched directly in the feature space. Conversely, \emph{REVISE}
-traverses the latent space of a variational autoencoder (VAE) fitted to
-the training data, where \(g(\cdot)\) corresponds to the decoder
-\cite{joshi2019realistic}. In addition to the distance penalty,
-\emph{ECCo} uses an additional penalty component that regularizes the
-energy associated with the counterfactual, \(\mathbf{x}^\prime\)
-\cite{altmeyer2024faithful}.
-
-\subsubsection{Convergence}\label{sec-app-conv}
-
-An important consideration when generating counterfactual explanations
-using gradient-based methods is how to define convergence. Two common
-choices are to 1) perform gradient descent over a fixed number of
-iterations \(T\), or 2) conclude the search as soon as the predicted
-probability for the target class has reached a pre-determined threshold,
-\(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\). We
-prefer the latter for our purposes, because it explicitly defines
-convergence in terms of the black-box model, \(\mathbf{M}(\mathbf{x})\).
-
-Defining convergence in this way allows for a more intuitive
-interpretation of the resulting counterfactual outcomes than with fixed
-\(T\). Specifically, it allows us to think of counterfactuals as
-explaining `high-confidence' predictions by the model for the target
-class \(y^+\). Depending on the context and application, different
-choices of \(\tau\) can be considered as representing `high-confidence'
-predictions.
-
-\subsection{Protecting Mutability Constraints with Linear
-Classifiers}\label{sec-app-constraints}
-
-In Section~\ref{sec-constraints} we explain that to avoid penalizing
-implausibility that arises due to mutability constraints, we impose a
-point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
-argue in Section~\ref{sec-constraints} that this approach induces models
-to be less sensitive to immutable features and demonstrate this
-empirically in Section~\ref{sec-experiments}. Below we derive the
-analytical results in Proposition~\ref{prp-mtblty}.
-
-\begin{proof}
-Let \(d_{\text{mtbl}}\) and \(d_{\text{immtbl}}\) denote some mutable
-and immutable feature, respectively. Suppose that
-\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
-\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
-\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
-class \(k\). In words, we assume that the immutable feature tends to
-take lower values for samples in the non-target class \(y^-\) than in
-the target class \(y^+\). We assume the opposite to hold for the mutable
-feature.
-
-Assuming multivariate Gaussian class densities with common diagonal
-covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), we
-have for the log likelihood ratio between any two classes
-\(k,m \in \mathcal{K}\) \cite{hastie2009elements}:
-
-\begin{equation}\phantomsection\label{eq-loglike}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
-}\end{equation}
-
-By independence of \(x_1,...,x_D\), the full log-likelihood ratio
-decomposes into:
-
-\begin{equation}\phantomsection\label{eq-loglike-decomp}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
-}\end{equation}
-
-By the properties of our classifier (\emph{multinomial logistic
-regression}), we have:
-
-\begin{equation}\phantomsection\label{eq-multi}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
-}\end{equation}
-
-where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
-\(d\) for class \(k\).
-
-Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
-can identify that
-\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
-the assumptions we made above. Hence, we have that
-\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
-and
-\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)
-
-Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
-class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
-target class. Then the partial derivative of the contrastive divergence
-penalty Equation~\ref{eq-div} with respect to coefficient
-\(\theta_{y^+,d}\) is equal to
-
-\begin{equation}\phantomsection\label{eq-grad}{
-\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
-}\end{equation}
-
-and equal to zero everywhere else.
-
-Since \((\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})\) we
-are more likely to have
-\((x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0\) than vice
-versa at initialization. Similarly, we are more likely to have
-\((x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0\) since
-\((\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})\).
-
-This implies that if we do not protect feature \(d_{\text{immtbl}}\),
-the contrastive divergence penalty will decrease
-\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
-effect
-\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
-In words, not protecting the immutable feature would have the
-undesirable effect of making the classifier more sensitive to this
-feature, in that it would be more likely to predict class \(y^-\) as
-opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).
-
-By the same rationale, the contrastive divergence penalty can generally
-be expected to increase \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
-\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
-In words, this has the effect of making the classifier more sensitive to
-the mutable feature, in that it would be more likely to predict class
-\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).
-
-Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
-has the net affect of decreasing the classifier's sensitivity to the
-immutable feature relative to the mutable feature (i.e.~no change in
-sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
-for \(d_{\text{mtbl}}\)).
-\end{proof}
-
-\subsection{Domain Constraints}\label{domain-constraints}
-
-We apply domain constraints on counterfactuals during training and
-evaluation. There are at least two good reasons for doing so. Firstly,
-within the context of explainability and algorithmic recourse,
-real-world attributes are often domain constrained: the \emph{age}
-feature, for example, is lower bounded by zero and upper bounded by the
-maximum human lifespan. Secondly, domain constraints help mitigate
-training instabilities commonly associated with energy-based modelling
-\cite{grathwohl2020your,altmeyer2024faithful}.
-
-For our image datasets, features are pixel values and hence the domain
-is constrained by the lower and upper bound of values that pixels can
-take depending on how they are scaled (in our case \([-1,1]\)). For all
-other features \(d\) in our synthetic and tabular datasets, we
-automatically infer domain constraints
-\([x_d^{\text{LB}},x_d^{\text{UB}}]\) as follows,
-
-\begin{equation}\phantomsection\label{eq-domain}{
-\begin{aligned}
-x_d^{\text{LB}} &= \arg\min_{x_d} \{\mu_d - n_{\sigma_d}\sigma_d, \arg \min_{x_d} x_d\} \\
-x_d^{\text{UB}} &= \arg\max_{x_d} \{\mu_d + n_{\sigma_d}\sigma_d, \arg \max_{x_d} x_d\} 
-\end{aligned}
-}\end{equation}
-
-where \(\mu_d\) and \(\sigma_d\) denote the sample mean and standard
-deviation of feature \(d\). We set \(n_{\sigma_d}=3\) across the board
-but higher values and hence wider bounds may be appropriate depending on
-the application.
-
-\subsection{Training Details}\label{sec-app-training}
-
-In this section, we describe the training procedure in detail. While the
-details laid out here are not crucial for understanding our proposed
-approach, they are of importance to anyone looking to implement
-counterfactual training.
-
-\section{Detailed Results}\label{detailed-results}
-
-\subsection{Qualitative Findings for Image
-Data}\label{qualitative-findings-for-image-data}
-
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, bottomrule=.15mm, coltitle=black, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomtitle=1mm, leftrule=.75mm, breakable, colframe=quarto-callout-note-color-frame, left=2mm, colback=white, toptitle=1mm, toprule=.15mm, rightrule=.15mm, opacitybacktitle=0.6, title={Note}]
-
-Figure~\ref{fig-mnist} shows much more plausible (faithful)
-counterfactuals for a model with CT than the model with conventional
-training (Figure~\ref{fig-mnist-vanilla}). In fact, this is not even
-using \emph{ECCo+} and still showing better results than the best
-results we achieved in our AAAI paper for JEM ensembles.
-
-\end{tcolorbox}
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp.png}}
-
-}
-
-\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
-counterfactual training. The underlying generator, \emph{ECCo}, aims to
-generate counterfactuals that are faithful to the model
-\cite{altmeyer2024faithful}.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp_vanilla.png}}
-
-}
-
-\caption{\label{fig-mnist-vanilla}Counterfactual images for \emph{MLP}
-with conventional training. The underlying generator, \emph{ECCo}, aims
-to generate counterfactuals that are faithful to the model
-\cite{altmeyer2024faithful}.}
-
-\end{figure}%
-
-\subsection{Grid Searches}\label{sec-app-grid}
-
-\subsubsection{Generator Parameters}\label{generator-parameters}
-
-The hyperparameter grid with varying generator parameters during
-training is shown in Note~\ref{nte-gen-params-final-run-train}. The
-corresponding evaluation grid used for these experiments is shown in
-Note~\ref{nte-gen-params-final-run-eval}.
-
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, bottomrule=.15mm, coltitle=black, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomtitle=1mm, leftrule=.75mm, breakable, colframe=quarto-callout-note-color-frame, left=2mm, colback=white, toptitle=1mm, toprule=.15mm, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]
-
-\quartocalloutnte{nte-gen-params-final-run-train}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Decision Threshold: \texttt{0.75,\ 0.9,\ 0.95}
-  \item
-    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 5.0,\ 10.0,\ 20.0}
-  \item
-    Maximum Iterations: \texttt{5,\ 25,\ 50}
-  \end{itemize}
-\item
-  Generator: \texttt{ecco,\ generic,\ revise}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, titlerule=0mm, bottomrule=.15mm, coltitle=black, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, bottomtitle=1mm, leftrule=.75mm, breakable, colframe=quarto-callout-note-color-frame, left=2mm, colback=white, toptitle=1mm, toprule=.15mm, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]
-
-\quartocalloutnte{nte-gen-params-final-run-eval}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\subsubsection{High-Level Findings}\label{high-level-findings}
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/percent_valid.png}}
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/logs/objective---lambda_energy---generator_type---maxiter/acc_val.png}}
-
-\subsubsection{Detailed Findings}\label{detailed-findings}
-
-\paragraph{Linearly Separable}\label{linearly-separable}
-
-\begin{longtable}{cccccc}
-
-\caption{\label{tbl-lin_sep-lambda_energy_exper}Results for Linearly
-Separable data by energy penalty.}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Decision Threshold} & \textbf{$\lambda_{\text{energy}}$} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Decision Threshold} & \textbf{$\lambda_{\text{energy}}$} & \textbf{Maximum Iterations} & \textbf{Generator} & \textbf{Value} & \textbf{Std} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{6}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  0.75 & 0.1 & 5 & \textit{ECCo} & -15.9 & 16.7 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{41.9}} & \color{Green}{\textbf{8.13}} \\
-  0.75 & 0.1 & 5 & \textit{REVISE} & -64.4 & 3.58 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{28.9}} & \color{Green}{\textbf{2.28}} \\
-  0.75 & 0.1 & 25 & \textit{Generic} & -11.3 & 11.5 \\
-  0.75 & 0.1 & 25 & \textit{REVISE} & -55.3 & 4.18 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{10.9}} & \color{Green}{\textbf{7.47}} \\
-  0.75 & 0.1 & 50 & \textit{Generic} & -32.3 & 16.6 \\
-  0.75 & 0.1 & 50 & \textit{REVISE} & -44.1 & 4.24 \\
-  0.9 & 0.1 & 5 & \textit{ECCo} & -13.7 & 13.7 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{22.3}} & \color{Green}{\textbf{4.87}} \\
-  0.9 & 0.1 & 5 & \textit{REVISE} & -27.8 & 6.26 \\
-  0.9 & 0.1 & 25 & \textit{ECCo} & 8.84 & 7.13 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{30.7}} & \color{Green}{\textbf{5.77}} \\
-  0.9 & 0.1 & 25 & \textit{REVISE} & -72.7 & 34.9 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{17.1}} & \color{Green}{\textbf{7.98}} \\
-  0.9 & 0.1 & 50 & \textit{Generic} & 3.82 & 8.86 \\
-  0.9 & 0.1 & 50 & \textit{REVISE} & -45.8 & 2.79 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{35.7}} & \color{Green}{\textbf{7.56}} \\
-  0.95 & 0.1 & 5 & \textit{Generic} & -13.3 & 5.73 \\
-  0.95 & 0.1 & 5 & \textit{REVISE} & -54.9 & 5.43 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{37}} & \color{Green}{\textbf{5.31}} \\
-  0.95 & 0.1 & 25 & \textit{Generic} & 29 & 8.6 \\
-  0.95 & 0.1 & 25 & \textit{REVISE} & -40.5 & 41 \\
-  0.95 & 0.1 & 50 & \textit{ECCo} & 13.2 & 4.89 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.1}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{35.5}} & \color{Green}{\textbf{4.27}} \\
-  0.95 & 0.1 & 50 & \textit{REVISE} & -60.4 & 45.8 \\
-  \color{Red}{\textbf{0.75}} & \color{Red}{\textbf{0.5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{ECCo}} & \color{Red}{\textbf{-0.94}} & \color{Red}{\textbf{17.6}} \\
-  0.75 & 0.5 & 5 & \textit{Generic} & -11.7 & 17.4 \\
-  0.75 & 0.5 & 5 & \textit{REVISE} & -37.7 & 3.92 \\
-  0.75 & 0.5 & 25 & \textit{ECCo} & 14.7 & 6.32 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{32.3}} & \color{Green}{\textbf{3.49}} \\
-  0.75 & 0.5 & 25 & \textit{REVISE} & -39.9 & 7.21 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{28.3}} & \color{Green}{\textbf{5.05}} \\
-  0.75 & 0.5 & 50 & \textit{Generic} & -9.72 & 9.19 \\
-  0.75 & 0.5 & 50 & \textit{REVISE} & -30 & 5.22 \\
-  0.9 & 0.5 & 5 & \textit{ECCo} & -7.38 & 13.6 \\
-  \color{Red}{\textbf{0.9}} & \color{Red}{\textbf{0.5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{Generic}} & \color{Red}{\textbf{-0.0952}} & \color{Red}{\textbf{17.1}} \\
-  0.9 & 0.5 & 5 & \textit{REVISE} & -50.6 & 4.77 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{39.1}} & \color{Green}{\textbf{8.3}} \\
-  0.9 & 0.5 & 25 & \textit{Generic} & 38.6 & 6.03 \\
-  0.9 & 0.5 & 25 & \textit{REVISE} & -56.9 & 4.28 \\
-  0.9 & 0.5 & 50 & \textit{ECCo} & -6.53 & 10.9 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{13.6}} & \color{Green}{\textbf{12.8}} \\
-  0.9 & 0.5 & 50 & \textit{REVISE} & -24.1 & 12.5 \\
-  0.95 & 0.5 & 5 & \textit{ECCo} & 0.596 & 1.11 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{4.57}} & \color{Green}{\textbf{5.88}} \\
-  0.95 & 0.5 & 5 & \textit{REVISE} & -34.3 & 6.09 \\
-  0.95 & 0.5 & 25 & \textit{ECCo} & 16.5 & 3.53 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{19.1}} & \color{Green}{\textbf{4.43}} \\
-  0.95 & 0.5 & 25 & \textit{REVISE} & -37.2 & 39.2 \\
-  0.95 & 0.5 & 50 & \textit{ECCo} & -31 & 14.2 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{0.5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{24.8}} & \color{Green}{\textbf{4.78}} \\
-  0.95 & 0.5 & 50 & \textit{REVISE} & -1.28 & 2.7 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{37}} & \color{Green}{\textbf{9.21}} \\
-  0.75 & 5 & 5 & \textit{Generic} & 0.977 & 21.1 \\
-  0.75 & 5 & 5 & \textit{REVISE} & -55.2 & 4.75 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{21.8}} & \color{Green}{\textbf{5.97}} \\
-  0.75 & 5 & 25 & \textit{Generic} & 17.8 & 5.33 \\
-  0.75 & 5 & 25 & \textit{REVISE} & -68.1 & 8.69 \\
-  0.75 & 5 & 50 & \textit{ECCo} & 0.864 & 9.51 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{8.56}} & \color{Green}{\textbf{13.6}} \\
-  0.75 & 5 & 50 & \textit{REVISE} & -71.8 & 3.18 \\
-  0.9 & 5 & 5 & \textit{ECCo} & -3.59 & 20.3 \\
-  \color{Red}{\textbf{0.9}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{Generic}} & \color{Red}{\textbf{-1.18}} & \color{Red}{\textbf{16.7}} \\
-  0.9 & 5 & 5 & \textit{REVISE} & -58.6 & 3.55 \\
-  0.9 & 5 & 25 & \textit{ECCo} & -1.73 & 9.07 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{21.9}} & \color{Green}{\textbf{7.48}} \\
-  0.9 & 5 & 25 & \textit{REVISE} & -44.4 & 4.23 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{30.8}} & \color{Green}{\textbf{10.7}} \\
-  0.9 & 5 & 50 & \textit{Generic} & 16.3 & 11.1 \\
-  0.9 & 5 & 50 & \textit{REVISE} & -53.3 & 6.73 \\
-  \color{Red}{\textbf{0.95}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{5}} & \color{Red}{\textbf{ECCo}} & \color{Red}{\textbf{-5.29}} & \color{Red}{\textbf{24.5}} \\
-  0.95 & 5 & 5 & \textit{Generic} & -5.71 & 17.4 \\
-  0.95 & 5 & 5 & \textit{REVISE} & -36.5 & 4.01 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{26.9}} & \color{Green}{\textbf{4.08}} \\
-  0.95 & 5 & 25 & \textit{Generic} & 19.5 & 5.49 \\
-  0.95 & 5 & 25 & \textit{REVISE} & -25 & 5 \\
-  0.95 & 5 & 50 & \textit{ECCo} & 4.6 & 7.46 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{16.6}} & \color{Green}{\textbf{4.3}} \\
-  0.95 & 5 & 50 & \textit{REVISE} & -50.4 & 46.4 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{2.47}} & \color{Green}{\textbf{8.57}} \\
-  0.75 & 10 & 5 & \textit{Generic} & -7.23 & 9.13 \\
-  0.75 & 10 & 5 & \textit{REVISE} & -67.2 & 5.75 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{29.9}} & \color{Green}{\textbf{15.8}} \\
-  0.75 & 10 & 25 & \textit{Generic} & 4.47 & 8.22 \\
-  0.75 & 10 & 25 & \textit{REVISE} & -52 & 5.51 \\
-  0.75 & 10 & 50 & \textit{ECCo} & 6.23 & 6.11 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{15.2}} & \color{Green}{\textbf{10.2}} \\
-  0.75 & 10 & 50 & \textit{REVISE} & -17 & 5.46 \\
-  0.9 & 10 & 5 & \textit{ECCo} & -6.93 & 14.7 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{6.88}} & \color{Green}{\textbf{5.11}} \\
-  0.9 & 10 & 5 & \textit{REVISE} & -45.6 & 47.8 \\
-  0.9 & 10 & 25 & \textit{ECCo} & 6.2 & 6.75 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{34.7}} & \color{Green}{\textbf{9.74}} \\
-  0.9 & 10 & 25 & \textit{REVISE} & -55.7 & 6.04 \\
-  0.9 & 10 & 50 & \textit{ECCo} & 30.1 & 14.2 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{35}} & \color{Green}{\textbf{7.11}} \\
-  0.9 & 10 & 50 & \textit{REVISE} & -48.8 & 3.3 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{35.7}} & \color{Green}{\textbf{3.83}} \\
-  0.95 & 10 & 5 & \textit{Generic} & 2.85 & 7.9 \\
-  0.95 & 10 & 5 & \textit{REVISE} & -34.9 & 4.56 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{33}} & \color{Green}{\textbf{6.45}} \\
-  0.95 & 10 & 25 & \textit{Generic} & 24.8 & 4.59 \\
-  0.95 & 10 & 25 & \textit{REVISE} & -55.7 & 44 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{10}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{8.07}} & \color{Green}{\textbf{7.68}} \\
-  0.95 & 10 & 50 & \textit{Generic} & -7.11 & 9.2 \\
-  0.95 & 10 & 50 & \textit{REVISE} & -26.1 & 18.6 \\
-  0.75 & 20 & 5 & \textit{ECCo} & -10.8 & 20.8 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{12.6}} & \color{Green}{\textbf{17.6}} \\
-  0.75 & 20 & 5 & \textit{REVISE} & -21.6 & 3.34 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{17.2}} & \color{Green}{\textbf{4.25}} \\
-  0.75 & 20 & 25 & \textit{Generic} & -10.3 & 9.59 \\
-  0.75 & 20 & 25 & \textit{REVISE} & -59 & 3.66 \\
-  \color{Green}{\textbf{0.75}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{15.6}} & \color{Green}{\textbf{8.38}} \\
-  0.75 & 20 & 50 & \textit{Generic} & 3.92 & 6.2 \\
-  0.75 & 20 & 50 & \textit{REVISE} & -70.9 & 5.16 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{26.3}} & \color{Green}{\textbf{6.62}} \\
-  0.9 & 20 & 5 & \textit{Generic} & 2.35 & 7.04 \\
-  0.9 & 20 & 5 & \textit{REVISE} & -64.8 & 4.31 \\
-  0.9 & 20 & 25 & \textit{ECCo} & 21.1 & 11.6 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{39}} & \color{Green}{\textbf{5.98}} \\
-  0.9 & 20 & 25 & \textit{REVISE} & -50.7 & 13.8 \\
-  \color{Green}{\textbf{0.9}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{15.1}} & \color{Green}{\textbf{7.46}} \\
-  0.9 & 20 & 50 & \textit{Generic} & 8.53 & 6.95 \\
-  0.9 & 20 & 50 & \textit{REVISE} & -36 & 6.14 \\
-  0.95 & 20 & 5 & \textit{ECCo} & -23.8 & 17.8 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{5}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{7.23}} & \color{Green}{\textbf{4.74}} \\
-  0.95 & 20 & 5 & \textit{REVISE} & -34 & 4.11 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{25}} & \color{Green}{\textbf{ECCo}} & \color{Green}{\textbf{19.3}} & \color{Green}{\textbf{14.6}} \\
-  0.95 & 20 & 25 & \textit{Generic} & -1.35 & 7.71 \\
-  0.95 & 20 & 25 & \textit{REVISE} & -52.8 & 5.11 \\
-  0.95 & 20 & 50 & \textit{ECCo} & 18.7 & 3.84 \\
-  \color{Green}{\textbf{0.95}} & \color{Green}{\textbf{20}} & \color{Green}{\textbf{50}} & \color{Green}{\textbf{Generic}} & \color{Green}{\textbf{37.3}} & \color{Green}{\textbf{7.78}} \\
-  0.95 & 20 & 50 & \textit{REVISE} & -28.3 & 15 \\\bottomrule
-
-\end{longtable}
-
-\section{Computation Details}\label{computation-details}
-
-\subsection{Hardware}\label{hardware}
-
-We performed our experiments on a high-performance cluster. Details
-about the cluster will be disclosed upon publication to avoid revealing
-information that might interfere with the double-blind review process.
-Since our experiments involve highly parallel tasks and rather small
-models by today's standard, we have relied on distributed computing
-across multiple central processing units (CPU). Graphical processing
-units (GPU) were not required. For larger grid searches we used up to
-but typically less than 100 CPUs.
-
-\subsection{Software}\label{software}
-
-All computations were performed in the Julia Programming Language
-\cite{bezanson2017julia}. We have developed a package for counterfactual
-training that leverages and extends the functionality provided by
-several existing packages, most notably
-\href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{CounterfactualExplanations.jl}
-\cite{altmeyer2023explaining} and the
-\href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library for deep
-learning \cite{innes2018fashionable,innes2018flux}. For data-wrangling
-and presentation-ready tables we relied on
-\href{https://dataframes.juliadata.org/v1.7/}{DataFrames.jl}
-\cite{milan2023dataframes} and
-\href{https://ronisbr.github.io/PrettyTables.jl/v2.4/}{PrettyTables.jl}
-\cite{chagas2024pretty}, respectively. For plots and visualizations we
-used both \href{https://docs.juliaplots.org/v1.40/}{Plots.jl}
-\cite{PlotsJL} and \href{https://docs.makie.org/v0.22/}{Makie.jl}
-\cite{danisch2021makie}, in particular
-\href{https://aog.makie.org/v0.9.3/}{AlgebraOfGraphics.jl}. To
-distribute computational tasks across multiple processors, we have
-relied on \href{https://juliaparallel.org/MPI.jl/v0.20/}{MPI.jl}
-\cite{byrne2021mpi}.
+\subsection*{References}\label{references}
+\addcontentsline{toc}{subsection}{References}
 

### 2025-03-12 paper/sections/experiments.qmd
Add first part of discussion (Section 5.1)

@@
 
-### Impact of hyperparameter settings
+### Impact of hyperparameter settings {#sec-hyperparameters}
 We extensively test the impact of three types of hyperparameters on the proposed training regime.
@@
 Second, we note that the choice of a CE generator has a major impact on the results.
-For example, *REVISE* tends to perform the worst, most likely because it attempts to explain the decisions of a VAE that is an imperfect surrogate of the original model.
-Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in any iteration.
+For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful].
+Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch.
 Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict.
 On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes.
+In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.
 
-***Hyperparameters for penalties*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs.
+***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs.
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
-While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values for @eq-obj penalties, this effect is not very pronounced.
+While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values for any of the other penalties in @eq-obj, this effect is not very pronounced.
 

### 2025-03-12 paper/sections/intro.qmd

@@
 
-Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning the representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
-Modern advances in computing have made it possible to provide such models with ever growing degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models.
-Unfortunately, in doing so, the models learn increasingly complex and highly sensitive representations that we can no longer easily interpret.
+Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
+Modern advances in computing have made it possible to provide such models with ever-growing degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models.
+Unfortunately, in doing so, models learn increasingly complex and highly sensitive representations that we can no longer easily interpret.
 
@@
 
-Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery tickets" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success].
+Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success].
 Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data.

### 2025-03-12 paper/sections/method.qmd

@@
 Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual.
-For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]
+For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see supplementary appendix.]
+
+<!-- @sec-app-ce -->
 

### 2025-03-12 paper/preprint/paper.tex
Add final subsection to discussion

--- 2025-03-12 paper/preprint/paper.tex
@@
 \usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{definition}
 \newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{plain}
@@
 \textbf{\emph{Interventions on features may have downstream impacts on
-fairness.}}\\
-- \href{https://arxiv.org/abs/2401.16088}{Equality of opportunity}?\\
-- \href{https://proceedings.mlr.press/v202/gao23d/gao23d.pdf}{Social
-segregation}?\\
-- Supporting AR within one context/system may still unfairly target
-recourse-affected individuals within other contexts/systems? In Example
-3.1 younger individuals gain access to a loan but their age could still
-lead to, e.g., enforcing stronger supervisory mechanisms?
+fairness.}} Related to the point above, we provide a tool that allows
+practitioners to modify the sensitivity of a model with respect to
+certain features, which may have implication for the fair and equitable
+treatment of individuals subject to automated decisions. As protecting a
+set of features leads the model to assign higher relative importance to
+unprotected features, model owners could misuse our solution by
+enforcing explanations based on features that are more difficult to
+modify by some (group of) individuals. For example, consider again the
+Adult dataset where features such as workclass or education may be more
+difficult to change for underpriviledged groups. When applied
+irresponsibly, counterfactual training could result in an unfairly
+assigned burden of recourse (e.g.,
+\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
+threatening the equality of opportunity in the system
+(\citeproc{ref-bell2024fairness}{Bell et al. 2024}) and potentially
+reinforcing social segregation (\citeproc{ref-gao2023impact}{Gao and
+Lakkaraju 2023}). Still, as the referenced publications indicate, such
+phenomena are not specific to CT; regrettably, all types of ADM
+solutions without strong external protections have been recognized to
+promote harmful power dynamics (\citeproc{ref-maas2023machine}{Maas
+2023}).
 
@@
 
+\bibitem[\citeproctext]{ref-bell2024fairness}
+Bell, Andrew, Joao Fonseca, Carlo Abrate, Francesco Bonchi, and Julia
+Stoyanovich. 2024. {``Fairness in Algorithmic Recourse Through the Lens
+of Substantive Equality of Opportunity.''}
+\url{https://arxiv.org/abs/2401.16088}.
+
 \bibitem[\citeproctext]{ref-du2019implicit}
@@
 
+\bibitem[\citeproctext]{ref-gao2023impact}
+Gao, Ruijiang, and Himabindu Lakkaraju. 2023. {``On the Impact of
+Algorithmic Recourse on Social Segregation.''} In \emph{Proceedings of
+the 40th International Conference on Machine Learning}. ICML'23.
+Honolulu, Hawaii, USA: JMLR.org.
+
 \bibitem[\citeproctext]{ref-goodfellow2014explaining}
@@
 Information and Computation}, 881--88.
+
+\bibitem[\citeproctext]{ref-maas2023machine}
+Maas, Jonne. 2023. {``Machine Learning and Power Relations.''} \emph{AI
+\& SOCIETY} 38 (4): 1493--1500.
 
@@
 
+\bibitem[\citeproctext]{ref-sharma2020certifai}
+Sharma, Shubham, Jette Henderson, and Joydeep Ghosh. 2020. {``CERTIFAI:
+A Common Framework to Provide Explanations and Analyse the Fairness and
+Robustness of Black-Box Models.''} In \emph{Proceedings of the AAAI/ACM
+Conference on AI, Ethics, and Society}, 166--72. AIES '20. New York, NY,
+USA: Association for Computing Machinery.
+\url{https://doi.org/10.1145/3375627.3375812}.
+
 \bibitem[\citeproctext]{ref-snoek2012practical}

### 2025-03-12 paper/sections/discussion.qmd
Add final subsection to discussion

--- 2025-03-12 paper/sections/discussion.qmd
@@
 
-***Interventions on features may have downstream impacts on fairness.***  
-- [Equality of opportunity](https://arxiv.org/abs/2401.16088)?  
-- [Social segregation](https://proceedings.mlr.press/v202/gao23d/gao23d.pdf)?  
-- Supporting AR within one context/system may still unfairly target recourse-affected individuals within other contexts/systems?
-In Example 3.1 younger individuals gain access to a loan but their age could still lead to, e.g., enforcing stronger supervisory mechanisms?
+***Interventions on features may have downstream impacts on fairness.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions.
+As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.
+For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups.
+When applied irresponsibly, counterfactual training could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact].
+Still, as the referenced publications indicate, such phenomena are not specific to CT; regrettably, all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].

### 2025-03-12 paper/quarto_ecml/paper.tex
Review abstract and introduction

@@
 \newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-% \usepackage{amsthm}
-% \theoremstyle{plain}
-% \newtheorem{proposition}{Proposition}[section]
-% \theoremstyle{plain}
-% \newtheorem{corollary}{Research Question}[section]
-% \theoremstyle{definition}
-% \newtheorem{definition}{Definition}[section]
-% \theoremstyle{definition}
-% \newtheorem{example}{Example}[section]
-% \theoremstyle{remark}
-% \AtBeginDocument{\renewcommand*{\proofname}{Proof}}
-% \newtheorem*{remark}{Remark}
-% \newtheorem*{solution}{Solution}
-% \newtheorem{refremark}{Remark}[section]
-% \newtheorem{refsolution}{Solution}[section]
 \makeatother
@@
 \begin{abstract}
-We propose a novel training regime called Counterfactual Training that
+We propose a novel training regime termed Counterfactual Training that
 leverages counterfactual explanations to increase the explanatory
@@
 decision-making systems, counterfactuals should be plausible with
-respect to the underlying data and actionable with respect to
+respect to the underlying data and actionable with respect to the
 stakeholder requirements. Much existing research has therefore focused
@@
 these desiderata. In this work, we instead hold models directly
-accountable for this desired end goal: Counterfactual Training employs
+accountable for the desired end goal: Counterfactual Training employs
 counterfactuals ad-hoc during the training phase to minimize the
@@
 Today's prominence of artificial intelligence (AI) has largely been
-driven by advances in \textbf{representation learning}: instead of
-relying on features and rules that are carefully hand-crafted by humans,
-modern machine learning (ML) models are tasked with learning
-representations directly from data, guided by narrow objectives such as
-predictive accuracy \cite{goodfellow2016deep}. Modern advances in
-computing have made it possible to provide such models with ever-growing
-degrees of freedom to achieve that task, which has often led them to
-outperform traditionally more parsimonious models. Unfortunately, in
-doing so, models learn increasingly complex and highly sensitive
-representations that we can no longer easily interpret.
+driven by \textbf{representation learning}: instead of relying on
+features and rules that are carefully hand-crafted by humans, modern
+machine learning (ML) models are tasked with learning representations
+directly from data, guided by narrow objectives such as predictive
+accuracy \cite{goodfellow2016deep}. Modern advances in computing have
+made it possible to provide such models with ever-growing degrees of
+freedom to achieve that task, which frequently allows them to outperform
+traditionally more parsimonious models. Unfortunately, in doing so,
+models learn increasingly complex and highly sensitive representations
+that humans can no longer easily interpret.
 
@@
 revolution, \cite{szegedy2013intriguing} showed that artificial neural
-networks (ANN) are sensitive to adversarial examples: counterfactuals of
-model inputs that yield vastly different model predictions despite being
-``imperceptible'' in that they are semantically indifferent from their
-factual counterparts. Although some partially effective mitigation
-strategies have been proposed, for example \textbf{adversarial training}
-\cite{goodfellow2014explaining}, truly robust deep learning (DL) remains
-unattainable even for models that are considered shallow by today's
-standards \cite{kolter2023keynote}.
+networks (ANN) are sensitive to adversarial examples: perturbed versions
+of data instances that yield vastly different model predictions despite
+being ``imperceptible'' in that they are semantically indifferent from
+their factual counterparts. Even though some partially effective
+mitigation strategies have been proposed---most notably
+\textbf{adversarial training} \cite{goodfellow2014explaining}---truly
+robust deep learning (DL) remains unattainable even for models that are
+considered shallow by today's standards \cite{kolter2023keynote}.
 
@@
 neural networks are so expressive that they can fit randomly labeled
-data. Thus, looking at the predictive performance, the solutions may
-seem to provide compelling explanations for the data, when in fact they
-are based on purely associative, semantically meaningless patterns. This
-poses two related challenges. Firstly, there is no dependable way to
-verify if such complex representations correspond to meaningful and
+data. Thus, looking at the predictive performance alone, the solutions
+may seem to provide compelling explanations for the data, when in fact
+they are based on purely associative, semantically meaningless patterns.
+This poses two related challenges. Firstly, there is no dependable way
+to verify if such complex representations correspond to meaningful and
 plausible explanations. Secondly, even if we could resolve the first
@@
 The first challenge has attracted an abundance of research on
-\textbf{explainable AI} (XAI) which aims to develop tools to derive
-explanations from complex model representations. This can mitigate a
-scenario in which we deploy opaque models and blindly rely on their
-predictions. On countless occasions, this scenario has occurred in
-practice and caused real harm to people who were affected adversely and
-often unfairly by automated decision-making (ADM) systems involving
-opaque models \cite{oneil2016weapons,mcgregor2021preventing}. Effective
-XAI tools can aid us in monitoring models and providing recourse to
-individuals to turn adverse outcomes (e.g., ``loan application
-rejected'') into positive ones (e.g., ``application accepted'').
-\cite{wachter2017counterfactual} propose \textbf{counterfactual
-explanations} (CE) as an effective approach to achieve this goal: CEs
-explain how factual inputs need to change in order for some fitted model
-to produce some desired output, typically involving minimal
-perturbations.
+\textbf{explainable AI} (XAI), a paradigm that focuses on the
+development of tools to derive (post-hoc) explanations from complex
+model representations. Such explanations should mitigate a scenario in
+which practitioners deploy opaque models and blindly rely on their
+predictions. On countless occasions, this has happened in practice and
+caused real harms to people who were adversely and unfairly affected by
+automated decision-making (ADM) systems involving opaque models
+\cite{oneil2016weapons,mcgregor2021preventing}. Effective XAI tools can
+aid us in monitoring models and providing recourse to individuals to
+turn negative outcomes (e.g., ``loan application rejected'') into
+positive ones (e.g., ``application accepted''). Our work builds upon
+\textbf{counterfactual explanations} (CE) proposed by
+\cite{wachter2017counterfactual} as an effective approach to achieve
+this goal. CEs prescribe minimal changes for factual inputs that, if
+implemented, would prompt some fitted model to produce a desired output.
 
-To our surprise, the second challenge has not yet attracted any major
-consolidated research effort. Specifically, there has been no concerted
-effort towards improving improving models' explanatory capacity, which
-we will henceforth simply call ``explainability'', defined as the degree
-to which learned representations correspond to explanations that are
-interpretable and deemed \textbf{plausible} by humans (see
-Definition~\ref{def-explainability}). Instead, the choice has typically
-been to improve the ability of XAI tools to identify the subset
+To our surprise, the second challenge has not yet attracted major
+research interest. Specifically, there has been no concerted effort
+towards improving the ``explanatory capacity'' of models, i.e., the
+degree to which learned representations correspond to explanations that
+are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
+Definition~\ref{def-explainability}). Instead, the choice has generally
+been to improve the ability of XAI tools to identify the subset of
 explanations that are both plausible and valid for any given model,
 independent of whether the learned representations are also compatible
-with implausible explanations \cite{altmeyer2024faithful}. Fortunately,
-recent findings indicate that explainability can arise as byproduct of
-regularization techniques aimed at other objectives such as robustness,
-generalization, and generative capacity
+with plausible explanations \cite{altmeyer2024faithful}. Fortunately,
+recent findings indicate that improved explanatory capacity can arise as
+a consequence of regularization techniques aimed at other training
+objectives such as robustness, generalization, and generative capacity
 \cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
+As further discussed in Section~\ref{sec-lit}, our work consolidates
+these findings within a single objective.
 
-Building on these findings, we introduce \textbf{counterfactual
-training}: a novel training regime explicitly geared towards aligning
-model representations with plausible explanations. Our contributions are
-as follows:
+\textbf{Specifically, we introduce counterfactual training}: a novel
+training regime explicitly meant to align learned representations with
+plausible explanations that comply with user requirements. Our
+contributions are as follows:
 
@@
 \item
-  We discuss existing related work on improving models and consolidate
-  it through the lens of counterfactual explanations
-  (Section~\ref{sec-lit}).
+  We present a novel methodological framework that leverages adversarial
+  examples and faithful counterfactual explanations during the training
+  phase to improve the explanatory capacity and robustness of machine
+  learning models (Section~\ref{sec-method}).\\
 \item
-  We present our proposed methodological framework that leverages
-  faithful counterfactual explanations during the training phase of
-  models to achieve the explainability objective
-  (Section~\ref{sec-method}).
+  We propose a method to enforce global actionability constraints by
+  preventing models from assigning importance to immutable features,
+  i.e., ones over which decision subjects have no control
+  (Section~\ref{sec-method}).\\
 \item
-  Through extensive experiments we demonstrate the counterfactual
-  training improve model explainability while maintaining high
-  predictive performance. We run ablation studies and grid searches to
-  understand how the underlying model components and hyperparameters
-  affect outcomes. (Section~\ref{sec-experiments}).
+  Through extensive experiments we demonstrate that counterfactual
+  training promotes explainability while preserving high predictive
+  performance. We run ablation studies and grid searches to understand
+  how the underlying model components and hyperparameters affect
+  outcomes. (Section~\ref{sec-experiments}).
 \end{itemize}
 
-Despite some limitations of our approach discussed in
-Section~\ref{sec-discussion}, we conclude in
-Section~\ref{sec-conclusion} that counterfactual training provides a
-practical framework for researchers and practitioners interested in
-making opaque models more trustworthy. We also believe that this work
+Despite some limitations discussed in Section~\ref{sec-discussion}, we
+conclude in Section~\ref{sec-conclusion} that counterfactual training
+provides a useful framework for researchers and practitioners interested
+in making opaque models more trustworthy. We also believe that this work
 serves as an opportunity for XAI researchers to re-evaluate the trend of
@@
   background information on gradient-based counterfactual search and
-  convergence see supplementary appendix.}
+  convergence see \textbf{?@sec-app-ce}.}
 
@@
 demanding datasets on a high-performance computing cluster with 34 2GB
-CPUs\footnote{See supplementary appendix for computational details.}.
-However, there are three factors that attenuate the impact of this
-limitation. First, CT provides counterfactual explanations for the
-training samples essentially for free, which may be beneficial in many
-ADM systems. Second, we find that CT can retain its value when used as a
-``fine-tuning'' training regime for conventionally-trained models.
-Third, in principle, CT yields itself to parallel execution, which we
-have leveraged for our own experiments.
+CPUs (see details in \textbf{?@sec-hardware}). However, there are three
+factors that attenuate the impact of this limitation. First, CT provides
+counterfactual explanations for the training samples essentially for
+free, which may be beneficial in many ADM systems. Second, we find that
+CT can retain its value when used as a ``fine-tuning'' training regime
+for conventionally-trained models. Third, in principle, CT yields itself
+to parallel execution, which we have leveraged for our own experiments.
 
@@
 \textbf{\emph{Interventions on features may have downstream impacts on
-fairness.}}\\
-- \href{https://arxiv.org/abs/2401.16088}{Equality of opportunity}?\\
-- \href{https://proceedings.mlr.press/v202/gao23d/gao23d.pdf}{Social
-segregation}?\\
-- Supporting AR within one context/system may still unfairly target
-recourse-affected individuals within other contexts/systems? In Example
-3.1 younger individuals gain access to a loan but their age could still
-lead to, e.g., enforcing stronger supervisory mechanisms?
+fairness.}} Related to the point above, we provide a tool that allows
+practitioners to modify the sensitivity of a model with respect to
+certain features, which may have implication for the fair and equitable
+treatment of individuals subject to automated decisions. As protecting a
+set of features leads the model to assign higher relative importance to
+unprotected features, model owners could misuse our solution by
+enforcing explanations based on features that are more difficult to
+modify by some (group of) individuals. For example, consider again the
+Adult dataset where features such as workclass or education may be more
+difficult to change for underpriviledged groups. When applied
+irresponsibly, counterfactual training could result in an unfairly
+assigned burden of recourse \cite{sharma2020certifai}, threatening the
+equality of opportunity in the system \cite{bell2024fairness} and
+potentially reinforcing social segregation \cite{gao2023impact}. Still,
+as the referenced publications indicate, such phenomena are not specific
+to CT; regrettably, all types of ADM solutions without strong external
+protections have been recognized to promote harmful power dynamics
+\cite{maas2023machine}.
 
@@
 
+\newpage{}
+
 \bibliographystyle{splncs04}

### 2025-03-12 paper/sections/intro.qmd
Review abstract and introduction

@@
 
-Today's prominence of artificial intelligence (AI) has largely been driven by advances in **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
-Modern advances in computing have made it possible to provide such models with ever-growing degrees of freedom to achieve that task, which has often led them to outperform traditionally more parsimonious models.
-Unfortunately, in doing so, models learn increasingly complex and highly sensitive representations that we can no longer easily interpret.
+Today's prominence of artificial intelligence (AI) has largely been driven by **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
+Modern advances in computing have made it possible to provide such models with ever-growing degrees of freedom to achieve that task, which frequently allows them to outperform traditionally more parsimonious models.
+Unfortunately, in doing so, models learn increasingly complex and highly sensitive representations that humans can no longer easily interpret.
 
 The trend towards complexity for the sake of performance has come under serious scrutiny in recent years.
-At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: counterfactuals of model inputs that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts.
-Although some partially effective mitigation strategies have been proposed, for example **adversarial training** [@goodfellow2014explaining], truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].
+At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: perturbed versions of data instances that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts.
+Even though some partially effective mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].
 
@@
 Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data.
-Thus, looking at the predictive performance, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns.
+Thus, looking at the predictive performance alone, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns.
 This poses two related challenges.
@@
 
-The first challenge has attracted an abundance of research on **explainable AI** (XAI) which aims to develop tools to derive explanations from complex model representations.
-This can mitigate a scenario in which we deploy opaque models and blindly rely on their predictions.
-On countless occasions, this scenario has occurred in practice and caused real harm to people who were affected adversely and often unfairly by automated decision-making (ADM) systems involving opaque models [@oneil2016weapons; @mcgregor2021preventing].
-Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn adverse outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted").
-@wachter2017counterfactual propose **counterfactual explanations** (CE) as an effective approach to achieve this goal: CEs explain how factual inputs need to change in order for some fitted model to produce some desired output, typically involving minimal perturbations.
+The first challenge has attracted an abundance of research on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations.
+Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions.
+On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@oneil2016weapons; @mcgregor2021preventing].
+Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted").
+Our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal.
+CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce a desired output.
 
-To our surprise, the second challenge has not yet attracted any major consolidated research effort.
-Specifically, there has been no concerted effort towards improving improving models' explanatory capacity, which we will henceforth simply call "explainability", defined as the degree to which learned representations correspond to explanations that are interpretable and deemed **plausible** by humans (see @def-explainability).
-Instead, the choice has typically been to improve the ability of XAI tools to identify the subset explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with implausible explanations [@altmeyer2024faithful].
-Fortunately, recent findings indicate that explainability can arise as byproduct of regularization techniques aimed at other objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful].
+To our surprise, the second challenge has not yet attracted major research interest.
+Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see @def-explainability).
+Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with plausible explanations [@altmeyer2024faithful].
+Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful].
+As further discussed in @sec-lit, our work consolidates these findings within a single objective.
 
-Building on these findings, we introduce **counterfactual training**: a novel training regime explicitly geared towards aligning model representations with plausible explanations.
+**Specifically, we introduce counterfactual training**: a novel training regime explicitly meant to align learned representations with plausible explanations that comply with user requirements.
 Our contributions are as follows:
 
-- We discuss existing related work on improving models and consolidate it through the lens of counterfactual explanations (@sec-lit).
-- We present our proposed methodological framework that leverages faithful counterfactual explanations during the training phase of models to achieve the explainability objective (@sec-method).
-- Through extensive experiments we demonstrate the counterfactual training improve model explainability while maintaining high predictive performance.
+- We present a novel methodological framework that leverages adversarial examples and faithful counterfactual explanations during the training phase to improve the explanatory capacity and robustness of machine learning models (@sec-method).
+- We propose a method to enforce global actionability constraints by preventing models from assigning importance to immutable features, i.e., ones over which decision subjects have no control (@sec-method).
+- Through extensive experiments we demonstrate that counterfactual training promotes explainability while preserving high predictive performance.
 We run ablation studies and grid searches to understand how the underlying model components and hyperparameters affect outcomes.
@@
 
-Despite some limitations of our approach discussed in @sec-discussion, we conclude in @sec-conclusion that counterfactual training provides a practical framework for researchers and practitioners interested in making opaque models more trustworthy.
+Despite some limitations discussed in @sec-discussion, we conclude in @sec-conclusion that counterfactual training provides a useful framework for researchers and practitioners interested in making opaque models more trustworthy.
 We also believe that this work serves as an opportunity for XAI researchers to re-evaluate the trend of improving XAI tools without improving the underlying models.

### 2025-03-12 paper/quarto_ecml/paper.tex
Review related literature

@@
 \newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
+\theoremstyle{plain}
+\newtheorem{corollary}{Research Question}[section]
+\theoremstyle{remark}
+\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
+\newtheorem*{remark}{Remark}
+\newtheorem*{solution}{Solution}
+\newtheorem{refremark}{Remark}[section]
+\newtheorem{refsolution}{Solution}[section]
 \makeatother
@@
 
-To the best of our knowledge, our proposed framework of counterfactual
+To the best of our knowledge, the proposed framework for counterfactual
 training represents the first attempt to use counterfactual explanations
@@
 derived for an opaque model are also deemed plausible with respect to
-the underlying data and stakeholder requirements. To make this more
+the underlying data and stakeholder requirements; the former means that
+the counterfactuals should comply with the distribution of the factual
+data, the latter means that they should respect arbitrary (global)
+actionability constraints. To make the desiderata for our framework more
 concrete, we follow \cite{augustin2020adversarial} in tying the concept
 of explainability to the quality of counterfactual explanations that we
-can generate for a given model. The authors show that counterfactual
-explanations---understood here as minimal input perturbations that yield
-some desired model prediction---are generally more meaningful if the
-underlying model is more robust to adversarial examples. We can make
-intuitive sense of this finding when looking at adversarial training
-(AT) through the lens of representation learning with high degrees of
-freedom: by inducing models to ``unlearn'' representations that are
-susceptible to worst-case counterfactuals (i.e., adversarial examples),
-AT effectively removes some implausible explanations from the solution
-space.
+can generate for a given model. The authors show that CEs---understood
+here as minimal input perturbations that yield some desired model
+prediction---are generally more meaningful if the underlying model is
+more robust to adversarial examples. We can make intuitive sense of this
+finding when looking at adversarial training (AT) through the lens of
+representation learning with high degrees of freedom. As argued before,
+learned representations may be sensitive to producing implausible
+explanations and mispredicting for worst-case counterfactuals (i.e.,
+adversarial examples). Thus, by inducing models to ``unlearn''
+susceptiblity to such examples, AT can effectively remove implausible
+explanations from the solution space.
 
@@
 This interpretation of the link between explainability through
-counterfactuals on one side, and robustness to adversarial examples on
-the other, is backed by empirical evidence.
+counterfactuals on one side and robustness to adversarial examples on
+the other is backed by empirical evidence.
 \cite{sauer2021counterfactual} demonstrate that using counterfactual
@@
 supervised settings where inputs may be reasonably mapped to multiple
-outputs. They, too, demonstrate the augmenting the training data of
-image classifiers can improve generalization. \cite{teney2020learning}
-propose an approach using counterfactuals in training that does not rely
-on data augmentation: they argue that counterfactual pairs typically
-already exist in training datasets. Specifically, their approach relies
-on identifying similar input samples with different annotations and
-ensuring that the gradient of the classifier aligns with the vector
-between such pairs of counterfactual inputs using the cosine distance as
-the loss function.
+outputs. They, too, demonstrate that augmenting the training data of
+image classifiers can improve generalization. Finally,
+\cite{teney2020learning} propose an approach using counterfactuals in
+training that does not rely on data augmentation: they argue that
+counterfactual pairs typically already exist in training datasets.
+Specifically, their approach relies on identifying similar input samples
+with different annotations and ensuring that the gradient of the
+classifier aligns with the vector between such pairs of counterfactual
+inputs using the cosine distance as the loss function.
 
 In the natural language processing (NLP) domain, counterfactuals have
-similarly been used to improve models through data augmentation:
-\cite{wu2021polyjuice}, propose \emph{Polyjuice}, a general-purpose
+similarly been used to improve models through data augmentation.
+\cite{wu2021polyjuice} propose \emph{Polyjuice}, a general-purpose
 counterfactual generator for language models. They demonstrate
-empirically that augmenting training data through \emph{Polyjuice}
-counterfactuals improves robustness in a number of NLP tasks.
-\cite{balashankar2023improving} also use \emph{Polyjuice} to augment NLP
-datasets through diverse counterfactuals and show that classifier
-robustness improves up to 20\%. Finally, \cite{luu2023counterfactual}
-introduce Counterfactual Adversarial Training (CAT), which also aims at
-improving generalization and robustness of language models.
-Specifically, they propose to proceed as follows: firstly, they identify
-training samples that are subject to high predictive uncertainty;
-secondly, they generate counterfactual explanations for those samples;
-and, finally, they fine-tune the given language model on the augmented
-dataset that includes the generated counterfactuals.
+empirically that the augmentation of training data through
+\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
+tasks. \cite{balashankar2023improving} similarly use \emph{Polyjuice} to
+augment NLP datasets through diverse counterfactuals and show that
+classifier robustness improves by up to 20\%. Finally,
+\cite{luu2023counterfactual} introduce Counterfactual Adversarial
+Training (CAT), which also aims at improving generalization and
+robustness of language models through a three-step procedure. First, the
+authors identify training samples that are subject to high predictive
+uncertainty. Second, they generate counterfactual explanations for those
+samples. Finally, they fine-tune the given language model on the
+augmented dataset that includes the generated counterfactuals.
 
@@
 between counterfactual explanations and adversarial examples (AE).
-Pointing to clear similarities in how CE and AE are generated,
+Pointing to clear similarities in how CEs and AEs are generated,
 \cite{freiesleben2022intriguing} makes the case for jointly studying the
-opaqueness and robustness problem in representation learning. Formally,
-AE can be seen as the subset of CE for which misclassification is
+opaqueness and robustness problems in representation learning. Formally,
+AEs can be seen as the subset of CEs for which misclassification is
 achieved \cite{freiesleben2022intriguing}. Similarly,
-\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
-certain conditions and derive theoretical upper bounds on the distances
+\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
+certain conditions and derive theoretical upper bounds on distances
 between them.
@@
 counterfactuals during training with the explicit goal of affecting
-certain properties of post-hoc counterfactual explanations. Firstly,
-\cite{ross2021learning} propose a way to train models that are
-guaranteed to provide recourse for individuals to move from an adverse
-outcome to some positive target class with high probability. Their
-approach builds on adversarial training, where in this context
-susceptibility to targeted adversarial examples for the positive class
-is explicitly induced. The proposed method allows for imposing a set of
-actionability constraints ex-ante: for example, users can specify that
-certain features (e.g., \emph{age}, \emph{gender}, \ldots) are
-immutable. Secondly, \cite{guo2023counternet} are the first to propose
-an end-to-end training pipeline that includes counterfactual
-explanations as part of the training procedure. In particular, they
-propose a specific network architecture that includes a predictor and CE
-generator network, where the parameters of the CE generator network are
-learnable. Counterfactuals are generated during each training iteration
-and fed back to the predictor network. In contrast to
-\cite{guo2023counternet}, we impose no restrictions on the neural
-network architecture at all.
+certain properties of the post-hoc counterfactual explanations. Firstly,
+\cite{ross2021learning} propose a way to train models that guarantee
+individual recourse to some positive target class with high probability.
+Their approach builds on adversarial training by explicitly inducing
+susceptibility to targeted adversarial examples for the positive class.
+Additionally, the proposed method allows for imposing a set of
+actionability constraints ex-ante. For example, users can specify that
+certain features (e.g., \emph{age}, \emph{gender}) are immutable.
+Secondly, \cite{guo2023counternet} are the first to propose an
+end-to-end training pipeline that includes counterfactual explanations
+as part of the training procedure. In particular, they propose a
+specific network architecture that includes a predictor and CE generator
+network, where the parameters of the CE generator network are learnable.
+Counterfactuals are generated during each training iteration and fed
+back to the predictor network. In contrast to \cite{guo2023counternet},
+we impose no restrictions on the neural network architecture at all.
 
@@
 objectives. The authors propose a way to generate counterfactuals that
-are maximally \textbf{faithful} to the model in that they are consistent
-with what the model has learned about the underlying data. Formally,
-they rely on tools from energy-based modelling to minimize the
-divergence between the distribution of counterfactuals and the
-conditional posterior over inputs learned by the model. Their proposed
+are maximally faithful to the model in that they are consistent with
+what the model has learned about the underlying data. Formally, they
+rely on tools from energy-based modelling to minimize the divergence
+between the distribution of counterfactuals and the conditional
+posterior over inputs learned by the model. Their proposed
 counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
 and only if the underlying model has learned representations that align
-with them. They find that both deep ensembles
+with them. The authors find that both deep ensembles
 \cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)

### 2025-03-12 paper/sections/related.qmd
Review related literature

@@
 
-To the best of our knowledge, our proposed framework of counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability.
-In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and stakeholder requirements.
-To make this more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of counterfactual explanations that we can generate for a given model.
-The authors show that counterfactual explanations---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples.
-We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom: by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e., adversarial examples), AT effectively removes some implausible explanations from the solution space.
+To the best of our knowledge, the proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability.
+In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and stakeholder requirements; the former means that the counterfactuals should comply with the distribution of the factual data, the latter means that they should respect arbitrary (global) actionability constraints.
+To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of counterfactual explanations that we can generate for a given model.
+The authors show that CEs---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples.
+We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom.
+As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., adversarial examples).
+Thus, by inducing models to "unlearn" susceptiblity to such examples, AT can effectively remove implausible explanations from the solution space.
 
@@
 
-This interpretation of the link between explainability through counterfactuals on one side, and robustness to adversarial examples on the other, is backed by empirical evidence.
+This interpretation of the link between explainability through counterfactuals on one side and robustness to adversarial examples on the other is backed by empirical evidence.
 @sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness.
 Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
-They, too, demonstrate the augmenting the training data of image classifiers can improve generalization.
-@teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
+They, too, demonstrate that augmenting the training data of image classifiers can improve generalization.
+Finally, @teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
 Specifically, their approach relies on identifying similar input samples with different annotations and ensuring that the gradient of the classifier aligns with the vector between such pairs of counterfactual inputs using the cosine distance as the loss function.
 
-In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose *Polyjuice*, a general-purpose counterfactual generator for language models.
-They demonstrate empirically that augmenting training data through *Polyjuice* counterfactuals improves robustness in a number of NLP tasks.
-@balashankar2023improving also use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves up to 20%.
-Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models.
-Specifically, they propose to proceed as follows: firstly, they identify training samples that are subject to high predictive uncertainty; secondly, they generate counterfactual explanations for those samples; and, finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.
+In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation.
+@wu2021polyjuice propose *Polyjuice*, a general-purpose counterfactual generator for language models.
+They demonstrate empirically that the augmentation of training data through *Polyjuice* counterfactuals improves robustness in a number of NLP tasks.
+@balashankar2023improving similarly use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%.
+Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models through a three-step procedure.
+First, the authors identify training samples that are subject to high predictive uncertainty.
+Second, they generate counterfactual explanations for those samples.
+Finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.
 
 There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples (AE).
-Pointing to clear similarities in how CE and AE are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problem in representation learning.
-Formally, AE can be seen as the subset of CE for which misclassification is achieved [@freiesleben2022intriguing].
-Similarly, @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive theoretical upper bounds on the distances between them.
+Pointing to clear similarities in how CEs and AEs are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problems in representation learning.
+Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing].
+Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions and derive theoretical upper bounds on distances between them.
 
-Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of post-hoc counterfactual explanations.
-Firstly, @ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals to move from an adverse outcome to some positive target class with high probability.
-Their approach builds on adversarial training, where in this context susceptibility to targeted adversarial examples for the positive class is explicitly induced.
-The proposed method allows for imposing a set of actionability constraints ex-ante: for example, users can specify that certain features (e.g., *age*, *gender*, ...) are immutable.
+Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of the post-hoc counterfactual explanations.
+Firstly, @ross2021learning propose a way to train models that guarantee individual recourse to some positive target class with high probability.
+Their approach builds on adversarial training by explicitly inducing susceptibility to targeted adversarial examples for the positive class.
+Additionally, the proposed method allows for imposing a set of actionability constraints ex-ante.
+For example, users can specify that certain features (e.g., *age*, *gender*) are immutable.
 Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes counterfactual explanations as part of the training procedure.
@@
 In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives.
-The authors propose a way to generate counterfactuals that are maximally **faithful** to the model in that they are consistent with what the model has learned about the underlying data.
+The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
 Formally, they rely on tools from energy-based modelling to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
 Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
-They find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.
+The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.
 

### 2025-03-12 paper/quarto_ecml/paper.tex
Review counterfactual training up to 3.4

@@
 \newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\usepackage{amsthm}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
-\theoremstyle{remark}
-\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
-\newtheorem*{remark}{Remark}
-\newtheorem*{solution}{Solution}
-\newtheorem{refremark}{Remark}[section]
-\newtheorem{refsolution}{Solution}[section]
+
 \makeatother
@@
 
-Counterfactual training combines ideas from adversarial training,
+Counterfactual training (CT) combines ideas from adversarial training,
 energy-based modelling and counterfactuals explanations with the
-explicit objective of aligning representations with plausible
-explanations that comply with user requirements. In the context of CEs,
-plausibility has broadly been defined as the degree to which
-counterfactuals comply with the underlying data generating process
+explicit goal of aligning representations with plausible explanations
+that comply with user requirements. In the context of CEs, plausibility
+has broadly been defined as the degree to which counterfactuals comply
+with the underlying data-generating process
 \cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
 Plausibility is a necessary but insufficient condition for using CEs to
-provide algorithmic recourse (AR) to individuals affected by opaque
-models in practice. This is because for recourse recommendations to be
-\textbf{actionable}, they need to not only result in plausible
-counterfactuals but also be attainable. A plausible CE for a rejected
-20-year-old loan applicant, for example, might reveal that their
-application would have been accepted, if only they were 20 years older.
-Ignoring all other features, this would comply with the definition of
-plausibility if 40-year-old individuals were in fact more credit-worthy
-on average than young adults. But of course this CE does not qualify for
-providing actionable recourse to the applicant since \emph{age} is not a
-(directly) mutable feature. For our intents and purposes, counterfactual
-training aims to improve model explainability by aligning models with
-counterfactuals that meet both desiderata, plausibility and
-actionability. Formally, we define explainability as follows:
+provide algorithmic recourse (AR) to individuals (negatively) affected
+by opaque models. For AR recommendations to be actionable, they need to
+not only result in plausible counterfactuals but also be attainable. A
+plausible CE for a rejected 20-year-old loan applicant, for example,
+might reveal that their application would have been accepted, if only
+they were 20 years older. Ignoring all other features, this would comply
+with the definition of plausibility if 40-year-old individuals were in
+fact more credit-worthy on average than young adults. But of course this
+CE does not qualify for providing actionable recourse to the applicant
+since \emph{age} is not a (directly) mutable feature. CT aims to improve
+model explainability by aligning models with counterfactuals that meet
+both desiderata: plausibility and actionability. Formally, we define
+explainability as follows:
 
@@
 We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
-that faithfully generated counterfactuals are plausible (i.e.~consistent
-with the data) and actionable. Formally, we define these properties as
-follows:
+that faithfully generated counterfactuals are plausible (i.e.,
+consistent with the data) and actionable. Formally, we define these
+properties as follows:
 
@@
 
-The definitions of faithfulness and plausibility in
-Definition~\ref{def-explainability} are the same as in
+The characterization of faithfulness and plausibility in
+Definition~\ref{def-explainability} is the same as in
 \cite{altmeyer2024faithful}, with adapted notation. Actionability
@@
 initially proposed by \cite{wachter2017counterfactual}. For our
-purposes, we let \(T\) vary and consider the counterfactual search as
+purposes, we let \(T\) vary and consider the counterfactual search to be
 converged as soon as the predicted probability for the target class has
-reached a pre-determined threshold, \(\tau\):
+reached a pre-determined threshold \(\tau\):
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
@@
 \begin{equation}\phantomsection\label{eq-obj}{
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
+\begin{split}
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
+\end{split}
 }\end{equation}
 
-where \(\text{yloss}(\cdot)\) is any conventional classification loss
-that induces discriminative performance (e.g., cross-entropy). The
-second and third terms in Equation~\ref{eq-obj} are explained in more
-detail below. For now, they can be sufficiently described as inducing
-explainability directly and indirectly by penalizing: (1) the
-contrastive divergence, \(\text{div}(\cdot)\), between mature
-counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
-(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
+where \(\text{yloss}(\cdot)\) is a classification loss that induces
+discriminative performance (e.g., cross-entropy). The second and third
+terms in Equation~\ref{eq-obj} are explained in detail below. For now,
+they can be sufficiently described as inducing explainability directly
+and indirectly by penalizing: (1) the contrastive divergence,
+\(\text{div}(\cdot)\), between mature counterfactuals
+\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
+adversarial loss, \(\text{advloss}(.)\), with respect to nascent
 counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
 \(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
-regularises the magnitude of the energy terms involved in
-\(\text{div}(\cdot)\) \cite{du2019implicit}. The tradeoff between the
-different components can be governed by adjusting the strengths of the
-penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
+regularizes the magnitude of the energy terms involved in
+\(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off between the
+components can be governed by adjusting the strengths of the penalties
+\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
 \(\lambda_{\text{reg}}\).
 
-\subsubsection{Directly Inducing Explainability through Contrastive
-Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}
+\subsection{Directly Inducing Explainability with Contrastive
+Divergence}\label{directly-inducing-explainability-with-contrastive-divergence}
 
@@
 samples from \(p(\mathbf{x})\) and the generated samples from
-\(p_\theta(\mathbf{x})\). They show that JEMs can be trained to perform
-well at both tasks by directly maximizing the joint log-likelihood
-factorized as
+\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
+perform well at both tasks by directly maximizing the joint
+log-likelihood factorized as
 \(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
-The first factor can be optimized using conventional cross-entropy as in
+The first term can be optimized using conventional cross-entropy as in
 Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
@@
 tend to do well with respect to the plausibility objective in
-Definition~\ref{def-explainability}. If we consider samples drawn from
-\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
-finding, because the JEM objective effectively minimizes the divergence
-between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
-To generate samples, \cite{grathwohl2020your} rely on Stochastic
-Gradient Langevin Dynamics (SGLD) using an uninformative prior for
-initialization. This is where we depart from their methodology: instead
-of SGLD, we propose to use counterfactual explainers to generate
-counterfactuals of observed training samples. Specifically, we have:
+Definition~\ref{def-explainability}. This follows directly if we
+consider samples drawn from \(p_\theta(\mathbf{x})\) as counterfactuals
+because the JEM objective effectively minimizes the divergence between
+the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To
+generate samples, \cite{grathwohl2020your} rely on Stochastic Gradient
+Langevin Dynamics (SGLD) using an uninformative prior for initialization
+but we depart from their methodology. Instead of SGLD, we propose to use
+counterfactual explainers to generate counterfactuals of observed
+training samples. Specifically, we have:
 
@@
 
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
-particular, we set
+where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
 \(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
 where \(y^+\) denotes the index of the randomly drawn target class,
-\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed data point
-sampled from target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\).
-Conditional on the target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes
-a mature counterfactual for a randomly sampled factual from a non-target
-class generated through a gradient-based counterfactual generator for at
-most \(T\) iterations. We define mature counterfactuals as those that
-have either exhausted \(T\) or reached convergence in terms of the
-pre-determined decision threshold earlier.
+\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed sample from
+target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\). Conditional on the
+target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes a mature
+counterfactual for a randomly sampled factual from a non-target class
+generated with a gradient-based CE generator for up to \(T\) iterations.
+Mature counterfactuals are ones that have either reached convergence
+wrt. the decision threshold \(\tau\) or exhausted \(T\).
 
 Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
-of observed training samples (positive samples) while at same time
-increasing the energy of counterfactuals (negative samples)
-\cite{du2019implicit}. As the generated counterfactuals get more
-plausible (Definition~\ref{def-explainability}) over the course of
-training, these two opposing effects gradually balance each out
-\cite{lippe2024uvadlc}.
+of observed training samples (positive samples) while increasing the
+energy of counterfactuals (negative samples) \cite{du2019implicit}. As
+the counterfactuals get more plausible
+(Definition~\ref{def-explainability}) during training, these opposing
+effects gradually balance each other out \cite{lippe2024uvadlc}.
 
@@
 explainers that have been proposed in the literature to meet different
-desiderata. Typically, these methods facilitate the imposition of domain
-and mutability constraints, for example. In principle, any existing
-approach for generating counterfactual explanations is viable, so long
-as it does not violate the faithfulness condition. Like JEMs
-\cite{murphy2022probabilistic}, counterfactual training can be
-considered as a form of contrastive representation learning.
+desiderata. For example, many methods facilitate the imposition of
+domain and mutability constraints. In principle, any existing approach
+for generating counterfactual explanations is viable, so long as it does
+not violate the faithfulness condition. Like JEMs
+\cite{murphy2022probabilistic}, CT can be considered a form of
+contrastive representation learning.
 
-\subsubsection{Indirectly Inducing Explainability through Adversarial
-Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}
+\subsection{Indirectly Inducing Explainability with Adversarial
+Robustness}\label{indirectly-inducing-explainability-with-adversarial-robustness}
 
@@
 \(\mathbf{x}^\prime\) can be repurposed as additional training samples
-\cite{luu2023counterfactual,balashankar2023improving} or adversarial
-examples \cite{freiesleben2022intriguing,pawelczyk2022exploring}. This
-leaves some flexibility with respect to the exact choice for
+\cite{luu2023counterfactual,balashankar2023improving} or AEs
+\cite{freiesleben2022intriguing,pawelczyk2022exploring}. This leaves
+some flexibility with respect to the choice for
 \(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
-functional form to use, though likely not the only reasonable choice, is
-inspired by adversarial training:
+functional form, but likely not the only sensible choice, is inspired by
+adversarial training:
 
@@
 Under this choice, we consider nascent counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
-magnitude of the perturbation to any individual feature is at most
-\(\varepsilon\). This is closely aligned with
-\cite{szegedy2013intriguing}, who define an adversarial attack as an
-``imperceptible non-random perturbation''. Thus, we choose to work with
-a different distinction between CE and AE than
-\cite{freiesleben2022intriguing}, who considers misclassification as the
-key distinguishing feature of AE. One of the key observations in this
-work is that we can leverage counterfactual explanations during training
-and get adversarial examples, essentially for free.
+\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
+perturbation to any single feature is at most \(\varepsilon\). This is
+closely aligned with \cite{szegedy2013intriguing} who define an
+adversarial attack as an ``imperceptible non-random perturbation''.
+Thus, we choose to work with a different distinction between CE and AE
+than \cite{freiesleben2022intriguing} who consider misclassification as
+the key distinguishing feature of AE. One of the key observations in
+this work is that we can leverage CEs during training and get
+adversarial examples essentially for free.
 
@@
 \(\lambda_{\text{reg}}\) may increase the variability in outcomes when
-combined with high values for any of the other penalties in
-Equation~\ref{eq-obj}, this effect is not very pronounced.
+combined with high values of \(\lambda_{\text{div}}\) or
+\(\lambda_{\text{adv}}\), this effect is not very pronounced.
 
@@
 find that it is not necessary to employ CT during the entire training
-phase to achieve the desired improvements in explainability. We have
-tested training models conventionally during the first half of training
-before switching to CT after this initial ``burn-in'' period and
-observed positive results. Put differently, CT may be a way to improve
-the explainability of trained models in a fine-tuning manner.
+phase to achieve the desired improvements in explainability. When
+training models conventionally during the first 50\% of epochs before
+switching to CT for the next 50\% of epochs, we observed positive
+results. Put differently, CT may be a way to improve the explainability
+of models in a fine-tuning manner.
 
@@
 benefit from investigating more sophisticated approaches towards
-hyperparameter tuning. Notably, counterfactual training is iterative
-which makes a variety of methods applicable, including Bayesian
-\cite{snoek2012practical} or gradient-based \cite{franceschi2017forward}
-optimization.
+hyperparameter tuning. Notably, CT is iterative which makes a variety of
+methods applicable, including Bayesian \cite{snoek2012practical} or
+gradient-based \cite{franceschi2017forward} optimization.
 
@@
 as the referenced publications indicate, such phenomena are not specific
-to CT; regrettably, all types of ADM solutions without strong external
-protections have been recognized to promote harmful power dynamics
+to CT; all types of ADM solutions without strong external protections
+have been recognized to promote harmful power dynamics
 \cite{maas2023machine}.

### 2025-03-12 paper/sections/discussion.qmd
Review counterfactual training up to 3.4

@@
 Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning.
-Notably, counterfactual training is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.
+Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.
 
@@
 When applied irresponsibly, counterfactual training could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact].
-Still, as the referenced publications indicate, such phenomena are not specific to CT; regrettably, all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].
+Still, as the referenced publications indicate, such phenomena are not specific to CT; all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].

### 2025-03-12 paper/sections/experiments.qmd
Review counterfactual training up to 3.4

@@
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
-While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values for any of the other penalties in @eq-obj, this effect is not very pronounced.
+While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
 
@@
 Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability.
-We have tested training models conventionally during the first half of training before switching to CT after this initial "burn-in" period and observed positive results.
-Put differently, CT may be a way to improve the explainability of trained models in a fine-tuning manner.
+When training models conventionally during the first 50% of epochs before switching to CT for the next 50% of epochs, we observed positive results.
+Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.

### 2025-03-12 paper/sections/method.qmd
Review counterfactual training up to 3.4

--- 2025-03-10 paper/sections/method.qmd
@@
 
-Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit objective of aligning representations with plausible explanations that comply with user requirements.
-In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful].
-Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals affected by opaque models in practice.
-This is because for recourse recommendations to be **actionable**, they need to not only result in plausible counterfactuals but also be attainable.
+Counterfactual training (CT) combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit goal of aligning representations with plausible explanations that comply with user requirements.
+In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful].
+Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models.
+For AR recommendations to be actionable, they need to not only result in plausible counterfactuals but also be attainable.
 A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older.
@@
 But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature.
-For our intents and purposes, counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata, plausibility and actionability.
+CT aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability.
 Formally, we define explainability as follows:
@@
 
-We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e.
-consistent with the data) and actionable.
+We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e., consistent with the data) and actionable.
 Formally, we define these properties as follows:
@@
 
-The definitions of faithfulness and plausibility in @def-explainability are the same as in @altmeyer2024faithful, with adapted notation.
+The characterization of faithfulness and plausibility in @def-explainability is the same as in @altmeyer2024faithful, with adapted notation.
 Actionability constraints in @def-explainability vary and depend on the context in which $\mathbf{M}_\theta$ is deployed.
@@
 Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual.
-For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]
+For our purposes, we let $T$ vary and consider the counterfactual search to be converged as soon as the predicted probability for the target class has reached a pre-determined threshold $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]
 
@@
 $$
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y})
+\begin{split}
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
+\end{split}
 $$ {#eq-obj}
 
-where $\text{yloss}(\cdot)$ is any conventional classification loss that induces discriminative performance (e.g., cross-entropy).
-The two additional components in @eq-obj are explained in more detail below.
+where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy).
+The second and third terms in @eq-obj are explained in detail below.
 For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
-The tradeoff between the different components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$.
+Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit].
+The trade-off between the components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.
 
-### Directly Inducing Explainability through Contrastive Divergence
+## Directly Inducing Explainability with Contrastive Divergence
 
 @grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
-They show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$.
-The first factor can be optimized using conventional cross-entropy as in @eq-obj.
+The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$.
+The first term can be optimized using conventional cross-entropy as in @eq-obj.
 Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.
@@
 A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability.
-If we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals, this is an expected finding, because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$.
-To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization.
-This is where we depart from their methodology: instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples.
+This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$.
+To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology.
+Instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples.
 Specifically, we have:
@@
 where $\mathcal{E}_\theta(\cdot)$ denotes the energy function.
-In particular, we set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$, and $\mathbf{x^+}$ denotes an observed data point sampled from target domain: $\mathbf{X}^+=\{\mathbf{x}:y=y^+\}$.
-Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated through a gradient-based counterfactual generator for at most $T$ iterations.
-We define mature counterfactuals as those that have either exhausted $T$ or reached convergence in terms of the pre-determined decision threshold earlier.
+We set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$, and $\mathbf{x^+}$ denotes an observed sample from target domain: $\mathbf{X}^+=\{\mathbf{x}:y=y^+\}$.
+Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations.
+Mature counterfactuals are ones that have either reached convergence wrt.
+the decision threshold $\tau$ or exhausted $T$.
 
-Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while at same time increasing the energy of counterfactuals (negative samples) [@du2019implicit].
-As the generated counterfactuals get more plausible (@def-explainability) over the course of training, these two opposing effects gradually balance each out [@lippe2024uvadlc].
+Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit].
+As the counterfactuals get more plausible (@def-explainability) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].
 
 The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
-Typically, these methods facilitate the imposition of domain and mutability constraints, for example.
+For example, many methods facilitate the imposition of domain and mutability constraints.
 In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition.
-Like JEMs [@murphy2022probabilistic], counterfactual training can be considered as a form of contrastive representation learning.
+Like JEMs [@murphy2022probabilistic], CT can be considered a form of contrastive representation learning.
 
-### Indirectly Inducing Explainability through Adversarial Robustness
+## Indirectly Inducing Explainability with Adversarial Robustness
 
-Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring].
-This leaves some flexibility with respect to the exact choice for $\text{advloss}(\cdot)$ in @eq-obj.
-An intuitive functional form to use, though likely not the only reasonable choice, is inspired by adversarial training:
+Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or AEs [@freiesleben2022intriguing;@pawelczyk2022exploring].
+This leaves some flexibility with respect to the choice for $\text{advloss}(\cdot)$ in @eq-obj.
+An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:
 
@@
 
-Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as adversarial examples as long as the magnitude of the perturbation to any individual feature is at most $\varepsilon$.
-This is closely aligned with @szegedy2013intriguing, who define an adversarial attack as an "imperceptible non-random perturbation".
-Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing, who considers misclassification as the key distinguishing feature of AE.
-One of the key observations in this work is that we can leverage counterfactual explanations during training and get adversarial examples, essentially for free.
+Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$.
+This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation".
+Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE.
+One of the key observations in this work is that we can leverage CEs during training and get adversarial examples essentially for free.
 

### 2025-03-13 paper/preprint/paper.tex
Review abstract and introduction
final table done

--- 2025-03-12 paper/preprint/paper.tex
@@
 \newtheorem{proposition}{Proposition}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
 \theoremstyle{plain}
 \newtheorem{corollary}{Research Question}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
 \theoremstyle{remark}
@@
 
-\section{Detailed Results}\label{detailed-results}
+\section{Details on Main Experiments}\label{details-on-main-experiments}
+
+\subsection{Final Hyperparameters}\label{final-hyperparameters}
+
+As discussed Section~\ref{sec-experiments}, CT is sensitive to certain
+hyperparameter choices. We study the effect of many hyperparameters
+extensively in Section~\ref{sec-app-grid}. For the main results, we tune
+a small set of key hyperparameters (Section~\ref{sec-app-tune}). The
+final choices for the main results are presented for each data set in
+Table~\ref{tbl-final-params} along with training, test and batch sizes.
+
+\begin{longtable}{ccccccc}
+
+\caption{\label{tbl-final-params}Final hyperparameters used for the main
+results for the different datasets.}
+
+\tabularnewline
+
+  \toprule
+  \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
+  \endfirsthead
+  \toprule
+  \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
+  \endhead
+  \bottomrule
+  \multicolumn{7}{r}{Continuing table below.}\\
+  \bottomrule
+  \endfoot
+  \endlastfoot
+  Adult & $2.6 \cdot 10^{4}$ & $5.01 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.75 & $5 \cdot 10^{3}$ & 0.25 \\
+  California Housing & $1.65 \cdot 10^{4}$ & $3.1 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
+  Circles & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.5 & $1 \cdot 10^{3}$ & 0.5 \\
+  Credit & $1.06 \cdot 10^{4}$ & $1.92 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
+  GMSC & $1.34 \cdot 10^{4}$ & $2.47 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.5 & $5 \cdot 10^{3}$ & 0.5 \\
+  Linearly Separable & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.5 & $1 \cdot 10^{3}$ & 0.25 \\
+  Moons & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.9 & $1 \cdot 10^{3}$ & 0.25 \\
+  Overlapping & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.5 & $1 \cdot 10^{3}$ & 0.25 \\\bottomrule
+
+\end{longtable}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\subsection{Grid Searches}\label{sec-app-grid}
+\section{Grid Searches}\label{sec-app-grid}
 
@@
 
-\subsubsection{Evaluation Details}\label{evaluation-details}
+\subsection{Evaluation Details}\label{evaluation-details}
 
@@
 
-\paragraph{Predictive Performance}\label{predictive-performance}
+\subsubsection{Predictive Performance}\label{predictive-performance}
 
@@
 
-\paragraph{Counterfactual Outcomes}\label{sec-app-grid-hl}
+\subsubsection{Counterfactual Outcomes}\label{sec-app-grid-hl}
 
@@
 broadly across datasets: plausibility is improved by up to
-\textasciitilde60 percentage points (ppts) for the \emph{Circles} data
-(e.g. Figure~\ref{fig-grid-gen_params-plaus-circles}),
-\textasciitilde25-30ppts for the \emph{Moons} data (e.g.
-Figure~\ref{fig-grid-gen_params-plaus-moons}) and
-\textasciitilde10-20ppts for the \emph{Linearly Separable} data (e.g.
+\textasciitilde60 percent (\%) for the \emph{Circles} data (e.g.
+Figure~\ref{fig-grid-gen_params-plaus-circles}), \textasciitilde25-30\%
+for the \emph{Moons} data (e.g.
+Figure~\ref{fig-grid-gen_params-plaus-moons}) and \textasciitilde10-20\%
+for the \emph{Linearly Separable} data (e.g.
 Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
 average costs of faithful counterfactuals are reduced in many cases by
-around \textasciitilde20-25ppts for \emph{Circles} (e.g.
+around \textasciitilde20-25\% for \emph{Circles} (e.g.
 Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
-\textasciitilde50ppts for \emph{Moons} (e.g.
+\textasciitilde50\% for \emph{Moons} (e.g.
 Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
 Separable} data, costs are generally increased although typically by
-less than 10ppts (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
+less than 10\% (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
 which reflects a common tradeoff between costs and plausibility
@@
 
-\subsubsection{Generator Parameters}\label{sec-app-grid-gen}
+\subsection{Generator Parameters}\label{sec-app-grid-gen}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\paragraph{Accuracy}\label{accuracy}
+\subsubsection{Accuracy}\label{accuracy}
 
@@
   Circles & F1-score & Vanilla & 0.998 & 0.000558 \\
-  Lin Sep & Accuracy & Full & 0.999 & 0.00201 \\
-  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
-  Lin Sep & F1-score & Full & 0.999 & 0.00201 \\
-  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Linearly Separable & Accuracy & Full & 0.999 & 0.00201 \\
+  Linearly Separable & Accuracy & Vanilla & 1 & 0 \\
+  Linearly Separable & F1-score & Full & 0.999 & 0.00201 \\
+  Linearly Separable & F1-score & Vanilla & 1 & 0 \\
   Moons & Accuracy & Full & 0.999 & 0.000696 \\
@@
   Moons & F1-score & Vanilla & 1 & 0.00111 \\
-  Over & Accuracy & Full & 0.915 & 0.00477 \\
-  Over & Accuracy & Vanilla & 0.917 & 0.00123 \\
-  Over & F1-score & Full & 0.915 & 0.00478 \\
-  Over & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule
+  Overlapping & Accuracy & Full & 0.915 & 0.00477 \\
+  Overlapping & Accuracy & Vanilla & 0.917 & 0.00123 \\
+  Overlapping & F1-score & Full & 0.915 & 0.00478 \\
+  Overlapping & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule
 
@@
 
-\paragraph{Plausibility}\label{plausibility-2}
+\subsubsection{Plausibility}\label{plausibility-2}
 
@@
 
-\paragraph{Cost}\label{cost}
+\subsubsection{Cost}\label{cost}
 
@@
 
-\subsubsection{Penalty Strengths}\label{sec-app-grid-pen}
+\subsection{Penalty Strengths}\label{sec-app-grid-pen}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\paragraph{Accuracy}\label{accuracy-1}
+\subsubsection{Accuracy}\label{accuracy-1}
 
@@
   Circles & F1-score & Vanilla & 0.998 & 0.000875 \\
-  Lin Sep & Accuracy & Full & 0.998 & 0.00772 \\
-  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
-  Lin Sep & F1-score & Full & 0.998 & 0.00773 \\
-  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Linearly Separable & Accuracy & Full & 0.998 & 0.00772 \\
+  Linearly Separable & Accuracy & Vanilla & 1 & 0 \\
+  Linearly Separable & F1-score & Full & 0.998 & 0.00773 \\
+  Linearly Separable & F1-score & Vanilla & 1 & 0 \\
   Moons & Accuracy & Full & 0.987 & 0.0351 \\
@@
   Moons & F1-score & Vanilla & 0.998 & 0.0102 \\
-  Over & Accuracy & Full & 0.911 & 0.0217 \\
-  Over & Accuracy & Vanilla & 0.916 & 0.00236 \\
-  Over & F1-score & Full & 0.911 & 0.0219 \\
-  Over & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule
+  Overlapping & Accuracy & Full & 0.911 & 0.0217 \\
+  Overlapping & Accuracy & Vanilla & 0.916 & 0.00236 \\
+  Overlapping & F1-score & Full & 0.911 & 0.0219 \\
+  Overlapping & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule
 
@@
 
-\paragraph{Plausibility}\label{plausibility-3}
+\subsubsection{Plausibility}\label{plausibility-3}
 
@@
 
-\paragraph{Cost}\label{cost-1}
+\subsubsection{Cost}\label{cost-1}
 
@@
 
-\subsubsection{Other Parameters}\label{sec-app-grid-train}
+\subsection{Other Parameters}\label{sec-app-grid-train}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\paragraph{Accuracy}\label{accuracy-2}
+\subsubsection{Accuracy}\label{accuracy-2}
 
@@
   Circles & F1-score & Vanilla & 0.998 & 0.000566 \\
-  Lin Sep & Accuracy & Full & 0.999 & 0.00231 \\
-  Lin Sep & Accuracy & Vanilla & 1 & 0 \\
-  Lin Sep & F1-score & Full & 0.999 & 0.00231 \\
-  Lin Sep & F1-score & Vanilla & 1 & 0 \\
+  Linearly Separable & Accuracy & Full & 0.999 & 0.00231 \\
+  Linearly Separable & Accuracy & Vanilla & 1 & 0 \\
+  Linearly Separable & F1-score & Full & 0.999 & 0.00231 \\
+  Linearly Separable & F1-score & Vanilla & 1 & 0 \\
   Moons & Accuracy & Full & 0.996 & 0.0136 \\
@@
   Moons & F1-score & Vanilla & 0.988 & 0.022 \\
-  Over & Accuracy & Full & 0.914 & 0.00563 \\
-  Over & Accuracy & Vanilla & 0.918 & 0.00116 \\
-  Over & F1-score & Full & 0.914 & 0.0057 \\
-  Over & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule
+  Overlapping & Accuracy & Full & 0.914 & 0.00563 \\
+  Overlapping & Accuracy & Vanilla & 0.918 & 0.00116 \\
+  Overlapping & F1-score & Full & 0.914 & 0.0057 \\
+  Overlapping & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule
 
@@
 
-\paragraph{Plausibility}\label{plausibility-4}
+\subsubsection{Plausibility}\label{plausibility-4}
 
@@
 
-\paragraph{Cost}\label{cost-2}
+\subsubsection{Cost}\label{cost-2}
 
@@
 
-\subsection{Hyperparameter Tuning}\label{sec-app-tune}
+\section{Tuning Key Parameters}\label{sec-app-tune}
 
@@
 \(\lambda_{\text{reg}}\), the strength for the engery regularization:
-plausibility is increased by up to \textasciitilde10ppt
+plausibility is increased by up to \textasciitilde10\%
 (Figure~\ref{fig-tune_lr-plaus-mnist}) and the proportion of mature
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-warning-color-frame, title={Package Version (Reproducibility)}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-warning-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Package Version (Reproducibility)}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-warning-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\subsubsection{Key Parameters}\label{sec-app-tune-key}
+\subsection{Key Parameters}\label{sec-app-tune-key}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\paragraph{Plausibility}\label{plausibility-5}
+\subsubsection{Plausibility}\label{plausibility-5}
 
@@
 
-\paragraph{Proportion of Mature CE}\label{proportion-of-mature-ce}
+\subsubsection{Proportion of Mature CE}\label{proportion-of-mature-ce}
 
@@
 
-\subsubsection{Learning Rate}\label{sec-app-tune-lr}
+\subsection{Learning Rate}\label{sec-app-tune-lr}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune_lr-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colback=white, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, breakable, toptitle=1mm, coltitle=black, arc=.35mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
 
@@
 
-\paragraph{Plausibility}\label{plausibility-6}
+\subsubsection{Plausibility}\label{plausibility-6}
 
@@
 
-\paragraph{Proportion of Mature CE}\label{proportion-of-mature-ce-1}
+\subsubsection{Proportion of Mature CE}\label{proportion-of-mature-ce-1}
 

### 2025-03-13 paper/sections/other/grids.qmd
final table done

@@
-## Grid Searches {#sec-app-grid}
+# Grid Searches {#sec-app-grid}
 
@@
 
-### Evaluation Details
+## Evaluation Details
 
@@
 
-#### Predictive Performance
+### Predictive Performance
 
@@
 
-#### Counterfactual Outcomes {#sec-app-grid-hl}
+### Counterfactual Outcomes {#sec-app-grid-hl}
 
-Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to ~60 percentage points (ppts) for the *Circles* data (e.g.
-@fig-grid-gen_params-plaus-circles), ~25-30ppts for the *Moons* data (e.g.
-@fig-grid-gen_params-plaus-moons) and ~10-20ppts for the *Linearly Separable* data (e.g.
+Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to ~60 percent (%) for the *Circles* data (e.g.
+@fig-grid-gen_params-plaus-circles), ~25-30% for the *Moons* data (e.g.
+@fig-grid-gen_params-plaus-moons) and ~10-20% for the *Linearly Separable* data (e.g.
 @fig-grid-gen_params-plaus-lin_sep).
-At the same time, the average costs of faithful counterfactuals are reduced in many cases by around ~20-25ppts for  *Circles* (e.g.
-@fig-grid-gen_params-cost-circles) and up to ~50ppts for *Moons* (e.g.
+At the same time, the average costs of faithful counterfactuals are reduced in many cases by around ~20-25% for  *Circles* (e.g.
+@fig-grid-gen_params-cost-circles) and up to ~50% for *Moons* (e.g.
 @fig-grid-gen_params-cost-moons).
-For the *Linearly Separable* data, costs are generally increased although typically by less than 10ppts (e.g.
+For the *Linearly Separable* data, costs are generally increased although typically by less than 10% (e.g.
 @fig-grid-gen_params-cost-lin_sep), which reflects a common tradeoff between costs and plausibility [@altmeyer2024faithful].
@@
 
-### Generator Parameters {#sec-app-grid-gen}
+## Generator Parameters {#sec-app-grid-gen}
 
@@
 
-#### Accuracy
+### Accuracy
 
@@
 #| output: asis
+
 df = CTExperiments.aggregate_performance(eval_grids; byvars=["objective"]) 
@@
 
-#### Plausibility
+### Plausibility
 
@@
 
-#### Cost
+### Cost
 
@@
 
-### Penalty Strengths {#sec-app-grid-pen}
+## Penalty Strengths {#sec-app-grid-pen}
 
@@
 
-### Other Parameters {#sec-app-grid-train}
+## Other Parameters {#sec-app-grid-train}
 

### 2025-03-13 paper/sections/other/tuning.qmd
final table done

@@
-## Hyperparameter Tuning {#sec-app-tune}
+# Tuning Key Parameters {#sec-app-tune}
 
@@
 For the given low decision threshold, we find that the learning rate has no discernable impact on the proportion of mature counterfactuals (@fig-tune_lr-mat-adult to @fig-tune_lr-mat-over).
-We do notice, however, that the results for MNIST are much improved when using a low value $\lambda_{\text{reg}}$, the strength for the engery regularization: plausibility is increased by up to ~10ppt (@fig-tune_lr-plaus-mnist) and the proportion of mature counterfactuals reaches 100%.
+We do notice, however, that the results for MNIST are much improved when using a low value $\lambda_{\text{reg}}$, the strength for the engery regularization: plausibility is increased by up to ~10% (@fig-tune_lr-plaus-mnist) and the proportion of mature counterfactuals reaches 100%.
 
@@
 
-### Key Parameters {#sec-app-tune-key}
+## Key Parameters {#sec-app-tune-key}
 
@@
 
-#### Proportion of Mature CE
+### Proportion of Mature CE
 
@@
 
-### Learning Rate {#sec-app-tune-lr}
+## Learning Rate {#sec-app-tune-lr}
 

### 2025-03-13 paper/experiments/CTExperiments/src/tables.jl

@@ -9,6 +9,7 @@ function tabulate_results(
     wrap_table=true,
     table_type=:longtable,
     longtable_footer="Continuing table below.",
+    save_name::Union{String,Nothing}=nothing,
     kwrgs...,
 )
     df = inputs[1]
@@ -16,36 +17,63 @@ function tabulate_results(
     if isa(other_inputs.backend, Val{:latex})
         formatters = PrettyTables.ft_latex_sn(3)
         tf = PrettyTables.tf_latex_booktabs
-        tab = pretty_table(
-            df;
-            tf=tf,
-            formatters=formatters,
-            wrap_table=wrap_table,
-            table_type=table_type,
-            longtable_footer=longtable_footer,
-            alignment=:c,
-            other_inputs...,
-            kwrgs...,
-        )
+        if isnothing(save_name)
+            tab = pretty_table(
+                df;
+                tf=tf,
+                formatters=formatters,
+                wrap_table=wrap_table,
+                table_type=table_type,
+                longtable_footer=longtable_footer,
+                alignment=:c,
+                other_inputs...,
+                kwrgs...,
+            )
+            return tab
+        else
+            open(save_name, "w") do io
+                pretty_table(
+                    io, df;
+                    tf=tf,
+                    formatters=formatters,
+                    wrap_table=wrap_table,
+                    table_type=table_type,
+                    longtable_footer=longtable_footer,
+                    alignment=:c,
+                    other_inputs...,
+                    kwrgs...,
+                )
+            end
+        end
     else
-        tab = pretty_table(df; alignment=:c, other_inputs..., kwrgs...)
+        if isnothing(save_name)
+            tab = pretty_table(df; alignment=:c, other_inputs..., kwrgs...)
+            return tab
+        else
+            open(save_name, "w") do io
+                pretty_table(io, df; alignment=:c, other_inputs..., kwrgs...)
+            end
+        end
     end
-    return tab
 end
 
 function get_table_inputs(
-    df::DataFrame, value_var::String="mean"; backend::Val=Val(:text), kwrgs...
+    df::DataFrame, value_var::Union{Nothing,String}="mean"; backend::Val=Val(:text), kwrgs...
 )
     df = deepcopy(df)
 
-    # Highlighters:
-    hls = value_highlighter(df, value_var; backend=backend, kwrgs...)
-    if "generator_type" in names(df)
-        # Filter out "omni":
[…]

### 2025-03-13 paper/preprint/paper.tex
Review counterfactual training

--- 2025-03-12 paper/preprint/paper.tex
@@
 \theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
+\newtheorem{proposition}{Proposition}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
 \theoremstyle{definition}
 \newtheorem{example}{Example}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
 \theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
+\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 Today's prominence of artificial intelligence (AI) has largely been
-driven by advances in \textbf{representation learning}: instead of
-relying on features and rules that are carefully hand-crafted by humans,
-modern machine learning (ML) models are tasked with learning
-representations directly from data, guided by narrow objectives such as
-predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
-Bengio, and Courville 2016}). Modern advances in computing have made it
-possible to provide such models with ever-growing degrees of freedom to
-achieve that task, which has often led them to outperform traditionally
-more parsimonious models. Unfortunately, in doing so, models learn
-increasingly complex and highly sensitive representations that we can no
-longer easily interpret.
+driven by \textbf{representation learning}: instead of relying on
+features and rules that are carefully hand-crafted by humans, modern
+machine learning (ML) models are tasked with learning representations
+directly from data, guided by narrow objectives such as predictive
+accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow, Bengio, and
+Courville 2016}). Modern advances in computing have made it possible to
+provide such models with ever-growing degrees of freedom to achieve that
+task, which frequently allows them to outperform traditionally more
+parsimonious models. Unfortunately, in doing so, models learn
+increasingly complex and highly sensitive representations that humans
+can no longer easily interpret.
 
@@
 showed that artificial neural networks (ANN) are sensitive to
-adversarial examples: counterfactuals of model inputs that yield vastly
-different model predictions despite being ``imperceptible'' in that they
-are semantically indifferent from their factual counterparts. Although
-some partially effective mitigation strategies have been proposed, for
-example \textbf{adversarial training}
+adversarial examples: perturbed versions of data instances that yield
+vastly different model predictions despite being ``imperceptible'' in
+that they are semantically indifferent from their factual counterparts.
+Even though some partially effective mitigation strategies have been
+proposed---most notably \textbf{adversarial training}
 (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow, Shlens, and
-Szegedy 2014}), truly robust deep learning (DL) remains unattainable
+Szegedy 2014})---truly robust deep learning (DL) remains unattainable
 even for models that are considered shallow by today's standards
@@
 state-of-the-art neural networks are so expressive that they can fit
-randomly labeled data. Thus, looking at the predictive performance, the
-solutions may seem to provide compelling explanations for the data, when
-in fact they are based on purely associative, semantically meaningless
-patterns. This poses two related challenges. Firstly, there is no
-dependable way to verify if such complex representations correspond to
-meaningful and plausible explanations. Secondly, even if we could
-resolve the first challenge, it remains undecided how to ensure that
-models can \emph{only} learn valuable explanations.
+randomly labeled data. Thus, looking at the predictive performance
+alone, the solutions may seem to provide compelling explanations for the
+data, when in fact they are based on purely associative, semantically
+meaningless patterns. This poses two related challenges. Firstly, there
+is no dependable way to verify if such complex representations
+correspond to meaningful and plausible explanations. Secondly, even if
+we could resolve the first challenge, it remains undecided how to ensure
+that models can \emph{only} learn valuable explanations.
 
 The first challenge has attracted an abundance of research on
-\textbf{explainable AI} (XAI) which aims to develop tools to derive
-explanations from complex model representations. This can mitigate a
-scenario in which we deploy opaque models and blindly rely on their
-predictions. On countless occasions, this scenario has occurred in
-practice and caused real harm to people who were affected adversely and
-often unfairly by automated decision-making (ADM) systems involving
-opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016};
+\textbf{explainable AI} (XAI), a paradigm that focuses on the
+development of tools to derive (post-hoc) explanations from complex
+model representations. Such explanations should mitigate a scenario in
+which practitioners deploy opaque models and blindly rely on their
+predictions. On countless occasions, this has happened in practice and
+caused real harms to people who were adversely and unfairly affected by
+automated decision-making (ADM) systems involving opaque models
+(\citeproc{ref-oneil2016weapons}{O'Neil 2016};
 \citeproc{ref-mcgregor2021preventing}{McGregor 2021}). Effective XAI
 tools can aid us in monitoring models and providing recourse to
-individuals to turn adverse outcomes (e.g., ``loan application
-rejected'') into positive ones (e.g., ``application accepted'').
+individuals to turn negative outcomes (e.g., ``loan application
+rejected'') into positive ones (e.g., ``application accepted''). Our
+work builds upon \textbf{counterfactual explanations} (CE) proposed by
 Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}) propose
-\textbf{counterfactual explanations} (CE) as an effective approach to
-achieve this goal: CEs explain how factual inputs need to change in
-order for some fitted model to produce some desired output, typically
-involving minimal perturbations.
-
-To our surprise, the second challenge has not yet attracted any major
-consolidated research effort. Specifically, there has been no concerted
-effort towards improving improving models' explanatory capacity, which
-we will henceforth simply call ``explainability'', defined as the degree
-to which learned representations correspond to explanations that are
-interpretable and deemed \textbf{plausible} by humans (see
-Definition~\ref{def-explainability}). Instead, the choice has typically
-been to improve the ability of XAI tools to identify the subset
+(\citeproc{ref-wachter2017counterfactual}{2017}) as an effective
+approach to achieve this goal. CEs prescribe minimal changes for factual
+inputs that, if implemented, would prompt some fitted model to produce a
+desired output.
+
+To our surprise, the second challenge has not yet attracted major
+research interest. Specifically, there has been no concerted effort
+towards improving the ``explanatory capacity'' of models, i.e., the
+degree to which learned representations correspond to explanations that
+are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
+Definition~\ref{def-explainability}). Instead, the choice has generally
+been to improve the ability of XAI tools to identify the subset of
 explanations that are both plausible and valid for any given model,
 independent of whether the learned representations are also compatible
-with implausible explanations
+with plausible explanations
 (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-Fortunately, recent findings indicate that explainability can arise as
-byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization, and generative capacity
-(\citeproc{ref-schut2021generating}{Schut et al. 2021};
+Fortunately, recent findings indicate that improved explanatory capacity
+can arise as a consequence of regularization techniques aimed at other
+training objectives such as robustness, generalization, and generative
+capacity (\citeproc{ref-schut2021generating}{Schut et al. 2021};
 \citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein 2020};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-
-Building on these findings, we introduce \textbf{counterfactual
-training}: a novel training regime explicitly geared towards aligning
-model representations with plausible explanations. Our contributions are
-as follows:
+\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). As further
+discussed in Section~\ref{sec-lit}, our work consolidates these findings
+within a single objective.
+
+\textbf{Specifically, we introduce counterfactual training}: a novel
+training regime explicitly meant to align learned representations with
+plausible explanations that comply with user requirements. Our
+contributions are as follows:
 
@@
 derived for an opaque model are also deemed plausible with respect to
-the underlying data and stakeholder requirements. To make this more
+the underlying data and stakeholder requirements; the former means that
+the counterfactuals should comply with the distribution of the factual
+data, the latter means that they should respect arbitrary (global)
+actionability constraints. To make the desiderata for our framework more
 concrete, we follow Augustin, Meinke, and Hein
@@
 explainability to the quality of counterfactual explanations that we can
-generate for a given model. The authors show that counterfactual
-explanations---understood here as minimal input perturbations that yield
-some desired model prediction---are generally more meaningful if the
-underlying model is more robust to adversarial examples. We can make
-intuitive sense of this finding when looking at adversarial training
-(AT) through the lens of representation learning with high degrees of
-freedom: by inducing models to ``unlearn'' representations that are
-susceptible to worst-case counterfactuals (i.e., adversarial examples),
-AT effectively removes some implausible explanations from the solution
-space.
+generate for a given model. The authors show that CEs---understood here
+as minimal input perturbations that yield some desired model
+prediction---are generally more meaningful if the underlying model is
+more robust to adversarial examples. We can make intuitive sense of this
+finding when looking at adversarial training (AT) through the lens of
+representation learning with high degrees of freedom. As argued before,
+learned representations may be sensitive to producing implausible
+explanations and mispredicting for worst-case counterfactuals (i.e.,
+adversarial examples). Thus, by inducing models to ``unlearn''
+susceptiblity to such examples, AT can effectively remove implausible
+explanations from the solution space.
 
@@
 This interpretation of the link between explainability through
-counterfactuals on one side, and robustness to adversarial examples on
-the other, is backed by empirical evidence. Sauer and Geiger
+counterfactuals on one side and robustness to adversarial examples on
+the other is backed by empirical evidence. Sauer and Geiger
 (\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
@@
 learning, especially in supervised settings where inputs may be
-reasonably mapped to multiple outputs. They, too, demonstrate the
+reasonably mapped to multiple outputs. They, too, demonstrate that
 augmenting the training data of image classifiers can improve
-generalization. Teney, Abbasnedjad, and Hengel
+generalization. Finally, Teney, Abbasnedjad, and Hengel
 (\citeproc{ref-teney2020learning}{2020}) propose an approach using
@@
 In the natural language processing (NLP) domain, counterfactuals have
-similarly been used to improve models through data augmentation: Wu et
-al. (\citeproc{ref-wu2021polyjuice}{2021}), propose \emph{Polyjuice}, a
+similarly been used to improve models through data augmentation. Wu et
+al. (\citeproc{ref-wu2021polyjuice}{2021}) propose \emph{Polyjuice}, a
 general-purpose counterfactual generator for language models. They
-demonstrate empirically that augmenting training data through
+demonstrate empirically that the augmentation of training data through
 \emph{Polyjuice} counterfactuals improves robustness in a number of NLP
 tasks. Balashankar et al.
-(\citeproc{ref-balashankar2023improving}{2023}) also use
+(\citeproc{ref-balashankar2023improving}{2023}) similarly use
 \emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
-and show that classifier robustness improves up to 20\%. Finally, Luu
+and show that classifier robustness improves by up to 20\%. Finally, Luu
 and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
 Counterfactual Adversarial Training (CAT), which also aims at improving
-generalization and robustness of language models. Specifically, they
-propose to proceed as follows: firstly, they identify training samples
-that are subject to high predictive uncertainty; secondly, they generate
-counterfactual explanations for those samples; and, finally, they
-fine-tune the given language model on the augmented dataset that
-includes the generated counterfactuals.
+generalization and robustness of language models through a three-step
+procedure. First, the authors identify training samples that are subject
+to high predictive uncertainty. Second, they generate counterfactual
+explanations for those samples. Finally, they fine-tune the given
+language model on the augmented dataset that includes the generated
+counterfactuals.
 
@@
 between counterfactual explanations and adversarial examples (AE).
-Pointing to clear similarities in how CE and AE are generated,
+Pointing to clear similarities in how CEs and AEs are generated,
 Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
-case for jointly studying the opaqueness and robustness problem in
-representation learning. Formally, AE can be seen as the subset of CE
+case for jointly studying the opaqueness and robustness problems in
+representation learning. Formally, AEs can be seen as the subset of CEs
 for which misclassification is achieved
@@
 Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
-CE and AE are equivalent under certain conditions and derive theoretical
-upper bounds on the distances between them.
+CEs and AEs are equivalent under certain conditions and derive
+theoretical upper bounds on distances between them.
 
@@
 counterfactuals during training with the explicit goal of affecting
-certain properties of post-hoc counterfactual explanations. Firstly,
+certain properties of the post-hoc counterfactual explanations. Firstly,
 Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
-propose a way to train models that are guaranteed to provide recourse
-for individuals to move from an adverse outcome to some positive target
-class with high probability. Their approach builds on adversarial
-training, where in this context susceptibility to targeted adversarial
-examples for the positive class is explicitly induced. The proposed
-method allows for imposing a set of actionability constraints ex-ante:
-for example, users can specify that certain features (e.g., \emph{age},
-\emph{gender}, \ldots) are immutable. Secondly, Guo, Nguyen, and Yadav
+propose a way to train models that guarantee individual recourse to some
+positive target class with high probability. Their approach builds on
+adversarial training by explicitly inducing susceptibility to targeted
+adversarial examples for the positive class. Additionally, the proposed
+method allows for imposing a set of actionability constraints ex-ante.
+For example, users can specify that certain features (e.g., \emph{age},
+\emph{gender}) are immutable. Secondly, Guo, Nguyen, and Yadav
 (\citeproc{ref-guo2023counternet}{2023}) are the first to propose an
@@
 authors propose a way to generate counterfactuals that are maximally
-\textbf{faithful} to the model in that they are consistent with what the
-model has learned about the underlying data. Formally, they rely on
-tools from energy-based modelling to minimize the divergence between the
+faithful to the model in that they are consistent with what the model
+has learned about the underlying data. Formally, they rely on tools from
+energy-based modelling to minimize the divergence between the
 distribution of counterfactuals and the conditional posterior over
@@
 \emph{ECCCo}, yields plausible explanations if and only if the
-underlying model has learned representations that align with them. They
-find that both deep ensembles
+underlying model has learned representations that align with them. The
+authors find that both deep ensembles
 (\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
@@
 
-Counterfactual training combines ideas from adversarial training,
+Counterfactual training (CT) combines ideas from adversarial training,
 energy-based modelling and counterfactuals explanations with the
-explicit objective of aligning representations with plausible
-explanations that comply with user requirements. In the context of CEs,
-plausibility has broadly been defined as the degree to which
-counterfactuals comply with the underlying data generating process
+explicit goal of aligning representations with plausible explanations
+that comply with user requirements. In the context of CEs, plausibility
+has broadly been defined as the degree to which counterfactuals comply
+with the underlying data-generating process
 (\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
@@
 is a necessary but insufficient condition for using CEs to provide
-algorithmic recourse (AR) to individuals affected by opaque models in
-practice. This is because for recourse recommendations to be
-\textbf{actionable}, they need to not only result in plausible
-counterfactuals but also be attainable. A plausible CE for a rejected
-20-year-old loan applicant, for example, might reveal that their
-application would have been accepted, if only they were 20 years older.
-Ignoring all other features, this would comply with the definition of
-plausibility if 40-year-old individuals were in fact more credit-worthy
-on average than young adults. But of course this CE does not qualify for
-providing actionable recourse to the applicant since \emph{age} is not a
-(directly) mutable feature. For our intents and purposes, counterfactual
-training aims to improve model explainability by aligning models with
-counterfactuals that meet both desiderata, plausibility and
-actionability. Formally, we define explainability as follows:
+algorithmic recourse (AR) to individuals (negatively) affected by opaque
+models. For AR recommendations to be actionable, they need to not only
+result in plausible counterfactuals but also be attainable. A plausible
+CE for a rejected 20-year-old loan applicant, for example, might reveal
+that their application would have been accepted, if only they were 20
+years older. Ignoring all other features, this would comply with the
+definition of plausibility if 40-year-old individuals were in fact more
+credit-worthy on average than young adults. But of course this CE does
+not qualify for providing actionable recourse to the applicant since
+\emph{age} is not a (directly) mutable feature. CT aims to improve model
+explainability by aligning models with counterfactuals that meet both
+desiderata: plausibility and actionability. Formally, we define
+explainability as follows:
 
@@
 
-The definitions of faithfulness and plausibility in
-Definition~\ref{def-explainability} are the same as in Altmeyer et al.
+The characterization of faithfulness and plausibility in
+Definition~\ref{def-explainability} is the same as in Altmeyer et al.
 (\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
@@
 (\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
-let \(T\) vary and consider the counterfactual search as converged as
+let \(T\) vary and consider the counterfactual search to be converged as
 soon as the predicted probability for the target class has reached a
-pre-determined threshold, \(\tau\):
+pre-determined threshold \(\tau\):
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
@@
 \begin{equation}\phantomsection\label{eq-obj}{
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
+\begin{split}
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
+\end{split}
 }\end{equation}
 
-where \(\text{yloss}(\cdot)\) is any conventional classification loss
-that induces discriminative performance (e.g., cross-entropy). The
-second and third terms in Equation~\ref{eq-obj} are explained in more
-detail below. For now, they can be sufficiently described as inducing
-explainability directly and indirectly by penalizing: (1) the
-contrastive divergence, \(\text{div}(\cdot)\), between mature
-counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
-(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
+where \(\text{yloss}(\cdot)\) is a classification loss that induces
+discriminative performance (e.g., cross-entropy). The second and third
+terms in Equation~\ref{eq-obj} are explained in detail below. For now,
+they can be sufficiently described as inducing explainability directly
+and indirectly by penalizing: (1) the contrastive divergence,
+\(\text{div}(\cdot)\), between mature counterfactuals
+\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
+adversarial loss, \(\text{advloss}(.)\), with respect to nascent
 counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
 \(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
-regularises the magnitude of the energy terms involved in
+regularizes the magnitude of the energy terms involved in
 \(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and Mordatch
-2020}). The tradeoff between the different components can be governed by
+2020}). The trade-off between the components can be governed by
 adjusting the strengths of the penalties \(\lambda_{\text{div}}\),
@@
 
-\subsubsection{Directly Inducing Explainability through Contrastive
-Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}
+\subsection{Directly Inducing Explainability with Contrastive
+Divergence}\label{directly-inducing-explainability-with-contrastive-divergence}
 
@@
 (training) samples from \(p(\mathbf{x})\) and the generated samples from
-\(p_\theta(\mathbf{x})\). They show that JEMs can be trained to perform
-well at both tasks by directly maximizing the joint log-likelihood
-factorized as
+\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
+perform well at both tasks by directly maximizing the joint
+log-likelihood factorized as
 \(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
-The first factor can be optimized using conventional cross-entropy as in
+The first term can be optimized using conventional cross-entropy as in
 Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
@@
 well with respect to the plausibility objective in
-Definition~\ref{def-explainability}. If we consider samples drawn from
-\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
-finding, because the JEM objective effectively minimizes the divergence
-between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
-To generate samples, Grathwohl et al.
+Definition~\ref{def-explainability}. This follows directly if we
+consider samples drawn from \(p_\theta(\mathbf{x})\) as counterfactuals
+because the JEM objective effectively minimizes the divergence between
+the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To
+generate samples, Grathwohl et al.
 (\citeproc{ref-grathwohl2020your}{2020}) rely on Stochastic Gradient
-Langevin Dynamics (SGLD) using an uninformative prior for
-initialization. This is where we depart from their methodology: instead
-of SGLD, we propose to use counterfactual explainers to generate
-counterfactuals of observed training samples. Specifically, we have:
+Langevin Dynamics (SGLD) using an uninformative prior for initialization
+but we depart from their methodology. Instead of SGLD, we propose to use
+counterfactual explainers to generate counterfactuals of observed
+training samples. Specifically, we have:
 
@@
 
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
-particular, we set
+where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
 \(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
 where \(y^+\) denotes the index of the randomly drawn target class,
-\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed data point
-sampled from target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\).
-Conditional on the target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes
-a mature counterfactual for a randomly sampled factual from a non-target
-class generated through a gradient-based counterfactual generator for at
-most \(T\) iterations. We define mature counterfactuals as those that
-have either exhausted \(T\) or reached convergence in terms of the
-pre-determined decision threshold earlier.
+\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed sample from
+target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\). Conditional on the
+target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes a mature
+counterfactual for a randomly sampled factual from a non-target class
+generated with a gradient-based CE generator for up to \(T\) iterations.
+Mature counterfactuals are ones that have either reached convergence
+wrt. the decision threshold \(\tau\) or exhausted \(T\).
 
 Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
-of observed training samples (positive samples) while at same time
-increasing the energy of counterfactuals (negative samples)
-(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
+of observed training samples (positive samples) while increasing the
+energy of counterfactuals (negative samples)
+(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the
 counterfactuals get more plausible (Definition~\ref{def-explainability})
-over the course of training, these two opposing effects gradually
-balance each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
+during training, these opposing effects gradually balance each other out
+(\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
 
@@
 explainers that have been proposed in the literature to meet different
-desiderata. Typically, these methods facilitate the imposition of domain
-and mutability constraints, for example. In principle, any existing
-approach for generating counterfactual explanations is viable, so long
-as it does not violate the faithfulness condition. Like JEMs
-(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), counterfactual
-training can be considered as a form of contrastive representation
-learning.
-
-\subsubsection{Indirectly Inducing Explainability through Adversarial
-Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}
+desiderata. For example, many methods facilitate the imposition of
+domain and mutability constraints. In principle, any existing approach
+for generating counterfactual explanations is viable, so long as it does
+not violate the faithfulness condition. Like JEMs
+(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), CT can be
+considered a form of contrastive representation learning.
+
+\subsection{Indirectly Inducing Explainability with Adversarial
+Robustness}\label{indirectly-inducing-explainability-with-adversarial-robustness}
 
@@
 (\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023};
-\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023}) or
-adversarial examples
+\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023}) or AEs
 (\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
 \citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
-leaves some flexibility with respect to the exact choice for
+leaves some flexibility with respect to the choice for
 \(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
-functional form to use, though likely not the only reasonable choice, is
-inspired by adversarial training:
+functional form, but likely not the only sensible choice, is inspired by
+adversarial training:
 
@@
 Under this choice, we consider nascent counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
-magnitude of the perturbation to any individual feature is at most
-\(\varepsilon\). This is closely aligned with Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2013}), who define an adversarial
+\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
+perturbation to any single feature is at most \(\varepsilon\). This is
+closely aligned with Szegedy et al.
+(\citeproc{ref-szegedy2013intriguing}{2013}) who define an adversarial
 attack as an ``imperceptible non-random perturbation''. Thus, we choose
 to work with a different distinction between CE and AE than Freiesleben
-(\citeproc{ref-freiesleben2022intriguing}{2022}), who considers
+(\citeproc{ref-freiesleben2022intriguing}{2022}) who consider
 misclassification as the key distinguishing feature of AE. One of the
-key observations in this work is that we can leverage counterfactual
-explanations during training and get adversarial examples, essentially
-for free.
+key observations in this work is that we can leverage CEs during
+training and get adversarial examples essentially for free.
 
@@
 Mutability constraints can similarly be enforced by setting partial
-derivatives to zero to ensure that features are only mutated in the
+derivatives to zero to ensure that features are only perturbed in the
 allowed direction, if at all.
 
-Since actionability constraints are binding at test time, we should also
-impose them when generating \(\mathbf{x}^\prime\) during each training
-iteration to align model representations with user requirements. Through
-their effect on \(\mathbf{x}^\prime\), both types of constraints
-influence model outcomes through Equation~\ref{eq-div}. Here it is
-crucial that we avoid penalizing implausibility that arises due to
-mutability constraints. For any mutability-constrained feature \(d\)
-this can be achieved by enforcing
-\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
-\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
-violate mutability constraints. Specifically, we set
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
+Since such actionability constraints are binding at test time, we should
+also impose them when generating \(\mathbf{x}^\prime\) during each
+training iteration to inform model representations. Through their effect
+on \(\mathbf{x}^\prime\), both types of constraints influence model
+outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
+penalizing implausibility that arises due to mutability constraints. For
+any mutability-constrained feature \(d\) this can be achieved by
+enforcing \(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever
+perturbing \(\mathbf{x}^\prime[d]\) in the direction of
+\(\mathbf{x}[d]\) would violate mutability constraints. Specifically, we
+set \(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
 
@@
 \item
-  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\) but feature \(d\) can
+  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\), but feature \(d\) can
   only be decreased in practice.
 \item
-  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\) but feature \(d\) can
+  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\), but feature \(d\) can
   only be increased in practice.
@@
 independence of features. This raises a valid concern about the effect
-of protecting immutable features in the presence of proxy features that
-remain unprotected. We discuss this limitation in
-Section~\ref{sec-discussion}.
-
-\subsection{Illustration}\label{illustration}
+of protecting immutable features in the presence of proxies that remain
+unprotected. We address this in Section~\ref{sec-discussion}.
+
+\subsection{Example: influence of counterfactual training on
+explainability}\label{example-influence-of-counterfactual-training-on-explainability}
 
@@
 previous loan takers comprised of a binary outcome variable
-(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
-(1) the subjects' \emph{age}, which we define as immutable, and (2) the
-subjects' existing level of \emph{debt}, which we define as mutable.
-
-We have simulated this scenario using synthetic data with independent
-features and Gaussian class-conditional densities in
+(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
+features: (1) the subjects' \emph{age}, which we define as immutable,
+and (2) the subjects' existing level of \emph{debt}, which we define as
+mutable.
+
+We have simulated this scenario using synthetic data with two
+independent features and Gaussian class-conditional densities in
 Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
@@
 architecture each time (a linear classifier). In each case, we show the
-linear decision boundary (green) and the training data colored according
-to their ground-truth label: orange points belong to the target class,
+decision boundary (in green) and the training data colored according to
+their ground-truth label: orange points belong to the target class,
 \(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
@@
 impose mutability constraints at test time. The generated
-counterfactuals are all valid, but not plausible: they are clearly
-distinguishable from the ground-truth data. In panel (b), we have
-trained our model with counterfactual training, once again not imposing
-mutability constraints at test time. We observe that the counterfactuals
-are clearly plausible, therefore meeting the first objective of
+counterfactuals are all valid, but not plausible: they do not comply
+with the distribution of the factual samples in the target class to the
+point where they are clearly distinguishable from the ground-truth data.
+In panel (b), we have trained our model with counterfactual training,
+once again without any mutability constraints. We observe that the
+counterfactuals are highly plausible, meeting the first objective of
 Definition~\ref{def-explainability}.
@@
 counterfactual paths are shorter on average in panel (d), where we have
-used counterfactual training and protected immutable features as
-described in Section~\ref{sec-constraints}. In particular, we observe
-that due to the classifier's lower sensitivity to \emph{age}, recourse
-recommendations with respect to \emph{debt} are much more homogenous, in
-that they do not disproportionately punish younger individuals. The
-counterfactuals are also plausible with respect to the mutable feature.
-Thus, we consider the model in panel (d) as the most explainable
-according to Definition~\ref{def-explainability}.
+used counterfactual training and protected the immutable feature as
+described in Section~\ref{sec-constraints}. We observe that due to the
+classifier's lower sensitivity to \emph{age}, recourse recommendations
+with respect to \emph{debt} are much more homogenous and do not
+disproportionately punish younger individuals. The counterfactuals are
+also plausible with respect to the mutable feature. Thus, we consider
+the model in panel (d) as the most explainable according to
+Definition~\ref{def-explainability}.
 
@@
 
-\caption{\label{fig-poc}Visual illustration of how counterfactual
-training improves explainability. See Example~\ref{exm-grad} for
-details.}
+\caption{\label{fig-poc}Illustration for Example~\ref{exm-grad} of how
+CT improves model explainability.}
 
@@
 
-Does our proposed counterfactual training objective
+To what extent does our proposed counterfactual training objective
 (Equation~\ref{eq-obj}) induce models to learn plausible explanations?
@@
 
-Does our proposed counterfactual training objective
+To what extent does our proposed counterfactual training objective
 (Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
@@
 
-What are the effects of different hyperparameter choices with respect to
+What are the effects of different hyperparameter choices wrt.
 Equation~\ref{eq-obj}?
@@
 benefit from investigating more sophisticated approaches towards
-hyperparameter tuning. Notably, counterfactual training is iterative
-which makes a variety of methods applicable, including Bayesian (e.g.,
+hyperparameter tuning. Notably, CT is iterative which makes a variety of
+methods applicable, including Bayesian (e.g.,
 \citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
@@
 Lakkaraju 2023}). Still, as the referenced publications indicate, such
-phenomena are not specific to CT; regrettably, all types of ADM
-solutions without strong external protections have been recognized to
-promote harmful power dynamics (\citeproc{ref-maas2023machine}{Maas
-2023}).
+phenomena are not specific to CT; all types of ADM solutions without
+strong external protections have been recognized to promote harmful
+power dynamics (\citeproc{ref-maas2023machine}{Maas 2023}).
 

### 2025-03-13 paper/quarto_ecml/paper.tex
Review counterfactual training

@@
 enforced by setting partial derivatives to zero to ensure that features
-are only mutated in the allowed direction, if at all.
+are only perturbed in the allowed direction, if at all.
 
-Since actionability constraints are binding at test time, we should also
-impose them when generating \(\mathbf{x}^\prime\) during each training
-iteration to align model representations with user requirements. Through
-their effect on \(\mathbf{x}^\prime\), both types of constraints
-influence model outcomes through Equation~\ref{eq-div}. Here it is
-crucial that we avoid penalizing implausibility that arises due to
-mutability constraints. For any mutability-constrained feature \(d\)
-this can be achieved by enforcing
-\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
-\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
-violate mutability constraints. Specifically, we set
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
+Since such actionability constraints are binding at test time, we should
+also impose them when generating \(\mathbf{x}^\prime\) during each
+training iteration to inform model representations. Through their effect
+on \(\mathbf{x}^\prime\), both types of constraints influence model
+outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
+penalizing implausibility that arises due to mutability constraints. For
+any mutability-constrained feature \(d\) this can be achieved by
+enforcing \(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever
+perturbing \(\mathbf{x}^\prime[d]\) in the direction of
+\(\mathbf{x}[d]\) would violate mutability constraints. Specifically, we
+set \(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
 
@@
 independence of features. This raises a valid concern about the effect
-of protecting immutable features in the presence of proxy features that
-remain unprotected. We discuss this limitation in
-Section~\ref{sec-discussion}.
+of protecting immutable features in the presence of proxies that remain
+unprotected. We address this in Section~\ref{sec-discussion}.
 
-\subsection{Illustration}\label{illustration}
+\subsection{Example: influence of counterfactual training on
+explainability}\label{example-influence-of-counterfactual-training-on-explainability}
 
@@
 previous loan takers comprised of a binary outcome variable
-(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
-(1) the subjects' \emph{age}, which we define as immutable, and (2) the
-subjects' existing level of \emph{debt}, which we define as mutable.
+(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
+features: (1) the subjects' \emph{age}, which we define as immutable,
+and (2) the subjects' existing level of \emph{debt}, which we define as
+mutable.
 
-We have simulated this scenario using synthetic data with independent
-features and Gaussian class-conditional densities in
+We have simulated this scenario using synthetic data with two
+independent features and Gaussian class-conditional densities in
 Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
@@
 
-What are the effects of different hyperparameter choices with respect to
+What are the effects of hyperparameter selection with respect to
 Equation~\ref{eq-obj}?

### 2025-03-13 paper/sections/experiments.qmd
final table done
Review counterfactual training

@@
 
-Does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?
+To what extent does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?
 
@@
 
-Does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+To what extent does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
 
@@
 
-What are the effects of different hyperparameter choices with respect to @eq-obj?
+What are the effects of hyperparameter selection with respect to @eq-obj?
 

### 2025-03-13 paper/sections/method.qmd
Review counterfactual training

@@
 In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain.
-Mutability constraints can similarly be enforced by setting partial derivatives to zero to ensure that features are only mutated in the allowed direction, if at all.
+Mutability constraints can similarly be enforced by setting partial derivatives to zero to ensure that features are only perturbed in the allowed direction, if at all.
 
-Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to align model representations with user requirements.
-Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes through @eq-div.
+Since such actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations.
+Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div.
 Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints.
@@
 1. Feature $d$ is strictly immutable in practice.
-2. We have $\mathbf{x}[d]>\mathbf{x}^\prime[d]$ but feature $d$ can only be decreased in practice.
-3. We have $\mathbf{x}[d]<\mathbf{x}^\prime[d]$ but feature $d$ can only be increased in practice.
+2. We have $\mathbf{x}[d]>\mathbf{x}^\prime[d]$, but feature $d$ can only be decreased in practice.
+3. We have $\mathbf{x}[d]<\mathbf{x}^\prime[d]$, but feature $d$ can only be increased in practice.
 
@@
 It is worth highlighting that @prp-mtblty assumes independence of features.
-This raises a valid concern about the effect of protecting immutable features in the presence of proxy features that remain unprotected.
-We discuss this limitation in @sec-discussion.
+This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected.
+We address this in @sec-discussion.
 
-## Illustration
+## Example: influence of counterfactual training on explainability
 
@@
 Suppose we are interested in predicting the likelihood that loan applicants default on their credit.
-We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.
+We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.
 
-We have simulated this scenario using synthetic data with independent features and Gaussian class-conditional densities in @fig-poc.
+We have simulated this scenario using synthetic data with two independent features and Gaussian class-conditional densities in @fig-poc.
 The four panels in @fig-poc show the outcomes for different training procedures using the same model architecture each time (a linear classifier).
-In each case, we show the linear decision boundary (green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$.
+In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$.
 Stars indicate counterfactuals in the target class generated at test time using generic gradient descent until convergence.
@@
 In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time.
-The generated counterfactuals are all valid, but not plausible: they are clearly distinguishable from the ground-truth data.
-In panel (b), we have trained our model with counterfactual training, once again not imposing mutability constraints at test time.
-We observe that the counterfactuals are clearly plausible, therefore meeting the first objective of @def-explainability.
+The generated counterfactuals are all valid, but not plausible: they do not comply with the distribution of the factual samples in the target class to the point where they are clearly distinguishable from the ground-truth data.
+In panel (b), we have trained our model with counterfactual training, once again without any mutability constraints.
+We observe that the counterfactuals are highly plausible, meeting the first objective of @def-explainability.
 
@@
 Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants).
-By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected immutable features as described in @sec-constraints.
-In particular, we observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous, in that they do not disproportionately punish younger individuals.
+By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected the immutable feature as described in @sec-constraints.
+We observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous and do not disproportionately punish younger individuals.
 The counterfactuals are also plausible with respect to the mutable feature.
@@
 
-![Visual illustration of how counterfactual training improves explainability. See @exm-grad for details.](/paper/figures/poc.svg){#fig-poc}
+![Illustration for @exm-grad of how CT improves model explainability.](/paper/figures/poc.svg){#fig-poc}

### 2025-03-13 paper/quarto_ecml/paper.tex
Review experiments

@@
 
-Our key outcome of interest is how well models perform with respect to
-explainability (Definition~\ref{def-explainability}): to this end, we
+Our key outcome of interest is how well do models perform with respect
+to explainability (Definition~\ref{def-explainability}). To this end, we
 focus primarily on the plausibility and cost of faithfully generated
@@
 population maximum mean discrepancy \cite{gretton2012kernel}. The metric
-in Equation~\ref{eq-impl-div} is equal to zero iff
-\(\mathbf{X}^\prime=\mathbf{X}^+\).
+in Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
+are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
 
@@
 
-We run the experiments with three CE generators: \emph{Generic} of
-\cite{wachter2017counterfactual} as a simple baseline approach,
-\emph{REVISE} \cite{joshi2019realistic} that aims to generate plausible
-counterfactuals using a surrogate Variational Autoencoder (VAE), and
-\emph{ECCo}---the generator of \cite{altmeyer2023faithful} but without
-the conformal prediction component---as a method that directly targets
-both faithfulness and plausibility of the CEs.
+We run the experiments with three gradient-based generators:
+\emph{Generic} of \cite{wachter2017counterfactual} as a simple baseline
+approach, \emph{REVISE} \cite{joshi2019realistic} that aims to generate
+plausible counterfactuals using a surrogate Variational Autoencoder
+(VAE), and \emph{ECCo}---the generator of \cite{altmeyer2023faithful}
+but without the conformal prediction component---as a method that
+directly targets both faithfulness and plausibility of the CEs.
 
@@
 
-We begin the discussion by addressing the direct extensions of the
-counterfactual training approach in Section~\ref{sec-approach}. Then, we
-look at its broader limitations and challenges in
-Section~\ref{sec-limitations}.
+We first address the direct extensions of the counterfactual training
+approach in Section~\ref{sec-approach}. Then, we look at its limitations
+and challenges in Section~\ref{sec-limitations}.
 

### 2025-03-13 paper/sections/discussion.qmd
Review experiments

@@
 
-We begin the discussion by addressing the direct extensions of the counterfactual training approach in @sec-approach. Then, we look at its broader limitations and challenges in @sec-limitations.
+We first address the direct extensions of the counterfactual training approach in @sec-approach. Then, we look at its limitations and challenges in @sec-limitations.
 

### 2025-03-13 paper/sections/experiments.qmd
Review experiments

@@
 
-Our key outcome of interest is how well models perform with respect to explainability (@def-explainability): to this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
+Our key outcome of interest is how well do models perform with respect to explainability (@def-explainability).
+To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
 To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.
@@
 where $\mathbf{X}^\prime$ denotes a set of multiple counterfactuals and $\text{MMD}(\cdot)$ is an unbiased estimate of the squared population maximum mean discrepancy [@gretton2012kernel].
-The metric in @eq-impl-div is equal to zero iff $\mathbf{X}^\prime=\mathbf{X}^+$.
+The metric in @eq-impl-div is equal to zero iff the two distributions are the same, $\mathbf{X}^\prime=\mathbf{X}^+$.
 
@@
 
-We run the experiments with three CE generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2023faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the CEs.
+We run the experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2023faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the CEs.
 

### 2025-03-13 paper/quarto_ecml/paper.tex
Review discussion

@@
 through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
-arbitrarily defined by the user, CT requires the output space
-\(\mathcal{Y}\) to be discrete. Thus, it applies to binary and
-multi-class classification but it is not well-defined for other ML tasks
-where the change in outcome with respect to a decision threshold
-\(\tau\) cannot be readily quantified. In fact, this is a common
-restriction in research on CEs and AR that predominantly focuses on
-classification models. Although other settings have attracted some
-interest \cite{spooner2021counterfactual,zhao2023counterfactual}, there
-is still no consensus on what constitutes a counterfactual in such
-settings.
+arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
+discrete. Thus, it does not apply to ML tasks where the change in
+outcome cannot be readily quantified. Focus on classification models is
+a common restriction in research on CEs and AR. Other settings have
+attracted some interest (e.g., regression in
+\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
+little consensus how to robustly extend the notion of counterfactuals.
 
@@
 a sufficiently large number of counterfactuals during each training
-epoch and including only mature counterfactuals for contrastive
+epoch, and including only mature counterfactuals for contrastive
 divergence.
 
-\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
-discussed in Section~\ref{sec-hyperparameters}, our method benefits from
-tuning certain key hyperparameters. In this work, we have relied
+\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
+benefits from tuning certain key hyperparameters (see
+Section~\ref{sec-hyperparameters}). In this work, we have relied
 exclusively on grid search for this task. Future work on CT could
@@
 factors that attenuate the impact of this limitation. First, CT provides
-counterfactual explanations for the training samples essentially for
-free, which may be beneficial in many ADM systems. Second, we find that
-CT can retain its value when used as a ``fine-tuning'' training regime
-for conventionally-trained models. Third, in principle, CT yields itself
-to parallel execution, which we have leveraged for our own experiments.
+CEs for the training samples for free, which may be beneficial in many
+ADM systems. Second, we find that CT can retain its value when used as a
+``fine-tuning'' training regime for conventionally-trained models.
+Third, in principle, CT yields itself to parallel execution, which we
+have leveraged for our own experiments.
 
-\textbf{\emph{Immutable features may have proxies.}} In
-Proposition~\ref{prp-mtblty} we define an approach to protect immutable
-features and thus increase the actionability of the generated
-counterfactuals. However, this approach requires that model owners
-define the mutability constraints for (all) features considered by the
-model. Even with sufficient domain knowledge to protect all immutable
-features---ones that cannot be changed at all and ones that cannot be
-reasonably expected to change---there may exist proxies that are
-theoretically mutable (and hence should not be protected) but preserve
-enough information about the principals to counteract the protections.
-As one example, consider the Adult dataset used in our experiments where
-the mutable education status is a proxy for the immutable age, in that
-the attainment of degrees is correlated with age. Delineating
-actionability is a major undecided challenge in the AR literature
+\textbf{\emph{Immutable features may have proxies.}} We propose an
+approach to protect immutable features and thus increase the
+actionability of the generated CEs. However, it requires that model
+owners define the mutability constraints for (all) features considered
+by the model. Even with sufficient domain knowledge to protect all
+immutable features, there may exist proxies that are theoretically
+mutable (and hence should not be protected) but preserve enough
+information about the principals to hinder the protections. As an
+example, consider the Adult dataset used in our experiments where the
+mutable education status is a proxy for the immutable age, in that the
+attainment of degrees is correlated with age. Delineating actionability
+is a major undecided challenge in the AR literature
 \cite{venkatasubramanian2020philosophical} impacting the capacity of CT
@@
 
-\textbf{\emph{Interventions on features may have downstream impacts on
-fairness.}} Related to the point above, we provide a tool that allows
+\textbf{\emph{Interventions on features may impact fairness
+downstream.}} Related to the point above, we provide a tool that allows
 practitioners to modify the sensitivity of a model with respect to
@@
 
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
-
 \newpage{}

### 2025-03-13 paper/sections/discussion.qmd
Review discussion

@@
 ***CT is defined only for classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj.
-While $y^{-}$ and $y^{+}$ can be arbitrarily defined by the user, CT requires the output space $\mathcal{Y}$ to be discrete.
-Thus, it applies to binary and multi-class classification but it is not well-defined for other ML tasks where the change in outcome with respect to a decision threshold $\tau$ cannot be readily quantified.
-In fact, this is a common restriction in research on CEs and AR that predominantly focuses on classification models.
-Although other settings have attracted some interest [e.g., regression in @spooner2021counterfactual; @zhao2023counterfactual], there is still no consensus on what constitutes a counterfactual in such settings.
+While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete.
+Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified.
+Focus on classification models is a common restriction in research on CEs and AR.
+Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of counterfactuals.
 
@@
 For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method.
-Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating a sufficiently large number of counterfactuals during each training epoch and including only mature counterfactuals for contrastive divergence.
+Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating a sufficiently large number of counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.
 
-***CT is sensitive to hyperparameter selection.*** As discussed in @sec-hyperparameters, our method benefits from tuning certain key hyperparameters.
+***CT is sensitive to hyperparameter selection.*** Our method benefits from tuning certain key hyperparameters (see @sec-hyperparameters).
 In this work, we have relied exclusively on grid search for this task.
@@
 However, there are three factors that attenuate the impact of this limitation.
-First, CT provides counterfactual explanations for the training samples essentially for free, which may be beneficial in many ADM systems.
+First, CT provides CEs for the training samples for free, which may be beneficial in many ADM systems.
 Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models.
@@
 
-***Immutable features may have proxies.*** In @prp-mtblty we define an approach to protect immutable features and thus increase the actionability of the generated counterfactuals.
-However, this approach requires that model owners define the mutability constraints for (all) features considered by the model.
-Even with sufficient domain knowledge to protect all immutable features---ones that cannot be changed at all and ones that cannot be reasonably expected to change---there may exist proxies that are theoretically mutable (and hence should not be protected) but preserve enough information about the principals to counteract the protections.
-As one example, consider the Adult dataset used in our experiments where the mutable education status is a proxy for the immutable age, in that the attainment of degrees is correlated with age.
+***Immutable features may have proxies.*** We propose an approach to protect immutable features and thus increase the actionability of the generated CEs.
+However, it requires that model owners define the mutability constraints for (all) features considered by the model.
+Even with sufficient domain knowledge to protect all immutable features, there may exist proxies that are theoretically mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections.
+As an example, consider the Adult dataset used in our experiments where the mutable education status is a proxy for the immutable age, in that the attainment of degrees is correlated with age.
 Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to increase the explainability of the model.
 
-***Interventions on features may have downstream impacts on fairness.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions.
+***Interventions on features may impact fairness downstream.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions.
 As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.

### 2025-03-13 paper/quarto_ecml/paper.tex
Add draft conclusions

@@
 
+State-of-the-art machine learning models are prone to learning complex
+representations that cannot be interpreted by humans. Although post-hoc
+explainability approaches have attracted major research interest, these
+cannot guarantee that the explanations agree with the opaque model's
+learned representation of data. As a step towards addressing this
+challenge, we introduced counterfactual training, a novel training
+regime that incentivizes highly-explainable models. Our approach leads
+to explanations that are both plausible---compliant with the underlying
+data-generating process---and actionable---compliant with user-specified
+mutability constraints---and thus meaningful to their recipients.
+Through extensive experiments we demonstrate that counterfactual
+training satisfies its objectives while preserving the predictive
+performance of the trained models. We also find that our approach can be
+used to fine-tune conventionally-trained models and achieve similar
+gains in explainability. Finally, this work showcases that it is
+practical to improve models \emph{and} their explanations at the same
+time.
+
 \newpage{}

### 2025-03-13 paper/sections/other/main_results.qmd

```{julia}
projectdir = splitpath(pwd()) |>
    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
cd(projectdir)

using CTExperiments
using CTExperiments.CSV
using CTExperiments.DataFrames
using CTExperiments.StatsBase

using DotEnv
DotEnv.load!()
```

```{julia}
res_dir = joinpath(ENV["FINAL_GRID_RESULTS"],"single")
```

# Details on Main Experiments {.appendix}

## Final Hyperparameters

As discussed @sec-experiments, CT is sensitive to certain hyperparameter choices. We study the effect of many hyperparameters extensively in @sec-app-grid. For the main results, we tune a small set of key hyperparameters (@sec-app-tune). The final choices for the main results are presented for each data set in @tbl-final-params along with training, test and batch sizes.

::: {#tbl-final-params}

::: {.content-hidden unless-format="pdf"}

```{julia}
#| output: asis

df = final_params(res_dir)
get_table_inputs(df, nothing; backend=Val(:latex)) |>
    inputs -> tabulate_results(inputs; table_type=:tabular)
```

:::

Final hyperparameters used for the main results for the different datasets.

:::

```{julia}
#| eval: false

df = final_table(res_dir)
get_table_inputs(df, nothing; backend=Val(:latex)) |>
    inputs -> tabulate_results(inputs; table_type=:tabular, save_name="paper/tables/main.tex")
```

## Qualitative Findings for Image Data

::: {.callout-note}

@fig-mnist shows much more plausible (faithful) counterfactuals for a model with CT than the model with conventional training (@fig-mnist-vanilla). In fact, this is not even using *ECCo+* and still showing better results than the best results we achieved in our AAAI paper for JEM ensembles.

:::

![Counterfactual images for *MLP* with counterfactual training. The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}

![Counterfactual images for *MLP* with conventional training. The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp_vanilla.png){#fig-mnist-vanilla}

### 2025-03-13 paper/quarto_ecml/paper.tex
main results finally done

--- 2025-03-13 paper/quarto_ecml/paper.tex
@@
 \newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-
+% \usepackage{amsthm}
+% \theoremstyle{plain}
+% \newtheorem{corollary}{Research Question}[section]
+% \theoremstyle{definition}
+% \newtheorem{example}{Example}[section]
+% \theoremstyle{definition}
+% \newtheorem{definition}{Definition}[section]
+% \theoremstyle{plain}
+% \newtheorem{proposition}{Proposition}[section]
+% \theoremstyle{remark}
+% \AtBeginDocument{\renewcommand*{\proofname}{Proof}}
+% \newtheorem*{remark}{Remark}
+% \newtheorem*{solution}{Solution}
+% \newtheorem{refremark}{Remark}[section]
+% \newtheorem{refsolution}{Solution}[section]
 \makeatother
@@
 \begin{abstract}
-We propose a novel training regime termed Counterfactual Training that
+We propose a novel training regime called Counterfactual Training that
 leverages counterfactual explanations to increase the explanatory
@@
 decision-making systems, counterfactuals should be plausible with
-respect to the underlying data and actionable with respect to the
+respect to the underlying data and actionable with respect to
 stakeholder requirements. Much existing research has therefore focused
@@
 these desiderata. In this work, we instead hold models directly
-accountable for the desired end goal: Counterfactual Training employs
+accountable for this desired end goal: Counterfactual Training employs
 counterfactuals ad-hoc during the training phase to minimize the
@@
 Today's prominence of artificial intelligence (AI) has largely been
-driven by \textbf{representation learning}: instead of relying on
-features and rules that are carefully hand-crafted by humans, modern
-machine learning (ML) models are tasked with learning representations
-directly from data, guided by narrow objectives such as predictive
-accuracy \cite{goodfellow2016deep}. Modern advances in computing have
-made it possible to provide such models with ever-growing degrees of
-freedom to achieve that task, which frequently allows them to outperform
-traditionally more parsimonious models. Unfortunately, in doing so,
-models learn increasingly complex and highly sensitive representations
-that humans can no longer easily interpret.
+driven by advances in \textbf{representation learning}: instead of
+relying on features and rules that are carefully hand-crafted by humans,
+modern machine learning (ML) models are tasked with learning
+representations directly from data, guided by narrow objectives such as
+predictive accuracy \cite{goodfellow2016deep}. Modern advances in
+computing have made it possible to provide such models with ever-growing
+degrees of freedom to achieve that task, which has often led them to
+outperform traditionally more parsimonious models. Unfortunately, in
+doing so, models learn increasingly complex and highly sensitive
+representations that we can no longer easily interpret.
 
@@
 revolution, \cite{szegedy2013intriguing} showed that artificial neural
-networks (ANN) are sensitive to adversarial examples: perturbed versions
-of data instances that yield vastly different model predictions despite
-being ``imperceptible'' in that they are semantically indifferent from
-their factual counterparts. Even though some partially effective
-mitigation strategies have been proposed---most notably
-\textbf{adversarial training} \cite{goodfellow2014explaining}---truly
-robust deep learning (DL) remains unattainable even for models that are
-considered shallow by today's standards \cite{kolter2023keynote}.
+networks (ANN) are sensitive to adversarial examples: counterfactuals of
+model inputs that yield vastly different model predictions despite being
+``imperceptible'' in that they are semantically indifferent from their
+factual counterparts. Although some partially effective mitigation
+strategies have been proposed, for example \textbf{adversarial training}
+\cite{goodfellow2014explaining}, truly robust deep learning (DL) remains
+unattainable even for models that are considered shallow by today's
+standards \cite{kolter2023keynote}.
 
@@
 neural networks are so expressive that they can fit randomly labeled
-data. Thus, looking at the predictive performance alone, the solutions
-may seem to provide compelling explanations for the data, when in fact
-they are based on purely associative, semantically meaningless patterns.
-This poses two related challenges. Firstly, there is no dependable way
-to verify if such complex representations correspond to meaningful and
+data. Thus, looking at the predictive performance, the solutions may
+seem to provide compelling explanations for the data, when in fact they
+are based on purely associative, semantically meaningless patterns. This
+poses two related challenges. Firstly, there is no dependable way to
+verify if such complex representations correspond to meaningful and
 plausible explanations. Secondly, even if we could resolve the first
@@
 The first challenge has attracted an abundance of research on
-\textbf{explainable AI} (XAI), a paradigm that focuses on the
-development of tools to derive (post-hoc) explanations from complex
-model representations. Such explanations should mitigate a scenario in
-which practitioners deploy opaque models and blindly rely on their
-predictions. On countless occasions, this has happened in practice and
-caused real harms to people who were adversely and unfairly affected by
-automated decision-making (ADM) systems involving opaque models
-\cite{oneil2016weapons,mcgregor2021preventing}. Effective XAI tools can
-aid us in monitoring models and providing recourse to individuals to
-turn negative outcomes (e.g., ``loan application rejected'') into
-positive ones (e.g., ``application accepted''). Our work builds upon
-\textbf{counterfactual explanations} (CE) proposed by
-\cite{wachter2017counterfactual} as an effective approach to achieve
-this goal. CEs prescribe minimal changes for factual inputs that, if
-implemented, would prompt some fitted model to produce a desired output.
-
-To our surprise, the second challenge has not yet attracted major
-research interest. Specifically, there has been no concerted effort
-towards improving the ``explanatory capacity'' of models, i.e., the
-degree to which learned representations correspond to explanations that
-are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
-Definition~\ref{def-explainability}). Instead, the choice has generally
-been to improve the ability of XAI tools to identify the subset of
+\textbf{explainable AI} (XAI) which aims to develop tools to derive
+explanations from complex model representations. This can mitigate a
+scenario in which we deploy opaque models and blindly rely on their
+predictions. On countless occasions, this scenario has occurred in
+practice and caused real harm to people who were affected adversely and
+often unfairly by automated decision-making (ADM) systems involving
+opaque models \cite{oneil2016weapons,mcgregor2021preventing}. Effective
+XAI tools can aid us in monitoring models and providing recourse to
+individuals to turn adverse outcomes (e.g., ``loan application
+rejected'') into positive ones (e.g., ``application accepted'').
+\cite{wachter2017counterfactual} propose \textbf{counterfactual
+explanations} (CE) as an effective approach to achieve this goal: CEs
+explain how factual inputs need to change in order for some fitted model
+to produce some desired output, typically involving minimal
+perturbations.
+
+To our surprise, the second challenge has not yet attracted any major
+consolidated research effort. Specifically, there has been no concerted
+effort towards improving improving models' explanatory capacity, which
+we will henceforth simply call ``explainability'', defined as the degree
+to which learned representations correspond to explanations that are
+interpretable and deemed \textbf{plausible} by humans (see
+Definition~\ref{def-explainability}). Instead, the choice has typically
+been to improve the ability of XAI tools to identify the subset
 explanations that are both plausible and valid for any given model,
 independent of whether the learned representations are also compatible
-with plausible explanations \cite{altmeyer2024faithful}. Fortunately,
-recent findings indicate that improved explanatory capacity can arise as
-a consequence of regularization techniques aimed at other training
-objectives such as robustness, generalization, and generative capacity
+with implausible explanations \cite{altmeyer2024faithful}. Fortunately,
+recent findings indicate that explainability can arise as byproduct of
+regularization techniques aimed at other objectives such as robustness,
+generalization, and generative capacity
 \cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
-As further discussed in Section~\ref{sec-lit}, our work consolidates
-these findings within a single objective.
-
-\textbf{Specifically, we introduce counterfactual training}: a novel
-training regime explicitly meant to align learned representations with
-plausible explanations that comply with user requirements. Our
-contributions are as follows:
+
+Building on these findings, we introduce \textbf{counterfactual
+training}: a novel training regime explicitly geared towards aligning
+model representations with plausible explanations. Our contributions are
+as follows:
 
@@
 \item
-  We present a novel methodological framework that leverages adversarial
-  examples and faithful counterfactual explanations during the training
-  phase to improve the explanatory capacity and robustness of machine
-  learning models (Section~\ref{sec-method}).\\
+  We discuss existing related work on improving models and consolidate
+  it through the lens of counterfactual explanations
+  (Section~\ref{sec-lit}).
 \item
-  We propose a method to enforce global actionability constraints by
-  preventing models from assigning importance to immutable features,
-  i.e., ones over which decision subjects have no control
-  (Section~\ref{sec-method}).\\
+  We present our proposed methodological framework that leverages
+  faithful counterfactual explanations during the training phase of
+  models to achieve the explainability objective
+  (Section~\ref{sec-method}).
 \item
-  Through extensive experiments we demonstrate that counterfactual
-  training promotes explainability while preserving high predictive
-  performance. We run ablation studies and grid searches to understand
-  how the underlying model components and hyperparameters affect
-  outcomes. (Section~\ref{sec-experiments}).
+  Through extensive experiments we demonstrate the counterfactual
+  training improve model explainability while maintaining high
+  predictive performance. We run ablation studies and grid searches to
+  understand how the underlying model components and hyperparameters
+  affect outcomes. (Section~\ref{sec-experiments}).
 \end{itemize}
 
-Despite some limitations discussed in Section~\ref{sec-discussion}, we
-conclude in Section~\ref{sec-conclusion} that counterfactual training
-provides a useful framework for researchers and practitioners interested
-in making opaque models more trustworthy. We also believe that this work
+Despite some limitations of our approach discussed in
+Section~\ref{sec-discussion}, we conclude in
+Section~\ref{sec-conclusion} that counterfactual training provides a
+practical framework for researchers and practitioners interested in
+making opaque models more trustworthy. We also believe that this work
 serves as an opportunity for XAI researchers to re-evaluate the trend of
@@
 
-To the best of our knowledge, the proposed framework for counterfactual
+To the best of our knowledge, our proposed framework of counterfactual
 training represents the first attempt to use counterfactual explanations
@@
 derived for an opaque model are also deemed plausible with respect to
-the underlying data and stakeholder requirements; the former means that
-the counterfactuals should comply with the distribution of the factual
-data, the latter means that they should respect arbitrary (global)
-actionability constraints. To make the desiderata for our framework more
+the underlying data and stakeholder requirements. To make this more
 concrete, we follow \cite{augustin2020adversarial} in tying the concept
 of explainability to the quality of counterfactual explanations that we
-can generate for a given model. The authors show that CEs---understood
-here as minimal input perturbations that yield some desired model
-prediction---are generally more meaningful if the underlying model is
-more robust to adversarial examples. We can make intuitive sense of this
-finding when looking at adversarial training (AT) through the lens of
-representation learning with high degrees of freedom. As argued before,
-learned representations may be sensitive to producing implausible
-explanations and mispredicting for worst-case counterfactuals (i.e.,
-adversarial examples). Thus, by inducing models to ``unlearn''
-susceptiblity to such examples, AT can effectively remove implausible
-explanations from the solution space.
+can generate for a given model. The authors show that counterfactual
+explanations---understood here as minimal input perturbations that yield
+some desired model prediction---are generally more meaningful if the
+underlying model is more robust to adversarial examples. We can make
+intuitive sense of this finding when looking at adversarial training
+(AT) through the lens of representation learning with high degrees of
+freedom: by inducing models to ``unlearn'' representations that are
+susceptible to worst-case counterfactuals (i.e., adversarial examples),
+AT effectively removes some implausible explanations from the solution
+space.
 
@@
 This interpretation of the link between explainability through
-counterfactuals on one side and robustness to adversarial examples on
-the other is backed by empirical evidence.
+counterfactuals on one side, and robustness to adversarial examples on
+the other, is backed by empirical evidence.
 \cite{sauer2021counterfactual} demonstrate that using counterfactual
@@
 supervised settings where inputs may be reasonably mapped to multiple
-outputs. They, too, demonstrate that augmenting the training data of
-image classifiers can improve generalization. Finally,
-\cite{teney2020learning} propose an approach using counterfactuals in
-training that does not rely on data augmentation: they argue that
-counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on identifying similar input samples
-with different annotations and ensuring that the gradient of the
-classifier aligns with the vector between such pairs of counterfactual
-inputs using the cosine distance as the loss function.
+outputs. They, too, demonstrate the augmenting the training data of
+image classifiers can improve generalization. \cite{teney2020learning}
+propose an approach using counterfactuals in training that does not rely
+on data augmentation: they argue that counterfactual pairs typically
+already exist in training datasets. Specifically, their approach relies
+on identifying similar input samples with different annotations and
+ensuring that the gradient of the classifier aligns with the vector
+between such pairs of counterfactual inputs using the cosine distance as
+the loss function.
 
 In the natural language processing (NLP) domain, counterfactuals have
-similarly been used to improve models through data augmentation.
-\cite{wu2021polyjuice} propose \emph{Polyjuice}, a general-purpose
+similarly been used to improve models through data augmentation:
+\cite{wu2021polyjuice}, propose \emph{Polyjuice}, a general-purpose
 counterfactual generator for language models. They demonstrate
-empirically that the augmentation of training data through
-\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
-tasks. \cite{balashankar2023improving} similarly use \emph{Polyjuice} to
-augment NLP datasets through diverse counterfactuals and show that
-classifier robustness improves by up to 20\%. Finally,
-\cite{luu2023counterfactual} introduce Counterfactual Adversarial
-Training (CAT), which also aims at improving generalization and
-robustness of language models through a three-step procedure. First, the
-authors identify training samples that are subject to high predictive
-uncertainty. Second, they generate counterfactual explanations for those
-samples. Finally, they fine-tune the given language model on the
-augmented dataset that includes the generated counterfactuals.
+empirically that augmenting training data through \emph{Polyjuice}
+counterfactuals improves robustness in a number of NLP tasks.
+\cite{balashankar2023improving} also use \emph{Polyjuice} to augment NLP
+datasets through diverse counterfactuals and show that classifier
+robustness improves up to 20\%. Finally, \cite{luu2023counterfactual}
+introduce Counterfactual Adversarial Training (CAT), which also aims at
+improving generalization and robustness of language models.
+Specifically, they propose to proceed as follows: firstly, they identify
+training samples that are subject to high predictive uncertainty;
+secondly, they generate counterfactual explanations for those samples;
+and, finally, they fine-tune the given language model on the augmented
+dataset that includes the generated counterfactuals.
 
@@
 between counterfactual explanations and adversarial examples (AE).
-Pointing to clear similarities in how CEs and AEs are generated,
+Pointing to clear similarities in how CE and AE are generated,
 \cite{freiesleben2022intriguing} makes the case for jointly studying the
-opaqueness and robustness problems in representation learning. Formally,
-AEs can be seen as the subset of CEs for which misclassification is
+opaqueness and robustness problem in representation learning. Formally,
+AE can be seen as the subset of CE for which misclassification is
 achieved \cite{freiesleben2022intriguing}. Similarly,
-\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
-certain conditions and derive theoretical upper bounds on distances
+\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
+certain conditions and derive theoretical upper bounds on the distances
 between them.
@@
 counterfactuals during training with the explicit goal of affecting
-certain properties of the post-hoc counterfactual explanations. Firstly,
-\cite{ross2021learning} propose a way to train models that guarantee
-individual recourse to some positive target class with high probability.
-Their approach builds on adversarial training by explicitly inducing
-susceptibility to targeted adversarial examples for the positive class.
-Additionally, the proposed method allows for imposing a set of
-actionability constraints ex-ante. For example, users can specify that
-certain features (e.g., \emph{age}, \emph{gender}) are immutable.
-Secondly, \cite{guo2023counternet} are the first to propose an
-end-to-end training pipeline that includes counterfactual explanations
-as part of the training procedure. In particular, they propose a
-specific network architecture that includes a predictor and CE generator
-network, where the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed
-back to the predictor network. In contrast to \cite{guo2023counternet},
-we impose no restrictions on the neural network architecture at all.
+certain properties of post-hoc counterfactual explanations. Firstly,
+\cite{ross2021learning} propose a way to train models that are
+guaranteed to provide recourse for individuals to move from an adverse
+outcome to some positive target class with high probability. Their
+approach builds on adversarial training, where in this context
+susceptibility to targeted adversarial examples for the positive class
+is explicitly induced. The proposed method allows for imposing a set of
+actionability constraints ex-ante: for example, users can specify that
+certain features (e.g., \emph{age}, \emph{gender}, \ldots) are
+immutable. Secondly, \cite{guo2023counternet} are the first to propose
+an end-to-end training pipeline that includes counterfactual
+explanations as part of the training procedure. In particular, they
+propose a specific network architecture that includes a predictor and CE
+generator network, where the parameters of the CE generator network are
+learnable. Counterfactuals are generated during each training iteration
+and fed back to the predictor network. In contrast to
+\cite{guo2023counternet}, we impose no restrictions on the neural
+network architecture at all.
 
@@
 objectives. The authors propose a way to generate counterfactuals that
-are maximally faithful to the model in that they are consistent with
-what the model has learned about the underlying data. Formally, they
-rely on tools from energy-based modelling to minimize the divergence
-between the distribution of counterfactuals and the conditional
-posterior over inputs learned by the model. Their proposed
+are maximally \textbf{faithful} to the model in that they are consistent
+with what the model has learned about the underlying data. Formally,
+they rely on tools from energy-based modelling to minimize the
+divergence between the distribution of counterfactuals and the
+conditional posterior over inputs learned by the model. Their proposed
 counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
 and only if the underlying model has learned representations that align
-with them. The authors find that both deep ensembles
+with them. They find that both deep ensembles
 \cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
@@
 
-Counterfactual training (CT) combines ideas from adversarial training,
+Counterfactual training combines ideas from adversarial training,
 energy-based modelling and counterfactuals explanations with the
-explicit goal of aligning representations with plausible explanations
-that comply with user requirements. In the context of CEs, plausibility
-has broadly been defined as the degree to which counterfactuals comply
-with the underlying data-generating process
+explicit objective of aligning representations with plausible
+explanations that comply with user requirements. In the context of CEs,
+plausibility has broadly been defined as the degree to which
+counterfactuals comply with the underlying data generating process
 \cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
 Plausibility is a necessary but insufficient condition for using CEs to
-provide algorithmic recourse (AR) to individuals (negatively) affected
-by opaque models. For AR recommendations to be actionable, they need to
-not only result in plausible counterfactuals but also be attainable. A
-plausible CE for a rejected 20-year-old loan applicant, for example,
-might reveal that their application would have been accepted, if only
-they were 20 years older. Ignoring all other features, this would comply
-with the definition of plausibility if 40-year-old individuals were in
-fact more credit-worthy on average than young adults. But of course this
-CE does not qualify for providing actionable recourse to the applicant
-since \emph{age} is not a (directly) mutable feature. CT aims to improve
-model explainability by aligning models with counterfactuals that meet
-both desiderata: plausibility and actionability. Formally, we define
-explainability as follows:
+provide algorithmic recourse (AR) to individuals affected by opaque
+models in practice. This is because for recourse recommendations to be
+\textbf{actionable}, they need to not only result in plausible
+counterfactuals but also be attainable. A plausible CE for a rejected
+20-year-old loan applicant, for example, might reveal that their
+application would have been accepted, if only they were 20 years older.
+Ignoring all other features, this would comply with the definition of
+plausibility if 40-year-old individuals were in fact more credit-worthy
+on average than young adults. But of course this CE does not qualify for
+providing actionable recourse to the applicant since \emph{age} is not a
+(directly) mutable feature. For our intents and purposes, counterfactual
+training aims to improve model explainability by aligning models with
+counterfactuals that meet both desiderata, plausibility and
+actionability. Formally, we define explainability as follows:
 
@@
 We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
-that faithfully generated counterfactuals are plausible (i.e.,
-consistent with the data) and actionable. Formally, we define these
-properties as follows:
+that faithfully generated counterfactuals are plausible (i.e.~consistent
+with the data) and actionable. Formally, we define these properties as
+follows:
 
@@
 
-The characterization of faithfulness and plausibility in
-Definition~\ref{def-explainability} is the same as in
+The definitions of faithfulness and plausibility in
+Definition~\ref{def-explainability} are the same as in
 \cite{altmeyer2024faithful}, with adapted notation. Actionability
@@
 initially proposed by \cite{wachter2017counterfactual}. For our
-purposes, we let \(T\) vary and consider the counterfactual search to be
+purposes, we let \(T\) vary and consider the counterfactual search as
 converged as soon as the predicted probability for the target class has
-reached a pre-determined threshold \(\tau\):
+reached a pre-determined threshold, \(\tau\):
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
@@
 \begin{equation}\phantomsection\label{eq-obj}{
-\begin{split}
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
-\end{split}
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
 }\end{equation}
 
-where \(\text{yloss}(\cdot)\) is a classification loss that induces
-discriminative performance (e.g., cross-entropy). The second and third
-terms in Equation~\ref{eq-obj} are explained in detail below. For now,
-they can be sufficiently described as inducing explainability directly
-and indirectly by penalizing: (1) the contrastive divergence,
-\(\text{div}(\cdot)\), between mature counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
-adversarial loss, \(\text{advloss}(.)\), with respect to nascent
+where \(\text{yloss}(\cdot)\) is any conventional classification loss
+that induces discriminative performance (e.g., cross-entropy). The
+second and third terms in Equation~\ref{eq-obj} are explained in more
+detail below. For now, they can be sufficiently described as inducing
+explainability directly and indirectly by penalizing: (1) the
+contrastive divergence, \(\text{div}(\cdot)\), between mature
+counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
+(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
 counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
 \(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
-regularizes the magnitude of the energy terms involved in
-\(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off between the
-components can be governed by adjusting the strengths of the penalties
-\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
+regularises the magnitude of the energy terms involved in
+\(\text{div}(\cdot)\) \cite{du2019implicit}. The tradeoff between the
+different components can be governed by adjusting the strengths of the
+penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
 \(\lambda_{\text{reg}}\).
 
-\subsection{Directly Inducing Explainability with Contrastive
-Divergence}\label{directly-inducing-explainability-with-contrastive-divergence}
+\subsubsection{Directly Inducing Explainability through Contrastive
+Divergence}\label{directly-inducing-explainability-through-contrastive-divergence}
 
@@
 samples from \(p(\mathbf{x})\) and the generated samples from
-\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
-perform well at both tasks by directly maximizing the joint
-log-likelihood factorized as
+\(p_\theta(\mathbf{x})\). They show that JEMs can be trained to perform
+well at both tasks by directly maximizing the joint log-likelihood
+factorized as
 \(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
-The first term can be optimized using conventional cross-entropy as in
+The first factor can be optimized using conventional cross-entropy as in
 Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
@@
 tend to do well with respect to the plausibility objective in
-Definition~\ref{def-explainability}. This follows directly if we
-consider samples drawn from \(p_\theta(\mathbf{x})\) as counterfactuals
-because the JEM objective effectively minimizes the divergence between
-the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To
-generate samples, \cite{grathwohl2020your} rely on Stochastic Gradient
-Langevin Dynamics (SGLD) using an uninformative prior for initialization
-but we depart from their methodology. Instead of SGLD, we propose to use
-counterfactual explainers to generate counterfactuals of observed
-training samples. Specifically, we have:
+Definition~\ref{def-explainability}. If we consider samples drawn from
+\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
+finding, because the JEM objective effectively minimizes the divergence
+between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
+To generate samples, \cite{grathwohl2020your} rely on Stochastic
+Gradient Langevin Dynamics (SGLD) using an uninformative prior for
+initialization. This is where we depart from their methodology: instead
+of SGLD, we propose to use counterfactual explainers to generate
+counterfactuals of observed training samples. Specifically, we have:
 
@@
 
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
+where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
+particular, we set
 \(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
 where \(y^+\) denotes the index of the randomly drawn target class,
-\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed sample from
-target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\). Conditional on the
-target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes a mature
-counterfactual for a randomly sampled factual from a non-target class
-generated with a gradient-based CE generator for up to \(T\) iterations.
-Mature counterfactuals are ones that have either reached convergence
-wrt. the decision threshold \(\tau\) or exhausted \(T\).
+\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed data point
+sampled from target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\).
+Conditional on the target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes
+a mature counterfactual for a randomly sampled factual from a non-target
+class generated through a gradient-based counterfactual generator for at
+most \(T\) iterations. We define mature counterfactuals as those that
+have either exhausted \(T\) or reached convergence in terms of the
+pre-determined decision threshold earlier.
 
 Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
-of observed training samples (positive samples) while increasing the
-energy of counterfactuals (negative samples) \cite{du2019implicit}. As
-the counterfactuals get more plausible
-(Definition~\ref{def-explainability}) during training, these opposing
-effects gradually balance each other out \cite{lippe2024uvadlc}.
+of observed training samples (positive samples) while at same time
+increasing the energy of counterfactuals (negative samples)
+\cite{du2019implicit}. As the generated counterfactuals get more
+plausible (Definition~\ref{def-explainability}) over the course of
+training, these two opposing effects gradually balance each out
+\cite{lippe2024uvadlc}.
 
@@
 explainers that have been proposed in the literature to meet different
-desiderata. For example, many methods facilitate the imposition of
-domain and mutability constraints. In principle, any existing approach
-for generating counterfactual explanations is viable, so long as it does
-not violate the faithfulness condition. Like JEMs
-\cite{murphy2022probabilistic}, CT can be considered a form of
-contrastive representation learning.
-
-\subsection{Indirectly Inducing Explainability with Adversarial
-Robustness}\label{indirectly-inducing-explainability-with-adversarial-robustness}
+desiderata. Typically, these methods facilitate the imposition of domain
+and mutability constraints, for example. In principle, any existing
+approach for generating counterfactual explanations is viable, so long
+as it does not violate the faithfulness condition. Like JEMs
+\cite{murphy2022probabilistic}, counterfactual training can be
+considered as a form of contrastive representation learning.
+
+\subsubsection{Indirectly Inducing Explainability through Adversarial
+Robustness}\label{indirectly-inducing-explainability-through-adversarial-robustness}
 
@@
 \(\mathbf{x}^\prime\) can be repurposed as additional training samples
-\cite{luu2023counterfactual,balashankar2023improving} or AEs
-\cite{freiesleben2022intriguing,pawelczyk2022exploring}. This leaves
-some flexibility with respect to the choice for
+\cite{luu2023counterfactual,balashankar2023improving} or adversarial
+examples \cite{freiesleben2022intriguing,pawelczyk2022exploring}. This
+leaves some flexibility with respect to the exact choice for
 \(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
-functional form, but likely not the only sensible choice, is inspired by
-adversarial training:
+functional form to use, though likely not the only reasonable choice, is
+inspired by adversarial training:
 
@@
 Under this choice, we consider nascent counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
-perturbation to any single feature is at most \(\varepsilon\). This is
-closely aligned with \cite{szegedy2013intriguing} who define an
-adversarial attack as an ``imperceptible non-random perturbation''.
-Thus, we choose to work with a different distinction between CE and AE
-than \cite{freiesleben2022intriguing} who consider misclassification as
-the key distinguishing feature of AE. One of the key observations in
-this work is that we can leverage CEs during training and get
-adversarial examples essentially for free.
+\(\mathbf{x}_{t\leq T}^\prime\) as adversarial examples as long as the
+magnitude of the perturbation to any individual feature is at most
+\(\varepsilon\). This is closely aligned with
+\cite{szegedy2013intriguing}, who define an adversarial attack as an
+``imperceptible non-random perturbation''. Thus, we choose to work with
+a different distinction between CE and AE than
+\cite{freiesleben2022intriguing}, who considers misclassification as the
+key distinguishing feature of AE. One of the key observations in this
+work is that we can leverage counterfactual explanations during training
+and get adversarial examples, essentially for free.
 
@@
 enforced by setting partial derivatives to zero to ensure that features
-are only perturbed in the allowed direction, if at all.
-
-Since such actionability constraints are binding at test time, we should
-also impose them when generating \(\mathbf{x}^\prime\) during each
-training iteration to inform model representations. Through their effect
-on \(\mathbf{x}^\prime\), both types of constraints influence model
-outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
-penalizing implausibility that arises due to mutability constraints. For
-any mutability-constrained feature \(d\) this can be achieved by
-enforcing \(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever
-perturbing \(\mathbf{x}^\prime[d]\) in the direction of
-\(\mathbf{x}[d]\) would violate mutability constraints. Specifically, we
-set \(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
+are only mutated in the allowed direction, if at all.
+
+Since actionability constraints are binding at test time, we should also
+impose them when generating \(\mathbf{x}^\prime\) during each training
+iteration to align model representations with user requirements. Through
+their effect on \(\mathbf{x}^\prime\), both types of constraints
+influence model outcomes through Equation~\ref{eq-div}. Here it is
+crucial that we avoid penalizing implausibility that arises due to
+mutability constraints. For any mutability-constrained feature \(d\)
+this can be achieved by enforcing
+\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
+\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
+violate mutability constraints. Specifically, we set
+\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
 
@@
 \item
-  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\), but feature \(d\) can
+  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\) but feature \(d\) can
   only be decreased in practice.
 \item
-  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\), but feature \(d\) can
+  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\) but feature \(d\) can
   only be increased in practice.
@@
 independence of features. This raises a valid concern about the effect
-of protecting immutable features in the presence of proxies that remain
-unprotected. We address this in Section~\ref{sec-discussion}.
-
-\subsection{Example: influence of counterfactual training on
-explainability}\label{example-influence-of-counterfactual-training-on-explainability}
+of protecting immutable features in the presence of proxy features that
+remain unprotected. We discuss this limitation in
+Section~\ref{sec-discussion}.
+
+\subsection{Illustration}\label{illustration}
 
@@
 previous loan takers comprised of a binary outcome variable
-(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
-features: (1) the subjects' \emph{age}, which we define as immutable,
-and (2) the subjects' existing level of \emph{debt}, which we define as
-mutable.
-
-We have simulated this scenario using synthetic data with two
-independent features and Gaussian class-conditional densities in
+(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
+(1) the subjects' \emph{age}, which we define as immutable, and (2) the
+subjects' existing level of \emph{debt}, which we define as mutable.
+
+We have simulated this scenario using synthetic data with independent
+features and Gaussian class-conditional densities in
 Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
@@
 architecture each time (a linear classifier). In each case, we show the
-decision boundary (in green) and the training data colored according to
-their ground-truth label: orange points belong to the target class,
+linear decision boundary (green) and the training data colored according
+to their ground-truth label: orange points belong to the target class,
 \(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
@@
 impose mutability constraints at test time. The generated
-counterfactuals are all valid, but not plausible: they do not comply
-with the distribution of the factual samples in the target class to the
-point where they are clearly distinguishable from the ground-truth data.
-In panel (b), we have trained our model with counterfactual training,
-once again without any mutability constraints. We observe that the
-counterfactuals are highly plausible, meeting the first objective of
+counterfactuals are all valid, but not plausible: they are clearly
+distinguishable from the ground-truth data. In panel (b), we have
+trained our model with counterfactual training, once again not imposing
+mutability constraints at test time. We observe that the counterfactuals
+are clearly plausible, therefore meeting the first objective of
 Definition~\ref{def-explainability}.
@@
 counterfactual paths are shorter on average in panel (d), where we have
-used counterfactual training and protected the immutable feature as
-described in Section~\ref{sec-constraints}. We observe that due to the
-classifier's lower sensitivity to \emph{age}, recourse recommendations
-with respect to \emph{debt} are much more homogenous and do not
-disproportionately punish younger individuals. The counterfactuals are
-also plausible with respect to the mutable feature. Thus, we consider
-the model in panel (d) as the most explainable according to
-Definition~\ref{def-explainability}.
+used counterfactual training and protected immutable features as
+described in Section~\ref{sec-constraints}. In particular, we observe
+that due to the classifier's lower sensitivity to \emph{age}, recourse
+recommendations with respect to \emph{debt} are much more homogenous, in
+that they do not disproportionately punish younger individuals. The
+counterfactuals are also plausible with respect to the mutable feature.
+Thus, we consider the model in panel (d) as the most explainable
+according to Definition~\ref{def-explainability}.
 
@@
 
-\caption{\label{fig-poc}Illustration for Example~\ref{exm-grad} of how
-CT improves model explainability.}
+\caption{\label{fig-poc}Visual illustration of how counterfactual
+training improves explainability. See Example~\ref{exm-grad} for
+details.}
 
@@
 
-To what extent does our proposed counterfactual training objective
+Does our proposed counterfactual training objective
 (Equation~\ref{eq-obj}) induce models to learn plausible explanations?
@@
 
-To what extent does our proposed counterfactual training objective
+Does our proposed counterfactual training objective
 (Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
@@
 
-What are the effects of hyperparameter selection with respect to
+What are the effects of different hyperparameter choices with respect to
 Equation~\ref{eq-obj}?
@@
 
-Our key outcome of interest is how well do models perform with respect
-to explainability (Definition~\ref{def-explainability}). To this end, we
+Our key outcome of interest is how well models perform with respect to
+explainability (Definition~\ref{def-explainability}): to this end, we
 focus primarily on the plausibility and cost of faithfully generated
@@
 \begin{equation}\phantomsection\label{eq-impl-dist}{
-\text{implaus}_{\text{dist}}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
 }\end{equation}
@@
 \begin{equation}\phantomsection\label{eq-impl-div}{
-\text{implaus}_{\text{div}}(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
+\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
 }\end{equation}
@@
 population maximum mean discrepancy \cite{gretton2012kernel}. The metric
-in Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
-are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
+in Equation~\ref{eq-impl-div} is equal to zero iff
+\(\mathbf{X}^\prime=\mathbf{X}^+\).
 
@@
 
-We run the experiments with three gradient-based generators:
-\emph{Generic} of \cite{wachter2017counterfactual} as a simple baseline
-approach, \emph{REVISE} \cite{joshi2019realistic} that aims to generate
-plausible counterfactuals using a surrogate Variational Autoencoder
-(VAE), and \emph{ECCo}---the generator of \cite{altmeyer2023faithful}
-but without the conformal prediction component---as a method that
-directly targets both faithfulness and plausibility of the CEs.
+We run the experiments with three CE generators: \emph{Generic} of
+\cite{wachter2017counterfactual} as a simple baseline approach,
+\emph{REVISE} \cite{joshi2019realistic} that aims to generate plausible
+counterfactuals using a surrogate Variational Autoencoder (VAE), and
+\emph{ECCo}---the generator of \cite{altmeyer2023faithful} but without
+the conformal prediction component---as a method that directly targets
+both faithfulness and plausibility of the CEs.
 
@@
 \subsubsection{Plausibility}\label{plausibility-1}
+
+Table~\ref{tbl-main} presents our main empirical findings. The top five
+rows show the percentage reduction in implausibility according to
+Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
+used for \emph{ECCo} at test time. The following row shows the reduction
+in implausibility as measured by Equation~\ref{eq-impl-div} and
+aggregated across all test specifications of \emph{ECCo}. The final two
+rows show the test accuracies for the model trained with CT and
+conventionally trained models (``vanilla'').
+
+We observe that for all datasets except \emph{OL} and across all test
+settings, the average distance of counterfactuals from observed samples
+in the target class is reduced, indicating improved plausibility. The
+magnitude of improvements varies by dataset: for the simple synthetic
+datasets, distance reductions range from around 20-40\% (\emph{LS},
+\emph{Moon}) to almost 60\% (\emph{Circ}). For the real-world tabular
+datasets, improvements are generally smaller but still substantial in
+many cases with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
+7-8\% for \emph{Cred} and around 3\% for \emph{Adult}. For our only
+vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
+results for our proposed divergence metric are qualitatively similar,
+but generally even more pronounced: for the \emph{Circ} dataset,
+implausibility is reduced by almost 94\% to virtually zero as we
+verified by looking at the absolute outcome. Improvements for other
+datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
+\emph{OL} the reduction is negative, consistent with the distance-based
+metric. The only dataset, for which our proposed metric disagrees with
+the distance-based metric is \emph{MNIST}.
+
+These broad and substantial improvements in plausibility generally do
+not come at the cost of decreased predictive performance: test accuracy
+for CT is virtually identical to the baseline for \emph{Adult},
+\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
+improved for \emph{Cred}. Exceptions to this general pattern are
+\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
+in test accuracy of 2, 5 and 15 percentage points, respectively. We note
+in this context, that we have not optimized our models for predictive
+performance at all and worked with very small networks. In summary, we
+find that CT can substantially improve the quality of explanations
+learned by models without generally sacrificing predictive accuracy.
+
+\begin{table}
+
+\caption{\label{tbl-main}Key plausibility and predictive performance
+metrics for all datasets. The top five rows show the percentage
+reduction in implausibility according to Equation~\ref{eq-impl-dist} for
+varying degrees of the energy penalty used for \emph{ECCo} at test time.
+The following row shows the reduction in implausibility as measured by
+Equation~\ref{eq-impl-div} and aggregated across all test specifications
+of \emph{ECCo}. The final two rows show the test accuracies for the
+model trained with CT and conventionally trained models (``vanilla'').}
+
+\centering{
+
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
+\input{tables/main.tex}
+\end{tabular*}
+
+}
+
+\end{table}%
 
@@
 \(\lambda_{\text{reg}}\) may increase the variability in outcomes when
-combined with high values of \(\lambda_{\text{div}}\) or
-\(\lambda_{\text{adv}}\), this effect is not very pronounced.
+combined with high values for any of the other penalties in
+Equation~\ref{eq-obj}, this effect is not very pronounced.
 
@@
 find that it is not necessary to employ CT during the entire training
-phase to achieve the desired improvements in explainability. When
-training models conventionally during the first 50\% of epochs before
-switching to CT for the next 50\% of epochs, we observed positive
-results. Put differently, CT may be a way to improve the explainability
-of models in a fine-tuning manner.
+phase to achieve the desired improvements in explainability. We have
+tested training models conventionally during the first half of training
+before switching to CT after this initial ``burn-in'' period and
+observed positive results. Put differently, CT may be a way to improve
+the explainability of trained models in a fine-tuning manner.
 
@@
 
-We first address the direct extensions of the counterfactual training
-approach in Section~\ref{sec-approach}. Then, we look at its limitations
-and challenges in Section~\ref{sec-limitations}.
+We begin the discussion by addressing the direct extensions of the
+counterfactual training approach in Section~\ref{sec-approach}. Then, we
+look at its broader limitations and challenges in
+Section~\ref{sec-limitations}.
 
@@
 benefit from investigating more sophisticated approaches towards
-hyperparameter tuning. Notably, CT is iterative which makes a variety of
-methods applicable, including Bayesian \cite{snoek2012practical} or
-gradient-based \cite{franceschi2017forward} optimization.
+hyperparameter tuning. Notably, counterfactual training is iterative
+which makes a variety of methods applicable, including Bayesian
+\cite{snoek2012practical} or gradient-based \cite{franceschi2017forward}
+optimization.
 
@@
 \textbf{\emph{Interventions on features may have downstream impacts on
-fairness.}} Related to the point above, we provide a tool that allows
-practitioners to modify the sensitivity of a model with respect to
-certain features, which may have implication for the fair and equitable
-treatment of individuals subject to automated decisions. As protecting a
-set of features leads the model to assign higher relative importance to
-unprotected features, model owners could misuse our solution by
-enforcing explanations based on features that are more difficult to
-modify by some (group of) individuals. For example, consider again the
-Adult dataset where features such as workclass or education may be more
-difficult to change for underpriviledged groups. When applied
-irresponsibly, counterfactual training could result in an unfairly
-assigned burden of recourse \cite{sharma2020certifai}, threatening the
-equality of opportunity in the system \cite{bell2024fairness} and
-potentially reinforcing social segregation \cite{gao2023impact}. Still,
-as the referenced publications indicate, such phenomena are not specific
-to CT; all types of ADM solutions without strong external protections
-have been recognized to promote harmful power dynamics
-\cite{maas2023machine}.
+fairness.}}\\
+- \href{https://arxiv.org/abs/2401.16088}{Equality of opportunity}?\\
+- \href{https://proceedings.mlr.press/v202/gao23d/gao23d.pdf}{Social
+segregation}?\\
+- Supporting AR within one context/system may still unfairly target
+recourse-affected individuals within other contexts/systems? In Example
+3.1 younger individuals gain access to a loan but their age could still
+lead to, e.g., enforcing stronger supervisory mechanisms?
 
@@
 
-\newpage{}
-
 \bibliographystyle{splncs04}

### 2025-03-13 paper/sections/experiments.qmd
main results finally done

--- 2025-03-12 paper/sections/experiments.qmd
@@
 $$
-\text{implaus}_{\text{dist}}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
 $$ {#eq-impl-dist}
@@
 $$
-\text{implaus}_{\text{div}}(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
+\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
 $$ {#eq-impl-div}
@@
 ### Plausibility
+
+@tbl-main presents our main empirical findings.
+The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time.
+The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*.
+The final two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
+
+We observe that for all datasets except *OL* and across all test settings, the average distance of counterfactuals from observed samples in the target class is reduced, indicating improved plausibility.
+The magnitude of improvements varies by dataset: for the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*).
+For the real-world tabular datasets, improvements are generally smaller but still substantial in many cases with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*.
+For our only vision dataset (*MNIST*), distances are reduced by up to 9%.
+The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by looking at the absolute outcome.
+Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*).
+For *OL* the reduction is negative, consistent with the distance-based metric.
+The only dataset, for which our proposed metric disagrees with the distance-based metric is *MNIST*.
+
+These broad and substantial improvements in plausibility generally do not come at the cost of decreased predictive performance: test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*.
+Exceptions to this general pattern are *MNIST*, *CH* and *GMSC*, for which we observe reduction in test accuracy of 2, 5 and 15 percentage points, respectively.
+We note in this context, that we have not optimized our models for predictive performance at all and worked with very small networks.
+In summary, we find that CT can substantially improve the quality of explanations learned by models without generally sacrificing predictive accuracy.
+
+::: {#tbl-main}
+
+```{=latex}
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
+\input{tables/main.tex}
+\end{tabular*}
+```
+
+Key plausibility and predictive performance metrics for all datasets.
+The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time.
+The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*.
+The final two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
+
+:::
 
@@
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
-While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
+While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values for any of the other penalties in @eq-obj, this effect is not very pronounced.
 
@@
 Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability.
-When training models conventionally during the first 50% of epochs before switching to CT for the next 50% of epochs, we observed positive results.
-Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.
+We have tested training models conventionally during the first half of training before switching to CT after this initial "burn-in" period and observed positive results.
+Put differently, CT may be a way to improve the explainability of trained models in a fine-tuning manner.

### 2025-03-13 paper/sections/other/main_results.qmd
main results finally done

@@
 df = final_table(res_dir)
-get_table_inputs(df, nothing; backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; table_type=:tabular, save_name="paper/tables/main.tex")
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+tabulate_results(inputs; table_type=:tabular, save_name="paper/quarto_ecml/tables/main.tex", wrap_table=false)
 ```

### 2025-03-13 paper/experiments/CTExperiments/src/presentation.jl

@@ -51,17 +51,20 @@ function filter_dict(dict::Dict; drop_fields=["name", "data", "data_params"])
 end
 
 global LatexReplacements = Dict(
-    "lambda_energy" => "\$\\lambda_{\\text{energy}}\$",
-    "lambda_cost" => "\$\\lambda_{\\text{cost}}\$",
+    "lambda_energy" => "\$\\lambda_{\\text{egy}}\$",
+    "lambda_cost" => "\$\\lambda_{\\text{cst}}\$",
     "lambda_adversarial" => "\$\\lambda_{\\text{adv}}\$",
     "lambda_energy_diff" => "\$\\lambda_{\\text{div}}\$",
     "lambda_energy_reg" => "\$\\lambda_{\\text{reg}}\$",
-    "lambda_class_loss" => "\$\\lambda_{\\text{yloss}}\$",
+    "lambda_class_loss" => "\$\\lambda_{\\text{clf}}\$",
     "gmsc" => get_name(GMSC(); pretty=true),
-    "lin_sep" => get_name(LinearlySeparable(); pretty=true),
-    "over" => get_name(Overlapping(); pretty=true),
-    "cali" => get_name(CaliHousing(); pretty=true),
-    "mnist" => get_name(MNIST(); pretty=true)
+    "lin_sep" => "LS",
+    "over" => "OL",
+    "cali" => "CH",
+    "mnist" => get_name(MNIST(); pretty=true),
+    "circles" => "Circ",
+    "moons" => "Moon",
+    "credit" => "Cred"
 )
 
 function format_header(s::String; replacements::Dict=LatexReplacements)
@@ -613,8 +616,8 @@ function get_img_command(data_names, full_paths, fig_labels; fig_caption="")
 end
 
 global LatexMetricReplacements = Dict(
-    "mmd" => "\$\\text{implaus}_{\\text{div}}\$",
-    "plausibility_distance_from_target" => "\$\\text{implaus}_{\\text{dist}}\$",
+    "mmd" => "\$ \\text{IP}^* \$",
+    "plausibility_distance_from_target" => "\$ \\text{IP} \$",
 )
 
 function format_metric(m::String)
@@ -634,7 +637,7 @@ function aggregate_ce_evaluation(res_dir::String; y="mmd", byvars=nothing, rebas
     df.objective .= CTExperiments.format_header.(df.objective)
     df.variable .= CTExperiments.format_metric.(y)
     if rebase
[…]

### 2025-03-13 paper/quarto_ecml/paper.tex

--- 2025-03-13 paper/quarto_ecml/paper.tex
@@
 \newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-
 \makeatother
@@
 are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
-Definition~\ref{def-explainability}). Instead, the choice has generally
-been to improve the ability of XAI tools to identify the subset of
+Def. \ref{def-explainability}). Instead, the choice has generally been
+to improve the ability of XAI tools to identify the subset of
 explanations that are both plausible and valid for any given model,
@@
 
-\begin{definition}[Model
-Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}
+\begin{definition}[Model Explainability]
+\label{def-explainability}
+Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$.
+Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.
 
-Let \(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a
-supervised classification model that maps from the \(D\)-dimensional
-input space \(\mathcal{X}\) to representations
-\(\phi(\mathbf{x};\theta)\) and finally to the \(K\)-dimensional output
-space \(\mathcal{Y}\). Assume that for any given input-output pair
-\(\{\mathbf{x},\mathbf{y}\}_i\) there exists a counterfactual
-\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
-where \(\arg\max_y{\mathbf{y}^{+}}=y^+\) and \(y^+\) denotes the index
-of the target class.
-
-We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
-that faithfully generated counterfactuals are plausible (i.e.,
-consistent with the data) and actionable. Formally, we define these
-properties as follows:
+We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. Formally, we define these properties as follows,
 
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  (Plausibility)
-  \(\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
-  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
-\item
-  (Actionability) Permutations \(\Delta\) are subject to some
-  actionability constraints.
+    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
+    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
+    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
 \end{enumerate}
 
-We consider counterfactuals as faithful to the extent that they are
-consistent with what the model has learned about the input data. Let
-\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
-over inputs, then formally:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  (Faithfulness)
-  \(\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
-  where \(A\) is defined as above.
-\end{enumerate}
-
+where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
 \end{definition}
 
-The characterization of faithfulness and plausibility in
-Definition~\ref{def-explainability} is the same as in
-\cite{altmeyer2024faithful}, with adapted notation. Actionability
-constraints in Definition~\ref{def-explainability} vary and depend on
-the context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
+The characterization of faithfulness and plausibility in Def.
+\ref{def-explainability} is the same as in \cite{altmeyer2024faithful},
+with adapted notation. Intuitively, plausible counterfactuals are
+consistent with the data and faithful counterfactuals are consistent
+with what the model has learned about input data. Actionability
+constraints in Def. \ref{def-explainability} vary and depend on the
+context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
 focus on domain and mutability constraints for individual features
@@
   background information on gradient-based counterfactual search and
-  convergence see \textbf{?@sec-app-ce}.}
+  convergence see supplementary appendix.}
 
-To train models with high explainability as defined in
-Definition~\ref{def-explainability}, we propose to leverage
-counterfactuals in the following objective:
+To train models with high explainability as defined in Def.
+\ref{def-explainability}, we propose to leverage counterfactuals in the
+following objective:
 
@@
 A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
-tend to do well with respect to the plausibility objective in
-Definition~\ref{def-explainability}. This follows directly if we
-consider samples drawn from \(p_\theta(\mathbf{x})\) as counterfactuals
-because the JEM objective effectively minimizes the divergence between
-the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To
-generate samples, \cite{grathwohl2020your} rely on Stochastic Gradient
-Langevin Dynamics (SGLD) using an uninformative prior for initialization
-but we depart from their methodology. Instead of SGLD, we propose to use
+tend to do well with respect to the plausibility objective in Def.
+\ref{def-explainability}. This follows directly if we consider samples
+drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
+objective effectively minimizes the divergence between the conditional
+posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
+\cite{grathwohl2020your} rely on Stochastic Gradient Langevin Dynamics
+(SGLD) using an uninformative prior for initialization but we depart
+from their methodology. Instead of SGLD, we propose to use
 counterfactual explainers to generate counterfactuals of observed
@@
 energy of counterfactuals (negative samples) \cite{du2019implicit}. As
-the counterfactuals get more plausible
-(Definition~\ref{def-explainability}) during training, these opposing
-effects gradually balance each other out \cite{lippe2024uvadlc}.
+the counterfactuals get more plausible (Def. \ref{def-explainability})
+during training, these opposing effects gradually balance each other out
+\cite{lippe2024uvadlc}.
 
@@
 
-\begin{proposition}[Protecting Immutable
-Features]\protect\hypertarget{prp-mtblty}{}\label{prp-mtblty}
-
-Let
-\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
-denote a linear classifier with softmax activation \(\mathcal{S}\)
-(i.e., \emph{multinomial logistic regression}) where
-\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
-we assume multivariate Gaussian class densities with common diagonal
-covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
-then protecting an immutable feature from the contrastive divergence
-penalty (Equation~\ref{eq-div}) will result in lower classifier
-sensitivity to that feature relative to the remaining features, provided
-that at least one of those is discriminative and mutable.
-
+\begin{proposition}[Protecting Immutable Features]
+\label{prp-mtblty}
+Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
+If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
 \end{proposition}
 
-It is worth highlighting that Proposition~\ref{prp-mtblty} assumes
-independence of features. This raises a valid concern about the effect
-of protecting immutable features in the presence of proxies that remain
-unprotected. We address this in Section~\ref{sec-discussion}.
+It is worth highlighting that Prp.\textasciitilde{}\ref{prp-mtblty}
+assumes independence of features. This raises a valid concern about the
+effect of protecting immutable features in the presence of proxies that
+remain unprotected. We address this in Section~\ref{sec-discussion}.
 
-\subsection{Example: influence of counterfactual training on
-explainability}\label{example-influence-of-counterfactual-training-on-explainability}
-
-To better convey the intuition underlying our proposed method, we
-illustrate different model outcomes in Example~\ref{exm-grad}.
-
-\begin{example}[Prediction of Consumer Credit
-Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}
+\subsection{Example (Prediction of Consumer Credit
+Default)}\label{example-prediction-of-consumer-credit-default}
 
@@
 counterfactuals are highly plausible, meeting the first objective of
-Definition~\ref{def-explainability}.
+Def. \ref{def-explainability}.
 
@@
 also plausible with respect to the mutable feature. Thus, we consider
-the model in panel (d) as the most explainable according to
-Definition~\ref{def-explainability}.
-
-\end{example}
+the model in panel (d) as the most explainable according to Def.
+\ref{def-explainability}.
 
@@
 
-\caption{\label{fig-poc}Illustration for Example~\ref{exm-grad} of how
-CT improves model explainability.}
+\caption{\label{fig-poc}Illustration of how CT improves model
+explainability.}
 
@@
 
-\begin{corollary}[Plausibility]\protect\hypertarget{cor-plaus}{}\label{cor-plaus}
-
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
   To what extent does our proposed counterfactual training objective
   (Equation~\ref{eq-obj}) induce models to learn plausible explanations?
-
-\end{corollary}
-
-\begin{corollary}[Actionability]\protect\hypertarget{cor-action}{}\label{cor-action}
-
+\item
   To what extent does our proposed counterfactual training objective
@@
   outcomes in the presence of actionability constraints?
-
-\end{corollary}
-
-Beyond this, we are also interested in understanding how robust our
-answers to RQ~\ref{cor-plaus} and RQ~\ref{cor-action} are:
-
-\begin{corollary}[Hyperparameters]\protect\hypertarget{cor-hyper}{}\label{cor-hyper}
-
+\item
   What are the effects of hyperparameter selection with respect to
   Equation~\ref{eq-obj}?
-
-\end{corollary}
+\end{enumerate}
 
@@
 Our key outcome of interest is how well do models perform with respect
-to explainability (Definition~\ref{def-explainability}). To this end, we
-focus primarily on the plausibility and cost of faithfully generated
+to explainability (Def. \ref{def-explainability}). To this end, we focus
+primarily on the plausibility and cost of faithfully generated
 counterfactuals at test time. To measure the cost of counterfactuals, we
@@
 
-\subsubsection{Plausibility}\label{plausibility-1}
+\subsubsection{Plausibility}\label{plausibility}
 
-\subsubsection{Actionability}\label{actionability-1}
+Table~\ref{tbl-main} presents our main empirical findings. The top five
+rows show the percentage reduction in implausibility according to
+Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
+used for \emph{ECCo} at test time. The following row shows the reduction
+in implausibility as measured by Equation~\ref{eq-impl-div} and
+aggregated across all test specifications of \emph{ECCo}. The final two
+rows show the test accuracies for the model trained with CT and
+conventionally trained models (``vanilla'').
+
+We observe that for all datasets except \emph{OL} and across all test
+settings, the average distance of counterfactuals from observed samples
+in the target class is reduced, indicating improved plausibility. The
+magnitude of improvements varies by dataset: for the simple synthetic
+datasets, distance reductions range from around 20-40\% (\emph{LS},
+\emph{Moon}) to almost 60\% (\emph{Circ}). For the real-world tabular
+datasets, improvements are generally smaller but still substantial in
+many cases with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
+7-8\% for \emph{Cred} and around 3\% for \emph{Adult}. For our only
+vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
+results for our proposed divergence metric are qualitatively similar,
+but generally even more pronounced: for the \emph{Circ} dataset,
+implausibility is reduced by almost 94\% to virtually zero as we
+verified by looking at the absolute outcome. Improvements for other
+datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
+\emph{OL} the reduction is negative, consistent with the distance-based
+metric. The only dataset, for which our proposed metric disagrees with
+the distance-based metric is \emph{MNIST}.
+
+These broad and substantial improvements in plausibility generally do
+not come at the cost of decreased predictive performance: test accuracy
+for CT is virtually identical to the baseline for \emph{Adult},
+\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
+improved for \emph{Cred}. Exceptions to this general pattern are
+\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
+in test accuracy of 2, 5 and 15 percentage points, respectively. We note
+in this context, that we have not optimized our models for predictive
+performance at all and worked with very small networks. In summary, we
+find that CT can substantially improve the quality of explanations
+learned by models without generally sacrificing predictive accuracy.
+
+\begin{table}
+
+\caption{\label{tbl-main}Key plausibility and predictive performance
+metrics for all datasets. The top five rows show the percentage
+reduction in implausibility according to Equation~\ref{eq-impl-dist} for
+varying degrees of the energy penalty used for \emph{ECCo} at test time.
+The following row shows the reduction in implausibility as measured by
+Equation~\ref{eq-impl-div} and aggregated across all test specifications
+of \emph{ECCo}. The final two rows show the test accuracies for the
+model trained with CT and conventionally trained models (``vanilla'').}
+
+\centering{
+
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
+\input{tables/main.tex}
+\end{tabular*}
+
+}
+
+\end{table}%
+
+\subsubsection{Actionability}\label{actionability}
 
@@
 demanding datasets on a high-performance computing cluster with 34 2GB
-CPUs (see details in \textbf{?@sec-hardware}). However, there are three
-factors that attenuate the impact of this limitation. First, CT provides
-CEs for the training samples for free, which may be beneficial in many
+CPUs\footnote{See supplementary appendix for computational details.}.
+However, there are three factors that attenuate the impact of this
+limitation. First, CT provides counterfactual explanations for the
+training samples essentially for free, which may be beneficial in many
 ADM systems. Second, we find that CT can retain its value when used as a

### 2025-03-13 paper/sections/appendix.qmd

@@
 
-\setcounter{section}{0}
+<!-- \setcounter{section}{0} -->
 \renewcommand{\thesection}{\Alph{section}}
 
-\setcounter{table}{0}
+<!-- \setcounter{table}{0} -->
 \renewcommand{\thetable}{A\arabic{table}}
 
-\setcounter{figure}{0}
+<!-- \setcounter{figure}{0} -->
 \renewcommand{\thefigure}{A\arabic{figure}}
@@
 
-# Technical Details of Our Approach {.appendix}
+{{< include /paper/sections/other/technical.qmd >}}
 
-{{< include /paper/sections/other/counterfactuals.qmd >}}
-
-{{< include /paper/sections/other/constraints.qmd >}}
-
-{{< include /paper/sections/other/training.qmd >}}
-
-# Detailed Results {.appendix}
-
-## Qualitative Findings for Image Data
-
-::: {.callout-note}
-
-@fig-mnist shows much more plausible (faithful) counterfactuals for a model with CT than the model with conventional training (@fig-mnist-vanilla).
-In fact, this is not even using *ECCo+* and still showing better results than the best results we achieved in our AAAI paper for JEM ensembles.
-
-:::
-
-![Counterfactual images for *MLP* with counterfactual training.
-The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}
-
-![Counterfactual images for *MLP* with conventional training.
-The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp_vanilla.png){#fig-mnist-vanilla}
+{{< include /paper/sections/other/main_results.qmd >}}
 
@@
 
-<!-- {{< include /paper/sections/other/initial_grids.qmd >}} -->
+{{< include /paper/sections/other/tuning.qmd >}}
 

### 2025-03-13 paper/sections/discussion.qmd

@@
 While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations.
-We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see details in @sec-hardware).
+We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs^[See supplementary appendix for computational details.].
 However, there are three factors that attenuate the impact of this limitation.
-First, CT provides CEs for the training samples for free, which may be beneficial in many ADM systems.
+First, CT provides counterfactual explanations for the training samples essentially for free, which may be beneficial in many ADM systems.
 Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models.

### 2025-03-13 paper/sections/experiments.qmd

@@
 
-::: {#cor-plaus}
-
-## Plausibility
-
-Does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?
-
-:::
-
-::: {#cor-action}
-
-## Actionability
-
-Does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
-
-:::
-
-Beyond this, we are also interested in understanding how robust our answers to @cor-plaus and @cor-action are:
-
-::: {#cor-hyper}
-
-## Hyperparameters
-
-What are the effects of different hyperparameter choices with respect to @eq-obj?
-
-:::
+1. To what extent does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?
+2. To what extent does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+3. What are the effects of hyperparameter selection with respect to @eq-obj?
 
@@
 
-Our key outcome of interest is how well models perform with respect to explainability (@def-explainability): to this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
+Our key outcome of interest is how well do models perform with respect to explainability (Def.
+\ref{def-explainability}).
+To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
 To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.

### 2025-03-13 paper/sections/intro.qmd

@@
 To our surprise, the second challenge has not yet attracted major research interest.
-Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see @def-explainability).
+Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see Def.
+\ref{def-explainability}).
 Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with plausible explanations [@altmeyer2024faithful].

### 2025-03-13 paper/sections/method.qmd

@@
 
-::: {#def-explainability}
-
-## Model Explainability
-
+\begin{definition}[Model Explainability]
+\label{def-explainability}
 Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$.
@@
 
-We say that $\mathbf{M}_\theta$ is **explainable** to the extent that faithfully generated counterfactuals are plausible (i.e., consistent with the data) and actionable.
-Formally, we define these properties as follows:
+We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. Formally, we define these properties as follows,
 
-1. (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
-2. (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
+\begin{enumerate}
+    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
+    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
+    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
+\end{enumerate}
 
-We consider counterfactuals as faithful to the extent that they are consistent with what the model has learned about the input data.
-Let $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denote the conditional posterior over inputs, then formally:
+where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
+\end{definition}
 
-3. (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
-
-:::
-
-The characterization of faithfulness and plausibility in @def-explainability is the same as in @altmeyer2024faithful, with adapted notation.
-Actionability constraints in @def-explainability vary and depend on the context in which $\mathbf{M}_\theta$ is deployed.
+The characterization of faithfulness and plausibility in Def.
+\ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation.
+Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about input data.
+Actionability constraints in Def.
+\ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed.
 In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$.
@@
 Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual.
-For our purposes, we let $T$ vary and consider the counterfactual search to be converged as soon as the predicted probability for the target class has reached a pre-determined threshold $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see @sec-app-ce.]
+For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see supplementary appendix.]
 
-To train models with high explainability as defined in @def-explainability, we propose to leverage counterfactuals in the following objective:
+<!-- @sec-app-ce -->
+
+To train models with high explainability as defined in Def. \ref{def-explainability}, we propose to leverage counterfactuals in the following objective:
 
@@
 
-A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in @def-explainability.
+A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in Def.
+\ref{def-explainability}.
 This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$.
@@
 Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit].
-As the counterfactuals get more plausible (@def-explainability) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].
+As the counterfactuals get more plausible (Def.
+\ref{def-explainability}) during training, these opposing effects gradually balance each other out [@lippe2024uvadlc].
 
@@
 
-::: {#prp-mtblty}
+\begin{proposition}[Protecting Immutable Features]
+\label{prp-mtblty}
+Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
+If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
+\end{proposition}
 
-## Protecting Immutable Features
-
-Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ (i.e., *multinomial logistic regression*) where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
-If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty (@eq-div) will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
-
-:::
-
-It is worth highlighting that @prp-mtblty assumes independence of features.
+It is worth highlighting that Prp.~\ref{prp-mtblty} assumes independence of features.
 This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected.
@@
 
-## Example: influence of counterfactual training on explainability
-
-To better convey the intuition underlying our proposed method, we illustrate different model outcomes in @exm-grad.
-
-::: {#exm-grad}
-
-## Prediction of Consumer Credit Default
+## Example (Prediction of Consumer Credit Default)
 
@@
 In panel (b), we have trained our model with counterfactual training, once again without any mutability constraints.
-We observe that the counterfactuals are highly plausible, meeting the first objective of @def-explainability.
+We observe that the counterfactuals are highly plausible, meeting the first objective of Def.
+\ref{def-explainability}.
 
@@
 The counterfactuals are also plausible with respect to the mutable feature.
-Thus, we consider the model in panel (d) as the most explainable according to @def-explainability.
+Thus, we consider the model in panel (d) as the most explainable according to Def.
+\ref{def-explainability}.
 
-:::
-
-![Illustration for @exm-grad of how CT improves model explainability.](/paper/figures/poc.svg){#fig-poc}
+![Illustration of how CT improves model explainability.](/paper/figures/poc.svg){#fig-poc}

### 2025-03-13 paper/sections/other/constraints.qmd

@@
 We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments.
-Below we derive the analytical results in @prp-mtblty.
+Below we derive the analytical results in Prp.~\ref{prp-mtblty}.
 
@@
 no change in sensitivity for $d_{\text{immtbl}}$ relative to increased sensitivity for $d_{\text{mtbl}}$).
-
-:::
-
-::: {.callout-warning}
-
-\@Cynthia, \@Arie, I have tentatively phrased the above in terms of a theorem and proof.
-This is something I've so far shied away from because I feel a bit out of my depth when it comes to mathematical proofs.
-The above makes intuitive sense to me, but I don't know for sure if it's correct.
 

### 2025-03-13 paper/sections/other/main_results.qmd
uff

--- 2025-03-13 paper/sections/other/main_results.qmd
@@
 df = final_table(res_dir)
-get_table_inputs(df, nothing; backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; table_type=:tabular, save_name="paper/tables/main.tex")
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+tabulate_results(inputs; table_type=:tabular, save_name="paper/quarto_ecml/tables/main.tex", wrap_table=false)
+tabulate_results(inputs; table_type=:tabular, save_name="paper/preprint/tables/main.tex", wrap_table=false)
 ```

### 2025-03-13 paper/preprint/paper.tex

--- 2025-03-13 paper/preprint/paper.tex
@@
 
+\usepackage{amsthm}
+\theoremstyle{plain}
+\newtheorem{proposition}{Proposition}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\theoremstyle{definition}
+\newtheorem{example}{Example}[section]
+\theoremstyle{plain}
 \usepackage{arxiv}
@@
 \usepackage{amsthm}
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{definition}
-\newtheorem{example}{Example}[section]
-\theoremstyle{plain}
-\newtheorem{corollary}{Research Question}[section]
 \theoremstyle{remark}
@@
 state-of-the-art neural networks are so expressive that they can fit
-randomly labeled data. Thus, looking at the predictive performance, the
-solutions may seem to provide compelling explanations for the data, when
-in fact they are based on purely associative, semantically meaningless
-patterns. This poses two related challenges. Firstly, there is no
-dependable way to verify if such complex representations correspond to
-meaningful and plausible explanations. Secondly, even if we could
-resolve the first challenge, it remains undecided how to ensure that
-models can \emph{only} learn valuable explanations.
+randomly labeled data. Thus, looking at the predictive performance
+alone, the solutions may seem to provide compelling explanations for the
+data, when in fact they are based on purely associative, semantically
+meaningless patterns. This poses two related challenges. Firstly, there
+is no dependable way to verify if such complex representations
+correspond to meaningful and plausible explanations. Secondly, even if
+we could resolve the first challenge, it remains undecided how to ensure
+that models can \emph{only} learn valuable explanations.
 
 The first challenge has attracted an abundance of research on
-\textbf{explainable AI} (XAI) which aims to develop tools to derive
-explanations from complex model representations. This can mitigate a
-scenario in which we deploy opaque models and blindly rely on their
-predictions. On countless occasions, this scenario has occurred in
-practice and caused real harm to people who were affected adversely and
-often unfairly by automated decision-making (ADM) systems involving
-opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016};
+\textbf{explainable AI} (XAI), a paradigm that focuses on the
+development of tools to derive (post-hoc) explanations from complex
+model representations. Such explanations should mitigate a scenario in
+which practitioners deploy opaque models and blindly rely on their
+predictions. On countless occasions, this has happened in practice and
+caused real harms to people who were adversely and unfairly affected by
+automated decision-making (ADM) systems involving opaque models
+(\citeproc{ref-oneil2016weapons}{O'Neil 2016};
 \citeproc{ref-mcgregor2021preventing}{McGregor 2021}). Effective XAI
 tools can aid us in monitoring models and providing recourse to
-individuals to turn adverse outcomes (e.g., ``loan application
-rejected'') into positive ones (e.g., ``application accepted'').
+individuals to turn negative outcomes (e.g., ``loan application
+rejected'') into positive ones (e.g., ``application accepted''). Our
+work builds upon \textbf{counterfactual explanations} (CE) proposed by
 Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}) propose
-\textbf{counterfactual explanations} (CE) as an effective approach to
-achieve this goal: CEs explain how factual inputs need to change in
-order for some fitted model to produce some desired output, typically
-involving minimal perturbations.
-
-To our surprise, the second challenge has not yet attracted any major
-consolidated research effort. Specifically, there has been no concerted
-effort towards improving improving models' explanatory capacity, which
-we will henceforth simply call ``explainability'', defined as the degree
-to which learned representations correspond to explanations that are
-interpretable and deemed \textbf{plausible} by humans (see
-Definition~\ref{def-explainability}). Instead, the choice has typically
-been to improve the ability of XAI tools to identify the subset
+(\citeproc{ref-wachter2017counterfactual}{2017}) as an effective
+approach to achieve this goal. CEs prescribe minimal changes for factual
+inputs that, if implemented, would prompt some fitted model to produce a
+desired output.
+
+To our surprise, the second challenge has not yet attracted major
+research interest. Specifically, there has been no concerted effort
+towards improving the ``explanatory capacity'' of models, i.e., the
+degree to which learned representations correspond to explanations that
+are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
+Def. \ref{def-explainability}). Instead, the choice has generally been
+to improve the ability of XAI tools to identify the subset of
 explanations that are both plausible and valid for any given model,
 independent of whether the learned representations are also compatible
-with implausible explanations
+with plausible explanations
 (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-Fortunately, recent findings indicate that explainability can arise as
-byproduct of regularization techniques aimed at other objectives such as
-robustness, generalization, and generative capacity
-(\citeproc{ref-schut2021generating}{Schut et al. 2021};
+Fortunately, recent findings indicate that improved explanatory capacity
+can arise as a consequence of regularization techniques aimed at other
+training objectives such as robustness, generalization, and generative
+capacity (\citeproc{ref-schut2021generating}{Schut et al. 2021};
 \citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein 2020};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-
-Building on these findings, we introduce \textbf{counterfactual
-training}: a novel training regime explicitly geared towards aligning
-model representations with plausible explanations. Our contributions are
-as follows:
+\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). As further
+discussed in Section~\ref{sec-lit}, our work consolidates these findings
+within a single objective.
+
+\textbf{Specifically, we introduce counterfactual training}: a novel
+training regime explicitly meant to align learned representations with
+plausible explanations that comply with user requirements. Our
+contributions are as follows:
 
@@
 \item
-  We discuss existing related work on improving models and consolidate
-  it through the lens of counterfactual explanations
-  (Section~\ref{sec-lit}).
-\item
-  We present our proposed methodological framework that leverages
-  faithful counterfactual explanations during the training phase of
-  models to achieve the explainability objective
-  (Section~\ref{sec-method}).
-\item
-  Through extensive experiments we demonstrate the counterfactual
-  training improve model explainability while maintaining high
-  predictive performance. We run ablation studies and grid searches to
-  understand how the underlying model components and hyperparameters
-  affect outcomes. (Section~\ref{sec-experiments}).
+  We present a novel methodological framework that leverages adversarial
+  examples and faithful counterfactual explanations during the training
+  phase to improve the explanatory capacity and robustness of machine
+  learning models (Section~\ref{sec-method}).\\
+\item
+  We propose a method to enforce global actionability constraints by
+  preventing models from assigning importance to immutable features,
+  i.e., ones over which decision subjects have no control
+  (Section~\ref{sec-method}).\\
+\item
+  Through extensive experiments we demonstrate that counterfactual
+  training promotes explainability while preserving high predictive
+  performance. We run ablation studies and grid searches to understand
+  how the underlying model components and hyperparameters affect
+  outcomes. (Section~\ref{sec-experiments}).
 \end{itemize}
 
-Despite some limitations of our approach discussed in
-Section~\ref{sec-discussion}, we conclude in
-Section~\ref{sec-conclusion} that counterfactual training provides a
-practical framework for researchers and practitioners interested in
-making opaque models more trustworthy. We also believe that this work
+Despite some limitations discussed in Section~\ref{sec-discussion}, we
+conclude in Section~\ref{sec-conclusion} that counterfactual training
+provides a useful framework for researchers and practitioners interested
+in making opaque models more trustworthy. We also believe that this work
 serves as an opportunity for XAI researchers to re-evaluate the trend of
@@
 is a necessary but insufficient condition for using CEs to provide
-algorithmic recourse (AR) to individuals affected by opaque models in
-practice. This is because for recourse recommendations to be
-\textbf{actionable}, they need to not only result in plausible
-counterfactuals but also be attainable. A plausible CE for a rejected
-20-year-old loan applicant, for example, might reveal that their
-application would have been accepted, if only they were 20 years older.
-Ignoring all other features, this would comply with the definition of
-plausibility if 40-year-old individuals were in fact more credit-worthy
-on average than young adults. But of course this CE does not qualify for
-providing actionable recourse to the applicant since \emph{age} is not a
-(directly) mutable feature. For our intents and purposes, counterfactual
-training aims to improve model explainability by aligning models with
-counterfactuals that meet both desiderata, plausibility and
-actionability. Formally, we define explainability as follows:
-
-\begin{definition}[Model
-Explainability]\protect\hypertarget{def-explainability}{}\label{def-explainability}
-
-Let \(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denote a
-supervised classification model that maps from the \(D\)-dimensional
-input space \(\mathcal{X}\) to representations
-\(\phi(\mathbf{x};\theta)\) and finally to the \(K\)-dimensional output
-space \(\mathcal{Y}\). Assume that for any given input-output pair
-\(\{\mathbf{x},\mathbf{y}\}_i\) there exists a counterfactual
-\(\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})\)
-where \(\arg\max_y{\mathbf{y}^{+}}=y^+\) and \(y^+\) denotes the index
-of the target class.
-
-We say that \(\mathbf{M}_\theta\) is \textbf{explainable} to the extent
-that faithfully generated counterfactuals are plausible (i.e.~consistent
-with the data) and actionable. Formally, we define these properties as
-follows:
+algorithmic recourse (AR) to individuals (negatively) affected by opaque
+models. For AR recommendations to be actionable, they need to not only
+result in plausible counterfactuals but also be attainable. A plausible
+CE for a rejected 20-year-old loan applicant, for example, might reveal
+that their application would have been accepted, if only they were 20
+years older. Ignoring all other features, this would comply with the
+definition of plausibility if 40-year-old individuals were in fact more
+credit-worthy on average than young adults. But of course this CE does
+not qualify for providing actionable recourse to the applicant since
+\emph{age} is not a (directly) mutable feature. CT aims to improve model
+explainability by aligning models with counterfactuals that meet both
+desiderata: plausibility and actionability. Formally, we define
+explainability as follows:
+
+\begin{definition}[Model Explainability]
+\label{def-explainability}
+Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$.
+Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.
+
+We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. Formally, we define these properties as follows,
 
 \begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  (Plausibility)
-  \(\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
-  where \(A\) is some small region around \(\mathbf{x}^{\prime}\).
-\item
-  (Actionability) Permutations \(\Delta\) are subject to some
-  actionability constraints.
+    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
+    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
+    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
 \end{enumerate}
 
-We consider counterfactuals as faithful to the extent that they are
-consistent with what the model has learned about the input data. Let
-\(p_\theta(\mathbf{x}|\mathbf{y}^{+})\) denote the conditional posterior
-over inputs, then formally:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{2}
-\tightlist
-\item
-  (Faithfulness)
-  \(\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1\)
-  where \(A\) is defined as above.
-\end{enumerate}
-
+where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
 \end{definition}
 
-The definitions of faithfulness and plausibility in
-Definition~\ref{def-explainability} are the same as in Altmeyer et al.
+The characterization of faithfulness and plausibility in Def.
+\ref{def-explainability} is the same as in Altmeyer et al.
 (\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
-Actionability constraints in Definition~\ref{def-explainability} vary
-and depend on the context in which \(\mathbf{M}_\theta\) is deployed. In
-this work, we focus on domain and mutability constraints for individual
-features \(x_d\) for \(d=1,...,D\). We limit ourselves to classification
-tasks for reasons discussed in Section~\ref{sec-discussion}.
+Intuitively, plausible counterfactuals are consistent with the data and
+faithful counterfactuals are consistent with what the model has learned
+about input data. Actionability constraints in Def.
+\ref{def-explainability} vary and depend on the context in which
+\(\mathbf{M}_\theta\) is deployed. In this work, we focus on domain and
+mutability constraints for individual features \(x_d\) for
+\(d=1,...,D\). We limit ourselves to classification tasks for reasons
+discussed in Section~\ref{sec-discussion}.
 
@@
 
-To train models with high explainability as defined in
-Definition~\ref{def-explainability}, we propose to leverage
-counterfactuals in the following objective:
+To train models with high explainability as defined in Def.
+\ref{def-explainability}, we propose to leverage counterfactuals in the
+following objective:
 
 \begin{equation}\phantomsection\label{eq-obj}{
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) + \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
+\begin{split}
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
+\end{split}
 }\end{equation}
 
-where \(\text{yloss}(\cdot)\) is any conventional classification loss
-that induces discriminative performance (e.g., cross-entropy). The
-second and third terms in Equation~\ref{eq-obj} are explained in more
-detail below. For now, they can be sufficiently described as inducing
-explainability directly and indirectly by penalizing: (1) the
-contrastive divergence, \(\text{div}(\cdot)\), between mature
-counterfactuals \(\mathbf{x}_T^\prime\) and observed samples \(x\) and,
-(2) the adversarial loss, \(\text{advloss}(.)\), with respect to nascent
+where \(\text{yloss}(\cdot)\) is a classification loss that induces
+discriminative performance (e.g., cross-entropy). The second and third
+terms in Equation~\ref{eq-obj} are explained in detail below. For now,
+they can be sufficiently described as inducing explainability directly
+and indirectly by penalizing: (1) the contrastive divergence,
+\(\text{div}(\cdot)\), between mature counterfactuals
+\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
+adversarial loss, \(\text{advloss}(.)\), with respect to nascent
 counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
 \(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
-regularises the magnitude of the energy terms involved in
+regularizes the magnitude of the energy terms involved in
 \(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and Mordatch
-2020}). The tradeoff between the different components can be governed by
+2020}). The trade-off between the components can be governed by
 adjusting the strengths of the penalties \(\lambda_{\text{div}}\),
@@
 (\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
-well with respect to the plausibility objective in
-Definition~\ref{def-explainability}. If we consider samples drawn from
-\(p_\theta(\mathbf{x})\) as counterfactuals, this is an expected
-finding, because the JEM objective effectively minimizes the divergence
-between the conditional posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\).
-To generate samples, Grathwohl et al.
-(\citeproc{ref-grathwohl2020your}{2020}) rely on Stochastic Gradient
-Langevin Dynamics (SGLD) using an uninformative prior for
-initialization. This is where we depart from their methodology: instead
+well with respect to the plausibility objective in Def.
+\ref{def-explainability}. This follows directly if we consider samples
+drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
+objective effectively minimizes the divergence between the conditional
+posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
+Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) rely on
+Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative
+prior for initialization but we depart from their methodology. Instead
 of SGLD, we propose to use counterfactual explainers to generate
@@
 
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. In
-particular, we set
+where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
 \(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
 where \(y^+\) denotes the index of the randomly drawn target class,
-\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed data point
-sampled from target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\).
-Conditional on the target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes
-a mature counterfactual for a randomly sampled factual from a non-target
-class generated through a gradient-based counterfactual generator for at
-most \(T\) iterations. We define mature counterfactuals as those that
-have either exhausted \(T\) or reached convergence in terms of the
-pre-determined decision threshold earlier.
+\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed sample from
+target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\). Conditional on the
+target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes a mature
+counterfactual for a randomly sampled factual from a non-target class
+generated with a gradient-based CE generator for up to \(T\) iterations.
+Mature counterfactuals are ones that have either reached convergence
+wrt. the decision threshold \(\tau\) or exhausted \(T\).
 
 Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
-of observed training samples (positive samples) while at same time
-increasing the energy of counterfactuals (negative samples)
-(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the generated
-counterfactuals get more plausible (Definition~\ref{def-explainability})
-over the course of training, these two opposing effects gradually
-balance each out (\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
+of observed training samples (positive samples) while increasing the
+energy of counterfactuals (negative samples)
+(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the
+counterfactuals get more plausible (Def. \ref{def-explainability})
+during training, these opposing effects gradually balance each other out
+(\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
 
@@
 
-\begin{proposition}[Protecting Immutable
-Features]\protect\hypertarget{prp-mtblty}{}\label{prp-mtblty}
-
-Let
-\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
-denote a linear classifier with softmax activation \(\mathcal{S}\)
-(i.e., \emph{multinomial logistic regression}) where
-\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
-we assume multivariate Gaussian class densities with common diagonal
-covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
-then protecting an immutable feature from the contrastive divergence
-penalty (Equation~\ref{eq-div}) will result in lower classifier
-sensitivity to that feature relative to the remaining features, provided
-that at least one of those is discriminative and mutable.
-
+\begin{proposition}[Protecting Immutable Features]
+\label{prp-mtblty}
+Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
+If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
 \end{proposition}
 
-It is worth highlighting that Proposition~\ref{prp-mtblty} assumes
-independence of features. This raises a valid concern about the effect
-of protecting immutable features in the presence of proxy features that
-remain unprotected. We discuss this limitation in
-Section~\ref{sec-discussion}.
-
-\subsection{Illustration}\label{illustration}
-
-To better convey the intuition underlying our proposed method, we
-illustrate different model outcomes in Example~\ref{exm-grad}.
-
-\begin{example}[Prediction of Consumer Credit
-Default]\protect\hypertarget{exm-grad}{}\label{exm-grad}
+It is worth highlighting that Prp.\textasciitilde{}\ref{prp-mtblty}
+assumes independence of features. This raises a valid concern about the
+effect of protecting immutable features in the presence of proxies that
+remain unprotected. We address this in Section~\ref{sec-discussion}.
+
+\subsection{Example (Prediction of Consumer Credit
+Default)}\label{example-prediction-of-consumer-credit-default}
 
@@
 impose mutability constraints at test time. The generated
-counterfactuals are all valid, but not plausible: they are clearly
-distinguishable from the ground-truth data. In panel (b), we have
-trained our model with counterfactual training, once again not imposing
-mutability constraints at test time. We observe that the counterfactuals
-are clearly plausible, therefore meeting the first objective of
-Definition~\ref{def-explainability}.
+counterfactuals are all valid, but not plausible: they do not comply
+with the distribution of the factual samples in the target class to the
+point where they are clearly distinguishable from the ground-truth data.
+In panel (b), we have trained our model with counterfactual training,
+once again without any mutability constraints. We observe that the
+counterfactuals are highly plausible, meeting the first objective of
+Def. \ref{def-explainability}.
 
@@
 counterfactual paths are shorter on average in panel (d), where we have
-used counterfactual training and protected immutable features as
-described in Section~\ref{sec-constraints}. In particular, we observe
-that due to the classifier's lower sensitivity to \emph{age}, recourse
-recommendations with respect to \emph{debt} are much more homogenous, in
-that they do not disproportionately punish younger individuals. The
-counterfactuals are also plausible with respect to the mutable feature.
-Thus, we consider the model in panel (d) as the most explainable
-according to Definition~\ref{def-explainability}.
-
-\end{example}
+used counterfactual training and protected the immutable feature as
+described in Section~\ref{sec-constraints}. We observe that due to the
+classifier's lower sensitivity to \emph{age}, recourse recommendations
+with respect to \emph{debt} are much more homogenous and do not
+disproportionately punish younger individuals. The counterfactuals are
+also plausible with respect to the mutable feature. Thus, we consider
+the model in panel (d) as the most explainable according to Def.
+\ref{def-explainability}.
 
@@
 
-\caption{\label{fig-poc}Visual illustration of how counterfactual
-training improves explainability. See Example~\ref{exm-grad} for
-details.}
+\caption{\label{fig-poc}Illustration of how CT improves model
+explainability.}
 
@@
 
-\begin{corollary}[Plausibility]\protect\hypertarget{cor-plaus}{}\label{cor-plaus}
-
-Does our proposed counterfactual training objective
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  To what extent does our proposed counterfactual training objective
   (Equation~\ref{eq-obj}) induce models to learn plausible explanations?
-
-\end{corollary}
-
-\begin{corollary}[Actionability]\protect\hypertarget{cor-action}{}\label{cor-action}
-
-Does our proposed counterfactual training objective
+\item
+  To what extent does our proposed counterfactual training objective
   (Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
   outcomes in the presence of actionability constraints?
-
-\end{corollary}
-
-Beyond this, we are also interested in understanding how robust our
-answers to RQ~\ref{cor-plaus} and RQ~\ref{cor-action} are:
-
-\begin{corollary}[Hyperparameters]\protect\hypertarget{cor-hyper}{}\label{cor-hyper}
-
-What are the effects of different hyperparameter choices with respect to
+\item
+  What are the effects of hyperparameter selection with respect to
   Equation~\ref{eq-obj}?
-
-\end{corollary}
+\end{enumerate}
 
@@
 
-Our key outcome of interest is how well models perform with respect to
-explainability (Definition~\ref{def-explainability}): to this end, we
-focus primarily on the plausibility and cost of faithfully generated
+Our key outcome of interest is how well do models perform with respect
+to explainability (Def. \ref{def-explainability}). To this end, we focus
+primarily on the plausibility and cost of faithfully generated
 counterfactuals at test time. To measure the cost of counterfactuals, we
@@
 (\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
-Equation~\ref{eq-impl-div} is equal to zero iff
-\(\mathbf{X}^\prime=\mathbf{X}^+\).
+Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
+are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
 
@@
 
-We run the experiments with three CE generators: \emph{Generic} of
-Wachter, Mittelstadt, and Russell
+We run the experiments with three gradient-based generators:
+\emph{Generic} of Wachter, Mittelstadt, and Russell
 (\citeproc{ref-wachter2017counterfactual}{2017}) as a simple baseline
@@
 
-\subsubsection{Plausibility}\label{plausibility-1}
-
-\subsubsection{Actionability}\label{actionability-1}
+\subsubsection{Plausibility}\label{plausibility}
+
+Table~\ref{tbl-main} presents our main empirical findings. The top five
+rows show the percentage reduction in implausibility according to
+Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
+used for \emph{ECCo} at test time. The following row shows the reduction
+in implausibility as measured by Equation~\ref{eq-impl-div} and
+aggregated across all test specifications of \emph{ECCo}. The final two
+rows show the test accuracies for the model trained with CT and
+conventionally trained models (``vanilla'').
+
+We observe that for all datasets except \emph{OL} and across all test
+settings, the average distance of counterfactuals from observed samples
+in the target class is reduced, indicating improved plausibility. The
+magnitude of improvements varies by dataset: for the simple synthetic
+datasets, distance reductions range from around 20-40\% (\emph{LS},
+\emph{Moon}) to almost 60\% (\emph{Circ}). For the real-world tabular
+datasets, improvements are generally smaller but still substantial in
+many cases with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
+7-8\% for \emph{Cred} and around 3\% for \emph{Adult}. For our only
+vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
+results for our proposed divergence metric are qualitatively similar,
+but generally even more pronounced: for the \emph{Circ} dataset,
+implausibility is reduced by almost 94\% to virtually zero as we
+verified by looking at the absolute outcome. Improvements for other
+datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
+\emph{OL} the reduction is negative, consistent with the distance-based
+metric. The only dataset, for which our proposed metric disagrees with
+the distance-based metric is \emph{MNIST}.
+
+These broad and substantial improvements in plausibility generally do
+not come at the cost of decreased predictive performance: test accuracy
+for CT is virtually identical to the baseline for \emph{Adult},
+\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
+improved for \emph{Cred}. Exceptions to this general pattern are
+\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
+in test accuracy of 2, 5 and 15 percentage points, respectively. We note
+in this context, that we have not optimized our models for predictive
+performance at all and worked with very small networks. In summary, we
+find that CT can substantially improve the quality of explanations
+learned by models without generally sacrificing predictive accuracy.
+
+\begin{table}
+
+\caption{\label{tbl-main}Key plausibility and predictive performance
+metrics for all datasets. The top five rows show the percentage
+reduction in implausibility according to Equation~\ref{eq-impl-dist} for
+varying degrees of the energy penalty used for \emph{ECCo} at test time.
+The following row shows the reduction in implausibility as measured by
+Equation~\ref{eq-impl-div} and aggregated across all test specifications
+of \emph{ECCo}. The final two rows show the test accuracies for the
+model trained with CT and conventionally trained models (``vanilla'').}
+
+\centering{
+
+\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
+\input{tables/main.tex}
+\end{tabular*}
+
+}
+
+\end{table}%
+
+\subsubsection{Actionability}\label{actionability}
 
@@
 through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
-arbitrarily defined by the user, CT requires the output space
-\(\mathcal{Y}\) to be discrete. Thus, it applies to binary and
-multi-class classification but it is not well-defined for other ML tasks
-where the change in outcome with respect to a decision threshold
-\(\tau\) cannot be readily quantified. In fact, this is a common
-restriction in research on CEs and AR that predominantly focuses on
-classification models. Although other settings have attracted some
-interest (e.g., regression in
-\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021};
+arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
+discrete. Thus, it does not apply to ML tasks where the change in
+outcome cannot be readily quantified. Focus on classification models is
+a common restriction in research on CEs and AR. Other settings have
+attracted some interest (e.g., regression in
+(\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021};
 \citeproc{ref-zhao2023counterfactual}{Zhao, Broelemann, and Kasneci
-2023}), there is still no consensus on what constitutes a counterfactual
-in such settings.
+2023})), but there is little consensus how to robustly extend the notion
+of counterfactuals.
 
@@
 (\(\lambda_{\text{reg}}\)), generating a sufficiently large number of
-counterfactuals during each training epoch and including only mature
+counterfactuals during each training epoch, and including only mature
 counterfactuals for contrastive divergence.
 
-\textbf{\emph{CT is sensitive to hyperparameter selection.}} As
-discussed in Section~\ref{sec-hyperparameters}, our method benefits from
-tuning certain key hyperparameters. In this work, we have relied
+\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
+benefits from tuning certain key hyperparameters (see
+Section~\ref{sec-hyperparameters}). In this work, we have relied
 exclusively on grid search for this task. Future work on CT could
 benefit from investigating more sophisticated approaches towards
-hyperparameter tuning. Notably, counterfactual training is iterative
-which makes a variety of methods applicable, including Bayesian (e.g.,
+hyperparameter tuning. Notably, CT is iterative which makes a variety of
+methods applicable, including Bayesian (e.g.,
 \citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
@@
 
-\textbf{\emph{Immutable features may have proxies.}} In
-Proposition~\ref{prp-mtblty} we define an approach to protect immutable
-features and thus increase the actionability of the generated
-counterfactuals. However, this approach requires that model owners
-define the mutability constraints for (all) features considered by the
-model. Even with sufficient domain knowledge to protect all immutable
-features---ones that cannot be changed at all and ones that cannot be
-reasonably expected to change---there may exist proxies that are
-theoretically mutable (and hence should not be protected) but preserve
-enough information about the principals to counteract the protections.
-As one example, consider the Adult dataset used in our experiments where
-the mutable education status is a proxy for the immutable age, in that
-the attainment of degrees is correlated with age. Delineating
-actionability is a major undecided challenge in the AR literature (see,
-e.g.,
+\textbf{\emph{Immutable features may have proxies.}} We propose an
+approach to protect immutable features and thus increase the
+actionability of the generated CEs. However, it requires that model
+owners define the mutability constraints for (all) features considered
+by the model. Even with sufficient domain knowledge to protect all
+immutable features, there may exist proxies that are theoretically
+mutable (and hence should not be protected) but preserve enough
+information about the principals to hinder the protections. As an
+example, consider the Adult dataset used in our experiments where the
+mutable education status is a proxy for the immutable age, in that the
+attainment of degrees is correlated with age. Delineating actionability
+is a major undecided challenge in the AR literature (see, e.g.,
 \citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
@@
 
-\textbf{\emph{Interventions on features may have downstream impacts on
-fairness.}}\\
-- \href{https://arxiv.org/abs/2401.16088}{Equality of opportunity}?\\
-- \href{https://proceedings.mlr.press/v202/gao23d/gao23d.pdf}{Social
-segregation}?\\
-- Supporting AR within one context/system may still unfairly target
-recourse-affected individuals within other contexts/systems? In Example
-3.1 younger individuals gain access to a loan but their age could still
-lead to, e.g., enforcing stronger supervisory mechanisms?
+\textbf{\emph{Interventions on features may impact fairness
+downstream.}} Related to the point above, we provide a tool that allows
+practitioners to modify the sensitivity of a model with respect to
+certain features, which may have implication for the fair and equitable
+treatment of individuals subject to automated decisions. As protecting a
+set of features leads the model to assign higher relative importance to
+unprotected features, model owners could misuse our solution by
+enforcing explanations based on features that are more difficult to
+modify by some (group of) individuals. For example, consider again the
+Adult dataset where features such as workclass or education may be more
+difficult to change for underpriviledged groups. When applied
+irresponsibly, counterfactual training could result in an unfairly
+assigned burden of recourse (e.g.,
+\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
+threatening the equality of opportunity in the system
+(\citeproc{ref-bell2024fairness}{Bell et al. 2024}) and potentially
+reinforcing social segregation (\citeproc{ref-gao2023impact}{Gao and
+Lakkaraju 2023}). Still, as the referenced publications indicate, such
+phenomena are not specific to CT; all types of ADM solutions without
+strong external protections have been recognized to promote harmful
+power dynamics (\citeproc{ref-maas2023machine}{Maas 2023}).
 
 \section{Conclusion}\label{sec-conclusion}
+
+State-of-the-art machine learning models are prone to learning complex
+representations that cannot be interpreted by humans. Although post-hoc
+explainability approaches have attracted major research interest, these
+cannot guarantee that the explanations agree with the opaque model's
+learned representation of data. As a step towards addressing this
+challenge, we introduced counterfactual training, a novel training
+regime that incentivizes highly-explainable models. Our approach leads
+to explanations that are both plausible---compliant with the underlying
+data-generating process---and actionable---compliant with user-specified
+mutability constraints---and thus meaningful to their recipients.
+Through extensive experiments we demonstrate that counterfactual
+training satisfies its objectives while preserving the predictive
+performance of the trained models. We also find that our approach can be
+used to fine-tune conventionally-trained models and achieve similar
+gains in explainability. Finally, this work showcases that it is
+practical to improve models \emph{and} their explanations at the same
+time.
 
@@
 
+\bibitem[\citeproctext]{ref-bell2024fairness}
+Bell, Andrew, Joao Fonseca, Carlo Abrate, Francesco Bonchi, and Julia
+Stoyanovich. 2024. {``Fairness in Algorithmic Recourse Through the Lens
+of Substantive Equality of Opportunity.''}
+\url{https://arxiv.org/abs/2401.16088}.
+
 \bibitem[\citeproctext]{ref-bezanson2017julia}
@@
 {Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.
+
+\bibitem[\citeproctext]{ref-sharma2020certifai}
+Sharma, Shubham, Jette Henderson, and Joydeep Ghosh. 2020. {``CERTIFAI:
+A Common Framework to Provide Explanations and Analyse the Fairness and
+Robustness of Black-Box Models.''} In \emph{Proceedings of the AAAI/ACM
+Conference on AI, Ethics, and Society}, 166--72. AIES '20. New York, NY,
+USA: Association for Computing Machinery.
+\url{https://doi.org/10.1145/3375627.3375812}.
 
@@
 empirically in Section~\ref{sec-experiments}. Below we derive the
-analytical results in Proposition~\ref{prp-mtblty}.
+analytical results in Prp.\textasciitilde{}\ref{prp-mtblty}.
 
@@
 
-\begin{longtable}{ccccccc}
+\begin{table}
 
@@
 
-\tabularnewline
-
+\centering{
+
+  \begin{tabular}{cccccccc}
     \toprule
-  \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Data} & \textbf{No. Train} & \textbf{No. Test} & \textbf{Batchsize} & \textbf{Decision Threshold} & \textbf{No. Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{7}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  Adult & $2.6 \cdot 10^{4}$ & $5.01 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.75 & $5 \cdot 10^{3}$ & 0.25 \\
-  California Housing & $1.65 \cdot 10^{4}$ & $3.1 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
-  Circles & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.5 & $1 \cdot 10^{3}$ & 0.5 \\
-  Credit & $1.06 \cdot 10^{4}$ & $1.92 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
-  GMSC & $1.34 \cdot 10^{4}$ & $2.47 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & 0.5 & $5 \cdot 10^{3}$ & 0.5 \\
-  Linearly Separable & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.5 & $1 \cdot 10^{3}$ & 0.25 \\
-  Moons & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.9 & $1 \cdot 10^{3}$ & 0.25 \\
-  Overlapping & $3.6 \cdot 10^{3}$ & 600 & 30 & 0.5 & $1 \cdot 10^{3}$ & 0.25 \\\bottomrule
-
-\end{longtable}
+    \textbf{Data} & \textbf{No.
+Train} & \textbf{No.
+Test} & \textbf{Batchsize} & \textbf{Domain} & \textbf{Decision Threshold} & \textbf{No.
+Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
+    Adult & $2.6 \cdot 10^{4}$ & $5.01 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & none & 0.75 & $5 \cdot 10^{3}$ & 0.25 \\
+    CH & $1.65 \cdot 10^{4}$ & $3.1 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & none & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
+    Circ & $3.6 \cdot 10^{3}$ & 600 & 30 & none & 0.5 & $1 \cdot 10^{3}$ & 0.5 \\
+    Cred & $1.06 \cdot 10^{4}$ & $1.92 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & none & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
+    GMSC & $1.34 \cdot 10^{4}$ & $2.47 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & none & 0.5 & $5 \cdot 10^{3}$ & 0.5 \\
+    LS & $3.6 \cdot 10^{3}$ & 600 & 30 & none & 0.5 & $1 \cdot 10^{3}$ & 0.01 \\
+    MNIST & $1.1 \cdot 10^{4}$ & $2 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & (-1.0, 1.0) & 0.5 & $5 \cdot 10^{3}$ & 0.01 \\
+    Moon & $3.6 \cdot 10^{3}$ & 600 & 30 & none & 0.9 & $1 \cdot 10^{3}$ & 0.25 \\
+    OL & $3.6 \cdot 10^{3}$ & 600 & 30 & none & 0.5 & $1 \cdot 10^{3}$ & 0.25 \\\bottomrule
+  \end{tabular}
+
+}
+
+\end{table}%
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
   \item
-    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 5.0,\ 10.0,\ 20.0}
+    \(\lambda_{\text{egy}}\): \texttt{0.1,\ 0.5,\ 5.0,\ 10.0,\ 20.0}
   \item
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
   \item
-    \(\lambda_{\text{energy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
+    \(\lambda_{\text{egy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
   \end{itemize}
@@
   \endlastfoot
-  Circles & Accuracy & Full & 0.997 & 0.00309 \\
-  Circles & Accuracy & Vanilla & 0.998 & 0.000557 \\
-  Circles & F1-score & Full & 0.997 & 0.00309 \\
-  Circles & F1-score & Vanilla & 0.998 & 0.000558 \\
-  Linearly Separable & Accuracy & Full & 0.999 & 0.00201 \\
-  Linearly Separable & Accuracy & Vanilla & 1 & 0 \\
-  Linearly Separable & F1-score & Full & 0.999 & 0.00201 \\
-  Linearly Separable & F1-score & Vanilla & 1 & 0 \\
-  Moons & Accuracy & Full & 0.999 & 0.000696 \\
-  Moons & Accuracy & Vanilla & 1 & 0.00111 \\
-  Moons & F1-score & Full & 0.999 & 0.000696 \\
-  Moons & F1-score & Vanilla & 1 & 0.00111 \\
-  Overlapping & Accuracy & Full & 0.915 & 0.00477 \\
-  Overlapping & Accuracy & Vanilla & 0.917 & 0.00123 \\
-  Overlapping & F1-score & Full & 0.915 & 0.00478 \\
-  Overlapping & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule
+  Circ & Accuracy & Full & 0.997 & 0.00309 \\
+  Circ & Accuracy & Vanilla & 0.998 & 0.000557 \\
+  Circ & F1-score & Full & 0.997 & 0.00309 \\
+  Circ & F1-score & Vanilla & 0.998 & 0.000558 \\
+  LS & Accuracy & Full & 0.999 & 0.00201 \\
+  LS & Accuracy & Vanilla & 1 & 0 \\
+  LS & F1-score & Full & 0.999 & 0.00201 \\
+  LS & F1-score & Vanilla & 1 & 0 \\
+  Moon & Accuracy & Full & 0.999 & 0.000696 \\
+  Moon & Accuracy & Vanilla & 1 & 0.00111 \\
+  Moon & F1-score & Full & 0.999 & 0.000696 \\
+  Moon & F1-score & Vanilla & 1 & 0.00111 \\
+  OL & Accuracy & Full & 0.915 & 0.00477 \\
+  OL & Accuracy & Vanilla & 0.917 & 0.00123 \\
+  OL & F1-score & Full & 0.915 & 0.00478 \\
+  OL & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule
 
@@
 
-\subsubsection{Plausibility}\label{plausibility-2}
+\subsubsection{Plausibility}\label{plausibility-1}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
   \endlastfoot
-  Circles & Accuracy & Full & 0.994 & 0.0144 \\
-  Circles & Accuracy & Vanilla & 0.998 & 0.000875 \\
-  Circles & F1-score & Full & 0.994 & 0.0145 \\
-  Circles & F1-score & Vanilla & 0.998 & 0.000875 \\
-  Linearly Separable & Accuracy & Full & 0.998 & 0.00772 \\
-  Linearly Separable & Accuracy & Vanilla & 1 & 0 \\
-  Linearly Separable & F1-score & Full & 0.998 & 0.00773 \\
-  Linearly Separable & F1-score & Vanilla & 1 & 0 \\
-  Moons & Accuracy & Full & 0.987 & 0.0351 \\
-  Moons & Accuracy & Vanilla & 0.998 & 0.0101 \\
-  Moons & F1-score & Full & 0.987 & 0.0352 \\
-  Moons & F1-score & Vanilla & 0.998 & 0.0102 \\
-  Overlapping & Accuracy & Full & 0.911 & 0.0217 \\
-  Overlapping & Accuracy & Vanilla & 0.916 & 0.00236 \\
-  Overlapping & F1-score & Full & 0.911 & 0.0219 \\
-  Overlapping & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule
+  Circ & Accuracy & Full & 0.994 & 0.0144 \\
+  Circ & Accuracy & Vanilla & 0.998 & 0.000875 \\
+  Circ & F1-score & Full & 0.994 & 0.0145 \\
+  Circ & F1-score & Vanilla & 0.998 & 0.000875 \\
+  LS & Accuracy & Full & 0.998 & 0.00772 \\
+  LS & Accuracy & Vanilla & 1 & 0 \\
+  LS & F1-score & Full & 0.998 & 0.00773 \\
+  LS & F1-score & Vanilla & 1 & 0 \\
+  Moon & Accuracy & Full & 0.987 & 0.0351 \\
+  Moon & Accuracy & Vanilla & 0.998 & 0.0101 \\
+  Moon & F1-score & Full & 0.987 & 0.0352 \\
+  Moon & F1-score & Vanilla & 0.998 & 0.0102 \\
+  OL & Accuracy & Full & 0.911 & 0.0217 \\
+  OL & Accuracy & Vanilla & 0.916 & 0.00236 \\
+  OL & F1-score & Full & 0.911 & 0.0219 \\
+  OL & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule
 
@@
 
-\subsubsection{Plausibility}\label{plausibility-3}
+\subsubsection{Plausibility}\label{plausibility-2}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
   \endlastfoot
-  Circles & Accuracy & Full & 0.995 & 0.00431 \\
-  Circles & Accuracy & Vanilla & 0.998 & 0.000566 \\
-  Circles & F1-score & Full & 0.995 & 0.00432 \\
-  Circles & F1-score & Vanilla & 0.998 & 0.000566 \\
-  Linearly Separable & Accuracy & Full & 0.999 & 0.00231 \\
-  Linearly Separable & Accuracy & Vanilla & 1 & 0 \\
-  Linearly Separable & F1-score & Full & 0.999 & 0.00231 \\
-  Linearly Separable & F1-score & Vanilla & 1 & 0 \\
-  Moons & Accuracy & Full & 0.996 & 0.0136 \\
-  Moons & Accuracy & Vanilla & 0.988 & 0.022 \\
-  Moons & F1-score & Full & 0.996 & 0.0136 \\
-  Moons & F1-score & Vanilla & 0.988 & 0.022 \\
-  Overlapping & Accuracy & Full & 0.914 & 0.00563 \\
-  Overlapping & Accuracy & Vanilla & 0.918 & 0.00116 \\
-  Overlapping & F1-score & Full & 0.914 & 0.0057 \\
-  Overlapping & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule
+  Circ & Accuracy & Full & 0.995 & 0.00431 \\
+  Circ & Accuracy & Vanilla & 0.998 & 0.000566 \\
+  Circ & F1-score & Full & 0.995 & 0.00432 \\
+  Circ & F1-score & Vanilla & 0.998 & 0.000566 \\
+  LS & Accuracy & Full & 0.999 & 0.00231 \\
+  LS & Accuracy & Vanilla & 1 & 0 \\
+  LS & F1-score & Full & 0.999 & 0.00231 \\
+  LS & F1-score & Vanilla & 1 & 0 \\
+  Moon & Accuracy & Full & 0.996 & 0.0136 \\
+  Moon & Accuracy & Vanilla & 0.988 & 0.022 \\
+  Moon & F1-score & Full & 0.996 & 0.0136 \\
+  Moon & F1-score & Vanilla & 0.988 & 0.022 \\
+  OL & Accuracy & Full & 0.914 & 0.00563 \\
+  OL & Accuracy & Vanilla & 0.918 & 0.00116 \\
+  OL & F1-score & Full & 0.914 & 0.0057 \\
+  OL & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule
 
@@
 
-\subsubsection{Plausibility}\label{plausibility-4}
+\subsubsection{Plausibility}\label{plausibility-3}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Package Version (Reproducibility)}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-warning-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Package Version (Reproducibility)}, colframe=quarto-callout-warning-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
 
-\subsubsection{Plausibility}\label{plausibility-5}
+\subsubsection{Plausibility}\label{plausibility-4}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune_lr-train}: Training Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, arc=.35mm, titlerule=0mm, bottomrule=.15mm, bottomtitle=1mm, breakable, left=2mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, colback=white, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
 
@@
 
-\subsubsection{Plausibility}\label{plausibility-6}
+\subsubsection{Plausibility}\label{plausibility-5}
 
@@
 
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/lin_sep/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-lin_sep}Average outcomes for the
+plausibility measure across key hyperparameters. Data: Linearly
+Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
 \pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
@@
 counterfactuals in each epoch. Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/lin_sep/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+
+}
+
+\caption{\label{fig-tune_lr-mat-lin_sep}Proportion of mature
+counterfactuals in each epoch. Data: Linearly Separable.}
 

### 2025-03-13 paper/sections/abstract.qmd

@@
 
-We propose a novel training regime called Counterfactual Training that leverages counterfactual explanations to increase the explanatory capacity of models.
+We propose a novel training regime termed Counterfactual Training that leverages counterfactual explanations to increase the explanatory capacity of models.
 Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output.
-To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to stakeholder requirements.
+To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the stakeholder requirements.
 Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata.
-In this work, we instead hold models directly accountable for this desired end goal: Counterfactual Training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
+In this work, we instead hold models directly accountable for the desired end goal: Counterfactual Training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
 We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

### 2025-03-13 paper/sections/conclusion.qmd

@@
 # Conclusion {#sec-conclusion}
-
-\begin{credits}
-\subsubsection{\ackname}
-
-- Aleks
-- DelftBlue
-- Julia packages: Julia paper itself, PrettyTables, Flux, Plots, AlgebraOfGraphics
-
-\subsubsection{\discintname}
-
-The authors have no competing interests to declare that are relevant to the content of this article.
-\end{credits}
+State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans.
+Although post-hoc explainability approaches have attracted major research interest, these cannot guarantee that the explanations agree with the opaque model's learned representation of data.
+As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models.
+Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients.
+Through extensive experiments we demonstrate that counterfactual training satisfies its objectives while preserving the predictive performance of the trained models.
+We also find that our approach can be used to fine-tune conventionally-trained models and achieve similar gains in explainability.
+Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

### 2025-03-13 paper/presenations/slides.qmd
some work on cutting content

@@
 ---
-title: Counterfactual Training
+title: counterfactual training
 subtitle: Update Meeting Dec 2024
@@
 
-Counterfactual Training (CT) combines ideas from Energy-Based Models and Adversarial Training:
+counterfactual training (CT) combines ideas from Energy-Based Models and Adversarial Training:
 

### 2025-03-13 paper/quarto_ecml/paper.tex
some work on cutting content

@@
 \makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
+
 \AtBeginDocument{%
@@
 \makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+
 \makeatother
@@
 \begin{abstract}
-We propose a novel training regime termed Counterfactual Training that
+We propose a novel training regime termed counterfactual training that
 leverages counterfactual explanations to increase the explanatory
@@
 these desiderata. In this work, we instead hold models directly
-accountable for the desired end goal: Counterfactual Training employs
+accountable for the desired end goal: counterfactual training employs
 counterfactuals ad-hoc during the training phase to minimize the
@@
   convention, where ``degrees of freedom'' refer to the number of
-  parameters estimated from data.} Indeed, recent work on the so called
+  parameters estimated from data.} Indeed, recent work on the so-called
 ``lottery ticket hypothesis'' suggests that modern neural networks can
@@
 
-\textbf{Specifically, we introduce counterfactual training}: a novel
-training regime explicitly meant to align learned representations with
-plausible explanations that comply with user requirements. Our
-contributions are as follows:
+\textbf{Specifically, we introduce counterfactual training (CT)}: a
+novel training regime explicitly meant to align learned representations
+with plausible explanations that comply with user requirements. The
+remainder of this paper is structured as follows. Section~\ref{sec-lit}
+presents related work, focusing in particular on the link between
+adversarial examples and counterfactual explanations. Then follow our
+main contributions:
 
-\begin{itemize}
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \tightlist
 \item
-  We present a novel methodological framework that leverages adversarial
-  examples and faithful counterfactual explanations during the training
-  phase to improve the explanatory capacity and robustness of machine
-  learning models (Section~\ref{sec-method}).\\
+  In Section~\ref{sec-method}, we introduce our methodological framework
+  and show theoretically that it can be used to enforce global
+  actionability constraints.\\
 \item
-  We propose a method to enforce global actionability constraints by
-  preventing models from assigning importance to immutable features,
-  i.e., ones over which decision subjects have no control
-  (Section~\ref{sec-method}).\\
-\item
-  Through extensive experiments we demonstrate that counterfactual
-  training promotes explainability while preserving high predictive
-  performance. We run ablation studies and grid searches to understand
-  how the underlying model components and hyperparameters affect
-  outcomes. (Section~\ref{sec-experiments}).
-\end{itemize}
+  Through extensive experiments we demonstrate that CT substantially
+  improves explainability without sacrificing predictive performance
+  (Section~\ref{sec-experiments}).
+\end{enumerate}
 
-Despite some limitations discussed in Section~\ref{sec-discussion}, we
-conclude in Section~\ref{sec-conclusion} that counterfactual training
-provides a useful framework for researchers and practitioners interested
-in making opaque models more trustworthy. We also believe that this work
-serves as an opportunity for XAI researchers to re-evaluate the trend of
-improving XAI tools without improving the underlying models.
+We discuss future research and challenges in
+Section~\ref{sec-discussion} and conclude in
+Section~\ref{sec-conclusion} that CT is a promising new approach towards
+making opaque models more trustworthy.
 
@@
 aligning models with plausible explanations and has inspired our
-proposed counterfactual training objective, as we explain in
-Section~\ref{sec-method}.
+proposed CT objective, as we explain in Section~\ref{sec-method}.
 
@@
 
-Counterfactual training (CT) combines ideas from adversarial training,
+Counterfactual training combines ideas from adversarial training,
 energy-based modelling and counterfactuals explanations with the
@@
 \begin{split}
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
 \end{split}
@@
 \(\text{div}(\cdot)\), between mature counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
-adversarial loss, \(\text{advloss}(.)\), with respect to nascent
-counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
-\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
-regularizes the magnitude of the energy terms involved in
-\(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off between the
-components can be governed by adjusting the strengths of the penalties
-\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
+\(\mathbf{x}_T^\prime\) and observed samples
+\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
+class \(y^+\), and, (2) the adversarial loss, \(\text{advloss}(.)\),
+with respect to nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\).
+Finally, \(\text{ridge}(\cdot)\) denotes a Ridge penalty
+(\(\ell_2\)-norm) that regularizes the magnitude of the energy terms
+involved in \(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off
+between the components can be governed by adjusting the strengths of the
+penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
 \(\lambda_{\text{reg}}\).
@@
 \begin{equation}\phantomsection\label{eq-div}{
-\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
+\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
 }\end{equation}
@@
 where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
-\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
+\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
 where \(y^+\) denotes the index of the randomly drawn target class,
-\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed sample from
-target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\). Conditional on the
-target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes a mature
-counterfactual for a randomly sampled factual from a non-target class
-generated with a gradient-based CE generator for up to \(T\) iterations.
-Mature counterfactuals are ones that have either reached convergence
-wrt. the decision threshold \(\tau\) or exhausted \(T\).
+\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
+\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
+sampled factual from a non-target class generated with a gradient-based
+CE generator for up to \(T\) iterations. Mature counterfactuals are ones
+that have either reached convergence wrt. the decision threshold
+\(\tau\) or exhausted \(T\).
 
@@
 any mutability-constrained feature \(d\) this can be achieved by
-enforcing \(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever
+enforcing \(\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0\) whenever
 perturbing \(\mathbf{x}^\prime[d]\) in the direction of
-\(\mathbf{x}[d]\) would violate mutability constraints. Specifically, we
-set \(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:
+\(\mathbf{x}^+[d]\) would violate mutability constraints. Specifically,
+we set \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) if:
 
@@
 \item
-  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\), but feature \(d\) can
-  only be decreased in practice.
+  We have \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but feature \(d\)
+  can only be decreased in practice.
 \item
-  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\), but feature \(d\) can
-  only be increased in practice.
+  We have \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but feature \(d\)
+  can only be increased in practice.
 \end{enumerate}
@@
 From a Bayesian perspective, setting
-\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
-a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
-Intuitively, we think of this simply in terms ignoring implausibility
-costs with respect to immutable features, which effectively forces the
-model to instead seek plausibility with respect to the remaining
-features. This in turn results in lower overall sensitivity to immutable
-features, which we demonstrate empirically for different classifiers in
-Section~\ref{sec-experiments}. Under certain conditions, this results
-holds theoretically:\footnote{For the proof, see the supplementary
-  appendix.}
+\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
+assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
+feature \(d\). Intuitively, we think of this simply in terms ignoring
+implausibility costs with respect to immutable features, which
+effectively forces the model to instead seek plausibility with respect
+to the remaining features. This in turn results in lower overall
+sensitivity to immutable features, which we demonstrate empirically for
+different classifiers in Section~\ref{sec-experiments}. Under certain
+conditions, this results holds theoretically:\footnote{For the proof,
+  see the supplementary appendix.}
 
@@
 point where they are clearly distinguishable from the ground-truth data.
-In panel (b), we have trained our model with counterfactual training,
-once again without any mutability constraints. We observe that the
-counterfactuals are highly plausible, meeting the first objective of
-Def. \ref{def-explainability}.
+In panel (b), we have trained our model with CT, once again without any
+mutability constraints. We observe that the counterfactuals are highly
+plausible, meeting the first objective of Def. \ref{def-explainability}.
 
@@
 counterfactual paths are shorter on average in panel (d), where we have
-used counterfactual training and protected the immutable feature as
-described in Section~\ref{sec-constraints}. We observe that due to the
-classifier's lower sensitivity to \emph{age}, recourse recommendations
-with respect to \emph{debt} are much more homogenous and do not
-disproportionately punish younger individuals. The counterfactuals are
-also plausible with respect to the mutable feature. Thus, we consider
-the model in panel (d) as the most explainable according to Def.
-\ref{def-explainability}.
+used CT and protected the immutable feature as described in
+Section~\ref{sec-constraints}. We observe that due to the classifier's
+lower sensitivity to \emph{age}, recourse recommendations with respect
+to \emph{debt} are much more homogenous and do not disproportionately
+punish younger individuals. The counterfactuals are also plausible with
+respect to the mutable feature. Thus, we consider the model in panel (d)
+as the most explainable according to Def. \ref{def-explainability}.
 
@@
 
-We first address the direct extensions of the counterfactual training
-approach in Section~\ref{sec-approach}. Then, we look at its limitations
-and challenges in Section~\ref{sec-limitations}.
+We first address the direct extensions of CT in
+Section~\ref{sec-approach}. Then, we look at its limitations and
+challenges in Section~\ref{sec-limitations}.
 
-\subsection{Future research}\label{sec-approach}
+\subsection{Future Research}\label{sec-approach}
 
@@
 
-\subsection{Limitations and challenges}\label{sec-limitations}
+\subsection{Limitations and Challenges}\label{sec-limitations}
 
@@
 difficult to change for underpriviledged groups. When applied
-irresponsibly, counterfactual training could result in an unfairly
-assigned burden of recourse \cite{sharma2020certifai}, threatening the
-equality of opportunity in the system \cite{bell2024fairness} and
-potentially reinforcing social segregation \cite{gao2023impact}. Still,
-as the referenced publications indicate, such phenomena are not specific
-to CT; all types of ADM solutions without strong external protections
-have been recognized to promote harmful power dynamics
-\cite{maas2023machine}.
+irresponsibly, CT could result in an unfairly assigned burden of
+recourse \cite{sharma2020certifai}, threatening the equality of
+opportunity in the system \cite{bell2024fairness} and potentially
+reinforcing social segregation \cite{gao2023impact}. Still, as the
+referenced publications indicate, such phenomena are not specific to CT;
+all types of ADM solutions without strong external protections have been
+recognized to promote harmful power dynamics \cite{maas2023machine}.
 
@@
 mutability constraints---and thus meaningful to their recipients.
-Through extensive experiments we demonstrate that counterfactual
-training satisfies its objectives while preserving the predictive
-performance of the trained models. We also find that our approach can be
-used to fine-tune conventionally-trained models and achieve similar
-gains in explainability. Finally, this work showcases that it is
-practical to improve models \emph{and} their explanations at the same
-time.
+Through extensive experiments we demonstrate that CT satisfies its
+objectives while preserving the predictive performance of the trained
+models. We also find that our approach can be used to fine-tune
+conventionally-trained models and achieve similar gains in
+explainability. Finally, this work showcases that it is practical to
+improve models \emph{and} their explanations at the same time.
 

### 2025-03-13 paper/sections/abstract.qmd
some work on cutting content

@@
 
-We propose a novel training regime termed Counterfactual Training that leverages counterfactual explanations to increase the explanatory capacity of models.
+We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models.
 Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output.
@@
 Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata.
-In this work, we instead hold models directly accountable for the desired end goal: Counterfactual Training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
+In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
 We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

### 2025-03-13 paper/sections/conclusion.qmd
some work on cutting content

@@
 Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients.
-Through extensive experiments we demonstrate that counterfactual training satisfies its objectives while preserving the predictive performance of the trained models.
+Through extensive experiments we demonstrate that CT satisfies its objectives while preserving the predictive performance of the trained models.
 We also find that our approach can be used to fine-tune conventionally-trained models and achieve similar gains in explainability.

### 2025-03-13 paper/sections/discussion.qmd
some work on cutting content

@@
 
-We first address the direct extensions of the counterfactual training approach in @sec-approach. Then, we look at its limitations and challenges in @sec-limitations.
+We first address the direct extensions of CT in @sec-approach. Then, we look at its limitations and challenges in @sec-limitations.
 
-## Future research {#sec-approach}
+## Future Research {#sec-approach}
 
@@
 
-## Limitations and challenges {#sec-limitations}
+## Limitations and Challenges {#sec-limitations}
 
@@
 For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups.
-When applied irresponsibly, counterfactual training could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact].
+When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact].
 Still, as the referenced publications indicate, such phenomena are not specific to CT; all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].

### 2025-03-13 paper/sections/intro.qmd
some work on cutting content

@@
 
-Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success].
+Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so-called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success].
 Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data.
@@
 
-**Specifically, we introduce counterfactual training**: a novel training regime explicitly meant to align learned representations with plausible explanations that comply with user requirements.
-Our contributions are as follows:
+**Specifically, we introduce counterfactual training (CT)**: a novel training regime explicitly meant to align learned representations with plausible explanations that comply with user requirements.
+The remainder of this paper is structured as follows.
+@sec-lit presents related work, focusing in particular on the link between adversarial examples and counterfactual explanations.
+Then follow our main contributions:
 
-- We present a novel methodological framework that leverages adversarial examples and faithful counterfactual explanations during the training phase to improve the explanatory capacity and robustness of machine learning models (@sec-method).
-- We propose a method to enforce global actionability constraints by preventing models from assigning importance to immutable features, i.e., ones over which decision subjects have no control (@sec-method).
-- Through extensive experiments we demonstrate that counterfactual training promotes explainability while preserving high predictive performance.
-We run ablation studies and grid searches to understand how the underlying model components and hyperparameters affect outcomes.
-(@sec-experiments).
+1. In @sec-method, we introduce our methodological framework and show theoretically that it can be used to enforce global actionability constraints.  
+2. Through extensive experiments we demonstrate that CT substantially improves explainability without sacrificing predictive performance (@sec-experiments).
 
-Despite some limitations discussed in @sec-discussion, we conclude in @sec-conclusion that counterfactual training provides a useful framework for researchers and practitioners interested in making opaque models more trustworthy.
-We also believe that this work serves as an opportunity for XAI researchers to re-evaluate the trend of improving XAI tools without improving the underlying models.
+We discuss future research and challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising new approach towards making opaque models more trustworthy.

### 2025-03-13 paper/sections/method.qmd
some work on cutting content

@@
 
-Counterfactual training (CT) combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit goal of aligning representations with plausible explanations that comply with user requirements.
+Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit goal of aligning representations with plausible explanations that comply with user requirements.
 In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful].
@@
 The second and third terms in @eq-obj are explained in detail below.
-For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $x$ and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
+For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
 Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit].
@@
 $$
-\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
+\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
 $$ {#eq-div}
@@
 where $\mathcal{E}_\theta(\cdot)$ denotes the energy function.
-We set $\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$, and $\mathbf{x^+}$ denotes an observed sample from target domain: $\mathbf{X}^+=\{\mathbf{x}:y=y^+\}$.
+We set $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$.
 Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations.
@@
 Here it is crucial that we avoid penalizing implausibility that arises due to mutability constraints.
-For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}[d]$ would violate mutability constraints.
-Specifically, we set $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ if:
+For any mutability-constrained feature $d$ this can be achieved by enforcing $\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0$ whenever perturbing $\mathbf{x}^\prime[d]$ in the direction of $\mathbf{x}^+[d]$ would violate mutability constraints.
+Specifically, we set $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ if:
 
 1. Feature $d$ is strictly immutable in practice.
-2. We have $\mathbf{x}[d]>\mathbf{x}^\prime[d]$, but feature $d$ can only be decreased in practice.
-3. We have $\mathbf{x}[d]<\mathbf{x}^\prime[d]$, but feature $d$ can only be increased in practice.
+2. We have $\mathbf{x}^+[d]>\mathbf{x}^\prime[d]$, but feature $d$ can only be decreased in practice.
+3. We have $\mathbf{x}^+[d]<\mathbf{x}^\prime[d]$, but feature $d$ can only be increased in practice.
 
-From a Bayesian perspective, setting $\mathbf{x}[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x})$ with respect to feature $d$.
+From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$.
 Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features.
@@
 The generated counterfactuals are all valid, but not plausible: they do not comply with the distribution of the factual samples in the target class to the point where they are clearly distinguishable from the ground-truth data.
-In panel (b), we have trained our model with counterfactual training, once again without any mutability constraints.
+In panel (b), we have trained our model with CT, once again without any mutability constraints.
 We observe that the counterfactuals are highly plausible, meeting the first objective of Def.
@@
 Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants).
-By comparison, counterfactual paths are shorter on average in panel (d), where we have used counterfactual training and protected the immutable feature as described in @sec-constraints.
+By comparison, counterfactual paths are shorter on average in panel (d), where we have used CT and protected the immutable feature as described in @sec-constraints.
 We observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous and do not disproportionately punish younger individuals.

### 2025-03-13 paper/sections/other/grids.qmd
some work on cutting content

@@
 
-Overall, we find that Counterfactual Training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to ~60 percent (%) for the *Circles* data (e.g.
+Overall, we find that counterfactual training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to ~60 percent (%) for the *Circles* data (e.g.
 @fig-grid-gen_params-plaus-circles), ~25-30% for the *Moons* data (e.g.

### 2025-03-13 paper/sections/related.qmd
some work on cutting content

@@
 The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your].
-This is closely related to the idea of aligning models with plausible explanations and has inspired our proposed counterfactual training objective, as we explain in @sec-method.
+This is closely related to the idea of aligning models with plausible explanations and has inspired our proposed CT objective, as we explain in @sec-method.

### 2025-03-13 paper/quarto_ecml/paper.tex
final push

@@
 \end{split}
-}\end{equation}
-
-where \(\text{yloss}(\cdot)\) is a classification loss that induces
-discriminative performance (e.g., cross-entropy). The second and third
-terms in Equation~\ref{eq-obj} are explained in detail below. For now,
-they can be sufficiently described as inducing explainability directly
-and indirectly by penalizing: (1) the contrastive divergence,
-\(\text{div}(\cdot)\), between mature counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples
+}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
+that induces discriminative performance (e.g., cross-entropy). The
+second and third terms in Equation~\ref{eq-obj} are explained in detail
+below. For now, they can be sufficiently described as inducing
+explainability directly and indirectly by penalizing: (1) the
+contrastive divergence, \(\text{div}(\cdot)\), between mature
+counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
 \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
@@
 \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
-}\end{equation}
-
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
+}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
+function. We set
 \(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
@@
 \end{aligned}
-}\end{equation}
-
-Under this choice, we consider nascent counterfactuals
+}\end{equation} Under this choice, we consider nascent counterfactuals
 \(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
@@
 \text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-}\end{equation}
-
-and introduce a novel divergence metric,
+}\end{equation} and introduce a novel divergence metric,
 
@@
 \text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
-}\end{equation}
-
-where \(\mathbf{X}^\prime\) denotes a set of multiple counterfactuals
-and \(\text{MMD}(\cdot)\) is an unbiased estimate of the squared
-population maximum mean discrepancy \cite{gretton2012kernel}. The metric
-in Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
-are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
+}\end{equation} where \(\mathbf{X}^\prime\) denotes a set of multiple
+counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
+squared population maximum mean discrepancy \cite{gretton2012kernel}.
+The metric in Equation~\ref{eq-impl-div} is equal to zero iff the two
+distributions are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
 

### 2025-03-13 paper/sections/experiments.qmd
final push

@@
 $$ {#eq-impl-dist}
-
 and introduce a novel divergence metric,
@@
 $$ {#eq-impl-div}
-
 where $\mathbf{X}^\prime$ denotes a set of multiple counterfactuals and $\text{MMD}(\cdot)$ is an unbiased estimate of the squared population maximum mean discrepancy [@gretton2012kernel].

### 2025-03-13 paper/sections/method.qmd
final push

@@
 $$ {#eq-obj}
-
 where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy).
@@
 $$ {#eq-div}
-
 where $\mathcal{E}_\theta(\cdot)$ denotes the energy function.
@@
 $$ {#eq-adv}
-
 Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$.

### 2025-03-13 paper/preprint/paper.tex
rerendered full paper

@@
 \newcommand{\runninghead}{A Preprint }
-\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
+\renewcommand{\runninghead}{counterfactual training (A Preprint) }
 \title{Counterfactual Training: Teaching Models Plausible and Actionable
@@
   to the number of parameters estimated from data.} Indeed, recent work
-on the so called ``lottery ticket hypothesis'' suggests that modern
+on the so-called ``lottery ticket hypothesis'' suggests that modern
 neural networks can be pruned by up to 90\% while preserving their
@@
 
-\textbf{Specifically, we introduce counterfactual training}: a novel
-training regime explicitly meant to align learned representations with
-plausible explanations that comply with user requirements. Our
-contributions are as follows:
-
-\begin{itemize}
+\textbf{Specifically, we introduce counterfactual training (CT)}: a
+novel training regime explicitly meant to align learned representations
+with plausible explanations that comply with user requirements. The
+remainder of this paper is structured as follows. Section~\ref{sec-lit}
+presents related work, focusing in particular on the link between
+adversarial examples and counterfactual explanations. Then follow our
+main contributions:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
 \tightlist
 \item
-  We present a novel methodological framework that leverages adversarial
-  examples and faithful counterfactual explanations during the training
-  phase to improve the explanatory capacity and robustness of machine
-  learning models (Section~\ref{sec-method}).\\
-\item
-  We propose a method to enforce global actionability constraints by
-  preventing models from assigning importance to immutable features,
-  i.e., ones over which decision subjects have no control
-  (Section~\ref{sec-method}).\\
-\item
-  Through extensive experiments we demonstrate that counterfactual
-  training promotes explainability while preserving high predictive
-  performance. We run ablation studies and grid searches to understand
-  how the underlying model components and hyperparameters affect
-  outcomes. (Section~\ref{sec-experiments}).
-\end{itemize}
-
-Despite some limitations discussed in Section~\ref{sec-discussion}, we
-conclude in Section~\ref{sec-conclusion} that counterfactual training
-provides a useful framework for researchers and practitioners interested
-in making opaque models more trustworthy. We also believe that this work
-serves as an opportunity for XAI researchers to re-evaluate the trend of
-improving XAI tools without improving the underlying models.
+  In Section~\ref{sec-method}, we introduce our methodological framework
+  and show theoretically that it can be used to enforce global
+  actionability constraints.\\
+\item
+  Through extensive experiments we demonstrate that CT substantially
+  improves explainability without sacrificing predictive performance
+  (Section~\ref{sec-experiments}).
+\end{enumerate}
+
+We discuss future research and challenges in
+Section~\ref{sec-discussion} and conclude in
+Section~\ref{sec-conclusion} that CT is a promising new approach towards
+making opaque models more trustworthy.
 
@@
 is closely related to the idea of aligning models with plausible
-explanations and has inspired our proposed counterfactual training
-objective, as we explain in Section~\ref{sec-method}.
+explanations and has inspired our proposed CT objective, as we explain
+in Section~\ref{sec-method}.
 
@@
 \begin{split}
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
 \end{split}
-}\end{equation}
-
-where \(\text{yloss}(\cdot)\) is a classification loss that induces
-discriminative performance (e.g., cross-entropy). The second and third
-terms in Equation~\ref{eq-obj} are explained in detail below. For now,
-they can be sufficiently described as inducing explainability directly
-and indirectly by penalizing: (1) the contrastive divergence,
-\(\text{div}(\cdot)\), between mature counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
-adversarial loss, \(\text{advloss}(.)\), with respect to nascent
-counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
-\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
-regularizes the magnitude of the energy terms involved in
-\(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and Mordatch
-2020}). The trade-off between the components can be governed by
+}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
+that induces discriminative performance (e.g., cross-entropy). The
+second and third terms in Equation~\ref{eq-obj} are explained in detail
+below. For now, they can be sufficiently described as inducing
+explainability directly and indirectly by penalizing: (1) the
+contrastive divergence, \(\text{div}(\cdot)\), between mature
+counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
+\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
+class \(y^+\), and, (2) the adversarial loss, \(\text{advloss}(.)\),
+with respect to nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\).
+Finally, \(\text{ridge}(\cdot)\) denotes a Ridge penalty
+(\(\ell_2\)-norm) that regularizes the magnitude of the energy terms
+involved in \(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and
+Mordatch 2020}). The trade-off between the components can be governed by
 adjusting the strengths of the penalties \(\lambda_{\text{div}}\),
@@
 \begin{equation}\phantomsection\label{eq-div}{
-\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
-}\end{equation}
-
-where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
-\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
+\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
+}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
+function. We set
+\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
 where \(y^+\) denotes the index of the randomly drawn target class,
-\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed sample from
-target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\). Conditional on the
-target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes a mature
-counterfactual for a randomly sampled factual from a non-target class
-generated with a gradient-based CE generator for up to \(T\) iterations.
-Mature counterfactuals are ones that have either reached convergence
-wrt. the decision threshold \(\tau\) or exhausted \(T\).
+\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
+\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
+sampled factual from a non-target class generated with a gradient-based
+CE generator for up to \(T\) iterations. Mature counterfactuals are ones
+that have either reached convergence wrt. the decision threshold
+\(\tau\) or exhausted \(T\).
 
@@
 \item
-  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\), but feature \(d\) can
-  only be decreased in practice.
-\item
-  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\), but feature \(d\) can
-  only be increased in practice.
+  We have \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but feature \(d\)
+  can only be decreased in practice.
+\item
+  We have \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but feature \(d\)
+  can only be increased in practice.
 \end{enumerate}
@@
 \text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
-}\end{equation}
-
-where \(\mathbf{X}^\prime\) denotes a set of multiple counterfactuals
-and \(\text{MMD}(\cdot)\) is an unbiased estimate of the squared
-population maximum mean discrepancy
+}\end{equation} where \(\mathbf{X}^\prime\) denotes a set of multiple
+counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
+squared population maximum mean discrepancy
 (\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
@@
 
-We first address the direct extensions of the counterfactual training
-approach in Section~\ref{sec-approach}. Then, we look at its limitations
-and challenges in Section~\ref{sec-limitations}.
-
-\subsection{Future research}\label{sec-approach}
+We first address the direct extensions of CT in
+Section~\ref{sec-approach}. Then, we look at its limitations and
+challenges in Section~\ref{sec-limitations}.
+
+\subsection{Future Research}\label{sec-approach}
 
@@
 difficult to change for underpriviledged groups. When applied
-irresponsibly, counterfactual training could result in an unfairly
-assigned burden of recourse (e.g.,
-\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
-threatening the equality of opportunity in the system
+irresponsibly, CT could result in an unfairly assigned burden of
+recourse (e.g., \citeproc{ref-sharma2020certifai}{Sharma, Henderson, and
+Ghosh 2020}), threatening the equality of opportunity in the system
 (\citeproc{ref-bell2024fairness}{Bell et al. 2024}) and potentially
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-Overall, we find that Counterfactual Training (CT) achieves it key
+Overall, we find that counterfactual training (CT) achieves it key
 objectives consistently across all hyperparameter settings and also
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Package Version (Reproducibility)}, colframe=quarto-callout-warning-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, colbacktitle=quarto-callout-warning-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Package Version (Reproducibility)}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, coltitle=black, breakable, rightrule=.15mm, opacityback=0, toptitle=1mm, left=2mm, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, titlerule=0mm, arc=.35mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
 

### 2025-03-13 paper/quarto_ecml/paper.tex
Fix references

@@
 plausible counterfactuals using a surrogate Variational Autoencoder
-(VAE), and \emph{ECCo}---the generator of \cite{altmeyer2023faithful}
+(VAE), and \emph{ECCo}---the generator of \cite{altmeyer2024faithful}
 but without the conformal prediction component---as a method that
@@
 underlying counterfactual explainers. For example,
-\cite{altmeyer2023faithful} recognize this to be a challenge for
+\cite{altmeyer2024faithful} recognize this to be a challenge for
 \emph{ECCCo} and so it may have downstream impacts on our proposed

### 2025-03-13 paper/sections/experiments.qmd
Fix references

@@
 
-We run the experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2023faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the CEs.
+We run the experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the CEs.
 

### 2025-03-13 paper/quarto_ecml/paper.tex
Down to 16 pages without major cutting of content

@@
 serious scrutiny in recent years. At the very cusp of the deep learning
-revolution, \cite{szegedy2013intriguing} showed that artificial neural
-networks (ANN) are sensitive to adversarial examples: perturbed versions
-of data instances that yield vastly different model predictions despite
-being ``imperceptible'' in that they are semantically indifferent from
-their factual counterparts. Even though some partially effective
-mitigation strategies have been proposed---most notably
-\textbf{adversarial training} \cite{goodfellow2014explaining}---truly
-robust deep learning (DL) remains unattainable even for models that are
-considered shallow by today's standards \cite{kolter2023keynote}.
+(DL) revolution, \cite{szegedy2013intriguing} showed that artificial
+neural networks (ANN) are sensitive to adversarial examples (AEs):
+perturbed versions of data instances that yield vastly different model
+predictions despite being ``imperceptible'' in that they are
+semantically indifferent from their factual counterparts. Even though
+some partially effective mitigation strategies have been proposed---most
+notably \textbf{adversarial training}
+\cite{goodfellow2014explaining}---truly robust deep learning remains
+unattainable even for models that are considered ``shallow'' by today's
+standards \cite{kolter2023keynote}.
 
@@
 be pruned by up to 90\% while preserving their predictive performance
-\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
-Similarly, \cite{zhang2021understanding} showed that state-of-the-art
-neural networks are so expressive that they can fit randomly labeled
-data. Thus, looking at the predictive performance alone, the solutions
-may seem to provide compelling explanations for the data, when in fact
-they are based on purely associative, semantically meaningless patterns.
-This poses two related challenges. Firstly, there is no dependable way
-to verify if such complex representations correspond to meaningful and
+\cite{frankle2018lottery}. Similarly, \cite{zhang2021understanding}
+showed that state-of-the-art neural networks are expressive enough to
+fit randomly labeled data. Thus, looking at the predictive performance
+alone, the solutions may seem to provide compelling explanations for the
+data, when in fact they are based on purely associative, semantically
+meaningless patterns. This poses two challenges. Firstly, there is no
+dependable way to verify if representations correspond to meaningful,
 plausible explanations. Secondly, even if we could resolve the first
@@
 
-\textbf{Specifically, we introduce counterfactual training (CT)}: a
-novel training regime explicitly meant to align learned representations
-with plausible explanations that comply with user requirements. The
-remainder of this paper is structured as follows. Section~\ref{sec-lit}
-presents related work, focusing in particular on the link between
-adversarial examples and counterfactual explanations. Then follow our
-main contributions:
+\textbf{Specifically, we propose counterfactual training (CT)}: a novel
+training regime that aligns learned representations with plausible
+explanations compliant with user requirements. The remainder of this
+paper is structured as follows. Section~\ref{sec-lit} presents related
+work, focusing on the link between adversarial examples and
+counterfactual explanations. Then follow our main contributions:
 
@@
 \item
-  Through extensive experiments we demonstrate that CT substantially
-  improves explainability without sacrificing predictive performance
-  (Section~\ref{sec-experiments}).
+  In Section~\ref{sec-experiments}, through extensive experiments we
+  demonstrate that CT substantially improves explainability without
+  sacrificing predictive performance.
 \end{enumerate}
 
-We discuss future research and challenges in
-Section~\ref{sec-discussion} and conclude in
-Section~\ref{sec-conclusion} that CT is a promising new approach towards
+We discuss the challenges in Section~\ref{sec-discussion} and conclude
+in Section~\ref{sec-conclusion} that CT is a promising approach towards
 making opaque models more trustworthy.
@@
 derived for an opaque model are also deemed plausible with respect to
-the underlying data and stakeholder requirements; the former means that
-the counterfactuals should comply with the distribution of the factual
-data, the latter means that they should respect arbitrary (global)
-actionability constraints. To make the desiderata for our framework more
-concrete, we follow \cite{augustin2020adversarial} in tying the concept
-of explainability to the quality of counterfactual explanations that we
-can generate for a given model. The authors show that CEs---understood
-here as minimal input perturbations that yield some desired model
-prediction---are generally more meaningful if the underlying model is
-more robust to adversarial examples. We can make intuitive sense of this
-finding when looking at adversarial training (AT) through the lens of
-representation learning with high degrees of freedom. As argued before,
-learned representations may be sensitive to producing implausible
-explanations and mispredicting for worst-case counterfactuals (i.e.,
-adversarial examples). Thus, by inducing models to ``unlearn''
+the underlying data and (global) actionability constraints. To make the
+desiderata for our framework more concrete, we follow
+\cite{augustin2020adversarial} in tying the concept of explainability to
+the quality of CEs that can be generated for a given model. The authors
+show that CEs---understood as minimal input perturbations that yield
+some desired model prediction---tend to be more meaningful if the
+underlying model is more robust to adversarial examples. We can make
+intuitive sense of this finding when looking at adversarial training
+(AT) through the lens of representation learning with high degrees of
+freedom. As argued before, learned representations may be sensitive to
+producing implausible explanations and mispredicting for worst-case
+counterfactuals (i.e., AEs). Thus, by inducing models to ``unlearn''
 susceptiblity to such examples, AT can effectively remove implausible
@@
 There have also been several attempts at formalizing the relationship
-between counterfactual explanations and adversarial examples (AE).
-Pointing to clear similarities in how CEs and AEs are generated,
+between counterfactual explanations and adversarial examples. Pointing
+to clear similarities in how CEs and AEs are generated,
 \cite{freiesleben2022intriguing} makes the case for jointly studying the
@@
 
-We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. Formally, we define these properties as follows,
+We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. We define these properties as follows:
 
@@
 \end{enumerate}
-
 where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
@@
 that induces discriminative performance (e.g., cross-entropy). The
-second and third terms in Equation~\ref{eq-obj} are explained in detail
-below. For now, they can be sufficiently described as inducing
-explainability directly and indirectly by penalizing: (1) the
-contrastive divergence, \(\text{div}(\cdot)\), between mature
-counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
-\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
-class \(y^+\), and, (2) the adversarial loss, \(\text{advloss}(.)\),
-with respect to nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\).
-Finally, \(\text{ridge}(\cdot)\) denotes a Ridge penalty
-(\(\ell_2\)-norm) that regularizes the magnitude of the energy terms
-involved in \(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off
-between the components can be governed by adjusting the strengths of the
-penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
+second and third terms are explained in detail below. For now, they can
+be summarized as inducing explainability directly and indirectly by
+penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
+between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
+samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
+target class \(y^+\), and, (2) the adversarial loss,
+\(\text{advloss}(.)\), with respect to nascent counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
+denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
+of the energy terms involved in \(\text{div}(\cdot)\)
+\cite{du2019implicit}. The trade-off between the components can governed
+through penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
 \(\lambda_{\text{reg}}\).
@@
 
-It is worth highlighting that Prp.\textasciitilde{}\ref{prp-mtblty}
-assumes independence of features. This raises a valid concern about the
-effect of protecting immutable features in the presence of proxies that
-remain unprotected. We address this in Section~\ref{sec-discussion}.
+It is worth highlighting that \mbox{Prp. \ref{prp-mtblty}} assumes
+independence of features. This raises a valid concern about the effect
+of protecting immutable features in the presence of proxies that remain
+unprotected. We address this in Section~\ref{sec-discussion}.
 
@@
 \item
-  To what extent does our proposed counterfactual training objective
-  (Equation~\ref{eq-obj}) induce models to learn plausible explanations?
+  To what extent does our proposed counterfactual training objective in
+  Equation~\ref{eq-obj} induce models to learn plausible explanations?
 \item
-  To what extent does our proposed counterfactual training objective
-  (Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
+  To what extent does our proposed counterfactual training objective in
+  Equation~\ref{eq-obj} yield more favorable algorithmic recourse
   outcomes in the presence of actionability constraints?
@@
 \emph{OL} the reduction is negative, consistent with the distance-based
-metric. The only dataset, for which our proposed metric disagrees with
-the distance-based metric is \emph{MNIST}.
+metric. \emph{MNIST} is the only dataset for which the two metrics
+disagree.
 
@@
 find that CT can substantially improve the quality of explanations
-learned by models without generally sacrificing predictive accuracy.
+learned by models without sacrificing predictive accuracy.
 
@@
 
-We extensively test the impact of three types of hyperparameters on the
-proposed training regime. Our complete results are available in the
-technical appendix; this section focuses on the main findings.
+We test in-depth the impact of three types of hyperparameters; our
+complete results are in the appendix.
 
@@
 that a higher number of training epochs is beneficial. Interestingly, we
-find that it is not necessary to employ CT during the entire training
-phase to achieve the desired improvements in explainability. When
-training models conventionally during the first 50\% of epochs before
-switching to CT for the next 50\% of epochs, we observed positive
-results. Put differently, CT may be a way to improve the explainability
-of models in a fine-tuning manner.
+observed desired improvements in explainability when CT was combined
+with conventional training and applied only for the final 50\% of epochs
+of the complete training process. Put differently, CT may be a way to
+improve the explainability of models in a fine-tuning manner.
 
@@
 
-\textbf{\emph{CT is subject to training instabilities.}} Joint
-energy-based models are susceptible to instabilities during training
-\cite{grathwohl2020your} and even though we depart from the SGLD-based
-sampling, we still encounter major variability in the outcomes. CT is
-exposed to two potential sources of instabilities: (1) the energy-based
-contrastive divergence term in Equation~\ref{eq-div}, and (2) the
-underlying counterfactual explainers. For example,
-\cite{altmeyer2024faithful} recognize this to be a challenge for
-\emph{ECCCo} and so it may have downstream impacts on our proposed
-method. Still, we find that training instabilities can be successfully
-mitigated by regularizing energy (\(\lambda_{\text{reg}}\)), generating
-a sufficiently large number of counterfactuals during each training
-epoch, and including only mature counterfactuals for contrastive
-divergence.
+\textbf{\emph{CT is subject to training instabilities.}} JEMs are
+susceptible to instabilities during training \cite{grathwohl2020your}
+and even though we depart from the SGLD-based sampling, we still
+encounter major variability in the outcomes. CT is exposed to two
+potential sources of instabilities: (1) the energy-based contrastive
+divergence term in Equation~\ref{eq-div}, and (2) the underlying
+counterfactual explainers. Still, we find that training instabilities
+can be successfully mitigated by regularizing energy
+(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
+during each training epoch, and including only mature counterfactuals
+for contrastive divergence.
 
 \textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
-benefits from tuning certain key hyperparameters (see
+benefits from the tuning of certain key hyperparameters (see
 Section~\ref{sec-hyperparameters}). In this work, we have relied
@@
 
-\subsection{Limitations and Challenges}\label{sec-limitations}
+\subsection{Current Limitations}\label{sec-limitations}
 
-\textbf{\emph{CT increases the training time of models.}} Counterfactual
-training promotes explainability through CEs and robustness through AEs
-at the cost of longer training times compared to conventional training
-regimes. While higher numbers of iterations and counterfactuals per
-iteration positively impact the quality of found solutions, they also
-increase the required amount of computations. We find that relatively
-small grids with 270 settings can take almost four hours for more
-demanding datasets on a high-performance computing cluster with 34 2GB
-CPUs\footnote{See supplementary appendix for computational details.}.
-However, there are three factors that attenuate the impact of this
-limitation. First, CT provides counterfactual explanations for the
-training samples essentially for free, which may be beneficial in many
-ADM systems. Second, we find that CT can retain its value when used as a
-``fine-tuning'' training regime for conventionally-trained models.
-Third, in principle, CT yields itself to parallel execution, which we
-have leveraged for our own experiments.
+\textbf{\emph{CT increases the training time of models.}} CT can be more
+time-consuming than conventional training regimes. While higher numbers
+of CEs per iteration positively impact the quality of solutions, they
+also increase the amount of computations. Relatively small grids with
+270 settings can take almost four hours for more demanding datasets on a
+high-performance computing cluster with 34 2GB CPUs.\footnote{See
+  supplementary appendix for computational details.} Three factors
+attenuate this effect. First, CT amortizes the cost of CEs for the
+training samples. Second, we find that it can retain its value when used
+as a ``fine-tuning'' technique for conventionally-trained models. Third,
+it yields itself to parallel execution, which we have leveraged for our
+own experiments.
 
@@
 owners define the mutability constraints for (all) features considered
-by the model. Even with sufficient domain knowledge to protect all
-immutable features, there may exist proxies that are theoretically
-mutable (and hence should not be protected) but preserve enough
-information about the principals to hinder the protections. As an
-example, consider the Adult dataset used in our experiments where the
-mutable education status is a proxy for the immutable age, in that the
-attainment of degrees is correlated with age. Delineating actionability
-is a major undecided challenge in the AR literature
-\cite{venkatasubramanian2020philosophical} impacting the capacity of CT
-to increase the explainability of the model.
+by the model. Even if all immutable features are protected, there may
+exist proxies that are mutable (and hence should not be protected) but
+preserve enough information about the principals to hinder the
+protections. Delineating actionability is a major undecided challenge in
+the AR literature \cite{venkatasubramanian2020philosophical} impacting
+the capacity of CT to increase the explainability of the model.
 
-\textbf{\emph{Interventions on features may impact fairness
-downstream.}} Related to the point above, we provide a tool that allows
-practitioners to modify the sensitivity of a model with respect to
-certain features, which may have implication for the fair and equitable
-treatment of individuals subject to automated decisions. As protecting a
+\textbf{\emph{Interventions on features may impact fairness.}} We
+provide a tool that allows practitioners to modify the sensitivity of a
+model with respect to certain features, which may have implication for
+the fair and equitable treatment of decision subjects. As protecting a
 set of features leads the model to assign higher relative importance to
@@
 enforcing explanations based on features that are more difficult to
-modify by some (group of) individuals. For example, consider again the
-Adult dataset where features such as workclass or education may be more
-difficult to change for underpriviledged groups. When applied
-irresponsibly, CT could result in an unfairly assigned burden of
-recourse \cite{sharma2020certifai}, threatening the equality of
-opportunity in the system \cite{bell2024fairness} and potentially
-reinforcing social segregation \cite{gao2023impact}. Still, as the
-referenced publications indicate, such phenomena are not specific to CT;
-all types of ADM solutions without strong external protections have been
-recognized to promote harmful power dynamics \cite{maas2023machine}.
+modify by some (group of) individuals. For example, consider the Adult
+dataset used in our experiments, where \emph{workclass} or
+\emph{education} may be more difficult to change for underpriviledged
+groups. When applied irresponsibly, CT could result in an unfairly
+assigned burden of recourse \cite{sharma2020certifai}, threatening the
+equality of opportunity in the system \cite{bell2024fairness}. Still,
+these phenomena are not specific to CT.
 
@@
 State-of-the-art machine learning models are prone to learning complex
-representations that cannot be interpreted by humans. Although post-hoc
-explainability approaches have attracted major research interest, these
-cannot guarantee that the explanations agree with the opaque model's
-learned representation of data. As a step towards addressing this
-challenge, we introduced counterfactual training, a novel training
-regime that incentivizes highly-explainable models. Our approach leads
-to explanations that are both plausible---compliant with the underlying
-data-generating process---and actionable---compliant with user-specified
-mutability constraints---and thus meaningful to their recipients.
-Through extensive experiments we demonstrate that CT satisfies its
-objectives while preserving the predictive performance of the trained
-models. We also find that our approach can be used to fine-tune
-conventionally-trained models and achieve similar gains in
+representations that cannot be interpreted by humans and existing
+post-hoc explainability approaches cannot guarantee that the
+explanations agree with the model's learned representation of data. As a
+step towards addressing this challenge, we introduced counterfactual
+training, a novel training regime that incentivizes highly-explainable
+models. Our approach leads to explanations that are both
+plausible---compliant with the underlying data-generating process---and
+actionable---compliant with user-specified mutability constraints---and
+thus meaningful to their recipients. Through extensive experiments we
+demonstrate that CT satisfies its objectives while preserving the
+predictive performance of the models. Our approach can also be used to
+fine-tune conventionally-trained models and achieve similar gains in
 explainability. Finally, this work showcases that it is practical to

### 2025-03-13 paper/sections/conclusion.qmd
Down to 16 pages without major cutting of content

@@
 # Conclusion {#sec-conclusion}
-State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans.
-Although post-hoc explainability approaches have attracted major research interest, these cannot guarantee that the explanations agree with the opaque model's learned representation of data.
+State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans and existing post-hoc explainability approaches cannot guarantee that the explanations agree with the model's learned representation of data.
 As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models.
 Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients.
-Through extensive experiments we demonstrate that CT satisfies its objectives while preserving the predictive performance of the trained models.
-We also find that our approach can be used to fine-tune conventionally-trained models and achieve similar gains in explainability.
+Through extensive experiments we demonstrate that CT satisfies its objectives while preserving the predictive performance of the models.
+Our approach can also be used to fine-tune conventionally-trained models and achieve similar gains in explainability.
 Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

### 2025-03-13 paper/sections/discussion.qmd
Fix references
Down to 16 pages without major cutting of content

@@
 
-***CT is subject to training instabilities.*** Joint energy-based models are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes.
+***CT is subject to training instabilities.*** JEMs are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes.
 CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers.
-For example, @altmeyer2023faithful recognize this to be a challenge for *ECCCo* and so it may have downstream impacts on our proposed method.
-Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating a sufficiently large number of counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.
+Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.
 
-***CT is sensitive to hyperparameter selection.*** Our method benefits from tuning certain key hyperparameters (see @sec-hyperparameters).
+***CT is sensitive to hyperparameter selection.*** Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters).
 In this work, we have relied exclusively on grid search for this task.
@@
 
-## Limitations and Challenges {#sec-limitations}
+## Current Limitations {#sec-limitations}
 
-***CT increases the training time of models.*** Counterfactual training promotes explainability through CEs and robustness through AEs at the cost of longer training times compared to conventional training regimes.
-While higher numbers of iterations and counterfactuals per iteration positively impact the quality of found solutions, they also increase the required amount of computations.
-We find that relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs^[See supplementary appendix for computational details.].
-However, there are three factors that attenuate the impact of this limitation.
-First, CT provides counterfactual explanations for the training samples essentially for free, which may be beneficial in many ADM systems.
-Second, we find that CT can retain its value when used as a "fine-tuning" training regime for conventionally-trained models.
-Third, in principle, CT yields itself to parallel execution, which we have leveraged for our own experiments.
+***CT increases the training time of models.*** CT can be more time-consuming than conventional training regimes.
+While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations.
+Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect.
+First, CT amortizes the cost of CEs for the training samples.
+Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models.
+Third, it yields itself to parallel execution, which we have leveraged for our own experiments.
 
@@
 However, it requires that model owners define the mutability constraints for (all) features considered by the model.
-Even with sufficient domain knowledge to protect all immutable features, there may exist proxies that are theoretically mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections.
-As an example, consider the Adult dataset used in our experiments where the mutable education status is a proxy for the immutable age, in that the attainment of degrees is correlated with age.
+Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections.
 Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to increase the explainability of the model.
 
-***Interventions on features may impact fairness downstream.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions.
+***Interventions on features may impact fairness.***  We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects.
 As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.
-For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups.
-When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact].
-Still, as the referenced publications indicate, such phenomena are not specific to CT; all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].
+For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups.
+When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness].
+Still, these phenomena are not specific to CT.

### 2025-03-13 paper/sections/experiments.qmd
Down to 16 pages without major cutting of content

@@
 
-1. To what extent does our proposed counterfactual training objective (@eq-obj) induce models to learn plausible explanations?
-2. To what extent does our proposed counterfactual training objective (@eq-obj) yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+1. To what extent does our proposed counterfactual training objective in @eq-obj induce models to learn plausible explanations?
+2. To what extent does our proposed counterfactual training objective in @eq-obj yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
 3. What are the effects of hyperparameter selection with respect to @eq-obj?
@@
 For *OL* the reduction is negative, consistent with the distance-based metric.
-The only dataset, for which our proposed metric disagrees with the distance-based metric is *MNIST*.
+*MNIST* is the only dataset for which the two metrics disagree.
 
@@
 We note in this context, that we have not optimized our models for predictive performance at all and worked with very small networks.
-In summary, we find that CT can substantially improve the quality of explanations learned by models without generally sacrificing predictive accuracy.
+In summary, we find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive accuracy.
 
@@
 ### Impact of hyperparameter settings {#sec-hyperparameters}
-We extensively test the impact of three types of hyperparameters on the proposed training regime.
-Our complete results are available in the technical appendix; this section focuses on the main findings.
+We test in-depth the impact of three types of hyperparameters; our complete results are in the appendix.
 
@@
 We also confirm that a higher number of training epochs is beneficial.
-Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability.
-When training models conventionally during the first 50% of epochs before switching to CT for the next 50% of epochs, we observed positive results.
+Interestingly, we observed desired improvements in explainability when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process.
 Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.

### 2025-03-13 paper/sections/intro.qmd
Down to 16 pages without major cutting of content

@@
 The trend towards complexity for the sake of performance has come under serious scrutiny in recent years.
-At the very cusp of the deep learning revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples: perturbed versions of data instances that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts.
-Even though some partially effective mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---truly robust deep learning (DL) remains unattainable even for models that are considered shallow by today's standards [@kolter2023keynote].
+At the very cusp of the deep learning (DL) revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples (AEs): perturbed versions of data instances that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts.
+Even though some partially effective mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---truly robust deep learning remains unattainable even for models that are considered "shallow" by today's standards [@kolter2023keynote].
 
-Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so-called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery] and generalizability [@morcos2019success].
-Similarly, @zhang2021understanding showed that state-of-the-art neural networks are so expressive that they can fit randomly labeled data.
+Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so-called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery].
+Similarly, @zhang2021understanding showed that state-of-the-art neural networks are expressive enough to fit randomly labeled data.
 Thus, looking at the predictive performance alone, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns.
-This poses two related challenges.
-Firstly, there is no dependable way to verify if such complex representations correspond to meaningful and plausible explanations.
+This poses two challenges.
+Firstly, there is no dependable way to verify if representations correspond to meaningful, plausible explanations.
 Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.
@@
 
-**Specifically, we introduce counterfactual training (CT)**: a novel training regime explicitly meant to align learned representations with plausible explanations that comply with user requirements.
+**Specifically, we propose counterfactual training (CT)**: a novel training regime that aligns learned representations with plausible explanations compliant with user requirements.
 The remainder of this paper is structured as follows.
-@sec-lit presents related work, focusing in particular on the link between adversarial examples and counterfactual explanations.
+@sec-lit presents related work, focusing on the link between adversarial examples and counterfactual explanations.
 Then follow our main contributions:
@@
 1. In @sec-method, we introduce our methodological framework and show theoretically that it can be used to enforce global actionability constraints.  
-2. Through extensive experiments we demonstrate that CT substantially improves explainability without sacrificing predictive performance (@sec-experiments).
+2. In @sec-experiments, through extensive experiments we demonstrate that CT substantially improves explainability without sacrificing predictive performance.
 
-We discuss future research and challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising new approach towards making opaque models more trustworthy.
+We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.

### 2025-03-13 paper/sections/method.qmd
Down to 16 pages without major cutting of content

@@
 where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy).
-The second and third terms in @eq-obj are explained in detail below.
-For now, they can be sufficiently described as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
+The second and third terms are explained in detail below.
+For now, they can be summarized as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
 Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit].
-The trade-off between the components can be governed by adjusting the strengths of the penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.
+The trade-off between the components can governed through penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.
 
@@
 
-It is worth highlighting that Prp.~\ref{prp-mtblty} assumes independence of features.
+It is worth highlighting that \mbox{Prp.
+\ref{prp-mtblty}} assumes independence of features.
 This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected.

### 2025-03-13 paper/sections/related.qmd
Down to 16 pages without major cutting of content

@@
 To the best of our knowledge, the proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability.
-In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and stakeholder requirements; the former means that the counterfactuals should comply with the distribution of the factual data, the latter means that they should respect arbitrary (global) actionability constraints.
-To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of counterfactual explanations that we can generate for a given model.
-The authors show that CEs---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples.
+In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and (global) actionability constraints.
+To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of CEs that can be generated for a given model.
+The authors show that CEs---understood as minimal input perturbations that yield some desired model prediction---tend to be more meaningful if the underlying model is more robust to adversarial examples.
 We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom.
-As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., adversarial examples).
+As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., AEs).
 Thus, by inducing models to "unlearn" susceptiblity to such examples, AT can effectively remove implausible explanations from the solution space.
@@
 
-There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples (AE).
+There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples.
 Pointing to clear similarities in how CEs and AEs are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problems in representation learning.

### 2025-03-14 paper/preprint/paper.tex
just rerendered

@@
 serious scrutiny in recent years. At the very cusp of the deep learning
-revolution, Szegedy et al. (\citeproc{ref-szegedy2013intriguing}{2013})
-showed that artificial neural networks (ANN) are sensitive to
-adversarial examples: perturbed versions of data instances that yield
-vastly different model predictions despite being ``imperceptible'' in
-that they are semantically indifferent from their factual counterparts.
-Even though some partially effective mitigation strategies have been
-proposed---most notably \textbf{adversarial training}
+(DL) revolution, Szegedy et al.
+(\citeproc{ref-szegedy2013intriguing}{2013}) showed that artificial
+neural networks (ANN) are sensitive to adversarial examples (AEs):
+perturbed versions of data instances that yield vastly different model
+predictions despite being ``imperceptible'' in that they are
+semantically indifferent from their factual counterparts. Even though
+some partially effective mitigation strategies have been proposed---most
+notably \textbf{adversarial training}
 (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow, Shlens, and
-Szegedy 2014})---truly robust deep learning (DL) remains unattainable
-even for models that are considered shallow by today's standards
+Szegedy 2014})---truly robust deep learning remains unattainable even
+for models that are considered ``shallow'' by today's standards
 (\citeproc{ref-kolter2023keynote}{Kolter 2023}).
@@
 predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
-Carbin 2019}) and generalizability
-(\citeproc{ref-morcos2019success}{Morcos et al. 2019}). Similarly, Zhang
-et al. (\citeproc{ref-zhang2021understanding}{2021}) showed that
-state-of-the-art neural networks are so expressive that they can fit
-randomly labeled data. Thus, looking at the predictive performance
-alone, the solutions may seem to provide compelling explanations for the
-data, when in fact they are based on purely associative, semantically
-meaningless patterns. This poses two related challenges. Firstly, there
-is no dependable way to verify if such complex representations
-correspond to meaningful and plausible explanations. Secondly, even if
-we could resolve the first challenge, it remains undecided how to ensure
-that models can \emph{only} learn valuable explanations.
+Carbin 2019}). Similarly, Zhang et al.
+(\citeproc{ref-zhang2021understanding}{2021}) showed that
+state-of-the-art neural networks are expressive enough to fit randomly
+labeled data. Thus, looking at the predictive performance alone, the
+solutions may seem to provide compelling explanations for the data, when
+in fact they are based on purely associative, semantically meaningless
+patterns. This poses two challenges. Firstly, there is no dependable way
+to verify if representations correspond to meaningful, plausible
+explanations. Secondly, even if we could resolve the first challenge, it
+remains undecided how to ensure that models can \emph{only} learn
+valuable explanations.
 
@@
 derived for an opaque model are also deemed plausible with respect to
-the underlying data and stakeholder requirements; the former means that
-the counterfactuals should comply with the distribution of the factual
-data, the latter means that they should respect arbitrary (global)
-actionability constraints. To make the desiderata for our framework more
-concrete, we follow Augustin, Meinke, and Hein
-(\citeproc{ref-augustin2020adversarial}{2020}) in tying the concept of
-explainability to the quality of counterfactual explanations that we can
-generate for a given model. The authors show that CEs---understood here
-as minimal input perturbations that yield some desired model
-prediction---are generally more meaningful if the underlying model is
-more robust to adversarial examples. We can make intuitive sense of this
-finding when looking at adversarial training (AT) through the lens of
-representation learning with high degrees of freedom. As argued before,
-learned representations may be sensitive to producing implausible
-explanations and mispredicting for worst-case counterfactuals (i.e.,
-adversarial examples). Thus, by inducing models to ``unlearn''
-susceptiblity to such examples, AT can effectively remove implausible
-explanations from the solution space.
+the underlying data and (global) actionability constraints. To make the
+desiderata for our framework more concrete, we follow Augustin, Meinke,
+and Hein (\citeproc{ref-augustin2020adversarial}{2020}) in tying the
+concept of explainability to the quality of CEs that can be generated
+for a given model. The authors show that CEs---understood as minimal
+input perturbations that yield some desired model prediction---tend to
+be more meaningful if the underlying model is more robust to adversarial
+examples. We can make intuitive sense of this finding when looking at
+adversarial training (AT) through the lens of representation learning
+with high degrees of freedom. As argued before, learned representations
+may be sensitive to producing implausible explanations and mispredicting
+for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to
+``unlearn'' susceptiblity to such examples, AT can effectively remove
+implausible explanations from the solution space.
 
@@
 There have also been several attempts at formalizing the relationship
-between counterfactual explanations and adversarial examples (AE).
-Pointing to clear similarities in how CEs and AEs are generated,
-Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
-case for jointly studying the opaqueness and robustness problems in
+between counterfactual explanations and adversarial examples. Pointing
+to clear similarities in how CEs and AEs are generated, Freiesleben
+(\citeproc{ref-freiesleben2022intriguing}{2022}) makes the case for
+jointly studying the opaqueness and robustness problems in
 representation learning. Formally, AEs can be seen as the subset of CEs
@@
 that induces discriminative performance (e.g., cross-entropy). The
-second and third terms in Equation~\ref{eq-obj} are explained in detail
-below. For now, they can be sufficiently described as inducing
-explainability directly and indirectly by penalizing: (1) the
-contrastive divergence, \(\text{div}(\cdot)\), between mature
-counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
-\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
-class \(y^+\), and, (2) the adversarial loss, \(\text{advloss}(.)\),
-with respect to nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\).
-Finally, \(\text{ridge}(\cdot)\) denotes a Ridge penalty
-(\(\ell_2\)-norm) that regularizes the magnitude of the energy terms
-involved in \(\text{div}(\cdot)\) (\citeproc{ref-du2019implicit}{Du and
-Mordatch 2020}). The trade-off between the components can be governed by
-adjusting the strengths of the penalties \(\lambda_{\text{div}}\),
-\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).
+second and third terms are explained in detail below. For now, they can
+be summarized as inducing explainability directly and indirectly by
+penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
+between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
+samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
+target class \(y^+\), and, (2) the adversarial loss,
+\(\text{advloss}(.)\), with respect to nascent counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
+denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
+of the energy terms involved in \(\text{div}(\cdot)\)
+(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-off
+between the components can governed through penalties
+\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
+\(\lambda_{\text{reg}}\).
 
@@
 \item
-  To what extent does our proposed counterfactual training objective
-  (Equation~\ref{eq-obj}) induce models to learn plausible explanations?
-\item
-  To what extent does our proposed counterfactual training objective
-  (Equation~\ref{eq-obj}) yield more favorable algorithmic recourse
+  To what extent does our proposed counterfactual training objective in
+  Equation~\ref{eq-obj} induce models to learn plausible explanations?
+\item
+  To what extent does our proposed counterfactual training objective in
+  Equation~\ref{eq-obj} yield more favorable algorithmic recourse
   outcomes in the presence of actionability constraints?
@@
 Variational Autoencoder (VAE), and \emph{ECCo}---the generator of
-Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023}) but without
+Altmeyer et al. (\citeproc{ref-altmeyer2024faithful}{2024}) but without
 the conformal prediction component---as a method that directly targets
@@
 
-\textbf{\emph{CT is subject to training instabilities.}} Joint
-energy-based models are susceptible to instabilities during training
+\textbf{\emph{CT is subject to training instabilities.}} JEMs are
+susceptible to instabilities during training
 (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
@@
 Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
-For example, Altmeyer et al. (\citeproc{ref-altmeyer2023faithful}{2023})
-recognize this to be a challenge for \emph{ECCCo} and so it may have
-downstream impacts on our proposed method. Still, we find that training
-instabilities can be successfully mitigated by regularizing energy
-(\(\lambda_{\text{reg}}\)), generating a sufficiently large number of
-counterfactuals during each training epoch, and including only mature
-counterfactuals for contrastive divergence.
+Still, we find that training instabilities can be successfully mitigated
+by regularizing energy (\(\lambda_{\text{reg}}\)), generating
+sufficiently many counterfactuals during each training epoch, and
+including only mature counterfactuals for contrastive divergence.
 
 \textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
-benefits from tuning certain key hyperparameters (see
+benefits from the tuning of certain key hyperparameters (see
 Section~\ref{sec-hyperparameters}). In this work, we have relied
@@
 
-\subsection{Limitations and Challenges}\label{sec-limitations}
-
-\textbf{\emph{CT increases the training time of models.}} Counterfactual
-training promotes explainability through CEs and robustness through AEs
-at the cost of longer training times compared to conventional training
-regimes. While higher numbers of iterations and counterfactuals per
-iteration positively impact the quality of found solutions, they also
-increase the required amount of computations. We find that relatively
-small grids with 270 settings can take almost four hours for more
-demanding datasets on a high-performance computing cluster with 34 2GB
-CPUs\footnote{See supplementary appendix for computational details.}.
-However, there are three factors that attenuate the impact of this
-limitation. First, CT provides counterfactual explanations for the
-training samples essentially for free, which may be beneficial in many
-ADM systems. Second, we find that CT can retain its value when used as a
-``fine-tuning'' training regime for conventionally-trained models.
-Third, in principle, CT yields itself to parallel execution, which we
-have leveraged for our own experiments.
+\subsection{Current Limitations}\label{sec-limitations}
+
+\textbf{\emph{CT increases the training time of models.}} CT can be more
+time-consuming than conventional training regimes. While higher numbers
+of CEs per iteration positively impact the quality of solutions, they
+also increase the amount of computations. Relatively small grids with
+270 settings can take almost four hours for more demanding datasets on a
+high-performance computing cluster with 34 2GB CPUs.\footnote{See
+  supplementary appendix for computational details.} Three factors
+attenuate this effect. First, CT amortizes the cost of CEs for the
+training samples. Second, we find that it can retain its value when used
+as a ``fine-tuning'' technique for conventionally-trained models. Third,
+it yields itself to parallel execution, which we have leveraged for our
+own experiments.
 
@@
 owners define the mutability constraints for (all) features considered
-by the model. Even with sufficient domain knowledge to protect all
-immutable features, there may exist proxies that are theoretically
-mutable (and hence should not be protected) but preserve enough
-information about the principals to hinder the protections. As an
-example, consider the Adult dataset used in our experiments where the
-mutable education status is a proxy for the immutable age, in that the
-attainment of degrees is correlated with age. Delineating actionability
-is a major undecided challenge in the AR literature (see, e.g.,
+by the model. Even if all immutable features are protected, there may
+exist proxies that are mutable (and hence should not be protected) but
+preserve enough information about the principals to hinder the
+protections. Delineating actionability is a major undecided challenge in
+the AR literature (see, e.g.,
 \citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
@@
 
-\textbf{\emph{Interventions on features may impact fairness
-downstream.}} Related to the point above, we provide a tool that allows
-practitioners to modify the sensitivity of a model with respect to
-certain features, which may have implication for the fair and equitable
-treatment of individuals subject to automated decisions. As protecting a
+\textbf{\emph{Interventions on features may impact fairness.}} We
+provide a tool that allows practitioners to modify the sensitivity of a
+model with respect to certain features, which may have implication for
+the fair and equitable treatment of decision subjects. As protecting a
 set of features leads the model to assign higher relative importance to
@@
 enforcing explanations based on features that are more difficult to
-modify by some (group of) individuals. For example, consider again the
-Adult dataset where features such as workclass or education may be more
-difficult to change for underpriviledged groups. When applied
-irresponsibly, CT could result in an unfairly assigned burden of
-recourse (e.g., \citeproc{ref-sharma2020certifai}{Sharma, Henderson, and
-Ghosh 2020}), threatening the equality of opportunity in the system
-(\citeproc{ref-bell2024fairness}{Bell et al. 2024}) and potentially
-reinforcing social segregation (\citeproc{ref-gao2023impact}{Gao and
-Lakkaraju 2023}). Still, as the referenced publications indicate, such
-phenomena are not specific to CT; all types of ADM solutions without
-strong external protections have been recognized to promote harmful
-power dynamics (\citeproc{ref-maas2023machine}{Maas 2023}).
+modify by some (group of) individuals. For example, consider the Adult
+dataset used in our experiments, where \emph{workclass} or
+\emph{education} may be more difficult to change for underpriviledged
+groups. When applied irresponsibly, CT could result in an unfairly
+assigned burden of recourse (e.g.,
+\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
+threatening the equality of opportunity in the system
+(\citeproc{ref-bell2024fairness}{Bell et al. 2024}). Still, these
+phenomena are not specific to CT.
 
@@
 
-\bibitem[\citeproctext]{ref-altmeyer2023faithful}
-Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia C.
-S. Liem. 2023. {``Faithful Model Explanations Through Energy-Constrained
-Conformal Counterfactuals.''} \url{https://arxiv.org/abs/2312.10648}.
-
 \bibitem[\citeproctext]{ref-altmeyer2024faithful}
@@
 
-\bibitem[\citeproctext]{ref-gao2023impact}
-Gao, Ruijiang, and Himabindu Lakkaraju. 2023. {``On the Impact of
-Algorithmic Recourse on Social Segregation.''} In \emph{Proceedings of
-the 40th International Conference on Machine Learning}. ICML'23.
-Honolulu, Hawaii, USA: JMLR.org.
-
 \bibitem[\citeproctext]{ref-goodfellow2014explaining}
@@
 
-\bibitem[\citeproctext]{ref-maas2023machine}
-Maas, Jonne. 2023. {``Machine Learning and Power Relations.''} \emph{AI
-\& SOCIETY} 38 (4): 1493--1500.
-
 \bibitem[\citeproctext]{ref-mcgregor2021preventing}
@@
 of the AAAI Conference on Artificial Intelligence}, 35:15458--63. 17.
-
-\bibitem[\citeproctext]{ref-morcos2019success}
-Morcos, Ari S., Haonan Yu, Michela Paganini, and Yuandong Tian. 2019.
-{``One Ticket to Win Them All: Generalizing Lottery Ticket
-Initializations Across Datasets and Optimizers.''} In \emph{Proceedings
-of the 33rd International Conference on Neural Information Processing
-Systems}. Red Hook, NY, USA: Curran Associates Inc.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-warning-color-frame, colbacktitle=quarto-callout-warning-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Package Version (Reproducibility)}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Package Version (Reproducibility)}, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, colframe=quarto-callout-warning-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, bottomtitle=1mm, titlerule=0mm, bottomrule=.15mm, opacityback=0, toprule=.15mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable, opacitybacktitle=0.6, colback=white, toptitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, arc=.35mm, coltitle=black, left=2mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
 

### 2025-03-14 paper/quarto_ecml/paper.tex
can now compute sensitivity to featues

@@
 assess how similar counterfactuals are to observed samples in the target
-domain. We rely on the distance-based metric used by
-\cite{altmeyer2024faithful},
+domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
+distance-based metric used by \cite{altmeyer2024faithful},
 
@@
 \text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
-}\end{equation} where \(\mathbf{X}^\prime\) denotes a set of multiple
+}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
 counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the

### 2025-03-14 paper/sections/experiments.qmd
can now compute sensitivity to featues

@@
 To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.
-For plausibility, we assess how similar counterfactuals are to observed samples in the target domain.
+For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$.
 We rely on the distance-based metric used by @altmeyer2024faithful,
@@
 $$ {#eq-impl-div}
-where $\mathbf{X}^\prime$ denotes a set of multiple counterfactuals and $\text{MMD}(\cdot)$ is an unbiased estimate of the squared population maximum mean discrepancy [@gretton2012kernel].
+where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is an unbiased estimate of the squared population maximum mean discrepancy [@gretton2012kernel].
 The metric in @eq-impl-div is equal to zero iff the two distributions are the same, $\mathbf{X}^\prime=\mathbf{X}^+$.

### 2025-03-14 paper/experiments/CTExperiments/src/globals.jl

@@ -1,10 +1,49 @@
-using CounterfactualExplanations.Evaluation: validity_strict
+using CounterfactualExplanations.Evaluation: validity_strict, feature_sensitivity, get_outid
 using JSON
+
+"""
+    add_sensitivity!(bmk::Benchmark)
+
+Adds a column indicating the index of feature `d` for which `feature_sensitivity` was computed.
+"""
+function add_sensitivity!(bmk::Benchmark)
+    sens = get_global_param("sensitivity", nothing)
+    if isnothing(sens)
+        return bmk
+    end
+    sens = parse_sens(sens)
+    feature_id = []
+    for varname in bmk.evaluation.variable
+        if !contains(varname, "sens")
+            push!(feature_id, missing)
+        else
+            push!(feature_id, sens[get_outid(varname)])
+        end
+    end
+    bmk.evaluation.feature_id .= feature_id
+    return bmk
+end
+
+"""
+    parse_sens(sensitivity)
+
+Parses the `--sensitivity` command line argument.
+"""
+function parse_sens(sensitivity)
+    return sensitivity = try
+        parse.(Int, split(sensitivity, ","))
+    catch
+        error(
+            "The `--sensitivity` argument should be either an integer (e.g. `--sensitivity=1`) or a comma-separated list of integers (e.g. `--sensitivity=1,2,5`)",
+        )
+    end
+end
 
 function get_ce_measures(;
     use_mmd=get_global_param("use_mmd", true),
     length_scale=get_global_param("length_scale", 5.0),
     compute_p=get_global_param("compute_p", nothing),
+    sensitivity=get_global_param("sensitivity", nothing),
 )
     measures = [
         validity,
@@ -23,6 +62,12 @@ function get_ce_measures(;
                 compute_p=compute_p,
             ),
         )
+    end
+
+    if !isnothing(sensitivity)
+        sensitivity = parse_sens(sensitivity)
+        sens(x; kwrgs...) = feature_sensitivity(x, sensitivity)
+        push!(measures, sens)
     end
 
     return measures

### 2025-03-14 paper/quarto_ecml/paper.tex
actionability done

@@
 punish younger individuals. The counterfactuals are also plausible with
-respect to the mutable feature. Thus, we consider the model in panel (d)
-as the most explainable according to Def. \ref{def-explainability}.
+respect to the mutable feature and shorter on average than in (c). Thus,
+we consider the model in panel (d) as the most explainable according to
+Def. \ref{def-explainability}.
 
@@
 
-\subsubsection{Plausibility}\label{plausibility}
+\subsubsection{Plausibility}\label{sec-plaus}
 
@@
 in implausibility as measured by Equation~\ref{eq-impl-div} and
-aggregated across all test specifications of \emph{ECCo}. The final two
-rows show the test accuracies for the model trained with CT and
+aggregated across all test specifications of \emph{ECCo}. The following
+two rows show the test accuracies for the model trained with CT and
 conventionally trained models (``vanilla'').
@@
 Equation~\ref{eq-impl-div} and aggregated across all test specifications
-of \emph{ECCo}. The final two rows show the test accuracies for the
-model trained with CT and conventionally trained models (``vanilla'').}
+of \emph{ECCo}. The following two rows show the test accuracies for the
+model trained with CT and conventionally trained models (``vanilla'').
+The final row present the average reduction in costs when imposing
+mutability constraints (Section~\ref{sec-act})}
 
@@
 
-\subsubsection{Actionability}\label{actionability}
+\subsubsection{Actionability}\label{sec-act}
+
+In Section~\ref{sec-method}, we show that our proposed way for encoding
+mutability constraints leads to lower classifier sensitivity wrt.
+immutable features for linear models, tilting the decision boundary in
+favour of mutable features instead. For binding constraints at test
+time, this has the effect of shorter counterfactual paths and hence
+smaller average costs to individuals. To extend this to the non-linear
+case, we test the effect of imposing mutability constraints empirically
+for our synthetic data using the same evaluation scheme as above. The
+final row in Table~\ref{tbl-main} reports the average reduction in costs
+for CT compared to the baseline, when imposing that either the first or
+the second feature is immutable. In all cases, costs are reduced
+substantially indicating that classifiers trained with CT are indeed
+more sensitive to mutable features.
 

### 2025-03-14 paper/sections/experiments.qmd
actionability done

@@
 
-### Plausibility
+### Plausibility {#sec-plaus}
 
@@
 The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*.
-The final two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
+The following two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
 
@@
 The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*.
-The final two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
+The following two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
+The final row present the average reduction in costs when imposing mutability constraints (@sec-act)
 
@@
 
-### Actionability
+### Actionability {#sec-act}
+
+In @sec-method, we show that our proposed way for encoding mutability constraints leads to lower classifier sensitivity wrt.
+immutable features for linear models, tilting the decision boundary in favour of mutable features instead.
+For binding constraints at test time, this has the effect of shorter counterfactual paths and hence smaller average costs to individuals.
+To extend this to the non-linear case, we test the effect of imposing mutability constraints empirically for our synthetic data using the same evaluation scheme as above.
+The final row in @tbl-main reports the average reduction in costs for CT compared to the baseline, when imposing that either the first or the second feature is immutable.
+In all cases, costs are reduced substantially indicating that classifiers trained with CT are indeed more sensitive to mutable features.
 

### 2025-03-14 paper/experiments/CTExperiments/src/CTExperiments.jl

@@ -41,7 +41,7 @@ export test_performance, evaluate_counterfactuals
 export generate_factual_target_pairs
 export to_toml
 export aggregate_logs, aggregate_ce_evaluation, aggregate_performance
-export final_table, final_results, final_params
+export final_table, final_results, final_params, final_mutability
 export PlotParams, useful_byvars
 export plot_errorbar_logs, plot_measure_ce, plot_ce
 export set_work_dir, get_work_dir, results_dir

### 2025-03-14 paper/sections/experiments.qmd
minor fixes

@@
 
-Key plausibility and predictive performance metrics for all datasets.
-The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time.
-The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*.
-The following two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
-The final row present the average reduction in costs when imposing mutability constraints (@sec-act)
+Key explainability and predictive performance metrics for all datasets.
+**Plausibility**: the top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time.
+The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications.
+**Accuracy**: The following two rows show the test accuracies for the models trained with CT and the baseline.
+**Actionability**: The final row present the average reduction in costs when imposing mutability constraints.
 

### 2025-03-14 paper/quarto_ecml/paper.tex

@@
 
-\caption{\label{tbl-main}Key plausibility and predictive performance
-metrics for all datasets. The top five rows show the percentage
-reduction in implausibility according to Equation~\ref{eq-impl-dist} for
-varying degrees of the energy penalty used for \emph{ECCo} at test time.
-The following row shows the reduction in implausibility as measured by
-Equation~\ref{eq-impl-div} and aggregated across all test specifications
-of \emph{ECCo}. The following two rows show the test accuracies for the
-model trained with CT and conventionally trained models (``vanilla'').
-The final row present the average reduction in costs when imposing
-mutability constraints (Section~\ref{sec-act})}
+\caption{\label{tbl-main}Key explainability and predictive performance
+metrics for all datasets. \textbf{Plausibility}: the top five rows show
+the percentage reduction in implausibility according to
+Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
+used for \emph{ECCo} at test time. The following row shows the reduction
+in implausibility as measured by Equation~\ref{eq-impl-div} and
+aggregated across all test specifications. \textbf{Accuracy}: The
+following two rows show the test accuracies for the models trained with
+CT and the baseline. \textbf{Actionability}: The final row present the
+average reduction in costs when imposing mutability constraints.}
 

### 2025-03-14 paper/sections/other/main_results.qmd

--- 2025-03-13 paper/sections/other/main_results.qmd
@@
 res_dir = joinpath(ENV["FINAL_GRID_RESULTS"],"single")
+mtbl_dir = joinpath(ENV["FINAL_GRID_RESULTS"],"mutability")
 ```
 
-# Details on Main Experiments {.appendix}
+# Details on Main Experiments {#sec-app-main .appendix}
 
@@
 
-df = final_table(res_dir)
-get_table_inputs(df, nothing; backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; table_type=:tabular, save_name="paper/tables/main.tex")
+tbl_mtbl = final_mutability(mtbl_dir; var=["distance"])
+df = final_table(res_dir; tbl_mtbl)
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+tabulate_results(inputs; table_type=:tabular, save_name="paper/quarto_ecml/tables/main.tex", wrap_table=false)
+tabulate_results(inputs; table_type=:tabular, save_name="paper/preprint/tables/main.tex", wrap_table=false)
 ```

### 2025-03-14 paper/sections/other/tuning.qmd

@@
 Based on the findings from our initial large grid searches (@sec-app-grid), we tune selected hyperparameters for all datasets: namely, the decision threshold $\tau$ and the strength of the energy regularization $\lambda_{\text{reg}}$.
-The final hyperparameter choices for each dataset are presented in **ADD TABLE**.
+The final hyperparameter choices for each dataset are presented in @tbl-final-params in @sec-app-main.
 Detailed results for each data set are shown in @fig-tune-plaus-adult to @fig-tune-mat-over.
-From **ADD TABLE**, we notice that the same decision threshold of $\tau=0.5$ is optimal for all but on dataset.
+From @tbl-final-params, we notice that the same decision threshold of $\tau=0.5$ is optimal for all but on dataset.
 We attribute this to the fact that a low decision threshold results in a higher share of mature counterfactuals and hence more opportunities for the model to learn from examples (@fig-tune-mat-adult to @fig-tune-mat-over).

### 2025-03-14 paper/quarto_ecml/paper.tex
Shorten to slightly less than 16 pages by restructuring

@@
 training regime that aligns learned representations with plausible
-explanations compliant with user requirements. The remainder of this
+explanations respecting actionability constraints. The remainder of this
 paper is structured as follows. Section~\ref{sec-lit} presents related
 work, focusing on the link between adversarial examples and
-counterfactual explanations. Then follow our main contributions:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  In Section~\ref{sec-method}, we introduce our methodological framework
-  and show theoretically that it can be used to enforce global
-  actionability constraints.\\
-\item
-  In Section~\ref{sec-experiments}, through extensive experiments we
+counterfactual explanations. Then follow our two principal
+contributions. In Section~\ref{sec-method}, we introduce our
+methodological framework and show theoretically that it can be employed
+to enforce global actionability constraints. In
+Section~\ref{sec-experiments}, through extensive experiments, we
 demonstrate that CT substantially improves explainability without
-  sacrificing predictive performance.
-\end{enumerate}
-
-We discuss the challenges in Section~\ref{sec-discussion} and conclude
-in Section~\ref{sec-conclusion} that CT is a promising approach towards
+sacrificing predictive performance. We discuss the challenges in
+Section~\ref{sec-discussion} and conclude in
+Section~\ref{sec-conclusion} that CT is a promising approach towards
 making opaque models more trustworthy.
@@
 
-In the natural language processing (NLP) domain, counterfactuals have
-similarly been used to improve models through data augmentation.
-\cite{wu2021polyjuice} propose \emph{Polyjuice}, a general-purpose
-counterfactual generator for language models. They demonstrate
-empirically that the augmentation of training data through
-\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
-tasks. \cite{balashankar2023improving} similarly use \emph{Polyjuice} to
-augment NLP datasets through diverse counterfactuals and show that
-classifier robustness improves by up to 20\%. Finally,
+In the natural language processing (NLP) domain, CEs have also been used
+to improve models through data augmentation. \cite{wu2021polyjuice}
+propose \emph{Polyjuice}, a general-purpose counterfactual generator for
+language models. The authors empirically demonstrate that the
+augmentation of training data with their method improves robustness in a
+number of NLP tasks. \cite{balashankar2023improving} similarly use
+\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
+and show that classifier robustness improves by up to 20\%. Finally,
 \cite{luu2023counterfactual} introduce Counterfactual Adversarial
 Training (CAT), which also aims at improving generalization and
-robustness of language models through a three-step procedure. First, the
+robustness of language models through a three-step procedure: the
 authors identify training samples that are subject to high predictive
-uncertainty. Second, they generate counterfactual explanations for those
-samples. Finally, they fine-tune the given language model on the
-augmented dataset that includes the generated counterfactuals.
+uncertainty, generate CEs for those samples, and then fine-tune the
+language model on a dataset augmented with the CEs.
 
@@
 actionability constraints ex-ante. For example, users can specify that
-certain features (e.g., \emph{age}, \emph{gender}) are immutable.
-Secondly, \cite{guo2023counternet} are the first to propose an
-end-to-end training pipeline that includes counterfactual explanations
+certain features are immutable. Secondly, \cite{guo2023counternet} are
+the first to propose an end-to-end training pipeline that includes CEs
 as part of the training procedure. In particular, they propose a
@@
 back to the predictor network. In contrast to \cite{guo2023counternet},
-we impose no restrictions on the neural network architecture at all.
+we impose no restrictions on the ANN architecture at all.
 
@@
 
-\cite{grathwohl2020your} observe that any classifier can be
+As observed by \cite{grathwohl2020your}, any classifier can be
 re-interpreted as a joint energy-based model (JEM) that learns to
@@
 different classifiers in Section~\ref{sec-experiments}. Under certain
-conditions, this results holds theoretically:\footnote{For the proof,
-  see the supplementary appendix.}
+conditions, this result holds theoretically:\footnote{For the proof, see
+  the supplementary appendix.}
 
@@
 
-It is worth highlighting that \mbox{Prp. \ref{prp-mtblty}} assumes
-independence of features. This raises a valid concern about the effect
-of protecting immutable features in the presence of proxies that remain
-unprotected. We address this in Section~\ref{sec-discussion}.
+It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}}
+assumes independence of features. This raises a valid concern about the
+effect of protecting immutable features in the presence of proxies that
+remain unprotected. We address this in Section~\ref{sec-discussion}.
 
@@
 independent features and Gaussian class-conditional densities in
-Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
-outcomes for different training procedures using the same model
-architecture each time (a linear classifier). In each case, we show the
-decision boundary (in green) and the training data colored according to
-their ground-truth label: orange points belong to the target class,
-\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
-indicate counterfactuals in the target class generated at test time
-using generic gradient descent until convergence.
+Figure~\ref{fig-poc}. The four panels show the outcomes for different
+training procedures using the same model architectures (a linear
+classifier). In panels (a) and (c) we have trained the models
+conventionally, while in panels (b) and (d) we used CT. Only in panels
+(c) and (d) do we impose the mutability constraint on \emph{age} at test
+time. In each case, we show the decision boundary (in green) and the
+training data colored according to their ground-truth label: orange
+points belong to the target class, \(y^+=2\), blue points belong to the
+non-target class, \(y^-=1\). Stars indicate CEs in the target class
+generated at test time using generic gradient descent until convergence.
 
-In panel (a), we have trained our model conventionally, and we do not
-impose mutability constraints at test time. The generated
-counterfactuals are all valid, but not plausible: they do not comply
-with the distribution of the factual samples in the target class to the
-point where they are clearly distinguishable from the ground-truth data.
-In panel (b), we have trained our model with CT, once again without any
-mutability constraints. We observe that the counterfactuals are highly
+In all cases the counterfactuals are valid but their quality differs. In
+panel (a), they are not plausible: they do not comply with the
+distribution of the factuals in \(y^+\) to the point where they form a
+clearly distinguishable cluster. In panel (b), they are highly
 plausible, meeting the first objective of Def. \ref{def-explainability}.
-
-In panel (c), we have used conventional training again, this time
-imposing the mutability constraint on \emph{age} at test time.
-Counterfactuals are valid but involve some substantial reductions in
-\emph{debt} for some individuals (very young applicants). By comparison,
-counterfactual paths are shorter on average in panel (d), where we have
-used CT and protected the immutable feature as described in
-Section~\ref{sec-constraints}. We observe that due to the classifier's
-lower sensitivity to \emph{age}, recourse recommendations with respect
-to \emph{debt} are much more homogenous and do not disproportionately
-punish younger individuals. The counterfactuals are also plausible with
-respect to the mutable feature and shorter on average than in (c). Thus,
-we consider the model in panel (d) as the most explainable according to
-Def. \ref{def-explainability}.
+In panel (c), the CEs involve substantial reductions in \emph{debt} for
+younger applicants. By comparison, counterfactual paths are shorter on
+average in panel (d) where we have protected the immutable \emph{age} as
+described in Section~\ref{sec-constraints}. Due to the classifier's
+lower sensitivity to \emph{age}, recommendations with respect to
+\emph{debt} are much more homogenous and do not unfairly punish younger
+individuals. These counterfactuals are also plausible with respect to
+the mutable feature. Thus, we consider the model in panel (d) as the
+most explainable according to Def. \ref{def-explainability}.
 
@@
 
-In this section, we present experiments that we have conducted in order
-to answer the following research questions:
+Here we present the experiments conducted to answer our research
+questions:
 
@@
 \item
-  To what extent does our proposed counterfactual training objective in
-  Equation~\ref{eq-obj} induce models to learn plausible explanations?
+  To what extent does our counterfactual training objective as defined
+  in Equation~\ref{eq-obj} induce models to learn plausible
+  explanations?
 \item
-  To what extent does our proposed counterfactual training objective in
-  Equation~\ref{eq-obj} yield more favorable algorithmic recourse
-  outcomes in the presence of actionability constraints?
+  To what extent does the CT objective produce more favorable
+  algorithmic recourse outcomes in the presence of actionability
+  constraints?
 \item
-  What are the effects of hyperparameter selection with respect to
-  Equation~\ref{eq-obj}?
+  What are the effects of hyperparameter selection wrt. the CT
+  objective?
 \end{enumerate}
@@
 \subsection{Experimental Setup}\label{experimental-setup}
-
-\subsubsection{Evaluation}\label{evaluation}
 
@@
 
-\subsubsection{Plausibility}\label{sec-plaus}
+\subsubsection{Plausibility.}\label{sec-plaus}
 
@@
 Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
-used for \emph{ECCo} at test time. The following row shows the reduction
-in implausibility as measured by Equation~\ref{eq-impl-div} and
-aggregated across all test specifications of \emph{ECCo}. The following
-two rows show the test accuracies for the model trained with CT and
-conventionally trained models (``vanilla'').
+used for \emph{ECCo} at test time. Row 6 shows the reduction in
+implausibility as measured by Equation~\ref{eq-impl-div} and aggregated
+across all test specifications of \emph{ECCo}. Rows 7 and 8 show the
+test accuracies for the model trained with CT and conventionally trained
+models (``vanilla'').
 
-We observe that for all datasets except \emph{OL} and across all test
-settings, the average distance of counterfactuals from observed samples
-in the target class is reduced, indicating improved plausibility. The
-magnitude of improvements varies by dataset: for the simple synthetic
-datasets, distance reductions range from around 20-40\% (\emph{LS},
-\emph{Moon}) to almost 60\% (\emph{Circ}). For the real-world tabular
-datasets, improvements are generally smaller but still substantial in
-many cases with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
-7-8\% for \emph{Cred} and around 3\% for \emph{Adult}. For our only
-vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
-results for our proposed divergence metric are qualitatively similar,
-but generally even more pronounced: for the \emph{Circ} dataset,
-implausibility is reduced by almost 94\% to virtually zero as we
-verified by looking at the absolute outcome. Improvements for other
-datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
+For all datasets except \emph{OL} and across all test settings, the
+average distance of CEs from observed samples in the target class is
+reduced, indicating improved plausibility. The magnitude of improvements
+varies. For the simple synthetic datasets, distance reductions range
+from around 20-40\% (\emph{LS}, \emph{Moon}) to almost 60\%
+(\emph{Circ}). For the real-world tabular datasets, improvements tend to
+be smaller but still substantial, with around 10-15\% for \emph{CH},
+11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred} and around 3\% for
+\emph{Adult}. For our only vision dataset (\emph{MNIST}), distances are
+reduced by up to 9\%. The results for our proposed divergence metric are
+qualitatively similar, but generally even more pronounced: for the
+\emph{Circ} dataset, implausibility is reduced by almost 94\% to
+virtually zero as we verified by the absolute outcome. Improvements for
+other datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
 \emph{OL} the reduction is negative, consistent with the distance-based
@@
 
-\subsubsection{Actionability}\label{sec-act}
+\subsubsection{Actionability.}\label{sec-act}
 
@@
 \subsubsection{Impact of hyperparameter
-settings}\label{sec-hyperparameters}
+settings.}\label{sec-hyperparameters}
 
-We test in-depth the impact of three types of hyperparameters; our
-complete results are in the appendix.
-
+We test the impact of three types of hyperparameters; our complete
+results are in the supplementary appendix.\\
 \textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
@@
 leads to optimal results because it is associated with high proportions
-of mature counterfactuals.
-
+of mature counterfactuals.\\
 \textbf{\emph{Hyperparameters for penalties.}} We find that the strength
@@
 combined with high values of \(\lambda_{\text{div}}\) or
-\(\lambda_{\text{adv}}\), this effect is not very pronounced.
-
+\(\lambda_{\text{adv}}\), this effect is not very pronounced.\\
 \textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
@@
 
-We first address the direct extensions of CT in
-Section~\ref{sec-approach}. Then, we look at its limitations and
-challenges in Section~\ref{sec-limitations}.
-
-\subsection{Future Research}\label{sec-approach}
-
-\textbf{\emph{CT is defined only for classification settings.}} Our
-formulation relies on the distinction between non-target class(es)
-\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
-through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
-arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
-discrete. Thus, it does not apply to ML tasks where the change in
-outcome cannot be readily quantified. Focus on classification models is
-a common restriction in research on CEs and AR. Other settings have
-attracted some interest (e.g., regression in
-\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
-little consensus how to robustly extend the notion of counterfactuals.
-
-\textbf{\emph{CT is subject to training instabilities.}} JEMs are
-susceptible to instabilities during training \cite{grathwohl2020your}
-and even though we depart from the SGLD-based sampling, we still
-encounter major variability in the outcomes. CT is exposed to two
-potential sources of instabilities: (1) the energy-based contrastive
-divergence term in Equation~\ref{eq-div}, and (2) the underlying
-counterfactual explainers. Still, we find that training instabilities
-can be successfully mitigated by regularizing energy
-(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
-during each training epoch, and including only mature counterfactuals
-for contrastive divergence.
-
-\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
-benefits from the tuning of certain key hyperparameters (see
-Section~\ref{sec-hyperparameters}). In this work, we have relied
-exclusively on grid search for this task. Future work on CT could
-benefit from investigating more sophisticated approaches towards
-hyperparameter tuning. Notably, CT is iterative which makes a variety of
-methods applicable, including Bayesian \cite{snoek2012practical} or
-gradient-based \cite{franceschi2017forward} optimization.
-
-\subsection{Current Limitations}\label{sec-limitations}
-
+As our results indicate, counterfactual training produces models that
+are more explainable. However, our approach is not without
+limitations.\\
 \textbf{\emph{CT increases the training time of models.}} CT can be more
@@
 it yields itself to parallel execution, which we have leveraged for our
-own experiments.
-
+own experiments.\\
 \textbf{\emph{Immutable features may have proxies.}} We propose an
@@
 the AR literature \cite{venkatasubramanian2020philosophical} impacting
-the capacity of CT to increase the explainability of the model.
-
+the capacity of CT to fulfill its intended goal.\\
 \textbf{\emph{Interventions on features may impact fairness.}} We
@@
 these phenomena are not specific to CT.
+
+We also highlight several interesting directions for future research.\\
+\textbf{\emph{Extending CT beyond classification settings.}} Our
+formulation relies on the distinction between non-target class(es)
+\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
+through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
+arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
+discrete. Thus, it does not apply to ML tasks where the change in
+outcome cannot be readily quantified. Focus on classification models is
+a common restriction in research on CEs and AR. Other settings have
+attracted some interest (e.g., regression in
+\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
+little consensus how to robustly extend the notion of CEs.\\
+\textbf{\emph{Addressing the training instabilities.}} JEMs are
+susceptible to instabilities during training \cite{grathwohl2020your}
+and even though we depart from the SGLD-based sampling, we still
+encounter major variability in the outcomes. CT is exposed to two
+potential sources of instabilities: (1) the energy-based contrastive
+divergence term in Equation~\ref{eq-div}, and (2) the underlying
+counterfactual explainers. Still, we find that training instabilities
+can be successfully mitigated by regularizing energy
+(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
+during each training epoch, and including only mature counterfactuals
+for contrastive divergence.\\
+\textbf{\emph{Improving hyperparameter selection procedures.}} Our
+method benefits from the tuning of certain key hyperparameters (see
+Section~\ref{sec-hyperparameters}). In this work, we have relied
+exclusively on grid search for this task. Future work on CT could
+benefit from investigating more sophisticated approaches towards
+hyperparameter tuning. Notably, CT is iterative which makes a variety of
+methods applicable, including Bayesian \cite{snoek2012practical} or
+gradient-based \cite{franceschi2017forward} optimization.
 

### 2025-03-14 paper/sections/discussion.qmd
Shorten to slightly less than 16 pages by restructuring

@@
 
-We first address the direct extensions of CT in @sec-approach. Then, we look at its limitations and challenges in @sec-limitations.
-
-## Future Research {#sec-approach}
-
-***CT is defined only for classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj.
-While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete.
-Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified.
-Focus on classification models is a common restriction in research on CEs and AR.
-Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of counterfactuals.
-
-***CT is subject to training instabilities.*** JEMs are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes.
-CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers.
-Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.
-
-***CT is sensitive to hyperparameter selection.*** Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters).
-In this work, we have relied exclusively on grid search for this task.
-Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning.
-Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.
-
-## Current Limitations {#sec-limitations}
-
+As our results indicate, counterfactual training produces models that are more explainable. However, our approach is not without limitations.  
 ***CT increases the training time of models.*** CT can be more time-consuming than conventional training regimes.
@@
 Third, it yields itself to parallel execution, which we have leveraged for our own experiments.
-
 ***Immutable features may have proxies.*** We propose an approach to protect immutable features and thus increase the actionability of the generated CEs.
@@
 Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections.
-Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to increase the explainability of the model.
-
+Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.
 ***Interventions on features may impact fairness.***  We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects.
@@
 Still, these phenomena are not specific to CT.
+
+We also highlight several interesting directions for future research.  
+***Extending CT beyond classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj.
+While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete.
+Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified.
+Focus on classification models is a common restriction in research on CEs and AR.
+Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of CEs.
+***Addressing the training instabilities.*** JEMs are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes.
+CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers.
+Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.
+***Improving hyperparameter selection procedures.*** Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters).
+In this work, we have relied exclusively on grid search for this task.
+Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning.
+Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.

### 2025-03-14 paper/sections/experiments.qmd
Shorten to slightly less than 16 pages by restructuring

@@
 
-In this section, we present experiments that we have conducted in order to answer the following research questions:
+Here we present the experiments conducted to answer our research questions:
 
-1. To what extent does our proposed counterfactual training objective in @eq-obj induce models to learn plausible explanations?
-2. To what extent does our proposed counterfactual training objective in @eq-obj yield more favorable algorithmic recourse outcomes in the presence of actionability constraints?
-3. What are the effects of hyperparameter selection with respect to @eq-obj?
+1. To what extent does our counterfactual training objective as defined in @eq-obj induce models to learn plausible explanations?
+2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+3. What are the effects of hyperparameter selection wrt. the CT objective?
 
 ## Experimental Setup
-
-### Evaluation
 
@@
 
-### Plausibility {#sec-plaus}
+### Plausibility. {#sec-plaus}
 
@@
 The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time.
-The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*.
-The following two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
+Row 6 shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*.
+Rows 7 and 8 show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
 
-We observe that for all datasets except *OL* and across all test settings, the average distance of counterfactuals from observed samples in the target class is reduced, indicating improved plausibility.
-The magnitude of improvements varies by dataset: for the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*).
-For the real-world tabular datasets, improvements are generally smaller but still substantial in many cases with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*.
+For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility.
+The magnitude of improvements varies.
+For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*).
+For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*.
 For our only vision dataset (*MNIST*), distances are reduced by up to 9%.
-The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by looking at the absolute outcome.
+The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome.
 Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*).
@@
 
-### Actionability {#sec-act}
+### Actionability. {#sec-act}
 
@@
 
-### Impact of hyperparameter settings {#sec-hyperparameters}
-We test in-depth the impact of three types of hyperparameters; our complete results are in the appendix.
-
+### Impact of hyperparameter settings. {#sec-hyperparameters}
+We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
 ***Hyperparameters of the CE generators.*** First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time.
@@
 In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.
-
 ***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs.
@@
 While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
-
 ***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.

### 2025-03-14 paper/sections/intro.qmd
Shorten to slightly less than 16 pages by restructuring

@@
 
-**Specifically, we propose counterfactual training (CT)**: a novel training regime that aligns learned representations with plausible explanations compliant with user requirements.
+**Specifically, we propose counterfactual training (CT)**: a novel training regime that aligns learned representations with plausible explanations respecting actionability constraints.
 The remainder of this paper is structured as follows.
 @sec-lit presents related work, focusing on the link between adversarial examples and counterfactual explanations.
-Then follow our main contributions:
-
-1. In @sec-method, we introduce our methodological framework and show theoretically that it can be used to enforce global actionability constraints.  
-2. In @sec-experiments, through extensive experiments we demonstrate that CT substantially improves explainability without sacrificing predictive performance.
-
+Then follow our two principal contributions.
+In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints.
+In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability without sacrificing predictive performance.
 We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.

### 2025-03-14 paper/sections/method.qmd
Shorten to slightly less than 16 pages by restructuring

@@
 
-@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
+As observed by @grathwohl2020your, any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
 The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$.
@@
 This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments.
-Under certain conditions, this results holds theoretically:^[For the proof, see the supplementary appendix.]
+Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]
 
@@
 
-It is worth highlighting that \mbox{Prp.
-\ref{prp-mtblty}} assumes independence of features.
+It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
 This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected.
@@
 We have simulated this scenario using synthetic data with two independent features and Gaussian class-conditional densities in @fig-poc.
-The four panels in @fig-poc show the outcomes for different training procedures using the same model architecture each time (a linear classifier).
+The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier).
+In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we used CT.
+Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time.
 In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$.
-Stars indicate counterfactuals in the target class generated at test time using generic gradient descent until convergence.
+Stars indicate CEs in the target class generated at test time using generic gradient descent until convergence.
 
-In panel (a), we have trained our model conventionally, and we do not impose mutability constraints at test time.
-The generated counterfactuals are all valid, but not plausible: they do not comply with the distribution of the factual samples in the target class to the point where they are clearly distinguishable from the ground-truth data.
-In panel (b), we have trained our model with CT, once again without any mutability constraints.
-We observe that the counterfactuals are highly plausible, meeting the first objective of Def.
+In all cases the counterfactuals are valid but their quality differs.
+In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster.
+In panel (b), they are highly plausible, meeting the first objective of Def.
 \ref{def-explainability}.
-
-In panel (c), we have used conventional training again, this time imposing the mutability constraint on *age* at test time.
-Counterfactuals are valid but involve some substantial reductions in *debt* for some individuals (very young applicants).
-By comparison, counterfactual paths are shorter on average in panel (d), where we have used CT and protected the immutable feature as described in @sec-constraints.
-We observe that due to the classifier's lower sensitivity to *age*, recourse recommendations with respect to *debt* are much more homogenous and do not disproportionately punish younger individuals.
-The counterfactuals are also plausible with respect to the mutable feature.
+In panel (c), the CEs involve substantial reductions in *debt* for younger applicants.
+By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints.
+Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals.
+These counterfactuals are also plausible with respect to the mutable feature.
 Thus, we consider the model in panel (d) as the most explainable according to Def.

### 2025-03-14 paper/sections/related.qmd
Shorten to slightly less than 16 pages by restructuring

@@
 
-In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation.
+In the natural language processing (NLP) domain, CEs have also been used to improve models through data augmentation.
 @wu2021polyjuice propose *Polyjuice*, a general-purpose counterfactual generator for language models.
-They demonstrate empirically that the augmentation of training data through *Polyjuice* counterfactuals improves robustness in a number of NLP tasks.
+The authors empirically demonstrate that the augmentation of training data with their method improves robustness in a number of NLP tasks.
 @balashankar2023improving similarly use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%.
-Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models through a three-step procedure.
-First, the authors identify training samples that are subject to high predictive uncertainty.
-Second, they generate counterfactual explanations for those samples.
-Finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.
+Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models through a three-step procedure: the authors identify training samples that are subject to high predictive uncertainty, generate CEs for those samples, and then fine-tune the language model on a dataset augmented with the CEs.
 
@@
 Additionally, the proposed method allows for imposing a set of actionability constraints ex-ante.
-For example, users can specify that certain features (e.g., *age*, *gender*) are immutable.
-Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes counterfactual explanations as part of the training procedure.
+For example, users can specify that certain features are immutable.
+Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes CEs as part of the training procedure.
 In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable.
 Counterfactuals are generated during each training iteration and fed back to the predictor network.
-In contrast to @guo2023counternet, we impose no restrictions on the neural network architecture at all.
+In contrast to @guo2023counternet, we impose no restrictions on the ANN architecture at all.
 

### 2025-03-14 paper/sections/other/main_results.qmd
ok

@@
 using CTExperiments.DataFrames
+using CTExperiments.Plots
 using CTExperiments.StatsBase
+using Random
 
@@
 
-::: {.callout-note}
+```{julia}
+#| eval: false
+
+Random.seed!(42)    # change seed for different outcome
+overwrite = false   # set to `true` if you want to overwrite the file
+fname = joinpath(res_dir,"mlp/mnist/grid_config.toml")
+output_dir = mkpath("paper/experiments/output/extra/")
+gen = CTExperiments.GeneratorParams(lambda_energy=25.0)
+conv = "max_iter"
+cfg = EvaluationConfig(
+    grid_file=fname, 
+    counterfactual_params=(conv=conv, generator_params=gen), 
+    save_dir=output_dir,
+    test_time=true
+)
+generate_factual_target_pairs(cfg; overwrite, nce=10)
+plts = plot_ce(MNIST(), cfg; byvars="objective")
+savefig(plts["objective"]["full"]["ecco"], "paper/figures/mnist_mlp.png")
+savefig(plts["objective"]["vanilla"]["ecco"], "paper/figures/mnist_mlp_vanilla.png")
+```
 
 @fig-mnist shows much more plausible (faithful) counterfactuals for a model with CT than the model with conventional training (@fig-mnist-vanilla).
-In fact, this is not even using *ECCo+* and still showing better results than the best results we achieved in our AAAI paper for JEM ensembles.
 
-:::
-
+::: {layout="[10,-2,10]" layout-valign="top"}
 ![Counterfactual images for *MLP* with counterfactual training.
+Factual images are shown on the diagonal, with the corresponding counterfactual for each target class (columns) in that same row.
 The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}
 
-![Counterfactual images for *MLP* with conventional training.
-The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp_vanilla.png){#fig-mnist-vanilla}
+![The same setup, factuals, model architecture and generator as in @fig-mnist, but the model was trained with CT.](/paper/figures/mnist_mlp_vanilla.png){#fig-mnist-vanilla}
+:::

### 2025-03-14 paper/experiments/CTExperiments/src/plots.jl

@@ -412,7 +412,7 @@ function plot_ce(data::MNIST, df::DataFrame, factual::Int, target::Int; axis=def
     if length(x) == 0
         plt = Plots.plot(; axis=([], false), size=values(axis))
     else
-        @assert length(x) == 1 "Expected 1 value, got $(length(x))."
+        # @assert length(x) == 1 "Expected 1 value, got $(length(x))."
         x = x[1]
         if target == factual
             title = "Factual"
@@ -482,6 +482,7 @@ function plot_ce(
     nsamples::Int=25,
     kwrgs...,
 )
+
     M = load_results(exper)[3]
     generator = CTExperiments.get_generator(params.generator_params)
     conv = CTExperiments.get_convergence(params)
@@ -489,11 +490,6 @@ function plot_ce(
     if isnothing(target)
         @info "No target supplied, choosing first label."
         target = data.y_levels[1]
-    end
-
-    # Cannot plot counterfactuals for OmnisionGenerator:
-    if isa(generator, OmniscientGenerator)
-        return Plots.plot(M, data; size=(500, 500), cb=false, target=target, kwrgs...)
     end
 
     candidates = findall(predict_label(M, data) .!= target)

### 2025-03-14 paper/preprint/paper.tex

@@
 
+\usepackage[title]{appendix}
+\usepackage{placeins}
 \usepackage{amsthm}
@@
 \newcommand{\runninghead}{A Preprint }
-\renewcommand{\runninghead}{counterfactual training (A Preprint) }
+\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
 \title{Counterfactual Training: Teaching Models Plausible and Actionable
@@
 directly from data, guided by narrow objectives such as predictive
-accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow, Bengio, and
+accuracy (\citeproc{ref-goodfellow2016deep}{Goodfellow, Bengio, and
 Courville 2016}). Modern advances in computing have made it possible to
@@
 (DL) revolution, Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2013}) showed that artificial
+(\citeproc{ref-szegedy2013intriguing}{2014}) showed that artificial
 neural networks (ANN) are sensitive to adversarial examples (AEs):
@@
 notably \textbf{adversarial training}
-(\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow, Shlens, and
-Szegedy 2014})---truly robust deep learning remains unattainable even
-for models that are considered ``shallow'' by today's standards
+(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
+2015})---truly robust deep learning remains unattainable even for models
+that are considered ``shallow'' by today's standards
 (\citeproc{ref-kolter2023keynote}{Kolter 2023}).
@@
 closely aligned with Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2013}) who define an adversarial
+(\citeproc{ref-szegedy2013intriguing}{2014}) who define an adversarial
 attack as an ``imperceptible non-random perturbation''. Thus, we choose
@@
 descent in the feature space for optimization
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
-In this context, domain constraints can be imposed by simply projecting
-counterfactuals back to the specified domain, if the previous gradient
-step resulted in updated feature values that were out-of-domain.
-Mutability constraints can similarly be enforced by setting partial
-derivatives to zero to ensure that features are only perturbed in the
-allowed direction, if at all.
+(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, and Liem
+2023}). In this context, domain constraints can be imposed by simply
+projecting counterfactuals back to the specified domain, if the previous
+gradient step resulted in updated feature values that were
+out-of-domain. Mutability constraints can similarly be enforced by
+setting partial derivatives to zero to ensure that features are only
+perturbed in the allowed direction, if at all.
 
@@
 assess how similar counterfactuals are to observed samples in the target
-domain. We rely on the distance-based metric used by Altmeyer et al.
+domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
+distance-based metric used by Altmeyer et al.
 (\citeproc{ref-altmeyer2024faithful}{2024}),
@@
 
-\caption{\label{tbl-main}Key plausibility and predictive performance
-metrics for all datasets. The top five rows show the percentage
-reduction in implausibility according to Equation~\ref{eq-impl-dist} for
-varying degrees of the energy penalty used for \emph{ECCo} at test time.
-The following row shows the reduction in implausibility as measured by
-Equation~\ref{eq-impl-div} and aggregated across all test specifications
-of \emph{ECCo}. The final two rows show the test accuracies for the
-model trained with CT and conventionally trained models (``vanilla'').}
+\caption{\label{tbl-main}Key explainability and predictive performance
+metrics for all datasets. \textbf{Plausibility}: the top five rows show
+the percentage reduction in implausibility according to
+Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
+used for \emph{ECCo} at test time. The following row shows the reduction
+in implausibility as measured by Equation~\ref{eq-impl-div} and
+aggregated across all test specifications. \textbf{Accuracy}: The
+following two rows show the test accuracies for the models trained with
+CT and the baseline. \textbf{Actionability}: The final row present the
+average reduction in costs when imposing mutability constraints.}
 
@@
 Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
-den Hengel. 2020. {``Counterfactual Vision and Language Learning.''} In
-\emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
+den Hengel. 2020. {``{Counterfactual Vision and Language Learning}.''}
+In \emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
 Recognition (CVPR)}, 10041--51.
@@
 \bibitem[\citeproctext]{ref-altmeyer2023explaining}
-Altmeyer, Patrick, Arie van Deursen, et al. 2023. {``Explaining
-Black-Box Models Through Counterfactuals.''} In \emph{Proceedings of the
-JuliaCon Conferences}, 1:130. 1.
+Altmeyer, Patrick, Arie van Deursen, and Cynthia C. S. Liem. 2023.
+{``{Explaining Black-Box Models through Counterfactuals}.''} In
+\emph{Proceedings of the JuliaCon Conferences}, 1:130. 1.
 
 \bibitem[\citeproctext]{ref-altmeyer2024faithful}
-Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia CS
-Liem. 2024. {``Faithful Model Explanations Through Energy-Constrained
-Conformal Counterfactuals.''} In \emph{Proceedings of the AAAI
-Conference on Artificial Intelligence}, 38:10829--37. 10.
+Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia C.
+S. Liem. 2024. {``{Faithful Model Explanations through
+Energy-Constrained Conformal Counterfactuals}.''} In \emph{Proceedings
+of the Thirty-Eighth AAAI Conference on Artificial Intelligence},
+38:10829--37. 10. \url{https://doi.org/10.1609/aaai.v38i10.28956}.
 
@@
 Augustin, Maximilian, Alexander Meinke, and Matthias Hein. 2020.
-{``Adversarial Robustness on in-and Out-Distribution Improves
-Explainability.''} In \emph{European Conference on Computer Vision},
-228--45. Springer.
+{``{Adversarial Robustness on In- and Out-Distribution Improves
+Explainability}.''} In \emph{Computer Vision -- ECCV 2020}, edited by
+Andrea Vedaldi, Horst Bischof, Thomas Brox, and Jan-Michael Frahm,
+228--45. Cham: Springer.
 
@@
 Balashankar, Ananth, Xuezhi Wang, Yao Qin, Ben Packer, Nithum Thain, Ed
-Chi, Jilin Chen, and Alex Beutel. 2023. {``Improving Classifier
-Robustness Through Active Generative Counterfactual Data
-Augmentation.''} In \emph{Findings of the Association for Computational
-Linguistics: EMNLP 2023}, 127--39.
+Chi, Jilin Chen, and Alex Beutel. 2023. {``{Improving Classifier
+Robustness through Active Generative Counterfactual Data
+Augmentation}.''} In \emph{Findings of the Association for Computational
+Linguistics: EMNLP 2023}, 127--39. ACL.
+\url{https://doi.org/10.18653/v1/2023.findings-emnlp.10}.
 
@@
 Bell, Andrew, Joao Fonseca, Carlo Abrate, Francesco Bonchi, and Julia
-Stoyanovich. 2024. {``Fairness in Algorithmic Recourse Through the Lens
-of Substantive Equality of Opportunity.''}
+Stoyanovich. 2024. {``{Fairness in Algorithmic Recourse Through the Lens
+of Substantive Equality of Opportunity}.''}
 \url{https://arxiv.org/abs/2401.16088}.
@@
 \bibitem[\citeproctext]{ref-du2019implicit}
-Du, Yilun, and Igor Mordatch. 2020. {``Implicit Generation and
-Generalization in Energy-Based Models.''}
+Du, Yilun, and Igor Mordatch. 2020. {``{Implicit Generation and
+Generalization in Energy-Based Models}.''}
 \url{https://arxiv.org/abs/1903.08689}.
@@
 Optimization}.''} In \emph{Proceedings of the 34th International
-Conference on Machine Learning}, edited by Doina Precup and Yee Whye
-Teh, 70:1165--73. Proceedings of Machine Learning Research. PMLR.
-\url{https://proceedings.mlr.press/v70/franceschi17a.html}.
+Conference on Machine Learning}, 1165--73. ICML'17. Sydney, NSW,
+Australia: JMLR.org.
 
 \bibitem[\citeproctext]{ref-frankle2018lottery}
-Frankle, Jonathan, and Michael Carbin. 2019. {``The Lottery Ticket
-Hypothesis: Finding Sparse, Trainable Neural Networks.''} In
+Frankle, Jonathan, and Michael Carbin. 2019. {``{The Lottery Ticket
+Hypothesis: Finding Sparse, Trainable Neural Networks}.''} In
 \emph{International Conference on Learning Representations}.
-\url{https://openreview.net/forum?id=rJl-b3RcF7}.
 
 \bibitem[\citeproctext]{ref-freiesleben2022intriguing}
-Freiesleben, Timo. 2022. {``The Intriguing Relation Between
-Counterfactual Explanations and Adversarial Examples.''} \emph{Minds and
-Machines} 32 (1): 77--109.
-
-\bibitem[\citeproctext]{ref-goodfellow2014explaining}
-Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
-{``Explaining and Harnessing Adversarial Examples.''}
-\url{https://arxiv.org/abs/1412.6572}.
+Freiesleben, Timo. 2022. {``{The Intriguing Relation Between
+Counterfactual Explanations and Adversarial Examples}.''} \emph{Minds
+and Machines} 32 (1): 77--109.
 
@@
 {Learning}}. {MIT Press}.
+
+\bibitem[\citeproctext]{ref-goodfellow2014explaining}
+Goodfellow, Ian, Jonathon Shlens, and Christian Szegedy. 2015.
+{``{Explaining and Harnessing Adversarial Examples}.''}
+\url{https://arxiv.org/abs/1412.6572}.
 
@@
 \bibitem[\citeproctext]{ref-guidotti2022counterfactual}
-Guidotti, Riccardo. 2022. {``Counterfactual Explanations and How to Find
-Them: Literature Review and Benchmarking.''} \emph{Data Mining and
-Knowledge Discovery}, 1--55.
+Guidotti, Riccardo. 2022. {``{Counterfactual Explanations and How to
+Find Them: Literature Review and Benchmarking}.''} \emph{Data Mining and
+Knowledge Discovery} 38 (5): 2770--2824.
+\url{https://doi.org/10.1007/s10618-022-00831-6}.
 
 \bibitem[\citeproctext]{ref-guo2023counternet}
-Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``CounterNet:
-End-to-End Training of Prediction Aware Counterfactual Explanations.''}
+Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``{CounterNet:
+End-to-End Training of Prediction Aware Counterfactual Explanations}.''}
 In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
-Discovery and Data Mining}, 577--89. KDD '23. New York, NY, USA:
+Discovery and Data Mining}, 577-\/-589. KDD '23. New York, NY, USA:
 Association for Computing Machinery.
@@
 Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
-Joydeep Ghosh. 2019. {``Towards Realistic Individual Recourse and
-Actionable Explanations in Black-Box Decision Making Systems.''}
+Joydeep Ghosh. 2019. {``{Towards Realistic Individual Recourse and
+Actionable Explanations in Black-Box Decision Making Systems}.''}
 \url{https://arxiv.org/abs/1907.09615}.
@@
 Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
-IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)},
-xvi--. Los Alamitos, CA, USA: IEEE Computer Society.
+IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)}. Los
+Alamitos, CA, USA: IEEE Computer Society.
 \url{https://doi.org/10.1109/SaTML54575.2023.00009}.
@@
 {``Simple and Scalable Predictive Uncertainty Estimation Using Deep
-Ensembles.''} \emph{Advances in Neural Information Processing Systems}
-30.
+Ensembles.''} In \emph{Proceedings of the 31st International Conference
+on Neural Information Processing Systems}, 6405--16. NIPS'17. Red Hook,
+NY, USA: Curran Associates Inc.
 
@@
 \bibitem[\citeproctext]{ref-luu2023counterfactual}
-Luu, Hoai Linh, and Naoya Inoue. 2023. {``Counterfactual Adversarial
-Training for Improving Robustness of Pre-Trained Language Models.''} In
+Luu, Hoai Linh, and Naoya Inoue. 2023. {``{Counterfactual Adversarial
+Training for Improving Robustness of Pre-trained Language Models}.''} In
 \emph{Proceedings of the 37th Pacific Asia Conference on Language,
-Information and Computation}, 881--88.
+Information and Computation}, 881--88. ACL.
+\url{https://aclanthology.org/2023.paclic-1.88/}.
 
@@
 \bibitem[\citeproctext]{ref-ross2021learning}
-Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024. {``Learning
-Models for Actionable Recourse.''} In \emph{Proceedings of the 35th
-International Conference on Neural Information Processing Systems}. NIPS
-'21. Red Hook, NY, USA: Curran Associates Inc.
+Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024.
+{``{Learning Models for Actionable Recourse}.''} In \emph{Proceedings of
+the 35th International Conference on Neural Information Processing
+Systems}. NIPS '21. Red Hook, NY, USA: Curran Associates Inc.
 
 \bibitem[\citeproctext]{ref-sauer2021counterfactual}
-Sauer, Axel, and Andreas Geiger. 2021. {``Counterfactual Generative
-Networks.''} \url{https://arxiv.org/abs/2101.06046}.
+Sauer, Axel, and Andreas Geiger. 2021. {``{Counterfactual Generative
+Networks}.''} \url{https://arxiv.org/abs/2101.06046}.
 
 \bibitem[\citeproctext]{ref-schut2021generating}
-Schut, Lisa, Oscar Key, Rory Mc Grath, Luca Costabello, Bogdan
-Sacaleanu, Yarin Gal, et al. 2021. {``Generating {Interpretable
-Counterfactual Explanations By Implicit Minimisation} of {Epistemic} and
-{Aleatoric Uncertainties}.''} In \emph{International {Conference} on
-{Artificial Intelligence} and {Statistics}}, 1756--64. {PMLR}.
+Schut, Lisa, Oscar Key, Rory McGrath, Luca Costabello, Bogdan Sacaleanu,
+Yarin Gal, et al. 2021. {``Generating {Interpretable Counterfactual
+Explanations By Implicit Minimisation} of {Epistemic} and {Aleatoric
+Uncertainties}.''} In \emph{International {Conference} on {Artificial
+Intelligence} and {Statistics}}, 1756--64. {PMLR}.
 
 \bibitem[\citeproctext]{ref-sharma2020certifai}
-Sharma, Shubham, Jette Henderson, and Joydeep Ghosh. 2020. {``CERTIFAI:
+Sharma, Shubham, Jette Henderson, and Joydeep Ghosh. 2020. {``{CERTIFAI:
 A Common Framework to Provide Explanations and Analyse the Fairness and
-Robustness of Black-Box Models.''} In \emph{Proceedings of the AAAI/ACM
+Robustness of Black-box Models}.''} In \emph{Proceedings of the AAAI/ACM
 Conference on AI, Ethics, and Society}, 166--72. AIES '20. New York, NY,
@@
 Bayesian Optimization of Machine Learning Algorithms}.''} In
-\emph{Advances in Neural Information Processing Systems}, edited by F.
-Pereira, C. J. Burges, L. Bottou, and K. Q. Weinberger. Vol. 25. Curran
+\emph{Proceedings of the 26th International Conference on Neural
+Information Processing Systems - Volume 2}, edited by F. Pereira, C. J.
+Burges, L. Bottou, and K. Q. Weinberger. Vol. 25. NIPS'12. Curran
 Associates, Inc.
-\url{https://proceedings.neurips.cc/paper_files/paper/2012/file/05311655a15b75fab86956663e1819cd-Paper.pdf}.
 
@@
 and Daniele Magazzeni. 2021. {``{Counterfactual Explanations for
-Arbitrary Regression Models}.''} \emph{CoRR} abs/2106.15212.
-\url{https://arxiv.org/abs/2106.15212}.
+Arbitrary Regression Models}.''} \url{https://arxiv.org/abs/2106.15212}.
 
@@
 Szegedy, Christian, Wojciech Zaremba, Ilya Sutskever, Joan Bruna,
-Dumitru Erhan, Ian Goodfellow, and Rob Fergus. 2013. {``Intriguing
+Dumitru Erhan, Ian Goodfellow, and Rob Fergus. 2014. {``Intriguing
 Properties of Neural Networks.''} \url{https://arxiv.org/abs/1312.6199}.
@@
 {``Learning What Makes a Difference from Counterfactual Examples and
-Gradient Supervision.''} In \emph{Computer Vision--ECCV 2020: 16th
-European Conference, Glasgow, UK, August 23--28, 2020, Proceedings, Part
-x 16}, 580--99. Springer.
+Gradient Supervision.''} In \emph{Computer Vision - ECCV 2020}, 580--99.
+Berlin, Heidelberg: Springer-Verlag.
+\url{https://doi.org/10.1007/978-3-030-58607-2_34}.
 
@@
 \bibitem[\citeproctext]{ref-wilson2020case}
-Wilson, Andrew Gordon. 2020. {``The Case for Bayesian Deep Learning.''}
-\url{https://arxiv.org/abs/2001.10995}.
+Wilson, Andrew Gordon. 2020. {``{The Case for Bayesian Deep
+Learning}.''} \url{https://arxiv.org/abs/2001.10995}.
 
@@
 (Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
-and Roberto Navigli, 6707--23. Online: Association for Computational
-Linguistics. \url{https://doi.org/10.18653/v1/2021.acl-long.523}.
+and Roberto Navigli, 6707--23. Online: ACL.
+\url{https://doi.org/10.18653/v1/2021.acl-long.523}.
 
@@
 
+\begin{appendices}
+
 \FloatBarrier
-
-\renewcommand{\thesection}{\Alph{section}}
-
-\renewcommand{\thetable}{A\arabic{table}}
-
-\renewcommand{\thefigure}{A\arabic{figure}}
 
@@
 }\end{equation}
+
+In our implementation, Equation~\ref{eq-mmd} is by default applied to
+the entire subset of the training data for which \(y=y^+\).
 
@@
 
-\section{Details on Main Experiments}\label{details-on-main-experiments}
+\section{Details on Main Experiments}\label{sec-app-main}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
-
 Figure~\ref{fig-mnist} shows much more plausible (faithful)
 counterfactuals for a model with CT than the model with conventional
-training (Figure~\ref{fig-mnist-vanilla}). In fact, this is not even
-using \emph{ECCo+} and still showing better results than the best
-results we achieved in our AAAI paper for JEM ensembles.
-
-\end{tcolorbox}
-
-\begin{figure}
+training (Figure~\ref{fig-mnist-vanilla}).
+
+\begin{figure}
+
+\begin{minipage}{0.46\linewidth}
+
+\begin{figure}[H]
 
@@
 \caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
-counterfactual training. The underlying generator, \emph{ECCo}, aims to
-generate counterfactuals that are faithful to the model
+counterfactual training. Factual images are shown on the diagonal, with
+the corresponding counterfactual for each target class (columns) in that
+same row. The underlying generator, \emph{ECCo}, aims to generate
+counterfactuals that are faithful to the model
 (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
@@
 
-\begin{figure}
+\end{minipage}%
+%
+\begin{minipage}{0.09\linewidth}
+~\end{minipage}%
+%
+\begin{minipage}{0.46\linewidth}
+
+\begin{figure}[H]
 
@@
 
-\caption{\label{fig-mnist-vanilla}Counterfactual images for \emph{MLP}
-with conventional training. The underlying generator, \emph{ECCo}, aims
-to generate counterfactuals that are faithful to the model
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
-
-\end{figure}%
+\caption{\label{fig-mnist-vanilla}The same setup, factuals, model
+architecture and generator as in Figure~\ref{fig-mnist}, but the model
+was trained with CT.}
+
+\end{figure}%
+
+\end{minipage}%
+
+\end{figure}%
+
+\FloatBarrier
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 
-\subsubsection{Accuracy}\label{accuracy}
+\subsubsection{Predictive Performance}\label{predictive-performance-1}
+
+Predictive performance measures for this grid search are shown in
+Table~\ref{tbl-acc-gen}.
 
@@
 
-\subsubsection{Plausibility}\label{plausibility-1}
+\subsubsection{Plausibility}\label{plausibility}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 
-\subsubsection{Accuracy}\label{accuracy-1}
+\subsubsection{Predictive Performance}\label{predictive-performance-2}
+
+Predictive performance measures for this grid search are shown in
+Table~\ref{tbl-acc-pen}.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 
-\subsubsection{Accuracy}\label{accuracy-2}
+\subsubsection{Predictive Performance}\label{predictive-performance-3}
+
+Predictive performance measures for this grid search are shown in
+Table~\ref{tbl-acc-train}.
 
@@
 \end{figure}%
+
+\FloatBarrier
 
@@
 the energy regularization \(\lambda_{\text{reg}}\). The final
-hyperparameter choices for each dataset are presented in \textbf{ADD
-TABLE}. Detailed results for each data set are shown in
-Figure~\ref{fig-tune-plaus-adult} to Figure~\ref{fig-tune-mat-over}.
-From \textbf{ADD TABLE}, we notice that the same decision threshold of
-\(\tau=0.5\) is optimal for all but on dataset. We attribute this to the
-fact that a low decision threshold results in a higher share of mature
-counterfactuals and hence more opportunities for the model to learn from
-examples (Figure~\ref{fig-tune-mat-adult} to
-Figure~\ref{fig-tune-mat-over}). This has played a role in particular
-for our real-world tabular datasets and MNIST, which suffered from low
-levels of maturity for higher decision thresholds. In cases where
-maturity is not an issue, as for \emph{Moons}, higher decision
-thresholds lead to better outcomes, which may have to do with the fact
-that the resulting counterfactuals are more faithful to the model.
-Concerning the regularization strength, we find somewhat high variation
-across datasets. Most notably, we find that relatively low levels of
-regularization are optimal for MNIST. We hypothesize that this finding
-may be attributed to the uniform scaling of all input features (digits).
+hyperparameter choices for each dataset are presented in
+Table~\ref{tbl-final-params} in Section~\ref{sec-app-main}. Detailed
+results for each data set are shown in Figure~\ref{fig-tune-plaus-adult}
+to Figure~\ref{fig-tune-mat-over}. From Table~\ref{tbl-final-params}, we
+notice that the same decision threshold of \(\tau=0.5\) is optimal for
+all but on dataset. We attribute this to the fact that a low decision
+threshold results in a higher share of mature counterfactuals and hence
+more opportunities for the model to learn from examples
+(Figure~\ref{fig-tune-mat-adult} to Figure~\ref{fig-tune-mat-over}).
+This has played a role in particular for our real-world tabular datasets
+and MNIST, which suffered from low levels of maturity for higher
+decision thresholds. In cases where maturity is not an issue, as for
+\emph{Moons}, higher decision thresholds lead to better outcomes, which
+may have to do with the fact that the resulting counterfactuals are more
+faithful to the model. Concerning the regularization strength, we find
+somewhat high variation across datasets. Most notably, we find that
+relatively low levels of regularization are optimal for MNIST. We
+hypothesize that this finding may be attributed to the uniform scaling
+of all input features (digits).
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Package Version (Reproducibility)}, toprule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, colframe=quarto-callout-warning-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Package Version (Reproducibility)}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, colbacktitle=quarto-callout-warning-color!10!white, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-tune-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, rightrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colframe=quarto-callout-note-color-frame, breakable, opacityback=0, colback=white, bottomtitle=1mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
 
@@
 \href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{CounterfactualExplanations.jl}
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023})
-and the \href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library for
-deep learning (\citeproc{ref-innes2018fashionable}{Michael Innes et al.
-2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). For
+(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, and Liem
+2023}) and the \href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library
+for deep learning (\citeproc{ref-innes2018fashionable}{Michael Innes et
+al. 2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). For
 data-wrangling and presentation-ready tables we relied on
@@
 
+\end{appendices}
+
 \end{document}

### 2025-03-14 paper/quarto_ecml/paper.tex
Fix indentation in hyperparameter results and discussion

@@
 results are in the supplementary appendix.\\
-\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
+\indent \emph{Hyperparameters of the CE generators.} First, we observe
 that CT is highly sensitive to hyperparameter settings but (a) there are
@@
 of mature counterfactuals.\\
-\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
+\indent \emph{Hyperparameters for penalties.} We find that the strength
 of the energy regularization, \(\lambda_{\text{reg}}\) is highly
@@
 \(\lambda_{\text{adv}}\) is much less evident. While high values of
-\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
-combined with high values of \(\lambda_{\text{div}}\) or
-\(\lambda_{\text{adv}}\), this effect is not very pronounced.\\
-\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
+\(\lambda_{\text{reg}}\) may increase the variability in outcomes
+(especially) when combined with high values of \(\lambda_{\text{div}}\)
+or \(\lambda_{\text{adv}}\), this effect is not very pronounced.\\
+\indent \emph{Other hyperparameters.} We observe that the effectiveness
 and stability of CT is positively associated with the number of
@@
 As our results indicate, counterfactual training produces models that
-are more explainable. However, our approach is not without
+are more explainable. Nonetheless, it brings about three important
 limitations.\\
-\textbf{\emph{CT increases the training time of models.}} CT can be more
+\indent \emph{CT increases the training time of models.} CT can be more
 time-consuming than conventional training regimes. While higher numbers
@@
 own experiments.\\
-\textbf{\emph{Immutable features may have proxies.}} We propose an
+\indent \emph{Immutable features may have proxies.} We propose an
 approach to protect immutable features and thus increase the
@@
 the capacity of CT to fulfill its intended goal.\\
-\textbf{\emph{Interventions on features may impact fairness.}} We
-provide a tool that allows practitioners to modify the sensitivity of a
-model with respect to certain features, which may have implication for
-the fair and equitable treatment of decision subjects. As protecting a
-set of features leads the model to assign higher relative importance to
+\indent \emph{Interventions on features may impact fairness.} We provide
+a tool that allows practitioners to modify the sensitivity of a model
+with respect to certain features, which may have implication for the
+fair and equitable treatment of decision subjects. As protecting a set
+of features leads the model to assign higher relative importance to
 unprotected features, model owners could misuse our solution by
@@
 
-We also highlight several interesting directions for future research.\\
-\textbf{\emph{Extending CT beyond classification settings.}} Our
-formulation relies on the distinction between non-target class(es)
-\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
-through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
-arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
-discrete. Thus, it does not apply to ML tasks where the change in
-outcome cannot be readily quantified. Focus on classification models is
-a common restriction in research on CEs and AR. Other settings have
-attracted some interest (e.g., regression in
+We also highlight several important directions for future research.
+Firstly, it is an interesting challenge to extend CT beyond
+classification settings. Our formulation relies on the distinction
+between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
+generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
+and \(y^{+}\) can be arbitrarily defined, CT requires the output space
+\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
+where the change in outcome cannot be readily quantified. Focus on
+classification models is a common restriction in research on CEs and AR.
+Other settings have attracted some interest (e.g., regression in
 \cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
 little consensus how to robustly extend the notion of CEs.\\
-\textbf{\emph{Addressing the training instabilities.}} JEMs are
-susceptible to instabilities during training \cite{grathwohl2020your}
-and even though we depart from the SGLD-based sampling, we still
-encounter major variability in the outcomes. CT is exposed to two
-potential sources of instabilities: (1) the energy-based contrastive
-divergence term in Equation~\ref{eq-div}, and (2) the underlying
-counterfactual explainers. Still, we find that training instabilities
-can be successfully mitigated by regularizing energy
-(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
-during each training epoch, and including only mature counterfactuals
-for contrastive divergence.\\
-\textbf{\emph{Improving hyperparameter selection procedures.}} Our
-method benefits from the tuning of certain key hyperparameters (see
-Section~\ref{sec-hyperparameters}). In this work, we have relied
-exclusively on grid search for this task. Future work on CT could
-benefit from investigating more sophisticated approaches towards
-hyperparameter tuning. Notably, CT is iterative which makes a variety of
-methods applicable, including Bayesian \cite{snoek2012practical} or
-gradient-based \cite{franceschi2017forward} optimization.
+\indent Secondly, our approach is susceptible to training instabilities.
+This problem has been recognized for JEMs \cite{grathwohl2020your} and
+even though we depart from the SGLD-based sampling, we still encounter
+considerable variability in the outcomes. CT is exposed to two potential
+sources of instabilities: (1) the energy-based contrastive divergence
+term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
+explainers. We find several promising ways to mitigate this problem:
+regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
+many counterfactuals during each epoch, and including only mature
+counterfactuals for contrastive divergence.\\
+\indent Finally, we believe that it is possible to substantially improve
+hyperparameter selection procedures. Our method benefits from the tuning
+of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
+In this work, we have relied exclusively on grid search for this task.
+Future work on CT could benefit from investigating more sophisticated
+approaches. Notably, CT is iterative which makes methods such as
+Bayesian (e.g., \cite{snoek2012practical}) or gradient-based (e.g.,
+\cite{franceschi2017forward}) optimization applicable.
 

### 2025-03-14 paper/sections/discussion.qmd
Fix indentation in hyperparameter results and discussion

@@
 
-As our results indicate, counterfactual training produces models that are more explainable. However, our approach is not without limitations.  
-***CT increases the training time of models.*** CT can be more time-consuming than conventional training regimes.
+As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
+`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes.
 While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations.
@@
 Third, it yields itself to parallel execution, which we have leveraged for our own experiments.
-***Immutable features may have proxies.*** We propose an approach to protect immutable features and thus increase the actionability of the generated CEs.
+`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs.
 However, it requires that model owners define the mutability constraints for (all) features considered by the model.
@@
 Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.
-***Interventions on features may impact fairness.***  We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects.
+`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects.
 As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.
@@
 
-We also highlight several interesting directions for future research.  
-***Extending CT beyond classification settings.*** Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj.
+We also highlight several important directions for future research.
+Firstly, it is an interesting challenge to extend CT beyond classification settings.
+Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj.
 While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete.
@@
 Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of CEs.
-***Addressing the training instabilities.*** JEMs are susceptible to instabilities during training [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter major variability in the outcomes.
+`\indent`{=latex} Secondly, our approach is susceptible to training instabilities.
+This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter considerable variability in the outcomes.
 CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers.
-Still, we find that training instabilities can be successfully mitigated by regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each training epoch, and including only mature counterfactuals for contrastive divergence.
-***Improving hyperparameter selection procedures.*** Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters).
+We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.
+`\indent`{=latex} Finally, we believe that it is possible to substantially improve hyperparameter selection procedures.
+Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters).
 In this work, we have relied exclusively on grid search for this task.
-Future work on CT could benefit from investigating more sophisticated approaches towards hyperparameter tuning.
-Notably, CT is iterative which makes a variety of methods applicable, including Bayesian [e.g., @snoek2012practical] or gradient-based [e.g., @franceschi2017forward] optimization.
+Future work on CT could benefit from investigating more sophisticated approaches.
+Notably, CT is iterative which makes methods such as Bayesian (e.g., [@snoek2012practical]) or gradient-based (e.g., [@franceschi2017forward]) optimization applicable.

### 2025-03-14 paper/sections/experiments.qmd
Fix indentation in hyperparameter results and discussion

@@
 We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
-***Hyperparameters of the CE generators.*** First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time.
+`\indent`{=latex} *Hyperparameters of the CE generators.* First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time.
 Second, we note that the choice of a CE generator has a major impact on the results.
@@
 In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.
-***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs.
+`\indent`{=latex} *Hyperparameters for penalties.* We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs.
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
-While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
-***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
+While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes (especially) when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
+`\indent`{=latex} *Other hyperparameters.* We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
 We also confirm that a higher number of training epochs is beneficial.

### 2025-03-14 paper/quarto_ecml/paper.tex
Add description of datasets
TODO Patrick: (1) put images of synthetic datasets in the appendix, (2) is "Cred" the Default of Credit Card Clients?

@@
 but without the conformal prediction component---as a method that
-directly targets both faithfulness and plausibility of the CEs.
+directly targets both faithfulness and plausibility of the
+counterfactuals.
+
+We make use of nine classification datasets common in the CE/AR
+literature. Four of them are synthetic with two classes and different
+characteristics: linearly separable clusters (\emph{LS}), overlapping
+clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
+moons (\emph{Moon}). These datasets are generated using the library of
+\cite{altmeyer2023explaining} and we present them in the supplementary
+appendix. Next, we have four real-world binary tabular datasets from the
+domain of economics: \emph{Adult} (a.k.a. Census data) of
+\cite{becker1996adult2}, California housing (\emph{CH}) of
+\cite{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
+\cite{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) of
+\cite{kaggle2011give2}. Finally, for the convenience of illustration, we
+use of the 10-class \emph{MNIST} vision dataset \cite{lecun1998mnist}.
 

### 2025-03-14 paper/sections/experiments.qmd
Add description of datasets
TODO Patrick: (1) put images of synthetic datasets in the appendix, (2) is "Cred" the Default of Credit Card Clients?

@@
 
-We run the experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the CEs.
+We run the experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.
+
+We make use of nine classification datasets common in the CE/AR literature.
+Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*).
+These datasets are generated using the library of [@altmeyer2023explaining] and we present them in the supplementary appendix.
+Next, we have four real-world binary tabular datasets from the domain of economics: *Adult* (a.k.a.
+Census data) of [@becker1996adult2], California housing (*CH*) of [@pace1997sparse], Default of Credit Card Clients (*Cred*) of [@yeh2016default], and Give Me Some Credit (*GMSC*) of [@kaggle2011give2].
+Finally, for the convenience of illustration, we use of the 10-class *MNIST* vision dataset [@lecun1998mnist].
 

### 2025-03-14 paper/quarto_ecml/paper.tex
Parse comments from the first document
No TODOs

@@
 turn negative outcomes (e.g., ``loan application rejected'') into
-positive ones (e.g., ``application accepted''). Our work builds upon
-\textbf{counterfactual explanations} (CE) proposed by
+positive ones (e.g., ``application accepted''). In line with this, our
+work builds upon \textbf{counterfactual explanations} (CE) proposed by
 \cite{wachter2017counterfactual} as an effective approach to achieve
@@
 explanations that are both plausible and valid for any given model,
-independent of whether the learned representations are also compatible
-with plausible explanations \cite{altmeyer2024faithful}. Fortunately,
-recent findings indicate that improved explanatory capacity can arise as
-a consequence of regularization techniques aimed at other training
-objectives such as robustness, generalization, and generative capacity
+independent of whether the learned representations are in fact
+compatible with plausible explanations \cite{altmeyer2024faithful}.
+Fortunately, recent findings indicate that improved explanatory capacity
+can arise as a consequence of regularization techniques aimed at other
+training objectives such as robustness, generalization, and generative
+capacity
 \cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
@@
 
-\textbf{Specifically, we propose counterfactual training (CT)}: a novel
-training regime that aligns learned representations with plausible
-explanations respecting actionability constraints. The remainder of this
-paper is structured as follows. Section~\ref{sec-lit} presents related
-work, focusing on the link between adversarial examples and
-counterfactual explanations. Then follow our two principal
-contributions. In Section~\ref{sec-method}, we introduce our
-methodological framework and show theoretically that it can be employed
-to enforce global actionability constraints. In
-Section~\ref{sec-experiments}, through extensive experiments, we
-demonstrate that CT substantially improves explainability without
-sacrificing predictive performance. We discuss the challenges in
-Section~\ref{sec-discussion} and conclude in
+\textbf{Specifically, we introduce Counterfactual Training (CT)}: a
+novel training regime explicitly geared towards improving the
+explainability of models. In high-level terms, we define this concept as
+as the extent to which valid explanations derived for an opaque model
+are also deemed plausible with respect to the underlying data and the
+global actionability constraints. To the best of our knowledge, our
+framework represents the first attempt to address this challenge by
+employing counterfactual explanations already in the training phase.
+
+The remainder of this manuscript is structured as follows.
+Section~\ref{sec-lit} presents related work, focusing on the link
+between AEs and CEs. Then follow our two principal contributions. In
+Section~\ref{sec-method}, we introduce our methodological framework and
+show theoretically that it can be employed to enforce global
+actionability constraints. In Section~\ref{sec-experiments}, through
+extensive experiments, we demonstrate that CT substantially improves
+explainability without sacrificing predictive performance. We discuss
+the challenges in Section~\ref{sec-discussion} and conclude in
 Section~\ref{sec-conclusion} that CT is a promising approach towards
@@
 
-To the best of our knowledge, the proposed framework for counterfactual
-training represents the first attempt to use counterfactual explanations
-during training to improve model explainability. In high-level terms, we
-define model explainability as the extent to which valid explanations
-derived for an opaque model are also deemed plausible with respect to
-the underlying data and (global) actionability constraints. To make the
-desiderata for our framework more concrete, we follow
+To make the desiderata for our framework more concrete, we follow
 \cite{augustin2020adversarial} in tying the concept of explainability to
@@
 counterfactuals (i.e., AEs). Thus, by inducing models to ``unlearn''
-susceptiblity to such examples, AT can effectively remove implausible
-explanations from the solution space.
+susceptiblity to such examples, adversarial training can effectively
+remove implausible explanations from the solution space.
 
@@
 \cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
-certain conditions and derive theoretical upper bounds on distances
-between them.
+certain conditions.
 
@@
 
-\subsection{Beyond Robustness}\label{beyond-robustness}
+\subsection{Aligning Representations with Plausible
+Explanations}\label{aligning-representations-with-plausible-explanations}
 
@@
 what the model has learned about the underlying data. Formally, they
-rely on tools from energy-based modelling to minimize the divergence
-between the distribution of counterfactuals and the conditional
-posterior over inputs learned by the model. Their proposed
-counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
-and only if the underlying model has learned representations that align
-with them. The authors find that both deep ensembles
-\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
-\cite{grathwohl2020your} tend to do well in this regard.
+rely on tools from energy-based modelling \cite{teh2003energy} to
+minimize the divergence between the distribution of counterfactuals and
+the conditional posterior over inputs learned by the model. Their
+proposed counterfactual explainer, \emph{ECCCo}, yields plausible
+explanations if and only if the underlying model has learned
+representations that align with them. The authors find that both deep
+ensembles \cite{lakshminarayanan2016simple} and joint energy-based
+models (JEMs) \cite{grathwohl2020your} tend to do well in this regard.
 
@@
 \cite{schut2021generating} similarly found that generating plausible
-(``interpretable'') counterfactual explanations is almost trivial for
-deep ensembles that have also undergone adversarial training. The case
-for JEMs is even clearer: they involve a hybrid objective that induces
-both high predictive performance and generative capacity
-\cite{grathwohl2020your}. This is closely related to the idea of
-aligning models with plausible explanations and has inspired our
-proposed CT objective, as we explain in Section~\ref{sec-method}.
+(``interpretable'') CEs is almost trivial for deep ensembles that have
+also undergone adversarial training. The case for JEMs is even clearer:
+they involve a hybrid objective that induces both high predictive
+performance and generative capacity \cite{grathwohl2020your}. This is
+closely related to the idea of aligning models with plausible
+explanations and has inspired our CT objective.
 
@@
 
-Counterfactual training combines ideas from adversarial training,
-energy-based modelling and counterfactuals explanations with the
-explicit goal of aligning representations with plausible explanations
-that comply with user requirements. In the context of CEs, plausibility
-has broadly been defined as the degree to which counterfactuals comply
-with the underlying data-generating process
+In this section we propose our novel counterfactual training objective.
+In CT, we combine ideas from adversarial training, counterfactual
+explanations, and energy-based modelling with the explicit goal of
+aligning representations with plausible explanations that comply with
+user-defined actionability constraints.
+
+In the context of counterfactual explanations, plausibility has broadly
+been defined as the degree to which counterfactuals comply with the
+underlying data-generating process
 \cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
@@
 provide algorithmic recourse (AR) to individuals (negatively) affected
-by opaque models. For AR recommendations to be actionable, they need to
-not only result in plausible counterfactuals but also be attainable. A
-plausible CE for a rejected 20-year-old loan applicant, for example,
-might reveal that their application would have been accepted, if only
-they were 20 years older. Ignoring all other features, this would comply
-with the definition of plausibility if 40-year-old individuals were in
-fact more credit-worthy on average than young adults. But of course this
-CE does not qualify for providing actionable recourse to the applicant
-since \emph{age} is not a (directly) mutable feature. CT aims to improve
-model explainability by aligning models with counterfactuals that meet
-both desiderata: plausibility and actionability. Formally, we define
-explainability as follows:
+by opaque models. To be actionable, AR recommendations must also be
+attainable. A plausible CE for a rejected 20-year-old loan applicant,
+for example, might reveal that their application would have been
+accepted, if only they were 20 years older. Ignoring all other features,
+this would comply with the definition of plausibility if 40-year-old
+individuals were in fact more credit-worthy on average than young
+adults. But of course this CE does not qualify for providing actionable
+recourse to the applicant since \emph{age} is not a (directly) mutable
+feature. Counterfactual training aims to improve model explainability by
+aligning models with counterfactuals that meet both desiderata:
+plausibility and actionability. Formally, we define explainability as
+follows:
 

### 2025-03-14 paper/sections/intro.qmd
Parse comments from the first document
No TODOs

@@
 Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted").
-Our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal.
+In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal.
 CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce a desired output.
@@
 \ref{def-explainability}).
-Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are also compatible with plausible explanations [@altmeyer2024faithful].
+Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are in fact compatible with plausible explanations [@altmeyer2024faithful].
 Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful].
@@
 
-**Specifically, we propose counterfactual training (CT)**: a novel training regime that aligns learned representations with plausible explanations respecting actionability constraints.
-The remainder of this paper is structured as follows.
-@sec-lit presents related work, focusing on the link between adversarial examples and counterfactual explanations.
+**Specifically, we introduce Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models.
+In high-level terms, we define this concept as as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and the global actionability constraints.
+To the best of our knowledge, our framework represents the first attempt to address this challenge by employing counterfactual explanations already in the training phase.
+
+ The remainder of this manuscript is structured as follows.
+@sec-lit presents related work, focusing on the link between AEs and CEs.
 Then follow our two principal contributions.

### 2025-03-14 paper/sections/method.qmd
Parse comments from the first document
No TODOs

@@
 
-Counterfactual training combines ideas from adversarial training, energy-based modelling and counterfactuals explanations with the explicit goal of aligning representations with plausible explanations that comply with user requirements.
-In the context of CEs, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful].
+In this section we propose our novel counterfactual training objective.
+In CT, we combine ideas from adversarial training, counterfactual explanations, and energy-based modelling with the explicit goal of aligning representations with plausible explanations that comply with user-defined actionability constraints.
+
+In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful].
 Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models.
-For AR recommendations to be actionable, they need to not only result in plausible counterfactuals but also be attainable.
+To be actionable, AR recommendations must also be attainable.
 A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older.
@@
 But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature.
-CT aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability.
+Counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability.
 Formally, we define explainability as follows:

### 2025-03-14 paper/sections/related.qmd
Parse comments from the first document
No TODOs

@@
 
-To the best of our knowledge, the proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability.
-In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and (global) actionability constraints.
 To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of CEs that can be generated for a given model.
@@
 As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., AEs).
-Thus, by inducing models to "unlearn" susceptiblity to such examples, AT can effectively remove implausible explanations from the solution space.
+Thus, by inducing models to "unlearn" susceptiblity to such examples, adversarial training can effectively remove implausible explanations from the solution space.
 
@@
 Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing].
-Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions and derive theoretical upper bounds on distances between them.
+Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions.
 
@@
 
-## Beyond Robustness
+## Aligning Representations with Plausible Explanations
 
@@
 The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
-Formally, they rely on tools from energy-based modelling to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
+Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
 Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
@@
 Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data.
-Previous work by @schut2021generating similarly found that generating plausible ("interpretable") counterfactual explanations is almost trivial for deep ensembles that have also undergone adversarial training.
+Previous work by @schut2021generating similarly found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have also undergone adversarial training.
 The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your].
-This is closely related to the idea of aligning models with plausible explanations and has inspired our proposed CT objective, as we explain in @sec-method.
+This is closely related to the idea of aligning models with plausible explanations and has inspired our CT objective.

### 2025-03-14 paper/quarto_ecml/paper.tex
Parse key comments from second/third document
TODO Patrick: remaining comments, namely (1) better labels for Fig. 1? (2) reorganizing Table 1? (3) potentially some extra explanation in results?

@@
 the other is backed by empirical evidence.
-\cite{sauer2021counterfactual} demonstrate that using counterfactual
+\cite{sauer2021counterfactual} demonstrates that using counterfactual
 images during classifier training improves model robustness. Similarly,
@@
 Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
-\cite{grathwohl2020your} minimize the contrastive divergence between
+\cite{grathwohl2020your} minimizes the contrastive divergence between
 these observed samples from \(p(\mathbf{x})\) and generated samples from
@@
 posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
-\cite{grathwohl2020your} rely on Stochastic Gradient Langevin Dynamics
+\cite{grathwohl2020your} relies on Stochastic Gradient Langevin Dynamics
 (SGLD) using an uninformative prior for initialization but we depart
@@
 }\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
-function. We set
+function defined as
 \(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
@@
 
-The departure from SGLD allows us to tap into the vast repertoire of
-explainers that have been proposed in the literature to meet different
-desiderata. For example, many methods facilitate the imposition of
-domain and mutability constraints. In principle, any existing approach
-for generating counterfactual explanations is viable, so long as it does
-not violate the faithfulness condition. Like JEMs
-\cite{murphy2022probabilistic}, CT can be considered a form of
-contrastive representation learning.
+The departure from SGLD of \cite{grathwohl2020your} allows us to tap
+into the vast repertoire of explainers that have been proposed in the
+literature to meet different desiderata. For example, many methods
+facilitate the imposition of domain and mutability constraints. In
+principle, any existing approach for generating counterfactual
+explanations is viable, so long as it does not violate the faithfulness
+condition. Like JEMs \cite{murphy2022probabilistic}, CT can be
+considered a form of contrastive representation learning.
 
@@
 \(\mathbf{x}^\prime\) can be repurposed as additional training samples
-\cite{luu2023counterfactual,balashankar2023improving} or AEs
-\cite{freiesleben2022intriguing,pawelczyk2022exploring}. This leaves
-some flexibility with respect to the choice for
-\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
-functional form, but likely not the only sensible choice, is inspired by
-adversarial training:
+\cite{luu2023counterfactual,balashankar2023improving} or adversarial
+examples \cite{freiesleben2022intriguing,pawelczyk2022exploring}. This
+leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
+Equation~\ref{eq-obj}. An intuitive functional form, but likely not the
+only sensible choice, is inspired by adversarial training:
 
@@
 than \cite{freiesleben2022intriguing} who consider misclassification as
-the key distinguishing feature of AE. One of the key observations in
+the key distinguishing feature of AE. One of the key observations of
 this work is that we can leverage CEs during training and get
-adversarial examples essentially for free.
+adversarial examples essentially for free, which can be used to reap the
+aforementioned benefits of adversarial training.
 
@@
 
-We have simulated this scenario using synthetic data with two
-independent features and Gaussian class-conditional densities in
+We have simulated this scenario using synthetic data with independent
+\emph{age} and \emph{debt}, and Gaussian class-conditional densities in
 Figure~\ref{fig-poc}. The four panels show the outcomes for different
@@
 
-Here we present the experiments conducted to answer our research
+In our experiments we seek to answer the following three research
 questions:
@@
 \item
-  To what extent does our counterfactual training objective as defined
-  in Equation~\ref{eq-obj} induce models to learn plausible
+  To what extent does the counterfactual training objective (as defined
+  in Equation~\ref{eq-obj}) induce models to learn plausible
   explanations?
@@
 
-Our key outcome of interest is how well do models perform with respect
-to explainability (Def. \ref{def-explainability}). To this end, we focus
+Our key outcome of interest is how well models perform with respect to
+explainability (Def. \ref{def-explainability}). To this end, we focus
 primarily on the plausibility and cost of faithfully generated
@@
 time, this has the effect of shorter counterfactual paths and hence
-smaller average costs to individuals. To extend this to the non-linear
-case, we test the effect of imposing mutability constraints empirically
-for our synthetic data using the same evaluation scheme as above. The
-final row in Table~\ref{tbl-main} reports the average reduction in costs
-for CT compared to the baseline, when imposing that either the first or
-the second feature is immutable. In all cases, costs are reduced
-substantially indicating that classifiers trained with CT are indeed
-more sensitive to mutable features.
+smaller average costs (\(\ell_1\)-norm) to individuals. To extend this
+to the non-linear case, we test the effect of imposing mutability
+constraints empirically for our synthetic data using the same evaluation
+scheme as above. The final row in Table~\ref{tbl-main} reports the
+average reduction in costs for CT compared to the baseline, when
+imposing that either the first or the second feature is immutable. In
+all cases, costs are reduced substantially, indicating that classifiers
+trained with CT are indeed more sensitive to mutable features.
 

### 2025-03-14 paper/sections/experiments.qmd
Parse key comments from second/third document
TODO Patrick: remaining comments, namely (1) better labels for Fig. 1? (2) reorganizing Table 1? (3) potentially some extra explanation in results?

@@
 
-Here we present the experiments conducted to answer our research questions:
+In our experiments we seek to answer the following three research questions:
 
-1. To what extent does our counterfactual training objective as defined in @eq-obj induce models to learn plausible explanations?
+1. To what extent does the counterfactual training objective (as defined in @eq-obj) induce models to learn plausible explanations?
 2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
@@
 
-Our key outcome of interest is how well do models perform with respect to explainability (Def.
+Our key outcome of interest is how well models perform with respect to explainability (Def.
 \ref{def-explainability}).
@@
 immutable features for linear models, tilting the decision boundary in favour of mutable features instead.
-For binding constraints at test time, this has the effect of shorter counterfactual paths and hence smaller average costs to individuals.
+For binding constraints at test time, this has the effect of shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals.
 To extend this to the non-linear case, we test the effect of imposing mutability constraints empirically for our synthetic data using the same evaluation scheme as above.
 The final row in @tbl-main reports the average reduction in costs for CT compared to the baseline, when imposing that either the first or the second feature is immutable.
-In all cases, costs are reduced substantially indicating that classifiers trained with CT are indeed more sensitive to mutable features.
+In all cases, costs are reduced substantially, indicating that classifiers trained with CT are indeed more sensitive to mutable features.
 

### 2025-03-14 paper/sections/method.qmd
Parse key comments from second/third document
TODO Patrick: remaining comments, namely (1) better labels for Fig. 1? (2) reorganizing Table 1? (3) potentially some extra explanation in results?

@@
 The first term can be optimized using conventional cross-entropy as in @eq-obj.
-Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimize the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.
+Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimizes the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.
 
@@
 This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$.
-To generate samples, @grathwohl2020your rely on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology.
+To generate samples, @grathwohl2020your relies on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology.
 Instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples.
@@
 $$ {#eq-div}
-where $\mathcal{E}_\theta(\cdot)$ denotes the energy function.
-We set $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$.
+where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$.
 Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations.
@@
 
-The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
+The departure from SGLD of [@grathwohl2020your] allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
 For example, many methods facilitate the imposition of domain and mutability constraints.
@@
 
-Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or AEs [@freiesleben2022intriguing;@pawelczyk2022exploring].
-This leaves some flexibility with respect to the choice for $\text{advloss}(\cdot)$ in @eq-obj.
+Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring].
+This leaves some flexibility wrt.
+the choice for $\text{advloss}(\cdot)$ in @eq-obj.
 An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:
@@
 Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE.
-One of the key observations in this work is that we can leverage CEs during training and get adversarial examples essentially for free.
+One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.
 
@@
 
-We have simulated this scenario using synthetic data with two independent features and Gaussian class-conditional densities in @fig-poc.
+We have simulated this scenario using synthetic data with independent *age* and *debt*, and Gaussian class-conditional densities in @fig-poc.
 The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier).

### 2025-03-14 paper/quarto_ecml/paper.tex
Back to (exactly) 16 pages!
TODO: Complete read-through

@@
 In the natural language processing (NLP) domain, CEs have also been used
-to improve models through data augmentation. \cite{wu2021polyjuice}
+to improve models through data augmentation. \cite{wu2021polyjuice2}
 propose \emph{Polyjuice}, a general-purpose counterfactual generator for
@@
 
-The characterization of faithfulness and plausibility in Def.
+\noindent The characterization of faithfulness and plausibility in Def.
 \ref{def-explainability} is the same as in \cite{altmeyer2024faithful},
@@
 The first term can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
+Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
 \cite{grathwohl2020your} minimizes the contrastive divergence between
-these observed samples from \(p(\mathbf{x})\) and generated samples from
+the observed samples from \(p(\mathbf{x})\) and generated samples from
 \(p_\theta(\mathbf{x})\).
@@
 
-From a Bayesian perspective, setting
+\noindent From a Bayesian perspective, setting
 \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
@@
 
-It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}}
-assumes independence of features. This raises a valid concern about the
-effect of protecting immutable features in the presence of proxies that
-remain unprotected. We address this in Section~\ref{sec-discussion}.
+\noindent It is worth highlighting that
+\mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
+This raises a valid concern about the effect of protecting immutable
+features in the presence of proxies that remain unprotected. We address
+this in Section~\ref{sec-discussion}.
 
@@
 
-These broad and substantial improvements in plausibility generally do
-not come at the cost of decreased predictive performance: test accuracy
-for CT is virtually identical to the baseline for \emph{Adult},
+We find that CT can substantially improve the quality of explanations
+learned by models without sacrificing predictive performance. Test
+accuracy for CT is virtually identical to the baseline for \emph{Adult},
 \emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
@@
 \emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
-in test accuracy of 2, 5 and 15 percentage points, respectively. We note
-in this context, that we have not optimized our models for predictive
-performance at all and worked with very small networks. In summary, we
-find that CT can substantially improve the quality of explanations
-learned by models without sacrificing predictive accuracy.
+in test accuracy of 2, 5 and 15 percentage points, respectively. In this
+context, we also note that we have not optimized the models (very small
+networks) for predictive performance at all.
 
@@
 
-\caption{\label{tbl-main}Key explainability and predictive performance
-metrics for all datasets. \textbf{Plausibility}: the top five rows show
-the percentage reduction in implausibility according to
-Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
-used for \emph{ECCo} at test time. The following row shows the reduction
-in implausibility as measured by Equation~\ref{eq-impl-div} and
-aggregated across all test specifications. \textbf{Accuracy}: The
-following two rows show the test accuracies for the models trained with
-CT and the baseline. \textbf{Actionability}: The final row present the
-average reduction in costs when imposing mutability constraints.}
+\caption{\label{tbl-main}Key performance metrics across all datasets.
+\textbf{Plausibility}: Rows 1-5 show the percentage reduction in
+implausibility via Equation~\ref{eq-impl-dist} for varying degrees of
+the energy penalty used for \emph{ECCo} at test time; row 6 shows the
+reduction in implausibility via Equation~\ref{eq-impl-div} and
+aggregated across all test specifications. \textbf{Accuracy} (rows 7-8):
+test accuracies for models trained with CT and the baseline.
+\textbf{Actionability} (row 9): average reduction in costs when imposing
+mutability constraints.}
 
@@
 favour of mutable features instead. For binding constraints at test
-time, this has the effect of shorter counterfactual paths and hence
-smaller average costs (\(\ell_1\)-norm) to individuals. To extend this
-to the non-linear case, we test the effect of imposing mutability
-constraints empirically for our synthetic data using the same evaluation
-scheme as above. The final row in Table~\ref{tbl-main} reports the
-average reduction in costs for CT compared to the baseline, when
+time, this leads to shorter counterfactual paths and hence smaller
+average costs (\(\ell_1\)-norm) to individuals. To extend this to the
+non-linear case, we test the effect of imposing mutability constraints
+empirically for our synthetic data using the same evaluation scheme as
+above. The final row in Table~\ref{tbl-main} reports the average
+reduction in costs for CT compared to the ``vanilla'' baseline, when
 imposing that either the first or the second feature is immutable. In
@@
 results are in the supplementary appendix.\\
-\indent \emph{Hyperparameters of the CE generators.} First, we observe
-that CT is highly sensitive to hyperparameter settings but (a) there are
-manageable patterns and (b) we can typically identify settings that
-improve either plausibility or cost, and commonly both of them at the
-same time. Second, we note that the choice of a CE generator has a major
-impact on the results. For example, \emph{REVISE} tends to perform the
-worst, most likely because it uses a surrogate VAE to generate
-counterfactuals which impedes faithfulness \cite{altmeyer2024faithful}.
-Third, increasing \(T\), the maximum number of steps, generally yields
-better outcomes because more CEs can mature in each training epoch.
-Fourth, the impact of \(\tau\), the required decision threshold is more
-difficult to predict. On ``harder'' datasets it may be difficult to
-satisfy high \(\tau\) for any given sample (i.e., also factuals) and so
-increasing this threshold does not seem to correlate with better
-outcomes. In fact, we have generally found that a choice of \(\tau=0.5\)
-leads to optimal results because it is associated with high proportions
-of mature counterfactuals.\\
-\indent \emph{Hyperparameters for penalties.} We find that the strength
-of the energy regularization, \(\lambda_{\text{reg}}\) is highly
-impactful; energy must be sufficiently regularized to avoid poor
-performance in terms of decreased plausibility and increased costs. The
-sensitivity with respect to \(\lambda_{\text{div}}\) and
-\(\lambda_{\text{adv}}\) is much less evident. While high values of
-\(\lambda_{\text{reg}}\) may increase the variability in outcomes
-(especially) when combined with high values of \(\lambda_{\text{div}}\)
-or \(\lambda_{\text{adv}}\), this effect is not very pronounced.\\
-\indent \emph{Other hyperparameters.} We observe that the effectiveness
-and stability of CT is positively associated with the number of
-counterfactuals generated during each training epoch. We also confirm
-that a higher number of training epochs is beneficial. Interestingly, we
-observed desired improvements in explainability when CT was combined
+\indent We note that CT is highly sensitive to the choice of a CE
+generator and its hyperparameters but (a) there are manageable patterns
+and (b) we can typically identify settings that improve either
+plausibility or cost, and commonly both of them at the same time. For
+example, \emph{REVISE} tends to perform the worst, most likely because
+it uses a surrogate VAE to generate counterfactuals which impedes
+faithfulness \cite{altmeyer2024faithful}. Increasing \(T\), the maximum
+number of steps, generally yields better outcomes because more CEs can
+mature in each training epoch. The impact of \(\tau\), the required
+decision threshold is more difficult to predict. On ``harder'' datasets
+it may be difficult to satisfy high \(\tau\) for any given sample (i.e.,
+also factuals) and so increasing this threshold does not seem to
+correlate with better outcomes. In fact, the choice of \(\tau=0.5\)
+generally leads to optimal results because it is associated with high
+proportions of mature counterfactuals.\\
+\indent The strength of the energy regularization,
+\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
+performance in terms of decreased plausibility and increased costs if
+insufficiently high. The sensitivity with respect to
+\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
+evident. While high values of \(\lambda_{\text{reg}}\) may increase the
+variability in outcomes when combined with high values of
+\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
+very pronounced.\\
+\indent The effectiveness and stability of CT is positively associated
+with the number of counterfactuals generated during each training epoch.
+We also confirm that a higher number of training epochs is beneficial.
+Interestingly, we observed desired improvements when CT was combined
 with conventional training and applied only for the final 50\% of epochs
-of the complete training process. Put differently, CT may be a way to
-improve the explainability of models in a fine-tuning manner.
+of the complete training process. Put differently, CT can improve the
+explainability of models in a fine-tuning manner.
 
@@
 Other settings have attracted some interest (e.g., regression in
-\cite{spooner2021counterfactual,zhao2023counterfactual}), but there is
-little consensus how to robustly extend the notion of CEs.\\
+\cite{spooner2021counterfactual}), but there is little consensus how to
+robustly extend the notion of CEs.\\
 \indent Secondly, our approach is susceptible to training instabilities.
@@
 approaches. Notably, CT is iterative which makes methods such as
-Bayesian (e.g., \cite{snoek2012practical}) or gradient-based (e.g.,
-\cite{franceschi2017forward}) optimization applicable.
+Bayesian or gradient-based optimization applicable (see
+\cite{bischl2023hyperparameter}).
 

### 2025-03-14 paper/sections/discussion.qmd
Back to (exactly) 16 pages!
TODO: Complete read-through

@@
 For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups.
-When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness].
+When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness].
 Still, these phenomena are not specific to CT.
@@
 Focus on classification models is a common restriction in research on CEs and AR.
-Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual; @zhao2023counterfactual]), but there is little consensus how to robustly extend the notion of CEs.
+Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual]), but there is little consensus how to robustly extend the notion of CEs.
 `\indent`{=latex} Secondly, our approach is susceptible to training instabilities.
@@
 Future work on CT could benefit from investigating more sophisticated approaches.
-Notably, CT is iterative which makes methods such as Bayesian (e.g., [@snoek2012practical]) or gradient-based (e.g., [@franceschi2017forward]) optimization applicable.
+Notably, CT is iterative which makes methods such as Bayesian or gradient-based optimization applicable (see [@bischl2023hyperparameter]).

### 2025-03-14 paper/sections/experiments.qmd
Back to (exactly) 16 pages!
TODO: Complete read-through

@@
 
-These broad and substantial improvements in plausibility generally do not come at the cost of decreased predictive performance: test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*.
+We find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive performance.
+Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*.
 Exceptions to this general pattern are *MNIST*, *CH* and *GMSC*, for which we observe reduction in test accuracy of 2, 5 and 15 percentage points, respectively.
-We note in this context, that we have not optimized our models for predictive performance at all and worked with very small networks.
-In summary, we find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive accuracy.
+In this context, we also note that we have not optimized the models (very small networks) for predictive performance at all.
 
@@
 
-Key explainability and predictive performance metrics for all datasets.
-**Plausibility**: the top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time.
-The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications.
-**Accuracy**: The following two rows show the test accuracies for the models trained with CT and the baseline.
-**Actionability**: The final row present the average reduction in costs when imposing mutability constraints.
-
+Key performance metrics across all datasets.
+**Plausibility**: Rows 1-5 show the percentage reduction in implausibility via @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time; row 6 shows the reduction in implausibility via @eq-impl-div and aggregated across all test specifications.
+**Accuracy** (rows 7-8): test accuracies for models trained with CT and the baseline.
+**Actionability** (row 9): average reduction in costs when imposing mutability constraints.
 :::
@@
 immutable features for linear models, tilting the decision boundary in favour of mutable features instead.
-For binding constraints at test time, this has the effect of shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals.
+For binding constraints at test time, this leads to shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals.
 To extend this to the non-linear case, we test the effect of imposing mutability constraints empirically for our synthetic data using the same evaluation scheme as above.
-The final row in @tbl-main reports the average reduction in costs for CT compared to the baseline, when imposing that either the first or the second feature is immutable.
+The final row in @tbl-main reports the average reduction in costs for CT compared to the "vanilla" baseline, when imposing that either the first or the second feature is immutable.
 In all cases, costs are reduced substantially, indicating that classifiers trained with CT are indeed more sensitive to mutable features.
@@
 We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
-`\indent`{=latex} *Hyperparameters of the CE generators.* First, we observe that CT is highly sensitive to hyperparameter settings but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time.
-Second, we note that the choice of a CE generator has a major impact on the results.
+`\indent`{=latex} We note that CT is highly sensitive to the choice of a CE generator and its hyperparameters but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time.
 For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful].
-Third, increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch.
-Fourth, the impact of $\tau$, the required decision threshold is more difficult to predict.
+Increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch.
+The impact of $\tau$, the required decision threshold is more difficult to predict.
 On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes.
-In fact, we have generally found that a choice of $\tau=0.5$ leads to optimal results because it is associated with high proportions of mature counterfactuals.
-`\indent`{=latex} *Hyperparameters for penalties.* We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs.
+In fact, the choice of $\tau=0.5$ generally leads to optimal results because it is associated with high proportions of mature counterfactuals.
+`\indent`{=latex} The strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high.
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
-While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes (especially) when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
-`\indent`{=latex} *Other hyperparameters.* We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
+While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
+`\indent`{=latex} The effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
 We also confirm that a higher number of training epochs is beneficial.
-Interestingly, we observed desired improvements in explainability when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process.
-Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.
+Interestingly, we observed desired improvements when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process.
+Put differently, CT can improve the explainability of models in a fine-tuning manner.

### 2025-03-14 paper/sections/method.qmd
Back to (exactly) 16 pages!
TODO: Complete read-through

@@
 
-The characterization of faithfulness and plausibility in Def.
+`\noindent`{=latex} The characterization of faithfulness and plausibility in Def.
 \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation.
@@
 The first term can be optimized using conventional cross-entropy as in @eq-obj.
-Then, to optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimizes the contrastive divergence between these observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.
+To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimizes the contrastive divergence between the observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.
 
@@
 
-From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$.
+`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$.
 Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features.
@@
 
-It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
+`\noindent`{=latex} It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
 This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected.

### 2025-03-14 paper/sections/related.qmd
Parse key comments from second/third document
TODO Patrick: remaining comments, namely (1) better labels for Fig. 1? (2) reorganizing Table 1? (3) potentially some extra explanation in results?
Back to (exactly) 16 pages!
TODO: Complete read-through

@@
 This interpretation of the link between explainability through counterfactuals on one side and robustness to adversarial examples on the other is backed by empirical evidence.
-@sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness.
+@sauer2021counterfactual demonstrates that using counterfactual images during classifier training improves model robustness.
 Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
@@
 In the natural language processing (NLP) domain, CEs have also been used to improve models through data augmentation.
-@wu2021polyjuice propose *Polyjuice*, a general-purpose counterfactual generator for language models.
+@wu2021polyjuice2 propose *Polyjuice*, a general-purpose counterfactual generator for language models.
 The authors empirically demonstrate that the augmentation of training data with their method improves robustness in a number of NLP tasks.

### 2025-03-14 paper/quarto_ecml/paper.tex
Slightly modify abstract for consistency with document, nicely reorder references, fix some typos
TODO: Should we remove eds. from references 4, 29 to be internally consistent for InProceedings entries?

@@
 to produce some desired output. To be useful in real-word
-decision-making systems, counterfactuals should be plausible with
-respect to the underlying data and actionable with respect to the
-stakeholder requirements. Much existing research has therefore focused
-on developing post-hoc methods to generate counterfactuals that meet
-these desiderata. In this work, we instead hold models directly
-accountable for the desired end goal: counterfactual training employs
-counterfactuals ad-hoc during the training phase to minimize the
-divergence between learned representations and plausible, actionable
-explanations. We demonstrate empirically and theoretically that our
-proposed method facilitates training models that deliver inherently
-desirable explanations while maintaining high predictive performance.
+decision-making systems, counterfactuals should be (1) plausible with
+respect to the underlying data and (2) actionable with respect to the
+user-defined mutability constraints. Much existing research has
+therefore focused on developing post-hoc methods to generate
+counterfactuals that meet these desiderata. In this work, we instead
+hold models directly accountable for the desired end goal:
+counterfactual training employs counterfactuals ad-hoc during the
+training phase to minimize the divergence between learned
+representations and plausible, actionable explanations. We demonstrate
+empirically and theoretically that our proposed method facilitates
+training models that deliver inherently desirable explanations while
+maintaining high predictive performance.
 
@@
 automated decision-making (ADM) systems involving opaque models
-\cite{oneil2016weapons,mcgregor2021preventing}. Effective XAI tools can
+\cite{mcgregor2021preventing,oneil2016weapons}. Effective XAI tools can
 aid us in monitoring models and providing recourse to individuals to
@@
 can arise as a consequence of regularization techniques aimed at other
-training objectives such as robustness, generalization, and generative
-capacity
-\cite{schut2021generating,augustin2020adversarial,altmeyer2024faithful}.
+training objectives such as generative capacity, generalization, or
+robustness
+\cite{altmeyer2024faithful,augustin2020adversarial,schut2021generating}.
 As further discussed in Section~\ref{sec-lit}, our work consolidates
@@
 underlying data-generating process
-\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
+\cite{altmeyer2024faithful,guidotti2022counterfactual,poyiadzi2020face}.
 Plausibility is a necessary but insufficient condition for using CEs to
@@
 The first term can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
+Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\),
 \cite{grathwohl2020your} minimizes the contrastive divergence between
@@
 \(\mathbf{x}^\prime\) can be repurposed as additional training samples
-\cite{luu2023counterfactual,balashankar2023improving} or adversarial
+\cite{balashankar2023improving,luu2023counterfactual} or adversarial
 examples \cite{freiesleben2022intriguing,pawelczyk2022exploring}. This
@@
 assigned burden of recourse \cite{sharma2020certifai}, threatening the
-equality of opportunity in the system \cite{bell2024fairness}. Still,
-these phenomena are not specific to CT.
+equality of opportunity in the system \cite{bell2024fairness}.
+Nonetheless, these phenomena are not specific to CT.
 
@@
 approaches. Notably, CT is iterative which makes methods such as
-Bayesian or gradient-based optimization applicable (see
+Bayesian or gradient-based optimization applicable (see, e.g.,
 \cite{bischl2023hyperparameter}).

### 2025-03-14 paper/sections/discussion.qmd
Slightly modify abstract for consistency with document, nicely reorder references, fix some typos
TODO: Should we remove eds. from references 4, 29 to be internally consistent for InProceedings entries?

@@
 When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness].
-Still, these phenomena are not specific to CT.
+Nonetheless, these phenomena are not specific to CT.
 
@@
 Future work on CT could benefit from investigating more sophisticated approaches.
-Notably, CT is iterative which makes methods such as Bayesian or gradient-based optimization applicable (see [@bischl2023hyperparameter]).
+Notably, CT is iterative which makes methods such as Bayesian or gradient-based optimization applicable (see, e.g., [@bischl2023hyperparameter]).

### 2025-03-14 paper/sections/intro.qmd
Slightly modify abstract for consistency with document, nicely reorder references, fix some typos
TODO: Should we remove eds. from references 4, 29 to be internally consistent for InProceedings entries?

@@
 Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions.
-On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@oneil2016weapons; @mcgregor2021preventing].
+On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@mcgregor2021preventing; @oneil2016weapons].
 Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted").
@@
 Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are in fact compatible with plausible explanations [@altmeyer2024faithful].
-Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as robustness, generalization, and generative capacity [@schut2021generating; @augustin2020adversarial; @altmeyer2024faithful].
+Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating].
 As further discussed in @sec-lit, our work consolidates these findings within a single objective.

### 2025-03-14 paper/sections/method.qmd
Slightly modify abstract for consistency with document, nicely reorder references, fix some typos
TODO: Should we remove eds. from references 4, 29 to be internally consistent for InProceedings entries?

@@
 
-In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@poyiadzi2020face;@guidotti2022counterfactual;@altmeyer2024faithful].
+In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@altmeyer2024faithful;@guidotti2022counterfactual;@poyiadzi2020face].
 Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models.
@@
 The first term can be optimized using conventional cross-entropy as in @eq-obj.
-To optimize $\log p_\theta(\mathbf{x})$ @grathwohl2020your minimizes the contrastive divergence between the observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.
+To optimize $\log p_\theta(\mathbf{x})$, @grathwohl2020your minimizes the contrastive divergence between the observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.
 
@@
 
-Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@luu2023counterfactual;@balashankar2023improving] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring].
+Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@balashankar2023improving;@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring].
 This leaves some flexibility wrt.

### 2025-03-15 paper/quarto_ecml/paper.tex
Fix referencing inconsistencies and some final typos
Fin?

@@
 to improve models through data augmentation. \cite{wu2021polyjuice2}
-propose \emph{Polyjuice}, a general-purpose counterfactual generator for
-language models. The authors empirically demonstrate that the
+proposes \emph{Polyjuice}, a general-purpose counterfactual generator
+for language models. The authors empirically demonstrate that the
 augmentation of training data with their method improves robustness in a
-number of NLP tasks. \cite{balashankar2023improving} similarly use
+number of NLP tasks. \cite{balashankar2023improving} similarly uses
 \emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
 and show that classifier robustness improves by up to 20\%. Finally,
-\cite{luu2023counterfactual} introduce Counterfactual Adversarial
-Training (CAT), which also aims at improving generalization and
-robustness of language models through a three-step procedure: the
-authors identify training samples that are subject to high predictive
-uncertainty, generate CEs for those samples, and then fine-tune the
-language model on a dataset augmented with the CEs.
+\cite{luu2023counterfactual} introduces Counterfactual Adversarial
+Training (CAT) that also aims to improve generalization and robustness
+of language models through a three-step procedure: the authors identify
+training samples that are subject to high predictive uncertainty,
+generate CEs for them, and fine-tune the language model on a dataset
+augmented with the CEs.
 
@@
 achieved \cite{freiesleben2022intriguing}. Similarly,
-\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
-certain conditions.
+\cite{pawelczyk2022exploring} shows that CEs and AEs are equivalent
+under certain conditions.
 
@@
 certain properties of the post-hoc counterfactual explanations. Firstly,
-\cite{ross2021learning} propose a way to train models that guarantee
+\cite{ross2021learning} proposes a way to train models that guarantee
 individual recourse to some positive target class with high probability.
@@
 actionability constraints ex-ante. For example, users can specify that
-certain features are immutable. Secondly, \cite{guo2023counternet} are
+certain features are immutable. Secondly, \cite{guo2023counternet} is
 the first to propose an end-to-end training pipeline that includes CEs
@@
 towards aligning representations with plausible explanations. In a work
-closely related to this one, \cite{altmeyer2024faithful} show that
+closely related to this one, \cite{altmeyer2024faithful} shows that
 explainability can be improved through model averaging and refined model
@@
 on a single locally optimal representations that corresponds to
-semantically meaningless explanations for the data. Previous work by
+semantically meaningless explanations for the data. Previous work of
 \cite{schut2021generating} similarly found that generating plausible
@@
 perturbation to any single feature is at most \(\varepsilon\). This is
-closely aligned with \cite{szegedy2013intriguing} who define an
+closely aligned with \cite{szegedy2013intriguing} that defines an
 adversarial attack as an ``imperceptible non-random perturbation''.
 Thus, we choose to work with a different distinction between CE and AE
-than \cite{freiesleben2022intriguing} who consider misclassification as
-the key distinguishing feature of AE. One of the key observations of
+than \cite{freiesleben2022intriguing} that considers misclassification
+as the key distinguishing feature of AE. One of the key observations of
 this work is that we can leverage CEs during training and get
@@
 
-In all cases the counterfactuals are valid but their quality differs. In
-panel (a), they are not plausible: they do not comply with the
-distribution of the factuals in \(y^+\) to the point where they form a
-clearly distinguishable cluster. In panel (b), they are highly
+In all cases the counterfactuals are all valid but their quality
+differs. In panel (a), they are not plausible: they do not comply with
+the distribution of the factuals in \(y^+\) to the point where they form
+a clearly distinguishable cluster. In panel (b), they are highly
 plausible, meeting the first objective of Def. \ref{def-explainability}.
@@
 \item
-  To what extent does the counterfactual training objective (as defined
-  in Equation~\ref{eq-obj}) induce models to learn plausible
+  To what extent does the counterfactual training objective as it is
+  defined in Equation~\ref{eq-obj} induce models to learn plausible
   explanations?
@@
 domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
-distance-based metric used by \cite{altmeyer2024faithful},
+distance-based metric used in \cite{altmeyer2024faithful},
 
@@
 be smaller but still substantial, with around 10-15\% for \emph{CH},
-11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred} and around 3\% for
+11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred}, and around 3\% for
 \emph{Adult}. For our only vision dataset (\emph{MNIST}), distances are
@@
 accuracy for CT is virtually identical to the baseline for \emph{Adult},
-\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
+\emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and even slightly
 improved for \emph{Cred}. Exceptions to this general pattern are
-\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
-in test accuracy of 2, 5 and 15 percentage points, respectively. In this
+\emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we observe reduction
+in test accuracy of 2, 5, and 15 percentage points respectively. In this
 context, we also note that we have not optimized the models (very small
@@
 protections. Delineating actionability is a major undecided challenge in
-the AR literature \cite{venkatasubramanian2020philosophical} impacting
-the capacity of CT to fulfill its intended goal.\\
+the AR literature (see, e.g.,
+\cite{venkatasubramanian2020philosophical}) impacting the capacity of CT
+to fulfill its intended goal.\\
 \indent \emph{Interventions on features may impact fairness.} We provide

### 2025-03-15 paper/sections/experiments.qmd
Fix referencing inconsistencies and some final typos
Fin?

@@
 
-1. To what extent does the counterfactual training objective (as defined in @eq-obj) induce models to learn plausible explanations?
+1. To what extent does the counterfactual training objective as it is defined in @eq-obj induce models to learn plausible explanations?
 2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
@@
 For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$.
-We rely on the distance-based metric used by @altmeyer2024faithful,
+We rely on the distance-based metric used in @altmeyer2024faithful,
 
@@
 For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*).
-For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*.
+For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred*, and around 3% for *Adult*.
 For our only vision dataset (*MNIST*), distances are reduced by up to 9%.
@@
 We find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive performance.
-Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*.
-Exceptions to this general pattern are *MNIST*, *CH* and *GMSC*, for which we observe reduction in test accuracy of 2, 5 and 15 percentage points, respectively.
+Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon*, and *OL*, and even slightly improved for *Cred*.
+Exceptions to this general pattern are *MNIST*, *CH*, and *GMSC*, for which we observe reduction in test accuracy of 2, 5, and 15 percentage points respectively.
 In this context, we also note that we have not optimized the models (very small networks) for predictive performance at all.

### 2025-03-15 paper/sections/related.qmd
Fix referencing inconsistencies and some final typos
Fin?

@@
 In the natural language processing (NLP) domain, CEs have also been used to improve models through data augmentation.
-@wu2021polyjuice2 propose *Polyjuice*, a general-purpose counterfactual generator for language models.
+@wu2021polyjuice2 proposes *Polyjuice*, a general-purpose counterfactual generator for language models.
 The authors empirically demonstrate that the augmentation of training data with their method improves robustness in a number of NLP tasks.
-@balashankar2023improving similarly use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%.
-Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models through a three-step procedure: the authors identify training samples that are subject to high predictive uncertainty, generate CEs for those samples, and then fine-tune the language model on a dataset augmented with the CEs.
+@balashankar2023improving similarly uses *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%.
+Finally, @luu2023counterfactual introduces Counterfactual Adversarial Training (CAT) that also aims to improve generalization and robustness of language models through a three-step procedure: the authors identify training samples that are subject to high predictive uncertainty, generate CEs for them, and fine-tune the language model on a dataset augmented with the CEs.
 
@@
 Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing].
-Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions.
+Similarly, @pawelczyk2022exploring shows that CEs and AEs are equivalent under certain conditions.
 
 Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of the post-hoc counterfactual explanations.
-Firstly, @ross2021learning propose a way to train models that guarantee individual recourse to some positive target class with high probability.
+Firstly, @ross2021learning proposes a way to train models that guarantee individual recourse to some positive target class with high probability.
 Their approach builds on adversarial training by explicitly inducing susceptibility to targeted adversarial examples for the positive class.
@@
 For example, users can specify that certain features are immutable.
-Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes CEs as part of the training procedure.
+Secondly, @guo2023counternet is the first to propose an end-to-end training pipeline that includes CEs as part of the training procedure.
 In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable.
@@
 Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations.
-In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives.
+In a work closely related to this one, @altmeyer2024faithful shows that explainability can be improved through model averaging and refined model objectives.
 The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
@@
 Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data.
-Previous work by @schut2021generating similarly found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have also undergone adversarial training.
+Previous work of @schut2021generating similarly found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have also undergone adversarial training.
 The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your].

### 2025-03-15 paper/preprint/paper.tex
rerendered full file

@@
 
-In the natural language processing (NLP) domain, counterfactuals have
-similarly been used to improve models through data augmentation. Wu et
-al. (\citeproc{ref-wu2021polyjuice}{2021}) propose \emph{Polyjuice}, a
-general-purpose counterfactual generator for language models. They
-demonstrate empirically that the augmentation of training data through
-\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
-tasks. Balashankar et al.
-(\citeproc{ref-balashankar2023improving}{2023}) similarly use
-\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
-and show that classifier robustness improves by up to 20\%. Finally, Luu
-and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
-Counterfactual Adversarial Training (CAT), which also aims at improving
-generalization and robustness of language models through a three-step
-procedure. First, the authors identify training samples that are subject
-to high predictive uncertainty. Second, they generate counterfactual
-explanations for those samples. Finally, they fine-tune the given
-language model on the augmented dataset that includes the generated
-counterfactuals.
+In the natural language processing (NLP) domain, CEs have also been used
+to improve models through data augmentation. Wu et al.
+(\citeproc{ref-wu2021polyjuice}{2021}) propose \emph{Polyjuice}, a
+general-purpose counterfactual generator for language models. The
+authors empirically demonstrate that the augmentation of training data
+with their method improves robustness in a number of NLP tasks.
+Balashankar et al. (\citeproc{ref-balashankar2023improving}{2023})
+similarly use \emph{Polyjuice} to augment NLP datasets through diverse
+counterfactuals and show that classifier robustness improves by up to
+20\%. Finally, Luu and Inoue
+(\citeproc{ref-luu2023counterfactual}{2023}) introduce Counterfactual
+Adversarial Training (CAT), which also aims at improving generalization
+and robustness of language models through a three-step procedure: the
+authors identify training samples that are subject to high predictive
+uncertainty, generate CEs for those samples, and then fine-tune the
+language model on a dataset augmented with the CEs.
 
@@
 method allows for imposing a set of actionability constraints ex-ante.
-For example, users can specify that certain features (e.g., \emph{age},
-\emph{gender}) are immutable. Secondly, Guo, Nguyen, and Yadav
+For example, users can specify that certain features are immutable.
+Secondly, Guo, Nguyen, and Yadav
 (\citeproc{ref-guo2023counternet}{2023}) are the first to propose an
-end-to-end training pipeline that includes counterfactual explanations
-as part of the training procedure. In particular, they propose a
-specific network architecture that includes a predictor and CE generator
-network, where the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed
-back to the predictor network. In contrast to Guo, Nguyen, and Yadav
+end-to-end training pipeline that includes CEs as part of the training
+procedure. In particular, they propose a specific network architecture
+that includes a predictor and CE generator network, where the parameters
+of the CE generator network are learnable. Counterfactuals are generated
+during each training iteration and fed back to the predictor network. In
+contrast to Guo, Nguyen, and Yadav
 (\citeproc{ref-guo2023counternet}{2023}), we impose no restrictions on
-the neural network architecture at all.
+the ANN architecture at all.
 
@@
 
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
-any classifier can be re-interpreted as a joint energy-based model (JEM)
-that learns to discriminate output classes conditional on the observed
-(training) samples from \(p(\mathbf{x})\) and the generated samples from
+As observed by Grathwohl et al.
+(\citeproc{ref-grathwohl2020your}{2020}), any classifier can be
+re-interpreted as a joint energy-based model (JEM) that learns to
+discriminate output classes conditional on the observed (training)
+samples from \(p(\mathbf{x})\) and the generated samples from
 \(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
@@
 independent features and Gaussian class-conditional densities in
-Figure~\ref{fig-poc}. The four panels in Figure~\ref{fig-poc} show the
-outcomes for different training procedures using the same model
-architecture each time (a linear classifier). In each case, we show the
-decision boundary (in green) and the training data colored according to
-their ground-truth label: orange points belong to the target class,
-\(y^+=2\), blue points belong to the non-target class, \(y^-=1\). Stars
-indicate counterfactuals in the target class generated at test time
-using generic gradient descent until convergence.
-
-In panel (a), we have trained our model conventionally, and we do not
-impose mutability constraints at test time. The generated
-counterfactuals are all valid, but not plausible: they do not comply
-with the distribution of the factual samples in the target class to the
-point where they are clearly distinguishable from the ground-truth data.
-In panel (b), we have trained our model with CT, once again without any
-mutability constraints. We observe that the counterfactuals are highly
+Figure~\ref{fig-poc}. The four panels show the outcomes for different
+training procedures using the same model architectures (a linear
+classifier). In panels (a) and (c) we have trained the models
+conventionally, while in panels (b) and (d) we used CT. Only in panels
+(c) and (d) do we impose the mutability constraint on \emph{age} at test
+time. In each case, we show the decision boundary (in green) and the
+training data colored according to their ground-truth label: orange
+points belong to the target class, \(y^+=2\), blue points belong to the
+non-target class, \(y^-=1\). Stars indicate CEs in the target class
+generated at test time using generic gradient descent until convergence.
+
+In all cases the counterfactuals are valid but their quality differs. In
+panel (a), they are not plausible: they do not comply with the
+distribution of the factuals in \(y^+\) to the point where they form a
+clearly distinguishable cluster. In panel (b), they are highly
 plausible, meeting the first objective of Def. \ref{def-explainability}.
-
-In panel (c), we have used conventional training again, this time
-imposing the mutability constraint on \emph{age} at test time.
-Counterfactuals are valid but involve some substantial reductions in
-\emph{debt} for some individuals (very young applicants). By comparison,
-counterfactual paths are shorter on average in panel (d), where we have
-used CT and protected the immutable feature as described in
-Section~\ref{sec-constraints}. We observe that due to the classifier's
-lower sensitivity to \emph{age}, recourse recommendations with respect
-to \emph{debt} are much more homogenous and do not disproportionately
-punish younger individuals. The counterfactuals are also plausible with
-respect to the mutable feature and shorter on average than in (c). Thus,
-we consider the model in panel (d) as the most explainable according to
-Def. \ref{def-explainability}.
+In panel (c), the CEs involve substantial reductions in \emph{debt} for
+younger applicants. By comparison, counterfactual paths are shorter on
+average in panel (d) where we have protected the immutable \emph{age} as
+described in Section~\ref{sec-constraints}. Due to the classifier's
+lower sensitivity to \emph{age}, recommendations with respect to
+\emph{debt} are much more homogenous and do not unfairly punish younger
+individuals. These counterfactuals are also plausible with respect to
+the mutable feature. Thus, we consider the model in panel (d) as the
+most explainable according to Def. \ref{def-explainability}.
 
@@
 \item
-  To what extent does our proposed counterfactual training objective in
-  Equation~\ref{eq-obj} induce models to learn plausible explanations?
-\item
-  To what extent does our proposed counterfactual training objective in
-  Equation~\ref{eq-obj} yield more favorable algorithmic recourse
-  outcomes in the presence of actionability constraints?
-\item
-  What are the effects of hyperparameter selection with respect to
-  Equation~\ref{eq-obj}?
+  To what extent does our counterfactual training objective as defined
+  in Equation~\ref{eq-obj} induce models to learn plausible
+  explanations?
+\item
+  To what extent does the CT objective produce more favorable
+  algorithmic recourse outcomes in the presence of actionability
+  constraints?
+\item
+  What are the effects of hyperparameter selection wrt. the CT
+  objective?
 \end{enumerate}
@@
 Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
-used for \emph{ECCo} at test time. The following row shows the reduction
-in implausibility as measured by Equation~\ref{eq-impl-div} and
-aggregated across all test specifications of \emph{ECCo}. The following
-two rows show the test accuracies for the model trained with CT and
-conventionally trained models (``vanilla'').
-
-We observe that for all datasets except \emph{OL} and across all test
-settings, the average distance of counterfactuals from observed samples
-in the target class is reduced, indicating improved plausibility. The
-magnitude of improvements varies by dataset: for the simple synthetic
-datasets, distance reductions range from around 20-40\% (\emph{LS},
-\emph{Moon}) to almost 60\% (\emph{Circ}). For the real-world tabular
-datasets, improvements are generally smaller but still substantial in
-many cases with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
-7-8\% for \emph{Cred} and around 3\% for \emph{Adult}. For our only
-vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
-results for our proposed divergence metric are qualitatively similar,
-but generally even more pronounced: for the \emph{Circ} dataset,
-implausibility is reduced by almost 94\% to virtually zero as we
-verified by looking at the absolute outcome. Improvements for other
-datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
+used for \emph{ECCo} at test time. Row 6 shows the reduction in
+implausibility as measured by Equation~\ref{eq-impl-div} and aggregated
+across all test specifications of \emph{ECCo}. Rows 7 and 8 show the
+test accuracies for the model trained with CT and conventionally trained
+models (``vanilla'').
+
+For all datasets except \emph{OL} and across all test settings, the
+average distance of CEs from observed samples in the target class is
+reduced, indicating improved plausibility. The magnitude of improvements
+varies. For the simple synthetic datasets, distance reductions range
+from around 20-40\% (\emph{LS}, \emph{Moon}) to almost 60\%
+(\emph{Circ}). For the real-world tabular datasets, improvements tend to
+be smaller but still substantial, with around 10-15\% for \emph{CH},
+11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred} and around 3\% for
+\emph{Adult}. For our only vision dataset (\emph{MNIST}), distances are
+reduced by up to 9\%. The results for our proposed divergence metric are
+qualitatively similar, but generally even more pronounced: for the
+\emph{Circ} dataset, implausibility is reduced by almost 94\% to
+virtually zero as we verified by the absolute outcome. Improvements for
+other datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
 \emph{OL} the reduction is negative, consistent with the distance-based
@@
 \subsubsection{Impact of hyperparameter
-settings}\label{sec-hyperparameters}
-
-We test in-depth the impact of three types of hyperparameters; our
-complete results are in the appendix.
-
+settings.}\label{sec-hyperparameters}
+
+We test the impact of three types of hyperparameters; our complete
+results are in the supplementary appendix.\\
 \textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
@@
 results because it is associated with high proportions of mature
-counterfactuals.
-
+counterfactuals.\\
 \textbf{\emph{Hyperparameters for penalties.}} We find that the strength
@@
 
-We first address the direct extensions of CT in
-Section~\ref{sec-approach}. Then, we look at its limitations and
-challenges in Section~\ref{sec-limitations}.
-
-\subsection{Future Research}\label{sec-approach}
-
-\textbf{\emph{CT is defined only for classification settings.}} Our
-formulation relies on the distinction between non-target class(es)
-\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
-through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
-arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
-discrete. Thus, it does not apply to ML tasks where the change in
-outcome cannot be readily quantified. Focus on classification models is
-a common restriction in research on CEs and AR. Other settings have
-attracted some interest (e.g., regression in
-(\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021};
-\citeproc{ref-zhao2023counterfactual}{Zhao, Broelemann, and Kasneci
-2023})), but there is little consensus how to robustly extend the notion
-of counterfactuals.
-
-\textbf{\emph{CT is subject to training instabilities.}} JEMs are
-susceptible to instabilities during training
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
-though we depart from the SGLD-based sampling, we still encounter major
-variability in the outcomes. CT is exposed to two potential sources of
-instabilities: (1) the energy-based contrastive divergence term in
-Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
-Still, we find that training instabilities can be successfully mitigated
-by regularizing energy (\(\lambda_{\text{reg}}\)), generating
-sufficiently many counterfactuals during each training epoch, and
-including only mature counterfactuals for contrastive divergence.
-
-\textbf{\emph{CT is sensitive to hyperparameter selection.}} Our method
-benefits from the tuning of certain key hyperparameters (see
-Section~\ref{sec-hyperparameters}). In this work, we have relied
-exclusively on grid search for this task. Future work on CT could
-benefit from investigating more sophisticated approaches towards
-hyperparameter tuning. Notably, CT is iterative which makes a variety of
-methods applicable, including Bayesian (e.g.,
-\citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
-gradient-based (e.g., \citeproc{ref-franceschi2017forward}{Franceschi et
-al. 2017}) optimization.
-
-\subsection{Current Limitations}\label{sec-limitations}
-
+As our results indicate, counterfactual training produces models that
+are more explainable. However, our approach is not without
+limitations.\\
 \textbf{\emph{CT increases the training time of models.}} CT can be more
@@
 \citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
-and Alfano 2020}) impacting the capacity of CT to increase the
-explainability of the model.
-
+and Alfano 2020}) impacting the capacity of CT to fulfill its intended
+goal.\\
 \textbf{\emph{Interventions on features may impact fairness.}} We
@@
 phenomena are not specific to CT.
+
+We also highlight several interesting directions for future research.\\
+\textbf{\emph{Extending CT beyond classification settings.}} Our
+formulation relies on the distinction between non-target class(es)
+\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
+through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
+arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
+discrete. Thus, it does not apply to ML tasks where the change in
+outcome cannot be readily quantified. Focus on classification models is
+a common restriction in research on CEs and AR. Other settings have
+attracted some interest (e.g., regression in
+(\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021};
+\citeproc{ref-zhao2023counterfactual}{Zhao, Broelemann, and Kasneci
+2023})), but there is little consensus how to robustly extend the notion
+of CEs.\\
+\textbf{\emph{Addressing the training instabilities.}} JEMs are
+susceptible to instabilities during training
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
+though we depart from the SGLD-based sampling, we still encounter major
+variability in the outcomes. CT is exposed to two potential sources of
+instabilities: (1) the energy-based contrastive divergence term in
+Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
+Still, we find that training instabilities can be successfully mitigated
+by regularizing energy (\(\lambda_{\text{reg}}\)), generating
+sufficiently many counterfactuals during each training epoch, and
+including only mature counterfactuals for contrastive divergence.\\
+\textbf{\emph{Improving hyperparameter selection procedures.}} Our
+method benefits from the tuning of certain key hyperparameters (see
+Section~\ref{sec-hyperparameters}). In this work, we have relied
+exclusively on grid search for this task. Future work on CT could
+benefit from investigating more sophisticated approaches towards
+hyperparameter tuning. Notably, CT is iterative which makes a variety of
+methods applicable, including Bayesian (e.g.,
+\citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
+gradient-based (e.g., \citeproc{ref-franceschi2017forward}{Franceschi et
+al. 2017}) optimization.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-train-final-run-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Package Version (Reproducibility)}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-warning-color-frame, colbacktitle=quarto-callout-warning-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-warning-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, coltitle=black, title={Package Version (Reproducibility)}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-tune-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-tune-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-tune_lr-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, opacitybacktitle=0.6, toprule=.15mm, rightrule=.15mm, colback=white, arc=.35mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, coltitle=black, opacityback=0, titlerule=0mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, colbacktitle=quarto-callout-note-color!10!white, breakable]
+\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
 

### 2025-03-15 paper/sections/abstract.qmd

@@
 Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output.
-To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the stakeholder requirements.
+To be useful in real-word decision-making systems, counterfactuals should be (1) plausible with respect to the underlying data and (2) actionable with respect to the user-defined mutability constraints.
 Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata.

### 2025-03-15 paper/sections/discussion.qmd

@@
 Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections.
-Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.
+Delineating actionability is a major undecided challenge in the AR literature (see, e.g., [@venkatasubramanian2020philosophical]) impacting the capacity of CT to fulfill its intended goal.
 `\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects.

### 2025-03-15 paper/sections/method.qmd

@@
 Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$.
-This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation".
-Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE.
+This is closely aligned with @szegedy2013intriguing that defines an adversarial attack as an "imperceptible non-random perturbation".
+Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing that considers misclassification as the key distinguishing feature of AE.
 One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.
@@
 
-In all cases the counterfactuals are valid but their quality differs.
+In all cases the counterfactuals are all valid but their quality differs.
 In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster.

### 2025-03-15 paper/sections/other/main_results.qmd
table fixed and I need a break after this latex nightmare

@@
 inputs = get_table_inputs(df, nothing; backend=Val(:latex))
-tabulate_results(inputs; table_type=:tabular, save_name="paper/quarto_ecml/tables/main.tex", wrap_table=false)
-tabulate_results(inputs; table_type=:tabular, save_name="paper/preprint/tables/main.tex", wrap_table=false)
+formatters = (
+    PrettyTables.ft_round(1,findall(.!contains.(names(df),"Acc"))),
+    PrettyTables.ft_round(2,findall(contains.(names(df),"Acc")))
+)
+for fname in ["paper/quarto_ecml/tables/main.tex","paper/preprint/tables/main.tex"]
+    tabulate_results(
+        inputs; 
+        table_type=:tabular, 
+        save_name=fname, 
+        wrap_table=false, 
+        alignment=[:c,fill(:r,size(df,2)-1)...],
+        formatters=formatters
+    )
+end
 ```

### 2025-03-15 paper/experiments/CTExperiments/src/tables.jl

@@ -10,12 +10,13 @@ function tabulate_results(
     table_type=:longtable,
     longtable_footer="Continuing table below.",
     save_name::Union{String,Nothing}=nothing,
+    formatters=PrettyTables.ft_round(2),
+    alignment=:c,
     kwrgs...,
 )
     df = inputs[1]
     other_inputs = inputs[2]
     if isa(other_inputs.backend, Val{:latex})
-        formatters = PrettyTables.ft_latex_sn(3)
         tf = PrettyTables.tf_latex_booktabs
         if isnothing(save_name)
             tab = pretty_table(
@@ -25,7 +26,7 @@ function tabulate_results(
                 wrap_table=wrap_table,
                 table_type=table_type,
                 longtable_footer=longtable_footer,
-                alignment=:c,
+                alignment,
                 other_inputs...,
                 kwrgs...,
             )
@@ -40,7 +41,7 @@ function tabulate_results(
                     wrap_table=wrap_table,
                     table_type=table_type,
                     longtable_footer=longtable_footer,
-                    alignment=:c,
+                    alignment,
                     other_inputs...,
                     kwrgs...,
                 )
@@ -48,11 +49,11 @@ function tabulate_results(
         end
     else
         if isnothing(save_name)
-            tab = pretty_table(df; alignment=:c, other_inputs..., kwrgs...)
+            tab = pretty_table(df; alignment, other_inputs..., kwrgs...)
             return tab
         else
             open(save_name, "w") do io
-                pretty_table(io, df; alignment=:c, other_inputs..., kwrgs...)
+                pretty_table(io, df; alignment, other_inputs..., kwrgs...)
             end
         end
     end
@@ -80,7 +81,8 @@ function get_table_inputs(
         hls = ()
     end
 
-    header = format_header.(names(df); replacements=LatexHeaderReplacements)
+    hs = multi_row_header(names(df))
+    header = [format_header.(h; replacements=LatexHeaderReplacements) for h in hs] |> x -> tuple(x...)
     return df, (; highlighters=hls, backend=backend, header=header)
 end
 

### 2025-03-15 paper/quarto_ecml/paper.tex
table fixed and I need a break after this latex nightmare
vamooooos

@@
 classifier). In panels (a) and (c) we have trained the models
-conventionally, while in panels (b) and (d) we used CT. Only in panels
-(c) and (d) do we impose the mutability constraint on \emph{age} at test
-time. In each case, we show the decision boundary (in green) and the
-training data colored according to their ground-truth label: orange
-points belong to the target class, \(y^+=2\), blue points belong to the
-non-target class, \(y^-=1\). Stars indicate CEs in the target class
-generated at test time using generic gradient descent until convergence.
+conventionally, while in panels (b) and (d) we used CT.
 
-In all cases the counterfactuals are all valid but their quality
-differs. In panel (a), they are not plausible: they do not comply with
-the distribution of the factuals in \(y^+\) to the point where they form
-a clearly distinguishable cluster. In panel (b), they are highly
-plausible, meeting the first objective of Def. \ref{def-explainability}.
-In panel (c), the CEs involve substantial reductions in \emph{debt} for
-younger applicants. By comparison, counterfactual paths are shorter on
-average in panel (d) where we have protected the immutable \emph{age} as
+In all cases, all counterfactuals (stars) are valid---they have cross
+the decision boundary (green)---but their quality differs. In panel (a),
+they are not plausible: they do not comply with the distribution of the
+factuals in \(y^+\) to the point where they form a clearly
+distinguishable cluster. In panel (b), they are highly plausible,
+meeting the first objective of Def. \ref{def-explainability}. In panel
+(c), the CEs involve substantial reductions in \emph{debt} for younger
+applicants. By comparison, counterfactual paths are shorter on average
+in panel (d) where we have protected the immutable \emph{age} as
 described in Section~\ref{sec-constraints}. Due to the classifier's
@@
 \caption{\label{fig-poc}Illustration of how CT improves model
-explainability.}
+explainability: (a) conventional training, all mutable; (b) CT, all
+mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
+immutable. The decision boundary is shown in green along with training
+data colored according to their ground-truth label: \(y^+=2\) (orange)
+and \(y^-=1\) (blue). Stars indicate CEs for the target class.}
 
@@
 
-Our key outcome of interest is how well models perform with respect to
-explainability (Def. \ref{def-explainability}). To this end, we focus
-primarily on the plausibility and cost of faithfully generated
-counterfactuals at test time. To measure the cost of counterfactuals, we
-follow the standard convention of using distances (\(\ell_1\)-norm)
-between factuals and counterfactuals as a proxy. For plausibility, we
-assess how similar counterfactuals are to observed samples in the target
-domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
-distance-based metric used in \cite{altmeyer2024faithful},
+Our key outcome of interest is improved explainability (Def.
+\ref{def-explainability}). To this end, we focus primarily on the
+plausibility and cost of faithfully generated counterfactuals at test
+time. To measure the cost of counterfactuals, we follow the standard
+convention of using distances (\(\ell_1\)-norm) between factuals and
+counterfactuals as a proxy. For plausibility, we assess how similar
+counterfactuals are to observed samples in the target domain,
+\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the distance-based
+metric used in \cite{altmeyer2024faithful},
 
@@
 
-In addition to cost and plausibility, we also compute other standard
-metrics to evaluate counterfactuals at test time including validity and
-redundancy. Finally, we also assess the predictive performance of models
-using standard metrics.
+In addition to cost and plausibility, we compute other standard metrics
+to evaluate counterfactuals including validity and redundancy. Finally,
+we also assess the predictive performance of models using standard
+metrics, including robust accuracy estimated on adversarially perturbed
+data using FGSM \cite{goodfellow2014explaining}.
 
@@
 
+To assess our proposed training regime, we investigate the improvements
+in performance metrics when using CT on top of a weak baseline (BL):
+specifically, a multilayer perceptron (\emph{MLP}). This is the best way
+to get a clear picture of how effective CT is and consistent with how
+assessment is done in the related literature
+\cite{ross2021learning,teney2020learning,goodfellow2014explaining}.
+
 \subsection{Experimental Results}\label{experimental-results}
 
-\subsubsection{Plausibility.}\label{sec-plaus}
+\subsubsection{Plausibility}\label{sec-plaus}
 
-Table~\ref{tbl-main} presents our main empirical findings. The top five
-rows show the percentage reduction in implausibility according to
-Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
-used for \emph{ECCo} at test time. Row 6 shows the reduction in
-implausibility as measured by Equation~\ref{eq-impl-div} and aggregated
-across all test specifications of \emph{ECCo}. Rows 7 and 8 show the
-test accuracies for the model trained with CT and conventionally trained
-models (``vanilla'').
+Table~\ref{tbl-main} presents our main empirical findings. For all
+datasets except \emph{OL} and across all test settings, the average
+distance of CEs from observed samples in the target class is reduced,
+indicating improved plausibility. The magnitude of improvements varies.
+For the simple synthetic datasets, distance reductions range from around
+20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
+real-world tabular datasets, improvements tend to be smaller but still
+substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
+7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For our only
+vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
+results for our proposed divergence metric are qualitatively similar,
+but generally even more pronounced: for the \emph{Circ} dataset,
+implausibility is reduced by almost 94\% to virtually zero as we
+verified by the absolute outcome. Improvements for other datasets range
+from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For \emph{OL} the
+reduction is negative, consistent with the distance-based metric.
+\emph{MNIST} is the only dataset for which the distance and divergence
+metrics disagree. Upon visual inspection of the image counterfactuals we
+find that CT clearly improves plausibility (see appendix).
 
-For all datasets except \emph{OL} and across all test settings, the
-average distance of CEs from observed samples in the target class is
-reduced, indicating improved plausibility. The magnitude of improvements
-varies. For the simple synthetic datasets, distance reductions range
-from around 20-40\% (\emph{LS}, \emph{Moon}) to almost 60\%
-(\emph{Circ}). For the real-world tabular datasets, improvements tend to
-be smaller but still substantial, with around 10-15\% for \emph{CH},
-11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred}, and around 3\% for
-\emph{Adult}. For our only vision dataset (\emph{MNIST}), distances are
-reduced by up to 9\%. The results for our proposed divergence metric are
-qualitatively similar, but generally even more pronounced: for the
-\emph{Circ} dataset, implausibility is reduced by almost 94\% to
-virtually zero as we verified by the absolute outcome. Improvements for
-other datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
-\emph{OL} the reduction is negative, consistent with the distance-based
-metric. \emph{MNIST} is the only dataset for which the two metrics
-disagree.
+\subsubsection{Predictive Performance}\label{sec-pred}
 
-We find that CT can substantially improve the quality of explanations
-learned by models without sacrificing predictive performance. Test
-accuracy for CT is virtually identical to the baseline for \emph{Adult},
-\emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and even slightly
-improved for \emph{Cred}. Exceptions to this general pattern are
-\emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we observe reduction
-in test accuracy of 2, 5, and 15 percentage points respectively. In this
-context, we also note that we have not optimized the models (very small
-networks) for predictive performance at all.
+Test accuracy for CT is virtually identical to the baseline for
+\emph{Adult}, \emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and
+even slightly improved for \emph{Cred}. Exceptions to this general
+pattern are \emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we
+observe a reduction in test accuracy of 2, 5, and 15 percentage points
+respectively. When looking at robust test accuracies (Acc.\(^*\)) for
+these datasets in particular, we find that CT strongly outperforms the
+baseline. In fact, we find that CT improves adversarial robustness on
+all datasets.
 
@@
 \caption{\label{tbl-main}Key performance metrics across all datasets.
-\textbf{Plausibility}: Rows 1-5 show the percentage reduction in
-implausibility via Equation~\ref{eq-impl-dist} for varying degrees of
-the energy penalty used for \emph{ECCo} at test time; row 6 shows the
-reduction in implausibility via Equation~\ref{eq-impl-div} and
-aggregated across all test specifications. \textbf{Accuracy} (rows 7-8):
-test accuracies for models trained with CT and the baseline.
-\textbf{Actionability} (row 9): average reduction in costs when imposing
-mutability constraints.}
+\textbf{Plausibility}: Columns 2-6 show the percentage reduction in
+implausibility (\(\text{IP}\)) for varying degrees of the energy penalty
+used for \emph{ECCo} at test time; column 7 shows the reduction in
+\(\text{IP}^*\) (\(\text{MMD}\)), aggregated across all test
+specifications. \textbf{Accuracy} (columns 8-11): test accuracies and
+robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
+\textbf{Actionability} (column 9): average reduction in costs when
+imposing mutability constraints.}
 
@@
 
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
 \input{tables/main.tex}
-\end{tabular*}
 
@@
 
-\subsubsection{Actionability.}\label{sec-act}
+\subsubsection{Actionability}\label{sec-act}
 

### 2025-03-15 paper/sections/experiments.qmd
table fixed and I need a break after this latex nightmare
vamooooos

@@
 
-Our key outcome of interest is how well models perform with respect to explainability (Def.
+Our key outcome of interest is improved explainability (Def.
 \ref{def-explainability}).
@@
 
-In addition to cost and plausibility, we also compute other standard metrics to evaluate counterfactuals at test time including validity and redundancy.
-Finally, we also assess the predictive performance of models using standard metrics.
+In addition to cost and plausibility, we compute other standard metrics to evaluate counterfactuals including validity and redundancy.
+Finally, we also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].
 
@@
 
+To assess our proposed training regime, we investigate the improvements in performance metrics when using CT on top of a weak baseline (BL): specifically, a multilayer perceptron (*MLP*).
+This is the best way to get a clear picture of how effective CT is and consistent with how assessment is done in the related literature [@ross2021learning;@teney2020learning;@goodfellow2014explaining].
+
 ## Experimental Results
 
-### Plausibility. {#sec-plaus}
+### Plausibility {#sec-plaus}
 
 @tbl-main presents our main empirical findings.
-The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time.
-Row 6 shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*.
-Rows 7 and 8 show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").
-
 For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility.
@@
 For *OL* the reduction is negative, consistent with the distance-based metric.
-*MNIST* is the only dataset for which the two metrics disagree.
+*MNIST* is the only dataset for which the distance and divergence metrics disagree.
+Upon visual inspection of the image counterfactuals we find that CT clearly improves plausibility (see appendix).
 
-We find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive performance.
+### Predictive Performance {#sec-pred}
+
 Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon*, and *OL*, and even slightly improved for *Cred*.
-Exceptions to this general pattern are *MNIST*, *CH*, and *GMSC*, for which we observe reduction in test accuracy of 2, 5, and 15 percentage points respectively.
-In this context, we also note that we have not optimized the models (very small networks) for predictive performance at all.
+Exceptions to this general pattern are *MNIST*, *CH*, and *GMSC*, for which we observe a reduction in test accuracy of 2, 5, and 15 percentage points respectively.
+When looking at robust test accuracies (Acc.$^*$) for these datasets in particular, we find that CT strongly outperforms the baseline.
+In fact, we find that CT improves adversarial robustness on all datasets.
 
@@
 ```{=latex}
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
 \input{tables/main.tex}
-\end{tabular*}
 ```
@@
 Key performance metrics across all datasets.
-**Plausibility**: Rows 1-5 show the percentage reduction in implausibility via @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time; row 6 shows the reduction in implausibility via @eq-impl-div and aggregated across all test specifications.
-**Accuracy** (rows 7-8): test accuracies for models trained with CT and the baseline.
-**Actionability** (row 9): average reduction in costs when imposing mutability constraints.
+**Plausibility**: Columns 2-6 show the percentage reduction in implausibility ($\text{IP}$) for varying degrees of the energy penalty used for *ECCo* at test time; column 7 shows the reduction in $\text{IP}^*$ ($\text{MMD}$), aggregated across all test specifications.
+**Accuracy** (columns 8-11): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL).
+**Actionability** (column 9): average reduction in costs when imposing mutability constraints.
 :::
 
-### Actionability. {#sec-act}
+### Actionability {#sec-act}
 

### 2025-03-15 paper/sections/method.qmd
vamooooos

@@
 In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we used CT.
-Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time.
+
+<!-- Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time.
 In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$.
 Stars indicate CEs in the target class generated at test time using generic gradient descent until convergence.
+-->
 
-In all cases the counterfactuals are all valid but their quality differs.
+In all cases, all counterfactuals (stars) are valid---they have cross the decision boundary (green)---but their quality differs.
 In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster.
@@
 
-![Illustration of how CT improves model explainability.](/paper/figures/poc.svg){#fig-poc}
+![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable.
+The decision boundary is shown in green along with training data colored according to their ground-truth label: $y^+=2$ (orange) and $y^-=1$ (blue).
+Stars indicate CEs for the target class.](/paper/figures/poc.svg){#fig-poc}

### 2025-03-15 paper/sections/other/main_results.qmd
vamooooos

--- 2025-03-14 paper/sections/other/main_results.qmd
@@
 inputs = get_table_inputs(df, nothing; backend=Val(:latex))
-tabulate_results(inputs; table_type=:tabular, save_name="paper/quarto_ecml/tables/main.tex", wrap_table=false)
-tabulate_results(inputs; table_type=:tabular, save_name="paper/preprint/tables/main.tex", wrap_table=false)
+formatters = (
+    PrettyTables.ft_round(1,findall(.!contains.(names(df),"Acc"))),
+    PrettyTables.ft_round(2,findall(contains.(names(df),"Acc")))
+)
+for fname in ["paper/quarto_ecml/tables/main.tex","paper/preprint/tables/main.tex"]
+    tabulate_results(
+        inputs; 
+        table_type=:tabular, 
+        save_name=fname, 
+        wrap_table=false, 
+        alignment=[:l,fill(:r,size(df,2)-1)...],
+        formatters=formatters
+    )
+end
 ```

### 2025-03-15 paper/preprint/paper.tex
vamooooos

@@
 to produce some desired output. To be useful in real-word
-decision-making systems, counterfactuals should be plausible with
-respect to the underlying data and actionable with respect to the
-stakeholder requirements. Much existing research has therefore focused
-on developing post-hoc methods to generate counterfactuals that meet
-these desiderata. In this work, we instead hold models directly
-accountable for the desired end goal: counterfactual training employs
-counterfactuals ad-hoc during the training phase to minimize the
-divergence between learned representations and plausible, actionable
-explanations. We demonstrate empirically and theoretically that our
-proposed method facilitates training models that deliver inherently
-desirable explanations while maintaining high predictive performance.
+decision-making systems, counterfactuals should be (1) plausible with
+respect to the underlying data and (2) actionable with respect to the
+user-defined mutability constraints. Much existing research has
+therefore focused on developing post-hoc methods to generate
+counterfactuals that meet these desiderata. In this work, we instead
+hold models directly accountable for the desired end goal:
+counterfactual training employs counterfactuals ad-hoc during the
+training phase to minimize the divergence between learned
+representations and plausible, actionable explanations. We demonstrate
+empirically and theoretically that our proposed method facilitates
+training models that deliver inherently desirable explanations while
+maintaining high predictive performance.
 \end{abstract}
@@
 automated decision-making (ADM) systems involving opaque models
-(\citeproc{ref-oneil2016weapons}{O'Neil 2016};
-\citeproc{ref-mcgregor2021preventing}{McGregor 2021}). Effective XAI
-tools can aid us in monitoring models and providing recourse to
-individuals to turn negative outcomes (e.g., ``loan application
-rejected'') into positive ones (e.g., ``application accepted''). Our
+(\citeproc{ref-mcgregor2021preventing}{McGregor 2021};
+\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI tools can
+aid us in monitoring models and providing recourse to individuals to
+turn negative outcomes (e.g., ``loan application rejected'') into
+positive ones (e.g., ``application accepted''). In line with this, our
 work builds upon \textbf{counterfactual explanations} (CE) proposed by
@@
 explanations that are both plausible and valid for any given model,
-independent of whether the learned representations are also compatible
-with plausible explanations
+independent of whether the learned representations are in fact
+compatible with plausible explanations
 (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
@@
 can arise as a consequence of regularization techniques aimed at other
-training objectives such as robustness, generalization, and generative
-capacity (\citeproc{ref-schut2021generating}{Schut et al. 2021};
+training objectives such as generative capacity, generalization, or
+robustness (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024};
 \citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein 2020};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). As further
+\citeproc{ref-schut2021generating}{Schut et al. 2021}). As further
 discussed in Section~\ref{sec-lit}, our work consolidates these findings
@@
 
-To the best of our knowledge, the proposed framework for counterfactual
-training represents the first attempt to use counterfactual explanations
-during training to improve model explainability. In high-level terms, we
-define model explainability as the extent to which valid explanations
-derived for an opaque model are also deemed plausible with respect to
-the underlying data and (global) actionability constraints. To make the
-desiderata for our framework more concrete, we follow Augustin, Meinke,
-and Hein (\citeproc{ref-augustin2020adversarial}{2020}) in tying the
-concept of explainability to the quality of CEs that can be generated
-for a given model. The authors show that CEs---understood as minimal
-input perturbations that yield some desired model prediction---tend to
-be more meaningful if the underlying model is more robust to adversarial
+To make the desiderata for our framework more concrete, we follow
+Augustin, Meinke, and Hein
+(\citeproc{ref-augustin2020adversarial}{2020}) in tying the concept of
+explainability to the quality of CEs that can be generated for a given
+model. The authors show that CEs---understood as minimal input
+perturbations that yield some desired model prediction---tend to be more
+meaningful if the underlying model is more robust to adversarial
 examples. We can make intuitive sense of this finding when looking at
@@
 for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to
-``unlearn'' susceptiblity to such examples, AT can effectively remove
-implausible explanations from the solution space.
+``unlearn'' susceptiblity to such examples, adversarial training can
+effectively remove implausible explanations from the solution space.
 
@@
 the other is backed by empirical evidence. Sauer and Geiger
-(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
+(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrates that using
 counterfactual images during classifier training improves model
@@
 to improve models through data augmentation. Wu et al.
-(\citeproc{ref-wu2021polyjuice}{2021}) propose \emph{Polyjuice}, a
+(\citeproc{ref-wu2021polyjuice2}{2021}) proposes \emph{Polyjuice}, a
 general-purpose counterfactual generator for language models. The
@@
 Balashankar et al. (\citeproc{ref-balashankar2023improving}{2023})
-similarly use \emph{Polyjuice} to augment NLP datasets through diverse
+similarly uses \emph{Polyjuice} to augment NLP datasets through diverse
 counterfactuals and show that classifier robustness improves by up to
 20\%. Finally, Luu and Inoue
-(\citeproc{ref-luu2023counterfactual}{2023}) introduce Counterfactual
-Adversarial Training (CAT), which also aims at improving generalization
-and robustness of language models through a three-step procedure: the
+(\citeproc{ref-luu2023counterfactual}{2023}) introduces Counterfactual
+Adversarial Training (CAT) that also aims to improve generalization and
+robustness of language models through a three-step procedure: the
 authors identify training samples that are subject to high predictive
-uncertainty, generate CEs for those samples, and then fine-tune the
-language model on a dataset augmented with the CEs.
+uncertainty, generate CEs for them, and fine-tune the language model on
+a dataset augmented with the CEs.
 
@@
 (\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
-Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
-CEs and AEs are equivalent under certain conditions and derive
-theoretical upper bounds on distances between them.
+Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) shows
+that CEs and AEs are equivalent under certain conditions.
 
@@
 Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
-propose a way to train models that guarantee individual recourse to some
-positive target class with high probability. Their approach builds on
-adversarial training by explicitly inducing susceptibility to targeted
-adversarial examples for the positive class. Additionally, the proposed
-method allows for imposing a set of actionability constraints ex-ante.
-For example, users can specify that certain features are immutable.
-Secondly, Guo, Nguyen, and Yadav
-(\citeproc{ref-guo2023counternet}{2023}) are the first to propose an
+proposes a way to train models that guarantee individual recourse to
+some positive target class with high probability. Their approach builds
+on adversarial training by explicitly inducing susceptibility to
+targeted adversarial examples for the positive class. Additionally, the
+proposed method allows for imposing a set of actionability constraints
+ex-ante. For example, users can specify that certain features are
+immutable. Secondly, Guo, Nguyen, and Yadav
+(\citeproc{ref-guo2023counternet}{2023}) is the first to propose an
 end-to-end training pipeline that includes CEs as part of the training
@@
 closely related to this one, Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}) show that explainability can
-be improved through model averaging and refined model objectives. The
-authors propose a way to generate counterfactuals that are maximally
+(\citeproc{ref-altmeyer2024faithful}{2024}) shows that explainability
+can be improved through model averaging and refined model objectives.
+The authors propose a way to generate counterfactuals that are maximally
 faithful to the model in that they are consistent with what the model
 has learned about the underlying data. Formally, they rely on tools from
-energy-based modelling to minimize the divergence between the
-distribution of counterfactuals and the conditional posterior over
-inputs learned by the model. Their proposed counterfactual explainer,
-\emph{ECCCo}, yields plausible explanations if and only if the
-underlying model has learned representations that align with them. The
-authors find that both deep ensembles
-(\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
-and Blundell 2017}) and joint energy-based models (JEMs)
+energy-based modelling (\citeproc{ref-teh2003energy}{Teh et al. 2003})
+to minimize the divergence between the distribution of counterfactuals
+and the conditional posterior over inputs learned by the model. Their
+proposed counterfactual explainer, \emph{ECCCo}, yields plausible
+explanations if and only if the underlying model has learned
+representations that align with them. The authors find that both deep
+ensembles (\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan,
+Pritzel, and Blundell 2017}) and joint energy-based models (JEMs)
 (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
@@
 representations that corresponds to semantically meaningless
-explanations for the data. Previous work by Schut et al.
+explanations for the data. Previous work of Schut et al.
 (\citeproc{ref-schut2021generating}{2021}) similarly found that
-generating plausible (``interpretable'') counterfactual explanations is
-almost trivial for deep ensembles that have also undergone adversarial
-training. The case for JEMs is even clearer: they involve a hybrid
-objective that induces both high predictive performance and generative
-capacity (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This
-is closely related to the idea of aligning models with plausible
-explanations and has inspired our proposed CT objective, as we explain
-in Section~\ref{sec-method}.
+generating plausible (``interpretable'') CEs is almost trivial for deep
+ensembles that have also undergone adversarial training. The case for
+JEMs is even clearer: they involve a hybrid objective that induces both
+high predictive performance and generative capacity
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This is
+closely related to the idea of aligning models with plausible
+explanations and has inspired our CT objective.
 
@@
 
-Counterfactual training combines ideas from adversarial training,
-energy-based modelling and counterfactuals explanations with the
-explicit goal of aligning representations with plausible explanations
-that comply with user requirements. In the context of CEs, plausibility
-has broadly been defined as the degree to which counterfactuals comply
-with the underlying data-generating process
-(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
+In this section we propose our novel counterfactual training objective.
+In CT, we combine ideas from adversarial training, counterfactual
+explanations, and energy-based modelling with the explicit goal of
+aligning representations with plausible explanations that comply with
+user-defined actionability constraints.
+
+In the context of counterfactual explanations, plausibility has broadly
+been defined as the degree to which counterfactuals comply with the
+underlying data-generating process
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024};
 \citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
-is a necessary but insufficient condition for using CEs to provide
+\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020}). Plausibility is
+a necessary but insufficient condition for using CEs to provide
 algorithmic recourse (AR) to individuals (negatively) affected by opaque
-models. For AR recommendations to be actionable, they need to not only
-result in plausible counterfactuals but also be attainable. A plausible
-CE for a rejected 20-year-old loan applicant, for example, might reveal
-that their application would have been accepted, if only they were 20
-years older. Ignoring all other features, this would comply with the
-definition of plausibility if 40-year-old individuals were in fact more
-credit-worthy on average than young adults. But of course this CE does
-not qualify for providing actionable recourse to the applicant since
-\emph{age} is not a (directly) mutable feature. CT aims to improve model
-explainability by aligning models with counterfactuals that meet both
-desiderata: plausibility and actionability. Formally, we define
-explainability as follows:
+models. To be actionable, AR recommendations must also be attainable. A
+plausible CE for a rejected 20-year-old loan applicant, for example,
+might reveal that their application would have been accepted, if only
+they were 20 years older. Ignoring all other features, this would comply
+with the definition of plausibility if 40-year-old individuals were in
+fact more credit-worthy on average than young adults. But of course this
+CE does not qualify for providing actionable recourse to the applicant
+since \emph{age} is not a (directly) mutable feature. Counterfactual
+training aims to improve model explainability by aligning models with
+counterfactuals that meet both desiderata: plausibility and
+actionability. Formally, we define explainability as follows:
 
@@
 
-The characterization of faithfulness and plausibility in Def.
+\noindent The characterization of faithfulness and plausibility in Def.
 \ref{def-explainability} is the same as in Altmeyer et al.
@@
 The first term can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. Then, to optimize \(\log p_\theta(\mathbf{x})\)
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
-contrastive divergence between these observed samples from
+Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\),
+Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimizes the
+contrastive divergence between the observed samples from
 \(p(\mathbf{x})\) and generated samples from \(p_\theta(\mathbf{x})\).
@@
 posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) rely on
+Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) relies on
 Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative
@@
 
-The departure from SGLD allows us to tap into the vast repertoire of
-explainers that have been proposed in the literature to meet different
-desiderata. For example, many methods facilitate the imposition of
-domain and mutability constraints. In principle, any existing approach
-for generating counterfactual explanations is viable, so long as it does
-not violate the faithfulness condition. Like JEMs
+The departure from SGLD of (\citeproc{ref-grathwohl2020your}{Grathwohl
+et al. 2020}) allows us to tap into the vast repertoire of explainers
+that have been proposed in the literature to meet different desiderata.
+For example, many methods facilitate the imposition of domain and
+mutability constraints. In principle, any existing approach for
+generating counterfactual explanations is viable, so long as it does not
+violate the faithfulness condition. Like JEMs
 (\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), CT can be
@@
 \(\mathbf{x}^\prime\) can be repurposed as additional training samples
-(\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023};
-\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023}) or AEs
-(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
+(\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023};
+\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023}) or adversarial
+examples (\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
 \citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
-leaves some flexibility with respect to the choice for
-\(\text{advloss}(\cdot)\) in Equation~\ref{eq-obj}. An intuitive
-functional form, but likely not the only sensible choice, is inspired by
-adversarial training:
+leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
+Equation~\ref{eq-obj}. An intuitive functional form, but likely not the
+only sensible choice, is inspired by adversarial training:
 
@@
 closely aligned with Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2014}) who define an adversarial
+(\citeproc{ref-szegedy2013intriguing}{2014}) that defines an adversarial
 attack as an ``imperceptible non-random perturbation''. Thus, we choose
 to work with a different distinction between CE and AE than Freiesleben
-(\citeproc{ref-freiesleben2022intriguing}{2022}) who consider
+(\citeproc{ref-freiesleben2022intriguing}{2022}) that considers
 misclassification as the key distinguishing feature of AE. One of the
-key observations in this work is that we can leverage CEs during
-training and get adversarial examples essentially for free.
+key observations of this work is that we can leverage CEs during
+training and get adversarial examples essentially for free, which can be
+used to reap the aforementioned benefits of adversarial training.
 
@@
 classifier). In panels (a) and (c) we have trained the models
-conventionally, while in panels (b) and (d) we used CT. Only in panels
-(c) and (d) do we impose the mutability constraint on \emph{age} at test
-time. In each case, we show the decision boundary (in green) and the
-training data colored according to their ground-truth label: orange
-points belong to the target class, \(y^+=2\), blue points belong to the
-non-target class, \(y^-=1\). Stars indicate CEs in the target class
-generated at test time using generic gradient descent until convergence.
-
-In all cases the counterfactuals are valid but their quality differs. In
-panel (a), they are not plausible: they do not comply with the
-distribution of the factuals in \(y^+\) to the point where they form a
-clearly distinguishable cluster. In panel (b), they are highly
-plausible, meeting the first objective of Def. \ref{def-explainability}.
-In panel (c), the CEs involve substantial reductions in \emph{debt} for
-younger applicants. By comparison, counterfactual paths are shorter on
-average in panel (d) where we have protected the immutable \emph{age} as
+conventionally, while in panels (b) and (d) we used CT.
+
+In all cases, all counterfactuals (stars) are valid---they have cross
+the decision boundary (green)---but their quality differs. In panel (a),
+they are not plausible: they do not comply with the distribution of the
+factuals in \(y^+\) to the point where they form a clearly
+distinguishable cluster. In panel (b), they are highly plausible,
+meeting the first objective of Def. \ref{def-explainability}. In panel
+(c), the CEs involve substantial reductions in \emph{debt} for younger
+applicants. By comparison, counterfactual paths are shorter on average
+in panel (d) where we have protected the immutable \emph{age} as
 described in Section~\ref{sec-constraints}. Due to the classifier's
@@
 \item
-  To what extent does our counterfactual training objective as defined
-  in Equation~\ref{eq-obj} induce models to learn plausible
+  To what extent does the counterfactual training objective as it is
+  defined in Equation~\ref{eq-obj} induce models to learn plausible
   explanations?
@@
 
-Our key outcome of interest is how well do models perform with respect
-to explainability (Def. \ref{def-explainability}). To this end, we focus
-primarily on the plausibility and cost of faithfully generated
-counterfactuals at test time. To measure the cost of counterfactuals, we
-follow the standard convention of using distances (\(\ell_1\)-norm)
-between factuals and counterfactuals as a proxy. For plausibility, we
-assess how similar counterfactuals are to observed samples in the target
-domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
-distance-based metric used by Altmeyer et al.
+Our key outcome of interest is improved explainability (Def.
+\ref{def-explainability}). To this end, we focus primarily on the
+plausibility and cost of faithfully generated counterfactuals at test
+time. To measure the cost of counterfactuals, we follow the standard
+convention of using distances (\(\ell_1\)-norm) between factuals and
+counterfactuals as a proxy. For plausibility, we assess how similar
+counterfactuals are to observed samples in the target domain,
+\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the distance-based
+metric used in Altmeyer et al.
 (\citeproc{ref-altmeyer2024faithful}{2024}),
@@
 
-In addition to cost and plausibility, we also compute other standard
-metrics to evaluate counterfactuals at test time including validity and
-redundancy. Finally, we also assess the predictive performance of models
-using standard metrics.
+In addition to cost and plausibility, we compute other standard metrics
+to evaluate counterfactuals including validity and redundancy. Finally,
+we also assess the predictive performance of models using standard
+metrics, including robust accuracy estimated on adversarially perturbed
+data using FGSM (\citeproc{ref-goodfellow2014explaining}{Goodfellow,
+Shlens, and Szegedy 2015}).
 
@@
 the conformal prediction component---as a method that directly targets
-both faithfulness and plausibility of the CEs.
+both faithfulness and plausibility of the counterfactuals.
+
+We make use of nine classification datasets common in the CE/AR
+literature. Four of them are synthetic with two classes and different
+characteristics: linearly separable clusters (\emph{LS}), overlapping
+clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
+moons (\emph{Moon}). These datasets are generated using the library of
+(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, and Liem
+2023}) and we present them in the supplementary appendix. Next, we have
+four real-world binary tabular datasets from the domain of economics:
+\emph{Adult} (a.k.a. Census data) of
+(\citeproc{ref-becker1996adult2}{Becker and Kohavi 1996}), California
+housing (\emph{CH}) of (\citeproc{ref-pace1997sparse}{Pace and Barry
+1997}), Default of Credit Card Clients (\emph{Cred}) of
+(\citeproc{ref-yeh2016default}{Yeh 2016}), and Give Me Some Credit
+(\emph{GMSC}) of (\citeproc{ref-kaggle2011give2}{Kaggle 2011}). Finally,
+for the convenience of illustration, we use of the 10-class \emph{MNIST}
+vision dataset (\citeproc{ref-lecun1998mnist}{LeCun 1998}).
+
+To assess our proposed training regime, we investigate the improvements
+in performance metrics when using CT on top of a weak baseline (BL):
+specifically, a multilayer perceptron (\emph{MLP}). This is the best way
+to get a clear picture of how effective CT is and consistent with how
+assessment is done in the related literature
+(\citeproc{ref-ross2021learning}{Ross, Lakkaraju, and Bastani 2024};
+\citeproc{ref-teney2020learning}{Teney, Abbasnedjad, and Hengel 2020};
+\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
+2015}).
 
@@
 
-\subsubsection{Plausibility.}\label{sec-plaus}
-
-Table~\ref{tbl-main} presents our main empirical findings. The top five
-rows show the percentage reduction in implausibility according to
-Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
-used for \emph{ECCo} at test time. Row 6 shows the reduction in
-implausibility as measured by Equation~\ref{eq-impl-div} and aggregated
-across all test specifications of \emph{ECCo}. Rows 7 and 8 show the
-test accuracies for the model trained with CT and conventionally trained
-models (``vanilla'').
-
-For all datasets except \emph{OL} and across all test settings, the
-average distance of CEs from observed samples in the target class is
-reduced, indicating improved plausibility. The magnitude of improvements
-varies. For the simple synthetic datasets, distance reductions range
-from around 20-40\% (\emph{LS}, \emph{Moon}) to almost 60\%
-(\emph{Circ}). For the real-world tabular datasets, improvements tend to
-be smaller but still substantial, with around 10-15\% for \emph{CH},
-11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred} and around 3\% for
-\emph{Adult}. For our only vision dataset (\emph{MNIST}), distances are
-reduced by up to 9\%. The results for our proposed divergence metric are
-qualitatively similar, but generally even more pronounced: for the
-\emph{Circ} dataset, implausibility is reduced by almost 94\% to
-virtually zero as we verified by the absolute outcome. Improvements for
-other datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
-\emph{OL} the reduction is negative, consistent with the distance-based
-metric. \emph{MNIST} is the only dataset for which the two metrics
-disagree.
-
-These broad and substantial improvements in plausibility generally do
-not come at the cost of decreased predictive performance: test accuracy
-for CT is virtually identical to the baseline for \emph{Adult},
-\emph{Circ}, \emph{LS}, \emph{Moon} and \emph{OL}, and even slightly
-improved for \emph{Cred}. Exceptions to this general pattern are
-\emph{MNIST}, \emph{CH} and \emph{GMSC}, for which we observe reduction
-in test accuracy of 2, 5 and 15 percentage points, respectively. We note
-in this context, that we have not optimized our models for predictive
-performance at all and worked with very small networks. In summary, we
-find that CT can substantially improve the quality of explanations
-learned by models without sacrificing predictive accuracy.
+\subsubsection{Plausibility}\label{sec-plaus}
+
+Table~\ref{tbl-main} presents our main empirical findings. For all
+datasets except \emph{OL} and across all test settings, the average
+distance of CEs from observed samples in the target class is reduced,
+indicating improved plausibility. The magnitude of improvements varies.
+For the simple synthetic datasets, distance reductions range from around
+20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
+real-world tabular datasets, improvements tend to be smaller but still
+substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
+7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For our only
+vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
+results for our proposed divergence metric are qualitatively similar,
+but generally even more pronounced: for the \emph{Circ} dataset,
+implausibility is reduced by almost 94\% to virtually zero as we
+verified by the absolute outcome. Improvements for other datasets range
+from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For \emph{OL} the
+reduction is negative, consistent with the distance-based metric.
+\emph{MNIST} is the only dataset for which the distance and divergence
+metrics disagree. Upon visual inspection of the image counterfactuals we
+find that CT clearly improves plausibility (see appendix).
+
+\subsubsection{Predictive Performance}\label{sec-pred}
+
+Test accuracy for CT is virtually identical to the baseline for
+\emph{Adult}, \emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and
+even slightly improved for \emph{Cred}. Exceptions to this general
+pattern are \emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we
+observe a reduction in test accuracy of 2, 5, and 15 percentage points
+respectively. When looking at robust test accuracies (Acc.\(^*\)) for
+these datasets in particular, we find that CT strongly outperforms the
+baseline. In fact, we find that CT improves adversarial robustness on
+all datasets.
 
@@
 
-\caption{\label{tbl-main}Key explainability and predictive performance
-metrics for all datasets. \textbf{Plausibility}: the top five rows show
-the percentage reduction in implausibility according to
-Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
-used for \emph{ECCo} at test time. The following row shows the reduction
-in implausibility as measured by Equation~\ref{eq-impl-div} and
-aggregated across all test specifications. \textbf{Accuracy}: The
-following two rows show the test accuracies for the models trained with
-CT and the baseline. \textbf{Actionability}: The final row present the
-average reduction in costs when imposing mutability constraints.}
-
-\centering{
-
-\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
+\caption{\label{tbl-main}Key performance metrics across all datasets.
+\textbf{Plausibility}: Columns 2-6 show the percentage reduction in
+implausibility (\(\text{IP}\)) for varying degrees of the energy penalty
+used for \emph{ECCo} at test time; column 7 shows the reduction in
+\(\text{IP}^*\) (\(\text{MMD}\)), aggregated across all test
+specifications. \textbf{Accuracy} (columns 8-11): test accuracies and
+robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
+\textbf{Actionability} (column 9): average reduction in costs when
+imposing mutability constraints.}
+
+\centering{
+
 \input{tables/main.tex}
-\end{tabular*}
 
@@
 favour of mutable features instead. For binding constraints at test
-time, this has the effect of shorter counterfactual paths and hence
-smaller average costs to individuals. To extend this to the non-linear
-case, we test the effect of imposing mutability constraints empirically
-for our synthetic data using the same evaluation scheme as above. The
-final row in Table~\ref{tbl-main} reports the average reduction in costs
-for CT compared to the baseline, when imposing that either the first or
-the second feature is immutable. In all cases, costs are reduced
-substantially indicating that classifiers trained with CT are indeed
-more sensitive to mutable features.
+time, this leads to shorter counterfactual paths and hence smaller
+average costs (\(\ell_1\)-norm) to individuals. To extend this to the
+non-linear case, we test the effect of imposing mutability constraints
+empirically for our synthetic data using the same evaluation scheme as
+above. The final row in Table~\ref{tbl-main} reports the average
+reduction in costs for CT compared to the ``vanilla'' baseline, when
+imposing that either the first or the second feature is immutable. In
+all cases, costs are reduced substantially, indicating that classifiers
+trained with CT are indeed more sensitive to mutable features.
 
@@
 results are in the supplementary appendix.\\
-\textbf{\emph{Hyperparameters of the CE generators.}} First, we observe
-that CT is highly sensitive to hyperparameter settings but (a) there are
-manageable patterns and (b) we can typically identify settings that
-improve either plausibility or cost, and commonly both of them at the
-same time. Second, we note that the choice of a CE generator has a major
-impact on the results. For example, \emph{REVISE} tends to perform the
-worst, most likely because it uses a surrogate VAE to generate
-counterfactuals which impedes faithfulness
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Third,
-increasing \(T\), the maximum number of steps, generally yields better
-outcomes because more CEs can mature in each training epoch. Fourth, the
+\indent We note that CT is highly sensitive to the choice of a CE
+generator and its hyperparameters but (a) there are manageable patterns
+and (b) we can typically identify settings that improve either
+plausibility or cost, and commonly both of them at the same time. For
+example, \emph{REVISE} tends to perform the worst, most likely because
+it uses a surrogate VAE to generate counterfactuals which impedes
+faithfulness (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al.
+2024}). Increasing \(T\), the maximum number of steps, generally yields
+better outcomes because more CEs can mature in each training epoch. The
 impact of \(\tau\), the required decision threshold is more difficult to
@@
 this threshold does not seem to correlate with better outcomes. In fact,
-we have generally found that a choice of \(\tau=0.5\) leads to optimal
-results because it is associated with high proportions of mature
-counterfactuals.\\
-\textbf{\emph{Hyperparameters for penalties.}} We find that the strength
-of the energy regularization, \(\lambda_{\text{reg}}\) is highly
-impactful; energy must be sufficiently regularized to avoid poor
-performance in terms of decreased plausibility and increased costs. The
-sensitivity with respect to \(\lambda_{\text{div}}\) and
-\(\lambda_{\text{adv}}\) is much less evident. While high values of
-\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
-combined with high values of \(\lambda_{\text{div}}\) or
-\(\lambda_{\text{adv}}\), this effect is not very pronounced.\\
-\textbf{\emph{Other hyperparameters.}} We observe that the effectiveness
-and stability of CT is positively associated with the number of
-counterfactuals generated during each training epoch. We also confirm
-that a higher number of training epochs is beneficial. Interestingly, we
-observed desired improvements in explainability when CT was combined
+the choice of \(\tau=0.5\) generally leads to optimal results because it
+is associated with high proportions of mature counterfactuals.\\
+\indent The strength of the energy regularization,
+\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
+performance in terms of decreased plausibility and increased costs if
+insufficiently high. The sensitivity with respect to
+\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
+evident. While high values of \(\lambda_{\text{reg}}\) may increase the
+variability in outcomes when combined with high values of
+\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
+very pronounced.\\
+\indent The effectiveness and stability of CT is positively associated
+with the number of counterfactuals generated during each training epoch.
+We also confirm that a higher number of training epochs is beneficial.
+Interestingly, we observed desired improvements when CT was combined
 with conventional training and applied only for the final 50\% of epochs
-of the complete training process. Put differently, CT may be a way to
-improve the explainability of models in a fine-tuning manner.
+of the complete training process. Put differently, CT can improve the
+explainability of models in a fine-tuning manner.
 
@@
 the AR literature (see, e.g.,
-\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
-and Alfano 2020}) impacting the capacity of CT to fulfill its intended
+(\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
+and Alfano 2020})) impacting the capacity of CT to fulfill its intended
 goal.\\
-\textbf{\emph{Interventions on features may impact fairness.}} We
-provide a tool that allows practitioners to modify the sensitivity of a
-model with respect to certain features, which may have implication for
-the fair and equitable treatment of decision subjects. As protecting a
-set of features leads the model to assign higher relative importance to
+\indent \emph{Interventions on features may impact fairness.} We provide
+a tool that allows practitioners to modify the sensitivity of a model
+with respect to certain features, which may have implication for the
+fair and equitable treatment of decision subjects. As protecting a set
+of features leads the model to assign higher relative importance to
 unprotected features, model owners could misuse our solution by
@@
 groups. When applied irresponsibly, CT could result in an unfairly
-assigned burden of recourse (e.g.,
-\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
-threatening the equality of opportunity in the system
-(\citeproc{ref-bell2024fairness}{Bell et al. 2024}). Still, these
-phenomena are not specific to CT.
-
-We also highlight several interesting directions for future research.\\
-\textbf{\emph{Extending CT beyond classification settings.}} Our
-formulation relies on the distinction between non-target class(es)
-\(y^{-}\) and target class(es) \(y^{+}\) to generate counterfactuals
-through Equation~\ref{eq-obj}. While \(y^{-}\) and \(y^{+}\) can be
-arbitrarily defined, CT requires the output space \(\mathcal{Y}\) to be
-discrete. Thus, it does not apply to ML tasks where the change in
-outcome cannot be readily quantified. Focus on classification models is
-a common restriction in research on CEs and AR. Other settings have
-attracted some interest (e.g., regression in
-(\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021};
-\citeproc{ref-zhao2023counterfactual}{Zhao, Broelemann, and Kasneci
-2023})), but there is little consensus how to robustly extend the notion
-of CEs.\\
-\textbf{\emph{Addressing the training instabilities.}} JEMs are
-susceptible to instabilities during training
+assigned burden of recourse (\citeproc{ref-sharma2020certifai}{Sharma,
+Henderson, and Ghosh 2020}), threatening the equality of opportunity in
+the system (\citeproc{ref-bell2024fairness}{Bell et al. 2024}).
+Nonetheless, these phenomena are not specific to CT.
+
+We also highlight several important directions for future research.
+Firstly, it is an interesting challenge to extend CT beyond
+classification settings. Our formulation relies on the distinction
+between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
+generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
+and \(y^{+}\) can be arbitrarily defined, CT requires the output space
+\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
+where the change in outcome cannot be readily quantified. Focus on
+classification models is a common restriction in research on CEs and AR.
+Other settings have attracted some interest (e.g., regression in
+(\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021})), but
+there is little consensus how to robustly extend the notion of CEs.\\
+\indent Secondly, our approach is susceptible to training instabilities.
+This problem has been recognized for JEMs
 (\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
-though we depart from the SGLD-based sampling, we still encounter major
-variability in the outcomes. CT is exposed to two potential sources of
-instabilities: (1) the energy-based contrastive divergence term in
-Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
-Still, we find that training instabilities can be successfully mitigated
-by regularizing energy (\(\lambda_{\text{reg}}\)), generating
-sufficiently many counterfactuals during each training epoch, and
-including only mature counterfactuals for contrastive divergence.\\
-\textbf{\emph{Improving hyperparameter selection procedures.}} Our
-method benefits from the tuning of certain key hyperparameters (see
-Section~\ref{sec-hyperparameters}). In this work, we have relied
-exclusively on grid search for this task. Future work on CT could
-benefit from investigating more sophisticated approaches towards
-hyperparameter tuning. Notably, CT is iterative which makes a variety of
-methods applicable, including Bayesian (e.g.,
-\citeproc{ref-snoek2012practical}{Snoek, Larochelle, and Adams 2012}) or
-gradient-based (e.g., \citeproc{ref-franceschi2017forward}{Franceschi et
-al. 2017}) optimization.
+though we depart from the SGLD-based sampling, we still encounter
+considerable variability in the outcomes. CT is exposed to two potential
+sources of instabilities: (1) the energy-based contrastive divergence
+term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
+explainers. We find several promising ways to mitigate this problem:
+regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
+many counterfactuals during each epoch, and including only mature
+counterfactuals for contrastive divergence.\\
+\indent Finally, we believe that it is possible to substantially improve
+hyperparameter selection procedures. Our method benefits from the tuning
+of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
+In this work, we have relied exclusively on grid search for this task.
+Future work on CT could benefit from investigating more sophisticated
+approaches. Notably, CT is iterative which makes methods such as
+Bayesian or gradient-based optimization applicable (see, e.g.,
+(\citeproc{ref-bischl2023hyperparameter}{Bischl et al. 2023})).
 
@@
 {``{Explaining Black-Box Models through Counterfactuals}.''} In
-\emph{Proceedings of the JuliaCon Conferences}, 1:130. 1.
+\emph{Proceedings of the JuliaCon Conferences}, 1:130.
 
@@
 
+\bibitem[\citeproctext]{ref-becker1996adult2}
+Becker, Barry, and Ronny Kohavi. 1996. {``{Adult}.''} UCI Machine
+Learning Repository.
+
 \bibitem[\citeproctext]{ref-bell2024fairness}
@@
 
+\bibitem[\citeproctext]{ref-bischl2023hyperparameter}
+Bischl, Bernd, Martin Binder, Michel Lang, Tobias Pielok, Jakob Richter,
+Stefan Coors, Janek Thomas, et al. 2023. {``{Hyperparameter
+optimization: Foundations, algorithms, best practices, and open
+challenges}.''} \emph{WIREs Data Mining and Knowledge Discovery} 13 (2):
+e1484. https://doi.org/\url{https://doi.org/10.1002/widm.1484}.
+
 \bibitem[\citeproctext]{ref-milan2023dataframes}
@@
 \url{https://arxiv.org/abs/1903.08689}.
-
-\bibitem[\citeproctext]{ref-franceschi2017forward}
-Franceschi, Luca, Michele Donini, Paolo Frasconi, and Massimiliano
-Pontil. 2017. {``{Forward and Reverse Gradient-Based Hyperparameter
-Optimization}.''} In \emph{Proceedings of the 34th International
-Conference on Machine Learning}, 1165--73. ICML'17. Sydney, NSW,
-Australia: JMLR.org.
 
@@
 
+\bibitem[\citeproctext]{ref-kaggle2011give2}
+Kaggle. 2011. {``Give Me Some Credit, {Improve} on the State of the Art
+in Credit Scoring by Predicting the Probability That Somebody Will
+Experience Financial Distress in the Next Two Years.''}
+https://www.kaggle.com/c/GiveMeSomeCredit; {Kaggle}.
+
 \bibitem[\citeproctext]{ref-kolter2023keynote}
@@
 
+\bibitem[\citeproctext]{ref-lecun1998mnist}
+LeCun, Yann. 1998. {``{The MNIST database of handwritten digits}.''}
+http://yann.lecun.com/exdb/mnist/.
+
 \bibitem[\citeproctext]{ref-lippe2024uvadlc}
@@
 Increases Inequality and Threatens Democracy}. {Crown}.
+
+\bibitem[\citeproctext]{ref-pace1997sparse}
+Pace, R Kelley, and Ronald Barry. 1997. {``Sparse Spatial
+Autoregressions.''} \emph{Statistics \& Probability Letters} 33 (3):
+291--97. \url{https://doi.org/10.1016/s0167-7152(96)00140-x}.
 
@@
 
-\bibitem[\citeproctext]{ref-snoek2012practical}
-Snoek, Jasper, Hugo Larochelle, and Ryan P. Adams. 2012. {``{Practical
-Bayesian Optimization of Machine Learning Algorithms}.''} In
-\emph{Proceedings of the 26th International Conference on Neural
-Information Processing Systems - Volume 2}, edited by F. Pereira, C. J.
-Burges, L. Bottou, and K. Q. Weinberger. Vol. 25. NIPS'12. Curran
-Associates, Inc.
-
 \bibitem[\citeproctext]{ref-spooner2021counterfactual}
@@
 Properties of Neural Networks.''} \url{https://arxiv.org/abs/1312.6199}.
+
+\bibitem[\citeproctext]{ref-teh2003energy}
+Teh, Yee Whye, Max Welling, Simon Osindero, and Geoffrey E. Hinton.
+2003. {``Energy-Based Models for Sparse Overcomplete Representations.''}
+\emph{J. Mach. Learn. Res.} 4 (null): 1235--60.
 
@@
 
-\bibitem[\citeproctext]{ref-wu2021polyjuice}
+\bibitem[\citeproctext]{ref-wu2021polyjuice2}
 Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
@@
 11th International Joint Conference on Natural Language Processing
-(Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
-and Roberto Navigli, 6707--23. Online: ACL.
+(Volume 1: Long Papers)}, 6707--23. ACL.
 \url{https://doi.org/10.18653/v1/2021.acl-long.523}.
+
+\bibitem[\citeproctext]{ref-yeh2016default}
+Yeh, I-Cheng. 2016. {``{Default of Credit Card Clients}.''} UCI Machine
+Learning Repository.
 
@@
 \url{https://doi.org/10.1145/3446776}.
-
-\bibitem[\citeproctext]{ref-zhao2023counterfactual}
-Zhao, Xuan, Klaus Broelemann, and Gjergji Kasneci. 2023.
-{``{Counterfactual Explanation for Regression via Disentanglement in
-Latent Space}.''} In \emph{2023 IEEE International Conference on Data
-Mining Workshops (ICDMW)}, 976--84. Los Alamitos, CA, USA: IEEE Computer
-Society. \url{https://doi.org/10.1109/ICDMW60847.2023.00130}.
 
@@
 \begin{equation}\phantomsection\label{eq-grad}{
-\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
+\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x}^+,\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x}^+)[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x^+_{d}
 }\end{equation}
@@
 are more likely to have
-\((x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0\) than vice
-versa at initialization. Similarly, we are more likely to have
-\((x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0\) since
+\((x_{d_{\text{immtbl}}}^\prime - x^+_{d_{\text{immtbl}}}) < 0\) than
+vice versa at initialization. Similarly, we are more likely to have
+\((x_{d_{\text{mtbl}}}^\prime - x^+_{d_{\text{mtbl}}}) > 0\) since
 \((\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})\).
@@
 
+\subsubsection{Default Hyperparameters}\label{default-hyperparameters}
+
+Note~\ref{nte-train-default} presents the default hyperparameters used
+during training.
+
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-train-default}: Training Phase}]
+
+\quartocalloutnte{nte-train-default}
+
+\begin{itemize}
+\tightlist
+\item
+  Meta Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Generator: \texttt{ecco}
+  \item
+    Model: \texttt{mlp}
+  \end{itemize}
+\item
+  Model:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Activation: \texttt{relu}
+  \item
+    No.~Hidden: \texttt{32}
+  \item
+    No.~Layers: \texttt{1}
+  \end{itemize}
+\item
+  Training Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Burnin: \texttt{0.0}
+  \item
+    Class Loss: \texttt{logitcrossentropy}
+  \item
+    Convergence: \texttt{threshold}
+  \item
+    Generator Parameters:
+
+    \begin{itemize}
+    \tightlist
+    \item
+      Decision Threshold: \texttt{0.75}
+    \item
+      \(\lambda_{\text{cst}}\): \texttt{0.001}
+    \item
+      \(\lambda_{\text{egy}}\): \texttt{5.0}
+    \item
+      Learning Rate: \texttt{0.25}
+    \item
+      Maximum Iterations: \texttt{30}
+    \item
+      Optimizer: \texttt{sgd}
+    \item
+      Type: \texttt{ECCo}
+    \end{itemize}
+  \item
+    \(\lambda_{\text{adv}}\): \texttt{0.25}
+  \item
+    \(\lambda_{\text{clf}}\): \texttt{1.0}
+  \item
+    \(\lambda_{\text{div}}\): \texttt{0.5}
+  \item
+    \(\lambda_{\text{reg}}\): \texttt{0.1}
+  \item
+    Learning Rate: \texttt{0.001}
+  \item
+    No.~Counterfactuals: \texttt{1000}
+  \item
+    No.~Epochs: \texttt{100}
+  \item
+    Objective: \texttt{full}
+  \item
+    Optimizer: \texttt{adam}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
+\subsection{Evaluation Details}\label{sec-app-eval}
+
+Note~\ref{nte-eval-default} presents the default hyperparameters used
+during evaluation.
+
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}]
+
+\quartocalloutnte{nte-eval-default}
+
+\begin{itemize}
+\tightlist
+\item
+  Counterfactual Parameters:
+
+  \begin{itemize}
+  \tightlist
+  \item
+    Convergence: \texttt{threshold}
+  \item
+    Decision Threshold: \texttt{0.95}
+  \item
+    Generator Parameters:
+
+    \begin{itemize}
+    \tightlist
+    \item
+      Decision Threshold: \texttt{0.75}
+    \item
+      \(\lambda_{\text{cst}}\): \texttt{0.001}
+    \item
+      \(\lambda_{\text{egy}}\): \texttt{5.0}
+    \item
+      Learning Rate: \texttt{0.25}
+    \item
+      Maximum Iterations: \texttt{30}
+    \item
+      Optimizer: \texttt{sgd}
+    \item
+      Type: \texttt{ECCo}
+    \end{itemize}
+  \item
+    Maximum Iterations: \texttt{50}
+  \item
+    No.~Individuals: \texttt{100}
+  \item
+    No.~Runs: \texttt{5}
+  \end{itemize}
+\end{itemize}
+
+\end{tcolorbox}
+
 \section{Details on Main Experiments}\label{sec-app-main}
@@
 \caption{\label{tbl-final-params}Final hyperparameters used for the main
-results for the different datasets.}
+results preseneted in Section~\ref{sec-experiments}. Any hyperparameter
+not shown here is set to its default value
+(Note~\ref{nte-train-default}).}
 
@@
 Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
-    Adult & $2.6 \cdot 10^{4}$ & $5.01 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & none & 0.75 & $5 \cdot 10^{3}$ & 0.25 \\
-    CH & $1.65 \cdot 10^{4}$ & $3.1 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & none & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
-    Circ & $3.6 \cdot 10^{3}$ & 600 & 30 & none & 0.5 & $1 \cdot 10^{3}$ & 0.5 \\
-    Cred & $1.06 \cdot 10^{4}$ & $1.92 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & none & 0.5 & $5 \cdot 10^{3}$ & 0.25 \\
-    GMSC & $1.34 \cdot 10^{4}$ & $2.47 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & none & 0.5 & $5 \cdot 10^{3}$ & 0.5 \\
-    LS & $3.6 \cdot 10^{3}$ & 600 & 30 & none & 0.5 & $1 \cdot 10^{3}$ & 0.01 \\
-    MNIST & $1.1 \cdot 10^{4}$ & $2 \cdot 10^{3}$ & $1 \cdot 10^{3}$ & (-1.0, 1.0) & 0.5 & $5 \cdot 10^{3}$ & 0.01 \\
-    Moon & $3.6 \cdot 10^{3}$ & 600 & 30 & none & 0.9 & $1 \cdot 10^{3}$ & 0.25 \\
-    OL & $3.6 \cdot 10^{3}$ & 600 & 30 & none & 0.5 & $1 \cdot 10^{3}$ & 0.25 \\\bottomrule
+    Adult & 26049 & 5010 & 1000 & none & 0.75 & 5000 & 0.25 \\
+    CH & 16504 & 3101 & 1000 & none & 0.5 & 5000 & 0.25 \\
+    Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
+    Cred & 10617 & 1923 & 1000 & none & 0.5 & 5000 & 0.25 \\
+    GMSC & 13371 & 2474 & 1000 & none & 0.5 & 5000 & 0.5 \\
+    LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
+    MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\
+    Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
+    OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\bottomrule
   \end{tabular}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]
 
@@
   \endlastfoot
-  Circ & Accuracy & Full & 0.997 & 0.00309 \\
-  Circ & Accuracy & Vanilla & 0.998 & 0.000557 \\
-  Circ & F1-score & Full & 0.997 & 0.00309 \\
-  Circ & F1-score & Vanilla & 0.998 & 0.000558 \\
-  LS & Accuracy & Full & 0.999 & 0.00201 \\
-  LS & Accuracy & Vanilla & 1 & 0 \\
-  LS & F1-score & Full & 0.999 & 0.00201 \\
-  LS & F1-score & Vanilla & 1 & 0 \\
-  Moon & Accuracy & Full & 0.999 & 0.000696 \\
-  Moon & Accuracy & Vanilla & 1 & 0.00111 \\
-  Moon & F1-score & Full & 0.999 & 0.000696 \\
-  Moon & F1-score & Vanilla & 1 & 0.00111 \\
-  OL & Accuracy & Full & 0.915 & 0.00477 \\
-  OL & Accuracy & Vanilla & 0.917 & 0.00123 \\
-  OL & F1-score & Full & 0.915 & 0.00478 \\
-  OL & F1-score & Vanilla & 0.917 & 0.00124 \\\bottomrule
+  Circ & Accuracy & Full & 1.0 & 0.0 \\
+  Circ & Accuracy & Vanilla & 1.0 & 0.0 \\
+  Circ & F1-score & Full & 1.0 & 0.0 \\
+  Circ & F1-score & Vanilla & 1.0 & 0.0 \\
+  LS & Accuracy & Full & 1.0 & 0.0 \\
+  LS & Accuracy & Vanilla & 1.0 & 0.0 \\
+  LS & F1-score & Full & 1.0 & 0.0 \\
+  LS & F1-score & Vanilla & 1.0 & 0.0 \\
+  Moon & Accuracy & Full & 1.0 & 0.0 \\
+  Moon & Accuracy & Vanilla & 1.0 & 0.0 \\
+  Moon & F1-score & Full & 1.0 & 0.0 \\
+  Moon & F1-score & Vanilla & 1.0 & 0.0 \\
+  OL & Accuracy & Full & 0.91 & 0.0 \\
+  OL & Accuracy & Vanilla & 0.92 & 0.0 \\
+  OL & F1-score & Full & 0.91 & 0.0 \\
+  OL & F1-score & Vanilla & 0.92 & 0.0 \\\bottomrule
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-gen_params-plaus-circles}Average outcomes for
-the plausibility measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+the plausibility measure across hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-gen_params-plaus-lin_sep}Average outcomes for
-the plausibility measure across hyperparameters. Data: Linearly
-Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+the plausibility measure across hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-gen_params-plaus-moons}Average outcomes for the
-plausibility measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-gen_params-plaus-over}Average outcomes for the
-plausibility measure across hyperparameters. Data: Overlapping.}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Overlapping.}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}
 
@@
 \caption{\label{fig-grid-gen_params-cost-circles}Average outcomes for
-the cost measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+the cost measure across hyperparameters. This shows the \% change from
+the baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}
 
@@
 \caption{\label{fig-grid-gen_params-cost-lin_sep}Average outcomes for
-the cost measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+the cost measure across hyperparameters. This shows the \% change from
+the baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data: Linearly
+Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}
 
@@
 \caption{\label{fig-grid-gen_params-cost-moons}Average outcomes for the
-cost measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}}
+cost measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}
 
@@
 \caption{\label{fig-grid-gen_params-cost-over}Average outcomes for the
-cost measure across hyperparameters. Data: Overlapping.}
+cost measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data:
+Overlapping.}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}]
 
@@
   \endlastfoot
-  Circ & Accuracy & Full & 0.994 & 0.0144 \\
-  Circ & Accuracy & Vanilla & 0.998 & 0.000875 \\
-  Circ & F1-score & Full & 0.994 & 0.0145 \\
-  Circ & F1-score & Vanilla & 0.998 & 0.000875 \\
-  LS & Accuracy & Full & 0.998 & 0.00772 \\
-  LS & Accuracy & Vanilla & 1 & 0 \\
-  LS & F1-score & Full & 0.998 & 0.00773 \\
-  LS & F1-score & Vanilla & 1 & 0 \\
-  Moon & Accuracy & Full & 0.987 & 0.0351 \\
-  Moon & Accuracy & Vanilla & 0.998 & 0.0101 \\
-  Moon & F1-score & Full & 0.987 & 0.0352 \\
-  Moon & F1-score & Vanilla & 0.998 & 0.0102 \\
-  OL & Accuracy & Full & 0.911 & 0.0217 \\
-  OL & Accuracy & Vanilla & 0.916 & 0.00236 \\
-  OL & F1-score & Full & 0.911 & 0.0219 \\
-  OL & F1-score & Vanilla & 0.916 & 0.00236 \\\bottomrule
+  Circ & Accuracy & Full & 0.99 & 0.01 \\
+  Circ & Accuracy & Vanilla & 1.0 & 0.0 \\
+  Circ & F1-score & Full & 0.99 & 0.01 \\
+  Circ & F1-score & Vanilla & 1.0 & 0.0 \\
+  LS & Accuracy & Full & 1.0 & 0.01 \\
+  LS & Accuracy & Vanilla & 1.0 & 0.0 \\
+  LS & F1-score & Full & 1.0 & 0.01 \\
+  LS & F1-score & Vanilla & 1.0 & 0.0 \\
+  Moon & Accuracy & Full & 0.99 & 0.04 \\
+  Moon & Accuracy & Vanilla & 1.0 & 0.01 \\
+  Moon & F1-score & Full & 0.99 & 0.04 \\
+  Moon & F1-score & Vanilla & 1.0 & 0.01 \\
+  OL & Accuracy & Full & 0.91 & 0.02 \\
+  OL & Accuracy & Vanilla & 0.92 & 0.0 \\
+  OL & F1-score & Full & 0.91 & 0.02 \\
+  OL & F1-score & Vanilla & 0.92 & 0.0 \\\bottomrule
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-pen-plaus-circles}Average outcomes for the
-plausibility measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-pen-plaus-lin_sep}Average outcomes for the
-plausibility measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-pen-plaus-moons}Average outcomes for the
-plausibility measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-pen-plaus-over}Average outcomes for the
-plausibility measure across hyperparameters. Data: Overlapping.}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Overlapping.}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}
 
@@
 \caption{\label{fig-grid-pen-cost-circles}Average outcomes for the cost
-measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}
 
@@
 \caption{\label{fig-grid-pen-cost-lin_sep}Average outcomes for the cost
-measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data: Linearly
+Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}
 
@@
 \caption{\label{fig-grid-pen-cost-moons}Average outcomes for the cost
-measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}}
+measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}
 
@@
 \caption{\label{fig-grid-pen-cost-over}Average outcomes for the cost
-measure across hyperparameters. Data: Overlapping.}
+measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data:
+Overlapping.}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-train-final-run-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}]
 
@@
   \endlastfoot
-  Circ & Accuracy & Full & 0.995 & 0.00431 \\
-  Circ & Accuracy & Vanilla & 0.998 & 0.000566 \\
-  Circ & F1-score & Full & 0.995 & 0.00432 \\
-  Circ & F1-score & Vanilla & 0.998 & 0.000566 \\
-  LS & Accuracy & Full & 0.999 & 0.00231 \\
-  LS & Accuracy & Vanilla & 1 & 0 \\
-  LS & F1-score & Full & 0.999 & 0.00231 \\
-  LS & F1-score & Vanilla & 1 & 0 \\
-  Moon & Accuracy & Full & 0.996 & 0.0136 \\
-  Moon & Accuracy & Vanilla & 0.988 & 0.022 \\
-  Moon & F1-score & Full & 0.996 & 0.0136 \\
-  Moon & F1-score & Vanilla & 0.988 & 0.022 \\
-  OL & Accuracy & Full & 0.914 & 0.00563 \\
-  OL & Accuracy & Vanilla & 0.918 & 0.00116 \\
-  OL & F1-score & Full & 0.914 & 0.0057 \\
-  OL & F1-score & Vanilla & 0.918 & 0.00116 \\\bottomrule
+  Circ & Accuracy & Full & 0.99 & 0.0 \\
+  Circ & Accuracy & Vanilla & 1.0 & 0.0 \\
+  Circ & F1-score & Full & 0.99 & 0.0 \\
+  Circ & F1-score & Vanilla & 1.0 & 0.0 \\
+  LS & Accuracy & Full & 1.0 & 0.0 \\
+  LS & Accuracy & Vanilla & 1.0 & 0.0 \\
+  LS & F1-score & Full & 1.0 & 0.0 \\
+  LS & F1-score & Vanilla & 1.0 & 0.0 \\
+  Moon & Accuracy & Full & 1.0 & 0.01 \\
+  Moon & Accuracy & Vanilla & 0.99 & 0.02 \\
+  Moon & F1-score & Full & 1.0 & 0.01 \\
+  Moon & F1-score & Vanilla & 0.99 & 0.02 \\
+  OL & Accuracy & Full & 0.91 & 0.01 \\
+  OL & Accuracy & Vanilla & 0.92 & 0.0 \\
+  OL & F1-score & Full & 0.91 & 0.01 \\
+  OL & F1-score & Vanilla & 0.92 & 0.0 \\\bottomrule
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-train-plaus-circles}Average outcomes for the
-plausibility measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-train-plaus-lin_sep}Average outcomes for the
-plausibility measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-train-plaus-moons}Average outcomes for the
-plausibility measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-grid-train-plaus-over}Average outcomes for the
-plausibility measure across hyperparameters. Data: Overlapping.}
+plausibility measure across hyperparameters. This shows the \% change
+from the baseline model for the distance-based implausibility metric
+(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Overlapping.}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}
 
@@
 \caption{\label{fig-grid-train-cost-circles}Average outcomes for the
-cost measure across hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+cost measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}
 
@@
 \caption{\label{fig-grid-train-cost-lin_sep}Average outcomes for the
-cost measure across hyperparameters. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+cost measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data: Linearly
+Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}
 
@@
 \caption{\label{fig-grid-train-cost-moons}Average outcomes for the cost
-measure across hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}}
+measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}
 
@@
 \caption{\label{fig-grid-train-cost-over}Average outcomes for the cost
-measure across hyperparameters. Data: Overlapping.}
+measure across hyperparameters. This shows the \% change from the
+baseline model for the distance-based cost metric
+(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
+Russell 2017}). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCo}). Data:
+Overlapping.}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-warning-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-warning-color-frame, rightrule=.15mm, coltitle=black, title={Package Version (Reproducibility)}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-warning-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-warning-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Package Version (Reproducibility)}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-tune-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}]
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune-plaus-adult}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Adult.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune-plaus-cali}Average outcomes for the
-plausibility measure across key hyperparameters. Data: California
-Housing.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: California Housing.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune-plaus-circles}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune-plaus-credit}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Credit.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune-plaus-gmsc}Average outcomes for the
-plausibility measure across key hyperparameters. Data: GMSC.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune-plaus-lin_sep}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Linearly
-Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune-plaus-mnist}Average outcomes for the
-plausibility measure across key hyperparameters. Data: MNIST.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune-plaus-moons}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune-plaus-over}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Overlapping.}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Overlapping.}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-tune_lr-train}: Training Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, left=2mm, leftrule=.75mm, opacityback=0, colback=white, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, toprule=.15mm, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, rightrule=.15mm, coltitle=black, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, titlerule=0mm, bottomtitle=1mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}]
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune_lr-plaus-adult}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Adult.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Adult.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune_lr-plaus-credit}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Credit.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Credit.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune_lr-plaus-gmsc}Average outcomes for the
-plausibility measure across key hyperparameters. Data: GMSC.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/lin_sep/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: GMSC.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/lin_sep/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune_lr-plaus-lin_sep}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Linearly
-Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Linearly Separable.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune_lr-plaus-mnist}Average outcomes for the
-plausibility measure across key hyperparameters. Data: MNIST.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
 
@@
 \caption{\label{fig-tune_lr-plaus-over}Average outcomes for the
-plausibility measure across key hyperparameters. Data: Overlapping.}
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Overlapping.}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/lin_sep/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/lin_sep/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}}
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
 

### 2025-03-15 paper/sections/other/constraints.qmd
vamooooos

@@
 $$
-\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x},\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x})[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x_{d}
+\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x}^+,\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x}^+)[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x^+_{d}
 $$ {#eq-grad}
@@
 
-Since $(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})$ we are more likely to have $(x_{d_{\text{immtbl}}}^\prime - x_{d_{\text{immtbl}}}) < 0$ than vice versa at initialization.
-Similarly, we are more likely to have $(x_{d_{\text{mtbl}}}^\prime - x_{d_{\text{mtbl}}}) > 0$ since $(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})$.
+Since $(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})$ we are more likely to have $(x_{d_{\text{immtbl}}}^\prime - x^+_{d_{\text{immtbl}}}) < 0$ than vice versa at initialization.
+Similarly, we are more likely to have $(x_{d_{\text{mtbl}}}^\prime - x^+_{d_{\text{mtbl}}}) > 0$ since $(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})$.
 

### 2025-03-15 paper/sections/other/grids.qmd
vamooooos

@@
 DotEnv.load!()
+
+img_width = 80
 ```
@@
 
-### Accuracy
+### Predictive Performance
+
+Predictive performance measures for this grid search are shown in @tbl-acc-gen.
 
@@
 imgfname = "plausibility_distance_from_target.png"
-fig_caption = "Average outcomes for the plausibility measure across hyperparameters."
+fig_caption = "Average outcomes for the plausibility measure across hyperparameters.
+This shows the % change from the baseline model for the distance-based implausibility metric (@eq-impl-dist).
+Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCo*)."
 full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
-include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption) 
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption, width=img_width) 
 _str = join(include_img_commands, "\n\n")
@@
 imgfname = "distance.png"
-fig_caption = "Average outcomes for the cost measure across hyperparameters."
+fig_caption = "Average outcomes for the cost measure across hyperparameters.
+This shows the % change from the baseline model for the distance-based cost metric [@wachter2017counterfactual].
+Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCo*)."
 full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
-include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption)
+include_img_commands = CTExperiments.get_img_command(data_names, full_paths, fig_labels; fig_caption, width=img_width)
 _str = join(include_img_commands, "\n\n")
@@
 
-### Accuracy
+### Predictive Performance
+
+Predictive performance measures for this grid search are shown in @tbl-acc-pen.
 
@@
 
-### Accuracy
+### Predictive Performance
+
+Predictive performance measures for this grid search are shown in @tbl-acc-train.
 

### 2025-03-15 paper/sections/other/main_results.qmd
vamooooos

@@
 df = final_params(res_dir)
+float_cols = ["decision_threshold", "lambda_energy_reg"]
 get_table_inputs(df, nothing; backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; table_type=:tabular)
+    inputs -> tabulate_results(inputs; table_type=:tabular, formatters=PrettyTables.ft_round(2, findall([x in float_cols for x in names(df)])))
 ```
@@
 
-Final hyperparameters used for the main results for the different datasets.
+Final hyperparameters used for the main results preseneted in @sec-experiments. Any hyperparameter not shown here is set to its default value (@nte-train-default).
 

### 2025-03-15 paper/sections/other/training.qmd
vamooooos

@@
+```{julia}
+projectdir = splitpath(pwd()) |>
+    ss -> joinpath(ss[1:findall([s == "CounterfactualTraining.jl" for s in ss])[1]]...) 
+cd(projectdir)
+
+using CTExperiments
+using CTExperiments.CSV
+using CTExperiments.DataFrames
+using CTExperiments.StatsBase
+using Logging
+
+using DotEnv
+DotEnv.load!()
+```
+
 ## Training Details {#sec-app-training}
@@
 While the details laid out here are not crucial for understanding our proposed approach, they are of importance to anyone looking to implement counterfactual training.
+
+### Default Hyperparameters
+
+@nte-train-default presents the default hyperparameters used during training.
+
+::: {#nte-train-default .callout-note}
+
+## Training Phase
+
+```{julia}
+#| output: asis
+
+dict = Logging.with_logger(Logging.NullLogger()) do 
+    CTExperiments.to_dict(Experiment()) 
+end
+println(CTExperiments.dict_to_quarto_markdown(dict, filter_empty=false))
+```
+
+:::
+
+## Evaluation Details {#sec-app-eval}
+
+@nte-eval-default presents the default hyperparameters used during evaluation.
+
+::: {#nte-eval-default .callout-note}
+
+## Evaluation Phase
+
+```{julia}
+#| output: asis
+
+dict = Logging.with_logger(Logging.NullLogger()) do  
+    CTExperiments.to_dict(EvaluationConfig(ExperimentGrid()))
+end
+println(CTExperiments.dict_to_quarto_markdown(dict, filter_empty=false))
+```
+
+:::

### 2025-03-15 paper/sections/other/tuning.qmd
vamooooos

@@
 imgfname = "plausibility_distance_from_target.png"
-fig_caption = "Average outcomes for the plausibility measure across key hyperparameters."
+fig_caption = "Average outcomes for the plausibility measure across key hyperparameters.
+This shows the % change from the baseline model for the distance-based implausibility metric (@eq-impl-dist).
+Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCo*)."
 full_paths = joinpath.(data_dirs, joinpath(ce_suffix,imgfname))

### 2025-03-15 paper/experiments/CTExperiments/src/presentation.jl

@@ -33,18 +33,37 @@ function is_empty_value(v::Any)
 end
 
 # Function to filter dictionary
-function filter_dict(dict::Dict; drop_fields=["name", "data", "data_params"])
+function filter_dict(
+    dict::Dict;
+    drop_fields=[
+        "name",
+        "data",
+        "data_params",
+        "experiment_name",
+        "dim_reduction",
+        "nneighbours",
+        "grid_file",
+    ],
+    filter_empty::Bool=true,
+)
 
     # Take care of nested dicts:
     for (k, v) in dict
         if v isa Dict
-            dict[k] = filter_dict(v; drop_fields)
+            dict[k] = filter_dict(v; drop_fields, filter_empty)
         end
     end
 
-    # Filter out empty values and specified fields
-    dict = filter(dict) do (k, v)
-        !is_empty_value(v) && !(k in drop_fields)
+    if filter_empty
+        # Filter out empty values and specified fields
+        dict = filter(dict) do (k, v)
+            !is_empty_value(v) && !(k in drop_fields)
+        end
+    else
+        # Filter out empty values and specified fields
+        dict = filter(dict) do (k, v)
+            !(k in drop_fields)
+        end
     end
 
     return dict
@@ -175,21 +194,22 @@ function format_header(s::String; replacements::Dict=LatexReplacements)
     return s
 end
 
-function to_mkd(dict::Dict, level::Int=0; header::Union{Nothing,String}=nothing)
-    drop_fields = [
-        "name",
-        "concatenate_output",
-        "parallelizer",
-        "store_ce",
-        "threaded",
-        "verbose",
-        "vertical_splits",
-        "grid_file",
-        "inherit",
-        "save_dir",
-        "test_time",
-        "ndiv",
-    ]
+global _drop_fields = [
+    "name",
+    "concatenate_output",
+    "parallelizer",
+    "store_ce",
+    "threaded",
+    "verbose",
+    "vertical_splits",
+    "grid_file",
+    "inherit",
+    "save_dir",
+    "test_time",
+    "ndiv",
+]
+
+function to_mkd(dict::Dict, level::Int=0; header::Union{Nothing,String}=nothing, drop_fields=_drop_fields)
     dict = filter(((k, v),) -> length(v) > 0 && !(k in drop_fields), dict)
 
     # Create indent string based on level
[…]

### 2025-03-15 paper/quarto_ecml/paper.tex
Back to 16 pages, emphasize robustness a bit more

@@
 caused real harms to people who were adversely and unfairly affected by
-automated decision-making (ADM) systems involving opaque models
-\cite{mcgregor2021preventing,oneil2016weapons}. Effective XAI tools can
-aid us in monitoring models and providing recourse to individuals to
-turn negative outcomes (e.g., ``loan application rejected'') into
-positive ones (e.g., ``application accepted''). In line with this, our
-work builds upon \textbf{counterfactual explanations} (CE) proposed by
+automated decision-making (ADM) systems involving opaque models (see,
+e.g., \cite{oneil2016weapons}). Effective XAI tools can aid us in
+monitoring models and providing recourse to individuals to turn negative
+outcomes (e.g., ``loan application rejected'') into positive ones (e.g.,
+``application accepted''). In line with this, our work builds upon
+\textbf{counterfactual explanations} (CE) proposed by
 \cite{wachter2017counterfactual} as an effective approach to achieve
@@
 extensive experiments, we demonstrate that CT substantially improves
-explainability without sacrificing predictive performance. We discuss
-the challenges in Section~\ref{sec-discussion} and conclude in
-Section~\ref{sec-conclusion} that CT is a promising approach towards
-making opaque models more trustworthy.
+explainability and positively contributes to the robustness of trained
+models without sacrificing predictive performance. Finally, in
+Section~\ref{sec-discussion}, we discuss the challenges and conclude
+that CT is a promising approach towards making opaque models more
+trustworthy.
 
@@
 mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
-immutable. The decision boundary is shown in green along with training
-data colored according to their ground-truth label: \(y^+=2\) (orange)
-and \(y^-=1\) (blue). Stars indicate CEs for the target class.}
+immutable. The linear decision boundary is shown in green along with
+training data colored according to their ground-truth label: \(y^-=1\)
+in blue and \(y^+=2\) in orange. Stars indicate counterfactuals in the
+target class.}
 
@@
 
-Our key outcome of interest is improved explainability (Def.
+Our key outcome of interest is improvement in explainability (Def.
 \ref{def-explainability}). To this end, we focus primarily on the
 plausibility and cost of faithfully generated counterfactuals at test
-time. To measure the cost of counterfactuals, we follow the standard
-convention of using distances (\(\ell_1\)-norm) between factuals and
-counterfactuals as a proxy. For plausibility, we assess how similar
-counterfactuals are to observed samples in the target domain,
-\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the distance-based
-metric used in \cite{altmeyer2024faithful},
+time. To measure the cost, we follow the standard convention of using
+distances (\(\ell_1\)-norm) between factuals and counterfactuals as a
+proxy. For plausibility, we assess how similar CEs are to the observed
+samples in the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We
+rely on the distance-based metric used in \cite{altmeyer2024faithful},
 
@@
 
-To assess our proposed training regime, we investigate the improvements
-in performance metrics when using CT on top of a weak baseline (BL):
-specifically, a multilayer perceptron (\emph{MLP}). This is the best way
-to get a clear picture of how effective CT is and consistent with how
-assessment is done in the related literature
-\cite{ross2021learning,teney2020learning,goodfellow2014explaining}.
+To assess CT, we investigate the improvements in performance metrics
+when using it on top of a weak baseline (BL): a multilayer perceptron
+(\emph{MLP}). This is the best way to get a clear picture of the
+effectiveness of CT, and it is consistent with how assessment is done in
+the related literature
+\cite{goodfellow2014explaining,ross2021learning,teney2020learning}.
 
@@
 substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
-7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For our only
-vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
-results for our proposed divergence metric are qualitatively similar,
-but generally even more pronounced: for the \emph{Circ} dataset,
+7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For the vision
+dataset (\emph{MNIST}), distances are reduced by up to 9\%. The results
+for our proposed divergence metric are qualitatively similar, but
+generally even more pronounced: for the \emph{Circ} dataset,
 implausibility is reduced by almost 94\% to virtually zero as we
 verified by the absolute outcome. Improvements for other datasets range
-from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For \emph{OL} the
+from 28\% (\emph{Moon}) up to 78\% (\emph{GMSC}). For \emph{OL} the
 reduction is negative, consistent with the distance-based metric.
@@
 metrics disagree. Upon visual inspection of the image counterfactuals we
-find that CT clearly improves plausibility (see appendix).
+find that CT clearly improves plausibility (see supplementary appendix).
 
@@
 
-\caption{\label{tbl-main}Key performance metrics across all datasets.
-\textbf{Plausibility}: Columns 2-6 show the percentage reduction in
-implausibility (\(\text{IP}\)) for varying degrees of the energy penalty
-used for \emph{ECCo} at test time; column 7 shows the reduction in
-\(\text{IP}^*\) (\(\text{MMD}\)), aggregated across all test
-specifications. \textbf{Accuracy} (columns 8-11): test accuracies and
-robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
-\textbf{Actionability} (column 9): average reduction in costs when
-imposing mutability constraints.}
+\caption{\label{tbl-main}Key performance metrics across all datasets
+(column 1). \textbf{Plausibility}: Columns 2-6 show the percentage
+reduction in implausibility (\(\text{IP}\)) for varying degrees of the
+energy penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test
+time; column 7 shows the reduction in \(\text{IP}^*\) (\(\text{MMD}\)),
+aggregated across all test specifications. \textbf{Accuracy} (columns
+8-11): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
+and the baseline (BL). \textbf{Actionability} (column 12): average
+reduction in costs when imposing mutability constraints reported for the
+four datasets for which we could identify meaningful features to
+protect.}
 
@@
 
-\section{Discussion}\label{sec-discussion}
+\section{Conclusions}\label{sec-discussion}
 
@@
   supplementary appendix for computational details.} Three factors
-attenuate this effect. First, CT amortizes the cost of CEs for the
-training samples. Second, we find that it can retain its value when used
-as a ``fine-tuning'' technique for conventionally-trained models. Third,
-it yields itself to parallel execution, which we have leveraged for our
-own experiments.\\
+attenuate this effect: (1) CT amortizes the cost of CEs for the training
+samples; (2) it can retain its value when used as a ``fine-tuning''
+technique for conventionally-trained models; and (3) it yields itself to
+parallel execution, which we have leveraged for our own experiments.\\
 \indent \emph{Immutable features may have proxies.} We propose an
@@
 preserve enough information about the principals to hinder the
-protections. Delineating actionability is a major undecided challenge in
-the AR literature (see, e.g.,
-\cite{venkatasubramanian2020philosophical}) impacting the capacity of CT
-to fulfill its intended goal.\\
+protections. Delineating actionability is a major open challenge in the
+AR literature (see, e.g., \cite{venkatasubramanian2020philosophical})
+impacting the capacity of CT to fulfill its intended goal.\\
 \indent \emph{Interventions on features may impact fairness.} We provide
@@
 with respect to certain features, which may have implication for the
-fair and equitable treatment of decision subjects. As protecting a set
-of features leads the model to assign higher relative importance to
-unprotected features, model owners could misuse our solution by
-enforcing explanations based on features that are more difficult to
-modify by some (group of) individuals. For example, consider the Adult
-dataset used in our experiments, where \emph{workclass} or
-\emph{education} may be more difficult to change for underpriviledged
-groups. When applied irresponsibly, CT could result in an unfairly
-assigned burden of recourse \cite{sharma2020certifai}, threatening the
-equality of opportunity in the system \cite{bell2024fairness}.
-Nonetheless, these phenomena are not specific to CT.
-
-We also highlight several important directions for future research.
-Firstly, it is an interesting challenge to extend CT beyond
-classification settings. Our formulation relies on the distinction
-between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
-generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
-and \(y^{+}\) can be arbitrarily defined, CT requires the output space
-\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
-where the change in outcome cannot be readily quantified. Focus on
-classification models is a common restriction in research on CEs and AR.
-Other settings have attracted some interest (e.g., regression in
-\cite{spooner2021counterfactual}), but there is little consensus how to
-robustly extend the notion of CEs.\\
+fair and equitable treatment of decision subjects. Model owners could
+misuse this solution by enforcing explanations based on features that
+are more difficult to modify by some (group of) individuals. For
+example, consider the \emph{Adult} dataset used in our experiments,
+where \emph{workclass} or \emph{education} may be more difficult to
+change for underpriviledged groups. When applied irresponsibly, CT could
+result in an unfairly assigned burden of recourse
+\cite{sharma2020certifai}, threatening the equality of opportunity in
+the system \cite{bell2024fairness}. Nonetheless, these phenomena are not
+specific to CT.\\
+\hfill \break \indent We also highlight several important directions for
+future research. Firstly, it is an interesting challenge to extend CT
+beyond classification settings. Our formulation relies on the
+distinction between non-target class(es) \(y^{-}\) and target class(es)
+\(y^{+}\) to generate counterfactuals through Equation~\ref{eq-obj}.
+While \(y^{-}\) and \(y^{+}\) can be arbitrarily defined, CT requires
+the output space \(\mathcal{Y}\) to be discrete. Thus, it does not apply
+to ML tasks where the change in outcome cannot be readily quantified.
+Focus on classification models is a common restriction in research on
+CEs and AR. Other settings have attracted some interest (e.g.,
+regression in \cite{spooner2021counterfactual}), but there is little
+consensus how to robustly extend the notion of CEs.\\
 \indent Secondly, our approach is susceptible to training instabilities.
@@
 Bayesian or gradient-based optimization applicable (see, e.g.,
-\cite{bischl2023hyperparameter}).
+\cite{bischl2023hyperparameter}). \hfill \break
 
-\section{Conclusion}\label{sec-conclusion}
-
-State-of-the-art machine learning models are prone to learning complex
-representations that cannot be interpreted by humans and existing
-post-hoc explainability approaches cannot guarantee that the
-explanations agree with the model's learned representation of data. As a
-step towards addressing this challenge, we introduced counterfactual
-training, a novel training regime that incentivizes highly-explainable
-models. Our approach leads to explanations that are both
-plausible---compliant with the underlying data-generating process---and
-actionable---compliant with user-specified mutability constraints---and
-thus meaningful to their recipients. Through extensive experiments we
-demonstrate that CT satisfies its objectives while preserving the
-predictive performance of the models. Our approach can also be used to
-fine-tune conventionally-trained models and achieve similar gains in
-explainability. Finally, this work showcases that it is practical to
-improve models \emph{and} their explanations at the same time.
+To conclude, state-of-the-art machine learning models are prone to
+learning complex representations that cannot be interpreted by humans
+and existing post-hoc explainability approaches cannot guarantee that
+the explanations agree with the model's learned representation of data.
+As a step towards addressing this challenge, we introduced
+counterfactual training, a novel training regime that incentivizes
+highly-explainable models. Our approach leads to explanations that are
+both plausible---compliant with the underlying data-generating
+process---and actionable---compliant with user-specified mutability
+constraints---and thus meaningful to their recipients. Through extensive
+experiments we demonstrate that CT satisfies its objective while
+promoting robustness and preserving the predictive performance of the
+models. It can also be used to fine-tune conventionally-trained models
+and achieve similar gains. Finally, this work showcases that it is
+practical to improve models \emph{and} their explanations at the same
+time.
 

### 2025-03-15 paper/sections/conclusion.qmd
Back to 16 pages, emphasize robustness a bit more

@@
-# Conclusion {#sec-conclusion}
-State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans and existing post-hoc explainability approaches cannot guarantee that the explanations agree with the model's learned representation of data.
+<!-- # Conclusion {#sec-conclusion} -->
+To conclude, state-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans and existing post-hoc explainability approaches cannot guarantee that the explanations agree with the model's learned representation of data.
 As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models.
 Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients.
-Through extensive experiments we demonstrate that CT satisfies its objectives while preserving the predictive performance of the models.
-Our approach can also be used to fine-tune conventionally-trained models and achieve similar gains in explainability.
+Through extensive experiments we demonstrate that CT satisfies its objective while promoting robustness and preserving the predictive performance of the models.
+It can also be used to fine-tune conventionally-trained models and achieve similar gains.
 Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

### 2025-03-15 paper/sections/discussion.qmd
Back to 16 pages, emphasize robustness a bit more

--- 2025-03-14 paper/sections/discussion.qmd
@@
-# Discussion {#sec-discussion}
+# Conclusions {#sec-discussion}
 
@@
 While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations.
-Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect.
-First, CT amortizes the cost of CEs for the training samples.
-Second, we find that it can retain its value when used as a "fine-tuning" technique for conventionally-trained models.
-Third, it yields itself to parallel execution, which we have leveraged for our own experiments.
+Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect: (1) CT amortizes the cost of CEs for the training samples; (2) it can retain its value when used as a "fine-tuning" technique for conventionally-trained models; and (3) it yields itself to parallel execution, which we have leveraged for our own experiments.
 `\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs.
@@
 Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections.
-Delineating actionability is a major undecided challenge in the AR literature [see, e.g., @venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.
+Delineating actionability is a major open challenge in the AR literature (see, e.g., [@venkatasubramanian2020philosophical]) impacting the capacity of CT to fulfill its intended goal.
 `\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects.
-As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.
-For example, consider the Adult dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups.
+Model owners could misuse this solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.
+For example, consider the *Adult* dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups.
 When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness].
 Nonetheless, these phenomena are not specific to CT.
-
-We also highlight several important directions for future research.
+`\hfill \break`{=latex}
+`\indent`{=latex} We also highlight several important directions for future research.
 Firstly, it is an interesting challenge to extend CT beyond classification settings.
@@
 Notably, CT is iterative which makes methods such as Bayesian or gradient-based optimization applicable (see, e.g., [@bischl2023hyperparameter]).
+`\hfill \break`{=latex}

### 2025-03-15 paper/sections/experiments.qmd
Back to 16 pages, emphasize robustness a bit more

@@
 
-Our key outcome of interest is improved explainability (Def.
+Our key outcome of interest is improvement in explainability (Def.
 \ref{def-explainability}).
 To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
-To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.
-For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$.
+To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.
+For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$.
 We rely on the distance-based metric used in @altmeyer2024faithful,
@@
 
-To assess our proposed training regime, we investigate the improvements in performance metrics when using CT on top of a weak baseline (BL): specifically, a multilayer perceptron (*MLP*).
-This is the best way to get a clear picture of how effective CT is and consistent with how assessment is done in the related literature [@ross2021learning;@teney2020learning;@goodfellow2014explaining].
+To assess CT, we investigate the improvements in performance metrics when using it on top of a weak baseline (BL): a multilayer perceptron (*MLP*).
+This is the best way to get a clear picture of the effectiveness of CT, and it is consistent with how assessment is done in the related literature [@goodfellow2014explaining;@ross2021learning;@teney2020learning].
 
@@
 For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred*, and around 3% for *Adult*.
-For our only vision dataset (*MNIST*), distances are reduced by up to 9%.
+For the vision dataset (*MNIST*), distances are reduced by up to 9%.
 The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome.
-Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*).
+Improvements for other datasets range from 28% (*Moon*) up to 78% (*GMSC*).
 For *OL* the reduction is negative, consistent with the distance-based metric.
 *MNIST* is the only dataset for which the distance and divergence metrics disagree.
-Upon visual inspection of the image counterfactuals we find that CT clearly improves plausibility (see appendix).
+Upon visual inspection of the image counterfactuals we find that CT clearly improves plausibility (see supplementary appendix).
 
@@
 
-Key performance metrics across all datasets.
-**Plausibility**: Columns 2-6 show the percentage reduction in implausibility ($\text{IP}$) for varying degrees of the energy penalty used for *ECCo* at test time; column 7 shows the reduction in $\text{IP}^*$ ($\text{MMD}$), aggregated across all test specifications.
+Key performance metrics across all datasets (column 1).
+**Plausibility**: Columns 2-6 show the percentage reduction in implausibility ($\text{IP}$) for varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time; column 7 shows the reduction in $\text{IP}^*$ ($\text{MMD}$), aggregated across all test specifications.
 **Accuracy** (columns 8-11): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL).
-**Actionability** (column 9): average reduction in costs when imposing mutability constraints.
+**Actionability** (column 12): average reduction in costs when imposing mutability constraints reported for the four datasets for which we could identify meaningful features to protect.
 :::

### 2025-03-15 paper/sections/method.qmd
Back to 16 pages, emphasize robustness a bit more

@@
 ![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable.
-The decision boundary is shown in green along with training data colored according to their ground-truth label: $y^+=2$ (orange) and $y^-=1$ (blue).
-Stars indicate CEs for the target class.](/paper/figures/poc.svg){#fig-poc}
+The linear decision boundary is shown in green along with training data colored according to their ground-truth label: $y^-=1$ in blue and $y^+=2$ in orange.
+Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc}

### 2025-03-15 paper/quarto_ecml/paper.tex
Add robustness to abstract

@@
 training models that deliver inherently desirable explanations while
-maintaining high predictive performance.
+promoting robustness and preserving high predictive performance.
 

### 2025-03-15 paper/quarto_ecml/paper.tex
Fix some final typos

@@
 images during classifier training improves model robustness. Similarly,
-\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
-potentially useful training data in machine learning, especially in
-supervised settings where inputs may be reasonably mapped to multiple
-outputs. They, too, demonstrate that augmenting the training data of
-image classifiers can improve generalization. Finally,
-\cite{teney2020learning} propose an approach using counterfactuals in
+\cite{abbasnejad2020counterfactual} argues that counterfactuals
+represent potentially useful training data in machine learning,
+especially in supervised settings where inputs may be reasonably mapped
+to multiple outputs. They, too, demonstrate that augmenting the training
+data of image classifiers can improve generalization. Finally,
+\cite{teney2020learning} proposes an approach using counterfactuals in
 training that does not rely on data augmentation: they argue that
@@
 function defined as
-\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
-where \(y^+\) denotes the index of the randomly drawn target class,
+\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
+with \(y^+\) denoting the index of the randomly drawn target class,
 \(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
@@
 We have simulated this scenario using synthetic data with independent
-\emph{age} and \emph{debt}, and Gaussian class-conditional densities in
-Figure~\ref{fig-poc}. The four panels show the outcomes for different
-training procedures using the same model architectures (a linear
-classifier). In panels (a) and (c) we have trained the models
+\emph{age} and \emph{debt} features, and Gaussian class-conditional
+densities in Figure~\ref{fig-poc}. The four panels show the outcomes for
+different training procedures using the same model architectures (a
+linear classifier). In panels (a) and (c) we have trained the models
 conventionally, while in panels (b) and (d) we used CT.
 
-In all cases, all counterfactuals (stars) are valid---they have cross
+In all cases, all counterfactuals (stars) are valid---they have crossed
 the decision boundary (green)---but their quality differs. In panel (a),
@@
 metrics disagree. Upon visual inspection of the image counterfactuals we
-find that CT clearly improves plausibility (see supplementary appendix).
+find that CT clearly improves plausibility (see supplementary appendix
+for images).
 

### 2025-03-15 paper/sections/related.qmd
Fix some final typos

@@
 @sauer2021counterfactual demonstrates that using counterfactual images during classifier training improves model robustness.
-Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
+Similarly, @abbasnejad2020counterfactual argues that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
 They, too, demonstrate that augmenting the training data of image classifiers can improve generalization.
-Finally, @teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
+Finally, @teney2020learning proposes an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
 Specifically, their approach relies on identifying similar input samples with different annotations and ensuring that the gradient of the classifier aligns with the vector between such pairs of counterfactual inputs using the cosine distance as the loss function.

### 2025-03-15 paper/preprint/paper.tex
done

@@
 
+Below we provide an overview of some notation used frequently throughout
+the paper:
+
 \begin{itemize}
@@
 \item
+  \(\mathbf{x}\): a single training sample.
+\item
+  \(\mathbf{x}^\prime\): a counterfactual.
+\item
+  \(\mathbf{x}^+\): a training sample in the target class
+  (ground-truth).
+\item
   \(\mathbf{y}^+\): The one-hot encoded output vector for the target
@@
   \(\Theta\): Matrix of parameters.
+\item
+  \(\mathbf{M}(\cdot)\): linear predictions (logits) of the classifier.
 \end{itemize}
@@
 \subsection{Other Technical Details}\label{other-technical-details}
+
+Maximum mean discrepancy is defined as follows,
 
@@
 \begin{aligned}
-MMD({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
+\text{MMD}({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
 \end{aligned}
@@
 
-In our implementation, Equation~\ref{eq-mmd} is by default applied to
-the entire subset of the training data for which \(y=y^+\).
+where \(k(\cdot,\cdot)\) is a kernel function
+(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). We make use of
+a Gaussian kernel with a constant length-scale parameter of \(0.5\). In
+our implementation, Equation~\ref{eq-mmd} is by default applied to the
+entire subset of the training data for which \(y=y^+\).
 
@@
 
-\subsection{Training Details}\label{sec-app-training}
-
-In this section, we describe the training procedure in detail. While the
-details laid out here are not crucial for understanding our proposed
-approach, they are of importance to anyone looking to implement
-counterfactual training.
-
-\subsubsection{Default Hyperparameters}\label{default-hyperparameters}
+\subsection{Training Hyperparameters}\label{sec-app-training}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-train-default}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-default}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-Note~\ref{nte-eval-default} presents the default hyperparameters used
-during evaluation.
-
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}]
+For all of our evaluations, we proceed as follows: for each experiment
+setting we generate multiple counterfactuals (``No.~Counterfactuals''),
+randomly choosing the factual and target class each time
+(Note~\ref{nte-eval-default}). We do this across multiple rounds
+(``No.~Runs'') with different random seeds to account for stochasticity
+(Note~\ref{nte-eval-default}). This is in line with standard practice in
+the related literature on CE. Note~\ref{nte-eval-default} presents the
+default hyperparameters used during evaluation. For our final results
+presented in the main paper, we rely on held out test sets to sample
+factuals (and outputs for our performance metrics). For tuning purposes
+we rely on training or validation sets.
+
+\subsubsection{Robust Accuracy}\label{robust-accuracy}
+
+To evaluate robust accuracy (Acc.\(^*\)), we use the Fast Gradient Sign
+Method (FGSM) to perturb test samples
+(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
+2015}). For the main results, we have set the perturbation size to
+\(\epsilon=0.03\). We have also tested other perturbation sizes, as well
+as randomly perturbed data. Although not reported here, we have
+consistently found strong outperformance of CT compared to the weak
+baseline.
+
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-eval-default}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-warning-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-warning-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Package Version (Reproducibility)}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-warning-color-frame, title={Package Version (Reproducibility)}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, opacityback=0, colframe=quarto-callout-note-color-frame, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, breakable, titlerule=0mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, coltitle=black, toptitle=1mm, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
 

### 2025-03-15 paper/sections/other/notation.qmd
done with grid search appendix
done

@@
 # Notation {.appendix}
+
+Below we provide an overview of some notation used frequently throughout the paper:
 
@@
 - $y^-$: The non-target class and also the index of non-the target class.
+- $\mathbf{x}$: a single training sample.
+- $\mathbf{x}^\prime$: a counterfactual. 
+- $\mathbf{x}^+$: a training sample in the target class (ground-truth).
 - $\mathbf{y}^+$: The one-hot encoded output vector for the target class. 
@@
 - $\Theta$: Matrix of parameters. 
+- $\mathbf{M}(\cdot)$: linear predictions (logits) of the classifier.
+
+## Other Technical Details
+
+Maximum mean discrepancy is defined as follows,
+
+$$
+\begin{aligned}
+\text{MMD}({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
+\end{aligned}
+$$ {#eq-mmd}
+
+where $k(\cdot,\cdot)$ is a kernel function [@gretton2012kernel].
+We make use of a Gaussian kernel with a constant length-scale parameter of $0.5$.
+In our implementation, @eq-mmd is by default applied to the entire subset of the training data for which $y=y^+$.

### 2025-03-15 paper/preprint/paper.tex
ok now

@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-default}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-train-default}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-eval-default}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-eval-default}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 \end{tcolorbox}
+
+\FloatBarrier
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-warning-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-warning-color-frame, title={Package Version (Reproducibility)}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-warning-color!10!white, colframe=quarto-callout-warning-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Package Version (Reproducibility)}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune_lr-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, leftrule=.75mm, colback=white, titlerule=0mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, left=2mm, toptitle=1mm, bottomrule=.15mm, opacitybacktitle=0.6, opacityback=0, breakable]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
 

### 2025-03-15 paper/preprint/paper.tex

@@
 training models that deliver inherently desirable explanations while
-maintaining high predictive performance.
+promoting robustness and preserving high predictive performance.
 \end{abstract}
@@
 caused real harms to people who were adversely and unfairly affected by
-automated decision-making (ADM) systems involving opaque models
-(\citeproc{ref-mcgregor2021preventing}{McGregor 2021};
-\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective XAI tools can
-aid us in monitoring models and providing recourse to individuals to
-turn negative outcomes (e.g., ``loan application rejected'') into
-positive ones (e.g., ``application accepted''). In line with this, our
-work builds upon \textbf{counterfactual explanations} (CE) proposed by
-Wachter, Mittelstadt, and Russell
+automated decision-making (ADM) systems involving opaque models (see,
+e.g., (\citeproc{ref-oneil2016weapons}{O'Neil 2016})). Effective XAI
+tools can aid us in monitoring models and providing recourse to
+individuals to turn negative outcomes (e.g., ``loan application
+rejected'') into positive ones (e.g., ``application accepted''). In line
+with this, our work builds upon \textbf{counterfactual explanations}
+(CE) proposed by Wachter, Mittelstadt, and Russell
 (\citeproc{ref-wachter2017counterfactual}{2017}) as an effective
@@
 robustness. Similarly, Abbasnejad et al.
-(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
+(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argues that
 counterfactuals represent potentially useful training data in machine
@@
 generalization. Finally, Teney, Abbasnedjad, and Hengel
-(\citeproc{ref-teney2020learning}{2020}) propose an approach using
+(\citeproc{ref-teney2020learning}{2020}) proposes an approach using
 counterfactuals in training that does not rely on data augmentation:
@@
 
-Our key outcome of interest is improved explainability (Def.
+Our key outcome of interest is improvement in explainability (Def.
 \ref{def-explainability}). To this end, we focus primarily on the
 plausibility and cost of faithfully generated counterfactuals at test
-time. To measure the cost of counterfactuals, we follow the standard
-convention of using distances (\(\ell_1\)-norm) between factuals and
-counterfactuals as a proxy. For plausibility, we assess how similar
-counterfactuals are to observed samples in the target domain,
-\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the distance-based
-metric used in Altmeyer et al.
+time. To measure the cost, we follow the standard convention of using
+distances (\(\ell_1\)-norm) between factuals and counterfactuals as a
+proxy. For plausibility, we assess how similar CEs are to the observed
+samples in the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We
+rely on the distance-based metric used in Altmeyer et al.
 (\citeproc{ref-altmeyer2024faithful}{2024}),
@@
 
-To assess our proposed training regime, we investigate the improvements
-in performance metrics when using CT on top of a weak baseline (BL):
-specifically, a multilayer perceptron (\emph{MLP}). This is the best way
-to get a clear picture of how effective CT is and consistent with how
-assessment is done in the related literature
-(\citeproc{ref-ross2021learning}{Ross, Lakkaraju, and Bastani 2024};
-\citeproc{ref-teney2020learning}{Teney, Abbasnedjad, and Hengel 2020};
-\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
-2015}).
+To assess CT, we investigate the improvements in performance metrics
+when using it on top of a weak baseline (BL): a multilayer perceptron
+(\emph{MLP}). This is the best way to get a clear picture of the
+effectiveness of CT, and it is consistent with how assessment is done in
+the related literature
+(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
+2015}; \citeproc{ref-ross2021learning}{Ross, Lakkaraju, and Bastani
+2024}; \citeproc{ref-teney2020learning}{Teney, Abbasnedjad, and Hengel
+2020}).
 
@@
 metrics disagree. Upon visual inspection of the image counterfactuals we
-find that CT clearly improves plausibility (see appendix).
+find that CT clearly improves plausibility (see supplementary appendix
+for images).
 
@@
 preserve enough information about the principals to hinder the
-protections. Delineating actionability is a major undecided challenge in
-the AR literature (see, e.g.,
+protections. Delineating actionability is a major open challenge in the
+AR literature (see, e.g.,
 (\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
@@
 with respect to certain features, which may have implication for the
-fair and equitable treatment of decision subjects. As protecting a set
-of features leads the model to assign higher relative importance to
-unprotected features, model owners could misuse our solution by
-enforcing explanations based on features that are more difficult to
-modify by some (group of) individuals. For example, consider the Adult
-dataset used in our experiments, where \emph{workclass} or
-\emph{education} may be more difficult to change for underpriviledged
-groups. When applied irresponsibly, CT could result in an unfairly
-assigned burden of recourse (\citeproc{ref-sharma2020certifai}{Sharma,
-Henderson, and Ghosh 2020}), threatening the equality of opportunity in
-the system (\citeproc{ref-bell2024fairness}{Bell et al. 2024}).
-Nonetheless, these phenomena are not specific to CT.
-
-We also highlight several important directions for future research.
-Firstly, it is an interesting challenge to extend CT beyond
-classification settings. Our formulation relies on the distinction
-between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
-generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
-and \(y^{+}\) can be arbitrarily defined, CT requires the output space
-\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
-where the change in outcome cannot be readily quantified. Focus on
-classification models is a common restriction in research on CEs and AR.
-Other settings have attracted some interest (e.g., regression in
-(\citeproc{ref-spooner2021counterfactual}{Spooner et al. 2021})), but
-there is little consensus how to robustly extend the notion of CEs.\\
+fair and equitable treatment of decision subjects. Model owners could
+misuse this solution by enforcing explanations based on features that
+are more difficult to modify by some (group of) individuals. For
+example, consider the \emph{Adult} dataset used in our experiments,
+where \emph{workclass} or \emph{education} may be more difficult to
+change for underpriviledged groups. When applied irresponsibly, CT could
+result in an unfairly assigned burden of recourse
+(\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
+threatening the equality of opportunity in the system
+(\citeproc{ref-bell2024fairness}{Bell et al. 2024}). Nonetheless, these
+phenomena are not specific to CT.\\
+\hfill \break \indent We also highlight several important directions for
+future research. Firstly, it is an interesting challenge to extend CT
+beyond classification settings. Our formulation relies on the
+distinction between non-target class(es) \(y^{-}\) and target class(es)
+\(y^{+}\) to generate counterfactuals through Equation~\ref{eq-obj}.
+While \(y^{-}\) and \(y^{+}\) can be arbitrarily defined, CT requires
+the output space \(\mathcal{Y}\) to be discrete. Thus, it does not apply
+to ML tasks where the change in outcome cannot be readily quantified.
+Focus on classification models is a common restriction in research on
+CEs and AR. Other settings have attracted some interest (e.g.,
+regression in (\citeproc{ref-spooner2021counterfactual}{Spooner et al.
+2021})), but there is little consensus how to robustly extend the notion
+of CEs.\\
 \indent Secondly, our approach is susceptible to training instabilities.
@@
 (\citeproc{ref-bischl2023hyperparameter}{Bischl et al. 2023})).
-
-\section{Conclusion}\label{sec-conclusion}
-
-State-of-the-art machine learning models are prone to learning complex
-representations that cannot be interpreted by humans and existing
-post-hoc explainability approaches cannot guarantee that the
-explanations agree with the model's learned representation of data. As a
-step towards addressing this challenge, we introduced counterfactual
-training, a novel training regime that incentivizes highly-explainable
-models. Our approach leads to explanations that are both
-plausible---compliant with the underlying data-generating process---and
-actionable---compliant with user-specified mutability constraints---and
-thus meaningful to their recipients. Through extensive experiments we
-demonstrate that CT satisfies its objectives while preserving the
-predictive performance of the models. Our approach can also be used to
-fine-tune conventionally-trained models and achieve similar gains in
-explainability. Finally, this work showcases that it is practical to
-improve models \emph{and} their explanations at the same time.
+\hfill \break
+
+To conclude, state-of-the-art machine learning models are prone to
+learning complex representations that cannot be interpreted by humans
+and existing post-hoc explainability approaches cannot guarantee that
+the explanations agree with the model's learned representation of data.
+As a step towards addressing this challenge, we introduced
+counterfactual training, a novel training regime that incentivizes
+highly-explainable models. Our approach leads to explanations that are
+both plausible---compliant with the underlying data-generating
+process---and actionable---compliant with user-specified mutability
+constraints---and thus meaningful to their recipients. Through extensive
+experiments we demonstrate that CT satisfies its objective while
+promoting robustness and preserving the predictive performance of the
+models. It can also be used to fine-tune conventionally-trained models
+and achieve similar gains. Finally, this work showcases that it is
+practical to improve models \emph{and} their explanations at the same
+time.
 
@@
 \url{https://aclanthology.org/2023.paclic-1.88/}.
-
-\bibitem[\citeproctext]{ref-mcgregor2021preventing}
-McGregor, Sean. 2021. {``{Preventing repeated real world AI failures by
-cataloging incidents: The AI incident database}.''} In \emph{Proceedings
-of the AAAI Conference on Artificial Intelligence}, 35:15458--63. 17.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-train-default}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-default}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-eval-default}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-eval-default}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-train-final-run-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-warning-color!10!white, colframe=quarto-callout-warning-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Package Version (Reproducibility)}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Package Version (Reproducibility)}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-warning-color!10!white, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune_lr-train}: Training Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune_lr-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, toprule=.15mm, breakable, bottomtitle=1mm, left=2mm, leftrule=.75mm, opacityback=0, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, arc=.35mm, opacitybacktitle=0.6, toptitle=1mm, colback=white, titlerule=0mm, rightrule=.15mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
 

### 2025-03-15 paper/sections/method.qmd

@@
 $$ {#eq-div}
-where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$ where $y^+$ denotes the index of the randomly drawn target class, $y^+ \sim p(y)$.
+where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$, with $y^+$ denoting the index of the randomly drawn target class, $y^+ \sim p(y)$.
 Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations.
@@
 
-We have simulated this scenario using synthetic data with independent *age* and *debt*, and Gaussian class-conditional densities in @fig-poc.
+We have simulated this scenario using synthetic data with independent *age* and *debt* features, and Gaussian class-conditional densities in @fig-poc.
 The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier).
@@
 
-In all cases, all counterfactuals (stars) are valid---they have cross the decision boundary (green)---but their quality differs.
+In all cases, all counterfactuals (stars) are valid---they have crossed the decision boundary (green)---but their quality differs.
 In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster.

### 2025-03-15 paper/preprint/paper.tex
final rerender of appendix

@@
 
-\usepackage{lineno}
-\linenumbers
 \newcommand{\runninghead}{A Preprint }
-\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
+\renewcommand{\runninghead}{Counterfactual Training }
 \title{Counterfactual Training: Teaching Models Plausible and Actionable
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-default}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-default}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-eval-default}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 \caption{\label{tbl-final-params}Final hyperparameters used for the main
-results preseneted in Section~\ref{sec-experiments}. Any hyperparameter
+results presented in Section~\ref{sec-experiments}. Any hyperparameter
 not shown here is set to its default value
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-final-run-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Package Version (Reproducibility)}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-warning-color!10!white, colframe=quarto-callout-warning-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-warning-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-warning-color!10!white, title={Package Version (Reproducibility)}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune_lr-train}: Training Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, rightrule=.15mm, left=2mm, toptitle=1mm, titlerule=0mm, opacitybacktitle=0.6, coltitle=black, toprule=.15mm, breakable, leftrule=.75mm, colback=white, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, bottomtitle=1mm, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
 

### 2025-03-17 paper/preprint/paper.tex

@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-default}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-default}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-warning-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-warning-color!10!white, title={Package Version (Reproducibility)}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
-
-Tuning was run using \texttt{v1.1.3} of \texttt{TaijaData}. The
-follow-up version \texttt{v1.1.4} introduced an option to split
-real-world tabular datasets into train and test set, ensuring that
-pre-processing steps like standardization is fit on the training set
-only. If you are rerunning the tuning experiments with a version of
-\texttt{TaijaData} that is higher than \texttt{v1.1.3}, than for the
-default parameters specified in the configuration files, you may end up
-with slightly different results, although we would not expect any
-changes in terms of qualitative findings. For exact reproducibility,
-please use \texttt{v1.1.3}.
-
-\end{tcolorbox}
-
 \subsection{Key Parameters}\label{sec-app-tune-key}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
 

### 2025-03-17 paper/sections/abstract.qmd

@@
 In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
-We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.
+We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while promoting robustness and preserving high predictive performance.

### 2025-03-17 paper/sections/intro.qmd

@@
 Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions.
-On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@mcgregor2021preventing; @oneil2016weapons].
+On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models (see, e.g., [@oneil2016weapons]).
 Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted").
@@
 In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints.
-In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability without sacrificing predictive performance.
-We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.
+In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability and positively contributes to the robustness of trained models without sacrificing predictive performance.
+Finally, in @sec-discussion, we discuss the challenges and conclude that CT is a promising approach towards making opaque models more trustworthy.

### 2025-03-17 paper/sections/other/training.qmd

@@
 
-## Training Details {#sec-app-training}
-
-In this section, we describe the training procedure in detail.
-While the details laid out here are not crucial for understanding our proposed approach, they are of importance to anyone looking to implement counterfactual training.
-
-### Default Hyperparameters
+## Training Hyperparameters {#sec-app-training}
 
@@
 
+For all of our evaluations, we proceed as follows: for each experiment setting we generate multiple counterfactuals ("No.
+Counterfactuals"), randomly choosing the factual and target class each time (@nte-eval-default).
+We do this across multiple rounds ("No.
+Runs") with different random seeds to account for stochasticity (@nte-eval-default).
+This is in line with standard practice in the related literature on CE.
 @nte-eval-default presents the default hyperparameters used during evaluation.
+For our final results presented in the main paper, we rely on held out test sets to sample factuals (and outputs for our performance metrics).
+For tuning purposes we rely on training or validation sets.
+
+### Robust Accuracy
+
+To evaluate robust accuracy (Acc.$^*$), we use the Fast Gradient Sign Method (FGSM) to perturb test samples [@goodfellow2014explaining].
+For the main results, we have set the perturbation size to $\epsilon=0.03$.
+We have also tested other perturbation sizes, as well as randomly perturbed data.
+Although not reported here, we have consistently found strong outperformance of CT compared to the weak baseline.
 

### 2025-03-17 v0.0.4/search_index.js
build based on 9977e2b

@@ -1,0 +1,3 @@
+var documenterSearchIndex = {"docs":
+[{"location":"","page":"Home","title":"Home","text":"CurrentModule = CounterfactualTraining","category":"page"},{"location":"#CounterfactualTraining","page":"Home","title":"CounterfactualTraining","text":"","category":"section"},{"location":"","page":"Home","title":"Home","text":"Documentation for CounterfactualTraining.","category":"page"},{"location":"","page":"Home","title":"Home","text":"","category":"page"},{"location":"","page":"Home","title":"Home","text":"Modules = [CounterfactualTraining]","category":"page"},{"location":"#CounterfactualTraining.AbstractObjective","page":"Home","title":"CounterfactualTraining.AbstractObjective","text":"Base type of training objectives.\n\n\n\n\n\n","category":"type"},{"location":"#CounterfactualTraining.AdversarialObjective","page":"Home","title":"CounterfactualTraining.AdversarialObjective","text":"AdversarialObjective <: AbstractObjective\n\nThe AdversarialObjective is a concrete implementation of the AbstractObjective abstract type that optimizes for:\n\nStandard classification objective (the discriminative task).\nAdversarial classification objective on the counterfactuals (the explainability task).\n\n\n\n\n\n","category":"type"},{"location":"#CounterfactualTraining.AdversarialObjective-2","page":"Home","title":"CounterfactualTraining.AdversarialObjective","text":"(obj::AdversarialObjective)(\n    yhat,\n    y,\n    energy_differential::Vector{<:AbstractFloat}=[0.0f0],\n    regularization::Vector{<:AbstractFloat}=[0.0f0],\n    adversarial_loss::Union{AbstractFloat,Vector{<:AbstractFloat}}=0.0f0;\n    agg=mean,\n    kwrgs...,\n)\n\nIf the adversarial_loss has been computed already, obj::AdversarialObjective can be called directly on predictions yhat and labels y. The different loss components are then added together with a weighting vector lambda.\n\n\n\n\n\n","category":"type"},{"location":"#CounterfactualTraining.AdversarialObjective-Tuple{}","page":"Home","title":"CounterfactualTraining.AdversarialObjective","text":"AdversarialObjective(;\n    class_loss::Function=Flux.Losses.logitcrossentropy,\n    lambda::Vector{<:AbstractFloat}=[1.0, 0.25]\n)\n\nOuter constructor for the Ad
[…]

### 2025-06-25 paper/quarto_aaai/body.tex
frustrating

--- 2025-03-15 paper/quarto_ecml/paper.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[runningheads]{llncs}
-
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[activate=false]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{2}
-
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
 \makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\@ifpackageloaded{hyperref}{}{%
+  \newcommand{\texorpdfstring}[2]{#1}%
+}
+\newcommand{\tightlist}
 \makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
-% Set default figure placement to htbp
-\def\fps@figure{htbp}
-\makeatother
-
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{booktabs}
-\usepackage[misc]{ifsym}
-\newcommand{\corr}{(\Letter)}
-\usepackage{placeins}
-\makeatletter
-
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Fig.}
-\else
-  \newcommand\figurename{Fig.}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-
-\makeatother
-
-\usepackage{bookmark}
-
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
-  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
-  pdfkeywords={Counterfactual Training, Counterfactual
-Explanations, Algorithmic Recourse, Explainable AI, Representation
-Learning},
-  colorlinks=true,
-  linkcolor={black},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-\hypersetup{nolinks=true}
-
 \title{Counterfactual Training: Teaching Models Plausible and Actionable
 Explanations}
-
-\titlerunning{Counterfactual Training}
-
-\author{Author information scrubbed for double-blind reviewing}
-
-\begin{document}
+\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
+Deursen \and Cynthia C. S. Liem}
+\date{}\begin{document}
 \maketitle
-
 \begin{abstract}
@@
 promoting robustness and preserving high predictive performance.
-
-\keywords{Counterfactual Training \and Counterfactual
-Explanations \and Algorithmic Recourse \and Explainable
-AI \and Representation Learning}
-
 \end{abstract}
-
 \section{Introduction}\label{introduction}
@@
 directly from data, guided by narrow objectives such as predictive
-accuracy \cite{goodfellow2016deep}. Modern advances in computing have
+accuracy \citep{goodfellow2016deep}. Modern advances in computing have
 made it possible to provide such models with ever-growing degrees of
@@
 serious scrutiny in recent years. At the very cusp of the deep learning
-(DL) revolution, \cite{szegedy2013intriguing} showed that artificial
+(DL) revolution, \citet{szegedy2013intriguing} showed that artificial
 neural networks (ANN) are sensitive to adversarial examples (AEs):
@@
 notably \textbf{adversarial training}
-\cite{goodfellow2014explaining}---truly robust deep learning remains
+\citep{goodfellow2014explaining}---truly robust deep learning remains
 unattainable even for models that are considered ``shallow'' by today's
-standards \cite{kolter2023keynote}.
+standards \citep{kolter2023keynote}.
 
@@
 many solutions that are locally optimal with respect to narrow
-objectives \cite{wilson2020case}.\footnote{We follow the standard ML
+objectives \citep{wilson2020case}.\footnote{We follow the standard ML
   convention, where ``degrees of freedom'' refer to the number of
@@
 be pruned by up to 90\% while preserving their predictive performance
-\cite{frankle2018lottery}. Similarly, \cite{zhang2021understanding}
+\citep{frankle2018lottery}. Similarly, \citet{zhang2021understanding}
 showed that state-of-the-art neural networks are expressive enough to
@@
 automated decision-making (ADM) systems involving opaque models (see,
-e.g., \cite{oneil2016weapons}). Effective XAI tools can aid us in
+e.g., \citep{oneil2016weapons}). Effective XAI tools can aid us in
 monitoring models and providing recourse to individuals to turn negative
@@
 \textbf{counterfactual explanations} (CE) proposed by
-\cite{wachter2017counterfactual} as an effective approach to achieve
+\citet{wachter2017counterfactual} as an effective approach to achieve
 this goal. CEs prescribe minimal changes for factual inputs that, if
@@
 independent of whether the learned representations are in fact
-compatible with plausible explanations \cite{altmeyer2024faithful}.
+compatible with plausible explanations \citep{altmeyer2024faithful}.
 Fortunately, recent findings indicate that improved explanatory capacity
@@
 robustness
-\cite{altmeyer2024faithful,augustin2020adversarial,schut2021generating}.
+\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
 As further discussed in Section~\ref{sec-lit}, our work consolidates
@@
 To make the desiderata for our framework more concrete, we follow
-\cite{augustin2020adversarial} in tying the concept of explainability to
-the quality of CEs that can be generated for a given model. The authors
-show that CEs---understood as minimal input perturbations that yield
-some desired model prediction---tend to be more meaningful if the
+\citet{augustin2020adversarial} in tying the concept of explainability
+to the quality of CEs that can be generated for a given model. The
+authors show that CEs---understood as minimal input perturbations that
+yield some desired model prediction---tend to be more meaningful if the
 underlying model is more robust to adversarial examples. We can make
@@
 the other is backed by empirical evidence.
-\cite{sauer2021counterfactual} demonstrates that using counterfactual
+\citet{sauer2021counterfactual} demonstrates that using counterfactual
 images during classifier training improves model robustness. Similarly,
-\cite{abbasnejad2020counterfactual} argues that counterfactuals
+\citet{abbasnejad2020counterfactual} argues that counterfactuals
 represent potentially useful training data in machine learning,
@@
 data of image classifiers can improve generalization. Finally,
-\cite{teney2020learning} proposes an approach using counterfactuals in
+\citet{teney2020learning} proposes an approach using counterfactuals in
 training that does not rely on data augmentation: they argue that
@@
 In the natural language processing (NLP) domain, CEs have also been used
-to improve models through data augmentation. \cite{wu2021polyjuice2}
+to improve models through data augmentation. \citet{wu2021polyjuice2}
 proposes \emph{Polyjuice}, a general-purpose counterfactual generator
@@
 augmentation of training data with their method improves robustness in a
-number of NLP tasks. \cite{balashankar2023improving} similarly uses
+number of NLP tasks. \citet{balashankar2023improving} similarly uses
 \emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
 and show that classifier robustness improves by up to 20\%. Finally,
-\cite{luu2023counterfactual} introduces Counterfactual Adversarial
+\citet{luu2023counterfactual} introduces Counterfactual Adversarial
 Training (CAT) that also aims to improve generalization and robustness
@@
 to clear similarities in how CEs and AEs are generated,
-\cite{freiesleben2022intriguing} makes the case for jointly studying the
-opaqueness and robustness problems in representation learning. Formally,
-AEs can be seen as the subset of CEs for which misclassification is
-achieved \cite{freiesleben2022intriguing}. Similarly,
-\cite{pawelczyk2022exploring} shows that CEs and AEs are equivalent
-under certain conditions.
+\citet{freiesleben2022intriguing} makes the case for jointly studying
+the opaqueness and robustness problems in representation learning.
+Formally, AEs can be seen as the subset of CEs for which
+misclassification is achieved \citep{freiesleben2022intriguing}.
+Similarly, \citet{pawelczyk2022exploring} shows that CEs and AEs are
+equivalent under certain conditions.
 
@@
 certain properties of the post-hoc counterfactual explanations. Firstly,
-\cite{ross2021learning} proposes a way to train models that guarantee
+\citet{ross2021learning} proposes a way to train models that guarantee
 individual recourse to some positive target class with high probability.
@@
 actionability constraints ex-ante. For example, users can specify that
-certain features are immutable. Secondly, \cite{guo2023counternet} is
+certain features are immutable. Secondly, \citet{guo2023counternet} is
 the first to propose an end-to-end training pipeline that includes CEs
@@
 Counterfactuals are generated during each training iteration and fed
-back to the predictor network. In contrast to \cite{guo2023counternet},
+back to the predictor network. In contrast to \citet{guo2023counternet},
 we impose no restrictions on the ANN architecture at all.
@@
 towards aligning representations with plausible explanations. In a work
-closely related to this one, \cite{altmeyer2024faithful} shows that
+closely related to this one, \citet{altmeyer2024faithful} shows that
 explainability can be improved through model averaging and refined model
@@
 what the model has learned about the underlying data. Formally, they
-rely on tools from energy-based modelling \cite{teh2003energy} to
+rely on tools from energy-based modelling \citep{teh2003energy} to
 minimize the divergence between the distribution of counterfactuals and
@@
 representations that align with them. The authors find that both deep
-ensembles \cite{lakshminarayanan2016simple} and joint energy-based
-models (JEMs) \cite{grathwohl2020your} tend to do well in this regard.
+ensembles \citep{lakshminarayanan2016simple} and joint energy-based
+models (JEMs) \citep{grathwohl2020your} tend to do well in this regard.
 
@@
 approximate Bayesian model averages, which are most called for when
-models are underspecified by the available data \cite{wilson2020case}.
+models are underspecified by the available data \citep{wilson2020case}.
 Averaging across solutions mitigates the aforementioned risk of relying
@@
 semantically meaningless explanations for the data. Previous work of
-\cite{schut2021generating} similarly found that generating plausible
+\citet{schut2021generating} similarly found that generating plausible
 (``interpretable'') CEs is almost trivial for deep ensembles that have
@@
 they involve a hybrid objective that induces both high predictive
-performance and generative capacity \cite{grathwohl2020your}. This is
+performance and generative capacity \citep{grathwohl2020your}. This is
 closely related to the idea of aligning models with plausible
@@
 underlying data-generating process
-\cite{altmeyer2024faithful,guidotti2022counterfactual,poyiadzi2020face}.
+\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
 Plausibility is a necessary but insufficient condition for using CEs to
@@
 \noindent The characterization of faithfulness and plausibility in Def.
-\ref{def-explainability} is the same as in \cite{altmeyer2024faithful},
+\ref{def-explainability} is the same as in \citet{altmeyer2024faithful},
 with adapted notation. Intuitively, plausible counterfactuals are
@@
 explanation generated through gradient descent over \(T\) iterations as
-initially proposed by \cite{wachter2017counterfactual}. For our
+initially proposed by \citet{wachter2017counterfactual}. For our
 purposes, we let \(T\) vary and consider the counterfactual search as
@@
 of the energy terms involved in \(\text{div}(\cdot)\)
-\cite{du2019implicit}. The trade-off between the components can governed
-through penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
-\(\lambda_{\text{reg}}\).
+\citep{du2019implicit}. The trade-off between the components can
+governed through penalties \(\lambda_{\text{div}}\),
+\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).
 
@@
 
-As observed by \cite{grathwohl2020your}, any classifier can be
+As observed by \citet{grathwohl2020your}, any classifier can be
 re-interpreted as a joint energy-based model (JEM) that learns to
@@
 Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\),
-\cite{grathwohl2020your} minimizes the contrastive divergence between
+\citet{grathwohl2020your} minimizes the contrastive divergence between
 the observed samples from \(p(\mathbf{x})\) and generated samples from
@@
 
-A key empirical finding in \cite{altmeyer2024faithful} was that JEMs
+A key empirical finding in \citet{altmeyer2024faithful} was that JEMs
 tend to do well with respect to the plausibility objective in Def.
@@
 posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
-\cite{grathwohl2020your} relies on Stochastic Gradient Langevin Dynamics
-(SGLD) using an uninformative prior for initialization but we depart
-from their methodology. Instead of SGLD, we propose to use
+\citet{grathwohl2020your} relies on Stochastic Gradient Langevin
+Dynamics (SGLD) using an uninformative prior for initialization but we
+depart from their methodology. Instead of SGLD, we propose to use
 counterfactual explainers to generate counterfactuals of observed
@@
 of observed training samples (positive samples) while increasing the
-energy of counterfactuals (negative samples) \cite{du2019implicit}. As
+energy of counterfactuals (negative samples) \citep{du2019implicit}. As
 the counterfactuals get more plausible (Def. \ref{def-explainability})
 during training, these opposing effects gradually balance each other out
-\cite{lippe2024uvadlc}.
+\citep{lippe2024uvadlc}.
 
-The departure from SGLD of \cite{grathwohl2020your} allows us to tap
+The departure from SGLD of \citep{grathwohl2020your} allows us to tap
 into the vast repertoire of explainers that have been proposed in the
@@
 explanations is viable, so long as it does not violate the faithfulness
-condition. Like JEMs \cite{murphy2022probabilistic}, CT can be
+condition. Like JEMs \citep{murphy2022probabilistic}, CT can be
 considered a form of contrastive representation learning.
@@
 \(\mathbf{x}^\prime\) can be repurposed as additional training samples
-\cite{balashankar2023improving,luu2023counterfactual} or adversarial
-examples \cite{freiesleben2022intriguing,pawelczyk2022exploring}. This
+\citep{balashankar2023improving, luu2023counterfactual} or adversarial
+examples \citep{freiesleben2022intriguing, pawelczyk2022exploring}. This
 leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
@@
 perturbation to any single feature is at most \(\varepsilon\). This is
-closely aligned with \cite{szegedy2013intriguing} that defines an
+closely aligned with \citet{szegedy2013intriguing} that defines an
 adversarial attack as an ``imperceptible non-random perturbation''.
 Thus, we choose to work with a different distinction between CE and AE
-than \cite{freiesleben2022intriguing} that considers misclassification
+than \citet{freiesleben2022intriguing} that considers misclassification
 as the key distinguishing feature of AE. One of the key observations of
@@
 descent in the feature space for optimization
-\cite{altmeyer2023explaining}. In this context, domain constraints can
+\citep{altmeyer2023explaining}. In this context, domain constraints can
 be imposed by simply projecting counterfactuals back to the specified
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
+\pandocbounded{\includegraphics[keepaspectratio]{body_files/mediabag/../../paper/figures/poc.pdf}}
 
@@
 samples in the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We
-rely on the distance-based metric used in \cite{altmeyer2024faithful},
+rely on the distance-based metric used in \citet{altmeyer2024faithful},
 
@@
 counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
-squared population maximum mean discrepancy \cite{gretton2012kernel}.
+squared population maximum mean discrepancy \citep{gretton2012kernel}.
 The metric in Equation~\ref{eq-impl-div} is equal to zero iff the two
@@
 metrics, including robust accuracy estimated on adversarially perturbed
-data using FGSM \cite{goodfellow2014explaining}.
+data using FGSM \citep{goodfellow2014explaining}.
 
 We run the experiments with three gradient-based generators:
-\emph{Generic} of \cite{wachter2017counterfactual} as a simple baseline
-approach, \emph{REVISE} \cite{joshi2019realistic} that aims to generate
+\emph{Generic} of \citet{wachter2017counterfactual} as a simple baseline
+approach, \emph{REVISE} \citep{joshi2019realistic} that aims to generate
 plausible counterfactuals using a surrogate Variational Autoencoder
-(VAE), and \emph{ECCo}---the generator of \cite{altmeyer2024faithful}
+(VAE), and \emph{ECCo}---the generator of \citet{altmeyer2024faithful}
 but without the conformal prediction component---as a method that
@@
 moons (\emph{Moon}). These datasets are generated using the library of
-\cite{altmeyer2023explaining} and we present them in the supplementary
+\citep{altmeyer2023explaining} and we present them in the supplementary
 appendix. Next, we have four real-world binary tabular datasets from the
 domain of economics: \emph{Adult} (a.k.a. Census data) of
-\cite{becker1996adult2}, California housing (\emph{CH}) of
-\cite{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
-\cite{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) of
-\cite{kaggle2011give2}. Finally, for the convenience of illustration, we
-use of the 10-class \emph{MNIST} vision dataset \cite{lecun1998mnist}.
+\citep{becker1996adult2}, California housing (\emph{CH}) of
+\citep{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
+\citep{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) of
+\citep{kaggle2011give2}. Finally, for the convenience of illustration,
+we use of the 10-class \emph{MNIST} vision dataset
+\citep{lecun1998mnist}.
 
@@
 the related literature
-\cite{goodfellow2014explaining,ross2021learning,teney2020learning}.
+\citep{goodfellow2014explaining, ross2021learning, teney2020learning}.
 
@@
 it uses a surrogate VAE to generate counterfactuals which impedes
-faithfulness \cite{altmeyer2024faithful}. Increasing \(T\), the maximum
+faithfulness \citep{altmeyer2024faithful}. Increasing \(T\), the maximum
 number of steps, generally yields better outcomes because more CEs can
@@
 protections. Delineating actionability is a major open challenge in the
-AR literature (see, e.g., \cite{venkatasubramanian2020philosophical})
+AR literature (see, e.g., \citep{venkatasubramanian2020philosophical})
 impacting the capacity of CT to fulfill its intended goal.\\
@@
 result in an unfairly assigned burden of recourse
-\cite{sharma2020certifai}, threatening the equality of opportunity in
-the system \cite{bell2024fairness}. Nonetheless, these phenomena are not
-specific to CT.\\
+\citep{sharma2020certifai}, threatening the equality of opportunity in
+the system \citep{bell2024fairness}. Nonetheless, these phenomena are
+not specific to CT.\\
 \hfill \break \indent We also highlight several important directions for
@@
 CEs and AR. Other settings have attracted some interest (e.g.,
-regression in \cite{spooner2021counterfactual}), but there is little
+regression in \citep{spooner2021counterfactual}), but there is little
 consensus how to robustly extend the notion of CEs.\\
 \indent Secondly, our approach is susceptible to training instabilities.
-This problem has been recognized for JEMs \cite{grathwohl2020your} and
+This problem has been recognized for JEMs \citep{grathwohl2020your} and
 even though we depart from the SGLD-based sampling, we still encounter
@@
 Bayesian or gradient-based optimization applicable (see, e.g.,
-\cite{bischl2023hyperparameter}). \hfill \break
+\citep{bischl2023hyperparameter}). \hfill \break
 
@@
 time.
-
-\bibliographystyle{splncs04}
-\bibliography{bibliography}
-
 \end{document}

### 2025-06-25 paper/quarto_aaai/paper.tex
frustrating

%File: anonymous-submission-latex-2026.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage[submission]{aaai2026}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
%
% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
\usepackage{algorithm}
\usepackage{algorithmic}

%
% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
\usepackage{newfloat}
\usepackage{listings}
\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
\lstset{%
	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
	aboveskip=0pt,belowskip=0pt,%
	showstringspaces=false,tabsize=2,breaklines=true}
\floatstyle{ruled}
\newfloat{listing}{tb}{lst}{}
\floatname{listing}{Listing}
%
% Keep the \pdfinfo as shown here. There's no need
% for you to add the /Title and /Author tags.
\pdfinfo{
/TemplateVersion (2026.1)
}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\providecommand\phantomsection{}
\providecommand\pandocbounded{}

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

\setcounter{secnumdepth}{0} %May be changed to 1 or 2 if section numbers are desired.

% The file aaai2026.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%

% Title

% Your title must be in mixed case, not sentence case.
% That means all verbs (including short verbs like be, is, using,and go),
% nouns, adverbs, adjectives should be capitalized, including both words in hyphenated terms, while
% articles, conjunctions, and prepositions are lower case unless they
% directly follow a colon or long dash
% \title{AAAI Press Anonymous Submission\\Instructions for Authors Using \LaTeX{}}
\author{
    %Authors
    % All authors must be in the same font size and format.
    Written by AAAI Press Staff\textsuperscript{\rm 1}\thanks{With help from the AAAI Publications Committee.}\\
    AAAI Style Contributions by Pater Patel Schneider,
    Sunil Issar,\\
    J. Scott Penberthy,
    George Ferguson,
    Hans Guesgen,
    Francisco Cruz\equalcontrib,
    Marc Pujol-Gonzalez\equalcontrib
}
\affiliations{
    %Afiliations
    \textsuperscript{\rm 1}Association for the Advancement of Artificial Intelligence\\
    % If you have multiple authors and multiple affiliations
    % use superscripts in text and roman font to identify them.
    % For example,

    % Sunil Issar\textsuperscript{\rm 2},
    % J. Scott Penberthy\textsuperscript{\rm 3},
    % George Ferguson\textsuperscript{\rm 4},
    % Hans Guesgen\textsuperscript{\rm 5}
    % Note that the comma should be placed after the superscript

    1101 Pennsylvania Ave, NW Suite 300\\
    Washington, DC 20004 USA\\
    % email address must be in roman text type, not monospace or sans serif
    proceedings-questions@aaai.org
%
% See more examples next
}

%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
\iffalse
\title{My Publication Title --- Single Author}
\author {
    Author Name
}
\affiliations{
    Affiliation\\
    Affiliation Line 2\\
    name@example.com
}
\fi

\iffalse
%Example, Multiple Authors, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \title{My Publication Title --- Multiple Authors}
\author {
    % Authors
    First Author Name\textsuperscript{\rm 1},
    Second Author Name\textsuperscript{\rm 2},
    Third Author Name\textsuperscript{\rm 1}
}
\affiliations {
    % Affiliations
    \textsuperscript{\rm 1}Affiliation 1\\
    \textsuperscript{\rm 2}Affiliation 2\\
    firstAuthor@affiliation1.com, secondAuthor@affilation2.com, thirdAuthor@affiliation1.com
}
\fi

\input{body.tex}

\bibliography{bibliography}

### 2025-06-25 paper/quarto_aaai/paper.tex
buff

@@
+\makeatletter
+\@ifpackageloaded{hyperref}{}{%
+  \newcommand{\texorpdfstring}[2]{#1}%
+}
+\newcommand{\tightlist}
+\makeatother
+\providecommand\phantomsection{}
+
 %File: anonymous-submission-latex-2026.tex
@@
 }
-
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{amsfonts}
-\theoremstyle{definition}
-\newtheorem{definition}{Definition}[section]
-\theoremstyle{plain}
-\newtheorem{proposition}{Proposition}[section]
-\providecommand\phantomsection{}
-\providecommand\pandocbounded{}
 
@@
 
-\setcounter{secnumdepth}{0} %May be changed to 1 or 2 if section numbers are desired.
+\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.
 
@@
 
-\input{body.tex}
+% REMOVE THIS: bibentry
+% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
+\usepackage{bibentry}
+% END REMOVE bibentry
+\usepackage{amsmath}
 
-\bibliography{bibliography}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
+\usepackage{graphicx}
+\makeatletter
+\newsavebox\pandoc@box
+\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
+  \sbox\pandoc@box{#1}%
+  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
+  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
+  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
+  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
+  \else\usebox{\pandoc@box}%
+  \fi%
+}
+% Set default figure placement to htbp
+\def\fps@figure{htbp}
+\makeatother
+
+\title{Counterfactual Training: Teaching Models Plausible and Actionable
+Explanations}
+\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
+Deursen \and Cynthia C. S. Liem}
+\date{}
+\begin{document}
+\maketitle
+\begin{abstract}
+We propose a novel training regime termed counterfactual training that
+leverages counterfactual explanations to increase the explanatory
+capacity of models. Counterfactual explanations have emerged as a
+popular post-hoc explanation method for opaque machine learning models:
+they inform how factual inputs would need to change in order for a model
+to produce some desired output. To be useful in real-word
+decision-making systems, counterfactuals should be (1) plausible with
+respect to the underlying data and (2) actionable with respect to the
+user-defined mutability constraints. Much existing research has
+therefore focused on developing post-hoc methods to generate
+counterfactuals that meet these desiderata. In this work, we instead
+hold models directly accountable for the desired end goal:
+counterfactual training employs counterfactuals ad-hoc during the
+training phase to minimize the divergence between learned
+representations and plausible, actionable explanations. We demonstrate
+empirically and theoretically that our proposed method facilitates
+training models that deliver inherently desirable explanations while
+promoting robustness and preserving high predictive performance.
+\end{abstract}
+\section{Introduction}\label{introduction}
+
+Today's prominence of artificial intelligence (AI) has largely been
+driven by \textbf{representation learning}: instead of relying on
+features and rules that are carefully hand-crafted by humans, modern
+machine learning (ML) models are tasked with learning representations
+directly from data, guided by narrow objectives such as predictive
+accuracy \citep{goodfellow2016deep}. Modern advances in computing have
+made it possible to provide such models with ever-growing degrees of
+freedom to achieve that task, which frequently allows them to outperform
+traditionally more parsimonious models. Unfortunately, in doing so,
+models learn increasingly complex and highly sensitive representations
+that humans can no longer easily interpret.
+
+The trend towards complexity for the sake of performance has come under
+serious scrutiny in recent years. At the very cusp of the deep learning
+(DL) revolution, \citet{szegedy2013intriguing} showed that artificial
+neural networks (ANN) are sensitive to adversarial examples (AEs):
+perturbed versions of data instances that yield vastly different model
+predictions despite being ``imperceptible'' in that they are
+semantically indifferent from their factual counterparts. Even though
+some partially effective mitigation strategies have been proposed---most
+notably \textbf{adversarial training}
+\citep{goodfellow2014explaining}---truly robust deep learning remains
+unattainable even for models that are considered ``shallow'' by today's
+standards \citep{kolter2023keynote}.
+
+Part of the problem is that the high degrees of freedom provide room for
+many solutions that are locally optimal with respect to narrow
+objectives \citep{wilson2020case}.\footnote{We follow the standard ML
+  convention, where ``degrees of freedom'' refer to the number of
+  parameters estimated from data.} Indeed, recent work on the so-called
+``lottery ticket hypothesis'' suggests that modern neural networks can
+be pruned by up to 90\% while preserving their predictive performance
+\citep{frankle2018lottery}. Similarly, \citet{zhang2021understanding}
+showed that state-of-the-art neural networks are expressive enough to
+fit randomly labeled data. Thus, looking at the predictive performance
+alone, the solutions may seem to provide compelling explanations for the
+data, when in fact they are based on purely associative, semantically
+meaningless patterns. This poses two challenges. Firstly, there is no
+dependable way to verify if representations correspond to meaningful,
+plausible explanations. Secondly, even if we could resolve the first
+challenge, it remains undecided how to ensure that models can
+\emph{only} learn valuable explanations.
+
+The first challenge has attracted an abundance of research on
+\textbf{explainable AI} (XAI), a paradigm that focuses on the
+development of tools to derive (post-hoc) explanations from complex
+model representations. Such explanations should mitigate a scenario in
+which practitioners deploy opaque models and blindly rely on their
+predictions. On countless occasions, this has happened in practice and
+caused real harms to people who were adversely and unfairly affected by
+automated decision-making (ADM) systems involving opaque models (see,
+e.g., \citep{oneil2016weapons}). Effective XAI tools can aid us in
+monitoring models and providing recourse to individuals to turn negative
+outcomes (e.g., ``loan application rejected'') into positive ones (e.g.,
+``application accepted''). In line with this, our work builds upon
+\textbf{counterfactual explanations} (CE) proposed by
+\citet{wachter2017counterfactual} as an effective approach to achieve
+this goal. CEs prescribe minimal changes for factual inputs that, if
+implemented, would prompt some fitted model to produce a desired output.
+
+To our surprise, the second challenge has not yet attracted major
+research interest. Specifically, there has been no concerted effort
+towards improving the ``explanatory capacity'' of models, i.e., the
+degree to which learned representations correspond to explanations that
+are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
+Def. \ref{def-explainability}). Instead, the choice has generally been
+to improve the ability of XAI tools to identify the subset of
+explanations that are both plausible and valid for any given model,
+independent of whether the learned representations are in fact
+compatible with plausible explanations \citep{altmeyer2024faithful}.
+Fortunately, recent findings indicate that improved explanatory capacity
+can arise as a consequence of regularization techniques aimed at other
+training objectives such as generative capacity, generalization, or
+robustness
+\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
+As further discussed in Section~\ref{sec-lit}, our work consolidates
+these findings within a single objective.
+
+\textbf{Specifically, we introduce Counterfactual Training (CT)}: a
+novel training regime explicitly geared towards improving the
+explainability of models. In high-level terms, we define this concept as
+as the extent to which valid explanations derived for an opaque model
+are also deemed plausible with respect to the underlying data and the
+global actionability constraints. To the best of our knowledge, our
+framework represents the first attempt to address this challenge by
+employing counterfactual explanations already in the training phase.
+
+The remainder of this manuscript is structured as follows.
+Section~\ref{sec-lit} presents related work, focusing on the link
+between AEs and CEs. Then follow our two principal contributions. In
+Section~\ref{sec-method}, we introduce our methodological framework and
+show theoretically that it can be employed to enforce global
+actionability constraints. In Section~\ref{sec-experiments}, through
+extensive experiments, we demonstrate that CT substantially improves
+explainability and positively contributes to the robustness of trained
+models without sacrificing predictive performance. Finally, in
+Section~\ref{sec-discussion}, we discuss the challenges and conclude
+that CT is a promising approach towards making opaque models more
+trustworthy.
+
+\section{Related Literature}\label{sec-lit}
+
+To make the desiderata for our framework more concrete, we follow
+\citet{augustin2020adversarial} in tying the concept of explainability
+to the quality of CEs that can be generated for a given model. The
+authors show that CEs---understood as minimal input perturbations that
+yield some desired model prediction---tend to be more meaningful if the
+underlying model is more robust to adversarial examples. We can make
+intuitive sense of this finding when looking at adversarial training
+(AT) through the lens of representation learning with high degrees of
+freedom. As argued before, learned representations may be sensitive to
+producing implausible explanations and mispredicting for worst-case
+counterfactuals (i.e., AEs). Thus, by inducing models to ``unlearn''
+susceptiblity to such examples, adversarial training can effectively
+remove implausible explanations from the solution space.
+
+\subsection{Adversarial Examples are Counterfactual
+Explanations}\label{adversarial-examples-are-counterfactual-explanations}
+
+This interpretation of the link between explainability through
+counterfactuals on one side and robustness to adversarial examples on
+the other is backed by empirical evidence.
+\citet{sauer2021counterfactual} demonstrates that using counterfactual
+images during classifier training improves model robustness. Similarly,
+\citet{abbasnejad2020counterfactual} argues that counterfactuals
+represent potentially useful training data in machine learning,
+especially in supervised settings where inputs may be reasonably mapped
+to multiple outputs. They, too, demonstrate that augmenting the training
+data of image classifiers can improve generalization. Finally,
+\citet{teney2020learning} proposes an approach using counterfactuals in
+training that does not rely on data augmentation: they argue that
+counterfactual pairs typically already exist in training datasets.
+Specifically, their approach relies on identifying similar input samples
+with different annotations and ensuring that the gradient of the
+classifier aligns with the vector between such pairs of counterfactual
+inputs using the cosine distance as the loss function.
+
+In the natural language processing (NLP) domain, CEs have also been used
+to improve models through data augmentation. \citet{wu2021polyjuice2}
+proposes \emph{Polyjuice}, a general-purpose counterfactual generator
+for language models. The authors empirically demonstrate that the
+augmentation of training data with their method improves robustness in a
+number of NLP tasks. \citet{balashankar2023improving} similarly uses
+\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
+and show that classifier robustness improves by up to 20\%. Finally,
+\citet{luu2023counterfactual} introduces Counterfactual Adversarial
+Training (CAT) that also aims to improve generalization and robustness
+of language models through a three-step procedure: the authors identify
+training samples that are subject to high predictive uncertainty,
+generate CEs for them, and fine-tune the language model on a dataset
+augmented with the CEs.
+
+There have also been several attempts at formalizing the relationship
+between counterfactual explanations and adversarial examples. Pointing
+to clear similarities in how CEs and AEs are generated,
+\citet{freiesleben2022intriguing} makes the case for jointly studying
+the opaqueness and robustness problems in representation learning.
+Formally, AEs can be seen as the subset of CEs for which
+misclassification is achieved \citep{freiesleben2022intriguing}.
+Similarly, \citet{pawelczyk2022exploring} shows that CEs and AEs are
+equivalent under certain conditions.
+
+Two recent works are closely related to ours in that they use
+counterfactuals during training with the explicit goal of affecting
+certain properties of the post-hoc counterfactual explanations. Firstly,
+\citet{ross2021learning} proposes a way to train models that guarantee
+individual recourse to some positive target class with high probability.
+Their approach builds on adversarial training by explicitly inducing
+susceptibility to targeted adversarial examples for the positive class.
+Additionally, the proposed method allows for imposing a set of
+actionability constraints ex-ante. For example, users can specify that
+certain features are immutable. Secondly, \citet{guo2023counternet} is
+the first to propose an end-to-end training pipeline that includes CEs
+as part of the training procedure. In particular, they propose a
+specific network architecture that includes a predictor and CE generator
+network, where the parameters of the CE generator network are learnable.
+Counterfactuals are generated during each training iteration and fed
+back to the predictor network. In contrast to \citet{guo2023counternet},
+we impose no restrictions on the ANN architecture at all.
+
+\subsection{Aligning Representations with Plausible
+Explanations}\label{aligning-representations-with-plausible-explanations}
+
+Improving the adversarial robustness of models is not the only path
+towards aligning representations with plausible explanations. In a work
+closely related to this one, \citet{altmeyer2024faithful} shows that
+explainability can be improved through model averaging and refined model
+objectives. The authors propose a way to generate counterfactuals that
+are maximally faithful to the model in that they are consistent with
+what the model has learned about the underlying data. Formally, they
+rely on tools from energy-based modelling \citep{teh2003energy} to
+minimize the divergence between the distribution of counterfactuals and
+the conditional posterior over inputs learned by the model. Their
+proposed counterfactual explainer, \emph{ECCCo}, yields plausible
+explanations if and only if the underlying model has learned
+representations that align with them. The authors find that both deep
+ensembles \citep{lakshminarayanan2016simple} and joint energy-based
+models (JEMs) \citep{grathwohl2020your} tend to do well in this regard.
+
+Once again it helps to look at these findings through the lens of
+representation learning with high degrees of freedom. Deep ensembles are
+approximate Bayesian model averages, which are most called for when
+models are underspecified by the available data \citep{wilson2020case}.
+Averaging across solutions mitigates the aforementioned risk of relying
+on a single locally optimal representations that corresponds to
+semantically meaningless explanations for the data. Previous work of
+\citet{schut2021generating} similarly found that generating plausible
+(``interpretable'') CEs is almost trivial for deep ensembles that have
+also undergone adversarial training. The case for JEMs is even clearer:
+they involve a hybrid objective that induces both high predictive
+performance and generative capacity \citep{grathwohl2020your}. This is
+closely related to the idea of aligning models with plausible
+explanations and has inspired our CT objective.
+
+\section{Counterfactual Training}\label{sec-method}
+
+In this section we propose our novel counterfactual training objective.
+In CT, we combine ideas from adversarial training, counterfactual
+explanations, and energy-based modelling with the explicit goal of
+aligning representations with plausible explanations that comply with
+user-defined actionability constraints.
+
+In the context of counterfactual explanations, plausibility has broadly
+been defined as the degree to which counterfactuals comply with the
+underlying data-generating process
+\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
+Plausibility is a necessary but insufficient condition for using CEs to
+provide algorithmic recourse (AR) to individuals (negatively) affected
+by opaque models. To be actionable, AR recommendations must also be
+attainable. A plausible CE for a rejected 20-year-old loan applicant,
+for example, might reveal that their application would have been
+accepted, if only they were 20 years older. Ignoring all other features,
+this would comply with the definition of plausibility if 40-year-old
+individuals were in fact more credit-worthy on average than young
+adults. But of course this CE does not qualify for providing actionable
+recourse to the applicant since \emph{age} is not a (directly) mutable
+feature. Counterfactual training aims to improve model explainability by
+aligning models with counterfactuals that meet both desiderata:
+plausibility and actionability. Formally, we define explainability as
+follows:
+
+\begin{definition}[Model Explainability]
+\label{def-explainability}
+Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$.
+Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.
+
+We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. We define these properties as follows:
+
+\begin{enumerate}
+    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
+    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
+    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
+\end{enumerate}
+where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
+\end{definition}
+
+\noindent The characterization of faithfulness and plausibility in Def.
+\ref{def-explainability} is the same as in \citet{altmeyer2024faithful},
+with adapted notation. Intuitively, plausible counterfactuals are
+consistent with the data and faithful counterfactuals are consistent
+with what the model has learned about input data. Actionability
+constraints in Def. \ref{def-explainability} vary and depend on the
+context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
+focus on domain and mutability constraints for individual features
+\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
+for reasons discussed in Section~\ref{sec-discussion}.
+
+\subsection{Our Proposed Objective}\label{our-proposed-objective}
+
+Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
+explanation generated through gradient descent over \(T\) iterations as
+initially proposed by \citet{wachter2017counterfactual}. For our
+purposes, we let \(T\) vary and consider the counterfactual search as
+converged as soon as the predicted probability for the target class has
+reached a pre-determined threshold, \(\tau\):
+\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
+where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
+  background information on gradient-based counterfactual search and
+  convergence see supplementary appendix.}
+
+To train models with high explainability as defined in Def.
+\ref{def-explainability}, we propose to leverage counterfactuals in the
+following objective:
+
+\begin{equation}\phantomsection\label{eq-obj}{
+\begin{split}
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
+\end{split}
+}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
+that induces discriminative performance (e.g., cross-entropy). The
+second and third terms are explained in detail below. For now, they can
+be summarized as inducing explainability directly and indirectly by
+penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
+between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
+samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
+target class \(y^+\), and, (2) the adversarial loss,
+\(\text{advloss}(.)\), with respect to nascent counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
+denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
+of the energy terms involved in \(\text{div}(\cdot)\)
+\citep{du2019implicit}. The trade-off between the components can
+governed through penalties \(\lambda_{\text{div}}\),
+\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).
+
+\subsection{Directly Inducing Explainability with Contrastive
+Divergence}\label{directly-inducing-explainability-with-contrastive-divergence}
+
+As observed by \citet{grathwohl2020your}, any classifier can be
+re-interpreted as a joint energy-based model (JEM) that learns to
+discriminate output classes conditional on the observed (training)
+samples from \(p(\mathbf{x})\) and the generated samples from
+\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
+perform well at both tasks by directly maximizing the joint
+log-likelihood factorized as
+\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
+The first term can be optimized using conventional cross-entropy as in
+Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\),
+\citet{grathwohl2020your} minimizes the contrastive divergence between
+the observed samples from \(p(\mathbf{x})\) and generated samples from
+\(p_\theta(\mathbf{x})\).
+
+A key empirical finding in \citet{altmeyer2024faithful} was that JEMs
+tend to do well with respect to the plausibility objective in Def.
+\ref{def-explainability}. This follows directly if we consider samples
+drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
+objective effectively minimizes the divergence between the conditional
+posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
+\citet{grathwohl2020your} relies on Stochastic Gradient Langevin
+Dynamics (SGLD) using an uninformative prior for initialization but we
+depart from their methodology. Instead of SGLD, we propose to use
+counterfactual explainers to generate counterfactuals of observed
+training samples. Specifically, we have:
+
+\begin{equation}\phantomsection\label{eq-div}{
+\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
+}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
+function defined as
+\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
+with \(y^+\) denoting the index of the randomly drawn target class,
+\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
+\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
+sampled factual from a non-target class generated with a gradient-based
+CE generator for up to \(T\) iterations. Mature counterfactuals are ones
+that have either reached convergence wrt. the decision threshold
+\(\tau\) or exhausted \(T\).
+
+Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
+of observed training samples (positive samples) while increasing the
+energy of counterfactuals (negative samples) \citep{du2019implicit}. As
+the counterfactuals get more plausible (Def. \ref{def-explainability})
+during training, these opposing effects gradually balance each other out
+\citep{lippe2024uvadlc}.
+
+The departure from SGLD of \citep{grathwohl2020your} allows us to tap
+into the vast repertoire of explainers that have been proposed in the
+literature to meet different desiderata. For example, many methods
+facilitate the imposition of domain and mutability constraints. In
+principle, any existing approach for generating counterfactual
+explanations is viable, so long as it does not violate the faithfulness
+condition. Like JEMs \citep{murphy2022probabilistic}, CT can be
+considered a form of contrastive representation learning.
+
+\subsection{Indirectly Inducing Explainability with Adversarial
+Robustness}\label{indirectly-inducing-explainability-with-adversarial-robustness}
+
+Based on our analysis in Section~\ref{sec-lit}, counterfactuals
+\(\mathbf{x}^\prime\) can be repurposed as additional training samples
+\citep{balashankar2023improving, luu2023counterfactual} or adversarial
+examples \citep{freiesleben2022intriguing, pawelczyk2022exploring}. This
+leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
+Equation~\ref{eq-obj}. An intuitive functional form, but likely not the
+only sensible choice, is inspired by adversarial training:
+
+\begin{equation}\phantomsection\label{eq-adv}{
+\begin{aligned}
+\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
+t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
+\end{aligned}
+}\end{equation} Under this choice, we consider nascent counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
+perturbation to any single feature is at most \(\varepsilon\). This is
+closely aligned with \citet{szegedy2013intriguing} that defines an
+adversarial attack as an ``imperceptible non-random perturbation''.
+Thus, we choose to work with a different distinction between CE and AE
+than \citet{freiesleben2022intriguing} that considers misclassification
+as the key distinguishing feature of AE. One of the key observations of
+this work is that we can leverage CEs during training and get
+adversarial examples essentially for free, which can be used to reap the
+aforementioned benefits of adversarial training.
+
+\subsection{Encoding Actionability Constraints}\label{sec-constraints}
+
+Many existing counterfactual explainers support domain and mutability
+constraints out-of-the-box. In fact, both types of constraints can be
+implemented for any counterfactual explainer that relies on gradient
+descent in the feature space for optimization
+\citep{altmeyer2023explaining}. In this context, domain constraints can
+be imposed by simply projecting counterfactuals back to the specified
+domain, if the previous gradient step resulted in updated feature values
+that were out-of-domain. Mutability constraints can similarly be
+enforced by setting partial derivatives to zero to ensure that features
+are only perturbed in the allowed direction, if at all.
+
+Since such actionability constraints are binding at test time, we should
+also impose them when generating \(\mathbf{x}^\prime\) during each
+training iteration to inform model representations. Through their effect
+on \(\mathbf{x}^\prime\), both types of constraints influence model
+outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
+penalizing implausibility that arises due to mutability constraints. For
+any mutability-constrained feature \(d\) this can be achieved by
+enforcing \(\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0\) whenever
+perturbing \(\mathbf{x}^\prime[d]\) in the direction of
+\(\mathbf{x}^+[d]\) would violate mutability constraints. Specifically,
+we set \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) if:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Feature \(d\) is strictly immutable in practice.
+\item
+  We have \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but feature \(d\)
+  can only be decreased in practice.
+\item
+  We have \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but feature \(d\)
+  can only be increased in practice.
+\end{enumerate}
+
+\noindent From a Bayesian perspective, setting
+\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
+assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
+feature \(d\). Intuitively, we think of this simply in terms ignoring
+implausibility costs with respect to immutable features, which
+effectively forces the model to instead seek plausibility with respect
+to the remaining features. This in turn results in lower overall
+sensitivity to immutable features, which we demonstrate empirically for
+different classifiers in Section~\ref{sec-experiments}. Under certain
+conditions, this result holds theoretically:\footnote{For the proof, see
+  the supplementary appendix.}
+
+\begin{proposition}[Protecting Immutable Features]
+\label{prp-mtblty}
+Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
+If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
+\end{proposition}
+
+\noindent It is worth highlighting that
+\mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
+This raises a valid concern about the effect of protecting immutable
+features in the presence of proxies that remain unprotected. We address
+this in Section~\ref{sec-discussion}.
+
+\subsection{Example (Prediction of Consumer Credit
+Default)}\label{example-prediction-of-consumer-credit-default}
+
+Suppose we are interested in predicting the likelihood that loan
+applicants default on their credit. We have access to historical data on
+previous loan takers comprised of a binary outcome variable
+(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
+features: (1) the subjects' \emph{age}, which we define as immutable,
+and (2) the subjects' existing level of \emph{debt}, which we define as
+mutable.
+
+We have simulated this scenario using synthetic data with independent
+\emph{age} and \emph{debt} features, and Gaussian class-conditional
+densities in Figure~\ref{fig-poc}. The four panels show the outcomes for
+different training procedures using the same model architectures (a
+linear classifier). In panels (a) and (c) we have trained the models
+conventionally, while in panels (b) and (d) we used CT.
+
+In all cases, all counterfactuals (stars) are valid---they have crossed
+the decision boundary (green)---but their quality differs. In panel (a),
+they are not plausible: they do not comply with the distribution of the
+factuals in \(y^+\) to the point where they form a clearly
+distinguishable cluster. In panel (b), they are highly plausible,
+meeting the first objective of Def. \ref{def-explainability}. In panel
+(c), the CEs involve substantial reductions in \emph{debt} for younger
+applicants. By comparison, counterfactual paths are shorter on average
+in panel (d) where we have protected the immutable \emph{age} as
+described in Section~\ref{sec-constraints}. Due to the classifier's
+lower sensitivity to \emph{age}, recommendations with respect to
+\emph{debt} are much more homogenous and do not unfairly punish younger
+individuals. These counterfactuals are also plausible with respect to
+the mutable feature. Thus, we consider the model in panel (d) as the
+most explainable according to Def. \ref{def-explainability}.
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
+
+}
+
+\caption{\label{fig-poc}Illustration of how CT improves model
+explainability: (a) conventional training, all mutable; (b) CT, all
+mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
+immutable. The linear decision boundary is shown in green along with
+training data colored according to their ground-truth label: \(y^-=1\)
+in blue and \(y^+=2\) in orange. Stars indicate counterfactuals in the
+target class.}
+
+\end{figure}%
+
+\section{Experiments}\label{sec-experiments}
+
+In our experiments we seek to answer the following three research
+questions:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  To what extent does the counterfactual training objective as it is
+  defined in Equation~\ref{eq-obj} induce models to learn plausible
+  explanations?
+\item
+  To what extent does the CT objective produce more favorable
+  algorithmic recourse outcomes in the presence of actionability
+  constraints?
+\item
+  What are the effects of hyperparameter selection wrt. the CT
+  objective?
+\end{enumerate}
+
+\subsection{Experimental Setup}\label{experimental-setup}
+
+Our key outcome of interest is improvement in explainability (Def.
+\ref{def-explainability}). To this end, we focus primarily on the
+plausibility and cost of faithfully generated counterfactuals at test
+time. To measure the cost, we follow the standard convention of using
+distances (\(\ell_1\)-norm) between factuals and counterfactuals as a
+proxy. For plausibility, we assess how similar CEs are to the observed
+samples in the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We
+rely on the distance-based metric used in \citet{altmeyer2024faithful},
+
+\begin{equation}\phantomsection\label{eq-impl-dist}{
+\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
+}\end{equation} and introduce a novel divergence metric,
+
+\begin{equation}\phantomsection\label{eq-impl-div}{
+\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
+}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
+counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
+squared population maximum mean discrepancy \citep{gretton2012kernel}.
+The metric in Equation~\ref{eq-impl-div} is equal to zero iff the two
+distributions are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
+
+In addition to cost and plausibility, we compute other standard metrics
+to evaluate counterfactuals including validity and redundancy. Finally,
+we also assess the predictive performance of models using standard
+metrics, including robust accuracy estimated on adversarially perturbed
+data using FGSM \citep{goodfellow2014explaining}.
+
+We run the experiments with three gradient-based generators:
+\emph{Generic} of \citet{wachter2017counterfactual} as a simple baseline
+approach, \emph{REVISE} \citep{joshi2019realistic} that aims to generate
+plausible counterfactuals using a surrogate Variational Autoencoder
+(VAE), and \emph{ECCo}---the generator of \citet{altmeyer2024faithful}
+but without the conformal prediction component---as a method that
+directly targets both faithfulness and plausibility of the
+counterfactuals.
+
+We make use of nine classification datasets common in the CE/AR
+literature. Four of them are synthetic with two classes and different
+characteristics: linearly separable clusters (\emph{LS}), overlapping
+clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
+moons (\emph{Moon}). These datasets are generated using the library of
+\citep{altmeyer2023explaining} and we present them in the supplementary
+appendix. Next, we have four real-world binary tabular datasets from the
+domain of economics: \emph{Adult} (a.k.a. Census data) of
+\citep{becker1996adult2}, California housing (\emph{CH}) of
+\citep{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
+\citep{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) of
+\citep{kaggle2011give2}. Finally, for the convenience of illustration,
+we use of the 10-class \emph{MNIST} vision dataset
+\citep{lecun1998mnist}.
+
+To assess CT, we investigate the improvements in performance metrics
+when using it on top of a weak baseline (BL): a multilayer perceptron
+(\emph{MLP}). This is the best way to get a clear picture of the
+effectiveness of CT, and it is consistent with how assessment is done in
+the related literature
+\citep{goodfellow2014explaining, ross2021learning, teney2020learning}.
+
+\subsection{Experimental Results}\label{experimental-results}
+
+\subsubsection{Plausibility}\label{sec-plaus}
+
+Table~\ref{tbl-main} presents our main empirical findings. For all
+datasets except \emph{OL} and across all test settings, the average
+distance of CEs from observed samples in the target class is reduced,
+indicating improved plausibility. The magnitude of improvements varies.
+For the simple synthetic datasets, distance reductions range from around
+20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
+real-world tabular datasets, improvements tend to be smaller but still
+substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
+7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For the vision
+dataset (\emph{MNIST}), distances are reduced by up to 9\%. The results
+for our proposed divergence metric are qualitatively similar, but
+generally even more pronounced: for the \emph{Circ} dataset,
+implausibility is reduced by almost 94\% to virtually zero as we
+verified by the absolute outcome. Improvements for other datasets range
+from 28\% (\emph{Moon}) up to 78\% (\emph{GMSC}). For \emph{OL} the
+reduction is negative, consistent with the distance-based metric.
+\emph{MNIST} is the only dataset for which the distance and divergence
+metrics disagree. Upon visual inspection of the image counterfactuals we
+find that CT clearly improves plausibility (see supplementary appendix
+for images).
+
+\subsubsection{Predictive Performance}\label{sec-pred}
+
+Test accuracy for CT is virtually identical to the baseline for
+\emph{Adult}, \emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and
+even slightly improved for \emph{Cred}. Exceptions to this general
+pattern are \emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we
+observe a reduction in test accuracy of 2, 5, and 15 percentage points
+respectively. When looking at robust test accuracies (Acc.\(^*\)) for
+these datasets in particular, we find that CT strongly outperforms the
+baseline. In fact, we find that CT improves adversarial robustness on
+all datasets.
+
+\begin{table}
+
+\caption{\label{tbl-main}Key performance metrics across all datasets
+(column 1). \textbf{Plausibility}: Columns 2-6 show the percentage
+reduction in implausibility (\(\text{IP}\)) for varying degrees of the
+energy penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test
+time; column 7 shows the reduction in \(\text{IP}^*\) (\(\text{MMD}\)),
+aggregated across all test specifications. \textbf{Accuracy} (columns
+8-11): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
+and the baseline (BL). \textbf{Actionability} (column 12): average
+reduction in costs when imposing mutability constraints reported for the
+four datasets for which we could identify meaningful features to
+protect.}
+
+\centering{
+
+\input{tables/main.tex}
+
+}
+
+\end{table}%
+
+\subsubsection{Actionability}\label{sec-act}
+
+In Section~\ref{sec-method}, we show that our proposed way for encoding
+mutability constraints leads to lower classifier sensitivity wrt.
+immutable features for linear models, tilting the decision boundary in
+favour of mutable features instead. For binding constraints at test
+time, this leads to shorter counterfactual paths and hence smaller
+average costs (\(\ell_1\)-norm) to individuals. To extend this to the
+non-linear case, we test the effect of imposing mutability constraints
+empirically for our synthetic data using the same evaluation scheme as
+above. The final row in Table~\ref{tbl-main} reports the average
+reduction in costs for CT compared to the ``vanilla'' baseline, when
+imposing that either the first or the second feature is immutable. In
+all cases, costs are reduced substantially, indicating that classifiers
+trained with CT are indeed more sensitive to mutable features.
+
+\subsubsection{Impact of hyperparameter
+settings.}\label{sec-hyperparameters}
+
+We test the impact of three types of hyperparameters; our complete
+results are in the supplementary appendix.\\
+\indent We note that CT is highly sensitive to the choice of a CE
+generator and its hyperparameters but (a) there are manageable patterns
+and (b) we can typically identify settings that improve either
+plausibility or cost, and commonly both of them at the same time. For
+example, \emph{REVISE} tends to perform the worst, most likely because
+it uses a surrogate VAE to generate counterfactuals which impedes
+faithfulness \citep{altmeyer2024faithful}. Increasing \(T\), the maximum
+number of steps, generally yields better outcomes because more CEs can
+mature in each training epoch. The impact of \(\tau\), the required
+decision threshold is more difficult to predict. On ``harder'' datasets
+it may be difficult to satisfy high \(\tau\) for any given sample (i.e.,
+also factuals) and so increasing this threshold does not seem to
+correlate with better outcomes. In fact, the choice of \(\tau=0.5\)
+generally leads to optimal results because it is associated with high
+proportions of mature counterfactuals.\\
+\indent The strength of the energy regularization,
+\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
+performance in terms of decreased plausibility and increased costs if
+insufficiently high. The sensitivity with respect to
+\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
+evident. While high values of \(\lambda_{\text{reg}}\) may increase the
+variability in outcomes when combined with high values of
+\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
+very pronounced.\\
+\indent The effectiveness and stability of CT is positively associated
+with the number of counterfactuals generated during each training epoch.
+We also confirm that a higher number of training epochs is beneficial.
+Interestingly, we observed desired improvements when CT was combined
+with conventional training and applied only for the final 50\% of epochs
+of the complete training process. Put differently, CT can improve the
+explainability of models in a fine-tuning manner.
+
+\section{Conclusions}\label{sec-discussion}
+
+As our results indicate, counterfactual training produces models that
+are more explainable. Nonetheless, it brings about three important
+limitations.\\
+\indent \emph{CT increases the training time of models.} CT can be more
+time-consuming than conventional training regimes. While higher numbers
+of CEs per iteration positively impact the quality of solutions, they
+also increase the amount of computations. Relatively small grids with
+270 settings can take almost four hours for more demanding datasets on a
+high-performance computing cluster with 34 2GB CPUs.\footnote{See
+  supplementary appendix for computational details.} Three factors
+attenuate this effect: (1) CT amortizes the cost of CEs for the training
+samples; (2) it can retain its value when used as a ``fine-tuning''
+technique for conventionally-trained models; and (3) it yields itself to
+parallel execution, which we have leveraged for our own experiments.\\
+\indent \emph{Immutable features may have proxies.} We propose an
+approach to protect immutable features and thus increase the
+actionability of the generated CEs. However, it requires that model
+owners define the mutability constraints for (all) features considered
+by the model. Even if all immutable features are protected, there may
+exist proxies that are mutable (and hence should not be protected) but
+preserve enough information about the principals to hinder the
+protections. Delineating actionability is a major open challenge in the
+AR literature (see, e.g., \citep{venkatasubramanian2020philosophical})
+impacting the capacity of CT to fulfill its intended goal.\\
+\indent \emph{Interventions on features may impact fairness.} We provide
+a tool that allows practitioners to modify the sensitivity of a model
+with respect to certain features, which may have implication for the
+fair and equitable treatment of decision subjects. Model owners could
+misuse this solution by enforcing explanations based on features that
+are more difficult to modify by some (group of) individuals. For
+example, consider the \emph{Adult} dataset used in our experiments,
+where \emph{workclass} or \emph{education} may be more difficult to
+change for underpriviledged groups. When applied irresponsibly, CT could
+result in an unfairly assigned burden of recourse
+\citep{sharma2020certifai}, threatening the equality of opportunity in
+the system \citep{bell2024fairness}. Nonetheless, these phenomena are
+not specific to CT.\\
+\hfill \break \indent We also highlight several important directions for
+future research. Firstly, it is an interesting challenge to extend CT
+beyond classification settings. Our formulation relies on the
+distinction between non-target class(es) \(y^{-}\) and target class(es)
+\(y^{+}\) to generate counterfactuals through Equation~\ref{eq-obj}.
+While \(y^{-}\) and \(y^{+}\) can be arbitrarily defined, CT requires
+the output space \(\mathcal{Y}\) to be discrete. Thus, it does not apply
+to ML tasks where the change in outcome cannot be readily quantified.
+Focus on classification models is a common restriction in research on
+CEs and AR. Other settings have attracted some interest (e.g.,
+regression in \citep{spooner2021counterfactual}), but there is little
+consensus how to robustly extend the notion of CEs.\\
+\indent Secondly, our approach is susceptible to training instabilities.
+This problem has been recognized for JEMs \citep{grathwohl2020your} and
+even though we depart from the SGLD-based sampling, we still encounter
+considerable variability in the outcomes. CT is exposed to two potential
+sources of instabilities: (1) the energy-based contrastive divergence
+term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
+explainers. We find several promising ways to mitigate this problem:
+regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
+many counterfactuals during each epoch, and including only mature
+counterfactuals for contrastive divergence.\\
+\indent Finally, we believe that it is possible to substantially improve
+hyperparameter selection procedures. Our method benefits from the tuning
+of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
+In this work, we have relied exclusively on grid search for this task.
+Future work on CT could benefit from investigating more sophisticated
+approaches. Notably, CT is iterative which makes methods such as
+Bayesian or gradient-based optimization applicable (see, e.g.,
+\citep{bischl2023hyperparameter}). \hfill \break
+
+To conclude, state-of-the-art machine learning models are prone to
+learning complex representations that cannot be interpreted by humans
+and existing post-hoc explainability approaches cannot guarantee that
+the explanations agree with the model's learned representation of data.
+As a step towards addressing this challenge, we introduced
+counterfactual training, a novel training regime that incentivizes
+highly-explainable models. Our approach leads to explanations that are
+both plausible---compliant with the underlying data-generating
+process---and actionable---compliant with user-specified mutability
+constraints---and thus meaningful to their recipients. Through extensive
+experiments we demonstrate that CT satisfies its objective while
+promoting robustness and preserving the predictive performance of the
+models. It can also be used to fine-tune conventionally-trained models
+and achieve similar gains. Finally, this work showcases that it is
+practical to improve models \emph{and} their explanations at the same
+time.
+\end{document}

### 2025-06-26 paper/quarto_aaai/paper.tex
template seems to be working
can't get full width table to work

@@
 % END REMOVE bibentry
+\usepackage{amsfonts}
 \usepackage{amsmath}
+\usepackage{amsthm}
+\newtheorem{proposition}{Proposition}[section]
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
 
@@
 \begin{split}
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
+\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
 \end{split}
@@
 
-\begin{figure}
+\begin{figure*}
 
@@
 
-\end{figure}%
+\end{figure*}%
 
@@
 time.
+
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
+\renewcommand{\bibsection}{}
+\bibliography{../quarto_ecml/bibliography.bib}
 \end{document}

### 2025-06-26 paper/experiments/CTExperiments/src/tables.jl

@@ -12,6 +12,7 @@ function tabulate_results(
     save_name::Union{String,Nothing}=nothing,
     formatters=PrettyTables.ft_round(2),
     alignment=:c,
+    wrap_table_environment="table*",
     kwrgs...,
 )
     df = inputs[1]
@@ -24,6 +25,7 @@ function tabulate_results(
                 tf=tf,
                 formatters=formatters,
                 wrap_table=wrap_table,
+                wrap_table_environment=wrap_table_environment,
                 table_type=table_type,
                 longtable_footer=longtable_footer,
                 alignment,
@@ -39,6 +41,7 @@ function tabulate_results(
                     tf=tf,
                     formatters=formatters,
                     wrap_table=wrap_table,
+                    wrap_table_environment=wrap_table_environment,
                     table_type=table_type,
                     longtable_footer=longtable_footer,
                     alignment,

### 2025-06-26 paper/preprint/paper.tex
postponing appendix

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
 ]{article}
-
+\usepackage{xcolor}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{5}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{5}
 % Make \paragraph and \subparagraph free-standing
@@
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatother
+
 % definitions for citeproc citations
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 \newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
-
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 
-\begin{table}
+\begin{table*}
 
@@
 
-\end{table}%
+\end{table*}%
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-default}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-default}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
+\begin{table*}
   \begin{tabular}{cccccccc}
@@
   \end{tabular}
+\end{table*}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
 
@@
 
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/cali/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-cali}Average outcomes for the
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: California Housing.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/circles/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-circles}Average outcomes for the
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
 \includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
@@
 
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/moons/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
+
+}
+
+\caption{\label{fig-tune_lr-plaus-moons}Average outcomes for the
+plausibility measure across key hyperparameters. This shows the \%
+change from the baseline model for the distance-based implausibility
+metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
+across evaluation runs and test settings (varying parameters for
+\emph{ECCo}). Data: Moons.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
 \includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
@@
 
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/cali/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
+
+}
+
+\caption{\label{fig-tune_lr-mat-cali}Proportion of mature
+counterfactuals in each epoch. Data: California Housing.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/circles/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
+
+}
+
+\caption{\label{fig-tune_lr-mat-circles}Proportion of mature
+counterfactuals in each epoch. Data: Circles.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
 \includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
@@
 counterfactuals in each epoch. Data: MNIST.}
+
+\end{figure}%
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/moons/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
+
+}
+
+\caption{\label{fig-tune_lr-mat-moons}Proportion of mature
+counterfactuals in each epoch. Data: Moons.}
 

### 2025-06-26 paper/sections/method.qmd

@@
 The linear decision boundary is shown in green along with training data colored according to their ground-truth label: $y^-=1$ in blue and $y^+=2$ in orange.
-Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc}
+Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

### 2025-06-27 paper/quarto_aaai/paper.tex
Adjust introduction

@@
 driven by \textbf{representation learning}: instead of relying on
-features and rules that are carefully hand-crafted by humans, modern
-machine learning (ML) models are tasked with learning representations
-directly from data, guided by narrow objectives such as predictive
-accuracy \citep{goodfellow2016deep}. Modern advances in computing have
-made it possible to provide such models with ever-growing degrees of
-freedom to achieve that task, which frequently allows them to outperform
-traditionally more parsimonious models. Unfortunately, in doing so,
-models learn increasingly complex and highly sensitive representations
-that humans can no longer easily interpret.
+features and rules hand-crafted by humans, modern machine learning (ML)
+models are tasked with learning representations directly from the data,
+guided by narrow objectives such as predictive accuracy
+\citep{goodfellow2016deep}. Advances in computing have made it possible
+to provide these models with ever-growing degrees of freedom to achieve
+this task, which often allows them to outperform traditionally
+parsimonious models. Unfortunately, in doing so, models learn
+increasingly complex, sensitive representations that humans can no
+longer easily interpret.
 
 The trend towards complexity for the sake of performance has come under
-serious scrutiny in recent years. At the very cusp of the deep learning
-(DL) revolution, \citet{szegedy2013intriguing} showed that artificial
-neural networks (ANN) are sensitive to adversarial examples (AEs):
-perturbed versions of data instances that yield vastly different model
-predictions despite being ``imperceptible'' in that they are
-semantically indifferent from their factual counterparts. Even though
-some partially effective mitigation strategies have been proposed---most
-notably \textbf{adversarial training}
-\citep{goodfellow2014explaining}---truly robust deep learning remains
-unattainable even for models that are considered ``shallow'' by today's
-standards \citep{kolter2023keynote}.
+scrutiny in recent years. At the very cusp of the deep learning (DL)
+revolution, \citet{szegedy2013intriguing} showed that artificial neural
+networks (ANN) are susceptible to adversarial examples (AEs): perturbed
+versions of data instances that yield vastly different model predictions
+despite being semantically indistinguishable from their factual
+counterparts. Some partial mitigation strategies have been
+proposed---most notably \textbf{adversarial training}
+\citep{goodfellow2014explaining}---but truly robust deep learning
+remains unattainable even for models that are considered ``shallow'' by
+today's standards \citep{kolter2023keynote}.
 
@@
   convention, where ``degrees of freedom'' refer to the number of
-  parameters estimated from data.} Indeed, recent work on the so-called
+  parameters estimated from data.} As one example, research on the
 ``lottery ticket hypothesis'' suggests that modern neural networks can
-be pruned by up to 90\% while preserving their predictive performance
-\citep{frankle2018lottery}. Similarly, \citet{zhang2021understanding}
-showed that state-of-the-art neural networks are expressive enough to
-fit randomly labeled data. Thus, looking at the predictive performance
-alone, the solutions may seem to provide compelling explanations for the
-data, when in fact they are based on purely associative, semantically
-meaningless patterns. This poses two challenges. Firstly, there is no
-dependable way to verify if representations correspond to meaningful,
-plausible explanations. Secondly, even if we could resolve the first
-challenge, it remains undecided how to ensure that models can
-\emph{only} learn valuable explanations.
-
-The first challenge has attracted an abundance of research on
+be pruned by up to 90\% without losing predictive performance
+\citep{frankle2018lottery}. Thus, looking at the predictive performance
+alone, found solutions may seem to provide compelling explanations for
+the data, when in fact they are based on purely associative and
+semantically meaningless patterns. This poses two related challenges.
+Firstly, there is no dependable way to verify if learned representations
+correspond to meaningful, plausible explanations. Secondly, even if we
+resolve this challenge, it remains undecided how to ensure that machine
+learning models can \emph{only} learn valuable explanations.
+
+The first challenge has attracted an abundance of work on
 \textbf{explainable AI} (XAI), a paradigm that focuses on the
 development of tools to derive (post-hoc) explanations from complex
-model representations. Such explanations should mitigate a scenario in
-which practitioners deploy opaque models and blindly rely on their
-predictions. On countless occasions, this has happened in practice and
-caused real harms to people who were adversely and unfairly affected by
-automated decision-making (ADM) systems involving opaque models (see,
-e.g., \citep{oneil2016weapons}). Effective XAI tools can aid us in
-monitoring models and providing recourse to individuals to turn negative
-outcomes (e.g., ``loan application rejected'') into positive ones (e.g.,
-``application accepted''). In line with this, our work builds upon
-\textbf{counterfactual explanations} (CE) proposed by
-\citet{wachter2017counterfactual} as an effective approach to achieve
-this goal. CEs prescribe minimal changes for factual inputs that, if
-implemented, would prompt some fitted model to produce a desired output.
+model representations, aiming to mitigate scenarios in which
+practitioners deploy opaque models and have to blindly rely on their
+predictions. On many occasions, this has happened in practice, causing
+harms to people who were adversely and unfairly affected by automated
+decision-making (ADM) systems involving opaque models (see, e.g.,
+\citet{oneil2016weapons}). Effective XAI tools can also aid in
+monitoring models and providing recourse, empowering people to turn
+negative outcomes (e.g., ``loan application rejected'') into positive
+ones (e.g., ``loan application accepted''). In line with this, our work
+builds upon \textbf{counterfactual explanations} (CE) proposed by
+\citet{wachter2017counterfactual}; CEs prescribe minimal changes for
+factual inputs that, if implemented, would prompt some fitted model to
+produce an alternative, more desirable output.
 
 To our surprise, the second challenge has not yet attracted major
-research interest. Specifically, there has been no concerted effort
-towards improving the ``explanatory capacity'' of models, i.e., the
-degree to which learned representations correspond to explanations that
-are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
-Def. \ref{def-explainability}). Instead, the choice has generally been
-to improve the ability of XAI tools to identify the subset of
-explanations that are both plausible and valid for any given model,
-independent of whether the learned representations are in fact
-compatible with plausible explanations \citep{altmeyer2024faithful}.
-Fortunately, recent findings indicate that improved explanatory capacity
-can arise as a consequence of regularization techniques aimed at other
-training objectives such as generative capacity, generalization, or
-robustness
+research interest. In particular, there has been no concerted effort
+towards improving the degree to which learned representations correspond
+to explanations that are \textbf{interpretable} and deemed
+\textbf{plausible} by humans, which we simply term ``explainability'' in
+this manuscript (see Def. \ref{def-explainability}). Instead, the choice
+has generally been to improve the ability of XAI tools to identify the
+subset of explanations that are both plausible and valid for any given
+model, independent of whether these explanations are compatible with the
+learned representations \citep{altmeyer2024faithful}. Fortunately,
+recent findings indicate that improved explainability can arise as a
+consequence of regularization techniques aimed at other training
+objectives such as generative capacity, generalization, or robustness
 \citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
-As further discussed in Section~\ref{sec-lit}, our work consolidates
-these findings within a single objective.
-
-\textbf{Specifically, we introduce Counterfactual Training (CT)}: a
-novel training regime explicitly geared towards improving the
-explainability of models. In high-level terms, we define this concept as
-as the extent to which valid explanations derived for an opaque model
-are also deemed plausible with respect to the underlying data and the
-global actionability constraints. To the best of our knowledge, our
-framework represents the first attempt to address this challenge by
-employing counterfactual explanations already in the training phase.
+Our work consolidates these findings within a single framework.
+
+Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
+training regime explicitly geared towards improving the explainability
+of models. In high-level terms, we define this concept as as the extent
+to which valid explanations derived for an opaque model are also deemed
+plausible with respect to the underlying data and the global
+actionability constraints. To our knowledge, our framework represents
+the first attempt to address this challenge by employing counterfactual
+explanations already during the training phase.
 
@@
 actionability constraints. In Section~\ref{sec-experiments}, through
-extensive experiments, we demonstrate that CT substantially improves
-explainability and positively contributes to the robustness of trained
-models without sacrificing predictive performance. Finally, in
-Section~\ref{sec-discussion}, we discuss the challenges and conclude
+extensive experiments, we empirically demonstrate that CT substantially
+improves explainability and positively contributes to the robustness of
+trained models without sacrificing predictive performance. Finally, in
+Section~\ref{sec-discussion}, we discuss open challenges and conclude
 that CT is a promising approach towards making opaque models more

### 2025-06-27 paper/sections/intro.qmd
Adjust introduction

--- 2025-03-14 paper/sections/intro.qmd
@@
 
-Today's prominence of artificial intelligence (AI) has largely been driven by **representation learning**: instead of relying on features and rules that are carefully hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
-Modern advances in computing have made it possible to provide such models with ever-growing degrees of freedom to achieve that task, which frequently allows them to outperform traditionally more parsimonious models.
-Unfortunately, in doing so, models learn increasingly complex and highly sensitive representations that humans can no longer easily interpret.
+Today's prominence of artificial intelligence (AI) has largely been driven by **representation learning**: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
+Advances in computing have made it possible to provide these models with ever-growing degrees of freedom to achieve this task, which often allows them to outperform traditionally parsimonious models.
+Unfortunately, in doing so, models learn increasingly complex, sensitive representations that humans can no longer easily interpret.
 
-The trend towards complexity for the sake of performance has come under serious scrutiny in recent years.
-At the very cusp of the deep learning (DL) revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are sensitive to adversarial examples (AEs): perturbed versions of data instances that yield vastly different model predictions despite being "imperceptible" in that they are semantically indifferent from their factual counterparts.
-Even though some partially effective mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---truly robust deep learning remains unattainable even for models that are considered "shallow" by today's standards [@kolter2023keynote].
+The trend towards complexity for the sake of performance has come under scrutiny in recent years.
+At the very cusp of the deep learning (DL) revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are susceptible to adversarial examples (AEs): perturbed versions of data instances that yield vastly different model predictions despite being semantically indistinguishable from their factual counterparts.
+Some partial mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---but truly robust deep learning remains unattainable even for models that are considered "shallow" by today's standards [@kolter2023keynote].
 
-Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] Indeed, recent work on the so-called "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% while preserving their predictive performance [@frankle2018lottery].
-Similarly, @zhang2021understanding showed that state-of-the-art neural networks are expressive enough to fit randomly labeled data.
-Thus, looking at the predictive performance alone, the solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative, semantically meaningless patterns.
-This poses two challenges.
-Firstly, there is no dependable way to verify if representations correspond to meaningful, plausible explanations.
-Secondly, even if we could resolve the first challenge, it remains undecided how to ensure that models can *only* learn valuable explanations.
+Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] As one example, research on the "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% without losing predictive performance [@frankle2018lottery].
+Thus, looking at the predictive performance alone, found solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative and semantically meaningless patterns.
+This poses two related challenges.
+Firstly, there is no dependable way to verify if learned representations correspond to meaningful, plausible explanations.
+Secondly, even if we resolve this challenge, it remains undecided how to ensure that machine learning models can *only* learn valuable explanations.
 
-The first challenge has attracted an abundance of research on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations.
-Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions.
-On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@mcgregor2021preventing; @oneil2016weapons].
-Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted").
-In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal.
-CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce a desired output.
+The first challenge has attracted an abundance of work on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations, aiming to mitigate scenarios in which practitioners deploy opaque models and have to blindly rely on their predictions.
+On many occasions, this has happened in practice, causing harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models (see, e.g., @oneil2016weapons).
+Effective XAI tools can also aid in monitoring models and providing recourse, empowering people to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "loan application accepted").
+In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual; CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output.
 
 To our surprise, the second challenge has not yet attracted major research interest.
-Specifically, there has been no concerted effort towards improving the "explanatory capacity" of models, i.e., the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans (see Def.
+In particular, there has been no concerted effort towards improving the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans, which we simply term "explainability" in this manuscript (see Def.
 \ref{def-explainability}).
-Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether the learned representations are in fact compatible with plausible explanations [@altmeyer2024faithful].
-Fortunately, recent findings indicate that improved explanatory capacity can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating].
-As further discussed in @sec-lit, our work consolidates these findings within a single objective.
+Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether these explanations are compatible with the learned representations [@altmeyer2024faithful].
+Fortunately, recent findings indicate that improved explainability can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating].
+Our work consolidates these findings within a single framework.
 
-**Specifically, we introduce Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models.
+Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models.
 In high-level terms, we define this concept as as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and the global actionability constraints.
-To the best of our knowledge, our framework represents the first attempt to address this challenge by employing counterfactual explanations already in the training phase.
+To our knowledge, our framework represents the first attempt to address this challenge by employing counterfactual explanations already during the training phase.
 
@@
 In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints.
-In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability without sacrificing predictive performance.
-We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.
+In @sec-experiments, through extensive experiments, we empirically demonstrate that CT substantially improves explainability and positively contributes to the robustness of trained models without sacrificing predictive performance.
+Finally, in @sec-discussion, we discuss open challenges and conclude that CT is a promising approach towards making opaque models more trustworthy.

### 2025-06-27 paper/quarto_aaai/paper.tex
Adjust framing of "explainability"

@@
 research interest. In particular, there has been no concerted effort
-towards improving the degree to which learned representations correspond
-to explanations that are \textbf{interpretable} and deemed
-\textbf{plausible} by humans, which we simply term ``explainability'' in
-this manuscript (see Def. \ref{def-explainability}). Instead, the choice
-has generally been to improve the ability of XAI tools to identify the
-subset of explanations that are both plausible and valid for any given
-model, independent of whether these explanations are compatible with the
-learned representations \citep{altmeyer2024faithful}. Fortunately,
-recent findings indicate that improved explainability can arise as a
-consequence of regularization techniques aimed at other training
-objectives such as generative capacity, generalization, or robustness
+towards improving the degree to which learned representations promote
+explanations that are both \textbf{interpretable} to and deemed
+\textbf{plausible} by humans. Instead, the typical choice has been to
+improve the ability of XAI tools to identify the subset of explanations
+that are plausible and valid for any given model, independent of whether
+these explanations are compatible with the learned representations
+\citep{altmeyer2024faithful}. Fortunately, recent findings indicate that
+improved ``explainability'' of a model can arise as a consequence of
+regularization techniques aimed at other training objectives such as
+generative capacity, generalization, or robustness
 \citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
-Our work consolidates these findings within a single framework.
+Our contribution consolidates these findings within a unified framework.
 
@@
 training regime explicitly geared towards improving the explainability
-of models. In high-level terms, we define this concept as as the extent
-to which valid explanations derived for an opaque model are also deemed
-plausible with respect to the underlying data and the global
-actionability constraints. To our knowledge, our framework represents
-the first attempt to address this challenge by employing counterfactual
-explanations already during the training phase.
+of models. In high-level terms, we posit that the explainability of a
+model depends on the extent to which valid explanations derived for this
+model are also deemed plausible with respect to the underlying data and
+the global actionability constraints (see Def.
+\ref{def-explainability}). To the best of our knowledge, Counterfactual
+Training represents the first attempt to achieve more explainable models
+by employing counterfactual explanations already in the training phase.
 
@@
 
-We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. We define these properties as follows:
+We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are also plausible and actionable. We define these properties as follows:
 

### 2025-06-27 paper/sections/intro.qmd
Adjust framing of "explainability"

@@
 To our surprise, the second challenge has not yet attracted major research interest.
-In particular, there has been no concerted effort towards improving the degree to which learned representations correspond to explanations that are **interpretable** and deemed **plausible** by humans, which we simply term "explainability" in this manuscript (see Def.
-\ref{def-explainability}).
-Instead, the choice has generally been to improve the ability of XAI tools to identify the subset of explanations that are both plausible and valid for any given model, independent of whether these explanations are compatible with the learned representations [@altmeyer2024faithful].
-Fortunately, recent findings indicate that improved explainability can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating].
-Our work consolidates these findings within a single framework.
+In particular, there has been no concerted effort towards improving the degree to which learned representations promote explanations that are both **interpretable** to and deemed **plausible** by humans.
+Instead, the typical choice has been to improve the ability of XAI tools to identify the subset of explanations that are plausible and valid for any given model, independent of whether these explanations are compatible with the learned representations [@altmeyer2024faithful].
+Fortunately, recent findings indicate that improved "explainability" of a model can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating].
+Our contribution consolidates these findings within a unified framework.
 
 Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models.
-In high-level terms, we define this concept as as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and the global actionability constraints.
-To our knowledge, our framework represents the first attempt to address this challenge by employing counterfactual explanations already during the training phase.
+In high-level terms, we posit that the explainability of a model depends on the extent to which valid explanations derived for this model are also deemed plausible with respect to the underlying data and the global actionability constraints (see Def.
+\ref{def-explainability}).
+To the best of our knowledge, Counterfactual Training represents the first attempt to achieve more explainable models by employing counterfactual explanations already in the training phase.
 

### 2025-06-27 paper/quarto_aaai/paper.tex
Adjust related work

@@
 \citep{altmeyer2024faithful}. Fortunately, recent findings indicate that
-improved ``explainability'' of a model can arise as a consequence of
-regularization techniques aimed at other training objectives such as
+improved ``explanatory capacity'' of a model can arise as a consequence
+of regularization techniques aimed at other training objectives such as
 generative capacity, generalization, or robustness
@@
 Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
-training regime explicitly geared towards improving the explainability
-of models. In high-level terms, we posit that the explainability of a
-model depends on the extent to which valid explanations derived for this
-model are also deemed plausible with respect to the underlying data and
-the global actionability constraints (see Def.
-\ref{def-explainability}). To the best of our knowledge, Counterfactual
+training regime explicitly geared towards improving the explanatory
+capacity of models that, in high-level terms, we define as the extent to
+which valid explanations derived for a model can be deemed plausible
+with respect to the underlying data and global actionability constraints
+(we refine this notion in Def. \ref{def-explainability}). For
+simplicity, we refer to models with high explanatory capacity as
+\emph{explainable}. To the best of our knowledge, Counterfactual
 Training represents the first attempt to achieve more explainable models
@@
 To make the desiderata for our framework more concrete, we follow
-\citet{augustin2020adversarial} in tying the concept of explainability
-to the quality of CEs that can be generated for a given model. The
-authors show that CEs---understood as minimal input perturbations that
-yield some desired model prediction---tend to be more meaningful if the
-underlying model is more robust to adversarial examples. We can make
-intuitive sense of this finding when looking at adversarial training
-(AT) through the lens of representation learning with high degrees of
-freedom. As argued before, learned representations may be sensitive to
-producing implausible explanations and mispredicting for worst-case
-counterfactuals (i.e., AEs). Thus, by inducing models to ``unlearn''
-susceptiblity to such examples, adversarial training can effectively
-remove implausible explanations from the solution space.
-
-\subsection{Adversarial Examples are Counterfactual
-Explanations}\label{adversarial-examples-are-counterfactual-explanations}
-
-This interpretation of the link between explainability through
-counterfactuals on one side and robustness to adversarial examples on
-the other is backed by empirical evidence.
-\citet{sauer2021counterfactual} demonstrates that using counterfactual
+\citet{augustin2020adversarial} in tying explainability to the quality
+of CEs that can be generated for a given model. The authors show that
+CEs (understood as minimal input perturbations that yield some desired
+model prediction) tend to be more meaningful if the underlying model is
+more robust to adversarial examples. We can make intuitive sense of this
+finding if we look at adversarial training (AT) through the lens of
+representation learning with high degrees of freedom. As argued before,
+learned representations may be sensitive to producing implausible
+explanations and mispredicting for worst-case counterfactuals (i.e.,
+AEs). Thus, by inducing models to ``unlearn'' susceptiblity to such
+examples, adversarial training can effectively remove implausible
+explanations from the solution space.
+
+\subsection{Adversarial Examples are
+Counterfactuals}\label{adversarial-examples-are-counterfactuals}
+
+The interpretation of the link between explainability through
+counterfactuals on the one side, and robustness to adversarial examples
+on the other is backed by empirical evidence.
+\citet{sauer2021counterfactual} demonstrate that using counterfactual
 images during classifier training improves model robustness. Similarly,
-\citet{abbasnejad2020counterfactual} argues that counterfactuals
+\citet{abbasnejad2020counterfactual} argue that counterfactuals
 represent potentially useful training data in machine learning,
 especially in supervised settings where inputs may be reasonably mapped
-to multiple outputs. They, too, demonstrate that augmenting the training
-data of image classifiers can improve generalization. Finally,
-\citet{teney2020learning} proposes an approach using counterfactuals in
-training that does not rely on data augmentation: they argue that
-counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on identifying similar input samples
-with different annotations and ensuring that the gradient of the
+to multiple outputs. They, too, show that augmenting the training data
+of (image) classifiers can improve generalization performance. Finally,
+\citet{teney2020learning} argue that counterfactual pairs tend to exist
+in training data. Hence, their approach aims to identify similar input
+samples with different annotations and ensure that the gradient of the
 classifier aligns with the vector between such pairs of counterfactual
-inputs using the cosine distance as the loss function.
-
-In the natural language processing (NLP) domain, CEs have also been used
-to improve models through data augmentation. \citet{wu2021polyjuice2}
-proposes \emph{Polyjuice}, a general-purpose counterfactual generator
-for language models. The authors empirically demonstrate that the
+inputs using a cosine distance loss function.
+
+CEs have also been used to improve models in the natural language
+processing (NLP) domain. First, \citet{wu2021polyjuice2} propose
+\emph{Polyjuice}, a general-purpose counterfactual generator for
+language models. The authors empirically demonstrate that the
 augmentation of training data with their method improves robustness in a
-number of NLP tasks. \citet{balashankar2023improving} similarly uses
+number of NLP tasks. Second, \citet{balashankar2023improving} use
 \emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
 and show that classifier robustness improves by up to 20\%. Finally,
-\citet{luu2023counterfactual} introduces Counterfactual Adversarial
-Training (CAT) that also aims to improve generalization and robustness
-of language models through a three-step procedure: the authors identify
-training samples that are subject to high predictive uncertainty,
-generate CEs for them, and fine-tune the language model on a dataset
-augmented with the CEs.
+\citet{luu2023counterfactual} introduce Counterfactual Adversarial
+Training (CAT) that aims to improve generalization and robustness of
+language models by generating counterfactuals for training samples that
+are subject to high predictive uncertainty.
 
@@
 misclassification is achieved \citep{freiesleben2022intriguing}.
-Similarly, \citet{pawelczyk2022exploring} shows that CEs and AEs are
+Similarly, \citet{pawelczyk2022exploring} show that CEs and AEs are
 equivalent under certain conditions.
 
-Two recent works are closely related to ours in that they use
+Two other works are closely related to ours in that they use
 counterfactuals during training with the explicit goal of affecting
 certain properties of the post-hoc counterfactual explanations. Firstly,
-\citet{ross2021learning} proposes a way to train models that guarantee
-individual recourse to some positive target class with high probability.
-Their approach builds on adversarial training by explicitly inducing
-susceptibility to targeted adversarial examples for the positive class.
-Additionally, the proposed method allows for imposing a set of
-actionability constraints ex-ante. For example, users can specify that
-certain features are immutable. Secondly, \citet{guo2023counternet} is
-the first to propose an end-to-end training pipeline that includes CEs
-as part of the training procedure. In particular, they propose a
-specific network architecture that includes a predictor and CE generator
-network, where the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed
-back to the predictor network. In contrast to \citet{guo2023counternet},
-we impose no restrictions on the ANN architecture at all.
-
-\subsection{Aligning Representations with Plausible
-Explanations}\label{aligning-representations-with-plausible-explanations}
+\citet{ross2021learning} propose a way to train models that guarantee
+recourse to a positive target class with high probability. Their
+approach builds on adversarial training by explicitly inducing
+susceptibility to targeted AEs for the positive class. Additionally, the
+method allows for imposing a set of actionability constraints ex-ante.
+For example, users can specify that certain features are immutable.
+Secondly, \citet{guo2023counternet} are the first to propose an
+end-to-end training pipeline that includes CEs as part of the training
+procedure. Their \emph{CounterNet} network architecture includes a
+predictor and a CE generator, where the parameters of the CE generator
+are learnable. Counterfactuals are generated during each training
+iteration and fed back to the predictor. In contrast, we impose no
+restrictions on the ANN architecture at all.
+
+\subsection{Aligning Representations with
+Explanations}\label{aligning-representations-with-explanations}
 
 Improving the adversarial robustness of models is not the only path
-towards aligning representations with plausible explanations. In a work
-closely related to this one, \citet{altmeyer2024faithful} shows that
+towards aligning representations with plausible explanations. In a
+closely related work, \citet{altmeyer2024faithful} show that
 explainability can be improved through model averaging and refined model
-objectives. The authors propose a way to generate counterfactuals that
-are maximally faithful to the model in that they are consistent with
-what the model has learned about the underlying data. Formally, they
-rely on tools from energy-based modelling \citep{teh2003energy} to
-minimize the divergence between the distribution of counterfactuals and
-the conditional posterior over inputs learned by the model. Their
-proposed counterfactual explainer, \emph{ECCCo}, yields plausible
-explanations if and only if the underlying model has learned
-representations that align with them. The authors find that both deep
-ensembles \citep{lakshminarayanan2016simple} and joint energy-based
-models (JEMs) \citep{grathwohl2020your} tend to do well in this regard.
+objectives. They propose a way to generate counterfactuals that are
+maximally faithful to the model in that they are consistent with what
+the model has learned about the underlying data. Formally, they rely on
+tools from energy-based modelling \citep{teh2003energy} to minimize the
+divergence between the distribution of counterfactuals and the
+conditional posterior over inputs learned by the model. Their
+counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
+and only if the underlying model has learned representations that align
+with them. The authors find that both deep ensembles
+\citep{lakshminarayanan2016simple} and joint energy-based models (JEMs)
+\citep{grathwohl2020your} tend to do well in this regard.
 
@@
 representation learning with high degrees of freedom. Deep ensembles are
-approximate Bayesian model averages, which are most called for when
-models are underspecified by the available data \citep{wilson2020case}.
-Averaging across solutions mitigates the aforementioned risk of relying
-on a single locally optimal representations that corresponds to
-semantically meaningless explanations for the data. Previous work of
-\citet{schut2021generating} similarly found that generating plausible
-(``interpretable'') CEs is almost trivial for deep ensembles that have
-also undergone adversarial training. The case for JEMs is even clearer:
-they involve a hybrid objective that induces both high predictive
-performance and generative capacity \citep{grathwohl2020your}. This is
-closely related to the idea of aligning models with plausible
-explanations and has inspired our CT objective.
+approximate Bayesian model averages, which are particularly effective
+when models are underspecified by the available data
+\citep{wilson2020case}. Averaging across solutions mitigates the
+aforementioned risk of overrelying on a single locally optimal
+representation that corresponds to semantically meaningless explanations
+for the data. Likewise, previous work of \citet{schut2021generating}
+found that generating plausible (``interpretable'') CEs is almost
+trivial for deep ensembles that have undergone adversarial training. The
+case for JEMs is even clearer: they optimize a hybrid objective that
+induces both high predictive performance and strong generative capacity
+\citep{grathwohl2020your}, which bears resemblance to the idea of
+aligning models with plausible explanations and has inspired our CT
+objective.
 
@@
 
-We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are also plausible and actionable. We define these properties as follows:
+We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated counterfactuals are also plausible and actionable. We define these properties as follows:
 
@@
 \end{enumerate}
-where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
+where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable}. 
 \end{definition}

### 2025-06-27 paper/sections/intro.qmd
Adjust related work

@@
 Instead, the typical choice has been to improve the ability of XAI tools to identify the subset of explanations that are plausible and valid for any given model, independent of whether these explanations are compatible with the learned representations [@altmeyer2024faithful].
-Fortunately, recent findings indicate that improved "explainability" of a model can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating].
+Fortunately, recent findings indicate that improved "explanatory capacity" of a model can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating].
 Our contribution consolidates these findings within a unified framework.
 
-Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explainability of models.
-In high-level terms, we posit that the explainability of a model depends on the extent to which valid explanations derived for this model are also deemed plausible with respect to the underlying data and the global actionability constraints (see Def.
+Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explanatory capacity of models that, in high-level terms, we define as the extent to which valid explanations derived for a model can be deemed plausible with respect to the underlying data and global actionability constraints (we refine this notion in Def.
 \ref{def-explainability}).
+For simplicity, we refer to models with high explanatory capacity as *explainable*.
 To the best of our knowledge, Counterfactual Training represents the first attempt to achieve more explainable models by employing counterfactual explanations already in the training phase.

### 2025-06-27 paper/sections/method.qmd
Adjust framing of "explainability"
Adjust related work

@@
 
-We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. We define these properties as follows:
+We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated counterfactuals are also plausible and actionable. We define these properties as follows:
 

### 2025-06-27 paper/sections/related.qmd
Adjust related work

--- 2025-03-14 paper/sections/related.qmd
@@
 
-To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of CEs that can be generated for a given model.
-The authors show that CEs---understood as minimal input perturbations that yield some desired model prediction---tend to be more meaningful if the underlying model is more robust to adversarial examples.
-We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom.
+To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying explainability to the quality of CEs that can be generated for a given model.
+The authors show that CEs (understood as minimal input perturbations that yield some desired model prediction) tend to be more meaningful if the underlying model is more robust to adversarial examples.
+We can make intuitive sense of this finding if we look at adversarial training (AT) through the lens of representation learning with high degrees of freedom.
 As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., AEs).
@@
 
-## Adversarial Examples are Counterfactual Explanations
+## Adversarial Examples are Counterfactuals
 
-This interpretation of the link between explainability through counterfactuals on one side and robustness to adversarial examples on the other is backed by empirical evidence.
-@sauer2021counterfactual demonstrates that using counterfactual images during classifier training improves model robustness.
+The interpretation of the link between explainability through counterfactuals on the one side, and robustness to adversarial examples on the other is backed by empirical evidence.
+@sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness.
 Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs.
-They, too, demonstrate that augmenting the training data of image classifiers can improve generalization.
-Finally, @teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets.
-Specifically, their approach relies on identifying similar input samples with different annotations and ensuring that the gradient of the classifier aligns with the vector between such pairs of counterfactual inputs using the cosine distance as the loss function.
+They, too, show that augmenting the training data of (image) classifiers can improve generalization performance.
+Finally, @teney2020learning argue that counterfactual pairs tend to exist in training data.
+Hence, their approach aims to identify similar input samples with different annotations and ensure that the gradient of the classifier aligns with the vector between such pairs of counterfactual inputs using a cosine distance loss function.
 
-In the natural language processing (NLP) domain, CEs have also been used to improve models through data augmentation.
-@wu2021polyjuice2 propose *Polyjuice*, a general-purpose counterfactual generator for language models.
+CEs have also been used to improve models in the natural language processing (NLP) domain.
+First, @wu2021polyjuice2 propose *Polyjuice*, a general-purpose counterfactual generator for language models.
 The authors empirically demonstrate that the augmentation of training data with their method improves robustness in a number of NLP tasks.
-@balashankar2023improving similarly use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%.
-Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models through a three-step procedure: the authors identify training samples that are subject to high predictive uncertainty, generate CEs for those samples, and then fine-tune the language model on a dataset augmented with the CEs.
+Second, @balashankar2023improving use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%.
+Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT) that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
 
@@
 
-Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of the post-hoc counterfactual explanations.
-Firstly, @ross2021learning propose a way to train models that guarantee individual recourse to some positive target class with high probability.
-Their approach builds on adversarial training by explicitly inducing susceptibility to targeted adversarial examples for the positive class.
-Additionally, the proposed method allows for imposing a set of actionability constraints ex-ante.
+Two other works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of the post-hoc counterfactual explanations.
+Firstly, @ross2021learning propose a way to train models that guarantee recourse to a positive target class with high probability.
+Their approach builds on adversarial training by explicitly inducing susceptibility to targeted AEs for the positive class.
+Additionally, the method allows for imposing a set of actionability constraints ex-ante.
 For example, users can specify that certain features are immutable.
 Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes CEs as part of the training procedure.
-In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable.
-Counterfactuals are generated during each training iteration and fed back to the predictor network.
-In contrast to @guo2023counternet, we impose no restrictions on the ANN architecture at all.
+Their *CounterNet* network architecture includes a predictor and a CE generator, where the parameters of the CE generator are learnable.
+Counterfactuals are generated during each training iteration and fed back to the predictor.
+In contrast, we impose no restrictions on the ANN architecture at all.
 
-## Aligning Representations with Plausible Explanations
+## Aligning Representations with Explanations
 
 Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations.
-In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives.
-The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
+In a closely related work, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives.
+They propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
 Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
-Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
+Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
 The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.
@@
 Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom.
-Deep ensembles are approximate Bayesian model averages, which are most called for when models are underspecified by the available data [@wilson2020case].
-Averaging across solutions mitigates the aforementioned risk of relying on a single locally optimal representations that corresponds to semantically meaningless explanations for the data.
-Previous work by @schut2021generating similarly found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have also undergone adversarial training.
-The case for JEMs is even clearer: they involve a hybrid objective that induces both high predictive performance and generative capacity [@grathwohl2020your].
-This is closely related to the idea of aligning models with plausible explanations and has inspired our CT objective.
+Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case].
+Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations for the data.
+Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training.
+The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which bears resemblance to the idea of aligning models with plausible explanations and has inspired our CT objective.

### 2025-06-28 paper/quarto_aaai/paper.tex
Adjust methods

@@
 
-In this section we propose our novel counterfactual training objective.
-In CT, we combine ideas from adversarial training, counterfactual
-explanations, and energy-based modelling with the explicit goal of
-aligning representations with plausible explanations that comply with
-user-defined actionability constraints.
+This section introduces the Counterfactual Training objective. CT
+combines ideas from adversarial training, counterfactual explanations,
+and energy-based modelling with the explicit goal of producing models
+whose learned representations align with plausible explanations that
+further comply with user-defined actionability constraints.
 
 In the context of counterfactual explanations, plausibility has broadly
-been defined as the degree to which counterfactuals comply with the
+been defined as the degree to which generated CEs comply with the
 underlying data-generating process
@@
 provide algorithmic recourse (AR) to individuals (negatively) affected
-by opaque models. To be actionable, AR recommendations must also be
-attainable. A plausible CE for a rejected 20-year-old loan applicant,
-for example, might reveal that their application would have been
-accepted, if only they were 20 years older. Ignoring all other features,
-this would comply with the definition of plausibility if 40-year-old
-individuals were in fact more credit-worthy on average than young
-adults. But of course this CE does not qualify for providing actionable
-recourse to the applicant since \emph{age} is not a (directly) mutable
-feature. Counterfactual training aims to improve model explainability by
-aligning models with counterfactuals that meet both desiderata:
-plausibility and actionability. Formally, we define explainability as
-follows:
+by opaque models. An AR recommendations must also be actionable, i.e.,
+possible to attain by the recipient. A plausible CE for a rejected
+20-year-old loan applicant, for example, might reveal that their
+application would have been accepted, if only they had been 20 years
+older. Ignoring all other features, this would comply with the
+definition of plausibility if 40-year-old individuals were in fact more
+credit-worthy on average than young adults. But of course this CE does
+not qualify for providing actionable recourse to the applicant since
+\emph{age} is not a (directly) mutable feature. Counterfactual training
+aims to improve model explainability by aligning models with
+counterfactuals that meet both desiderata: plausibility and
+actionability. Formally, we define explainability as follows:
 
@@
 Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$.
-Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.
+Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$, where $\arg\max_y{\mathbf{y}^{+}}=y^+$ is the index of the target class.
 
@@
 \begin{enumerate}
-    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
+    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$, where $A$ is some arbitrarily small region around $\mathbf{x}^{\prime}$.
+    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
     \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
-    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
 \end{enumerate}
-where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs. For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable}. 
+and $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs.
+For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript.
 \end{definition}
@@
 consistent with the data and faithful counterfactuals are consistent
-with what the model has learned about input data. Actionability
+with what the model has learned about the input data. Actionability
 constraints in Def. \ref{def-explainability} vary and depend on the
 context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
-focus on domain and mutability constraints for individual features
-\(x_d\) for \(d=1,...,D\). We limit ourselves to classification tasks
-for reasons discussed in Section~\ref{sec-discussion}.
+choose to only consider domain and mutability constraints for individual
+features \(x_d\) for \(d=1,...,D\). We also limit ourselves to
+classification tasks for reasons discussed in
+Section~\ref{sec-discussion}.
 
@@
 
-To train models with high explainability as defined in Def.
-\ref{def-explainability}, we propose to leverage counterfactuals in the
-following objective:
+To train explainable models as defined in Def. \ref{def-explainability},
+we propose to leverage counterfactuals in the following objective:
 
@@
 be summarized as inducing explainability directly and indirectly by
-penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
-between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
-samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
-target class \(y^+\), and, (2) the adversarial loss,
+penalizing (in 2\textsuperscript{nd} term) the contrastive divergence,
+\(\text{div}(\cdot)\), between mature counterfactuals
+\(\mathbf{x}_T^\prime\) and observed samples
+\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
+class \(y^+\), and (in 3\textsuperscript{rd} term) the adversarial loss,
 \(\text{advloss}(.)\), with respect to nascent counterfactuals
@@
 
-As observed by \citet{grathwohl2020your}, any classifier can be
-re-interpreted as a joint energy-based model (JEM) that learns to
-discriminate output classes conditional on the observed (training)
-samples from \(p(\mathbf{x})\) and the generated samples from
-\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
-perform well at both tasks by directly maximizing the joint
-log-likelihood factorized as
-\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
-The first term can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\),
-\citet{grathwohl2020your} minimizes the contrastive divergence between
-the observed samples from \(p(\mathbf{x})\) and generated samples from
+\citet{grathwohl2020your} observed that any classifier can be
+re-interpreted as a joint energy-based model that learns to discriminate
+output classes conditional on the observed (training) samples from
+\(p(\mathbf{x})\) and the generated samples from
+\(p_\theta(\mathbf{x})\). \citet{grathwohl2020your} show that JEMs can
+be trained to perform well at both tasks by directly maximizing the
+joint log-likelihood:
+\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\),
+where the first term can be optimized using cross-entropy as in
+Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\), they
+minimize the contrastive divergence between the observed samples from
+\(p(\mathbf{x})\) and the generated samples from
 \(p_\theta(\mathbf{x})\).
 
-A key empirical finding in \citet{altmeyer2024faithful} was that JEMs
-tend to do well with respect to the plausibility objective in Def.
+A key empirical finding of \citet{altmeyer2024faithful} was that JEMs
+perform well on the plausibility objective in Def.
 \ref{def-explainability}. This follows directly if we consider samples
-drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
+drawn from \(p_\theta(\mathbf{x})\) as counterfactuals --- the JEM
 objective effectively minimizes the divergence between the conditional
 posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
-\citet{grathwohl2020your} relies on Stochastic Gradient Langevin
-Dynamics (SGLD) using an uninformative prior for initialization but we
-depart from their methodology. Instead of SGLD, we propose to use
-counterfactual explainers to generate counterfactuals of observed
-training samples. Specifically, we have:
+\citet{grathwohl2020your} use Stochastic Gradient Langevin Dynamics
+(SGLD) with an uninformative prior for initialization but we depart from
+their methodology. Instead we propose to leverage counterfactual
+explainers to generate counterfactuals of observed training samples.
+Specifically, we have:
 
@@
 
-The departure from SGLD of \citep{grathwohl2020your} allows us to tap
-into the vast repertoire of explainers that have been proposed in the
-literature to meet different desiderata. For example, many methods
-facilitate the imposition of domain and mutability constraints. In
-principle, any existing approach for generating counterfactual
-explanations is viable, so long as it does not violate the faithfulness
-condition. Like JEMs \citep{murphy2022probabilistic}, CT can be
+The departure from SGLD allows us to tap into the vast repertoire of
+explainers that have been proposed in the literature to meet different
+desiderata. For example, many methods support domain and mutability
+constraints. In principle, any existing approach for generating CEs is
+viable, so long as it does not violate the faithfulness condition. Like
+JEMs \citep{murphy2022probabilistic}, Counterfactual Training can be
 considered a form of contrastive representation learning.
@@
 examples \citep{freiesleben2022intriguing, pawelczyk2022exploring}. This
-leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
-Equation~\ref{eq-obj}. An intuitive functional form, but likely not the
-only sensible choice, is inspired by adversarial training:
+leaves some flexibility with regards to the choice for the
+\(\text{advloss}(\cdot)\) term in Equation~\ref{eq-obj}. An intuitive
+functional form, but likely not the only sensible choice, is inspired by
+adversarial training:
 
@@
 perturbation to any single feature is at most \(\varepsilon\). This is
-closely aligned with \citet{szegedy2013intriguing} that defines an
+closely aligned with \citet{szegedy2013intriguing} who define an
 adversarial attack as an ``imperceptible non-random perturbation''.
 Thus, we choose to work with a different distinction between CE and AE
-than \citet{freiesleben2022intriguing} that considers misclassification
-as the key distinguishing feature of AE. One of the key observations of
+than \citet{freiesleben2022intriguing} who consider misclassification as
+the key distinguishing feature of AE. One of the key observations of
 this work is that we can leverage CEs during training and get
@@
 constraints out-of-the-box. In fact, both types of constraints can be
-implemented for any counterfactual explainer that relies on gradient
-descent in the feature space for optimization
-\citep{altmeyer2023explaining}. In this context, domain constraints can
-be imposed by simply projecting counterfactuals back to the specified
-domain, if the previous gradient step resulted in updated feature values
-that were out-of-domain. Mutability constraints can similarly be
-enforced by setting partial derivatives to zero to ensure that features
-are only perturbed in the allowed direction, if at all.
-
-Since such actionability constraints are binding at test time, we should
-also impose them when generating \(\mathbf{x}^\prime\) during each
-training iteration to inform model representations. Through their effect
-on \(\mathbf{x}^\prime\), both types of constraints influence model
+implemented for any explainer that relies on gradient descent in the
+feature space for optimization \citep{altmeyer2023explaining}. In this
+context, domain constraints can be imposed by simply projecting
+counterfactuals back to the specified domain, if the previous gradient
+step resulted in updated feature values that were out-of-domain.
+Similarly, mutability constraints can be enforced by setting partial
+derivatives to zero to ensure that features are only perturbed in the
+allowed direction, if at all.
+
+Since actionability constraints are binding at test time, we should also
+impose them when generating \(\mathbf{x}^\prime\) during each training
+iteration to inform model representations. Through their effect on
+\(\mathbf{x}^\prime\), both types of constraints influence model
 outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
@@
 \item
-  We have \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but feature \(d\)
-  can only be decreased in practice.
+  \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but \(d\) can only be
+  decreased in practice.
 \item
-  We have \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but feature \(d\)
-  can only be increased in practice.
+  \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but \(d\) can only be
+  increased in practice.
 \end{enumerate}
@@
 assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
-feature \(d\). Intuitively, we think of this simply in terms ignoring
-implausibility costs with respect to immutable features, which
-effectively forces the model to instead seek plausibility with respect
-to the remaining features. This in turn results in lower overall
-sensitivity to immutable features, which we demonstrate empirically for
-different classifiers in Section~\ref{sec-experiments}. Under certain
-conditions, this result holds theoretically:\footnote{For the proof, see
-  the supplementary appendix.}
+feature \(d\). Intuitively, we think of this as ignoring implausibility
+costs with respect to immutable features, which effectively forces the
+model to instead seek plausibility with respect to the remaining
+features. This in turn results in lower overall sensitivity to immutable
+features, which we demonstrate empirically for different classifiers in
+Section~\ref{sec-experiments}. Under certain conditions, this result
+holds theoretically:\footnote{For the proof, see the supplementary
+  appendix.}
 
@@
 Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
-If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
+Assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
 \end{proposition}
 
-\noindent It is worth highlighting that
-\mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
-This raises a valid concern about the effect of protecting immutable
-features in the presence of proxies that remain unprotected. We address
-this in Section~\ref{sec-discussion}.
+\noindent It is worth noting that \mbox{Proposition \ref{prp-mtblty}}
+assumes independence of features. This raises a valid concern about the
+effect of protecting immutable features in the presence of proxies that
+remain unprotected. We address this in Section~\ref{sec-discussion}.
 
@@
 and (2) the subjects' existing level of \emph{debt}, which we define as
-mutable.
-
-We have simulated this scenario using synthetic data with independent
-\emph{age} and \emph{debt} features, and Gaussian class-conditional
-densities in Figure~\ref{fig-poc}. The four panels show the outcomes for
-different training procedures using the same model architectures (a
-linear classifier). In panels (a) and (c) we have trained the models
-conventionally, while in panels (b) and (d) we used CT.
-
-In all cases, all counterfactuals (stars) are valid---they have crossed
-the decision boundary (green)---but their quality differs. In panel (a),
-they are not plausible: they do not comply with the distribution of the
-factuals in \(y^+\) to the point where they form a clearly
-distinguishable cluster. In panel (b), they are highly plausible,
-meeting the first objective of Def. \ref{def-explainability}. In panel
-(c), the CEs involve substantial reductions in \emph{debt} for younger
-applicants. By comparison, counterfactual paths are shorter on average
-in panel (d) where we have protected the immutable \emph{age} as
-described in Section~\ref{sec-constraints}. Due to the classifier's
-lower sensitivity to \emph{age}, recommendations with respect to
-\emph{debt} are much more homogenous and do not unfairly punish younger
-individuals. These counterfactuals are also plausible with respect to
-the mutable feature. Thus, we consider the model in panel (d) as the
-most explainable according to Def. \ref{def-explainability}.
+mutable. We have simulated this scenario using synthetic data with
+independent \emph{age} and \emph{debt} features, and Gaussian
+class-conditional densities in Figure~\ref{fig-poc}. The four panels
+show the outcomes for different training procedures using the same model
+architecture (a linear classifier). In panels (a) and (c) we have
+trained the models conventionally, while in panels (b) and (d) we have
+applied CT.
+
+In all four cases, all counterfactuals (stars) are valid---they have
+crossed the decision boundary (green) into the target class---but their
+quality differs. In panel (a), they are not plausible: they do not
+comply with the distribution of the factuals in \(y^+\) to the point
+where they form a clearly discernible cluster. In panel (b), they are
+highly plausible, meeting the first objective of Def.
+\ref{def-explainability}. In panel (c), the CEs involve substantial
+reductions in \emph{debt} for younger applicants. By comparison,
+counterfactual paths are shorter on average in panel (d) where we have
+protected the immutable \emph{age} as described in
+Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
+to \emph{age}, recommendations with respect to \emph{debt} are much more
+homogenous and do not unfairly punish younger individuals. These
+counterfactuals are also plausible with respect to the mutable feature.
+Thus, we consider the model in panel (d) as the most explainable
+according to Def. \ref{def-explainability}.
 
@@
 immutable. The linear decision boundary is shown in green along with
-training data colored according to their ground-truth label: \(y^-=1\)
-in blue and \(y^+=2\) in orange. Stars indicate counterfactuals in the
-target class.}
+training data colored according to ground-truth labels: \(y^-=1\) (blue)
+and \(y^+=2\) (orange). Stars indicate counterfactuals in the target
+class.}
 

### 2025-06-28 paper/sections/method.qmd
Adjust methods

@@
 
-In this section we propose our novel counterfactual training objective.
-In CT, we combine ideas from adversarial training, counterfactual explanations, and energy-based modelling with the explicit goal of aligning representations with plausible explanations that comply with user-defined actionability constraints.
+This section introduces the Counterfactual Training objective.
+CT combines ideas from adversarial training, counterfactual explanations, and energy-based modelling with the explicit goal of producing models whose learned representations align with plausible explanations that further comply with user-defined actionability constraints.
 
-In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which counterfactuals comply with the underlying data-generating process [@altmeyer2024faithful;@guidotti2022counterfactual;@poyiadzi2020face].
+In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which generated CEs comply with the underlying data-generating process [@altmeyer2024faithful;@guidotti2022counterfactual;@poyiadzi2020face].
 Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models.
-To be actionable, AR recommendations must also be attainable.
-A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they were 20 years older.
+An AR recommendations must also be actionable, i.e., possible to attain by the recipient.
+A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they had been 20 years older.
 Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults.
@@
 \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation.
-Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about input data.
+Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data.
 Actionability constraints in Def.
 \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed.
-In this work, we focus on domain and mutability constraints for individual features $x_d$ for $d=1,...,D$.
-We limit ourselves to classification tasks for reasons discussed in @sec-discussion.
+In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$.
+We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.
 
@@
 
-To train models with high explainability as defined in Def. \ref{def-explainability}, we propose to leverage counterfactuals in the following objective:
+To train explainable models as defined in Def. \ref{def-explainability}, we propose to leverage counterfactuals in the following objective:
 
@@
 The second and third terms are explained in detail below.
-For now, they can be summarized as inducing explainability directly and indirectly by penalizing: (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and, (2) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
+For now, they can be summarized as inducing explainability directly and indirectly by penalizing (in 2\textsuperscript{nd} term) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (in 3\textsuperscript{rd} term) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
 Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit].
@@
 
-As observed by @grathwohl2020your, any classifier can be re-interpreted as a joint energy-based model (JEM) that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
-The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood factorized as $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$.
-The first term can be optimized using conventional cross-entropy as in @eq-obj.
-To optimize $\log p_\theta(\mathbf{x})$, @grathwohl2020your minimizes the contrastive divergence between the observed samples from $p(\mathbf{x})$ and generated samples from $p_\theta(\mathbf{x})$.
+@grathwohl2020your observed that any classifier can be re-interpreted as a joint energy-based model that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
+@grathwohl2020your show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood: $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$, where the first term can be optimized using cross-entropy as in @eq-obj.
+To optimize $\log p_\theta(\mathbf{x})$, they minimize the contrastive divergence between the observed samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
 
-A key empirical finding in @altmeyer2024faithful was that JEMs tend to do well with respect to the plausibility objective in Def.
+A key empirical finding of @altmeyer2024faithful was that JEMs perform well on the plausibility objective in Def.
 \ref{def-explainability}.
-This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals because the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$.
-To generate samples, @grathwohl2020your relies on Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative prior for initialization but we depart from their methodology.
-Instead of SGLD, we propose to use counterfactual explainers to generate counterfactuals of observed training samples.
+This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals --- the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$.
+To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology.
+Instead we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples.
 Specifically, we have:
@@
 
-The departure from SGLD of [@grathwohl2020your] allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
-For example, many methods facilitate the imposition of domain and mutability constraints.
-In principle, any existing approach for generating counterfactual explanations is viable, so long as it does not violate the faithfulness condition.
-Like JEMs [@murphy2022probabilistic], CT can be considered a form of contrastive representation learning.
+The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
+For example, many methods support domain and mutability constraints.
+In principle, any existing approach for generating CEs is viable, so long as it does not violate the faithfulness condition.
+Like JEMs [@murphy2022probabilistic], Counterfactual Training can be considered a form of contrastive representation learning.
 
@@
 Based on our analysis in @sec-lit, counterfactuals $\mathbf{x}^\prime$ can be repurposed as additional training samples [@balashankar2023improving;@luu2023counterfactual] or adversarial examples [@freiesleben2022intriguing;@pawelczyk2022exploring].
-This leaves some flexibility wrt.
-the choice for $\text{advloss}(\cdot)$ in @eq-obj.
+This leaves some flexibility with regards to the choice for the $\text{advloss}(\cdot)$ term in @eq-obj.
 An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:
@@
 Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$.
-This is closely aligned with @szegedy2013intriguing that defines an adversarial attack as an "imperceptible non-random perturbation".
-Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing that considers misclassification as the key distinguishing feature of AE.
+This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation".
+Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE.
 One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.
@@
 Many existing counterfactual explainers support domain and mutability constraints out-of-the-box.
-In fact, both types of constraints can be implemented for any counterfactual explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining].
+In fact, both types of constraints can be implemented for any explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining].
 In this context, domain constraints can be imposed by simply projecting counterfactuals back to the specified domain, if the previous gradient step resulted in updated feature values that were out-of-domain.
-Mutability constraints can similarly be enforced by setting partial derivatives to zero to ensure that features are only perturbed in the allowed direction, if at all.
+Similarly, mutability constraints can be enforced by setting partial derivatives to zero to ensure that features are only perturbed in the allowed direction, if at all.
 
-Since such actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations.
+Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations.
 Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div.
@@
 1. Feature $d$ is strictly immutable in practice.
-2. We have $\mathbf{x}^+[d]>\mathbf{x}^\prime[d]$, but feature $d$ can only be decreased in practice.
-3. We have $\mathbf{x}^+[d]<\mathbf{x}^\prime[d]$, but feature $d$ can only be increased in practice.
+2. $\mathbf{x}^+[d]>\mathbf{x}^\prime[d]$, but $d$ can only be decreased in practice.
+3. $\mathbf{x}^+[d]<\mathbf{x}^\prime[d]$, but $d$ can only be increased in practice.
 
 `\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$.
-Intuitively, we think of this simply in terms ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features.
+Intuitively, we think of this as ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features.
 This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments.
@@
 Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
-If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
+Assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
 \end{proposition}
 
-`\noindent`{=latex} It is worth highlighting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
+`\noindent`{=latex} It is worth noting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
 This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected.
@@
 We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.
-
 We have simulated this scenario using synthetic data with independent *age* and *debt* features, and Gaussian class-conditional densities in @fig-poc.
-The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier).
-In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we used CT.
+The four panels show the outcomes for different training procedures using the same model architecture (a linear classifier).
+In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT.
 
@@
 
-In all cases, all counterfactuals (stars) are valid---they have crossed the decision boundary (green)---but their quality differs.
-In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster.
+In all four cases, all counterfactuals (stars) are valid---they have crossed the decision boundary (green) into the target class---but their quality differs.
+In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly discernible cluster.
 In panel (b), they are highly plausible, meeting the first objective of Def.
@@
 ![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable.
-The linear decision boundary is shown in green along with training data colored according to their ground-truth label: $y^-=1$ in blue and $y^+=2$ in orange.
+The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange).
 Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

### 2025-06-28 paper/quarto_aaai/paper.tex
Adjust experiments

@@
 
-\subsection{Example (Prediction of Consumer Credit
-Default)}\label{example-prediction-of-consumer-credit-default}
+\subsection{Example: Prediction of Credit
+Defaults}\label{example-prediction-of-credit-defaults}
 
@@
 
-In our experiments we seek to answer the following three research
-questions:
+We seek to answer the following three research questions:
 
@@
 \item
-  To what extent does the CT objective produce more favorable
-  algorithmic recourse outcomes in the presence of actionability
-  constraints?
+  To what extent does CT produce more favorable AR outcomes in the
+  presence of actionability constraints?
 \item
-  What are the effects of hyperparameter selection wrt. the CT
-  objective?
+  What are the effects of hyperparameter selection on CT?
 \end{enumerate}
@@
 Our key outcome of interest is improvement in explainability (Def.
-\ref{def-explainability}). To this end, we focus primarily on the
-plausibility and cost of faithfully generated counterfactuals at test
-time. To measure the cost, we follow the standard convention of using
-distances (\(\ell_1\)-norm) between factuals and counterfactuals as a
-proxy. For plausibility, we assess how similar CEs are to the observed
-samples in the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We
-rely on the distance-based metric used in \citet{altmeyer2024faithful},
+\ref{def-explainability}). Thus, we focus primarily on the plausibility
+and cost of faithfully generated counterfactuals at test time. To
+measure the cost, we follow the standard convention of using distances
+(\(\ell_1\)-norm) between factuals and counterfactuals as a proxy. For
+plausibility, we assess how similar CEs are to the observed samples in
+the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
+metric used by \citet{altmeyer2024faithful},
 
@@
 }\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
-counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
-squared population maximum mean discrepancy \citep{gretton2012kernel}.
-The metric in Equation~\ref{eq-impl-div} is equal to zero iff the two
-distributions are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
+counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
+the squared population maximum mean discrepancy, proposed by
+\citet{gretton2012kernel}. The metric in Equation~\ref{eq-impl-div} is
+equal to zero if and only if the two distributions are exactly the same,
+\(\mathbf{X}^\prime=\mathbf{X}^+\).
 
 In addition to cost and plausibility, we compute other standard metrics
-to evaluate counterfactuals including validity and redundancy. Finally,
-we also assess the predictive performance of models using standard
-metrics, including robust accuracy estimated on adversarially perturbed
-data using FGSM \citep{goodfellow2014explaining}.
+to evaluate counterfactuals such as validity and redundancy. Finally, we
+also assess the predictive performance of models using standard metrics,
+including robust accuracy estimated on adversarially perturbed data
+using FGSM \citep{goodfellow2014explaining}.
 
@@
 (\emph{MLP}). This is the best way to get a clear picture of the
-effectiveness of CT, and it is consistent with how assessment is done in
+effectiveness of CT, and it is consistent with evaluation practices in
 the related literature
@@
 reduction is negative, consistent with the distance-based metric.
-\emph{MNIST} is the only dataset for which the distance and divergence
-metrics disagree. Upon visual inspection of the image counterfactuals we
-find that CT clearly improves plausibility (see supplementary appendix
-for images).
-
-\subsubsection{Predictive Performance}\label{sec-pred}
+\emph{MNIST} is the only dataset for which the two metrics disagree.
+However, upon visual inspection of the image counterfactuals (available
+in the supplementary appendix), we find that CT clearly improves
+plausibility.
+
+\subsubsection{Predictive Performance.}\label{sec-pred}
 
@@
 these datasets in particular, we find that CT strongly outperforms the
-baseline. In fact, we find that CT improves adversarial robustness on
+baseline. In fact, we observe that CT improves adversarial robustness on
 all datasets.
@@
 
-\subsubsection{Actionability}\label{sec-act}
-
-In Section~\ref{sec-method}, we show that our proposed way for encoding
-mutability constraints leads to lower classifier sensitivity wrt.
-immutable features for linear models, tilting the decision boundary in
-favour of mutable features instead. For binding constraints at test
-time, this leads to shorter counterfactual paths and hence smaller
-average costs (\(\ell_1\)-norm) to individuals. To extend this to the
-non-linear case, we test the effect of imposing mutability constraints
-empirically for our synthetic data using the same evaluation scheme as
-above. The final row in Table~\ref{tbl-main} reports the average
-reduction in costs for CT compared to the ``vanilla'' baseline, when
-imposing that either the first or the second feature is immutable. In
-all cases, costs are reduced substantially, indicating that classifiers
-trained with CT are indeed more sensitive to mutable features.
-
-\subsubsection{Impact of hyperparameter
-settings.}\label{sec-hyperparameters}
-
-We test the impact of three types of hyperparameters; our complete
-results are in the supplementary appendix.\\
-\indent We note that CT is highly sensitive to the choice of a CE
-generator and its hyperparameters but (a) there are manageable patterns
-and (b) we can typically identify settings that improve either
-plausibility or cost, and commonly both of them at the same time. For
+\subsubsection{Actionability.}\label{sec-act}
+
+In Section~\ref{sec-method}, we show that our method of encoding
+mutability constraints decreases sensitivity to immutable features for
+linear models, tilting the decision boundary in favour of mutable
+features. For constraints binding at test time, this leads to shorter
+counterfactual paths and hence smaller average costs (\(\ell_1\)-norm)
+to individuals. To extend this to the non-linear case, we test the
+effect of imposing mutability constraints for the synthetic datasets
+using the same evaluation scheme as above. The final row in
+Table~\ref{tbl-main} reports the average reduction in costs for CT
+compared to the ``vanilla'' baseline, when imposing that either the
+first or the second feature is immutable. In all cases, costs are
+reduced substantially, indicating that classifiers trained with CT are
+indeed more sensitive to mutable features.
+
+\subsubsection{Hyperparameter settings.}\label{sec-hyperparameters}
+
+We test the impact of three types of hyperparameters. Here we focus on
+the highlights; full results are available in the supplementary
+appendix.\\
+\indent First, we note that CT is highly sensitive to the choice of a CE
+generator and its hyperparameters but (1) there are manageable patterns,
+and (2) we can usually identify settings that improve either
+plausibility or cost, and often both of them at the same time. For
 example, \emph{REVISE} tends to perform the worst, most likely because
@@
 also factuals) and so increasing this threshold does not seem to
-correlate with better outcomes. In fact, the choice of \(\tau=0.5\)
-generally leads to optimal results because it is associated with high
-proportions of mature counterfactuals.\\
-\indent The strength of the energy regularization,
+correlate with better outcomes. In fact, \(\tau=0.5\) generally leads to
+optimal results as it is associated with high proportions of mature
+counterfactuals.\\
+\indent Second, the strength of the energy regularization,
 \(\lambda_{\text{reg}}\) is highly impactful and leads to poor
@@
 very pronounced.\\
-\indent The effectiveness and stability of CT is positively associated
-with the number of counterfactuals generated during each training epoch.
-We also confirm that a higher number of training epochs is beneficial.
+\indent Third, the effectiveness and stability of CT is positively
+associated with the number of counterfactuals generated during each
+training epoch. A higher number of training epochs is also beneficial.
 Interestingly, we observed desired improvements when CT was combined

### 2025-06-28 paper/sections/experiments.qmd
Fix some final typos
can't get full width table to work
Adjust experiments

@@
 
-In our experiments we seek to answer the following three research questions:
+We seek to answer the following three research questions:
 
 1. To what extent does the counterfactual training objective as it is defined in @eq-obj induce models to learn plausible explanations?
-2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
-3. What are the effects of hyperparameter selection wrt. the CT objective?
+2. To what extent does CT produce more favorable AR outcomes in the presence of actionability constraints?
+3. What are the effects of hyperparameter selection on CT?
 
@@
 \ref{def-explainability}).
-To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
+Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
 To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.
 For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$.
-We rely on the distance-based metric used in @altmeyer2024faithful,
+We rely on the metric used by @altmeyer2024faithful,
 
@@
 $$ {#eq-impl-div}
-where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is an unbiased estimate of the squared population maximum mean discrepancy [@gretton2012kernel].
-The metric in @eq-impl-div is equal to zero iff the two distributions are the same, $\mathbf{X}^\prime=\mathbf{X}^+$.
+where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel.
+The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.
 
-In addition to cost and plausibility, we compute other standard metrics to evaluate counterfactuals including validity and redundancy.
+In addition to cost and plausibility, we compute other standard metrics to evaluate counterfactuals such as validity and redundancy.
 Finally, we also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].
@@
 To assess CT, we investigate the improvements in performance metrics when using it on top of a weak baseline (BL): a multilayer perceptron (*MLP*).
-This is the best way to get a clear picture of the effectiveness of CT, and it is consistent with how assessment is done in the related literature [@goodfellow2014explaining;@ross2021learning;@teney2020learning].
+This is the best way to get a clear picture of the effectiveness of CT, and it is consistent with evaluation practices in the related literature [@goodfellow2014explaining;@ross2021learning;@teney2020learning].
 
@@
 For *OL* the reduction is negative, consistent with the distance-based metric.
-*MNIST* is the only dataset for which the distance and divergence metrics disagree.
-Upon visual inspection of the image counterfactuals we find that CT clearly improves plausibility (see supplementary appendix).
+*MNIST* is the only dataset for which the two metrics disagree.
+However, upon visual inspection of the image counterfactuals (available in the supplementary appendix), we find that CT clearly improves plausibility.
 
-### Predictive Performance {#sec-pred}
+### Predictive Performance. {#sec-pred}
 
@@
 When looking at robust test accuracies (Acc.$^*$) for these datasets in particular, we find that CT strongly outperforms the baseline.
-In fact, we find that CT improves adversarial robustness on all datasets.
+In fact, we observe that CT improves adversarial robustness on all datasets.
 
@@
 ```{=latex}
+\begin{table*}
 \input{tables/main.tex}
+\end{table*}
 ```
@@
 
-### Actionability {#sec-act}
+### Actionability. {#sec-act}
 
-In @sec-method, we show that our proposed way for encoding mutability constraints leads to lower classifier sensitivity wrt.
-immutable features for linear models, tilting the decision boundary in favour of mutable features instead.
-For binding constraints at test time, this leads to shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals.
-To extend this to the non-linear case, we test the effect of imposing mutability constraints empirically for our synthetic data using the same evaluation scheme as above.
+In @sec-method, we show that our method of encoding mutability constraints decreases sensitivity to immutable features for linear models, tilting the decision boundary in favour of mutable features.
+For constraints binding at test time, this leads to shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals.
+To extend this to the non-linear case, we test the effect of imposing mutability constraints for the synthetic datasets using the same evaluation scheme as above.
 The final row in @tbl-main reports the average reduction in costs for CT compared to the "vanilla" baseline, when imposing that either the first or the second feature is immutable.
@@
 
-### Impact of hyperparameter settings. {#sec-hyperparameters}
-We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
-`\indent`{=latex} We note that CT is highly sensitive to the choice of a CE generator and its hyperparameters but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time.
+### Hyperparameter settings. {#sec-hyperparameters}
+We test the impact of three types of hyperparameters. Here we focus on the highlights; full results are available in the supplementary appendix.  
+`\indent`{=latex} First, we note that CT is highly sensitive to the choice of a CE generator and its hyperparameters but (1) there are manageable patterns, and (2) we can usually identify settings that improve either plausibility or cost, and often both of them at the same time.
 For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful].
@@
 On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes.
-In fact, the choice of $\tau=0.5$ generally leads to optimal results because it is associated with high proportions of mature counterfactuals.
-`\indent`{=latex} The strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high.
+In fact, $\tau=0.5$ generally leads to optimal results as it is associated with high proportions of mature counterfactuals.
+`\indent`{=latex} Second, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high.
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
 While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
-`\indent`{=latex} The effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
-We also confirm that a higher number of training epochs is beneficial.
+`\indent`{=latex} Third, the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
+A higher number of training epochs is also beneficial.
 Interestingly, we observed desired improvements when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process.

### 2025-06-28 paper/quarto_aaai/paper.tex
Adjust conclusion, currently at 7 pages of text

@@
 As our results indicate, counterfactual training produces models that
-are more explainable. Nonetheless, it brings about three important
-limitations.\\
-\indent \emph{CT increases the training time of models.} CT can be more
+are more explainable. Nonetheless, these advantages come at the cost of
+two important limitations.
+
+\emph{CT increases the training time of models.} CT can be more
 time-consuming than conventional training regimes. While higher numbers
 of CEs per iteration positively impact the quality of solutions, they
-also increase the amount of computations. Relatively small grids with
-270 settings can take almost four hours for more demanding datasets on a
+also increase the amount of computations. In our experiments, grids with
+270 settings took up to four hours for more demanding datasets on a
 high-performance computing cluster with 34 2GB CPUs.\footnote{See
   supplementary appendix for computational details.} Three factors
-attenuate this effect: (1) CT amortizes the cost of CEs for the training
-samples; (2) it can retain its value when used as a ``fine-tuning''
-technique for conventionally-trained models; and (3) it yields itself to
-parallel execution, which we have leveraged for our own experiments.\\
-\indent \emph{Immutable features may have proxies.} We propose an
-approach to protect immutable features and thus increase the
-actionability of the generated CEs. However, it requires that model
-owners define the mutability constraints for (all) features considered
-by the model. Even if all immutable features are protected, there may
-exist proxies that are mutable (and hence should not be protected) but
-preserve enough information about the principals to hinder the
-protections. Delineating actionability is a major open challenge in the
-AR literature (see, e.g., \citep{venkatasubramanian2020philosophical})
-impacting the capacity of CT to fulfill its intended goal.\\
-\indent \emph{Interventions on features may impact fairness.} We provide
-a tool that allows practitioners to modify the sensitivity of a model
-with respect to certain features, which may have implication for the
-fair and equitable treatment of decision subjects. Model owners could
-misuse this solution by enforcing explanations based on features that
-are more difficult to modify by some (group of) individuals. For
-example, consider the \emph{Adult} dataset used in our experiments,
-where \emph{workclass} or \emph{education} may be more difficult to
-change for underpriviledged groups. When applied irresponsibly, CT could
-result in an unfairly assigned burden of recourse
-\citep{sharma2020certifai}, threatening the equality of opportunity in
-the system \citep{bell2024fairness}. Nonetheless, these phenomena are
-not specific to CT.\\
-\hfill \break \indent We also highlight several important directions for
-future research. Firstly, it is an interesting challenge to extend CT
-beyond classification settings. Our formulation relies on the
-distinction between non-target class(es) \(y^{-}\) and target class(es)
-\(y^{+}\) to generate counterfactuals through Equation~\ref{eq-obj}.
-While \(y^{-}\) and \(y^{+}\) can be arbitrarily defined, CT requires
-the output space \(\mathcal{Y}\) to be discrete. Thus, it does not apply
-to ML tasks where the change in outcome cannot be readily quantified.
-Focus on classification models is a common restriction in research on
-CEs and AR. Other settings have attracted some interest (e.g.,
-regression in \citep{spooner2021counterfactual}), but there is little
-consensus how to robustly extend the notion of CEs.\\
-\indent Secondly, our approach is susceptible to training instabilities.
-This problem has been recognized for JEMs \citep{grathwohl2020your} and
-even though we depart from the SGLD-based sampling, we still encounter
-considerable variability in the outcomes. CT is exposed to two potential
-sources of instabilities: (1) the energy-based contrastive divergence
-term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
-explainers. We find several promising ways to mitigate this problem:
-regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
-many counterfactuals during each epoch, and including only mature
+attenuate this effect: (1) CT yields itself to parallel execution, which
+we have leveraged for our own experiments; (2) it amortizes the cost of
+CEs for the training samples; (3) as demonstrated, it retains its value
+when used as a ``fine-tuning'' technique for conventionally-trained
+models.
+
+\emph{Interventions on features have implications for fairness.} We
+provide a tool that allows practitioners to modify the sensitivity of a
+model with respect to certain features. Model owners can use our
+solution to support the fair and equitable treatment of decision
+subjects, but they could also misuse it by enforcing explanations based
+on features that are more difficult to modify by some (group of)
+individuals. When applied irresponsibly, CT could result in an unfairly
+assigned burden of recourse \citep{sharma2020certifai}, threatening the
+equality of opportunity in the system \citep{bell2024fairness}.
+Additionally, CT requires that model owners define the mutability
+constraints for the features considered by the model. Even if all
+immutable features are protected, there may exist proxies that are
+mutable --- and hence should not be protected --- but preserve
+sufficient information about the principals to hinder the protections.
+Deciding on actionability is still a major open challenge in the AR
+literature \citep{venkatasubramanian2020philosophical} impacting the
+capacity of CT to fulfill its intended goal. As such, neither of these
+phenomena is ultimately specific to CT.
+
+We also highlight several important directions for future research.
+Firstly, it is an interesting challenge to extend CT beyond
+classification settings. Our formulation relies on the distinction
+between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
+generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
+and \(y^{+}\) can be arbitrarily defined, CT requires the output space
+\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
+where the change in outcome cannot be readily discretized. Focus on
+classification is a common choice in research on CEs and AR. Other
+settings have attracted some interest, e.g., regression
+\citep{spooner2021counterfactual}, but there is little consensus how to
+robustly extend the notion of CEs.\\
+Secondly, CT is susceptible to instabilities in training. This problem
+has been recognized for JEMs \citep{grathwohl2020your} and even though
+we depart from the SGLD-based sampling, we still encounter variability
+in outcomes. CT is exposed to two potential sources of training
+instabilities: (1) the energy-based contrastive divergence term in
+Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
+We find several promising ways to mitigate this problem: regularizing
+energy (\(\lambda_{\text{reg}}\)), generating sufficiently many
+counterfactuals during each epoch, and including only mature
 counterfactuals for contrastive divergence.\\
-\indent Finally, we believe that it is possible to substantially improve
-hyperparameter selection procedures. Our method benefits from the tuning
-of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
-In this work, we have relied exclusively on grid search for this task.
-Future work on CT could benefit from investigating more sophisticated
-approaches. Notably, CT is iterative which makes methods such as
-Bayesian or gradient-based optimization applicable (see, e.g.,
-\citep{bischl2023hyperparameter}). \hfill \break
+Finally, we believe that it is possible to substantially improve
+hyperparameter selection procedures, and thus performance. In this work,
+we have relied exclusively on grid searches for this task. Future work
+on CT could benefit from investigating more sophisticated approaches.
+Notably, our method is iterative which makes methods such as Bayesian or
+gradient-based optimization applicable \citep{bischl2023hyperparameter}.
 
 To conclude, state-of-the-art machine learning models are prone to
-learning complex representations that cannot be interpreted by humans
-and existing post-hoc explainability approaches cannot guarantee that
-the explanations agree with the model's learned representation of data.
-As a step towards addressing this challenge, we introduced
-counterfactual training, a novel training regime that incentivizes
-highly-explainable models. Our approach leads to explanations that are
-both plausible---compliant with the underlying data-generating
-process---and actionable---compliant with user-specified mutability
-constraints---and thus meaningful to their recipients. Through extensive
-experiments we demonstrate that CT satisfies its objective while
-promoting robustness and preserving the predictive performance of the
-models. It can also be used to fine-tune conventionally-trained models
-and achieve similar gains. Finally, this work showcases that it is
-practical to improve models \emph{and} their explanations at the same
-time.
+learning complex representations that cannot be interpreted by humans.
+Existing explainability solutions cannot guarantee that explanations
+agree with the model's learned representation of data. As a step towards
+addressing this challenge, we introduced counterfactual training, a
+novel training regime that incentivizes highly-explainable models. Our
+approach leads to explanations that are both plausible (compliant with
+the underlying data-generating process) and actionable (compliant with
+user-specified mutability constraints), and thus meaningful to their
+recipients. Through extensive experiments, we demonstrated that CT
+satisfies its objective while promoting robustness and preserving the
+predictive performance of models. It can also be used to fine-tune
+conventionally-trained models and achieve similar gains. Thus, our work
+showcases that it is practical to simultaneously improve models and
+their explanations.
 

### 2025-06-28 paper/sections/discussion.qmd
Adjust conclusion, currently at 7 pages of text

@@
 
-As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, it brings about three important limitations.  
-`\indent`{=latex} *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes.
+As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, these advantages come at the cost of two important limitations.
+
+*CT increases the training time of models.* CT can be more time-consuming than conventional training regimes.
 While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations.
-Relatively small grids with 270 settings can take almost four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect: (1) CT amortizes the cost of CEs for the training samples; (2) it can retain its value when used as a "fine-tuning" technique for conventionally-trained models; and (3) it yields itself to parallel execution, which we have leveraged for our own experiments.
-`\indent`{=latex} *Immutable features may have proxies.* We propose an approach to protect immutable features and thus increase the actionability of the generated CEs.
-However, it requires that model owners define the mutability constraints for (all) features considered by the model.
-Even if all immutable features are protected, there may exist proxies that are mutable (and hence should not be protected) but preserve enough information about the principals to hinder the protections.
-Delineating actionability is a major open challenge in the AR literature (see, e.g., [@venkatasubramanian2020philosophical]) impacting the capacity of CT to fulfill its intended goal.
-`\indent`{=latex} *Interventions on features may impact fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of decision subjects.
-Model owners could misuse this solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.
-For example, consider the *Adult* dataset used in our experiments, where *workclass* or *education* may be more difficult to change for underpriviledged groups.
+In our experiments, grids with 270 settings took up to four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect: (1) CT yields itself to parallel execution, which we have leveraged for our own experiments; (2) it amortizes the cost of CEs for the training samples; (3) as demonstrated, it retains its value when used as a "fine-tuning" technique for conventionally-trained models.
+
+*Interventions on features have implications for fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features.
+Model owners can use our solution to support the fair and equitable treatment of decision subjects, but they could also misuse it by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.
 When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness].
-Nonetheless, these phenomena are not specific to CT.
-`\hfill \break`{=latex}
-`\indent`{=latex} We also highlight several important directions for future research.
+Additionally, CT requires that model owners define the mutability constraints for the features considered by the model.
+Even if all immutable features are protected, there may exist proxies that are mutable --- and hence should not be protected --- but preserve sufficient information about the principals to hinder the protections.
+Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.
+As such, neither of these phenomena is ultimately specific to CT.
+
+We also highlight several important directions for future research.
 Firstly, it is an interesting challenge to extend CT beyond classification settings.
@@
 While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete.
-Thus, it does not apply to ML tasks where the change in outcome cannot be readily quantified.
-Focus on classification models is a common restriction in research on CEs and AR.
-Other settings have attracted some interest (e.g., regression in [@spooner2021counterfactual]), but there is little consensus how to robustly extend the notion of CEs.
-`\indent`{=latex} Secondly, our approach is susceptible to training instabilities.
-This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter considerable variability in the outcomes.
-CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers.
+Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized.
+Focus on classification is a common choice in research on CEs and AR.
+Other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.
+Secondly, CT is susceptible to instabilities in training.
+This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter variability in outcomes.
+CT is exposed to two potential sources of training instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers.
 We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.
-`\indent`{=latex} Finally, we believe that it is possible to substantially improve hyperparameter selection procedures.
-Our method benefits from the tuning of certain key hyperparameters (see @sec-hyperparameters).
-In this work, we have relied exclusively on grid search for this task.
+Finally, we believe that it is possible to substantially improve hyperparameter selection procedures, and thus performance.
+In this work, we have relied exclusively on grid searches for this task.
 Future work on CT could benefit from investigating more sophisticated approaches.
-Notably, CT is iterative which makes methods such as Bayesian or gradient-based optimization applicable (see, e.g., [@bischl2023hyperparameter]).
-`\hfill \break`{=latex}
+Notably, our method is iterative which makes methods such as Bayesian or gradient-based optimization applicable [@bischl2023hyperparameter].

### 2025-06-28 paper/quarto_aaai/paper.tex
Add introduction to counterfactual generation, still at 7 pages

@@
 CEs have also been used to improve models in the natural language
-processing (NLP) domain. First, \citet{wu2021polyjuice2} propose
-\emph{Polyjuice}, a general-purpose counterfactual generator for
-language models. The authors empirically demonstrate that the
-augmentation of training data with their method improves robustness in a
-number of NLP tasks. Second, \citet{balashankar2023improving} use
-\emph{Polyjuice} to augment NLP datasets through diverse counterfactuals
-and show that classifier robustness improves by up to 20\%. Finally,
-\citet{luu2023counterfactual} introduce Counterfactual Adversarial
-Training (CAT) that aims to improve generalization and robustness of
-language models by generating counterfactuals for training samples that
-are subject to high predictive uncertainty.
+processing domain. For example, \citet{wu2021polyjuice2} propose
+\emph{Polyjuice}, a general-purpose CE generator for language models and
+demonstrate that the augmentation of training data with \emph{Polyjuice}
+improves robustness in a number of tasks, while
+\citet{luu2023counterfactual} introduce the \emph{Counterfactual
+Adversarial Training} (CAT) framework that aims to improve
+generalization and robustness of language models by generating
+counterfactuals for training samples that are subject to high predictive
+uncertainty.
 
@@
 
-This section introduces the Counterfactual Training objective. CT
+This section introduces the Counterfactual Training framework. CT
 combines ideas from adversarial training, counterfactual explanations,
@@
 Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
-explanation generated through gradient descent over \(T\) iterations as
-initially proposed by \citet{wachter2017counterfactual}. For our
-purposes, we let \(T\) vary and consider the counterfactual search as
+generated through gradient descent over \(T\) iterations as originally
+proposed by \citet{wachter2017counterfactual}. In broad terms,
+generating CEs using gradient descent entails solving some form of an
+objective that balances (1) the classification loss
+\(\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})\), and (2) one
+or more penalty terms \(\lambda_{i}\text{cost}_{i}(\cdot)\). The exact
+specification of these penalties induces various properties in the
+counterfactual outcomes, and tends to be the key feature that
+distinguishes various gradient-based ``generators'' or ``explainers''
+(CE algorithms) from the literature \citep{altmeyer2023explaining}.
+
+For our purposes, we let \(T\) vary and consider the search to be
 converged as soon as the predicted probability for the target class has
@@
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
-where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
-  background information on gradient-based counterfactual search and
-  convergence see supplementary appendix.}
-
-To train explainable models as defined in Def. \ref{def-explainability},
-we propose to leverage counterfactuals in the following objective:
+where \(\mathcal{S}\) is the softmax function.
+
+CT leverages gradient-based CE search already in training to produce
+perturbed versions of the factual samples \(\mathbf{x}\). Initially,
+until \(\tau\) is reached or when \(t\leq T\), generated samples
+\(\mathbf{x}_t^\prime\) serve as adversarial examples; we term them
+\emph{nascent} counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Over
+multiple training iterations up to time \(T\), these nascent
+counterfactuals move towards the decision boundary and turn into
+\emph{mature} counterfactuals \(\mathbf{x}_T^\prime\) when they cross
+into the target class or when \(t > T\).
+
+Formally, we propose the following counterfactual training objective to
+train explainable (as in Def. \ref{def-explainability}) models:
 
@@
 \end{split}
-}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
+}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
 that induces discriminative performance (e.g., cross-entropy). The
@@
 be summarized as inducing explainability directly and indirectly by
-penalizing (in 2\textsuperscript{nd} term) the contrastive divergence,
-\(\text{div}(\cdot)\), between mature counterfactuals
-\(\mathbf{x}_T^\prime\) and observed samples
+penalizing the contrastive divergence, \(\text{div}(\cdot)\), between
+mature counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
 \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
-class \(y^+\), and (in 3\textsuperscript{rd} term) the adversarial loss,
-\(\text{advloss}(.)\), with respect to nascent counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
-denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
-of the energy terms involved in \(\text{div}(\cdot)\)
-\citep{du2019implicit}. The trade-off between the components can
-governed through penalties \(\lambda_{\text{div}}\),
+class \(y^+\), and the adversarial loss, \(\text{advloss}(.)\), wrt.
+nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
+\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
+regularizes the magnitude of the energy terms involved in
+\(\text{div}(\cdot)\) \citep{du2019implicit}. The trade-off between the
+components are governed through \(\lambda_{\text{div}}\),
 \(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).
@@
 
-\noindent It is worth noting that \mbox{Proposition \ref{prp-mtblty}}
-assumes independence of features. This raises a valid concern about the
-effect of protecting immutable features in the presence of proxies that
-remain unprotected. We address this in Section~\ref{sec-discussion}.
+\noindent Proposition \ref{prp-mtblty} assumes independent features,
+which raises a concern about the effect of protecting immutable features
+in the presence of proxies. We address it in
+Section~\ref{sec-discussion}.
 
@@
 previous loan takers comprised of a binary outcome variable
-(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
+\(y\in\{1=\text{default},2=\text{no default}\}\) with two input
 features: (1) the subjects' \emph{age}, which we define as immutable,
@@
 \emph{CT increases the training time of models.} CT can be more
-time-consuming than conventional training regimes. While higher numbers
-of CEs per iteration positively impact the quality of solutions, they
-also increase the amount of computations. In our experiments, grids with
-270 settings took up to four hours for more demanding datasets on a
-high-performance computing cluster with 34 2GB CPUs.\footnote{See
-  supplementary appendix for computational details.} Three factors
-attenuate this effect: (1) CT yields itself to parallel execution, which
-we have leveraged for our own experiments; (2) it amortizes the cost of
-CEs for the training samples; (3) as demonstrated, it retains its value
-when used as a ``fine-tuning'' technique for conventionally-trained
-models.
+time-consuming than conventional training regimes. Higher numbers of CEs
+per iteration positively impact the quality of solutions, but they also
+increase the amount of computations. In our experiments, grids with 270
+settings took up to four hours for more demanding datasets on a
+high-performance computing cluster with 34 2GB CPUs (see supplementary
+appendix for details). Three factors attenuate this effect: (1) CT
+inherently yields itself to parallel execution; (2) it amortizes the
+cost of CEs for the training samples; (3) it retains its value when used
+to fine-tune conventionally-trained models.
 
@@
 on features that are more difficult to modify by some (group of)
-individuals. When applied irresponsibly, CT could result in an unfairly
+individuals. When used irresponsibly, CT could result in an unfairly
 assigned burden of recourse \citep{sharma2020certifai}, threatening the
-equality of opportunity in the system \citep{bell2024fairness}.
-Additionally, CT requires that model owners define the mutability
-constraints for the features considered by the model. Even if all
-immutable features are protected, there may exist proxies that are
-mutable --- and hence should not be protected --- but preserve
-sufficient information about the principals to hinder the protections.
-Deciding on actionability is still a major open challenge in the AR
-literature \citep{venkatasubramanian2020philosophical} impacting the
-capacity of CT to fulfill its intended goal. As such, neither of these
-phenomena is ultimately specific to CT.
+equality of opportunity \citep{bell2024fairness}. Additionally, CT
+requires mutability constraints for the features considered by the
+model. Even if all immutable features are protected, there may exist
+proxies that are mutable, and hence should not be protected, but
+preserve sufficient information about the principals to hinder the
+protections. Deciding on actionability is still a major open challenge
+in the AR literature \citep{venkatasubramanian2020philosophical}
+impacting the capacity of CT to fulfill its intended goal.
 
@@
 robustly extend the notion of CEs.\\
-Secondly, CT is susceptible to instabilities in training. This problem
-has been recognized for JEMs \citep{grathwohl2020your} and even though
-we depart from the SGLD-based sampling, we still encounter variability
-in outcomes. CT is exposed to two potential sources of training
-instabilities: (1) the energy-based contrastive divergence term in
-Equation~\ref{eq-div}, and (2) the underlying counterfactual explainers.
-We find several promising ways to mitigate this problem: regularizing
-energy (\(\lambda_{\text{reg}}\)), generating sufficiently many
-counterfactuals during each epoch, and including only mature
-counterfactuals for contrastive divergence.\\
+Secondly, CT is susceptible to training instabilities. This problem has
+been recognized for JEMs \citep{grathwohl2020your} and even though we
+depart from the SGLD sampling, we still encounter variability in
+outcomes. CT is exposed to two potential sources of instabilities: (1)
+the energy-based contrastive divergence term, \(\text{div}(\cdot)\), in
+Equation~\ref{eq-div}, and (2) the underlying explainers. We find
+several promising ways to mitigate this problem: regularizing energy
+(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
+during each epoch, and including only mature counterfactuals in
+\(\text{div}(\cdot)\).\\
 Finally, we believe that it is possible to substantially improve

### 2025-06-28 paper/sections/discussion.qmd
Add introduction to counterfactual generation, still at 7 pages

@@
 *CT increases the training time of models.* CT can be more time-consuming than conventional training regimes.
-While higher numbers of CEs per iteration positively impact the quality of solutions, they also increase the amount of computations.
-In our experiments, grids with 270 settings took up to four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs.^[See supplementary appendix for computational details.] Three factors attenuate this effect: (1) CT yields itself to parallel execution, which we have leveraged for our own experiments; (2) it amortizes the cost of CEs for the training samples; (3) as demonstrated, it retains its value when used as a "fine-tuning" technique for conventionally-trained models.
+Higher numbers of CEs per iteration positively impact the quality of solutions, but they also increase the amount of computations.
+In our experiments, grids with 270 settings took up to four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see supplementary appendix for details).
+Three factors attenuate this effect: (1) CT inherently yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; (3) it retains its value when used to fine-tune conventionally-trained models.
 
@@
 Model owners can use our solution to support the fair and equitable treatment of decision subjects, but they could also misuse it by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.
-When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness].
-Additionally, CT requires that model owners define the mutability constraints for the features considered by the model.
-Even if all immutable features are protected, there may exist proxies that are mutable --- and hence should not be protected --- but preserve sufficient information about the principals to hinder the protections.
+When used irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity [@bell2024fairness].
+Additionally, CT requires mutability constraints for the features considered by the model.
+Even if all immutable features are protected, there may exist proxies that are mutable, and hence should not be protected, but preserve sufficient information about the principals to hinder the protections.
 Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.
-As such, neither of these phenomena is ultimately specific to CT.
 
@@
 Other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.
-Secondly, CT is susceptible to instabilities in training.
-This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD-based sampling, we still encounter variability in outcomes.
-CT is exposed to two potential sources of training instabilities: (1) the energy-based contrastive divergence term in @eq-div, and (2) the underlying counterfactual explainers.
-We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals for contrastive divergence.
+Secondly, CT is susceptible to training instabilities.
+This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD sampling, we still encounter variability in outcomes.
+CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers.
+We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.
 Finally, we believe that it is possible to substantially improve hyperparameter selection procedures, and thus performance.

### 2025-06-28 paper/sections/method.qmd
Adjust experiments
Add introduction to counterfactual generation, still at 7 pages

@@
 
-This section introduces the Counterfactual Training objective.
+This section introduces the Counterfactual Training framework.
 CT combines ideas from adversarial training, counterfactual explanations, and energy-based modelling with the explicit goal of producing models whose learned representations align with plausible explanations that further comply with user-defined actionability constraints.
@@
 
-Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual explanation generated through gradient descent over $T$ iterations as initially proposed by @wachter2017counterfactual.
-For our purposes, we let $T$ vary and consider the counterfactual search as converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.^[For detailed background information on gradient-based counterfactual search and convergence see supplementary appendix.]
+Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual.
+In broad terms, generating CEs using gradient descent entails solving some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$.
+The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" (CE algorithms) from the literature [@altmeyer2023explaining].
 
-<!-- @sec-app-ce -->
+For our purposes, we let $T$ vary and consider the search to be converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.
 
-To train explainable models as defined in Def. \ref{def-explainability}, we propose to leverage counterfactuals in the following objective:
+CT leverages gradient-based CE search already in training to produce perturbed versions of the factual samples $\mathbf{x}$.
+Initially, until $\tau$ is reached or when $t\leq T$, generated samples $\mathbf{x}_t^\prime$ serve as adversarial examples; we term them *nascent* counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
+Over multiple training iterations up to time $T$, these nascent counterfactuals move towards the decision boundary and turn into *mature* counterfactuals $\mathbf{x}_T^\prime$ when they cross into the target class or when $t > T$.
+
+Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:
 
@@
 $$ {#eq-obj}
-where $\text{yloss}(\cdot)$ is a classification loss that induces discriminative performance (e.g., cross-entropy).
+where $\text{yloss}(\cdot)$ is any classification loss that induces discriminative performance (e.g., cross-entropy).
 The second and third terms are explained in detail below.
-For now, they can be summarized as inducing explainability directly and indirectly by penalizing (in 2\textsuperscript{nd} term) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (in 3\textsuperscript{rd} term) the adversarial loss, $\text{advloss}(.)$, with respect to nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
+For now, they can be summarized as inducing explainability directly and indirectly by penalizing the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and the adversarial loss, $\text{advloss}(.)$, wrt.
+nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
 Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit].
-The trade-off between the components can governed through penalties $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.
+The trade-off between the components are governed through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.
 
@@
 
-`\noindent`{=latex} It is worth noting that \mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
-This raises a valid concern about the effect of protecting immutable features in the presence of proxies that remain unprotected.
-We address this in @sec-discussion.
+`\noindent`{=latex} Proposition \ref{prp-mtblty} assumes independent features, which raises a concern about the effect of protecting immutable features in the presence of proxies.
+We address it in @sec-discussion.
 
-## Example (Prediction of Consumer Credit Default)
+## Example: Prediction of Credit Defaults
 
 Suppose we are interested in predicting the likelihood that loan applicants default on their credit.
-We have access to historical data on previous loan takers comprised of a binary outcome variable ($y\in\{1=\text{default},2=\text{no default}\}$) with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.
+We have access to historical data on previous loan takers comprised of a binary outcome variable $y\in\{1=\text{default},2=\text{no default}\}$ with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.
 We have simulated this scenario using synthetic data with independent *age* and *debt* features, and Gaussian class-conditional densities in @fig-poc.

### 2025-06-29 paper/quarto_aaai/paper.tex
Fix overview of CT

@@
     \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
-    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
+    \item (Actionability) Perturbations $\Delta$ are subject to some actionability constraints.
 \end{enumerate}
@@
 generated through gradient descent over \(T\) iterations as originally
-proposed by \citet{wachter2017counterfactual}. In broad terms,
-generating CEs using gradient descent entails solving some form of an
+proposed by \citet{wachter2017counterfactual}. In broad terms, searching
+for CEs using gradient descent entails optimizing some form of an
 objective that balances (1) the classification loss
@@
 
-CT leverages gradient-based CE search already in training to produce
-perturbed versions of the factual samples \(\mathbf{x}\). Initially,
-until \(\tau\) is reached or when \(t\leq T\), generated samples
-\(\mathbf{x}_t^\prime\) serve as adversarial examples; we term them
-\emph{nascent} counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Over
-multiple training iterations up to time \(T\), these nascent
-counterfactuals move towards the decision boundary and turn into
-\emph{mature} counterfactuals \(\mathbf{x}_T^\prime\) when they cross
-into the target class or when \(t > T\).
+CT adapts gradient-based CE search already in training to generate
+\(\mathbf{x}^\prime\), the perturbed versions of the factual samples
+\(\mathbf{x}\). We use the term \emph{nascent} to denote counterfactuals
+for which the search has not converged, \(\mathbf{x}_{t\leq T}^\prime\).
+Initially, when the perturbations are small, nascent counterfactuals
+serve as AEs. Over multiple iterations, they move towards the decision
+boundary and turn into \emph{mature} counterfactuals
+\(\mathbf{x}_T^\prime\) when they cross into the target class (becoming
+CEs) or when \(t > T\).
 

### 2025-06-29 paper/sections/method.qmd
Fix overview of CT

@@
 Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual.
-In broad terms, generating CEs using gradient descent entails solving some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$.
+In broad terms, searching for CEs using gradient descent entails optimizing some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$.
 The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" (CE algorithms) from the literature [@altmeyer2023explaining].
@@
 
-CT leverages gradient-based CE search already in training to produce perturbed versions of the factual samples $\mathbf{x}$.
-Initially, until $\tau$ is reached or when $t\leq T$, generated samples $\mathbf{x}_t^\prime$ serve as adversarial examples; we term them *nascent* counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
-Over multiple training iterations up to time $T$, these nascent counterfactuals move towards the decision boundary and turn into *mature* counterfactuals $\mathbf{x}_T^\prime$ when they cross into the target class or when $t > T$.
+CT adapts gradient-based CE search already in training to generate $\mathbf{x}^\prime$, the perturbed versions of the factual samples $\mathbf{x}$.
+We use the term *nascent* to denote counterfactuals for which the search has not converged, $\mathbf{x}_{t\leq T}^\prime$.
+Initially, when the perturbations are small, nascent counterfactuals serve as AEs.
+Over multiple iterations, they move towards the decision boundary and turn into *mature* counterfactuals $\mathbf{x}_T^\prime$ when they cross into the target class (becoming CEs) or when $t > T$.
 

### 2025-06-30 paper/quarto_aaai/paper.tex
Fix some awkward references after reformatting for AAAI

@@
 harms to people who were adversely and unfairly affected by automated
-decision-making (ADM) systems involving opaque models (see, e.g.,
-\citet{oneil2016weapons}). Effective XAI tools can also aid in
-monitoring models and providing recourse, empowering people to turn
-negative outcomes (e.g., ``loan application rejected'') into positive
-ones (e.g., ``loan application accepted''). In line with this, our work
-builds upon \textbf{counterfactual explanations} (CE) proposed by
+decision-making (ADM) systems involving opaque models; see, e.g.,
+\citet{oneil2016weapons}. Effective XAI tools can also aid in monitoring
+models and providing recourse, empowering people to turn negative
+outcomes (e.g., ``loan application rejected'') into positive ones (e.g.,
+``loan application accepted''). In line with this, our work builds upon
+\textbf{counterfactual explanations} (CE) proposed by
 \citet{wachter2017counterfactual}; CEs prescribe minimal changes for
@@
 
-\citet{grathwohl2020your} observed that any classifier can be
+\citet{grathwohl2020your} observe that any classifier can be
 re-interpreted as a joint energy-based model that learns to discriminate
@@
 \(p(\mathbf{x})\) and the generated samples from
-\(p_\theta(\mathbf{x})\). \citet{grathwohl2020your} show that JEMs can
-be trained to perform well at both tasks by directly maximizing the
-joint log-likelihood:
+\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
+perform well at both tasks by directly maximizing the joint
+log-likelihood:
 \(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\),
@@
 adversarial attack as an ``imperceptible non-random perturbation''.
-Thus, we choose to work with a different distinction between CE and AE
-than \citet{freiesleben2022intriguing} who consider misclassification as
-the key distinguishing feature of AE. One of the key observations of
-this work is that we can leverage CEs during training and get
-adversarial examples essentially for free, which can be used to reap the
-aforementioned benefits of adversarial training.
+Thus, we work with a different distinction between CE and AE than
+\citet{freiesleben2022intriguing} who considers misclassification as the
+distinguishing feature of adversarial examples. One of the key
+observations of this work is that we can leverage CEs during training
+and get AEs essentially for free to reap the aforementioned benefits of
+adversarial training.
 
@@
 moons (\emph{Moon}). These datasets are generated using the library of
-\citep{altmeyer2023explaining} and we present them in the supplementary
+\citet{altmeyer2023explaining} and we present them in the supplementary
 appendix. Next, we have four real-world binary tabular datasets from the
 domain of economics: \emph{Adult} (a.k.a. Census data) of
-\citep{becker1996adult2}, California housing (\emph{CH}) of
-\citep{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
-\citep{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) of
-\citep{kaggle2011give2}. Finally, for the convenience of illustration,
-we use of the 10-class \emph{MNIST} vision dataset
-\citep{lecun1998mnist}.
+\citet{becker1996adult2}, California housing (\emph{CH}) of
+\citet{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
+\citet{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) from
+\citet{kaggle2011give2}. Finally, for the convenience of illustration,
+we use the 10-class \emph{MNIST} dataset \citep{lecun1998mnist}.
 
@@
 proxies that are mutable, and hence should not be protected, but
-preserve sufficient information about the principals to hinder the
+preserve sufficient information about the principals to hinder these
 protections. Deciding on actionability is still a major open challenge

### 2025-06-30 paper/sections/discussion.qmd
Fix some awkward references after reformatting for AAAI

@@
 Additionally, CT requires mutability constraints for the features considered by the model.
-Even if all immutable features are protected, there may exist proxies that are mutable, and hence should not be protected, but preserve sufficient information about the principals to hinder the protections.
+Even if all immutable features are protected, there may exist proxies that are mutable, and hence should not be protected, but preserve sufficient information about the principals to hinder these protections.
 Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.

### 2025-06-30 paper/sections/experiments.qmd
Fix some awkward references after reformatting for AAAI

@@
 Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*).
-These datasets are generated using the library of [@altmeyer2023explaining] and we present them in the supplementary appendix.
+These datasets are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix.
 Next, we have four real-world binary tabular datasets from the domain of economics: *Adult* (a.k.a.
-Census data) of [@becker1996adult2], California housing (*CH*) of [@pace1997sparse], Default of Credit Card Clients (*Cred*) of [@yeh2016default], and Give Me Some Credit (*GMSC*) of [@kaggle2011give2].
-Finally, for the convenience of illustration, we use of the 10-class *MNIST* vision dataset [@lecun1998mnist].
+Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give2.
+Finally, for the convenience of illustration, we use the 10-class *MNIST* dataset [@lecun1998mnist].
 

### 2025-06-30 paper/sections/intro.qmd
Fix some awkward references after reformatting for AAAI

@@
 The first challenge has attracted an abundance of work on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations, aiming to mitigate scenarios in which practitioners deploy opaque models and have to blindly rely on their predictions.
-On many occasions, this has happened in practice, causing harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models (see, e.g., @oneil2016weapons).
+On many occasions, this has happened in practice, causing harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models; see, e.g., @oneil2016weapons.
 Effective XAI tools can also aid in monitoring models and providing recourse, empowering people to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "loan application accepted").

### 2025-06-30 paper/sections/method.qmd
Fix some awkward references after reformatting for AAAI

@@
 
-@grathwohl2020your observed that any classifier can be re-interpreted as a joint energy-based model that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
-@grathwohl2020your show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood: $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$, where the first term can be optimized using cross-entropy as in @eq-obj.
+@grathwohl2020your observe that any classifier can be re-interpreted as a joint energy-based model that learns to discriminate output classes conditional on the observed (training) samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
+The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood: $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$, where the first term can be optimized using cross-entropy as in @eq-obj.
 To optimize $\log p_\theta(\mathbf{x})$, they minimize the contrastive divergence between the observed samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
@@
 This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation".
-Thus, we choose to work with a different distinction between CE and AE than @freiesleben2022intriguing who consider misclassification as the key distinguishing feature of AE.
-One of the key observations of this work is that we can leverage CEs during training and get adversarial examples essentially for free, which can be used to reap the aforementioned benefits of adversarial training.
+Thus, we work with a different distinction between CE and AE than @freiesleben2022intriguing who considers misclassification as the distinguishing feature of adversarial examples.
+One of the key observations of this work is that we can leverage CEs during training and get AEs essentially for free to reap the aforementioned benefits of adversarial training.
 

### 2025-07-01 paper/quarto_aaai/paper.tex
Improve formatting of equations

@@
 % This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
-\usepackage{bibentry}
+% \usepackage{bibentry}
 % END REMOVE bibentry
@@
 \newtheorem{definition}{Definition}[section]
+% For mathematical environments, you may reduce fontsize but not below 6.5 point.
+\DeclareMathSizes{10}{9}{7}{6.5}
 
@@
 train explainable (as in Def. \ref{def-explainability}) models:
-
 \begin{equation}\phantomsection\label{eq-obj}{
-\begin{split}
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
-\end{split}
+\begin{aligned}
+&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
+\end{aligned}
 }\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
@@
 explainers to generate counterfactuals of observed training samples.
-Specifically, we have:
-
-\begin{equation}\phantomsection\label{eq-div}{
+Specifically, we have: \begin{equation}\phantomsection\label{eq-div}{
 \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
@@
 functional form, but likely not the only sensible choice, is inspired by
-adversarial training:
-
-\begin{equation}\phantomsection\label{eq-adv}{
+adversarial training: \begin{equation}\phantomsection\label{eq-adv}{
 \begin{aligned}
@@
 metric used by \citet{altmeyer2024faithful},
-
 \begin{equation}\phantomsection\label{eq-impl-dist}{
@@
 }\end{equation} and introduce a novel divergence metric,
-
 \begin{equation}\phantomsection\label{eq-impl-div}{

### 2025-07-01 paper/sections/experiments.qmd
Improve formatting of equations

@@
 We rely on the metric used by @altmeyer2024faithful,
-
 $$
@@
 and introduce a novel divergence metric,
-
 $$

### 2025-07-01 paper/sections/other/main_results.qmd
ok now
getting a little stuck here

@@
 ```
+
+\FloatBarrier
 
@@
 
-Final hyperparameters used for the main results preseneted in @sec-experiments. Any hyperparameter not shown here is set to its default value (@nte-train-default).
+Final hyperparameters used for the main results presented in @sec-experiments. Any hyperparameter not shown here is set to its default value (@nte-train-default).
 
@@
 
-![The same setup, factuals, model architecture and generator as in @fig-mnist, but the model was trained with CT.](/paper/figures/mnist_mlp_vanilla.png){#fig-mnist-vanilla}
+![The same setup, factuals, model architecture and generator as in @fig-mnist, but the model was trained conventionally.](/paper/figures/mnist_mlp_vanilla.png){#fig-mnist-vanilla}
 :::

### 2025-07-01 paper/experiments/CTExperiments/src/experiment.jl

@@ -163,6 +163,10 @@ function train_val_split(data::Dataset, ce_data::CounterfactualData, val_size)
         train_test_split(ce_data; test_size=val_size, keep_class_ratio=true)
     )
 
+    # Reduce float precision:
+    Xtrain = Float32.(Xtrain)
+    Xval = Float32.(Xval)
+
     train_set = Flux.DataLoader((Xtrain, ytrain); batchsize=data.batchsize)
     val_set = if data.n_validation > 0
         Flux.DataLoader((Xval, yval); batchsize=data.batchsize)

### 2025-07-01 paper/experiments/CTExperiments/src/model_and_data.jl

@@ -182,15 +182,21 @@ Helper function to get the mutability constraints for the dataset. If `data.muta
 """
 function get_mutability(data::Dataset)
     mtblty = data.mutability
+
+    # Single string:
     if mtblty isa String
         if mtblty == "none"
             mtblty = nothing
         else
             mtblty = fill(Symbol(mtblty), input_dim(data))
         end
-    else
+    end
+
+    # Vector of strings:
+    if mtblty isa Vector{String}
         mtblty = Symbol.(mtblty)
     end
+    
     return mtblty
 end
 

### 2025-07-01 src/counterfactuals.jl

@@ -57,14 +57,13 @@ function generate!(
         callback=callback,
     )
 
-    # counterfactuals = (ce -> ce.counterfactual).(ces)                                   # get actual counterfactuals
-    advexms = (ce -> ce.search[:last_valid_ae]).(ces)                                   # get adversarial example
+    advexms = (ce -> Float32.(ce.search[:last_valid_ae])).(ces)                                   # get adversarial example
     targets = (ce -> ce.target).(ces)                                                   # get targets
-    neighbours = (ce -> find_potential_neighbours(ce, counterfactual_data, 1)).(ces)    # randomly draw a sample from the target class
+    neighbours = (ce -> Float32.(find_potential_neighbours(ce, counterfactual_data, 1))).(ces)    # randomly draw a sample from the target class
     validities = (ce -> ce.search[:converged]).(ces)
     # Extract counterfactual if converged, else use neighbour:
     counterfactuals = [
-        validities[i] ? ce.counterfactual : neighbours[i] for (i, ce) in enumerate(ces)
+    validities[i] ? Float32.(ce.counterfactual) : neighbours[i] for (i, ce) in enumerate(ces)
     ]
 
     protect_immutable!(neighbours, counterfactuals, counterfactual_data.mutability)     # adjust for mutability

### 2025-07-01 paper/sections/method.qmd

@@
 Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:
-
 $$
-\begin{split}
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
-\end{split}
+\begin{aligned}
+&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
+\end{aligned}
 $$ {#eq-obj}
@@
 Specifically, we have:
-
 $$
@@
 An intuitive functional form, but likely not the only sensible choice, is inspired by adversarial training:
-
 $$

### 2025-07-01 paper/sections/related.qmd

@@
 
-CEs have also been used to improve models in the natural language processing (NLP) domain.
-First, @wu2021polyjuice2 propose *Polyjuice*, a general-purpose counterfactual generator for language models.
-The authors empirically demonstrate that the augmentation of training data with their method improves robustness in a number of NLP tasks.
-Second, @balashankar2023improving use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%.
-Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT) that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
+CEs have also been used to improve models in the natural language processing domain.
+For example, @wu2021polyjuice2 propose *Polyjuice*, a general-purpose CE generator for language models and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks, while @luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
 

### 2025-07-01 paper/sections/other/main_results.qmd
wowo

@@
 
-tbl_mtbl = final_mutability(mtbl_dir; var=["distance"])
-df = final_table(res_dir; tbl_mtbl)
+df = final_table(res_dir)
 inputs = get_table_inputs(df, nothing; backend=Val(:latex))

### 2025-07-01 paper/experiments/CTExperiments/src/presentation.jl

@@ -610,6 +610,7 @@ function aggregate_ce_evaluation(
         byvars_must_include = byvars_must_include[[
             x in names(df) for x in byvars_must_include
         ]]
+
         df_agg = aggregate_data(df, y, byvars; byvars_must_include=byvars_must_include)
 
         if !isnothing(agg_further_vars)
@@ -626,8 +627,8 @@ function aggregate_ce_evaluation(
                 byvars = setdiff(byvars, ["objective"])
                 df_agg =
                     groupby(df_agg, byvars) |>
-                    df -> combine(df, :ratio => (y -> (mean=-(mean(y)-1)*100, 
-                    se=100*std(y)/sqrt(length(y)))) => AsTable)
+                    df -> combine(df, :ratio => (y -> (mean=-(mean(skipmissing(y))-1)*100, 
+                        se=100*std(skipmissing(y))/sqrt(length(y)))) => AsTable)
                 return df_agg
             end
 
@@ -832,7 +833,7 @@ function final_table(
     tbl_mtbl::Union{Nothing,DataFrame}=nothing,
     ce_var=["plausibility_distance_from_target", "mmd"],
     perf_var=["acc"],
-    agg_further_vars=[["run"], ["run", "lambda_energy_eval"]],
+    agg_further_vars=[["run", "lambda_energy_eval"], ["run", "lambda_energy_eval"]],
     longformat::Bool=true,
 )
     # CE:
@@ -863,7 +864,6 @@ function final_table(
         df = vcat(df, tbl_mtbl; cols=:union)
     end
 
-    println(df)
 
     # Missing:
     df = coalesce.(df, "")
@@ -871,8 +871,10 @@ function final_table(
     select!(df, :measure, Not([:measure]))
 
     if longformat
-        df.measure = combine_header.(df.measure, string.(df.lambda_energy_eval))
-        select!(df, Not(:lambda_energy_eval))
+        if "lambda_energy_eval" in names(df)
+            df.measure = combine_header.(df.measure, string.(df.lambda_energy_eval))
+            select!(df, Not(:lambda_energy_eval))
+        end
         println(df)
         df =
             DataFrames.stack(df, Not(:measure)) |>
@@ -907,7 +909,7 @@ function final_mutability(
     df_ce = DataFrame()
     for (i, y) in enumerate(var)
         df = aggregate_ce_evaluation(
[…]

### 2025-07-03 paper/sections/other/main_results.qmd
buff

@@
 
+## Final Results {#sec-app-final-results}
+
 ```{julia}
@@
 )
-for fname in ["paper/quarto_ecml/tables/main.tex","paper/preprint/tables/main.tex"]
+for fname in ["paper/quarto_aaai/tables/main.tex","paper/preprint/tables/main.tex"]
+    tabulate_results(
+        inputs; 
+        table_type=:tabular, 
+        save_name=fname, 
+        wrap_table=false, 
+        alignment=[:l,fill(:r,size(df,2)-1)...],
+        formatters=formatters
+    )
+end
+```
+
+```{julia}
+#| eval: false
+
+# Detailed table with varying levels of hyperparameter:
+df = final_table(res_dir; agg_further_vars=[["run"], ["run"]])
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+formatters = (
+    PrettyTables.ft_round(1,findall(.!contains.(names(df),"Acc"))),
+    PrettyTables.ft_round(2,findall(contains.(names(df),"Acc")))
+)
+for fname in ["paper/quarto_aaai/tables/main_detailed.tex","paper/preprint/tables/main_detailed.tex"]
+    tabulate_results(
+        inputs; 
+        table_type=:tabular, 
+        save_name=fname, 
+        wrap_table=false, 
+        alignment=[:l,fill(:r,size(df,2)-1)...],
+        formatters=formatters
+    )
+end
+```
+
+```{julia}
+#| eval: false
+
+# Standard errors with hyperparameter uncertainty marginalized out:
+df = final_table(res_dir, total_uncertainty=false)
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+formatters = (
+    PrettyTables.ft_round(1,findall(.!contains.(names(df),"Acc"))),
+    PrettyTables.ft_round(2,findall(contains.(names(df),"Acc")))
+)
+for fname in ["paper/quarto_aaai/tables/main_marginalized.tex","paper/preprint/tables/main_marginalized.tex"]
     tabulate_results(

### 2025-07-03 paper/experiments/CTExperiments/src/presentation.jl

@@ -626,7 +626,7 @@ function aggregate_ce_evaluation(
             # Computing across-fold averages and between fold standard errors for ratios:
             if ratio
                 @assert sort(unique(df_agg.objective)) == ["full", "vanilla"] "Ratio calculation only works when comparing `full` vs. `vanilla`"
-                
+
                 if total_uncertainty
                     # Include uncertainty around lambda_energy_eval in standard error.
                     # This means that standard error also reflects possibly different hyperparameter
@@ -647,6 +647,7 @@ function aggregate_ce_evaluation(
                 # Final aggregation and standard errors:
                 df_agg = DataFrames.unstack(df_agg, :objective, :mean)
                 df_agg.ratio .= df_agg.full ./ df_agg.vanilla
+
                 byvars = setdiff(byvars, ["objective"])
                 df_agg =
                     groupby(df_agg, byvars) |>
@@ -880,7 +881,7 @@ function final_table(
     df_adv_perf = aggregate_performance(res_dir; measure=perf_var, adversarial=true, bootstrap)    # adversarial
     df_perf = vcat(df_perf, df_adv_perf)
     df = vcat(df_ce, df_perf; cols=:union) |> 
-        df -> transform!(df, [:mean, :se] => ((m, s) -> [isnan(si) ? "$(round(mi, digits=2))" : "$(round(mi, digits=2))+-$(round(si, digits=2))" for (mi, si) in zip(m, s)]) => :mean) |>
+        df -> transform!(df, [:mean, :se] => ((m, s) -> [isnan(si) ? "$(round(mi, digits=2))" : PrettyTables.LatexCell("$(round(mi, digits=2))\\pm$(round(si, digits=2))") for (mi, si) in zip(m, s)]) => :mean) |>
         df -> select!(df, Not(:se)) |>
         df -> DataFrames.unstack(df, :dataset, :mean)
 
@@ -893,6 +894,9 @@ function final_table(
     # Missing:
     df = coalesce.(df, "")
     rename!(df, :variable => :measure)
+    if !("lambda_energy_eval" in names(df))
+        df.measure .= (x -> x isa PrettyTables.LatexCell ? x.data : x).(df.measure)
+    end
     select!(df, :measure, Not([:measure]))
 
     if longformat

### 2025-07-03 paper/experiments/CTExperiments/src/presentation.jl
table done

@@ -393,7 +393,7 @@ function aggregate_data(
         df_agg = combine(df_agg, y => (y -> (mean=y, se=y)) => AsTable)
         return df_agg
     else
-        df_agg = combine(df_agg, y => (y -> (mean=agg_fun(y), se=std(y)/sqrt(length(y)))) => AsTable)
+        df_agg = combine(df_agg, y => (y -> (mean=agg_fun(y), se=std(y))) => AsTable)
         return sort(df_agg)
     end
 end
@@ -652,7 +652,7 @@ function aggregate_ce_evaluation(
                 df_agg =
                     groupby(df_agg, byvars) |>
                     df -> combine(df, :ratio => (y -> (mean=-(mean(skipmissing(y))-1)*100, 
-                        se=100*std(skipmissing(y))/sqrt(length(y)))) => AsTable)
+                        se=100*std(skipmissing(y)))) => AsTable)
                 return df_agg
             end
 
@@ -860,7 +860,7 @@ function final_table(
     agg_further_vars=[["run", "lambda_energy_eval"], ["run", "lambda_energy_eval"]],
     longformat::Bool=true,
     bootstrap::Int=100,
-    total_uncertainty::Bool=true,
+    total_uncertainty::Bool=false,
 )
     # CE:
     df_ce = DataFrame()
@@ -881,7 +881,7 @@ function final_table(
     df_adv_perf = aggregate_performance(res_dir; measure=perf_var, adversarial=true, bootstrap)    # adversarial
     df_perf = vcat(df_perf, df_adv_perf)
     df = vcat(df_ce, df_perf; cols=:union) |> 
-        df -> transform!(df, [:mean, :se] => ((m, s) -> [isnan(si) ? "$(round(mi, digits=2))" : PrettyTables.LatexCell("$(round(mi, digits=2))\\pm$(round(si, digits=2))") for (mi, si) in zip(m, s)]) => :mean) |>
+        df -> transform!(df, [:mean, :se] => ((m, s) -> [isnan(si) ? "$(round(mi, digits=2))" : PrettyTables.LatexCell("$(round(mi, digits=2))\\pm$(round(2si, digits=2))") for (mi, si) in zip(m, s)]) => :mean) |>
         df -> select!(df, Not(:se)) |>
         df -> DataFrames.unstack(df, :dataset, :mean)
 

### 2025-07-03 paper/sections/other/main_results.qmd
now done

--- 2025-07-01 paper/sections/other/main_results.qmd
@@
 
+## Final Results {#sec-app-final-results}
+
+Plus/minus two standard deviations of bootstrap estimates.
+
+<!-- Standard deviation of bootstrap is the standard error. -->
+
 ```{julia}
@@
 )
-for fname in ["paper/quarto_ecml/tables/main.tex","paper/preprint/tables/main.tex"]
+for fname in ["paper/quarto_aaai/tables/main.tex","paper/preprint/tables/main.tex"]
+    tabulate_results(
+        inputs; 
+        table_type=:tabular, 
+        save_name=fname, 
+        wrap_table=false, 
+        alignment=[:l,fill(:r,size(df,2)-1)...],
+        formatters=formatters
+    )
+end
+```
+
+```{julia}
+#| eval: false
+
+# Standard errors with hyperparameter uncertainty marginalized out:
+df = final_table(res_dir, total_uncertainty=true)
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+formatters = (
+    PrettyTables.ft_round(1,findall(.!contains.(names(df),"Acc"))),
+    PrettyTables.ft_round(2,findall(contains.(names(df),"Acc")))
+)
+for fname in ["paper/quarto_aaai/tables/main_total.tex","paper/preprint/tables/main_total.tex"]
     tabulate_results(

### 2025-07-03 paper/experiments/CTExperiments/src/presentation.jl

@@ -869,6 +869,7 @@ function final_table(
             res_dir; y, agg_further_vars=agg_further_vars[i], 
             rebase=false, 
             ratio=true,
+            total_uncertainty
         )
         df_ce = vcat(df_ce, df; cols=:union)
     end

### 2025-07-07 paper/quarto_aaai/paper.tex
buff
De-emphasize training times as limitation of CT

@@
 
-\subsection{Our Proposed Objective}\label{our-proposed-objective}
+\subsection{Proposed Objective}\label{proposed-objective}
 
@@
 distinguishes various gradient-based ``generators'' or ``explainers''
-(CE algorithms) from the literature \citep{altmeyer2023explaining}.
+(CE algorithms) from the literature \citep{altmeyer2023explaining},
+including all generators used in our experiments. We refer the reader to
+the supplementary appendix for details.
 
@@
 
-\emph{CT increases the training time of models.} CT can be more
-time-consuming than conventional training regimes. Higher numbers of CEs
-per iteration positively impact the quality of solutions, but they also
-increase the amount of computations. In our experiments, grids with 270
-settings took up to four hours for more demanding datasets on a
-high-performance computing cluster with 34 2GB CPUs (see supplementary
-appendix for details). Three factors attenuate this effect: (1) CT
-inherently yields itself to parallel execution; (2) it amortizes the
-cost of CEs for the training samples; (3) it retains its value when used
-to fine-tune conventionally-trained models.
-
 \emph{Interventions on features have implications for fairness.} We
@@
 impacting the capacity of CT to fulfill its intended goal.
+
+\emph{CT increases the training time of models.} Just like adversarial
+training, CT is more resource-intensive than conventional training
+regimes. Higher numbers of CEs per iteration improve the quality of
+learned representations, but they also increase the number of
+computations. As our codebase is not performance optimized, grids of 270
+settings took up to four hours for more demanding datasets in our
+experiments, when ran on 34 2GB CPUs (see supplementary appendix). Other
+than optimization, three factors mitigate this effect: (1) CT yields
+itself to parallel execution; (2) it amortizes the cost of CEs for the
+training samples; (3) it can be used to fine-tune conventionally-trained
+models.
 

### 2025-07-09 paper/sections/other/main_results.qmd
final table

@@
 using Random
+using Serialization
 
@@
 
-df = final_table(res_dir)
+df_mtbl = aggregate_ce_evaluation(
+    mtbl_dir; 
+    ratio=true, 
+    rebase=false, 
+    y="distance", 
+    agg_further_vars=["run", "lambda_energy_eval"], 
+    total_uncertainty=false, 
+    drop_models=["linear"]
+)
+df = final_table(res_dir; tbl_mtbl=df_mtbl)
 inputs = get_table_inputs(df, nothing; backend=Val(:latex))

### 2025-07-09 paper/experiments/CTExperiments/src/presentation.jl

@@ -768,13 +768,13 @@ function format_metric(m::String)
 end
 
 function aggregate_ce_evaluation(
-    res_dir::String; y="mmd", byvars=nothing, rebase=true, ratio=true, drop_models=[], kwrgs...
+    res_dir::String; y="mmd", byvars=nothing, rebase=true, ratio=true, drop_models::Vector{String}=String[], kwrgs...
 )
     byvars = gather_byvars(byvars, "data")
     eval_grids, _ = final_results(res_dir; drop_models)
     df = DataFrame()
     for (i, cfg) in enumerate(eval_grids)
-        @info "Evaluating model+data $i ..."
+        @info "Evaluating model+data $i/$(length(eval_grids)) ..."
         df_i = aggregate_ce_evaluation(cfg; y, byvars, rebase, ratio, kwrgs...)
         df = vcat(df, df_i)
     end
@@ -796,7 +796,7 @@ end
 global allowed_perf_measures = Dict("acc" => accuracy, "f1" => multiclass_f1score)
 
 function aggregate_performance(
-    res_dir::String; measure::Vector=["acc"], adversarial::Bool=false, bootstrap::Union{Nothing,Int}=nothing, drop_models=[],
+    res_dir::String; measure::Vector=["acc"], adversarial::Bool=false, bootstrap::Union{Nothing,Int}=nothing, drop_models::Vector{String}=String[],
 )
 
     # Get measures:
@@ -815,7 +815,7 @@ function aggregate_performance(
     return df
 end
 
-function final_results(res_dir::String; drop_models::Vector{String}=[])
+function final_results(res_dir::String; drop_models::Vector{String}=String[])
 
     # Get model and data directories:
     model_dirs = joinpath.(res_dir, readdir(res_dir)) |> x -> x[isdir.(x)]
@@ -845,8 +845,8 @@ function final_table(
     agg_further_vars=[["run", "lambda_energy_eval"], ["run", "lambda_energy_eval"]],
     longformat::Bool=true,
     bootstrap::Int=100,
-    total_uncertainty::Bool=false,
-    drop_models=[]
+    total_uncertainty::Bool=true,
+    drop_models::Vector{String}=String[]
 )
     # CE:
     df_ce = DataFrame()
@@ -864,22 +864,25 @@ function final_table(
     # Missing:
     df_ce = coalesce.(df_ce, "(agg.)")
 
-    display(df_ce)
-
[…]

### 2025-07-09 paper/sections/other/main_results.qmd
buuff

@@
 df = final_table(res_dir; tbl_mtbl=df_mtbl)
+Serialization.serialize("paper/experiments/output/final_table.jls", df)
+```
+
+```{julia}
+#| eval: false
+
+df = Serialization.deserialize("paper/experiments/output/final_table.jls")
 inputs = get_table_inputs(df, nothing; backend=Val(:latex))

### 2025-07-09 paper/sections/experiments.qmd
all done here

@@
 
+<!-- Final Results -->
+<!---->
+<!-- Columns 1-3 show mean outcome +- TWO (2) bootstrapped standard errors. -->
+<!---->
+<!-- - IP: significant and in many cases substantial reduction in implausibility, except for OL and Adult (both insfignficant) -->
+<!-- - IP*: significant and in many cases substantial reduction (up to 90%) in implausibility, except for MNIST and OL. -->
+<!-- - Cost: significant and in some cases substantial reduction, except for Adult, MNIST (both insignificant) and Credit (significant increase in costs). -->
+
 ::: {#tbl-main}

### 2025-07-10 paper/sections/experiments.qmd
addressing cynthias comment

--- 2025-06-30 paper/sections/experiments.qmd
@@
 
-We seek to answer the following three research questions:
+In our experiments we seek to answer the following three research questions:
 
-1. To what extent does the counterfactual training objective as it is defined in @eq-obj induce models to learn plausible explanations?
-2. To what extent does CT produce more favorable AR outcomes in the presence of actionability constraints?
-3. What are the effects of hyperparameter selection on CT?
+1. To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
+2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+3. What are the effects of hyperparameter selection wrt. the CT objective?
 
@@
 \ref{def-explainability}).
-Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
+To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
 To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.
 For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$.
-We rely on the metric used by @altmeyer2024faithful,
+We rely on the distance-based metric used in @altmeyer2024faithful,
 
@@
 $$ {#eq-impl-div}
-where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is the unbiased estimate of the squared population maximum mean discrepancy, proposed by @gretton2012kernel.
-The metric in @eq-impl-div is equal to zero if and only if the two distributions are exactly the same, $\mathbf{X}^\prime=\mathbf{X}^+$.
+where $\mathbf{X}^\prime$ denotes a collection of counterfactuals and $\text{MMD}(\cdot)$ is an unbiased estimate of the squared population maximum mean discrepancy [@gretton2012kernel].
+The metric in @eq-impl-div is equal to zero iff the two distributions are the same, $\mathbf{X}^\prime=\mathbf{X}^+$.
 
-In addition to cost and plausibility, we compute other standard metrics to evaluate counterfactuals such as validity and redundancy.
+In addition to cost and plausibility, we compute other standard metrics to evaluate counterfactuals including validity and redundancy.
 Finally, we also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].
@@
 Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*).
-These datasets are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix.
+These datasets are generated using the library of [@altmeyer2023explaining] and we present them in the supplementary appendix.
 Next, we have four real-world binary tabular datasets from the domain of economics: *Adult* (a.k.a.
-Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give2.
-Finally, for the convenience of illustration, we use the 10-class *MNIST* dataset [@lecun1998mnist].
+Census data) of [@becker1996adult2], California housing (*CH*) of [@pace1997sparse], Default of Credit Card Clients (*Cred*) of [@yeh2016default], and Give Me Some Credit (*GMSC*) of [@kaggle2011give2].
+Finally, for the convenience of illustration, we use of the 10-class *MNIST* vision dataset [@lecun1998mnist].
 
 To assess CT, we investigate the improvements in performance metrics when using it on top of a weak baseline (BL): a multilayer perceptron (*MLP*).
-This is the best way to get a clear picture of the effectiveness of CT, and it is consistent with evaluation practices in the related literature [@goodfellow2014explaining;@ross2021learning;@teney2020learning].
+This is the best way to get a clear picture of the effectiveness of CT, and it is consistent with how assessment is done in the related literature [@goodfellow2014explaining;@ross2021learning;@teney2020learning].
 
@@
 
-### Plausibility. {#sec-plaus}
+### Plausibility {#sec-plaus}
 
@@
 For *OL* the reduction is negative, consistent with the distance-based metric.
-*MNIST* is the only dataset for which the two metrics disagree.
-However, upon visual inspection of the image counterfactuals (available in the supplementary appendix), we find that CT clearly improves plausibility.
+*MNIST* is the only dataset for which the distance and divergence metrics disagree.
+Upon visual inspection of the image counterfactuals we find that CT clearly improves plausibility (see supplementary appendix for images).
 
-### Predictive Performance. {#sec-pred}
+### Predictive Performance {#sec-pred}
 
@@
 When looking at robust test accuracies (Acc.$^*$) for these datasets in particular, we find that CT strongly outperforms the baseline.
-In fact, we observe that CT improves adversarial robustness on all datasets.
+In fact, we find that CT improves adversarial robustness on all datasets.
 
@@
 
-### Actionability. {#sec-act}
+### Actionability {#sec-act}
 
-In @sec-method, we show that our method of encoding mutability constraints decreases sensitivity to immutable features for linear models, tilting the decision boundary in favour of mutable features.
-For constraints binding at test time, this leads to shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals.
-To extend this to the non-linear case, we test the effect of imposing mutability constraints for the synthetic datasets using the same evaluation scheme as above.
+In @sec-method, we show that our proposed way for encoding mutability constraints leads to lower classifier sensitivity wrt.
+immutable features for linear models, tilting the decision boundary in favour of mutable features instead.
+For binding constraints at test time, this leads to shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals.
+To extend this to the non-linear case, we test the effect of imposing mutability constraints empirically for our synthetic data using the same evaluation scheme as above.
 The final row in @tbl-main reports the average reduction in costs for CT compared to the "vanilla" baseline, when imposing that either the first or the second feature is immutable.
@@
 
-### Hyperparameter settings. {#sec-hyperparameters}
-We test the impact of three types of hyperparameters. Here we focus on the highlights; full results are available in the supplementary appendix.  
-`\indent`{=latex} First, we note that CT is highly sensitive to the choice of a CE generator and its hyperparameters but (1) there are manageable patterns, and (2) we can usually identify settings that improve either plausibility or cost, and often both of them at the same time.
+### Impact of hyperparameter settings. {#sec-hyperparameters}
+We test the impact of three types of hyperparameters; our complete results are in the supplementary appendix.  
+`\indent`{=latex} We note that CT is highly sensitive to the choice of a CE generator and its hyperparameters but (a) there are manageable patterns and (b) we can typically identify settings that improve either plausibility or cost, and commonly both of them at the same time.
 For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful].
@@
 On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes.
-In fact, $\tau=0.5$ generally leads to optimal results as it is associated with high proportions of mature counterfactuals.
-`\indent`{=latex} Second, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high.
+In fact, the choice of $\tau=0.5$ generally leads to optimal results because it is associated with high proportions of mature counterfactuals.
+`\indent`{=latex} The strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high.
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
 While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
-`\indent`{=latex} Third, the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
-A higher number of training epochs is also beneficial.
+`\indent`{=latex} The effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
+We also confirm that a higher number of training epochs is beneficial.
 Interestingly, we observed desired improvements when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process.

### 2025-07-10 paper/quarto_aaai/paper.tex

@@
 % This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
-% \usepackage{bibentry}
+\usepackage{bibentry}
 % END REMOVE bibentry
@@
 \newtheorem{definition}{Definition}[section]
-% For mathematical environments, you may reduce fontsize but not below 6.5 point.
-\DeclareMathSizes{10}{9}{7}{6.5}
+\usepackage{placeins}
+\usepackage{siunitx}
+\sisetup{uncertainty-mode = separate}
 
@@
 \item
-  To what extent does the counterfactual training objective as it is
-  defined in Equation~\ref{eq-obj} induce models to learn plausible
-  explanations?
+  To what extent does the CT objective in Equation 1 induce models to
+  learn plausible explanations?
 \item
-  To what extent does CT produce more favorable AR outcomes in the
-  presence of actionability constraints?
+  To what extent does the CT objective produce more favorable
+  algorithmic recourse outcomes in the presence of actionability
+  constraints?
 \item
-  What are the effects of hyperparameter selection on CT?
+  What are the effects of hyperparameter selection wrt. the CT
+  objective?
 \end{enumerate}

### 2025-07-10 paper/sections/conclusion.qmd

@@
-<!-- # Conclusion {#sec-conclusion} -->
-To conclude, state-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans and existing post-hoc explainability approaches cannot guarantee that the explanations agree with the model's learned representation of data.
+To conclude, state-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans.
+Existing explainability solutions cannot guarantee that explanations agree with the model's learned representation of data.
 As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models.
-Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients.
-Through extensive experiments we demonstrate that CT satisfies its objective while promoting robustness and preserving the predictive performance of the models.
+Our approach leads to explanations that are both plausible (compliant with the underlying data-generating process) and actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients.
+Through extensive experiments, we demonstrated that CT satisfies its objective while promoting robustness and preserving the predictive performance of models.
 It can also be used to fine-tune conventionally-trained models and achieve similar gains.
-Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.
+Thus, our work showcases that it is practical to simultaneously improve models and their explanations.

### 2025-07-10 paper/sections/discussion.qmd

@@
 As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, these advantages come at the cost of two important limitations.
-
-*CT increases the training time of models.* CT can be more time-consuming than conventional training regimes.
-Higher numbers of CEs per iteration positively impact the quality of solutions, but they also increase the amount of computations.
-In our experiments, grids with 270 settings took up to four hours for more demanding datasets on a high-performance computing cluster with 34 2GB CPUs (see supplementary appendix for details).
-Three factors attenuate this effect: (1) CT inherently yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; (3) it retains its value when used to fine-tune conventionally-trained models.
 
@@
 Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.
+
+*CT increases the training time of models.* Just like adversarial training, CT is more resource-intensive than conventional training regimes.
+Higher numbers of CEs per iteration improve the quality of learned representations, but they also increase the number of computations.
+As our codebase is not performance optimized, grids of 270 settings took up to four hours for more demanding datasets in our experiments, when ran on 34 2GB CPUs (see supplementary appendix).
+Other than optimization, three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; (3) it can be used to fine-tune conventionally-trained models.
 

### 2025-07-10 paper/sections/experiments.qmd

--- 2025-07-09 paper/sections/experiments.qmd
@@
 
-1. To what extent does the counterfactual training objective as it is defined in @eq-obj induce models to learn plausible explanations?
-2. To what extent does CT produce more favorable AR outcomes in the presence of actionability constraints?
-3. What are the effects of hyperparameter selection on CT?
+1. To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
+2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+3. What are the effects of hyperparameter selection wrt. the CT objective?
 

### 2025-07-10 paper/sections/method.qmd

@@
 
-## Our Proposed Objective
+## Proposed Objective
 
@@
 In broad terms, searching for CEs using gradient descent entails optimizing some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$.
-The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" (CE algorithms) from the literature [@altmeyer2023explaining].
+The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" (CE algorithms) from the literature [@altmeyer2023explaining], including all generators used in our experiments.
+We refer the reader to the supplementary appendix for details.
 

### 2025-07-10 paper/quarto_aaai/paper.tex
results section updated

@@
 model to instead seek plausibility with respect to the remaining
-features. This in turn results in lower overall sensitivity to immutable
-features, which we demonstrate empirically for different classifiers in
+features. This can be expected to result in lower overall sensitivity to
+immutable features, which we investigate empirically in
 Section~\ref{sec-experiments}. Under certain conditions, this result
@@
 \end{proposition}
-
-\noindent Proposition \ref{prp-mtblty} assumes independent features,
-which raises a concern about the effect of protecting immutable features
-in the presence of proxies. We address it in
-Section~\ref{sec-discussion}.
-
-\subsection{Example: Prediction of Credit
-Defaults}\label{example-prediction-of-credit-defaults}
-
-Suppose we are interested in predicting the likelihood that loan
-applicants default on their credit. We have access to historical data on
-previous loan takers comprised of a binary outcome variable
-\(y\in\{1=\text{default},2=\text{no default}\}\) with two input
-features: (1) the subjects' \emph{age}, which we define as immutable,
-and (2) the subjects' existing level of \emph{debt}, which we define as
-mutable. We have simulated this scenario using synthetic data with
-independent \emph{age} and \emph{debt} features, and Gaussian
-class-conditional densities in Figure~\ref{fig-poc}. The four panels
-show the outcomes for different training procedures using the same model
-architecture (a linear classifier). In panels (a) and (c) we have
-trained the models conventionally, while in panels (b) and (d) we have
-applied CT.
-
-In all four cases, all counterfactuals (stars) are valid---they have
-crossed the decision boundary (green) into the target class---but their
-quality differs. In panel (a), they are not plausible: they do not
-comply with the distribution of the factuals in \(y^+\) to the point
-where they form a clearly discernible cluster. In panel (b), they are
-highly plausible, meeting the first objective of Def.
-\ref{def-explainability}. In panel (c), the CEs involve substantial
-reductions in \emph{debt} for younger applicants. By comparison,
-counterfactual paths are shorter on average in panel (d) where we have
-protected the immutable \emph{age} as described in
-Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
-to \emph{age}, recommendations with respect to \emph{debt} are much more
-homogenous and do not unfairly punish younger individuals. These
-counterfactuals are also plausible with respect to the mutable feature.
-Thus, we consider the model in panel (d) as the most explainable
-according to Def. \ref{def-explainability}.
-
-\begin{figure*}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
-
-}
-
-\caption{\label{fig-poc}Illustration of how CT improves model
-explainability: (a) conventional training, all mutable; (b) CT, all
-mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
-immutable. The linear decision boundary is shown in green along with
-training data colored according to ground-truth labels: \(y^-=1\) (blue)
-and \(y^+=2\) (orange). Stars indicate counterfactuals in the target
-class.}
-
-\end{figure*}%
 
@@
 \ref{def-explainability}). Thus, we focus primarily on the plausibility
-and cost of faithfully generated counterfactuals at test time. To
-measure the cost, we follow the standard convention of using distances
-(\(\ell_1\)-norm) between factuals and counterfactuals as a proxy. For
-plausibility, we assess how similar CEs are to the observed samples in
-the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
-metric used by \citet{altmeyer2024faithful},
+and cost of faithfully generated counterfactuals at test
+time\footnote{Additional metrics such as validity and redundancy are
+  reported in the appendix.}. To measure the cost, we follow the
+standard convention of using distances (\(\ell_1\)-norm) between
+factuals and counterfactuals as a proxy. For plausibility, we assess how
+similar CEs are to the observed samples in the target domain,
+\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
+\citet{altmeyer2024faithful},
 \begin{equation}\phantomsection\label{eq-impl-dist}{
@@
 
-In addition to cost and plausibility, we compute other standard metrics
-to evaluate counterfactuals such as validity and redundancy. Finally, we
-also assess the predictive performance of models using standard metrics,
-including robust accuracy estimated on adversarially perturbed data
-using FGSM \citep{goodfellow2014explaining}.
-
-We run the experiments with three gradient-based generators:
-\emph{Generic} of \citet{wachter2017counterfactual} as a simple baseline
-approach, \emph{REVISE} \citep{joshi2019realistic} that aims to generate
-plausible counterfactuals using a surrogate Variational Autoencoder
-(VAE), and \emph{ECCo}---the generator of \citet{altmeyer2024faithful}
-but without the conformal prediction component---as a method that
-directly targets both faithfulness and plausibility of the
-counterfactuals.
+We also assess the predictive performance of models using standard
+metrics, including robust accuracy estimated on adversarially perturbed
+data using FGSM \citep{goodfellow2014explaining}.
+
+We run experiments with three gradient-based generators: \emph{Generic}
+of \citet{wachter2017counterfactual} as a simple baseline approach,
+\emph{REVISE} \citep{joshi2019realistic} that aims to generate plausible
+counterfactuals using a surrogate Variational Autoencoder (VAE), and
+\emph{ECCo}---the generator of \citet{altmeyer2024faithful} but without
+the conformal prediction component---as a method that directly targets
+both faithfulness and plausibility of the counterfactuals.
 
@@
 
+Our main quantitative results for the \emph{MLP} are summarised in
+Table~\ref{tbl-main}, which shows average outcomes along with
+bootstrapped standard errors for key metrics. The following example
+illustrates how to read Table~\ref{tbl-main}.
+
+\begin{figure*}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
+
+}
+
+\caption{\label{fig-poc}Illustration of how CT improves model
+explainability: (a) conventional training, all mutable; (b) CT, all
+mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
+immutable. The linear decision boundary is shown in green along with
+training data colored according to ground-truth labels: \(y^-=1\) (blue)
+and \(y^+=2\) (orange). Stars indicate counterfactuals in the target
+class.}
+
+\end{figure*}%
+
+\subsubsection{Synthetic Example: Prediction of Credit
+Defaults}\label{synthetic-example-prediction-of-credit-defaults}
+
+Figure~\ref{fig-poc} presents results for a linear classifier fitted to
+\emph{LS}, which complies with the data assumptions in Proposition
+\ref{prp-mtblty}. The four panels show the outcomes for different
+training procedures: in panels (a) and (c) we have trained the models
+conventionally, while in panels (b) and (d) we have applied CT. For
+illustrative purposes, suppose the first feature represents \emph{debt}
+(mutable) and the second feature represents \emph{age} (immutable) of
+loan applicants seeking counterfactual explanations for moving to the
+target class: loan provided (orange).
+
+In all four cases, it is possible to generate valid counterfactuals
+(stars) for unsuccessful applicants (blue). They cross the decision
+boundary (green) into the target class, but their quality differs. In
+panel (a), they are not plausible: they do not comply with the
+distribution of the factuals in \(y^+\) to the point where they form a
+clearly discernible cluster. In panel (b), they are highly plausible,
+meeting the first objective of Def. \ref{def-explainability}. This
+difference in outcomes is quantified for the non-linear MLP in the first
+two columns of Table~\ref{tbl-main} as the \(\%\)-reduction in
+implausibility: it is substantial and statistically significant for
+\emph{LS} across both metrics, the distance-based \(IP\) (29\%) and
+divergence-based \(IP^{*}\) (55\%).
+
+In panel (c) of Figure~\ref{fig-poc}, the CEs involve substantial
+reductions in \emph{debt} for younger applicants. By comparison,
+counterfactual paths are shorter on average in panel (d) where we have
+protected the immutable \emph{age} as described in
+Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
+to \emph{age}, recommendations with respect to \emph{debt} are much more
+homogenous and do not unfairly punish younger individuals. These
+counterfactuals are also plausible with respect to the mutable feature,
+despite requiring smaller debt reductions on average, resulting in
+smaller costs to individuals. This result is quantified for the
+non-linear case in column three of Table~\ref{tbl-main}, which shows the
+\(\%\)-reduction in costs averaged across valid counterfactuals. Once
+again, the impact of CT is statistically significant and substantial
+(14\%). Thus, we consider the model in panel (d) as the most explainable
+according to Def. \ref{def-explainability}. In the following, we present
+the results for all remaining datasets.
+
 \subsubsection{Plausibility.}\label{sec-plaus}
 
-Table~\ref{tbl-main} presents our main empirical findings. For all
-datasets except \emph{OL} and across all test settings, the average
-distance of CEs from observed samples in the target class is reduced,
-indicating improved plausibility. The magnitude of improvements varies.
-For the simple synthetic datasets, distance reductions range from around
-20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
-real-world tabular datasets, improvements tend to be smaller but still
-substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
-7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For the vision
-dataset (\emph{MNIST}), distances are reduced by up to 9\%. The results
-for our proposed divergence metric are qualitatively similar, but
-generally even more pronounced: for the \emph{Circ} dataset,
-implausibility is reduced by almost 94\% to virtually zero as we
-verified by the absolute outcome. Improvements for other datasets range
-from 28\% (\emph{Moon}) up to 78\% (\emph{GMSC}). For \emph{OL} the
-reduction is negative, consistent with the distance-based metric.
-\emph{MNIST} is the only dataset for which the two metrics disagree.
-However, upon visual inspection of the image counterfactuals (available
-in the supplementary appendix), we find that CT clearly improves
-plausibility.
+We find that CT generally leads to substantial and statistically
+significant improvements in plausibility: average reductions in \(IP\)
+range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
+for the real-world tabular datasets they are around 12\% for both
+\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
+\emph{Adult} and \emph{OL} we find no significant impact of CT on
+\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
+statistically significant, although the average degree of uncertainty is
+higher than for \(IP\): average reductions range from around 20\%
+(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
+for \emph{OL} and \emph{MNIST} are statistically insignificant and, in
+the latter case, do not align with qualitative findings.
+
+\subsubsection{Actionability.}\label{sec-act}
+
+We also find that CT can reduce sensitivity to immutable, protected
+features and thus lead to less costly counterfactual outcomes as
+illustrated in Figure~\ref{fig-poc}. In column three of
+Table~\ref{tbl-main}, we impose mutability constraints on selected
+features and compute the reduction in average costs of counterfactuals
+associated with CT compared to the weak baseline: for synthetic
+datasets, we always protect the first feature; for all real-world
+tabular datasets we could identify and protect an \emph{age} variable;
+for \emph{MNIST}, we protect the five top and bottom rows of digits.
+Reductions in costs are overwhelmingly positive and significant of up to
+nearly 60\% for \emph{GMSC}. While the estimated cost reductions for
+\emph{Adult} and \emph{MNIST} are not significant, FIGURE demonstrates
+that CT does have the expected effect: sensitivity to protected features
+as proxied by integrated gradients is drastically reduced\footnote{Additional
+  findings for integrated gradients are reported in the appendix.}. In
+the case of \emph{Cred}, average costs increase, likely because any
+potential benefits from protecting the \emph{age} are outweighed by the
+increase in costs required for greater plausibility.
 
@@
 
-\caption{\label{tbl-main}Key performance metrics across all datasets
-(column 1). \textbf{Plausibility}: Columns 2-6 show the percentage
-reduction in implausibility (\(\text{IP}\)) for varying degrees of the
-energy penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test
-time; column 7 shows the reduction in \(\text{IP}^*\) (\(\text{MMD}\)),
-aggregated across all test specifications. \textbf{Accuracy} (columns
-8-11): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
-and the baseline (BL). \textbf{Actionability} (column 12): average
-reduction in costs when imposing mutability constraints reported for the
-four datasets for which we could identify meaningful features to
-protect.}
+\caption{\label{tbl-main}Key performance metrics and bootstrapped
+standard errors for all datasets. \textbf{Plausibility} (columns 1-2):
+percentage reduction in implausibility for \(IP\) and \(IP^*\),
+respectively; \textbf{Actionability} (column 3): percentage reduction in
+costs with protected features. \textbf{Accuracy} (columns 4-7): test
+accuracies and robust accuracies (\(\text{Acc}^*\)) for CT and the
+baseline (BL). Counterfactual outcomes in columns 1-3 are aggregated
+across bootstrap samples and varying degrees of the energy penalty
+\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Standard
+errors for accuracy are bootstrapped from the test set.}
 
@@
 \end{table*}%
-
-\subsubsection{Actionability.}\label{sec-act}
-
-In Section~\ref{sec-method}, we show that our method of encoding
-mutability constraints decreases sensitivity to immutable features for
-linear models, tilting the decision boundary in favour of mutable
-features. For constraints binding at test time, this leads to shorter
-counterfactual paths and hence smaller average costs (\(\ell_1\)-norm)
-to individuals. To extend this to the non-linear case, we test the
-effect of imposing mutability constraints for the synthetic datasets
-using the same evaluation scheme as above. The final row in
-Table~\ref{tbl-main} reports the average reduction in costs for CT
-compared to the ``vanilla'' baseline, when imposing that either the
-first or the second feature is immutable. In all cases, costs are
-reduced substantially, indicating that classifiers trained with CT are
-indeed more sensitive to mutable features.
 

### 2025-07-10 paper/sections/experiments.qmd
results section updated

@@
 \ref{def-explainability}).
-Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time.
+Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time^[Additional metrics such as validity and redundancy are reported in the appendix.].
 To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.
@@
 
-In addition to cost and plausibility, we compute other standard metrics to evaluate counterfactuals such as validity and redundancy.
-Finally, we also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].
+We also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].
 
-We run the experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.
+We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.
 
@@
 
+Our main quantitative results for the *MLP* are summarised in @tbl-main, which shows average outcomes along with bootstrapped standard errors for key metrics.
+The following example illustrates how to read @tbl-main.
+
+![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable.
+The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange).
+Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}
+
+### Synthetic Example: Prediction of Credit Defaults
+
+@fig-poc presents results for a linear classifier fitted to *LS*, which complies with the data assumptions in Proposition \ref{prp-mtblty}.
+The four panels show the outcomes for different training procedures: in panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT.
+For illustrative purposes, suppose the first feature represents *debt* (mutable) and the second feature represents *age* (immutable) of loan applicants seeking counterfactual explanations for moving to the target class: loan provided (orange).
+
+In all four cases, it is possible to generate valid counterfactuals (stars) for unsuccessful applicants (blue).
+They cross the decision boundary (green) into the target class, but their quality differs.
+In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly discernible cluster.
+In panel (b), they are highly plausible, meeting the first objective of Def.
+\ref{def-explainability}.
+This difference in outcomes is quantified for the non-linear MLP in the first two columns of @tbl-main as the $\%$-reduction in implausibility: it is substantial and statistically significant for *LS* across both metrics, the distance-based $IP$ (29%) and divergence-based $IP^{*}$ (55%).
+
+In panel (c) of @fig-poc, the CEs involve substantial reductions in *debt* for younger applicants.
+By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints.
+Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals.
+These counterfactuals are also plausible with respect to the mutable feature, despite requiring smaller debt reductions on average, resulting in smaller costs to individuals.
+This result is quantified for the non-linear case in column three of @tbl-main, which shows the $\%$-reduction in costs averaged across valid counterfactuals.
+Once again, the impact of CT is statistically significant and substantial (14%).
+Thus, we consider the model in panel (d) as the most explainable according to Def.
+\ref{def-explainability}.
+In the following, we present the results for all remaining datasets.
+
 ### Plausibility. {#sec-plaus}
 
-@tbl-main presents our main empirical findings.
-For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility.
-The magnitude of improvements varies.
-For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*).
-For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred*, and around 3% for *Adult*.
-For the vision dataset (*MNIST*), distances are reduced by up to 9%.
-The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome.
-Improvements for other datasets range from 28% (*Moon*) up to 78% (*GMSC*).
-For *OL* the reduction is negative, consistent with the distance-based metric.
-*MNIST* is the only dataset for which the two metrics disagree.
-However, upon visual inspection of the image counterfactuals (available in the supplementary appendix), we find that CT clearly improves plausibility.
+We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $IP$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $IP$.
+Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*).
+The only negative findings for *OL* and *MNIST* are statistically insignificant and, in the latter case, do not align with qualitative findings.
+
+<!-- TODO: Add figure for MNIST -->
+
+### Actionability. {#sec-act}
+
+We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as illustrated in @fig-poc.
+In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of counterfactuals associated with CT compared to the weak baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits.
+Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*.
+While the estimated cost reductions for *Adult* and *MNIST* are not significant, FIGURE demonstrates that CT does have the expected effect: sensitivity to protected features as proxied by integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.].
+In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
+
+<!-- TODO: Add figure for MNIST -->
 
@@
 
-Key performance metrics across all datasets (column 1).
-**Plausibility**: Columns 2-6 show the percentage reduction in implausibility ($\text{IP}$) for varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time; column 7 shows the reduction in $\text{IP}^*$ ($\text{MMD}$), aggregated across all test specifications.
-**Accuracy** (columns 8-11): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL).
-**Actionability** (column 12): average reduction in costs when imposing mutability constraints reported for the four datasets for which we could identify meaningful features to protect.
+Key performance metrics and bootstrapped standard errors for all datasets.
+**Plausibility** (columns 1-2): percentage reduction in implausibility for $IP$ and $IP^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features.
+**Accuracy** (columns 4-7): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL).
+Counterfactual outcomes in columns 1-3 are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time.
+Standard errors for accuracy are bootstrapped from the test set.
 :::
-
-### Actionability. {#sec-act}
-
-In @sec-method, we show that our method of encoding mutability constraints decreases sensitivity to immutable features for linear models, tilting the decision boundary in favour of mutable features.
-For constraints binding at test time, this leads to shorter counterfactual paths and hence smaller average costs ($\ell_1$-norm) to individuals.
-To extend this to the non-linear case, we test the effect of imposing mutability constraints for the synthetic datasets using the same evaluation scheme as above.
-The final row in @tbl-main reports the average reduction in costs for CT compared to the "vanilla" baseline, when imposing that either the first or the second feature is immutable.
-In all cases, costs are reduced substantially, indicating that classifiers trained with CT are indeed more sensitive to mutable features.
 

### 2025-07-10 paper/sections/method.qmd
results section updated

@@
 Intuitively, we think of this as ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features.
-This in turn results in lower overall sensitivity to immutable features, which we demonstrate empirically for different classifiers in @sec-experiments.
+This can be expected to result in lower overall sensitivity to immutable features, which we investigate empirically in @sec-experiments.
 Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]
@@
 \end{proposition}
-
-`\noindent`{=latex} Proposition \ref{prp-mtblty} assumes independent features, which raises a concern about the effect of protecting immutable features in the presence of proxies.
-We address it in @sec-discussion.
-
-## Example: Prediction of Credit Defaults
-
-Suppose we are interested in predicting the likelihood that loan applicants default on their credit.
-We have access to historical data on previous loan takers comprised of a binary outcome variable $y\in\{1=\text{default},2=\text{no default}\}$ with two input features: (1) the subjects' *age*, which we define as immutable, and (2) the subjects' existing level of *debt*, which we define as mutable.
-We have simulated this scenario using synthetic data with independent *age* and *debt* features, and Gaussian class-conditional densities in @fig-poc.
-The four panels show the outcomes for different training procedures using the same model architecture (a linear classifier).
-In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT.
-
-<!-- Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time.
-In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$.
-Stars indicate CEs in the target class generated at test time using generic gradient descent until convergence.
--->
-
-In all four cases, all counterfactuals (stars) are valid---they have crossed the decision boundary (green) into the target class---but their quality differs.
-In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly discernible cluster.
-In panel (b), they are highly plausible, meeting the first objective of Def.
-\ref{def-explainability}.
-In panel (c), the CEs involve substantial reductions in *debt* for younger applicants.
-By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints.
-Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals.
-These counterfactuals are also plausible with respect to the mutable feature.
-Thus, we consider the model in panel (d) as the most explainable according to Def.
-\ref{def-explainability}.
-
-![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable.
-The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange).
-Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

### 2025-07-10 paper/quarto_aaai/paper.tex
minor adjustments to 3.1

@@
 counterfactual outcomes, and tends to be the key feature that
-distinguishes various gradient-based ``generators'' or ``explainers''
-(CE algorithms) from the literature \citep{altmeyer2023explaining},
-including all generators used in our experiments. We refer the reader to
-the supplementary appendix for details.
-
-For our purposes, we let \(T\) vary and consider the search to be
-converged as soon as the predicted probability for the target class has
-reached a pre-determined threshold, \(\tau\):
+distinguishes various gradient-based ``generators'' or ``explainers'' in
+the literature \citep{altmeyer2023explaining}, including all generators
+used in our experiments. We refer the reader to the supplementary
+appendix for details.
+
+CT adopts gradient-based CE search during training to generate
+on-the-fly model explanations \(\mathbf{x}^\prime\) for training
+samples. We use the term \emph{nascent} to denote counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\) that are not yet valid where \(t\)
+indicates the last iteration before the label is flipped. We store and
+use these interim counterfactuals as adversarial examples. Conversely,
+we consider counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature}
+explanations if they have either exhausted all \(T\) iterations or
+converged in terms reaching a pre-specified threshold, \(\tau\), for the
+predicted probability of the target class:
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
 where \(\mathcal{S}\) is the softmax function.
-
-CT adapts gradient-based CE search already in training to generate
-\(\mathbf{x}^\prime\), the perturbed versions of the factual samples
-\(\mathbf{x}\). We use the term \emph{nascent} to denote counterfactuals
-for which the search has not converged, \(\mathbf{x}_{t\leq T}^\prime\).
-Initially, when the perturbations are small, nascent counterfactuals
-serve as AEs. Over multiple iterations, they move towards the decision
-boundary and turn into \emph{mature} counterfactuals
-\(\mathbf{x}_T^\prime\) when they cross into the target class (becoming
-CEs) or when \(t > T\).
 

### 2025-07-10 paper/sections/method.qmd
minor adjustments to 3.1

@@
 In broad terms, searching for CEs using gradient descent entails optimizing some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$.
-The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" (CE algorithms) from the literature [@altmeyer2023explaining], including all generators used in our experiments.
+The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" in the literature [@altmeyer2023explaining], including all generators used in our experiments.
 We refer the reader to the supplementary appendix for details.
 
-For our purposes, we let $T$ vary and consider the search to be converged as soon as the predicted probability for the target class has reached a pre-determined threshold, $\tau$: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.
-
-CT adapts gradient-based CE search already in training to generate $\mathbf{x}^\prime$, the perturbed versions of the factual samples $\mathbf{x}$.
-We use the term *nascent* to denote counterfactuals for which the search has not converged, $\mathbf{x}_{t\leq T}^\prime$.
-Initially, when the perturbations are small, nascent counterfactuals serve as AEs.
-Over multiple iterations, they move towards the decision boundary and turn into *mature* counterfactuals $\mathbf{x}_T^\prime$ when they cross into the target class (becoming CEs) or when $t > T$.
+CT adopts gradient-based CE search during training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for training samples.
+We use the term *nascent* to denote counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that are not yet valid where $t$ indicates the last iteration before the label is flipped.
+We store and use these interim counterfactuals as adversarial examples.
+Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged in terms reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.
 

### 2025-07-10 paper/quarto_aaai/paper.tex
MNIST image added, though stil missing IG

@@
 for \emph{OL} and \emph{MNIST} are statistically insignificant and, in
-the latter case, do not align with qualitative findings.
+the latter case, do not align with qualitative findings, which are much
+more plausible for CT (Figure~\ref{fig-mnist}).
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_body.png}}
+
+}
+
+\caption{\label{fig-mnist}Explanations for \emph{MNIST} for BL (top) and
+CT (bottom). First columns is a randomly chosen factual 0. Columns 2
+through 5 are \emph{ECCo} counterfactuals for that factual in target
+classes 1 to 4. Columns 6 trough 10 show integrated gradients averaged
+over all test images in the corresponding classes (5-9).}
+
+\end{figure}%
 
@@
 nearly 60\% for \emph{GMSC}. While the estimated cost reductions for
-\emph{Adult} and \emph{MNIST} are not significant, FIGURE demonstrates
-that CT does have the expected effect: sensitivity to protected features
-as proxied by integrated gradients is drastically reduced\footnote{Additional
-  findings for integrated gradients are reported in the appendix.}. In
-the case of \emph{Cred}, average costs increase, likely because any
-potential benefits from protecting the \emph{age} are outweighed by the
-increase in costs required for greater plausibility.
+\emph{Adult} and \emph{MNIST} are not significant,
+Figure~\ref{fig-mnist} demonstrates that CT does have the expected
+effect: sensitivity to protected features as proxied by integrated
+gradients is drastically reduced\footnote{Additional findings for
+  integrated gradients are reported in the appendix.}. In the case of
+\emph{Cred}, average costs increase, likely because any potential
+benefits from protecting the \emph{age} are outweighed by the increase
+in costs required for greater plausibility.
 

### 2025-07-10 paper/sections/experiments.qmd
MNIST image added, though stil missing IG

@@
 Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*).
-The only negative findings for *OL* and *MNIST* are statistically insignificant and, in the latter case, do not align with qualitative findings.
+The only negative findings for *OL* and *MNIST* are statistically insignificant and, in the latter case, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).
 
-<!-- TODO: Add figure for MNIST -->
+![Explanations for *MNIST* for BL (top) and CT (bottom).
+First columns is a randomly chosen factual 0.
+Columns 2 through 5 are *ECCo* counterfactuals for that factual in target classes 1 to 4.
+Columns 6 trough 10 show integrated gradients averaged over all test images in the corresponding classes (5-9).](/paper/figures/mnist_body.png){#fig-mnist}
 
@@
 Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*.
-While the estimated cost reductions for *Adult* and *MNIST* are not significant, FIGURE demonstrates that CT does have the expected effect: sensitivity to protected features as proxied by integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.].
+While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist demonstrates that CT does have the expected effect: sensitivity to protected features as proxied by integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.].
 In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
-
-<!-- TODO: Add figure for MNIST -->
 

### 2025-07-10 paper/quarto_aaai/paper.tex
adjusted caption

@@
 \caption{\label{fig-mnist}Explanations for \emph{MNIST} for BL (top) and
-CT (bottom). First columns is a randomly chosen factual 0. Columns 2
-through 5 are \emph{ECCo} counterfactuals for that factual in target
-classes 1 to 4. Columns 6 trough 10 show integrated gradients averaged
-over all test images in the corresponding classes (5-9).}
+CT (bottom). First column is a random factual 0. Columns 2 to 5 are
+corresponding \emph{ECCo} counterfactuals in target classes 1 to 4.
+Columns 6 to 10 show integrated gradients averaged over all test images
+in classes 5 to 9.}
 

### 2025-07-10 paper/sections/experiments.qmd
adjusted caption

@@
 ![Explanations for *MNIST* for BL (top) and CT (bottom).
-First columns is a randomly chosen factual 0.
-Columns 2 through 5 are *ECCo* counterfactuals for that factual in target classes 1 to 4.
-Columns 6 trough 10 show integrated gradients averaged over all test images in the corresponding classes (5-9).](/paper/figures/mnist_body.png){#fig-mnist}
+First column is a random factual 0.
+Columns 2 to 5 are corresponding *ECCo* counterfactuals in target classes 1 to 4.
+Columns 6 to 10 show integrated gradients averaged over all test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}
 

### 2025-07-10 paper/quarto_aaai/paper.tex
adjusted reference to caption

@@
 \caption{\label{fig-mnist}Explanations for \emph{MNIST} for BL (top) and
-CT (bottom). First column is a random factual 0. Columns 2 to 5 are
-corresponding \emph{ECCo} counterfactuals in target classes 1 to 4.
+CT (bottom). First column is a random factual 0 (blue). Columns 2 to 5
+are corresponding \emph{ECCo} counterfactuals in target classes 1 to 4.
 Columns 6 to 10 show integrated gradients averaged over all test images
@@
 \emph{Adult} and \emph{MNIST} are not significant,
-Figure~\ref{fig-mnist} demonstrates that CT does have the expected
-effect: sensitivity to protected features as proxied by integrated
+Figure~\ref{fig-mnist} (columns 6-10) demonstrates that CT does have the
+expected effect: sensitivity to protected features as per integrated
 gradients is drastically reduced\footnote{Additional findings for

### 2025-07-10 paper/sections/experiments.qmd

@@
 ![Explanations for *MNIST* for BL (top) and CT (bottom).
-First column is a random factual 0.
+First column is a random factual 0 (blue).
 Columns 2 to 5 are corresponding *ECCo* counterfactuals in target classes 1 to 4.
@@
 Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*.
-While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist demonstrates that CT does have the expected effect: sensitivity to protected features as proxied by integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.].
+While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.].
 In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
@@
 In fact, we observe that CT improves adversarial robustness on all datasets.
-
-<!-- Final Results -->
-<!---->
-<!-- Columns 1-3 show mean outcome +- TWO (2) bootstrapped standard errors. -->
-<!---->
-<!-- - IP: significant and in many cases substantial reduction in implausibility, except for OL and Adult (both insfignficant) -->
-<!-- - IP*: significant and in many cases substantial reduction (up to 90%) in implausibility, except for MNIST and OL. -->
-<!-- - Cost: significant and in some cases substantial reduction, except for Adult, MNIST (both insignificant) and Credit (significant increase in costs). -->
 

### 2025-07-10 paper/sections/other/main_results.qmd

@@
 output_dir = mkpath("paper/experiments/output/extra/")
-gen = CTExperiments.GeneratorParams(lambda_energy=25.0)
+gen = CTExperiments.GeneratorParams(lambda_energy=25.0)     # instead of a high penalty, you can also use e.g. 1.0 and many iterations e.g 1000
 conv = "max_iter"

### 2025-07-10 paper/quarto_aaai/paper.tex
Trim to 7 pages

@@
 \fi
-
-% REMOVE THIS: bibentry
-% This is only needed to show inline citations in the guidelines document. You should not need it and can safely delete it.
-\usepackage{bibentry}
-% END REMOVE bibentry
 \usepackage{amsfonts}
@@
 \sisetup{uncertainty-mode = separate}
+\DeclareMathSizes{10}{9}{7}{6.5}
 
@@
 
-Part of the problem is that the high degrees of freedom provide room for
-many solutions that are locally optimal with respect to narrow
-objectives \citep{wilson2020case}.\footnote{We follow the standard ML
-  convention, where ``degrees of freedom'' refer to the number of
-  parameters estimated from data.} As one example, research on the
-``lottery ticket hypothesis'' suggests that modern neural networks can
-be pruned by up to 90\% without losing predictive performance
-\citep{frankle2018lottery}. Thus, looking at the predictive performance
-alone, found solutions may seem to provide compelling explanations for
-the data, when in fact they are based on purely associative and
-semantically meaningless patterns. This poses two related challenges.
-Firstly, there is no dependable way to verify if learned representations
-correspond to meaningful, plausible explanations. Secondly, even if we
-resolve this challenge, it remains undecided how to ensure that machine
-learning models can \emph{only} learn valuable explanations.
+Part of the problem is that the high degrees of freedom---high number of
+parameters estimated from data---provide room for many solutions that
+are locally optimal with respect to narrow objectives
+\citep{wilson2020case}. As one example, research on the ``lottery ticket
+hypothesis'' suggests that modern neural networks can be pruned by up to
+90\% without losing predictive performance \citep{frankle2018lottery}.
+Thus, looking at the predictive performance alone, found solutions may
+seem to provide compelling explanations for the data, when in fact they
+are based on purely associative and semantically meaningless patterns.
+This poses two related challenges. Firstly, there is no dependable way
+to verify if learned representations correspond to meaningful, plausible
+explanations. Secondly, even if we resolve this challenge, it remains
+undecided how to ensure that machine learning models can \emph{only}
+learn valuable explanations.
 
@@
 \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
-assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
-feature \(d\). Intuitively, we think of this as ignoring implausibility
-costs with respect to immutable features, which effectively forces the
-model to instead seek plausibility with respect to the remaining
-features. This can be expected to result in lower overall sensitivity to
-immutable features, which we investigate empirically in
-Section~\ref{sec-experiments}. Under certain conditions, this result
-holds theoretically:\footnote{For the proof, see the supplementary
-  appendix.}
+assuming a point mass prior for \(p(\mathbf{x}^+)\) wrt. feature \(d\).
+Intuitively, we think of this as ignoring implausibility costs of
+immutable features, which effectively forces the model to instead seek
+plausibility through the remaining features. This can be expected to
+result in lower overall sensitivity to immutable features, which we
+investigate empirically in Section~\ref{sec-experiments}. Under certain
+conditions, this result holds theoretically; for the proof, see the
+supplementary appendix:
 
@@
 \item
-  To what extent does the CT objective produce more favorable
-  algorithmic recourse outcomes in the presence of actionability
-  constraints?
+  To what extent does CT lead to more favorable AR outcomes in the
+  presence of actionability constraints?
 \item
-  What are the effects of hyperparameter selection wrt. the CT
-  objective?
+  What are the effects of hyperparameter selection on CT?
 \end{enumerate}
@@
 
-Our key outcome of interest is improvement in explainability (Def.
-\ref{def-explainability}). Thus, we focus primarily on the plausibility
-and cost of faithfully generated counterfactuals at test
-time\footnote{Additional metrics such as validity and redundancy are
-  reported in the appendix.}. To measure the cost, we follow the
-standard convention of using distances (\(\ell_1\)-norm) between
-factuals and counterfactuals as a proxy. For plausibility, we assess how
-similar CEs are to the observed samples in the target domain,
+Our focus is the improvement in explainability (Def.
+\ref{def-explainability}). Thus, we primarily look at the plausibility
+and cost of faithfully generated counterfactuals at test time. Other
+metrics, such as validity and redundancy, are reported in the
+supplementary appendix. To measure the cost, we follow the standard
+proxy of distances (\(\ell_1\)-norm) between factuals and
+counterfactuals. For plausibility, we assess how similar CEs are to the
+observed samples in the target domain,
 \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
@@
 
-We also assess the predictive performance of models using standard
-metrics, including robust accuracy estimated on adversarially perturbed
-data using FGSM \citep{goodfellow2014explaining}.
+For predictive performance, we use standard metrics, such as robust
+accuracy estimated on adversarially perturbed data using FGSM
+\citep{goodfellow2014explaining}.
 
@@
 counterfactuals using a surrogate Variational Autoencoder (VAE), and
-\emph{ECCo}---the generator of \citet{altmeyer2024faithful} but without
-the conformal prediction component---as a method that directly targets
-both faithfulness and plausibility of the counterfactuals.
+\emph{ECCo}---the generator of \citet{altmeyer2024faithful} without the
+conformal prediction component---as a method that directly targets both
+faithfulness and plausibility of the counterfactuals.
 
@@
 clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
-moons (\emph{Moon}). These datasets are generated using the library of
+moons (\emph{Moon}). They are generated using the library of
 \citet{altmeyer2023explaining} and we present them in the supplementary
-appendix. Next, we have four real-world binary tabular datasets from the
-domain of economics: \emph{Adult} (a.k.a. Census data) of
-\citet{becker1996adult2}, California housing (\emph{CH}) of
-\citet{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
-\citet{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) from
-\citet{kaggle2011give2}. Finally, for the convenience of illustration,
-we use the 10-class \emph{MNIST} dataset \citep{lecun1998mnist}.
+appendix. Next, we have four real-world binary tabular datasets:
+\emph{Adult} (Census data) of \citet{becker1996adult2}, California
+housing (\emph{CH}) of \citet{pace1997sparse}, Default of Credit Card
+Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me Some Credit
+(\emph{GMSC}) from \citet{kaggle2011give2}. Finally, for the convenience
+of illustration, we use the 10-class \emph{MNIST}
+\citep{lecun1998mnist}.
 
@@
 
-Our main quantitative results for the \emph{MLP} are summarised in
-Table~\ref{tbl-main}, which shows average outcomes along with
-bootstrapped standard errors for key metrics. The following example
+Our main quantitative results for \emph{MLP} models are summarised in
+Table~\ref{tbl-main}, which presents average outcomes along with
+bootstrapped two standard errors. The following example motivates CT and
 illustrates how to read Table~\ref{tbl-main}.
@@
 
-\subsubsection{Synthetic Example: Prediction of Credit
-Defaults}\label{synthetic-example-prediction-of-credit-defaults}
+\subsubsection{Synthetic Example: Prediction of Credit Card
+Defaults.}\label{synthetic-example-prediction-of-credit-card-defaults.}
 
 Figure~\ref{fig-poc} presents results for a linear classifier fitted to
-\emph{LS}, which complies with the data assumptions in Proposition
+\emph{LS} that complies with the data assumptions in Proposition
 \ref{prp-mtblty}. The four panels show the outcomes for different
@@
 to \emph{age}, recommendations with respect to \emph{debt} are much more
-homogenous and do not unfairly punish younger individuals. These
-counterfactuals are also plausible with respect to the mutable feature,
-despite requiring smaller debt reductions on average, resulting in
-smaller costs to individuals. This result is quantified for the
-non-linear case in column three of Table~\ref{tbl-main}, which shows the
-\(\%\)-reduction in costs averaged across valid counterfactuals. Once
-again, the impact of CT is statistically significant and substantial
-(14\%). Thus, we consider the model in panel (d) as the most explainable
-according to Def. \ref{def-explainability}. In the following, we present
-the results for all remaining datasets.
+homogenous and do not unfairly punish younger individuals. These CEs are
+also plausible with respect to the mutable feature, despite requiring
+smaller debt reductions on average, resulting in smaller costs to
+individuals. This result is quantified for the non-linear case in column
+three of Table~\ref{tbl-main}, which shows the \(\%\)-reduction in costs
+averaged across valid counterfactuals. Once again, the impact of CT is
+statistically significant and substantial (14\%). Thus, we consider the
+model in panel (d) as the most explainable according to Def.
+\ref{def-explainability}. Next, we present the results for all remaining
+datasets.
 
@@
 (\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
-for \emph{OL} and \emph{MNIST} are statistically insignificant and, in
-the latter case, do not align with qualitative findings, which are much
-more plausible for CT (Figure~\ref{fig-mnist}).
+for \emph{OL} and \emph{MNIST} are statistically insignificant and, for
+MNIST, do not align with qualitative findings, which are much more
+plausible for CT (Figure~\ref{fig-mnist}).
 
@@
 
-\caption{\label{fig-mnist}Explanations for \emph{MNIST} for BL (top) and
-CT (bottom). First column is a random factual 0 (blue). Columns 2 to 5
-are corresponding \emph{ECCo} counterfactuals in target classes 1 to 4.
-Columns 6 to 10 show integrated gradients averaged over all test images
-in classes 5 to 9.}
+\caption{\label{fig-mnist}Sample explanations for \emph{MNIST} for BL
+(top) and CT (bottom). First column is a random factual 0 (blue).
+Columns 2 to 5 are corresponding \emph{ECCo} counterfactuals in target
+classes 1 to 4. Columns 6 to 10 show integrated gradients averaged over
+all test images in classes 5 to 9.}
 
@@
 We also find that CT can reduce sensitivity to immutable, protected
-features and thus lead to less costly counterfactual outcomes as
-illustrated in Figure~\ref{fig-poc}. In column three of
-Table~\ref{tbl-main}, we impose mutability constraints on selected
-features and compute the reduction in average costs of counterfactuals
-associated with CT compared to the weak baseline: for synthetic
-datasets, we always protect the first feature; for all real-world
-tabular datasets we could identify and protect an \emph{age} variable;
-for \emph{MNIST}, we protect the five top and bottom rows of digits.
-Reductions in costs are overwhelmingly positive and significant of up to
-nearly 60\% for \emph{GMSC}. While the estimated cost reductions for
-\emph{Adult} and \emph{MNIST} are not significant,
-Figure~\ref{fig-mnist} (columns 6-10) demonstrates that CT does have the
-expected effect: sensitivity to protected features as per integrated
-gradients is drastically reduced\footnote{Additional findings for
-  integrated gradients are reported in the appendix.}. In the case of
-\emph{Cred}, average costs increase, likely because any potential
-benefits from protecting the \emph{age} are outweighed by the increase
-in costs required for greater plausibility.
+features and thus lead to less costly counterfactual outcomes as shown
+in Figure~\ref{fig-poc}. In column three of Table~\ref{tbl-main}, we
+impose mutability constraints on selected features and compute the
+reduction in average costs of CEs associated with CT compared to the
+baseline: for synthetic datasets, we always protect the first feature;
+for all real-world tabular datasets we could identify and protect an
+\emph{age} variable; for \emph{MNIST}, we protect the five top and
+bottom rows of digits. Reductions in costs are overwhelmingly positive
+and significant of up to nearly 60\% for \emph{GMSC}. While the
+estimated cost reductions for \emph{Adult} and \emph{MNIST} are not
+significant, Figure~\ref{fig-mnist} (columns 6-10) demonstrates that CT
+does have the expected effect: sensitivity to protected features as per
+integrated gradients is drastically reduced; details of this experiment
+are reported in the supplementary appendix. In the case of \emph{Cred},
+average costs increase, likely because any potential benefits from
+protecting the \emph{age} are outweighed by the increase in costs
+required for greater plausibility.
 
@@
 \caption{\label{tbl-main}Key performance metrics and bootstrapped
-standard errors for all datasets. \textbf{Plausibility} (columns 1-2):
-percentage reduction in implausibility for \(IP\) and \(IP^*\),
-respectively; \textbf{Actionability} (column 3): percentage reduction in
-costs with protected features. \textbf{Accuracy} (columns 4-7): test
-accuracies and robust accuracies (\(\text{Acc}^*\)) for CT and the
-baseline (BL). Counterfactual outcomes in columns 1-3 are aggregated
-across bootstrap samples and varying degrees of the energy penalty
-\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Standard
-errors for accuracy are bootstrapped from the test set.}
+\emph{two} standard errors for all datasets. \textbf{Plausibility}
+(columns 1-2): percentage reduction in implausibility for \(IP\) and
+\(IP^*\), respectively; \textbf{Actionability} (column 3): percentage
+reduction in costs with protected features. \textbf{Accuracy} (columns
+4-7): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
+and the baseline (BL). Counterfactual outcomes in columns 1-3 are
+aggregated across bootstrap samples and varying degrees of the energy
+penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
+Standard errors for accuracy are bootstrapped from the test set.}
 
@@
 
-\emph{CT increases the training time of models.} Just like adversarial
-training, CT is more resource-intensive than conventional training
-regimes. Higher numbers of CEs per iteration improve the quality of
-learned representations, but they also increase the number of
-computations. As our codebase is not performance optimized, grids of 270
-settings took up to four hours for more demanding datasets in our
-experiments, when ran on 34 2GB CPUs (see supplementary appendix). Other
-than optimization, three factors mitigate this effect: (1) CT yields
-itself to parallel execution; (2) it amortizes the cost of CEs for the
-training samples; (3) it can be used to fine-tune conventionally-trained
-models.
-
-We also highlight several important directions for future research.
+\emph{CT increases the training times.} Like adversarial training, CT is
+more resource-intensive than conventional regimes. Higher numbers of CEs
+improve the quality of learned representations but they also increase
+the number of computations. As our codebase is not performance
+optimized, grids of 270 settings for the largest datasets in our
+experiments took up to four hours using 34 2GB CPUs (see supplementary
+appendix). Other than optimization, three factors mitigate this effect:
+(1) CT yields itself to parallel execution; (2) it amortizes the cost of
+CEs for the training samples; and (3) it can be used to fine-tune
+conventionally-trained models.
+
+We also highlight three important directions for future research.
 Firstly, it is an interesting challenge to extend CT beyond
@@
 \(\text{div}(\cdot)\).\\
-Finally, we believe that it is possible to substantially improve
-hyperparameter selection procedures, and thus performance. In this work,
-we have relied exclusively on grid searches for this task. Future work
-on CT could benefit from investigating more sophisticated approaches.
-Notably, our method is iterative which makes methods such as Bayesian or
-gradient-based optimization applicable \citep{bischl2023hyperparameter}.
+Finally, we believe that it is possible to considerably improve
+hyperparameter selection procedures, and thus performance. We have
+relied exclusively on grid searches, but future research could benefit
+from more sophisticated approaches.
 
@@
 predictive performance of models. It can also be used to fine-tune
-conventionally-trained models and achieve similar gains. Thus, our work
-showcases that it is practical to simultaneously improve models and
-their explanations.
+conventionally-trained models and achieve similar gains. Lastly, our
+work highlights the value of simultaneously improving models and their
+explanations.
 

### 2025-07-10 paper/sections/discussion.qmd
Trim to 7 pages

--- 2025-06-30 paper/sections/discussion.qmd
@@
 
-We also highlight several important directions for future research.
+*CT increases the training times.* Like adversarial training, CT is more resource-intensive than conventional regimes.
+Higher numbers of CEs improve the quality of learned representations but they also increase the number of computations.
+As our codebase is not performance optimized, grids of 270 settings for the largest datasets in our experiments took up to four hours using 34 2GB CPUs (see supplementary appendix).
+Other than optimization, three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.
+
+We also highlight three important directions for future research.
 Firstly, it is an interesting challenge to extend CT beyond classification settings.
@@
 We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.
-Finally, we believe that it is possible to substantially improve hyperparameter selection procedures, and thus performance.
-In this work, we have relied exclusively on grid searches for this task.
-Future work on CT could benefit from investigating more sophisticated approaches.
-Notably, our method is iterative which makes methods such as Bayesian or gradient-based optimization applicable [@bischl2023hyperparameter].
+Finally, we believe that it is possible to considerably improve hyperparameter selection procedures, and thus performance.
+We have relied exclusively on grid searches, but future research could benefit from more sophisticated approaches.

### 2025-07-10 paper/sections/method.qmd
Trim to 7 pages

@@
 
-`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ with respect to feature $d$.
-Intuitively, we think of this as ignoring implausibility costs with respect to immutable features, which effectively forces the model to instead seek plausibility with respect to the remaining features.
+`\noindent`{=latex} From a Bayesian perspective, setting $\mathbf{x}^+[d] := \mathbf{x}^\prime[d]$ can be understood as assuming a point mass prior for $p(\mathbf{x}^+)$ wrt.
+feature $d$.
+Intuitively, we think of this as ignoring implausibility costs of immutable features, which effectively forces the model to instead seek plausibility through the remaining features.
 This can be expected to result in lower overall sensitivity to immutable features, which we investigate empirically in @sec-experiments.
-Under certain conditions, this result holds theoretically:^[For the proof, see the supplementary appendix.]
+Under certain conditions, this result holds theoretically; for the proof, see the supplementary appendix:
 

### 2025-07-14 paper/sections/experiments.qmd

@@
 1. To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
-2. To what extent does the CT objective produce more favorable algorithmic recourse outcomes in the presence of actionability constraints?
-3. What are the effects of hyperparameter selection wrt. the CT objective?
+2. To what extent does CT lead to more favorable AR outcomes in the presence of actionability constraints?
+3. What are the effects of hyperparameter selection on CT?
 
@@
 
-Our key outcome of interest is improvement in explainability (Def.
+Our focus is the improvement in explainability (Def.
 \ref{def-explainability}).
-Thus, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time^[Additional metrics such as validity and redundancy are reported in the appendix.].
-To measure the cost, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy.
+Thus, we primarily look at the plausibility and cost of faithfully generated counterfactuals at test time.
+Other metrics, such as validity and redundancy, are reported in the supplementary appendix.
+To measure the cost, we follow the standard proxy of distances ($\ell_1$-norm) between factuals and counterfactuals.
 For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$.
@@
 
-We also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].
+For predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].
 
-We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful but without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.
+We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.
 
@@
 Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*).
-These datasets are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix.
-Next, we have four real-world binary tabular datasets from the domain of economics: *Adult* (a.k.a.
-Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give2.
-Finally, for the convenience of illustration, we use the 10-class *MNIST* dataset [@lecun1998mnist].
+They are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix.
+Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give2.
+Finally, for the convenience of illustration, we use the 10-class *MNIST* [@lecun1998mnist].
 
@@
 
-Our main quantitative results for the *MLP* are summarised in @tbl-main, which shows average outcomes along with bootstrapped standard errors for key metrics.
-The following example illustrates how to read @tbl-main.
+Our main quantitative results for *MLP* models are summarised in @tbl-main, which presents average outcomes along with bootstrapped two standard errors.
+The following example motivates CT and illustrates how to read @tbl-main.
 
@@
 
-### Synthetic Example: Prediction of Credit Defaults
+### Synthetic Example: Prediction of Credit Card Defaults.
 
-@fig-poc presents results for a linear classifier fitted to *LS*, which complies with the data assumptions in Proposition \ref{prp-mtblty}.
+@fig-poc presents results for a linear classifier fitted to *LS* that complies with the data assumptions in Proposition \ref{prp-mtblty}.
 The four panels show the outcomes for different training procedures: in panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT.
@@
 Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals.
-These counterfactuals are also plausible with respect to the mutable feature, despite requiring smaller debt reductions on average, resulting in smaller costs to individuals.
+These CEs are also plausible with respect to the mutable feature, despite requiring smaller debt reductions on average, resulting in smaller costs to individuals.
 This result is quantified for the non-linear case in column three of @tbl-main, which shows the $\%$-reduction in costs averaged across valid counterfactuals.
@@
 \ref{def-explainability}.
-In the following, we present the results for all remaining datasets.
+Next, we present the results for all remaining datasets.
 
@@
 Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*).
-The only negative findings for *OL* and *MNIST* are statistically insignificant and, in the latter case, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).
+The only negative findings for *OL* and *MNIST* are statistically insignificant and, for MNIST, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).
 
-![Explanations for *MNIST* for BL (top) and CT (bottom).
+![Sample explanations for *MNIST* for BL (top) and CT (bottom).
 First column is a random factual 0 (blue).
@@
 
-We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as illustrated in @fig-poc.
-In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of counterfactuals associated with CT compared to the weak baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits.
+We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as shown in @fig-poc.
+In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of CEs associated with CT compared to the baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits.
 Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*.
-While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.].
+While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix.
 In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
@@
 
-Key performance metrics and bootstrapped standard errors for all datasets.
+Key performance metrics and bootstrapped *two* standard errors for all datasets.
 **Plausibility** (columns 1-2): percentage reduction in implausibility for $IP$ and $IP^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features.

### 2025-07-14 paper/sections/intro.qmd

@@
 
-Part of the problem is that the high degrees of freedom provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].^[We follow the standard ML convention, where "degrees of freedom" refer to the number of parameters estimated from data.] As one example, research on the "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% without losing predictive performance [@frankle2018lottery].
+Part of the problem is that the high degrees of freedom---high number of parameters estimated from data---provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].
+As one example, research on the "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% without losing predictive performance [@frankle2018lottery].
 Thus, looking at the predictive performance alone, found solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative and semantically meaningless patterns.

### 2025-07-14 paper/sections/intro.qmd
all addressed except stat sign

@@
 
-Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explanatory capacity of models that, in high-level terms, we define as the extent to which valid explanations derived for a model can be deemed plausible with respect to the underlying data and global actionability constraints (we refine this notion in Def.
+Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explanatory capacity of models that, in high-level terms, we define as the extent to which valid counterfactual explanations derived for a model can be deemed plausible with respect to the underlying data and global actionability constraints (we refine this notion in Def.
 \ref{def-explainability}).
-For simplicity, we refer to models with high explanatory capacity as *explainable*.
 To the best of our knowledge, Counterfactual Training represents the first attempt to achieve more explainable models by employing counterfactual explanations already in the training phase.

### 2025-07-14 paper/quarto_aaai/paper.tex
one more render after updated quarto template

@@
 \sisetup{uncertainty-mode = separate}
-\DeclareMathSizes{10}{9}{7}{6.5}
 
@@
 capacity of models that, in high-level terms, we define as the extent to
-which valid explanations derived for a model can be deemed plausible
-with respect to the underlying data and global actionability constraints
-(we refine this notion in Def. \ref{def-explainability}). For
-simplicity, we refer to models with high explanatory capacity as
-\emph{explainable}. To the best of our knowledge, Counterfactual
+which valid counterfactual explanations derived for a model can be
+deemed plausible with respect to the underlying data and global
+actionability constraints (we refine this notion in Def.
+\ref{def-explainability}). To the best of our knowledge, Counterfactual
 Training represents the first attempt to achieve more explainable models
@@
 definition of plausibility if 40-year-old individuals were in fact more
-credit-worthy on average than young adults. But of course this CE does
-not qualify for providing actionable recourse to the applicant since
-\emph{age} is not a (directly) mutable feature. Counterfactual training
-aims to improve model explainability by aligning models with
-counterfactuals that meet both desiderata: plausibility and
-actionability. Formally, we define explainability as follows:
+credit-worthy on average than young adults. But to the applicant, this
+CE does not provide actionable recourse, as age is not a (directly)
+mutable feature. Counterfactual training aims to improve model
+explainability by aligning models with counterfactuals that meet both
+desiderata: plausibility and actionability. Formally, we define
+explainability as follows:
 
@@
 for all real-world tabular datasets we could identify and protect an
-\emph{age} variable; for \emph{MNIST}, we protect the five top and
-bottom rows of digits. Reductions in costs are overwhelmingly positive
-and significant of up to nearly 60\% for \emph{GMSC}. While the
-estimated cost reductions for \emph{Adult} and \emph{MNIST} are not
-significant, Figure~\ref{fig-mnist} (columns 6-10) demonstrates that CT
-does have the expected effect: sensitivity to protected features as per
-integrated gradients is drastically reduced; details of this experiment
-are reported in the supplementary appendix. In the case of \emph{Cred},
-average costs increase, likely because any potential benefits from
-protecting the \emph{age} are outweighed by the increase in costs
-required for greater plausibility.
+\emph{age} variable; for \emph{MNIST}, we protect the five upper and
+lower pixel rows of the full image. Reductions in costs are
+overwhelmingly positive and significant of up to nearly 60\% for
+\emph{GMSC}. While the estimated cost reductions for \emph{Adult} and
+\emph{MNIST} are not significant, Figure~\ref{fig-mnist} (columns 6-10)
+demonstrates that CT does have the expected effect: sensitivity to
+protected features as per integrated gradients is drastically reduced;
+details of this experiment are reported in the supplementary appendix.
+In the case of \emph{Cred}, average costs increase, likely because any
+potential benefits from protecting the \emph{age} are outweighed by the
+increase in costs required for greater plausibility.
 
@@
 \emph{two} standard errors for all datasets. \textbf{Plausibility}
-(columns 1-2): percentage reduction in implausibility for \(IP\) and
-\(IP^*\), respectively; \textbf{Actionability} (column 3): percentage
-reduction in costs with protected features. \textbf{Accuracy} (columns
-4-7): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
-and the baseline (BL). Counterfactual outcomes in columns 1-3 are
+(columns 1-2): percentage reduction in implausibility for \(\text{IP}\)
+and \(\text{IP}^*\), respectively; \textbf{Actionability} (column 3):
+percentage reduction in costs with protected features. \textbf{Accuracy}
+(columns 4-7): test accuracies and robust accuracies (\(\text{Acc}^*\))
+for CT and the baseline (BL). Counterfactual outcomes in columns 1-3 are
 aggregated across bootstrap samples and varying degrees of the energy

### 2025-07-14 paper/sections/experiments.qmd

@@
 We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as shown in @fig-poc.
-In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of CEs associated with CT compared to the baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits.
+In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of CEs associated with CT compared to the baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image.
 Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*.
@@
 Key performance metrics and bootstrapped *two* standard errors for all datasets.
-**Plausibility** (columns 1-2): percentage reduction in implausibility for $IP$ and $IP^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features.
+**Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features.
 **Accuracy** (columns 4-7): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL).

### 2025-07-15 paper/quarto_aaai/paper.tex
minor adjustments in text

@@
 implausibility: it is substantial and statistically significant for
-\emph{LS} across both metrics, the distance-based \(IP\) (29\%) and
-divergence-based \(IP^{*}\) (55\%).
+\emph{LS} across both metrics, the distance-based \(\text{IP}\) (29\%)
+and divergence-based \(\text{IP}^*\) (55\%).
 
@@
 We find that CT generally leads to substantial and statistically
-significant improvements in plausibility: average reductions in \(IP\)
-range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
-for the real-world tabular datasets they are around 12\% for both
-\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
+significant improvements in plausibility: average reductions in
+\(\text{IP}\) range from around 7\% for \emph{MNIST} to almost 60\% for
+\emph{Circ}; for the real-world tabular datasets they are around 12\%
+for both \emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
 \emph{Adult} and \emph{OL} we find no significant impact of CT on
-\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
-statistically significant, although the average degree of uncertainty is
-higher than for \(IP\): average reductions range from around 20\%
-(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
-for \emph{OL} and \emph{MNIST} are statistically insignificant and, for
-MNIST, do not align with qualitative findings, which are much more
-plausible for CT (Figure~\ref{fig-mnist}).
+\(\text{IP}\). Reductions in \(\text{IP}^*\) are even more substantial
+and generally statistically significant, although the average degree of
+uncertainty is higher than for \(\text{IP}\): average reductions range
+from around 20\% (\emph{Moons}) to almost 90\% (\emph{Circ}). The only
+negative findings for \emph{OL} and \emph{MNIST} are statistically
+insignificant and, for MNIST, do not align with qualitative findings,
+which are much more plausible for CT (Figure~\ref{fig-mnist}).
 

### 2025-07-15 paper/sections/experiments.qmd
minor adjustments in text

@@
 \ref{def-explainability}.
-This difference in outcomes is quantified for the non-linear MLP in the first two columns of @tbl-main as the $\%$-reduction in implausibility: it is substantial and statistically significant for *LS* across both metrics, the distance-based $IP$ (29%) and divergence-based $IP^{*}$ (55%).
+This difference in outcomes is quantified for the non-linear MLP in the first two columns of @tbl-main as the $\%$-reduction in implausibility: it is substantial and statistically significant for *LS* across both metrics, the distance-based $\text{IP}$ (29%) and divergence-based $\text{IP}^*$ (55%).
 
@@
 
-We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $IP$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $IP$.
-Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*).
+We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$.
+Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*).
 The only negative findings for *OL* and *MNIST* are statistically insignificant and, for MNIST, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).

### 2025-07-16 paper/quarto_aaai/paper.tex
asterisks added and rerendered

@@
 \caption{\label{tbl-main}Key performance metrics and bootstrapped
-\emph{two} standard errors for all datasets. \textbf{Plausibility}
-(columns 1-2): percentage reduction in implausibility for \(\text{IP}\)
-and \(\text{IP}^*\), respectively; \textbf{Actionability} (column 3):
-percentage reduction in costs with protected features. \textbf{Accuracy}
-(columns 4-7): test accuracies and robust accuracies (\(\text{Acc}^*\))
-for CT and the baseline (BL). Counterfactual outcomes in columns 1-3 are
-aggregated across bootstrap samples and varying degrees of the energy
-penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
-Standard errors for accuracy are bootstrapped from the test set.}
+standard errors (omitted if essentially zero) for all datasets.
+\textbf{Plausibility} (columns 1-2): percentage reduction in
+implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
+\textbf{Actionability} (column 3): percentage reduction in costs with
+protected features. \textbf{Accuracy} (columns 4-7): test accuracies and
+robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
+Counterfactual outcomes in columns 1-3 are aggregated across bootstrap
+samples and varying degrees of the energy penalty
+\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Asterisks
+(\(^*\)) in columns 1-3 indicate that difference is statistically
+significant at 1\%-level. Standard errors for accuracy are bootstrapped
+from the test set.}
 

### 2025-07-16 paper/experiments/CTExperiments/src/presentation.jl

@@ -606,7 +606,7 @@ function aggregate_ce_evaluation(
     valid_only::Bool=true,
     protected_only::Bool=false,
     ct_only::Bool=false,
-    conf_int::Union{Nothing,Vector{<:AbstractFloat}}=[0.95,0.99],
+    conf_int::Union{Nothing,Vector{<:AbstractFloat}}=[0.99],
     return_sig_level::Bool= true
 )
 
@@ -887,7 +887,7 @@ function final_table(
     bootstrap::Int=100,
     total_uncertainty::Bool=false,
     drop_models::Vector{String}=String[],
-    conf_int::Vector{<:AbstractFloat}=[0.95,0.99]
+    conf_int::Vector{<:AbstractFloat}=[0.99]
 )
     # CE:
     df_ce = DataFrame()
@@ -928,7 +928,7 @@ function final_table(
     # Post-process
     df = DataFrames.transform(
         df, 
-        [:mean, :se, :sig_level] => ((m, s, stars) -> [isnan(si) ? "$(round(mi, digits=2))" : PrettyTables.LatexCell("$(round(mi, digits=2))\\pm$(round(2si, digits=2)) ^{$star}") for (mi, si, star) in zip(m, s, stars)]) => :mean
+        [:mean, :se, :sig_level] => ((m, s, stars) -> [isnan(si) ? "$(round(mi, digits=2))" : PrettyTables.LatexCell("$(round(mi, digits=2))\\pm$(round(si, digits=2)) \$^{$star}\$") for (mi, si, star) in zip(m, s, stars)]) => :mean
     ) |>
         df -> select!(df, Not(:se, :sig_level)) |>
         df -> DataFrames.unstack(df, :dataset, :mean)

### 2025-07-16 paper/quarto_aaai/paper.tex
minor adjustments ot table 1 caption

@@
 robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
-Counterfactual outcomes in columns 1-3 are aggregated across bootstrap
+Counterfactual outcomes (col.~1-3) are aggregated across bootstrap
 samples and varying degrees of the energy penalty
 \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Asterisks
-(\(^*\)) in columns 1-3 indicate that difference is statistically
-significant at 1\%-level. Standard errors for accuracy are bootstrapped
-from the test set.}
+(\(^*\)) in columns 1-3 indicate that bootstrapped 99\%-confidence
+interval of differences in mean outcomes does \emph{not} include zero.
+Standard errors for accuracy (col.~4-7) are bootstrapped from the test
+set.}
 

### 2025-07-16 paper/sections/experiments.qmd
asterisks added and rerendered
minor adjustments ot table 1 caption

@@
 
-Key performance metrics and bootstrapped *two* standard errors for all datasets.
+Key performance metrics and bootstrapped standard errors (omitted if essentially zero) for all datasets.
 **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features.
 **Accuracy** (columns 4-7): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL).
-Counterfactual outcomes in columns 1-3 are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time.
-Standard errors for accuracy are bootstrapped from the test set.
+Counterfactual outcomes (col.
+1-3) are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time.
+Asterisks ($^*$) in columns 1-3 indicate that bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
+Standard errors for accuracy (col.
+4-7) are bootstrapped from the test set.
 :::

### 2025-07-17 paper/sections/intro.qmd
a very straigh-to-the-point intro now

@@
+---
+bibliography: bibliography.bib
+---
+
 # Introduction
 
-Today's prominence of artificial intelligence (AI) has largely been driven by **representation learning**: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
-Advances in computing have made it possible to provide these models with ever-growing degrees of freedom to achieve this task, which often allows them to outperform traditionally parsimonious models.
-Unfortunately, in doing so, models learn increasingly complex, sensitive representations that humans can no longer easily interpret.
+Counterfactual explanations (CE) have become an integral part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable].
+Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output.
+This is useful and necessary to not only understand how opaque AI models make predictions, but also to provide recourse to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque AI decision-making system (ADM).
 
-The trend towards complexity for the sake of performance has come under scrutiny in recent years.
-At the very cusp of the deep learning (DL) revolution, @szegedy2013intriguing showed that artificial neural networks (ANN) are susceptible to adversarial examples (AEs): perturbed versions of data instances that yield vastly different model predictions despite being semantically indistinguishable from their factual counterparts.
-Some partial mitigation strategies have been proposed---most notably **adversarial training** [@goodfellow2014explaining]---but truly robust deep learning remains unattainable even for models that are considered "shallow" by today's standards [@kolter2023keynote].
+For feedback to be meaningful, counterfactual explanations need to fulfill certain desiderata [@verma2020counterfactual; @karimi2020survey]: they should be faithful to the model [@altmeyer2024faithful], plausible [@joshi2019realistic] and actionable [@ustun2019actionable].
+Plausibility is typically understood as counterfactuals being *in-domain*: unsuccessful loan applicants that implement recourse should end up with a credit profile that is genuinely similar to that of individuals who have successfully repaid their loans in the past.
+Actionable explanations comply with practical constraints: a young, unsuccessful loan applicant cannot increase their age in an instance.
 
-Part of the problem is that the high degrees of freedom---high number of parameters estimated from data---provide room for many solutions that are locally optimal with respect to narrow objectives [@wilson2020case].
-As one example, research on the "lottery ticket hypothesis" suggests that modern neural networks can be pruned by up to 90% without losing predictive performance [@frankle2018lottery].
-Thus, looking at the predictive performance alone, found solutions may seem to provide compelling explanations for the data, when in fact they are based on purely associative and semantically meaningless patterns.
-This poses two related challenges.
-Firstly, there is no dependable way to verify if learned representations correspond to meaningful, plausible explanations.
-Secondly, even if we resolve this challenge, it remains undecided how to ensure that machine learning models can *only* learn valuable explanations.
+Existing state-of-the-art (SOTA) approaches in the field have largely focused on designing model-agnostic CE methods that identify subsets of counterfactuals, which comply with specific desiderata.
+This is problematic, because a narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful].
+In this work, we therefore follow the paradigm that models---as opposed to explanation methods---should be held accountable for plausible and actionable explanations.
+While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models.
 
-The first challenge has attracted an abundance of work on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations, aiming to mitigate scenarios in which practitioners deploy opaque models and have to blindly rely on their predictions.
-On many occasions, this has happened in practice, causing harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models; see, e.g., @oneil2016weapons.
-Effective XAI tools can also aid in monitoring models and providing recourse, empowering people to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "loan application accepted").
-In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual; CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output.
+Specifically, **we propose Counterfactual Training (CT)**: a novel training regime for differentiable models that leverage counterfactual explanations on-the-fly to ensure that models learn plausible and actionable explanations for the data.
+In essence, we achieve this by contrasting faithful explanations with ground-truth data, while at the same time protecting features that are subject to actionability constraints.
+To the best of our knowledge, CT represents the first attempt in this direction.
 
-To our surprise, the second challenge has not yet attracted major research interest.
-In particular, there has been no concerted effort towards improving the degree to which learned representations promote explanations that are both **interpretable** to and deemed **plausible** by humans.
-Instead, the typical choice has been to improve the ability of XAI tools to identify the subset of explanations that are plausible and valid for any given model, independent of whether these explanations are compatible with the learned representations [@altmeyer2024faithful].
-Fortunately, recent findings indicate that improved "explanatory capacity" of a model can arise as a consequence of regularization techniques aimed at other training objectives such as generative capacity, generalization, or robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating].
-Our contribution consolidates these findings within a unified framework.
+In our experiments, we show that thanks to counterfactual training, (1) implausibility decreases with 16% on average; (2) the cost of reaching counterfactuals with protected features decreases with 15%; and (3) robust accuracy increases for all datasets with up to .42.
 
-Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explanatory capacity of models that, in high-level terms, we define as the extent to which valid counterfactual explanations derived for a model can be deemed plausible with respect to the underlying data and global actionability constraints (we refine this notion in Def.
-\ref{def-explainability}).
-To the best of our knowledge, Counterfactual Training represents the first attempt to achieve more explainable models by employing counterfactual explanations already in the training phase.
+__________________________________________________________
 

### 2025-07-17 paper/sections/intro.qmd
major adjustments to intro and related work

@@
----
-bibliography: bibliography.bib
----
-
 # Introduction
 
-Counterfactual explanations (CE) have become an integral part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable].
-Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output.
-This is useful and necessary to not only understand how opaque AI models make predictions, but also to provide recourse to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque AI decision-making system (ADM).
+Today’s prominence of artificial intelligence (AI) has largely been driven by the success of representation learning with large degrees of freedom: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning highly complex representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
+These models tend to be so complex and opaque that humans cannot easily interpret their decision-making.
+
+Counterfactual explanations (CE) have become an integral part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable] that can be employed to make sense of this complexity.
+Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output (@fig-poc).
+This is useful and necessary to not only understand how opaque models make predictions, but also to provide recourse to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque AI decision-making system (ADM).
 
@@
 In this work, we therefore follow the paradigm that models---as opposed to explanation methods---should be held accountable for plausible and actionable explanations.
-While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models.
+While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models to improve their overall explanatory capacity.
 
-Specifically, **we propose Counterfactual Training (CT)**: a novel training regime for differentiable models that leverage counterfactual explanations on-the-fly to ensure that models learn plausible and actionable explanations for the data.
-In essence, we achieve this by contrasting faithful explanations with ground-truth data, while at the same time protecting features that are subject to actionability constraints.
-To the best of our knowledge, CT represents the first attempt in this direction.
-
-In our experiments, we show that thanks to counterfactual training, (1) implausibility decreases with 16% on average; (2) the cost of reaching counterfactuals with protected features decreases with 15%; and (3) robust accuracy increases for all datasets with up to .42.
-
-__________________________________________________________
+Specifically, **we propose Counterfactual Training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE).
+@fig-poc illustrates the outcome of CT compared to conventionally trained models: in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange); in panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals; in panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features; by contrast, the decision boundary in panel (c) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature.
+To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful CE with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals.
+To the best of our knowledge, CT represents the first venture in this direction with promising results, both empirical and theoretical.
 
 The remainder of this manuscript is structured as follows.
-@sec-lit presents related work, focusing on the link between AEs and CEs.
+@sec-lit presents related work, focusing on the links to contrastive and robust learning.
 Then follow our two principal contributions.
 In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints.
-In @sec-experiments, through extensive experiments, we empirically demonstrate that CT substantially improves explainability and positively contributes to the robustness of trained models without sacrificing predictive performance.
-Finally, in @sec-discussion, we discuss open challenges and conclude that CT is a promising approach towards making opaque models more trustworthy.
+In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) implausibility of CE decreases by 16% on average; (2) the cost of reaching valid counterfactuals with protected features decreases by 15%; and (3) robust accuracy increases for all datasets by up to 42 percentage points.
+Finally, in @sec-discussion, we discuss open challenges and conclude.
+
+![Counterfactual explanations (stars) for linear classifiers trained under different regimes on synthetic data: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable.
+The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange).
+Stars indicate counterfactuals in the target class.
+Synthetic features are annotated as *debt* and *age* for illustrative purposes.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

### 2025-07-17 paper/sections/related.qmd
major adjustments to intro and related work

@@
+---
+bibliography: bibliography.bib
+---
+
 # Related Literature {#sec-lit}
 
-To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying explainability to the quality of CEs that can be generated for a given model.
-The authors show that CEs (understood as minimal input perturbations that yield some desired model prediction) tend to be more meaningful if the underlying model is more robust to adversarial examples.
-We can make intuitive sense of this finding if we look at adversarial training (AT) through the lens of representation learning with high degrees of freedom.
-As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., AEs).
-Thus, by inducing models to "unlearn" susceptiblity to such examples, adversarial training can effectively remove implausible explanations from the solution space.
+To make the desiderata for our framework more concrete, we follow previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial].
+For simplicity, we refer to 'explanatory capacity' simply as 'explainability' in what follows.
 
-## Adversarial Examples are Counterfactuals
+## Explainability and Contrastive Learning
 
-The interpretation of the link between explainability through counterfactuals on the one side, and robustness to adversarial examples on the other is backed by empirical evidence.
+In a closely related work, @altmeyer2024faithful show that model averaging and, in particular, contrastive model objectives can produce more explainable and hence trustworthy models.
+To do so, the authors first propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
+Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
+Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
+The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your]---a form of constrastive learning---tend to do well in this regard.
+
+It helps to look at these findings through the lens of representation learning with high degrees of freedom.
+Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case].
+Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations for the data.
+Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training.
+The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which bears resemblance to the idea of aligning models with plausible explanations and has inspired our CT objective.
+
+## Explainability and Robust Learning
+
+@augustin2020adversarial show that CEs tend to be more meaningful ("explainable") if the underlying model is more robust to adversarial examples.
+Once again, we can make intuitive sense of this finding if we look at adversarial training (AT) through the lens of representation learning with high degrees of freedom: highly complex and flexible models may learn representations that make them sensitive to implausible or even adversarial examples [@szegedy2013intriguing].
+Thus, by inducing models to "unlearn" susceptibility to such examples, adversarial training can effectively remove implausible explanations from the solution space.
+
+This interpretation of the link between explainability through counterfactuals on the one side, and robustness to adversarial examples on the other is backed by empirical evidence.
 @sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness.
@@
 In contrast, we impose no restrictions on the ANN architecture at all.
-
-## Aligning Representations with Explanations
-
-Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations.
-In a closely related work, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives.
-They propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
-Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
-Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
-The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.
-
-Once again it helps to look at these findings through the lens of representation learning with high degrees of freedom.
-Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case].
-Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations for the data.
-Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training.
-The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which bears resemblance to the idea of aligning models with plausible explanations and has inspired our CT objective.

### 2025-07-17 paper/quarto_aaai/paper.tex
removed figure and synthetic example from experiments

@@
 Today's prominence of artificial intelligence (AI) has largely been
-driven by \textbf{representation learning}: instead of relying on
-features and rules hand-crafted by humans, modern machine learning (ML)
-models are tasked with learning representations directly from the data,
-guided by narrow objectives such as predictive accuracy
-\citep{goodfellow2016deep}. Advances in computing have made it possible
-to provide these models with ever-growing degrees of freedom to achieve
-this task, which often allows them to outperform traditionally
-parsimonious models. Unfortunately, in doing so, models learn
-increasingly complex, sensitive representations that humans can no
-longer easily interpret.
-
-The trend towards complexity for the sake of performance has come under
-scrutiny in recent years. At the very cusp of the deep learning (DL)
-revolution, \citet{szegedy2013intriguing} showed that artificial neural
-networks (ANN) are susceptible to adversarial examples (AEs): perturbed
-versions of data instances that yield vastly different model predictions
-despite being semantically indistinguishable from their factual
-counterparts. Some partial mitigation strategies have been
-proposed---most notably \textbf{adversarial training}
-\citep{goodfellow2014explaining}---but truly robust deep learning
-remains unattainable even for models that are considered ``shallow'' by
-today's standards \citep{kolter2023keynote}.
-
-Part of the problem is that the high degrees of freedom---high number of
-parameters estimated from data---provide room for many solutions that
-are locally optimal with respect to narrow objectives
-\citep{wilson2020case}. As one example, research on the ``lottery ticket
-hypothesis'' suggests that modern neural networks can be pruned by up to
-90\% without losing predictive performance \citep{frankle2018lottery}.
-Thus, looking at the predictive performance alone, found solutions may
-seem to provide compelling explanations for the data, when in fact they
-are based on purely associative and semantically meaningless patterns.
-This poses two related challenges. Firstly, there is no dependable way
-to verify if learned representations correspond to meaningful, plausible
-explanations. Secondly, even if we resolve this challenge, it remains
-undecided how to ensure that machine learning models can \emph{only}
-learn valuable explanations.
-
-The first challenge has attracted an abundance of work on
-\textbf{explainable AI} (XAI), a paradigm that focuses on the
-development of tools to derive (post-hoc) explanations from complex
-model representations, aiming to mitigate scenarios in which
-practitioners deploy opaque models and have to blindly rely on their
-predictions. On many occasions, this has happened in practice, causing
-harms to people who were adversely and unfairly affected by automated
-decision-making (ADM) systems involving opaque models; see, e.g.,
-\citet{oneil2016weapons}. Effective XAI tools can also aid in monitoring
-models and providing recourse, empowering people to turn negative
-outcomes (e.g., ``loan application rejected'') into positive ones (e.g.,
-``loan application accepted''). In line with this, our work builds upon
-\textbf{counterfactual explanations} (CE) proposed by
-\citet{wachter2017counterfactual}; CEs prescribe minimal changes for
-factual inputs that, if implemented, would prompt some fitted model to
-produce an alternative, more desirable output.
-
-To our surprise, the second challenge has not yet attracted major
-research interest. In particular, there has been no concerted effort
-towards improving the degree to which learned representations promote
-explanations that are both \textbf{interpretable} to and deemed
-\textbf{plausible} by humans. Instead, the typical choice has been to
-improve the ability of XAI tools to identify the subset of explanations
-that are plausible and valid for any given model, independent of whether
-these explanations are compatible with the learned representations
-\citep{altmeyer2024faithful}. Fortunately, recent findings indicate that
-improved ``explanatory capacity'' of a model can arise as a consequence
-of regularization techniques aimed at other training objectives such as
-generative capacity, generalization, or robustness
-\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating}.
-Our contribution consolidates these findings within a unified framework.
+driven by the success of representation learning with large degrees of
+freedom: instead of relying on features and rules hand-crafted by
+humans, modern machine learning (ML) models are tasked with learning
+highly complex representations directly from the data, guided by narrow
+objectives such as predictive accuracy \citep{goodfellow2016deep}. These
+models tend to be so complex and opaque that humans cannot easily
+interpret their decision-making.
+
+Counterfactual explanations (CE) have become an integral part of the
+broader explainable AI (XAI) toolkit \citep{molnar2022interpretable}
+that can be employed to make sense of this complexity. Originally
+proposed by \citet{wachter2017counterfactual}, CEs prescribe minimal
+changes for factual inputs that, if implemented, would prompt some
+fitted model to produce an alternative, more desirable output
+(Figure~\ref{fig-poc}). This is useful and necessary to not only
+understand how opaque models make predictions, but also to provide
+recourse to individuals subjected to them: a retail bank, for example,
+could use CE to provide meaningful feedback to unsuccessful loan
+applicants that were rejected based on an opaque AI decision-making
+system (ADM).
+
+For feedback to be meaningful, counterfactual explanations need to
+fulfill certain desiderata
+\citep{verma2020counterfactual, karimi2020survey}: they should be
+faithful to the model \citep{altmeyer2024faithful}, plausible
+\citep{joshi2019realistic} and actionable \citep{ustun2019actionable}.
+Plausibility is typically understood as counterfactuals being
+\emph{in-domain}: unsuccessful loan applicants that implement recourse
+should end up with a credit profile that is genuinely similar to that of
+individuals who have successfully repaid their loans in the past.
+Actionable explanations comply with practical constraints: a young,
+unsuccessful loan applicant cannot increase their age in an instance.
+
+Existing state-of-the-art (SOTA) approaches in the field have largely
+focused on designing model-agnostic CE methods that identify subsets of
+counterfactuals, which comply with specific desiderata. This is
+problematic, because a narrow focus on any specific desideratum can
+adversely affect others: it is possible, for example, to generate
+plausible counterfactuals for models that are also highly vulnerable to
+implausible, possibly adversarial counterfactuals
+\citep{altmeyer2024faithful}. In this work, we therefore follow the
+paradigm that models---as opposed to explanation methods---should be
+held accountable for plausible and actionable explanations. While
+previous work has shown that at least plausibility can be indirectly
+achieved through existing techniques aimed at models' generative
+capacity, generalization and robustness
+\citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
+we directly incorporate both plausibility and actionability in the
+training objective of models to improve their overall explanatory
+capacity.
 
 Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
-training regime explicitly geared towards improving the explanatory
-capacity of models that, in high-level terms, we define as the extent to
-which valid counterfactual explanations derived for a model can be
-deemed plausible with respect to the underlying data and global
-actionability constraints (we refine this notion in Def.
-\ref{def-explainability}). To the best of our knowledge, Counterfactual
-Training represents the first attempt to achieve more explainable models
-by employing counterfactual explanations already in the training phase.
+training regime that leverages counterfactual explanations on-the-fly to
+ensure that differentiable models learn plausible and actionable
+explanations for the underlying data, while at the same time also being
+more robust to adversarial examples (AE). Figure~\ref{fig-poc}
+illustrates the outcome of CT compared to conventionally trained models:
+in panel (a), faithful and valid counterfactuals end up near the
+decision boundary forming a clearly distinguishable cluster in the
+target class (orange); in panel (b), CT is applied to the same
+underlying linear classifier architecture resulting in much more
+plausible counterfactuals; in panel (c), the classifier is again trained
+conventionally and we have introduced a mutability constraint on the
+\emph{age} feature at test time---counterfactuals are valid but the
+classifier is roughly equally sensitive to both features; by contrast,
+the decision boundary in panel (c) has titled, making the model trained
+with CT relatively less sensitive to the immutable \emph{age} feature.
+To achieve these outcomes, CT draws inspiration from the literature on
+contrastive and robust learning: we contrast faithful CE with
+ground-truth data while protecting immutable features, and capitalize on
+methodological links between CE and AE by penalizing the model's
+adversarial loss on interim (\emph{nascent}) counterfactuals. To the
+best of our knowledge, CT represents the first venture in this direction
+with promising results, both empirical and theoretical.
 
 The remainder of this manuscript is structured as follows.
-Section~\ref{sec-lit} presents related work, focusing on the link
-between AEs and CEs. Then follow our two principal contributions. In
-Section~\ref{sec-method}, we introduce our methodological framework and
-show theoretically that it can be employed to enforce global
-actionability constraints. In Section~\ref{sec-experiments}, through
-extensive experiments, we empirically demonstrate that CT substantially
-improves explainability and positively contributes to the robustness of
-trained models without sacrificing predictive performance. Finally, in
-Section~\ref{sec-discussion}, we discuss open challenges and conclude
-that CT is a promising approach towards making opaque models more
-trustworthy.
+Section~\ref{sec-lit} presents related work, focusing on the links to
+contrastive and robust learning. Then follow our two principal
+contributions. In Section~\ref{sec-method}, we introduce our
+methodological framework and show theoretically that it can be employed
+to enforce global actionability constraints. In our experiments
+(Section~\ref{sec-experiments}), we find that thanks to counterfactual
+training, (1) implausibility of CE decreases by 16\% on average; (2) the
+cost of reaching valid counterfactuals with protected features decreases
+by 15\%; and (3) robust accuracy increases for all datasets by up to 42
+percentage points. Finally, in Section~\ref{sec-discussion}, we discuss
+open challenges and conclude.
+
+\begin{figure*}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
+
+}
+
+\caption{\label{fig-poc}Counterfactual explanations (stars) for linear
+classifiers trained under different regimes on synthetic data: (a)
+conventional training, all mutable; (b) CT, all mutable; (c)
+conventional, \emph{age} immutable; (d) CT, \emph{age} immutable. The
+linear decision boundary is shown in green along with training data
+colored according to ground-truth labels: \(y^-=1\) (blue) and \(y^+=2\)
+(orange). Stars indicate counterfactuals in the target class. Synthetic
+features are annotated as \emph{debt} and \emph{age} for illustrative
+purposes.}
+
+\end{figure*}%
 
@@
 To make the desiderata for our framework more concrete, we follow
-\citet{augustin2020adversarial} in tying explainability to the quality
-of CEs that can be generated for a given model. The authors show that
-CEs (understood as minimal input perturbations that yield some desired
-model prediction) tend to be more meaningful if the underlying model is
-more robust to adversarial examples. We can make intuitive sense of this
-finding if we look at adversarial training (AT) through the lens of
-representation learning with high degrees of freedom. As argued before,
-learned representations may be sensitive to producing implausible
-explanations and mispredicting for worst-case counterfactuals (i.e.,
-AEs). Thus, by inducing models to ``unlearn'' susceptiblity to such
-examples, adversarial training can effectively remove implausible
-explanations from the solution space.
-
-\subsection{Adversarial Examples are
-Counterfactuals}\label{adversarial-examples-are-counterfactuals}
-
-The interpretation of the link between explainability through
+previous work in tying the explanatory capacity of models to the quality
+of CEs that can be generated for them
+\citep{altmeyer2024faithful, augustin2020adversarial}. For simplicity,
+we refer to `explanatory capacity' simply as `explainability' in what
+follows (see Def. \ref{def-explainability} below for a more formal
+exposition).
+
+\subsection{Explainability and Contrastive
+Learning}\label{explainability-and-contrastive-learning}
+
+In a closely related work, \citet{altmeyer2024faithful} show that model
+averaging and, in particular, contrastive model objectives can produce
+more explainable and hence trustworthy models. To do so, the authors
+first propose a way to generate counterfactuals that are maximally
+faithful to the model in that they are consistent with what the model
+has learned about the underlying data. Formally, they rely on tools from
+energy-based modelling \citep{teh2003energy} to minimize the divergence
+between the distribution of counterfactuals and the conditional
+posterior over inputs learned by the model. Their counterfactual
+explainer, \emph{ECCCo}, yields plausible explanations if and only if
+the underlying model has learned representations that align with them.
+The authors find that both deep ensembles
+\citep{lakshminarayanan2016simple} and joint energy-based models (JEMs)
+\citep{grathwohl2020your}---a form of constrastive learning---tend to do
+well in this regard.
+
+It helps to look at these findings through the lens of representation
+learning with high degrees of freedom. Deep ensembles are approximate
+Bayesian model averages, which are particularly effective when models
+are underspecified by the available data \citep{wilson2020case}.
+Averaging across solutions mitigates the aforementioned risk of
+overrelying on a single locally optimal representation that corresponds
+to semantically meaningless explanations for the data. Likewise,
+previous work of \citet{schut2021generating} found that generating
+plausible (``interpretable'') CEs is almost trivial for deep ensembles
+that have undergone adversarial training. The case for JEMs is even
+clearer: they optimize a hybrid objective that induces both high
+predictive performance and strong generative capacity
+\citep{grathwohl2020your}, which bears resemblance to the idea of
+aligning models with plausible explanations and has inspired our CT
+objective.
+
+\subsection{Explainability and Robust
+Learning}\label{explainability-and-robust-learning}
+
+\citet{augustin2020adversarial} show that CEs tend to be more meaningful
+(``explainable'') if the underlying model is more robust to adversarial
+examples. Once again, we can make intuitive sense of this finding if we
+look at adversarial training (AT) through the lens of representation
+learning with high degrees of freedom: highly complex and flexible
+models may learn representations that make them sensitive to implausible
+or even adversarial examples \citep{szegedy2013intriguing}. Thus, by
+inducing models to ``unlearn'' susceptibility to such examples,
+adversarial training can effectively remove implausible explanations
+from the solution space.
+
+This interpretation of the link between explainability through
 counterfactuals on the one side, and robustness to adversarial examples
@@
 restrictions on the ANN architecture at all.
-
-\subsection{Aligning Representations with
-Explanations}\label{aligning-representations-with-explanations}
-
-Improving the adversarial robustness of models is not the only path
-towards aligning representations with plausible explanations. In a
-closely related work, \citet{altmeyer2024faithful} show that
-explainability can be improved through model averaging and refined model
-objectives. They propose a way to generate counterfactuals that are
-maximally faithful to the model in that they are consistent with what
-the model has learned about the underlying data. Formally, they rely on
-tools from energy-based modelling \citep{teh2003energy} to minimize the
-divergence between the distribution of counterfactuals and the
-conditional posterior over inputs learned by the model. Their
-counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
-and only if the underlying model has learned representations that align
-with them. The authors find that both deep ensembles
-\citep{lakshminarayanan2016simple} and joint energy-based models (JEMs)
-\citep{grathwohl2020your} tend to do well in this regard.
-
-Once again it helps to look at these findings through the lens of
-representation learning with high degrees of freedom. Deep ensembles are
-approximate Bayesian model averages, which are particularly effective
-when models are underspecified by the available data
-\citep{wilson2020case}. Averaging across solutions mitigates the
-aforementioned risk of overrelying on a single locally optimal
-representation that corresponds to semantically meaningless explanations
-for the data. Likewise, previous work of \citet{schut2021generating}
-found that generating plausible (``interpretable'') CEs is almost
-trivial for deep ensembles that have undergone adversarial training. The
-case for JEMs is even clearer: they optimize a hybrid objective that
-induces both high predictive performance and strong generative capacity
-\citep{grathwohl2020your}, which bears resemblance to the idea of
-aligning models with plausible explanations and has inspired our CT
-objective.
 
@@
 illustrates how to read Table~\ref{tbl-main}.
-
-\begin{figure*}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
-
-}
-
-\caption{\label{fig-poc}Illustration of how CT improves model
-explainability: (a) conventional training, all mutable; (b) CT, all
-mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
-immutable. The linear decision boundary is shown in green along with
-training data colored according to ground-truth labels: \(y^-=1\) (blue)
-and \(y^+=2\) (orange). Stars indicate counterfactuals in the target
-class.}
-
-\end{figure*}%
-
-\subsubsection{Synthetic Example: Prediction of Credit Card
-Defaults.}\label{synthetic-example-prediction-of-credit-card-defaults.}
-
-Figure~\ref{fig-poc} presents results for a linear classifier fitted to
-\emph{LS} that complies with the data assumptions in Proposition
-\ref{prp-mtblty}. The four panels show the outcomes for different
-training procedures: in panels (a) and (c) we have trained the models
-conventionally, while in panels (b) and (d) we have applied CT. For
-illustrative purposes, suppose the first feature represents \emph{debt}
-(mutable) and the second feature represents \emph{age} (immutable) of
-loan applicants seeking counterfactual explanations for moving to the
-target class: loan provided (orange).
-
-In all four cases, it is possible to generate valid counterfactuals
-(stars) for unsuccessful applicants (blue). They cross the decision
-boundary (green) into the target class, but their quality differs. In
-panel (a), they are not plausible: they do not comply with the
-distribution of the factuals in \(y^+\) to the point where they form a
-clearly discernible cluster. In panel (b), they are highly plausible,
-meeting the first objective of Def. \ref{def-explainability}. This
-difference in outcomes is quantified for the non-linear MLP in the first
-two columns of Table~\ref{tbl-main} as the \(\%\)-reduction in
-implausibility: it is substantial and statistically significant for
-\emph{LS} across both metrics, the distance-based \(\text{IP}\) (29\%)
-and divergence-based \(\text{IP}^*\) (55\%).
-
-In panel (c) of Figure~\ref{fig-poc}, the CEs involve substantial
-reductions in \emph{debt} for younger applicants. By comparison,
-counterfactual paths are shorter on average in panel (d) where we have
-protected the immutable \emph{age} as described in
-Section~\ref{sec-constraints}. Due to the classifier's lower sensitivity
-to \emph{age}, recommendations with respect to \emph{debt} are much more
-homogenous and do not unfairly punish younger individuals. These CEs are
-also plausible with respect to the mutable feature, despite requiring
-smaller debt reductions on average, resulting in smaller costs to
-individuals. This result is quantified for the non-linear case in column
-three of Table~\ref{tbl-main}, which shows the \(\%\)-reduction in costs
-averaged across valid counterfactuals. Once again, the impact of CT is
-statistically significant and substantial (14\%). Thus, we consider the
-model in panel (d) as the most explainable according to Def.
-\ref{def-explainability}. Next, we present the results for all remaining
-datasets.
 
@@
 \renewcommand{\bibsection}{}
-\bibliography{../quarto_ecml/bibliography.bib}
+\bibliography{../quarto_aaai/bibliography.bib}
 \end{document}

### 2025-07-17 paper/quarto_aaai/paper.tex
shorted one sentence
Draft adjustments in discussions/conclusions

--- 2025-07-16 paper/quarto_aaai/paper.tex
@@
 capacity of models that, in high-level terms, we define as the extent to
-which valid counterfactual explanations derived for a model can be
-deemed plausible with respect to the underlying data and global
-actionability constraints (we refine this notion in Def.
-\ref{def-explainability}). To the best of our knowledge, Counterfactual
+which valid explanations derived for a model can be deemed plausible
+with respect to the underlying data and global actionability constraints
+(we refine this notion in Def. \ref{def-explainability}). For
+simplicity, we refer to models with high explanatory capacity as
+\emph{explainable}. To the best of our knowledge, Counterfactual
 Training represents the first attempt to achieve more explainable models
@@
 improves explainability and positively contributes to the robustness of
-trained models without sacrificing predictive performance. Finally, in
-Section~\ref{sec-discussion}, we discuss open challenges and conclude
-that CT is a promising approach towards making opaque models more
-trustworthy.
+trained models without sacrificing predictive performance. In
+Section~\ref{sec-discussion}, we discuss open challenges and, in
+\textbf{?@sec-conclusion}, conclude that CT is a promising approach
+towards making opaque models more trustworthy.
 
@@
 definition of plausibility if 40-year-old individuals were in fact more
-credit-worthy on average than young adults. But to the applicant, this
-CE does not provide actionable recourse, as age is not a (directly)
-mutable feature. Counterfactual training aims to improve model
-explainability by aligning models with counterfactuals that meet both
-desiderata: plausibility and actionability. Formally, we define
-explainability as follows:
+credit-worthy on average than young adults. But of course this CE does
+not qualify for providing actionable recourse to the applicant since
+\emph{age} is not a (directly) mutable feature. Counterfactual training
+aims to improve model explainability by aligning models with
+counterfactuals that meet both desiderata: plausibility and
+actionability. Formally, we define explainability as follows:
 
@@
 implausibility: it is substantial and statistically significant for
-\emph{LS} across both metrics, the distance-based \(\text{IP}\) (29\%)
-and divergence-based \(\text{IP}^*\) (55\%).
+\emph{LS} across both metrics, the distance-based \(IP\) (29\%) and
+divergence-based \(IP^{*}\) (55\%).
 
@@
 We find that CT generally leads to substantial and statistically
-significant improvements in plausibility: average reductions in
-\(\text{IP}\) range from around 7\% for \emph{MNIST} to almost 60\% for
-\emph{Circ}; for the real-world tabular datasets they are around 12\%
-for both \emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
+significant improvements in plausibility: average reductions in \(IP\)
+range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
+for the real-world tabular datasets they are around 12\% for both
+\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
 \emph{Adult} and \emph{OL} we find no significant impact of CT on
-\(\text{IP}\). Reductions in \(\text{IP}^*\) are even more substantial
-and generally statistically significant, although the average degree of
-uncertainty is higher than for \(\text{IP}\): average reductions range
-from around 20\% (\emph{Moons}) to almost 90\% (\emph{Circ}). The only
-negative findings for \emph{OL} and \emph{MNIST} are statistically
-insignificant and, for MNIST, do not align with qualitative findings,
-which are much more plausible for CT (Figure~\ref{fig-mnist}).
+\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
+statistically significant, although the average degree of uncertainty is
+higher than for \(IP\): average reductions range from around 20\%
+(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
+for \emph{OL} and \emph{MNIST} are statistically insignificant and, for
+MNIST, do not align with qualitative findings, which are much more
+plausible for CT (Figure~\ref{fig-mnist}).
 
@@
 for all real-world tabular datasets we could identify and protect an
-\emph{age} variable; for \emph{MNIST}, we protect the five upper and
-lower pixel rows of the full image. Reductions in costs are
-overwhelmingly positive and significant of up to nearly 60\% for
-\emph{GMSC}. While the estimated cost reductions for \emph{Adult} and
-\emph{MNIST} are not significant, Figure~\ref{fig-mnist} (columns 6-10)
-demonstrates that CT does have the expected effect: sensitivity to
-protected features as per integrated gradients is drastically reduced;
-details of this experiment are reported in the supplementary appendix.
-In the case of \emph{Cred}, average costs increase, likely because any
-potential benefits from protecting the \emph{age} are outweighed by the
-increase in costs required for greater plausibility.
+\emph{age} variable; for \emph{MNIST}, we protect the five top and
+bottom rows of digits. Reductions in costs are overwhelmingly positive
+and significant of up to nearly 60\% for \emph{GMSC}. While the
+estimated cost reductions for \emph{Adult} and \emph{MNIST} are not
+significant, Figure~\ref{fig-mnist} (columns 6-10) demonstrates that CT
+does have the expected effect: sensitivity to protected features as per
+integrated gradients is drastically reduced; details of this experiment
+are reported in the supplementary appendix. In the case of \emph{Cred},
+average costs increase, likely because any potential benefits from
+protecting the \emph{age} are outweighed by the increase in costs
+required for greater plausibility.
 
@@
 \caption{\label{tbl-main}Key performance metrics and bootstrapped
-standard errors (omitted if essentially zero) for all datasets.
-\textbf{Plausibility} (columns 1-2): percentage reduction in
-implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
-\textbf{Actionability} (column 3): percentage reduction in costs with
-protected features. \textbf{Accuracy} (columns 4-7): test accuracies and
-robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
-Counterfactual outcomes in columns 1-3 are aggregated across bootstrap
-samples and varying degrees of the energy penalty
-\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Asterisks
-(\(^*\)) in columns 1-3 indicate that difference is statistically
-significant at 1\%-level. Standard errors for accuracy are bootstrapped
-from the test set.}
+\emph{two} standard errors for all datasets. \textbf{Plausibility}
+(columns 1-2): percentage reduction in implausibility for \(IP\) and
+\(IP^*\), respectively; \textbf{Actionability} (column 3): percentage
+reduction in costs with protected features. \textbf{Accuracy} (columns
+4-7): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
+and the baseline (BL). Counterfactual outcomes in columns 1-3 are
+aggregated across bootstrap samples and varying degrees of the energy
+penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
+Standard errors for accuracy are bootstrapped from the test set.}
 
@@
 
-\section{Conclusions}\label{sec-discussion}
+\section{Discussion}\label{sec-discussion}
 
@@
 \emph{Interventions on features have implications for fairness.} We
-provide a tool that allows practitioners to modify the sensitivity of a
-model with respect to certain features. Model owners can use our
-solution to support the fair and equitable treatment of decision
-subjects, but they could also misuse it by enforcing explanations based
-on features that are more difficult to modify by some (group of)
-individuals. When used irresponsibly, CT could result in an unfairly
-assigned burden of recourse \citep{sharma2020certifai}, threatening the
-equality of opportunity \citep{bell2024fairness}. Additionally, CT
-requires mutability constraints for the features considered by the
-model. Even if all immutable features are protected, there may exist
-proxies that are mutable, and hence should not be protected, but
-preserve sufficient information about the principals to hinder these
-protections. Deciding on actionability is still a major open challenge
-in the AR literature \citep{venkatasubramanian2020philosophical}
-impacting the capacity of CT to fulfill its intended goal.
+provide a method to modify the sensitivity of a model to certain
+features, which can be misused by enforcing explanations based on
+features that are more difficult to modify by a (group of) decision
+subjects. Such abuse could result in an unfairly assigned burden of
+recourse \citep{sharma2020certifai}, threatening the equality of
+opportunity \citep{bell2024fairness}. Also, even if all immutable
+features are protected, there may exist proxies that are theoretically
+mutable, but preserve sufficient information about the principals to
+hinder these protections. Indeed, deciding on the actionability of
+features remains a major open challenge in the AR literature
+\citep{venkatasubramanian2020philosophical}.
 
@@
 more resource-intensive than conventional regimes. Higher numbers of CEs
-improve the quality of learned representations but they also increase
-the number of computations. As our codebase is not performance
-optimized, grids of 270 settings for the largest datasets in our
+improve the quality of the learned representations but they also
+increase the number of computations. As our codebase is not performance
+optimized, grids of around 300 settings for the largest datasets in our
 experiments took up to four hours using 34 2GB CPUs (see supplementary
@@
 
-We also highlight three important directions for future research.
-Firstly, it is an interesting challenge to extend CT beyond
-classification settings. Our formulation relies on the distinction
-between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
-generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
-and \(y^{+}\) can be arbitrarily defined, CT requires the output space
-\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
-where the change in outcome cannot be readily discretized. Focus on
-classification is a common choice in research on CEs and AR. Other
-settings have attracted some interest, e.g., regression
-\citep{spooner2021counterfactual}, but there is little consensus how to
-robustly extend the notion of CEs.\\
-Secondly, CT is susceptible to training instabilities. This problem has
-been recognized for JEMs \citep{grathwohl2020your} and even though we
-depart from the SGLD sampling, we still encounter variability in
-outcomes. CT is exposed to two potential sources of instabilities: (1)
-the energy-based contrastive divergence term, \(\text{div}(\cdot)\), in
-Equation~\ref{eq-div}, and (2) the underlying explainers. We find
-several promising ways to mitigate this problem: regularizing energy
-(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
-during each epoch, and including only mature counterfactuals in
-\(\text{div}(\cdot)\).\\
+We also highlight three key directions for future research. Firstly, it
+is an interesting challenge to extend CT beyond classification settings.
+Our formulation relies on the distinction between non-target class(es)
+and target class(es), requiring the output space to be discrete. Thus,
+it does not apply to ML tasks where the change in outcome cannot be
+readily discretized. Focus on classification is a common choice in
+research on CEs and AR; other settings have attracted some interest,
+e.g., regression \citep{spooner2021counterfactual}, but there is little
+consensus how to robustly extend the notion of CEs.\\
+Secondly, CT is susceptible to training instabilities, as also
+recognized for the related JEMs \citep{grathwohl2020your}. CT is exposed
+to two sources of instabilities: (1) the energy-based contrastive
+divergence term, \(\text{div}(\cdot)\), in Equation~\ref{eq-div}, and
+(2) the underlying explainers. We find several promising ways to
+mitigate this problem: regularizing energy (\(\lambda_{\text{reg}}\)),
+generating sufficiently many counterfactuals during each epoch, and
+including only mature counterfactuals in \(\text{div}(\cdot)\).\\
 Finally, we believe that it is possible to considerably improve
 hyperparameter selection procedures, and thus performance. We have
-relied exclusively on grid searches, but future research could benefit
-from more sophisticated approaches.
-
-To conclude, state-of-the-art machine learning models are prone to
-learning complex representations that cannot be interpreted by humans.
-Existing explainability solutions cannot guarantee that explanations
-agree with the model's learned representation of data. As a step towards
-addressing this challenge, we introduced counterfactual training, a
-novel training regime that incentivizes highly-explainable models. Our
-approach leads to explanations that are both plausible (compliant with
-the underlying data-generating process) and actionable (compliant with
-user-specified mutability constraints), and thus meaningful to their
-recipients. Through extensive experiments, we demonstrated that CT
-satisfies its objective while promoting robustness and preserving the
-predictive performance of models. It can also be used to fine-tune
-conventionally-trained models and achieve similar gains. Lastly, our
-work highlights the value of simultaneously improving models and their
-explanations.
+relied exclusively on grid searches, but future work could benefit from
+more sophisticated approaches.
+
+\section{Conclusions}\label{sec-conclusions}
+
+State-of-the-art machine learning models are prone to learning complex
+representations that cannot be interpreted by humans. Existing
+explainability solutions cannot guarantee that explanations agree with
+these learned representation. As a step towards addressing this
+challenge, we introduce counterfactual training, a novel training regime
+that incentivizes highly-explainable models. Through extensive
+experiments, we demonstrate that CT satisfies this objective while
+promoting robustness and preserving the predictive performance of
+models. Explanations generated from CT-based models are both more
+plausible (compliant with the underlying data-generating process) and
+more actionable (compliant with user-specified mutability constraints),
+and thus meaningful to their recipients. We also show that CT can be
+used to fine-tune conventionally-trained models and achieve similar
+gains. Lastly, our work highlights the value of simultaneously improving
+models and their explanations.
 

### 2025-07-17 paper/sections/conclusion.qmd
Trim to 7 pages
Draft adjustments in discussions/conclusions

@@
-To conclude, state-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans.
-Existing explainability solutions cannot guarantee that explanations agree with the model's learned representation of data.
-As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models.
-Our approach leads to explanations that are both plausible (compliant with the underlying data-generating process) and actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients.
-Through extensive experiments, we demonstrated that CT satisfies its objective while promoting robustness and preserving the predictive performance of models.
-It can also be used to fine-tune conventionally-trained models and achieve similar gains.
-Thus, our work showcases that it is practical to simultaneously improve models and their explanations.
+# Conclusions {#sec-conclusions}
+
+State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans.
+Existing explainability solutions cannot guarantee that explanations agree with these learned representation.
+As a step towards addressing this challenge, we introduce counterfactual training, a novel training regime that incentivizes highly-explainable models.
+Through extensive experiments, we demonstrate that CT satisfies this objective while promoting robustness and preserving the predictive performance of models.
+Explanations generated from CT-based models are both more plausible (compliant with the underlying data-generating process) and more actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients.
+We also show that CT can be used to fine-tune conventionally-trained models and achieve similar gains.
+Lastly, our work highlights the value of simultaneously improving models and their explanations.

### 2025-07-17 paper/sections/discussion.qmd
Draft adjustments in discussions/conclusions

@@
-# Conclusions {#sec-discussion}
+# Discussion {#sec-discussion}
 
@@
 
-*Interventions on features have implications for fairness.* We provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features.
-Model owners can use our solution to support the fair and equitable treatment of decision subjects, but they could also misuse it by enforcing explanations based on features that are more difficult to modify by some (group of) individuals.
-When used irresponsibly, CT could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity [@bell2024fairness].
-Additionally, CT requires mutability constraints for the features considered by the model.
-Even if all immutable features are protected, there may exist proxies that are mutable, and hence should not be protected, but preserve sufficient information about the principals to hinder these protections.
-Deciding on actionability is still a major open challenge in the AR literature [@venkatasubramanian2020philosophical] impacting the capacity of CT to fulfill its intended goal.
+*Interventions on features have implications for fairness.* We provide a method to modify the sensitivity of a model to certain features, which can be misused by enforcing explanations based on features that are more difficult to modify by a (group of) decision subjects.
+Such abuse could result in an unfairly assigned burden of recourse [@sharma2020certifai], threatening the equality of opportunity [@bell2024fairness].
+Also, even if all immutable features are protected, there may exist proxies that are theoretically mutable, but preserve sufficient information about the principals to hinder these protections.
+Indeed, deciding on the actionability of features remains a major open challenge in the AR literature [@venkatasubramanian2020philosophical].
 
 *CT increases the training times.* Like adversarial training, CT is more resource-intensive than conventional regimes.
-Higher numbers of CEs improve the quality of learned representations but they also increase the number of computations.
-As our codebase is not performance optimized, grids of 270 settings for the largest datasets in our experiments took up to four hours using 34 2GB CPUs (see supplementary appendix).
+Higher numbers of CEs improve the quality of the learned representations but they also increase the number of computations.
+As our codebase is not performance optimized, grids of around 300 settings for the largest datasets in our experiments took up to four hours using 34 2GB CPUs (see supplementary appendix).
 Other than optimization, three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.
 
-We also highlight three important directions for future research.
+We also highlight three key directions for future research.
 Firstly, it is an interesting challenge to extend CT beyond classification settings.
-Our formulation relies on the distinction between non-target class(es) $y^{-}$ and target class(es) $y^{+}$ to generate counterfactuals through @eq-obj.
-While $y^{-}$ and $y^{+}$ can be arbitrarily defined, CT requires the output space $\mathcal{Y}$ to be discrete.
+Our formulation relies on the distinction between non-target class(es) and target class(es), requiring the output space to be discrete.
 Thus, it does not apply to ML tasks where the change in outcome cannot be readily discretized.
-Focus on classification is a common choice in research on CEs and AR.
-Other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.
-Secondly, CT is susceptible to training instabilities.
-This problem has been recognized for JEMs [@grathwohl2020your] and even though we depart from the SGLD sampling, we still encounter variability in outcomes.
-CT is exposed to two potential sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers.
+Focus on classification is a common choice in research on CEs and AR; other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.
+Secondly, CT is susceptible to training instabilities, as also recognized for the related JEMs [@grathwohl2020your].
+CT is exposed to two sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers.
 We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.
 Finally, we believe that it is possible to considerably improve hyperparameter selection procedures, and thus performance.
-We have relied exclusively on grid searches, but future research could benefit from more sophisticated approaches.
+We have relied exclusively on grid searches, but future work could benefit from more sophisticated approaches.

### 2025-07-17 paper/sections/intro.qmd
Draft adjustments in discussions/conclusions

--- 2025-07-14 paper/sections/intro.qmd
@@
 
-Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explanatory capacity of models that, in high-level terms, we define as the extent to which valid counterfactual explanations derived for a model can be deemed plausible with respect to the underlying data and global actionability constraints (we refine this notion in Def.
+Specifically, **we propose Counterfactual Training (CT)**: a novel training regime explicitly geared towards improving the explanatory capacity of models that, in high-level terms, we define as the extent to which valid explanations derived for a model can be deemed plausible with respect to the underlying data and global actionability constraints (we refine this notion in Def.
 \ref{def-explainability}).
+For simplicity, we refer to models with high explanatory capacity as *explainable*.
 To the best of our knowledge, Counterfactual Training represents the first attempt to achieve more explainable models by employing counterfactual explanations already in the training phase.
@@
 In @sec-experiments, through extensive experiments, we empirically demonstrate that CT substantially improves explainability and positively contributes to the robustness of trained models without sacrificing predictive performance.
-Finally, in @sec-discussion, we discuss open challenges and conclude that CT is a promising approach towards making opaque models more trustworthy.
+In @sec-discussion, we discuss open challenges and, in @sec-conclusion, conclude that CT is a promising approach towards making opaque models more trustworthy.

### 2025-07-17 paper/sections/experiments.qmd

@@
 The following example motivates CT and illustrates how to read @tbl-main.
-
-![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable.
-The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange).
-Stars indicate counterfactuals in the target class.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}
-
-### Synthetic Example: Prediction of Credit Card Defaults.
-
-@fig-poc presents results for a linear classifier fitted to *LS* that complies with the data assumptions in Proposition \ref{prp-mtblty}.
-The four panels show the outcomes for different training procedures: in panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we have applied CT.
-For illustrative purposes, suppose the first feature represents *debt* (mutable) and the second feature represents *age* (immutable) of loan applicants seeking counterfactual explanations for moving to the target class: loan provided (orange).
-
-In all four cases, it is possible to generate valid counterfactuals (stars) for unsuccessful applicants (blue).
-They cross the decision boundary (green) into the target class, but their quality differs.
-In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly discernible cluster.
-In panel (b), they are highly plausible, meeting the first objective of Def.
-\ref{def-explainability}.
-This difference in outcomes is quantified for the non-linear MLP in the first two columns of @tbl-main as the $\%$-reduction in implausibility: it is substantial and statistically significant for *LS* across both metrics, the distance-based $\text{IP}$ (29%) and divergence-based $\text{IP}^*$ (55%).
-
-In panel (c) of @fig-poc, the CEs involve substantial reductions in *debt* for younger applicants.
-By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints.
-Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals.
-These CEs are also plausible with respect to the mutable feature, despite requiring smaller debt reductions on average, resulting in smaller costs to individuals.
-This result is quantified for the non-linear case in column three of @tbl-main, which shows the $\%$-reduction in costs averaged across valid counterfactuals.
-Once again, the impact of CT is statistically significant and substantial (14%).
-Thus, we consider the model in panel (d) as the most explainable according to Def.
-\ref{def-explainability}.
-Next, we present the results for all remaining datasets.
 

### 2025-07-17 paper/sections/intro.qmd

--- 2025-07-17 paper/sections/intro.qmd
@@
 Today’s prominence of artificial intelligence (AI) has largely been driven by the success of representation learning with large degrees of freedom: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning highly complex representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
-These models tend to be so complex and opaque that humans cannot easily interpret their decision-making.
+These models tend to be so complex that humans cannot easily interpret their decision-making.
 
-Counterfactual explanations (CE) have become an integral part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable] that can be employed to make sense of this complexity.
+Counterfactual explanations (CE) have become a key part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable] that can be applied to make sense of this complexity.
 Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output (@fig-poc).
-This is useful and necessary to not only understand how opaque models make predictions, but also to provide recourse to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque AI decision-making system (ADM).
+This is useful and necessary to not only understand how opaque models make their predictions, but also to provide algorithmic recourse (AR) to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque automated decision-making (ADM) system.
 
-For feedback to be meaningful, counterfactual explanations need to fulfill certain desiderata [@verma2020counterfactual; @karimi2020survey]: they should be faithful to the model [@altmeyer2024faithful], plausible [@joshi2019realistic] and actionable [@ustun2019actionable].
-Plausibility is typically understood as counterfactuals being *in-domain*: unsuccessful loan applicants that implement recourse should end up with a credit profile that is genuinely similar to that of individuals who have successfully repaid their loans in the past.
-Actionable explanations comply with practical constraints: a young, unsuccessful loan applicant cannot increase their age in an instance.
+For such feedback to be meaningful, counterfactual explanations need to fulfill certain desiderata [@verma2020counterfactual; @karimi2020survey]---they should be faithful to the model [@altmeyer2024faithful], plausible [@joshi2019realistic] and actionable [@ustun2019actionable].
+Plausibility is typically understood as counterfactuals being *in-domain*: unsuccessful loan applicants that implement the provided recourse should end up with credit profiles that are genuinely similar to that of individuals who have successfully repaid their loans in the past.
+Actionable explanations comply with practical constraints: a young, unsuccessful loan applicant cannot increase their age in an instant.
 
@@
 This is problematic, because a narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful].
-In this work, we therefore follow the paradigm that models---as opposed to explanation methods---should be held accountable for plausible and actionable explanations.
+In this work, we therefore follow the paradigm that models---as opposed to explanation methods---should be held accountable for explanations that are plausible and actionable.
 While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models to improve their overall explanatory capacity.
@@
 @fig-poc illustrates the outcome of CT compared to conventionally trained models: in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange); in panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals; in panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features; by contrast, the decision boundary in panel (c) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature.
-To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful CE with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals.
-To the best of our knowledge, CT represents the first venture in this direction with promising results, both empirical and theoretical.
+To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful (*mature*) CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals.
+To the best of our knowledge, CT represents the first venture in this direction with promising empirical and theoretical results.
 
@@
 In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints.
-In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) implausibility of CE decreases by 16% on average; (2) the cost of reaching valid counterfactuals with protected features decreases by 15%; and (3) robust accuracy increases for all datasets by up to 42 percentage points.
-Finally, in @sec-discussion, we discuss open challenges and conclude.
+In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by 16% on average; (2) the cost of reaching valid counterfactuals with protected features decreases by 15%; and (3) the robust accuracy increases by up to 42 percentage points.
+Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.
 
@@
 Stars indicate counterfactuals in the target class.
-Synthetic features are annotated as *debt* and *age* for illustrative purposes.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}
+For illustration, synthetic features are annotated as *debt* and *age*.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

### 2025-07-17 paper/sections/other/main_results.qmd

@@
 
-```{julia}
-#| eval: false
-
-df = Serialization.deserialize("paper/experiments/output/final_table.jls")
-inputs = get_table_inputs(df, nothing; backend=Val(:latex))
-formatters = (
-    PrettyTables.ft_round(1,findall(.!contains.(names(df),"Acc"))),
-    PrettyTables.ft_round(2,findall(contains.(names(df),"Acc")))
-)
-for fname in ["paper/quarto_aaai/tables/main.tex","paper/preprint/tables/main.tex"]
-    tabulate_results(
-        inputs; 
-        table_type=:tabular, 
-        save_name=fname, 
-        wrap_table=false, 
-        alignment=[:l,fill(:r,size(df,2)-1)...],
-        formatters=formatters
-    )
-end
-```
+## Confidence Intervals
 
 ```{julia}
-#| eval: false
+lvls = [0.90, 0.95, 0.99]
 
-# Standard errors with hyperparameter uncertainty marginalized out:
-df = final_table(res_dir, total_uncertainty=true)
-inputs = get_table_inputs(df, nothing; backend=Val(:latex))
-formatters = (
-    PrettyTables.ft_round(1,findall(.!contains.(names(df),"Acc"))),
-    PrettyTables.ft_round(2,findall(contains.(names(df),"Acc")))
+# Mutability:
+y = "distance"
+df_mtbl = aggregate_ce_evaluation(
+    mtbl_dir; 
+    y,
+    return_sig_level=false, 
+    conf_int=lvls,
+    agg_further_vars=["run", "lambda_energy_eval"],
+    drop_models=["linear"]
 )
-for fname in ["paper/quarto_aaai/tables/main_total.tex","paper/preprint/tables/main_total.tex"]
-    tabulate_results(
-        inputs; 
-        table_type=:tabular, 
-        save_name=fname, 
-        wrap_table=false, 
-        alignment=[:l,fill(:r,size(df,2)-1)...],
-        formatters=formatters
+df_mtbl.measure .= y
+
+# Distance-based:
+y = "plausibility_distance_from_target"
+df_ip = aggregate_ce_evaluation(
+    res_dir;
+    y,
+    return_sig_level=false, 
+    conf_int=lvls,
+    agg_further_vars=["run", "lambda_energy_eval"],
 ) 
-end
+df_ip.measure .= y
+
+# Divergence-based:
+y = "mmd"
+df_mmd = aggregate_ce_evaluation(
+    res_dir;
+    y
+    return_sig_level=false, 
+    conf_int=lvls,
+    agg_further_vars=["run", "lambda_energy_eval"],
+)
+df_mmd.measure .= y
+df = innerjoin(df_mtbl, df_ip, df_mmd; on=:dataset)
 ```

### 2025-07-17 paper/sections/related.qmd

@@
----
-bibliography: bibliography.bib
----
-
 # Related Literature {#sec-lit}
 
-To make the desiderata for our framework more concrete, we follow previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial].
-For simplicity, we refer to 'explanatory capacity' simply as 'explainability' in what follows.
+To make the desiderata for our framework more concrete, we follow the previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial].
+For simplicity, we refer to "explanatory capacity" as "explainability" in the rest of this manuscript (see Def.
+\ref{def-explainability} below for a more formal exposition).
 
@@
 In a closely related work, @altmeyer2024faithful show that model averaging and, in particular, contrastive model objectives can produce more explainable and hence trustworthy models.
-To do so, the authors first propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
+The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
 Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
 Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
-The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your]---a form of constrastive learning---tend to do well in this regard.
+The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your], a form of constrastive learning, tend to do well in this regard.
 

### 2025-07-17 paper/quarto_aaai/paper.tex

--- 2025-07-17 paper/quarto_aaai/paper.tex
@@
 objectives such as predictive accuracy \citep{goodfellow2016deep}. These
-models tend to be so complex and opaque that humans cannot easily
-interpret their decision-making.
+models tend to be so complex that humans cannot easily interpret their
+decision-making.
 
-Counterfactual explanations (CE) have become an integral part of the
-broader explainable AI (XAI) toolkit \citep{molnar2022interpretable}
-that can be employed to make sense of this complexity. Originally
-proposed by \citet{wachter2017counterfactual}, CEs prescribe minimal
-changes for factual inputs that, if implemented, would prompt some
-fitted model to produce an alternative, more desirable output
-(Figure~\ref{fig-poc}). This is useful and necessary to not only
-understand how opaque models make predictions, but also to provide
-recourse to individuals subjected to them: a retail bank, for example,
-could use CE to provide meaningful feedback to unsuccessful loan
-applicants that were rejected based on an opaque AI decision-making
-system (ADM).
+Counterfactual explanations (CE) have become a key part of the broader
+explainable AI (XAI) toolkit \citep{molnar2022interpretable} that can be
+applied to make sense of this complexity. Originally proposed by
+\citet{wachter2017counterfactual}, CEs prescribe minimal changes for
+factual inputs that, if implemented, would prompt some fitted model to
+produce an alternative, more desirable output (Figure~\ref{fig-poc}).
+This is useful and necessary to not only understand how opaque models
+make their predictions, but also to provide algorithmic recourse (AR) to
+individuals subjected to them: a retail bank, for example, could use CE
+to provide meaningful feedback to unsuccessful loan applicants that were
+rejected based on an opaque automated decision-making (ADM) system.
 
-For feedback to be meaningful, counterfactual explanations need to
+For such feedback to be meaningful, counterfactual explanations need to
 fulfill certain desiderata
-\citep{verma2020counterfactual, karimi2020survey}: they should be
+\citep{verma2020counterfactual, karimi2020survey}---they should be
 faithful to the model \citep{altmeyer2024faithful}, plausible
@@
 Plausibility is typically understood as counterfactuals being
-\emph{in-domain}: unsuccessful loan applicants that implement recourse
-should end up with a credit profile that is genuinely similar to that of
-individuals who have successfully repaid their loans in the past.
-Actionable explanations comply with practical constraints: a young,
-unsuccessful loan applicant cannot increase their age in an instance.
+\emph{in-domain}: unsuccessful loan applicants that implement the
+provided recourse should end up with credit profiles that are genuinely
+similar to that of individuals who have successfully repaid their loans
+in the past. Actionable explanations comply with practical constraints:
+a young, unsuccessful loan applicant cannot increase their age in an
+instant.
 
@@
 paradigm that models---as opposed to explanation methods---should be
-held accountable for plausible and actionable explanations. While
-previous work has shown that at least plausibility can be indirectly
-achieved through existing techniques aimed at models' generative
-capacity, generalization and robustness
+held accountable for explanations that are plausible and actionable.
+While previous work has shown that at least plausibility can be
+indirectly achieved through existing techniques aimed at models'
+generative capacity, generalization and robustness
 \citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
@@
 To achieve these outcomes, CT draws inspiration from the literature on
-contrastive and robust learning: we contrast faithful CE with
-ground-truth data while protecting immutable features, and capitalize on
-methodological links between CE and AE by penalizing the model's
-adversarial loss on interim (\emph{nascent}) counterfactuals. To the
-best of our knowledge, CT represents the first venture in this direction
-with promising results, both empirical and theoretical.
+contrastive and robust learning: we contrast faithful (\emph{mature})
+CEs with ground-truth data while protecting immutable features, and
+capitalize on methodological links between CE and AE by penalizing the
+model's adversarial loss on interim (\emph{nascent}) counterfactuals. To
+the best of our knowledge, CT represents the first venture in this
+direction with promising empirical and theoretical results.
 
@@
 (Section~\ref{sec-experiments}), we find that thanks to counterfactual
-training, (1) implausibility of CE decreases by 16\% on average; (2) the
-cost of reaching valid counterfactuals with protected features decreases
-by 15\%; and (3) robust accuracy increases for all datasets by up to 42
-percentage points. Finally, in Section~\ref{sec-discussion}, we discuss
-open challenges and conclude.
+training, (1) the implausibility of CEs decreases by 16\% on average;
+(2) the cost of reaching valid counterfactuals with protected features
+decreases by 15\%; and (3) the robust accuracy increases by up to 42
+percentage points. Finally, we discuss open challenges in
+Section~\ref{sec-discussion} and conclude in
+Section~\ref{sec-conclusion}.
 
@@
 colored according to ground-truth labels: \(y^-=1\) (blue) and \(y^+=2\)
-(orange). Stars indicate counterfactuals in the target class. Synthetic
-features are annotated as \emph{debt} and \emph{age} for illustrative
-purposes.}
+(orange). Stars indicate counterfactuals in the target class. For
+illustration, synthetic features are annotated as \emph{debt} and
+\emph{age}.}
 
@@
 
-To make the desiderata for our framework more concrete, we follow
+To make the desiderata for our framework more concrete, we follow the
 previous work in tying the explanatory capacity of models to the quality
@@
 \citep{altmeyer2024faithful, augustin2020adversarial}. For simplicity,
-we refer to `explanatory capacity' simply as `explainability' in what
-follows (see Def. \ref{def-explainability} below for a more formal
-exposition).
+we refer to ``explanatory capacity'' as ``explainability'' in the rest
+of this manuscript (see Def. \ref{def-explainability} below for a more
+formal exposition).
 
@@
 averaging and, in particular, contrastive model objectives can produce
-more explainable and hence trustworthy models. To do so, the authors
-first propose a way to generate counterfactuals that are maximally
-faithful to the model in that they are consistent with what the model
-has learned about the underlying data. Formally, they rely on tools from
-energy-based modelling \citep{teh2003energy} to minimize the divergence
-between the distribution of counterfactuals and the conditional
-posterior over inputs learned by the model. Their counterfactual
-explainer, \emph{ECCCo}, yields plausible explanations if and only if
-the underlying model has learned representations that align with them.
-The authors find that both deep ensembles
-\citep{lakshminarayanan2016simple} and joint energy-based models (JEMs)
-\citep{grathwohl2020your}---a form of constrastive learning---tend to do
-well in this regard.
+more explainable and hence trustworthy models. The authors propose a way
+to generate counterfactuals that are maximally faithful to the model in
+that they are consistent with what the model has learned about the
+underlying data. Formally, they rely on tools from energy-based
+modelling \citep{teh2003energy} to minimize the divergence between the
+distribution of counterfactuals and the conditional posterior over
+inputs learned by the model. Their counterfactual explainer,
+\emph{ECCCo}, yields plausible explanations if and only if the
+underlying model has learned representations that align with them. The
+authors find that both deep ensembles \citep{lakshminarayanan2016simple}
+and joint energy-based models (JEMs) \citep{grathwohl2020your}, a form
+of constrastive learning, tend to do well in this regard.
 
@@
 
-We also highlight three important directions for future research.
-Firstly, it is an interesting challenge to extend CT beyond
-classification settings. Our formulation relies on the distinction
-between non-target class(es) \(y^{-}\) and target class(es) \(y^{+}\) to
-generate counterfactuals through Equation~\ref{eq-obj}. While \(y^{-}\)
-and \(y^{+}\) can be arbitrarily defined, CT requires the output space
-\(\mathcal{Y}\) to be discrete. Thus, it does not apply to ML tasks
-where the change in outcome cannot be readily discretized. Focus on
-classification is a common choice in research on CEs and AR. Other
-settings have attracted some interest, e.g., regression
-\citep{spooner2021counterfactual}, but there is little consensus how to
-robustly extend the notion of CEs.\\
-Secondly, CT is susceptible to training instabilities. This problem has
-been recognized for JEMs \citep{grathwohl2020your} and even though we
-depart from the SGLD sampling, we still encounter variability in
-outcomes. CT is exposed to two potential sources of instabilities: (1)
-the energy-based contrastive divergence term, \(\text{div}(\cdot)\), in
-Equation~\ref{eq-div}, and (2) the underlying explainers. We find
-several promising ways to mitigate this problem: regularizing energy
-(\(\lambda_{\text{reg}}\)), generating sufficiently many counterfactuals
-during each epoch, and including only mature counterfactuals in
-\(\text{div}(\cdot)\).\\
+We also highlight three key directions for future research. Firstly, it
+is an interesting challenge to extend CT beyond classification settings.
+Our formulation relies on the distinction between non-target class(es)
+and target class(es), requiring the output space to be discrete. Thus,
+it does not apply to ML tasks where the change in outcome cannot be
+readily discretized. Focus on classification is a common choice in
+research on CEs and AR; other settings have attracted some interest,
+e.g., regression \citep{spooner2021counterfactual}, but there is little
+consensus how to robustly extend the notion of CEs.\\
+Secondly, CT is susceptible to training instabilities, as also
+recognized for the related JEMs \citep{grathwohl2020your}. CT is exposed
+to two sources of instabilities: (1) the energy-based contrastive
+divergence term, \(\text{div}(\cdot)\), in Equation~\ref{eq-div}, and
+(2) the underlying explainers. We find several promising ways to
+mitigate this problem: regularizing energy (\(\lambda_{\text{reg}}\)),
+generating sufficiently many counterfactuals during each epoch, and
+including only mature counterfactuals in \(\text{div}(\cdot)\).\\
 Finally, we believe that it is possible to considerably improve
 hyperparameter selection procedures, and thus performance. We have
-relied exclusively on grid searches, but future research could benefit
-from more sophisticated approaches.
+relied exclusively on grid searches, but future work could benefit from
+more sophisticated approaches.
 
-To conclude, state-of-the-art machine learning models are prone to
-learning complex representations that cannot be interpreted by humans.
-Existing explainability solutions cannot guarantee that explanations
-agree with the model's learned representation of data. As a step towards
-addressing this challenge, we introduced counterfactual training, a
-novel training regime that incentivizes highly-explainable models. Our
-approach leads to explanations that are both plausible (compliant with
-the underlying data-generating process) and actionable (compliant with
-user-specified mutability constraints), and thus meaningful to their
-recipients. Through extensive experiments, we demonstrated that CT
-satisfies its objective while promoting robustness and preserving the
-predictive performance of models. It can also be used to fine-tune
-conventionally-trained models and achieve similar gains. Lastly, our
-work highlights the value of simultaneously improving models and their
-explanations.
+\section{Conclusion}\label{sec-conclusion}
+
+State-of-the-art machine learning models are prone to learning complex
+representations that cannot be interpreted by humans. Existing
+explainability solutions cannot guarantee that explanations agree with
+these learned representation. As a step towards addressing this
+challenge, we introduce counterfactual training, a novel training regime
+that incentivizes highly-explainable models. Through extensive
+experiments, we demonstrate that CT satisfies this objective while
+promoting robustness and preserving the predictive performance of
+models. Explanations generated from CT-based models are both more
+plausible (compliant with the underlying data-generating process) and
+more actionable (compliant with user-specified mutability constraints),
+and thus meaningful to their recipients. We also show that CT can be
+used to fine-tune conventionally-trained models and achieve similar
+gains. Lastly, our work highlights the value of simultaneously improving
+models and their explanations.
 

### 2025-07-17 paper/sections/other/main_results.qmd
table updates

@@
 Serialization.serialize("paper/experiments/output/final_table.jls", df)
+```
+
+```{julia}
+#| eval: false
+
+df = Serialization.deserialize("paper/experiments/output/final_table.jls")
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+formatters = (
+    PrettyTables.ft_round(1,findall(.!contains.(names(df),"Acc"))),
+    PrettyTables.ft_round(2,findall(contains.(names(df),"Acc")))
+)
+for fname in ["paper/quarto_aaai/tables/main.tex","paper/preprint/tables/main.tex"]
+    tabulate_results(
+        inputs; 
+        table_type=:tabular, 
+        save_name=fname, 
+        wrap_table=false, 
+        alignment=[:l,fill(:r,size(df,2)-1)...],
+        formatters=formatters
+    )
+end
+```
+
+### Robust Performance Plots
+
+```{julia}
+plt = CTExperiments.plot_performance(res_dir, eps=range(0.0, 0.1, length=10) |> collect, adversarial=true, byvars=["objective", "eps"]) |>
+    plt -> draw(plt, figure=(size=(800, 190),), axis=(yticks=[0.0,0.5,1.0], limits=(nothing, (0,1)), xticklabelsvisible=false, xticksvisible=false))
+save("paper/figures/acc.png", plt; px_per_unit=3)
 ```

### 2025-07-17 paper/experiments/CTExperiments/src/presentation.jl

@@ -869,7 +869,7 @@ function aggregate_performance(
     return df
 end
 
-function plot_performance(res_dir; measure::Vector=["acc"], adversarial::Bool=false, bootstrap::Union{Nothing,Int}=nothing, drop_models::Vector{String}=String[], eps::Vector{<:AbstractFloat}=[0.03], byvars=["objective"]
+function plot_performance(res_dir; measure::Vector=["acc"], adversarial::Bool=false, bootstrap::Union{Nothing,Int}=nothing, drop_models::Vector{String}=String[], eps::Vector{<:AbstractFloat}=[0.03], byvars=["objective"], drop_synthetic::Bool=true
 )
 
     # Get measures:
@@ -879,8 +879,12 @@ function plot_performance(res_dir; measure::Vector=["acc"], adversarial::Bool=fa
 
     eval_grids, _ = final_results(res_dir; drop_models)
     df = aggregate_performance(eval_grids; measure, adversarial, bootstrap, eps, byvars)
+    df.objective .= ifelse.(df.objective .== "Full", "CT", "BL")
+    if drop_synthetic
+        filter!(df -> !(df.dataset in ["LS", "OL", "Circ", "Moon"]), df)
+    end
 
-    plt = data(df) * mapping(:eps, :mean, color=:objective, col=:dataset) * visual(ScatterLines)
+    plt = data(df) * mapping(:eps => L"\text{Perturbation Size: }\varepsilon \in [0,0.1]", :mean => L"\text{Accuracy}", color=:objective => "Model", marker=:objective => "Model", col=:dataset) * visual(ScatterLines)
     return plt
 end
 
@@ -916,7 +920,8 @@ function final_table(
     bootstrap::Int=100,
     total_uncertainty::Bool=false,
     drop_models::Vector{String}=String[],
-    conf_int::Vector{<:AbstractFloat}=[0.99]
+    conf_int::Vector{<:AbstractFloat}=[0.99],
+    include_performance::Bool=false,
 )
     # CE:
     df_ce = DataFrame()
@@ -935,16 +940,19 @@ function final_table(
     # Missing:
     df_ce = coalesce.(df_ce, "(agg.)")
 
-    # Performance:
-    df_perf = aggregate_performance(res_dir; measure=perf_var, bootstrap, drop_models)                          # unperturbed
-    df_adv_perf = aggregate_performance(res_dir; measure=perf_var, adversarial=true, bootstrap, drop_models)    # adversarial
[…]

### 2025-07-17 paper/sections/experiments.qmd
not sure what the hell quarto is doing now

@@
 
+::: {#tbl-main}
+
+```{=latex}
+\input{tables/main.tex}
+```
+
+Key counterfactual evaluation metrics and bootstrapped standard errors for all datasets.
+**Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features.
+Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time.
+Asterisks ($^*$) indicate that bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
+
+:::
+
 ### Plausibility. {#sec-plaus}
@@
 
-::: {#tbl-main}
-
-```{=latex}
-\begin{table*}
-\input{tables/main.tex}
-\end{table*}
-```
-
-Key performance metrics and bootstrapped standard errors (omitted if essentially zero) for all datasets.
-**Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features.
-**Accuracy** (columns 4-7): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL).
-Counterfactual outcomes (col.
-1-3) are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time.
-Asterisks ($^*$) in columns 1-3 indicate that bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
-Standard errors for accuracy (col.
-4-7) are bootstrapped from the test set.
-:::
+![Test accuracies on adversarially perturbed data with varying perturbation sizes for all non-synthetic datasets.](paper/figures/acc.png){#fig-acc fig-env="figure*"}
 

### 2025-07-17 paper/quarto_aaai/paper.tex
not sure what the hell quarto is doing now
bloody hell

--- 2025-07-17 paper/quarto_aaai/paper.tex
@@
 objectives such as predictive accuracy \citep{goodfellow2016deep}. These
-models tend to be so complex and opaque that humans cannot easily
-interpret their decision-making.
+models tend to be so complex that humans cannot easily interpret their
+decision-making.
 
@@
 
+\begin{table}
+
+\caption{\label{tbl-main}Key counterfactual evaluation metrics and
+bootstrapped standard errors for all datasets. \textbf{Plausibility}
+(columns 1-2): percentage reduction in implausibility for \(\text{IP}\)
+and \(\text{IP}^*\), respectively; \textbf{Actionability} (column 3):
+percentage reduction in costs with protected features. Outcomes are
+aggregated across bootstrap samples and varying degrees of the energy
+penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
+Asterisks (\(^*\)) indicate that bootstrapped 99\%-confidence interval
+of differences in mean outcomes does \emph{not} include zero.}
+
+\centering{
+
+\input{tables/main.tex}
+
+}
+
+\end{table}%
+
 \subsubsection{Plausibility.}\label{sec-plaus}
@@
 
-\begin{table*}
-
-\caption{\label{tbl-main}Key performance metrics and bootstrapped
-standard errors (omitted if essentially zero) for all datasets.
-\textbf{Plausibility} (columns 1-2): percentage reduction in
-implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
-\textbf{Actionability} (column 3): percentage reduction in costs with
-protected features. \textbf{Accuracy} (columns 4-7): test accuracies and
-robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
-Counterfactual outcomes (col.~1-3) are aggregated across bootstrap
-samples and varying degrees of the energy penalty
-\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Asterisks
-(\(^*\)) in columns 1-3 indicate that bootstrapped 99\%-confidence
-interval of differences in mean outcomes does \emph{not} include zero.
-Standard errors for accuracy (col.~4-7) are bootstrapped from the test
-set.}
+\begin{figure*}
 
@@
 
-\input{tables/main.tex}
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/acc.png}}
 
@@
 
-\end{table*}%
+\caption{\label{fig-acc}Test accuracies on adversarially perturbed data
+with varying perturbation sizes for all non-synthetic datasets.}
+
+\end{figure*}%
 

### 2025-07-17 paper/sections/experiments.qmd
bloody hell

@@
 
-![Test accuracies on adversarially perturbed data with varying perturbation sizes for all non-synthetic datasets.](paper/figures/acc.png){#fig-acc fig-env="figure*"}
+![Test accuracies on adversarially perturbed data with varying perturbation sizes for all non-synthetic datasets.](/paper/figures/acc.png){#fig-acc fig-env="figure*"}
 

### 2025-07-18 paper/sections/other/main_results.qmd
accuracy chart including synthetic data

@@
 using CTExperiments
+using CTExperiments.CairoMakie
 using CTExperiments.CSV
@@
 ```{julia}
-plt = CTExperiments.plot_performance(res_dir, eps=range(0.0, 0.1, length=10) |> collect, adversarial=true, byvars=["objective", "eps"]) |>
-    plt -> draw(plt, figure=(size=(800, 190),), axis=(yticks=[0.0,0.5,1.0], limits=(nothing, (0,1)), xticklabelsvisible=false, xticksvisible=false))
-save("paper/figures/acc.png", plt; px_per_unit=3)
+plt = CTExperiments.plot_performance(res_dir, eps=range(0.0, 0.1, length=10) |> collect, adversarial=true, byvars=["objective", "eps"], drop_synthetic=false)
+plt_out = draw(plt, figure=(size=(800, 130),), axis=(yticks=[0.0,0.5,1.0], limits=(nothing, (0,1)), xticklabelsvisible=false, xticksvisible=false))
+save("paper/figures/acc.png", plt_out; px_per_unit=3)
 ```

### 2025-07-18 paper/experiments/CTExperiments/src/presentation.jl

@@ -1,4 +1,5 @@
 using AlgebraOfGraphics
+using CategoricalArrays
 using CounterfactualExplanations
 using CairoMakie
 using DataFrames
@@ -86,6 +87,7 @@ global LatexReplacements = Dict(
     "circles" => "Circ",
     "moons" => "Moon",
     "credit" => "Cred",
+    "adult" => "Adult",
 )
 
 function split_at_parentheses_precise(s::String)
@@ -869,6 +871,18 @@ function aggregate_performance(
     return df
 end
 
+const ds_order = [
+    LatexReplacements["lin_sep"],
+    LatexReplacements["circles"],
+    LatexReplacements["moons"],
+    LatexReplacements["over"],
+    LatexReplacements["adult"],
+    LatexReplacements["cali"],
+    LatexReplacements["credit"],
+    LatexReplacements["gmsc"],
+    LatexReplacements["mnist"],
+]
+
 function plot_performance(res_dir; measure::Vector=["acc"], adversarial::Bool=false, bootstrap::Union{Nothing,Int}=nothing, drop_models::Vector{String}=String[], eps::Vector{<:AbstractFloat}=[0.03], byvars=["objective"], drop_synthetic::Bool=true
 )
 
@@ -880,6 +894,7 @@ function plot_performance(res_dir; measure::Vector=["acc"], adversarial::Bool=fa
     eval_grids, _ = final_results(res_dir; drop_models)
     df = aggregate_performance(eval_grids; measure, adversarial, bootstrap, eps, byvars)
     df.objective .= ifelse.(df.objective .== "Full", "CT", "BL")
+    df.dataset .= categorical(df.dataset, levels=ds_order)
     if drop_synthetic
         filter!(df -> !(df.dataset in ["LS", "OL", "Circ", "Moon"]), df)
     end

### 2025-07-18 paper/quarto_aaai/paper.tex
table adjusted

@@
 
+\small
+\centering
 \input{tables/main.tex}
@@
 
-Test accuracy for CT is virtually identical to the baseline for
-\emph{Adult}, \emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and
-even slightly improved for \emph{Cred}. Exceptions to this general
-pattern are \emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we
-observe a reduction in test accuracy of 2, 5, and 15 percentage points
-respectively. When looking at robust test accuracies (Acc.\(^*\)) for
-these datasets in particular, we find that CT strongly outperforms the
-baseline. In fact, we observe that CT improves adversarial robustness on
-all datasets.
+Test accuracies on adversarially perturbed data are shown in
+Figure~\ref{fig-acc}. The perturbations size,
+\(\varepsilon \in [0,0.1]\), increases along the horizontal axis and
+includes zero, corresponding to standard test accuracy for non-perturbed
+data. We observe that for all synthetic datasets, predictive performance
+of CT is virtually identical to the baseline and unaffected by
+adversarial perturbations. For all real-world datasets, we find that CT
+substantially improves robustness: while in some cases baseline
+accuracies drop to essentially zero for large enough perturbation sizes,
+accuracies for CT are remarkably robust to gradient-based adversarial
+attacks.
 

### 2025-07-18 paper/sections/experiments.qmd
table adjusted

@@
 ```{=latex}
+\begin{table}[h]
+\small
+\centering
 \input{tables/main.tex}
+\end{table}
 ```
@@
 
-Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon*, and *OL*, and even slightly improved for *Cred*.
-Exceptions to this general pattern are *MNIST*, *CH*, and *GMSC*, for which we observe a reduction in test accuracy of 2, 5, and 15 percentage points respectively.
-When looking at robust test accuracies (Acc.$^*$) for these datasets in particular, we find that CT strongly outperforms the baseline.
-In fact, we observe that CT improves adversarial robustness on all datasets.
+Test accuracies on adversarially perturbed data are shown in @fig-acc.
+The perturbations size, $\varepsilon \in [0,0.1]$, increases along the horizontal axis and includes zero, corresponding to standard test accuracy for non-perturbed data.
+We observe that for all synthetic datasets, predictive performance of CT is virtually identical to the baseline and unaffected by adversarial perturbations.
+For all real-world datasets, we find that CT substantially improves robustness: while in some cases baseline accuracies drop to essentially zero for large enough perturbation sizes, accuracies for CT are remarkably robust to gradient-based adversarial attacks.
 

### 2025-07-18 paper/sections/other/main_results.qmd
table adjusted

--- 2025-07-17 paper/sections/other/main_results.qmd
@@
 Serialization.serialize("paper/experiments/output/final_table.jls", df)
+```
+
+```{julia}
+#| eval: false
+
+df = Serialization.deserialize("paper/experiments/output/final_table.jls")
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+for fname in ["paper/quarto_aaai/tables/main.tex","paper/preprint/tables/main.tex"]
+    tabulate_results(
+        inputs; 
+        table_type=:tabular, 
+        save_name=fname, 
+        wrap_table=false, 
+    )
+end
+```
+
+### Robust Performance Plots
+
+```{julia}
+plt = CTExperiments.plot_performance(res_dir, eps=range(0.0, 0.1, length=10) |> collect, adversarial=true, byvars=["objective", "eps"], drop_synthetic=false)
+plt_out = draw(plt, figure=(size=(1000, 150),), axis=(yticks=[0.0,0.5,1.0], limits=(nothing, (0,1)), xticklabelsvisible=false, xticksvisible=false))
+save("paper/figures/acc.png", plt_out; px_per_unit=3)
 ```

### 2025-07-18 paper/experiments/CTExperiments/src/presentation.jl

@@ -937,6 +937,7 @@ function final_table(
     drop_models::Vector{String}=String[],
     conf_int::Vector{<:AbstractFloat}=[0.99],
     include_performance::Bool=false,
+    add_aggregates::Bool=true,
 )
     # CE:
     df_ce = DataFrame()
@@ -975,6 +976,18 @@ function final_table(
         df.sig_level .= ""
     else
         df.sig_level .=  coalesce.(df.sig_level, "")
+    end
+    
+    if add_aggregates
+        df_agg =combine(groupby(df, :variable), :mean => mean => :mean)
+        df_agg.dataset .= "Avg."
+        df_agg.se .= NaN
+        df = vcat(df, df_agg; cols=:union)
+        df = coalesce.(df, "")
+        df.var_id = (x -> x.data).(df.variable)
+        sort!(df, [:var_id, :dataset])
+        select!(df, Not(:var_id))
+        display(df)
     end
 
     # Post-process

### 2025-07-18 paper/quarto_aaai/paper.tex
final adjustments to text

@@
 
-Specifically, \textbf{we propose Counterfactual Training (CT)}: a novel
+Specifically, we propose \textbf{counterfactual training (CT)}: a novel
 training regime that leverages counterfactual explanations on-the-fly to
@@
 \text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-}\end{equation} and introduce a novel divergence metric,
+}\end{equation} and introduce a novel divergence-based adaption,
 \begin{equation}\phantomsection\label{eq-impl-div}{
@@
 
-For predictive performance, we use standard metrics, such as robust
-accuracy estimated on adversarially perturbed data using FGSM
+To assess outcomes with respect to actionability for non-linear models,
+we look at the average costs of valid counterfactuals in terms of their
+distances from factual starting points. While this an imperfect proxy of
+sensitivity, we hypothesize that CT can reduce these costs by teaching
+models to seek plausibility with respect to mutable features, much like
+we observe in Figure~\ref{fig-poc} (panel (d) vs panel (c)). We
+supplement this analysis with qualitative findings for integrated
+gradients \citep{sundararajan2017ig}. Finally, for predictive
+performance, we use standard metrics, such as robust accuracy estimated
+on adversarially perturbed data using FGSM
 \citep{goodfellow2014explaining}.
@@
 
-Our main quantitative results for \emph{MLP} models are summarised in
-Table~\ref{tbl-main}, which presents average outcomes along with
-bootstrapped two standard errors. The following example motivates CT and
-illustrates how to read Table~\ref{tbl-main}.
+Our main results for plausibility and actionability for \emph{MLP}
+models are summarised in Table~\ref{tbl-main}, which presents
+counterfactual outcomes grouped by dataset along with standard errors
+averaged across bootstrap samples.
+
+The first two columns (\(\text{IP}\) and \(\text{IP}^*\)) show the
+percentage reduction in implausibility for our two metrics when using CT
+on top of the weak baseline. Focusing on the first row for \emph{LS} for
+a moment, the observed positive values indicate that faithful
+counterfactuals are around 30-55\% more plausible for models trained
+with CT, in line with our observations in Figure~\ref{fig-poc} (panel
+(b) vs.~(a)). Asterisks indicate that reductions are statistically
+significant: bootstrapped 99\%-confidence intervals of differences in
+mean outcomes do \emph{not} include zero.
+
+The third column shows results for the scenario when mutability
+constraints are imposed on selected features. Once again we are
+comparing CT to the baseline, so reductions in the positive direction
+imply that valid counterfactuals are cheaper when using CT with feature
+protection. Relating this back to Figure~\ref{fig-poc}, the third column
+represents the reduction in distances travelled by counterfactuals in
+panel (d) compared to (c). In the following, we summarize the remaining
+results.
 
@@
 
-\caption{\label{tbl-main}Key counterfactual evaluation metrics and
-bootstrapped standard errors for all datasets. \textbf{Plausibility}
-(columns 1-2): percentage reduction in implausibility for \(\text{IP}\)
-and \(\text{IP}^*\), respectively; \textbf{Actionability} (column 3):
-percentage reduction in costs with protected features. Outcomes are
-aggregated across bootstrap samples and varying degrees of the energy
-penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time.
-Asterisks (\(^*\)) indicate that bootstrapped 99\%-confidence interval
-of differences in mean outcomes does \emph{not} include zero.}
+\caption{\label{tbl-main}Key evaluation metrics for valid counterfactual
+along with bootstrapped standard errors for all datasets.
+\textbf{Plausibility} (columns 1-2): percentage reduction in
+implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
+\textbf{Actionability} (column 3): percentage reduction in costs with
+protected features. Outcomes are aggregated across bootstrap samples and
+varying degrees of the energy penalty \(\lambda_{\text{egy}}\) used for
+\emph{ECCo} at test time. Asterisks (\(^*\)) indicate that bootstrapped
+99\%-confidence interval of differences in mean outcomes does \emph{not}
+include zero.}
 
@@
 
-We also find that CT can reduce sensitivity to immutable, protected
-features and thus lead to less costly counterfactual outcomes as shown
-in Figure~\ref{fig-poc}. In column three of Table~\ref{tbl-main}, we
-impose mutability constraints on selected features and compute the
-reduction in average costs of CEs associated with CT compared to the
-baseline: for synthetic datasets, we always protect the first feature;
-for all real-world tabular datasets we could identify and protect an
-\emph{age} variable; for \emph{MNIST}, we protect the five upper and
-lower pixel rows of the full image. Reductions in costs are
-overwhelmingly positive and significant of up to nearly 60\% for
-\emph{GMSC}. While the estimated cost reductions for \emph{Adult} and
-\emph{MNIST} are not significant, Figure~\ref{fig-mnist} (columns 6-10)
-demonstrates that CT does have the expected effect: sensitivity to
-protected features as per integrated gradients is drastically reduced;
-details of this experiment are reported in the supplementary appendix.
-In the case of \emph{Cred}, average costs increase, likely because any
-potential benefits from protecting the \emph{age} are outweighed by the
-increase in costs required for greater plausibility.
+We also find that CT can improve actionability in the presence of
+immutable features (column 3 in Table~\ref{tbl-main}): for synthetic
+datasets, we always protect the first feature; for all real-world
+tabular datasets we could identify and protect an \emph{age} variable;
+for \emph{MNIST}, we protect the five upper and lower pixel rows of the
+full image. Reductions in costs are overwhelmingly positive and
+significant of up to nearly 60\% for \emph{GMSC}. While the estimated
+cost reductions for \emph{Adult} and \emph{MNIST} are not significant,
+Figure~\ref{fig-mnist} (columns 6-10) demonstrates that in the latter
+case, CT does have the expected effect: sensitivity to protected
+features as per integrated gradients is drastically reduced; details of
+this experiment are reported in the supplementary appendix. In the case
+of \emph{Cred}, average costs increase, likely because any potential
+benefits from protecting the \emph{age} are outweighed by the increase
+in costs required for greater plausibility.
 

### 2025-07-18 paper/sections/experiments.qmd
final adjustments to text

@@
 $$ {#eq-impl-dist}
-and introduce a novel divergence metric,
+and introduce a novel divergence-based adaption,
 $$
@@
 
-For predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].
+To assess outcomes with respect to actionability for non-linear models, we look at the average costs of valid counterfactuals in terms of their distances from factual starting points.
+While this an imperfect proxy of sensitivity, we hypothesize that CT can reduce these costs by teaching models to seek plausibility with respect to mutable features, much like we observe in @fig-poc (panel (d) vs panel (c)).
+We supplement this analysis with qualitative findings for integrated gradients [@sundararajan2017ig].
+Finally, for predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].
 
@@
 
-Our main quantitative results for *MLP* models are summarised in @tbl-main, which presents average outcomes along with bootstrapped two standard errors.
-The following example motivates CT and illustrates how to read @tbl-main.
+Our main results for plausibility and actionability for *MLP* models are summarised in @tbl-main, which presents counterfactual outcomes grouped by dataset along with standard errors averaged across bootstrap samples.
+
+The first two columns ($\text{IP}$ and $\text{IP}^*$) show the percentage reduction in implausibility for our two metrics when using CT on top of the weak baseline.
+Focusing on the first row for *LS* for a moment, the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc (panel (b) vs.
+(a)).
+Asterisks indicate that reductions are statistically significant: bootstrapped 99%-confidence intervals of differences in mean outcomes do *not* include zero.
+
+The third column shows results for the scenario when mutability constraints are imposed on selected features.
+Once again we are comparing CT to the baseline, so reductions in the positive direction imply that valid counterfactuals are cheaper when using CT with feature protection.
+Relating this back to @fig-poc, the third column represents the reduction in distances travelled by counterfactuals in panel (d) compared to (c).
+In the following, we summarize the remaining results.
 
@@
 
-Key counterfactual evaluation metrics and bootstrapped standard errors for all datasets.
+Key evaluation metrics for valid counterfactual along with bootstrapped standard errors for all datasets.
 **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features.
@@
 
-We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as shown in @fig-poc.
-In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of CEs associated with CT compared to the baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image.
+We also find that CT can improve actionability in the presence of immutable features (column 3 in @tbl-main): for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image.
 Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*.
-While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix.
+While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that in the latter case, CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix.
 In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

### 2025-07-18 paper/sections/intro.qmd
final adjustments to text

@@
 
-Specifically, **we propose Counterfactual Training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE).
+Specifically, we propose **counterfactual training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE).
 @fig-poc illustrates the outcome of CT compared to conventionally trained models: in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange); in panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals; in panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features; by contrast, the decision boundary in panel (c) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature.

### 2025-07-18 paper/quarto_aaai/paper.tex
final adjustments to text

@@
 (Section~\ref{sec-experiments}), we find that thanks to counterfactual
-training, (1) the implausibility of CEs decreases by 16\% on average;
-(2) the cost of reaching valid counterfactuals with protected features
-decreases by 15\%; and (3) the robust accuracy increases by up to 42
-percentage points. Finally, we discuss open challenges in
+training, (1) the implausibility of CEs decreases by up to 90\%; (2) the
+cost of reaching valid counterfactuals with protected features decreases
+by 19\% on average; and (3) adversarial robustness improves across the
+board. Finally, we discuss open challenges in
 Section~\ref{sec-discussion} and conclude in

### 2025-07-18 paper/sections/intro.qmd
final adjustments to text

@@
 In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints.
-In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by 16% on average; (2) the cost of reaching valid counterfactuals with protected features decreases by 15%; and (3) the robust accuracy increases by up to 42 percentage points.
+In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) adversarial robustness improves across the board.
 Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.

### 2025-07-18 paper/quarto_aaai/paper.tex
Emphasize RQ answers

@@
 \sisetup{uncertainty-mode = separate}
+\usepackage{enumitem}
 
@@
 
-We seek to answer the following three research questions:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  To what extent does the CT objective in Equation 1 induce models to
-  learn plausible explanations?
-\item
-  To what extent does CT lead to more favorable AR outcomes in the
-  presence of actionability constraints?
-\item
-  What are the effects of hyperparameter selection on CT?
+We seek to answer the following four research questions:
+
+\begin{enumerate}[label={(\makebox[2em][c]{RQ\arabic*})}, leftmargin=3.5em]
+    \item To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
+    \item To what extent does CT result in more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+    \item To what extent does CT influence the adversarial robustness of trained models?
+    \item What are the effects of hyperparameter selection on counterfactual training?
 \end{enumerate}
@@
 Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me Some Credit
-(\emph{GMSC}) from \citet{kaggle2011give2}. Finally, for the convenience
+(\emph{GMSC}) from \citet{kaggle2011give}. Finally, for the convenience
 of illustration, we use the 10-class \emph{MNIST}
@@
 counterfactual outcomes grouped by dataset along with standard errors
-averaged across bootstrap samples.
+averaged across bootstrap samples. Asterisks (\(^*\)) indicate that the
+observed effects are statistically significant: bootstrapped
+99\%-confidence intervals of differences in mean outcomes do \emph{not}
+include zero.
 
@@
 percentage reduction in implausibility for our two metrics when using CT
-on top of the weak baseline. Focusing on the first row for \emph{LS} for
-a moment, the observed positive values indicate that faithful
+on top of the weak baseline. As an example, consider the first row
+(\emph{LS}): the observed positive values indicate that faithful
 counterfactuals are around 30-55\% more plausible for models trained
-with CT, in line with our observations in Figure~\ref{fig-poc} (panel
-(b) vs.~(a)). Asterisks indicate that reductions are statistically
-significant: bootstrapped 99\%-confidence intervals of differences in
-mean outcomes do \emph{not} include zero.
-
-The third column shows results for the scenario when mutability
-constraints are imposed on selected features. Once again we are
+with CT, in line with our observations in Figure~\ref{fig-poc} in panel
+(b) compared to panel (a).
+
+The third column shows the results for a scenario when mutability
+constraints are imposed on the selected features. Again, we are
 comparing CT to the baseline, so reductions in the positive direction
-imply that valid counterfactuals are cheaper when using CT with feature
-protection. Relating this back to Figure~\ref{fig-poc}, the third column
-represents the reduction in distances travelled by counterfactuals in
-panel (d) compared to (c). In the following, we summarize the remaining
-results.
+imply that valid counterfactuals are ``cheaper'' (more actionable) when
+using CT with feature protection. Relating this back to
+Figure~\ref{fig-poc}, the third column represents the reduction in
+distances travelled by counterfactuals in panel (d) compared to panel
+(c). In the following paragraphs, we summarize the results for all
+datasets.
 
@@
 implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
-\textbf{Actionability} (column 3): percentage reduction in costs with
-protected features. Outcomes are aggregated across bootstrap samples and
-varying degrees of the energy penalty \(\lambda_{\text{egy}}\) used for
-\emph{ECCo} at test time. Asterisks (\(^*\)) indicate that bootstrapped
-99\%-confidence interval of differences in mean outcomes does \emph{not}
-include zero.}
+\textbf{Cost} / \textbf{Actionability} (column 3): percentage reduction
+in costs when selected features are protected. Outcomes are aggregated
+across bootstrap samples and varying degrees of the energy penalty
+\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Asterisks
+(\(^*\)) indicate that the bootstrapped 99\%-confidence interval of
+differences in mean outcomes does \emph{not} include zero.}
 
@@
 
-\subsubsection{Plausibility.}\label{sec-plaus}
-
-We find that CT generally leads to substantial and statistically
-significant improvements in plausibility: average reductions in
-\(\text{IP}\) range from around 7\% for \emph{MNIST} to almost 60\% for
-\emph{Circ}; for the real-world tabular datasets they are around 12\%
-for both \emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
-\emph{Adult} and \emph{OL} we find no significant impact of CT on
-\(\text{IP}\). Reductions in \(\text{IP}^*\) are even more substantial
-and generally statistically significant, although the average degree of
-uncertainty is higher than for \(\text{IP}\): average reductions range
-from around 20\% (\emph{Moons}) to almost 90\% (\emph{Circ}). The only
-negative findings for \emph{OL} and \emph{MNIST} are statistically
-insignificant and, for MNIST, do not align with qualitative findings,
-which are much more plausible for CT (Figure~\ref{fig-mnist}).
+\subsubsection{Plausibility (RQ1).}\label{sec-plaus}
+
+\emph{CT generally produces substantial and statistically significant
+improvements in plausibility.}\\
+Average reductions in \(\text{IP}\) range from around 7\% for
+\emph{MNIST} to almost 60\% for \emph{Circ}. For the real-world tabular
+datasets they are around 12\% for \emph{CH} and \emph{Cred} and almost
+25\% for \emph{GMSC}; for \emph{Adult} and \emph{OL} we find no
+significant impact of CT on \(\text{IP}\). Reductions in \(\text{IP}^*\)
+are even more substantial and generally statistically significant,
+although the average degree of uncertainty is higher than for
+\(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
+almost 90\% (\emph{Circ}). The only negative findings for \emph{OL} and
+\emph{MNIST} are not statistically significant and, for MNIST, do not
+align with qualitative findings, which are much more plausible for CT
+(Figure~\ref{fig-mnist}).
 
@@
 
-\subsubsection{Actionability.}\label{sec-act}
-
-We also find that CT can improve actionability in the presence of
-immutable features (column 3 in Table~\ref{tbl-main}): for synthetic
-datasets, we always protect the first feature; for all real-world
-tabular datasets we could identify and protect an \emph{age} variable;
-for \emph{MNIST}, we protect the five upper and lower pixel rows of the
-full image. Reductions in costs are overwhelmingly positive and
-significant of up to nearly 60\% for \emph{GMSC}. While the estimated
-cost reductions for \emph{Adult} and \emph{MNIST} are not significant,
-Figure~\ref{fig-mnist} (columns 6-10) demonstrates that in the latter
-case, CT does have the expected effect: sensitivity to protected
-features as per integrated gradients is drastically reduced; details of
-this experiment are reported in the supplementary appendix. In the case
-of \emph{Cred}, average costs increase, likely because any potential
-benefits from protecting the \emph{age} are outweighed by the increase
-in costs required for greater plausibility.
-
-\subsubsection{Predictive Performance.}\label{sec-pred}
-
-Test accuracies on adversarially perturbed data are shown in
-Figure~\ref{fig-acc}. The perturbations size,
-\(\varepsilon \in [0,0.1]\), increases along the horizontal axis and
-includes zero, corresponding to standard test accuracy for non-perturbed
-data. We observe that for all synthetic datasets, predictive performance
-of CT is virtually identical to the baseline and unaffected by
-adversarial perturbations. For all real-world datasets, we find that CT
-substantially improves robustness: while in some cases baseline
-accuracies drop to essentially zero for large enough perturbation sizes,
-accuracies for CT are remarkably robust to gradient-based adversarial
-attacks.
+\subsubsection{Actionability (RQ2).}\label{sec-act}
+
+\emph{CT tends to improve actionability in the presence of immutable
+features, but the exact effects can be challenging to establish due to
+feature interactions.}
+
+For synthetic datasets, we always protect the first feature; for all
+real-world tabular datasets we could identify and protect an \emph{age}
+variable; for \emph{MNIST}, we protect the five upper and lower pixel
+rows of the full image. Reductions in costs are overwhelmingly positive
+and significant of up to around 60\% for \emph{GMSC}. While the
+estimated cost increase for \emph{Adult} and \emph{MNIST}, these effects
+are not significant and Figure~\ref{fig-mnist} (columns 6-10)
+demonstrates that in the latter case, CT does have the expected effect:
+sensitivity to protected features as per integrated gradients is
+drastically reduced; details of this experiment are reported in the
+supplementary appendix. In the case of \emph{Cred}, average costs
+increase, likely because any potential benefits from protecting the
+\emph{age} are outweighed by the increase in costs required for greater
+plausibility.
 
@@
 
-\subsubsection{Hyperparameter settings.}\label{sec-hyperparameters}
-
-We test the impact of three types of hyperparameters. Here we focus on
-the highlights; full results are available in the supplementary
-appendix.\\
-\indent First, we note that CT is highly sensitive to the choice of a CE
-generator and its hyperparameters but (1) there are manageable patterns,
-and (2) we can usually identify settings that improve either
-plausibility or cost, and often both of them at the same time. For
-example, \emph{REVISE} tends to perform the worst, most likely because
-it uses a surrogate VAE to generate counterfactuals which impedes
-faithfulness \citep{altmeyer2024faithful}. Increasing \(T\), the maximum
-number of steps, generally yields better outcomes because more CEs can
-mature in each training epoch. The impact of \(\tau\), the required
-decision threshold is more difficult to predict. On ``harder'' datasets
-it may be difficult to satisfy high \(\tau\) for any given sample (i.e.,
-also factuals) and so increasing this threshold does not seem to
-correlate with better outcomes. In fact, \(\tau=0.5\) generally leads to
-optimal results as it is associated with high proportions of mature
-counterfactuals.\\
-\indent Second, the strength of the energy regularization,
+\subsubsection{Predictive Performance (RQ3).}\label{sec-pred}
+
+\emph{Models trained with CT are substantially more robust to
+gradient-based adversarial attacks than conventionally-trained
+baselines.}
+
+Test accuracies on adversarially perturbed data are shown in
+Figure~\ref{fig-acc}. The perturbations size, \(\varepsilon\in[0,0.1]\),
+increases along the horizontal axis and includes zero, corresponding to
+standard test accuracy for non-perturbed data. For all synthetic
+datasets, predictive performance of CT is virtually identical to the
+baseline and unaffected by perturbations. For all real-world datasets,
+we find that CT substantially improves robustness: while in some cases
+baseline accuracies drop to essentially zero for large enough
+perturbation sizes, accuracies of CT models are remarkably robust to
+gradient-based adversarial attacks.
+
+\subsubsection{Hyperparameter settings
+(RQ4).}\label{sec-hyperparameters}
+
+\emph{CT is highly sensitive to the choice of a CE generator and its
+hyperparameters but (1) we observe manageable patterns, and (2) we can
+usually identify settings that improve either plausibility or
+actionability, and typically both of them at the same time.}
+
+We evaluate the impacts of three types of hyperparameters on CT. In this
+section we focus on the highlights and make the full results available
+in the supplementary appendix.
+
+For example, \emph{REVISE} tends to perform the worst, most likely
+because it uses a surrogate VAE to generate counterfactuals which
+impedes faithfulness \citep{altmeyer2024faithful}. Increasing \(T\), the
+maximum number of steps, generally yields better outcomes because more
+CEs can mature in each training epoch. The impact of \(\tau\), the
+required decision threshold is more difficult to predict. On ``harder''
+datasets it may be difficult to satisfy high \(\tau\) for any given
+sample (i.e., also factuals) and so increasing this threshold does not
+seem to correlate with better outcomes. In fact, \(\tau=0.5\) generally
+leads to optimal results as it is associated with high proportions of
+mature counterfactuals.
+
+Second, the strength of the energy regularization,
 \(\lambda_{\text{reg}}\) is highly impactful and leads to poor
@@
 \(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
-very pronounced.\\
-\indent Third, the effectiveness and stability of CT is positively
-associated with the number of counterfactuals generated during each
-training epoch. A higher number of training epochs is also beneficial.
-Interestingly, we observed desired improvements when CT was combined
-with conventional training and applied only for the final 50\% of epochs
-of the complete training process. Put differently, CT can improve the
-explainability of models in a fine-tuning manner.
+particularly pronounced.
+
+Third, the effectiveness and stability of CT is positively associated
+with the number of counterfactuals generated during each training epoch.
+A higher number of training epochs is also beneficial. Interestingly, we
+observed desired improvements when CT was combined with conventional
+training and applied only for the final 50\% of epochs of the complete
+training process. Put differently, CT can improve the explainability of
+models in a fine-tuning manner.
 
@@
 e.g., regression \citep{spooner2021counterfactual}, but there is little
-consensus how to robustly extend the notion of CEs.\\
-Secondly, CT is susceptible to training instabilities, as also
-recognized for the related JEMs \citep{grathwohl2020your}. CT is exposed
-to two sources of instabilities: (1) the energy-based contrastive
+consensus how to robustly extend the notion of CEs.
+
+Secondly, CT is susceptible to training instabilities, as is also the
+case for the related JEMs \citep{grathwohl2020your}. CT is exposed to
+two sources of instabilities: (1) the energy-based contrastive
 divergence term, \(\text{div}(\cdot)\), in Equation~\ref{eq-div}, and
@@
 generating sufficiently many counterfactuals during each epoch, and
-including only mature counterfactuals in \(\text{div}(\cdot)\).\\
+including only mature counterfactuals in \(\text{div}(\cdot)\).
+
 Finally, we believe that it is possible to considerably improve

### 2025-07-18 paper/quarto_aaai/paper.tex
methodology adjusted and algorithm added

--- 2025-07-18 paper/quarto_aaai/paper.tex
@@
 CEs have also been used to improve models in the natural language
-processing domain. For example, \citet{wu2021polyjuice2} propose
+processing domain. For example, \citet{wu2021polyjuice} propose
 \emph{Polyjuice}, a general-purpose CE generator for language models and
@@
 
-This section introduces the Counterfactual Training framework. CT
-combines ideas from adversarial training, counterfactual explanations,
-and energy-based modelling with the explicit goal of producing models
-whose learned representations align with plausible explanations that
-further comply with user-defined actionability constraints.
-
-In the context of counterfactual explanations, plausibility has broadly
-been defined as the degree to which generated CEs comply with the
-underlying data-generating process
-\citep{altmeyer2024faithful, guidotti2022counterfactual, poyiadzi2020face}.
-Plausibility is a necessary but insufficient condition for using CEs to
-provide algorithmic recourse (AR) to individuals (negatively) affected
-by opaque models. An AR recommendations must also be actionable, i.e.,
-possible to attain by the recipient. A plausible CE for a rejected
-20-year-old loan applicant, for example, might reveal that their
-application would have been accepted, if only they had been 20 years
-older. Ignoring all other features, this would comply with the
-definition of plausibility if 40-year-old individuals were in fact more
-credit-worthy on average than young adults. But to the applicant, this
-CE does not provide actionable recourse, as age is not a (directly)
-mutable feature. Counterfactual training aims to improve model
-explainability by aligning models with counterfactuals that meet both
-desiderata: plausibility and actionability. Formally, we define
-explainability as follows:
+This section introduces the counterfactual training framework. As
+mentioned above, CT applies ideas from contrastive and robust learning
+to counterfactual explanations in order to produce models whose learned
+representations align with plausible explanations that further comply
+with user-defined actionability constraints.
+
+Counterfactual explanations are typically generated by solving
+variations of the following optimization problem,
+\begin{equation}\phantomsection\label{eq-general}{
+\begin{aligned}
+\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
+\end{aligned}
+}\end{equation} where
+\(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denotes a
+classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
+\(D\) features and \(\mathbf{y}^+\) denotes some target class. The
+\(\text{yloss}(\cdot)\) function quantifies the discrepancy between
+current model predictions for \(\mathbf{x}^{\prime}\) and the target
+class (a conventional choice is cross-entropy). Finally, we use
+\(\text{reg}(\cdot)\) to denote any form of regularization used to
+induce certain properties on the counterfactual. In their seminal paper,
+\citet{wachter2017counterfactual} propose regularizing the distance
+between counterfactuals and their original factual values, to ensure
+that individuals seeking recourse through CE face minimal costs in terms
+of feature changes. Different variations of Equation~\ref{eq-general}
+have been proposed in the literature to address various other desiderata
+including those already introduced above (faithfulness, plausibility and
+actionability). Like \citet{wachter2017counterfactual}, most of these
+approaches rely on gradient descent to optimize
+Equation~\ref{eq-general}. For more details on various approaches tested
+in this work, we refer the reader to the supplementary appendix. In the
+following, we describe in detail how counterfactuals are generated and
+used in CT.
+
+\subsection{Proposed Training
+Objective}\label{proposed-training-objective}
+
+The principal goal of CT is to improve model explainability by aligning
+models with faithful explanations that are plausible and actionable.
+Formally, we define explainability as follows:
 
@@
 
-\subsection{Proposed Objective}\label{proposed-objective}
-
 Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
 generated through gradient descent over \(T\) iterations as originally
-proposed by \citet{wachter2017counterfactual}. In broad terms, searching
-for CEs using gradient descent entails optimizing some form of an
-objective that balances (1) the classification loss
-\(\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})\), and (2) one
-or more penalty terms \(\lambda_{i}\text{cost}_{i}(\cdot)\). The exact
-specification of these penalties induces various properties in the
-counterfactual outcomes, and tends to be the key feature that
-distinguishes various gradient-based ``generators'' or ``explainers'' in
-the literature \citep{altmeyer2023explaining}, including all generators
-used in our experiments. We refer the reader to the supplementary
-appendix for details.
-
-CT adopts gradient-based CE search during training to generate
-on-the-fly model explanations \(\mathbf{x}^\prime\) for training
-samples. We use the term \emph{nascent} to denote counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\) that are not yet valid where \(t\)
-indicates the last iteration before the label is flipped. We store and
-use these interim counterfactuals as adversarial examples. Conversely,
-we consider counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature}
-explanations if they have either exhausted all \(T\) iterations or
-converged in terms reaching a pre-specified threshold, \(\tau\), for the
-predicted probability of the target class:
+proposed by \citet{wachter2017counterfactual}. CT adopts gradient-based
+CE search during training to generate on-the-fly model explanations
+\(\mathbf{x}^\prime\) for training samples. We use the term
+\emph{nascent} to denote interim counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\) that have not yet converged. As we
+explain below, these nascent counterfactuals can be stored and
+repurposed as adversarial examples. Conversely, we consider
+counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
+they have either exhausted all \(T\) iterations or converged in terms
+reaching a pre-specified threshold, \(\tau\), for the predicted
+probability of the target class:
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
@@
 components are governed through \(\lambda_{\text{div}}\),
-\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\).
+\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\). The full training
+regime is sketched out in Algorithm \ref{alg-experiment}.
+
+\begin{algorithm}[h]
+  \caption{Counterfactual Training}
+    \label{alg-experiment}
+    \begin{algorithmic}[1]
+    \REQUIRE Training dataset $\mathcal{D}$, initial model $\mathbf{M}_{\theta_0}$
+    \WHILE{not converged}
+        \STATE Sample $\mathbf{x}$ and $\mathbf{y}$ from dataset $\mathcal{D}$.
+        \STATE Sample $\mathbf{x}^{\prime}_0$, $\mathbf{y}^+$ and $\mathbf{x}^+$.
+        \FOR{$t = 1$ to $T$}
+            \STATE Backpropagate $\nabla_{\mathbf{x}^\prime}$ through Equation \ref{eq-general}. Store $\mathbf{x}_t^\prime$.
+        \ENDFOR
+        \STATE Backpropagate $\nabla_{\theta}$ through Equation \ref{eq-obj}.
+    \ENDWHILE
+    \RETURN $\mathbf{M}_\theta$
+    \end{algorithmic}
+\end{algorithm}
 

### 2025-07-18 paper/sections/experiments.qmd
methodology adjusted and algorithm added

@@
 They are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix.
-Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give2.
+Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give.
 Finally, for the convenience of illustration, we use the 10-class *MNIST* [@lecun1998mnist].

### 2025-07-18 paper/sections/method.qmd
methodology adjusted and algorithm added

@@
 
-This section introduces the Counterfactual Training framework.
-CT combines ideas from adversarial training, counterfactual explanations, and energy-based modelling with the explicit goal of producing models whose learned representations align with plausible explanations that further comply with user-defined actionability constraints.
+This section introduces the counterfactual training framework.
+As mentioned above, CT applies ideas from contrastive and robust learning to counterfactual explanations in order to produce models whose learned representations align with plausible explanations that further comply with user-defined actionability constraints.
 
-In the context of counterfactual explanations, plausibility has broadly been defined as the degree to which generated CEs comply with the underlying data-generating process [@altmeyer2024faithful;@guidotti2022counterfactual;@poyiadzi2020face].
-Plausibility is a necessary but insufficient condition for using CEs to provide algorithmic recourse (AR) to individuals (negatively) affected by opaque models.
-An AR recommendations must also be actionable, i.e., possible to attain by the recipient.
-A plausible CE for a rejected 20-year-old loan applicant, for example, might reveal that their application would have been accepted, if only they had been 20 years older.
-Ignoring all other features, this would comply with the definition of plausibility if 40-year-old individuals were in fact more credit-worthy on average than young adults.
-But of course this CE does not qualify for providing actionable recourse to the applicant since *age* is not a (directly) mutable feature.
-Counterfactual training aims to improve model explainability by aligning models with counterfactuals that meet both desiderata: plausibility and actionability.
-Formally, we define explainability as follows:
+Counterfactual explanations are typically generated by solving variations of the following optimization problem,
+$$
+\begin{aligned}
+\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
+\end{aligned}
+$$ {#eq-general}
+where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+$ denotes some target class.
+The $\text{yloss}(\cdot)$ function quantifies the discrepancy between current model predictions for $\mathbf{x}^{\prime}$ and the target class (a conventional choice is cross-entropy).
+Finally, we use $\text{reg}(\cdot)$ to denote any form of regularization used to induce certain properties on the counterfactual.
+In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values, to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes.
+Different variations of @eq-general have been proposed in the literature to address various other desiderata including those already introduced above (faithfulness, plausibility and actionability).
+Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general.
+For more details on various approaches tested in this work, we refer the reader to the supplementary appendix.
+In the following, we describe in detail how counterfactuals are generated and used in CT.
+
+## Proposed Training Objective
+
+The principal goal of CT is to improve model explainability by aligning models with faithful explanations that are plausible and actionable. Formally, we define explainability as follows:
 
@@
 
-## Proposed Objective
-
 Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual.
-In broad terms, searching for CEs using gradient descent entails optimizing some form of an objective that balances (1) the classification loss $\text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y})$, and (2) one or more penalty terms $\lambda_{i}\text{cost}_{i}(\cdot)$.
-The exact specification of these penalties induces various properties in the counterfactual outcomes, and tends to be the key feature that distinguishes various gradient-based "generators" or "explainers" in the literature [@altmeyer2023explaining], including all generators used in our experiments.
-We refer the reader to the supplementary appendix for details.
-
 CT adopts gradient-based CE search during training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for training samples.
-We use the term *nascent* to denote counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that are not yet valid where $t$ indicates the last iteration before the label is flipped.
-We store and use these interim counterfactuals as adversarial examples.
+We use the term *nascent* to denote interim counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that have not yet converged.
+As we explain below, these nascent counterfactuals can be stored and repurposed as adversarial examples.
 Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged in terms reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.
@@
 The trade-off between the components are governed through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.
+The full training regime is sketched out in Algorithm \ref{alg-experiment}.
+
+\begin{algorithm}[h]
+  \caption{Counterfactual Training}
+    \label{alg-experiment}
+    \begin{algorithmic}[1]
+    \REQUIRE Training dataset $\mathcal{D}$, initial model $\mathbf{M}_{\theta_0}$
+    \WHILE{not converged}
+        \STATE Sample $\mathbf{x}$ and $\mathbf{y}$ from dataset $\mathcal{D}$.
+        \STATE Sample $\mathbf{x}^{\prime}_0$, $\mathbf{y}^+$ and $\mathbf{x}^+$.
+        \FOR{$t = 1$ to $T$}
+            \STATE Backpropagate $\nabla_{\mathbf{x}^\prime}$ through Equation \ref{eq-general}. Store $\mathbf{x}_t^\prime$.
+        \ENDFOR
+        \STATE Backpropagate $\nabla_{\theta}$ through Equation \ref{eq-obj}.
+    \ENDWHILE
+    \RETURN $\mathbf{M}_\theta$
+    \end{algorithmic}
+\end{algorithm}
 

### 2025-07-18 paper/sections/related.qmd
methodology adjusted and algorithm added

@@
 CEs have also been used to improve models in the natural language processing domain.
-For example, @wu2021polyjuice2 propose *Polyjuice*, a general-purpose CE generator for language models and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks, while @luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
+For example, @wu2021polyjuice propose *Polyjuice*, a general-purpose CE generator for language models and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks, while @luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
 

### 2025-07-18 paper/quarto_aaai/paper.tex

--- 2025-07-18 paper/quarto_aaai/paper.tex
@@
 \sisetup{uncertainty-mode = separate}
+\DeclareMathSizes{10}{9}{7}{6.5}
 \usepackage{enumitem}

### 2025-07-18 paper/sections/experiments.qmd

@@
 
-We seek to answer the following three research questions:
+We seek to answer the following four research questions:
 
-1. To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
-2. To what extent does CT lead to more favorable AR outcomes in the presence of actionability constraints?
-3. What are the effects of hyperparameter selection on CT?
+\begin{enumerate}[label={(\makebox[2em][c]{RQ\arabic*})}, leftmargin=3.5em]
+    \item To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
+    \item To what extent does CT result in more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+    \item To what extent does CT influence the adversarial robustness of trained models?
+    \item What are the effects of hyperparameter selection on counterfactual training?
+\end{enumerate}
 
@@
 Our main results for plausibility and actionability for *MLP* models are summarised in @tbl-main, which presents counterfactual outcomes grouped by dataset along with standard errors averaged across bootstrap samples.
+Asterisks ($^*$) indicate that the observed effects are statistically significant: bootstrapped 99%-confidence intervals of differences in mean outcomes do *not* include zero.
 
 The first two columns ($\text{IP}$ and $\text{IP}^*$) show the percentage reduction in implausibility for our two metrics when using CT on top of the weak baseline.
-Focusing on the first row for *LS* for a moment, the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc (panel (b) vs.
-(a)).
-Asterisks indicate that reductions are statistically significant: bootstrapped 99%-confidence intervals of differences in mean outcomes do *not* include zero.
+As an example, consider the first row (*LS*): the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc in panel (b) compared to panel (a).
 
-The third column shows results for the scenario when mutability constraints are imposed on selected features.
-Once again we are comparing CT to the baseline, so reductions in the positive direction imply that valid counterfactuals are cheaper when using CT with feature protection.
-Relating this back to @fig-poc, the third column represents the reduction in distances travelled by counterfactuals in panel (d) compared to (c).
-In the following, we summarize the remaining results.
+The third column shows the results for a scenario when mutability constraints are imposed on the selected features.
+Again, we are comparing CT to the baseline, so reductions in the positive direction imply that valid counterfactuals are "cheaper" (more actionable) when using CT with feature protection.
+Relating this back to @fig-poc, the third column represents the reduction in distances travelled by counterfactuals in panel (d) compared to panel (c).
+In the following paragraphs, we summarize the results for all datasets.
 
 ::: {#tbl-main}
-
 ```{=latex}
@@
 ```
-
 Key evaluation metrics for valid counterfactual along with bootstrapped standard errors for all datasets.
-**Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Actionability** (column 3): percentage reduction in costs with protected features.
+**Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Cost** / **Actionability** (column 3): percentage reduction in costs when selected features are protected.
 Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time.
-Asterisks ($^*$) indicate that bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
-
+Asterisks ($^*$) indicate that the bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
 :::
 
-### Plausibility. {#sec-plaus}
-
-We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$.
-Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*).
-The only negative findings for *OL* and *MNIST* are statistically insignificant and, for MNIST, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).
+### Plausibility (RQ1). {#sec-plaus}
+*CT generally produces substantial and statistically significant improvements in plausibility.*  
+Average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*.
+For the real-world tabular datasets they are around 12% for *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $\text{IP}$.
+Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: reductions range from around 20% (*Moons*) to almost 90% (*Circ*).
+The only negative findings for *OL* and *MNIST* are not statistically significant and, for MNIST, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).
 
@@
 
-### Actionability. {#sec-act}
+### Actionability (RQ2). {#sec-act}
+*CT tends to improve actionability in the presence of immutable features, but the exact effects can be challenging to establish due to feature interactions.*
 
-We also find that CT can improve actionability in the presence of immutable features (column 3 in @tbl-main): for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image.
-Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*.
-While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist (columns 6-10) demonstrates that in the latter case, CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix.
+For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image.
+Reductions in costs are overwhelmingly positive and significant of up to around 60% for *GMSC*.
+While the estimated cost increase for *Adult* and *MNIST*, these effects are not significant and @fig-mnist (columns 6-10) demonstrates that in the latter case, CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix.
 In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
-
-### Predictive Performance. {#sec-pred}
-
-Test accuracies on adversarially perturbed data are shown in @fig-acc.
-The perturbations size, $\varepsilon \in [0,0.1]$, increases along the horizontal axis and includes zero, corresponding to standard test accuracy for non-perturbed data.
-We observe that for all synthetic datasets, predictive performance of CT is virtually identical to the baseline and unaffected by adversarial perturbations.
-For all real-world datasets, we find that CT substantially improves robustness: while in some cases baseline accuracies drop to essentially zero for large enough perturbation sizes, accuracies for CT are remarkably robust to gradient-based adversarial attacks.
 
@@
 
-### Hyperparameter settings. {#sec-hyperparameters}
-We test the impact of three types of hyperparameters. Here we focus on the highlights; full results are available in the supplementary appendix.  
-`\indent`{=latex} First, we note that CT is highly sensitive to the choice of a CE generator and its hyperparameters but (1) there are manageable patterns, and (2) we can usually identify settings that improve either plausibility or cost, and often both of them at the same time.
+### Predictive Performance (RQ3). {#sec-pred}
+*Models trained with CT are substantially more robust to gradient-based adversarial attacks than conventionally-trained baselines.*
+
+Test accuracies on adversarially perturbed data are shown in @fig-acc.
+The perturbations size, $\varepsilon\in[0,0.1]$, increases along the horizontal axis and includes zero, corresponding to standard test accuracy for non-perturbed data.
+For all synthetic datasets, predictive performance of CT is virtually identical to the baseline and unaffected by perturbations.
+For all real-world datasets, we find that CT substantially improves robustness: while in some cases baseline accuracies drop to essentially zero for large enough perturbation sizes, accuracies of CT models are remarkably robust to gradient-based adversarial attacks.
+
+### Hyperparameter settings (RQ4). {#sec-hyperparameters}
+*CT is highly sensitive to the choice of a CE generator and its hyperparameters but (1) we observe manageable patterns, and (2) we can usually identify settings that improve either plausibility or actionability, and typically both of them at the same time.*
+
+We evaluate the impacts of three types of hyperparameters on CT. In this section we focus on the highlights and make the full results available in the supplementary appendix.
+
 For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful].
@@
 In fact, $\tau=0.5$ generally leads to optimal results as it is associated with high proportions of mature counterfactuals.
-`\indent`{=latex} Second, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high.
+
+Second, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high.
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
-While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.
-`\indent`{=latex} Third, the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
+While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not particularly pronounced.
+
+Third, the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
 A higher number of training epochs is also beneficial.

### 2025-07-18 paper/quarto_aaai/paper.tex
Cut down to 7 pages

@@
 we refer to ``explanatory capacity'' as ``explainability'' in the rest
-of this manuscript (see Def. \ref{def-explainability} below for a more
-formal exposition).
+of this manuscript (see Def. \ref{def-explainability}).
 
@@
 overrelying on a single locally optimal representation that corresponds
-to semantically meaningless explanations for the data. Likewise,
-previous work of \citet{schut2021generating} found that generating
-plausible (``interpretable'') CEs is almost trivial for deep ensembles
-that have undergone adversarial training. The case for JEMs is even
-clearer: they optimize a hybrid objective that induces both high
-predictive performance and strong generative capacity
-\citep{grathwohl2020your}, which bears resemblance to the idea of
-aligning models with plausible explanations and has inspired our CT
-objective.
+to semantically meaningless explanations. Likewise, previous work of
+\citet{schut2021generating} found that generating plausible
+(``interpretable'') CEs is almost trivial for deep ensembles that have
+undergone adversarial training. The case for JEMs is even clearer: they
+optimize a hybrid objective that induces both high predictive
+performance and strong generative capacity \citep{grathwohl2020your},
+which resembles the idea of aligning models with plausible explanations
+and has inspired CT.
 
@@
 have been proposed in the literature to address various other desiderata
-including those already introduced above (faithfulness, plausibility and
+including those introduced above (faithfulness, plausibility and
 actionability). Like \citet{wachter2017counterfactual}, most of these
@@
 
-The principal goal of CT is to improve model explainability by aligning
-models with faithful explanations that are plausible and actionable.
-Formally, we define explainability as follows:
+The goal of CT is to improve model explainability by aligning models
+with faithful explanations that are plausible and actionable. Formally,
+we define explainability as follows:
 
@@
 
-\noindent The characterization of faithfulness and plausibility in Def.
+The characterization of faithfulness and plausibility in Def.
 \ref{def-explainability} is the same as in \citet{altmeyer2024faithful},
@@
 baseline accuracies drop to essentially zero for large enough
-perturbation sizes, accuracies of CT models are remarkably robust to
-gradient-based adversarial attacks.
+perturbation sizes, accuracies of CT models remain remarkably robust.
 
@@
 
-For example, \emph{REVISE} tends to perform the worst, most likely
-because it uses a surrogate VAE to generate counterfactuals which
-impedes faithfulness \citep{altmeyer2024faithful}. Increasing \(T\), the
-maximum number of steps, generally yields better outcomes because more
-CEs can mature in each training epoch. The impact of \(\tau\), the
-required decision threshold is more difficult to predict. On ``harder''
-datasets it may be difficult to satisfy high \(\tau\) for any given
-sample (i.e., also factuals) and so increasing this threshold does not
-seem to correlate with better outcomes. In fact, \(\tau=0.5\) generally
-leads to optimal results as it is associated with high proportions of
-mature counterfactuals.
+First, we look at hyperparameters that guide the gradient-based
+counterfactual search. Increasing \(T\), the maximum number of steps,
+generally yields better outcomes because more CEs can mature in each
+training epoch. The impact of \(\tau\), the required decision threshold
+is more difficult to predict. On ``harder'' datasets it may be difficult
+to satisfy high \(\tau\) for any given sample (i.e., also factuals) and
+so increasing this threshold does not seem to correlate with better
+outcomes. In fact, \(\tau=0.5\) generally leads to optimal results as it
+is associated with high proportions of mature counterfactuals.
 

### 2025-07-18 paper/sections/experiments.qmd
Cut down to 7 pages

@@
 For all synthetic datasets, predictive performance of CT is virtually identical to the baseline and unaffected by perturbations.
-For all real-world datasets, we find that CT substantially improves robustness: while in some cases baseline accuracies drop to essentially zero for large enough perturbation sizes, accuracies of CT models are remarkably robust to gradient-based adversarial attacks.
+For all real-world datasets, we find that CT substantially improves robustness: while in some cases baseline accuracies drop to essentially zero for large enough perturbation sizes, accuracies of CT models remain remarkably robust.
 
@@
 
-For example, *REVISE* tends to perform the worst, most likely because it uses a surrogate VAE to generate counterfactuals which impedes faithfulness [@altmeyer2024faithful].
+First, we look at hyperparameters that guide the gradient-based counterfactual search.
 Increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch.

### 2025-07-18 paper/sections/method.qmd
Cut down to 7 pages

@@
 In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values, to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes.
-Different variations of @eq-general have been proposed in the literature to address various other desiderata including those already introduced above (faithfulness, plausibility and actionability).
+Different variations of @eq-general have been proposed in the literature to address various other desiderata including those introduced above (faithfulness, plausibility and actionability).
 Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general.
@@
 
-The principal goal of CT is to improve model explainability by aligning models with faithful explanations that are plausible and actionable. Formally, we define explainability as follows:
+The goal of CT is to improve model explainability by aligning models with faithful explanations that are plausible and actionable. Formally, we define explainability as follows:
 
@@
 
-`\noindent`{=latex} The characterization of faithfulness and plausibility in Def.
+The characterization of faithfulness and plausibility in Def.
 \ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation.

### 2025-07-18 paper/sections/related.qmd
Cut down to 7 pages

@@
 For simplicity, we refer to "explanatory capacity" as "explainability" in the rest of this manuscript (see Def.
-\ref{def-explainability} below for a more formal exposition).
+\ref{def-explainability}).
 
@@
 Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case].
-Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations for the data.
+Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations.
 Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training.
-The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which bears resemblance to the idea of aligning models with plausible explanations and has inspired our CT objective.
+The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which resembles the idea of aligning models with plausible explanations and has inspired CT.
 

### 2025-07-18 paper/quarto_aaai/paper.tex
Editorial pass up until S4.2

@@
 to produce some desired output. To be useful in real-word
-decision-making systems, counterfactuals should be (1) plausible with
-respect to the underlying data and (2) actionable with respect to the
+decision-making systems, counterfactuals should be plausible with
+respect to the underlying data and actionable with respect to the
 user-defined mutability constraints. Much existing research has
@@
 hold models directly accountable for the desired end goal:
-counterfactual training employs counterfactuals ad-hoc during the
-training phase to minimize the divergence between learned
-representations and plausible, actionable explanations. We demonstrate
-empirically and theoretically that our proposed method facilitates
-training models that deliver inherently desirable explanations while
-promoting robustness and preserving high predictive performance.
+counterfactual training employs counterfactuals during the training
+phase to minimize the divergence between learned representations and
+plausible, actionable explanations. We demonstrate empirically and
+theoretically that our proposed method facilitates training models that
+deliver inherently desirable explanations while promoting robustness and
+preserving high predictive performance of the models.
 \end{abstract}
@@
 Today's prominence of artificial intelligence (AI) has largely been
-driven by the success of representation learning with large degrees of
+driven by the success of representation learning with high degrees of
 freedom: instead of relying on features and rules hand-crafted by
@@
 models tend to be so complex that humans cannot easily interpret their
-decision-making.
+decision logic.
 
@@
 counterfactuals, which comply with specific desiderata. This is
-problematic, because a narrow focus on any specific desideratum can
+problematic, because the narrow focus on any specific desideratum can
 adversely affect others: it is possible, for example, to generate
@@
 implausible, possibly adversarial counterfactuals
-\citep{altmeyer2024faithful}. In this work, we therefore follow the
+\citep{altmeyer2024faithful}. In this work, we therefore embrace the
 paradigm that models---as opposed to explanation methods---should be
@@
 more robust to adversarial examples (AE). Figure~\ref{fig-poc}
-illustrates the outcome of CT compared to conventionally trained models:
-in panel (a), faithful and valid counterfactuals end up near the
-decision boundary forming a clearly distinguishable cluster in the
-target class (orange); in panel (b), CT is applied to the same
+illustrates the outcomes of CT compared to a conventionally trained
+model. First, in panel (a), faithful and valid counterfactuals end up
+near the decision boundary forming a clearly distinguishable cluster in
+the target class (orange). In panel (b), CT is applied to the same
 underlying linear classifier architecture resulting in much more
-plausible counterfactuals; in panel (c), the classifier is again trained
+plausible counterfactuals. In panel (c), the classifier is again trained
 conventionally and we have introduced a mutability constraint on the
 \emph{age} feature at test time---counterfactuals are valid but the
-classifier is roughly equally sensitive to both features; by contrast,
-the decision boundary in panel (c) has titled, making the model trained
+classifier is roughly equally sensitive to both features. By contrast,
+the decision boundary in panel (d) has titled, making the model trained
 with CT relatively less sensitive to the immutable \emph{age} feature.
 To achieve these outcomes, CT draws inspiration from the literature on
-contrastive and robust learning: we contrast faithful (\emph{mature})
-CEs with ground-truth data while protecting immutable features, and
-capitalize on methodological links between CE and AE by penalizing the
-model's adversarial loss on interim (\emph{nascent}) counterfactuals. To
-the best of our knowledge, CT represents the first venture in this
-direction with promising empirical and theoretical results.
+contrastive and robust learning: we contrast faithful CEs with
+ground-truth data while protecting immutable features, and capitalize on
+methodological links between CE and AE by penalizing the model's
+adversarial loss on interim (\emph{nascent}) counterfactuals. To the
+best of our knowledge, CT represents the first venture in this direction
+with promising empirical and theoretical results.
 
@@
 cost of reaching valid counterfactuals with protected features decreases
-by 19\% on average; and (3) adversarial robustness improves across the
-board. Finally, we discuss open challenges in
+by 19\% on average; and (3) the adversarial robustness improves across
+the board. Finally, we discuss open challenges in
 Section~\ref{sec-discussion} and conclude in
@@
 \citet{wachter2017counterfactual} propose regularizing the distance
-between counterfactuals and their original factual values, to ensure
-that individuals seeking recourse through CE face minimal costs in terms
-of feature changes. Different variations of Equation~\ref{eq-general}
-have been proposed in the literature to address various other desiderata
-including those introduced above (faithfulness, plausibility and
-actionability). Like \citet{wachter2017counterfactual}, most of these
-approaches rely on gradient descent to optimize
-Equation~\ref{eq-general}. For more details on various approaches tested
-in this work, we refer the reader to the supplementary appendix. In the
-following, we describe in detail how counterfactuals are generated and
-used in CT.
+between counterfactuals and their original factual values to ensure that
+individuals seeking recourse through CE face minimal costs in terms of
+feature changes. Different variations of Equation~\ref{eq-general} have
+been proposed in the literature to address many desiderata including the
+ones discussed above (faithfulness, plausibility and actionability).
+Like \citet{wachter2017counterfactual}, most of these approaches rely on
+gradient descent to optimize Equation~\ref{eq-general}. For more details
+on the approaches tested in this work, we refer the reader to the
+supplementary appendix. In the following, we describe in detail how
+counterfactuals are generated and used in counterfactual training.
 
@@
 
-\begin{enumerate}
-    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$, where $A$ is some arbitrarily small region around $\mathbf{x}^{\prime}$.
+\begin{itemize}
+    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ is an arbitrarily small region around $\mathbf{x}^{\prime}$.
     \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
-    \item (Actionability) Perturbations $\Delta$ are subject to some actionability constraints.
-\end{enumerate}
+    \item (Actionability) Perturbations $\Delta$ may be subject to some actionability constraints.
+\end{itemize}
 and $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs.
@@
 proposed by \citet{wachter2017counterfactual}. CT adopts gradient-based
-CE search during training to generate on-the-fly model explanations
-\(\mathbf{x}^\prime\) for training samples. We use the term
+CE search in training to generate on-the-fly model explanations
+\(\mathbf{x}^\prime\) for the training samples. We use the term
 \emph{nascent} to denote interim counterfactuals
@@
 counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
-they have either exhausted all \(T\) iterations or converged in terms
-reaching a pre-specified threshold, \(\tau\), for the predicted
-probability of the target class:
+they have either exhausted all \(T\) iterations or converged by reaching
+a pre-specified threshold, \(\tau\), for the predicted probability of
+the target class:
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
@@
 be summarized as inducing explainability directly and indirectly by
-penalizing the contrastive divergence, \(\text{div}(\cdot)\), between
-mature counterfactuals \(\mathbf{x}_T^\prime\) and observed samples
-\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
-class \(y^+\), and the adversarial loss, \(\text{advloss}(.)\), wrt.
-nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
-\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
-regularizes the magnitude of the energy terms involved in
-\(\text{div}(\cdot)\) \citep{du2019implicit}. The trade-off between the
-components are governed through \(\lambda_{\text{div}}\),
-\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\). The full training
-regime is sketched out in Algorithm \ref{alg-experiment}.
+penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
+between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
+samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
+target class \(y^+\), and (2) the adversarial loss,
+\(\text{advloss}(.)\), wrt. nascent counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
+denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
+of the energy terms involved in \(\text{div}(\cdot)\)
+\citep{du2019implicit}. The trade-offs between these components are
+adjusted through \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
+\(\lambda_{\text{reg}}\). The full training regime is sketched out in
+Algorithm \ref{alg-experiment}.
 
@@
 \text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-}\end{equation} and introduce a novel divergence-based adaption,
+}\end{equation} and introduce a novel divergence-based adaptation,
 \begin{equation}\phantomsection\label{eq-impl-div}{
@@
 models to seek plausibility with respect to mutable features, much like
-we observe in Figure~\ref{fig-poc} (panel (d) vs panel (c)). We
+we observe in Figure~\ref{fig-poc} in panel (d) compared to (c). We
 supplement this analysis with qualitative findings for integrated

### 2025-07-18 paper/sections/abstract.qmd
Editorial pass up until S4.2

--- 2025-03-13 paper/sections/abstract.qmd
@@
 Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output.
-To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the stakeholder requirements.
+To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the user-defined mutability constraints.
 Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata.
-In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
-We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.
+In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
+We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while promoting robustness and preserving high predictive performance of the models.

### 2025-07-18 paper/sections/experiments.qmd
Editorial pass up until S4.2

@@
 $$ {#eq-impl-dist}
-and introduce a novel divergence-based adaption,
+and introduce a novel divergence-based adaptation,
 $$
@@
 To assess outcomes with respect to actionability for non-linear models, we look at the average costs of valid counterfactuals in terms of their distances from factual starting points.
-While this an imperfect proxy of sensitivity, we hypothesize that CT can reduce these costs by teaching models to seek plausibility with respect to mutable features, much like we observe in @fig-poc (panel (d) vs panel (c)).
+While this an imperfect proxy of sensitivity, we hypothesize that CT can reduce these costs by teaching models to seek plausibility with respect to mutable features, much like we observe in @fig-poc in panel (d) compared to (c).
 We supplement this analysis with qualitative findings for integrated gradients [@sundararajan2017ig].

### 2025-07-18 paper/sections/intro.qmd
Editorial pass up until S4.2

@@
 
-Today’s prominence of artificial intelligence (AI) has largely been driven by the success of representation learning with large degrees of freedom: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning highly complex representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
-These models tend to be so complex that humans cannot easily interpret their decision-making.
+Today’s prominence of artificial intelligence (AI) has largely been driven by the success of representation learning with high degrees of freedom: instead of relying on features and rules hand-crafted by humans, modern machine learning (ML) models are tasked with learning highly complex representations directly from the data, guided by narrow objectives such as predictive accuracy [@goodfellow2016deep].
+These models tend to be so complex that humans cannot easily interpret their decision logic.
 
@@
 Existing state-of-the-art (SOTA) approaches in the field have largely focused on designing model-agnostic CE methods that identify subsets of counterfactuals, which comply with specific desiderata.
-This is problematic, because a narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful].
-In this work, we therefore follow the paradigm that models---as opposed to explanation methods---should be held accountable for explanations that are plausible and actionable.
+This is problematic, because the narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful].
+In this work, we therefore embrace the paradigm that models---as opposed to explanation methods---should be held accountable for explanations that are plausible and actionable.
 While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models to improve their overall explanatory capacity.
@@
 Specifically, we propose **counterfactual training (CT)**: a novel training regime that leverages counterfactual explanations on-the-fly to ensure that differentiable models learn plausible and actionable explanations for the underlying data, while at the same time also being more robust to adversarial examples (AE).
-@fig-poc illustrates the outcome of CT compared to conventionally trained models: in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange); in panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals; in panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features; by contrast, the decision boundary in panel (c) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature.
-To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful (*mature*) CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals.
+@fig-poc illustrates the outcomes of CT compared to a conventionally trained model.
+First, in panel (a), faithful and valid counterfactuals end up near the decision boundary forming a clearly distinguishable cluster in the target class (orange).
+In panel (b), CT is applied to the same underlying linear classifier architecture resulting in much more plausible counterfactuals.
+In panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features.
+By contrast, the decision boundary in panel (d) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature.
+To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals.
 To the best of our knowledge, CT represents the first venture in this direction with promising empirical and theoretical results.
@@
 In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints.
-In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) adversarial robustness improves across the board.
+In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) the adversarial robustness improves across the board.
 Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.

### 2025-07-18 paper/sections/method.qmd
Editorial pass up until S4.2

@@
 Finally, we use $\text{reg}(\cdot)$ to denote any form of regularization used to induce certain properties on the counterfactual.
-In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values, to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes.
-Different variations of @eq-general have been proposed in the literature to address various other desiderata including those introduced above (faithfulness, plausibility and actionability).
+In their seminal paper, @wachter2017counterfactual propose regularizing the distance between counterfactuals and their original factual values to ensure that individuals seeking recourse through CE face minimal costs in terms of feature changes.
+Different variations of @eq-general have been proposed in the literature to address many desiderata including the ones discussed above (faithfulness, plausibility and actionability).
 Like @wachter2017counterfactual, most of these approaches rely on gradient descent to optimize @eq-general.
-For more details on various approaches tested in this work, we refer the reader to the supplementary appendix.
-In the following, we describe in detail how counterfactuals are generated and used in CT.
+For more details on the approaches tested in this work, we refer the reader to the supplementary appendix.
+In the following, we describe in detail how counterfactuals are generated and used in counterfactual training.
 
@@
 Let $\mathbf{x}_t^\prime$ for $t=0,...,T$ denote a counterfactual generated through gradient descent over $T$ iterations as originally proposed by @wachter2017counterfactual.
-CT adopts gradient-based CE search during training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for training samples.
+CT adopts gradient-based CE search in training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for the training samples.
 We use the term *nascent* to denote interim counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that have not yet converged.
 As we explain below, these nascent counterfactuals can be stored and repurposed as adversarial examples.
-Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged in terms reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.
+Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged by reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.
 
@@
 The second and third terms are explained in detail below.
-For now, they can be summarized as inducing explainability directly and indirectly by penalizing the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and the adversarial loss, $\text{advloss}(.)$, wrt.
+For now, they can be summarized as inducing explainability directly and indirectly by penalizing (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (2) the adversarial loss, $\text{advloss}(.)$, wrt.
 nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
 Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit].
-The trade-off between the components are governed through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.
+The trade-offs between these components are adjusted through $\lambda_{\text{div}}$, $\lambda_{\text{adv}}$ and $\lambda_{\text{reg}}$.
 The full training regime is sketched out in Algorithm \ref{alg-experiment}.

### 2025-07-18 paper/quarto_aaai/paper.tex
Finish editorial pass

@@
 
-This section introduces the counterfactual training framework. As
-mentioned above, CT applies ideas from contrastive and robust learning
-to counterfactual explanations in order to produce models whose learned
-representations align with plausible explanations that further comply
-with user-defined actionability constraints.
+This section introduces the counterfactual training framework, applying
+ideas from contrastive and robust learning to counterfactual
+explanations. CT produces models whose learned representations align
+with plausible explanations that comply with user-defined actionability
+constraints.
 
@@
 Our main results for plausibility and actionability for \emph{MLP}
-models are summarised in Table~\ref{tbl-main}, which presents
+models are summarised in Table~\ref{tbl-main} that presents
 counterfactual outcomes grouped by dataset along with standard errors
-averaged across bootstrap samples. Asterisks (\(^*\)) indicate that the
-observed effects are statistically significant: bootstrapped
-99\%-confidence intervals of differences in mean outcomes do \emph{not}
-include zero.
+averaged across bootstrap samples. Asterisks (\(^*\)) are used when the
+bootstrapped 99\%-confidence interval of differences in mean outcomes
+does \emph{not} include zero, so the observed effects are statistically
+significant at the 0.01 level.
 
@@
 percentage reduction in implausibility for our two metrics when using CT
-on top of the weak baseline. As an example, consider the first row
-(\emph{LS}): the observed positive values indicate that faithful
+on top of the weak baseline. As an example, consider the first row for
+\emph{LS} data: the observed positive values indicate that faithful
 counterfactuals are around 30-55\% more plausible for models trained
@@
 \emph{CT tends to improve actionability in the presence of immutable
-features, but the exact effects can be challenging to establish due to
-feature interactions.}
+features, but the exact effect is sometimes hard to establish due to,
+e.g., feature interactions.}
 
@@
 rows of the full image. Reductions in costs are overwhelmingly positive
-and significant of up to around 60\% for \emph{GMSC}. While the
+and significant, of up to around 60\% for \emph{GMSC}. While the
 estimated cost increase for \emph{Adult} and \emph{MNIST}, these effects
@@
 challenge, we introduce counterfactual training, a novel training regime
-that incentivizes highly-explainable models. Through extensive
-experiments, we demonstrate that CT satisfies this objective while
-promoting robustness and preserving the predictive performance of
-models. Explanations generated from CT-based models are both more
-plausible (compliant with the underlying data-generating process) and
-more actionable (compliant with user-specified mutability constraints),
-and thus meaningful to their recipients. We also show that CT can be
-used to fine-tune conventionally-trained models and achieve similar
-gains. Lastly, our work highlights the value of simultaneously improving
-models and their explanations.
+that integrates recent advances in contrastive learning, adversarial
+robustness, and counterfactual explanations to incentivize
+highly-explainable models. Through extensive experiments, we demonstrate
+that CT satisfies this goal while preserving the predictive performance
+and promoting robustness of models. Explanations generated from CT-based
+models are both more plausible (compliant with the underlying
+data-generating process) and more actionable (compliant with
+user-specified mutability constraints), and thus meaningful to their
+recipients. In turn, our work highlights the value of simultaneously
+improving models and their explanations.
 

### 2025-07-18 paper/sections/experiments.qmd
Finish editorial pass

@@
 
-Our main results for plausibility and actionability for *MLP* models are summarised in @tbl-main, which presents counterfactual outcomes grouped by dataset along with standard errors averaged across bootstrap samples.
-Asterisks ($^*$) indicate that the observed effects are statistically significant: bootstrapped 99%-confidence intervals of differences in mean outcomes do *not* include zero.
+Our main results for plausibility and actionability for *MLP* models are summarised in @tbl-main that presents counterfactual outcomes grouped by dataset along with standard errors averaged across bootstrap samples.
+Asterisks ($^*$) are used when the bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero, so the observed effects are statistically significant at the 0.01 level.
 
 The first two columns ($\text{IP}$ and $\text{IP}^*$) show the percentage reduction in implausibility for our two metrics when using CT on top of the weak baseline.
-As an example, consider the first row (*LS*): the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc in panel (b) compared to panel (a).
+As an example, consider the first row for *LS* data: the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc in panel (b) compared to panel (a).
 
@@
 ### Actionability (RQ2). {#sec-act}
-*CT tends to improve actionability in the presence of immutable features, but the exact effects can be challenging to establish due to feature interactions.*
+*CT tends to improve actionability in the presence of immutable features, but the exact effect is sometimes hard to establish due to, e.g., feature interactions.*
 
 For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image.
-Reductions in costs are overwhelmingly positive and significant of up to around 60% for *GMSC*.
+Reductions in costs are overwhelmingly positive and significant, of up to around 60% for *GMSC*.
 While the estimated cost increase for *Adult* and *MNIST*, these effects are not significant and @fig-mnist (columns 6-10) demonstrates that in the latter case, CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix.

### 2025-07-18 paper/sections/method.qmd
Finish editorial pass

@@
 
-This section introduces the counterfactual training framework.
-As mentioned above, CT applies ideas from contrastive and robust learning to counterfactual explanations in order to produce models whose learned representations align with plausible explanations that further comply with user-defined actionability constraints.
+This section introduces the counterfactual training framework, applying ideas from contrastive and robust learning to counterfactual explanations.
+CT produces models whose learned representations align with plausible explanations that comply with user-defined actionability constraints.
 

### 2025-07-19 paper/sections/other/grids.qmd
adjusted Algorithm 1 and ensured consistet capitalization for CT, AR, CE etc.

@@
 
-Overall, we find that counterfactual training (CT) achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to ~60 percent (%) for the *Circles* data (e.g.
-@fig-grid-gen_params-plaus-circles), ~25-30% for the *Moons* data (e.g.
-@fig-grid-gen_params-plaus-moons) and ~10-20% for the *Linearly Separable* data (e.g.
+Overall, we find that counterfactual training achieves it key objectives consistently across all hyperparameter settings and also broadly across datasets: plausibility is improved by up to 60 percent (%) for the *Circles* data (e.g.
+@fig-grid-gen_params-plaus-circles), 25-30% for the *Moons* data (e.g.
+@fig-grid-gen_params-plaus-moons) and 10-20% for the *Linearly Separable* data (e.g.
 @fig-grid-gen_params-plaus-lin_sep).
-At the same time, the average costs of faithful counterfactuals are reduced in many cases by around ~20-25% for  *Circles* (e.g.
-@fig-grid-gen_params-cost-circles) and up to ~50% for *Moons* (e.g.
+At the same time, the average costs of faithful counterfactuals are reduced in many cases by around 20-25% for  *Circles* (e.g.
+@fig-grid-gen_params-cost-circles) and up to 50% for *Moons* (e.g.
 @fig-grid-gen_params-cost-moons).

### 2025-07-19 paper/quarto_aaai/paper.tex
adjusted Algorithm 1 and ensured consistet capitalization for CT, AR, CE etc.
trying to address comments on figure 2

@@
 cost of reaching valid counterfactuals with protected features decreases
-by 19\% on average; and (3) the adversarial robustness improves across
-the board. Finally, we discuss open challenges in
+by 19\% on average; and (3) models' adversarial robustness improves
+across the board. Finally, we discuss open challenges in
 Section~\ref{sec-discussion} and conclude in
@@
     \begin{algorithmic}[1]
-    \REQUIRE Training dataset $\mathcal{D}$, initial model $\mathbf{M}_{\theta_0}$
+    \REQUIRE Training dataset $\mathcal{D}$, initialize model $\mathbf{M}_{\theta}$
     \WHILE{not converged}
@@
 viable, so long as it does not violate the faithfulness condition. Like
-JEMs \citep{murphy2022probabilistic}, Counterfactual Training can be
+JEMs \citep{murphy2022probabilistic}, counterfactual training can be
 considered a form of contrastive representation learning.
@@
 counterfactuals using a surrogate Variational Autoencoder (VAE), and
-\emph{ECCo}---the generator of \citet{altmeyer2024faithful} without the
-conformal prediction component---as a method that directly targets both
-faithfulness and plausibility of the counterfactuals.
+\emph{ECCCo} \citep{altmeyer2024faithful}, which targets faithfulness.
 
@@
 across bootstrap samples and varying degrees of the energy penalty
-\(\lambda_{\text{egy}}\) used for \emph{ECCo} at test time. Asterisks
+\(\lambda_{\text{egy}}\) used for \emph{ECCCo} at test time. Asterisks
 (\(^*\)) indicate that the bootstrapped 99\%-confidence interval of
@@
 \(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
-almost 90\% (\emph{Circ}). The only negative findings for \emph{OL} and
-\emph{MNIST} are not statistically significant and, for MNIST, do not
-align with qualitative findings, which are much more plausible for CT
-(Figure~\ref{fig-mnist}).
+almost 90\% (\emph{Circ}). The only negative findings are for OL and
+MNIST, but they are not statistically significant. A qualitative
+inspection of the counterfactuals in Figure~\ref{fig-mnist} (columns
+2-5) suggests recognizable digits 1-4 for the model trained with CT
+(bottom row), unlike the baseline (top row).
 
@@
 
-\caption{\label{fig-mnist}Sample explanations for \emph{MNIST} for BL
-(top) and CT (bottom). First column is a random factual 0 (blue).
-Columns 2 to 5 are corresponding \emph{ECCo} counterfactuals in target
-classes 1 to 4. Columns 6 to 10 show integrated gradients averaged over
-all test images in classes 5 to 9.}
+\caption{\label{fig-mnist}Visual explanations for \emph{MNIST} for BL
+(top) and CT (bottom). \textbf{Plausibility}: col.~1 is a random factual
+0 (blue); cols. 2-5 are corresponding \emph{ECCCo} counterfactuals in
+target classes 1 to 4. \textbf{Actionability}: cols. 6-10 show
+integrated gradients averaged over test images in classes 5 to 9.}
 
@@
 variable; for \emph{MNIST}, we protect the five upper and lower pixel
-rows of the full image. Reductions in costs are overwhelmingly positive
-and significant, of up to around 60\% for \emph{GMSC}. While the
-estimated cost increase for \emph{Adult} and \emph{MNIST}, these effects
-are not significant and Figure~\ref{fig-mnist} (columns 6-10)
-demonstrates that in the latter case, CT does have the expected effect:
-sensitivity to protected features as per integrated gradients is
-drastically reduced; details of this experiment are reported in the
-supplementary appendix. In the case of \emph{Cred}, average costs
-increase, likely because any potential benefits from protecting the
-\emph{age} are outweighed by the increase in costs required for greater
-plausibility.
+rows of the full image. Statistically significant reductions in costs
+overwhelmingly point in the expected positive direction reaching up to
+around 60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average
+costs increase, likely because any potential benefits from protecting
+the \emph{age} are outweighed by the increase in costs required for
+greater plausibility. The findings for \emph{Adult} and \emph{MNIST} are
+not significant. A qualitative inspection of the class-conditional
+integrated gradients in Figure~\ref{fig-mnist} (columns 6-10) suggests
+that CT still has the expected effect: the model (bottom) is insensitive
+(blue) to the protected rows of pixels; details of this experiment are
+reported in the supplementary appendix.
 

### 2025-07-19 paper/sections/experiments.qmd
trying to address comments on figure 2

@@
 
-We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCo*---the generator of @altmeyer2024faithful without the conformal prediction component---as a method that directly targets both faithfulness and plausibility of the counterfactuals.
+We run experiments with three gradient-based generators: *Generic* of @wachter2017counterfactual as a simple baseline approach, *REVISE* [@joshi2019realistic] that aims to generate plausible counterfactuals using a surrogate Variational Autoencoder (VAE), and *ECCCo* [@altmeyer2024faithful], which targets faithfulness.
 
@@
 **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Cost** / **Actionability** (column 3): percentage reduction in costs when selected features are protected.
-Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCo* at test time.
+Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCCo* at test time.
 Asterisks ($^*$) indicate that the bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.
@@
 Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: reductions range from around 20% (*Moons*) to almost 90% (*Circ*).
-The only negative findings for *OL* and *MNIST* are not statistically significant and, for MNIST, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).
+The only negative findings are for OL and MNIST, but they are not statistically significant.
+A qualitative inspection of the counterfactuals in @fig-mnist (columns 2-5) suggests recognizable digits 1-4 for the model trained with CT (bottom row), unlike the baseline (top row).
 
-![Sample explanations for *MNIST* for BL (top) and CT (bottom).
-First column is a random factual 0 (blue).
-Columns 2 to 5 are corresponding *ECCo* counterfactuals in target classes 1 to 4.
-Columns 6 to 10 show integrated gradients averaged over all test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}
+![Visual explanations for *MNIST* for BL (top) and CT (bottom).
+**Plausibility**: col.
+1 is a random factual 0 (blue); cols.
+2-5 are corresponding *ECCCo* counterfactuals in target classes 1 to 4.
+**Actionability**: cols.
+6-10 show integrated gradients averaged over test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}
 
@@
 For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image.
-Reductions in costs are overwhelmingly positive and significant, of up to around 60% for *GMSC*.
-While the estimated cost increase for *Adult* and *MNIST*, these effects are not significant and @fig-mnist (columns 6-10) demonstrates that in the latter case, CT does have the expected effect: sensitivity to protected features as per integrated gradients is drastically reduced; details of this experiment are reported in the supplementary appendix.
-In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
+Statistically significant reductions in costs overwhelmingly point in the expected positive direction reaching up to around 60% for *GMSC*.
+Only in the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
+The findings for *Adult* and *MNIST* are not significant.
+A qualitative inspection of the class-conditional integrated gradients in @fig-mnist (columns 6-10) suggests that CT still has the expected effect: the model (bottom) is insensitive (blue) to the protected rows of pixels; details of this experiment are reported in the supplementary appendix.
 

### 2025-07-19 paper/sections/intro.qmd
trying to address comments on figure 2

@@
 In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints.
-In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) the adversarial robustness improves across the board.
+In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) models' adversarial robustness improves across the board.
 Finally, we discuss open challenges in @sec-discussion and conclude in @sec-conclusion.

### 2025-07-19 paper/quarto_aaai/paper.tex
editorial pass until section 2

@@
 respect to the underlying data and actionable with respect to the
-user-defined mutability constraints. Much existing research has
-therefore focused on developing post-hoc methods to generate
-counterfactuals that meet these desiderata. In this work, we instead
-hold models directly accountable for the desired end goal:
-counterfactual training employs counterfactuals during the training
-phase to minimize the divergence between learned representations and
-plausible, actionable explanations. We demonstrate empirically and
-theoretically that our proposed method facilitates training models that
-deliver inherently desirable explanations while promoting robustness and
-preserving high predictive performance of the models.
+feature mutability constraints. Much existing research has therefore
+focused on developing post-hoc methods to generate counterfactuals that
+meet these desiderata. In this work, we instead hold models directly
+accountable for the desired end goal: counterfactual training employs
+counterfactuals during the training phase to minimize the divergence
+between learned representations and plausible, actionable explanations.
+We demonstrate empirically and theoretically that our proposed method
+facilitates training models that deliver inherently desirable
+explanations and have improved adversarial robustness.
 \end{abstract}
@@
 factual inputs that, if implemented, would prompt some fitted model to
-produce an alternative, more desirable output (Figure~\ref{fig-poc}).
-This is useful and necessary to not only understand how opaque models
-make their predictions, but also to provide algorithmic recourse (AR) to
+produce an alternative, more desirable output. This is useful and
+necessary to not only understand how opaque models make their
+predictions, but also to provide algorithmic recourse (AR) to
 individuals subjected to them: a retail bank, for example, could use CE
 to provide meaningful feedback to unsuccessful loan applicants that were
-rejected based on an opaque automated decision-making (ADM) system.
+rejected based on an opaque automated decision-making (ADM) system
+(Figure~\ref{fig-poc}).
 
@@
 a young, unsuccessful loan applicant cannot increase their age in an
-instant.
+instance.
 
@@
 \citep{altmeyer2024faithful}. In this work, we therefore embrace the
-paradigm that models---as opposed to explanation methods---should be
-held accountable for explanations that are plausible and actionable.
-While previous work has shown that at least plausibility can be
-indirectly achieved through existing techniques aimed at models'
-generative capacity, generalization and robustness
+paradigm that models (as opposed to explanation methods) should be held
+accountable for explanations that are plausible and actionable. While
+previous work has shown that at least plausibility can be indirectly
+achieved through existing techniques aimed at models' generative
+capacity, generalization and robustness
 \citep{altmeyer2024faithful, augustin2020adversarial, schut2021generating},
@@
 methodological framework and show theoretically that it can be employed
-to enforce global actionability constraints. In our experiments
+to respect global actionability constraints. In our experiments
 (Section~\ref{sec-experiments}), we find that thanks to counterfactual
@@
 linear decision boundary is shown in green along with training data
-colored according to ground-truth labels: \(y^-=1\) (blue) and \(y^+=2\)
-(orange). Stars indicate counterfactuals in the target class. For
-illustration, synthetic features are annotated as \emph{debt} and
-\emph{age}.}
+colored according to ground-truth labels: \(y^-=\text{"loan withheld"}\)
+(blue) and \(y^+=\text{"loan provided"}\) (orange). Class and feature
+annotations (\emph{debt} and \emph{age}) are for illustrative purposes.}
 
@@
 
-To make the desiderata for our framework more concrete, we follow the
+To make the desiderata for our framework more concrete, we follow
 previous work in tying the explanatory capacity of models to the quality

### 2025-07-19 paper/sections/abstract.qmd
editorial pass until section 2

@@
 Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output.
-To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the user-defined mutability constraints.
+To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the feature mutability constraints.
 Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata.
 In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
-We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while promoting robustness and preserving high predictive performance of the models.
+We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations and have improved adversarial robustness.

### 2025-07-19 paper/sections/related.qmd
editorial pass until section 2

@@
 
-To make the desiderata for our framework more concrete, we follow the previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial].
+To make the desiderata for our framework more concrete, we follow previous work in tying the explanatory capacity of models to the quality of CEs that can be generated for them [@altmeyer2024faithful; @augustin2020adversarial].
 For simplicity, we refer to "explanatory capacity" as "explainability" in the rest of this manuscript (see Def.

### 2025-07-21 paper/quarto_aaai/paper.tex
editorial pass until section 3.4

@@
 more explainable and hence trustworthy models. The authors propose a way
-to generate counterfactuals that are maximally faithful to the model in
-that they are consistent with what the model has learned about the
-underlying data. Formally, they rely on tools from energy-based
-modelling \citep{teh2003energy} to minimize the divergence between the
+to generate counterfactuals that are maximally faithful models in that
+they are consistent with what models have learned about the underlying
+data. Formally, they rely on tools from energy-based modelling
+\citep{teh2003energy} to minimize the contrastive divergence between the
 distribution of counterfactuals and the conditional posterior over
-inputs learned by the model. Their counterfactual explainer,
-\emph{ECCCo}, yields plausible explanations if and only if the
-underlying model has learned representations that align with them. The
-authors find that both deep ensembles \citep{lakshminarayanan2016simple}
-and joint energy-based models (JEMs) \citep{grathwohl2020your}, a form
-of constrastive learning, tend to do well in this regard.
+inputs learned by a model. Their counterfactual explainer, \emph{ECCCo},
+yields plausible explanations if and only if the underlying model has
+learned representations that align with them. The authors find that both
+deep ensembles \citep{lakshminarayanan2016simple} and joint energy-based
+models (JEMs) \citep{grathwohl2020your}, a form of constrastive
+learning, tend to do well in this regard.
 
@@
 are underspecified by the available data \citep{wilson2020case}.
-Averaging across solutions mitigates the aforementioned risk of
-overrelying on a single locally optimal representation that corresponds
-to semantically meaningless explanations. Likewise, previous work of
+Averaging across solutions mitigates the risk of overrelying on a single
+locally optimal representation that corresponds to semantically
+meaningless explanations. Likewise, previous work of
 \citet{schut2021generating} found that generating plausible
@@
 classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
-\(D\) features and \(\mathbf{y}^+\) denotes some target class. The
-\(\text{yloss}(\cdot)\) function quantifies the discrepancy between
-current model predictions for \(\mathbf{x}^{\prime}\) and the target
-class (a conventional choice is cross-entropy). Finally, we use
+\(D\) features and \(\mathbf{y}^+\in\mathcal{Y}\) denotes some target
+class. The \(\text{yloss}(\cdot)\) function quantifies the discrepancy
+between current model predictions for \(\mathbf{x}^{\prime}\) and the
+target class (a conventional choice is cross-entropy). Finally, we use
 \(\text{reg}(\cdot)\) to denote any form of regularization used to
@@
 
-We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated counterfactuals are also plausible and actionable. We define these properties as follows:
+We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated, valid counterfactuals are also plausible and actionable.
+We define these properties as follows:
 
@@
 \end{itemize}
-and $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs.
+Here, $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs.
 For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript.
@@
 The characterization of faithfulness and plausibility in Def.
-\ref{def-explainability} is the same as in \citet{altmeyer2024faithful},
-with adapted notation. Intuitively, plausible counterfactuals are
-consistent with the data and faithful counterfactuals are consistent
-with what the model has learned about the input data. Actionability
-constraints in Def. \ref{def-explainability} vary and depend on the
-context in which \(\mathbf{M}_\theta\) is deployed. In this work, we
-choose to only consider domain and mutability constraints for individual
-features \(x_d\) for \(d=1,...,D\). We also limit ourselves to
-classification tasks for reasons discussed in
-Section~\ref{sec-discussion}.
+\ref{def-explainability} follows \citet{altmeyer2024faithful}, with
+adapted notation. Intuitively, plausible counterfactuals are consistent
+with the data and faithful counterfactuals are consistent with what the
+model has learned about the input data. Ac tionability constraints in
+Def. \ref{def-explainability} vary and depend on the context in which
+\(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
+consider mutability constraints for individual features \(x_d\) for
+\(d=1,...,D\). We also limit ourselves to classification tasks for
+reasons discussed in Section~\ref{sec-discussion}.
 
@@
 Formally, we propose the following counterfactual training objective to
-train explainable (as in Def. \ref{def-explainability}) models:
+train explainable (as in Def. \ref{def-explainability}) models,
 \begin{equation}\phantomsection\label{eq-obj}{
@@
 minimize the contrastive divergence between the observed samples from
-\(p(\mathbf{x})\) and the generated samples from
-\(p_\theta(\mathbf{x})\).
-
-A key empirical finding of \citet{altmeyer2024faithful} was that JEMs
-perform well on the plausibility objective in Def.
-\ref{def-explainability}. This follows directly if we consider samples
-drawn from \(p_\theta(\mathbf{x})\) as counterfactuals --- the JEM
-objective effectively minimizes the divergence between the conditional
-posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
-\citet{grathwohl2020your} use Stochastic Gradient Langevin Dynamics
-(SGLD) with an uninformative prior for initialization but we depart from
-their methodology. Instead we propose to leverage counterfactual
-explainers to generate counterfactuals of observed training samples.
-Specifically, we have: \begin{equation}\phantomsection\label{eq-div}{
+\(p(\mathbf{x})\) and samples generated from \(p_\theta(\mathbf{x})\).
+
+To generate samples, \citet{grathwohl2020your} use Stochastic Gradient
+Langevin Dynamics (SGLD) with an uninformative prior for initialization
+but we depart from their methodology: we propose to leverage
+counterfactual explainers to generate counterfactuals of observed
+training samples. Specifically, we have:
+\begin{equation}\phantomsection\label{eq-div}{
 \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
@@
 sampled factual from a non-target class generated with a gradient-based
-CE generator for up to \(T\) iterations. Mature counterfactuals are ones
-that have either reached convergence wrt. the decision threshold
-\(\tau\) or exhausted \(T\).
-
-Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
-of observed training samples (positive samples) while increasing the
-energy of counterfactuals (negative samples) \citep{du2019implicit}. As
-the counterfactuals get more plausible (Def. \ref{def-explainability})
-during training, these opposing effects gradually balance each other out
+CE generator for up to \(T\) iterations. Intuitively, the gradient of
+Equation~\ref{eq-div} decreases the energy of observed training samples
+(positive samples) while increasing the energy of counterfactuals
+(negative samples) \citep{du2019implicit}. As the counterfactuals get
+more plausible (Def. \ref{def-explainability}) during training, these
+opposing effects gradually balance each other out
 \citep{lippe2024uvadlc}.
 
-The departure from SGLD allows us to tap into the vast repertoire of
-explainers that have been proposed in the literature to meet different
-desiderata. For example, many methods support domain and mutability
-constraints. In principle, any existing approach for generating CEs is
-viable, so long as it does not violate the faithfulness condition. Like
-JEMs \citep{murphy2022probabilistic}, counterfactual training can be
+Since maturity of counterfactuals in terms of a probability threshold is
+often reached before \(T\), this form of sampling is not only more
+closely aligned with Def. \ref{def-explainability}., but can also speed
+up training times compared to SGLD. The departure from SGLD also allows
+us to tap into the vast repertoire of explainers that have been proposed
+in the literature to meet different desiderata. For example, many
+methods support domain and mutability constraints. In principle, any
+existing approach for generating CEs is viable, so long as it does not
+violate the faithfulness condition. Like JEMs
+\citep{murphy2022probabilistic}, counterfactual training can be
 considered a form of contrastive representation learning.

### 2025-07-21 paper/sections/method.qmd
adjusted Algorithm 1 and ensured consistet capitalization for CT, AR, CE etc.
editorial pass until section 3.4

@@
 $$ {#eq-general}
-where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+$ denotes some target class.
+where $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denotes a classifier, $\mathbf{x}^{\prime}$ denotes the counterfactual with $D$ features and $\mathbf{y}^+\in\mathcal{Y}$ denotes some target class.
 The $\text{yloss}(\cdot)$ function quantifies the discrepancy between current model predictions for $\mathbf{x}^{\prime}$ and the target class (a conventional choice is cross-entropy).
@@
 The characterization of faithfulness and plausibility in Def.
-\ref{def-explainability} is the same as in @altmeyer2024faithful, with adapted notation.
+\ref{def-explainability} follows @altmeyer2024faithful, with adapted notation.
 Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data.
-Actionability constraints in Def.
+Ac tionability constraints in Def.
 \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed.
-In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$.
+In this work, we choose to only consider mutability constraints for individual features $x_d$ for $d=1,...,D$.
 We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.
@@
 
-Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models:
+Formally, we propose the following counterfactual training objective to train explainable (as in Def. \ref{def-explainability}) models,
 $$
@@
 The authors show that JEMs can be trained to perform well at both tasks by directly maximizing the joint log-likelihood: $\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})$, where the first term can be optimized using cross-entropy as in @eq-obj.
-To optimize $\log p_\theta(\mathbf{x})$, they minimize the contrastive divergence between the observed samples from $p(\mathbf{x})$ and the generated samples from $p_\theta(\mathbf{x})$.
+To optimize $\log p_\theta(\mathbf{x})$, they minimize the contrastive divergence between the observed samples from $p(\mathbf{x})$ and samples generated from $p_\theta(\mathbf{x})$.
 
-A key empirical finding of @altmeyer2024faithful was that JEMs perform well on the plausibility objective in Def.
-\ref{def-explainability}.
-This follows directly if we consider samples drawn from $p_\theta(\mathbf{x})$ as counterfactuals --- the JEM objective effectively minimizes the divergence between the conditional posterior and $p(\mathbf{x}|\mathbf{y}^{+})$.
-To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology.
-Instead we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples.
+To generate samples, @grathwohl2020your use Stochastic Gradient Langevin Dynamics (SGLD) with an uninformative prior for initialization but we depart from their methodology: we propose to leverage counterfactual explainers to generate counterfactuals of observed training samples.
 Specifically, we have:
@@
 Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations.
-Mature counterfactuals are ones that have either reached convergence wrt.
-the decision threshold $\tau$ or exhausted $T$.
-
 Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit].
@@
 
-The departure from SGLD allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
+Since maturity of counterfactuals in terms of a probability threshold is often reached before $T$, this form of sampling is not only more closely aligned with Def.
+\ref{def-explainability}., but can also speed up training times compared to SGLD.
+The departure from SGLD also allows us to tap into the vast repertoire of explainers that have been proposed in the literature to meet different desiderata.
 For example, many methods support domain and mutability constraints.
 In principle, any existing approach for generating CEs is viable, so long as it does not violate the faithfulness condition.
-Like JEMs [@murphy2022probabilistic], Counterfactual Training can be considered a form of contrastive representation learning.
+Like JEMs [@murphy2022probabilistic], counterfactual training can be considered a form of contrastive representation learning.
 

### 2025-07-21 paper/sections/related.qmd
editorial pass until section 3.4

@@
 In a closely related work, @altmeyer2024faithful show that model averaging and, in particular, contrastive model objectives can produce more explainable and hence trustworthy models.
-The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data.
-Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model.
+The authors propose a way to generate counterfactuals that are maximally faithful models in that they are consistent with what models have learned about the underlying data.
+Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the contrastive divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by a model.
 Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
@@
 Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case].
-Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations.
+Averaging across solutions mitigates the risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations.
 Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training.

### 2025-07-21 paper/quarto_aaai/paper.tex
done with editorail pass

@@
 \(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
-consider mutability constraints for individual features \(x_d\) for
-\(d=1,...,D\). We also limit ourselves to classification tasks for
-reasons discussed in Section~\ref{sec-discussion}.
+consider domain and mutability constraints for individual features
+\(x_d\) for \(d=1,...,D\). We also limit ourselves to classification
+tasks for reasons discussed in Section~\ref{sec-discussion}.
 
@@
 Many existing counterfactual explainers support domain and mutability
-constraints out-of-the-box. In fact, both types of constraints can be
-implemented for any explainer that relies on gradient descent in the
-feature space for optimization \citep{altmeyer2023explaining}. In this
-context, domain constraints can be imposed by simply projecting
-counterfactuals back to the specified domain, if the previous gradient
-step resulted in updated feature values that were out-of-domain.
-Similarly, mutability constraints can be enforced by setting partial
-derivatives to zero to ensure that features are only perturbed in the
-allowed direction, if at all.
-
-Since actionability constraints are binding at test time, we should also
-impose them when generating \(\mathbf{x}^\prime\) during each training
+constraints. In fact, both types of constraints can be implemented for
+any explainer that relies on gradient descent in the feature space for
+optimization \citep{altmeyer2023explaining}. In this context, domain
+constraints can be imposed by simply projecting counterfactuals back to
+the specified domain, if the previous gradient step resulted in updated
+feature values that were out-of-domain. Similarly, mutability
+constraints can be enforced by setting partial derivatives to zero to
+ensure that features are only perturbed in the allowed direction, if at
+all.
+
+Since actionability constraints are binding at test time, we also impose
+them when generating \(\mathbf{x}^\prime\) during each training
 iteration to inform model representations. Through their effect on
@@
 plausibility through the remaining features. This can be expected to
-result in lower overall sensitivity to immutable features, which we
-investigate empirically in Section~\ref{sec-experiments}. Under certain
-conditions, this result holds theoretically; for the proof, see the
-supplementary appendix:
+result in relatively lower sensitivity to immutable features; and higher
+relative sensitivity to mutable features should make
+mutability-constrained recourse less costly
+(Section~\ref{sec-experiments}). Under certain conditions, this result
+holds theoretically; for the proof, see the supplementary appendix:
 
@@
 proxy of distances (\(\ell_1\)-norm) between factuals and
-counterfactuals. For plausibility, we assess how similar CEs are to the
+counterfactuals. For plausibility, we assess how similar CEs are to
 observed samples in the target domain,
@@
 clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
-moons (\emph{Moon}). They are generated using the library of
-\citet{altmeyer2023explaining} and we present them in the supplementary
-appendix. Next, we have four real-world binary tabular datasets:
-\emph{Adult} (Census data) of \citet{becker1996adult2}, California
-housing (\emph{CH}) of \citet{pace1997sparse}, Default of Credit Card
-Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me Some Credit
-(\emph{GMSC}) from \citet{kaggle2011give}. Finally, for the convenience
-of illustration, we use the 10-class \emph{MNIST}
+moons (\emph{Moon}). Next, we have four real-world binary tabular
+datasets: \emph{Adult} (Census data) of \citet{becker1996adult2},
+California housing (\emph{CH}) of \citet{pace1997sparse}, Default of
+Credit Card Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me
+Some Credit (\emph{GMSC}) from \citet{kaggle2011give}. Finally, for the
+convenience of illustration, we use the 10-class \emph{MNIST}
 \citep{lecun1998mnist}.
@@
 counterfactuals are around 30-55\% more plausible for models trained
-with CT, in line with our observations in Figure~\ref{fig-poc} in panel
-(b) compared to panel (a).
+with CT, in line with our observations in panel (b) of
+Figure~\ref{fig-poc} compared to panel (a).
 
@@
 \emph{CT generally produces substantial and statistically significant
-improvements in plausibility.}\\
+improvements in plausibility.}
+
 Average reductions in \(\text{IP}\) range from around 7\% for
@@
 \emph{CT tends to improve actionability in the presence of immutable
-features, but the exact effect is sometimes hard to establish due to,
-e.g., feature interactions.}
+features, but this is not guaranteed if the assumptions in Prp.
+\ref{prp-mtblty} are violated.}
 
@@
 
-First, we look at hyperparameters that guide the gradient-based
-counterfactual search. Increasing \(T\), the maximum number of steps,
-generally yields better outcomes because more CEs can mature in each
-training epoch. The impact of \(\tau\), the required decision threshold
-is more difficult to predict. On ``harder'' datasets it may be difficult
-to satisfy high \(\tau\) for any given sample (i.e., also factuals) and
-so increasing this threshold does not seem to correlate with better
-outcomes. In fact, \(\tau=0.5\) generally leads to optimal results as it
-is associated with high proportions of mature counterfactuals.
-
-Second, the strength of the energy regularization,
-\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
-performance in terms of decreased plausibility and increased costs if
-insufficiently high. The sensitivity with respect to
-\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
-evident. While high values of \(\lambda_{\text{reg}}\) may increase the
-variability in outcomes when combined with high values of
-\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
-particularly pronounced.
-
-Third, the effectiveness and stability of CT is positively associated
-with the number of counterfactuals generated during each training epoch.
-A higher number of training epochs is also beneficial. Interestingly, we
-observed desired improvements when CT was combined with conventional
-training and applied only for the final 50\% of epochs of the complete
-training process. Put differently, CT can improve the explainability of
-models in a fine-tuning manner.
+Firstly, we find that optimal results are generally obtained when using
+\emph{ECCCo} to generate counterfactuals. Conversely, using a generator
+that may inhibit faithfulness (\emph{REVISE}), tends to yield poor
+results. Concerning hyperparameters that guide the gradient-based
+counterfactual search, we find that increasing \(T\), the maximum number
+of steps, generally yields better outcomes because more CEs can mature.
+Relatedly, we also find that the effectiveness and stability of CT is
+positively associated with the total number of counterfactuals generated
+during each training epoch. The impact of \(\tau\), the decision
+threshold, is more difficult to predict. On ``harder'' datasets it may
+be difficult to satisfy high \(\tau\) for any given sample (i.e., also
+factuals) and so increasing this threshold does not seem to correlate
+with better outcomes. In fact, \(\tau=0.5\) generally leads to optimal
+results as it is associated with high proportions of mature
+counterfactuals.
+
+Secondly, the strength of the energy regularization,
+\(\lambda_{\text{reg}}\) is highly impactful and should be set
+sufficiently high to avoid common problems associated with exploding
+gradients. The sensitivity with respect to \(\lambda_{\text{div}}\) and
+\(\lambda_{\text{adv}}\) is much less evident. While high values of
+\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
+combined with high values of \(\lambda_{\text{div}}\) or
+\(\lambda_{\text{adv}}\), this effect is not particularly pronounced.
+
+Finally, we also observe desired improvements when CT was combined with
+conventional training and applied only for the final 50\% of epochs of
+the complete training process. Put differently, CT can improve the
+explainability of models in a post-hoc, fine-tuning manner.
 
@@
 
-\emph{CT increases the training times.} Like adversarial training, CT is
-more resource-intensive than conventional regimes. Higher numbers of CEs
-improve the quality of the learned representations but they also
-increase the number of computations. As our codebase is not performance
-optimized, grids of around 300 settings for the largest datasets in our
-experiments took up to four hours using 34 2GB CPUs (see supplementary
-appendix). Other than optimization, three factors mitigate this effect:
-(1) CT yields itself to parallel execution; (2) it amortizes the cost of
-CEs for the training samples; and (3) it can be used to fine-tune
-conventionally-trained models.
+\emph{Plausibility is costly.} As noted elsewhere
+\citep{altmeyer2024faithful}, more plausible counterfactuals are
+inevitably more costly. Thus, while CT improves plausibility and
+robustness, it can impact average costs and validity as cheap,
+implausible and adversarial explanations are removed from the solution
+space.
+
+\emph{CT increases the training times.} Like contrastive and robust
+learning, CT is more resource-intensive than conventional regimes. Three
+factors mitigate this effect: (1) CT yields itself to parallel
+execution; (2) it amortizes the cost of CEs for the training samples;
+and (3) it can be used to fine-tune conventionally-trained models.
 
@@
 
-Secondly, CT is susceptible to training instabilities, as is also the
-case for the related JEMs \citep{grathwohl2020your}. CT is exposed to
-two sources of instabilities: (1) the energy-based contrastive
-divergence term, \(\text{div}(\cdot)\), in Equation~\ref{eq-div}, and
-(2) the underlying explainers. We find several promising ways to
-mitigate this problem: regularizing energy (\(\lambda_{\text{reg}}\)),
-generating sufficiently many counterfactuals during each epoch, and
-including only mature counterfactuals in \(\text{div}(\cdot)\).
+Secondly, it is worth extending the analysis done here to additional CE
+generators including those that do not rely on computationally costly
+gradient-based optimization. This would reduce training costs while
+potentially maintaining benefits of CT.
 

### 2025-07-21 paper/sections/discussion.qmd
done with editorail pass

@@
 
-*CT increases the training times.* Like adversarial training, CT is more resource-intensive than conventional regimes.
-Higher numbers of CEs improve the quality of the learned representations but they also increase the number of computations.
-As our codebase is not performance optimized, grids of around 300 settings for the largest datasets in our experiments took up to four hours using 34 2GB CPUs (see supplementary appendix).
-Other than optimization, three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.
+*Plausibility is costly.* As noted elsewhere [@altmeyer2024faithful], more plausible counterfactuals are inevitably more costly.
+Thus, while CT improves plausibility and robustness, it can impact average costs and validity as cheap, implausible and adversarial explanations are removed from the solution space.
+
+*CT increases the training times.* Like contrastive and robust learning, CT is more resource-intensive than conventional regimes.
+Three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.
 
@@
 Focus on classification is a common choice in research on CEs and AR; other settings have attracted some interest, e.g., regression [@spooner2021counterfactual], but there is little consensus how to robustly extend the notion of CEs.
-Secondly, CT is susceptible to training instabilities, as also recognized for the related JEMs [@grathwohl2020your].
-CT is exposed to two sources of instabilities: (1) the energy-based contrastive divergence term, $\text{div}(\cdot)$, in @eq-div, and (2) the underlying explainers.
-We find several promising ways to mitigate this problem: regularizing energy ($\lambda_{\text{reg}}$), generating sufficiently many counterfactuals during each epoch, and including only mature counterfactuals in $\text{div}(\cdot)$.
+
+Secondly, it is worth extending the analysis done here to additional CE generators including those that do not rely on computationally costly gradient-based optimization.
+This would reduce training costs while potentially maintaining benefits of CT.
+
 Finally, we believe that it is possible to considerably improve hyperparameter selection procedures, and thus performance.

### 2025-07-21 paper/sections/experiments.qmd
done with editorail pass

@@
 To measure the cost, we follow the standard proxy of distances ($\ell_1$-norm) between factuals and counterfactuals.
-For plausibility, we assess how similar CEs are to the observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$.
+For plausibility, we assess how similar CEs are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$.
 We rely on the metric used by @altmeyer2024faithful,
@@
 Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*).
-They are generated using the library of @altmeyer2023explaining and we present them in the supplementary appendix.
 Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give.
@@
 The first two columns ($\text{IP}$ and $\text{IP}^*$) show the percentage reduction in implausibility for our two metrics when using CT on top of the weak baseline.
-As an example, consider the first row for *LS* data: the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in @fig-poc in panel (b) compared to panel (a).
+As an example, consider the first row for *LS* data: the observed positive values indicate that faithful counterfactuals are around 30-55% more plausible for models trained with CT, in line with our observations in panel (b) of @fig-poc compared to panel (a).
 
@@
 *CT generally produces substantial and statistically significant improvements in plausibility.*
+
 Average reductions in $\text{IP}$ range from around 7% for *MNIST* to almost 60% for *Circ*.
@@
 ### Actionability (RQ2). {#sec-act}
-*CT tends to improve actionability in the presence of immutable features, but the exact effect is sometimes hard to establish due to, e.g., feature interactions.*
+*CT tends to improve actionability in the presence of immutable features, but this is not guaranteed if the assumptions in Prp. \ref{prp-mtblty} are violated.*
 
@@
 
-First, we look at hyperparameters that guide the gradient-based counterfactual search.
-Increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature in each training epoch.
-The impact of $\tau$, the required decision threshold is more difficult to predict.
+Firstly, we find that optimal results are generally obtained when using *ECCCo* to generate counterfactuals.
+Conversely, using a generator that may inhibit faithfulness (*REVISE*), tends to yield poor results.
+Concerning hyperparameters that guide the gradient-based counterfactual search, we find that increasing $T$, the maximum number of steps, generally yields better outcomes because more CEs can mature.
+Relatedly, we also find that the effectiveness and stability of CT is positively associated with the total number of counterfactuals generated during each training epoch.
+The impact of $\tau$, the decision threshold, is more difficult to predict.
 On "harder" datasets it may be difficult to satisfy high $\tau$ for any given sample (i.e., also factuals) and so increasing this threshold does not seem to correlate with better outcomes.
@@
 
-Second, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and leads to poor performance in terms of decreased plausibility and increased costs if insufficiently high.
+Secondly, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and should be set sufficiently high to avoid common problems associated with exploding gradients.
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.
@@
 
-Third, the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch.
-A higher number of training epochs is also beneficial.
-Interestingly, we observed desired improvements when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process.
-Put differently, CT can improve the explainability of models in a fine-tuning manner.
+Finally, we also observe desired improvements when CT was combined with conventional training and applied only for the final 50% of epochs of the complete training process.
+Put differently, CT can improve the explainability of models in a post-hoc, fine-tuning manner.

### 2025-07-21 paper/sections/method.qmd
done with editorail pass

@@
 \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed.
-In this work, we choose to only consider mutability constraints for individual features $x_d$ for $d=1,...,D$.
+In this work, we choose to only consider domain and mutability constraints for individual features $x_d$ for $d=1,...,D$.
 We also limit ourselves to classification tasks for reasons discussed in @sec-discussion.
@@
 
-Many existing counterfactual explainers support domain and mutability constraints out-of-the-box.
+Many existing counterfactual explainers support domain and mutability constraints.
 In fact, both types of constraints can be implemented for any explainer that relies on gradient descent in the feature space for optimization [@altmeyer2023explaining].
@@
 
-Since actionability constraints are binding at test time, we should also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations.
+Since actionability constraints are binding at test time, we also impose them when generating $\mathbf{x}^\prime$ during each training iteration to inform model representations.
 Through their effect on $\mathbf{x}^\prime$, both types of constraints influence model outcomes via @eq-div.
@@
 Intuitively, we think of this as ignoring implausibility costs of immutable features, which effectively forces the model to instead seek plausibility through the remaining features.
-This can be expected to result in lower overall sensitivity to immutable features, which we investigate empirically in @sec-experiments.
+This can be expected to result in relatively lower sensitivity to immutable features; and higher relative sensitivity to mutable features should make mutability-constrained recourse less costly (@sec-experiments).
 Under certain conditions, this result holds theoretically; for the proof, see the supplementary appendix:

### 2025-07-21 paper/quarto_aaai/paper.tex
Clarify some sentences

@@
 facilitates training models that deliver inherently desirable
-explanations and have improved adversarial robustness.
+counterfactual explanations and exhibit greatly improved adversarial
+robustness.
 \end{abstract}
@@
 more explainable and hence trustworthy models. The authors propose a way
-to generate counterfactuals that are maximally faithful models in that
-they are consistent with what models have learned about the underlying
-data. Formally, they rely on tools from energy-based modelling
+to generate counterfactuals that are maximally faithful in that they are
+consistent with what models have learned about the underlying data.
+Formally, they rely on tools from energy-based modelling
 \citep{teh2003energy} to minimize the contrastive divergence between the
 distribution of counterfactuals and the conditional posterior over
-inputs learned by a model. Their counterfactual explainer, \emph{ECCCo},
-yields plausible explanations if and only if the underlying model has
-learned representations that align with them. The authors find that both
-deep ensembles \citep{lakshminarayanan2016simple} and joint energy-based
-models (JEMs) \citep{grathwohl2020your}, a form of constrastive
-learning, tend to do well in this regard.
+inputs learned by a model. Their algorithm, \emph{ECCCo}, yields
+plausible counterfactual explanations if and only if the underlying
+model has learned representations that align with them. The authors find
+that both deep ensembles \citep{lakshminarayanan2016simple} and joint
+energy-based models (JEMs) \citep{grathwohl2020your}, a form of
+constrastive learning, tend to do well in this regard.
 
@@
 
-We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated, valid counterfactuals are also plausible and actionable.
-We define these properties as follows:
+We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated, valid counterfactuals are also plausible and actionable. We define these properties as:
 
@@
 \emph{CT tends to improve actionability in the presence of immutable
-features, but this is not guaranteed if the assumptions in Prp.
+features, but this is not guaranteed if the assumptions in Proposition
 \ref{prp-mtblty} are violated.}
@@
 
-\emph{Plausibility is costly.} As noted elsewhere
-\citep{altmeyer2024faithful}, more plausible counterfactuals are
-inevitably more costly. Thus, while CT improves plausibility and
-robustness, it can impact average costs and validity as cheap,
-implausible and adversarial explanations are removed from the solution
-space.
-
-\emph{CT increases the training times.} Like contrastive and robust
+\emph{Plausibility is costly.} As noted by \citet{altmeyer2024faithful},
+more plausible counterfactuals are inevitably more costly. CT improves
+plausibility and robustness, but it can impact average costs and
+validity when cheap, implausible and adversarial explanations are
+removed from the solution space.
+
+\emph{CT increases the training times.} Just like contrastive and robust
 learning, CT is more resource-intensive than conventional regimes. Three
@@
 
-Secondly, it is worth extending the analysis done here to additional CE
-generators including those that do not rely on computationally costly
-gradient-based optimization. This would reduce training costs while
-potentially maintaining benefits of CT.
+Secondly, our analysis covers CE generators with different
+characteristics, but it is interesting to extend it to more algorithms,
+including ones that do not rely on computationally costly gradient-based
+optimization. This should reduce training costs while possibly
+preserving the benefits of CT.
 

### 2025-07-21 paper/sections/experiments.qmd
Clarify some sentences

@@
 ### Actionability (RQ2). {#sec-act}
-*CT tends to improve actionability in the presence of immutable features, but this is not guaranteed if the assumptions in Prp. \ref{prp-mtblty} are violated.*
+*CT tends to improve actionability in the presence of immutable features, but this is not guaranteed if the assumptions in Proposition \ref{prp-mtblty} are violated.*
 

### 2025-07-21 paper/quarto_aaai/paper.tex
adjsutments to remove Quarto generated latex

@@
 \centering
-\input{tables/main.tex}
+\begin{tabular}{
+  l
+  S[table-format=2.2(1.2)]
+  S[table-format=3.2(3.2)]
+  S[table-format=3.2(1.2)]
+}
+  \toprule
+  \textbf{Data} & \textbf{$ \text{IP} $ $(-\%)$} & \textbf{$ \text{IP}^* $ $(-\%)$} & \textbf{Cost $(-\%)$} \\\midrule
+  LS & 29.05\pm0.67 $^{*}$ & 55.33\pm2.03 $^{*}$ & 14.07\pm0.6 $^{*}$ \\
+  Circ & 56.29\pm0.44 $^{*}$ & 89.38\pm9.3 $^{*}$ & 45.55\pm0.76 $^{*}$ \\
+  Moon & 20.62\pm0.69 $^{*}$ & 19.26\pm8.12 $^{*}$ & 2.86\pm1.03 $^{*}$ \\
+  OL & -1.13\pm0.88 $^{}$ & -24.52\pm14.52 $^{}$ & 38.39\pm2.21 $^{*}$ \\\midrule
+  Adult & 0.77\pm1.34 $^{}$ & 32.29\pm6.87 $^{*}$ & -2.82\pm4.88 $^{}$ \\
+  CH & 12.05\pm1.41 $^{*}$ & 70.27\pm3.72 $^{*}$ & 40.71\pm1.55 $^{*}$ \\
+  Cred & 12.31\pm1.84 $^{*}$ & 54.89\pm11.21 $^{*}$ & -17.43\pm5.17 $^{*}$ \\
+  GMSC & 23.44\pm1.99 $^{*}$ & 73.31\pm4.83 $^{*}$ & 62.64\pm2.04 $^{*}$ \\
+  MNIST & 7.05\pm1.8 $^{*}$ & -25.09\pm109.05 $^{}$ & -12.34\pm6.52 $^{}$ \\\midrule
+  Avg. & 17.83 & 38.35 & 19.07 \\\bottomrule
+\end{tabular}
 
@@
 
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\renewcommand{\bibsection}{}
 \bibliography{../quarto_aaai/bibliography.bib}

### 2025-07-21 paper/sections/experiments.qmd
adjsutments to remove Quarto generated latex

@@
 \centering
-\input{tables/main.tex}
+\begin{tabular}{
+  l
+  S[table-format=2.2(1.2)]
+  S[table-format=3.2(3.2)]
+  S[table-format=3.2(1.2)]
+}
+  \toprule
+  \textbf{Data} & \textbf{$ \text{IP} $ $(-\%)$} & \textbf{$ \text{IP}^* $ $(-\%)$} & \textbf{Cost $(-\%)$} \\\midrule
+  LS & 29.05\pm0.67 $^{*}$ & 55.33\pm2.03 $^{*}$ & 14.07\pm0.6 $^{*}$ \\
+  Circ & 56.29\pm0.44 $^{*}$ & 89.38\pm9.3 $^{*}$ & 45.55\pm0.76 $^{*}$ \\
+  Moon & 20.62\pm0.69 $^{*}$ & 19.26\pm8.12 $^{*}$ & 2.86\pm1.03 $^{*}$ \\
+  OL & -1.13\pm0.88 $^{}$ & -24.52\pm14.52 $^{}$ & 38.39\pm2.21 $^{*}$ \\\midrule
+  Adult & 0.77\pm1.34 $^{}$ & 32.29\pm6.87 $^{*}$ & -2.82\pm4.88 $^{}$ \\
+  CH & 12.05\pm1.41 $^{*}$ & 70.27\pm3.72 $^{*}$ & 40.71\pm1.55 $^{*}$ \\
+  Cred & 12.31\pm1.84 $^{*}$ & 54.89\pm11.21 $^{*}$ & -17.43\pm5.17 $^{*}$ \\
+  GMSC & 23.44\pm1.99 $^{*}$ & 73.31\pm4.83 $^{*}$ & 62.64\pm2.04 $^{*}$ \\
+  MNIST & 7.05\pm1.8 $^{*}$ & -25.09\pm109.05 $^{}$ & -12.34\pm6.52 $^{}$ \\\midrule
+  Avg. & 17.83 & 38.35 & 19.07 \\\bottomrule
+\end{tabular}
 \end{table}

### 2025-07-22 paper/quarto_aaai/paper.tex

@@
 in costs when selected features are protected. Outcomes are aggregated
-across bootstrap samples and varying degrees of the energy penalty
-\(\lambda_{\text{egy}}\) used for \emph{ECCCo} at test time. Asterisks
-(\(^*\)) indicate that the bootstrapped 99\%-confidence interval of
-differences in mean outcomes does \emph{not} include zero.}
+across bootstrap samples (100 rounds) and varying degrees of the energy
+penalty \(\lambda_{\text{egy}}\) used for \emph{ECCCo} at test time.
+Asterisks (\(^*\)) indicate that the bootstrapped 99\%-confidence
+interval of differences in mean outcomes does \emph{not} include zero.}
 

### 2025-07-22 paper/sections/abstract.qmd

@@
 In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals during the training phase to minimize the divergence between learned representations and plausible, actionable explanations.
-We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations and have improved adversarial robustness.
+We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable counterfactual explanations and exhibit greatly improved adversarial robustness.

### 2025-07-22 paper/sections/discussion.qmd

@@
 
-*Plausibility is costly.* As noted elsewhere [@altmeyer2024faithful], more plausible counterfactuals are inevitably more costly.
-Thus, while CT improves plausibility and robustness, it can impact average costs and validity as cheap, implausible and adversarial explanations are removed from the solution space.
+*Plausibility is costly.* As noted by @altmeyer2024faithful, more plausible counterfactuals are inevitably more costly.
+CT improves plausibility and robustness, but it can impact average costs and validity when cheap, implausible and adversarial explanations are removed from the solution space.
 
-*CT increases the training times.* Like contrastive and robust learning, CT is more resource-intensive than conventional regimes.
+*CT increases the training times.* Just like contrastive and robust learning, CT is more resource-intensive than conventional regimes.
 Three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.
@@
 
-Secondly, it is worth extending the analysis done here to additional CE generators including those that do not rely on computationally costly gradient-based optimization.
-This would reduce training costs while potentially maintaining benefits of CT.
+Secondly, our analysis covers CE generators with different characteristics, but it is interesting to extend it to more algorithms, including ones that do not rely on computationally costly gradient-based optimization.
+This should reduce training costs while possibly preserving the benefits of CT.
 

### 2025-07-22 paper/sections/experiments.qmd

@@
 ```
+
 Key evaluation metrics for valid counterfactual along with bootstrapped standard errors for all datasets.
 **Plausibility** (columns 1-2): percentage reduction in implausibility for $\text{IP}$ and $\text{IP}^*$, respectively; **Cost** / **Actionability** (column 3): percentage reduction in costs when selected features are protected.
-Outcomes are aggregated across bootstrap samples and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCCo* at test time.
+Outcomes are aggregated across bootstrap samples (100 rounds) and varying degrees of the energy penalty $\lambda_{\text{egy}}$ used for *ECCCo* at test time.
 Asterisks ($^*$) indicate that the bootstrapped 99%-confidence interval of differences in mean outcomes does *not* include zero.

### 2025-07-22 paper/sections/related.qmd

@@
 In a closely related work, @altmeyer2024faithful show that model averaging and, in particular, contrastive model objectives can produce more explainable and hence trustworthy models.
-The authors propose a way to generate counterfactuals that are maximally faithful models in that they are consistent with what models have learned about the underlying data.
+The authors propose a way to generate counterfactuals that are maximally faithful in that they are consistent with what models have learned about the underlying data.
 Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the contrastive divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by a model.
-Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them.
+Their algorithm, *ECCCo*, yields plausible counterfactual explanations if and only if the underlying model has learned representations that align with them.
 The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your], a form of constrastive learning, tend to do well in this regard.

### 2025-07-22 paper/sections/other/main_results.qmd
nearly there

@@
     res_dir;
-    y
+    y,
     return_sig_level=false, 
@@
 df_mmd.measure .= y
-df = innerjoin(df_mtbl, df_ip, df_mmd; on=:dataset)
+df = vcat(df_mtbl, df_ip, df_mmd)
+Serialization.serialize("paper/experiments/output/ci.jls", df)
+```
+
+```{julia}
+#| output: asis
+#| label: tbl-ci
+#| tbl-cap: "Confidence Intervals"
+
+df = Serialization.deserialize("paper/experiments/output/ci.jls")
+df = select(df, :variable, :dataset, :full, :vanilla, "99.0%")
+df = transform(df, "99.0%" => AsTable) |>
+    df -> select(df, Not("99.0%"))
+rename!(df, :dataset => :data, :x1 => :LB, :x2 => :UB)
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+tabulate_results(
+    inputs; 
+    table_type=:tabular,
+    alignment = [:l, fill(:c, ncol(df)-1)...]
+)
+
 ```

### 2025-07-22 paper/preprint/appendix.qmd
finished table and rerendered preprint

---
title: Supplementary Appendix
engine: julia
julia: 
  exeflags: ["--project=../experiments/"]
format:
  arxiv-pdf:
    keep-tex: true  
    linenumbers: false
    doublespacing: false
    runninghead: "Counterfactual Training"
    authorcols: true
    include-in-header:
      text: |
        \usepackage[title]{appendix}
        \usepackage{placeins}
        \usepackage{amsthm}
        \theoremstyle{plain}
        \newtheorem{proposition}{Proposition}[section]
        \theoremstyle{definition}
        \newtheorem{definition}{Definition}[section]
        \theoremstyle{definition}
        \newtheorem{example}{Example}[section]
        \theoremstyle{plain}
---

```{=latex}
\begin{appendices}
```

{{< include /paper/sections/appendix.qmd >}}

```{=latex}
\end{appendices}
```

### 2025-07-22 paper/preprint/appendix.tex
finished table and rerendered preprint

--- 2025-06-26 paper/preprint/paper.tex
@@
 \usepackage{placeins}
+\usepackage{siunitx}
 \makeatletter
@@
 \hypersetup{
-  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
+  pdftitle={Supplementary Appendix},
   pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
@@
 \renewcommand{\runninghead}{Counterfactual Training }
-\title{Counterfactual Training: Teaching Models Plausible and Actionable
-Explanations}
+\title{Supplementary Appendix}
 \def\asep{\\\\\\ } % default: all authors on same column
@@
 \maketitle
-\begin{abstract}
-We propose a novel training regime termed counterfactual training that
-leverages counterfactual explanations to increase the explanatory
-capacity of models. Counterfactual explanations have emerged as a
-popular post-hoc explanation method for opaque machine learning models:
-they inform how factual inputs would need to change in order for a model
-to produce some desired output. To be useful in real-word
-decision-making systems, counterfactuals should be (1) plausible with
-respect to the underlying data and (2) actionable with respect to the
-user-defined mutability constraints. Much existing research has
-therefore focused on developing post-hoc methods to generate
-counterfactuals that meet these desiderata. In this work, we instead
-hold models directly accountable for the desired end goal:
-counterfactual training employs counterfactuals ad-hoc during the
-training phase to minimize the divergence between learned
-representations and plausible, actionable explanations. We demonstrate
-empirically and theoretically that our proposed method facilitates
-training models that deliver inherently desirable explanations while
-promoting robustness and preserving high predictive performance.
-\end{abstract}
 {\bfseries \emph Keywords}
@@
 
-\section{Introduction}\label{introduction}
-
-Today's prominence of artificial intelligence (AI) has largely been
-driven by \textbf{representation learning}: instead of relying on
-features and rules that are carefully hand-crafted by humans, modern
-machine learning (ML) models are tasked with learning representations
-directly from data, guided by narrow objectives such as predictive
-accuracy (\citeproc{ref-goodfellow2016deep}{Goodfellow, Bengio, and
-Courville 2016}). Modern advances in computing have made it possible to
-provide such models with ever-growing degrees of freedom to achieve that
-task, which frequently allows them to outperform traditionally more
-parsimonious models. Unfortunately, in doing so, models learn
-increasingly complex and highly sensitive representations that humans
-can no longer easily interpret.
-
-The trend towards complexity for the sake of performance has come under
-serious scrutiny in recent years. At the very cusp of the deep learning
-(DL) revolution, Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2014}) showed that artificial
-neural networks (ANN) are sensitive to adversarial examples (AEs):
-perturbed versions of data instances that yield vastly different model
-predictions despite being ``imperceptible'' in that they are
-semantically indifferent from their factual counterparts. Even though
-some partially effective mitigation strategies have been proposed---most
-notably \textbf{adversarial training}
-(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
-2015})---truly robust deep learning remains unattainable even for models
-that are considered ``shallow'' by today's standards
-(\citeproc{ref-kolter2023keynote}{Kolter 2023}).
-
-Part of the problem is that the high degrees of freedom provide room for
-many solutions that are locally optimal with respect to narrow
-objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
-  follow the standard ML convention, where ``degrees of freedom'' refer
-  to the number of parameters estimated from data.} Indeed, recent work
-on the so-called ``lottery ticket hypothesis'' suggests that modern
-neural networks can be pruned by up to 90\% while preserving their
-predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
-Carbin 2019}). Similarly, Zhang et al.
-(\citeproc{ref-zhang2021understanding}{2021}) showed that
-state-of-the-art neural networks are expressive enough to fit randomly
-labeled data. Thus, looking at the predictive performance alone, the
-solutions may seem to provide compelling explanations for the data, when
-in fact they are based on purely associative, semantically meaningless
-patterns. This poses two challenges. Firstly, there is no dependable way
-to verify if representations correspond to meaningful, plausible
-explanations. Secondly, even if we could resolve the first challenge, it
-remains undecided how to ensure that models can \emph{only} learn
-valuable explanations.
-
-The first challenge has attracted an abundance of research on
-\textbf{explainable AI} (XAI), a paradigm that focuses on the
-development of tools to derive (post-hoc) explanations from complex
-model representations. Such explanations should mitigate a scenario in
-which practitioners deploy opaque models and blindly rely on their
-predictions. On countless occasions, this has happened in practice and
-caused real harms to people who were adversely and unfairly affected by
-automated decision-making (ADM) systems involving opaque models (see,
-e.g., (\citeproc{ref-oneil2016weapons}{O'Neil 2016})). Effective XAI
-tools can aid us in monitoring models and providing recourse to
-individuals to turn negative outcomes (e.g., ``loan application
-rejected'') into positive ones (e.g., ``application accepted''). In line
-with this, our work builds upon \textbf{counterfactual explanations}
-(CE) proposed by Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}) as an effective
-approach to achieve this goal. CEs prescribe minimal changes for factual
-inputs that, if implemented, would prompt some fitted model to produce a
-desired output.
-
-To our surprise, the second challenge has not yet attracted major
-research interest. Specifically, there has been no concerted effort
-towards improving the ``explanatory capacity'' of models, i.e., the
-degree to which learned representations correspond to explanations that
-are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
-Def. \ref{def-explainability}). Instead, the choice has generally been
-to improve the ability of XAI tools to identify the subset of
-explanations that are both plausible and valid for any given model,
-independent of whether the learned representations are in fact
-compatible with plausible explanations
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-Fortunately, recent findings indicate that improved explanatory capacity
-can arise as a consequence of regularization techniques aimed at other
-training objectives such as generative capacity, generalization, or
-robustness (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024};
-\citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein 2020};
-\citeproc{ref-schut2021generating}{Schut et al. 2021}). As further
-discussed in Section~\ref{sec-lit}, our work consolidates these findings
-within a single objective.
-
-\textbf{Specifically, we introduce Counterfactual Training (CT)}: a
-novel training regime explicitly geared towards improving the
-explainability of models. In high-level terms, we define this concept as
-as the extent to which valid explanations derived for an opaque model
-are also deemed plausible with respect to the underlying data and the
-global actionability constraints. To the best of our knowledge, our
-framework represents the first attempt to address this challenge by
-employing counterfactual explanations already in the training phase.
-
-The remainder of this manuscript is structured as follows.
-Section~\ref{sec-lit} presents related work, focusing on the link
-between AEs and CEs. Then follow our two principal contributions. In
-Section~\ref{sec-method}, we introduce our methodological framework and
-show theoretically that it can be employed to enforce global
-actionability constraints. In Section~\ref{sec-experiments}, through
-extensive experiments, we demonstrate that CT substantially improves
-explainability and positively contributes to the robustness of trained
-models without sacrificing predictive performance. Finally, in
-Section~\ref{sec-discussion}, we discuss the challenges and conclude
-that CT is a promising approach towards making opaque models more
-trustworthy.
-
-\section{Related Literature}\label{sec-lit}
-
-To make the desiderata for our framework more concrete, we follow
-Augustin, Meinke, and Hein
-(\citeproc{ref-augustin2020adversarial}{2020}) in tying the concept of
-explainability to the quality of CEs that can be generated for a given
-model. The authors show that CEs---understood as minimal input
-perturbations that yield some desired model prediction---tend to be more
-meaningful if the underlying model is more robust to adversarial
-examples. We can make intuitive sense of this finding when looking at
-adversarial training (AT) through the lens of representation learning
-with high degrees of freedom. As argued before, learned representations
-may be sensitive to producing implausible explanations and mispredicting
-for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to
-``unlearn'' susceptiblity to such examples, adversarial training can
-effectively remove implausible explanations from the solution space.
-
-\subsection{Adversarial Examples are Counterfactual
-Explanations}\label{adversarial-examples-are-counterfactual-explanations}
-
-This interpretation of the link between explainability through
-counterfactuals on one side and robustness to adversarial examples on
-the other is backed by empirical evidence. Sauer and Geiger
-(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrates that using
-counterfactual images during classifier training improves model
-robustness. Similarly, Abbasnejad et al.
-(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argues that
-counterfactuals represent potentially useful training data in machine
-learning, especially in supervised settings where inputs may be
-reasonably mapped to multiple outputs. They, too, demonstrate that
-augmenting the training data of image classifiers can improve
-generalization. Finally, Teney, Abbasnedjad, and Hengel
-(\citeproc{ref-teney2020learning}{2020}) proposes an approach using
-counterfactuals in training that does not rely on data augmentation:
-they argue that counterfactual pairs typically already exist in training
-datasets. Specifically, their approach relies on identifying similar
-input samples with different annotations and ensuring that the gradient
-of the classifier aligns with the vector between such pairs of
-counterfactual inputs using the cosine distance as the loss function.
-
-In the natural language processing (NLP) domain, CEs have also been used
-to improve models through data augmentation. Wu et al.
-(\citeproc{ref-wu2021polyjuice2}{2021}) proposes \emph{Polyjuice}, a
-general-purpose counterfactual generator for language models. The
-authors empirically demonstrate that the augmentation of training data
-with their method improves robustness in a number of NLP tasks.
-Balashankar et al. (\citeproc{ref-balashankar2023improving}{2023})
-similarly uses \emph{Polyjuice} to augment NLP datasets through diverse
-counterfactuals and show that classifier robustness improves by up to
-20\%. Finally, Luu and Inoue
-(\citeproc{ref-luu2023counterfactual}{2023}) introduces Counterfactual
-Adversarial Training (CAT) that also aims to improve generalization and
-robustness of language models through a three-step procedure: the
-authors identify training samples that are subject to high predictive
-uncertainty, generate CEs for them, and fine-tune the language model on
-a dataset augmented with the CEs.
-
-There have also been several attempts at formalizing the relationship
-between counterfactual explanations and adversarial examples. Pointing
-to clear similarities in how CEs and AEs are generated, Freiesleben
-(\citeproc{ref-freiesleben2022intriguing}{2022}) makes the case for
-jointly studying the opaqueness and robustness problems in
-representation learning. Formally, AEs can be seen as the subset of CEs
-for which misclassification is achieved
-(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
-Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) shows
-that CEs and AEs are equivalent under certain conditions.
-
-Two recent works are closely related to ours in that they use
-counterfactuals during training with the explicit goal of affecting
-certain properties of the post-hoc counterfactual explanations. Firstly,
-Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
-proposes a way to train models that guarantee individual recourse to
-some positive target class with high probability. Their approach builds
-on adversarial training by explicitly inducing susceptibility to
-targeted adversarial examples for the positive class. Additionally, the
-proposed method allows for imposing a set of actionability constraints
-ex-ante. For example, users can specify that certain features are
-immutable. Secondly, Guo, Nguyen, and Yadav
-(\citeproc{ref-guo2023counternet}{2023}) is the first to propose an
-end-to-end training pipeline that includes CEs as part of the training
-procedure. In particular, they propose a specific network architecture
-that includes a predictor and CE generator network, where the parameters
-of the CE generator network are learnable. Counterfactuals are generated
-during each training iteration and fed back to the predictor network. In
-contrast to Guo, Nguyen, and Yadav
-(\citeproc{ref-guo2023counternet}{2023}), we impose no restrictions on
-the ANN architecture at all.
-
-\subsection{Aligning Representations with Plausible
-Explanations}\label{aligning-representations-with-plausible-explanations}
-
-Improving the adversarial robustness of models is not the only path
-towards aligning representations with plausible explanations. In a work
-closely related to this one, Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}) shows that explainability
-can be improved through model averaging and refined model objectives.
-The authors propose a way to generate counterfactuals that are maximally
-faithful to the model in that they are consistent with what the model
-has learned about the underlying data. Formally, they rely on tools from
-energy-based modelling (\citeproc{ref-teh2003energy}{Teh et al. 2003})
-to minimize the divergence between the distribution of counterfactuals
-and the conditional posterior over inputs learned by the model. Their
-proposed counterfactual explainer, \emph{ECCCo}, yields plausible
-explanations if and only if the underlying model has learned
-representations that align with them. The authors find that both deep
-ensembles (\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan,
-Pritzel, and Blundell 2017}) and joint energy-based models (JEMs)
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
-well in this regard.
-
-Once again it helps to look at these findings through the lens of
-representation learning with high degrees of freedom. Deep ensembles are
-approximate Bayesian model averages, which are most called for when
-models are underspecified by the available data
-(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
-mitigates the aforementioned risk of relying on a single locally optimal
-representations that corresponds to semantically meaningless
-explanations for the data. Previous work of Schut et al.
-(\citeproc{ref-schut2021generating}{2021}) similarly found that
-generating plausible (``interpretable'') CEs is almost trivial for deep
-ensembles that have also undergone adversarial training. The case for
-JEMs is even clearer: they involve a hybrid objective that induces both
-high predictive performance and generative capacity
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This is
-closely related to the idea of aligning models with plausible
-explanations and has inspired our CT objective.
-
-\section{Counterfactual Training}\label{sec-method}
-
-In this section we propose our novel counterfactual training objective.
-In CT, we combine ideas from adversarial training, counterfactual
-explanations, and energy-based modelling with the explicit goal of
-aligning representations with plausible explanations that comply with
-user-defined actionability constraints.
-
-In the context of counterfactual explanations, plausibility has broadly
-been defined as the degree to which counterfactuals comply with the
-underlying data-generating process
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024};
-\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
-\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020}). Plausibility is
-a necessary but insufficient condition for using CEs to provide
-algorithmic recourse (AR) to individuals (negatively) affected by opaque
-models. To be actionable, AR recommendations must also be attainable. A
-plausible CE for a rejected 20-year-old loan applicant, for example,
-might reveal that their application would have been accepted, if only
-they were 20 years older. Ignoring all other features, this would comply
-with the definition of plausibility if 40-year-old individuals were in
-fact more credit-worthy on average than young adults. But of course this
-CE does not qualify for providing actionable recourse to the applicant
-since \emph{age} is not a (directly) mutable feature. Counterfactual
-training aims to improve model explainability by aligning models with
-counterfactuals that meet both desiderata: plausibility and
-actionability. Formally, we define explainability as follows:
-
-\begin{definition}[Model Explainability]
-\label{def-explainability}
-Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$.
-Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.
-
-We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. We define these properties as follows:
-
-\begin{enumerate}
-    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
-    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
-    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
-\end{enumerate}
-where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
-\end{definition}
-
-\noindent The characterization of faithfulness and plausibility in Def.
-\ref{def-explainability} is the same as in Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
-Intuitively, plausible counterfactuals are consistent with the data and
-faithful counterfactuals are consistent with what the model has learned
-about input data. Actionability constraints in Def.
-\ref{def-explainability} vary and depend on the context in which
-\(\mathbf{M}_\theta\) is deployed. In this work, we focus on domain and
-mutability constraints for individual features \(x_d\) for
-\(d=1,...,D\). We limit ourselves to classification tasks for reasons
-discussed in Section~\ref{sec-discussion}.
-
-\subsection{Our Proposed Objective}\label{our-proposed-objective}
-
-Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
-explanation generated through gradient descent over \(T\) iterations as
-initially proposed by Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
-let \(T\) vary and consider the counterfactual search as converged as
-soon as the predicted probability for the target class has reached a
-pre-determined threshold, \(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
-where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
-  background information on gradient-based counterfactual search and
-  convergence see supplementary appendix.}
-
-To train models with high explainability as defined in Def.
-\ref{def-explainability}, we propose to leverage counterfactuals in the
-following objective:
-
-\begin{equation}\phantomsection\label{eq-obj}{
-\begin{split}
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
-\end{split}
-}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
-that induces discriminative performance (e.g., cross-entropy). The
-second and third terms are explained in detail below. For now, they can
-be summarized as inducing explainability directly and indirectly by
-penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
-between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
-samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
-target class \(y^+\), and, (2) the adversarial loss,
-\(\text{advloss}(.)\), with respect to nascent counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
-denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
-of the energy terms involved in \(\text{div}(\cdot)\)
-(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-off
-between the components can governed through penalties
-\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
-\(\lambda_{\text{reg}}\).
-
-\subsection{Directly Inducing Explainability with Contrastive
-Divergence}\label{directly-inducing-explainability-with-contrastive-divergence}
-
-As observed by Grathwohl et al.
-(\citeproc{ref-grathwohl2020your}{2020}), any classifier can be
-re-interpreted as a joint energy-based model (JEM) that learns to
-discriminate output classes conditional on the observed (training)
-samples from \(p(\mathbf{x})\) and the generated samples from
-\(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
-perform well at both tasks by directly maximizing the joint
-log-likelihood factorized as
-\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
-The first term can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\),
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimizes the
-contrastive divergence between the observed samples from
-\(p(\mathbf{x})\) and generated samples from \(p_\theta(\mathbf{x})\).
-
-A key empirical finding in Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
-well with respect to the plausibility objective in Def.
-\ref{def-explainability}. This follows directly if we consider samples
-drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
-objective effectively minimizes the divergence between the conditional
-posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) relies on
-Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative
-prior for initialization but we depart from their methodology. Instead
-of SGLD, we propose to use counterfactual explainers to generate
-counterfactuals of observed training samples. Specifically, we have:
-
-\begin{equation}\phantomsection\label{eq-div}{
-\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
-}\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
-function defined as
-\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\),
-with \(y^+\) denoting the index of the randomly drawn target class,
-\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
-\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
-sampled factual from a non-target class generated with a gradient-based
-CE generator for up to \(T\) iterations. Mature counterfactuals are ones
-that have either reached convergence wrt. the decision threshold
-\(\tau\) or exhausted \(T\).
-
-Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
-of observed training samples (positive samples) while increasing the
-energy of counterfactuals (negative samples)
-(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the
-counterfactuals get more plausible (Def. \ref{def-explainability})
-during training, these opposing effects gradually balance each other out
-(\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
-
-The departure from SGLD of (\citeproc{ref-grathwohl2020your}{Grathwohl
-et al. 2020}) allows us to tap into the vast repertoire of explainers
-that have been proposed in the literature to meet different desiderata.
-For example, many methods facilitate the imposition of domain and
-mutability constraints. In principle, any existing approach for
-generating counterfactual explanations is viable, so long as it does not
-violate the faithfulness condition. Like JEMs
-(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), CT can be
-considered a form of contrastive representation learning.
-
-\subsection{Indirectly Inducing Explainability with Adversarial
-Robustness}\label{indirectly-inducing-explainability-with-adversarial-robustness}
-
-Based on our analysis in Section~\ref{sec-lit}, counterfactuals
-\(\mathbf{x}^\prime\) can be repurposed as additional training samples
-(\citeproc{ref-balashankar2023improving}{Balashankar et al. 2023};
-\citeproc{ref-luu2023counterfactual}{Luu and Inoue 2023}) or adversarial
-examples (\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022};
-\citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
-leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
-Equation~\ref{eq-obj}. An intuitive functional form, but likely not the
-only sensible choice, is inspired by adversarial training:
-
-\begin{equation}\phantomsection\label{eq-adv}{
-\begin{aligned}
-\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
-t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
-\end{aligned}
-}\end{equation} Under this choice, we consider nascent counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
-perturbation to any single feature is at most \(\varepsilon\). This is
-closely aligned with Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2014}) that defines an adversarial
-attack as an ``imperceptible non-random perturbation''. Thus, we choose
-to work with a different distinction between CE and AE than Freiesleben
-(\citeproc{ref-freiesleben2022intriguing}{2022}) that considers
-misclassification as the key distinguishing feature of AE. One of the
-key observations of this work is that we can leverage CEs during
-training and get adversarial examples essentially for free, which can be
-used to reap the aforementioned benefits of adversarial training.
-
-\subsection{Encoding Actionability Constraints}\label{sec-constraints}
-
-Many existing counterfactual explainers support domain and mutability
-constraints out-of-the-box. In fact, both types of constraints can be
-implemented for any counterfactual explainer that relies on gradient
-descent in the feature space for optimization
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, and Liem
-2023}). In this context, domain constraints can be imposed by simply
-projecting counterfactuals back to the specified domain, if the previous
-gradient step resulted in updated feature values that were
-out-of-domain. Mutability constraints can similarly be enforced by
-setting partial derivatives to zero to ensure that features are only
-perturbed in the allowed direction, if at all.
-
-Since such actionability constraints are binding at test time, we should
-also impose them when generating \(\mathbf{x}^\prime\) during each
-training iteration to inform model representations. Through their effect
-on \(\mathbf{x}^\prime\), both types of constraints influence model
-outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
-penalizing implausibility that arises due to mutability constraints. For
-any mutability-constrained feature \(d\) this can be achieved by
-enforcing \(\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0\) whenever
-perturbing \(\mathbf{x}^\prime[d]\) in the direction of
-\(\mathbf{x}^+[d]\) would violate mutability constraints. Specifically,
-we set \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) if:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Feature \(d\) is strictly immutable in practice.
-\item
-  We have \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but feature \(d\)
-  can only be decreased in practice.
-\item
-  We have \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but feature \(d\)
-  can only be increased in practice.
-\end{enumerate}
-
-\noindent From a Bayesian perspective, setting
-\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
-assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
-feature \(d\). Intuitively, we think of this simply in terms ignoring
-implausibility costs with respect to immutable features, which
-effectively forces the model to instead seek plausibility with respect
-to the remaining features. This in turn results in lower overall
-sensitivity to immutable features, which we demonstrate empirically for
-different classifiers in Section~\ref{sec-experiments}. Under certain
-conditions, this result holds theoretically:\footnote{For the proof, see
-  the supplementary appendix.}
-
-\begin{proposition}[Protecting Immutable Features]
-\label{prp-mtblty}
-Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
-If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
-\end{proposition}
-
-\noindent It is worth highlighting that
-\mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
-This raises a valid concern about the effect of protecting immutable
-features in the presence of proxies that remain unprotected. We address
-this in Section~\ref{sec-discussion}.
-
-\subsection{Example (Prediction of Consumer Credit
-Default)}\label{example-prediction-of-consumer-credit-default}
-
-Suppose we are interested in predicting the likelihood that loan
-applicants default on their credit. We have access to historical data on
-previous loan takers comprised of a binary outcome variable
-(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
-features: (1) the subjects' \emph{age}, which we define as immutable,
-and (2) the subjects' existing level of \emph{debt}, which we define as
-mutable.
-
-We have simulated this scenario using synthetic data with independent
-\emph{age} and \emph{debt} features, and Gaussian class-conditional
-densities in Figure~\ref{fig-poc}. The four panels show the outcomes for
-different training procedures using the same model architectures (a
-linear classifier). In panels (a) and (c) we have trained the models
-conventionally, while in panels (b) and (d) we used CT.
-
-In all cases, all counterfactuals (stars) are valid---they have crossed
-the decision boundary (green)---but their quality differs. In panel (a),
-they are not plausible: they do not comply with the distribution of the
-factuals in \(y^+\) to the point where they form a clearly
-distinguishable cluster. In panel (b), they are highly plausible,
-meeting the first objective of Def. \ref{def-explainability}. In panel
-(c), the CEs involve substantial reductions in \emph{debt} for younger
-applicants. By comparison, counterfactual paths are shorter on average
-in panel (d) where we have protected the immutable \emph{age} as
-described in Section~\ref{sec-constraints}. Due to the classifier's
-lower sensitivity to \emph{age}, recommendations with respect to
-\emph{debt} are much more homogenous and do not unfairly punish younger
-individuals. These counterfactuals are also plausible with respect to
-the mutable feature. Thus, we consider the model in panel (d) as the
-most explainable according to Def. \ref{def-explainability}.
-
-\begin{figure*}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
-
-}
-
-\caption{\label{fig-poc}Illustration of how CT improves model
-explainability: (a) conventional training, all mutable; (b) CT, all
-mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
-immutable. The linear decision boundary is shown in green along with
-training data colored according to their ground-truth label: \(y^-=1\)
-in blue and \(y^+=2\) in orange. Stars indicate counterfactuals in the
-target class.}
-
-\end{figure*}%
-
-\section{Experiments}\label{sec-experiments}
-
-In our experiments we seek to answer the following three research
-questions:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  To what extent does the counterfactual training objective as it is
-  defined in Equation~\ref{eq-obj} induce models to learn plausible
-  explanations?
-\item
-  To what extent does the CT objective produce more favorable
-  algorithmic recourse outcomes in the presence of actionability
-  constraints?
-\item
-  What are the effects of hyperparameter selection wrt. the CT
-  objective?
-\end{enumerate}
-
-\subsection{Experimental Setup}\label{experimental-setup}
-
-Our key outcome of interest is improvement in explainability (Def.
-\ref{def-explainability}). To this end, we focus primarily on the
-plausibility and cost of faithfully generated counterfactuals at test
-time. To measure the cost, we follow the standard convention of using
-distances (\(\ell_1\)-norm) between factuals and counterfactuals as a
-proxy. For plausibility, we assess how similar CEs are to the observed
-samples in the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We
-rely on the distance-based metric used in Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}),
-
-\begin{equation}\phantomsection\label{eq-impl-dist}{
-\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-}\end{equation} and introduce a novel divergence metric,
-
-\begin{equation}\phantomsection\label{eq-impl-div}{
-\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
-}\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
-counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
-squared population maximum mean discrepancy
-(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
-Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
-are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
-
-In addition to cost and plausibility, we compute other standard metrics
-to evaluate counterfactuals including validity and redundancy. Finally,
-we also assess the predictive performance of models using standard
-metrics, including robust accuracy estimated on adversarially perturbed
-data using FGSM (\citeproc{ref-goodfellow2014explaining}{Goodfellow,
-Shlens, and Szegedy 2015}).
-
-We run the experiments with three gradient-based generators:
-\emph{Generic} of Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}) as a simple baseline
-approach, \emph{REVISE} (\citeproc{ref-joshi2019realistic}{Joshi et al.
-2019}) that aims to generate plausible counterfactuals using a surrogate
-Variational Autoencoder (VAE), and \emph{ECCo}---the generator of
-Altmeyer et al. (\citeproc{ref-altmeyer2024faithful}{2024}) but without
-the conformal prediction component---as a method that directly targets
-both faithfulness and plausibility of the counterfactuals.
-
-We make use of nine classification datasets common in the CE/AR
-literature. Four of them are synthetic with two classes and different
-characteristics: linearly separable clusters (\emph{LS}), overlapping
-clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
-moons (\emph{Moon}). These datasets are generated using the library of
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, and Liem
-2023}) and we present them in the supplementary appendix. Next, we have
-four real-world binary tabular datasets from the domain of economics:
-\emph{Adult} (a.k.a. Census data) of
-(\citeproc{ref-becker1996adult2}{Becker and Kohavi 1996}), California
-housing (\emph{CH}) of (\citeproc{ref-pace1997sparse}{Pace and Barry
-1997}), Default of Credit Card Clients (\emph{Cred}) of
-(\citeproc{ref-yeh2016default}{Yeh 2016}), and Give Me Some Credit
-(\emph{GMSC}) of (\citeproc{ref-kaggle2011give2}{Kaggle 2011}). Finally,
-for the convenience of illustration, we use of the 10-class \emph{MNIST}
-vision dataset (\citeproc{ref-lecun1998mnist}{LeCun 1998}).
-
-To assess CT, we investigate the improvements in performance metrics
-when using it on top of a weak baseline (BL): a multilayer perceptron
-(\emph{MLP}). This is the best way to get a clear picture of the
-effectiveness of CT, and it is consistent with how assessment is done in
-the related literature
-(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
-2015}; \citeproc{ref-ross2021learning}{Ross, Lakkaraju, and Bastani
-2024}; \citeproc{ref-teney2020learning}{Teney, Abbasnedjad, and Hengel
-2020}).
-
-\subsection{Experimental Results}\label{experimental-results}
-
-\subsubsection{Plausibility}\label{sec-plaus}
-
-Table~\ref{tbl-main} presents our main empirical findings. For all
-datasets except \emph{OL} and across all test settings, the average
-distance of CEs from observed samples in the target class is reduced,
-indicating improved plausibility. The magnitude of improvements varies.
-For the simple synthetic datasets, distance reductions range from around
-20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
-real-world tabular datasets, improvements tend to be smaller but still
-substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
-7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For the vision
-dataset (\emph{MNIST}), distances are reduced by up to 9\%. The results
-for our proposed divergence metric are qualitatively similar, but
-generally even more pronounced: for the \emph{Circ} dataset,
-implausibility is reduced by almost 94\% to virtually zero as we
-verified by the absolute outcome. Improvements for other datasets range
-from 28\% (\emph{Moon}) up to 78\% (\emph{GMSC}). For \emph{OL} the
-reduction is negative, consistent with the distance-based metric.
-\emph{MNIST} is the only dataset for which the distance and divergence
-metrics disagree. Upon visual inspection of the image counterfactuals we
-find that CT clearly improves plausibility (see supplementary appendix
-for images).
-
-\subsubsection{Predictive Performance}\label{sec-pred}
-
-Test accuracy for CT is virtually identical to the baseline for
-\emph{Adult}, \emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and
-even slightly improved for \emph{Cred}. Exceptions to this general
-pattern are \emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we
-observe a reduction in test accuracy of 2, 5, and 15 percentage points
-respectively. When looking at robust test accuracies (Acc.\(^*\)) for
-these datasets in particular, we find that CT strongly outperforms the
-baseline. In fact, we find that CT improves adversarial robustness on
-all datasets.
-
-\begin{table*}
-
-\caption{\label{tbl-main}Key performance metrics across all datasets
-(column 1). \textbf{Plausibility}: Columns 2-6 show the percentage
-reduction in implausibility (\(\text{IP}\)) for varying degrees of the
-energy penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test
-time; column 7 shows the reduction in \(\text{IP}^*\) (\(\text{MMD}\)),
-aggregated across all test specifications. \textbf{Accuracy} (columns
-8-11): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
-and the baseline (BL). \textbf{Actionability} (column 12): average
-reduction in costs when imposing mutability constraints reported for the
-four datasets for which we could identify meaningful features to
-protect.}
-
-\centering{
-
-\input{tables/main.tex}
-
-}
-
-\end{table*}%
-
-\subsubsection{Actionability}\label{sec-act}
-
-In Section~\ref{sec-method}, we show that our proposed way for encoding
-mutability constraints leads to lower classifier sensitivity wrt.
-immutable features for linear models, tilting the decision boundary in
-favour of mutable features instead. For binding constraints at test
-time, this leads to shorter counterfactual paths and hence smaller
-average costs (\(\ell_1\)-norm) to individuals. To extend this to the
-non-linear case, we test the effect of imposing mutability constraints
-empirically for our synthetic data using the same evaluation scheme as
-above. The final row in Table~\ref{tbl-main} reports the average
-reduction in costs for CT compared to the ``vanilla'' baseline, when
-imposing that either the first or the second feature is immutable. In
-all cases, costs are reduced substantially, indicating that classifiers
-trained with CT are indeed more sensitive to mutable features.
-
-\subsubsection{Impact of hyperparameter
-settings.}\label{sec-hyperparameters}
-
-We test the impact of three types of hyperparameters; our complete
-results are in the supplementary appendix.\\
-\indent We note that CT is highly sensitive to the choice of a CE
-generator and its hyperparameters but (a) there are manageable patterns
-and (b) we can typically identify settings that improve either
-plausibility or cost, and commonly both of them at the same time. For
-example, \emph{REVISE} tends to perform the worst, most likely because
-it uses a surrogate VAE to generate counterfactuals which impedes
-faithfulness (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al.
-2024}). Increasing \(T\), the maximum number of steps, generally yields
-better outcomes because more CEs can mature in each training epoch. The
-impact of \(\tau\), the required decision threshold is more difficult to
-predict. On ``harder'' datasets it may be difficult to satisfy high
-\(\tau\) for any given sample (i.e., also factuals) and so increasing
-this threshold does not seem to correlate with better outcomes. In fact,
-the choice of \(\tau=0.5\) generally leads to optimal results because it
-is associated with high proportions of mature counterfactuals.\\
-\indent The strength of the energy regularization,
-\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
-performance in terms of decreased plausibility and increased costs if
-insufficiently high. The sensitivity with respect to
-\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
-evident. While high values of \(\lambda_{\text{reg}}\) may increase the
-variability in outcomes when combined with high values of
-\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
-very pronounced.\\
-\indent The effectiveness and stability of CT is positively associated
-with the number of counterfactuals generated during each training epoch.
-We also confirm that a higher number of training epochs is beneficial.
-Interestingly, we observed desired improvements when CT was combined
-with conventional training and applied only for the final 50\% of epochs
-of the complete training process. Put differently, CT can improve the
-explainability of models in a fine-tuning manner.
-
-\section{Conclusions}\label{sec-discussion}
-
-As our results indicate, counterfactual training produces models that
-are more explainable. Nonetheless, it brings about three important
-limitations.\\
-\indent \emph{CT increases the training time of models.} CT can be more
-time-consuming than conventional training regimes. While higher numbers
-of CEs per iteration positively impact the quality of solutions, they
-also increase the amount of computations. Relatively small grids with
-270 settings can take almost four hours for more demanding datasets on a
-high-performance computing cluster with 34 2GB CPUs.\footnote{See
-  supplementary appendix for computational details.} Three factors
-attenuate this effect: (1) CT amortizes the cost of CEs for the training
-samples; (2) it can retain its value when used as a ``fine-tuning''
-technique for conventionally-trained models; and (3) it yields itself to
-parallel execution, which we have leveraged for our own experiments.\\
-\indent \emph{Immutable features may have proxies.} We propose an
-approach to protect immutable features and thus increase the
-actionability of the generated CEs. However, it requires that model
-owners define the mutability constraints for (all) features considered
-by the model. Even if all immutable features are protected, there may
-exist proxies that are mutable (and hence should not be protected) but
-preserve enough information about the principals to hinder the
-protections. Delineating actionability is a major open challenge in the
-AR literature (see, e.g.,
-(\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
-and Alfano 2020})) impacting the capacity of CT to fulfill its intended
-goal.\\
-\indent \emph{Interventions on features may impact fairness.} We provide
-a tool that allows practitioners to modify the sensitivity of a model
-with respect to certain features, which may have implication for the
-fair and equitable treatment of decision subjects. Model owners could
-misuse this solution by enforcing explanations based on features that
-are more difficult to modify by some (group of) individuals. For
-example, consider the \emph{Adult} dataset used in our experiments,
-where \emph{workclass} or \emph{education} may be more difficult to
-change for underpriviledged groups. When applied irresponsibly, CT could
-result in an unfairly assigned burden of recourse
-(\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
-threatening the equality of opportunity in the system
-(\citeproc{ref-bell2024fairness}{Bell et al. 2024}). Nonetheless, these
-phenomena are not specific to CT.\\
-\hfill \break \indent We also highlight several important directions for
-future research. Firstly, it is an interesting challenge to extend CT
-beyond classification settings. Our formulation relies on the
-distinction between non-target class(es) \(y^{-}\) and target class(es)
-\(y^{+}\) to generate counterfactuals through Equation~\ref{eq-obj}.
-While \(y^{-}\) and \(y^{+}\) can be arbitrarily defined, CT requires
-the output space \(\mathcal{Y}\) to be discrete. Thus, it does not apply
-to ML tasks where the change in outcome cannot be readily quantified.
-Focus on classification models is a common restriction in research on
-CEs and AR. Other settings have attracted some interest (e.g.,
-regression in (\citeproc{ref-spooner2021counterfactual}{Spooner et al.
-2021})), but there is little consensus how to robustly extend the notion
-of CEs.\\
-\indent Secondly, our approach is susceptible to training instabilities.
-This problem has been recognized for JEMs
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
-though we depart from the SGLD-based sampling, we still encounter
-considerable variability in the outcomes. CT is exposed to two potential
-sources of instabilities: (1) the energy-based contrastive divergence
-term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
-explainers. We find several promising ways to mitigate this problem:
-regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
-many counterfactuals during each epoch, and including only mature
-counterfactuals for contrastive divergence.\\
-\indent Finally, we believe that it is possible to substantially improve
-hyperparameter selection procedures. Our method benefits from the tuning
-of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
-In this work, we have relied exclusively on grid search for this task.
-Future work on CT could benefit from investigating more sophisticated
-approaches. Notably, CT is iterative which makes methods such as
-Bayesian or gradient-based optimization applicable (see, e.g.,
-(\citeproc{ref-bischl2023hyperparameter}{Bischl et al. 2023})).
-\hfill \break
-
-To conclude, state-of-the-art machine learning models are prone to
-learning complex representations that cannot be interpreted by humans
-and existing post-hoc explainability approaches cannot guarantee that
-the explanations agree with the model's learned representation of data.
-As a step towards addressing this challenge, we introduced
-counterfactual training, a novel training regime that incentivizes
-highly-explainable models. Our approach leads to explanations that are
-both plausible---compliant with the underlying data-generating
-process---and actionable---compliant with user-specified mutability
-constraints---and thus meaningful to their recipients. Through extensive
-experiments we demonstrate that CT satisfies its objective while
-promoting robustness and preserving the predictive performance of the
-models. It can also be used to fine-tune conventionally-trained models
-and achieve similar gains. Finally, this work showcases that it is
-practical to improve models \emph{and} their explanations at the same
-time.
-
-\section*{References}\label{references}
-\addcontentsline{toc}{section}{References}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-abbasnejad2020counterfactual}
-Abbasnejad, Ehsan, Damien Teney, Amin Parvaneh, Javen Shi, and Anton van
-den Hengel. 2020. {``{Counterfactual Vision and Language Learning}.''}
-In \emph{2020 IEEE/CVF Conference on Computer Vision and Pattern
-Recognition (CVPR)}, 10041--51.
-\url{https://doi.org/10.1109/CVPR42600.2020.01006}.
-
-\bibitem[\citeproctext]{ref-altmeyer2023explaining}
-Altmeyer, Patrick, Arie van Deursen, and Cynthia C. S. Liem. 2023.
-{``{Explaining Black-Box Models through Counterfactuals}.''} In
-\emph{Proceedings of the JuliaCon Conferences}, 1:130.
-
-\bibitem[\citeproctext]{ref-altmeyer2024faithful}
-Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia C.
-S. Liem. 2024. {``{Faithful Model Explanations through
-Energy-Constrained Conformal Counterfactuals}.''} In \emph{Proceedings
-of the Thirty-Eighth AAAI Conference on Artificial Intelligence},
-38:10829--37. 10. \url{https://doi.org/10.1609/aaai.v38i10.28956}.
-
-\bibitem[\citeproctext]{ref-augustin2020adversarial}
-Augustin, Maximilian, Alexander Meinke, and Matthias Hein. 2020.
-{``{Adversarial Robustness on In- and Out-Distribution Improves
-Explainability}.''} In \emph{Computer Vision -- ECCV 2020}, edited by
-Andrea Vedaldi, Horst Bischof, Thomas Brox, and Jan-Michael Frahm,
-228--45. Cham: Springer.
-
-\bibitem[\citeproctext]{ref-balashankar2023improving}
-Balashankar, Ananth, Xuezhi Wang, Yao Qin, Ben Packer, Nithum Thain, Ed
-Chi, Jilin Chen, and Alex Beutel. 2023. {``{Improving Classifier
-Robustness through Active Generative Counterfactual Data
-Augmentation}.''} In \emph{Findings of the Association for Computational
-Linguistics: EMNLP 2023}, 127--39. ACL.
-\url{https://doi.org/10.18653/v1/2023.findings-emnlp.10}.
-
-\bibitem[\citeproctext]{ref-becker1996adult2}
-Becker, Barry, and Ronny Kohavi. 1996. {``{Adult}.''} UCI Machine
-Learning Repository.
-
-\bibitem[\citeproctext]{ref-bell2024fairness}
-Bell, Andrew, Joao Fonseca, Carlo Abrate, Francesco Bonchi, and Julia
-Stoyanovich. 2024. {``{Fairness in Algorithmic Recourse Through the Lens
-of Substantive Equality of Opportunity}.''}
-\url{https://arxiv.org/abs/2401.16088}.
-
-\bibitem[\citeproctext]{ref-bezanson2017julia}
-Bezanson, Jeff, Alan Edelman, Stefan Karpinski, and Viral B Shah. 2017.
-{``Julia: A Fresh Approach to Numerical Computing.''} \emph{SIAM Review}
-59 (1): 65--98. \url{https://doi.org/10.1137/141000671}.
-
-\bibitem[\citeproctext]{ref-bischl2023hyperparameter}
-Bischl, Bernd, Martin Binder, Michel Lang, Tobias Pielok, Jakob Richter,
-Stefan Coors, Janek Thomas, et al. 2023. {``{Hyperparameter
-optimization: Foundations, algorithms, best practices, and open
-challenges}.''} \emph{WIREs Data Mining and Knowledge Discovery} 13 (2):
-e1484. https://doi.org/\url{https://doi.org/10.1002/widm.1484}.
-
-\bibitem[\citeproctext]{ref-milan2023dataframes}
-Bouchet-Valat, Milan, and Bogumił Kamiński. 2023. {``DataFrames.jl:
-Flexible and Fast Tabular Data in Julia.''} \emph{Journal of Statistical
-Software} 107 (4): 1--32. \url{https://doi.org/10.18637/jss.v107.i04}.
-
-\bibitem[\citeproctext]{ref-byrne2021mpi}
-Byrne, Simon, Lucas C. Wilcox, and Valentin Churavy. 2021. {``MPI.jl:
-Julia Bindings for the Message Passing Interface.''} \emph{Proceedings
-of the JuliaCon Conferences} 1 (1): 68.
-\url{https://doi.org/10.21105/jcon.00068}.
-
-\bibitem[\citeproctext]{ref-chagas2024pretty}
-Chagas, Ronan Arraes Jardim, Ben Baumgold, Glen Hertz, Hendrik Ranocha,
-Mark Wells, Nathan Boyer, Nicholas Ritchie, et al. 2024.
-{``Ronisbr/PrettyTables.jl: V2.4.0.''} Zenodo.
-\url{https://doi.org/10.5281/zenodo.13835553}.
-
-\bibitem[\citeproctext]{ref-PlotsJL}
-Christ, Simon, Daniel Schwabeneder, Christopher Rackauckas, Michael
-Krabbe Borregaard, and Thomas Breloff. 2023. {``Plots.jl -- a User
-Extendable Plotting API for the Julia Programming Language.''}
-https://doi.org/\url{https://doi.org/10.5334/jors.431}.
-
-\bibitem[\citeproctext]{ref-danisch2021makie}
-Danisch, Simon, and Julius Krumbiegel. 2021. {``{Makie.jl}: Flexible
-High-Performance Data Visualization for {Julia}.''} \emph{Journal of
-Open Source Software} 6 (65): 3349.
-\url{https://doi.org/10.21105/joss.03349}.
-
-\bibitem[\citeproctext]{ref-du2019implicit}
-Du, Yilun, and Igor Mordatch. 2020. {``{Implicit Generation and
-Generalization in Energy-Based Models}.''}
-\url{https://arxiv.org/abs/1903.08689}.
-
-\bibitem[\citeproctext]{ref-frankle2018lottery}
-Frankle, Jonathan, and Michael Carbin. 2019. {``{The Lottery Ticket
-Hypothesis: Finding Sparse, Trainable Neural Networks}.''} In
-\emph{International Conference on Learning Representations}.
-
-\bibitem[\citeproctext]{ref-freiesleben2022intriguing}
-Freiesleben, Timo. 2022. {``{The Intriguing Relation Between
-Counterfactual Explanations and Adversarial Examples}.''} \emph{Minds
-and Machines} 32 (1): 77--109.
-
-\bibitem[\citeproctext]{ref-goodfellow2016deep}
-Goodfellow, Ian, Yoshua Bengio, and Aaron Courville. 2016. \emph{Deep
-{Learning}}. {MIT Press}.
-
-\bibitem[\citeproctext]{ref-goodfellow2014explaining}
-Goodfellow, Ian, Jonathon Shlens, and Christian Szegedy. 2015.
-{``{Explaining and Harnessing Adversarial Examples}.''}
-\url{https://arxiv.org/abs/1412.6572}.
-
-\bibitem[\citeproctext]{ref-grathwohl2020your}
-Grathwohl, Will, Kuan-Chieh Wang, Joern-Henrik Jacobsen, David Duvenaud,
-Mohammad Norouzi, and Kevin Swersky. 2020. {``Your Classifier Is
-Secretly an Energy Based Model and You Should Treat It Like One.''} In
-\emph{International Conference on Learning Representations}.
-
-\bibitem[\citeproctext]{ref-gretton2012kernel}
-Gretton, Arthur, Karsten M Borgwardt, Malte J Rasch, Bernhard Schölkopf,
-and Alexander Smola. 2012. {``A Kernel Two-Sample Test.''} \emph{The
-Journal of Machine Learning Research} 13 (1): 723--73.
-
-\bibitem[\citeproctext]{ref-guidotti2022counterfactual}
-Guidotti, Riccardo. 2022. {``{Counterfactual Explanations and How to
-Find Them: Literature Review and Benchmarking}.''} \emph{Data Mining and
-Knowledge Discovery} 38 (5): 2770--2824.
-\url{https://doi.org/10.1007/s10618-022-00831-6}.
-
-\bibitem[\citeproctext]{ref-guo2023counternet}
-Guo, Hangzhi, Thanh H. Nguyen, and Amulya Yadav. 2023. {``{CounterNet:
-End-to-End Training of Prediction Aware Counterfactual Explanations}.''}
-In \emph{Proceedings of the 29th ACM SIGKDD Conference on Knowledge
-Discovery and Data Mining}, 577-\/-589. KDD '23. New York, NY, USA:
-Association for Computing Machinery.
-\url{https://doi.org/10.1145/3580305.3599290}.
-
-\bibitem[\citeproctext]{ref-hastie2009elements}
-Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
-Elements of Statistical Learning}. Springer New York.
-\url{https://doi.org/10.1007/978-0-387-84858-7}.
-
-\bibitem[\citeproctext]{ref-innes2018fashionable}
-Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco
-Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and
-Viral Shah. 2018. {``Fashionable Modelling with Flux.''}
-\url{https://arxiv.org/abs/1811.01457}.
-
-\bibitem[\citeproctext]{ref-innes2018flux}
-Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
-\emph{Journal of Open Source Software} 3 (25): 602.
-\url{https://doi.org/10.21105/joss.00602}.
-
-\bibitem[\citeproctext]{ref-joshi2019realistic}
-Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
-Joydeep Ghosh. 2019. {``{Towards Realistic Individual Recourse and
-Actionable Explanations in Black-Box Decision Making Systems}.''}
-\url{https://arxiv.org/abs/1907.09615}.
-
-\bibitem[\citeproctext]{ref-kaggle2011give2}
-Kaggle. 2011. {``Give Me Some Credit, {Improve} on the State of the Art
-in Credit Scoring by Predicting the Probability That Somebody Will
-Experience Financial Distress in the Next Two Years.''}
-https://www.kaggle.com/c/GiveMeSomeCredit; {Kaggle}.
-
-\bibitem[\citeproctext]{ref-kolter2023keynote}
-Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
-IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)}. Los
-Alamitos, CA, USA: IEEE Computer Society.
-\url{https://doi.org/10.1109/SaTML54575.2023.00009}.
-
-\bibitem[\citeproctext]{ref-lakshminarayanan2016simple}
-Lakshminarayanan, Balaji, Alexander Pritzel, and Charles Blundell. 2017.
-{``Simple and Scalable Predictive Uncertainty Estimation Using Deep
-Ensembles.''} In \emph{Proceedings of the 31st International Conference
-on Neural Information Processing Systems}, 6405--16. NIPS'17. Red Hook,
-NY, USA: Curran Associates Inc.
-
-\bibitem[\citeproctext]{ref-lecun1998mnist}
-LeCun, Yann. 1998. {``{The MNIST database of handwritten digits}.''}
-http://yann.lecun.com/exdb/mnist/.
-
-\bibitem[\citeproctext]{ref-lippe2024uvadlc}
-Lippe, Phillip. 2024. {``{UvA Deep Learning Tutorials}.''}
-\url{https://uvadlc-notebooks.readthedocs.io/en/latest/}.
-
-\bibitem[\citeproctext]{ref-luu2023counterfactual}
-Luu, Hoai Linh, and Naoya Inoue. 2023. {``{Counterfactual Adversarial
-Training for Improving Robustness of Pre-trained Language Models}.''} In
-\emph{Proceedings of the 37th Pacific Asia Conference on Language,
-Information and Computation}, 881--88. ACL.
-\url{https://aclanthology.org/2023.paclic-1.88/}.
-
-\bibitem[\citeproctext]{ref-murphy2022probabilistic}
-Murphy, Kevin P. 2022. \emph{Probabilistic {Machine Learning}: {An}
-Introduction}. {MIT Press}.
-
-\bibitem[\citeproctext]{ref-oneil2016weapons}
-O'Neil, Cathy. 2016. \emph{Weapons of Math Destruction: {How} Big Data
-Increases Inequality and Threatens Democracy}. {Crown}.
-
-\bibitem[\citeproctext]{ref-pace1997sparse}
-Pace, R Kelley, and Ronald Barry. 1997. {``Sparse Spatial
-Autoregressions.''} \emph{Statistics \& Probability Letters} 33 (3):
-291--97. \url{https://doi.org/10.1016/s0167-7152(96)00140-x}.
-
-\bibitem[\citeproctext]{ref-pawelczyk2022exploring}
-Pawelczyk, Martin, Chirag Agarwal, Shalmali Joshi, Sohini Upadhyay, and
-Himabindu Lakkaraju. 2022. {``Exploring Counterfactual Explanations
-Through the Lens of Adversarial Examples: A Theoretical and Empirical
-Analysis.''} In \emph{Proceedings of the 25th International Conference
-on Artificial Intelligence and Statistics}, edited by Gustau
-Camps-Valls, Francisco J. R. Ruiz, and Isabel Valera, 151:4574--94.
-Proceedings of Machine Learning Research. PMLR.
-\url{https://proceedings.mlr.press/v151/pawelczyk22a.html}.
-
-\bibitem[\citeproctext]{ref-poyiadzi2020face}
-Poyiadzi, Rafael, Kacper Sokol, Raul Santos-Rodriguez, Tijl De Bie, and
-Peter Flach. 2020. {``{FACE}: {Feasible} and Actionable Counterfactual
-Explanations.''} In \emph{Proceedings of the {AAAI}/{ACM Conference} on
-{AI}, {Ethics}, and {Society}}, 344--50.
-
-\bibitem[\citeproctext]{ref-ross2021learning}
-Ross, Alexis, Himabindu Lakkaraju, and Osbert Bastani. 2024.
-{``{Learning Models for Actionable Recourse}.''} In \emph{Proceedings of
-the 35th International Conference on Neural Information Processing
-Systems}. NIPS '21. Red Hook, NY, USA: Curran Associates Inc.
-
-\bibitem[\citeproctext]{ref-sauer2021counterfactual}
-Sauer, Axel, and Andreas Geiger. 2021. {``{Counterfactual Generative
-Networks}.''} \url{https://arxiv.org/abs/2101.06046}.
-
-\bibitem[\citeproctext]{ref-schut2021generating}
-Schut, Lisa, Oscar Key, Rory McGrath, Luca Costabello, Bogdan Sacaleanu,
-Yarin Gal, et al. 2021. {``Generating {Interpretable Counterfactual
-Explanations By Implicit Minimisation} of {Epistemic} and {Aleatoric
-Uncertainties}.''} In \emph{International {Conference} on {Artificial
-Intelligence} and {Statistics}}, 1756--64. {PMLR}.
-
-\bibitem[\citeproctext]{ref-sharma2020certifai}
-Sharma, Shubham, Jette Henderson, and Joydeep Ghosh. 2020. {``{CERTIFAI:
-A Common Framework to Provide Explanations and Analyse the Fairness and
-Robustness of Black-box Models}.''} In \emph{Proceedings of the AAAI/ACM
-Conference on AI, Ethics, and Society}, 166--72. AIES '20. New York, NY,
-USA: Association for Computing Machinery.
-\url{https://doi.org/10.1145/3375627.3375812}.
-
-\bibitem[\citeproctext]{ref-spooner2021counterfactual}
-Spooner, Thomas, Danial Dervovic, Jason Long, Jon Shepard, Jiahao Chen,
-and Daniele Magazzeni. 2021. {``{Counterfactual Explanations for
-Arbitrary Regression Models}.''} \url{https://arxiv.org/abs/2106.15212}.
-
-\bibitem[\citeproctext]{ref-szegedy2013intriguing}
-Szegedy, Christian, Wojciech Zaremba, Ilya Sutskever, Joan Bruna,
-Dumitru Erhan, Ian Goodfellow, and Rob Fergus. 2014. {``Intriguing
-Properties of Neural Networks.''} \url{https://arxiv.org/abs/1312.6199}.
-
-\bibitem[\citeproctext]{ref-teh2003energy}
-Teh, Yee Whye, Max Welling, Simon Osindero, and Geoffrey E. Hinton.
-2003. {``Energy-Based Models for Sparse Overcomplete Representations.''}
-\emph{J. Mach. Learn. Res.} 4 (null): 1235--60.
-
-\bibitem[\citeproctext]{ref-teney2020learning}
-Teney, Damien, Ehsan Abbasnedjad, and Anton van den Hengel. 2020.
-{``Learning What Makes a Difference from Counterfactual Examples and
-Gradient Supervision.''} In \emph{Computer Vision - ECCV 2020}, 580--99.
-Berlin, Heidelberg: Springer-Verlag.
-\url{https://doi.org/10.1007/978-3-030-58607-2_34}.
-
-\bibitem[\citeproctext]{ref-venkatasubramanian2020philosophical}
-Venkatasubramanian, Suresh, and Mark Alfano. 2020. {``The Philosophical
-Basis of Algorithmic Recourse.''} In \emph{Proceedings of the 2020
-Conference on Fairness, Accountability, and Transparency}, 284--93. FAT*
-'20. New York, NY, USA: Association for Computing Machinery.
-\url{https://doi.org/10.1145/3351095.3372876}.
-
-\bibitem[\citeproctext]{ref-wachter2017counterfactual}
-Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
-{``Counterfactual Explanations Without Opening the Black Box:
-{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
-841. \url{https://doi.org/10.2139/ssrn.3063289}.
-
-\bibitem[\citeproctext]{ref-wilson2020case}
-Wilson, Andrew Gordon. 2020. {``{The Case for Bayesian Deep
-Learning}.''} \url{https://arxiv.org/abs/2001.10995}.
-
-\bibitem[\citeproctext]{ref-wu2021polyjuice2}
-Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
-2021. {``Polyjuice: Generating Counterfactuals for Explaining,
-Evaluating, and Improving Models.''} In \emph{Proceedings of the 59th
-Annual Meeting of the Association for Computational Linguistics and the
-11th International Joint Conference on Natural Language Processing
-(Volume 1: Long Papers)}, 6707--23. ACL.
-\url{https://doi.org/10.18653/v1/2021.acl-long.523}.
-
-\bibitem[\citeproctext]{ref-yeh2016default}
-Yeh, I-Cheng. 2016. {``{Default of Credit Card Clients}.''} UCI Machine
-Learning Repository.
-
-\bibitem[\citeproctext]{ref-zhang2021understanding}
-Zhang, Chiyuan, Samy Bengio, Moritz Hardt, Benjamin Recht, and Oriol
-Vinyals. 2021. {``Understanding Deep Learning (Still) Requires
-Rethinking Generalization.''} \emph{Commun. ACM} 64 (3): 107--15.
-\url{https://doi.org/10.1145/3446776}.
-
-\end{CSLReferences}
-
-\newpage{}
-
 \begin{appendices}
-
-\FloatBarrier
 
@@
 
-In Section~\ref{sec-constraints} we explain that to avoid penalizing
+In \textbf{?@sec-constraints} we explain that to avoid penalizing
 implausibility that arises due to mutability constraints, we impose a
 point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
-argue in Section~\ref{sec-constraints} that this approach induces models
-to be less sensitive to immutable features and demonstrate this
-empirically in Section~\ref{sec-experiments}. Below we derive the
-analytical results in Prp.\textasciitilde{}\ref{prp-mtblty}.
+argue in \textbf{?@sec-constraints} that this approach induces models to
+be less sensitive to immutable features and demonstrate this empirically
+in \textbf{?@sec-experiments}. Below we derive the analytical results in
+Prp.\textasciitilde{}\ref{prp-mtblty}.
 
@@
 target class. Then the partial derivative of the contrastive divergence
-penalty Equation~\ref{eq-div} with respect to coefficient
-\(\theta_{y^+,d}\) is equal to
+penalty \textbf{?@eq-div} with respect to coefficient \(\theta_{y^+,d}\)
+is equal to
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-default}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-train-default}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 
-As discussed Section~\ref{sec-experiments}, CT is sensitive to certain
+As discussed \textbf{?@sec-experiments}, CT is sensitive to certain
 hyperparameter choices. We study the effect of many hyperparameters
@@
 \caption{\label{tbl-final-params}Final hyperparameters used for the main
-results presented in Section~\ref{sec-experiments}. Any hyperparameter
-not shown here is set to its default value
-(Note~\ref{nte-train-default}).}
-
-\centering{
-
-\begin{table*}
+results presented in \textbf{?@sec-experiments}. Any hyperparameter not
+shown here is set to its default value (Note~\ref{nte-train-default}).}
+
+\centering{
+
 \begin{tabular}{cccccccc}
@@
 Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
+  LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
+  Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
+  Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
+  OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\midrule
   Adult & 26049 & 5010 & 1000 & none & 0.75 & 5000 & 0.25 \\
   CH & 16504 & 3101 & 1000 & none & 0.5 & 5000 & 0.25 \\
-    Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
   Cred & 10617 & 1923 & 1000 & none & 0.5 & 5000 & 0.25 \\
   GMSC & 13371 & 2474 & 1000 & none & 0.5 & 5000 & 0.5 \\
-    LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
   MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\
-    Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
-    OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\bottomrule
 \end{tabular}
-\end{table*}
+
+}
+
+\end{table}%
+
+\subsection{Final Results}\label{sec-app-final-results}
+
+Plus/minus two standard deviations of bootstrap estimates.
+
+\subsubsection{Robust Performance Plots}\label{robust-performance-plots}
+
+\subsection{Confidence Intervals}\label{confidence-intervals}
+
+\begin{table}
+
+\caption{\label{tbl-ci}Confidence intervals}
+
+\centering{
+
+\begin{tabular}{lccccc}
+  \toprule
+  \textbf{Variable} & \textbf{Data} & \textbf{Full} & \textbf{Vanilla} & \textbf{LB} & \textbf{UB} \\\midrule
+  Cost $(-\%)$ & LS & 3.82 & 4.44 & -0.7 & -0.56 \\
+  $ \text{IP}^* $ $(-\%)$ & LS & 0.1 & 0.23 & -0.14 & -0.12 \\
+  $ \text{IP} $ $(-\%)$ & LS & 2.41 & 3.4 & -1.04 & -0.94 \\
+  Cost $(-\%)$ & Circ & 0.67 & 1.23 & -0.58 & -0.53 \\\midrule
+  $ \text{IP}^* $ $(-\%)$ & Circ & 0.0 & 0.0 & -0.01 & -0.0 \\
+  $ \text{IP} $ $(-\%)$ & Circ & 1.03 & 2.36 & -1.37 & -1.29 \\
+  Cost $(-\%)$ & Moon & 1.55 & 1.6 & -0.08 & -0.01 \\
+  $ \text{IP}^* $ $(-\%)$ & Moon & 0.02 & 0.02 & -0.01 & -0.0 \\
+  $ \text{IP} $ $(-\%)$ & Moon & 1.36 & 1.71 & -0.38 & -0.32 \\
+  Cost $(-\%)$ & OL & 1.62 & 2.63 & -1.15 & -0.81 \\\midrule
+  $ \text{IP}^* $ $(-\%)$ & OL & 0.12 & 0.09 & -0.01 & 0.05 \\
+  $ \text{IP} $ $(-\%)$ & OL & 4.49 & 4.44 & -0.03 & 0.13 \\
+  Cost $(-\%)$ & Adult & 2.26 & 2.2 & -0.22 & 0.28 \\
+  $ \text{IP}^* $ $(-\%)$ & Adult & 0.07 & 0.11 & -0.06 & -0.01 \\
+  $ \text{IP} $ $(-\%)$ & Adult & 15.03 & 15.15 & -0.68 & 0.26 \\
+  Cost $(-\%)$ & CH & 1.46 & 2.46 & -1.1 & -0.89 \\
+  $ \text{IP}^* $ $(-\%)$ & CH & 0.02 & 0.06 & -0.06 & -0.04 \\
+  $ \text{IP} $ $(-\%)$ & CH & 6.61 & 7.52 & -1.17 & -0.63 \\
+  Cost $(-\%)$ & Cred & 2.68 & 2.29 & 0.16 & 0.63 \\
+  $ \text{IP}^* $ $(-\%)$ & Cred & 0.03 & 0.06 & -0.05 & -0.01 \\
+  $ \text{IP} $ $(-\%)$ & Cred & 19.31 & 22.03 & -3.69 & -1.74 \\
+  Cost $(-\%)$ & GMSC & 1.14 & 3.05 & -2.45 & -1.77 \\
+  $ \text{IP}^* $ $(-\%)$ & GMSC & 0.02 & 0.07 & -0.06 & -0.04 \\
+  $ \text{IP} $ $(-\%)$ & GMSC & 6.19 & 8.09 & -2.4 & -1.49 \\
+  Cost $(-\%)$ & MNIST & 77.04 & 68.67 & -3.47 & 18.34 \\
+  $ \text{IP}^* $ $(-\%)$ & MNIST & 0.04 & 0.04 & -0.1 & 0.09 \\
+  $ \text{IP} $ $(-\%)$ & MNIST & 258.83 & 278.54 & -30.49 & -7.64 \\
+\end{tabular}
 
@@
 architecture and generator as in Figure~\ref{fig-mnist}, but the model
-was trained with CT.}
+was trained conventionally.}
 
@@
   Investigates the effect of changing the penalty strengths in out
-  proposed objective (Equation~\ref{eq-obj}).
+  proposed objective (\textbf{?@eq-obj}).
 \item
@@
 (im)plausibility and cost of counterfactuals at test time. Since the
-computation of our proposed divergence metric
-(Equation~\ref{eq-impl-div}) is memory-intensive, we rely on the
-distance-based metric for the grid searches. For the counterfactual
-evaluation, we draw factual samples from the training data for the grid
-searches to avoid data leakage with respect to our final results
-reported in the body of the paper. Specifically, we want to avoid
-choosing our default hyperparameters based on results on the test data.
-Since we are optimizing for explainability, not predictive performance,
-we still present test accuracy and F1-scores.
+computation of our proposed divergence metric (\textbf{?@eq-impl-div})
+is memory-intensive, we rely on the distance-based metric for the grid
+searches. For the counterfactual evaluation, we draw factual samples
+from the training data for the grid searches to avoid data leakage with
+respect to our final results reported in the body of the paper.
+Specifically, we want to avoid choosing our default hyperparameters
+based on results on the test data. Since we are optimizing for
+explainability, not predictive performance, we still present test
+accuracy and F1-scores.
 
@@
 
-Overall, we find that counterfactual training (CT) achieves it key
-objectives consistently across all hyperparameter settings and also
-broadly across datasets: plausibility is improved by up to
-\textasciitilde60 percent (\%) for the \emph{Circles} data (e.g.
-Figure~\ref{fig-grid-gen_params-plaus-circles}), \textasciitilde25-30\%
-for the \emph{Moons} data (e.g.
-Figure~\ref{fig-grid-gen_params-plaus-moons}) and \textasciitilde10-20\%
-for the \emph{Linearly Separable} data (e.g.
+Overall, we find that counterfactual training achieves it key objectives
+consistently across all hyperparameter settings and also broadly across
+datasets: plausibility is improved by up to 60 percent (\%) for the
+\emph{Circles} data (e.g.
+Figure~\ref{fig-grid-gen_params-plaus-circles}), 25-30\% for the
+\emph{Moons} data (e.g. Figure~\ref{fig-grid-gen_params-plaus-moons})
+and 10-20\% for the \emph{Linearly Separable} data (e.g.
 Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
 average costs of faithful counterfactuals are reduced in many cases by
-around \textasciitilde20-25\% for \emph{Circles} (e.g.
-Figure~\ref{fig-grid-gen_params-cost-circles}) and up to
-\textasciitilde50\% for \emph{Moons} (e.g.
-Figure~\ref{fig-grid-gen_params-cost-moons}). For the \emph{Linearly
-Separable} data, costs are generally increased although typically by
-less than 10\% (e.g. Figure~\ref{fig-grid-gen_params-cost-lin_sep}),
-which reflects a common tradeoff between costs and plausibility
+around 20-25\% for \emph{Circles} (e.g.
+Figure~\ref{fig-grid-gen_params-cost-circles}) and up to 50\% for
+\emph{Moons} (e.g. Figure~\ref{fig-grid-gen_params-cost-moons}). For the
+\emph{Linearly Separable} data, costs are generally increased although
+typically by less than 10\% (e.g.
+Figure~\ref{fig-grid-gen_params-cost-lin_sep}), which reflects a common
+tradeoff between costs and plausibility
 (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
   \toprule
-  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Se} \\\midrule
   \endfirsthead
   \toprule
-  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Std} \\\midrule
+  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Se} \\\midrule
   \endhead
@@
 change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Circles.}
+metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Circles.}
 
@@
 change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Linearly Separable.}
+metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Linearly Separable.}
 
@@
 from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
+(\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
 evaluation runs and test settings (varying parameters for \emph{ECCo}).
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Adult.}
+metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Adult.}
 
@@
 change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: California Housing.}
+metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: California Housing.}
 
@@
 change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Credit.}
+metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Credit.}
 
@@
 change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: GMSC.}
+metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: GMSC.}
 
@@
 change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: MNIST.}
+metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: MNIST.}
 
@@
 change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Moons.}
+metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Moons.}
 
@@
 change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Overlapping.}
+metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCo}).
+Data: Overlapping.}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
 
@@
 time to completion varied by dataset for reasons discussed in
-Section~\ref{sec-discussion}: 0h49m (\emph{Moons}), 1h4m (\emph{Linearly
+\textbf{?@sec-discussion}: 0h49m (\emph{Moons}), 1h4m (\emph{Linearly
 Separable}), 1h49m (\emph{Circles}), 3h52m (\emph{Overlapping}). Model
@@
 
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-altmeyer2023explaining}
+Altmeyer, Patrick, Arie van Deursen, and Cynthia C. S. Liem. 2023.
+{``{Explaining Black-Box Models through Counterfactuals}.''} In
+\emph{Proceedings of the JuliaCon Conferences}, 1:130.
+
+\bibitem[\citeproctext]{ref-altmeyer2024faithful}
+Altmeyer, Patrick, Mojtaba Farmanbar, Arie van Deursen, and Cynthia C.
+S. Liem. 2024. {``{Faithful Model Explanations through
+Energy-Constrained Conformal Counterfactuals}.''} In \emph{Proceedings
+of the Thirty-Eighth AAAI Conference on Artificial Intelligence},
+38:10829--37. 10. \url{https://doi.org/10.1609/aaai.v38i10.28956}.
+
+\bibitem[\citeproctext]{ref-bezanson2017julia}
+Bezanson, Jeff, Alan Edelman, Stefan Karpinski, and Viral B Shah. 2017.
+{``Julia: A Fresh Approach to Numerical Computing.''} \emph{SIAM Review}
+59 (1): 65--98. \url{https://doi.org/10.1137/141000671}.
+
+\bibitem[\citeproctext]{ref-milan2023dataframes}
+Bouchet-Valat, Milan, and Bogumił Kamiński. 2023. {``DataFrames.jl:
+Flexible and Fast Tabular Data in Julia.''} \emph{Journal of Statistical
+Software} 107 (4): 1--32. \url{https://doi.org/10.18637/jss.v107.i04}.
+
+\bibitem[\citeproctext]{ref-byrne2021mpi}
+Byrne, Simon, Lucas C. Wilcox, and Valentin Churavy. 2021. {``MPI.jl:
+Julia Bindings for the Message Passing Interface.''} \emph{Proceedings
+of the JuliaCon Conferences} 1 (1): 68.
+\url{https://doi.org/10.21105/jcon.00068}.
+
+\bibitem[\citeproctext]{ref-chagas2024pretty}
+Chagas, Ronan Arraes Jardim, Ben Baumgold, Glen Hertz, Hendrik Ranocha,
+Mark Wells, Nathan Boyer, Nicholas Ritchie, et al. 2024.
+{``Ronisbr/PrettyTables.jl: V2.4.0.''} Zenodo.
+\url{https://doi.org/10.5281/zenodo.13835553}.
+
+\bibitem[\citeproctext]{ref-PlotsJL}
+Christ, Simon, Daniel Schwabeneder, Christopher Rackauckas, Michael
+Krabbe Borregaard, and Thomas Breloff. 2023. {``Plots.jl -- a User
+Extendable Plotting API for the Julia Programming Language.''}
+https://doi.org/\url{https://doi.org/10.5334/jors.431}.
+
+\bibitem[\citeproctext]{ref-danisch2021makie}
+Danisch, Simon, and Julius Krumbiegel. 2021. {``{Makie.jl}: Flexible
+High-Performance Data Visualization for {Julia}.''} \emph{Journal of
+Open Source Software} 6 (65): 3349.
+\url{https://doi.org/10.21105/joss.03349}.
+
+\bibitem[\citeproctext]{ref-goodfellow2014explaining}
+Goodfellow, Ian, Jonathon Shlens, and Christian Szegedy. 2015.
+{``{Explaining and Harnessing Adversarial Examples}.''}
+\url{https://arxiv.org/abs/1412.6572}.
+
+\bibitem[\citeproctext]{ref-grathwohl2020your}
+Grathwohl, Will, Kuan-Chieh Wang, Joern-Henrik Jacobsen, David Duvenaud,
+Mohammad Norouzi, and Kevin Swersky. 2020. {``Your Classifier Is
+Secretly an Energy Based Model and You Should Treat It Like One.''} In
+\emph{International Conference on Learning Representations}.
+
+\bibitem[\citeproctext]{ref-gretton2012kernel}
+Gretton, Arthur, Karsten M Borgwardt, Malte J Rasch, Bernhard Schölkopf,
+and Alexander Smola. 2012. {``A Kernel Two-Sample Test.''} \emph{The
+Journal of Machine Learning Research} 13 (1): 723--73.
+
+\bibitem[\citeproctext]{ref-hastie2009elements}
+Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
+Elements of Statistical Learning}. Springer New York.
+\url{https://doi.org/10.1007/978-0-387-84858-7}.
+
+\bibitem[\citeproctext]{ref-innes2018fashionable}
+Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco
+Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and
+Viral Shah. 2018. {``Fashionable Modelling with Flux.''}
+\url{https://arxiv.org/abs/1811.01457}.
+
+\bibitem[\citeproctext]{ref-innes2018flux}
+Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
+\emph{Journal of Open Source Software} 3 (25): 602.
+\url{https://doi.org/10.21105/joss.00602}.
+
+\bibitem[\citeproctext]{ref-joshi2019realistic}
+Joshi, Shalmali, Oluwasanmi Koyejo, Warut Vijitbenjaronk, Been Kim, and
+Joydeep Ghosh. 2019. {``{Towards Realistic Individual Recourse and
+Actionable Explanations in Black-Box Decision Making Systems}.''}
+\url{https://arxiv.org/abs/1907.09615}.
+
+\bibitem[\citeproctext]{ref-wachter2017counterfactual}
+Wachter, Sandra, Brent Mittelstadt, and Chris Russell. 2017.
+{``Counterfactual Explanations Without Opening the Black Box:
+{Automated} Decisions and the {GDPR}.''} \emph{Harv. JL \& Tech.} 31:
+841. \url{https://doi.org/10.2139/ssrn.3063289}.
+
+\end{CSLReferences}
+
 \end{document}

### 2025-07-22 paper/preprint/paper.tex
finished table and rerendered preprint

@@
 \theoremstyle{plain}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\usepackage{amsfonts}
+\usepackage{amsmath}
+\usepackage{siunitx}
+\sisetup{uncertainty-mode = separate}
+\DeclareMathSizes{10}{9}{7}{6.5}
+\usepackage{enumitem}
 \usepackage{arxiv}
@@
 to produce some desired output. To be useful in real-word
-decision-making systems, counterfactuals should be (1) plausible with
-respect to the underlying data and (2) actionable with respect to the
-user-defined mutability constraints. Much existing research has
-therefore focused on developing post-hoc methods to generate
-counterfactuals that meet these desiderata. In this work, we instead
-hold models directly accountable for the desired end goal:
-counterfactual training employs counterfactuals ad-hoc during the
-training phase to minimize the divergence between learned
-representations and plausible, actionable explanations. We demonstrate
-empirically and theoretically that our proposed method facilitates
-training models that deliver inherently desirable explanations while
-promoting robustness and preserving high predictive performance.
+decision-making systems, counterfactuals should be plausible with
+respect to the underlying data and actionable with respect to the
+feature mutability constraints. Much existing research has therefore
+focused on developing post-hoc methods to generate counterfactuals that
+meet these desiderata. In this work, we instead hold models directly
+accountable for the desired end goal: counterfactual training employs
+counterfactuals during the training phase to minimize the divergence
+between learned representations and plausible, actionable explanations.
+We demonstrate empirically and theoretically that our proposed method
+facilitates training models that deliver inherently desirable
+counterfactual explanations and exhibit greatly improved adversarial
+robustness.
 \end{abstract}
@@
 Today's prominence of artificial intelligence (AI) has largely been
-driven by \textbf{representation learning}: instead of relying on
-features and rules that are carefully hand-crafted by humans, modern
-machine learning (ML) models are tasked with learning representations
-directly from data, guided by narrow objectives such as predictive
-accuracy (\citeproc{ref-goodfellow2016deep}{Goodfellow, Bengio, and
-Courville 2016}). Modern advances in computing have made it possible to
-provide such models with ever-growing degrees of freedom to achieve that
-task, which frequently allows them to outperform traditionally more
-parsimonious models. Unfortunately, in doing so, models learn
-increasingly complex and highly sensitive representations that humans
-can no longer easily interpret.
-
-The trend towards complexity for the sake of performance has come under
-serious scrutiny in recent years. At the very cusp of the deep learning
-(DL) revolution, Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2014}) showed that artificial
-neural networks (ANN) are sensitive to adversarial examples (AEs):
-perturbed versions of data instances that yield vastly different model
-predictions despite being ``imperceptible'' in that they are
-semantically indifferent from their factual counterparts. Even though
-some partially effective mitigation strategies have been proposed---most
-notably \textbf{adversarial training}
-(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
-2015})---truly robust deep learning remains unattainable even for models
-that are considered ``shallow'' by today's standards
-(\citeproc{ref-kolter2023keynote}{Kolter 2023}).
-
-Part of the problem is that the high degrees of freedom provide room for
-many solutions that are locally optimal with respect to narrow
-objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
-  follow the standard ML convention, where ``degrees of freedom'' refer
-  to the number of parameters estimated from data.} Indeed, recent work
-on the so-called ``lottery ticket hypothesis'' suggests that modern
-neural networks can be pruned by up to 90\% while preserving their
-predictive performance (\citeproc{ref-frankle2018lottery}{Frankle and
-Carbin 2019}). Similarly, Zhang et al.
-(\citeproc{ref-zhang2021understanding}{2021}) showed that
-state-of-the-art neural networks are expressive enough to fit randomly
-labeled data. Thus, looking at the predictive performance alone, the
-solutions may seem to provide compelling explanations for the data, when
-in fact they are based on purely associative, semantically meaningless
-patterns. This poses two challenges. Firstly, there is no dependable way
-to verify if representations correspond to meaningful, plausible
-explanations. Secondly, even if we could resolve the first challenge, it
-remains undecided how to ensure that models can \emph{only} learn
-valuable explanations.
-
-The first challenge has attracted an abundance of research on
-\textbf{explainable AI} (XAI), a paradigm that focuses on the
-development of tools to derive (post-hoc) explanations from complex
-model representations. Such explanations should mitigate a scenario in
-which practitioners deploy opaque models and blindly rely on their
-predictions. On countless occasions, this has happened in practice and
-caused real harms to people who were adversely and unfairly affected by
-automated decision-making (ADM) systems involving opaque models (see,
-e.g., (\citeproc{ref-oneil2016weapons}{O'Neil 2016})). Effective XAI
-tools can aid us in monitoring models and providing recourse to
-individuals to turn negative outcomes (e.g., ``loan application
-rejected'') into positive ones (e.g., ``application accepted''). In line
-with this, our work builds upon \textbf{counterfactual explanations}
-(CE) proposed by Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}) as an effective
-approach to achieve this goal. CEs prescribe minimal changes for factual
-inputs that, if implemented, would prompt some fitted model to produce a
-desired output.
-
-To our surprise, the second challenge has not yet attracted major
-research interest. Specifically, there has been no concerted effort
-towards improving the ``explanatory capacity'' of models, i.e., the
-degree to which learned representations correspond to explanations that
-are \textbf{interpretable} and deemed \textbf{plausible} by humans (see
-Def. \ref{def-explainability}). Instead, the choice has generally been
-to improve the ability of XAI tools to identify the subset of
-explanations that are both plausible and valid for any given model,
-independent of whether the learned representations are in fact
-compatible with plausible explanations
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-Fortunately, recent findings indicate that improved explanatory capacity
-can arise as a consequence of regularization techniques aimed at other
-training objectives such as generative capacity, generalization, or
+driven by the success of representation learning with high degrees of
+freedom: instead of relying on features and rules hand-crafted by
+humans, modern machine learning (ML) models are tasked with learning
+highly complex representations directly from the data, guided by narrow
+objectives such as predictive accuracy
+(\citeproc{ref-goodfellow2016deep}{Goodfellow, Bengio, and Courville
+2016}). These models tend to be so complex that humans cannot easily
+interpret their decision logic.
+
+Counterfactual explanations (CE) have become a key part of the broader
+explainable AI (XAI) toolkit
+(\citeproc{ref-molnar2022interpretable}{Molnar 2022}) that can be
+applied to make sense of this complexity. Originally proposed by
+Wachter, Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}), CEs prescribe minimal
+changes for factual inputs that, if implemented, would prompt some
+fitted model to produce an alternative, more desirable output. This is
+useful and necessary to not only understand how opaque models make their
+predictions, but also to provide algorithmic recourse (AR) to
+individuals subjected to them: a retail bank, for example, could use CE
+to provide meaningful feedback to unsuccessful loan applicants that were
+rejected based on an opaque automated decision-making (ADM) system
+(Figure~\ref{fig-poc}).
+
+For such feedback to be meaningful, counterfactual explanations need to
+fulfill certain desiderata (\citeproc{ref-verma2020counterfactual}{Verma
+et al. 2022}; \citeproc{ref-karimi2020survey}{Karimi et al.
+2021})---they should be faithful to the model
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}), plausible
+(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}) and actionable
+(\citeproc{ref-ustun2019actionable}{Ustun, Spangher, and Liu 2019}).
+Plausibility is typically understood as counterfactuals being
+\emph{in-domain}: unsuccessful loan applicants that implement the
+provided recourse should end up with credit profiles that are genuinely
+similar to that of individuals who have successfully repaid their loans
+in the past. Actionable explanations comply with practical constraints:
+a young, unsuccessful loan applicant cannot increase their age in an
+instance.
+
+Existing state-of-the-art (SOTA) approaches in the field have largely
+focused on designing model-agnostic CE methods that identify subsets of
+counterfactuals, which comply with specific desiderata. This is
+problematic, because the narrow focus on any specific desideratum can
+adversely affect others: it is possible, for example, to generate
+plausible counterfactuals for models that are also highly vulnerable to
+implausible, possibly adversarial counterfactuals
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). In this
+work, we therefore embrace the paradigm that models (as opposed to
+explanation methods) should be held accountable for explanations that
+are plausible and actionable. While previous work has shown that at
+least plausibility can be indirectly achieved through existing
+techniques aimed at models' generative capacity, generalization and
 robustness (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024};
 \citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein 2020};
-\citeproc{ref-schut2021generating}{Schut et al. 2021}). As further
-discussed in Section~\ref{sec-lit}, our work consolidates these findings
-within a single objective.
-
-\textbf{Specifically, we introduce Counterfactual Training (CT)}: a
-novel training regime explicitly geared towards improving the
-explainability of models. In high-level terms, we define this concept as
-as the extent to which valid explanations derived for an opaque model
-are also deemed plausible with respect to the underlying data and the
-global actionability constraints. To the best of our knowledge, our
-framework represents the first attempt to address this challenge by
-employing counterfactual explanations already in the training phase.
+\citeproc{ref-schut2021generating}{Schut et al. 2021}), we directly
+incorporate both plausibility and actionability in the training
+objective of models to improve their overall explanatory capacity.
+
+Specifically, we propose \textbf{counterfactual training (CT)}: a novel
+training regime that leverages counterfactual explanations on-the-fly to
+ensure that differentiable models learn plausible and actionable
+explanations for the underlying data, while at the same time also being
+more robust to adversarial examples (AE). Figure~\ref{fig-poc}
+illustrates the outcomes of CT compared to a conventionally trained
+model. First, in panel (a), faithful and valid counterfactuals end up
+near the decision boundary forming a clearly distinguishable cluster in
+the target class (orange). In panel (b), CT is applied to the same
+underlying linear classifier architecture resulting in much more
+plausible counterfactuals. In panel (c), the classifier is again trained
+conventionally and we have introduced a mutability constraint on the
+\emph{age} feature at test time---counterfactuals are valid but the
+classifier is roughly equally sensitive to both features. By contrast,
+the decision boundary in panel (d) has titled, making the model trained
+with CT relatively less sensitive to the immutable \emph{age} feature.
+To achieve these outcomes, CT draws inspiration from the literature on
+contrastive and robust learning: we contrast faithful CEs with
+ground-truth data while protecting immutable features, and capitalize on
+methodological links between CE and AE by penalizing the model's
+adversarial loss on interim (\emph{nascent}) counterfactuals. To the
+best of our knowledge, CT represents the first venture in this direction
+with promising empirical and theoretical results.
 
 The remainder of this manuscript is structured as follows.
-Section~\ref{sec-lit} presents related work, focusing on the link
-between AEs and CEs. Then follow our two principal contributions. In
-Section~\ref{sec-method}, we introduce our methodological framework and
-show theoretically that it can be employed to enforce global
-actionability constraints. In Section~\ref{sec-experiments}, through
-extensive experiments, we demonstrate that CT substantially improves
-explainability and positively contributes to the robustness of trained
-models without sacrificing predictive performance. Finally, in
-Section~\ref{sec-discussion}, we discuss the challenges and conclude
-that CT is a promising approach towards making opaque models more
-trustworthy.
+Section~\ref{sec-lit} presents related work, focusing on the links to
+contrastive and robust learning. Then follow our two principal
+contributions. In Section~\ref{sec-method}, we introduce our
+methodological framework and show theoretically that it can be employed
+to respect global actionability constraints. In our experiments
+(Section~\ref{sec-experiments}), we find that thanks to counterfactual
+training, (1) the implausibility of CEs decreases by up to 90\%; (2) the
+cost of reaching valid counterfactuals with protected features decreases
+by 19\% on average; and (3) models' adversarial robustness improves
+across the board. Finally, we discuss open challenges in
+Section~\ref{sec-discussion} and conclude in
+Section~\ref{sec-conclusion}.
+
+\begin{figure*}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
+
+}
+
+\caption{\label{fig-poc}Counterfactual explanations (stars) for linear
+classifiers trained under different regimes on synthetic data: (a)
+conventional training, all mutable; (b) CT, all mutable; (c)
+conventional, \emph{age} immutable; (d) CT, \emph{age} immutable. The
+linear decision boundary is shown in green along with training data
+colored according to ground-truth labels: \(y^-=\text{"loan withheld"}\)
+(blue) and \(y^+=\text{"loan provided"}\) (orange). Class and feature
+annotations (\emph{debt} and \emph{age}) are for illustrative purposes.}
+
+\end{figure*}%
 
@@
 To make the desiderata for our framework more concrete, we follow
+previous work in tying the explanatory capacity of models to the quality
+of CEs that can be generated for them
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024};
+\citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein
+2020}). For simplicity, we refer to ``explanatory capacity'' as
+``explainability'' in the rest of this manuscript (see Def.
+\ref{def-explainability}).
+
+\subsection{Explainability and Contrastive
+Learning}\label{explainability-and-contrastive-learning}
+
+In a closely related work, Altmeyer et al.
+(\citeproc{ref-altmeyer2024faithful}{2024}) show that model averaging
+and, in particular, contrastive model objectives can produce more
+explainable and hence trustworthy models. The authors propose a way to
+generate counterfactuals that are maximally faithful in that they are
+consistent with what models have learned about the underlying data.
+Formally, they rely on tools from energy-based modelling
+(\citeproc{ref-teh2003energy}{Teh et al. 2003}) to minimize the
+contrastive divergence between the distribution of counterfactuals and
+the conditional posterior over inputs learned by a model. Their
+algorithm, \emph{ECCCo}, yields plausible counterfactual explanations if
+and only if the underlying model has learned representations that align
+with them. The authors find that both deep ensembles
+(\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan, Pritzel,
+and Blundell 2017}) and joint energy-based models (JEMs)
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}), a form of
+constrastive learning, tend to do well in this regard.
+
+It helps to look at these findings through the lens of representation
+learning with high degrees of freedom. Deep ensembles are approximate
+Bayesian model averages, which are particularly effective when models
+are underspecified by the available data
+(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
+mitigates the risk of overrelying on a single locally optimal
+representation that corresponds to semantically meaningless
+explanations. Likewise, previous work of Schut et al.
+(\citeproc{ref-schut2021generating}{2021}) found that generating
+plausible (``interpretable'') CEs is almost trivial for deep ensembles
+that have undergone adversarial training. The case for JEMs is even
+clearer: they optimize a hybrid objective that induces both high
+predictive performance and strong generative capacity
+(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}), which
+resembles the idea of aligning models with plausible explanations and
+has inspired CT.
+
+\subsection{Explainability and Robust
+Learning}\label{explainability-and-robust-learning}
+
 Augustin, Meinke, and Hein
-(\citeproc{ref-augustin2020adversarial}{2020}) in tying the concept of
-explainability to the quality of CEs that can be generated for a given
-model. The authors show that CEs---understood as minimal input
-perturbations that yield some desired model prediction---tend to be more
-meaningful if the underlying model is more robust to adversarial
-examples. We can make intuitive sense of this finding when looking at
-adversarial training (AT) through the lens of representation learning
-with high degrees of freedom. As argued before, learned representations
-may be sensitive to producing implausible explanations and mispredicting
-for worst-case counterfactuals (i.e., AEs). Thus, by inducing models to
-``unlearn'' susceptiblity to such examples, adversarial training can
-effectively remove implausible explanations from the solution space.
-
-\subsection{Adversarial Examples are Counterfactual
-Explanations}\label{adversarial-examples-are-counterfactual-explanations}
+(\citeproc{ref-augustin2020adversarial}{2020}) show that CEs tend to be
+more meaningful (``explainable'') if the underlying model is more robust
+to adversarial examples. Once again, we can make intuitive sense of this
+finding if we look at adversarial training (AT) through the lens of
+representation learning with high degrees of freedom: highly complex and
+flexible models may learn representations that make them sensitive to
+implausible or even adversarial examples
+(\citeproc{ref-szegedy2013intriguing}{Szegedy et al. 2014}). Thus, by
+inducing models to ``unlearn'' susceptibility to such examples,
+adversarial training can effectively remove implausible explanations
+from the solution space.
 
 This interpretation of the link between explainability through
-counterfactuals on one side and robustness to adversarial examples on
-the other is backed by empirical evidence. Sauer and Geiger
-(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrates that using
+counterfactuals on the one side, and robustness to adversarial examples
+on the other is backed by empirical evidence. Sauer and Geiger
+(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
 counterfactual images during classifier training improves model
 robustness. Similarly, Abbasnejad et al.
-(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argues that
+(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
 counterfactuals represent potentially useful training data in machine
 learning, especially in supervised settings where inputs may be
-reasonably mapped to multiple outputs. They, too, demonstrate that
-augmenting the training data of image classifiers can improve
-generalization. Finally, Teney, Abbasnedjad, and Hengel
-(\citeproc{ref-teney2020learning}{2020}) proposes an approach using
-counterfactuals in training that does not rely on data augmentation:
-they argue that counterfactual pairs typically already exist in training
-datasets. Specifically, their approach relies on identifying similar
-input samples with different annotations and ensuring that the gradient
-of the classifier aligns with the vector between such pairs of
-counterfactual inputs using the cosine distance as the loss function.
-
-In the natural language processing (NLP) domain, CEs have also been used
-to improve models through data augmentation. Wu et al.
-(\citeproc{ref-wu2021polyjuice2}{2021}) proposes \emph{Polyjuice}, a
-general-purpose counterfactual generator for language models. The
-authors empirically demonstrate that the augmentation of training data
-with their method improves robustness in a number of NLP tasks.
-Balashankar et al. (\citeproc{ref-balashankar2023improving}{2023})
-similarly uses \emph{Polyjuice} to augment NLP datasets through diverse
-counterfactuals and show that classifier robustness improves by up to
-20\%. Finally, Luu and Inoue
-(\citeproc{ref-luu2023counterfactual}{2023}) introduces Counterfactual
-Adversarial Training (CAT) that also aims to improve generalization and
-robustness of language models through a three-step procedure: the
-authors identify training samples that are subject to high predictive
-uncertainty, generate CEs for them, and fine-tune the language model on
-a dataset augmented with the CEs.
+reasonably mapped to multiple outputs. They, too, show that augmenting
+the training data of (image) classifiers can improve generalization
+performance. Finally, Teney, Abbasnedjad, and Hengel
+(\citeproc{ref-teney2020learning}{2020}) argue that counterfactual pairs
+tend to exist in training data. Hence, their approach aims to identify
+similar input samples with different annotations and ensure that the
+gradient of the classifier aligns with the vector between such pairs of
+counterfactual inputs using a cosine distance loss function.
+
+CEs have also been used to improve models in the natural language
+processing domain. For example, Wu et al.
+(\citeproc{ref-wu2021polyjuice}{2021}) propose \emph{Polyjuice}, a
+general-purpose CE generator for language models and demonstrate that
+the augmentation of training data with \emph{Polyjuice} improves
+robustness in a number of tasks, while Luu and Inoue
+(\citeproc{ref-luu2023counterfactual}{2023}) introduce the
+\emph{Counterfactual Adversarial Training} (CAT) framework that aims to
+improve generalization and robustness of language models by generating
+counterfactuals for training samples that are subject to high predictive
+uncertainty.
 
@@
 (\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
-Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) shows
-that CEs and AEs are equivalent under certain conditions.
-
-Two recent works are closely related to ours in that they use
+Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
+CEs and AEs are equivalent under certain conditions.
+
+Two other works are closely related to ours in that they use
 counterfactuals during training with the explicit goal of affecting
@@
 Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
-proposes a way to train models that guarantee individual recourse to
-some positive target class with high probability. Their approach builds
-on adversarial training by explicitly inducing susceptibility to
-targeted adversarial examples for the positive class. Additionally, the
-proposed method allows for imposing a set of actionability constraints
-ex-ante. For example, users can specify that certain features are
-immutable. Secondly, Guo, Nguyen, and Yadav
-(\citeproc{ref-guo2023counternet}{2023}) is the first to propose an
+propose a way to train models that guarantee recourse to a positive
+target class with high probability. Their approach builds on adversarial
+training by explicitly inducing susceptibility to targeted AEs for the
+positive class. Additionally, the method allows for imposing a set of
+actionability constraints ex-ante. For example, users can specify that
+certain features are immutable. Secondly, Guo, Nguyen, and Yadav
+(\citeproc{ref-guo2023counternet}{2023}) are the first to propose an
 end-to-end training pipeline that includes CEs as part of the training
-procedure. In particular, they propose a specific network architecture
-that includes a predictor and CE generator network, where the parameters
-of the CE generator network are learnable. Counterfactuals are generated
-during each training iteration and fed back to the predictor network. In
-contrast to Guo, Nguyen, and Yadav
-(\citeproc{ref-guo2023counternet}{2023}), we impose no restrictions on
-the ANN architecture at all.
-
-\subsection{Aligning Representations with Plausible
-Explanations}\label{aligning-representations-with-plausible-explanations}
-
-Improving the adversarial robustness of models is not the only path
-towards aligning representations with plausible explanations. In a work
-closely related to this one, Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}) shows that explainability
-can be improved through model averaging and refined model objectives.
-The authors propose a way to generate counterfactuals that are maximally
-faithful to the model in that they are consistent with what the model
-has learned about the underlying data. Formally, they rely on tools from
-energy-based modelling (\citeproc{ref-teh2003energy}{Teh et al. 2003})
-to minimize the divergence between the distribution of counterfactuals
-and the conditional posterior over inputs learned by the model. Their
-proposed counterfactual explainer, \emph{ECCCo}, yields plausible
-explanations if and only if the underlying model has learned
-representations that align with them. The authors find that both deep
-ensembles (\citeproc{ref-lakshminarayanan2016simple}{Lakshminarayanan,
-Pritzel, and Blundell 2017}) and joint energy-based models (JEMs)
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) tend to do
-well in this regard.
-
-Once again it helps to look at these findings through the lens of
-representation learning with high degrees of freedom. Deep ensembles are
-approximate Bayesian model averages, which are most called for when
-models are underspecified by the available data
-(\citeproc{ref-wilson2020case}{Wilson 2020}). Averaging across solutions
-mitigates the aforementioned risk of relying on a single locally optimal
-representations that corresponds to semantically meaningless
-explanations for the data. Previous work of Schut et al.
-(\citeproc{ref-schut2021generating}{2021}) similarly found that
-generating plausible (``interpretable'') CEs is almost trivial for deep
-ensembles that have also undergone adversarial training. The case for
-JEMs is even clearer: they involve a hybrid objective that induces both
-high predictive performance and generative capacity
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}). This is
-closely related to the idea of aligning models with plausible
-explanations and has inspired our CT objective.
+procedure. Their \emph{CounterNet} network architecture includes a
+predictor and a CE generator, where the parameters of the CE generator
+are learnable. Counterfactuals are generated during each training
+iteration and fed back to the predictor. In contrast, we impose no
+restrictions on the ANN architecture at all.
 
@@
 
-In this section we propose our novel counterfactual training objective.
-In CT, we combine ideas from adversarial training, counterfactual
-explanations, and energy-based modelling with the explicit goal of
-aligning representations with plausible explanations that comply with
-user-defined actionability constraints.
-
-In the context of counterfactual explanations, plausibility has broadly
-been defined as the degree to which counterfactuals comply with the
-underlying data-generating process
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024};
-\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
-\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020}). Plausibility is
-a necessary but insufficient condition for using CEs to provide
-algorithmic recourse (AR) to individuals (negatively) affected by opaque
-models. To be actionable, AR recommendations must also be attainable. A
-plausible CE for a rejected 20-year-old loan applicant, for example,
-might reveal that their application would have been accepted, if only
-they were 20 years older. Ignoring all other features, this would comply
-with the definition of plausibility if 40-year-old individuals were in
-fact more credit-worthy on average than young adults. But of course this
-CE does not qualify for providing actionable recourse to the applicant
-since \emph{age} is not a (directly) mutable feature. Counterfactual
-training aims to improve model explainability by aligning models with
-counterfactuals that meet both desiderata: plausibility and
-actionability. Formally, we define explainability as follows:
+This section introduces the counterfactual training framework, applying
+ideas from contrastive and robust learning to counterfactual
+explanations. CT produces models whose learned representations align
+with plausible explanations that comply with user-defined actionability
+constraints.
+
+Counterfactual explanations are typically generated by solving
+variations of the following optimization problem,
+\begin{equation}\phantomsection\label{eq-general}{
+\begin{aligned}
+\min_{\mathbf{X}^\prime \in \mathcal{X}^D} \left\{  {\text{yloss}(\mathbf{M}_\theta(\mathbf{x}^{\prime}),\mathbf{y}^+)}+ \lambda {\text{reg}(\mathbf{x}^{\prime}) }  \right\} 
+\end{aligned}
+}\end{equation} where
+\(\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}\) denotes a
+classifier, \(\mathbf{x}^{\prime}\) denotes the counterfactual with
+\(D\) features and \(\mathbf{y}^+\in\mathcal{Y}\) denotes some target
+class. The \(\text{yloss}(\cdot)\) function quantifies the discrepancy
+between current model predictions for \(\mathbf{x}^{\prime}\) and the
+target class (a conventional choice is cross-entropy). Finally, we use
+\(\text{reg}(\cdot)\) to denote any form of regularization used to
+induce certain properties on the counterfactual. In their seminal paper,
+Wachter, Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}) propose regularizing
+the distance between counterfactuals and their original factual values
+to ensure that individuals seeking recourse through CE face minimal
+costs in terms of feature changes. Different variations of
+Equation~\ref{eq-general} have been proposed in the literature to
+address many desiderata including the ones discussed above
+(faithfulness, plausibility and actionability). Like Wachter,
+Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}), most of these
+approaches rely on gradient descent to optimize
+Equation~\ref{eq-general}. For more details on the approaches tested in
+this work, we refer the reader to the supplementary appendix. In the
+following, we describe in detail how counterfactuals are generated and
+used in counterfactual training.
+
+\subsection{Proposed Training
+Objective}\label{proposed-training-objective}
+
+The goal of CT is to improve model explainability by aligning models
+with faithful explanations that are plausible and actionable. Formally,
+we define explainability as follows:
 
@@
 Let $\mathbf{M}_\theta: \mathcal{X} \mapsto \mathcal{Y}$ denote a supervised classification model that maps from the $D$-dimensional input space $\mathcal{X}$ to representations $\phi(\mathbf{x};\theta)$ and finally to the $K$-dimensional output space $\mathcal{Y}$.
-Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$ where $\arg\max_y{\mathbf{y}^{+}}=y^+$ and $y^+$ denotes the index of the target class.
-
-We say that $\mathbf{M}_\theta$ is \textbf{explainable} to the extent that faithfully generated counterfactuals are plausible and actionable. We define these properties as follows:
-
-\begin{enumerate}
-    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is some small region around $\mathbf{x}^{\prime}$.
-    \item (Actionability) Permutations $\Delta$ are subject to some actionability constraints.
-    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$ where $A$ is defined as above.
-\end{enumerate}
-where $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior over inputs. 
+Assume that for any given input-output pair $\{\mathbf{x},\mathbf{y}\}_i$ there exists a counterfactual $\mathbf{x}^{\prime} = \mathbf{x} + \Delta: \mathbf{M}_\theta(\mathbf{x}^{\prime}) = \mathbf{y}^{+} \neq \mathbf{y} = \mathbf{M}_\theta(\mathbf{x})$, where $\arg\max_y{\mathbf{y}^{+}}=y^+$ is the index of the target class.
+
+We say that $\mathbf{M}_\theta$ has an \textbf{explanatory capacity} to the extent that faithfully generated, valid counterfactuals are also plausible and actionable. We define these properties as:
+
+\begin{itemize}
+    \item (Faithfulness) $\int^{A} p_\theta(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ is an arbitrarily small region around $\mathbf{x}^{\prime}$.
+    \item (Plausibility) $\int^{A} p(\mathbf{x}^\prime|\mathbf{y}^{+})d\mathbf{x} \rightarrow 1$; $A$ as specified above.
+    \item (Actionability) Perturbations $\Delta$ may be subject to some actionability constraints.
+\end{itemize}
+Here, $p_\theta(\mathbf{x}|\mathbf{y}^{+})$ denotes the conditional posterior distribution over inputs.
+For simplicity, we refer to a model with high explanatory capacity as \textbf{explainable} in this manuscript.
 \end{definition}
 
-\noindent The characterization of faithfulness and plausibility in Def.
-\ref{def-explainability} is the same as in Altmeyer et al.
+The characterization of faithfulness and plausibility in Def.
+\ref{def-explainability} follows Altmeyer et al.
 (\citeproc{ref-altmeyer2024faithful}{2024}), with adapted notation.
@@
 faithful counterfactuals are consistent with what the model has learned
-about input data. Actionability constraints in Def.
+about the input data. Ac tionability constraints in Def.
 \ref{def-explainability} vary and depend on the context in which
-\(\mathbf{M}_\theta\) is deployed. In this work, we focus on domain and
-mutability constraints for individual features \(x_d\) for
-\(d=1,...,D\). We limit ourselves to classification tasks for reasons
-discussed in Section~\ref{sec-discussion}.
-
-\subsection{Our Proposed Objective}\label{our-proposed-objective}
+\(\mathbf{M}_\theta\) is deployed. In this work, we choose to only
+consider domain and mutability constraints for individual features
+\(x_d\) for \(d=1,...,D\). We also limit ourselves to classification
+tasks for reasons discussed in Section~\ref{sec-discussion}.
 
 Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
-explanation generated through gradient descent over \(T\) iterations as
-initially proposed by Wachter, Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
-let \(T\) vary and consider the counterfactual search as converged as
-soon as the predicted probability for the target class has reached a
-pre-determined threshold, \(\tau\):
+generated through gradient descent over \(T\) iterations as originally
+proposed by Wachter, Mittelstadt, and Russell
+(\citeproc{ref-wachter2017counterfactual}{2017}). CT adopts
+gradient-based CE search in training to generate on-the-fly model
+explanations \(\mathbf{x}^\prime\) for the training samples. We use the
+term \emph{nascent} to denote interim counterfactuals
+\(\mathbf{x}_{t\leq T}^\prime\) that have not yet converged. As we
+explain below, these nascent counterfactuals can be stored and
+repurposed as adversarial examples. Conversely, we consider
+counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
+they have either exhausted all \(T\) iterations or converged by reaching
+a pre-specified threshold, \(\tau\), for the predicted probability of
+the target class:
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
-where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
-  background information on gradient-based counterfactual search and
-  convergence see supplementary appendix.}
-
-To train models with high explainability as defined in Def.
-\ref{def-explainability}, we propose to leverage counterfactuals in the
-following objective:
-
+where \(\mathcal{S}\) is the softmax function.
+
+Formally, we propose the following counterfactual training objective to
+train explainable (as in Def. \ref{def-explainability}) models,
 \begin{equation}\phantomsection\label{eq-obj}{
-\begin{split}
-\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) &+ \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
-\end{split}
-}\end{equation} where \(\text{yloss}(\cdot)\) is a classification loss
+\begin{aligned}
+&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
+\end{aligned}
+}\end{equation} where \(\text{yloss}(\cdot)\) is any classification loss
 that induces discriminative performance (e.g., cross-entropy). The
@@
 be summarized as inducing explainability directly and indirectly by
-penalizing: (1) the contrastive divergence, \(\text{div}(\cdot)\),
+penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
 between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
 samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
-target class \(y^+\), and, (2) the adversarial loss,
-\(\text{advloss}(.)\), with respect to nascent counterfactuals
+target class \(y^+\), and (2) the adversarial loss,
+\(\text{advloss}(.)\), wrt. nascent counterfactuals
 \(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
@@
 of the energy terms involved in \(\text{div}(\cdot)\)
-(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-off
-between the components can governed through penalties
-\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
-\(\lambda_{\text{reg}}\).
+(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). The trade-offs
+between these components are adjusted through \(\lambda_{\text{div}}\),
+\(\lambda_{\text{adv}}\) and \(\lambda_{\text{reg}}\). The full training
+regime is sketched out in Algorithm \ref{alg-experiment}.
+
+\begin{algorithm}[h]
+  \caption{Counterfactual Training}
+    \label{alg-experiment}
+    \begin{algorithmic}[1]
+    \REQUIRE Training dataset $\mathcal{D}$, initialize model $\mathbf{M}_{\theta}$
+    \WHILE{not converged}
+        \STATE Sample $\mathbf{x}$ and $\mathbf{y}$ from dataset $\mathcal{D}$.
+        \STATE Sample $\mathbf{x}^{\prime}_0$, $\mathbf{y}^+$ and $\mathbf{x}^+$.
+        \FOR{$t = 1$ to $T$}
+            \STATE Backpropagate $\nabla_{\mathbf{x}^\prime}$ through Equation \ref{eq-general}. Store $\mathbf{x}_t^\prime$.
+        \ENDFOR
+        \STATE Backpropagate $\nabla_{\theta}$ through Equation \ref{eq-obj}.
+    \ENDWHILE
+    \RETURN $\mathbf{M}_\theta$
+    \end{algorithmic}
+\end{algorithm}
 
@@
 
-As observed by Grathwohl et al.
-(\citeproc{ref-grathwohl2020your}{2020}), any classifier can be
-re-interpreted as a joint energy-based model (JEM) that learns to
-discriminate output classes conditional on the observed (training)
-samples from \(p(\mathbf{x})\) and the generated samples from
+Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
+any classifier can be re-interpreted as a joint energy-based model that
+learns to discriminate output classes conditional on the observed
+(training) samples from \(p(\mathbf{x})\) and the generated samples from
 \(p_\theta(\mathbf{x})\). The authors show that JEMs can be trained to
 perform well at both tasks by directly maximizing the joint
-log-likelihood factorized as
-\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
-The first term can be optimized using conventional cross-entropy as in
-Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\),
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimizes the
-contrastive divergence between the observed samples from
-\(p(\mathbf{x})\) and generated samples from \(p_\theta(\mathbf{x})\).
-
-A key empirical finding in Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}) was that JEMs tend to do
-well with respect to the plausibility objective in Def.
-\ref{def-explainability}. This follows directly if we consider samples
-drawn from \(p_\theta(\mathbf{x})\) as counterfactuals because the JEM
-objective effectively minimizes the divergence between the conditional
-posterior and \(p(\mathbf{x}|\mathbf{y}^{+})\). To generate samples,
-Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) relies on
-Stochastic Gradient Langevin Dynamics (SGLD) using an uninformative
-prior for initialization but we depart from their methodology. Instead
-of SGLD, we propose to use counterfactual explainers to generate
-counterfactuals of observed training samples. Specifically, we have:
-
+log-likelihood:
+\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\),
+where the first term can be optimized using cross-entropy as in
+Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\), they
+minimize the contrastive divergence between the observed samples from
+\(p(\mathbf{x})\) and samples generated from \(p_\theta(\mathbf{x})\).
+
+To generate samples, Grathwohl et al.
+(\citeproc{ref-grathwohl2020your}{2020}) use Stochastic Gradient
+Langevin Dynamics (SGLD) with an uninformative prior for initialization
+but we depart from their methodology: we propose to leverage
+counterfactual explainers to generate counterfactuals of observed
+training samples. Specifically, we have:
 \begin{equation}\phantomsection\label{eq-div}{
@@
 sampled factual from a non-target class generated with a gradient-based
-CE generator for up to \(T\) iterations. Mature counterfactuals are ones
-that have either reached convergence wrt. the decision threshold
-\(\tau\) or exhausted \(T\).
-
-Intuitively, the gradient of Equation~\ref{eq-div} decreases the energy
-of observed training samples (positive samples) while increasing the
-energy of counterfactuals (negative samples)
-(\citeproc{ref-du2019implicit}{Du and Mordatch 2020}). As the
-counterfactuals get more plausible (Def. \ref{def-explainability})
-during training, these opposing effects gradually balance each other out
-(\citeproc{ref-lippe2024uvadlc}{Lippe 2024}).
-
-The departure from SGLD of (\citeproc{ref-grathwohl2020your}{Grathwohl
-et al. 2020}) allows us to tap into the vast repertoire of explainers
-that have been proposed in the literature to meet different desiderata.
-For example, many methods facilitate the imposition of domain and
-mutability constraints. In principle, any existing approach for
-generating counterfactual explanations is viable, so long as it does not
+CE generator for up to \(T\) iterations. Intuitively, the gradient of
+Equation~\ref{eq-div} decreases the energy of observed training samples
+(positive samples) while increasing the energy of counterfactuals
+(negative samples) (\citeproc{ref-du2019implicit}{Du and Mordatch
+2020}). As the counterfactuals get more plausible (Def.
+\ref{def-explainability}) during training, these opposing effects
+gradually balance each other out (\citeproc{ref-lippe2024uvadlc}{Lippe
+2024}).
+
+Since maturity of counterfactuals in terms of a probability threshold is
+often reached before \(T\), this form of sampling is not only more
+closely aligned with Def. \ref{def-explainability}., but can also speed
+up training times compared to SGLD. The departure from SGLD also allows
+us to tap into the vast repertoire of explainers that have been proposed
+in the literature to meet different desiderata. For example, many
+methods support domain and mutability constraints. In principle, any
+existing approach for generating CEs is viable, so long as it does not
 violate the faithfulness condition. Like JEMs
-(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), CT can be
-considered a form of contrastive representation learning.
+(\citeproc{ref-murphy2022probabilistic}{Murphy 2022}), counterfactual
+training can be considered a form of contrastive representation
+learning.
 
@@
 \citeproc{ref-pawelczyk2022exploring}{Pawelczyk et al. 2022}). This
-leaves some flexibility wrt. the choice for \(\text{advloss}(\cdot)\) in
-Equation~\ref{eq-obj}. An intuitive functional form, but likely not the
-only sensible choice, is inspired by adversarial training:
-
-\begin{equation}\phantomsection\label{eq-adv}{
+leaves some flexibility with regards to the choice for the
+\(\text{advloss}(\cdot)\) term in Equation~\ref{eq-obj}. An intuitive
+functional form, but likely not the only sensible choice, is inspired by
+adversarial training: \begin{equation}\phantomsection\label{eq-adv}{
 \begin{aligned}
@@
 closely aligned with Szegedy et al.
-(\citeproc{ref-szegedy2013intriguing}{2014}) that defines an adversarial
-attack as an ``imperceptible non-random perturbation''. Thus, we choose
-to work with a different distinction between CE and AE than Freiesleben
-(\citeproc{ref-freiesleben2022intriguing}{2022}) that considers
-misclassification as the key distinguishing feature of AE. One of the
-key observations of this work is that we can leverage CEs during
-training and get adversarial examples essentially for free, which can be
-used to reap the aforementioned benefits of adversarial training.
+(\citeproc{ref-szegedy2013intriguing}{2014}) who define an adversarial
+attack as an ``imperceptible non-random perturbation''. Thus, we work
+with a different distinction between CE and AE than Freiesleben
+(\citeproc{ref-freiesleben2022intriguing}{2022}) who considers
+misclassification as the distinguishing feature of adversarial examples.
+One of the key observations of this work is that we can leverage CEs
+during training and get AEs essentially for free to reap the
+aforementioned benefits of adversarial training.
 
@@
 Many existing counterfactual explainers support domain and mutability
-constraints out-of-the-box. In fact, both types of constraints can be
-implemented for any counterfactual explainer that relies on gradient
-descent in the feature space for optimization
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, and Liem
-2023}). In this context, domain constraints can be imposed by simply
-projecting counterfactuals back to the specified domain, if the previous
-gradient step resulted in updated feature values that were
-out-of-domain. Mutability constraints can similarly be enforced by
+constraints. In fact, both types of constraints can be implemented for
+any explainer that relies on gradient descent in the feature space for
+optimization (\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen,
+and Liem 2023}). In this context, domain constraints can be imposed by
+simply projecting counterfactuals back to the specified domain, if the
+previous gradient step resulted in updated feature values that were
+out-of-domain. Similarly, mutability constraints can be enforced by
 setting partial derivatives to zero to ensure that features are only
@@
 
-Since such actionability constraints are binding at test time, we should
-also impose them when generating \(\mathbf{x}^\prime\) during each
-training iteration to inform model representations. Through their effect
-on \(\mathbf{x}^\prime\), both types of constraints influence model
+Since actionability constraints are binding at test time, we also impose
+them when generating \(\mathbf{x}^\prime\) during each training
+iteration to inform model representations. Through their effect on
+\(\mathbf{x}^\prime\), both types of constraints influence model
 outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
@@
 \item
-  We have \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but feature \(d\)
-  can only be decreased in practice.
-\item
-  We have \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but feature \(d\)
-  can only be increased in practice.
+  \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but \(d\) can only be
+  decreased in practice.
+\item
+  \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but \(d\) can only be
+  increased in practice.
 \end{enumerate}
@@
 \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
-assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
-feature \(d\). Intuitively, we think of this simply in terms ignoring
-implausibility costs with respect to immutable features, which
-effectively forces the model to instead seek plausibility with respect
-to the remaining features. This in turn results in lower overall
-sensitivity to immutable features, which we demonstrate empirically for
-different classifiers in Section~\ref{sec-experiments}. Under certain
-conditions, this result holds theoretically:\footnote{For the proof, see
-  the supplementary appendix.}
+assuming a point mass prior for \(p(\mathbf{x}^+)\) wrt. feature \(d\).
+Intuitively, we think of this as ignoring implausibility costs of
+immutable features, which effectively forces the model to instead seek
+plausibility through the remaining features. This can be expected to
+result in relatively lower sensitivity to immutable features; and higher
+relative sensitivity to mutable features should make
+mutability-constrained recourse less costly
+(Section~\ref{sec-experiments}). Under certain conditions, this result
+holds theoretically; for the proof, see the supplementary appendix:
 
@@
 Let $f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})$ denote a linear classifier with softmax activation $\mathcal{S}$ where $y\in\{1,...,K\}=\mathcal{K}$ and $\mathbf{x} \in \mathbb{R}^D$.
-If we assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
+Assume multivariate Gaussian class densities with common diagonal covariance matrix $\Sigma_k=\Sigma$ for all $k \in \mathcal{K}$, then protecting an immutable feature from the contrastive divergence penalty will result in lower classifier sensitivity to that feature relative to the remaining features, provided that at least one of those is discriminative and mutable.
 \end{proposition}
 
-\noindent It is worth highlighting that
-\mbox{Proposition \ref{prp-mtblty}} assumes independence of features.
-This raises a valid concern about the effect of protecting immutable
-features in the presence of proxies that remain unprotected. We address
-this in Section~\ref{sec-discussion}.
-
-\subsection{Example (Prediction of Consumer Credit
-Default)}\label{example-prediction-of-consumer-credit-default}
-
-Suppose we are interested in predicting the likelihood that loan
-applicants default on their credit. We have access to historical data on
-previous loan takers comprised of a binary outcome variable
-(\(y\in\{1=\text{default},2=\text{no default}\}\)) with two input
-features: (1) the subjects' \emph{age}, which we define as immutable,
-and (2) the subjects' existing level of \emph{debt}, which we define as
-mutable.
-
-We have simulated this scenario using synthetic data with independent
-\emph{age} and \emph{debt} features, and Gaussian class-conditional
-densities in Figure~\ref{fig-poc}. The four panels show the outcomes for
-different training procedures using the same model architectures (a
-linear classifier). In panels (a) and (c) we have trained the models
-conventionally, while in panels (b) and (d) we used CT.
-
-In all cases, all counterfactuals (stars) are valid---they have crossed
-the decision boundary (green)---but their quality differs. In panel (a),
-they are not plausible: they do not comply with the distribution of the
-factuals in \(y^+\) to the point where they form a clearly
-distinguishable cluster. In panel (b), they are highly plausible,
-meeting the first objective of Def. \ref{def-explainability}. In panel
-(c), the CEs involve substantial reductions in \emph{debt} for younger
-applicants. By comparison, counterfactual paths are shorter on average
-in panel (d) where we have protected the immutable \emph{age} as
-described in Section~\ref{sec-constraints}. Due to the classifier's
-lower sensitivity to \emph{age}, recommendations with respect to
-\emph{debt} are much more homogenous and do not unfairly punish younger
-individuals. These counterfactuals are also plausible with respect to
-the mutable feature. Thus, we consider the model in panel (d) as the
-most explainable according to Def. \ref{def-explainability}.
-
-\begin{figure*}
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
-
-}
-
-\caption{\label{fig-poc}Illustration of how CT improves model
-explainability: (a) conventional training, all mutable; (b) CT, all
-mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
-immutable. The linear decision boundary is shown in green along with
-training data colored according to their ground-truth label: \(y^-=1\)
-in blue and \(y^+=2\) in orange. Stars indicate counterfactuals in the
-target class.}
-
-\end{figure*}%
-
 \section{Experiments}\label{sec-experiments}
 
-In our experiments we seek to answer the following three research
-questions:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  To what extent does the counterfactual training objective as it is
-  defined in Equation~\ref{eq-obj} induce models to learn plausible
-  explanations?
-\item
-  To what extent does the CT objective produce more favorable
-  algorithmic recourse outcomes in the presence of actionability
-  constraints?
-\item
-  What are the effects of hyperparameter selection wrt. the CT
-  objective?
+We seek to answer the following four research questions:
+
+\begin{enumerate}[label={(\makebox[2em][c]{RQ\arabic*})}, leftmargin=3.5em]
+    \item To what extent does the CT objective in Equation 1 induce models to learn plausible explanations?
+    \item To what extent does CT result in more favorable algorithmic recourse outcomes in the presence of actionability constraints?
+    \item To what extent does CT influence the adversarial robustness of trained models?
+    \item What are the effects of hyperparameter selection on counterfactual training?
 \end{enumerate}
@@
 
-Our key outcome of interest is improvement in explainability (Def.
-\ref{def-explainability}). To this end, we focus primarily on the
-plausibility and cost of faithfully generated counterfactuals at test
-time. To measure the cost, we follow the standard convention of using
-distances (\(\ell_1\)-norm) between factuals and counterfactuals as a
-proxy. For plausibility, we assess how similar CEs are to the observed
-samples in the target domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We
-rely on the distance-based metric used in Altmeyer et al.
-(\citeproc{ref-altmeyer2024faithful}{2024}),
-
+Our focus is the improvement in explainability (Def.
+\ref{def-explainability}). Thus, we primarily look at the plausibility
+and cost of faithfully generated counterfactuals at test time. Other
+metrics, such as validity and redundancy, are reported in the
+supplementary appendix. To measure the cost, we follow the standard
+proxy of distances (\(\ell_1\)-norm) between factuals and
+counterfactuals. For plausibility, we assess how similar CEs are to
+observed samples in the target domain,
+\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the metric used by
+Altmeyer et al. (\citeproc{ref-altmeyer2024faithful}{2024}),
 \begin{equation}\phantomsection\label{eq-impl-dist}{
 \text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
-}\end{equation} and introduce a novel divergence metric,
-
+}\end{equation} and introduce a novel divergence-based adaptation,
 \begin{equation}\phantomsection\label{eq-impl-div}{
@@
 }\end{equation} where \(\mathbf{X}^\prime\) denotes a collection of
-counterfactuals and \(\text{MMD}(\cdot)\) is an unbiased estimate of the
-squared population maximum mean discrepancy
-(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). The metric in
-Equation~\ref{eq-impl-div} is equal to zero iff the two distributions
-are the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
-
-In addition to cost and plausibility, we compute other standard metrics
-to evaluate counterfactuals including validity and redundancy. Finally,
-we also assess the predictive performance of models using standard
-metrics, including robust accuracy estimated on adversarially perturbed
-data using FGSM (\citeproc{ref-goodfellow2014explaining}{Goodfellow,
-Shlens, and Szegedy 2015}).
-
-We run the experiments with three gradient-based generators:
-\emph{Generic} of Wachter, Mittelstadt, and Russell
+counterfactuals and \(\text{MMD}(\cdot)\) is the unbiased estimate of
+the squared population maximum mean discrepancy, proposed by Gretton et
+al. (\citeproc{ref-gretton2012kernel}{2012}). The metric in
+Equation~\ref{eq-impl-div} is equal to zero if and only if the two
+distributions are exactly the same, \(\mathbf{X}^\prime=\mathbf{X}^+\).
+
+To assess outcomes with respect to actionability for non-linear models,
+we look at the average costs of valid counterfactuals in terms of their
+distances from factual starting points. While this an imperfect proxy of
+sensitivity, we hypothesize that CT can reduce these costs by teaching
+models to seek plausibility with respect to mutable features, much like
+we observe in Figure~\ref{fig-poc} in panel (d) compared to (c). We
+supplement this analysis with qualitative findings for integrated
+gradients (\citeproc{ref-sundararajan2017ig}{Sundararajan, Taly, and Yan
+2017}). Finally, for predictive performance, we use standard metrics,
+such as robust accuracy estimated on adversarially perturbed data using
+FGSM (\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and
+Szegedy 2015}).
+
+We run experiments with three gradient-based generators: \emph{Generic}
+of Wachter, Mittelstadt, and Russell
 (\citeproc{ref-wachter2017counterfactual}{2017}) as a simple baseline
@@
 2019}) that aims to generate plausible counterfactuals using a surrogate
-Variational Autoencoder (VAE), and \emph{ECCo}---the generator of
-Altmeyer et al. (\citeproc{ref-altmeyer2024faithful}{2024}) but without
-the conformal prediction component---as a method that directly targets
-both faithfulness and plausibility of the counterfactuals.
+Variational Autoencoder (VAE), and \emph{ECCCo}
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}), which
+targets faithfulness.
 
@@
 clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
-moons (\emph{Moon}). These datasets are generated using the library of
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, and Liem
-2023}) and we present them in the supplementary appendix. Next, we have
-four real-world binary tabular datasets from the domain of economics:
-\emph{Adult} (a.k.a. Census data) of
-(\citeproc{ref-becker1996adult2}{Becker and Kohavi 1996}), California
-housing (\emph{CH}) of (\citeproc{ref-pace1997sparse}{Pace and Barry
-1997}), Default of Credit Card Clients (\emph{Cred}) of
-(\citeproc{ref-yeh2016default}{Yeh 2016}), and Give Me Some Credit
-(\emph{GMSC}) of (\citeproc{ref-kaggle2011give2}{Kaggle 2011}). Finally,
-for the convenience of illustration, we use of the 10-class \emph{MNIST}
-vision dataset (\citeproc{ref-lecun1998mnist}{LeCun 1998}).
+moons (\emph{Moon}). Next, we have four real-world binary tabular
+datasets: \emph{Adult} (Census data) of Becker and Kohavi
+(\citeproc{ref-becker1996adult2}{1996}), California housing (\emph{CH})
+of Pace and Barry (\citeproc{ref-pace1997sparse}{1997}), Default of
+Credit Card Clients (\emph{Cred}) of Yeh
+(\citeproc{ref-yeh2016default}{2016}), and Give Me Some Credit
+(\emph{GMSC}) from Kaggle (\citeproc{ref-kaggle2011give}{2011}).
+Finally, for the convenience of illustration, we use the 10-class
+\emph{MNIST} (\citeproc{ref-lecun1998mnist}{LeCun 1998}).
 
@@
 
-\subsubsection{Plausibility}\label{sec-plaus}
-
-Table~\ref{tbl-main} presents our main empirical findings. For all
-datasets except \emph{OL} and across all test settings, the average
-distance of CEs from observed samples in the target class is reduced,
-indicating improved plausibility. The magnitude of improvements varies.
-For the simple synthetic datasets, distance reductions range from around
-20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
-real-world tabular datasets, improvements tend to be smaller but still
-substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
-7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For the vision
-dataset (\emph{MNIST}), distances are reduced by up to 9\%. The results
-for our proposed divergence metric are qualitatively similar, but
-generally even more pronounced: for the \emph{Circ} dataset,
-implausibility is reduced by almost 94\% to virtually zero as we
-verified by the absolute outcome. Improvements for other datasets range
-from 28\% (\emph{Moon}) up to 78\% (\emph{GMSC}). For \emph{OL} the
-reduction is negative, consistent with the distance-based metric.
-\emph{MNIST} is the only dataset for which the distance and divergence
-metrics disagree. Upon visual inspection of the image counterfactuals we
-find that CT clearly improves plausibility (see supplementary appendix
-for images).
-
-\subsubsection{Predictive Performance}\label{sec-pred}
-
-Test accuracy for CT is virtually identical to the baseline for
-\emph{Adult}, \emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and
-even slightly improved for \emph{Cred}. Exceptions to this general
-pattern are \emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we
-observe a reduction in test accuracy of 2, 5, and 15 percentage points
-respectively. When looking at robust test accuracies (Acc.\(^*\)) for
-these datasets in particular, we find that CT strongly outperforms the
-baseline. In fact, we find that CT improves adversarial robustness on
-all datasets.
-
-\begin{table*}
-
-\caption{\label{tbl-main}Key performance metrics across all datasets
-(column 1). \textbf{Plausibility}: Columns 2-6 show the percentage
-reduction in implausibility (\(\text{IP}\)) for varying degrees of the
-energy penalty \(\lambda_{\text{egy}}\) used for \emph{ECCo} at test
-time; column 7 shows the reduction in \(\text{IP}^*\) (\(\text{MMD}\)),
-aggregated across all test specifications. \textbf{Accuracy} (columns
-8-11): test accuracies and robust accuracies (\(\text{Acc}^*\)) for CT
-and the baseline (BL). \textbf{Actionability} (column 12): average
-reduction in costs when imposing mutability constraints reported for the
-four datasets for which we could identify meaningful features to
-protect.}
-
-\centering{
-
-\input{tables/main.tex}
-
-}
-
-\end{table*}%
-
-\subsubsection{Actionability}\label{sec-act}
-
-In Section~\ref{sec-method}, we show that our proposed way for encoding
-mutability constraints leads to lower classifier sensitivity wrt.
-immutable features for linear models, tilting the decision boundary in
-favour of mutable features instead. For binding constraints at test
-time, this leads to shorter counterfactual paths and hence smaller
-average costs (\(\ell_1\)-norm) to individuals. To extend this to the
-non-linear case, we test the effect of imposing mutability constraints
-empirically for our synthetic data using the same evaluation scheme as
-above. The final row in Table~\ref{tbl-main} reports the average
-reduction in costs for CT compared to the ``vanilla'' baseline, when
-imposing that either the first or the second feature is immutable. In
-all cases, costs are reduced substantially, indicating that classifiers
-trained with CT are indeed more sensitive to mutable features.
-
-\subsubsection{Impact of hyperparameter
-settings.}\label{sec-hyperparameters}
-
-We test the impact of three types of hyperparameters; our complete
-results are in the supplementary appendix.\\
-\indent We note that CT is highly sensitive to the choice of a CE
-generator and its hyperparameters but (a) there are manageable patterns
-and (b) we can typically identify settings that improve either
-plausibility or cost, and commonly both of them at the same time. For
-example, \emph{REVISE} tends to perform the worst, most likely because
-it uses a surrogate VAE to generate counterfactuals which impedes
-faithfulness (\citeproc{ref-altmeyer2024faithful}{Altmeyer et al.
-2024}). Increasing \(T\), the maximum number of steps, generally yields
-better outcomes because more CEs can mature in each training epoch. The
-impact of \(\tau\), the required decision threshold is more difficult to
-predict. On ``harder'' datasets it may be difficult to satisfy high
-\(\tau\) for any given sample (i.e., also factuals) and so increasing
-this threshold does not seem to correlate with better outcomes. In fact,
-the choice of \(\tau=0.5\) generally leads to optimal results because it
-is associated with high proportions of mature counterfactuals.\\
-\indent The strength of the energy regularization,
-\(\lambda_{\text{reg}}\) is highly impactful and leads to poor
-performance in terms of decreased plausibility and increased costs if
-insufficiently high. The sensitivity with respect to
-\(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\) is much less
-evident. While high values of \(\lambda_{\text{reg}}\) may increase the
-variability in outcomes when combined with high values of
-\(\lambda_{\text{div}}\) or \(\lambda_{\text{adv}}\), this effect is not
-very pronounced.\\
-\indent The effectiveness and stability of CT is positively associated
-with the number of counterfactuals generated during each training epoch.
-We also confirm that a higher number of training epochs is beneficial.
-Interestingly, we observed desired improvements when CT was combined
-with conventional training and applied only for the final 50\% of epochs
-of the complete training process. Put differently, CT can improve the
-explainability of models in a fine-tuning manner.
-
-\section{Conclusions}\label{sec-discussion}
+Our main results for plausibility and actionability for \emph{MLP}
+models are summarised in Table~\ref{tbl-main} that presents
+counterfactual outcomes grouped by dataset along with standard errors
+averaged across bootstrap samples. Asterisks (\(^*\)) are used when the
+bootstrapped 99\%-confidence interval of differences in mean outcomes
+does \emph{not} include zero, so the observed effects are statistically
+significant at the 0.01 level.
+
+The first two columns (\(\text{IP}\) and \(\text{IP}^*\)) show the
+percentage reduction in implausibility for our two metrics when using CT
+on top of the weak baseline. As an example, consider the first row for
+\emph{LS} data: the observed positive values indicate that faithful
+counterfactuals are around 30-55\% more plausible for models trained
+with CT, in line with our observations in panel (b) of
+Figure~\ref{fig-poc} compared to panel (a).
+
+The third column shows the results for a scenario when mutability
+constraints are imposed on the selected features. Again, we are
+comparing CT to the baseline, so reductions in the positive direction
+imply that valid counterfactuals are ``cheaper'' (more actionable) when
+using CT with feature protection. Relating this back to
+Figure~\ref{fig-poc}, the third column represents the reduction in
+distances travelled by counterfactuals in panel (d) compared to panel
+(c). In the following paragraphs, we summarize the results for all
+datasets.
+
+\begin{table}
+
+\caption{\label{tbl-main}Key evaluation metrics for valid counterfactual
+along with bootstrapped standard errors for all datasets.
+\textbf{Plausibility} (columns 1-2): percentage reduction in
+implausibility for \(\text{IP}\) and \(\text{IP}^*\), respectively;
+\textbf{Cost} / \textbf{Actionability} (column 3): percentage reduction
+in costs when selected features are protected. Outcomes are aggregated
+across bootstrap samples (100 rounds) and varying degrees of the energy
+penalty \(\lambda_{\text{egy}}\) used for \emph{ECCCo} at test time.
+Asterisks (\(^*\)) indicate that the bootstrapped 99\%-confidence
+interval of differences in mean outcomes does \emph{not} include zero.}
+
+\centering{
+
+\small
+\centering
+\begin{tabular}{
+  l
+  S[table-format=2.2(1.2)]
+  S[table-format=3.2(3.2)]
+  S[table-format=3.2(1.2)]
+}
+  \toprule
+  \textbf{Data} & \textbf{$ \text{IP} $ $(-\%)$} & \textbf{$ \text{IP}^* $ $(-\%)$} & \textbf{Cost $(-\%)$} \\\midrule
+  LS & 29.05\pm0.67 $^{*}$ & 55.33\pm2.03 $^{*}$ & 14.07\pm0.6 $^{*}$ \\
+  Circ & 56.29\pm0.44 $^{*}$ & 89.38\pm9.3 $^{*}$ & 45.55\pm0.76 $^{*}$ \\
+  Moon & 20.62\pm0.69 $^{*}$ & 19.26\pm8.12 $^{*}$ & 2.86\pm1.03 $^{*}$ \\
+  OL & -1.13\pm0.88 $^{}$ & -24.52\pm14.52 $^{}$ & 38.39\pm2.21 $^{*}$ \\\midrule
+  Adult & 0.77\pm1.34 $^{}$ & 32.29\pm6.87 $^{*}$ & -2.82\pm4.88 $^{}$ \\
+  CH & 12.05\pm1.41 $^{*}$ & 70.27\pm3.72 $^{*}$ & 40.71\pm1.55 $^{*}$ \\
+  Cred & 12.31\pm1.84 $^{*}$ & 54.89\pm11.21 $^{*}$ & -17.43\pm5.17 $^{*}$ \\
+  GMSC & 23.44\pm1.99 $^{*}$ & 73.31\pm4.83 $^{*}$ & 62.64\pm2.04 $^{*}$ \\
+  MNIST & 7.05\pm1.8 $^{*}$ & -25.09\pm109.05 $^{}$ & -12.34\pm6.52 $^{}$ \\\midrule
+  Avg. & 17.83 & 38.35 & 19.07 \\\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\subsubsection{Plausibility (RQ1).}\label{sec-plaus}
+
+\emph{CT generally produces substantial and statistically significant
+improvements in plausibility.}
+
+Average reductions in \(\text{IP}\) range from around 7\% for
+\emph{MNIST} to almost 60\% for \emph{Circ}. For the real-world tabular
+datasets they are around 12\% for \emph{CH} and \emph{Cred} and almost
+25\% for \emph{GMSC}; for \emph{Adult} and \emph{OL} we find no
+significant impact of CT on \(\text{IP}\). Reductions in \(\text{IP}^*\)
+are even more substantial and generally statistically significant,
+although the average degree of uncertainty is higher than for
+\(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
+almost 90\% (\emph{Circ}). The only negative findings are for OL and
+MNIST, but they are not statistically significant. A qualitative
+inspection of the counterfactuals in Figure~\ref{fig-mnist} (columns
+2-5) suggests recognizable digits 1-4 for the model trained with CT
+(bottom row), unlike the baseline (top row).
+
+\begin{figure}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_body.png}}
+
+}
+
+\caption{\label{fig-mnist}Visual explanations for \emph{MNIST} for BL
+(top) and CT (bottom). \textbf{Plausibility}: col.~1 is a random factual
+0 (blue); cols. 2-5 are corresponding \emph{ECCCo} counterfactuals in
+target classes 1 to 4. \textbf{Actionability}: cols. 6-10 show
+integrated gradients averaged over test images in classes 5 to 9.}
+
+\end{figure}%
+
+\subsubsection{Actionability (RQ2).}\label{sec-act}
+
+\emph{CT tends to improve actionability in the presence of immutable
+features, but this is not guaranteed if the assumptions in Proposition
+\ref{prp-mtblty} are violated.}
+
+For synthetic datasets, we always protect the first feature; for all
+real-world tabular datasets we could identify and protect an \emph{age}
+variable; for \emph{MNIST}, we protect the five upper and lower pixel
+rows of the full image. Statistically significant reductions in costs
+overwhelmingly point in the expected positive direction reaching up to
+around 60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average
+costs increase, likely because any potential benefits from protecting
+the \emph{age} are outweighed by the increase in costs required for
+greater plausibility. The findings for \emph{Adult} and \emph{MNIST} are
+not significant. A qualitative inspection of the class-conditional
+integrated gradients in Figure~\ref{fig-mnist} (columns 6-10) suggests
+that CT still has the expected effect: the model (bottom) is insensitive
+(blue) to the protected rows of pixels; details of this experiment are
+reported in the supplementary appendix.
+
+\begin{figure*}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/acc.png}}
+
+}
+
+\caption{\label{fig-acc}Test accuracies on adversarially perturbed data
+with varying perturbation sizes for all non-synthetic datasets.}
+
+\end{figure*}%
+
+\subsubsection{Predictive Performance (RQ3).}\label{sec-pred}
+
+\emph{Models trained with CT are substantially more robust to
+gradient-based adversarial attacks than conventionally-trained
+baselines.}
+
+Test accuracies on adversarially perturbed data are shown in
+Figure~\ref{fig-acc}. The perturbations size, \(\varepsilon\in[0,0.1]\),
+increases along the horizontal axis and includes zero, corresponding to
+standard test accuracy for non-perturbed data. For all synthetic
+datasets, predictive performance of CT is virtually identical to the
+baseline and unaffected by perturbations. For all real-world datasets,
+we find that CT substantially improves robustness: while in some cases
+baseline accuracies drop to essentially zero for large enough
+perturbation sizes, accuracies of CT models remain remarkably robust.
+
+\subsubsection{Hyperparameter settings
+(RQ4).}\label{sec-hyperparameters}
+
+\emph{CT is highly sensitive to the choice of a CE generator and its
+hyperparameters but (1) we observe manageable patterns, and (2) we can
+usually identify settings that improve either plausibility or
+actionability, and typically both of them at the same time.}
+
+We evaluate the impacts of three types of hyperparameters on CT. In this
+section we focus on the highlights and make the full results available
+in the supplementary appendix.
+
+Firstly, we find that optimal results are generally obtained when using
+\emph{ECCCo} to generate counterfactuals. Conversely, using a generator
+that may inhibit faithfulness (\emph{REVISE}), tends to yield poor
+results. Concerning hyperparameters that guide the gradient-based
+counterfactual search, we find that increasing \(T\), the maximum number
+of steps, generally yields better outcomes because more CEs can mature.
+Relatedly, we also find that the effectiveness and stability of CT is
+positively associated with the total number of counterfactuals generated
+during each training epoch. The impact of \(\tau\), the decision
+threshold, is more difficult to predict. On ``harder'' datasets it may
+be difficult to satisfy high \(\tau\) for any given sample (i.e., also
+factuals) and so increasing this threshold does not seem to correlate
+with better outcomes. In fact, \(\tau=0.5\) generally leads to optimal
+results as it is associated with high proportions of mature
+counterfactuals.
+
+Secondly, the strength of the energy regularization,
+\(\lambda_{\text{reg}}\) is highly impactful and should be set
+sufficiently high to avoid common problems associated with exploding
+gradients. The sensitivity with respect to \(\lambda_{\text{div}}\) and
+\(\lambda_{\text{adv}}\) is much less evident. While high values of
+\(\lambda_{\text{reg}}\) may increase the variability in outcomes when
+combined with high values of \(\lambda_{\text{div}}\) or
+\(\lambda_{\text{adv}}\), this effect is not particularly pronounced.
+
+Finally, we also observe desired improvements when CT was combined with
+conventional training and applied only for the final 50\% of epochs of
+the complete training process. Put differently, CT can improve the
+explainability of models in a post-hoc, fine-tuning manner.
+
+\section{Discussion}\label{sec-discussion}
 
 As our results indicate, counterfactual training produces models that
-are more explainable. Nonetheless, it brings about three important
-limitations.\\
-\indent \emph{CT increases the training time of models.} CT can be more
-time-consuming than conventional training regimes. While higher numbers
-of CEs per iteration positively impact the quality of solutions, they
-also increase the amount of computations. Relatively small grids with
-270 settings can take almost four hours for more demanding datasets on a
-high-performance computing cluster with 34 2GB CPUs.\footnote{See
-  supplementary appendix for computational details.} Three factors
-attenuate this effect: (1) CT amortizes the cost of CEs for the training
-samples; (2) it can retain its value when used as a ``fine-tuning''
-technique for conventionally-trained models; and (3) it yields itself to
-parallel execution, which we have leveraged for our own experiments.\\
-\indent \emph{Immutable features may have proxies.} We propose an
-approach to protect immutable features and thus increase the
-actionability of the generated CEs. However, it requires that model
-owners define the mutability constraints for (all) features considered
-by the model. Even if all immutable features are protected, there may
-exist proxies that are mutable (and hence should not be protected) but
-preserve enough information about the principals to hinder the
-protections. Delineating actionability is a major open challenge in the
-AR literature (see, e.g.,
+are more explainable. Nonetheless, these advantages come at the cost of
+two important limitations.
+
+\emph{Interventions on features have implications for fairness.} We
+provide a method to modify the sensitivity of a model to certain
+features, which can be misused by enforcing explanations based on
+features that are more difficult to modify by a (group of) decision
+subjects. Such abuse could result in an unfairly assigned burden of
+recourse (\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh
+2020}), threatening the equality of opportunity
+(\citeproc{ref-bell2024fairness}{Bell et al. 2024}). Also, even if all
+immutable features are protected, there may exist proxies that are
+theoretically mutable, but preserve sufficient information about the
+principals to hinder these protections. Indeed, deciding on the
+actionability of features remains a major open challenge in the AR
+literature
 (\citeproc{ref-venkatasubramanian2020philosophical}{Venkatasubramanian
-and Alfano 2020})) impacting the capacity of CT to fulfill its intended
-goal.\\
-\indent \emph{Interventions on features may impact fairness.} We provide
-a tool that allows practitioners to modify the sensitivity of a model
-with respect to certain features, which may have implication for the
-fair and equitable treatment of decision subjects. Model owners could
-misuse this solution by enforcing explanations based on features that
-are more difficult to modify by some (group of) individuals. For
-example, consider the \emph{Adult} dataset used in our experiments,
-where \emph{workclass} or \emph{education} may be more difficult to
-change for underpriviledged groups. When applied irresponsibly, CT could
-result in an unfairly assigned burden of recourse
-(\citeproc{ref-sharma2020certifai}{Sharma, Henderson, and Ghosh 2020}),
-threatening the equality of opportunity in the system
-(\citeproc{ref-bell2024fairness}{Bell et al. 2024}). Nonetheless, these
-phenomena are not specific to CT.\\
-\hfill \break \indent We also highlight several important directions for
-future research. Firstly, it is an interesting challenge to extend CT
-beyond classification settings. Our formulation relies on the
-distinction between non-target class(es) \(y^{-}\) and target class(es)
-\(y^{+}\) to generate counterfactuals through Equation~\ref{eq-obj}.
-While \(y^{-}\) and \(y^{+}\) can be arbitrarily defined, CT requires
-the output space \(\mathcal{Y}\) to be discrete. Thus, it does not apply
-to ML tasks where the change in outcome cannot be readily quantified.
-Focus on classification models is a common restriction in research on
-CEs and AR. Other settings have attracted some interest (e.g.,
-regression in (\citeproc{ref-spooner2021counterfactual}{Spooner et al.
-2021})), but there is little consensus how to robustly extend the notion
-of CEs.\\
-\indent Secondly, our approach is susceptible to training instabilities.
-This problem has been recognized for JEMs
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020}) and even
-though we depart from the SGLD-based sampling, we still encounter
-considerable variability in the outcomes. CT is exposed to two potential
-sources of instabilities: (1) the energy-based contrastive divergence
-term in Equation~\ref{eq-div}, and (2) the underlying counterfactual
-explainers. We find several promising ways to mitigate this problem:
-regularizing energy (\(\lambda_{\text{reg}}\)), generating sufficiently
-many counterfactuals during each epoch, and including only mature
-counterfactuals for contrastive divergence.\\
-\indent Finally, we believe that it is possible to substantially improve
-hyperparameter selection procedures. Our method benefits from the tuning
-of certain key hyperparameters (see Section~\ref{sec-hyperparameters}).
-In this work, we have relied exclusively on grid search for this task.
-Future work on CT could benefit from investigating more sophisticated
-approaches. Notably, CT is iterative which makes methods such as
-Bayesian or gradient-based optimization applicable (see, e.g.,
-(\citeproc{ref-bischl2023hyperparameter}{Bischl et al. 2023})).
-\hfill \break
-
-To conclude, state-of-the-art machine learning models are prone to
-learning complex representations that cannot be interpreted by humans
-and existing post-hoc explainability approaches cannot guarantee that
-the explanations agree with the model's learned representation of data.
-As a step towards addressing this challenge, we introduced
-counterfactual training, a novel training regime that incentivizes
-highly-explainable models. Our approach leads to explanations that are
-both plausible---compliant with the underlying data-generating
-process---and actionable---compliant with user-specified mutability
-constraints---and thus meaningful to their recipients. Through extensive
-experiments we demonstrate that CT satisfies its objective while
-promoting robustness and preserving the predictive performance of the
-models. It can also be used to fine-tune conventionally-trained models
-and achieve similar gains. Finally, this work showcases that it is
-practical to improve models \emph{and} their explanations at the same
-time.
+and Alfano 2020}).
+
+\emph{Plausibility is costly.} As noted by Altmeyer et al.
+(\citeproc{ref-altmeyer2024faithful}{2024}), more plausible
+counterfactuals are inevitably more costly. CT improves plausibility and
+robustness, but it can impact average costs and validity when cheap,
+implausible and adversarial explanations are removed from the solution
+space.
+
+\emph{CT increases the training times.} Just like contrastive and robust
+learning, CT is more resource-intensive than conventional regimes. Three
+factors mitigate this effect: (1) CT yields itself to parallel
+execution; (2) it amortizes the cost of CEs for the training samples;
+and (3) it can be used to fine-tune conventionally-trained models.
+
+We also highlight three key directions for future research. Firstly, it
+is an interesting challenge to extend CT beyond classification settings.
+Our formulation relies on the distinction between non-target class(es)
+and target class(es), requiring the output space to be discrete. Thus,
+it does not apply to ML tasks where the change in outcome cannot be
+readily discretized. Focus on classification is a common choice in
+research on CEs and AR; other settings have attracted some interest,
+e.g., regression (\citeproc{ref-spooner2021counterfactual}{Spooner et
+al. 2021}), but there is little consensus how to robustly extend the
+notion of CEs.
+
+Secondly, our analysis covers CE generators with different
+characteristics, but it is interesting to extend it to more algorithms,
+including ones that do not rely on computationally costly gradient-based
+optimization. This should reduce training costs while possibly
+preserving the benefits of CT.
+
+Finally, we believe that it is possible to considerably improve
+hyperparameter selection procedures, and thus performance. We have
+relied exclusively on grid searches, but future work could benefit from
+more sophisticated approaches.
+
+\section{Conclusion}\label{sec-conclusion}
+
+State-of-the-art machine learning models are prone to learning complex
+representations that cannot be interpreted by humans. Existing
+explainability solutions cannot guarantee that explanations agree with
+these learned representation. As a step towards addressing this
+challenge, we introduce counterfactual training, a novel training regime
+that integrates recent advances in contrastive learning, adversarial
+robustness, and counterfactual explanations to incentivize
+highly-explainable models. Through extensive experiments, we demonstrate
+that CT satisfies this goal while preserving the predictive performance
+and promoting robustness of models. Explanations generated from CT-based
+models are both more plausible (compliant with the underlying
+data-generating process) and more actionable (compliant with
+user-specified mutability constraints), and thus meaningful to their
+recipients. In turn, our work highlights the value of simultaneously
+improving models and their explanations.
 
@@
 
-\bibitem[\citeproctext]{ref-bischl2023hyperparameter}
-Bischl, Bernd, Martin Binder, Michel Lang, Tobias Pielok, Jakob Richter,
-Stefan Coors, Janek Thomas, et al. 2023. {``{Hyperparameter
-optimization: Foundations, algorithms, best practices, and open
-challenges}.''} \emph{WIREs Data Mining and Knowledge Discovery} 13 (2):
-e1484. https://doi.org/\url{https://doi.org/10.1002/widm.1484}.
-
 \bibitem[\citeproctext]{ref-milan2023dataframes}
@@
 
-\bibitem[\citeproctext]{ref-frankle2018lottery}
-Frankle, Jonathan, and Michael Carbin. 2019. {``{The Lottery Ticket
-Hypothesis: Finding Sparse, Trainable Neural Networks}.''} In
-\emph{International Conference on Learning Representations}.
-
 \bibitem[\citeproctext]{ref-freiesleben2022intriguing}
@@
 Journal of Machine Learning Research} 13 (1): 723--73.
-
-\bibitem[\citeproctext]{ref-guidotti2022counterfactual}
-Guidotti, Riccardo. 2022. {``{Counterfactual Explanations and How to
-Find Them: Literature Review and Benchmarking}.''} \emph{Data Mining and
-Knowledge Discovery} 38 (5): 2770--2824.
-\url{https://doi.org/10.1007/s10618-022-00831-6}.
 
@@
 
-\bibitem[\citeproctext]{ref-kaggle2011give2}
+\bibitem[\citeproctext]{ref-kaggle2011give}
 Kaggle. 2011. {``Give Me Some Credit, {Improve} on the State of the Art
@@
 https://www.kaggle.com/c/GiveMeSomeCredit; {Kaggle}.
-
-\bibitem[\citeproctext]{ref-kolter2023keynote}
-Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
-IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)}. Los
-Alamitos, CA, USA: IEEE Computer Society.
-\url{https://doi.org/10.1109/SaTML54575.2023.00009}.
+\url{https://www.kaggle.com/c/GiveMeSomeCredit}.
+
+\bibitem[\citeproctext]{ref-karimi2020survey}
+Karimi, Amir-Hossein, Gilles Barthe, Bernhard Schölkopf, and Isabel
+Valera. 2021. {``A Survey of Algorithmic Recourse: Definitions,
+Formulations, Solutions, and Prospects.''}
+\url{https://arxiv.org/abs/2010.04050}.
 
@@
 
+\bibitem[\citeproctext]{ref-molnar2022interpretable}
+Molnar, Christoph. 2022. \emph{Interpretable Machine Learning: A Guide
+for Making Black Box Models Explainable}. 2nd ed.
+\url{https://christophm.github.io/interpretable-ml-book}.
+
 \bibitem[\citeproctext]{ref-murphy2022probabilistic}
@@
 Introduction}. {MIT Press}.
-
-\bibitem[\citeproctext]{ref-oneil2016weapons}
-O'Neil, Cathy. 2016. \emph{Weapons of Math Destruction: {How} Big Data
-Increases Inequality and Threatens Democracy}. {Crown}.
 
@@
 
-\bibitem[\citeproctext]{ref-poyiadzi2020face}
-Poyiadzi, Rafael, Kacper Sokol, Raul Santos-Rodriguez, Tijl De Bie, and
-Peter Flach. 2020. {``{FACE}: {Feasible} and Actionable Counterfactual
-Explanations.''} In \emph{Proceedings of the {AAAI}/{ACM Conference} on
-{AI}, {Ethics}, and {Society}}, 344--50.
-
 \bibitem[\citeproctext]{ref-ross2021learning}
@@
 
+\bibitem[\citeproctext]{ref-sundararajan2017ig}
+Sundararajan, Mukund, Ankur Taly, and Qiqi Yan. 2017. {``Axiomatic
+Attribution for Deep Networks.''}
+\url{https://arxiv.org/abs/1703.01365}.
+
 \bibitem[\citeproctext]{ref-szegedy2013intriguing}
@@
 
+\bibitem[\citeproctext]{ref-ustun2019actionable}
+Ustun, Berk, Alexander Spangher, and Yang Liu. 2019. {``Actionable
+Recourse in Linear Classification.''} In \emph{Proceedings of the
+{Conference} on {Fairness}, {Accountability}, and {Transparency}},
+10--19. \url{https://doi.org/10.1145/3287560.3287566}.
+
 \bibitem[\citeproctext]{ref-venkatasubramanian2020philosophical}
@@
 
+\bibitem[\citeproctext]{ref-verma2020counterfactual}
+Verma, Sahil, Varich Boonsanong, Minh Hoang, Keegan E. Hines, John P.
+Dickerson, and Chirag Shah. 2022. {``Counterfactual Explanations and
+Algorithmic Recourses for Machine Learning: A Review.''}
+\url{https://arxiv.org/abs/2010.10596}.
+
 \bibitem[\citeproctext]{ref-wachter2017counterfactual}
@@
 
-\bibitem[\citeproctext]{ref-wu2021polyjuice2}
+\bibitem[\citeproctext]{ref-wu2021polyjuice}
 Wu, Tongshuang, Marco Tulio Ribeiro, Jeffrey Heer, and Daniel Weld.
@@
 11th International Joint Conference on Natural Language Processing
-(Volume 1: Long Papers)}, 6707--23. ACL.
+(Volume 1: Long Papers)}, edited by Chengqing Zong, Fei Xia, Wenjie Li,
+and Roberto Navigli, 6707--23. Online: ACL.
 \url{https://doi.org/10.18653/v1/2021.acl-long.523}.
@@
 \begin{appendices}
-
-\FloatBarrier
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-default}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-train-default}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{table*}
 \begin{tabular}{cccccccc}
@@
 Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
+  LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
+  Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
+  Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
+  OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\midrule
   Adult & 26049 & 5010 & 1000 & none & 0.75 & 5000 & 0.25 \\
   CH & 16504 & 3101 & 1000 & none & 0.5 & 5000 & 0.25 \\
-    Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
   Cred & 10617 & 1923 & 1000 & none & 0.5 & 5000 & 0.25 \\
   GMSC & 13371 & 2474 & 1000 & none & 0.5 & 5000 & 0.5 \\
-    LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
   MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\
-    Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
-    OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\bottomrule
 \end{tabular}
-\end{table*}
+
+}
+
+\end{table}%
+
+\subsection{Final Results}\label{sec-app-final-results}
+
+Plus/minus two standard deviations of bootstrap estimates.
+
+\subsubsection{Robust Performance Plots}\label{robust-performance-plots}
+
+\subsection{Confidence Intervals}\label{confidence-intervals}
+
+\begin{table}
+
+\caption{\label{tbl-ci}Mean outcomes for \textbf{CT} and \textbf{BL}
+along with bootstrapped confidence intervals (99\%) for difference in
+mean outcomes grouped by dataset and evaluation metric. Column
+\textbf{LB} and \textbf{UB} show the lower and upper bound of the
+intervals, respectively, and computed using the percentile method. The
+underlying counterfactual evaluations are the same as the ones used to
+produce Table~\ref{tbl-main}.}
+
+\centering{
+
+\begin{tabular}{lccccc}
+  \toprule
+  \textbf{Variable} & \textbf{Data} & \textbf{CT} & \textbf{BL} & \textbf{LB} & \textbf{UB} \\\midrule
+  Cost & Adult & 2.26 & 2.2 & -0.22 & 0.28 \\
+  Cost & CH & 1.46 & 2.46 & -1.1 & -0.89 \\
+  Cost & Circ & 0.67 & 1.23 & -0.58 & -0.53 \\
+  Cost & Cred & 2.68 & 2.29 & 0.16 & 0.63 \\
+  Cost & GMSC & 1.14 & 3.05 & -2.45 & -1.77 \\
+  Cost & LS & 3.82 & 4.44 & -0.7 & -0.56 \\
+  Cost & MNIST & 77.04 & 68.67 & -3.47 & 18.34 \\
+  Cost & Moon & 1.55 & 1.6 & -0.08 & -0.01 \\
+  Cost & OL & 1.62 & 2.63 & -1.15 & -0.81 \\
+  $ \text{IP}^* $ & Adult & 0.07 & 0.11 & -0.06 & -0.01 \\
+  $ \text{IP}^* $ & CH & 0.02 & 0.06 & -0.06 & -0.04 \\
+  $ \text{IP}^* $ & Circ & 0.0 & 0.0 & -0.01 & -0.0 \\
+  $ \text{IP}^* $ & Cred & 0.03 & 0.06 & -0.05 & -0.01 \\
+  $ \text{IP}^* $ & GMSC & 0.02 & 0.07 & -0.06 & -0.04 \\
+  $ \text{IP}^* $ & LS & 0.1 & 0.23 & -0.14 & -0.12 \\
+  $ \text{IP}^* $ & MNIST & 0.04 & 0.04 & -0.1 & 0.09 \\
+  $ \text{IP}^* $ & Moon & 0.02 & 0.02 & -0.01 & -0.0 \\
+  $ \text{IP}^* $ & OL & 0.12 & 0.09 & -0.01 & 0.05 \\
+  $ \text{IP} $ & Adult & 15.03 & 15.15 & -0.68 & 0.26 \\
+  $ \text{IP} $ & CH & 6.61 & 7.52 & -1.17 & -0.63 \\
+  $ \text{IP} $ & Circ & 1.03 & 2.36 & -1.37 & -1.29 \\
+  $ \text{IP} $ & Cred & 19.31 & 22.03 & -3.69 & -1.74 \\
+  $ \text{IP} $ & GMSC & 6.19 & 8.09 & -2.4 & -1.49 \\
+  $ \text{IP} $ & LS & 2.41 & 3.4 & -1.04 & -0.94 \\
+  $ \text{IP} $ & MNIST & 258.83 & 278.54 & -30.49 & -7.64 \\
+  $ \text{IP} $ & Moon & 1.36 & 1.71 & -0.38 & -0.32 \\
+  $ \text{IP} $ & OL & 4.49 & 4.44 & -0.03 & 0.13 \\\bottomrule
+\end{tabular}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colframe=quarto-callout-note-color-frame, colback=white, titlerule=0mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacitybacktitle=0.6, breakable, coltitle=black, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, left=2mm, toprule=.15mm, arc=.35mm, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
 

### 2025-07-22 paper/sections/other/main_results.qmd
finished table and rerendered preprint

@@
 using CTExperiments.Plots
+using CTExperiments.PrettyTables
 using CTExperiments.StatsBase
@@
 get_table_inputs(df, nothing; backend=Val(:latex)) |>
-    inputs -> tabulate_results(inputs; table_type=:tabular, formatters=PrettyTables.ft_round(2, findall([x in float_cols for x in names(df)])))
+    inputs -> tabulate_results(inputs; wrap_table=false, table_type=:tabular, formatters=PrettyTables.ft_round(2, findall([x in float_cols for x in names(df)])))
 ```
@@
 ```{julia}
+#| eval: false
+
 plt = CTExperiments.plot_performance(res_dir, eps=range(0.0, 0.1, length=10) |> collect, adversarial=true, byvars=["objective", "eps"], drop_synthetic=false)
@@
 ```{julia}
+#| eval: false
+
 lvls = [0.90, 0.95, 0.99]
@@
 
+::: {#tbl-ci}
+
 ```{julia}
 #| output: asis
-#| label: tbl-ci
-#| tbl-cap: "Confidence Intervals"
 
 df = Serialization.deserialize("paper/experiments/output/ci.jls")
-df = select(df, :variable, :dataset, :full, :vanilla, "99.0%")
+df = select(df, :variable, :measure, :dataset, :full, :vanilla, "99.0%")
+df.variable = (x -> LatexCell(replace(x.data, " \$(-%)\$" => ""))).(df.variable)
+sort!(df, [:measure, :dataset])
+df = select(df, Not(:measure))
 df = transform(df, "99.0%" => AsTable) |>
     df -> select(df, Not("99.0%"))
-rename!(df, :dataset => :data, :x1 => :LB, :x2 => :UB)
+rename!(df, :dataset => :data, :x1 => :LB, :x2 => :UB, :full => :CT, :vanilla => :BL)
 inputs = get_table_inputs(df, nothing; backend=Val(:latex))
@@
     table_type=:tabular,
-    alignment = [:l, fill(:c, ncol(df)-1)...]
+    alignment = [:l, fill(:c, ncol(df)-1)...],
+    wrap_table = false,
+    sort_by_data = false,
+    hlines = :none,
 )
@@
 ```
+
+Mean outcomes for **CT** and **BL** along with bootstrapped confidence intervals (99%) for difference in mean outcomes grouped by dataset and evaluation metric.
+Column **LB** and **UB** show the lower and upper bound of the intervals, respectively, and computed using the percentile method.
+The underlying counterfactual evaluations are the same as the ones used to produce @tbl-main.
+
+:::
 

### 2025-07-22 paper/experiments/CTExperiments/src/tables.jl

@@ -14,20 +14,31 @@ function tabulate_results(
     formatters=PrettyTables.ft_round(2),
     alignment=:c,
     wrap_table_environment="table*",
+    sort_by_data::Bool=true,
+    hlines::Union{Nothing,Symbol}=nothing,
     kwrgs...,
 )
     df = inputs[1]
 
     # Group averages:
-    if "Avg." in df.data
-        _ds_order = [ds_order..., "Avg."]
-        hlines = [0, 1, 5, length(ds_order)+1, length(ds_order)+2]
-    else
-        _ds_order = ds_order
-        hlines = [0, 1, 5, length(ds_order)+2]
+    if "data" in names(df) && hlines != :none
+        if "Avg." in df.data
+            _ds_order = [ds_order..., "Avg."]
+            hlines = [0, 1, 5, length(ds_order)+1, length(ds_order)+2]
+        else
+            _ds_order = ds_order
+            hlines = [0, 1, 5, length(ds_order)+2]
+        end
+        df.data = categorical(df.data; levels=_ds_order)
+        if sort_by_data
+            sort!(df, :data)
+        end
     end
-    df.data = categorical(df.data; levels=_ds_order)
-    sort!(df, :data)
+
+    if hlines == :none
+    hlines = nothing
+    end
+
     other_inputs = inputs[2]
     if isa(other_inputs.backend, Val{:latex})
         tf = PrettyTables.tf_latex_booktabs

### 2025-07-22 paper/sections/other/main_results.qmd
finished tables

@@
 :::
+
+## Validity
+
+```{julia}
+#| eval: false
+
+df_mtbl = aggregate_ce_evaluation(
+    mtbl_dir; 
+    ratio=true, 
+    rebase=false, 
+    y="validity", 
+    agg_further_vars=["run", "lambda_energy_eval"], 
+    total_uncertainty=false, 
+    drop_models=["linear"]
+)
+df_mtbl.variable .= "val"
+df = final_table(
+    res_dir; 
+    ce_var=["validity"],
+    agg_further_vars=[["run", "lambda_energy_eval"]],
+    tbl_mtbl=df_mtbl
+)
+Serialization.serialize("paper/experiments/output/validity.jls", df)
+```

### 2025-07-22 paper/sections/other/main_results.qmd
tables added

--- 2025-07-22 paper/sections/other/main_results.qmd
@@
 
-## Confidence Intervals
+### Confidence Intervals
 
@@
 
-## Qualitative Findings for Image Data
+### Qualitative Findings for Image Data
 
@@
 :::
+
+### Costs
+
+```{julia}
+#| eval: false
+
+df = final_table(
+    res_dir; 
+    ce_var=["distance"], 
+    agg_further_vars=[["run", "lambda_energy_eval"]], 
+    total_uncertainty=false, 
+)
+Serialization.serialize("paper/experiments/output/costs.jls", df)
+```
+
+::: {#tbl-costs}
+
+```{julia}
+#| output: asis
+#| eval: true
+
+df = Serialization.deserialize("paper/experiments/output/costs.jls")
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+tabulate_results(
+    inputs; 
+    table_type=:tabular,
+    alignment = [:l, fill(:c, ncol(df)-1)...],
+    wrap_table = false,
+)
+
+```
+
+Costs
+
+:::
+
+### Validity
+
+```{julia}
+#| eval: false
+
+# Unconstrained:
+df = aggregate_ce_evaluation(
+    res_dir; 
+    ratio=false, 
+    rebase=false, 
+    y="validity", 
+    agg_further_vars=["run", "lambda_energy_eval"], 
+    total_uncertainty=false, 
+    drop_models=["linear"]
+)
+df = select(df, Not(:se, :variable)) |> df -> unstack(df, :objective, :mean)
+rename!(df, :dataset => :data, :Full => :CT, :Vanilla => :BL)
+Serialization.serialize("paper/experiments/output/validity.jls", df)
+
+# Mutability constrained:
+df = aggregate_ce_evaluation(
+    mtbl_dir; 
+    ratio=false, 
+    rebase=false, 
+    y="validity", 
+    agg_further_vars=["run", "lambda_energy_eval"], 
+    total_uncertainty=false, 
+    drop_models=["linear"]
+)
+df = select(df, Not(:se, :variable)) |> df -> unstack(df, :objective, :mean)
+rename!(df, :dataset => :data, :Full => :CT, :Vanilla => :BL)
+Serialization.serialize("paper/experiments/output/validity_mtbl.jls", df)
+```
+
+::: {#tbl-val}
+
+```{julia}
+#| output: asis
+#| eval: true
+
+df = Serialization.deserialize("paper/experiments/output/validity.jls")
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+tabulate_results(
+    inputs; 
+    table_type=:tabular,
+    alignment = [:l, fill(:c, ncol(df)-1)...],
+    wrap_table = false,
+)
+
+```
+
+Validity
+
+:::
+
+::: {#tbl-val-mtbl}
+
+```{julia}
+#| output: asis
+#| eval: true
+
+df = Serialization.deserialize("paper/experiments/output/validity_mtbl.jls")
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+tabulate_results(
+    inputs; 
+    table_type=:tabular,
+    alignment = [:l, fill(:c, ncol(df)-1)...],
+    wrap_table = false,
+)
+
+```
+
+Validity
+
+:::

### 2025-07-22 paper/preprint/paper.tex

@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-train-default}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\subsection{Confidence Intervals}\label{confidence-intervals}
+\subsubsection{Confidence Intervals}\label{confidence-intervals}
 
@@
 
-\subsection{Qualitative Findings for Image
+\subsubsection{Qualitative Findings for Image
 Data}\label{qualitative-findings-for-image-data}
@@
 \end{figure}%
+
+\subsubsection{Costs}\label{costs}
+
+\begin{table}
+
+\caption{\label{tbl-costs}Costs}
+
+\centering{
+
+\begin{tabular}{lc}
+  \toprule
+  \textbf{Data} & \textbf{Cost $(-\%)$} \\\midrule
+  LS & -26.82\pm0.86 $^{*}$ \\
+  Circ & 40.97\pm0.82 $^{*}$ \\
+  Moon & 33.83\pm0.98 $^{*}$ \\
+  OL & 10.35\pm1.28 $^{*}$ \\\midrule
+  Adult & 1.16\pm3.53 $^{}$ \\
+  CH & -34.89\pm2.31 $^{*}$ \\
+  Cred & 28.24\pm1.08 $^{*}$ \\
+  GMSC & 3.54\pm5.78 $^{}$ \\
+  MNIST & -31.67\pm7.72 $^{*}$ \\\midrule
+  Avg. & 2.75 \\\bottomrule
+\end{tabular}
+
+}
+
+\end{table}%
+
+\subsubsection{Validity}\label{validity}
+
+\begin{table}
+
+\caption{\label{tbl-val}Validity}
+
+\centering{
+
+\begin{tabular}{lcc}
+  \toprule
+  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
+  LS & 1.0 & 1.0 \\
+  Circ & 0.97 & 0.52 \\
+  Moon & 1.0 & 1.0 \\
+  OL & 0.87 & 0.98 \\\midrule
+  Adult & 0.61 & 0.99 \\
+  CH & 0.96 & 1.0 \\
+  Cred & 0.7 & 1.0 \\
+  GMSC & 0.63 & 1.0 \\
+  MNIST & 1.0 & 1.0 \\
+\end{tabular}
+
+}
+
+\end{table}%
+
+\begin{table}
+
+\caption{\label{tbl-val-mtbl}Validity}
+
+\centering{
+
+\begin{tabular}{lcc}
+  \toprule
+  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
+  LS & 1.0 & 1.0 \\
+  Circ & 0.67 & 0.49 \\
+  Moon & 0.99 & 0.98 \\
+  OL & 0.37 & 0.57 \\\midrule
+  Adult & 0.56 & 0.99 \\
+  CH & 0.96 & 1.0 \\
+  Cred & 0.67 & 1.0 \\
+  GMSC & 0.38 & 1.0 \\
+  MNIST & 1.0 & 1.0 \\
+\end{tabular}
+
+}
+
+\end{table}%
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, opacitybacktitle=0.6, rightrule=.15mm, titlerule=0mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, arc=.35mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, toptitle=1mm, left=2mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, leftrule=.75mm, opacityback=0, colback=white, coltitle=black]
+\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
 

### 2025-07-23 paper/sections/other/main_results.qmd
table added

@@
 
+### Integrated Gradients
+
+```{julia}
+#| output: asis
+
+df = Serialization.deserialize(joinpath(mtbl_dir, "ig.jls")) |>
+    df -> transform(df, [:mean, :se] => ByRow((m, s) -> LatexCell("$(round(m, digits=2)) \\pm $(round(s, digits=2))")) => :mean) |>
+    df -> select(df, Not(:se)) |>
+    df -> DataFrames.unstack(df, :objective, :mean) |>
+    df -> rename(df, :full => :CT, :vanilla => :BL) |>
+    df -> transform(df, :data => ByRow(d  -> CTExperiments.format_header(d)) => :data)
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+tabulate_results(
+    inputs; 
+    table_type=:tabular,
+    alignment = [:l, fill(:c, ncol(df)-1)...],
+    wrap_table = false,
+)
+
+```
+
 ### Costs

### 2025-07-23 paper/preprint/paper.tex
going with Alek's suggestions for standardization and baselines

@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-train-default}: Training Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
+\subsubsection{Integrated Gradients}\label{integrated-gradients}
+
+\begin{table}
+
+\caption{\label{tbl-ig}Integrated gradients.}
+
+\centering{
+
+\begin{tabular}{
+l
+S[table-format=2.2(3.2)]
+S[table-format=2.2(2.2)]
+}
+  \toprule
+  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
+  LS & 0.03 \pm 0.02 & 275.23 \pm 2389.26 \\
+  Circ & 9.58 \pm 32.17 & 34.76 \pm 154.26 \\
+  Moon & 15.05 \pm 44.79 & 0.69 \pm 0.5 \\
+  OL & 1.38 \pm 2.24 & 7.9 \pm 12.03 \\\midrule
+  Adult & 0.43 \pm 0.04 & 0.97 \pm 0.05 \\
+  CH & 0.09 \pm 0.03 & 0.22 \pm 0.06 \\
+  Cred & 0.0 \pm 0.01 & 0.43 \pm 0.06 \\
+  GMSC & 1.0 \pm 0.0 & 0.26 \pm 0.13 \\
+  MNIST & 0.18 \pm 0.02 & 0.43 \pm 0.06 \\
+\end{tabular}
+
+}
+
+\end{table}%
+
 \subsubsection{Costs}\label{costs}
@@
 
-\begin{tabular}{lc}
+\begin{tabular}{
+l
+S[table-format=2.2(1.2)]
+}
   \toprule
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune-train}: Training Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, left=2mm, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, breakable, bottomrule=.15mm, titlerule=0mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colback=white, opacityback=0, bottomtitle=1mm, coltitle=black, toptitle=1mm, leftrule=.75mm, rightrule=.15mm]
 

### 2025-07-23 paper/sections/other/main_results.qmd
going with Alek's suggestions for standardization and baselines

--- 2025-07-22 paper/sections/other/main_results.qmd
@@
 
+### Integrated Gradients
+
+:::{#tbl-ig}
+
+```{julia}
+#| output: asis
+
+df = Serialization.deserialize(joinpath(mtbl_dir, "ig.jls")) |>
+    df -> transform(df, [:mean, :se] => ByRow((m, s) -> LatexCell("$(round(m, digits=2)) \\pm $(round(s, digits=2))")) => :mean) |>
+    df -> select(df, Not(:se)) |>
+    df -> DataFrames.unstack(df, :objective, :mean) |>
+    df -> rename(df, :full => :CT, :vanilla => :BL) |>
+    df -> transform(df, :data => ByRow(d  -> CTExperiments.format_header(d)) => :data)
+inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+al = [
+    "l", 
+    "S[table-format=2.2(3.2)]", 
+    "S[table-format=2.2(2.2)]"
+] 
+tabulate_results(
+    inputs,
+    al
+    ; 
+    table_type=:tabular,
+    alignment = [:l, fill(:c, ncol(df)-1)...],
+    wrap_table = false,
+)
+
+```
+
+Integrated gradients.
+
+:::
+
 ### Costs
@@
 inputs = get_table_inputs(df, nothing; backend=Val(:latex))
+al = [
+    "l", 
+    "S[table-format=2.2(1.2)]", 
+]
 tabulate_results(
-    inputs; 
+    inputs, al; 
     table_type=:tabular,

### 2025-07-23 paper/experiments/CTExperiments/src/tables.jl

@@ -5,7 +5,8 @@ using PrettyTables
 using StatsBase
 
 function tabulate_results(
-    inputs;
+    inputs, al::Nothing=nothing;
+    io::IO=stdout,
     tf=PrettyTables.tf_latex_booktabs,
     wrap_table=true,
     table_type=:longtable,
@@ -44,6 +45,7 @@ function tabulate_results(
         tf = PrettyTables.tf_latex_booktabs
         if isnothing(save_name)
             tab = pretty_table(
+                io,
                 df;
                 tf=tf,
                 formatters=formatters,
@@ -77,7 +79,7 @@ function tabulate_results(
         end
     else
         if isnothing(save_name)
-            tab = pretty_table(df; alignment, hlines, other_inputs..., kwrgs...)
+            tab = pretty_table(io, df; alignment, hlines, other_inputs..., kwrgs...)
             return tab
         else
             open(save_name, "w") do io
@@ -85,6 +87,25 @@ function tabulate_results(
             end
         end
     end
+end
+
+function tabulate_results(inputs, al::Vector{String}; kwargs...)
+
+    if isnothing(al)
+        return tabulate_results(inputs; kwargs...)
+    end
+
+    # Capture output
+    io = IOBuffer()
+    tabulate_results(inputs; io=io, kwargs...)
+    output = String(take!(io))
+    
+    # Replace alignment with siunitx format
+    modified = replace(output, 
+        r"\\begin\{tabular\}\{[^}]+\}" => 
+                       "\\begin{tabular}{\n$( reduce((x,y) -> "$x\n$y", al))\n}")
+    
+    print(modified)
 end
 
 function get_table_inputs(

### 2025-07-23 paper/experiments/run_integrated_gradients.jl

@@ -51,17 +51,17 @@ for (i, exper_list) in enumerate(expers)
         exper_name = exper.meta_params.experiment_name
         obj = exper.training_params.objective
 
-        if data == "mnist"
-            bl = -ones(size(X,1),1)
-            igs = CTExperiments.integrated_gradients(exper; idx=idx, nrounds=nrounds, comm=comm, max_entropy=false, baseline=bl, verbose)
-        else
-            igs = CTExperiments.integrated_gradients(exper; idx=idx, nrounds=nrounds, comm=comm, max_entropy=false, baseline_type="random", verbose)
-        end
+        igs = CTExperiments.integrated_gradients(exper; idx=idx, nrounds=nrounds, comm=comm, max_entropy=false, baseline_type="random", verbose)
         
         if rank == 0
 
             # Collect:
-            igs = (ig -> (abs.(ig)) ./ (maximum(ig) .- minimum(ig) ) ).(igs)    # compute normalized contributions
+            if size(X,1) == 2
+                igs = (ig -> (abs.(ig)) ./ (maximum(ig) .- minimum(ig) ) ).(igs)    # compute normalized contributions
+            else
+                igs = (ig -> (ig .- minimum(ig)) ./ (maximum(ig) .- minimum(ig) ) ).(igs)    # compute normalized contributions
+            end
+            
             igs = (ig -> ig[exper.data.mutability,1]).(igs) |> igs -> reduce(hcat, igs)
 
             # Aggregate:

### 2025-07-23 paper/preprint/paper.tex
done

--- 2025-07-22 paper/preprint/paper.tex
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-default}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-eval-default}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\subsubsection{Costs}\label{costs}
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\subsubsection{Integrated Gradients}\label{integrated-gradients}
+
+\begin{table}[H]
+
+\caption{\label{tbl-ig}Integrated gradients.}
+
+\centering{
+
+\begin{tabular}{
+l
+S[table-format=2.2(3.2)]
+S[table-format=2.2(2.2)]
+}
+  \toprule
+  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
+  LS & 0.03 \pm 0.0 & 10.24 \pm 2.4 \\
+  Circ & 3.2 \pm 0.67 & 149.76 \pm 842.75 \\
+  Moon & 60.84 \pm 128.51 & 0.55 \pm 0.06 \\
+  OL & 0.78 \pm 0.12 & 4.81 \pm 1.08 \\\midrule
+  Adult & 0.43 \pm 0.01 & 1.0 \pm 0.0 \\
+  CH & 0.08 \pm 0.01 & 0.23 \pm 0.01 \\
+  Cred & 0.0 \pm 0.0 & 0.43 \pm 0.01 \\
+  GMSC & 1.0 \pm 0.0 & 0.21 \pm 0.03 \\
+  MNIST & 0.18 \pm 0.01 & 0.41 \pm 0.01 \\
+\end{tabular}
+
+}
+
+\end{table}%
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\begin{figure}[H]
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}
+
+}
+
+\caption{\label{fig-mnist-ig}Class-conditional integrated gradients.}
+
+\end{figure}%
+
+\end{minipage}%
+
+\end{figure}%
+
+\subsubsection{Costs and Validity}\label{costs-and-validity}
 
@@
 
-\caption{\label{tbl-costs}Costs}
-
-\centering{
-
-\begin{tabular}{lc}
+\caption{\label{tbl-panel}Costs and validity.}
+
+\begin{minipage}{0.33\linewidth}
+
+\subcaption{\label{tbl-costs}Costs}
+
+\centering{
+
+\begin{tabular}{
+l
+S[table-format=2.2(1.2)]
+}
   \toprule
@@
 
-\end{table}%
-
-\subsubsection{Validity}\label{validity}
-
-\begin{table}
-
-\caption{\label{tbl-val}Validity}
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\subcaption{\label{tbl-val}Validity}
 
@@
 
-\end{table}%
-
-\begin{table}
-
-\caption{\label{tbl-val-mtbl}Validity}
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\subcaption{\label{tbl-val-mtbl}Validity}
 
@@
 }
+
+\end{minipage}%
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
+\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, leftrule=.75mm, arc=.35mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, coltitle=black, titlerule=0mm, rightrule=.15mm, bottomtitle=1mm, breakable, bottomrule=.15mm, left=2mm, opacityback=0]
 

### 2025-07-23 _extensions/mikemahoney218/arxiv/partials/before-body.tex
formatting improved and toc added (to this end, turned appendix into standalone document).

$-- Implements the frontmatter, title page, and abstract.
$--
$-- %%%% TODO %%%%%
$-- Customize is needed, otherwise remove this partials to use Quarto default one
$-- %%%%%%%%%%%%%%%%
$if(has-frontmatter)$
\frontmatter
$endif$
$if(title)$
$if(beamer)$
\frame{\titlepage}
$else$
\maketitle
$endif$
$if(abstract)$
\begin{abstract}
$abstract$
\end{abstract}
$endif$
$endif$
$if(keywords)$
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
$for(keywords/allbutlast)$$keywords$ \sep $endfor$
$for(keywords/last)$$keywords$$endfor$

$endif$
\pagebreak

### 2025-07-23 paper/preprint/appendix.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

@@
 title: Supplementary Appendix
+subtitle: "Counterfactual Training: Teaching Models Plausible and Actionable Explanations"
+abstract: |
+  This is the supplementary appendix to our paper titled *Counterfactual Training: Teaching Models Plausible and Actionable Explanations*.
+It provides helpful details on mathematical notations and formulas, our proposed training regime, extended empirical findings, hyperparameter tuning and grid searches as well as software and computations.
 engine: julia
@@
   arxiv-pdf:
+    toc: true
     keep-tex: true  
@@
 
+{{< pagebreak >}}
+
 ```{=latex}
@@
 ```
+
+# References {-}
+
+::: {#refs}
+:::

### 2025-07-23 paper/preprint/appendix.tex
formatting improved and toc added (to this end, turned appendix into standalone document).

@@
   pdftitle={Supplementary Appendix},
-  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
   pdfkeywords={Counterfactual Training, Counterfactual
@@
 \title{Supplementary Appendix}
+\usepackage{etoolbox}
+\makeatletter
+\providecommand{\subtitle}[1]{% add subtitle to \maketitle
+  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
+}
+\makeatother
+\subtitle{Counterfactual Training: Teaching Models Plausible and
+Actionable Explanations}
 \def\asep{\\\\\\ } % default: all authors on same column
 \def\asep{\And }
-\author{\textbf{Patrick
-Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
-Engineering, Mathematics and Computer Science\\Delft University of
-Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
-Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\\asep\textbf{Arie van
-Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
-Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\}
+\author{}
 \date{}
@@
 \maketitle
+\begin{abstract}
+This is the supplementary appendix to our paper titled
+\emph{Counterfactual Training: Teaching Models Plausible and Actionable
+Explanations}. It provides helpful details on mathematical notations and
+formulas, our proposed training regime, extended empirical findings,
+hyperparameter tuning and grid searches as well as software and
+computations.
+\end{abstract}
 {\bfseries \emph Keywords}
@@
 Representation Learning
+
+\pagebreak
+\renewcommand*\contentsname{Table of contents}
+{
+\hypersetup{linkcolor=}
+\setcounter{tocdepth}{3}
+\tableofcontents
+}
+
+\newpage{}
 
@@
 distance (\(\ell_1\)-norm) of counterfactuals from their original
-factual state. For \emph{Generic} and \emph{ECCo}, we have
+factual state. For \emph{Generic} and \emph{ECCCo}, we have
 \(\mathcal{Z}:=\mathcal{X}\) and
@@
 (\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
-the distance penalty, \emph{ECCo} uses an additional penalty component
+the distance penalty, \emph{ECCCo} uses an additional penalty component
 that regularizes the energy associated with the counterfactual,
@@
 
-In \textbf{?@sec-constraints} we explain that to avoid penalizing
-implausibility that arises due to mutability constraints, we impose a
-point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
-argue in \textbf{?@sec-constraints} that this approach induces models to
-be less sensitive to immutable features and demonstrate this empirically
-in \textbf{?@sec-experiments}. Below we derive the analytical results in
-Prp.\textasciitilde{}\ref{prp-mtblty}.
+In the main paper, we explain that to avoid penalizing implausibility
+that arises due to mutability constraints, we impose a point mass prior
+on \(p(\mathbf{x})\) for the corresponding feature. We argue that this
+approach induces models to be relatively less sensitive to immutable
+features, propose a theoretical result supporting this and provide
+empirical evidence that strengthens our argument (both in the main paper
+and additional findings in this appendix). Below we derive the
+analytical results in Prp.\textasciitilde{}\ref{prp-mtblty}.
 
@@
 target class. Then the partial derivative of the contrastive divergence
-penalty \textbf{?@eq-div} with respect to coefficient \(\theta_{y^+,d}\)
-is equal to
+penalty with respect to coefficient \(\theta_{y^+,d}\) is equal to
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-train-default}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-default}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-eval-default}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 
-As discussed \textbf{?@sec-experiments}, CT is sensitive to certain
-hyperparameter choices. We study the effect of many hyperparameters
-extensively in Section~\ref{sec-app-grid}. For the main results, we tune
-a small set of key hyperparameters (Section~\ref{sec-app-tune}). The
-final choices for the main results are presented for each data set in
+As discussed the main paper, CT is sensitive to certain hyperparameter
+choices. We study the effect of many hyperparameters extensively in
+Section~\ref{sec-app-grid}. For the main results, we tune a small set of
+key hyperparameters (Section~\ref{sec-app-tune}). The final choices for
+the main results are presented for each data set in
 Table~\ref{tbl-final-params} along with training, test and batch sizes.
@@
 \caption{\label{tbl-final-params}Final hyperparameters used for the main
-results presented in \textbf{?@sec-experiments}. Any hyperparameter not
-shown here is set to its default value (Note~\ref{nte-train-default}).}
+results presented in the main paper. Any hyperparameter not shown here
+is set to its default value (Note~\ref{nte-train-default}).}
 
@@
 
-\caption{\label{tbl-ci}Confidence intervals}
+\caption{\label{tbl-ci}Mean outcomes for \textbf{CT} and \textbf{BL}
+along with bootstrapped confidence intervals (99\%) for difference in
+mean outcomes grouped by dataset and evaluation metric. Column
+\textbf{LB} and \textbf{UB} show the lower and upper bound of the
+intervals, respectively, and computed using the percentile method. The
+underlying counterfactual evaluations are the same as the ones used to
+produce the main table in the paper.}
 
@@
   \toprule
-  \textbf{Variable} & \textbf{Data} & \textbf{Full} & \textbf{Vanilla} & \textbf{LB} & \textbf{UB} \\\midrule
-  Cost $(-\%)$ & LS & 3.82 & 4.44 & -0.7 & -0.56 \\
-  $ \text{IP}^* $ $(-\%)$ & LS & 0.1 & 0.23 & -0.14 & -0.12 \\
-  $ \text{IP} $ $(-\%)$ & LS & 2.41 & 3.4 & -1.04 & -0.94 \\
-  Cost $(-\%)$ & Circ & 0.67 & 1.23 & -0.58 & -0.53 \\\midrule
-  $ \text{IP}^* $ $(-\%)$ & Circ & 0.0 & 0.0 & -0.01 & -0.0 \\
-  $ \text{IP} $ $(-\%)$ & Circ & 1.03 & 2.36 & -1.37 & -1.29 \\
-  Cost $(-\%)$ & Moon & 1.55 & 1.6 & -0.08 & -0.01 \\
-  $ \text{IP}^* $ $(-\%)$ & Moon & 0.02 & 0.02 & -0.01 & -0.0 \\
-  $ \text{IP} $ $(-\%)$ & Moon & 1.36 & 1.71 & -0.38 & -0.32 \\
-  Cost $(-\%)$ & OL & 1.62 & 2.63 & -1.15 & -0.81 \\\midrule
-  $ \text{IP}^* $ $(-\%)$ & OL & 0.12 & 0.09 & -0.01 & 0.05 \\
-  $ \text{IP} $ $(-\%)$ & OL & 4.49 & 4.44 & -0.03 & 0.13 \\
-  Cost $(-\%)$ & Adult & 2.26 & 2.2 & -0.22 & 0.28 \\
-  $ \text{IP}^* $ $(-\%)$ & Adult & 0.07 & 0.11 & -0.06 & -0.01 \\
-  $ \text{IP} $ $(-\%)$ & Adult & 15.03 & 15.15 & -0.68 & 0.26 \\
-  Cost $(-\%)$ & CH & 1.46 & 2.46 & -1.1 & -0.89 \\
-  $ \text{IP}^* $ $(-\%)$ & CH & 0.02 & 0.06 & -0.06 & -0.04 \\
-  $ \text{IP} $ $(-\%)$ & CH & 6.61 & 7.52 & -1.17 & -0.63 \\
-  Cost $(-\%)$ & Cred & 2.68 & 2.29 & 0.16 & 0.63 \\
-  $ \text{IP}^* $ $(-\%)$ & Cred & 0.03 & 0.06 & -0.05 & -0.01 \\
-  $ \text{IP} $ $(-\%)$ & Cred & 19.31 & 22.03 & -3.69 & -1.74 \\
-  Cost $(-\%)$ & GMSC & 1.14 & 3.05 & -2.45 & -1.77 \\
-  $ \text{IP}^* $ $(-\%)$ & GMSC & 0.02 & 0.07 & -0.06 & -0.04 \\
-  $ \text{IP} $ $(-\%)$ & GMSC & 6.19 & 8.09 & -2.4 & -1.49 \\
-  Cost $(-\%)$ & MNIST & 77.04 & 68.67 & -3.47 & 18.34 \\
-  $ \text{IP}^* $ $(-\%)$ & MNIST & 0.04 & 0.04 & -0.1 & 0.09 \\
-  $ \text{IP} $ $(-\%)$ & MNIST & 258.83 & 278.54 & -30.49 & -7.64 \\
+  \textbf{Variable} & \textbf{Data} & \textbf{CT} & \textbf{BL} & \textbf{LB} & \textbf{UB} \\\midrule
+  Cost & Adult & 2.26 & 2.2 & -0.22 & 0.28 \\
+  Cost & CH & 1.46 & 2.46 & -1.1 & -0.89 \\
+  Cost & Circ & 0.67 & 1.23 & -0.58 & -0.53 \\
+  Cost & Cred & 2.68 & 2.29 & 0.16 & 0.63 \\
+  Cost & GMSC & 1.14 & 3.05 & -2.45 & -1.77 \\
+  Cost & LS & 3.82 & 4.44 & -0.7 & -0.56 \\
+  Cost & MNIST & 77.04 & 68.67 & -3.47 & 18.34 \\
+  Cost & Moon & 1.55 & 1.6 & -0.08 & -0.01 \\
+  Cost & OL & 1.62 & 2.63 & -1.15 & -0.81 \\
+  $ \text{IP}^* $ & Adult & 0.07 & 0.11 & -0.06 & -0.01 \\
+  $ \text{IP}^* $ & CH & 0.02 & 0.06 & -0.06 & -0.04 \\
+  $ \text{IP}^* $ & Circ & 0.0 & 0.0 & -0.01 & -0.0 \\
+  $ \text{IP}^* $ & Cred & 0.03 & 0.06 & -0.05 & -0.01 \\
+  $ \text{IP}^* $ & GMSC & 0.02 & 0.07 & -0.06 & -0.04 \\
+  $ \text{IP}^* $ & LS & 0.1 & 0.23 & -0.14 & -0.12 \\
+  $ \text{IP}^* $ & MNIST & 0.04 & 0.04 & -0.1 & 0.09 \\
+  $ \text{IP}^* $ & Moon & 0.02 & 0.02 & -0.01 & -0.0 \\
+  $ \text{IP}^* $ & OL & 0.12 & 0.09 & -0.01 & 0.05 \\
+  $ \text{IP} $ & Adult & 15.03 & 15.15 & -0.68 & 0.26 \\
+  $ \text{IP} $ & CH & 6.61 & 7.52 & -1.17 & -0.63 \\
+  $ \text{IP} $ & Circ & 1.03 & 2.36 & -1.37 & -1.29 \\
+  $ \text{IP} $ & Cred & 19.31 & 22.03 & -3.69 & -1.74 \\
+  $ \text{IP} $ & GMSC & 6.19 & 8.09 & -2.4 & -1.49 \\
+  $ \text{IP} $ & LS & 2.41 & 3.4 & -1.04 & -0.94 \\
+  $ \text{IP} $ & MNIST & 258.83 & 278.54 & -30.49 & -7.64 \\
+  $ \text{IP} $ & Moon & 1.36 & 1.71 & -0.38 & -0.32 \\
+  $ \text{IP} $ & OL & 4.49 & 4.44 & -0.03 & 0.13 \\\bottomrule
 \end{tabular}
@@
 the corresponding counterfactual for each target class (columns) in that
-same row. The underlying generator, \emph{ECCo}, aims to generate
+same row. The underlying generator, \emph{ECCCo}, aims to generate
 counterfactuals that are faithful to the model
@@
 
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\subsubsection{Integrated Gradients}\label{integrated-gradients}
+
+\begin{table}[H]
+
+\caption{\label{tbl-ig}Integrated gradients.}
+
+\centering{
+
+\begin{tabular}{
+l
+S[table-format=2.2(3.2)]
+S[table-format=2.2(2.2)]
+}
+  \toprule
+  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
+  LS & 0.03 \pm 0.0 & 10.24 \pm 2.4 \\
+  Circ & 3.2 \pm 0.67 & 149.76 \pm 842.75 \\
+  Moon & 60.84 \pm 128.51 & 0.55 \pm 0.06 \\
+  OL & 0.78 \pm 0.12 & 4.81 \pm 1.08 \\\midrule
+  Adult & 0.43 \pm 0.01 & 1.0 \pm 0.0 \\
+  CH & 0.08 \pm 0.01 & 0.23 \pm 0.01 \\
+  Cred & 0.0 \pm 0.0 & 0.43 \pm 0.01 \\
+  GMSC & 1.0 \pm 0.0 & 0.21 \pm 0.03 \\
+  MNIST & 0.18 \pm 0.01 & 0.41 \pm 0.01 \\
+\end{tabular}
+
+}
+
+\end{table}%
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\begin{figure}[H]
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}
+
+}
+
+\caption{\label{fig-mnist-ig}Class-conditional integrated gradients.}
+
+\end{figure}%
+
+\end{minipage}%
+
+\end{figure}%
+
+\subsubsection{Costs and Validity}\label{costs-and-validity}
+
+\begin{table}
+
+\caption{\label{tbl-panel}Costs and validity.}
+
+\begin{minipage}{0.33\linewidth}
+
+\subcaption{\label{tbl-costs}Costs}
+
+\centering{
+
+\begin{tabular}{
+l
+S[table-format=2.2(1.2)]
+}
+  \toprule
+  \textbf{Data} & \textbf{Cost $(-\%)$} \\\midrule
+  LS & -26.82\pm0.86 $^{*}$ \\
+  Circ & 40.97\pm0.82 $^{*}$ \\
+  Moon & 33.83\pm0.98 $^{*}$ \\
+  OL & 10.35\pm1.28 $^{*}$ \\\midrule
+  Adult & 1.16\pm3.53 $^{}$ \\
+  CH & -34.89\pm2.31 $^{*}$ \\
+  Cred & 28.24\pm1.08 $^{*}$ \\
+  GMSC & 3.54\pm5.78 $^{}$ \\
+  MNIST & -31.67\pm7.72 $^{*}$ \\\midrule
+  Avg. & 2.75 \\\bottomrule
+\end{tabular}
+
+}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\subcaption{\label{tbl-val}Validity}
+
+\centering{
+
+\begin{tabular}{lcc}
+  \toprule
+  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
+  LS & 1.0 & 1.0 \\
+  Circ & 0.97 & 0.52 \\
+  Moon & 1.0 & 1.0 \\
+  OL & 0.87 & 0.98 \\\midrule
+  Adult & 0.61 & 0.99 \\
+  CH & 0.96 & 1.0 \\
+  Cred & 0.7 & 1.0 \\
+  GMSC & 0.63 & 1.0 \\
+  MNIST & 1.0 & 1.0 \\
+\end{tabular}
+
+}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\subcaption{\label{tbl-val-mtbl}Validity}
+
+\centering{
+
+\begin{tabular}{lcc}
+  \toprule
+  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
+  LS & 1.0 & 1.0 \\
+  Circ & 0.67 & 0.49 \\
+  Moon & 0.99 & 0.98 \\
+  OL & 0.37 & 0.57 \\\midrule
+  Adult & 0.56 & 0.99 \\
+  CH & 0.96 & 1.0 \\
+  Cred & 0.67 & 1.0 \\
+  GMSC & 0.38 & 1.0 \\
+  MNIST & 1.0 & 1.0 \\
+\end{tabular}
+
+}
+
+\end{minipage}%
+
+\end{table}%
+
 \FloatBarrier
@@
   \textbf{Penalty Strengths} (Section~\ref{sec-app-grid-pen}):
-  Investigates the effect of changing the penalty strengths in out
-  proposed objective (\textbf{?@eq-obj}).
+  Investigates the effect of changing the penalty strengths in our
+  proposed training objective.
 \item
@@
 (im)plausibility and cost of counterfactuals at test time. Since the
-computation of our proposed divergence metric (\textbf{?@eq-impl-div})
+computation of our proposed divergence-based adaption (\(\text{IP}^*\))
 is memory-intensive, we rely on the distance-based metric for the grid
@@
 VAE. In other words, counterfactuals generated by \emph{REVISE} are less
-faithful to the model that \emph{ECCo} and \emph{Generic}, and hence we
+faithful to the model that \emph{ECCCo} and \emph{Generic}, and hence we
 would expect them to be a less effective and, in fact, potentially
@@
 combination with higher decision thresholds (\(\tau\)) yields the best
-results when using \emph{ECCo}. But depending on the degree of class
+results when using \emph{ECCCo}. But depending on the degree of class
 separability of the underlying data, a high decision-threshold can also
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 change from the baseline model for the distance-based implausibility
-metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
+metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCCo}).
 Data: Circles.}
@@
 change from the baseline model for the distance-based implausibility
-metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
+metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCCo}).
 Data: Linearly Separable.}
@@
 plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Moons.}
+from the baseline model for the distance-based implausibility metric (\$
+ext\{IP\}\$). Boxplots indicate the variation across evaluation runs and
+test settings (varying parameters for \emph{ECCCo}). Data: Moons.}
 
@@
 plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Overlapping.}
+from the baseline model for the distance-based implausibility metric (\$
+ext\{IP\}\$). Boxplots indicate the variation across evaluation runs and
+test settings (varying parameters for \emph{ECCCo}). Data: Overlapping.}
 
@@
 Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Circles.}
+and test settings (varying parameters for \emph{ECCCo}). Data: Circles.}
 
@@
 Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Linearly
+and test settings (varying parameters for \emph{ECCCo}). Data: Linearly
 Separable.}
@@
 Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Moons.}
+and test settings (varying parameters for \emph{ECCCo}). Data: Moons.}
 
@@
 Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data:
+and test settings (varying parameters for \emph{ECCCo}). Data:
 Overlapping.}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Circles.}
+from the baseline model for the distance-based implausibility metric (\$
+ext\{IP\}\$). Boxplots indicate the variation across evaluation runs and
+test settings (varying parameters for \emph{ECCCo}). Data: Circles.}
 
@@
 plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Linearly Separable.}
+from the baseline model for the distance-based implausibility metric (\$
+ext\{IP\}\$). Boxplots indicate the variation across evaluation runs and
+test settings (varying parameters for \emph{ECCCo}). Data: Linearly
+Separable.}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 change from the baseline model for the distance-based implausibility
-metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
+metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCCo}).
 Data: Adult.}
@@
 change from the baseline model for the distance-based implausibility
-metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
+metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCCo}).
 Data: California Housing.}
@@
 change from the baseline model for the distance-based implausibility
-metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
+metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCCo}).
 Data: Credit.}
@@
 change from the baseline model for the distance-based implausibility
-metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
+metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCCo}).
 Data: GMSC.}
@@
 change from the baseline model for the distance-based implausibility
-metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
+metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCCo}).
 Data: MNIST.}
@@
 change from the baseline model for the distance-based implausibility
-metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
+metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCCo}).
 Data: Moons.}
@@
 change from the baseline model for the distance-based implausibility
-metric (\textbf{?@eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
+metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+evaluation runs and test settings (varying parameters for \emph{ECCCo}).
 Data: Overlapping.}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toptitle=1mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm, coltitle=black, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacityback=0, titlerule=0mm, rightrule=.15mm, leftrule=.75mm, breakable, left=2mm, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
 
@@
 combinations was parallelized across 34 CPUs with 2GB memory each. The
-time to completion varied by dataset for reasons discussed in
-\textbf{?@sec-discussion}: 0h49m (\emph{Moons}), 1h4m (\emph{Linearly
-Separable}), 1h49m (\emph{Circles}), 3h52m (\emph{Overlapping}). Model
-evaluations for large grid searches were parallelized across 20 CPUs
-with 3GB memory each. Evaluations for all data sets took less than one
-hour (\textless1h) to complete.
+time to completion varied by dataset: 0h49m (\emph{Moons}), 1h4m
+(\emph{Linearly Separable}), 1h49m (\emph{Circles}), 3h52m
+(\emph{Overlapping}). Model evaluations for large grid searches were
+parallelized across 20 CPUs with 3GB memory each. Evaluations for all
+data sets took less than one hour (\textless1h) to complete.
 
@@
 
+\section*{References}\label{references}
+\addcontentsline{toc}{section}{References}
+
 \phantomsection\label{refs}

### 2025-07-23 paper/sections/other/constraints.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

@@
 
-In @sec-constraints we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature.
-We argue in @sec-constraints that this approach induces models to be less sensitive to immutable features and demonstrate this empirically in @sec-experiments.
+In the main paper, we explain that to avoid penalizing implausibility that arises due to mutability constraints, we impose a point mass prior on $p(\mathbf{x})$ for the corresponding feature.
+We argue that this approach induces models to be relatively less sensitive to immutable features, propose a theoretical result supporting this and provide empirical evidence that strengthens our argument (both in the main paper and additional findings in this appendix).
 Below we derive the analytical results in Prp.~\ref{prp-mtblty}.
@@
 Let $\mathbf{x}^\prime$ denote some randomly chosen individual from class $y^-$ and let $y^+ \sim p(y)$ denote the randomly chosen target class.
-Then the partial derivative of the contrastive divergence penalty [@eq-div] with respect to coefficient $\theta_{y^+,d}$ is equal to
+Then the partial derivative of the contrastive divergence penalty with respect to coefficient $\theta_{y^+,d}$ is equal to
 

### 2025-07-23 paper/sections/other/counterfactuals.qmd
trying to address comments on figure 2
formatting improved and toc added (to this end, turned appendix into standalone document).

@@
 All generators also penalize the distance ($\ell_1$-norm) of counterfactuals from their original factual state.
-For *Generic* and *ECCo*, we have $\mathcal{Z}:=\mathcal{X}$ and $g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}$, that is counterfactual are searched directly in the feature space.
+For *Generic* and *ECCCo*, we have $\mathcal{Z}:=\mathcal{X}$ and $g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}$, that is counterfactual are searched directly in the feature space.
 Conversely, *REVISE* traverses the latent space of a variational autoencoder (VAE) fitted to the training data, where $g(\cdot)$ corresponds to the decoder [@joshi2019realistic].
-In addition to the distance penalty, *ECCo* uses an additional penalty component that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful].
+In addition to the distance penalty, *ECCCo* uses an additional penalty component that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful].
+
+<!-- TODO: Add note on why we omit the conformal prediction component. -->
 

### 2025-07-23 paper/sections/other/grids.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

@@
 1. **Generator Parameters** (@sec-app-grid-gen): Investigates the effect of changing hyperparameters that affect the counterfactual outcomes during the training phase.
-2. **Penalty Strengths** (@sec-app-grid-pen): Investigates the effect of changing the penalty strengths in out proposed objective (@eq-obj).
+2. **Penalty Strengths** (@sec-app-grid-pen): Investigates the effect of changing the penalty strengths in our proposed training objective.
 3. **Other Parameters** (@sec-app-grid-train): Investigates the effect of changing other training parameters, including the total number of generated counterfactuals in each epoch.
@@
 With respect to explanatory performance, we report here our findings for the (im)plausibility and cost of counterfactuals at test time.
-Since the computation of our proposed divergence metric (@eq-impl-div) is memory-intensive, we rely on the distance-based metric for the grid searches.
+Since the computation of our proposed divergence-based adaption ($\text{IP}^*$) is memory-intensive, we rely on the distance-based metric for the grid searches.
 For the counterfactual evaluation, we draw factual samples from the training data for the grid searches to avoid data leakage with respect to our final results reported in the body of the paper.
@@
 This finding can be attributed to the fact that *REVISE* effectively assigns the task of learning plausible explanations from the model itself to a surrogate VAE.
-In other words, counterfactuals generated by *REVISE* are less faithful to the model that *ECCo* and *Generic*, and hence we would expect them to be a less effective and, in fact, potentially detrimental role in our training regime.
+In other words, counterfactuals generated by *REVISE* are less faithful to the model that *ECCCo* and *Generic*, and hence we would expect them to be a less effective and, in fact, potentially detrimental role in our training regime.
 Secondly, we observe that allowing for a higher number of maximum steps $T$ for the counterfactual search generally yields better outcomes.
 This is intuitive, because it allows more counterfactuals to reach maturity in any given iteration.
-Looking in particular at the results for *Linearly Separable*, it seems that higher values for $T$ in combination with higher decision thresholds ($\tau$) yields the best results when using *ECCo*.
+Looking in particular at the results for *Linearly Separable*, it seems that higher values for $T$ in combination with higher decision thresholds ($\tau$) yields the best results when using *ECCCo*.
 But depending on the degree of class separability of the underlying data, a high decision-threshold can also affect results adversely, as evident from the results for the *Overlapping* data (@fig-grid-gen_params-plaus-over): here we find that CT generally fails to achieve its objective because only a tiny proportion of counterfactuals ever reaches maturity.
@@
 fig_caption = "Average outcomes for the plausibility measure across hyperparameters.
-This shows the % change from the baseline model for the distance-based implausibility metric (@eq-impl-dist).
-Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCo*)."
+This shows the % change from the baseline model for the distance-based implausibility metric (\$\text{IP}\$).
+Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCCo*)."
 full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))
@@
 This shows the % change from the baseline model for the distance-based cost metric [@wachter2017counterfactual].
-Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCo*)."
+Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCCo*)."
 full_paths = joinpath.(data_dirs, joinpath(suffix,imgfname))

### 2025-07-23 paper/sections/other/initial_grids.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

@@
 
-- **Energy Penalty** (@tbl-lin_sep-lambda_energy_exper): *ECCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training.
+- **Energy Penalty** (@tbl-lin_sep-lambda_energy_exper): *ECCCo* generally does yield better results than *Vanilla* for higher choices of the energy penalty (10,15) during training.
 *Generic* performs poorly accross the board.
-*Omni* seems to have an anchoring effect, in that it never performs terribly but also never as good as the best *ECCo* results.
+*Omni* seems to have an anchoring effect, in that it never performs terribly but also never as good as the best *ECCCo* results.
 *REVISE* performs poorly across the board.
 - **Cost** (@tbl-lin_sep-lambda_cost_exper): Results for all generators (except *Omni*) are quite bad, which can likely be attributed to extremely bad results for some choices of the **Energy Penalty** (results here are averaged).
-For *ECCo* and *Generic*, higher cost values generally lead to worse results.
+For *ECCCo* and *Generic*, higher cost values generally lead to worse results.
 - **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
-- **Validity**: *ECCo* almost always valid except for very low values during training and high values at evaluation time. *Generic* often has poor validity.
+- **Validity**: *ECCCo* almost always valid except for very low values during training and high values at evaluation time. *Generic* often has poor validity.
 - **Accuracy**: Seems largely unaffected.
@@
 
-- **Energy Penalty** (@tbl-moons-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, except for very low choices of the energy penalty during training for which it performs abismal.
+- **Energy Penalty** (@tbl-moons-lambda_energy_exper): *ECCCo* consistently yields better results than *Vanilla*, except for very low choices of the energy penalty during training for which it performs abismal.
 *Generic* performs quite badly across the board for high enough choices of the energy penalty at evaluation time.
@@
 *REVISE* performs poorly across the board.
-- **Cost (distance penalty)**: *Generic* generally does better for higher values, while *ECCo* does better for lower values.
+- **Cost (distance penalty)**: *Generic* generally does better for higher values, while *ECCCo* does better for lower values.
 - **Maximum Iterations**: No clear patterns recognizable, so it seems that smaller choices are ok. 
-- **Validity**: *ECCo* generally achieves full validity except for very low choices the energy penalty during training and high choices at evaluation time.
+- **Validity**: *ECCCo* generally achieves full validity except for very low choices the energy penalty during training and high choices at evaluation time.
 *Generic* performs poorly for high choices of the energy penalty during evaluation.
-- **Accuracy**: Largely unaffected although *ECCo* suffers a bit for very low choices the energy penalty during training. *REVISE* suffers a lot in general (around 10 percentage points).
+- **Accuracy**: Largely unaffected although *ECCCo* suffers a bit for very low choices the energy penalty during training. *REVISE* suffers a lot in general (around 10 percentage points).
 
@@
 
-- **Energy Penalty** (@tbl-circles-lambda_energy_exper): *ECCo* consistently yields better results than *Vanilla*, though primarily for low to medium choices of the energy penalty (<=5) during training.
-The same goes for *Generic*, which sometimes outperforms *ECCo* (for small energy penalty at evaluation time).
+- **Energy Penalty** (@tbl-circles-lambda_energy_exper): *ECCCo* consistently yields better results than *Vanilla*, though primarily for low to medium choices of the energy penalty (<=5) during training.
+The same goes for *Generic*, which sometimes outperforms *ECCCo* (for small energy penalty at evaluation time).
 *Omni* does alright for lower energy penalty at evaluation time, but loses out for higher choices.
 *REVISE* performs poorly across the board (except very low choices at evaluation time).
-- **Cost (distance penalty)**: *ECCo* and *Generic* generally achieve the best results when no cost penalty is used during training. Both *Omni* and *REVISE* are largely unaffected.
-- **Maximum Iterations**: *ECCo* consistently yields better results for higher numbers of iterations. *Generic* generally does best for a medium number (50). *Omni* is sometimes invalid (**???**).
-- **Validity**: *ECCo* tends to outperform its *Vanilla* counterpart, though primarily for low to medium choices of the energy penalty (<=5) during training and evaluation.
+- **Cost (distance penalty)**: *ECCCo* and *Generic* generally achieve the best results when no cost penalty is used during training. Both *Omni* and *REVISE* are largely unaffected.
+- **Maximum Iterations**: *ECCCo* consistently yields better results for higher numbers of iterations. *Generic* generally does best for a medium number (50). *Omni* is sometimes invalid (**???**).
+- **Validity**: *ECCCo* tends to outperform its *Vanilla* counterpart, though primarily for low to medium choices of the energy penalty (<=5) during training and evaluation.
 *Vanilla* typically worse across the board.
-- **Accuracy**: Mostly unaffected, but *REVISE* again consistently some deterioration and *ECCo* deteriorates for high choices of energy penalty during training, reflecting other outcomes above.
+- **Accuracy**: Mostly unaffected, but *REVISE* again consistently some deterioration and *ECCCo* deteriorates for high choices of energy penalty during training, reflecting other outcomes above.
 

### 2025-07-23 paper/sections/other/main_results.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

@@
 
-As discussed @sec-experiments, CT is sensitive to certain hyperparameter choices.
+As discussed the main paper, CT is sensitive to certain hyperparameter choices.
 We study the effect of many hyperparameters extensively in @sec-app-grid.
@@
 
-Final hyperparameters used for the main results presented in @sec-experiments. Any hyperparameter not shown here is set to its default value (@nte-train-default).
+Final hyperparameters used for the main results presented in the main paper. Any hyperparameter not shown here is set to its default value (@nte-train-default).
 
@@
 Column **LB** and **UB** show the lower and upper bound of the intervals, respectively, and computed using the percentile method.
-The underlying counterfactual evaluations are the same as the ones used to produce @tbl-main.
+The underlying counterfactual evaluations are the same as the ones used to produce the main table in the paper.
 
@@
 Factual images are shown on the diagonal, with the corresponding counterfactual for each target class (columns) in that same row.
-The underlying generator, *ECCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}
+The underlying generator, *ECCCo*, aims to generate counterfactuals that are faithful to the model [@altmeyer2024faithful].](/paper/figures/mnist_mlp.png){#fig-mnist}
 
@@
 :::
+
+::: {layout-ncol=2}
 
@@
 
-### Costs
+![Class-conditional integrated gradients.](/paper/figures/mnist_ig.png){#fig-mnist-ig}
+
+:::
+
+### Costs and Validity
+
+::: {#tbl-panel layout-ncol=3}
 
@@
 :::
-
-### Validity
 
@@
 :::
+
+Costs and validity.
+
+:::

### 2025-07-23 paper/sections/other/tuning.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

@@
 
-::: {.callout-warning}
+<!-- ::: {.callout-warning}
 
@@
 
-:::
+::: -->
 
@@
 fig_caption = "Average outcomes for the plausibility measure across key hyperparameters.
-This shows the % change from the baseline model for the distance-based implausibility metric (@eq-impl-dist).
-Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCo*)."
+This shows the % change from the baseline model for the distance-based implausibility metric (\$\text{IP}\$).
+Boxplots indicate the variation across evaluation runs and test settings (varying parameters for *ECCCo*)."
 full_paths = joinpath.(data_dirs, joinpath(ce_suffix,imgfname))

### 2025-07-23 paper/sections/experiments.qmd
add done

@@
 
-To assess outcomes with respect to actionability for non-linear models, we look at the average costs of valid counterfactuals in terms of their distances from factual starting points.
+To assess outcomes with respect to actionability for non-linear models, we look at the costs of (just) valid counterfactuals in terms of their distances from factual starting points with $\tau=0.5$.
 While this an imperfect proxy of sensitivity, we hypothesize that CT can reduce these costs by teaching models to seek plausibility with respect to mutable features, much like we observe in @fig-poc in panel (d) compared to (c).
-We supplement this analysis with qualitative findings for integrated gradients [@sundararajan2017ig].
+We supplement this analysis with estimates for integrated gradients [@sundararajan2017ig].
 Finally, for predictive performance, we use standard metrics, such as robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

### 2025-07-23 paper/sections/other/counterfactuals.qmd
add done

@@
 
-In this section, we provide some background on gradient-based counterfactual generators (@sec-app-ce-background) and discuss how we define convergence in this context (@sec-app-conv).
+In this section, we provide some additional background on gradient-based counterfactual generators (@sec-app-ce-background) and discuss how we define convergence in this context (@sec-app-conv).
 
@@
 
-Gradient-based counterfactual search was originally proposed by @wachter2017counterfactual. It generally solves the following unconstrained objective,
+Gradient-based counterfactual search was originally proposed by @wachter2017counterfactual. A more general version of the counterfactual search objective presented in the paper is as follows,
 
@@
 \begin{aligned}
-\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
+\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{reg}(g(\mathbf{z}^\prime)) }  \right\} 
 \end{aligned} 
@@
 
-where $g: \mathcal{Z} \mapsto \mathcal{X}$ is an invertible function that maps from the $L$-dimensional counterfactual state space to the feature space and $\text{cost}(\cdot)$ denotes one or more penalties that are used to induce certain properties of the counterfactual outcome.
+where $g: \mathcal{Z} \mapsto \mathcal{X}$ is an invertible function that maps from the $L$-dimensional counterfactual state space to the feature space and $\text{reg}(\cdot)$ denotes one or more penalties that are used to induce certain properties of the counterfactual outcome.
 As above, $\mathbf{y}^+$ denotes the target output and $\mathbf{M}_{\theta}(\mathbf{x})$ returns the logit predictions of the underlying classifier for $\mathbf{x}=g(\mathbf{z})$.
@@
 Conversely, *REVISE* traverses the latent space of a variational autoencoder (VAE) fitted to the training data, where $g(\cdot)$ corresponds to the decoder [@joshi2019realistic].
-In addition to the distance penalty, *ECCCo* uses an additional penalty component that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful].
-
-<!-- TODO: Add note on why we omit the conformal prediction component. -->
+In addition to the distance penalty, *ECCCo* uses a penalty that regularizes the energy associated with the counterfactual, $\mathbf{x}^\prime$ [@altmeyer2024faithful].
+We omit the conformal set size penalty proposed in the original paper, since a) the authors found faithfulness to primarily depend on the energy penalty and hence this alleviates us from one additional hyperparameter.
 

### 2025-07-23 paper/sections/other/main_results.qmd
add done

@@
 As discussed the main paper, CT is sensitive to certain hyperparameter choices.
-We study the effect of many hyperparameters extensively in @sec-app-grid.
+We study the effect of many hyperparameters extensively in @sec-app-grid of this appendix.
 For the main results, we tune a small set of key hyperparameters (@sec-app-tune).
@@
 
-## Final Results {#sec-app-final-results}
-
-Plus/minus two standard deviations of bootstrap estimates.
-
-<!-- Standard deviation of bootstrap is the standard error. -->
+<!-- NOTE: Computing final results below -->
 
@@
 
-### Robust Performance Plots
+<!-- NOTE: Generating robust performance plot below -->
 
@@
 ### Confidence Intervals
+
+@tbl-ci present the exact confidence intervals (99%) for the difference in mean outcomes on which we base our assessment of statistical significance in the main paper.
+Grouped by evaluation metrics (Variable) and dataset (Data), the table presents the mean outcomes for CT and BT and finally the lower bound (LB) and upper bound (UB) of the confidence interval.
+To compute the intervals, we used the percentile method for bootstrapped confidence intervals: the lower and upper bound represent the $\alpha/2$- and $(1-\alpha / 2)-$quantile of the bootstrap distribution, respectively, for $\alpha=0.01$.
 
@@
 
-Mean outcomes for **CT** and **BL** along with bootstrapped confidence intervals (99%) for difference in mean outcomes grouped by dataset and evaluation metric.
-Column **LB** and **UB** show the lower and upper bound of the intervals, respectively, and computed using the percentile method.
+Mean outcomes for CT and BL along with bootstrapped confidence intervals (99%) for difference in mean outcomes grouped by dataset and evaluation metric.
+Column LB and UB show the lower and upper bound of the intervals, respectively, and computed using the percentile method.
 The underlying counterfactual evaluations are the same as the ones used to produce the main table in the paper.
@@
 
-::: {#tbl-panel layout-ncol=3}
+In @tbl-panel, we present additional outcomes for common evaluation metrics: @tbl-costs presents the average reduction in costs of counterfactuals for CT vs.
+BL with no mutability constraints, i.e.
+corresponding to the first two columns in the main table of the paper; @tbl-val shows the corresponding average validities; finally, @tbl-val-mtbl shows average validities for the case with mutability constraints, i.e.
+corresponding to the third columns in the main table of the paper.
+
+As noted in the discussion section of the main paper, we observe mixed results results here.
+Average costs in terms of distances from factual values decrease for most datasets, which is positively surprising since improved plausibility requires counterfactuals to travel further into the target domain than minimum distance counterfactuals.
+It appears that in these cases faithful counterfactuals for the baseline model still end up far away from their initial starting points, but not close enough for samples in the target domain to be plausible.
+In that sense, CT can be seen to improve both plausibility and costs for faithful CE.
+In some cases though (*LS*, *CH*, *MNIST*), we do seem to observe the tradeoff between plausibility and costs play out, as we would expect (compare panels (a) and (b) of Figure 1 in the main paper for reference).
+
+Concerning validity, we find that can lead to substantial reductions and only increases average validity compared to the baseline in one case (*Circ*).
+As noted in the discussion section of the main paper, this result does not surprise us: by design, CT shrinks the solution space for valid counterfactual explanations, thus making it "harder" to reach validity compared to the baseline model.
+Note that for a number of reasons this should not be seen as problematic:
+
+1.
+Validity of gradient-based CE is a function on the number of steps and the step size which we both kept fixed during evaluation: simply adjusting $T=50$ to higher values or choosing a larger step size will lead to higher rates of validity.
+2.
+Even though reaching validity is sometimes "harder" in terms of the necessary number of steps for a given step size, we have already shown that the average distances that counterfactuals need to travel decrease for most datasets.
+Users care about costs in terms of feature distances, not search iteration steps.
+3.
+From a philosophical perspective on algorithmic recourse, validity in and off itself is not a sufficient desideratum for counterfactuals.
+In fact, @ven2020philosophical propose introducing an upper bound on costs of the flipset (i.e.
+the set of valid CE), arguing that valid but highly costly counterfactuals are not useful to individuals in practice.
+In a similar fashion, it could be argued that there should be an upper bound on the implausibility of counterfactuals in the flipset.
+
+::: {#tbl-panel layout="[[40,-20,40], [100]]"}
 
@@
 
-Costs
+Reduction in average costs for CT vs. the baseline. Results correspond to the case with no mutability constraints in the main table of the paper.
 
@@
 
-Validity
+Average validities of counterfactuals for CT and BL. Unconstrained case.
 
@@
 
-Validity
+Average validities of counterfactuals for CT and BL. Mutability constrained case.
 

### 2025-07-23 paper/sections/other/training.qmd
add done

--- 2025-03-15 paper/sections/other/training.qmd
@@
 
-@nte-eval-default presents the default hyperparameters used during evaluation.
+### Counterfactual Outcomes
+
+For all of our counterfactual evaluations, we proceed as follows: for each dataset we run $J$ bootstrap rounds ("No.
+Runs") to account for stochasticity (@nte-eval-default); for each bootstrap round, we randomly draw factual and target pairs; then, for each model, we draw samples from the test set (with replacement) for which the model predicts the randomly chosen factual class; finally, we generate multiple counterfactuals ("No.
+Counterfactuals") and evaluate the outcomes (@nte-eval-default).
+This is in line with standard practice in the related literature on CE.
+For our final results presented in the main paper, we rely on held-out test sets for evaluation.
+For tuning purposes we rely on training and/or validation sets.
+
+@nte-eval-default presents the default hyperparameters used during evaluation for tuning purposes. For the main results presented in the paper, we use larger evaluations, specifically:
+
+- "No. Runs": We set the number of bootstrap rounds to $J=100$ for all datasets. 
+- "No.
+Individuals": In each round we draw 1,250, 500 and 125 samples for synthetic datasets, real-world tabular datasets and *MNIST*, respectively, across five different values for the strength of the energy penalty of *ECCCo* at test time, $\lambda_{\text{egy}}\in\{0.1, 0.5, 1.0, 5.0, 10.0\}$.
 
@@
 dict = Logging.with_logger(Logging.NullLogger()) do  
-    CTExperiments.to_dict(EvaluationConfig(ExperimentGrid()))
+    CTExperiments.to_dict(EvaluationGrid(ExperimentGrid()))
 end
-println(CTExperiments.dict_to_quarto_markdown(dict, filter_empty=false))
+println(CTExperiments.dict_to_quarto_markdown(dict["counterfactual_params"], filter_empty=false))
 ```
@@
 :::
+
+### Predictive Performance
+
+To assess (robust) predictive performance, we evaluate model accuracy on (adversarially perturbed) test data.
+To generate adversarial examples we use the Fast Gradient Sign Method (FGSM) [@goodfellow2014explaining].
+For the main results in the paper, we choose a range of values $\epsilon=[0.0,0.1]$.
+In some places of this appendix, you will also find predictive performance evaluations in terms of the F1-score.

### 2025-07-23 paper/experiments/CTExperiments/src/presentation.jl

@@ -261,18 +261,18 @@ end
 
 # Function to create final Markdown string
 function dict_to_markdown(
-    dict::Dict; header::Union{Nothing,String}=nothing, filter_empty::Bool=true
+    dict::Dict; header::Union{Nothing,String}=nothing, filter_empty::Bool=true, kwrgs...
 )
     filtered_dict = filter_dict(dict; filter_empty)
-    return "md\"\"\"\n$(to_mkd(filtered_dict; header=header))\n\"\"\""
+    return "md\"\"\"\n$(to_mkd(filtered_dict; header=header, kwrgs...))\n\"\"\""
 end
 
 # New function specifically for Quarto output
 function dict_to_quarto_markdown(
-    dict::Dict; header::Union{Nothing,String}=nothing, filter_empty::Bool=true
+    dict::Dict; header::Union{Nothing,String}=nothing, filter_empty::Bool=true, kwrgs...
 )
     filtered_dict = filter_dict(dict; filter_empty)
-    return "$(to_mkd(filtered_dict; header=header))\n"
+    return "$(to_mkd(filtered_dict; header=header, kwrgs...))\n"
 end
 
 function adjust_plot_var(x::Union{Nothing,String}, cfg::CTExperiments.EvalConfigOrGrid)

### 2025-07-23 paper/experiments/CTExperiments/src/tables.jl

@@ -25,10 +25,10 @@ function tabulate_results(
     if "data" in names(df) && hlines != :none
         if "Avg." in df.data
             _ds_order = [ds_order..., "Avg."]
-            hlines = [0, 1, 5, length(ds_order)+1, length(ds_order)+2]
+            hlines = [0, 1, 5, length(_ds_order), length(_ds_order)+1]
         else
             _ds_order = ds_order
-            hlines = [0, 1, 5, length(ds_order)+2]
+            hlines = [0, 1, 5, length(_ds_order)+1]
         end
         df.data = categorical(df.data; levels=_ds_order)
         if sort_by_data

### 2025-07-23 paper/preprint/appendix.tex

@@
 
-\newcommand{\runninghead}{A Preprint }
+\newcommand{\runninghead}{}
 \renewcommand{\runninghead}{Counterfactual Training }
@@
 
+\subsection{Variables and Parameters}\label{variables-and-parameters}
+
 Below we provide an overview of some notation used frequently throughout
@@
 \item
+  \(\mathcal{Y}\): The output domain.
+\item
   \(y^+\): The target class and also the index of the target class.
@@
 \item
+  \(\mathcal{X}\): The input domain.
+\item
   \(\mathbf{x}\): a single training sample.
@@
   \(\mathbf{x}^\prime\): a counterfactual.
+\item
+  \(t=1,...,T\): Step indicator for counterfactual search iterations.
+\item
+  \(\mathbf{x}^\prime_{t \leq T}\): a nascent counterfactual, defined as
+  a counterfactual that has not yet converged.
+\item
+  \(\mathbf{x}^\prime_{T}\): a mature counterfactual, defined as a
+  counterfactual that has either passed a threshold probability or
+  exhausted all \(T\) steps.
 \item
@@
 
-\subsection{Other Technical Details}\label{other-technical-details}
+\subsection{Formulas}\label{formulas}
+
+\subsubsection{Maximum Mean Discrepancy}\label{maximum-mean-discrepancy}
 
@@
 
+\subsection{Other Conventions}\label{other-conventions}
+
+In some place of this appendix, we use the terms \emph{full/Full}
+(i.e.~the full CT objective) and \emph{vanilla/Vanilla} (i.e.~vanilla
+training objective) to refer to models trained with counterfactual
+training (\emph{CT}) and the baseline (\emph{BL}), respectively.
+
 \section{Technical Details of Our
@@
 
-In this section, we provide some background on gradient-based
+In this section, we provide some additional background on gradient-based
 counterfactual generators (Section~\ref{sec-app-ce-background}) and
@@
 Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}). It generally solves
-the following unconstrained objective,
+(\citeproc{ref-wachter2017counterfactual}{2017}). A more general version
+of the counterfactual search objective presented in the paper is as
+follows,
 
@@
 that maps from the \(L\)-dimensional counterfactual state space to the
-feature space and \(\text{cost}(\cdot)\) denotes one or more penalties
+feature space and \(\text{reg}(\cdot)\) denotes one or more penalties
 that are used to induce certain properties of the counterfactual
@@
 (\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
-the distance penalty, \emph{ECCCo} uses an additional penalty component
-that regularizes the energy associated with the counterfactual,
-\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
-al. 2024}).
+the distance penalty, \emph{ECCCo} uses a penalty that regularizes the
+energy associated with the counterfactual, \(\mathbf{x}^\prime\)
+(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). We omit the
+conformal set size penalty proposed in the original paper, since a) the
+authors found faithfulness to primarily depend on the energy penalty and
+hence this alleviates us from one additional hyperparameter.
 
@@
 and additional findings in this appendix). Below we derive the
-analytical results in Prp.\textasciitilde{}\ref{prp-mtblty}.
+analytical results in Proposition in the main paper.
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-default}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-train-default}: Training Phase}]
 
@@
 
-For all of our evaluations, we proceed as follows: for each experiment
-setting we generate multiple counterfactuals (``No.~Counterfactuals''),
-randomly choosing the factual and target class each time
-(Note~\ref{nte-eval-default}). We do this across multiple rounds
-(``No.~Runs'') with different random seeds to account for stochasticity
-(Note~\ref{nte-eval-default}). This is in line with standard practice in
-the related literature on CE. Note~\ref{nte-eval-default} presents the
-default hyperparameters used during evaluation. For our final results
-presented in the main paper, we rely on held out test sets to sample
-factuals (and outputs for our performance metrics). For tuning purposes
-we rely on training or validation sets.
-
-\subsubsection{Robust Accuracy}\label{robust-accuracy}
-
-To evaluate robust accuracy (Acc.\(^*\)), we use the Fast Gradient Sign
-Method (FGSM) to perturb test samples
+\subsubsection{Counterfactual Outcomes}\label{counterfactual-outcomes}
+
+For all of our counterfactual evaluations, we proceed as follows: for
+each dataset we run \(J\) bootstrap rounds (``No.~Runs'') to account for
+stochasticity (Note~\ref{nte-eval-default}); for each bootstrap round,
+we randomly draw factual and target pairs; then, for each model, we draw
+samples from the test set (with replacement) for which the model
+predicts the randomly chosen factual class; finally, we generate
+multiple counterfactuals (``No.~Counterfactuals'') and evaluate the
+outcomes (Note~\ref{nte-eval-default}). This is in line with standard
+practice in the related literature on CE. For our final results
+presented in the main paper, we rely on held-out test sets for
+evaluation. For tuning purposes we rely on training and/or validation
+sets.
+
+Note~\ref{nte-eval-default} presents the default hyperparameters used
+during evaluation for tuning purposes. For the main results presented in
+the paper, we use larger evaluations, specifically:
+
+\begin{itemize}
+\tightlist
+\item
+  ``No.~Runs'': We set the number of bootstrap rounds to \(J=100\) for
+  all datasets.
+\item
+  ``No.~Individuals'': In each round we draw 1,250, 500 and 125 samples
+  for synthetic datasets, real-world tabular datasets and \emph{MNIST},
+  respectively, across five different values for the strength of the
+  energy penalty of \emph{ECCCo} at test time,
+  \(\lambda_{\text{egy}}\in\{0.1, 0.5, 1.0, 5.0, 10.0\}\).
+\end{itemize}
+
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}]
+
+\quartocalloutnte{nte-eval-default}
+
+\begin{itemize}
+\tightlist
+\item
+  Convergence: \texttt{threshold}
+\item
+  Decision Threshold: \texttt{0.95}
+\item
+  Maximum Iterations: \texttt{50}
+\item
+  No.~Individuals: \texttt{100}
+\item
+  No.~Runs: \texttt{5}
+\end{itemize}
+
+\end{tcolorbox}
+
+\subsubsection{Predictive Performance}\label{predictive-performance}
+
+To assess (robust) predictive performance, we evaluate model accuracy on
+(adversarially perturbed) test data. To generate adversarial examples we
+use the Fast Gradient Sign Method (FGSM)
 (\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
-2015}). For the main results, we have set the perturbation size to
-\(\epsilon=0.03\). We have also tested other perturbation sizes, as well
-as randomly perturbed data. Although not reported here, we have
-consistently found strong outperformance of CT compared to the weak
-baseline.
-
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-eval-default}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
-
-\quartocalloutnte{nte-eval-default}
-
-\begin{itemize}
-\tightlist
-\item
-  Counterfactual Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Convergence: \texttt{threshold}
-  \item
-    Decision Threshold: \texttt{0.95}
-  \item
-    Generator Parameters:
-
-    \begin{itemize}
-    \tightlist
-    \item
-      Decision Threshold: \texttt{0.75}
-    \item
-      \(\lambda_{\text{cst}}\): \texttt{0.001}
-    \item
-      \(\lambda_{\text{egy}}\): \texttt{5.0}
-    \item
-      Learning Rate: \texttt{0.25}
-    \item
-      Maximum Iterations: \texttt{30}
-    \item
-      Optimizer: \texttt{sgd}
-    \item
-      Type: \texttt{ECCo}
-    \end{itemize}
-  \item
-    Maximum Iterations: \texttt{50}
-  \item
-    No.~Individuals: \texttt{100}
-  \item
-    No.~Runs: \texttt{5}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
+2015}). For the main results in the paper, we choose a range of values
+\(\epsilon=[0.0,0.1]\). In some places of this appendix, you will also
+find predictive performance evaluations in terms of the F1-score.
 
@@
 choices. We study the effect of many hyperparameters extensively in
-Section~\ref{sec-app-grid}. For the main results, we tune a small set of
-key hyperparameters (Section~\ref{sec-app-tune}). The final choices for
-the main results are presented for each data set in
-Table~\ref{tbl-final-params} along with training, test and batch sizes.
+Section~\ref{sec-app-grid} of this appendix. For the main results, we
+tune a small set of key hyperparameters (Section~\ref{sec-app-tune}).
+The final choices for the main results are presented for each data set
+in Table~\ref{tbl-final-params} along with training, test and batch
+sizes.
 
@@
   GMSC & 13371 & 2474 & 1000 & none & 0.5 & 5000 & 0.5 \\
-  MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\
+  MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\\bottomrule
 \end{tabular}
@@
 
-\subsection{Final Results}\label{sec-app-final-results}
-
-Plus/minus two standard deviations of bootstrap estimates.
-
-\subsubsection{Robust Performance Plots}\label{robust-performance-plots}
-
 \subsubsection{Confidence Intervals}\label{confidence-intervals}
 
+Table~\ref{tbl-ci} present the exact confidence intervals (99\%) for the
+difference in mean outcomes on which we base our assessment of
+statistical significance in the main paper. Grouped by evaluation
+metrics (Variable) and dataset (Data), the table presents the mean
+outcomes for CT and BT and finally the lower bound (LB) and upper bound
+(UB) of the confidence interval. To compute the intervals, we used the
+percentile method for bootstrapped confidence intervals: the lower and
+upper bound represent the \(\alpha/2\)- and \((1-\alpha / 2)-\)quantile
+of the bootstrap distribution, respectively, for \(\alpha=0.01\).
+
 \begin{table}
 
-\caption{\label{tbl-ci}Mean outcomes for \textbf{CT} and \textbf{BL}
-along with bootstrapped confidence intervals (99\%) for difference in
-mean outcomes grouped by dataset and evaluation metric. Column
-\textbf{LB} and \textbf{UB} show the lower and upper bound of the
-intervals, respectively, and computed using the percentile method. The
-underlying counterfactual evaluations are the same as the ones used to
-produce the main table in the paper.}
+\caption{\label{tbl-ci}Mean outcomes for CT and BL along with
+bootstrapped confidence intervals (99\%) for difference in mean outcomes
+grouped by dataset and evaluation metric. Column LB and UB show the
+lower and upper bound of the intervals, respectively, and computed using
+the percentile method. The underlying counterfactual evaluations are the
+same as the ones used to produce the main table in the paper.}
 
@@
   GMSC & 1.0 \pm 0.0 & 0.21 \pm 0.03 \\
-  MNIST & 0.18 \pm 0.01 & 0.41 \pm 0.01 \\
+  MNIST & 0.18 \pm 0.01 & 0.41 \pm 0.01 \\\bottomrule
 \end{tabular}
@@
 
+In Table~\ref{tbl-panel}, we present additional outcomes for common
+evaluation metrics: Table~\ref{tbl-costs} presents the average reduction
+in costs of counterfactuals for CT vs.~BL with no mutability
+constraints, i.e.~corresponding to the first two columns in the main
+table of the paper; Table~\ref{tbl-val} shows the corresponding average
+validities; finally, Table~\ref{tbl-val-mtbl} shows average validities
+for the case with mutability constraints, i.e.~corresponding to the
+third columns in the main table of the paper.
+
+As noted in the discussion section of the main paper, we observe mixed
+results results here. Average costs in terms of distances from factual
+values decrease for most datasets, which is positively surprising since
+improved plausibility requires counterfactuals to travel further into
+the target domain than minimum distance counterfactuals. It appears that
+in these cases faithful counterfactuals for the baseline model still end
+up far away from their initial starting points, but not close enough for
+samples in the target domain to be plausible. In that sense, CT can be
+seen to improve both plausibility and costs for faithful CE. In some
+cases though (\emph{LS}, \emph{CH}, \emph{MNIST}), we do seem to observe
+the tradeoff between plausibility and costs play out, as we would expect
+(compare panels (a) and (b) of Figure 1 in the main paper for
+reference).
+
+Concerning validity, we find that can lead to substantial reductions and
+only increases average validity compared to the baseline in one case
+(\emph{Circ}). As noted in the discussion section of the main paper,
+this result does not surprise us: by design, CT shrinks the solution
+space for valid counterfactual explanations, thus making it ``harder''
+to reach validity compared to the baseline model. Note that for a number
+of reasons this should not be seen as problematic:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Validity of gradient-based CE is a function on the number of steps and
+  the step size which we both kept fixed during evaluation: simply
+  adjusting \(T=50\) to higher values or choosing a larger step size
+  will lead to higher rates of validity.
+\item
+  Even though reaching validity is sometimes ``harder'' in terms of the
+  necessary number of steps for a given step size, we have already shown
+  that the average distances that counterfactuals need to travel
+  decrease for most datasets. Users care about costs in terms of feature
+  distances, not search iteration steps.
+\item
+  From a philosophical perspective on algorithmic recourse, validity in
+  and off itself is not a sufficient desideratum for counterfactuals. In
+  fact, Venkatasubramanian and Alfano
+  (\citeproc{ref-ven2020philosophical}{2020}) propose introducing an
+  upper bound on costs of the flipset (i.e.~the set of valid CE),
+  arguing that valid but highly costly counterfactuals are not useful to
+  individuals in practice. In a similar fashion, it could be argued that
+  there should be an upper bound on the implausibility of
+  counterfactuals in the flipset.
+\end{enumerate}
+
 \begin{table}
@@
 
-\begin{minipage}{0.33\linewidth}
-
-\subcaption{\label{tbl-costs}Costs}
+\begin{minipage}{0.40\linewidth}
+
+\subcaption{\label{tbl-costs}Reduction in average costs for CT vs.~the
+baseline. Results correspond to the case with no mutability constraints
+in the main table of the paper.}
 
@@
 %
-\begin{minipage}{0.33\linewidth}
-
-\subcaption{\label{tbl-val}Validity}
+\begin{minipage}{0.20\linewidth}
+~\end{minipage}%
+%
+\begin{minipage}{0.40\linewidth}
+
+\subcaption{\label{tbl-val}Average validities of counterfactuals for CT
+and BL. Unconstrained case.}
 
@@
   GMSC & 0.63 & 1.0 \\
-  MNIST & 1.0 & 1.0 \\
+  MNIST & 1.0 & 1.0 \\\bottomrule
 \end{tabular}
@@
 \end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\subcaption{\label{tbl-val-mtbl}Validity}
+\newline
+\begin{minipage}{\linewidth}
+
+\subcaption{\label{tbl-val-mtbl}Average validities of counterfactuals
+for CT and BL. Mutability constrained case.}
 
@@
   GMSC & 0.38 & 1.0 \\
-  MNIST & 1.0 & 1.0 \\
+  MNIST & 1.0 & 1.0 \\\bottomrule
 \end{tabular}
@@
 
-\subsubsection{Predictive Performance}\label{predictive-performance}
+\subsubsection{Predictive Performance}\label{predictive-performance-1}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]
 
@@
 
-\subsubsection{Predictive Performance}\label{predictive-performance-1}
+\subsubsection{Predictive Performance}\label{predictive-performance-2}
 
@@
 change from the baseline model for the distance-based implausibility
-metric (\$ ext\{IP\}\$). Boxplots indicate the variation across
+metric (\(\text{IP}\)). Boxplots indicate the variation across
 evaluation runs and test settings (varying parameters for \emph{ECCCo}).
@@
 plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric (\$
-ext\{IP\}\$). Boxplots indicate the variation across evaluation runs and
-test settings (varying parameters for \emph{ECCCo}). Data: Moons.}
+from the baseline model for the distance-based implausibility metric
+(\(\text{IP}\)). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCCo}). Data: Moons.}
 
@@
 plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric (\$
-ext\{IP\}\$). Boxplots indicate the variation across evaluation runs and
-test settings (varying parameters for \emph{ECCCo}). Data: Overlapping.}
+from the baseline model for the distance-based implausibility metric
+(\(\text{IP}\)). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCCo}). Data:
+Overlapping.}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}]
 
@@
 
-\subsubsection{Predictive Performance}\label{predictive-performance-2}
+\subsubsection{Predictive Performance}\label{predictive-performance-3}
 
@@
 plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric (\$
-ext\{IP\}\$). Boxplots indicate the variation across evaluation runs and
-test settings (varying parameters for \emph{ECCCo}). Data: Circles.}
+from the baseline model for the distance-based implausibility metric
+(\(\text{IP}\)). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCCo}). Data: Circles.}
 
@@
 plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric (\$
-ext\{IP\}\$). Boxplots indicate the variation across evaluation runs and
-test settings (varying parameters for \emph{ECCCo}). Data: Linearly
+from the baseline model for the distance-based implausibility metric
+(\(\text{IP}\)). Boxplots indicate the variation across evaluation runs
+and test settings (varying parameters for \emph{ECCCo}). Data: Linearly
 Separable.}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}]
 
@@
 
-\subsubsection{Predictive Performance}\label{predictive-performance-3}
+\subsubsection{Predictive Performance}\label{predictive-performance-4}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}]
 
@@
 
+\FloatBarrier
+
 \subsection{Learning Rate}\label{sec-app-tune-lr}
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-train}: Training Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, rightrule=.15mm, bottomtitle=1mm, colframe=quarto-callout-note-color-frame, colback=white, toptitle=1mm, arc=.35mm, left=2mm, toprule=.15mm, coltitle=black, opacitybacktitle=0.6, breakable, titlerule=0mm, opacityback=0, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, bottomrule=.15mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}]
 
@@
 across multiple central processing units (CPU). Graphical processing
-units (GPU) were not required.
+units (GPU) were \emph{not} used.
 
@@
 parallelized across 20 CPUs with 3GB memory each. Evaluations for all
-data sets took less than one hour (\textless1h) to complete.
+data sets took less than one hour (\textless1h) to complete but were
+generally more memory-intensive (see Section~\ref{sec-app-software} for
+additional details)
 
@@
 
-\subsection{Software}\label{software}
+\subsection{Software}\label{sec-app-software}
 
@@
 for deep learning (\citeproc{ref-innes2018fashionable}{Michael Innes et
-al. 2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). For
-data-wrangling and presentation-ready tables we relied on
+al. 2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). We chose to
+work with
+\href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{CounterfactualExplanations.jl}
+because it currently appears to be the most comprehensive and extensible
+package for counterfactual explanations. Despite its good interplay with
+Flux.jl, the package is not, however, optimized to be used in training.
+This has caused some issues with memory management and bottlenecked
+performance. All of our code will be provided to reviewers of this work
+and open-sourced upon publication.
+
+For data-wrangling and presentation-ready tables we relied on
 \href{https://dataframes.juliadata.org/v1.7/}{DataFrames.jl}
@@
 
+\subsection{Reproducibility}\label{reproducibility}
+
+We have taken care to set random seeds for reproducibility.
+
 \end{appendices}
@@
 
+\bibitem[\citeproctext]{ref-ven2020philosophical}
+Venkatasubramanian, Suresh, and Mark Alfano. 2020. {``The Philosophical
+Basis of Algorithmic Recourse.''} In \emph{Proceedings of the 2020
+Conference on Fairness, Accountability, and Transparency}, 284--93.
+
 \bibitem[\citeproctext]{ref-wachter2017counterfactual}

### 2025-07-23 paper/sections/other/compute.qmd
formatting improved and toc added (to this end, turned appendix into standalone document).

@@
 <!-- [@DHPC2022] -->
-## Hardware
+## Hardware {#sec-app-hardware}
 
@@
 Since our experiments involve highly parallel tasks and rather small models by today's standard, we have relied on distributed computing across multiple central processing units (CPU).
-Graphical processing units (GPU) were not required.
+Graphical processing units (GPU) were *not* used.
 
@@
 Model training for the largest grid searches with 270 unique parameter combinations was parallelized across 34 CPUs with 2GB memory each.
-The time to completion varied by dataset for reasons discussed in @sec-discussion: 0h49m (*Moons*), 1h4m (*Linearly Separable*), 1h49m (*Circles*), 3h52m (*Overlapping*).
+The time to completion varied by dataset: 0h49m (*Moons*), 1h4m (*Linearly Separable*), 1h49m (*Circles*), 3h52m (*Overlapping*).
 Model evaluations for large grid searches were parallelized across 20 CPUs with 3GB memory each.
-Evaluations for all data sets took less than one hour (<1h) to complete.
+Evaluations for all data sets took less than one hour (<1h) to complete but were generally more memory-intensive (see @sec-app-software for additional details)
 
@@
 
-## Software
+## Software {#sec-app-software}
 
@@
 We have developed a package for counterfactual training that leverages and extends the functionality provided by several existing packages, most notably [CounterfactualExplanations.jl](https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl) [@altmeyer2023explaining] and the [Flux.jl](https://fluxml.ai/Flux.jl/v0.16/) library for deep learning [@innes2018fashionable;@innes2018flux].
+We chose to work with [CounterfactualExplanations.jl](https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl) because it currently appears to be the most comprehensive and extensible package for counterfactual explanations.
+Despite its good interplay with Flux.jl, the package is not, however, optimized to be used in training.
+This has caused some issues with memory management and bottlenecked performance.
+All of our code will be provided to reviewers of this work and open-sourced upon publication.
+
 For data-wrangling and presentation-ready tables we relied on [DataFrames.jl](https://dataframes.juliadata.org/v1.7/) [@milan2023dataframes] and [PrettyTables.jl](https://ronisbr.github.io/PrettyTables.jl/v2.4/) [@chagas2024pretty], respectively.
@@
 To distribute computational tasks across multiple processors, we have relied on [MPI.jl](https://juliaparallel.org/MPI.jl/v0.20/) [@byrne2021mpi].
+
+## Reproducibility
+
+We have taken care to set random seeds for reproducibility.

### 2025-07-23 README.qmd
adjustments to README

@@
 
-# CounterfactualTraining
+# CounterfactualTraining.jl
 
-[![Stable](https://img.shields.io/badge/docs-stable-blue.svg)](https://JuliaTrustworthyAI.github.io/CounterfactualTraining.jl/stable/)
-[![Dev](https://img.shields.io/badge/docs-dev-blue.svg)](https://JuliaTrustworthyAI.github.io/CounterfactualTraining.jl/dev/)
-[![Build Status](https://github.com/JuliaTrustworthyAI/CounterfactualTraining.jl/actions/workflows/CI.yml/badge.svg?branch=main)](https://github.com/JuliaTrustworthyAI/CounterfactualTraining.jl/actions/workflows/CI.yml?query=branch%3Amain)
-[![Coverage](https://codecov.io/gh/JuliaTrustworthyAI/CounterfactualTraining.jl/branch/main/graph/badge.svg)](https://codecov.io/gh/JuliaTrustworthyAI/CounterfactualTraining.jl)
-[![Code Style: Blue](https://img.shields.io/badge/code%20style-blue-4495d1.svg)](https://github.com/invenia/BlueStyle)
-[![Aqua](https://raw.githubusercontent.com/JuliaTesting/Aqua.jl/master/badge.svg)](https://github.com/JuliaTesting/Aqua.jl)
+This repository contains all the code used to generate the experiments for the paper *Counterfactual Training: Teaching Models Plausible and Actionable Explanations*.
+
+## Structure
+
+The repository is structured like a standard Julia package with the main code base in `src/`, the documentation in `docs/` and the test suite in `test/`^[The latter two are both still incomplete].
+Everything relating specifically to the paper lives under `paper/`.
+
+### Code Base in `src/`
+
+The main code base for the CounterfactualTraining.jl package in `src/` is relatively simple and small.
+It implements the core functionality for counterfactual training.
+The code is commented with reference to relevant parts in the paper.
+
+### Code in `paper/`
+
+The majority of code scripts live in `paper/experiments/`, which is a self-sufficient Julia project that depends on CounterfactualTraining.jl in the root folder.
+The code scripts in `paper/experiments` include Julia and bash scripts used to implement and run the experiments specific to this paper.
+The project depends on another self-contained helper package that lives in `paper/experiments/CTExperiments/`.
+This helper package makes custom functions, structs and other definitions reusable in the project (`paper/experiments/`).
+All package dependencies are automatically handled in scripts using Julia's package manager, Pkg.jl.
+
+## Reproducibility
+
+If you want to reproduce the results presented in our paper and the supplementary appendix, please see the [paper/experiments/README](/paper/experiments/README.md) for details.

### 2025-07-23 paper/experiments/README.qmd
adjustments to README

@@
   exeflags: ["--project=./"]
+execute:
+  eval: false
+  echo: true
 ---
@@
 
-Next, we load the `CTExperiments` package---a small package that ships some additional functionality relevant only to the experiments run for the paper.
+Next, we load the `CTExperiments` package---a helper package that ships some additional functionality relevant only to the experiments run for the paper.
 
@@
 :::{.callout-warning}
-The extra arguments `--data=mydata --model=mymodel` should be consistent, i.e. they should be passed consitently to `determine_resources` and the whole `sbatch` command. 
+The extra arguments `--data=mydata --model=mymodel` should be consistent, i.e. they should be passed consistently to `determine_resources` and the whole `sbatch` command. 
 :::

### 2025-07-23 README.qmd
READMEs filled with content

@@
 It implements the core functionality for counterfactual training.
-The code is commented with reference to relevant parts in the paper.
+The code is commented with reference to relevant parts in the paper (look for `# ----- PAPER REF -----`).
 
@@
 The majority of code scripts live in `paper/experiments/`, which is a self-sufficient Julia project that depends on CounterfactualTraining.jl in the root folder.
-The code scripts in `paper/experiments` include Julia and bash scripts used to implement and run the experiments specific to this paper.
+The code scripts in `paper/experiments` include Julia, bash and TOML scripts used to implement, run and configure the experiments specific to this paper.
 The project depends on another self-contained helper package that lives in `paper/experiments/CTExperiments/`.

### 2025-07-23 paper/experiments/README.qmd
READMEs filled with content

@@
 
-## Command Line Usage
+## Running Experiments
+
+The Julia functions that live directly under `paper/experiments/` can be used to run the experiments. The main scripts of interest are:
+
+- `run_grid.jl` and `run_grid_sequentially.jl` which can be used to train model architectures on data according to specifications in configuration files (TOML).
+- `run_evaluation_grid.jl` and  `run_evaluation_grid_sequentially.jl` which can be used to evaluate previously run experiments.
+
+The `_sequentially` suffix indicates that experiments are run sequentially and tasks underlying experiments are distributed across processes (as opposed to distributed experiments across processes).
+
+### Configurations
+
+To configure and keep track of hyperparameters for differenct experiments, we use TOML files.
+The configurations used for our various experiments can be found in `paper/experiments/configs/`.
+The following files are of particular relevance:
+
+- `single.toml`: configures the main experiments that produced the main results without mutability constraints.
+- `mutability_*.toml`: configures the main experiments that produced the main results with mutability constraints.
+
+The remaining configuration files include those for extensive grid searches used for hyperparameter tuning.
+
+### Command Line Usage
+
+The Julia scripts introduced above look for certain environment variables pertaining to the configurations of experiments and evaluations, as well as paths to directories that will used to store results.
+Before running batch jobs, you therefore want to specify these variables as illustrated in the following example:
 
 ```
-sbatch --ntasks=$(determine_resources) --cpus-per-task=$NTHREADS myfile.sh
+export CONFIG=paper/experiments/configs/single.toml   # path to configuration file
+export OUTPUT_SUBDIR="final_run"                      # path to store outputs (root directory is /paper/experiments/output by default)
 ```
 
+You can then run scripts directly from the command line like so:
+
 ```
-sbatch --ntasks=$(determine_resources --data=mydata --model=mymodel) --cpus-per-task=$NTHREADS myfile.sh --data=mydata --model=mymodel
+julia --project=$EXPERIMENT_DIR $EXPERIMENT_DIR/run_grid.jl --config=$CONFIG --data="lin_sep" --model="mlp"
 ```
 
-:::{.callout-warning}
-The extra arguments `--data=mydata --model=mymodel` should be consistent, i.e. they should be passed consistently to `determine_resources` and the whole `sbatch` command. 
-:::
+Note that it is possible to provide extra arguments (here `--data` or `--model`) to overwrite specifications in the configuration file.
+To understand what arguments you can provide, we recommend familiarizing yourself with the code base in `paper/experiments/CTExperiments/`.
 
-## Interactive Usage
+## Other Things
 
-### Running a single experiment
+Some anecdotal or illustrative results were produced in simple Julia scripts or interactive notebooks.
+In fact, many of the Quarto notebooks (.qmd) in `paper/sections/other/` contain code cells that produce tables and charts.
+Note that many of these depend on experiment output being available that is not shipped with this repository and therefore first needs to be created.
+We will release the experiment outputs upon publication in a designated data repository, fit to handle large amounts of data.
 
-The `CTExperiments` package ships with a helper function that can be used to set up the default experiment configuration:
+### Figure 1
 
-```{julia}
-config_file = generate_template()
-```
+The chart in Figure 1 of the paper is produced in [paper/sections/other/constraints.qmd](paper/sections/other/constraints.qmd).
 
-An experiment can be generated directly from that configuration file:
+### Figure 2
 
-```{julia}
-experiment = Experiment(config_file)
-```
+The chart in Figure 2 of the paper can be reproduced using the following script: [paper/experiments/mnist_chart.jls](/paper/experiments/mnist_chart.jls).
+
+### Figure 3
+
+The chart in Figure 3 of the paper is produced in [paper/sections/other/main_results.qmd](paper/sections/other/main_results.qmd).

### 2025-07-23 paper/experiments/CTExperiments/src/tables.jl

@@ -5,7 +5,8 @@ using PrettyTables
 using StatsBase
 
 function tabulate_results(
-    inputs, al::Nothing=nothing;
+    inputs,
+    al::Nothing=nothing;
     io::IO=stdout,
     tf=PrettyTables.tf_latex_booktabs,
     wrap_table=true,
@@ -37,7 +38,7 @@ function tabulate_results(
     end
 
     if hlines == :none
-    hlines = nothing
+        hlines = nothing
     end
 
     other_inputs = inputs[2]
@@ -90,7 +91,6 @@ function tabulate_results(
 end
 
 function tabulate_results(inputs, al::Vector{String}; kwargs...)
-
     if isnothing(al)
         return tabulate_results(inputs; kwargs...)
     end
@@ -99,12 +99,13 @@ function tabulate_results(inputs, al::Vector{String}; kwargs...)
     io = IOBuffer()
     tabulate_results(inputs; io=io, kwargs...)
     output = String(take!(io))
-    
+
     # Replace alignment with siunitx format
-    modified = replace(output, 
-        r"\\begin\{tabular\}\{[^}]+\}" => 
-                       "\\begin{tabular}{\n$( reduce((x,y) -> "$x\n$y", al))\n}")
-    
+    modified = replace(
+        output,
+        r"\\begin\{tabular\}\{[^}]+\}" => "\\begin{tabular}{\n$( reduce((x,y) -> "$x\n$y", al))\n}",
+    )
+
     print(modified)
 end
 

### 2025-07-23 paper/experiments/mnist_chart.jl

@@ -24,7 +24,9 @@ Images.save("paper/figures/mnist_ig.png", img_ig_full)
 
 chosen_digits = 5:9
 
-img_ig = mosaicview(imgs_bl[chosen_digits .+ 1]..., imgs_ct[chosen_digits .+ 1]...; nrow=2, rowmajor=true)
+img_ig = mosaicview(
+    imgs_bl[chosen_digits .+ 1]..., imgs_ct[chosen_digits .+ 1]...; nrow=2, rowmajor=true
+)
 
 ## Counterfactuals:
 Random.seed!(42)    # change seed for different outcome
@@ -68,6 +70,5 @@ imgs_ct = imgs[1]
 imgs_bl = imgs[2]
 img_ce = mosaicview(imgs_bl..., imgs_ct...; nrow=2, rowmajor=true)
 
-img = mosaicview(img_ce, img_ig, nrow=1)
+img = mosaicview(img_ce, img_ig; nrow=1)
 Images.save("paper/figures/mnist_body.png", img)
-

### 2025-07-23 paper/preprint/appendix.tex
one final render:

@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-train-default}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-default}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-eval-default}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-tune-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, coltitle=black, arc=.35mm, titlerule=0mm, colback=white, leftrule=.75mm, opacityback=0, colframe=quarto-callout-note-color-frame, breakable, toprule=.15mm, left=2mm, opacitybacktitle=0.6, bottomrule=.15mm, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}]
+\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
 
@@
 performance. All of our code will be provided to reviewers of this work
-and open-sourced upon publication.
+and open-sourced upon publication. The code is commented with clearly
+marked references to the paper (look for
+\texttt{\#\ -\/-\/-\/-\/-\ PAPER\ REF\ -\/-\/-\/-\/-}).
 

### 2025-07-23 paper/preprint/appendix.tex
one final note on random seeds in appendix

@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-default}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-default}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-eval-default}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-eval-default}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-train}: Training Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, rightrule=.15mm, left=2mm, coltitle=black, colback=white, breakable, leftrule=.75mm, bottomtitle=1mm, arc=.35mm, opacityback=0, titlerule=0mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, bottomrule=.15mm, toptitle=1mm]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
 
@@
 
-We have taken care to set random seeds for reproducibility.
+We have taken care to set random seeds for reproducibility using Julia's
+Random.jl package from the standard library. A global seed and (if
+applicable or wanted) dataset-specific seeds can be specified in TOML
+configuration files, environment variables or in interactive Julia
+sessions. Additional details can be found in the code base.
 

### 2025-07-23 paper/sections/other/compute.qmd
READMEs filled with content
one final note on random seeds in appendix

@@
 All of our code will be provided to reviewers of this work and open-sourced upon publication.
+The code is commented with clearly marked references to the paper (look for `# ----- PAPER REF -----`).
 
@@
 
-We have taken care to set random seeds for reproducibility.
+We have taken care to set random seeds for reproducibility using Julia's Random.jl package from the standard library.
+A global seed and (if applicable or wanted) dataset-specific seeds can be specified in TOML configuration files, environment variables or in interactive Julia sessions.
+Additional details can be found in the code base.

### 2025-07-24 paper/preprint/appendix.tex
added text for IG

@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-default}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-default}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-eval-default}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
+\subsubsection{Integrated Gradients}\label{integrated-gradients}
+
+We make use of integrated gradients (IG) proposed by Sundararajan, Taly,
+and Yan (\citeproc{ref-sundararajan2017ig}{2017}) to empirically
+evaluate the feature protection mechanism in CT. We choose this approach
+because it produces theoretically sound results, works well for
+non-linear models, and remains relatively inexpensive.
+
+IG calculates the contribution of each input feature towards a specific
+prediction by approximating the integral of the model output with
+respect to its input, using a set of samples that linearly interpolate
+between a test instance and some baseline instance
+(\citeproc{ref-sundararajan2017ig}{Sundararajan, Taly, and Yan 2017}).
+This process produces a vector of real numbers, one per input feature,
+which informs about the contribution of each feature to the prediction.
+For example:
+
+\begin{itemize}
+\tightlist
+\item
+  a large positive value indicates that a feature has strong positive
+  influence on the classification (i.e., increases the score for a
+  class);
+\item
+  a small negative value indicates that a feature has weak negative
+  influence on the classification (i.e., decreases the score for a
+  class).
+\end{itemize}
+
+To calculate the contributions, IG compares the output to a baseline.
+The selection of an appropriate baseline is an important design decision
+--- it should produce a ``neutral'' prediction to avoid capturing
+effects that cannot be directly attributed to the model
+(\citeproc{ref-sundararajan2017ig}{Sundararajan, Taly, and Yan 2017};
+\citeproc{ref-sturmfels2020visualizing}{Sturmfels, Lundberg, and Lee
+2020}). To remain consistent in our evaluations, we use a baseline drawn
+at random from a uniform distribution, \(\mathcal{U}(-1,1)\) all
+datasets. This aligns with standard evaluation practices for IG.
+
+We run IG on models trained on all datasets to compare their sensitivity
+to features that were protected using CT:
+
+\begin{itemize}
+\tightlist
+\item
+  for synthetic datasets, this is always the first feature
+\item
+  for real-world tabular datasets, this is always \emph{age}
+\item
+  for MNIST, this is first five and last five rows of pixels
+\end{itemize}
+
+As IG outputs are not bounded (i.e., they are arbitrary real numbers),
+it becomes a challenge to meaningfully compare IG outputs of different
+models --- ones that are trained conventionally, and ones that underwent
+counterfactual training. For our purposes, we observe that the direction
+of contribution (sign) does not matter. If protecting a feature flips
+the sign of the corresponding output, then CT „overcorrected'' for this
+feature, but as long as the magnitude of its contribution decreases,
+then CT achieved its objective according to our formulation. Thus, we
+focus on the magnitude of IG outputs in the evaluation. Next, to
+accommodate for variable ranges of outputs, we standardize the outputs.
+
+Let \(\mathbb{g}_d\) denote the estimated IG for feature \(d\). Then in
+the case of 2D synthetic datasets we find that taking the absolute value
+of the outputs, \(|\mathbb{g}_d|\), and then dividing them by a
+\(\max(\mathbb{g}) - \min(\mathbb{g})\) term allows us to make the most
+meaningful comparison. In the case of real-world datasets we choose to
+normalize the values to a 0-1 range instead.
+
+We compare the (average) sensitivity to the features that were protected
+for CT models. We again use bootstrapping (100 rounds, 2500 samples per
+round) to establish the significance of our results. These are presented
+in Table~\ref{tbl-ig}.
+
+We observe substantial and statistically significant improvements for
+seven out of nine datasets. For the synthetic datasets we observe
+reduction in sensitivity to the protected feature in 3 out of 4 cases.
+The only dataset negatively impacted by CT --- albeit with highly
+variable results --- is \emph{Moon}, which clearly violates the
+assumptions in Proposition 3.1. For the real-world datasets, the
+reductions are over two-fold for \emph{Adult} and \emph{CH} and almost
+three-fold for \emph{MNIST}. In case of \emph{Cred} we can fully prevent
+the CT model from considering \emph{age} as a factor in classification.
+Only for \emph{GMSC} we observe negative impacts of CT which we believe
+is due to any or all of the following: a) data assumptions are violated;
+b) the impact of other components of the CT objective outweighs expected
+effects of feature protection; or c) the default baseline choice applied
+consistently to all datasets is not appropriate for GMSC.
+
 \begin{figure}
@@
 \begin{minipage}{0.50\linewidth}
-
-\subsubsection{Integrated Gradients}\label{integrated-gradients}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, left=2mm, opacitybacktitle=0.6, breakable, titlerule=0mm, arc=.35mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, coltitle=black, leftrule=.75mm, bottomtitle=1mm, toptitle=1mm, toprule=.15mm, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
 
@@
 
+\bibitem[\citeproctext]{ref-sturmfels2020visualizing}
+Sturmfels, Pascal, Scott Lundberg, and Su-In Lee. 2020. {``Visualizing
+the Impact of Feature Attribution Baselines.''} \emph{Distill} 5 (1):
+e22.
+
+\bibitem[\citeproctext]{ref-sundararajan2017ig}
+Sundararajan, Mukund, Ankur Taly, and Qiqi Yan. 2017. {``Axiomatic
+Attribution for Deep Networks.''}
+\url{https://arxiv.org/abs/1703.01365}.
+
 \bibitem[\citeproctext]{ref-ven2020philosophical}

### 2025-07-24 paper/sections/other/main_results.qmd
added text for IG

@@
 
+### Integrated Gradients
+
+We make use of integrated gradients (IG) proposed by @sundararajan2017ig to empirically evaluate the feature protection mechanism in CT.
+We choose this approach because it produces theoretically sound results, works well for non-linear models, and remains relatively inexpensive.
+
+IG calculates the contribution of each input feature towards a specific prediction by approximating the integral of the model output with respect to its input, using a set of samples that linearly interpolate between a test instance and some baseline instance [@sundararajan2017ig].
+This process produces a vector of real numbers, one per input feature, which informs about the contribution of each feature to the prediction.
+For example:
+
+- a large positive value indicates that a feature has strong positive influence on the classification (i.e., increases the score for a class);
+- a small negative value indicates that a feature has weak negative influence on the classification (i.e., decreases the score for a class).
+
+To calculate the contributions, IG compares the output to a baseline.
+The selection of an appropriate baseline is an important design decision — it should produce a "neutral" prediction to avoid capturing effects that cannot be directly attributed to the model [@sundararajan2017ig; @sturmfels2020visualizing].
+To remain consistent in our evaluations, we use a baseline drawn at random from a uniform distribution, $\mathcal{U}(-1,1)$ all datasets.
+This aligns with standard evaluation practices for IG.
+
+We run IG on models trained on all datasets to compare their sensitivity to features that were protected using CT:
+
+- for synthetic datasets, this is always the first feature
+- for real-world tabular datasets, this is always *age*
+- for MNIST, this is first five and last five rows of pixels
+
+As IG outputs are not bounded (i.e., they are arbitrary real numbers), it becomes a challenge to meaningfully compare IG outputs of different models — ones that are trained conventionally, and ones that underwent counterfactual training.
+For our purposes, we observe that the direction of contribution (sign) does not matter.
+If protecting a feature flips the sign of the corresponding output, then CT „overcorrected” for this feature, but as long as the magnitude of its contribution decreases, then CT achieved its objective according to our formulation.
+Thus, we focus on the magnitude of IG outputs in the evaluation.
+Next, to accommodate for variable ranges of outputs, we standardize the outputs.
+
+Let $\mathbb{g}_d$ denote the estimated IG for feature $d$.
+Then in the case of 2D synthetic datasets we find that taking the absolute value of the outputs, $|\mathbb{g}_d|$, and then dividing them by a $\max(\mathbb{g}) - \min(\mathbb{g})$ term allows us to make the most meaningful comparison.
+In the case of real-world datasets we choose to normalize the values to a 0-1 range instead.
+
+We compare the (average) sensitivity to the features that were protected for CT models.
+We again use bootstrapping (100 rounds, 2500 samples per round) to establish the significance of our results.
+These are presented in @tbl-ig.
+
+We observe substantial and statistically significant improvements for seven out of nine datasets.
+For the synthetic datasets we observe reduction in sensitivity to the protected feature in 3 out of 4 cases.
+The only dataset negatively impacted by CT — albeit with highly variable results — is *Moon*, which clearly violates the assumptions in Proposition 3.1.
+For the real-world datasets, the reductions are over two-fold for *Adult* and *CH* and almost three-fold for *MNIST*.
+In case of *Cred* we can fully prevent the CT model from considering *age* as a factor in classification.
+Only for *GMSC* we observe negative impacts of CT which we believe is due to any or all of the following: a) data assumptions are violated; b) the impact of other components of the CT objective outweighs expected effects of feature protection; or c) the default baseline choice applied consistently to all datasets is not appropriate for GMSC.
+
 ::: {layout-ncol=2}
-
-### Integrated Gradients
 

### 2025-07-24 paper/experiments/README.qmd
done

@@
 The Julia scripts introduced above look for certain environment variables pertaining to the configurations of experiments and evaluations, as well as paths to directories that will used to store results.
+Some of the environment variables specified in a `.env` file.
 Before running batch jobs, you therefore want to specify these variables as illustrated in the following example:
@@
 ```
+source .env                                           # declares a few environment variables
 export CONFIG=paper/experiments/configs/single.toml   # path to configuration file
@@
 
+### Cluster Jobs
+
+To run jobs on a cluster, you can use any of bash scripts under `paper/experiments/jobs/`.
+Assuming you have already exported environment variables as discussed above, you can run cluster jobs as follows:
+
+```
+sbatch --time=00:30:00 --ntasks=20 --cpus-per-task=$NTHREADS $RUN_GRID_SEQ --model=mlp --tau=0.5 --reg_strength=0.01 --data=lin_sep
+```
+
+Here we have specified a run time, `--time`, the number of processes to be used, `--ntasks`, and how many threads to use (defaults to 1).
+The `$RUN_GRID_SEQ` variable points to `paper/experiments/jobs/run_grid_sequentially.sh`.
+The remaining variables are supplied as `ARGS` to Julia and can be used to configure the experiment, as already discussed above.
+
 ## Other Things
@@
 
-The chart in Figure 1 of the paper is produced in [paper/sections/other/constraints.qmd](paper/sections/other/constraints.qmd).
+The chart in Figure 1 of the paper is produced in [paper/sections/other/constraints.qmd](../sections/other/constraints.qmd).
 
@@
 
-The chart in Figure 2 of the paper can be reproduced using the following script: [paper/experiments/mnist_chart.jls](/paper/experiments/mnist_chart.jls).
+The chart in Figure 2 of the paper can be reproduced using the following script: [paper/experiments/mnist_chart.jls](mnist_chart.jls).
 
@@
 
-The chart in Figure 3 of the paper is produced in [paper/sections/other/main_results.qmd](paper/sections/other/main_results.qmd).
+The chart in Figure 3 of the paper is produced in [paper/sections/other/main_results.qmd](../sections/other/main_results.qmd).
+Integrated gradients were computed using [paper/experiments/integrated_gradients_mnist.jl](integrated_gradients_mnist.jl)
+
+### Integrated Gradients
+
+The results for integrated gradients presented in the appendix can be reproduced using [paper/experiments/run_integrated_gradients.jl](run_integrated_gradients.jl).

### 2025-07-24 paper/preprint/appendix.tex
done

@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-default}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-default}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-Let \(\mathbb{g}_d\) denote the estimated IG for feature \(d\). Then in
+Let \(\mathbf{g}_d\) denote the estimated IG for feature \(d\). Then in
 the case of 2D synthetic datasets we find that taking the absolute value
-of the outputs, \(|\mathbb{g}_d|\), and then dividing them by a
-\(\max(\mathbb{g}) - \min(\mathbb{g})\) term allows us to make the most
+of the outputs, \(|\mathbf{g}_d|\), and then dividing them by a
+\(\max(\mathbf{g}) - \min(\mathbf{g})\) term allows us to make the most
 meaningful comparison. In the case of real-world datasets we choose to
@@
 
-\caption{\label{tbl-ig}Integrated gradients.}
+\caption{\label{tbl-ig}Average sensitivity \(\pm\) bootstrapped standard
+errors of protected features as per standardized integrated gradients.}
 
@@
 
-\caption{\label{fig-mnist-ig}Class-conditional integrated gradients.}
+\caption{\label{fig-mnist-ig}Class-conditional standardized integrated
+gradients for MNIST. Top: BL. Bottom: CT. Mutability constraints were
+imposed on the five top and bottom rows of pixels.}
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, leftrule=.75mm, coltitle=black, colback=white, toprule=.15mm, opacityback=0, colbacktitle=quarto-callout-note-color!10!white, breakable, toptitle=1mm, bottomrule=.15mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, rightrule=.15mm, arc=.35mm, titlerule=0mm, left=2mm, opacitybacktitle=0.6]
+\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
 

### 2025-07-24 paper/sections/other/main_results.qmd
done

@@
 
-Let $\mathbb{g}_d$ denote the estimated IG for feature $d$.
-Then in the case of 2D synthetic datasets we find that taking the absolute value of the outputs, $|\mathbb{g}_d|$, and then dividing them by a $\max(\mathbb{g}) - \min(\mathbb{g})$ term allows us to make the most meaningful comparison.
-In the case of real-world datasets we choose to normalize the values to a 0-1 range instead.
-
+Let $\mathbf{g}_d$ denote the estimated IG for feature $d$.
+Then in the case of 2D synthetic datasets we find that taking the absolute value of the outputs, $|\mathbf{g}_d|$, and then dividing them by a $\max(\mathbf{g}) - \min(\mathbf{g})$ term allows us to make the most meaningful comparison.
+In the case of real-world datasets we choose to normalize the values to a $[0,1]$ range instead.
 We compare the (average) sensitivity to the features that were protected for CT models.
-We again use bootstrapping (100 rounds, 2500 samples per round) to establish the significance of our results.
+Once again we use bootstrapping (100 rounds, 2500 samples per round) to establish the significance of our results.
 These are presented in @tbl-ig.
@@
 
-Integrated gradients.
+Average sensitivity $\pm$ bootstrapped standard errors of protected features as per standardized integrated gradients.
 
@@
 
-![Class-conditional integrated gradients.](/paper/figures/mnist_ig.png){#fig-mnist-ig}
+![Class-conditional standardized integrated gradients for MNIST.
+Top: BL.
+Bottom: CT.
+Mutability constraints were imposed on the five top and bottom rows of pixels.](/paper/figures/mnist_ig.png){#fig-mnist-ig}
 

### 2025-07-24 paper/preprint/paper.tex

--- 2025-07-22 paper/preprint/paper.tex
@@
 \newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\usepackage{amsthm}
-\theoremstyle{remark}
-\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
-\newtheorem*{remark}{Remark}
-\newtheorem*{solution}{Solution}
-\newtheorem{refremark}{Remark}[section]
-\newtheorem{refsolution}{Solution}[section]
 \makeatother
@@
 \makeatother
-\newcounter{quartocalloutnteno}
-\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}
 \usepackage{bookmark}
@@
   pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
-  pdfauthor={Patrick Altmeyer; Aleksander Buszydlik; Arie van Deursen; Cynthia C. S. Liem},
   pdfkeywords={Counterfactual Training, Counterfactual
@@
 
-\newcommand{\runninghead}{A Preprint }
+\newcommand{\runninghead}{}
 \renewcommand{\runninghead}{Counterfactual Training }
 \title{Counterfactual Training: Teaching Models Plausible and Actionable
-Explanations}
+Explanations\thanks{A Preprint.}}
 \def\asep{\\\\\\ } % default: all authors on same column
 \def\asep{\And }
-\author{\textbf{Patrick
-Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
-Engineering, Mathematics and Computer Science\\Delft University of
-Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Aleksander
-Buszydlik}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\\asep\textbf{Arie van
-Deursen}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\\asep\textbf{Cynthia C. S.
-Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
-Science\\Delft University of Technology\\\\}
+\author{}
 \date{}
@@
 Representation Learning
+
+\pagebreak
+
+\twocolumn
 
@@
 To assess outcomes with respect to actionability for non-linear models,
-we look at the average costs of valid counterfactuals in terms of their
-distances from factual starting points. While this an imperfect proxy of
-sensitivity, we hypothesize that CT can reduce these costs by teaching
-models to seek plausibility with respect to mutable features, much like
-we observe in Figure~\ref{fig-poc} in panel (d) compared to (c). We
-supplement this analysis with qualitative findings for integrated
-gradients (\citeproc{ref-sundararajan2017ig}{Sundararajan, Taly, and Yan
-2017}). Finally, for predictive performance, we use standard metrics,
-such as robust accuracy estimated on adversarially perturbed data using
-FGSM (\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and
-Szegedy 2015}).
+we look at the costs of (just) valid counterfactuals in terms of their
+distances from factual starting points with \(\tau=0.5\). While this an
+imperfect proxy of sensitivity, we hypothesize that CT can reduce these
+costs by teaching models to seek plausibility with respect to mutable
+features, much like we observe in Figure~\ref{fig-poc} in panel (d)
+compared to (c). We supplement this analysis with estimates for
+integrated gradients (\citeproc{ref-sundararajan2017ig}{Sundararajan,
+Taly, and Yan 2017}). Finally, for predictive performance, we use
+standard metrics, such as robust accuracy estimated on adversarially
+perturbed data using FGSM
+(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
+2015}).
 
@@
 \url{https://doi.org/10.1145/3580305.3599290}.
-
-\bibitem[\citeproctext]{ref-hastie2009elements}
-Hastie, Trevor, Robert Tibshirani, and Jerome Friedman. 2009. \emph{The
-Elements of Statistical Learning}. Springer New York.
-\url{https://doi.org/10.1007/978-0-387-84858-7}.
-
-\bibitem[\citeproctext]{ref-innes2018fashionable}
-Innes, Michael, Elliot Saba, Keno Fischer, Dhairya Gandhi, Marco
-Concetto Rudilosso, Neethu Mariya Joy, Tejan Karmali, Avik Pal, and
-Viral Shah. 2018. {``Fashionable Modelling with Flux.''}
-\url{https://arxiv.org/abs/1811.01457}.
-
-\bibitem[\citeproctext]{ref-innes2018flux}
-Innes, Mike. 2018. {``Flux: {Elegant} Machine Learning with {Julia}.''}
-\emph{Journal of Open Source Software} 3 (25): 602.
-\url{https://doi.org/10.21105/joss.00602}.
 
@@
 
-\newpage{}
-
-\begin{appendices}
-
-\section{Notation}\label{notation}
-
-Below we provide an overview of some notation used frequently throughout
-the paper:
-
-\begin{itemize}
-\tightlist
-\item
-  \(y^+\): The target class and also the index of the target class.
-\item
-  \(y^-\): The non-target class and also the index of non-the target
-  class.
-\item
-  \(\mathbf{x}\): a single training sample.
-\item
-  \(\mathbf{x}^\prime\): a counterfactual.
-\item
-  \(\mathbf{x}^+\): a training sample in the target class
-  (ground-truth).
-\item
-  \(\mathbf{y}^+\): The one-hot encoded output vector for the target
-  class.
-\item
-  \(\theta\): Model parameters (unspecified).
-\item
-  \(\Theta\): Matrix of parameters.
-\item
-  \(\mathbf{M}(\cdot)\): linear predictions (logits) of the classifier.
-\end{itemize}
-
-\subsection{Other Technical Details}\label{other-technical-details}
-
-Maximum mean discrepancy is defined as follows,
-
-\begin{equation}\phantomsection\label{eq-mmd}{
-\begin{aligned}
-\text{MMD}({X}^\prime,\tilde{X}^\prime) &= \frac{1}{m(m-1)}\sum_{i=1}^m\sum_{j\neq i}^m k(x_i,x_j) \\ &+ \frac{1}{n(n-1)}\sum_{i=1}^n\sum_{j\neq i}^n k(\tilde{x}_i,\tilde{x}_j) \\ &- \frac{2}{mn}\sum_{i=1}^m\sum_{j=1}^n k(x_i,\tilde{x}_j)
-\end{aligned}
-}\end{equation}
-
-where \(k(\cdot,\cdot)\) is a kernel function
-(\citeproc{ref-gretton2012kernel}{Gretton et al. 2012}). We make use of
-a Gaussian kernel with a constant length-scale parameter of \(0.5\). In
-our implementation, Equation~\ref{eq-mmd} is by default applied to the
-entire subset of the training data for which \(y=y^+\).
-
-\section{Technical Details of Our
-Approach}\label{technical-details-of-our-approach}
-
-\subsection{Generating Counterfactuals through Gradient
-Descent}\label{sec-app-ce}
-
-In this section, we provide some background on gradient-based
-counterfactual generators (Section~\ref{sec-app-ce-background}) and
-discuss how we define convergence in this context
-(Section~\ref{sec-app-conv}).
-
-\subsubsection{Background}\label{sec-app-ce-background}
-
-Gradient-based counterfactual search was originally proposed by Wachter,
-Mittelstadt, and Russell
-(\citeproc{ref-wachter2017counterfactual}{2017}). It generally solves
-the following unconstrained objective,
-
-\[
-\begin{aligned}
-\min_{\mathbf{z}^\prime \in \mathcal{Z}^L} \left\{  {\text{yloss}(\mathbf{M}_{\theta}(g(\mathbf{z}^\prime)),\mathbf{y}^+)}+ \lambda {\text{cost}(g(\mathbf{z}^\prime)) }  \right\} 
-\end{aligned} 
-\]
-
-where \(g: \mathcal{Z} \mapsto \mathcal{X}\) is an invertible function
-that maps from the \(L\)-dimensional counterfactual state space to the
-feature space and \(\text{cost}(\cdot)\) denotes one or more penalties
-that are used to induce certain properties of the counterfactual
-outcome. As above, \(\mathbf{y}^+\) denotes the target output and
-\(\mathbf{M}_{\theta}(\mathbf{x})\) returns the logit predictions of the
-underlying classifier for \(\mathbf{x}=g(\mathbf{z})\).
-
-For all generators used in this work we use standard logit crossentropy
-loss for \(\text{yloss}(\cdot)\). All generators also penalize the
-distance (\(\ell_1\)-norm) of counterfactuals from their original
-factual state. For \emph{Generic} and \emph{ECCo}, we have
-\(\mathcal{Z}:=\mathcal{X}\) and
-\(g(\mathbf{z})=g(\mathbf{z})^{-1}=\mathbf{z}\), that is counterfactual
-are searched directly in the feature space. Conversely, \emph{REVISE}
-traverses the latent space of a variational autoencoder (VAE) fitted to
-the training data, where \(g(\cdot)\) corresponds to the decoder
-(\citeproc{ref-joshi2019realistic}{Joshi et al. 2019}). In addition to
-the distance penalty, \emph{ECCo} uses an additional penalty component
-that regularizes the energy associated with the counterfactual,
-\(\mathbf{x}^\prime\) (\citeproc{ref-altmeyer2024faithful}{Altmeyer et
-al. 2024}).
-
-\subsubsection{Convergence}\label{sec-app-conv}
-
-An important consideration when generating counterfactual explanations
-using gradient-based methods is how to define convergence. Two common
-choices are to 1) perform gradient descent over a fixed number of
-iterations \(T\), or 2) conclude the search as soon as the predicted
-probability for the target class has reached a pre-determined threshold,
-\(\tau\):
-\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\). We
-prefer the latter for our purposes, because it explicitly defines
-convergence in terms of the black-box model, \(\mathbf{M}(\mathbf{x})\).
-
-Defining convergence in this way allows for a more intuitive
-interpretation of the resulting counterfactual outcomes than with fixed
-\(T\). Specifically, it allows us to think of counterfactuals as
-explaining `high-confidence' predictions by the model for the target
-class \(y^+\). Depending on the context and application, different
-choices of \(\tau\) can be considered as representing `high-confidence'
-predictions.
-
-\subsection{Protecting Mutability Constraints with Linear
-Classifiers}\label{sec-app-constraints}
-
-In Section~\ref{sec-constraints} we explain that to avoid penalizing
-implausibility that arises due to mutability constraints, we impose a
-point mass prior on \(p(\mathbf{x})\) for the corresponding feature. We
-argue in Section~\ref{sec-constraints} that this approach induces models
-to be less sensitive to immutable features and demonstrate this
-empirically in Section~\ref{sec-experiments}. Below we derive the
-analytical results in Prp.\textasciitilde{}\ref{prp-mtblty}.
-
-\begin{proof}
-Let \(d_{\text{mtbl}}\) and \(d_{\text{immtbl}}\) denote some mutable
-and immutable feature, respectively. Suppose that
-\(\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}}\) and
-\(\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}}\), where
-\(\mu_{k,d}\) denotes the conditional sample mean of feature \(d\) in
-class \(k\). In words, we assume that the immutable feature tends to
-take lower values for samples in the non-target class \(y^-\) than in
-the target class \(y^+\). We assume the opposite to hold for the mutable
-feature.
-
-Assuming multivariate Gaussian class densities with common diagonal
-covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\), we
-have for the log likelihood ratio between any two classes
-\(k,m \in \mathcal{K}\) (\citeproc{ref-hastie2009elements}{Hastie,
-Tibshirani, and Friedman 2009}):
-
-\begin{equation}\phantomsection\label{eq-loglike}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})}=\mathbf{x}^\intercal \Sigma^{-1}(\mu_{k}-\mu_{m})  + \text{const}
-}\end{equation}
-
-By independence of \(x_1,...,x_D\), the full log-likelihood ratio
-decomposes into:
-
-\begin{equation}\phantomsection\label{eq-loglike-decomp}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \frac{\mu_{k,d}-\mu_{m,d}}{\sigma_{d}^2} x_{d} + \text{const}
-}\end{equation}
-
-By the properties of our classifier (\emph{multinomial logistic
-regression}), we have:
-
-\begin{equation}\phantomsection\label{eq-multi}{
-\log \frac{p(k|\mathbf{x})}{p(m|\mathbf{x})} = \sum_{d=1}^D \left( \theta_{k,d} - \theta_{m,d} \right)x_d + \text{const}
-}\end{equation}
-
-where \(\theta_{k,d}=\Theta[k,d]\) denotes the coefficient on feature
-\(d\) for class \(k\).
-
-Based on Equation~\ref{eq-loglike-decomp} and Equation~\ref{eq-multi} we
-can identify that
-\((\mu_{k,d}-\mu_{m,d}) \propto (\theta_{k,d} - \theta_{m,d})\) under
-the assumptions we made above. Hence, we have that
-\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\)
-and
-\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)
-
-Let \(\mathbf{x}^\prime\) denote some randomly chosen individual from
-class \(y^-\) and let \(y^+ \sim p(y)\) denote the randomly chosen
-target class. Then the partial derivative of the contrastive divergence
-penalty Equation~\ref{eq-div} with respect to coefficient
-\(\theta_{y^+,d}\) is equal to
-
-\begin{equation}\phantomsection\label{eq-grad}{
-\frac{\partial}{\partial\theta_{y^+,d}} \left(\text{div}(\mathbf{x}^+,\mathbf{x^\prime},\mathbf{y};\theta)\right) = \frac{\partial}{\partial\theta_{y^+,d}} \left( \left(-\mathbf{M}_\theta(\mathbf{x}^+)[y^+]\right) - \left(-\mathbf{M}_\theta(\mathbf{x}^\prime)[y^+]\right) \right) = x_{d}^\prime - x^+_{d}
-}\end{equation}
-
-and equal to zero everywhere else.
-
-Since \((\mu_{y^-,d_{\text{immtbl}}} < \mu_{y^+,d_{\text{immtbl}}})\) we
-are more likely to have
-\((x_{d_{\text{immtbl}}}^\prime - x^+_{d_{\text{immtbl}}}) < 0\) than
-vice versa at initialization. Similarly, we are more likely to have
-\((x_{d_{\text{mtbl}}}^\prime - x^+_{d_{\text{mtbl}}}) > 0\) since
-\((\mu_{y^-,d_{\text{mtbl}}} > \mu_{y^+,d_{\text{mtbl}}})\).
-
-This implies that if we do not protect feature \(d_{\text{immtbl}}\),
-the contrastive divergence penalty will decrease
-\(\theta_{y^-,d_{\text{immtbl}}}\) thereby exacerbating the existing
-effect
-\((\theta_{y^-,d_{\text{immtbl}}} - \theta_{y^+,d_{\text{immtbl}}}) < 0\).
-In words, not protecting the immutable feature would have the
-undesirable effect of making the classifier more sensitive to this
-feature, in that it would be more likely to predict class \(y^-\) as
-opposed to \(y^+\) for lower values of \(d_{\text{immtbl}}\).
-
-By the same rationale, the contrastive divergence penalty can generally
-be expected to increase \(\theta_{y^-,d_{\text{mtbl}}}\) exacerbating
-\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
-In words, this has the effect of making the classifier more sensitive to
-the mutable feature, in that it would be more likely to predict class
-\(y^-\) as opposed to \(y^+\) for higher values of \(d_{\text{mtbl}}\).
-
-Thus, our proposed approach of protecting feature \(d_{\text{immtbl}}\)
-has the net affect of decreasing the classifier's sensitivity to the
-immutable feature relative to the mutable feature (i.e.~no change in
-sensitivity for \(d_{\text{immtbl}}\) relative to increased sensitivity
-for \(d_{\text{mtbl}}\)).
-\end{proof}
-
-\subsection{Domain Constraints}\label{domain-constraints}
-
-We apply domain constraints on counterfactuals during training and
-evaluation. There are at least two good reasons for doing so. Firstly,
-within the context of explainability and algorithmic recourse,
-real-world attributes are often domain constrained: the \emph{age}
-feature, for example, is lower bounded by zero and upper bounded by the
-maximum human lifespan. Secondly, domain constraints help mitigate
-training instabilities commonly associated with energy-based modelling
-(\citeproc{ref-grathwohl2020your}{Grathwohl et al. 2020};
-\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-
-For our image datasets, features are pixel values and hence the domain
-is constrained by the lower and upper bound of values that pixels can
-take depending on how they are scaled (in our case \([-1,1]\)). For all
-other features \(d\) in our synthetic and tabular datasets, we
-automatically infer domain constraints
-\([x_d^{\text{LB}},x_d^{\text{UB}}]\) as follows,
-
-\begin{equation}\phantomsection\label{eq-domain}{
-\begin{aligned}
-x_d^{\text{LB}} &= \arg\min_{x_d} \{\mu_d - n_{\sigma_d}\sigma_d, \arg \min_{x_d} x_d\} \\
-x_d^{\text{UB}} &= \arg\max_{x_d} \{\mu_d + n_{\sigma_d}\sigma_d, \arg \max_{x_d} x_d\} 
-\end{aligned}
-}\end{equation}
-
-where \(\mu_d\) and \(\sigma_d\) denote the sample mean and standard
-deviation of feature \(d\). We set \(n_{\sigma_d}=3\) across the board
-but higher values and hence wider bounds may be appropriate depending on
-the application.
-
-\subsection{Training Hyperparameters}\label{sec-app-training}
-
-Note~\ref{nte-train-default} presents the default hyperparameters used
-during training.
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-train-default}
-
-\begin{itemize}
-\tightlist
-\item
-  Meta Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Generator: \texttt{ecco}
-  \item
-    Model: \texttt{mlp}
-  \end{itemize}
-\item
-  Model:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Activation: \texttt{relu}
-  \item
-    No.~Hidden: \texttt{32}
-  \item
-    No.~Layers: \texttt{1}
-  \end{itemize}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Burnin: \texttt{0.0}
-  \item
-    Class Loss: \texttt{logitcrossentropy}
-  \item
-    Convergence: \texttt{threshold}
-  \item
-    Generator Parameters:
-
-    \begin{itemize}
-    \tightlist
-    \item
-      Decision Threshold: \texttt{0.75}
-    \item
-      \(\lambda_{\text{cst}}\): \texttt{0.001}
-    \item
-      \(\lambda_{\text{egy}}\): \texttt{5.0}
-    \item
-      Learning Rate: \texttt{0.25}
-    \item
-      Maximum Iterations: \texttt{30}
-    \item
-      Optimizer: \texttt{sgd}
-    \item
-      Type: \texttt{ECCo}
-    \end{itemize}
-  \item
-    \(\lambda_{\text{adv}}\): \texttt{0.25}
-  \item
-    \(\lambda_{\text{clf}}\): \texttt{1.0}
-  \item
-    \(\lambda_{\text{div}}\): \texttt{0.5}
-  \item
-    \(\lambda_{\text{reg}}\): \texttt{0.1}
-  \item
-    Learning Rate: \texttt{0.001}
-  \item
-    No.~Counterfactuals: \texttt{1000}
-  \item
-    No.~Epochs: \texttt{100}
-  \item
-    Objective: \texttt{full}
-  \item
-    Optimizer: \texttt{adam}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\subsection{Evaluation Details}\label{sec-app-eval}
-
-For all of our evaluations, we proceed as follows: for each experiment
-setting we generate multiple counterfactuals (``No.~Counterfactuals''),
-randomly choosing the factual and target class each time
-(Note~\ref{nte-eval-default}). We do this across multiple rounds
-(``No.~Runs'') with different random seeds to account for stochasticity
-(Note~\ref{nte-eval-default}). This is in line with standard practice in
-the related literature on CE. Note~\ref{nte-eval-default} presents the
-default hyperparameters used during evaluation. For our final results
-presented in the main paper, we rely on held out test sets to sample
-factuals (and outputs for our performance metrics). For tuning purposes
-we rely on training or validation sets.
-
-\subsubsection{Robust Accuracy}\label{robust-accuracy}
-
-To evaluate robust accuracy (Acc.\(^*\)), we use the Fast Gradient Sign
-Method (FGSM) to perturb test samples
-(\citeproc{ref-goodfellow2014explaining}{Goodfellow, Shlens, and Szegedy
-2015}). For the main results, we have set the perturbation size to
-\(\epsilon=0.03\). We have also tested other perturbation sizes, as well
-as randomly perturbed data. Although not reported here, we have
-consistently found strong outperformance of CT compared to the weak
-baseline.
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-eval-default}
-
-\begin{itemize}
-\tightlist
-\item
-  Counterfactual Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Convergence: \texttt{threshold}
-  \item
-    Decision Threshold: \texttt{0.95}
-  \item
-    Generator Parameters:
-
-    \begin{itemize}
-    \tightlist
-    \item
-      Decision Threshold: \texttt{0.75}
-    \item
-      \(\lambda_{\text{cst}}\): \texttt{0.001}
-    \item
-      \(\lambda_{\text{egy}}\): \texttt{5.0}
-    \item
-      Learning Rate: \texttt{0.25}
-    \item
-      Maximum Iterations: \texttt{30}
-    \item
-      Optimizer: \texttt{sgd}
-    \item
-      Type: \texttt{ECCo}
-    \end{itemize}
-  \item
-    Maximum Iterations: \texttt{50}
-  \item
-    No.~Individuals: \texttt{100}
-  \item
-    No.~Runs: \texttt{5}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\FloatBarrier
-
-\section{Details on Main Experiments}\label{sec-app-main}
-
-\subsection{Final Hyperparameters}\label{final-hyperparameters}
-
-As discussed Section~\ref{sec-experiments}, CT is sensitive to certain
-hyperparameter choices. We study the effect of many hyperparameters
-extensively in Section~\ref{sec-app-grid}. For the main results, we tune
-a small set of key hyperparameters (Section~\ref{sec-app-tune}). The
-final choices for the main results are presented for each data set in
-Table~\ref{tbl-final-params} along with training, test and batch sizes.
-
-\begin{table}
-
-\caption{\label{tbl-final-params}Final hyperparameters used for the main
-results presented in Section~\ref{sec-experiments}. Any hyperparameter
-not shown here is set to its default value
-(Note~\ref{nte-train-default}).}
-
-\centering{
-
-\begin{tabular}{cccccccc}
-  \toprule
-  \textbf{Data} & \textbf{No.
-Train} & \textbf{No.
-Test} & \textbf{Batchsize} & \textbf{Domain} & \textbf{Decision Threshold} & \textbf{No.
-Counterfactuals} & \textbf{$\lambda_{\text{reg}}$} \\\midrule
-  LS & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.01 \\
-  Circ & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.5 \\
-  Moon & 3600 & 600 & 30 & none & 0.9 & 1000 & 0.25 \\
-  OL & 3600 & 600 & 30 & none & 0.5 & 1000 & 0.25 \\\midrule
-  Adult & 26049 & 5010 & 1000 & none & 0.75 & 5000 & 0.25 \\
-  CH & 16504 & 3101 & 1000 & none & 0.5 & 5000 & 0.25 \\
-  Cred & 10617 & 1923 & 1000 & none & 0.5 & 5000 & 0.25 \\
-  GMSC & 13371 & 2474 & 1000 & none & 0.5 & 5000 & 0.5 \\
-  MNIST & 11000 & 2000 & 1000 & (-1.0, 1.0) & 0.5 & 5000 & 0.01 \\
-\end{tabular}
-
-}
-
-\end{table}%
-
-\subsection{Final Results}\label{sec-app-final-results}
-
-Plus/minus two standard deviations of bootstrap estimates.
-
-\subsubsection{Robust Performance Plots}\label{robust-performance-plots}
-
-\subsubsection{Confidence Intervals}\label{confidence-intervals}
-
-\begin{table}
-
-\caption{\label{tbl-ci}Mean outcomes for \textbf{CT} and \textbf{BL}
-along with bootstrapped confidence intervals (99\%) for difference in
-mean outcomes grouped by dataset and evaluation metric. Column
-\textbf{LB} and \textbf{UB} show the lower and upper bound of the
-intervals, respectively, and computed using the percentile method. The
-underlying counterfactual evaluations are the same as the ones used to
-produce Table~\ref{tbl-main}.}
-
-\centering{
-
-\begin{tabular}{lccccc}
-  \toprule
-  \textbf{Variable} & \textbf{Data} & \textbf{CT} & \textbf{BL} & \textbf{LB} & \textbf{UB} \\\midrule
-  Cost & Adult & 2.26 & 2.2 & -0.22 & 0.28 \\
-  Cost & CH & 1.46 & 2.46 & -1.1 & -0.89 \\
-  Cost & Circ & 0.67 & 1.23 & -0.58 & -0.53 \\
-  Cost & Cred & 2.68 & 2.29 & 0.16 & 0.63 \\
-  Cost & GMSC & 1.14 & 3.05 & -2.45 & -1.77 \\
-  Cost & LS & 3.82 & 4.44 & -0.7 & -0.56 \\
-  Cost & MNIST & 77.04 & 68.67 & -3.47 & 18.34 \\
-  Cost & Moon & 1.55 & 1.6 & -0.08 & -0.01 \\
-  Cost & OL & 1.62 & 2.63 & -1.15 & -0.81 \\
-  $ \text{IP}^* $ & Adult & 0.07 & 0.11 & -0.06 & -0.01 \\
-  $ \text{IP}^* $ & CH & 0.02 & 0.06 & -0.06 & -0.04 \\
-  $ \text{IP}^* $ & Circ & 0.0 & 0.0 & -0.01 & -0.0 \\
-  $ \text{IP}^* $ & Cred & 0.03 & 0.06 & -0.05 & -0.01 \\
-  $ \text{IP}^* $ & GMSC & 0.02 & 0.07 & -0.06 & -0.04 \\
-  $ \text{IP}^* $ & LS & 0.1 & 0.23 & -0.14 & -0.12 \\
-  $ \text{IP}^* $ & MNIST & 0.04 & 0.04 & -0.1 & 0.09 \\
-  $ \text{IP}^* $ & Moon & 0.02 & 0.02 & -0.01 & -0.0 \\
-  $ \text{IP}^* $ & OL & 0.12 & 0.09 & -0.01 & 0.05 \\
-  $ \text{IP} $ & Adult & 15.03 & 15.15 & -0.68 & 0.26 \\
-  $ \text{IP} $ & CH & 6.61 & 7.52 & -1.17 & -0.63 \\
-  $ \text{IP} $ & Circ & 1.03 & 2.36 & -1.37 & -1.29 \\
-  $ \text{IP} $ & Cred & 19.31 & 22.03 & -3.69 & -1.74 \\
-  $ \text{IP} $ & GMSC & 6.19 & 8.09 & -2.4 & -1.49 \\
-  $ \text{IP} $ & LS & 2.41 & 3.4 & -1.04 & -0.94 \\
-  $ \text{IP} $ & MNIST & 258.83 & 278.54 & -30.49 & -7.64 \\
-  $ \text{IP} $ & Moon & 1.36 & 1.71 & -0.38 & -0.32 \\
-  $ \text{IP} $ & OL & 4.49 & 4.44 & -0.03 & 0.13 \\\bottomrule
-\end{tabular}
-
-}
-
-\end{table}%
-
-\subsubsection{Qualitative Findings for Image
-Data}\label{qualitative-findings-for-image-data}
-
-Figure~\ref{fig-mnist} shows much more plausible (faithful)
-counterfactuals for a model with CT than the model with conventional
-training (Figure~\ref{fig-mnist-vanilla}).
-
-\begin{figure}
-
-\begin{minipage}{0.46\linewidth}
-
-\begin{figure}[H]
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp.png}}
-
-}
-
-\caption{\label{fig-mnist}Counterfactual images for \emph{MLP} with
-counterfactual training. Factual images are shown on the diagonal, with
-the corresponding counterfactual for each target class (columns) in that
-same row. The underlying generator, \emph{ECCo}, aims to generate
-counterfactuals that are faithful to the model
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).}
-
-\end{figure}%
-
-\end{minipage}%
-%
-\begin{minipage}{0.09\linewidth}
-~\end{minipage}%
-%
-\begin{minipage}{0.46\linewidth}
-
-\begin{figure}[H]
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_mlp_vanilla.png}}
-
-}
-
-\caption{\label{fig-mnist-vanilla}The same setup, factuals, model
-architecture and generator as in Figure~\ref{fig-mnist}, but the model
-was trained conventionally.}
-
-\end{figure}%
-
-\end{minipage}%
-
-\end{figure}%
-
-\subsubsection{Costs}\label{costs}
-
-\begin{table}
-
-\caption{\label{tbl-costs}Costs}
-
-\centering{
-
-\begin{tabular}{lc}
-  \toprule
-  \textbf{Data} & \textbf{Cost $(-\%)$} \\\midrule
-  LS & -26.82\pm0.86 $^{*}$ \\
-  Circ & 40.97\pm0.82 $^{*}$ \\
-  Moon & 33.83\pm0.98 $^{*}$ \\
-  OL & 10.35\pm1.28 $^{*}$ \\\midrule
-  Adult & 1.16\pm3.53 $^{}$ \\
-  CH & -34.89\pm2.31 $^{*}$ \\
-  Cred & 28.24\pm1.08 $^{*}$ \\
-  GMSC & 3.54\pm5.78 $^{}$ \\
-  MNIST & -31.67\pm7.72 $^{*}$ \\\midrule
-  Avg. & 2.75 \\\bottomrule
-\end{tabular}
-
-}
-
-\end{table}%
-
-\subsubsection{Validity}\label{validity}
-
-\begin{table}
-
-\caption{\label{tbl-val}Validity}
-
-\centering{
-
-\begin{tabular}{lcc}
-  \toprule
-  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
-  LS & 1.0 & 1.0 \\
-  Circ & 0.97 & 0.52 \\
-  Moon & 1.0 & 1.0 \\
-  OL & 0.87 & 0.98 \\\midrule
-  Adult & 0.61 & 0.99 \\
-  CH & 0.96 & 1.0 \\
-  Cred & 0.7 & 1.0 \\
-  GMSC & 0.63 & 1.0 \\
-  MNIST & 1.0 & 1.0 \\
-\end{tabular}
-
-}
-
-\end{table}%
-
-\begin{table}
-
-\caption{\label{tbl-val-mtbl}Validity}
-
-\centering{
-
-\begin{tabular}{lcc}
-  \toprule
-  \textbf{Data} & \textbf{CT} & \textbf{BL} \\\midrule
-  LS & 1.0 & 1.0 \\
-  Circ & 0.67 & 0.49 \\
-  Moon & 0.99 & 0.98 \\
-  OL & 0.37 & 0.57 \\\midrule
-  Adult & 0.56 & 0.99 \\
-  CH & 0.96 & 1.0 \\
-  Cred & 0.67 & 1.0 \\
-  GMSC & 0.38 & 1.0 \\
-  MNIST & 1.0 & 1.0 \\
-\end{tabular}
-
-}
-
-\end{table}%
-
-\FloatBarrier
-
-\section{Grid Searches}\label{sec-app-grid}
-
-To assess the hyperparameter sensitivity of our proposed training regime
-we ran multiple large grid searches for all of our synthetic datasets.
-We have grouped these grid searches into multiple categories:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  \textbf{Generator Parameters} (Section~\ref{sec-app-grid-gen}):
-  Investigates the effect of changing hyperparameters that affect the
-  counterfactual outcomes during the training phase.
-\item
-  \textbf{Penalty Strengths} (Section~\ref{sec-app-grid-pen}):
-  Investigates the effect of changing the penalty strengths in out
-  proposed objective (Equation~\ref{eq-obj}).
-\item
-  \textbf{Other Parameters} (Section~\ref{sec-app-grid-train}):
-  Investigates the effect of changing other training parameters,
-  including the total number of generated counterfactuals in each epoch.
-\end{enumerate}
-
-We begin by summarizing the high-level findings in
-Section~\ref{sec-app-grid-hl}. For each of the categories,
-Section~\ref{sec-app-grid-gen} to Section~\ref{sec-app-grid-train} then
-present all details including the exact parameter grids, average
-predictive performance outcomes and key evaluation metrics for the
-generated counterfactuals.
-
-\subsection{Evaluation Details}\label{evaluation-details}
-
-To measure predictive performance, we compute the accuracy and F1-score
-for all models on test data (Table~\ref{tbl-acc-gen},
-Table~\ref{tbl-acc-pen}, Table~\ref{tbl-acc-train}). With respect to
-explanatory performance, we report here our findings for the
-(im)plausibility and cost of counterfactuals at test time. Since the
-computation of our proposed divergence metric
-(Equation~\ref{eq-impl-div}) is memory-intensive, we rely on the
-distance-based metric for the grid searches. For the counterfactual
-evaluation, we draw factual samples from the training data for the grid
-searches to avoid data leakage with respect to our final results
-reported in the body of the paper. Specifically, we want to avoid
-choosing our default hyperparameters based on results on the test data.
-Since we are optimizing for explainability, not predictive performance,
-we still present test accuracy and F1-scores.
-
-\subsubsection{Predictive Performance}\label{predictive-performance}
-
-We find that CT is associated with little to no decrease in average
-predictive performance for our synthetic datasets: test accuracy and
-F1-scores decrease by at most \textasciitilde1 percentage point, but
-generally much less (Table~\ref{tbl-acc-gen}, Table~\ref{tbl-acc-pen},
-Table~\ref{tbl-acc-train}). Variation across hyperparameters is
-negligible as indicated by small standard deviations for these metrics
-across the board.
-
-\subsubsection{Counterfactual Outcomes}\label{sec-app-grid-hl}
-
-Overall, we find that counterfactual training achieves it key objectives
-consistently across all hyperparameter settings and also broadly across
-datasets: plausibility is improved by up to 60 percent (\%) for the
-\emph{Circles} data (e.g.
-Figure~\ref{fig-grid-gen_params-plaus-circles}), 25-30\% for the
-\emph{Moons} data (e.g. Figure~\ref{fig-grid-gen_params-plaus-moons})
-and 10-20\% for the \emph{Linearly Separable} data (e.g.
-Figure~\ref{fig-grid-gen_params-plaus-lin_sep}). At the same time, the
-average costs of faithful counterfactuals are reduced in many cases by
-around 20-25\% for \emph{Circles} (e.g.
-Figure~\ref{fig-grid-gen_params-cost-circles}) and up to 50\% for
-\emph{Moons} (e.g. Figure~\ref{fig-grid-gen_params-cost-moons}). For the
-\emph{Linearly Separable} data, costs are generally increased although
-typically by less than 10\% (e.g.
-Figure~\ref{fig-grid-gen_params-cost-lin_sep}), which reflects a common
-tradeoff between costs and plausibility
-(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
-
-We do observe strong sensitivity to certain hyperparameters, with clear
-an manageable patterns. Concerning generator parameters, we firstly find
-that using \emph{REVISE} to generate counterfactuals during training
-typically yields the worst outcomes out of all generators, often leading
-to a substantial decrease in plausibility. This finding can be
-attributed to the fact that \emph{REVISE} effectively assigns the task
-of learning plausible explanations from the model itself to a surrogate
-VAE. In other words, counterfactuals generated by \emph{REVISE} are less
-faithful to the model that \emph{ECCo} and \emph{Generic}, and hence we
-would expect them to be a less effective and, in fact, potentially
-detrimental role in our training regime. Secondly, we observe that
-allowing for a higher number of maximum steps \(T\) for the
-counterfactual search generally yields better outcomes. This is
-intuitive, because it allows more counterfactuals to reach maturity in
-any given iteration. Looking in particular at the results for
-\emph{Linearly Separable}, it seems that higher values for \(T\) in
-combination with higher decision thresholds (\(\tau\)) yields the best
-results when using \emph{ECCo}. But depending on the degree of class
-separability of the underlying data, a high decision-threshold can also
-affect results adversely, as evident from the results for the
-\emph{Overlapping} data (Figure~\ref{fig-grid-gen_params-plaus-over}):
-here we find that CT generally fails to achieve its objective because
-only a tiny proportion of counterfactuals ever reaches maturity.
-
-Regarding penalty strengths, we find that the strength of the energy
-regularization, \(\lambda_{\text{reg}}\) is a key hyperparameter, while
-sensitivity with respect to \(\lambda_{\text{div}}\) and
-\(\lambda_{\text{adv}}\) is much less evident. In particular, we observe
-that not regularizing energy enough or at all typically leads to poor
-performance in terms of decreased plausibility and increased costs, in
-particular for \emph{Circles} (Figure~\ref{fig-grid-pen-plaus-circles}),
-\emph{Linearly Separable} (Figure~\ref{fig-grid-pen-plaus-lin_sep}) and
-\emph{Overlapping} (Figure~\ref{fig-grid-pen-plaus-over}). High values
-of \(\lambda_{\text{reg}}\) can increase the variability in outcomes, in
-particular when combined with high values for \(\lambda_{\text{div}}\)
-and \(\lambda_{\text{adv}}\), but this effect is less pronounced.
-
-Finally, concerning other hyperparameters we observe that the
-effectiveness and stability of CT is positively associated with the
-number of counterfactuals generated during each training epoch, in
-particular for \emph{Circles}
-(Figure~\ref{fig-grid-train-plaus-circles}) and \emph{Moons}
-(Figure~\ref{fig-grid-train-plaus-moons}). We further find that a higher
-number of training epochs is beneficial as expected, where we tested
-training models for 50 and 100 epochs. Interestingly, we find that it is
-not necessary to employ CT during the entire training phase to achieve
-the desired improvements in explainability: specifically, we have tested
-training models conventionally during the first half of training before
-switching to CT after this initial burn-in period.
-
-\subsection{Generator Parameters}\label{sec-app-grid-gen}
-
-The hyperparameter grid with varying generator parameters during
-training is shown in Note~\ref{nte-gen-params-final-run-train}. The
-corresponding evaluation grid used for these experiments is shown in
-Note~\ref{nte-gen-params-final-run-eval}.
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-gen-params-final-run-train}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Decision Threshold: \texttt{0.75,\ 0.9,\ 0.95}
-  \item
-    \(\lambda_{\text{egy}}\): \texttt{0.1,\ 0.5,\ 5.0,\ 10.0,\ 20.0}
-  \item
-    Maximum Iterations: \texttt{5,\ 25,\ 50}
-  \end{itemize}
-\item
-  Generator: \texttt{ecco,\ generic,\ revise}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-gen-params-final-run-eval}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{egy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\subsubsection{Predictive Performance}\label{predictive-performance-1}
-
-Predictive performance measures for this grid search are shown in
-Table~\ref{tbl-acc-gen}.
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-acc-gen}Predictive performance measures by dataset
-and objective averaged across training-phase parameters
-(Note~\ref{nte-gen-params-final-run-train}) and evaluation-phase
-parameters (Note~\ref{nte-gen-params-final-run-eval}).}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Se} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Se} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  Circ & Accuracy & Full & 1.0 & 0.0 \\
-  Circ & Accuracy & Vanilla & 1.0 & 0.0 \\
-  Circ & F1-score & Full & 1.0 & 0.0 \\
-  Circ & F1-score & Vanilla & 1.0 & 0.0 \\
-  LS & Accuracy & Full & 1.0 & 0.0 \\
-  LS & Accuracy & Vanilla & 1.0 & 0.0 \\
-  LS & F1-score & Full & 1.0 & 0.0 \\
-  LS & F1-score & Vanilla & 1.0 & 0.0 \\
-  Moon & Accuracy & Full & 1.0 & 0.0 \\
-  Moon & Accuracy & Vanilla & 1.0 & 0.0 \\
-  Moon & F1-score & Full & 1.0 & 0.0 \\
-  Moon & F1-score & Vanilla & 1.0 & 0.0 \\
-  OL & Accuracy & Full & 0.91 & 0.0 \\
-  OL & Accuracy & Vanilla & 0.92 & 0.0 \\
-  OL & F1-score & Full & 0.91 & 0.0 \\
-  OL & F1-score & Vanilla & 0.92 & 0.0 \\\bottomrule
-
-\end{longtable}
-
-\subsubsection{Plausibility}\label{plausibility}
-
-The results with respect to the plausibility measure are shown in
-Figure~\ref{fig-grid-gen_params-plaus-circles} to
-Figure~\ref{fig-grid-gen_params-plaus-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-gen_params-plaus-circles}Average outcomes for
-the plausibility measure across hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-gen_params-plaus-lin_sep}Average outcomes for
-the plausibility measure across hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-gen_params-plaus-moons}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-gen_params-plaus-over}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Overlapping.}
-
-\end{figure}%
-
-\subsubsection{Cost}\label{cost}
-
-The results with respect to the cost measure are shown in
-Figure~\ref{fig-grid-gen_params-cost-circles} to
-Figure~\ref{fig-grid-gen_params-cost-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/circles/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}
-
-}
-
-\caption{\label{fig-grid-gen_params-cost-circles}Average outcomes for
-the cost measure across hyperparameters. This shows the \% change from
-the baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/lin_sep/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}
-
-}
-
-\caption{\label{fig-grid-gen_params-cost-lin_sep}Average outcomes for
-the cost measure across hyperparameters. This shows the \% change from
-the baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Linearly
-Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/moons/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}
-
-}
-
-\caption{\label{fig-grid-gen_params-cost-moons}Average outcomes for the
-cost measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/gen_params/mlp/over/evaluation/results/ce/decision_threshold_exper---lambda_energy_exper---maxiter_exper---maxiter---decision_threshold_exper/distance.png}
-
-}
-
-\caption{\label{fig-grid-gen_params-cost-over}Average outcomes for the
-cost measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data:
-Overlapping.}
-
-\end{figure}%
-
-\subsection{Penalty Strengths}\label{sec-app-grid-pen}
-
-The hyperparameter grid with varying penalty strengths during training
-is shown in Note~\ref{nte-pen-final-run-train}. The corresponding
-evaluation grid used for these experiments is shown in
-Note~\ref{nte-pen-final-run-eval}.
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-pen-final-run-train}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator: \texttt{ecco,\ generic,\ revise}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{adv}}\): \texttt{0.1,\ 0.25,\ 1.0}
-  \item
-    \(\lambda_{\text{div}}\): \texttt{0.01,\ 0.1,\ 1.0}
-  \item
-    \(\lambda_{\text{reg}}\): \texttt{0.0,\ 0.01,\ 0.1,\ 0.25,\ 0.5}
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-pen-final-run-eval}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{egy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\subsubsection{Predictive Performance}\label{predictive-performance-2}
-
-Predictive performance measures for this grid search are shown in
-Table~\ref{tbl-acc-pen}.
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-acc-pen}Predictive performance measures by dataset
-and objective averaged across training-phase parameters
-(Note~\ref{nte-pen-final-run-train}) and evaluation-phase parameters
-(Note~\ref{nte-pen-final-run-eval}).}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Se} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Se} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  Circ & Accuracy & Full & 0.99 & 0.01 \\
-  Circ & Accuracy & Vanilla & 1.0 & 0.0 \\
-  Circ & F1-score & Full & 0.99 & 0.01 \\
-  Circ & F1-score & Vanilla & 1.0 & 0.0 \\
-  LS & Accuracy & Full & 1.0 & 0.01 \\
-  LS & Accuracy & Vanilla & 1.0 & 0.0 \\
-  LS & F1-score & Full & 1.0 & 0.01 \\
-  LS & F1-score & Vanilla & 1.0 & 0.0 \\
-  Moon & Accuracy & Full & 0.99 & 0.04 \\
-  Moon & Accuracy & Vanilla & 1.0 & 0.01 \\
-  Moon & F1-score & Full & 0.99 & 0.04 \\
-  Moon & F1-score & Vanilla & 1.0 & 0.01 \\
-  OL & Accuracy & Full & 0.91 & 0.02 \\
-  OL & Accuracy & Vanilla & 0.92 & 0.0 \\
-  OL & F1-score & Full & 0.91 & 0.02 \\
-  OL & F1-score & Vanilla & 0.92 & 0.0 \\\bottomrule
-
-\end{longtable}
-
-\subsubsection{Plausibility}\label{plausibility-1}
-
-The results with respect to the plausibility measure are shown in
-Figure~\ref{fig-grid-pen-plaus-circles} to
-Figure~\ref{fig-grid-pen-plaus-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-pen-plaus-circles}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-pen-plaus-lin_sep}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-pen-plaus-moons}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-pen-plaus-over}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Overlapping.}
-
-\end{figure}%
-
-\subsubsection{Cost}\label{cost-1}
-
-The results with respect to the cost measure are shown in
-Figure~\ref{fig-grid-pen-cost-circles} to
-Figure~\ref{fig-grid-pen-cost-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/circles/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}
-
-}
-
-\caption{\label{fig-grid-pen-cost-circles}Average outcomes for the cost
-measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/lin_sep/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}
-
-}
-
-\caption{\label{fig-grid-pen-cost-lin_sep}Average outcomes for the cost
-measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Linearly
-Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/moons/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}
-
-}
-
-\caption{\label{fig-grid-pen-cost-moons}Average outcomes for the cost
-measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/penalties/mlp/over/evaluation/results/ce/lambda_adversarial---lambda_energy_reg---lambda_energy_diff---lambda_adversarial---lambda_adversarial/distance.png}
-
-}
-
-\caption{\label{fig-grid-pen-cost-over}Average outcomes for the cost
-measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data:
-Overlapping.}
-
-\end{figure}%
-
-\subsection{Other Parameters}\label{sec-app-grid-train}
-
-The hyperparameter grid with other varying training parameters is shown
-in Note~\ref{nte-train-final-run-train}. The corresponding evaluation
-grid used for these experiments is shown in
-Note~\ref{nte-train-final-run-eval}.
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-train-final-run-train}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator: \texttt{ecco,\ generic,\ revise}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Burnin: \texttt{0.0,\ 0.5}
-  \item
-    No.~Counterfactuals: \texttt{100,\ 1000}
-  \item
-    No.~Epochs: \texttt{50,\ 100}
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-train-final-run-eval}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{egy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\subsubsection{Predictive Performance}\label{predictive-performance-3}
-
-Predictive performance measures for this grid search are shown in
-Table~\ref{tbl-acc-train}.
-
-\begin{longtable}{ccccc}
-
-\caption{\label{tbl-acc-train}Predictive performance measures by dataset
-and objective averaged across training-phase parameters
-(Note~\ref{nte-train-final-run-train}) and evaluation-phase parameters
-(Note~\ref{nte-train-final-run-eval}).}
-
-\tabularnewline
-
-  \toprule
-  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Se} \\\midrule
-  \endfirsthead
-  \toprule
-  \textbf{Dataset} & \textbf{Variable} & \textbf{Objective} & \textbf{Mean} & \textbf{Se} \\\midrule
-  \endhead
-  \bottomrule
-  \multicolumn{5}{r}{Continuing table below.}\\
-  \bottomrule
-  \endfoot
-  \endlastfoot
-  Circ & Accuracy & Full & 0.99 & 0.0 \\
-  Circ & Accuracy & Vanilla & 1.0 & 0.0 \\
-  Circ & F1-score & Full & 0.99 & 0.0 \\
-  Circ & F1-score & Vanilla & 1.0 & 0.0 \\
-  LS & Accuracy & Full & 1.0 & 0.0 \\
-  LS & Accuracy & Vanilla & 1.0 & 0.0 \\
-  LS & F1-score & Full & 1.0 & 0.0 \\
-  LS & F1-score & Vanilla & 1.0 & 0.0 \\
-  Moon & Accuracy & Full & 1.0 & 0.01 \\
-  Moon & Accuracy & Vanilla & 0.99 & 0.02 \\
-  Moon & F1-score & Full & 1.0 & 0.01 \\
-  Moon & F1-score & Vanilla & 0.99 & 0.02 \\
-  OL & Accuracy & Full & 0.91 & 0.01 \\
-  OL & Accuracy & Vanilla & 0.92 & 0.0 \\
-  OL & F1-score & Full & 0.91 & 0.01 \\
-  OL & F1-score & Vanilla & 0.92 & 0.0 \\\bottomrule
-
-\end{longtable}
-
-\subsubsection{Plausibility}\label{plausibility-2}
-
-The results with respect to the plausibility measure are shown in
-Figure~\ref{fig-grid-train-plaus-circles} to
-Figure~\ref{fig-grid-train-plaus-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-train-plaus-circles}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-train-plaus-lin_sep}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-train-plaus-moons}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-grid-train-plaus-over}Average outcomes for the
-plausibility measure across hyperparameters. This shows the \% change
-from the baseline model for the distance-based implausibility metric
-(Equation~\ref{eq-impl-dist}). Boxplots indicate the variation across
-evaluation runs and test settings (varying parameters for \emph{ECCo}).
-Data: Overlapping.}
-
-\end{figure}%
-
-\subsubsection{Cost}\label{cost-2}
-
-The results with respect to the cost measure are shown in
-Figure~\ref{fig-grid-train-cost-circles} to
-Figure~\ref{fig-grid-train-cost-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/circles/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}
-
-}
-
-\caption{\label{fig-grid-train-cost-circles}Average outcomes for the
-cost measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/lin_sep/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}
-
-}
-
-\caption{\label{fig-grid-train-cost-lin_sep}Average outcomes for the
-cost measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Linearly
-Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/moons/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}
-
-}
-
-\caption{\label{fig-grid-train-cost-moons}Average outcomes for the cost
-measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/training_params/mlp/over/evaluation/results/ce/burnin---nce---nepochs---burnin---burnin/distance.png}
-
-}
-
-\caption{\label{fig-grid-train-cost-over}Average outcomes for the cost
-measure across hyperparameters. This shows the \% change from the
-baseline model for the distance-based cost metric
-(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
-Russell 2017}). Boxplots indicate the variation across evaluation runs
-and test settings (varying parameters for \emph{ECCo}). Data:
-Overlapping.}
-
-\end{figure}%
-
-\FloatBarrier
-
-\section{Tuning Key Parameters}\label{sec-app-tune}
-
-Based on the findings from our initial large grid searches
-(Section~\ref{sec-app-grid}), we tune selected hyperparameters for all
-datasets: namely, the decision threshold \(\tau\) and the strength of
-the energy regularization \(\lambda_{\text{reg}}\). The final
-hyperparameter choices for each dataset are presented in
-Table~\ref{tbl-final-params} in Section~\ref{sec-app-main}. Detailed
-results for each data set are shown in Figure~\ref{fig-tune-plaus-adult}
-to Figure~\ref{fig-tune-mat-over}. From Table~\ref{tbl-final-params}, we
-notice that the same decision threshold of \(\tau=0.5\) is optimal for
-all but on dataset. We attribute this to the fact that a low decision
-threshold results in a higher share of mature counterfactuals and hence
-more opportunities for the model to learn from examples
-(Figure~\ref{fig-tune-mat-adult} to Figure~\ref{fig-tune-mat-over}).
-This has played a role in particular for our real-world tabular datasets
-and MNIST, which suffered from low levels of maturity for higher
-decision thresholds. In cases where maturity is not an issue, as for
-\emph{Moons}, higher decision thresholds lead to better outcomes, which
-may have to do with the fact that the resulting counterfactuals are more
-faithful to the model. Concerning the regularization strength, we find
-somewhat high variation across datasets. Most notably, we find that
-relatively low levels of regularization are optimal for MNIST. We
-hypothesize that this finding may be attributed to the uniform scaling
-of all input features (digits).
-
-Finally, to increase the proportion of mature counterfactuals for some
-datasets, we have also investigated the effect on the learning rate
-\(\eta\) for the counterfactual search and even smaller regularization
-strengths for a fixed decision threshold of 0.5
-(Figure~\ref{fig-tune_lr-plaus-adult} to
-Figure~\ref{fig-tune_lr-plaus-over}). For the given low decision
-threshold, we find that the learning rate has no discernable impact on
-the proportion of mature counterfactuals
-(Figure~\ref{fig-tune_lr-mat-adult} to
-Figure~\ref{fig-tune_lr-mat-over}). We do notice, however, that the
-results for MNIST are much improved when using a low value
-\(\lambda_{\text{reg}}\), the strength for the engery regularization:
-plausibility is increased by up to \textasciitilde10\%
-(Figure~\ref{fig-tune_lr-plaus-mnist}) and the proportion of mature
-counterfactuals reaches 100\%.
-
-One consideration worth exploring is to combine high decision thresholds
-with high learning rates, which we have not investigated here.
-
-\subsection{Key Parameters}\label{sec-app-tune-key}
-
-The hyperparameter grid for tuning key parameters is shown in
-Note~\ref{nte-tune-train}. The corresponding evaluation grid used for
-these experiments is shown in Note~\ref{nte-tune-eval}.
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-tune-train}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Decision Threshold: \texttt{0.5,\ 0.75,\ 0.9}
-  \end{itemize}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{reg}}\): \texttt{0.1,\ 0.25,\ 0.5}
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-tune-eval}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{egy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\subsubsection{Plausibility}\label{plausibility-3}
-
-The results with respect to the plausibility measure are shown in
-Figure~\ref{fig-tune-plaus-adult} to Figure~\ref{fig-tune-plaus-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune-plaus-adult}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Adult.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune-plaus-cali}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: California Housing.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune-plaus-circles}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune-plaus-credit}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Credit.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune-plaus-gmsc}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: GMSC.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune-plaus-lin_sep}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune-plaus-mnist}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: MNIST.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune-plaus-moons}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/ce/lambda_energy_reg---decision_threshold_exper---lambda_energy_eval---lambda_energy_reg---decision_threshold_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune-plaus-over}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Overlapping.}
-
-\end{figure}%
-
-\subsubsection{Proportion of Mature CE}\label{proportion-of-mature-ce}
-
-The results with respect to the proportion of mature counterfactuals in
-each epoch are shown in Figure~\ref{fig-tune-mat-adult} to
-Figure~\ref{fig-tune-mat-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/adult/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune-mat-adult}Proportion of mature counterfactuals
-in each epoch. Data: Adult.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/cali/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune-mat-cali}Proportion of mature counterfactuals
-in each epoch. Data: California Housing.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/circles/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune-mat-circles}Proportion of mature
-counterfactuals in each epoch. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/credit/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune-mat-credit}Proportion of mature counterfactuals
-in each epoch. Data: Credit.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/gmsc/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune-mat-gmsc}Proportion of mature counterfactuals
-in each epoch. Data: GMSC.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/lin_sep/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune-mat-lin_sep}Proportion of mature
-counterfactuals in each epoch. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/mnist/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune-mat-mnist}Proportion of mature counterfactuals
-in each epoch. Data: MNIST.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/moons/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune-mat-moons}Proportion of mature counterfactuals
-in each epoch. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune/mlp/over/evaluation/results/logs/objective---decision_threshold---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune-mat-over}Proportion of mature counterfactuals
-in each epoch. Data: Overlapping.}
-
-\end{figure}%
-
-\subsection{Learning Rate}\label{sec-app-tune-lr}
-
-The hyperparameter grid for tuning the learning rate is shown in
-Note~\ref{nte-tune_lr-train}. The corresponding evaluation grid used for
-these experiments is shown in Note~\ref{nte-tune_lr-eval}.
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-tune_lr-train}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    Learning Rate: \texttt{0.1,\ 0.5,\ 1.0}
-  \end{itemize}
-\item
-  Model: \texttt{mlp}
-\item
-  Training Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{reg}}\): \texttt{0.01,\ 0.1,\ 0.5}
-  \item
-    Objective: \texttt{full,\ vanilla}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\begin{tcolorbox}[enhanced jigsaw, opacityback=0, bottomtitle=1mm, arc=.35mm, colback=white, coltitle=black, bottomrule=.15mm, breakable, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, toprule=.15mm, colframe=quarto-callout-note-color-frame, toptitle=1mm, opacitybacktitle=0.6, left=2mm, titlerule=0mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white]
-
-\quartocalloutnte{nte-tune_lr-eval}
-
-\begin{itemize}
-\tightlist
-\item
-  Generator Parameters:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    \(\lambda_{\text{egy}}\): \texttt{0.1,\ 0.5,\ 1.0,\ 5.0,\ 10.0}
-  \end{itemize}
-\end{itemize}
-
-\end{tcolorbox}
-
-\subsubsection{Plausibility}\label{plausibility-4}
-
-The results with respect to the plausibility measure are shown in
-Figure~\ref{fig-tune_lr-plaus-adult} to
-Figure~\ref{fig-tune_lr-plaus-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune_lr-plaus-adult}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Adult.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/cali/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune_lr-plaus-cali}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: California Housing.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/circles/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune_lr-plaus-circles}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune_lr-plaus-credit}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Credit.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune_lr-plaus-gmsc}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: GMSC.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/lin_sep/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune_lr-plaus-lin_sep}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune_lr-plaus-mnist}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: MNIST.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/moons/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune_lr-plaus-moons}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/ce/lambda_energy_reg---lr_exper---lambda_energy_eval---lambda_energy_reg---lr_exper/plausibility_distance_from_target.png}
-
-}
-
-\caption{\label{fig-tune_lr-plaus-over}Average outcomes for the
-plausibility measure across key hyperparameters. This shows the \%
-change from the baseline model for the distance-based implausibility
-metric (Equation~\ref{eq-impl-dist}). Boxplots indicate the variation
-across evaluation runs and test settings (varying parameters for
-\emph{ECCo}). Data: Overlapping.}
-
-\end{figure}%
-
-\subsubsection{Proportion of Mature CE}\label{proportion-of-mature-ce-1}
-
-The results with respect to the proportion of mature counterfactuals in
-each epoch are shown in Figure~\ref{fig-tune_lr-mat-adult} to
-Figure~\ref{fig-tune_lr-mat-over}.
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/adult/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune_lr-mat-adult}Proportion of mature
-counterfactuals in each epoch. Data: Adult.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/cali/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune_lr-mat-cali}Proportion of mature
-counterfactuals in each epoch. Data: California Housing.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/circles/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune_lr-mat-circles}Proportion of mature
-counterfactuals in each epoch. Data: Circles.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/credit/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune_lr-mat-credit}Proportion of mature
-counterfactuals in each epoch. Data: Credit.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/gmsc/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune_lr-mat-gmsc}Proportion of mature
-counterfactuals in each epoch. Data: GMSC.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/lin_sep/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune_lr-mat-lin_sep}Proportion of mature
-counterfactuals in each epoch. Data: Linearly Separable.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/mnist/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune_lr-mat-mnist}Proportion of mature
-counterfactuals in each epoch. Data: MNIST.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/moons/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune_lr-mat-moons}Proportion of mature
-counterfactuals in each epoch. Data: Moons.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{../../paper/experiments/output/final_run/tune_lr/mlp/over/evaluation/results/logs/objective---lr---lambda_energy_reg/percent_valid.png}
-
-}
-
-\caption{\label{fig-tune_lr-mat-over}Proportion of mature
-counterfactuals in each epoch. Data: Overlapping.}
-
-\end{figure}%
-
-\FloatBarrier
-
-\section{Computation Details}\label{computation-details}
-
-\subsection{Hardware}\label{sec-app-hardware}
-
-We performed our experiments on a high-performance cluster. Details
-about the cluster will be disclosed upon publication to avoid revealing
-information that might interfere with the double-blind review process.
-Since our experiments involve highly parallel tasks and rather small
-models by today's standard, we have relied on distributed computing
-across multiple central processing units (CPU). Graphical processing
-units (GPU) were not required.
-
-\subsubsection{Grid Searches}\label{grid-searches}
-
-Model training for the largest grid searches with 270 unique parameter
-combinations was parallelized across 34 CPUs with 2GB memory each. The
-time to completion varied by dataset for reasons discussed in
-Section~\ref{sec-discussion}: 0h49m (\emph{Moons}), 1h4m (\emph{Linearly
-Separable}), 1h49m (\emph{Circles}), 3h52m (\emph{Overlapping}). Model
-evaluations for large grid searches were parallelized across 20 CPUs
-with 3GB memory each. Evaluations for all data sets took less than one
-hour (\textless1h) to complete.
-
-\subsubsection{Tuning}\label{tuning}
-
-For tuning of selected hyperparameters, we distributed the task of
-generating counterfactuals during training across 40 CPUs with 2GB
-memory each for all tabular datasets. Except for the \emph{Adult}
-dataset, all training runs were completed in less that half an hour
-(\textless0h30m). The \emph{Adult} dataset took around 0h35m to
-complete. Evaluations across 20 CPUs with 3GB memory each generally took
-less than 0h30m to complete. For \emph{MNIST}, we relied on 100 CPUs
-with 2GB memory each. For the \emph{MLP}, training of all models could
-be completed in 1h30m, while the evaluation across 20 CPUs (6GB memory)
-took 4h12m. For the \emph{CNN}, training of all models took
-\textasciitilde8h, with conventionally trained models taking
-\textasciitilde0h15m each and model with CT taking
-\textasciitilde0h30m-0h45m each.
-
-\subsection{Software}\label{software}
-
-All computations were performed in the Julia Programming Language
-(\citeproc{ref-bezanson2017julia}{Bezanson et al. 2017}). We have
-developed a package for counterfactual training that leverages and
-extends the functionality provided by several existing packages, most
-notably
-\href{https://github.com/JuliaTrustworthyAI/CounterfactualExplanations.jl}{CounterfactualExplanations.jl}
-(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, and Liem
-2023}) and the \href{https://fluxml.ai/Flux.jl/v0.16/}{Flux.jl} library
-for deep learning (\citeproc{ref-innes2018fashionable}{Michael Innes et
-al. 2018}; \citeproc{ref-innes2018flux}{Mike Innes 2018}). For
-data-wrangling and presentation-ready tables we relied on
-\href{https://dataframes.juliadata.org/v1.7/}{DataFrames.jl}
-(\citeproc{ref-milan2023dataframes}{Bouchet-Valat and Kamiński 2023})
-and
-\href{https://ronisbr.github.io/PrettyTables.jl/v2.4/}{PrettyTables.jl}
-(\citeproc{ref-chagas2024pretty}{Chagas et al. 2024}), respectively. For
-plots and visualizations we used both
-\href{https://docs.juliaplots.org/v1.40/}{Plots.jl}
-(\citeproc{ref-PlotsJL}{Christ et al. 2023}) and
-\href{https://docs.makie.org/v0.22/}{Makie.jl}
-(\citeproc{ref-danisch2021makie}{Danisch and Krumbiegel 2021}), in
-particular \href{https://aog.makie.org/v0.9.3/}{AlgebraOfGraphics.jl}.
-To distribute computational tasks across multiple processors, we have
-relied on \href{https://juliaparallel.org/MPI.jl/v0.20/}{MPI.jl}
-(\citeproc{ref-byrne2021mpi}{Byrne, Wilcox, and Churavy 2021}).
-
-\end{appendices}
-
 \end{document}

### 2025-07-24 paper/presenations/slides.qmd

@@
 subtitle: Update Meeting Dec 2024
-author: 
-  - name: Patrick Altmeyer
-    url: https://www.patalt.org/
-  - name: Arie van Deursen
-  - name: Cynthia C. S. Liem
 institute: Delft University of Technology
@@
 ---
-
-## Status
-
-- *Code base*: In place and streamlined for reproducibility and configuration. 
-- *Experiments*: Lots of work done and results largely supportive of idea.
-    - Ran into problems on DelftBlue, which has set me back about 2 weeks. 
-- *Paper*: Still bare-bones. 
-- *ICML*: Potentially still possible to submit something, but this will be rushed and not "finished".
-
-## Problems on Cluster
-
-- Trying to distribute: 
-    1. Models/experiments across processes.
-    2. For each model/experiment distribute the counterfactual search across processes.
-- Out-of-memory issues, data races, ...
-- Multi-processing for models & multi-threading for counterfactual search: low CPU efficiency on DelftBlue (jobs get cancelled).
-
-## The Big Oversight
-
-- **Core problem**: `CounterfactualExplanation` objects are *huge*. They store `X` (the input), `y` (the output), etc. in their own memory.
-- Can't easily fit `CounterfactualExplanation` objects on cluster memory and pass them around processes. 
-- **Good news**: Relatively straightforward to add `FlattenedCE` with gazzilion times lower memory footprint (done!).
-- Nested parallelization issue remains and I will not spend more time trying to make it work.
 
@@
 
-- All counterfactuals at test time generated using *ECCo*.
+- All counterfactuals at test time generated using *ECCCo*.
 - Penalty on energy differential increases from *l.* to *r.*
@@
 :::{.column width='50%'}
-![Counterfactual explanations for model trained with CT (*ECCo*).](www/moons_ce_ecco.png){#fig-ce-ecco}
+![Counterfactual explanations for model trained with CT (*ECCCo*).](www/moons_ce_ecco.png){#fig-ce-ecco}
 :::
@@
 
-## MNIST (Vanilla vs *ECCo*)
+## MNIST (Vanilla vs *ECCCo*)
 
@@
 :::{.column width='50%'}
-![Faithful counterfactuals for same MLP with CT (*ECCo*).](www/poc_model_ct_ecco.png){#fig-mnist-model-ct-ecco}
+![Faithful counterfactuals for same MLP with CT (*ECCCo*).](www/poc_model_ct_ecco.png){#fig-mnist-model-ct-ecco}
 :::
@@
 ::::
-
-# Planning Ahead
-
-## Planned Contributions
-
-- Focus on enhancing trustworthiness and applicability of small opaque models. 
-    - Add more datasets and neuro-tree models (relatively straightforward).
-- Do not pitch as state-of-the-art approach to generative modelling similar to JEMs, but rather as an extension of JEMs to XAI and AR. 
-- Limit use of image datasets to illustrate arguments:
-    - Example: what happens if digit values can only be increased/decreased?
-
-## Timeline
-
-- Code base is already streamlined very well to allow for grid searches with easy configuration.
-- Lots of open tasks and questions (see [issue](https://github.com/JuliaTrustworthyAI/CounterfactualTraining.jl/issues/8))
-    1. Add MMD to measure plausibility.
-    2. Evaluate adversarial robustness and generative capacity. Compare to AT, JEM.
-    3. ... 
-- Still aiming for ICML to commit to a timeline, but it seems unlikely to be in the shape I would like.

### 2025-07-24 paper/quarto_aaai/paper.tex

@@
 Explanations}
-\author{Patrick Altmeyer \and Aleksander Buszydlik \and Arie van
-Deursen \and Cynthia C. S. Liem}
+\author{}
 \date{}
@@
 To assess outcomes with respect to actionability for non-linear models,
-we look at the average costs of valid counterfactuals in terms of their
-distances from factual starting points. While this an imperfect proxy of
-sensitivity, we hypothesize that CT can reduce these costs by teaching
-models to seek plausibility with respect to mutable features, much like
-we observe in Figure~\ref{fig-poc} in panel (d) compared to (c). We
-supplement this analysis with qualitative findings for integrated
-gradients \citep{sundararajan2017ig}. Finally, for predictive
+we look at the costs of (just) valid counterfactuals in terms of their
+distances from factual starting points with \(\tau=0.5\). While this an
+imperfect proxy of sensitivity, we hypothesize that CT can reduce these
+costs by teaching models to seek plausibility with respect to mutable
+features, much like we observe in Figure~\ref{fig-poc} in panel (d)
+compared to (c). We supplement this analysis with estimates for
+integrated gradients \citep{sundararajan2017ig}. Finally, for predictive
 performance, we use standard metrics, such as robust accuracy estimated

### 2025-07-24 paper/sections/conclusion.qmd

@@
-# Conclusions {#sec-conclusions}
+# Conclusion {#sec-conclusion}
 
@@
 Existing explainability solutions cannot guarantee that explanations agree with these learned representation.
-As a step towards addressing this challenge, we introduce counterfactual training, a novel training regime that incentivizes highly-explainable models.
-Through extensive experiments, we demonstrate that CT satisfies this objective while promoting robustness and preserving the predictive performance of models.
+As a step towards addressing this challenge, we introduce counterfactual training, a novel training regime that integrates recent advances in contrastive learning, adversarial robustness, and counterfactual explanations to incentivize highly-explainable models.
+Through extensive experiments, we demonstrate that CT satisfies this goal while preserving the predictive performance and promoting robustness of models.
 Explanations generated from CT-based models are both more plausible (compliant with the underlying data-generating process) and more actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients.
-We also show that CT can be used to fine-tune conventionally-trained models and achieve similar gains.
-Lastly, our work highlights the value of simultaneously improving models and their explanations.
+In turn, our work highlights the value of simultaneously improving models and their explanations.

### 2025-07-24 paper/sections/intro.qmd

@@
 Counterfactual explanations (CE) have become a key part of the broader explainable AI (XAI) toolkit [@molnar2022interpretable] that can be applied to make sense of this complexity.
-Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output (@fig-poc).
-This is useful and necessary to not only understand how opaque models make their predictions, but also to provide algorithmic recourse (AR) to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque automated decision-making (ADM) system.
+Originally proposed by @wachter2017counterfactual, CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce an alternative, more desirable output.
+This is useful and necessary to not only understand how opaque models make their predictions, but also to provide algorithmic recourse (AR) to individuals subjected to them: a retail bank, for example, could use CE to provide meaningful feedback to unsuccessful loan applicants that were rejected based on an opaque automated decision-making (ADM) system (@fig-poc).
 
@@
 Plausibility is typically understood as counterfactuals being *in-domain*: unsuccessful loan applicants that implement the provided recourse should end up with credit profiles that are genuinely similar to that of individuals who have successfully repaid their loans in the past.
-Actionable explanations comply with practical constraints: a young, unsuccessful loan applicant cannot increase their age in an instant.
+Actionable explanations comply with practical constraints: a young, unsuccessful loan applicant cannot increase their age in an instance.
 
@@
 This is problematic, because the narrow focus on any specific desideratum can adversely affect others: it is possible, for example, to generate plausible counterfactuals for models that are also highly vulnerable to implausible, possibly adversarial counterfactuals [@altmeyer2024faithful].
-In this work, we therefore embrace the paradigm that models---as opposed to explanation methods---should be held accountable for explanations that are plausible and actionable.
+In this work, we therefore embrace the paradigm that models (as opposed to explanation methods) should be held accountable for explanations that are plausible and actionable.
 While previous work has shown that at least plausibility can be indirectly achieved through existing techniques aimed at models' generative capacity, generalization and robustness [@altmeyer2024faithful; @augustin2020adversarial; @schut2021generating], we directly incorporate both plausibility and actionability in the training objective of models to improve their overall explanatory capacity.
@@
 Then follow our two principal contributions.
-In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints.
+In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to respect global actionability constraints.
 In our experiments (@sec-experiments), we find that thanks to counterfactual training, (1) the implausibility of CEs decreases by up to 90%; (2) the cost of reaching valid counterfactuals with protected features decreases by 19% on average; and (3) models' adversarial robustness improves across the board.
@@
 ![Counterfactual explanations (stars) for linear classifiers trained under different regimes on synthetic data: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable.
-The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=1$ (blue) and $y^+=2$ (orange).
-Stars indicate counterfactuals in the target class.
-For illustration, synthetic features are annotated as *debt* and *age*.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}
+The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=\text{"loan withheld"}$ (blue) and $y^+=\text{"loan provided"}$ (orange).
+Class and feature annotations (*debt* and *age*) are for illustrative purposes.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}

### 2025-07-24 paper/sections/other/notation.qmd

@@
 # Notation {.appendix}
+
+## Variables and Parameters
 
@@
 
+- $\mathcal{Y}$: The output domain.
 - $y^+$: The target class and also the index of the target class.
 - $y^-$: The non-target class and also the index of non-the target class.
+- $\mathcal{X}$: The input domain.
 - $\mathbf{x}$: a single training sample.
 - $\mathbf{x}^\prime$: a counterfactual.
+- $t=1,...,T$: Step indicator for counterfactual search iterations.
+- $\mathbf{x}^\prime_{t \leq T}$: a nascent counterfactual, defined as a counterfactual that has not yet converged.
+- $\mathbf{x}^\prime_{T}$: a mature counterfactual, defined as a counterfactual that has either passed a threshold probability or exhausted all $T$ steps.
 - $\mathbf{x}^+$: a training sample in the target class (ground-truth).
@@
 
-## Other Technical Details
+## Formulas
+
+### Maximum Mean Discrepancy
 
@@
 In our implementation, @eq-mmd is by default applied to the entire subset of the training data for which $y=y^+$.
+
+## Other Conventions
+
+In some place of this appendix, we use the terms *full/Full* (i.e.
+the full CT objective) and *vanilla/Vanilla* (i.e.
+vanilla training objective) to refer to models trained with counterfactual training (*CT*) and the baseline (*BL*), respectively.

### 2025-07-25 paper/quarto_aaai/paper.tex
mnist figure untangled

@@
 \sisetup{uncertainty-mode = separate}
+\usepackage{enumitem}
 \DeclareMathSizes{10}{9}{7}{6.5}
-\usepackage{enumitem}
+\usepackage{subcaption}
 
@@
 
-\pandocbounded{\includegraphics[keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}}
+\includegraphics[width=0.85\linewidth,height=\textheight,keepaspectratio]{paper_files/mediabag/../../paper/figures/poc.pdf}
 
@@
 \(\text{IP}\): reductions range from around 20\% (\emph{Moons}) to
-almost 90\% (\emph{Circ}). The only negative findings are for OL and
-MNIST, but they are not statistically significant. A qualitative
-inspection of the counterfactuals in Figure~\ref{fig-mnist} (columns
-2-5) suggests recognizable digits 1-4 for the model trained with CT
-(bottom row), unlike the baseline (top row).
+almost 90\% (\emph{Circ}). The only negative findings are for \emph{OL}
+and \emph{MNIST}, but they are insignificant. A qualitative inspection
+of the counterfactuals in Figure~\ref{fig-mnist-ce} suggests
+recognizable digits for the model trained with CT (bottom row), unlike
+the baseline (top row).
 
@@
 
+\begin{minipage}{\linewidth}
+
 \centering{
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_body.png}}
-
-}
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ce.png}}
+
+}
+
+\subcaption{\label{fig-mnist-ce}\emph{Plausibility}: faithful
+counterfactuals (\emph{ECCCo}).}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}
+
+}
+
+\subcaption{\label{fig-mnist-ig}\emph{Actionability}: class-cond.
+integrated gradients.}
+
+\end{minipage}%
 
 \caption{\label{fig-mnist}Visual explanations for \emph{MNIST} for BL
-(top) and CT (bottom). \textbf{Plausibility}: col.~1 is a random factual
-0 (blue); cols. 2-5 are corresponding \emph{ECCCo} counterfactuals in
-target classes 1 to 4. \textbf{Actionability}: cols. 6-10 show
-integrated gradients averaged over test images in classes 5 to 9.}
+(top rows) and CT (bottom rows).}
 
@@
 real-world tabular datasets we could identify and protect an \emph{age}
-variable; for \emph{MNIST}, we protect the five upper and lower pixel
+variable; for \emph{MNIST}, we protect the five top and bottom pixel
 rows of the full image. Statistically significant reductions in costs
-overwhelmingly point in the expected positive direction reaching up to
-around 60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average
-costs increase, likely because any potential benefits from protecting
-the \emph{age} are outweighed by the increase in costs required for
-greater plausibility. The findings for \emph{Adult} and \emph{MNIST} are
-not significant. A qualitative inspection of the class-conditional
-integrated gradients in Figure~\ref{fig-mnist} (columns 6-10) suggests
-that CT still has the expected effect: the model (bottom) is insensitive
-(blue) to the protected rows of pixels; details of this experiment are
-reported in the supplementary appendix.
+overwhelmingly point in the positive direction reaching up to around
+60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average costs
+increase, likely because any benefits from protecting \emph{age} are
+outweighed by an increase in costs required for greater plausibility.
+The findings for \emph{Adult} and \emph{MNIST} are insignificant. A
+qualitative inspection of the class-conditional integrated gradients in
+Figure~\ref{fig-mnist-ig} suggests that CT still has the expected
+effect: the model (bottom) is insensitive (blue) to the protected rows
+of pixels; details are reported in the appendix.
 
@@
 As our results indicate, counterfactual training produces models that
-are more explainable. Nonetheless, these advantages come at the cost of
-two important limitations.
+are more explainable. Nonetheless, these advantages come with certain
+limitations.
 
@@
 State-of-the-art machine learning models are prone to learning complex
-representations that cannot be interpreted by humans. Existing
-explainability solutions cannot guarantee that explanations agree with
-these learned representation. As a step towards addressing this
-challenge, we introduce counterfactual training, a novel training regime
-that integrates recent advances in contrastive learning, adversarial
-robustness, and counterfactual explanations to incentivize
-highly-explainable models. Through extensive experiments, we demonstrate
-that CT satisfies this goal while preserving the predictive performance
-and promoting robustness of models. Explanations generated from CT-based
-models are both more plausible (compliant with the underlying
-data-generating process) and more actionable (compliant with
-user-specified mutability constraints), and thus meaningful to their
-recipients. In turn, our work highlights the value of simultaneously
-improving models and their explanations.
+representations that cannot be interpreted by humans. Existing work on
+counterfactual explanations has largely focused on designing tools to
+generate plausible and actionable for any model. In this work, we
+instead hold models accountable for delivering such explanations. We
+introduce counterfactual training: a novel training regime that
+integrates recent advances in contrastive learning, adversarial
+robustness, and CE to incentivize highly-explainable models. Through
+theoretical results and extensive experiments, we demonstrate that CT
+satisfies this goal while promoting adversarial robustness of models.
+Explanations generated from CT-based models are both more plausible
+(compliant with the underlying data-generating process) and more
+actionable (compliant with user-specified mutability constraints), and
+thus meaningful to their recipients. In turn, our work highlights the
+value of simultaneously improving models and their explanations.
 

### 2025-07-25 paper/sections/conclusion.qmd
mnist figure untangled

@@
 State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans.
-Existing explainability solutions cannot guarantee that explanations agree with these learned representation.
-As a step towards addressing this challenge, we introduce counterfactual training, a novel training regime that integrates recent advances in contrastive learning, adversarial robustness, and counterfactual explanations to incentivize highly-explainable models.
-Through extensive experiments, we demonstrate that CT satisfies this goal while preserving the predictive performance and promoting robustness of models.
+Existing work on counterfactual explanations has largely focused on designing tools to generate plausible and actionable for any model.
+In this work, we instead hold models accountable for delivering such explanations.
+We introduce counterfactual training: a novel training regime that integrates recent advances in contrastive learning, adversarial robustness, and CE to incentivize highly-explainable models.
+Through theoretical results and extensive experiments, we demonstrate that CT satisfies this goal while promoting adversarial robustness of models.
 Explanations generated from CT-based models are both more plausible (compliant with the underlying data-generating process) and more actionable (compliant with user-specified mutability constraints), and thus meaningful to their recipients.

### 2025-07-25 paper/sections/experiments.qmd
mnist figure untangled

@@
 Reductions in $\text{IP}^*$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $\text{IP}$: reductions range from around 20% (*Moons*) to almost 90% (*Circ*).
-The only negative findings are for OL and MNIST, but they are not statistically significant.
-A qualitative inspection of the counterfactuals in @fig-mnist (columns 2-5) suggests recognizable digits 1-4 for the model trained with CT (bottom row), unlike the baseline (top row).
+The only negative findings are for *OL* and *MNIST*, but they are insignificant.
+A qualitative inspection of the counterfactuals in @fig-mnist-ce suggests recognizable digits for the model trained with CT (bottom row), unlike the baseline (top row).
 
-![Visual explanations for *MNIST* for BL (top) and CT (bottom).
-**Plausibility**: col.
-1 is a random factual 0 (blue); cols.
-2-5 are corresponding *ECCCo* counterfactuals in target classes 1 to 4.
-**Actionability**: cols.
-6-10 show integrated gradients averaged over test images in classes 5 to 9.](/paper/figures/mnist_body.png){#fig-mnist}
+::: {#fig-mnist layout-nrow=2 layout-ncol=1}
+
+![*Plausibility*: faithful counterfactuals (*ECCCo*).](/paper/figures/mnist_ce.png){#fig-mnist-ce}
+
+![*Actionability*: class-cond. integrated gradients.](/paper/figures/mnist_ig.png){#fig-mnist-ig}
+
+Visual explanations for *MNIST* for BL (top rows) and CT (bottom rows).
+
+:::
 
@@
 
-For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five upper and lower pixel rows of the full image.
-Statistically significant reductions in costs overwhelmingly point in the expected positive direction reaching up to around 60% for *GMSC*.
-Only in the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
-The findings for *Adult* and *MNIST* are not significant.
-A qualitative inspection of the class-conditional integrated gradients in @fig-mnist (columns 6-10) suggests that CT still has the expected effect: the model (bottom) is insensitive (blue) to the protected rows of pixels; details of this experiment are reported in the supplementary appendix.
+For synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom pixel rows of the full image.
+Statistically significant reductions in costs overwhelmingly point in the positive direction reaching up to around 60% for *GMSC*.
+Only in the case of *Cred*, average costs increase, likely because any benefits from protecting *age* are outweighed by an increase in costs required for greater plausibility.
+The findings for *Adult* and *MNIST* are insignificant.
+A qualitative inspection of the class-conditional integrated gradients in @fig-mnist-ig suggests that CT still has the expected effect: the model (bottom) is insensitive (blue) to the protected rows of pixels; details are reported in the appendix.
 

### 2025-07-25 paper/experiments/mnist_chart.jl

@@ -44,7 +44,7 @@ factual = chosen_digits[1]
 both_predict_factual = false
 while !both_predict_factual
     chosen = rand(findall(data.output_encoder.labels .== factual))
-    x = select_factual(data, chosen)
+    global x = select_factual(data, chosen)
     both_predict_factual =
         predict_label(models[1], data, x)[1] ==
         predict_label(models[2], data, x)[1] ==
@@ -56,7 +56,7 @@ imgs = []
 conv = CounterfactualExplanations.Convergence.MaxIterConvergence(1000)
 for M in models
     imgs_M = [CTExperiments.convert2mnist(x; blue=true)]
-    for target in chosen_digits[2:end]
+    for target in all_digits[2:end]
         img =
             generate_counterfactual(x, target, data, M, gen; convergence=conv) |>
             CounterfactualExplanations.counterfactual |>
@@ -69,6 +69,10 @@ end
 imgs_ct = imgs[1]
 imgs_bl = imgs[2]
 img_ce = mosaicview(imgs_bl..., imgs_ct...; nrow=2, rowmajor=true)
+Images.save("paper/figures/mnist_ce.png",img_ce)
 
+# Combined image:
+idx = [findall(all_digits .== x)[1] for x in chosen_digits]
+img_ce = mosaicview(imgs_bl[idx]..., imgs_ct[idx]...; nrow=2, rowmajor=true)
 img = mosaicview(img_ce, img_ig; nrow=1)
 Images.save("paper/figures/mnist_body.png", img)

### 2025-07-25 paper/quarto_aaai/paper.tex
final adjustments

@@
 
-\caption{\label{fig-mnist}Visual explanations for \emph{MNIST} for BL
-(top rows) and CT (bottom rows).}
+\caption{\label{fig-mnist}Visual explanations for \emph{MNIST}. Top and
+bottom rows (of images) show the results for BL and CT, respectively.
+Mutability constraints are imposed on the five top and bottom rows (of
+pixels). CT produces plausible counterfactuals (a) and is insensitive to
+protected features (b).}
 

### 2025-07-25 paper/sections/experiments.qmd
final adjustments

@@
 
-Visual explanations for *MNIST* for BL (top rows) and CT (bottom rows).
+Visual explanations for *MNIST*.
+Top and bottom rows (of images) show the results for BL and CT, respectively.
+Mutability constraints are imposed on the five top and bottom rows (of pixels).
+CT produces plausible counterfactuals (a) and is insensitive to protected features (b).
 

### 2025-07-25 paper/quarto_aaai/paper.tex
ready to submit

@@
 
-\subcaption{\label{fig-mnist-ce}\emph{Plausibility}: faithful
-counterfactuals (\emph{ECCCo}).}
+\subcaption{\label{fig-mnist-ce}\emph{Plausibility}: \emph{ECCCo}
+counterfactuals for factual `0' (blue).}
 
@@
 Mutability constraints are imposed on the five top and bottom rows (of
-pixels). CT produces plausible counterfactuals (a) and is insensitive to
-protected features (b).}
+pixels). CT produces plausible counterfactuals (a) and is less sensitive
+to protected features (b).}
 
@@
 Figure~\ref{fig-mnist-ig} suggests that CT still has the expected
-effect: the model (bottom) is insensitive (blue) to the protected rows
-of pixels; details are reported in the appendix.
+effect: the model (bottom) is less sensitive (blue) to the protected
+rows of pixels; details are reported in the appendix.
 

### 2025-07-25 paper/sections/other/main_results.qmd
ready to submit

@@
 The selection of an appropriate baseline is an important design decision — it should produce a "neutral" prediction to avoid capturing effects that cannot be directly attributed to the model [@sundararajan2017ig; @sturmfels2020visualizing].
-To remain consistent in our evaluations, we use a baseline drawn at random from a uniform distribution, $\mathcal{U}(-1,1)$ all datasets.
+To remain consistent in our evaluations, we use a baseline drawn at random from a uniform distribution, $\mathcal{U}(-1,1)$, for all datasets.
 This aligns with standard evaluation practices for IG.
@@
 As IG outputs are not bounded (i.e., they are arbitrary real numbers), it becomes a challenge to meaningfully compare IG outputs of different models — ones that are trained conventionally, and ones that underwent counterfactual training.
-For our purposes, we observe that the direction of contribution (sign) does not matter.
-If protecting a feature flips the sign of the corresponding output, then CT „overcorrected” for this feature, but as long as the magnitude of its contribution decreases, then CT achieved its objective according to our formulation.
-Thus, we focus on the magnitude of IG outputs in the evaluation.
-Next, to accommodate for variable ranges of outputs, we standardize the outputs.
+For our purposes, we observe with reference to our Proposition, that we are interested estimating changes in the relative contribution of protected features compared to mutable ones.
+Thus, to meaningfully compare integrated gradients for different models and to accommodate for variable ranges of outputs in absolute terms, we standardize the integrated gradients across features.
 
@@
 Only for *GMSC* we observe negative impacts of CT which we believe is due to any or all of the following: a) data assumptions are violated; b) the impact of other components of the CT objective outweighs expected effects of feature protection; or c) the default baseline choice applied consistently to all datasets is not appropriate for GMSC.
-
-::: {layout-ncol=2}
 
@@
 Average sensitivity $\pm$ bootstrapped standard errors of protected features as per standardized integrated gradients.
-
-:::
-
-![Class-conditional standardized integrated gradients for MNIST.
-Top: BL.
-Bottom: CT.
-Mutability constraints were imposed on the five top and bottom rows of pixels.](/paper/figures/mnist_ig.png){#fig-mnist-ig}
 

### 2025-07-25 paper/preprint/appendix.tex

@@
 and
-\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\)
+\((\theta_{y^-,d_{\text{mtbl}}} - \theta_{y^+,d_{\text{mtbl}}}) > 0\).
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-default}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacityback=0, colback=white]
 
@@
 2020}). To remain consistent in our evaluations, we use a baseline drawn
-at random from a uniform distribution, \(\mathcal{U}(-1,1)\) all
+at random from a uniform distribution, \(\mathcal{U}(-1,1)\), for all
 datasets. This aligns with standard evaluation practices for IG.
@@
 models --- ones that are trained conventionally, and ones that underwent
-counterfactual training. For our purposes, we observe that the direction
-of contribution (sign) does not matter. If protecting a feature flips
-the sign of the corresponding output, then CT „overcorrected'' for this
-feature, but as long as the magnitude of its contribution decreases,
-then CT achieved its objective according to our formulation. Thus, we
-focus on the magnitude of IG outputs in the evaluation. Next, to
-accommodate for variable ranges of outputs, we standardize the outputs.
+counterfactual training. For our purposes, we observe with reference to
+our Proposition, that we are interested estimating changes in the
+relative contribution of protected features compared to mutable ones.
+Thus, to meaningfully compare integrated gradients for different models
+and to accommodate for variable ranges of outputs in absolute terms, we
+standardize the integrated gradients across features.
 
@@
 meaningful comparison. In the case of real-world datasets we choose to
-normalize the values to a 0-1 range instead.
-
-We compare the (average) sensitivity to the features that were protected
-for CT models. We again use bootstrapping (100 rounds, 2500 samples per
-round) to establish the significance of our results. These are presented
-in Table~\ref{tbl-ig}.
+normalize the values to a \([0,1]\) range instead. We compare the
+(average) sensitivity to the features that were protected for CT models.
+Once again we use bootstrapping (100 rounds, 2500 samples per round) to
+establish the significance of our results. These are presented in
+Table~\ref{tbl-ig}.
 
@@
 
-\begin{figure}
-
-\begin{minipage}{0.50\linewidth}
-
-\begin{table}[H]
+\begin{table}
 
@@
 \end{table}%
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\begin{figure}[H]
-
-\centering{
-
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}
-
-}
-
-\caption{\label{fig-mnist-ig}Class-conditional standardized integrated
-gradients for MNIST. Top: BL. Bottom: CT. Mutability constraints were
-imposed on the five top and bottom rows of pixels.}
-
-\end{figure}%
-
-\end{minipage}%
-
-\end{figure}%
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacityback=0, colback=white]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, rightrule=.15mm, opacitybacktitle=0.6, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, arc=.35mm, breakable, toprule=.15mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, colback=white, toptitle=1mm, bottomrule=.15mm, leftrule=.75mm, opacityback=0]
+\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacityback=0, colback=white]
 

### 2025-07-25 paper/preprint/paper.tex

@@
 
+\begin{minipage}{\linewidth}
+
 \centering{
 
-\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_body.png}}
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ce.png}}
 
@@
 
-\caption{\label{fig-mnist}Visual explanations for \emph{MNIST} for BL
-(top) and CT (bottom). \textbf{Plausibility}: col.~1 is a random factual
-0 (blue); cols. 2-5 are corresponding \emph{ECCCo} counterfactuals in
-target classes 1 to 4. \textbf{Actionability}: cols. 6-10 show
-integrated gradients averaged over test images in classes 5 to 9.}
+\subcaption{\label{fig-mnist-ce}\emph{Plausibility}: \emph{ECCCo}
+counterfactuals for factual `0' (blue).}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_ig.png}}
+
+}
+
+\subcaption{\label{fig-mnist-ig}\emph{Actionability}: class-cond.
+integrated gradients.}
+
+\end{minipage}%
+
+\caption{\label{fig-mnist}Visual explanations for \emph{MNIST}. Top and
+bottom rows (of images) show the results for BL and CT, respectively.
+Mutability constraints are imposed on the five top and bottom rows (of
+pixels). CT produces plausible counterfactuals (a) and is less sensitive
+to protected features (b).}
 
@@
 real-world tabular datasets we could identify and protect an \emph{age}
-variable; for \emph{MNIST}, we protect the five upper and lower pixel
+variable; for \emph{MNIST}, we protect the five top and bottom pixel
 rows of the full image. Statistically significant reductions in costs
-overwhelmingly point in the expected positive direction reaching up to
-around 60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average
-costs increase, likely because any potential benefits from protecting
-the \emph{age} are outweighed by the increase in costs required for
-greater plausibility. The findings for \emph{Adult} and \emph{MNIST} are
-not significant. A qualitative inspection of the class-conditional
-integrated gradients in Figure~\ref{fig-mnist} (columns 6-10) suggests
-that CT still has the expected effect: the model (bottom) is insensitive
-(blue) to the protected rows of pixels; details of this experiment are
-reported in the supplementary appendix.
+overwhelmingly point in the positive direction reaching up to around
+60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average costs
+increase, likely because any benefits from protecting \emph{age} are
+outweighed by an increase in costs required for greater plausibility.
+The findings for \emph{Adult} and \emph{MNIST} are insignificant. A
+qualitative inspection of the class-conditional integrated gradients in
+Figure~\ref{fig-mnist-ig} suggests that CT still has the expected
+effect: the model (bottom) is less sensitive (blue) to the protected
+rows of pixels; details are reported in the appendix.
 

### 2025-07-25 paper/presenations/slides.qmd

@@
 subtitle: Update Meeting Dec 2024
-institute: Delft University of Technology
 date: today

### 2025-07-25 paper/sections/experiments.qmd

@@
 
-![*Plausibility*: faithful counterfactuals (*ECCCo*).](/paper/figures/mnist_ce.png){#fig-mnist-ce}
+![*Plausibility*: *ECCCo* counterfactuals for factual '0' (blue).](/paper/figures/mnist_ce.png){#fig-mnist-ce}
 
@@
 Mutability constraints are imposed on the five top and bottom rows (of pixels).
-CT produces plausible counterfactuals (a) and is insensitive to protected features (b).
+CT produces plausible counterfactuals (a) and is less sensitive to protected features (b).
 
@@
 The findings for *Adult* and *MNIST* are insignificant.
-A qualitative inspection of the class-conditional integrated gradients in @fig-mnist-ig suggests that CT still has the expected effect: the model (bottom) is insensitive (blue) to the protected rows of pixels; details are reported in the appendix.
+A qualitative inspection of the class-conditional integrated gradients in @fig-mnist-ig suggests that CT still has the expected effect: the model (bottom) is less sensitive (blue) to the protected rows of pixels; details are reported in the appendix.
 

### 2025-07-29 paper/preprint/appendix.tex
final pass-through done

@@
 \item
-  \(\mathbf{x}^\prime_{t \leq T}\): a nascent counterfactual, defined as
-  a counterfactual that has not yet converged.
-\item
-  \(\mathbf{x}^\prime_{T}\): a mature counterfactual, defined as a
-  counterfactual that has either passed a threshold probability or
+  \(\mathbf{x}_{\text{AE}}^\prime\): a nascent counterfactual, defined
+  as a counterfactual that has not yet converged.
+\item
+  \(\mathbf{x}_{\text{CE}}^\prime\): a mature counterfactual, defined as
+  a counterfactual that has either passed a threshold probability or
   exhausted all \(T\) steps.
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-default}: Training Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-default}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 outcomes (Note~\ref{nte-eval-default}). This is in line with standard
-practice in the related literature on CE. For our final results
+practice in the related literature on CE (see e.g. Schut et al.
+(\citeproc{ref-schut2021generating}{2021})). For our final results
 presented in the main paper, we rely on held-out test sets for
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-eval-default}: Evaluation Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-eval-default}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-final-run-train}: Training Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune-train}: Training Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune_lr-train}: Training Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-train}: Training Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
-\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, breakable, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm, opacitybacktitle=0.6, toprule=.15mm, toptitle=1mm, rightrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, left=2mm, bottomrule=.15mm, leftrule=.75mm, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, opacityback=0, colback=white]
+\begin{tcolorbox}[enhanced jigsaw, breakable, bottomrule=.15mm, opacitybacktitle=0.6, colback=white, arc=.35mm, toprule=.15mm, coltitle=black, left=2mm, colbacktitle=quarto-callout-note-color!10!white, toptitle=1mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, leftrule=.75mm, rightrule=.15mm, bottomtitle=1mm, opacityback=0]
 
@@
 
+\bibitem[\citeproctext]{ref-schut2021generating}
+Schut, Lisa, Oscar Key, Rory McGrath, Luca Costabello, Bogdan Sacaleanu,
+Yarin Gal, et al. 2021. {``Generating {Interpretable Counterfactual
+Explanations By Implicit Minimisation} of {Epistemic} and {Aleatoric
+Uncertainties}.''} In \emph{International {Conference} on {Artificial
+Intelligence} and {Statistics}}, 1756--64. {PMLR}.
+
 \bibitem[\citeproctext]{ref-sturmfels2020visualizing}

### 2025-07-29 paper/preprint/paper.tex
final pass-through done

@@
 classifier is roughly equally sensitive to both features. By contrast,
-the decision boundary in panel (d) has titled, making the model trained
+the decision boundary in panel (d) has tilted, making the model trained
 with CT relatively less sensitive to the immutable \emph{age} feature.
@@
 (\citeproc{ref-wu2021polyjuice}{2021}) propose \emph{Polyjuice}, a
-general-purpose CE generator for language models and demonstrate that
+general-purpose CE generator for language models, and demonstrate that
 the augmentation of training data with \emph{Polyjuice} improves
-robustness in a number of tasks, while Luu and Inoue
+robustness in a number of tasks. Luu and Inoue
 (\citeproc{ref-luu2023counterfactual}{2023}) introduce the
@@
 faithful counterfactuals are consistent with what the model has learned
-about the input data. Ac tionability constraints in Def.
+about the input data. Actionability constraints in Def.
 \ref{def-explainability} vary and depend on the context in which
@@
 term \emph{nascent} to denote interim counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\) that have not yet converged. As we
+\(\mathbf{x}_{\text{AE}}^\prime\) that have not yet converged. As we
 explain below, these nascent counterfactuals can be stored and
 repurposed as adversarial examples. Conversely, we consider
-counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
-they have either exhausted all \(T\) iterations or converged by reaching
-a pre-specified threshold, \(\tau\), for the predicted probability of
-the target class:
+counterfactuals \(\mathbf{x}_{\text{CE}}^\prime\) as \emph{mature}
+explanations if they have converged within the \(T\) iterations by
+reaching a pre-specified threshold, \(\tau\), for the predicted
+probability of the target class:
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
@@
 \begin{aligned}
-&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
+&\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta) \\+ &\lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{\text{AE}}^\prime),\mathbf{y}) + \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta)
 \end{aligned}
@@
 penalizing (1) the contrastive divergence, \(\text{div}(\cdot)\),
-between mature counterfactuals \(\mathbf{x}_T^\prime\) and observed
-samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the
-target class \(y^+\), and (2) the adversarial loss,
+between mature counterfactuals \(\mathbf{x}_{\text{CE}}^\prime\) and
+observed samples \(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\)
+in the target class \(y^+\), and (2) the adversarial loss,
 \(\text{advloss}(.)\), wrt. nascent counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\). Finally, \(\text{ridge}(\cdot)\)
+\(\mathbf{x}_{\text{AE}}^\prime\). Finally, \(\text{ridge}(\cdot)\)
 denotes a Ridge penalty (\(\ell_2\)-norm) that regularizes the magnitude
@@
 \begin{equation}\phantomsection\label{eq-div}{
-\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
+\text{div}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_{\text{CE}}^\prime,y)
 }\end{equation} where \(\mathcal{E}_\theta(\cdot)\) denotes the energy
@@
 \(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
-\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
-sampled factual from a non-target class generated with a gradient-based
-CE generator for up to \(T\) iterations. Intuitively, the gradient of
-Equation~\ref{eq-div} decreases the energy of observed training samples
-(positive samples) while increasing the energy of counterfactuals
-(negative samples) (\citeproc{ref-du2019implicit}{Du and Mordatch
-2020}). As the counterfactuals get more plausible (Def.
+\(\mathbf{x}_{\text{CE}}^\prime\) denotes a mature counterfactual for a
+randomly sampled factual from a non-target class generated with a
+gradient-based CE generator for up to \(T\) iterations. Intuitively, the
+gradient of Equation~\ref{eq-div} decreases the energy of observed
+training samples (positive samples) while increasing the energy of
+counterfactuals (negative samples) (\citeproc{ref-du2019implicit}{Du and
+Mordatch 2020}). As the counterfactuals get more plausible (Def.
 \ref{def-explainability}) during training, these opposing effects
@@
 \begin{aligned}
-\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
+\text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{\text{AE}}^\prime),\mathbf{y};\varepsilon)&=\text{yloss}(\mathbf{M}_\theta(\mathbf{x}_{t_\varepsilon}^\prime),\mathbf{y}) \\
 t_\varepsilon &= \max_t \{t: ||\Delta_t||_\infty < \varepsilon\}
@@
 }\end{equation} Under this choice, we consider nascent counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\) as AEs as long as the magnitude of the
+\(\mathbf{x}_{\text{AE}}^\prime\) as AEs as long as the magnitude of the
 perturbation to any single feature is at most \(\varepsilon\). This is
@@
 literature. Four of them are synthetic with two classes and different
-characteristics: linearly separable clusters (\emph{LS}), overlapping
-clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
-moons (\emph{Moon}). Next, we have four real-world binary tabular
-datasets: \emph{Adult} (Census data) of Becker and Kohavi
+characteristics: linearly separable Gaussian clusters (\emph{LS}),
+overlapping clusters (\emph{OL}), concentric circles (\emph{Circ}), and
+interlocking moons (\emph{Moon}). Next, we have four real-world binary
+tabular datasets: \emph{Adult} (Census data) of Becker and Kohavi
 (\citeproc{ref-becker1996adult2}{1996}), California housing (\emph{CH})
@@
 Test accuracies on adversarially perturbed data are shown in
-Figure~\ref{fig-acc}. The perturbations size, \(\varepsilon\in[0,0.1]\),
+Figure~\ref{fig-acc}. The perturbation size, \(\varepsilon\in[0,0.1]\),
 increases along the horizontal axis and includes zero, corresponding to
@@
 We evaluate the impacts of three types of hyperparameters on CT. In this
-section we focus on the highlights and make the full results available
+section, we focus on the highlights and make the full results available
 in the supplementary appendix.
@@
 Secondly, the strength of the energy regularization,
-\(\lambda_{\text{reg}}\) is highly impactful and should be set
+\(\lambda_{\text{reg}}\), is highly impactful and should be set
 sufficiently high to avoid common problems associated with exploding
@@
 
-\emph{CT increases the training times.} Just like contrastive and robust
+\emph{CT increases training times.} Just like contrastive and robust
 learning, CT is more resource-intensive than conventional regimes. Three
@@
 counterfactual explanations has largely focused on designing tools to
-generate plausible and actionable for any model. In this work, we
-instead hold models accountable for delivering such explanations. We
-introduce counterfactual training: a novel training regime that
-integrates recent advances in contrastive learning, adversarial
-robustness, and CE to incentivize highly-explainable models. Through
-theoretical results and extensive experiments, we demonstrate that CT
-satisfies this goal while promoting adversarial robustness of models.
-Explanations generated from CT-based models are both more plausible
-(compliant with the underlying data-generating process) and more
-actionable (compliant with user-specified mutability constraints), and
-thus meaningful to their recipients. In turn, our work highlights the
-value of simultaneously improving models and their explanations.
+generate plausible and actionable explanations for any model. In this
+work, we instead hold models accountable for delivering such
+explanations. We introduce counterfactual training: a novel training
+regime that integrates recent advances in contrastive learning,
+adversarial robustness, and CE to incentivize highly explainable models.
+Through theoretical results and extensive experiments, we demonstrate
+that CT satisfies this goal while promoting adversarial robustness of
+models. Explanations generated from CT-based models are both more
+plausible (compliant with the underlying data-generating process) and
+more actionable (compliant with user-specified mutability constraints),
+and thus meaningful to their recipients. In turn, our work highlights
+the value of simultaneously improving models and their explanations.
 

### 2025-07-29 paper/sections/conclusion.qmd
final pass-through done

@@
 State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans.
-Existing work on counterfactual explanations has largely focused on designing tools to generate plausible and actionable for any model.
+Existing work on counterfactual explanations has largely focused on designing tools to generate plausible and actionable explanations for any model.
 In this work, we instead hold models accountable for delivering such explanations.
-We introduce counterfactual training: a novel training regime that integrates recent advances in contrastive learning, adversarial robustness, and CE to incentivize highly-explainable models.
+We introduce counterfactual training: a novel training regime that integrates recent advances in contrastive learning, adversarial robustness, and CE to incentivize highly explainable models.
 Through theoretical results and extensive experiments, we demonstrate that CT satisfies this goal while promoting adversarial robustness of models.

### 2025-07-29 paper/sections/discussion.qmd
mnist figure untangled
final pass-through done

@@
 
-As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, these advantages come at the cost of two important limitations.
+As our results indicate, counterfactual training produces models that are more explainable. Nonetheless, these advantages come with certain limitations.
 
@@
 
-*CT increases the training times.* Just like contrastive and robust learning, CT is more resource-intensive than conventional regimes.
+*CT increases training times.* Just like contrastive and robust learning, CT is more resource-intensive than conventional regimes.
 Three factors mitigate this effect: (1) CT yields itself to parallel execution; (2) it amortizes the cost of CEs for the training samples; and (3) it can be used to fine-tune conventionally-trained models.

### 2025-07-29 paper/sections/intro.qmd
mnist figure untangled
final pass-through done

@@
 In panel (c), the classifier is again trained conventionally and we have introduced a mutability constraint on the *age* feature at test time---counterfactuals are valid but the classifier is roughly equally sensitive to both features.
-By contrast, the decision boundary in panel (d) has titled, making the model trained with CT relatively less sensitive to the immutable *age* feature.
+By contrast, the decision boundary in panel (d) has tilted, making the model trained with CT relatively less sensitive to the immutable *age* feature.
 To achieve these outcomes, CT draws inspiration from the literature on contrastive and robust learning: we contrast faithful CEs with ground-truth data while protecting immutable features, and capitalize on methodological links between CE and AE by penalizing the model's adversarial loss on interim (*nascent*) counterfactuals.
@@
 The linear decision boundary is shown in green along with training data colored according to ground-truth labels: $y^-=\text{"loan withheld"}$ (blue) and $y^+=\text{"loan provided"}$ (orange).
-Class and feature annotations (*debt* and *age*) are for illustrative purposes.](/paper/figures/poc.svg){#fig-poc fig-env="figure*"}
+Class and feature annotations (*debt* and *age*) are for illustrative purposes.](/paper/figures/poc.svg){#fig-poc fig-env="figure*" width=85%}

### 2025-07-29 paper/sections/method.qmd
final pass-through done

@@
 Intuitively, plausible counterfactuals are consistent with the data and faithful counterfactuals are consistent with what the model has learned about the input data.
-Ac tionability constraints in Def.
+Actionability constraints in Def.
 \ref{def-explainability} vary and depend on the context in which $\mathbf{M}_\theta$ is deployed.
@@
 CT adopts gradient-based CE search in training to generate on-the-fly model explanations $\mathbf{x}^\prime$ for the training samples.
-We use the term *nascent* to denote interim counterfactuals $\mathbf{x}_{t\leq T}^\prime$ that have not yet converged.
+We use the term *nascent* to denote interim counterfactuals $\mathbf{x}_{\text{AE}}^\prime$ that have not yet converged.
 As we explain below, these nascent counterfactuals can be stored and repurposed as adversarial examples.
-Conversely, we consider counterfactuals $\mathbf{x}_T^\prime$ as *mature* explanations if they have either exhausted all $T$ iterations or converged by reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.
+Conversely, we consider counterfactuals $\mathbf{x}_{\text{CE}}^\prime$ as *mature* explanations if they have converged within the $T$ iterations by reaching a pre-specified threshold, $\tau$, for the predicted probability of the target class: $\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau$, where $\mathcal{S}$ is the softmax function.
 
@@
 The second and third terms are explained in detail below.
-For now, they can be summarized as inducing explainability directly and indirectly by penalizing (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_T^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (2) the adversarial loss, $\text{advloss}(.)$, wrt.
-nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$.
+For now, they can be summarized as inducing explainability directly and indirectly by penalizing (1) the contrastive divergence, $\text{div}(\cdot)$, between mature counterfactuals $\mathbf{x}_{\text{CE}}^\prime$ and observed samples $\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}$ in the target class $y^+$, and (2) the adversarial loss, $\text{advloss}(.)$, wrt.
+nascent counterfactuals $\mathbf{x}_{\text{AE}}^\prime$.
 Finally, $\text{ridge}(\cdot)$ denotes a Ridge penalty ($\ell_2$-norm) that regularizes the magnitude of the energy terms involved in $\text{div}(\cdot)$ [@du2019implicit].
@@
 $$
-\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
+\text{div}(\mathbf{x}^+,\mathbf{x}_{\text{CE}}^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_{\text{CE}}^\prime,y)
 $$ {#eq-div}
 where $\mathcal{E}_\theta(\cdot)$ denotes the energy function defined as $\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]$, with $y^+$ denoting the index of the randomly drawn target class, $y^+ \sim p(y)$.
-Conditional on the target class $y^+$, $\mathbf{x}_T^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations.
+Conditional on the target class $y^+$, $\mathbf{x}_{\text{CE}}^\prime$ denotes a mature counterfactual for a randomly sampled factual from a non-target class generated with a gradient-based CE generator for up to $T$ iterations.
 Intuitively, the gradient of @eq-div decreases the energy of observed training samples (positive samples) while increasing the energy of counterfactuals (negative samples) [@du2019implicit].
@@
 $$ {#eq-adv}
-Under this choice, we consider nascent counterfactuals $\mathbf{x}_{t\leq T}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$.
+Under this choice, we consider nascent counterfactuals $\mathbf{x}_{\text{AE}}^\prime$ as AEs as long as the magnitude of the perturbation to any single feature is at most $\varepsilon$.
 This is closely aligned with @szegedy2013intriguing who define an adversarial attack as an "imperceptible non-random perturbation".

### 2025-07-29 paper/sections/related.qmd
final pass-through done

@@
 CEs have also been used to improve models in the natural language processing domain.
-For example, @wu2021polyjuice propose *Polyjuice*, a general-purpose CE generator for language models and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks, while @luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
+For example, @wu2021polyjuice propose *Polyjuice*, a general-purpose CE generator for language models, and demonstrate that the augmentation of training data with *Polyjuice* improves robustness in a number of tasks.
+@luu2023counterfactual introduce the *Counterfactual Adversarial Training* (CAT) framework that aims to improve generalization and robustness of language models by generating counterfactuals for training samples that are subject to high predictive uncertainty.
 

### 2025-07-29 src/counterfactuals.jl

@@ -74,7 +74,7 @@ function generate!(
         ces
     )   # randomly draw a sample from the target class
     validities = (ce -> ce.search[:converged]).(ces)
-    # Extract counterfactual if converged, else use neighbour:
+    # Extract counterfactual if converged, else use neighbour (no penalty):
     counterfactuals = [
         validities[i] ? eltype(xs[1]).(ce.counterfactual) : neighbours[i] for
         (i, ce) in enumerate(ces)

### 2025-07-29 paper/quarto_aaai/paper.tex

@@
 processing domain. For example, \citet{wu2021polyjuice} propose
-\emph{Polyjuice}, a general-purpose CE generator for language models and
-demonstrate that the augmentation of training data with \emph{Polyjuice}
-improves robustness in a number of tasks, while
+\emph{Polyjuice}, a general-purpose CE generator for language models,
+and demonstrate that the augmentation of training data with
+\emph{Polyjuice} improves robustness in a number of tasks.
 \citet{luu2023counterfactual} introduce the \emph{Counterfactual
@@
 with the data and faithful counterfactuals are consistent with what the
-model has learned about the input data. Ac tionability constraints in
+model has learned about the input data. Actionability constraints in
 Def. \ref{def-explainability} vary and depend on the context in which
@@
 \emph{nascent} to denote interim counterfactuals
-\(\mathbf{x}_{t\leq T}^\prime\) that have not yet converged. As we
+\(\mathbf{x}_{\text{AE}}^\prime\) that have not yet converged. As we
 explain below, these nascent counterfactuals can be stored and
 repurposed as adversarial examples. Conversely, we consider
-counterfactuals \(\mathbf{x}_T^\prime\) as \emph{mature} explanations if
-they have either exhausted all \(T\) iterations or converged by reaching
-a pre-specified threshold, \(\tau\), for the predicted probability of
-the target class:
+counterfactuals \(\mathbf{x}_{\text{CE}}^\prime\) as \emph{mature}
+explanations if they have converged within the \(T\) iterations by
+reaching a pre-specified threshold, \(\tau\), for the predicted
+probability of the target class:
 \(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
@@
 \(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
-\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
-sampled factual from a non-target class generated with a gradient-based
-CE generator for up to \(T\) iterations. Intuitively, the gradient of
-Equation~\ref{eq-div} decreases the energy of observed training samples
-(positive samples) while increasing the energy of counterfactuals
-(negative samples) \citep{du2019implicit}. As the counterfactuals get
-more plausible (Def. \ref{def-explainability}) during training, these
-opposing effects gradually balance each other out
+\(\mathbf{x}_{\text{CE}}^\prime\) denotes a mature counterfactual for a
+randomly sampled factual from a non-target class generated with a
+gradient-based CE generator for up to \(T\) iterations. Intuitively, the
+gradient of Equation~\ref{eq-div} decreases the energy of observed
+training samples (positive samples) while increasing the energy of
+counterfactuals (negative samples) \citep{du2019implicit}. As the
+counterfactuals get more plausible (Def. \ref{def-explainability})
+during training, these opposing effects gradually balance each other out
 \citep{lippe2024uvadlc}.
@@
 literature. Four of them are synthetic with two classes and different
-characteristics: linearly separable clusters (\emph{LS}), overlapping
-clusters (\emph{OL}), concentric circles (\emph{Circ}), and interlocking
-moons (\emph{Moon}). Next, we have four real-world binary tabular
-datasets: \emph{Adult} (Census data) of \citet{becker1996adult2},
-California housing (\emph{CH}) of \citet{pace1997sparse}, Default of
-Credit Card Clients (\emph{Cred}) of \citet{yeh2016default}, and Give Me
-Some Credit (\emph{GMSC}) from \citet{kaggle2011give}. Finally, for the
-convenience of illustration, we use the 10-class \emph{MNIST}
-\citep{lecun1998mnist}.
+characteristics: linearly separable Gaussian clusters (\emph{LS}),
+overlapping clusters (\emph{OL}), concentric circles (\emph{Circ}), and
+interlocking moons (\emph{Moon}). Next, we have four real-world binary
+tabular datasets: \emph{Adult} (Census data) of
+\citet{becker1996adult2}, California housing (\emph{CH}) of
+\citet{pace1997sparse}, Default of Credit Card Clients (\emph{Cred}) of
+\citet{yeh2016default}, and Give Me Some Credit (\emph{GMSC}) from
+\citet{kaggle2011give}. Finally, for the convenience of illustration, we
+use the 10-class \emph{MNIST} \citep{lecun1998mnist}.
 

### 2025-07-29 paper/sections/experiments.qmd

@@
 We make use of nine classification datasets common in the CE/AR literature.
-Four of them are synthetic with two classes and different characteristics: linearly separable clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*).
+Four of them are synthetic with two classes and different characteristics: linearly separable Gaussian clusters (*LS*), overlapping clusters (*OL*), concentric circles (*Circ*), and interlocking moons (*Moon*).
 Next, we have four real-world binary tabular datasets: *Adult* (Census data) of @becker1996adult2, California housing (*CH*) of @pace1997sparse, Default of Credit Card Clients (*Cred*) of @yeh2016default, and Give Me Some Credit (*GMSC*) from @kaggle2011give.
@@
 Test accuracies on adversarially perturbed data are shown in @fig-acc.
-The perturbations size, $\varepsilon\in[0,0.1]$, increases along the horizontal axis and includes zero, corresponding to standard test accuracy for non-perturbed data.
+The perturbation size, $\varepsilon\in[0,0.1]$, increases along the horizontal axis and includes zero, corresponding to standard test accuracy for non-perturbed data.
 For all synthetic datasets, predictive performance of CT is virtually identical to the baseline and unaffected by perturbations.
@@
 
-We evaluate the impacts of three types of hyperparameters on CT. In this section we focus on the highlights and make the full results available in the supplementary appendix.
+We evaluate the impacts of three types of hyperparameters on CT. In this section, we focus on the highlights and make the full results available in the supplementary appendix.
 
@@
 
-Secondly, the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful and should be set sufficiently high to avoid common problems associated with exploding gradients.
+Secondly, the strength of the energy regularization, $\lambda_{\text{reg}}$, is highly impactful and should be set sufficiently high to avoid common problems associated with exploding gradients.
 The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident.

### 2025-07-29 paper/sections/other/notation.qmd

@@
 - $t=1,...,T$: Step indicator for counterfactual search iterations.
-- $\mathbf{x}^\prime_{t \leq T}$: a nascent counterfactual, defined as a counterfactual that has not yet converged.
-- $\mathbf{x}^\prime_{T}$: a mature counterfactual, defined as a counterfactual that has either passed a threshold probability or exhausted all $T$ steps.
+- $\mathbf{x}_{\text{AE}}^\prime$: a nascent counterfactual, defined as a counterfactual that has not yet converged.
+- $\mathbf{x}_{\text{CE}}^\prime$: a mature counterfactual, defined as a counterfactual that has either passed a threshold probability or exhausted all $T$ steps.
 - $\mathbf{x}^+$: a training sample in the target class (ground-truth).

### 2025-11-28 paper/sections/other/main_results.qmd
updated to v1.12

@@
 using CTExperiments
-using CTExperiments.CairoMakie
+using CTExperiments.Makie
 using CTExperiments.CSV

### 2025-11-28 paper/experiments/CTExperiments/src/presentation.jl

@@ -1,7 +1,7 @@
 using AlgebraOfGraphics
 using CategoricalArrays
 using CounterfactualExplanations
-using CairoMakie
+using Makie
 using DataFrames
 using HypothesisTests
 using Plots: Plots, PlotMeasures

### 2025-11-28 paper/experiments/ar_plot.jl

@@ -1,6 +1,6 @@
 using CTExperiments
 using CTExperiments.AlgebraOfGraphics
-using CTExperiments.CairoMakie
+using CTExperiments.Makie
 
 using DotEnv
 DotEnv.load!()

### 2025-11-28 paper/experiments/final_table_ablation.jl

@@ -3,7 +3,7 @@ using Pkg;
 Pkg.status();
 
 using CTExperiments
-using CTExperiments.CairoMakie
+using CTExperiments.Makie
 using CTExperiments.CSV
 using CTExperiments.DataFrames
 using CTExperiments.Plots
